### 2022-01-26 chapters/fairness/fvgm/abstract.tex
adding two fairness papers

\section{abstract}
In recent years, machine learning (ML) algorithms have been deployed in safety-critical and high-stake decision-making, where the \textit{fairness} of algorithms is of paramount importance. Fairness in ML  centers on detecting bias towards certain demographic populations induced by an ML classifier and proposes algorithmic solutions to mitigate the bias with respect to different fairness definitions.  To this end, several \textit{fairness verifiers} have been proposed that compute the bias in the prediction of an ML classifier\textemdash essentially beyond a finite dataset\textemdash given the probability distribution of input features. In the context of verifying \textit{linear classifiers}, existing fairness verifiers are limited by \emph{accuracy} due to imprecise modeling of correlations among features and \emph{scalability} due to restrictive formulations of the classifiers as SSAT/SMT formulas or by sampling.

In this paper, we propose an efficient fairness verifier, called \fvgm, that encodes the correlations among features as a Bayesian network. In contrast to existing verifiers, \fvgm~proposes a \textit{stochastic subset-sum} based approach for verifying linear classifiers. Experimentally, we show that  \fvgm~leads to an \textit{accurate} and \textit{scalable} assessment for more diverse families of fairness-enhancing algorithms, fairness attacks, and group/causal fairness metrics than the state-of-the-art. We also demonstrate that {\fvgm} facilitates the computation of fairness influence functions as a stepping stone to detect the source of bias induced by subsets of features.

%that encodes the correlations among features as a Bayesian network and \red{enables disentangled measurement of bias induced by the data and the classifier}.

### 2022-01-26 chapters/fairness/fvgm/appendix.tex
adding two fairness papers

\clearpage
\appendix

\clearpage

		\section{Broader Impact}
		We have presented an efficient framework for verifying fairness of linear classifiers. This work does not bear any negative societal impact. The sole purpose of our framework is to detect bias of machine learning models that are deployed in high-stake decision making.

		\section{Background}

		\subsection{Fairness Metrics}
		In this section, we state the fairness metrics verified in this paper in further details.
		\subsubsection{Group Fairness.} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\framework} verifies independence and separation metrics. 
		The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups. Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier $\alg$, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in A$.
		Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.

		\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, a classifier satisfies $(1 - \epsilon)$-disparate impact if, for $\epsilon \in [0,1] $,
		\[
		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}'} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'].
		\]
		Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of PPV among sensitive groups, and prescribe this to be near zero. Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $, 
		\[
		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}']| \le \epsilon.
		\]
		For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier $\alg$.

		In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive features $\sensitive$ given the class labels $Y$. In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound sensitive groups. 
		Lower value of equalized odds indicates better fairness.
		A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound sensitive groups $\mathbf{a}, \mathbf{a}' \in A$,
		$ \max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 0]| \le \epsilon, $ and $
		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 1]| \le \epsilon.
		$

		\subsubsection{Path-specific Causal Fairness.}
		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.  For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as 
		$
		\max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}', \mediator ]| \le \epsilon
		$.

		Therefore, PCF constrains that $ \hat{Y} $ is not directly dependent of $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through $ \mediator $. PCF is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.

		\begin{example}
			Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator). It is practical to consider that gender $ \in $ \{male, female\} can affect the college-participation of individuals, and all three features are determining factors for the hiring process. Let `male' be the most favored group by the classifier, for instance. Path-specific causal fairness (PCF) ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $). She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.  Therefore, PCF measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
		\end{example}

%		\newpage
		\section{Proofs of Theoretical Results}
		\iffalse
		\begin{lemmarep}
			$ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.\footnote{\blue{This is the rectified version of Lemma~\ref{lm:property-subset-sum}. We note that this correction does not change any claim in Section~\ref{sec:fvgm} except the proposed heuristic in Section~\ref{sec:dp_formulation}, which we address in final version of the paper.}}
		\end{lemmarep}
		\begin{proof}
		The quantifiers in $\mathbf{B}$ can be either universal, existential, or random. The sum of weights for existential and universal variables in $\mathbf{B}$ are calculated as $ w_{\exists} =  \sum_{i|q_i = \exists} \max\{w_i, 0\}$ and  $ w_{\forall} =  \sum_{i|q_i = \forall} \min\{w_i, 0\}  $, respectively by following Eq.~\eqref{eq:dp_recurse}.   Since the contributions of $ w_\exists $ and $ w_\forall $ in $S(B,\tau)$ are fixed, it suffices to prove that $ S(B,\tau) $ does not rely on the order of \textit{random variables} in $\mathbf{B}$.

		Let  $ \tau' =  \tau - w_\exists - w_\forall$ be the revised threshold for random variables $ B' = \{B_i | q_i = \R\} \subseteq\mathbf{B}$. Then $ S(B,\tau) $ is equivalent to solving $ S(B', \tau') $, and we show that $ S(B', \tau') $ does not rely on the order of $ B' $.

		We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ based on its assignment $\mathbf{B}$. 
\begin{align*}
		f_w(i, b) = \begin{cases}
	w_i &\text{ if }\mathbf{B}= 1, \\
	0 &\text{ if }\mathbf{B}= 0.
	\end{cases}
\end{align*}
		Let $ \sigma \in \{0,1\}^{|B'|} $ be a vector of  assignments of variables in $ B' $. Then, $ S(B', \tau') $ is computed as
		\[ S(B', \tau') = \sum_{\sigma \in \{0,1\}^{|B'|}}  \mathds{1}[\sum_{i=1}^{|B'|} f_w(i, \sigma_i) \ge \tau'] \prod_{i=1}^{|B'|}\Pr[B_i = \sigma_i]. \]
		 The aforementioned equation represents a stochastic subset-sum problem, where for each assignment $ \sigma $, we compute the decision version of the subset-sum problem corresponding to $ \sigma $ and multiply with probabilities of random-variables. We observe that $ S(B',\tau') $ does not rely on the order of $ B' $, because both multiplication and addition operations are commutative for real numbers. 
			\end{proof}
		\iffalse
		\begin{lemmarep}
			Let $ Q = \{q_i\} $ be the set of quantifiers of $\mathbf{B}$. If $ |Q| = 1$, then $ Q $ is called homogeneous, and otherwise $ Q $ is called heterogeneous. If $ Q $ is homogeneous, then $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$. Otherwise, $ S(B, \tau) $ is order-specific to $\mathbf{B}$.
		\end{lemmarep}

		\begin{proof}
			When $ |Q|  = 1 $, all variables in $\mathbf{B}$ are either existential, universal or random. When all variables are existential, $ S(B,\tau) $ is solved linearly in $ |B| $ as $ S(B, \tau) = \mathds{1}[\sum_i \max\{w_i, 0\} \ge \tau] $. Similarly, when all variables are universal, $ S(B, \tau) = \mathds{1}[\sum_i \min\{w_i, 0\} \ge \tau] $. In both cases,  the sum of $ |B| $ integers does not depend on the order of integers.

			We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ depending on its assignment $\mathbf{B}$. 
			\[
			f_w(i, b) = \begin{cases}
			w_i \text{ if }\mathbf{B}= 1 \\
			0 \text{ if }\mathbf{B}= 0
			\end{cases}
			\]
			Let $ \sigma \in \{0,1\}^{|B|} $ be a vector of  assignments of $\mathbf{B}$. When all variables are random, $ S(B, \tau) $ is computed as
			\[ S(B, \tau) = \sum_{\sigma \in \{0,1\}^{|B|}}  \mathds{1}[\sum_{i=1}^{|B|} f_w(i, \sigma_i) \ge \tau] \prod_{i=1}^{|B|}\Pr[B_i = \sigma_i], \]
			which also does not rely on the order of $\mathbf{B}$. Therefore, if $ Q $ is homogeneous, $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.

			\red{Lemma is not correct}

		\end{proof}
	\fi

		\begin{lemmarep}
			The decision version of the fairness verification problem for linear classifiers is in $\mathrm{NP^{PP}}$.
		\end{lemmarep}

		\begin{proof}

			\textbf{Case 1: Subset-sum problem is $ \mathrm{NP} $-complete by a reduction to $ 3 $-SAT.}

			We show that subset-sum problem is a $ \mathrm{NP} $-complete problem. In particular, we show that computing $ \sum_{i} w_iB_i = \tau $ for $ B_i \in\mathbf{B}$ is $ \mathrm{NP} $-complete through a reduction to $ 3 $-SAT~\cite{kleinberg2006algorithm}. $ 3 $-SAT problem computes a satisfying assignment of  a CNF formula, say $ \phi $, defining over $ n $ variables $ X_1, \dots, X_n $ and $ m $ clauses $ C_1, \dots, C_m $ where each clause has three literals. We start by defining two integer weights $ a_i $ and $ b_i $ associated with each variable $ X_i $, where including $ a_i $ corresponds to assigning $ X_i $ to true and including $ b_i $ corresponds to assigning $ X_i $ to false. We compute $ a_i $ and $ b_i $ in the following. 
			\begin{align*}
				&a_i = 10^{m+i} + \sum_{j|X_i \in C_j} 10^j\\
				&b_i = 10^{m+i} + \sum_{j|\neg X_i \in C_j} 10^j
			\end{align*}

			We construct a subset-sum problem of $ 2n $ variables with weights $ W =  \cup_{i=1}^n\{a_i, b_i\} $. Concretely, let $\mathbf{B}= \{B_{1,1}, B_{1,2}, \dots, B_{n,1}, B_{n,2}\} $ be the set of variables such that $ w(B_{i,1}) = a_i $ and $ w(B_{i,2}) = b_i $ and our goal is to compute an assignment of $\mathbf{B}$ such that $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $.

			Let $ \sigma $ be a satisfying assignment of $ 3 $-SAT formula $ \phi $. Given the assignment $\sigma$, we consider a \textit{solution} of the subset-sum problem by assigning $ B_{i, 1} = 1 $ and $ B_{i, 2} = 0 $ whenever $ \sigma(X_i) = 1 $, and $ B_{i, 1} = 0 $ and $ B_{i, 2} = 1 $ whenever $ \sigma(X_i) = 10 $. Intuitively, 
			we consider a subset of weights from $ W $ by including 
			 $ a_i $ if $ X_i $ is true and otherwise we include $ b_i $.  
			 The sum of \textit{included} weights has following properties. 
			\begin{itemize}
				\item  the sum has $ 1 $ in leftmost $ n $ digits, corresponding to $ 10^{m+i} $ for $ i = 1, \dots, n $ as we include one of $ a_i $ or $ b_i $ for each $ i $. 
				\item the sum has $ 1, 2 $, or $ 3 $ in the next $ m $ digits. The digit $ 10^j $ is the number of true literals in clause $ C_j $ and it is non-zero since $ \sigma $ satisfies each clause in  $ \phi $.
				\item $ 0 $ in the rightmost digit.
			\end{itemize}

			In the subset-sum problem, we set the threshold $ \tau = \sum_{i=1}^n 10^{m+i} + 3 \sum_{j=1}^m 10^j $. We additionally consider integer weights $ c_j =  d_j = 10^j $, which will be added to $ W $ as discussed next. For brevity of presentation, we are not assigning Boolean variables in $\mathbf{B}$ for additional weights $ c_j $ and $ d_j $.
			 We then prove that  $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ if and only if $ \phi $ is satisfiable.

			\begin{itemize}
				\item If $ \phi $ is satisfiable, then we select weights $  a _i $ if $ \sigma(X_i) = 1 $, else $ \mathbf{B}_i $. As discussed above, the sum of selected weights has leftmost $ n $ digits of $ \tau $ correct. To make remaining digits correct, we may need to add both of $ c_j $ and $ d_j $, one of $ c_j $ and $ d_j $, or none of $ c_j $ and $ d_j $ depending  if the digit $ 10^j $ in the sum is $ 1,2, $ or $ 3 $. We emphasis that this addition is deterministic as it relies on the number of true literals in $ C_j $ for the assignment $ \sigma $. Thus, for a satisfying assignment of $ 3 $-SAT, there exists a solution of the subset-sum problem.

				\item Finally, we prove that a solution of the subset-sum problem corresponds to a satisfying assignment of $ \phi $.  First we observe that in any digit $ 10^k $, for $ k \le m $, there are at most five weights with $ 1 $ in digit $ 10^k $: the three literals corresponding to clause $ k $ and $ c_k $ and $ d_k $ (otherwise when $ k > m $, there are at most $ 2 $  weights with $ 1 $ in digit $ 10^k $).  Since there are at most $ 5 $ ones in any position, a subset of these weights will not cause any carries in addition. Hence, we can only get a total of $ \tau $ by including the right number of $ 1 $s in any digit. To do this, one needs to include exactly one of $ a_i $ and $ b_i $, and the corresponding truth assignment $ \sigma $ needs to satisfy $ \phi $ so that in each position $ 1 \le j \le m $, we get at least one $ 1 $ in digit $ 10^j $. We can then add $ c_j $ and $ d_j $ to increase the sum of weights to $ \tau $, as discussed previously. This, for a solution of the subset-sum problem, there is a corresponding satisfying assignment of $ 3 $-SAT
			\end{itemize}

			In the above analysis, the construction of $ a_i $ and $ b_i $ has polynomial complexity with respect to $ n $ and $ m $. Therefore, the subset-sum problem $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ and equivalently, $ \sum_{i} w_iB_i = \tau $ is $ \mathrm{NP} $-complete because $ 3 $-SAT is also $ \mathrm{NP} $-complete. We further claim that computing $ \sum_{i} w_iB_i \ge \tau $ is at least as hard as computing $ \sum_{i} w_iB_i = \tau $, which is required for analyzing the complexity of stochastic subset-sum problem, as discussed next.

			\textbf{Case 2: Stochastic subset-sum problem is $ \mathrm{PSPACE} $-complete by a reduction to $ 3 $-SSAT.}

			We observe that a $ 3 $-SSAT formula that is based on a $ 3 $-SAT formula $ \phi $ and an additional specification of quantifiers for each variable in $ \phi $. Concretely, we define a $ 3 $-SSAT formula $$ \Phi \triangleq (\{q(X_i)\}_{i=1}^{n}, \phi) \text{ , where } q(X_i) \in \{\exists, \forall, \R^{p}\} .$$ To analyze the complexity of a stochastic subset-sum problem, we discuss a procedure in the following that is based on a polynomial reduction to $ 3 $-SSAT.

			We apply the aforementioned construction procedure ($ 3 $-SAT to subset-sum problem) to translate  a $ 3 $-SSAT formula to a stochastic sub-set sum problem. In particular, for each quantified variable in $ \phi $, we now discuss the quantifier of variables in the stochastic subset-sum problem.  If $ X_i $ is $ \exists $ (resp.\ $ \forall $) quantified, associated $ B_{i,1} $ and $ B_{i,2} $ are also $ \exists $ (resp.\ $ \forall $) quantified. Moreover, if $ X_i $ is randomized ($ \R^{p_i} $) quantified with $ p_i = \Pr[X_i = 1] $, then both $ B_{i,1} $ and $ B_{i,2} $ are randomized quantified with probability $ \Pr[B_{i,1}] = p_i $ and $ \Pr[B_{i,2}] = 1 - p_i $. Since the decision version of $ 3 $-SSAT is $ \mathrm{PSPACE} $-complete~\cite{littman2001stochastic}  and the presented reduction procedure has polynomial complexity, the decision version of stochastic subset-sum if also $ \mathrm{PSPACE} $-complete.

			\textbf{Case 3: Fairness verification for linear classifiers is in
			$\mathrm{NP^{PP}}$ by a reduction to E-MAJSAT.}

			E-MAJSAT is a specific quantifier arrangement of SSAT problems where existential (or universal) variables appear before random variables. Formally, a E-MAJSAT formula 
			$$ \Phi =  (\{q(X_1) = \exists,\cdots,q(X_i) =\exists, q(X_{i+1}) =\R^{p_{i+1}}, \cdots, q(X_n) = \R^{p_{n}} \}, \phi).$$ The decision version of E-MAJSAT problem has complexity in $\mathrm{NP^{PP}}$~\cite{littman2001stochastic}. Furthermore, the decision version of SSAT problems with universal quantifiers followed by random-quantifiers are also in $\mathrm{NP^{PP}}$ as this a dual problem of E-MAJSAT~\cite{littman2001stochastic}. 
			In Section~\ref{sec:fvgm}, we present fairness verification of linear classifiers defined over Boolean variables using a formulation of stochastic subset-sum problem. Similar to E-MAJSAT,  existential (or universal) variables appear before random variables in our formulation.  Thus, the decision version of fairness verification of linear classifiers is in $\mathrm{NP^{PP}}$.
			\iffalse 
			\red{We further highlight that verification of linear classifiers on continuous real-valued variables is at least as hard as verifying classifiers on Boolean variables.}\fi
		\end{proof}
		\fi

		\begin{lemmarep}
			Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $. Let $ w_{\exists} = \sum_{B_i \in \mathbf{B} | q_i = \exists} \max\{w_i, 0\}$  and $ w_{\forall} = \sum_{B_i \in \mathbf{B} | q_i = \forall} \min\{w_i, 0\}$ be the considered sum of weights of existential and universal variables, respectively. We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $. The space complexity is  $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
		\end{lemmarep}

		\begin{proof}
			\textbf{Case 1: All $n$ variables in $\mathbf{B}$ have randomized quantifiers.}

			At step-$i$ of the dynamic programming (Eq.~\eqref{eq:dp_recurse}), we modify the residual threshold of that step, namely $\tau_i$, either by subtracting $w_i$ or by retaining it.
			Now, we observe that the residual threshold $\tau_i$ for any $i \in \lbrace1,\ldots,n\rbrace$ will be in $[0, \tau+|w_{neg}|]$. This holds because if $\tau_i$ crosses these bounds, the dynamic programming is terminated as shown in Equation~\eqref{eq:dp_terminus}.
			Since all weights of $\{w_i\}_{i=1}^{n}$ are integers, the maximum number of values that the residual threshold can take, is $ (\tau + |w_{neg}|) $.
			Thus, we need to store at most $ n(\tau + |w_{neg}|)$ values in the memory for performing dynamic programming with $n$ variables and $ (\tau + |w_{neg}|) $ number of possible weights.
			Thus, the space complexity  is $ \mathcal{O}(n(\tau + |w_{neg}|)) $.

			In order to construct the dynamic programming table, we have to call the $\mathsf{dp}$ function $ \mathcal{O}(n(\tau + |w_{neg}|)) $ times, in the worst-case.
			Thus, the time complexity of our method is $ \mathcal{O}(n(\tau + |w_{neg}|) $.

			\textbf{Case 2: $n'$ variables have existential or universal quantifiers and $n-n'$ variables have randomized quantifiers in $\mathbf{B}$.}

			According to Eq.~\eqref{eq:dp_recurse}, $ w_\exists $ and $ w_\forall $ are the fixed weights of all existential and universal variables, respectively. Therefore, we need to consider at most $ \tau + |w_{neg}| -  w_\exists - w_\forall $ values of residual weights for random variables. By applying analysis in Case 1, the space and time complexity is derived as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.

			We note that there is an additional time complexity of $ \mathcal{O}(n') $ for existential and universal variables in Eq.~\eqref{eq:dp_recurse}. Thus the time complexity becomes $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $. We, however, do not require to store any entry for existential and universal variables in $ \mathsf{dp} $  function and thus, the space complexity remains the same as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
		\end{proof}

		\begin{lemmarep}
			Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $. Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $. To exactly compute {\stochastic} with correlated variables in dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
		\end{lemmarep}

		\input{tables/sanity_check.tex}

		\begin{proof}
			We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $. For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{sec:dp_with_BN}.

			For variables in $\mathbf{B}\setminus \mathbf{V} $, we apply analysis from  Lemma~\ref{lemma:complexity_sss}, where we consider $ (n - n'' - |\mathbf{V}|) $ random variables, $ n'' $ existential/universal variables, and residual weights can take at most $ (\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) $ values. Hence, time complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $, and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $

			Combining two cases, overall time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $. 
		\end{proof}

%	\newpage	
	\section{Extended Experimental Evaluations}
	\label{appendix:experiments}
	Each experiment is performed on Intel Xeon E$ 7-8857 $ v$2 $ CPUs with $ 16 $GB memory, $ 64 $bit Linux distribution based on Debian OS and clock speed $ 3 $ GHz. In the following, we discuss extended experimental results.

	\subsection{Accuracy Comparison Among Different Verifiers}
	We have considered a synthetic problem for comparing accuracy among different verifiers. For Example~\ref{example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $. For `income' feature ($ I $), we consider two Gaussian distributions $ \Pr[I | \text{age} \ge 40] \sim \mathcal{N}(0.6, 0.1) $ and $ \Pr[I | \text{age} < 40] \sim \mathcal{N}(0.4, 0.1) $ separated by two age groups. Moreover, for `fitness' feature ($ F $), we consider two Gaussian distributions $ \Pr[F | \text{age} \ge 40] \sim \mathcal{N}(0.7, 0.1) $ and $ \Pr[F | \text{age} < 40] \sim \mathcal{N}(0.3, 0.1) $. On this data, the trained LR and SVM classifier has decision boundary as $ 7.26I + 7.4F - 1.34A \ge 6.62 $ and $ 9.37I + 9.75F - 0.34A \ge 9.4 $, respectively.

	In Figure~\ref{fig:synthetic_bn} we show the Bayesian Network on discretized features, in particular for income and fitness features. In Figure~\ref{fig:synthetic_ppv_max_LR} \ to Figure~\ref{fig:synthetic_ppv_min_SVM}, we show PPVs of different classifiers computed by different verifiers, where  {\framework} outputs closest to exactly computed values, in comparison with Justicia. In Figure~\ref{fig:synthetic_sample_size_LR} to Figure~\ref{fig:synthetic_sample_size_SVM}, we show the effect of sample size on {\framework} in measuring fairness metrics: disparate impact and statistical parity, where with increasing sample size, the estimate becomes more accurate.

		\begin{figure}[!t]
		\begin{center}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_SVM}}	\\		
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_SVM}}\\

			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_SVM}}		\\
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_SVM}}\\

			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_LR}}
			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_SVM}}

			\caption{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
			\label{fig:time_vary_features}

		\end{center}

	\end{figure}

	\subsection{Scalability Comparison Among Different Verifiers}

	In Figure~\ref{fig:time_vary_features}, we present the runtime of different fairness verifiers while varying the number of features in different datasets. We observe that with an increase of features, the runtime increases in general.

	\input{tables/fairness_attacks.tex}
	%\input{tables/verifier_scalability.tex}
	%\input{tables/ssat_vs_subset_sum.tex}
	%\input{tables/fairness_algorithms.tex}
	\input{tables/fairness_metrics_on_compound_sensitive_groups.tex}	
	\input{tables/dag_structure_complexity.tex}

	\subsection{Verifying Fairness Algorithms on Multiple Fairness Metrics}

	We show extended results on verifying fairness attack in Figure~\ref{fig:attack_extended} for two fairness metrics: disparate impact (DI) and statistical parity (SP). We observe that {\framework} can detect poisoning attack for both metrics.

	In Figure~\ref{fig:multiple_metrics} we show verification results on compound sensitive groups with respect to multiple fairness metrics. In this figure, we observe that with an increase in the number of groups, fairness metrics worsens\textemdash disparate impact decreases and other three metrics increases. Additionally, in Table~\ref{fig:extended_verifying_fairness_algorithms}, we show extended results for German and COMPAS dataset on verifying fairness algorithms: reweighing and optimized pre-processing algorithms.

	\subsection{Performance Analysis of Bayesian Network}
	In Figure~\ref{fig:results_DAG_complexity}, we analyze the performance of encoding Bayesian Networks of differing complexity. We define the complexity of the network as  $ \frac{|V|}{|\nonsensitive \cup \sensitive|} $, which is the ratio between the number of features appearing in the network and total features. In this figure, as the ratio increases, both computation time of {\framework} and learning time of Bayesian Network increase.

	\begin{table}[!t]

	\centering
	\small
	\caption{Extended results for verification of different fairness enhancing algorithms using {\framework}. Numbers in bold refer to fairness improvement w.r.t. different fairness metrics. RW and OP refer to reweighing and optimized-preprocessing algorithms respectively.  }
	\label{fig:extended_verifying_fairness_algorithms}
	\setlength{\tabcolsep}{.5em}

	\begin{tabular}{lllrrrrrrrrrrrrr}

		\toprule
		Dataset & Sensitive & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta$ EO\\
		\midrule

		\multirow{4}{*}{Adult}&\multirow{2}{*}{race}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
		&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
		&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\

		\midrule
		\multirow{4}{*}{COMPAS}&\multirow{2}{*}{race}&RW&$ \textbf{0.13} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.06} $\\
		&&OP&$ \textbf{0.15} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ \textbf{-0.05} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ \textbf{0.1} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ 0.04 $\\
		&&OP&$ \textbf{0.09} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ \textbf{-0.03} $\\

		\midrule
		\multirow{4}{*}{German}&\multirow{2}{*}{age}&RW&$ \textbf{0.52} $&$ \textbf{-0.53} $&$ \textbf{-0.52} $&$ \textbf{-0.47} $\\
		&&OP&$ \textbf{0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.51} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ -0.06 $&$ 0.06 $&$ 0.06 $&$ 0.02 $\\
		&&OP&$ -0.12 $&$ 0.12 $&$ 0.12 $&$ 0.07 $\\

		\bottomrule
	\end{tabular}
\end{table}

\iffalse
\begin{table}[h!]
		\centering
		\caption{Extended results for runtime of different verifiers in seconds.  `\textemdash'~ refers to timeout ($ 900 $ seconds). Bold number refers to best scalability results. We additionally show results for decision tree classifiers (DT). {\framework} takes comparatively more time for decision tree classifiers because of encoding feature-correlations represented as a Bayesian network.}
			\label{tab:scalability_extended}	
		\begin{tabular}{llrrrrrrrrrrrrrr}

			\toprule
			Dataset & Classifier & FairSquare & VeriFair & Justicia & {\framework} \\
			\midrule

			\multirow{3}{*}{Ricci} & LR & \textemdash &  $ 2.0 $  &  $ 2.2 $  &  $ \textbf{0.1} $  \\ 
			& SVM & \textemdash &  $ 1.8 $  &  $ 2.2 $  &  $ \textbf{0.1} $  \\ 
			& DT &  $ 4.6 $  &  $ 1.9 $  &  $ \textbf{0.1} $  &  $ \textbf{0.1} $  \\

			\midrule
			\multirow{3}{*}{Titanic} & LR & \textemdash &  $ 0.5 $  &  $ 0.3 $  &  $ \textbf{0.1} $  \\ 
			& SVM & \textemdash &  $ 0.4 $  &  $ 0.2 $  &  $ \textbf{0.1} $  \\ 
			& DT &  $ 432.9 $  &  $ 17.2 $  &  $ \textbf{0.3} $  &  $ 1.6 $  \\

			\midrule
			\multirow{3}{*}{Compas} & LR & \textemdash &  $ 12.2 $  & \textemdash &  $ \textbf{0.4} $  \\ 
			& SVM & \textemdash &  $ 11.6 $  & \textemdash &  $ \textbf{0.5} $  \\ 
			& DT & \textemdash &  $ 377.1 $  &  $ \textbf{0.3} $  &  $ 121.7 $  \\

			\midrule
			\multirow{3}{*}{Adult} & LR & \textemdash &  $ 7.3 $  & \textemdash &  $ \textbf{0.9} $  \\ 
			& SVM & \textemdash &  $ 21.7 $  &  $ \textbf{1.7} $  &  $ 1.8 $  \\ 
			& DT & \textemdash &  $ 57.9 $  &  $ 0.4 $  &  $ \textbf{0.3} $  \\

			\midrule
			\multirow{3}{*}{German} & LR & \textemdash &  $ 19.2 $  & \textemdash &  $ \textbf{1.4} $  \\ 
			& SVM & \textemdash &  $ 28.8 $  & \textemdash &  $ \textbf{1.4} $  \\ 
			& DT & \textemdash &  $ 78.5 $  &  $ \textbf{0.2} $  &  $ 2.3 $  \\

			\bottomrule

		\end{tabular}

	\end{table}	
\fi

%\newpage
\input{sections/fairness_influence_functions.tex}

\iffalse
\clearpage
\paragraph{Limitations of existing SSAT solvers.}
Since there is no state of the art SSAT solver that can handle universal quantification, we cannot consider conditional probabilities of universal variables while encoding the Bayesian network. In {\framework}, universal quantification appears while learning the least favored group in the learning encoding. When we do not consider conditional probabilities on universal variables, earlier approach could solve this using an existential SSAT solver by first negating the CNF and then subtracting computed probability from $ 1 $ as a solution to the original universal SSAT problem. This approach does not work when conditional probabilities are involved.

For example, consider a clause $ a \vee\mathbf{B}\vee c $ defined on three Boolean variables where $ a $ is existential and $ b,c $ are random. Without considering conditional probabilities on $ a $, it is assigned true trivially. However, with conditional probabilities $ a $ can be false depending on the probability. Similar argument holds when $ a $ is universal. Hence, when conditional probabilities are considered for universal variables, a dedicated SSAT solver is also required to solve them. 
\fi

%\noindent\makebox[\linewidth]{\rule{\textwidth}{0.4pt}}

\iffalse
\clearpage

\section{Progress}

\begin{enumerate}
	\item \textbf{Encoding conditional probabilities:} In our experiments, we have studied two algorithms for learning Bayesian network from data: Notears and pgmpy. After learning a DAG structure, we post-process it before feeding to {\framework}.

	\begin{itemize}
		\item \textbf{Modifying in-degree edges for sensitive variables.} In the SSAT-based encoding, sensitive variables have existential (or universal) quantification because we are interested in knowing which assignment to these variables result in the most (or least) favored group.

		In a Bayesian network, a directed edge $ (a,b) $ between variable $ a $ and $\mathbf{B}$ refers to the conditional dependence of $\mathbf{B}$ on $ a $.  We now consider a case where variable $\mathbf{B}$ is a sensitive variable. Since an existential (or universal) quantification on variable $\mathbf{B}$ fixes its assignment, we first reverse the direction of the edge to $ (b,a) $ in order to capture the dependence between two variables. We highlight that {\framework} does not have a strict restriction for DAG structure as input. Therefore, even if reversing an edge may violate DAG structure, {\framework} is not affected by this modification.

		If both $ a $ and $\mathbf{B}$ refers to different categories for a sensitive attribute, more specifically they belong to the one-hot encoded variables for a sensitive attribute, we remove the edge $ (a, b) $ from the learned DAG.

	\end{itemize}

	\item \textbf{Verifying Path-specific Causal Fairness.}

	\item \textbf{Fairness Attack Verification:} We apply {\framework} to verify fairness poisoning attack \cite{?}. This attack algorithm generates poisoned data, which together with initial training data results in a ML classifier with decaying fairness results. With {\framework} we verify the effectiveness of fairness poisoning attacks.

	\item \textbf{Verifying pre-processing and post-processing fairness algorithms:} We apply {\framework} to verify the  effectiveness of fairness enhancing algorithms, in particular pre-processing and post-processing algorithms.

	\begin{itemize}
		\item \textbf{Pre-processing fairness algorithms.} Pre-processing algorithms process a  dataset such that any model learned on the dataset has better fairness results, computed in terms of disparate impact, statistical parity and so on. To verify the effectiveness of a pre-processing algorithm, we first train a classifier on unprocessed data and later another classifier on processed data. Finally, we compute the difference in different fairness metrics using {\framework}. For an effective fairness algorithm, disparate impact increases and statistical parity decreases for the processed data.

		\item \textbf{Post-processing fairness algorithms.} 
		Post processing fairness algorithms works on the output of any already trained classifier so that the final output by the fairness algorithm improves fairness of the classifier. \red{In order to verify a post-processing fairness algorithm, we learn a new classifier on the processed data\footnote{Learning a new classifier is not a requirement for approaches that can compute fairness on a dataset (includes class-label). However, our framework requires a classifier, which learns the mapping between input features and class-label. } and report difference in fairness metrics. The new classifier should have high accuracy on the processed data, otherwise the verification result by {\framework} is not reliable.  }

	\end{itemize}

	\item \textbf{Feature Influence.} We employ {\framework} to compute feature-influence of each feature in the data. For a distribution $ \mathcal{D} $, a finite sample dataset $ D \sim \mathcal{D} $, and feature $ x $, we define its influence in terms of the positive predictive value (PPV) of the classifier.  
	\[
	q(x) \triangleq  \Pr[\hat{Y}_{D_{-x}} = 1 | \mathcal{D}_{-x}] - \Pr[\hat{Y}_D = 1 | \mathcal{D}]
	\]

	In this equation, $ \mathcal{D}_{-x} $ (resp.\ $ D_{-x} $) refers to the distribution (resp.\ dataset) where all but feature $ x $ are present.
	Intuitively, the influence of feature $ x $ is the difference in positive predictive value of the classifier as if $ x $ does and does not participate in the task of classification. Note that, the classifier is learned on dataset $ D $ whereas $ q(x) $ is computed for the distribution $ \mathcal{D} $.

	\textbf{Outline:}
	\begin{enumerate}
		\item Train a classifier $ M $ on a dataset $ D $
		\item From the dataset, compute product distribution of features $ \mathcal{D} $. This is handled by {\framework} \red{Can we implement an interface where the distribution is passed as inputs? }
		\item Compute base values of different fairness metrics. In our experiments, we have computed disparate impact and statistical parity. '
		\item To compute the influence of an attribute, we replace that attribute with random value on the dataset $ D' $. For a continuous attribute, we randomly generate attribute value within the maximum and minimum of original attribute-values. For a categorical attribute, we randomly select a category. 
		\item We again calculate distribution from $ D' $ and run {\framework} on the new distribution.
		\item Since random-value replacement involves randomness, we have repeated the process for $ 5 $ iterations. 
	\end{enumerate}

	\textbf{Result Analysis}
	We first learn dependency among features using Bayesian structure learning algorithm. For the dependency structure, we compute feature-influence and report in Figure~\ref{fig:feature_influence_compas}. In COMPAS dataset, we consider `race' as a sensitive attribute where race = 1 denotes Caucasian. In all three reported classifiers, prior-counts and juvenile-count are top influential features for both sensitive group. Other non-crime features such as race, age are also found to be influential for logistic regression classifier and CNF classifier. Moreover, depending on sensitive group, age and race have differentiating influence. 
	\red{Incomplete analysis}

\end{enumerate}

For the learning encoding, sensitive variables $ \sensitive $ are existentially quantified in $ \mathsf{prefix_{ERE}} $ and universally quantified in $ \mathsf{prefix_{URE}} $. These variables are followed by randomized quantified variables $ \lambda_{X_i}, \lambda_{x,\mathbf{u}} $ and $ X_j \in \nonsensitive \setminus V $. Finally, remaining variables such as $ \lambda_{\hat{Y}} $ and $ X_i $ for each $ X_i \in V(G) $ are existentially quantified.

\paragraph{Extra}
To this end, we consider all features $ \nonsensitive \cup \sensitive $ to be Boolean. We additionally constrain the classifier $ \hat{Y} $  to be a Boolean formula $ \phi_{\hat{Y}} $ in CNF defined over $ \nonsensitive \cup \sensitive $ such that, for an assignment of $ \nonsensitive \cup \sensitive $, $ \hat{Y} = 1 $ whenever $ \phi_{\hat{Y}} $ is true, and $ \hat{Y} = 0 $ otherwise. Additionally, we construct another CNF formula $ \phi_\BN $ to encode the joint distribution of $ \BN $, which we discuss shortly.

One approach to verify different fairness metrics such as disparate impact, statistical parity, equalized odds, and path-specific causal fairness is to compute the most favored group and the least favored group based on the group specific PPV of the classifier~\cite{ghosh2020justicia}. As such, we can either (i) enumerate computation of group specific PPV for all sensitive groups or (ii) learn the most favored group and the least favored group, to compute these metrics. 	Provided that we have  $ \phi_{\hat{Y}} $ and $ \phi_\BN $ representing the classifier and the Bayesian network, we now discuss two encodings in the following.

\begin{enumerate}
	\item \textbf{Enumeration.} We consider an RE-SSAT (Randomized Exponential Stochastic SAT) instance with random quantifier over Boolean variables to compute $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \BN] $. 
	\begin{align}
	\Phi_{\mathbf{a}} = \mathsf{prefix_{Enum}}, \phi_{\hat{Y}} \wedge \phi_\BN \wedge (\sensitive = \mathbf{a})
	\end{align}
	where $ \mathsf{prefix_{Enum}} $ is an ordered sequence of randomized quantified Boolean variables and $ \sensitive = \mathbf{a} $ encodes the sensitive group $ \mathbf{a} \in \sensitive $ as a CNF. We discuss quantifiers in detail in Section~\ref{?}. An RE-SSAT formula can be solved by a projected model counter where we project model-counting (number of satisfying assignments in a CNF) on randomized quantified variables. Solving $ \Phi_{\mathbf{a}} $,  the probability of satisfaction $ \Pr[\Phi_{\mathbf{a}}] $ exactly computes group-specific PPV of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \BN] $. Finally, we enumerate this computation for all sensitive groups $ \forall \mathsf{a} \in A $ and use  $ \max_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $  and $ \min_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $\textemdash the PPV of the most and the least favored group, respectively\textemdash  to compute different fairness metrics.

	\item \textbf{Learning.} The second encoding efficiently learns the most favored group and the least favored group by instantiating two SSAT formulas: ERE (Existential-Random-Existential) and URE (Universal-Random-Existential) SSAT formulas, respectively. Both SSAT formulas have the same CNF formula and only difference is in the prefix. 
	\begin{align}
	&\Phi_\mathsf{ERE} = \mathsf{prefix_{ERE}}, \phi_{\hat{Y}} \wedge \phi_\BN\\
	&\Phi_\mathsf{URE} = \mathsf{prefix_{URE}}, \phi_{\hat{Y}} \wedge \phi_\BN
	\end{align}
	$ \mathsf{prefix_{ERE}} $ (resp.\ $ \mathsf{prefix_{URE}} $) is an ordered sequence of existentially (resp.\ universally) quantified variables followed by randomized quantified variables.
	We apply existential quantifier ($ \exists $) on sensitive features $ A $ in $ \Phi_\mathsf{ERE} $ using $ \mathsf{prefix_{ERE}} $. Therefore, according to rule(??) in Section\ref{??}, we learn an assignment to $ A $, say the most favored sensitive group,  resulting in the maximum satisfying probability of $ \Phi_\mathsf{ERE} $. Formally, $ \Pr[\Phi_\mathsf{ER}] $ is equal to $ \max_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $ in the enumeration encoding. We similarly apply universal quantifiers ($ \forall $) on $ A $ to learn the least favored sensitive group with $ \Pr[\Phi_\mathsf{URE}] $ equalizing to $ \min_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $. 
\end{enumerate}

\red{Complexity analysis has to be verified.}
Both enumeration and learning encoding output same results and their worst case complexity is same. In particular, solving an  RE-SSAT formula has complexity $ \text{PP}^{\text{NP}} $ and we enumerate for all sensitive groups, which is exponentially many in $ A $. Similarly, solving both ER-SSAT and URE-SSAT is $ \text{NP}^{\text{PP}} $ hard.

We now discuss the construction of the CNF formula, particularly $ \phi_{\hat{Y}} $ and $ \phi_\BN $. Ghosh et al.~\cite{ghosh2020justicia} have proposed techniques to encode decision trees and linear classifiers as a CNF formula $ \phi_{\hat{Y}} $. For linear classifiers, the decision function is viewed as a pseudo-Boolean constraint and can be efficiently encoded to CNF by introducing auxiliary variables, denoted by $ \lambda_{\hat{Y}} $.  Support-vector classifiers with linear kernels can also be encoded to CNF in the same approach to linear classifiers. Hence, we now focus on the construction of $ \phi_\BN $ to encode the Bayesian network to CNF.

\red{Additional}
\begin{enumerate}
	\item Advance terminate when $ S(W,\tau, Q) \le w_{neg} $, where $ w_{neg} $ is the sum of negative weights in $ W $.
	\item Within similar quantifiers, sort based on descending weights to leverage early termination.
	\item We can learn maximum and minimum of $ S(W,\tau,Q) $ in one go (should be explained in dynamic programming). It is not required to convert weights to positive. One benefit of assuming all positive weights is that we then have one sub-problem with selection of weights (and this sub-problem has maximum value). However, to compute minimum of $ S $, we again need to consider the sub-problem with not selection of weights. But we can propose a dynamic approach where both max and min can be solved in one go. 
\end{enumerate}

\section{Experimental Evaluations}

\begin{figure}
	\begin{center}
		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Edges}}
		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Time_(DAG)}}\\
		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Disparate_impact}}
		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Statistical_parity}}\\
		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_ppv}}\\
	\end{center}
	\caption{Experimenting the effect of regularizer in DAG learner inside {\framework}. With increasing values of regularizer, learned DAGs become sparser by undertaking less learning time. For different DAGs, we evaluate positive predictive value (PPV) for a fixed classification algorithm. We occasionally observe that disparate impact  increases and statistical parity decreases for sparser DAG. We show break down result for two different sensitive groups where the group `Age $ = 1 $' is often (but not always) more discriminated than `Age $ = 0 $'  in majority values of regularizer.  Therefore, {\framework} can verify bias induced by data (in form of Bayesian Network)  for the fixed classification algorithm.}
\end{figure}
\fi

\input{sections/cnf_correlations}

%\clearpage
%	\bibliographystyle{plainnat.bst}
%	\bibliography{ref.bib}

### 2022-01-26 chapters/fairness/fvgm/applications.tex
adding two fairness papers

\section{Applications of {\fvgm}}\label{fairness_fvgm_sec:applications}
In this section, we apply {\fvgm} for verifying fairness-enhancing algorithms and depreciation attacks. We also show that {\fvgm} facilitates computation of fairness influence functions enabling detection of bias due to individual features.

\iffalse
\begin{figure}[t!]
	%\vspace{-1em}

	\centering
	\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/disp_fairness_attack}}
	\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/stat_fairness_attack}}
	%\vspace{-0.5em}
	\caption{Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
\end{figure}
\fi

\paragraph{Verifying Fairness-enhancing Algorithms.} We deploy {\fvgm} in verifying the effectiveness of fairness-enhancing algorithms designed to ameliorate bias. For example, fairness 
pre-processing algorithms can be validated by applying {\fvgm} on the unprocessed and the processed data separately and comparing different fairness metrics. In Table~\ref{fairness_fvgm_tab:fair_algo_verification}, we report the effect of fairness algorithms w.r.t. four fairness metrics: disparate impact (DI), path-specific causal fairness (PCF), statistical parity (SP), and equalized odds (EO). Note that, fairness is improved if DI increases and the rest of the metrics decrease. In most instances for Adult dataset, reweighing (RW)~\cite{kamiran2012data} and optimized pre-processing (OP)~\cite{calmon2017optimized} algorithms are successful in improving fairness. Except for the unfairness regarding the sensitive feature `sex', where OP algorithm fails. Thus, \textit{{\fvgm} verifies the enhancement and decrement in fairness by  fairness-enhancing algorithms. }

\paragraph{Verifying Fairness Attacks.}
We apply {\fvgm} in verifying a fairness poisoning-attack algorithm. This algorithm injects a small fraction of poisoned samples into the training data such that the classifier becomes relatively unfair~\cite{solans2020poisoning}. We apply this attack to add $\{1, 5, \dots, 160\}$ poisoned samples  and measure the corresponding DI's. In Figure~\ref{fairness_fvgm_fig:fairness_attacks},  {\fvgm} verifies that DI of the classifier decreases, i.e.\textit{ the classifier becomes more unfair},\textit{ as the number of poisoned samples increases}. Therefore, {\fvgm} shows the potential of being deployed in safety-critical applications to detect fairness attacks. For example, if we set $0.9$ as an acceptable threshold of DI, {\fvgm} can raise an alarm once $160$ poisoned samples are added.

\iffalse
\begin{table}[!t]

	\centering
	\small
	\caption{Verification of fairness algorithms using {\fvgm}. Numbers in bold refer to fairness improvement. RW and OP refer to reweighing and optimized-preprocessing algorithms.  }\label{fairness_fvgm_tab:fair_algo_verification}
	\setlength{\tabcolsep}{.4em}

	\begin{tabular}{lllrrrrrrrrrrrrr}

		\toprule
		Dataset & Sensitive & Algo. & $ \Delta $ DI &  $ \Delta $ PCF & $ \Delta $ SP & $ \Delta $ EO\\
		\midrule

		\multirow{4}{*}{Adult}&\multirow{2}{*}{race}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
		&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
		&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\

		\midrule
		\multirow{4}{*}{COMPAS}&\multirow{2}{*}{race}&RW&$ \textbf{0.13} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.06} $\\
		&&OP&$ \textbf{0.15} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ \textbf{-0.05} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ \textbf{0.1} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ 0.04 $\\
		&&OP&$ \textbf{0.09} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ \textbf{-0.03} $\\

		\midrule
		\multirow{4}{*}{German}&\multirow{2}{*}{age}&RW&$ \textbf{0.52} $&$ \textbf{-0.53} $&$ \textbf{-0.52} $&$ \textbf{-0.47} $\\
		&&OP&$ \textbf{0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.51} $\\
		\cmidrule{2-7}
		&\multirow{2}{*}{sex}&RW&$ -0.06 $&$ 0.06 $&$ 0.06 $&$ 0.02 $\\
		&&OP&$ -0.12 $&$ 0.12 $&$ 0.12 $&$ 0.07 $\\

		\bottomrule
	\end{tabular}
\end{table}
\fi

\paragraph{Fairness Influence Function (FIF): Tracing Sources of Unfairness.}
Another application of \fvgm{} as a fairness verifier is to quantify the effect of a subset of features on fairness. Thus, we define fairness influence function (FIF)  that computes the effect of a subset of features $\mathbf{S}$ on the PPV of a classifier given a specific sensitive group $ \sensitive = \mathbf{a}$,
$
	\mathsf{FIF}(\mathbf{S}) \triangleq \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mathcal{D}] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a},  \mathcal{D}_{-\mathbf{S}}].
$
FIF allows us to explain the sources of unfairness in the classifier. In practice, we compute FIF of $\mathbf{S}$ by replacing the probability distribution of $\mathbf{S}$ with a uniformly random distribution, referred to as  $ \mathcal{D}_{-\mathbf{S}} $, and reporting differences in the conditional PPVs. Further details of FIF are in Appendix.

In Figure~\ref{fairness_fvgm_fig:influence function}, we compute FIF for all features in COMPAS dataset, separately for two sensitive groups: Female (`sex' $ = 1 $) and Male. This dataset concerns the likelihood of a person of re-offending crimes within the next two years. We first observe that the base PPV values are different for the two groups ($ 0.46 $ vs $ 0.61 $ for Female and Male), thereby showing Male as more probable to re-offend crimes than Female. Moreover, FIF of the feature `age' is comparatively higher in magnitude for Male than Female. 
This implies that while deciding recidivism the algorithm assumes that Female individuals across different ages re-offend crimes with almost the same probability and the probability of re-offending for Male individuals highly depends on age. Thus, applying {\fvgm} to compute FIF provides us insights about sources of bias and thus, indicators to improve fairness.

\begin{figure}[t!]	
	\centering
	\vspace{-2ex}
	\subfloat{\includegraphics[scale=0.23]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_1}}
	\subfloat{\includegraphics[scale=0.23]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_0}}
	%\vspace{-0.5em}
%	\vspace{-2ex}
	\caption{\footnotesize FIF computation for COMPAS dataset. For Female (left plot) and Male, influence of `age' decreases and the PPV of the classifier increases  by different magnitudes.}\label{fairness_fvgm_fig:influence function}
%	\vspace{-2ex}
\end{figure}

### 2022-01-26 chapters/fairness/fvgm/background.tex
adding two fairness papers

	\section{Background}
	In this section, we define different fairness metrics proposed for classification. Following that, we state basics of stochastic subset sum and Bayesian networks that are the main components of our methodology.

	\textbf{Fairness in ML.}
	We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of triples $ (\nonsensitive, \sensitive, Y) $ generated from an underlying distribution $\mathcal{D}$. $ \nonsensitive \triangleq \{X_1, \dots, X_{m_1}\} $ are non-sensitive features whereas $ \sensitive \triangleq \{A_1, \dots, A_{m_2}\} $ are categorical sensitive features.  $Y \in \{0,1\}$ is the binary label (or class) of $(\nonsensitive,\sensitive)$. Each non-sensitive feature $ X_i$ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each sensitive feature $ A_j \in \{0, \dots, N_j\}  $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. 
%	\red{Therefore, the distribution $\mathcal{D}$ is the joint distribution on $(\nonsensitive,\sensitive)$, where  $\mathcal{X}_i$ and $\mathcal{A}_i$ are the marginal distributions.} 
	%Therefore, $ \mathcal{D} = \prod_{i=1}^{m} \mathcal{X}_i \prod_{j=1}^{n} \mathcal{A}_j $ denotes the product distribution of sensitive and non-sensitive features and thus considered the underlying distribution for the dataset $ D $.  Additionally, $ Y \in \{0,1\} $ denotes the binary class label of $ (\nonsensitive, \sensitive) $. Let $ \alg : (\nonsensitive, \sensitive) \rightarrow \hat{Y} $  be a binary classier trained on $ D $, where $ \hat{Y} \in \{0,1\} $ is the predicted class. The positive predictive value (PPV) of $ \alg $, denoted as $ \Pr[\hat{Y} = 1] $, is the probability of positive outcome.
 	We use $ (\mathbf{x}, \mathbf{a}) $ to denote the feature-values of  $ (\nonsensitive, \sensitive) $.  For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{m_2}] $ is called a \textit{compound sensitive group}. For example, consider $ \sensitive = $ \{race, sex\} where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}. Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group. 
 	We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \hat{Y} $. Here, $\hat{Y} \in \{0,1\}$ is the predicted class of $ (\nonsensitive, \sensitive) $.

 	We now discuss different fairness metrics in the literature based on the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.  	
 	In this paper, {\fvgm} verifies two families of fairness metrics: group fairness (first three in the following) and path-specific causal fairness.

 	\begin{enumerate}[leftmargin=*]
 		\itemsep0em 
 		\item \textit{Disparate impact} (DI): A classifier  satisfies $(1 - \epsilon)$-disparate impact if for $\epsilon \in [0,1] $,
 		$
 		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}'} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'].
 		$
 		\item \textit{Statistical Parity} (SP): A classifier satisfies $\epsilon$-statistical parity if for $\epsilon \in [0,1] $, 
 		$
 		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}']| \le \epsilon.
 		$
 		\item \textit{Equalized Odds} (EO): 	A classifier satisfies $\epsilon$-equalized odds if for $\epsilon \in [0,1] $,
 		$ \max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 0]| \le \epsilon, $ and $
 		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 1]| \le \epsilon.
 		$
 		\item \textit{Path-specific Causal Fairness} (PCF): 
 		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.  For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as 
 		$
 		\max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}', \mediator ]| \le \epsilon
 		$.
 	\end{enumerate}

 	  For all of the above metrics, lower value of $\epsilon$ indicates higher fairness demonstrated by the classifier $\mathcal{M}$. Following the observation of~\cite{ghosh2020justicia},  computing all of the aforementioned fairness metrics is equivalent to computing the maximum and minimum of the \textit{Positive Predictive Value} (PPV) of the classifier, denoted as $\Pr[\hat{Y}=1|\mathbf{A} =\mathbf{a}]$, for all compound sensitive groups $\mathbf{a}$ from $ \sensitive $. Thus, in Section~\ref{fairness_fvgm_sec:fvgm}, we focus on computing the maximum and minimum PPVs and then extend it to assess corresponding fairness metrics. We call the group for which PPV is maximum (minimum) as the \textit{most (least) favored group}.

% 	\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, 
% 	Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of PPV among sensitive groups, and prescribe this to be near zero. Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $, 
% 	\[
% 	\max_{\mathbf{a}_i, \mathbf{a}_j \in A}|\Pr[\hat{Y} =1 | A = \mathbf{a}_i] - \Pr [\hat{Y} = 1| A = \mathbf{a}_j]| \le \epsilon.
% 	\] 	
% 	
% 	In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive features $A$ given the actual class labels $Y$. In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound sensitive groups. 
% 	Lower value of equalized odds indicates better fairness.
% 	A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound sensitive groups $\mathbf{a}, \mathbf{a} \in A$,
% 	\begin{equation}
% 	\begin{split}
% 	|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1| A = \mathbf{a}, Y = 0]| &\le \epsilon, \\ 
% 	|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1| A = \mathbf{a}, Y = 1]| &\le \epsilon.\notag
% 	\end{split}
% 	\end{equation}
% 	
% 	Path-specific causal fairness (PCF) states that the outcome $ \hat{Y} $ should not depend directly on a sensitive feature, but may depend indirectly on the sensitive feature through other \textit{mediator} features[\cite{?}]. 
% 	
% 	
% 	
% 	\red{Need to clarify, group fairness and causal fairness?}
% 	
% 	
% 	\blue{It may be space consuming to discuss different fairness metrics, in which case we can mention that computing PPV is the core component of estimating different fairness metrics.}
%

	\textbf{Stochastic Subset Sum Problem ({\stochastic}).} 
	Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ w_i \in \mathbb{Z} $ be the weight of $ B_i $. Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} w_i B_i = \tau $, for a constant $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$. Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}. The \textit{counting} version of the subset-sum problem counts all $ \mathbf{b} $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $. In this paper, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\mathbf{B}|} w_i B_i \ge \tau $ where variables $ B_i $'s are stochastic. More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} w_iB_i \ge \tau] $.    Details of {\stochastic} are in Section~\ref{fairness_fvgm_sec:stochastic_sum_set_sum}.

	\textbf{Bayesian Network.}
	In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables. In this paper, we leverage an access to a Bayesian network on $(\nonsensitive, \sensitive)$ that represents the joint distribution on them. 
	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a Directed Acyclic Graph (DAG), and $\theta$ is a set of parameters that encodes the conditional probabilities induced by the joint distribution under investigation. Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable. Edges $ \mathbf{E} \in \mathbf{V} \times \mathbf{V} $ imply conditional dependencies among variables. For each variable $ V_i \in \mathbf{V} $, let $ \parent(V_i) \subseteq \mathbf{V} \setminus \{V_i\} $ denote the set of parents of $ V_i $. Given $\parent(V_i)$ and parameters $\theta$, $ V_i $ is independent of its other non-descendant variables in $\graph$. Thus, for the assignment $ v_i $ of $ V_i $ and $ \mathbf{u} $ of $ \parent(V_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\mathbf{V}$ as:

	\begin{align}
	\Pr[V_1=v_1, \dots, V_{|\mathbf{V}|}=v_{|\mathbf{V}|}] = \prod_{i=1}^{|\mathbf{V}|} \Pr[V_i = v_i | \parent(V_i) = \mathbf{u}; \theta].
	\label{fairness_fvgm_eq:BN}
	\end{align}

	\iffalse	
	$ \factors $ contains a factor $ \factor $ representing a function over the  assignment or valuation of  $ \{V_i\} \cup \parent(V_i)  $. Let $ v $ and $ \mathbf{u} $ denote an assignment of $ V_i $ and $ \parent(V_i) $, respectively. 	Formally a factor $ \factor(V_i = v, \parent(V_i) = \mathbf{u}) \triangleq  \Pr[V_i = v | \parent(V_i) = \mathbf{u}] $ denotes the conditional probability of $ V_i = v $ given $ \parent(V_i) = \mathbf{u} $. Avoiding notation clutter, we drop $ v $ and $ \mathbf{u} $ and use $ \factor(V_i, \parent(V_i)) \triangleq  \Pr[V_i| \parent(V_i)] $, when it is clear from the context.
	\fi

### 2022-01-26 chapters/fairness/fvgm/cnf_correlations.tex
adding two fairness papers

	%	\newpage

\section{Fairness Verification of CNF Classifiers with Feature-correlations} 
\label{sec:CNF_feature_correlation}
Existing verifiers such as Justicia focuses mainly on classifiers expressible as a CNF formula. But none of the existing verifiers consider correlation among the features while computing fairness metrics. In this section,  we address fairness verification of CNF-based classifiers with explicit consideration of correlations among features. In particular, we present how to encode a Bayesian network, which captures correlated features, into a SSAT-based formulation that is tailored for verifying CNF classifiers. To this end, we first discuss the basics of SSAT followed by an elaboration of the proposed methodology.

\subsection{Background: Stochastic Boolean Satisfiability (SSAT)}\label{sec:ssat}
Let $ \mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables taking assignment in $ \{0,1\}^m $. A \textit{literal} $ l_i $ is a variable $ \bool_i $ or its complement $ \neg \bool_i $. A propositional formula $\phi$ defined over $\mathbf{B}$ is in CNF if $\phi \triangleq \wedge_i C_i $   is  a conjunction of clauses and each clause $ C_i \triangleq \vee_j l_j $ is a disjunction of literals. A CNF formula $ \phi $ is satisfied (called SAT) if there is an assignment $ \sigma $ over $\mathbf{B}$ that evaluates formula $ \phi $ to $ 1 $ (or true)\textemdash at least one literal in each clause is true and all clauses are true. We additionally consider a quantification over each Boolean variable $ B_i $, denoted by $ q_i \in \{\exists, \forall, \R^{p_i}\}$, where $ \exists $ is existential, $ \forall $ is universal and $ \R $ is a randomized quantifier with $ p_i = \Pr[B_i = 1] $. The SSAT problem takes \textit{an ordered set of quantifiers} over  $\mathbf{B}$ and a CNF formula $ \phi $ and computes the \textit{probability of satisfaction} of $ \phi $ given quantifiers.  Formally, a SSAT formula $ \Phi \triangleq (\{q_i\}_{i=1}^{m}, \phi) $ and the SSAT problem computes $ \Pr[\Phi]  \in [0,1]$. The semantics of SSAT formula is in the following.

\begin{enumerate}
	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
	\item $ \Pr [\Phi] = \max_{\bool_1} \{\Pr[\Phi|_{\bool_1}], \Pr[\Phi|_{\neg \bool_1}]\}$ if $ \bool_1 $ is existentially quantified ($ \exists $), 
	\item $ \Pr [\Phi] = \min_{\bool_1} \{\Pr[\Phi|_{\bool_1}], \Pr[\Phi|_{\neg \bool_1}]\} $ if $ \bool_1 $ is universally quantified ($ \forall $), 
	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool_1}] + (1-p) \Pr[\Phi|_{\neg \bool_1}] $ if $ \bool_1 $ is randomized quantified ($\R^{p}$) with probability $p = \Pr[\bool_1 = 1]$,
\end{enumerate}

In SSAT, we recursively solve for each Boolean variable in $\mathbf{B}$ starting with the first variable $ B_1 $
where $ \Phi|_{\bool_1} $ (resp.\ $ \Phi|_{\neg \bool_1} $) is the SSAT formula with CNF $ \phi $ substituted by an assignment of $ \bool_1 $ as true (resp.\ false) and quantifiers $ \{q_i\}_{i=2}^{m} $.

In this paper, we are interested in two specific SSAT formulations: exists-random (ER) SSAT formulas and universal-random (UR) SSAT formulas. In ER-SSAT (resp.\ UR-SSAT), $ \{q_i\} $ is constructed such that  existential (resp.\ universal) quantified variables is followed by randomized quantified variables. For more details on SSAT formulas,  we refer to \cite{ghosh2020justicia,lee2017solving, lee2018solving}.

%This correspondence of existential and universal quantifiers of choice variables with the maximum and minimum probabilities in case of SSAT, motivate us to assign the choice (or sensitive) variables similar quantifiers while computing the maximum and minimum PPVs for linear classifiers.

\subsection{Methodology}
For CNF classifiers~\cite{GMM20}, SSAT is a natural choice as it computes the probability of a CNF formula  given quantifiers of variables\textemdash equivalently, the PPV of a CNF classifier. \cite{ghosh2020justicia} has proposed a SSAT based formulation for verifying a CNF classifier $ \phi_{\hat{Y}} $ with a fundamental limitation of ignoring feature-correlations, which we address now. Let $ \phi_\BN $ be a CNF formula that encodes the Bayesian Network $ \BN $. Following~\cite{ghosh2020justicia}, we discuss an encoding that compute the PPV of the classifier for the most favored sensitive group with additional consideration of features-correlation. Let $\mathbf{Q}$ denote an ordered set of quantifiers over variables in the conjoined CNF $ \phi_{\hat{Y}} \wedge \phi_\BN $. We then construct a SSAT formula $ \Phi =  (\mathbf{Q},  \phi_{\hat{Y}} \wedge \phi_\BN) $ and solve it for computing $ \max_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $. For computing $ \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ for the least favored group ,only difference is in the construction of $\mathbf{Q}$. We next discuss the construction of both $ \phi_\BN $ and $\mathbf{Q}$.

\paragraph{Encoding a Bayesian Network as a CNF Formula.}\label{sec:BN_to_CNF}
Our goal is to encode the Bayesian network $ \BN = (\graph, \factors) $ into a CNF formula $ \phi_\BN $ such that \textit{the weighted model count} of $ \phi_\BN $ exactly computes a joint probability distribution~\cite{chavira2008probabilistic}.  We note that SSAT does not allow conditional probabilities of randomized quantified variables trivially. Hence, $ \phi_\BN $ contains \textit{additional variables} to capture the conditional probabilities, as discussed next.

Let  $ G = (\mathbf{V}, E) $ where $  $  $ \mathbf{V} \subseteq \nonsensitive \cup \sensitive $ and $ \mathbf{E} \subseteq \mathbf{V} \times \mathbf{V} $. 	For each network variable $ V_i \in \mathbf{V} $, we define a Boolean \textit{indicator}  variable $ \lambda_{V_i} $ such that $ \Pr[\lambda_{V_i}] \triangleq \Pr[V_i] $. We add following constraint in $ \phi_\BN $ to establish the relation between $ \lambda_{V_i} $ and $ V_i $. 
\begin{align}
	\lambda_{V_i} \leftrightarrow V_i,
	\label{eq:indicator_constraint}
\end{align}
Intuitively, both $ \lambda_{V_i} $ and $ V_i $ are either true or false. This constraint can be trivially translated to clauses in CNF using the equivalence rule $ a \leftrightarrow b\equiv (\neg a \vee b)  \wedge (a \vee \neg a) $ for Boolean variables $ a, b$.

We now show encoding of conditional probabilities induced by parameter $ \theta $. Let $ V_i \in  \mathbf{V}  $  be a vertex in $ G $ where $ \parent(V_i) \ne \emptyset $ be $ V_i $'s parents and $ |\parent(V_i)| = k $. Additionally, let $ v $ and $ \mathbf{u} \triangleq [u_1,.., u_k] $ be an assignment of  $ V_i $ and $ \parent(V_i)  $, respectively.  To encode $ \Pr[V_i = v| \parent(V_i) = \mathbf{u}]$, we introduce auxiliary variable $ \lambda_{v,\mathbf{u}} $ and add following constraints in $ \phi_{\BN} $.
\begin{align}
	\lambda_{v,\mathbf{u}}  \wedge \bigwedge_{j=1}^{k} \lambda_{u_j} \rightarrow \lambda_v
	\label{eq:factor_pos}
\end{align}
\begin{align}
	\neg \lambda_{v,\mathbf{u}}  \wedge \bigwedge_{j=1}^{k} \lambda_{u_j} \rightarrow \neg \lambda_v
	\label{eq:factor_neg}
\end{align}

where $ \lambda_v \equiv \lambda_{V_i} $. Moreover, $ \lambda_{u_j} $ is the indicator variable corresponding to the $ j^\text{th} $ parent in $ \parent(V_i) $. In the above two constraints,	for a fixed assignment $ \mathbf{u} $ of parents $ \parent(V_i) $, both $ \lambda_v $ and $ \lambda_{v,\mathbf{u}} $ are either true or false.  Hence, these two constraints encode the conditional probability of $ V_i = v $ given $ \parent(V_i) = \mathbf{u} $ using $ \Pr[\lambda_{v,\mathbf{u}}] = \Pr[V_i = v| \parent(V_i) = \mathbf{u}]$. Both constraints can be translated to CNF clauses trivially. For example, Eq.~\ref{eq:factor_pos} is translated as $ \neg \lambda_{v,\mathbf{u}}  \vee \bigvee_{j=1}^{k} \neg \lambda_{u_j} \vee \lambda_v $. We next analyze the complexity of $ \phi_\BN $ in terms of the number of variables and clauses.

\begin{lemma}
	For a Bayesian network $ \BN = (\graph, \factors) $ defined over $ n $ Boolean network variables, the encoded CNF formula $ \phi_\BN $ has $ n + |\factors| $  variables and $ 2(n + |\factors|) $ clauses. 
\end{lemma}
\begin{proof}
	Since the DAG in the Bayesian network has $ n $ vertices, we consider $ n $ indicator variables. Moreover, for encoding conditional probabilities, we consider $ |\theta| $ auxiliary variables where parameter $ \theta $ denotes the number of distinct conditional probabilities in the network. Hence, total variables in $ \phi_\BN $ is $ n + |\factors|  $.

	According to Eq.~\eqref{eq:indicator_constraint},~\eqref{eq:factor_pos},~\eqref{eq:factor_neg}, there are $2( n + |\theta|) $ clauses in $ \phi_\BN $ as discussed in Section~\ref{sec:BN_to_CNF}.

\end{proof}

\begin{figure}[!t]
	\begin{center}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_DT}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_DT}}\\
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_DT}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_DT}}

	\end{center}

	\caption{
		Computing PPV of Decision Tree (DT) classifiers using {\framework} and Justicia. {\framework} incorporates correlated features represented as a Bayesian network in Figure~\ref{fig:synthetic_bn}. We also present the effect of sample size on different fairness metrics such as DI and SP for DT classifier. }
	\label{fig:synthetic_results}
\end{figure}

\paragraph{Constructing Quantifiers $ \mathbf{Q} $.}
We now discuss the ordered set of quantifiers for a SSAT formula containing CNF $ \phi_{\hat{Y}} \wedge \phi_\BN $\textemdash the solution of which constitutes the maximum (minimum) PPV of a CNF classifier. $ \phi_{\hat{Y}} \wedge \phi_\BN $ contains four different variables : (i) sensitive variables $ \sensitive $, (ii) non-sensitive variables $ \nonsensitive $, (iii) indicator variables $  \lambda_{V_i} $, and (iv) auxiliary variables $ \lambda_{v,\mathbf{u}} $. Among them, (iii) and (iv) are associated with $ \phi_\BN $ and the rest for $ \phi_{\hat{Y}} $. For computing the maximum PPV of the classifier, we construct  an exists-random-exists (ERE) SSAT formula with quantifiers $\mathbf{Q}$ as follows: we set sensitive features $ \sensitive $ with existential quantifiers in the beginning of $ \mathbf{Q} $ followed by $  \lambda_{V_i}, \lambda_{v,\mathbf{u}} $ and $ X_j \in \nonsensitive \setminus \mathbf{V} $ with randomized quantifiers. Since sensitive variables are existentially quantified\textemdash similar to stochastic subset-sum problem\textemdash the solution of SSAT formula is maximized.
Finally, remaining variables appearing in the Bayesian network such as $ X_i \in \mathbf{V} $ are existentially quantified in $\mathbf{Q}$ as their assignment is fixed by indicator variables $ \lambda_{V_i} $. In contrast, for computing the minimum PPV of the classifier, we consider an universal-random-exists (URE) SSAT formula  where we set sensitive features $ \sensitive $ as universal quantifiers with all other quantifiers remaining same.

\iffalse
\subsection{Verifying Classifiers with Non-Boolean Features} {\framework} can verify beyond non-Boolean features. For binary decision tree classifiers, real-valued features are discretized in each node of the tree by comparing the feature against a constant threshold. Thus, each node constitutes a Boolean feature and the tree itself forms a CNF formula, as shown by~\cite{ghosh2020justicia}.
\fi

### 2022-01-26 chapters/fairness/fvgm/conclusion.tex
adding two fairness papers

\section{Conclusion}
In this paper, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem. {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately. We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. 
As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}. %We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF). 
Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.

### 2022-01-26 chapters/fairness/fvgm/dp_illustration.tex
adding two fairness papers

\begin{figure*}[t!]
	\centering

	\subfloat[Known marginal  probabilities.]{
%		\centering
		\scalebox{0.75}{
		\begin{tikzpicture}[>=stealth',shorten >=1pt, on grid,initial/.style={}]
			\node[state, align=center, fill=existential] (T1) {$ 1, 2 $\\$ 0.55 $};
			\node[state, align=center, label={[align=left,right]0:$ 0.85\Pr[Q] + 0.35\Pr[\neg Q] $\\$ =0.55 $}] (T2) [below = 1.6 cm of T1] {$2,1$\\$ 0.55 $};
			\node[state, align=center] (T3) [below left = 1.5 cm and 1.8 cm of T2] {$ 3,0 $\\$ 0.85$};
			\node[state, align=center, fill=terminate] (T4) [below left = 1.5 cm and 1 cm of T3] {$ 4, -1$\\$ 1 $};
			\node[state, align=center] (T5) [below right = 1.5 cm and 1 cm of T3] {$ 4, 0$\\$ 0.7 $};
			\node[state, align=center, fill=terminate] (T6) [below left = 1.5 cm and 1 cm of T5] {$ 5,1 $\\ $0 $};
			\node[state, align=center, fill=terminate] (T7) [below right = 1.5 cm and 1 cm of T5] {$ 5,0 $\\$ 1 $};
			\node[state, align=center] (T8) [below right = 1.5 cm and 1.6 cm of T2] {$ 3,1 $\\$ 0.35 $};
			\node[state, align=center, fill=collision] (T9) [below left = 1.5 cm and 1 cm of T8] {$ 4, 0$\\$ 0.7$};
			\node[state, align=center] (T10) [below right = 1.5 cm and 1 cm of T8] {$ 4, 1$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T11) [below left = 1.5 cm and 1 cm of T10] {$ 5, 2$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T12) [below right = 1.5 cm and 1 cm of T10] {$ 5,1 $\\$ 0 $};

			\tikzset{every node/.style={fill=white}}
			\path (T1) edge [right] node {$P$}  (T2);
			\path (T2) edge [left] node {$Q$}  (T3);
			\path (T3) edge [left] node {$R$}  (T4);
			\path (T3) edge [right] node {$\neg R$}  (T5);
			\path (T5) edge [left] node {$S$}  (T6);
			\path (T5) edge [right] node {$\neg S$}  (T7);
			\path (T2) edge [right] node {$\neg Q$}  (T8);
			\path (T8) edge [left] node {$R$}  (T9);
			\path (T8) edge [left] node {$\neg R$}  (T10);
			\path (T10) edge [left] node {$S$}  (T11);
			\path (T10) edge [left] node {$\neg S$}  (T12);
			\end{tikzpicture}
		}
\label{fairness_fvgm_fig:example_dp}}
\hspace*{-0.5cm}
\subfloat[Probabilities computed with Bayesian network.]
{  
%		\centering
		\scalebox{0.75}{
			\begin{tikzpicture}[>=stealth',shorten >=1pt, on grid,initial/.style={}]
			\node[state, align=center, fill=existential, accepting] (T1) {$ 1, 2 $\\$ 0.65 $};
			\node[state, align=center, accepting, label={[align=left,right]0:$ 0.85\Pr[Q|P] + $ \\ $   0.35\Pr[\neg Q|P] $\\$ =0.65 $}] (T2) [below left = 1.6 cm and 2 cm of T1] {$2,1$\\$ 0.65 $};
			\node[state, align=center] (T3) [below left = 1.5 cm and 1.8 cm of T2] {$ 3,0 $\\$ 0.85$};
			\node[state, align=center, fill=terminate] (T4) [below left = 1.5 cm and 1 cm of T3] {$ 4, -1$\\$ 1 $};
			\node[state, align=center] (T5) [below right = 1.5 cm and 1 cm of T3] {$ 4, 0$\\$ 0.7 $};
			\node[state, align=center, fill=terminate] (T6) [below left = 1.5 cm and 1 cm of T5] {$ 5,1 $\\ $0 $};
			\node[state, align=center, fill=terminate] (T7) [below right = 1.5 cm and 1 cm of T5] {$ 5,0 $\\$ 1 $};
			\node[state, align=center] (T8) [below right = 1.5 cm and 1.6 cm of T2] {$ 3,1 $\\$ 0.35 $};
			\node[state, align=center, fill=collision] (T9) [below left = 1.5 cm and 1 cm of T8] {$ 4, 0$\\$ 0.7$};
			\node[state, align=center] (T10) [below right = 1.5 cm and 1 cm of T8] {$ 4, 1$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T11) [below left = 1.5 cm and 1 cm of T10] {$ 5, 2$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T12) [below right = 1.5 cm and 1 cm of T10] {$ 5,1 $\\$ 0 $};

			\node[state, align=center, accepting] (T13) [below right = 1.6 and 2 cm of T1] {$2,2$\\$ 0.11 $};
			\node[state, align=center, fill=collision] (T14) [below left = 1.5 and 1 cm of T13] {$3,1$\\$ 0.35 $};
			\node[state, align=center] (T15) [below right = 1.5 and 1 cm of T13] {$3,2$\\$ 0 $};
			\node[state, align=center, fill=collision] (T16) [below left = 1.5 and 1 cm of T15] {$4,1$\\$ 0 $};
			\node[state, align=center] (T17) [below right = 1.5 and 1 cm of T15] {$4,2$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T18) [below left = 1.5 and 1 cm of T17] {$5,3$\\$ 0 $};
			\node[state, align=center, fill=terminate] (T19) [below right = 1.5 and 1 cm of T17] {$5,2$\\$ 0 $};

			\node[state, align=center] (P) [right = 5 cm of T1] {$P$};
			\node[state, align=center] (Q) [right = 2 cm of T15] {$Q$};

			\tikzset{every node/.style={fill=white}}
			\path (T1) edge [right] node {$P$}  (T2);
			\path (T2) edge [left] node {$Q$}  (T3);
			\path (T3) edge [left] node {$R$}  (T4);
			\path (T3) edge [right] node {$\neg R$}  (T5);
			\path (T5) edge [left] node {$S$}  (T6);
			\path (T5) edge [right] node {$\neg S$}  (T7);
			\path (T2) edge [left] node {$\neg Q$}  (T8);
			\path (T8) edge [left] node {$R$}  (T9);
			\path (T8) edge [left] node {$\neg R$}  (T10);
			\path (T10) edge [left] node {$S$}  (T11);
			\path (T10) edge [left] node {$\neg S$}  (T12);
			\path (T1) edge [right] node {$\neg P$}  (T13);
			\path (T13) edge [right] node {$Q$}  (T14);
			\path (T13) edge [right] node {$\neg Q$}  (T15);
			\path (T15) edge [right] node {$R$}  (T16);
			\path (T15) edge [right] node {$\neg R$}  (T17);
			\path (T17) edge [right] node {$S$}  (T18);
			\path (T17) edge [right] node {$\neg S$}  (T19);

			\path (P) edge [->] node [left=0.1cm] {\parbox{01.2cm}{Bayesian\\network}}(Q);
		\end{tikzpicture}
	}
	\label{fairness_fvgm_fig:example_BN}
}

\caption{\footnotesize Search tree representation of {\stochastic} for computing the maximum PPV of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ . Each node is labeled by $ (i,\tau') $, where $ i $ is the index of $ \mathbf{B} $ and $ \tau' $ is the residual threshold. The tree is explored using Depth-First Search (DFS) starting with left child. Within a node, the value in the bottom denotes $ \mathsf{dp}(i, \tau') $ that is solved recursively based on sub-problems $ \mathsf{dp}(i+1, \cdot) $ in child nodes. 	Yellow nodes denote \textit{existential} variables and all other nodes are  \textit{random} variables. Additionally, a green node denotes a collision, in which case a previously computed $ \mathsf{dp} $ solution is returned. Leaf nodes (gray) are computed based on terminating conditions in Eq.~\ref{fairness_fvgm_eq:dp_terminus}. In Figure~\ref{fairness_fvgm_fig:example_BN},  nodes with double circles, such as $ \{(1,2), (2,1), (2,2)\} $,  are enumerated exponentially to compute conditional probabilities from the Bayesian network.}
\label{fairness_fvgm_fig:example_tree_exploration}
\end{figure*}

### 2022-01-26 chapters/fairness/fvgm/evaluations.tex
adding two fairness papers

\section{Empirical Performance Analysis}
\label{fairness_fvgm_sec:experiments}
In this section, we empirically evaluate the performance of {\fvgm}. We first present the experimental setup and the objective of our experiments, followed by experimental results.\footnote{Due to page limit, we present additional experimental results, proof of lemmas, and miscellaneous details in Appendix.}

\begin{figure}[t!]
	\begin{center}		
		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_LR_time_}}
		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_SVM_time_}}		
	\end{center}
	%	\vspace{-2ex}
	\caption{\footnotesize A cactus plot to present the scalability of different fairness verifiers. The number of solved benchmarks are on the $ X $-axis and the time taken on the $ Y $-axis; a point $ (x,y) $ implies that a verifier takes less than or equal to $ y $ seconds to compute fairness metrics of $ x $ many benchmarks. We consider $ 100 $ benchmarks generated from $ 5 $ real-world datasets using $ 5 $-fold cross-validation. In each fold, we consider $\{25, 50, 75, 100\} $ percent of non-sensitive features.}\label{fairness_fvgm_fig:scalability_exp}
	%	\vspace{-2ex}
\end{figure}

\textbf{Experimental Setup.}
We  implement a  prototype of {\fvgm} in Python (version 3.8).  
We deploy the Scikit-learn library for learning linear classifiers such as Logistic Regression (LR) and Support Vector Machine (SVM) with linear kernels. We perform five-fold cross-validation. While the classifier is trained on continuous features, we discretize them to Boolean features to be invoked by {\fvgm}. While discretizing, we apply a gird-search to estimate the best bin-size within a maximum bin of $ 10 $. To convert the coefficients of features into integers, we employ another grid-search to choose the best multiplier within $ \{1,2, \dots, 100\} $. For learning a Bayesian network on the converted Boolean data, we deploy the PGMPY library~\cite{ankan2015pgmpy}. For network learning, we apply a Hill-climbing search algorithm that learns a DAG structure by optimizing K2 score~\cite{koller2009probabilistic}. For estimating parameters of the network, we use Maximum Likelihood Estimation (MLE) algorithm.

We compare {\fvgm} with three existing fairness verifiers: Justicia~\cite{ghosh2020justicia}, FairSquare~\cite{albarghouthi2017fairsquare}, and VeriFair~\cite{bastani2019probabilistic}. 
%Additionally, we have verified two fairness algorithms: reweighing algorithm (RW)~\cite{kamiran2012data} and the optimized pre-processing algorithm (OP)~\cite{calmon2017optimized}, and a fairness poisoning-attack algorithm~\cite{solans2020poisoning}.

\subsection{Scalability Analysis}
\textbf{Benchmarks.} We perform the scalability analysis on five real-world datasets studied in fair ML literature: UCI Adult, German-credit~\cite{DK2017}, COMPAS~\cite{angwin2016machine}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}). We consider $ 100 $ benchmarks generated from 5 real-world datasets and report the computation times (for DI and SP) of different verifiers.

\textbf{Results.} In Figure~\ref{fairness_fvgm_fig:scalability_exp}, we present the scalability results of different verifiers. First, we observe that FairSquare often times out ($ =900 $ seconds) and can solve $ \le 5 $ benchmarks. This indicates that SMT-based reduction for linear classifiers cannot scale. Similarly, SSAT-based verifier Justicia that performs pseudo-Boolean to CNF translation for linear classifiers, times out for around $  20 $ out of $ 100 $ benchmarks. Sampling-based framework, VeriFair, has comparatively better scalability than SMT/SSAT based frameworks and can solve more than $ 90 $ benchmarks. Finally, {\fvgm} achieves impressive scalability by solving all $ 100 $ benchmarks with $ 1 $ to $ 2 $ orders of magnitude runtime improvements than existing verifiers. Therefore,\textit{ {\stochastic}-based framework {\fvgm} proves to be highly scalable in verifying fairness properties of linear classifiers than the state-of-the-art.}

\begin{figure}[t!]
	\begin{center}	
		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_LR_01}}
		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_SVM_01}}		
		%		\subfloat[]{\includegraphics[scale=0.25]{figures/fairness/fvgm/sanity_DI_LR_005}}
		%		\subfloat[]{\includegraphics[scale=0.25]{figures/fairness/fvgm/sanity_DI_SVM_005}}
		%				
	\end{center}
	%	\vspace{-2ex}
	\caption{\footnotesize Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features. {\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of DI for LR and SVM classifiers.}\label{fairness_fvgm_fig:sanity_exp}
	%	\vspace{-2ex}
\end{figure}
\subsection{Accuracy Analysis}
\noindent\textbf{Benchmark Generation.} To perform accuracy analysis, we require the ground truth, which is not available for real-world instances. Therefore, we focus on generating synthetic benchmarks for analytically computing the ground truth of different fairness metrics, such as DI, from the known distribution of features. In each benchmark, we consider $ n \in \{2, 3, 4, 5\} $ features including one Boolean sensitive feature, say $ A $, generated from a Bernoulli distribution with mean $ 0.5 $.  We generate non-sensitive features $ X_i $ from Gaussian distributions such that   $ \Pr[X_i | A = 1] \sim \mathcal{N}(\mu_i, \sigma^2) $ and $ \Pr[X_i | A = 0] \sim \mathcal{N}(\mu_i', \sigma^2) $, where $ \mu_i, \mu_i' \in [0,1] $, $ \sigma = 0.1 $, and $ \mu_i, \mu_i' $ are chosen from a uniform distribution in $ [0,1] $. Finally, we create label $ Y = \mathds{1}[ \sum_{i=1}^{n-1} X_i \ge 0.5 \sum_{i=1}^{n-1} (\mu_i + \mu_i')] $ such that $ Y $ does not directly depend on the sensitive feature. For each $ n $, we generate $ 100 $ random benchmarks, learn LR and SVM classifiers on them, and compute DI\footnote{  
To analytically compute DI, let the coefficients of the classifier be $ w_i $ for $ X_i $ and $ w_A $ for $ A $, and bias be $ \tau $. Since all non-sensitive features are from Gaussian distributions, we compute the probability of the predicted class $ \Pr[\hat{Y} | A = 1]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i, \sigma_{\hat{Y}}^2) $ and $ \Pr[\hat{Y} | A = 0]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i', \sigma_{\hat{Y}}^2) $ with $ \sigma_{\hat{Y}}^2 =   (\sum_{i=1}^{n-1}w_i^2)\sigma^2 $. Hence, the PPV of the classifier  is $  1 - \mathsf{CDF}_{\hat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\hat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function. Finally, we compute DI by taking the ratio of the minimum and the maximum of the two PPVs.}
using different verifiers.

\textbf{Results.} 
%\begin{enumerate}[leftmargin=*]
%	\itemsep0em 
%	\item How \textit{accurate} is {\fvgm} w.r.t. existing verifiers?
%	\item How \textit{scalable} is {\fvgm} w.r.t. existing verifiers?
%	%\item Can {\fvgm} be applied to fairness problems such as \textit{verifying} different \textit{fairness metric}s for different \textit{fair ML algorithms} and \textit{fairness attacks}, and also \textit{detecting the sources of bias} due to different subset of features?
%\end{enumerate}
%To summarize our experimental results, {\fvgm} is more accurate and scalable than existing fairness verifiers in verifying linear classifiers on benchmark datasets. %{\fvgm} can verify diverse set of fairness metrics for multiple fairness-enhancing algorithms and depreciation attacks. {\fvgm} also computes fairness influence functions that can detect bias in the level of individual features. 
%\red{Due to limited space, we discuss additional experiments including verification of CNF-based classifiers with feature correlations and performance analysis of encoding Bayesian networks in Appendix.}
 We  assess the accuracy of the competing verifiers in estimating fairness metrics, specifically DI with LR and SVM classifiers. In Figure~\ref{fairness_fvgm_fig:sanity_exp}, {\fvgm} computes DI closest to the \textit{Exact} value for different number of features and both type of classifiers. In contrast, Justicia, FairSquare, and VeriFair measure DI far from the \textit{Exact} because of ignoring correlations among the features. For example, for SVM classifier with  $ n = 5 $ (right plot in Figure~\ref{fairness_fvgm_fig:sanity_exp}), \textit{Exact} DI is $ 0.089 $ (average over 100 random benchmarks). Here, {\fvgm} computes DI as $ 0.094 $, while all other verifiers compute DI as at least $ 0.233 $. Therefore, \textit{{\fvgm} is more accurate than existing verifiers as it explicitly considers correlations among features}.

\iffalse
\begin{table}[!t]

		\centering
		\small
		\caption{Verification of fairness algorithms using {\fvgm}. Numbers in bold refer to fairness improvement. RW and OP refer to reweighing and optimized-preprocessing algorithms respectively.  }\label{fairness_fvgm_tab:fair_algo_verification}
		\setlength{\tabcolsep}{.2em}

			\begin{tabular}{lllrrrrrrrrrrrrr}

				\toprule
				Dataset & Sensitive & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta$ EO\\
				\midrule

				\multirow{4}{*}{Adult}&\multirow{2}{*}{race}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
				&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
				\cmidrule{2-7}
				&\multirow{2}{*}{sex}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
				&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\

				\midrule
				\multirow{4}{*}{COMPAS}&\multirow{2}{*}{race}&RW&$ \textbf{0.13} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.06} $\\
				&&OP&$ \textbf{0.15} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ \textbf{-0.05} $\\
				\cmidrule{2-7}
				&\multirow{2}{*}{sex}&RW&$ \textbf{0.1} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ 0.04 $\\
				&&OP&$ \textbf{0.09} $&$ \textbf{-0.04} $&$ \textbf{-0.04} $&$ \textbf{-0.03} $\\

				\midrule
				\multirow{4}{*}{German}&\multirow{2}{*}{age}&RW&$ \textbf{0.52} $&$ \textbf{-0.53} $&$ \textbf{-0.52} $&$ \textbf{-0.47} $\\
				&&OP&$ \textbf{0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.53} $&$ \textbf{-0.51} $\\
				\cmidrule{2-7}
				&\multirow{2}{*}{sex}&RW&$ -0.06 $&$ 0.06 $&$ 0.06 $&$ 0.02 $\\
				&&OP&$ -0.12 $&$ 0.12 $&$ 0.12 $&$ 0.07 $\\

				\bottomrule
		\end{tabular}
\end{table}
\fi

\iffalse
\textbf{Verification: Verifying Fairness Attacks and Fairness-enhancing Algorithms.}
We now demonstrate the applications of {\fvgm} starting with verifying fairness attacks. We instantiate a poisoning-attack algorithm that injects a small fraction of poisoned samples to the training data such that the classifier becomes relatively unfair~\cite{solans2020poisoning}. We use this algorithm to add $x, y, z$ number of poisoned samples and to measure the corresponding DI and SP. From Figure~\ref{fairness_fvgm_fig:fairness_attacks}, we observe that {\fvgm} verifies that DI of the classifier decreases and SP increases, i.e.\textit{ the classifier becomes more unfair},\textit{ as the number of poisoned samples increases}. Therefore, {\fvgm} shows the potential of being deployed in safety-critical applications to detect fairness attacks. For example, if we set $DI=0.9$ as an acceptable threshold of DI, {\fvgm} can raise an alarm once $xx$ poisoned samples are added.

We have employed {\fvgm} in verifying fairness-enhancing algorithms that pre-process the dataset to ameliorate bias.  In Table~\ref{fairness_fvgm_tab:fair_algo_verification}, We report the effect of fairness algorithms w.r.t. four fairness metrics: disparate impact (DI), path-specific causal fairness (PCF), statistical parity (SP) and equalized odds (EO). Note that, fairness is improved if DI increases and rest of the metrics decrease. We observe that in most instances for Adult and COMPAS dataset, reweighing (RW) and optimized pre-processing (OP) algorithms are successful in improving fairness. Except the unfairness regarding the sensitive feature `sex', where both of the algorithms fail. Thus, {\fvgm} verifies the enhancement and decrement in fairness of a classifier on a dataset as promised. 
%\iffalse
%More precisely, {\fvgm} guarantees that while these algorithms improve fairness on  a finite dataset, say a training dataset, the improvement also generalizes to the distribution of the dataset.
%\fi

\textbf{Tracing Sources of Unfairness: Computing Fairness Influence Function (FIF).}
In Figure~\ref{fairness_fvgm_fig:influence function}, we have computed the fairness influence function for all features in COMPAS dataset, separately for two sensitive groups: Female (`sex' $ = 1 $) and Male. This dataset concerns with the likelihood of a person of re-offending crimes within next two years. We first observe that the base PPV values are different for the two groups ($ 0.46 $ vs $ 0.61 $ for Female and Male), thereby showing Male as more probable to re-offend crimes than Female. Moreover, FIF of the feature `age' is comparatively higher in magnitude for Male than Female. 
This implies while deciding recidivism the algorithm assumes that Female individuals across different ages re-offend crimes with almost same probability and the probability of re-offending for Male individuals highly depends on age.
Thus, applying {\fvgm} to compute provides us insights about sources of bias and thus, indicators to improve fairness.
\fi

\iffalse
\begin{figure}[t!]
		\centering
		\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/disp_fairness_attack}}
		\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/stat_fairness_attack}}
		\caption{Verifying fairness poisoning attack using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
\end{figure}
\fi

\begin{table}[!t]
	\begin{minipage}{0.6\columnwidth}
		\centering
		\small

		\setlength{\tabcolsep}{.25em}

		\begin{tabular}{lllrrrrrrrrrrrrr}

			\toprule
			$\mathbf{D}$ & $ \sensitive $ & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta $EO\\
			\midrule

			\multirow{4}{*}{\rotatebox[origin=c]{90}{Adult}}&\multirow{2}{*}{\rotatebox[origin=c]{90}{race}}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
			&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
			\cmidrule{2-7}
			&\multirow{2}{*}{\rotatebox[origin=c]{90}{sex}}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
			&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\

			\bottomrule
		\end{tabular}

		\caption{\footnotesize Verification of fairness algorithms using {\fvgm}. $ \mathbf{D} $ and $ \sensitive $ denote dataset and sensitive features.  RW and OP refer to reweighing and optimized-preprocessing algorithms. Numbers in bold refer to fairness improvement.   }\label{fairness_fvgm_tab:fair_algo_verification}

	\end{minipage}\hspace*{1em}
	\begin{minipage}{0.4\columnwidth}
		\centering
		\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/disp_fairness_attack}}
		\captionof{figure}{\footnotesize Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
	\end{minipage}\hspace*{1em}
%	\vspace{-2ex}

\end{table}

### 2022-01-26 chapters/fairness/fvgm/fairness_influence_functions.tex
adding two fairness papers

%\clearpage
\section{Fairness Influence Functions (FIF)}
Now, we present an elaborate discussion on computing fairness influence functions.
A Fairness Influence Function, denoted as $ \mathsf{FIF}(\cdot) $, is computed with respect to a \textit{quantity of interest}, for example, the  PPV of the classifier or different fairness metrics such as DI and SP. Let $ \mathbf{S}  \subseteq \nonsensitive $ be a set of non-sensitive features, for which we are interested in computing their influence. A general approach to compute $ \mathsf{FIF}(\mathbf{S}) $ is to replace each feature in $ \mathbf{S} $ with random values and report differences in the quantity of interest~\cite{datta2016algorithmic}. Let $ \mathcal{X}'_i $ denote the modified marginal distribution where we replace feature $ X_i \in \mathbf{S} $ with random values. For example, $ \mathcal{X}'_i $ can be viewed as an uniform distribution within the support of $ X_i $. We then define the modified product distribution, denoted as $ \mathcal{D}_{-\mathbf{S}} $ by combining all marginal distributions, where each feature in $ \mathbf{S} $ has a modified distribution. 
\[
\mathcal{D}_{-\mathbf{S}} = \prod_{i | X_i \in \mathbf{S}} \mathcal{X}'_i \prod_{i | X_i \in X \setminus \mathbf{S}} \mathcal{X}_i \prod_{j=1}^{n} \mathcal{A}_j 
\]
We first give a general definition of influence function of $ \mathbf{S} $ on a quantity, say $ Q $, in the following. 
\[
\mathsf{FIF}(\mathbf{S}) \triangleq Q(\mathcal{D}) - Q(\mathcal{D}_{-\mathbf{S}})
\]
Intuitively, influence of $ \mathbf{S} $ is the difference in a quantity computed for the original distribution $ \mathcal{D} $ and the modified distribution $ \mathcal{D}_{-\mathbf{S}} $. In the following, we define influence function in terms of PPV of the classifier. 
\[
\mathsf{FIF}_{\mathsf{PPV}}(\mathbf{S}) \triangleq \Pr[\hat{Y} = 1 | \mathcal{D}] - \Pr[\hat{Y} = 1 | \mathcal{D}_{-\mathbf{S}}]
\]
Influence function also generalizes to PPV specific to compound sensitive groups. For a sensitive group $ \mathbf{a} \in \sensitive $, we define group-specific influence as follows.

\[
\mathsf{FIF}_{\mathsf{PPV},\; \mathbf{a}}(\mathbf{S}) \triangleq \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mathcal{D}] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a},  \mathcal{D}_{-\mathbf{S}}]
\]

\paragraph{Experimental Analysis.} We empirically instantiate computation of FIF and its consequences for a logistic regression classifier on COMPAS and Adult datasets.

For both the datasets, we consider biological `sex' as the sensitive features.
In both cases, we denote the sensitive groups `male' and `female' as `sex=0' and `sex=1', respectively. 
FIFs computed for these sensitive groups show influence of different features and their relevant disparities. We illustrate the results in Figure~\ref{fig:fif}.

Figure~\ref{fig:fif_a} illustrates that the PPV value for the `male' population in the dataset is $0.61$. If we now replace the feature `age' with uniformly random values, the PPV for the same group becomes $0.46$, Thus, the FIF of the feature `age' for `male' group is $0.15$.
Similarly, all the red bars in Figures~\ref{fig:fif_a}-\ref{fig:fif_b}, ~\ref{fig:fif_c}-\ref{fig:fif_d} indicate a positive influence of that feature and the green bars indicate a negative influence of the feature on the individual getting classified to $\hat{Y}=1$.

In Figures~\ref{fig:fif_c} and~\ref{fig:fif_f}, we illustrate the influence of different features on the disparate impact between the two groups `male' and `female'. 
The green bar indicates that removing a feature causes an increase in DI, i.e. fairness in classification, and the red bar indicates the opposite. Alternatively, we can conclude that bigger is the green bar for a feature, higher is the bias-inducing effect of it.

Thus, from Figure~\ref{fig:fif_c}, we conclude that `age' is the most bias-inducing feature among the two groups `male' and `female'.
From Fig~\ref{fig:fif_a} and~\ref{fig:fif_b}, we observe that `age' is a decisive feature for `male' while it is comparatively insignificant for `female'.

In case of Adult dataset, the PPV for `male' and `female' groups are 0.45 and 0.3 respectively. Among all the features, we observe that `race' has a positive impact on the classification for both the groups. This means removing the `race' information decreases the chance of getting classified to higher economic group. On the other hand, `capital-gain' has the highest influence on the classification for both groups. It is also the most biased-inducing feature as it varies most disparately among the two groups as the `capital-gain' differs significantly between two groups. 
\begin{figure}[t!]
	\begin{center}

		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_0}\label{fig:fif_a}}
		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_1}\label{fig:fif_b}}\\
		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_0}\label{fig:fif_d}}
		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_1}\label{fig:fif_e}}\\
		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_DI}\label{fig:fif_c}}
		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_DI}\label{fig:fif_f}}\\

	\end{center}
	\caption{Extended results on computing feature influence functions (FIF) for COMPAS  and Adult dataset.}\label{fig:fif}
\end{figure}

### 2022-01-26 chapters/fairness/fvgm/framework.tex
adding two fairness papers

\section{{\fvgm}: Fairness Verification with Graphical Models}\label{fairness_fvgm_sec:fvgm}

In this section, we present {\fvgm}, a fairness verification framework for linear classifiers that accounts for correlated features represented as a graphical model. The core idea of verifying fairness of a classifier is to compute the PPV of the classifier with respect to all compound sensitive groups. To this end, {\fvgm} solves a stochastic subset sum problem, {\stochastic}, that is equivalent to computing the PPV of the classifier for the most (and the least) favored sensitive group. In this section, we first define {\stochastic} and present an efficient dynamic programming solution for {\stochastic}. We then extend {\stochastic} to consider correlated features as input. We conclude by discussing fairness verification based on the solution of {\stochastic} in practice.

\textbf{Problem Formulation.}	
Given a linear classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \hat{Y} $ and a probability distribution $ \mathcal{D} $ of $ \nonsensitive \cup \sensitive $, our objective is to compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ with respect to $ \mathcal{D} $. In this study, we express a linear classifier $\mathcal{M}$ as

\[	\hat{Y} = \mathds{1}\Big[\sum_{i} w_{X_i}X_i + \sum_{j} w_{A_j}A_j \ge \tau\Big].\]

Here, $ w $ denotes the weight (or coefficients) of a feature, $ \tau $ denotes the bias of the classifier, and $\mathds{1}$ is an indicator function. Hence, the prediction $ \hat{Y} =1 $ if and only if the inner inequality holds.
Thus, computing the maximum (resp.\ minimum) PPV is equivalent to finding out the assignment of $A_j$'s for which the probability of satisfying the inner inequality is highest (resp.\ lowest). We reduce this computation into an instance of {\stochastic}. To perform this reduction, we assume  weights $ w $ and bias $ \tau $ as integers, and features $\nonsensitive \cup \sensitive $ as Boolean. In Sec.~\ref{fairness_fvgm_sec:practical}, we relax these assumptions and extend to the practical settings.

\subsection{$ \stochastic $: Stochastic Subset Sum Problem}\label{fairness_fvgm_sec:stochastic_sum_set_sum}
Now, we formally describe the specification and semantics of {\stochastic}.
{\stochastic} operates on  a set of Boolean variables $ \mathbf{B} = \{B_i\}_{i=1}^{n} \in \{0,1\}^{n} $, where $ w_i \in \mathbb{Z} $ is the weight of $ B_i $, and $n \triangleq |\mathbf{B}|$. Given a constant threshold $ \tau \in \mathbb{Z} $, {\stochastic} computes the \textit{probability} of a subset of $ \mathbf{B} $ with sum of weights of non-zero variables to be at least $ \tau $. Formally,

\[S(\mathbf{B}, \tau) \triangleq \Pr\Big[\sum_{i} w_iB_i \ge \tau \Big] \in [0,1].\]

Aligning with terminologies in stochastic satisfiability (SSAT)~\cite{littman2001stochastic}, we categorize the variables $ \mathbf{B} $ into two types: (i) \textit{chance variables} that are stochastic and have an associated probability of being assigned to $ 1 $ and (ii) \textit{choice variables} that we  optimize while computing $ S(\mathbf{B}, \tau) $.  To specify the category of variables, we consider a \textit{quantifier} $ q_i \in \{\R^{p_i}, \exists, \forall\} $ for each $ B_i $. Elaborately, $ \R^{p} $ is a \textit{random quantifier} corresponding to a chance variable $ B $, where  $ p\triangleq \Pr[B = 1]$. In contrast, $ \exists $ is an \textit{existential quantifier} corresponding to a choice variable $ B $ such that a Boolean assignment of $ B $  \textit{maximizes}  $ S(\mathbf{B}, \tau) $. Finally, $ \forall $ is an \textit{universal quantifier} for a choice variable $ B $ that fixes an assignment to $ B $ that \textit{minimizes} $ S(\mathbf{B}, \tau) $.

Now, we formally present the semantics of $ S(\mathbf{B}, \tau) $ provided that each variable $ B_i $ has weight $ w_i $ and quantifier $ q_i $. Let  $ \mathbf{B}[2:n] \triangleq \{B_j\}_{j=2}^{n} $ be the subset of $\mathbf{B}$ without the first variable $ B_1 $. Then $ S(\mathbf{B}, \tau) $ is recursively defined as:

\begin{align}\label{fairness_fvgm_eq:semantics_recurse}
 \hspace*{-1.2em} S(\mathbf{B},\tau) =
 \begin{cases}
 \mathds{1}[ \tau \le 0 ], \; \text{if } \textbf{B} = \emptyset \\
 S(\mathbf{B}[2:n],\tau - \max\{w_1, 0\}), \; \text{if } q_1 = \exists\\
 S(\mathbf{B}[2:n],\tau - \min\{w_1, 0\}), \; \text{if } q_1 = \forall\\
 p_1\times S(\mathbf{B}[2:n],\tau - w_1) +\\ (1-p_1)\times S(\mathbf{B}[2:n],\tau),\; \text{if } q_1 = \R^{p_1}
 \end{cases}
\end{align}

%\begin{enumerate}
% 	\item $ S(\mathbf{B},\tau) = \mathds{1}[ \tau \le 0 ] $, if $ \textbf{B} = \emptyset $, %Here $ \mathds{1}[\text{true}] = 1 $ and $ \mathds{1}[\text{false}] = 0 $,
% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \max\{w_1, 0\}) $, if $ q_1 =  \exists $,
% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \min\{w_1, 0\}) $, if $ q_1 =  \forall $,
% 	\item $ S(\mathbf{B},\tau) = p_1S(\mathbf{B}_{-1},\tau - w_1) + (1-p_1) S(\mathbf{B}_{-1},\tau) $, if $ q_1 = \R^{p_1}  $.
%\end{enumerate} 
%KSM: Don't use semantic 1, semantic 2, -- very confusing

Observe that when $ \textbf{B} $ is empty, $S$ is computed as $ 1 $ if $ \tau \le 0 $, and  $ S = 0 $ otherwise. For  existential and universal quantifiers, we compute $ S $ based on the weight. Specifically, if $ q_1 = \exists $, we decrement the threshold $ \tau $ by the maximum between $ w_1 $ and $ 0 $. For example, if $ w_1 > 0 $, $ B_1 $ is assigned $ 1 $, and assigned $ 0 $ otherwise. Therefore, by solving for an existential variable, we  maximize $ S $. In contrast, when if $ q_1 = \forall $, we fix an assignment of $ B_1 $ that minimizes $ S $ by choosing between the minimum of $ w_1 $ and $ 0 $. Finally, for random quantifiers, we decompose the computation of $S$ into two sub-problems: one sub-problem where $ B_1 = 1 $ and the updated threshold becomes $ \tau - w_1 $ and another sub-problem where $ B_1 = 0 $ and the updated threshold remains the same. Herein, we compute $ S $ as the expected output of both sub-problems.

%KSM: Lemma is a very strong term for such a straightforward observation
\begin{remark}
	\label{fairness_fvgm_lm:property_s3p}
	$ S(\mathbf{B},\tau) $ does not depend on the order of $ \mathbf{B} $.
\end{remark}

\textbf{Computing Minimum and Maximum PPVs of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.  The coefficients $ w_{A_j} $ and $ w_{X_i} $ of the classifier become weights of $ \mathbf{B} $. Also, we get $n=m_1 +m_2$. For non-sensitive variables $ X_i $, which are chance variables, we derive their marginal probability $ p_i = \Pr[X_i = 1] $ from the distribution $ \mathcal{D} $.  According to semantic of {\stochastic}, setting $ \exists $ quantifiers on $ \sensitive $ computes the maximum value of $ S(\mathbf{B}, \tau) $ that equalizes the maximum PPV of the classifier. In this case, the \textit{inferred} assignment of $ \sensitive $ implies the most favored group $ \mathbf{a}_{{\max}} =  \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $. In contrast, to compute the minimum PPV, we instead assign each variable  $ A_j $ a universal  quantifier  while keeping random quantifiers over $ X_i $, and infer the least favored group $ \mathbf{a}_{{\min}} =  \argmin_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.

\iffalse
\red{The decision version of computing the maximum (minimum) PPV is to decide whether there is an assignment of \textit{sensitive} or \textit{choice variables}, for which the \textit{non-sensitive} or \textit{chance variables} yield a PPV greater or less than $\alpha \in [0,1]$. Now, we formally state the hardness of verifying linear classifiers followed by an efficient dynamic programming solution.
\begin{lemma}
	\label{fairness_fvgm_lm:hardness}
	The decision version of the fairness verification problem for linear classifiers is in $\mathrm{NP^{PP}}$.
\end{lemma}}
\fi

 \subsection{A Dynamic Programming Solution for {\stochastic}}
 \label{fairness_fvgm_sec:dp_formulation}
 %\dbcomment{Use $ \mathsf{dp}(i, \tau) $ given $ B $}

We propose a dynamic programming approach~\cite{pisinger1999linear,woeginger1992equal} to solve {\stochastic} as the problem has overlapping sub-problem properties. 
For example, $S(\mathbf{B}, \tau)$ can be solved by solving $S(\mathbf{B}[2:n], \tau')$, where the updated threshold $\tau'$, called the \textit{residual threshold}, depends on the original threshold $\tau$ and the assignment of $B_1$ as shown in Eq.~\eqref{fairness_fvgm_eq:semantics_recurse}.
%\red{For example, the residual threshold $ \tau $ can be equal for a fixed subset of $ \mathbf{B} $, say $ \mathbf{B}[i:n] \triangleq \{B_j\}_{j=i}^{n} $, derived from different assignments of $ B_1, \dots, B_{i-1} $.} 
Building on this observation, we propose a recursion and terminating condition leading to our dynamic programming algorithm.

\textit{Recursion.} We consider a function $ \mathsf{dp}(i, \tau) $ that solves the sub-problem $ S(\mathbf{B}[i:n],\tau)  $, for $ i \in \{1,\ldots,n\}  $. The semantics of  $ S(\mathbf{B},\tau) $ in Eq.~\eqref{fairness_fvgm_eq:semantics_recurse} induces the recursive definition of $ \mathsf{dp}(i,\tau) $ as:

\begin{align}\label{fairness_fvgm_eq:dp_recurse}
 \mathsf{dp}(i,\tau)=
 \begin{cases}
 \mathsf{dp}(i+1, \tau-\max\{w_i, 0\}), \; \text{if } q_i = \exists\\
 \mathsf{dp}(i+1, \tau-\min\{w_i, 0\}), \; \text{if } q_i = \forall\\
 p_i \times \mathsf{dp}(i+1, \tau-w_i) + \\ (1- p_i) \times \mathsf{dp}(i+1, \tau) ,\; \text{if } q_i = \R^{p_i}
 \end{cases}
\end{align}

Eq.~\eqref{fairness_fvgm_eq:dp_recurse} shows that $ S(\mathbf{B},\tau) $ can be solved by instantiating $ \mathbf{dp}(1, \tau) $, which includes all the variables in $ \mathbf{B} $.

\textit{Terminating Condition.} %We now define the \textit{terminating conditions} of the recursion. 
Let $ w_{neg} $, $ w_{pos} $, and $ w_{all} $ be the sum of negative, positive, and all weights of $ \mathbf{B} $, respectively. We observe that $ w_{neg} \le w_{all} \le w_{pos}$. Thus, for any $ i $, if the {residual} threshold $ \tau \le w_{neg}$, there is always a subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $. Conversely, when $ \tau > w_{pos}$, there is no subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.	We leverage this bound and tighten the terminating conditions of $ \mathsf{dp}(i, \tau) $ in Eq.~\eqref{fairness_fvgm_eq:dp_terminus}.

\begin{align}\label{fairness_fvgm_eq:dp_terminus}
 \mathsf{dp}(i, \tau) =
 \begin{cases}
 1\text{ if }\tau \le w_{neg}\\
 0\text{ if } \tau > w_{pos}\\
 \mathds{1}[\tau \le 0] \text{ if } i=n + 1
 \end{cases} 
\end{align}

Eq.~\eqref{fairness_fvgm_eq:dp_recurse} and~\eqref{fairness_fvgm_eq:dp_terminus} together define our dynamic programming algorithm. While deploying the algorithm, we store $ \mathsf{dp}(i, \tau) $ in memory to avoid repetitive computations.  This allows us to achieve a pseudo-polynomial algorithm (Lemma~\ref{fairness_fvgm_lemma:complexity_sss}) instead of a na\"ive exponential algorithm enumerating all possible assignments.

\begin{lemma}\label{fairness_fvgm_lemma:complexity_sss}
 	Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $. Let $ w_{\exists} = \sum_{B_i \in \mathbf{B} | q_i = \exists} \max\{w_i, 0\}$  and $ w_{\forall} = \sum_{B_i \in \mathbf{B} | q_i = \forall} \min\{w_i, 0\}$ be the considered sum of weights of existential and universal variables, respectively. We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $. The space complexity is  $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
\end{lemma}

\textbf{A Heuristic for Faster Computation.} 
We propose two improvements for a faster computation of the dynamic programming solution. Firstly, we observe that in Eq.~\eqref{fairness_fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights. Hence, we reorder $ \mathbf{B} $ such that existential/universal variables appear earlier in $ \mathbf{B} $ than random variables. This allows us to avoid unnecessary repeated exploration of existential/universal variables in $ \mathsf{dp} $. Moreover, according to the remark in Section~\ref{fairness_fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}. Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables. In particular, if $ \tau \le 0.5(w_{pos} - w_{neg}) $, $ \tau $ is closer to $ w_{pos} $ than $ w_{neg} $. Hence, we sort each cluster in descending order of weights. Otherwise,  we sort in ascending order. We illustrate our dynamic programming approach in Example~\ref{fairness_fvgm_example:subset-sum}.

\begin{example}\label{fairness_fvgm_example:subset-sum}
We consider a linear classifier $ P + Q + R - S \ge 2$. Herein, $ P $ is a Boolean sensitive feature, and $ Q, R, S $ are Boolean non-sensitive features with $ \Pr[Q] = 0.4,  \Pr[R] = 0.5 $, and $ \Pr[S] = 0.3 $. To compute the maximum PPV of the classifier,  we impose an existential quantifier on $P$ and randomized quantifiers on others. This leads us to the observation that $ P = 1 $ is the optimal assignment as $ w_P = 1 > 0 $. We now require to compute $ \Pr[Q + R - S \ge 1] $, which by dynamic programming, is computed as $ 0.55 $. The solution is visualized as a search tree in Figure~\ref{fairness_fvgm_fig:example_dp}, where we observe that storing solution of sub-problems in the memory avoids repetitive computation, such as exploring the node $ (4,0) $. Similarly, the minimum PPV  of the classifier is $ 0.14 $ (not shown in Figure~\ref{fairness_fvgm_fig:example_dp}) where we impose a universal quantifier on $P$ to obtain $ P = 0 $ as the optimal assignment. 
\end{example}

### 2022-01-26 chapters/fairness/fvgm/framework2.tex
adding two fairness papers

\input{chapters/fairness/fvgm/dp_illustration}	
\subsection{{\stochastic} with Correlated Variables} 
\label{sec:dp_with_BN}
In {\stochastic} presented in Section~\ref{fairness_fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent. This independence assumption often leads to an \textit{inaccurate estimate} of PPVs because both sensitive and non-sensitive features can be correlated in practical fairness problems. Therefore, we extend {\stochastic} to include correlations among variables.

We consider a Bayesian network $ \BN = (\graph, \theta) $ to represent correlated variables, where $ G \triangleq (\mathbf{V}, \mathbf{E}) $, $ \mathbf{V} \subseteq \mathbf{B} $, $ \mathbf{E} \subseteq \mathbf{V} \times \mathbf{V}  $, and $ \theta $ is the parameter of the network.  In  $ \BN $, we constrain that there is no conditional probability of choice (i.e., existential and universal) variables as we optimize their assignment in {\stochastic}. Choice variables, however, can impose conditions on chance (i.e., random) variables. In practice, we achieve this by allowing no incoming edge on choice variables while learning $ \BN $ (ref. Section~\ref{fairness_fvgm_sec:experiments}).

For a chance variable $ B_i \in \mathbf{V} $, let $ \parent(B_i) $ denote its parents. According to Eq.~\eqref{fairness_fvgm_eq:BN},  for an assignment $ \mathbf{u} $ of $ \parent(B_i) $, $ \BN $ ensures $ B_i $ to be independent of other non-descendant variables in $ \mathbf{V} $. Hence, in the recursion of Eq.~\eqref{fairness_fvgm_eq:dp_recurse}, we substitute  $ p_i $  with  $ \Pr[B_i = 1| \parent(B_i) = \mathbf{u}] $. In order to explicate the dependence on $ \mathbf{u} $, we denote the expected solution of $ S(\mathbf{B}[i:n], \tau) $ as 
$ \mathsf{dp}(i, \tau, \mathbf{u}) $, which for $ B_i \in \mathbf{V} $ is modified as:

\begin{align*}
	\mathsf{dp}(i,  &\tau, \mathbf{u}) = \Pr[B_i = 0| \parent(B_i) = \mathbf{u}] \mathsf{dp}(i+1, \tau, \mathbf{u} \cup \{0\})  \\
	& + \Pr[B_i = 1| \parent(B_i) = \mathbf{u}]  \mathsf{dp}(i+1, \tau-w_i, \mathbf{u} \cup \{1\}).
\end{align*}

Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to know the assignment of parents before computing $ \mathsf{dp} $ on the child. Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}') $ for all $ \mathbf{u}' \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities $ \Pr[B_i = 1|\parent(B_i) = \mathbf{u}'] $ into $ \stochastic $.  For this enumeration, we do not store $ \mathsf{dp}(i, \tau, \mathbf{u}) $ in memory. However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Section~\ref{fairness_fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.  Lemma~\ref{fairness_fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fairness_fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).

\begin{lemma}
	\label{fairness_fvgm_lm:complexity_dp_with_bn}
	Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $. Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $. To exactly compute {\stochastic} with correlated variables in dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
\end{lemma}

\textbf{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $. For computing PPVs of a linear classifier with correlated features, we consider a heuristic to sort variables in $ \mathbf{B} = \sensitive \cup \nonsensitive $. Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the network and $ \mathbf{V}^c = \mathbf{B} \setminus \mathbf{V} $. In this heuristic, we sort sensitive variables $ \sensitive $ by positioning $ \sensitive \cap \mathbf{V} $ in the beginning followed by $ \sensitive \cap \mathbf{V}^c $. Then we order the variables $ \mathbf{B} $ such that variables in $ \sensitive $ precedes those in $ \nonsensitive \cap \mathbf{V} $, and the variables in $ \nonsensitive \cap \mathbf{V}^c $ follows the ones in $ \nonsensitive \cap \mathbf{V} $. This sorting allows us to avoid repetitive enumeration of variables in $ \mathbf{V} \subseteq \mathbf{B} $ as they are placed earlier in $ \mathbf{B} $.

	\begin{example}
		We extend Example~\ref{fairness_fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $. Parameters $ \theta $ imply conditional probabilities $ \Pr[Q|P] = 0.6 $ and $ \Pr[Q|\neg P] = 0.3 $. 	In Figure~\ref{fairness_fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $. We, however, observe that the dynamic programming solution in Section~\ref{fairness_fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $. Hence following the calculation in Figure~\ref{fairness_fvgm_fig:example_BN}, we obtain the maximum PPV  as $ 0.65 $ for $ P = 1 $. The minimum PPV (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $. 
	\end{example}

	\subsection{Fairness Verification with Computed PPVs} 
	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair ($\epsilon \in [0,1]$) is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$. We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum PPV $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum PPV $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.

	For measuring fairness metric SP, we compute the difference $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $. We, however, deploy {\fvgm} twice while measuring EO: one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $. 
	In each case, we compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of EO.  
	For measuring causal metric PCF, we compute  $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ conditioned on mediator features $ \mathbf{Z} $ and take their difference. 
	To measure DI, we compute the ratio $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] / \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $. In contrast to other fairness metrics, DI closer to 1 indicates higher fairness level. Thus, we verify whether a classifier achieves $(1 - \epsilon)$-DI by checking $ \mathds{1}[DI(\alg| \mathcal{D}) \ge 1 - \epsilon] $.

	\subsection{Extension to Practical Settings}\label{fairness_fvgm_sec:practical}
	For verifying linear classifiers with real-valued features and coefficients, we preprocess them so that {\fvgm} can be invoked. Let $ X_c \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ w_c \in \mathbb{R} $ in the classifier. We discretize $ X_c $ to a set of $ k $ Boolean variables $ \mathbf{B}_{c} $ based on the binning-based discretization and assign a Boolean variable to each bin. Hence, $ B_i \in \mathbf{B}_{c} $ becomes $ 1 $, when $ X_c $ belongs to the $ i^\text{th} $ bin. Let $ \mu_i $ denote the mean of feature-values within $ i^\text{th} $ bin. We then set the coefficient of $ B_i $ as $ w_c\mu_i $. By law of large numbers, $ X_c \approx \sum_i \mu_iB_i $ for infinitely many bins~\cite{grimmett2020probability}. Finally, we multiply the coefficients of discretized variables by $ l \in \mathbb{N} \setminus \{0\} $ and round to an integer. Accuracy of the preprocessing step relies on the number of bins $ k $ and the multiplier $ l $. Therefore, we empirically fine-tune both $ k $ and $ l $ by comparing  the processed classifier with the initial classifier on a validation dataset.
\begin{comment}
	\begin{figure}
		\centering
		\subfloat[Exploration tree of stochastic sub-set sum problem.]{	\includegraphics[scale=0.3]{figures/exploration_tree_DP}\label{fig:example_dp}}
		\subfloat[Bayesian network]{\includegraphics[scale=.4]{figures/example_BN}\label{fig:example_BN}}
		\subfloat[Exploration tree of stochastic sub-set sum problem with Bayesian network]{\includegraphics[scale=.4]{figures/exploration_tree_DP_BN}	\label{fig:example_dp_BN}
		}

	\end{figure}
\end{comment}

### 2022-01-26 chapters/fairness/fvgm/introduction.tex
adding two fairness papers

\section{Introduction}
	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}. ML algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
%	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction. Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
	 The following example illustrates a plausible case of unfairness induced by ML algorithms. 
	\begin{example}\label{fairness_fvgm_example:intro}
		Following~\cite[Example 1.]{ghosh2020justicia}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fairness_fvgm_fig:example1}). Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions. We generate 1000 samples from these distributions and use them to train a Support Vector Machine (SVM) classifier. The decision boundary of this classifier is $9.37I + 9.75F - 0.34A \ge 9.4$, where $A = 1$ denotes the sensitive group `age $ \ge $ 40'. This classifier selects an individual above and below $40$ years of age with probabilities $0.24$ and $0.86$, respectively. This illustrates a disparate treatment of individuals of two age groups by the SVM classifier. 
	\end{example}
%	The above example illustrates that even in simple scenarios, ML-based solutions can lead to disparate treatment. Motivated by studies indicating that such scenarios are rather not uncommon, there is a thrust in the research on the improvement of fairness in ML by proposing different notions of fairness definitions and fairness algorithms~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.

%KSM: Make the two plots side by side
\begin{figure}[h!]
	%		{r}{0.25\textwidth}%\vspace*{-2.4em}
	\begin{center}
		\includegraphics[scale = 0.4]{figures/fairness/fvgm/sanity_probability}
	\end{center}
	\vspace*{-1em}
	\caption{\footnotesize Age-dependent distributions of income and fitness in Example~\ref{fairness_fvgm_example:intro}.}\label{fairness_fvgm_fig:example1}%\vspace*{-1.6em}
\end{figure}

	In order to identify and mitigate the bias of ML classifiers, different fairness definitions and fairness algorithms have been proposed~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.	In this paper, we focus on two families of fairness definitions: \textit{group} and \textit{causal} fairness. Group fairness metrics, such as disparate impact and equalized odds constrain the probability of the positive prediction of the classifier to be (almost) equal among different sensitive groups~\cite{dwork2012fairness,feldman2015certifying}.	On the other hand, causal fairness metrics assess the difference in positive predictions  if every feature in the causal relation  remains identical except the sensitive feature~\cite{nabi2018fair,zhang2018fairness}. The early works on \textit{fairness verification} focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}. Naturally, such techniques were limited in enhancing confidence of users for wide deployment. Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare, bastani2019probabilistic, ghosh2020justicia}.  More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution. For Example~\ref{fairness_fvgm_example:intro}, a fairness verifier takes the SVM classifier and the distribution of features $ I, F, A $ as an input and outputs the probability of positive prediction of the classifier for  different sensitive groups.

%	\red{Despite extensive efforts over the past few years, the scalability and accuracy of fairness verifiers remains a major challenge. In particular, the existing techniques fail to scale beyond {\em toy} examples. In this work, we seek to remedy the situation. As a first step, we focus on  \textit{linear classifiers}, which has attracted a significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. At this point, it is worth highlighting that our empirical evaluation demonstrates that the existing techniques fail to scale beyond small examples or provide highly inaccurate estimates for comparatively {\em small} linear classifiers. }

	In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly, \cite{ghosh2020justicia} and \cite{albarghouthi2017fairsquare} propose formal methods that reduce the problem into a solution of an SSAT or an SMT~ formula respectively.	Secondly, \cite{bastani2019probabilistic} propose a sampling-based approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost. Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t. sensitive features. As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.

	 In this work, we seek to remedy the aforementioned situation. As a first step, we focus on  \textit{linear classifiers}, which has attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}. At this point, it is worth highlighting that our empirical evaluation demonstrates that the existing techniques fail to scale beyond small examples or provide highly inaccurate estimates for comparatively {\em small} linear classifiers.

%	In this paper, we argue that the above solution is not practical in verifying linear classifiers for two reasons: (i) the encoding of a linear classifier into an SSAT or an SMT formula or using a sampling approach suffer from scalability and (ii) ignoring correlations among all features results in an inaccurate estimate of fairness metrics.

%	Given multiple fairness definitions, fairness algorithms either \textit{preprocess} the training data~\cite{calmon2017optimized,kamiran2012data,zemel2013learning}, \textit{in-process} the classifier during training~\cite{zhang2018mitigating}, or \textit{post-process} the prediction of the trained classifiers~\cite{hardt2016equality,kamiran2012decision} to improve fairness. In contrast, fairness-attack algorithms try to manipulate the data or the classifier to worsen its fairness~\cite{solans2020poisoning}.

%	\red{Among different ML classifiers, researcher This observation motivates us in asking the following research question: \textit{Can we design an efficient fairness verification framework for linear classifiers that can verify different fairness definitions and algorithms accurately?}

%	
%	Considering the surge of research in ML fairness based on linear classifiers, we focus on designing a fairness verifier particularly tailored for linear classifiers. We further show novel applications of the verifier in practical fairness tasks, such as detecting fairness attacks and computing the source of unfairness in the level of input features.
%

	\textbf{Our Contributions.} In this paper, we propose a fairness verification framework, namely {\fvgm}, for accurately and efficiently verifying linear classifiers. {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming. To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).  
	%Finally, {\fvgm} encodes the Bayesian Network in a SSAT-based formulation to verify classifiers in CNF and in the stochastic subset-sum based formulation to verify linear classifiers.  
	Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms. We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.

### 2022-01-26 chapters/fairness/justicia/abstract.tex
adding two fairness papers

\section{Abstract}
	As a technology ML is oblivious to societal good or bad, and thus, the field of fair machine learning has stepped up to propose multiple mathematical definitions, algorithms, and systems to ensure different notions of fairness in ML applications.
	Given the multitude of propositions, it has become imperative to formally verify the fairness metrics satisfied by different algorithms on different datasets.
	In this paper, we propose a \textit{stochastic satisfiability} (SSAT) framework, {\justicia}, that formally verifies different fairness measures of supervised learning algorithms with respect to the underlying data distribution.
	We instantiate {\justicia} on multiple classification and bias mitigation algorithms, and datasets to verify different fairness metrics, such as disparate impact, statistical parity, and equalized odds.
	{\justicia} is scalable, accurate, and operates on non-Boolean and compound sensitive attributes unlike existing distribution-based verifiers, such as FairSquare and VeriFair.
	Being distribution-based by design, {\justicia} is more robust than the verifiers, such as AIF360, that operate on specific test samples.
	We also theoretically bound the finite-sample error of the verified fairness measure.

### 2022-01-26 chapters/fairness/justicia/analysis.tex
adding two fairness papers

\section{Theoretical analysis}

We now discuss theoretical analysis of our framework. Our analysis considers the following points.

\begin{enumerate}
	\item \textbf{Total error due to finite sample and thus, error in probability computation.} We compute the probability of the Boolean variables on finite sample (e.g., a given dataset). Hence the computed probability may carry error that is forwarded to the computation of the positive prediction rate and eventually to measuring fairness metrics.

	Let us consider that $ p_i $ is the computed probability of a Boolean variable $ b_i $ being assigned to $ \top $ on a finite sample $ D $ and $ \hat{p_i} $ is the probability  on the actual distribution $ \mathcal{D} $. If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.

	\red{I may be wrong!!} The satisfying probability $ p $ of $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $. 
	\[
		p = \sum_\sigma \prod_{b_i \in \sigma}p_i	
	\]
	This probability is computed according to $ D $. However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
	\[
		\hat{p} = \sum_\sigma \prod_{b_i \in \sigma}p_i \pm \epsilon	
	\]

	We need to find a tight bound of $ |p - \hat{p}| $.

	\item \textbf{Error due to thresholding of the features.} For the linear classifier, we convert the decision function to CNF using pseudo-Boolean encoding. This encoding results in a large CNF with many variables and clauses and results in more computational effort while solving the SSAT formula. In the linear classifier, the coefficients carry the weights/importance of the attributes. After normalizing all the coefficients, we can exclude attributes with coefficients less than a threshold $ \lambda $, i.e., we set the coefficient to $ 0 $ if it is less than $ \lambda $. Such thresholding would result in a smaller-size CNF. However, it  incur errors while computing the positive prediction rate of the classifier.

	\item \textbf{Error due to discretization (histogram based) of continuous attributes.} For continuous real-valued attributes, we can discretize them initially and learn a linear classifier on the Boolean discretized attributes. But this poses a limitation of {\framework} where the linear classifier must have Boolean attributes only. Instead we can apply discretization of the real-valued attributes after training, which we discuss in the following.

	Let $ x $ be a continuous attribute and $ w $ be its weight learned during training. We want to discretize $ x $ to a set $ \mathbf{B} $ of Boolean attributes. For discretization, we simply consider interval-based approach where for each interval (or bin) in the continuous space of $ x $, we consider a Boolean variable $ b_i \in \mathbf{B} $ that is assigned $ \top $ when the value of $ x $ lies within the interval and $ b_i $ is assigned $ \bot $ otherwise. We now compute the revised weights of the  discretized variables in $ \mathbf{B} $. Let $ \mu_i $ be the mean of the interval where  $ b_i $ is active. Then we set the revised weight of $ b_i $ to be $ \mu_i\cdot w $. We can show trivially that if we consider infinite number of intervals, this revised weight is equivalent to the actual weight of the continuous attribute, i.e., $ x \approx \sum_i \mu_ib_i $.

	We now need to measure error when considering finite number of intervals.

	\item \textbf{Error due to indistinguishably of probabilities below a threshold} Not implemented yet. 
\end{enumerate}

\todo[inline]{
	$x_1, \ldots, x_n$ OR, $p_1, \ldots, p_n$

	$P(\phi) = 1- \prod_{i=1}^{n} (1- p_i)$.

	$|\hat{p}_i - p_i| \leq \epsilon$ w.p $1 - 1/e^2$

	$P(\phi) = 1- \prod_{i=1}^{n} (1- p_i - \epsilon)$

	$\log (1-P(\phi)) = \sum_{i=1}^{n} \log(1- p_i - \epsilon) \leq \sum_{i=1}^{n} \log(1- p_i) + n\log(\epsilon)$
}

### 2022-01-26 chapters/fairness/justicia/appendix.tex
adding two fairness papers

\appendix
\section{Proofs}
\begin{lemma}
	\label{thm:complexity}
	Solving the ER-SSAT and RE-SSAT problems are $\mathrm{NP}^{\mathrm{PP}}$ hard~\cite{littman2001stochastic}.
\end{lemma}

\begin{proof}[Proof of Lemma~\ref{thm:complexity}]
	The decision version of ER-SSAT problem is 
	\begin{equation}
		\Phi := \exists a_1,\dots, \exists a_n, \R^{p_{x_1}}x_1, \dots, \R^{p_{x_m}}x_m. \; \Pr[\phi_{\hat{y}}] \geq t,\notag
	\end{equation}
	where $t$ is a threshold in $[0,1]$.
	It is exactly the an E-MAJSAT (or threshold SAT) problem which is $NP^{PP}$ hard~\cite{littman2001stochastic}.
	If there's no random variable and $t=1$, ER-SSAT reduces to a SAT problem, which is NP-hard. If there's no existential variable, ER-SSAT reduces to a MAJSAT problem, which is PP-hard.
	Similar arguments also hold for RE-SSAT problem.
\end{proof}

\begin{equation}
\Phi_{\mathsf{UR}} := \forall \sensitive_1,\dots, \forall \sensitive_n,
\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
\label{eq:ar}
\end{equation}
\begin{equation}
\Phi'_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \neg \phi_{\hat{Y}}.
\label{eq:er_complement}
\end{equation}
\begin{lemma}\label{thm:dual}
	Given Eq.~\eqref{eq:ar} and~\eqref{eq:er_complement},	$ \Pr[\Phi_{\mathsf{UR}}] = 1 - \Pr[\Phi'_{\mathsf{ER}}]  $.
\end{lemma}

\begin{proof}[Proof of Lemma~\ref{thm:dual}]
	Both $ \Phi_{\mathsf{UR}} $ and $ \Phi'_{\mathsf{ER}} $ have  random quantified variables in the identical order in the prefix. According to the definition of SSAT formulas,
\begin{equation}
\Pr[\Phi_{\mathsf{UR}}] = \min\limits_{a_1, \dots,a_n} \Pr[\phi_{\hat{Y}}] \text{ and } \Pr[\Phi'_{\mathsf{ER}}] = \max\limits_{a_1, \dots,a_n} \Pr[\neg\phi_{\hat{Y}}].\notag
\end{equation}
We can show the following duality between ER-SSAT and UR-SSAT,
\begin{equation}
\begin{split}
\Pr[\Phi'_{\mathsf{ER}}] &= \max\limits_{a_1, \dots,a_n} \Pr[\neg\phi_{\hat{Y}}]  \\
&= \min\limits_{a_1, \dots,a_n} (1 - \Pr[\phi_{\hat{Y}}])\\
&= 1 - \min\limits_{a_1, \dots,a_n}  \Pr[\phi_{\hat{Y}}]\\
&= 1 - \Pr[\Phi_{\mathsf{UR}}].\notag
\end{split}
\end{equation}
\end{proof}

\begin{lemma}
	\label{lm:equivalence}
	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound protected group $ \mathbf{a} \in A $. If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
	$\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   
	and
	$\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.   
\end{lemma}

\begin{proof}[Proof of Lemma~\ref{lm:equivalence}] 
	\blue{It is trivial} that the PPV of most favored group $ \mathbf{a}_{\mathsf{fav}} $ is the maximum PPV of all compound groups $ \mathbf{a} \in A $. Similarly, the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}} $ is the minimum PPV of all compound groups $ \mathbf{a} \in A $.

	By construction of the SSAT formulas, the PPV of $ \mathbf{a}_{\mathsf{fav}} $ and $ \mathbf{a}_{\mathsf{unfav}} $ are $ \Pr[\Phi_{\mathsf{ER}}] $ and $ \Pr[\Phi_{\mathsf{UR}}] $ respectively. Since $ \Pr[\Phi_{\mathbf{a}}] $ is the PPV of the compound group $ \mathbf{a} $, 
	\begin{equation}
	\begin{split}
		\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]
	\text{ and }
	\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}].\notag
	\end{split}
	\end{equation}
\end{proof}

\section{Practical Settings}
In this section, we  relax assumptions of Boolean classifiers and Boolean attributes and extend {\framework} to verify fairness metrics in a more practical setting. 
We first discuss the input classifiers of {\framework} in the following.

\subsubsection{Beyond CNF Classifiers.}
In the presented SSAT approach for verifying fairness, we assume the classifier $ \hat{Y} $ to be represented as a CNF formula.  In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental,yu2020computing}.  However, {\framework} can be extended beyond CNF classifiers, in particular to decision trees and linear classifiers that are widely adopted in the ML fairness studies~\cite{zemel2013learning,zafar2017fairness,xu2019achieving,zhang2019faht,raff2018fair,friedler2019comparative}.

\textbf{Encoding Decision Trees as CNF.} Existing  rule-based classifiers, for example, binary decision trees  can be trivially encoded as  CNF formulas.  In the binary decision tree, each node in the tree is a literal. A path  from the root to the leaf is a conjunction of literals (hence, a path is a clause) and the tree itself is a disjunction of all paths (or  clauses). In order to derive a CNF representation $ \phi $ of the decision tree, we first construct a DNF by considering all paths terminating at leaves with negative class label ($ \hat{y} = 0 $) and then complement it to a CNF using De Morgan's rule. Therefore, for any input that is classified positive by the decision tree satisfies $ \phi $ and vice versa. In {\frameworklearn} for learning the least favored group, we can construct a negated CNF classifier in Eq.~\ref{eq:er_complement} by only including paths terminating on positive labeled leaves.

\textbf{Encoding Linear Classifiers as CNF.} Linear classifiers on Boolean attributes can  be encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}. We consider a linear classifier  $ W\cdot X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathcal{R}^{|X}| $ and bias $ b \in \mathcal{R} $.  We first normalize $ W$ and $b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint, e.g., \textit{at-least} $ k $ constraint.  We then apply  pseudo-Boolean constraints to CNF translation to encode the decision boundary to CNF. This encoding usually introduces additional Boolean variables and results in large CNF. In order to generate a smaller CNF, we can apply thresholding techniques on the weights $ W $ to consider attributes with higher weights only. For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda $ and $ w_i \in W $, we can set $ w_i = 0 $. Thus,  lower weighted (hence less important) attributes do not appear in the encoded CNF.  Finally, to construct the negated classifier in the SSAT formula in Eq.~\ref{eq:er_complement}, we encode $ W\cdot X + b < 0 $ to CNF using \textit{at-most} $ k $ encoding.

In practical problems, attributes are generally real-valued or categorical. We next discuss how {\framework} can work beyond Boolean attributes.

\subsubsection{Beyond Boolean Attributes.}
Classifiers that are already represented in CNF are usually trained on a Boolean abstraction of the input attributes where  each categorical attribute is one-hot encoded  and each real-valued attribute is discretized into a set of Boolean attributes~\cite{LKCL2019,GMM20}. Thus, {\framework} can verify CNF classifiers readily.

\textbf{Decision Trees.} In case of binary decision tree classifiers, the input attributes are numerical or categorical, but each attribute is compared against a constant in each internal node of the tree. Hence, we fix a Boolean variable for each internal node where the Boolean assignment to the variable decides one of the two branches to choose from the current node.

\textbf{Linear Classifiers.} Linear classifiers are generally trained on numerical attributes where we apply following discretization. Consider a numerical attribute $ x $ where $ w $ is its weight. We want to discretize $ x $ to a set $ \mathbf{B} $ of Boolean attributes and recalculate the weights of the variables in $ \mathbf{B} $ from $ w $. For discretization, we simply consider interval-based approach where for each interval (or bin) in the continuous space of $ x $, we consider a Boolean variable $ b_i \in \mathbf{B} $ such that $ b_i $ is assigned $ \top $ (or $ 1 $) when the attribute-value of $ x $ lies within the interval and $ b_i $ is assigned $ \bot $ (or $ 0 $) otherwise. Let $ \mu_i $ be the mean of the interval where  $ b_i $ can be $ \top $. We then fix the revised weight of $ b_i $ to be $ \mu_i\cdot w $. We can show trivially that if we consider infinite number of intervals, $ x \approx \sum_i \mu_ib_i $.

\begin{example}[ER-SSAT encoding]
	\label{example:er_ssat}
	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple protected groups. As the example in Figure~\ref{fig:fair_example} is degenerate for this purpose, we introduce another protected group `sex $ \in $ \{male, female\}'. Consider a Boolean variable $ S $ for `sex' where the literal $ S $ denotes `sex = male'. With this new protected attribute, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ F,I,J $ have same distributions as discussed in Example~\ref{example:re_ssat}. 
	Hence, we obtain the ER-SSAT formula of $\alg$ to learn the most favored group:
	\begin{equation}
	\Phi_{\mathsf{ER}} :=  \exists S,\exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; (\neg F \vee I \vee S) \wedge (F \vee J).\notag
	\end{equation}
	As we solve $ \Phi_{\mathsf{ER}} $, we learn that the assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, i.e. `male individuals with age $ < 40 $' is the most favored group with PPV computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$. Similarly, to learn the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
	\begin{equation}
	\Phi_{\mathsf{ER'}} :=  \exists S, \exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; \neg((\neg F \vee I \vee S) \wedge (F \vee J)).\notag
	\end{equation}
	Solving $ \Phi_{\mathsf{ER'}} $, we learn the assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi_{\mathsf{ER'}}] = 0.57 $. Thus, `female individuals with age $ < 40 $' constitute the least favored group with PPV:  $ 1-0.57 = 0.43$. 
	Thus, {\frameworklearn} allows us to learn the most and least favored groups and the corresponding discrimination.
\end{example}

\section{Experiments}
\subsection{Experimental Setup}
Since both {\framework}  and FairSquare take a  probability distribution of the attributes as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution.

\subsection{Comparative Evaluation of Two Encodings}
\iffalse
\red{Accuracy is removed}
\textbf{Accuracy.} In Section~\ref{sec:framework},  we have theoretically shown the equivalence between RE and ER encodings  and experimentally validate this in Figure~\ref{fig:equivalence} where both encodings output the same value for different fairness metrics. In the RE encoding, we consider two variations: namely `RE' and `RE(cor)' where the former does not consider the correlation between the protected and the non-protected attributes and the later includes correlation. \blue{In Figure~\ref{fig:equivalence}, we find that disparate impact usually decreases when the correlation is considered. This result is justified by the implicit dependence between the protected and non-protected attributes that the classifiers learn during training.}
\fi

%\textbf{Execution Time.} 
While both {\frameworkenum} and {\frameworklearn}  have the same output, the {\frameworklearn} encoding  improves exponentially  in runtime  than {\frameworkenum}  on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fig:runtime_diff_encodings}. This analysis justifies that the na\"ive enumeration-based approach cannot verify large-scale fairness problems containing multiple protected attributes. 
\begin{figure}[t!]
	\begin{center}
		\includegraphics[scale=.32]{figures/encoding_runtime_Adult_DT.pdf}
		\includegraphics[scale=.32]{figures/encoding_runtime_Adult_IMLI.pdf}
		\hfill
		\caption{Runtime comparison of different encodings for varying total protected groups in the Adult dataset}
		\label{fig:runtime_diff_encodings}
	\end{center}
\end{figure}\textbf{}

\begin{table*}[t!]        
	\centering
	\caption{Verification of different fairness enhancing algorithms for multiple datasets and classifiers using {\framework}. Numbers in bold refer to fairness improvement  compared against the unprocessed (orig.) dataset. RW and OP refer to reweighing and optimized-preprocessing algorithm respectively.}\label{tab:fair_algo_verification_appendix}
	\vspace*{-.2em}
	%        \setlength{\tabcolsep}{.3em}
	\begin{tabular}{ll
			ccc
			ccc
			%            				ccc
			%            				ccc
			ccc
			ccc}
		\toprule
		\multirow{3}{*}{Classifier}& Dataset $ \rightarrow $   & 
		\multicolumn{6}{c}{German} \\
		\cmidrule(lr){3-8}
		& Protected  $ \rightarrow $ & 
                \multicolumn{3}{c}{Age}   & \multicolumn{3}{c}{Sex}  
		\\ 
		\cmidrule(lr){3-5}
		\cmidrule(lr){6-8}

		& Algorithm  $ \rightarrow $ &  
		orig. & RW & OP &
		orig. & RW & OP \\ 
		\midrule

		\multirow{3}{*}{\shortstack{Logistic \\ regression}}
		& Disparte impact&  $ 0.00 $ &  $ 0.00 $ &  $ \mathbf{0.31} $ &  $ 0.27 $ &  $ \mathbf{0.46} $ &  $ 0.17 $  \\
		& Stat. parity&  $ 0.45 $ &  $ \mathbf{0.03} $ &  $ \mathbf{0.12} $ &  $ 0.03 $ &  $ \mathbf{0.02} $ &  $ 0.07 $  \\
		& Equalized odds&  $ 0.65 $ &  $ \mathbf{0.04} $ &  $ \mathbf{0.14} $ &  $ 0.10 $ &  $ \mathbf{0.08} $ &  $ 0.13 $  \\
		\midrule
		\multirow{3}{*}{\shortstack{Decision \\ tree}}
		& Disparte impact&  $ 0.00 $ &  $ \mathbf{0.56} $ &  $ \mathbf{0.12} $ &  $ 0.35 $ &  $ \mathbf{0.37} $ &  $ \mathbf{0.38} $  \\
		& Stat. parity&  $ 0.35 $ &  $ \mathbf{0.02} $ &  $ \mathbf{0.22} $ &  $ 0.05 $ &  $ 0.10 $ &  $ 0.11 $  \\
		& Equalized odds&  $ 0.36 $ &  $ \mathbf{0.05} $ &  $ \mathbf{0.28} $ &  $ 0.06 $ &  $ 0.16 $ &  $ 0.17 $  \\

		\bottomrule
	\end{tabular}
\end{table*}

\begin{comment}
\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_op_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_op_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_op_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_op_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying optimized preprocessing (op) algorithm on Adult dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_op_DT_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_op_DT_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_op_LR_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_op_LR_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying optimized preprocessing (op) algorithm on German-credit dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_op_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_op_DT_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_op_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_op_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying optimized preprocessing (op) algorithm on Compas dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_DT_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_DT_age.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_LR_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_LR_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_rw_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_rw_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_rw_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_rw_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Adult_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Adult_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying reweighing (rw) algorithm on Adult dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_rw_DT_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_rw_DT_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_rw_LR_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_rw_LR_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_German_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_German_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying reweighing (rw) algorithm on German-credit dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_rw_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_rw_DT_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_rw_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_rw_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/DI_Compas_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/SPD_Compas_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Result for applying reweighing (rw) algorithm on Compas dataset. We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR). If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm. RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_DT_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_DT_race.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_LR_race.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_LR_race.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\begin{center}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_DT_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_DT_age.pdf}
		} \hfil  
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_LR_age.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_LR_age.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_DT_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_DT_sex.pdf}
		} \hfil

		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_LR_sex.pdf}
		}
		\subfloat{
			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_LR_sex.pdf}
		} \hfil

	\end{center}

	\caption{Effect of sampling.} 
\end{figure*}

\begin{figure*}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, sex.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_sex.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, sex.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_sex.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, sex.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_sex, age.pdf}
	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_sex.pdf}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, sex.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_sex.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, sex.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_sex.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, sex.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_sex, age.pdf}
\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_sex.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_sex.pdf}

\end{figure*}

\begin{figure*}
		\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_age, sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_sex.pdf}

\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex.pdf}

\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, passenger class, age.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, passenger class.pdf}
	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_dt.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_lr.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_mlic.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_dt.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_lr.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_mlic.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_dt.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_lr.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_mlic.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_dt.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_lr.pdf}
	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_mlic.pdf}
\end{figure*}

\begin{figure*}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_mlic_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_mlic_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_mlic_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_mlic_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_mlic_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_dt_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_lr_RE.pdf}
\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_mlic_RE.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_mlic_RE(cor).pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_mlic_RE.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_mlic_RE.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_mlic_RE(cor).pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_mlic_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_dt_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_dt_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_lr_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_lr_RE.pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_mlic_RE(cor).pdf}
	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_mlic_RE.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_sex.pdf}

\end{figure*}

\begin{figure*}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_age, sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_age, sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_sex.pdf}
\end{figure*}

\begin{figure*}

	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Position.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Race, Position.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Race.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Position.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Race, Position.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Race.pdf}
\end{figure*}

\begin{figure*}

	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_passenger class, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_passenger class.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, passenger class, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, passenger class.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_passenger class, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_passenger class.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, passenger class, age.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, passenger class.pdf}
	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex.pdf}
\end{figure*}

\end{comment}

### 2022-01-26 chapters/fairness/justicia/conclusion.tex
adding two fairness papers

\section{Discussion and Future Work}
In this paper, we have presented {\framework}, a unified formal framework based on SSAT to verify different fairness metrics in presence of compound protected groups. In our empirical studies, {\framework} achieves higher accuracy and impressive scalability than the existing distribution-based fairness verifiers FairSquare and VeriFair. Moreover, {\framework} verifies different fairness-enhancing algorithms in multiple datasets and classifiers. Additionally, {\framework} is more robust than the sample-based verifier AIF360. As a future work, we incorporate attribute-correlation in the presented learning-based encoding.

### 2022-01-26 chapters/fairness/justicia/discussion.tex
adding two fairness papers

\section{Discussion and Future Work}
Though formal verification of different fairness metrics of an ML algorithm for different datasets is an important question, existing verifiers are not scalable, accurate, and extendable to non-Boolean protected attributes. We propose a stochastic SAT-based approach, {\justicia}, that formally verifies independence and separation metrics of fairness for different classifiers and distributions for compound protected groups.
Experimental evaluations demonstrate that {\justicia} achieves \textit{higher accuracy} and \textit{scalability} in comparison to the state-of-the-art verifiers, FairSquare and VeriFair, while yielding \textit{higher robustness} than the sample-based tools, such as AIF360.

Our work opens up several new directions of research. One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs. Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
%Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

%In this paper, we focus on formal verification of whether a given model satisfies a given fairness criterion. To this end, we proposed a stochastic SAT-based approach, {\justicia}, that formally verifies independence and separation metrics of fairness for different classifiers and distributions. We demonstrates that {\justicia} achieves higher accuracy and scalability in comparison to the prior state of the art approaches while achieving higher robustness in comparison to purely dataset based approaches such as AIF360. Our work opens up several directions of new research. Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2022-01-26 chapters/fairness/justicia/experiments.tex
adding two fairness papers

\section{Empirical Performance Analysis}
\label{fairness_justicia_sec:experiments}
In this section, we discuss the empirical studies to evaluate the performance of {\justicia} in verifying different fairness metrics. We first discuss the experimental setup and the objective of the experiments and then evaluate the experimental results.
\subsection{Experimental Setup}
We have implemented a prototype of {\justicia} in Python (version $ 3.7.3 $). The core computation of {\justicia} relies on solving SSAT formulas using an off-the-shelf SSAT solver. To this end, we employ the state of the art RE-SSAT solver of~\cite{lee2017solving} and the ER-SSAT solver of~\cite{lee2018solving}. Both solvers output the exact  satisfying probability of the SSAT formula.

For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art distributional verifiers FairSquare and VeriFair, and also a sample-based fairness measuring tool: AIF360. 
In the experiments, we have studied three type of classifiers: CNF learner, decision trees and logistic regression classifier.
Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner IMLI of~\cite{ghosh19incremental}. We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula.
We have also verified two fairness-enhancing algorithms: reweighing algorithm~\cite{kamiran2012data} and the optimized pre-processing  algorithm~\cite{calmon2017optimized}. 
We have experimented on multiple datasets containing multiple protected attributes: the UCI\footnote{\url{ http://archive.ics.uci.edu/ml}} Adult and German-credit dataset,  ProPublica’s COMPAS recidivism dataset~\cite{angwin2016machine}, Ricci dataset~\cite{mcginley2010ricci}, and Titanic dataset\footnote{\url{https://www.kaggle.com/c/titanic}}.
\iffalse 
Since both {\justicia}  and FairSquare take a  probability distribution of the attributes as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution. 
\fi
%that pre-process the dataset to mitigate its bias

Our empirical studies have the following objectives:

\begin{enumerate}
	\item How accurate and scalable {\justicia} is with respect to existing fairness verifiers, FairSquare and VeriFair?
	\item Can {\justicia} verify the effectiveness of different fairness-enhancing algorithms on different datasets?
	\item Can {\justicia} verify fairness in the presence of compound sensitive groups?
	\item How robust is {\justicia} in comparison to sample-based tools like AIF360 for varying sample sizes?
	\item How do the computational efficiencies of {\justicialearn} and {\justiciaenum} compare?
\end{enumerate}

%\begin{enumerate}
%	\item How accurate and scalable {\justicia} is with respect to existing fairness verifiers, FairSquare and VeriFair?
%	\item Can {\justicia} verify the effectiveness of different fairness-enhancing algorithms on different datasets?
%	\item Can {\justicia} verify fairness in the presence of compound sensitive groups?
%	\item How robust is {\justicia} in comparison to empirical tools like AIF360 for varying sample sizes?
%\end{enumerate}

Our experimental studies validate that {\justicia} is more accurate and scalable than the state-of-the-art verifiers FairSquare and VeriFair. {\justicia} is able to verify the effectiveness of different fairness-enhancing algorithms for multiple fairness metrics, and datasets. {\justicia} achieves scalable performance in the presence of compound sensitive groups that the existing verifiers cannot handle.  {\justicia} is also more robust than the sample-based tools such as AIF360.
Finally, {\justicialearn} is significantly efficient in terms of runtime than {\justiciaenum}.

\input{chapters/fairness/justicia/table_synthetic.tex}
\input{chapters/fairness/justicia/table_scalability.tex}
\input{chapters/fairness/justicia/table_fairness_algorithm.tex}

%\red{Add lines related to research question 5.}

%\subsection{Experimental Results}
%We now discuss our empirical findings.
\begin{figure*}[t!]
		\begin{minipage}{0.48\textwidth}
		\centering
		\begin{minipage}{0.48\columnwidth}
			\includegraphics[scale=.15]{figures/fairness/justicia/sensitive_attribute_race_di_Adult_DT_RE.pdf}
		\end{minipage}
		\begin{minipage}{0.48\columnwidth}
			\includegraphics[scale=.15]{figures/fairness/justicia/sensitive_attribute_race_spd_Adult_DT_RE.pdf}
		\end{minipage}
%		\begin{minipage}{0.32\columnwidth}
%			\includegraphics[scale=.14]{figures/fairness/justicia/sensitive_attribute_race_eqo_Adult_DT_RE.pdf}
%		\end{minipage}
		\caption{Fairness metrics measured by {\justicia} for different protected groups in the Adult dataset. The number within parenthesis in the xticks denotes total compound groups.}
		\label{fairness_justicia_fig:protected_groups}
	\end{minipage}
	\hfill
	\begin{minipage}{0.48\textwidth}
		\includegraphics[scale=.15]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}
		\hfill
		\includegraphics[scale=.15]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}
		\caption{Standard deviation in estimation of disparate impact (DI) and stat. parity (SP)  for different sample sizes. {\justicia} is more robust with variation of sample size than  AIF360. }
		\label{fairness_justicia_fig:sample-size}
	\end{minipage}
\end{figure*}

\subsection{Experimental Analysis}

\begin{comment}
\begin{figure}[t]
\centering
\includegraphics[scale=.32]{figures/fairness/justicia/sanity_DI_DT.pdf}
\caption{Results of different verifiers on  a synthetic problem.
}
\label{fairness_justicia_fig:synthetic}
\end{figure}
\end{comment}

%\subsubsection{Performance of Different Verifiers.}
\subsubsection{Accuracy: Less Than $  1\%$-error.} 
In order to assess the accuracy of different verifiers, we have considered the decision tree in Figure~\ref{fairness_justicia_fig:fair_example} for which the fairness metrics  are analytically computable. 
In Table~\ref{fairness_justicia_tab:synthetic}, we show the computed fairness metrics by {\justicia}, FairSquare, VeriFair, and AIF360. We observe that {\justicia} and AIF360  yield more accurate estimates of DI and SP compared against the ground truth with less than $1\%$ error.
FairSquare and VeriFair  estimate the disparate impact to be $0.99$ and thus, being unable to verify the fairness violation. 
Thus, {\justicia} is significantly accurate than the existing formal verifiers: FairSquare and VeriFair. 
%First we observe that both RE and ER encoding result in the same disparate impact, thereby showing the equivalence between the two encodings.

\iffalse
\begin{figure}[t]
	\begin{center}
		\subfloat{
			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_DT.pdf}
		}
		\subfloat{
			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_LR.pdf}
		} \hfill
		%		\subfloat{
		%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_DT.pdf}
		%		}
		%		\subfloat{
		%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_LR.pdf}
		%		} \hfill
		\subfloat{
			\includegraphics[scale=.32]{figures/fairness/justicia/time_ratio_FS_DT.pdf}
		}
		\subfloat{
			\includegraphics[scale=.32]{figures/fairness/justicia/time_ratio_FS_LR.pdf}
		} \hfill

		\caption{Comparison of {\justicia} (RE(cor) encoding) with FairSquare (FS). The top two figures show measured disparate impact and the bottom two figures show ratio of runtime of the two verifiers for decision tree (DT) and logistic regression (LR).}
		\label{fairness_justicia_fig:fairsquare}
	\end{center}
\end{figure}
\fi

\subsubsection{Scalability: $ 1 $ to $ 3 $ Orders of Magnitude Speed-up.} 
%Since {\justicia} appears to be more accurate than FairSquare and VeriFair in the synthetic benchmark, 
We have tested the scalability of {\justicia}, FairSquare, and VeriFair on practical benchmarks with a timeout of $900$ seconds and reported the execution time of these verifiers on decision tree and logistic regression in Table~\ref{fairness_justicia_tab:FS_VF_Justicia}. We observe that {\justicia} shows impressive scalability than the competing verifiers. Particularly, {\justicia} is $ 1 $ to $ 2 $ orders of magnitude faster than FairSquare and  $ 1 $ to $ 3 $ orders of magnitude faster than VeriFair. Additionally, FairSquare times out in most  benchmarks.
Thus, {\justicia} is not only accurate but also scalable than the existing verifiers.

\begin{comment}
We now discuss the performance of  {\justicia} compared against FairSquare in practical benchmarks. In general, we have observed that FairSquare cannot  scale to benchmarks with large number of attributes. Hence we show result on smaller benchmarks e.g.,  Ricci and Titanic dataset. In this context, we emphasize that FairSquare only handles single binary protected attribute, whereas {\justicia} does not have such a restriction. In Figure~\ref{fairness_justicia_fig:fairsquare}, we see that Both {\justicia} (RE(cor) encoding) and FairSquare (FS) certify higher disparate impact for Ricci dataset and lower disparate impact for  Titanic dataset on decision tree classifier (DT).  When we consider logistic regression (LR), both {\justicia} and FairSquare certify that disparate impact decreases for Ricci dataset. In this context, FairSquare cannot scale to Titanic dataset for the logistic regression classifier within the cut off time of 400 seconds. Moving focus on the  runtime, we have observed that FairSquare takes one to two magnitude higher time for decision tree  and two to three magnitude higher time for logistic regression, thereby demonstrating the poor scalability of FairSquare in verifying fairness. 
\end{comment}

\subsubsection{Verification: Detecting Compounded Discrimination in Protected Groups.}
We have tested {\justicia} for datasets consisting of multiple protected attributes and reported the results in Figure~\ref{fairness_justicia_fig:protected_groups}. {\justicia} operates on datasets with even 40 compound protected groups and can potentially scale more than that while the state-of-the-art fairness verifiers (e.g., FairSquare and VeriFair) consider a single protected attribute.
Thus, {\justicia} removes an important limitation in practical fairness verification. 
Additionally, we observe in most datasets the disparate impact decreases and thus, discrimination increases as more compound protected groups are considered. For instance, when we increase the total  groups from $ 5 $ to $ 40 $ in the Adult dataset, disparate impact decreases from around $ 0.9 $ to $ 0.3 $, thereby detecting higher discrimination. Thus, {\justicia} detects that the marginalized individuals of a specific type (e.g., `race')  are even more discriminated and marginalized when they also belong to a marginalized group of another type (e.g., `sex').

\subsubsection{Verification: Fairness of Algorithms on Datasets.}
We have experimented with two fairness-enhancing algorithms: the reweighing (RW) algorithm and the optimized-preprocessing (OP) algorithm.
Both of them pre-process to remove statistical bias from the dataset. 
We study the effectiveness of these algorithms using {\justicia} on three datasets each with two different protected attributes.  
In Table~\ref{fairness_justicia_tab:fair_algo_verification}, we report different fairness metrics on logistic regression and decision tree. We observe that {\justicia} verifies fairness improvement as the bias mitigating algorithms are applied.  For example, for the Adult dataset with `race' as the protected attribute, disparate impact increases from $ 0.23 $ to $ 0.85 $ for applying the reweighing algorithm on logistic regression classifier. In addition, statistical parity decreases from $ 0.09 $ to $ 0.01 $, and equalized odds decreases from $ 0.13 $ to $ 0.03 $, thereby showing the effectiveness of reweighing algorithm in all three fairness metrics. 
{\justicia} also finds instances where the fairness algorithms fail, specially when considering the decision tree classifier. 
Thus, {\justicia} enables verification of different fairness enhancing algorithms in literature.

\subsubsection{Robustness: Stability to Sample Size.} 
We have compared the robustness of {\justicia} with AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics. 
In Figure~\ref{fairness_justicia_fig:sample-size}, AIF360 shows higher standard deviation for lower sample-size and the value decreases as  the sample-size increases. 
In contrast, {\justicia} shows significantly lower ($\sim10\times$ to $100\times$) standard deviation for different sample-sizes. 
The reason is that AIF360 empirically measures on a fixed test dataset whereas {\justicia} provides estimates over the data generating distribution.
Thus, {\justicia} is more robust than the sample-based verifier AIF360.

\iffalse
\begin{figure}[t]
	\begin{center}
		\begin{minipage}{.32\columnwidth}
			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_di_Adult_DT_race__sex.pdf}
		\end{minipage}
		\begin{minipage}{.32\columnwidth}
			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_spd_Adult_DT_race__sex.pdf}
		\end{minipage}
		\begin{minipage}{.32\columnwidth}
			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_eqo_Adult_DT_race__sex.pdf}
		\end{minipage}
		%		\subfloat{
		%			\includegraphics[scale=.32]{figures/fairness/justicia/equivalence_di_adult_lr_sex.pdf}
		%		}
		%		\subfloat{
		%			\includegraphics[scale=.32]{figures/fairness/justicia/equivalence_spd_adult_lr_sex.pdf}
		%		} \hfil
		\caption{Fairness metrics measured by different encodings. Both RE and ER encodings result in same fairness value while RE(cor) encoding usually shows lower fairness value for considering correlation of the protected attributes.\red{Can be removed}}
		\label{fairness_justicia_fig:equivalence}
	\end{center}
\end{figure}
\fi

\iffalse
\begin{figure}
	\subfloat{
		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Adult_op_LR_sex.pdf}
	}
	\subfloat{
		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Adult_rw_LR_sex.pdf}
	}\\
	\subfloat{
		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Compas_op_LR_sex.pdf}
	}
	\subfloat{
		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Compas_rw_LR_sex.pdf}
	}
	\caption{Fairness metrics for applying optimized pre-processing (op) and reweighing (rw) algorithm on the Adult dataset for the protected attribute sex. As we apply these algorithms, fairness is usually enhanced certified by SSAT based verifiers {\justicia} (e.g.,  RE(cor) and RE encoding) and the statistical tool AIF360 (AIF). In each pair of columns, the first and second column refer to before and after applying the algorithm, respectively.}
	\label{fairness_justicia_fig:fairness_algorithm_aif}
\end{figure}
\fi

%\begin{figure}
%	\begin{center}
%		\subfloat{
%			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_DT.pdf}
%		} 
%		\subfloat{
%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_DT.pdf}
%		} \hfil
%		\subfloat{
%			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_LR.pdf}
%		} 
%		\subfloat{
%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_LR.pdf}
%		} \hfil
%			
%		\caption{Comparison with FairSquare.}
%	\end{center}
%\end{figure}

%\begin{figure}
%	\begin{center}
%		\subfloat{
%			\includegraphics[scale=.32]{figures/fairness/justicia/threshold_adult_lr_RE(cor)_age.pdf}
%		} 
%			\subfloat{
%		\includegraphics[scale=.32]{figures/fairness/justicia/threshold_ricci_lr_RE(cor)_Race.pdf}
%		} \hfil
%	
%		\caption{Effect of thresholding.}
%	\end{center}
%\end{figure}

\begin{figure}[t!]
	\begin{center}
		\includegraphics[scale=.32]{figures/fairness/justicia/encoding_runtime_Adult_DT.pdf}
		\includegraphics[scale=.32]{figures/fairness/justicia/encoding_runtime_Adult_IMLI.pdf}
		\hfill
		\caption{Runtime comparison of different encodings while varying total protected groups in the Adult dataset.}
		\label{fairness_justicia_fig:runtime_diff_encodings}
	\end{center}
\end{figure}

\subsubsection{Comparative Evaluation of Different Encodings.}
While both {\justiciaenum} and {\justicialearn}  have the same output according to Lemma~\ref{fairness_justicia_lm:equivalence},  {\justicialearn} encoding  improves exponentially  in runtime  than {\justiciaenum} encoding on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fairness_justicia_fig:runtime_diff_encodings}. {\justiciacond} (conditional probabilities w.r.t. protected groups) also has an exponential trend in runtime similar to {\justiciaenum}.  This analysis justifies that the na\"ive enumeration-based approach cannot verify large-scale fairness problems containing multiple protected attributes, and {\justicialearn} is a more efficient approach for practical use.

%The runtime efficiency of {\justicialearn} posits it as a more scalable and practical approach to verify fairness than {\justiciaenum}.
%\red{Justify the use of }.

### 2022-01-26 chapters/fairness/justicia/extra.tex
adding two fairness papers

\clearpage
\section{Verifying Fairness}
With an increasing applications of different algorithmic decision processes in our day to day life, it becomes crucial to evaluate if the algorithm is fair with respect to  all concerned ethnic groups.  In our study we focus on machine learning (ML) algorithms that are trained on a given training set. Now the decision may be unfair for one of the following two reasons or both. 
\begin{enumerate}
	\item \label{data-bias} The underlying data on which the algorithm is trained is biased towards a particular ethnic group(s) and the ML algorithm essentially learns to be unfair during its training. 
	\item \label{model-bias} The training dataset is not biased, however, the ML algorithm is showing bias in its decision. Various factors can cause such an unfairness. For instance, the ML algorithm typically considers different hyper-parameters and optimizes them during training. Such an optimization (towards higher accuracy) may be the reason why the ML algorithm becomes unfair although the training set is not biased. 
\end{enumerate}

In this study, we refer \ref{data-bias} as data bias and \ref{model-bias} as model bias. Now the goal of this study is to verify the fairness of a prediction system that may be vulnerable to be unfair in its decision either for data-bias or model-bias or both.

We now discuss the set of challenges to address in this study. Firstly, there are different definitions of fairness in the literature, of which we primarily focus on counterfactual fairness. \blue{[We need to give reasons as to why we focus on counterfactual fairness.]} Secondly,  we make an assumption that the causal graph of the training dataset is available. Although this assumption may not be always met due to the unavailability of causal graph for a particular dataset or the higher computational expense of calculating the exact causal graph for a given dataset.  Thirdly, we assume the ML algorithm to be a black-box meaning that the knowledge of the inner working of the prediction model is unknown. However, the model can be queried to get the classification of an input.

Since in the practical situation, only the ML model is given but not the training dataset, we still assume that the specification of the dataset is known in advance so that we can generate random inputs meeting the specification and query the model to get its class label. By specification, we mean the description of input features, their ranges and other meta information of the features, i.e., whether a feature is a sensitive feature or not. In this context, the sensitive feature carry the same definition as discussed in fairness literature.

We now give a formal definition of the problem. Let $ X $ and $ A $ be the set of non-sensitive features and sensitive features, respectively and $ M $ be the ML model that we treat as a black-box. $ M $ implements a classification function $ Y=f(X,A) $  and for binary classification $ Y \in \{0,1\} $. Therefore, $ Y $ can be utilized to answer membership queries of input $ x\circ a  $ where $ x \circ a $ is the valuation of features in $ X \cup A $. Here $ \circ $ denotes the concatenation of two vector $ x $ and $ a $.   We are then interested in verifying whether $ Y $ is fair or not depending on the definition of counterfactual fairness, which we introduce next.

\textbf{Counterfactual fairness:} Consider that $ X $, $ A $, and $ Y $ are three random variables on non-sensitive features, sensitive features, and predictor, respectively. Now a predictor $ Y $ is counterfactually fair if the following condition holds.

\begin{equation}
\label{eq:counterfactual-fairness}
\mathsf{Pr}(Y = y| X=x, A=a) = \mathsf{Pr}(Y=y| X=x, A=a')
\end{equation}

In words,  a predictor $ Y $ is counterfactually fair if the $ Y $  is independent of the  valuation of the sensitive feature of the input.

Given a predictor $ Y $, we want to verify if $ Y $ satisfies Eq.~\ref{eq:counterfactual-fairness}.

\red{[Now introduce $ \epsilon, \delta $] definition of fairness verification.}

\textbf{Independence.} e.g. demographic parity, statistical parity, group parity

\textbf{Separation.} e.g. equalizing the odds over groups

\textbf{Sufficiency.} e.g. Counterfactual fairness

### 2022-01-26 chapters/fairness/justicia/framework.tex
adding two fairness papers

\section{{\justicia}: An SSAT Framework to Verify Fairness Metrics}
\label{fairness_justicia_sec:framework}
In this section, we present the primary contribution of this paper, {\justicia}, which is an SSAT-based framework for verifying independence and separation metrics of fairness.

Given a binary classifier $\alg$ and a probability distribution over dataset $(X,A,Y) \sim \mathcal{D} $, our goal is to verify whether $\alg$ achieves independence and separation metrics with respect to the distribution $\mathcal{D}$. We  focus on a classifier that can be translated to a CNF formula of Boolean variables $\mathbf{B} $. 
The probability $ p_i $ of $\bool_i \in \mathbf{B}$ being assigned to $1$ is induced by the data generating distribution $\mathcal{D}$. 
In order to verify fairness metrics in compound protected groups, we discuss an enumeration-based approach and an equivalent learning-based approach. 
%We conclude this section by proposing a conditional distribution based enumeration for compound protected groups in Section~\ref{fairness_justicia_sec:cond_ssat}. 
We then provide a theoretical analysis for a high-probability error bound on the fairness metric and conclude with extension of {\justicia} in practical settings.

\iffalse
In this section, we present the main contribution of this paper, {\justicia}, which is an SSAT framework for verifying independence and separation metrics of fairness. 
We first state the problem formally in Section~\ref{fairness_justicia_sec:problem_statement}. 
To verify fairness metrics in compound protected groups, we discuss an enumeration approach in Section~\ref{fairness_justicia_sec:enumeration_ssat} and an equivalent but more efficient learning approach in Section~\ref{fairness_justicia_sec:learn_ssat}. 
We conclude this section by proposing a conditional distribution based enumeration for compound protected groups in Section~\ref{fairness_justicia_sec:cond_ssat}.

\subsection{Problem Statement}
\label{fairness_justicia_sec:problem_statement}
Given a binary classifier $\alg$ and a probability distribution over dataset $(X,A,Y) \sim \mathcal{D} $, our goal is to verify whether $\alg$ achieves independence and separation metrics with respect to the distribution $\mathcal{D}$. We  focus on a classifier that can be translated to a CNF formula of Boolean variables $\mathbf{B} $. 
The probability $ p_i $ of $\bool_i \in \mathbf{B}$ being assigned to $1$ is induced by the data generating distribution $\mathcal{D}$. 
In our contribution, we reduce the verification problem to solving appropriately designed SSAT instances.
\fi

\subsection{Evaluating Fairness with RE-SSAT Encoding}
\label{fairness_justicia_sec:enumeration_ssat}
In order to verify independence and separation metrics, the core component of {\justicia} is to compute the positive predictive value $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$ for a compound protected group $\mathbf{a}$.  For simplicity, we  initially make some assumptions and discuss their practical relaxations later in this section.   
We first assume the classifier $\alg$ is representable as a CNF formula, namely $\phi_{\hat{Y}}$, such that $ \hat{Y} = 1 $ when $ \phi_{\hat{Y}}$ is satisfied and  $\hat{Y} =0$ otherwise. Since a Boolean CNF classifier is defined over Boolean variables, we  assume all attributes in $X$ and $A$ to be Boolean. Finally, we assume independence of non-protected attributes on protected attributes and $p_i $ is the  probability of the attribute $ \nonsensitive_i $ being assigned to $ 1 $ for any  $\nonsensitive_i \in X  $.

Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$. In the prefix of $ \Phi_{\mathbf{a}} $,  all non-protected Boolean attributes in $X$ are assigned randomized quantification and they are followed by the protected Boolean attributes in $ A $ with existential quantification. The CNF formula $ \phi $ in $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\hat{Y} = 1 | A = \mathbf{a}] $. In order to encode the conditional $ A = \mathbf{a} $, we take the conjunction of the Boolean variables in $ A $ that symbolically specifies the compound protected group $ \mathbf{a} $. For example, we represent two protected attributes: race $ \in $ \{White, Colour\} and sex $ \in $ \{male, female\} by the Boolean variables $ R $ and $ S $  respectively. Hence, the compound groups $\{\textrm{White}, \textrm{male}\}$ and $\{\textrm{Colour}, \textrm{female}\}$ are represented by $ R \wedge S $ and $ \neg R \wedge \neg S $, respectively. Thus, the RE-SSAT formula for computing the probability  $ \Pr[\hat{Y} = 1 | A = \mathbf{a}] $ is
\begin{equation}	\label{fairness_justicia_eq:re}
\begin{split}
	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m}_{\text{non-protected attributes}},  &\underbrace{\exists \sensitive_1,\dots, \exists \sensitive_n}_{\text{protected attributes}},
\phi_{\hat{Y}} \wedge (A=\mathbf{a}).\notag
\end{split}
\end{equation}
In $ \Phi_{\mathbf{a}} $, the existentially quantified variables $ \sensitive_1, \dots, \sensitive_n $ are assigned values  according  to the constraint $ A=\mathbf{a} $. \footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Therefore, by solving the SSAT formula $ \Phi_{\mathbf{a}} $,  the SSAT solver finds the probability $ \Pr[\Phi_{\mathbf{a}}] $ for the protected group $ A=\mathbf{a} $ given the random values of $ \nonsensitive_1, \dots, \nonsensitive_m $, which is the PPV of the protected group $\mathbf{a} $ for the distribution $ \mathcal{D} $ and algorithm $\alg$.

For simplicity, we have described computing the PPV of each compound protected group without considering the correlation between the protected and non-protected attributes. In reality, correlation exists between the protected and non-protected attributes. Thus, the non-protected attributes may have different conditional distributions for different protected groups. We incorporate these conditional distributions in RE-SSAT encoding by evaluating the conditional probability $ p_i = \Pr[\nonsensitive_i =\text{TRUE}| A=\mathbf{a}] $ instead of the independent probability $\Pr[\nonsensitive_i =\text{TRUE}]$ for any $\nonsensitive_i\in \nonsensitive$. We illustrate this method in Example~\ref{fairness_justicia_example:re_ssat}.
%This relaxes the independence requirement.

\begin{example}[RE-SSAT encoding]
	\label{fairness_justicia_example:re_ssat}
	Here, we illustrate the RE-SSAT formula for calculating the PPV for the protected group `age $ \ge 40 $' in the decision tree of Figure~\ref{fairness_justicia_fig:fair_example}. We assign three Boolean variables $ F,I,J $ for the three nodes in the tree such that the literal $ F,I,J $ denote `fitness $ \ge 0.61 $', `income $ \ge 0.29 $', and `income $ \ge 0.69 $', respectively. We consider another Boolean variable $A$  where the literal $ A $ represents the protected group `age $ \ge 40 $'. Thus, the CNF formula  for the decision tree is $ (\neg F \vee I) \wedge (F \vee J) $. From the distribution in Figure~\ref{fairness_justicia_fig:fair_example}, we get $ \Pr[F] = 0.41, \Pr[I] = 0.93 $, and $ \Pr[J] = 0.09 $. Given this information, we calculate the PPV for the protected group `age $ \ge 40 $' by solving the RE-SSAT formula:
	\begin{equation}
	\Phi_A := \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge A.\notag
	\end{equation}
	From the solution to this SSAT formula, we get $ \Pr[\Phi_A] = 0.43 $. Similarly, to calculate the PPV for the group `age $ < 40 $', we replace the unit (single-literal) clause $ A $ with $ \neg A $ in the CNF in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $ where $ \Pr[\Phi_{\neg A}] = 0.43 $. 
	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of $A$ and $\neg A$, both age groups demonstrate equal PPV as the protected attribute is not explicitly present in the classifier. 
	However, there is an implicit bias in the data distribution for different protected groups and the classifier unintentionally learns it. 
	To capture this implicit bias, we calculate the conditional probabilities  $ \Pr[F|A] = 0.01, \Pr[I|A] = 0.99 $, and $ \Pr[J|A] = 0.18 $ from the distribution. Using the conditional probabilities in  $\Phi_A $, we find that $ \Pr[\Phi_A] = 0.18 $ for `age $ \ge 40 $'. For `age $ < 40 $',  we similarly obtain $ \Pr[F|\neg A] = 0.82, \Pr[I|\neg A] = 0.88 $, and $ \Pr[J|\neg A] = 0.01 $, and thus  $ \Pr[\Phi_{\neg A}] = 0.72 $. 
	Therefore, presented RE-SSAT encoding detects the discrimination of the classifier among different protected groups. An astute reader would observe that $I$ and $J$ are not independent. Following~\cite{chavira2008probabilistic}, we can simply capture relationship between the variables using constraints and if needed, auxiliary variables. In this case, it suffices to add the the constraint $J \rightarrow I$. 
\end{example}

\subsubsection{Measuring Fairness Metrics.}
As we compute the probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. 
For that, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ that requires solving exponential (with $n$) number of SSAT instances. 
We elaborate this enumeration approach, namely {\justiciaenum}, in Algorithm~\ref{fairness_justicia_algo:enum}  (Line~\ref{fairness_justicia_algo:justicia_enum_begin}--\ref{fairness_justicia_algo:justicia_enum_end}).

\begin{algorithm}[t!]
	\caption{\justicia: SSAT-based Fairness Verifier}
	\label{fairness_justicia_algo:enum}
	\footnotesize
	\begin{algorithmic}[1]
		\Function{{\justiciaenum}}{$ X,A,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_enum_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y} = 1) $
		%\State $ p_{i} = \mathsf{CalculateProb}(\nonsensitive_i), \forall \nonsensitive_i \in X $
		\ForAll{$\mathbf{a} \in A$ }
		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(\nonsensitive_i | \mathbf{a}), \forall \nonsensitive_i \in X $
		\State $ \phi := \phi_{\hat{Y}} \wedge (A=\mathbf{a}) $
		\State $  \Phi_\mathbf{a} := \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \exists \sensitive_1,\dots, \exists \sensitive_n,  \phi $
		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns a probability}
		\EndFor
		\State \Return $ \max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] $
		\label{fairness_justicia_algo:justicia_enum_end}
		\EndFunction

		\Function{{\justicialearn}}{$ X,A,\hat{Y} $}
		\label{fairness_justicia_algo:justicia_learn_begin}
		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(\nonsensitive_i), \forall \nonsensitive_i \in X $
		\State $  \Phi_\mathbf{ER} := \exists \sensitive_1,\dots, \exists \sensitive_n, \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \phi_{\hat{Y}} $
		\State $  \Phi'_\mathbf{ER} := \exists \sensitive_1,\dots, \exists \sensitive_n, \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \neg \phi_{\hat{Y}} $
		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
		\label{fairness_justicia_algo:justicia_learn_end}
		\EndFunction
	\end{algorithmic}

\end{algorithm}

We calculate the ratio of the minimum and the maximum probabilities according to the definition of disparate impact. 
We compute statistical parity by taking the difference between the maximum and the minimum probabilities of all $ \Pr[\Phi_{\mathbf{a}}] $.
Moreover, to measure equalized odds, we compute two SSAT instances for each compound group with modified values of $ p_i $. 
Specifically, to compute TPR, we use the conditional probability $ p_i = \Pr[\nonsensitive_i|Y=1] $ on samples with class label $ Y = 1 $ and take the difference between the maximum and the minimum probabilities of all compound groups. In addition, to compute FPR, we use the conditional probability $ p_i = \Pr[\nonsensitive_i|Y=0] $ on samples with $ Y = 0 $ and take the difference similarly.
Thus, {\justiciaenum} allows us to compute different fairness metrics using a unified algorithmic framework.

\subsection{Learning Fairness with ER-SSAT Encoding}
\label{fairness_justicia_sec:learn_ssat}
In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the protected attributes. 
Therefore, the enumeration approach may suffer from scalability issues. 
Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their PPVs to measure different fairness metrics.

\subsubsection{Learning the Most Favored Group.}
In an SSAT formula $ \Phi $, the order of quantification of the Boolean variables in the prefix  carries distinct interpretation of the satisfying probability of $ \Phi $.  
In ER-SSAT formula, the probability of satisfying $ \Phi $ is the \textit{maximum} satisfying probability over the existentially quantified variables given the randomized quantified variables (by Rule 2, Sec.~\ref{fairness_justicia_sec:ssat}). 
In this paper, we leverage this property to compute the most favored group with the highest PPV. 
We consider the following ER-SSAT formula. 
\begin{equation}
\Phi_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
 \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
 \label{fairness_justicia_eq:er}
\end{equation}
%In the prefix of the ER-SSAT formula $ \Phi_{\mathsf{ER}} $, the protected variables are  existentially quantified and they are followed by randomized quantified non-protected variables, which is in reverse order  in Eq.~\eqref{fairness_justicia_eq:re}. 
The CNF formula $\phi_{\hat{Y}}$ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound protected group.  Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_n = a^{\max}_n $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum. 
Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_n \}$ achieving the highest PPV. 
%Since an assignment to $\{\sensitive_1, \dots, \sensitive_n\}$ uniquely denotes a compound protected group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.

\subsubsection{Learning the Least Favored Group.}
In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-protected variables $ \nonsensitive_1, \dots, \nonsensitive_m $. In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the protected variables and randomized quantification over the non-protected variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
\begin{equation}
\Phi_{\mathsf{UR}} := \forall \sensitive_1,\dots, \forall \sensitive_n,
\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
\label{fairness_justicia_eq:ar}
\end{equation}
A UR-SSAT formula returns the minimum satisfying probability of $ \phi $ over the universally quantified variables in contrast to the ER-SSAT formula that returns the maximum satisfying  probability over the existentially quantified variables.  
Due to practical issues to solve UR-SSAT formula, in this paper, we leverage the \textit{duality} between UR-SSAT (Eq.~\eqref{fairness_justicia_eq:ar}) and ER-SSAT formulas (Eq.~\eqref{fairness_justicia_eq:er_complement}) 
\begin{equation}
\Phi'_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \neg \phi_{\hat{Y}}.
\label{fairness_justicia_eq:er_complement}
\end{equation}
%\begin{equation}
%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_n}_{\text{protected attributes}},
%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m}_{\text{non-protected attributes}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
%\label{fairness_justicia_eq:er_complement}
%\end{equation}
and solve the UR-SSAT formula on the CNF $ \phi $ using the ER-SSAT formula on the complemented CNF $ \neg \phi $~\cite{littman2001stochastic}. Lemma~\ref{fairness_justicia_thm:dual} encodes this duality.
%We apply such a duality because of the practical unavailability of an UR-SSAT solver (to the best of our knowledge).  Formally, we solve the following ER-SSAT formula to find the least favored group. 
\begin{lemma}\label{fairness_justicia_thm:dual}
Given Eq.~\eqref{fairness_justicia_eq:ar} and~\eqref{fairness_justicia_eq:er_complement},	$ \Pr[\Phi_{\mathsf{UR}}] = 1 - \Pr[\Phi'_{\mathsf{ER}}]  $.
\end{lemma}
As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the protected attributes $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_n\}$ that maximizes $\Phi'_{\mathsf{ER}}$. 
If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$. We present the algorithm for this learning approach, namely {\justicialearn} in Algorithm~\ref{fairness_justicia_algo:enum} (Line~\ref{fairness_justicia_algo:justicia_learn_begin}--\ref{fairness_justicia_algo:justicia_learn_end}).

\iffalse
\begin{algorithm}[t!]
	\caption{\justicialearn: Learning ER-SSAT Encoding}
	\label{fairness_justicia_algo:learn}
	\begin{algorithmic}[1]
		\Function{{\justicialearn}}{$ X,A,\hat{Y} $}
		\State $ \phi_{\hat{Y}} = \mathsf{CNF}(\hat{Y}  = 1) $
		\State $ p_{i} = \mathsf{CalculateProb}(x_i), \forall x_i \in X $
		\State $  \Phi_\mathbf{ER} = \exists a_1,\dots, \exists a_n, \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m. \; \phi_{\hat{Y}} $
		\State $  \Phi'_\mathbf{ER} = \exists a_1,\dots, \exists a_n, \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m. \; \neg \phi_{\hat{Y}} $
		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
		\EndFunction
	\end{algorithmic}
\end{algorithm}
\fi

In ER-SSAT formula of Eq.~\eqref{fairness_justicia_eq:er_complement}, we need to negate the classifier $ \phi_{\hat{Y}} $ to another CNF formula $ \neg \phi_{\hat{Y}} $. The na\"ive approach of negating a CNF to another CNF generates exponential number of new clauses. Here, we can apply Tseitin transformation that increases the clauses linearly while introducing linear number of new variables~\cite{tseitin1983complexity}. As an alternative, we also directly encode the classifier $\alg$ for the negative class label $\hat{Y} = 0$ as a CNF formula and pass it to $\Phi'_{\mathsf{ER}} $, if possible. The last approach is generally more efficient than the other approaches as the resulting CNF is often smaller.

\begin{example}[ER-SSAT encoding]
	\label{fairness_justicia_example:er_ssat}
	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple protected groups. As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another protected group `sex $ \in $ \{male, female\}'. Consider a Boolean variable $ S $ for `sex' where the literal $ S $ denotes `sex = male'. With this new protected attribute, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ A,F,I,J $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}. 
	Hence, we obtain the ER-SSAT formula of $\alg$ to learn the most favored group:
	$ \Phi_{\mathsf{ER}} =  \exists S,\exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; (\neg F \vee I \vee S) \wedge (F \vee J).
	$

	As we solve $ \Phi_{\mathsf{ER}} $, we learn that the assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, i.e. `male individuals with age $ < 40 $' is the most favored group with PPV computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$. Similarly, to learn the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
	$	\Phi_{\mathsf{ER'}} =  \exists S, \exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; \neg((\neg F \vee I \vee S) \wedge (F \vee J)).
	$

	Solving $ \Phi_{\mathsf{ER'}} $, we learn the assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi_{\mathsf{ER'}}] = 0.57 $. Thus, `female individuals with age $ < 40 $' constitute the least favored group with PPV:  $ 1-0.57 = 0.43$. 
	Thus, {\justicialearn} allows us to learn the most and least favored groups and the corresponding discrimination.
\end{example}
We use the PPVs of the most and least favored groups to compute different fairness metrics. We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.
\begin{lemma}
	\label{fairness_justicia_lm:equivalence}
	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound protected group $ \mathbf{a} \in A $. If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
	$\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   
	and
	$\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.   
\end{lemma}
\iffalse
\begin{theorem}
	For an ER-SSAT problem, the na\"ive sample complexity is given by 
	\[ k = O\left(\frac{1}{\epsilon^2} (n + \ln(1/\delta))  \right)\]
	where $\hat{p} - p \leq \epsilon$ with probability $1-\delta$.
\end{theorem}
\red{For a simple sampling based algo after evaluating $k$ assignments, the error between estimates probability and actual probability would differ by $\epsilon$ with probability $1- 2^{n+1} e^{-2\epsilon^2 k}$. So the sample complexity of a naive algo is $k = O(\epsilon^{-2} (\log(1/\delta) + n))$.}
\fi

%\subsection{Conditionally Evaluating an SSAT Encoding}
%\label{fairness_justicia_sec:cond_ssat}

%\begin{algorithm}[t!]
%	\caption{\justiciacond: Conditional Evaluation of RE Encoding \red{Can be removed, only difference is in line 4}}
%	\begin{algorithmic}[1]
%		\Function{{\justiciacond}}{$ X,A,\hat{Y} $}
%		\State $ \phi_{\hat{Y}} = \mathsf{CNF}(\hat{Y}  = 1) $
%		\ForAll{$\mathbf{a} \in A$ }
%		\State $ p_{i} = \mathsf{\blue{calculateCondProb}}(x_i, \mathbf{a}), \forall x_i \in X $
%		\State $  \Phi_\mathbf{a} = \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m, \exists \sensitive_1,\dots, \exists \sensitive_n. \; \phi_{\hat{Y}} \wedge \mathsf{CNF}(A=\mathbf{a}) $
%		\State $ \Pr[\Phi_\mathbf{a}]  = \ma
%thsf{SSAT}(\Phi_\mathbf{a}) $
%		\EndFor
%		\State \Return $ \max\limits_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}], \min\limits_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] $
%		\EndFunction		
%	\end{algorithmic}
%\end{algorithm}

\iffalse
\red{While it is trivial to use conditional probability distribution for the RE encoding, it is not straightforward for the presented ER-encoding, which we leave as a future work. 
}
\fi

\begin{comment}
\red{Let $ \bool_i $ and $ \bool_j $ be two Boolean (non-protected) attributes and we want to encode their pair-wise correlation in {\justicia}. For an assignment to $ \bool_i, \bool_j $, we consider a new Boolean variable $ v $ and add a constraint $ v \leftrightarrow \bool_i \wedge \bool_j $ in the CNF $ \phi $ of the SSAT formula $ \Phi $. Additionally, $ v $ is given a randomized quantification in the prefix of $ \Phi $ and the probability of $ v $ is calculated as the probability of both $ \bool_i $ and $ \bool_j $ assigning to $ 1 $ in the distribution $ \mathcal{D} $. Note that, to encode the total correlation of all Boolean attributes, the mentioned approach introduces exponentially (with the number of attributes) many new variables and add an exponential number of constraints to $ \phi $ with an aim of computing a more precise satisfying probability of $ \Phi $. Optionally, one can encode the correlation among a selected set of attributes of interest in {\justicia}. }
\end{comment}

\subsection{Theoretical Analysis: Error Bounds}\label{fairness_justicia_sec:theory}
We access the data generating distribution through  finite number of samples observed from it. These finite sample set introduce errors in the computed probabilities of the randomised quantifiers being $1$. These finite-sample errors in computed probabilities induce further errors in the computed positive predictive value (PPV) and fairness metrics. We next provide a bound on this finite-sample error.

Let us consider that $\hat{p_i}$ is the estimated probability of a Boolean variable $ \bool_i $ being assigned to $ 1 $ from $k$-samples and $p_i$ is the true probability according to $ \mathcal{D} $. 
%If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.  
Thus, the true satisfying probability $p$ of $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{\bool_i \in \sigma}p_i$.
This probability is estimated as $\hat{p}$ using $k$-samples from the data generating distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$. 
%However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
%\[
%\hat{p} = \sum_\sigma \prod_{\bool_i \in \sigma}\hat{p_i}  \leq \sum_\sigma \prod_{\bool_i \in \sigma}(1 \pm \epsilon_i) p_i = \prod_{i =1}^m (1 \pm \epsilon_i) p = \epsilon_0 p
%\]
\iffalse
If we consider $\epsilon_i = \frac{\ln \epsilon_0}{2^i}$, we obtain
\begin{equation*}
\begin{split}
\prod_{i =1}^m (1 \pm \epsilon_i) &\leq \left(\frac{1}{n} \sum_{i} (1 + \epsilon_i)\right)^n \\
&= (\frac{1}{n} \sum_{i} (1+\frac{\ln \epsilon_0}{2^i})^n\\
&\leq (1  +\frac{\ln \epsilon_0}{n})^n\\
&\leq e^{\ln \epsilon_0} = \epsilon_0.\notag
\end{split}
\end{equation*}
\fi
\begin{theorem}\label{fairness_justicia_thm:sample}
	For an ER-SSAT problem, the sample complexity is given by 
	$ k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0} \right),$
	where $\frac{\hat{p}}{p} \leq \epsilon_0$ with probability $1-\delta$ such that $\epsilon_0 \geq 1$.
\end{theorem}
%Theorem~\ref{fairness_justicia_thm:sample} states that for the prescribed number of samples the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ would also satisfy $\hat{DI} \leq  \epsilon_0 DI, $ and $\hat{SP} \leq 2\epsilon_0 SP$.
%Apply Hoeffding inequality for each term and then use Union bound for $2^n$ existential variables.
\begin{corollary}
	If $k$ samples are considered from the data-generating distribution in {\justicia} such that 
	$
	k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0}\right),
	$
	the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ satisfy, with probability $1-\delta$,
	$
	\hat{DI} \leq  \epsilon_0 DI, \quad \text{and} \quad \hat{SP} \leq 2\epsilon_0 SP.
	$
\end{corollary}
This implies that given a classifier $ \alg \triangleq \Pr(\hat{Y}|X, A) $ represented as a CNF formula and a data-generating distribution $ (X,A,Y) \sim \mathcal{D} $, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
Thus, {\justicia} is a sound framework of fairness verification with high probability.
%\red{Proof of all theorems for arxiv version}
%Use definition of $GF = p_1/p_2$.

\iffalse

\begin{corollary}[Hypothesis]
	If $k$ samples are considered in \justicia and the estimated group fairness value $\hat{GF}$ satisfies
	\begin{equation*}
	1- e^{\epsilon_0} \leq \frac{\hat{GF}}{GF} \leq 1 + e^{\epsilon_0}
	\end{equation*}
	with probability $1-\delta$, then
	\begin{equation*}
	k = O\left((n+ \ln(1/\delta))\frac{m + \ln m}{\ln \epsilon_0}\right).
	\end{equation*}
\end{corollary}
\fi

### 2022-01-26 chapters/fairness/justicia/introduction.tex
adding two fairness papers

\section{Introduction}
%\paragraph{Motivation: ML + Fairness}
Machine learning (ML) is becoming the omnipresent technology of our time. 
ML algorithms are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions.
Thus, human lives are now pervasively influenced by data, ML, and their inherent bias.

\begin{example}\label{fairness_justicia_example:intro}
Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding eligibility for health insurance depending on the fitness and income of the individuals of different age groups (20-40 and 40-60).
Typically, incomes of individuals increase as their ages increase while their fitness deteriorates.
We assume relation of income and fitness depends on the age as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example}.
Now, if we train a decision tree~\cite{narodytska2018learning} on these fitness and income indicators to decide the eligibility of an individual to get a health insurance, we observe that the `optimal' decision tree (ref. Figure~\ref{fairness_justicia_fig:fair_example}) selects a person above and below $40$ years with probabilities $0.18$ and $0.72$ respectively.
This simple example demonstrates that even if an ML algorithm does not explicitly learn to differentiate on the basis of a sensitive attribute, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
\end{example}

\begin{figure}[t!]
	\centering
	\begin{minipage}{.18\columnwidth}
		\centering
		\scalebox{0.5}{	
			\begin{tikzpicture}[x=1.5cm,y=1.8cm]
			% Define nodes
			\node[latent,scale=1.5] (a1) {$\textrm{age}$} ; %
			\node[obs, scale=1.5, below=of a1, xshift=-.6cm] (h) {$\textrm{fitness}$}; %
			\node[obs, scale=1.5, below=of a1, xshift=.6cm] (i) {$\textrm{income}$}; %
			\node[obs, scale=1.5, below=of h, xshift=.6cm] (p) {$\hat{Y}$}; %			
			%%add edge
			\edge[] {a1} {h,i} ;
			\edge[] {h,i} {p} ;
			\end{tikzpicture}
		}
	\end{minipage}\hfill
	\begin{minipage}{.2\columnwidth}
		\centering
		\includegraphics[scale=0.35]{figures/fairness/justicia/example.pdf}
	\end{minipage}\hfill
	\begin{minipage}{.5\columnwidth}
		\scalebox{0.37}{	
			\begin{tikzpicture}[x=1cm,y=1.8cm]
			\node [box, scale=2]                                    (p)      {fitness $\geq 0.61$};
			\node [scale=2, above= -0.1cm of p]  (a)    {Trained Decision Tree};
			\node [scale=2, box, below=  -1cm of p, xshift=-1.4cm]    (a1)    {income\\ $\geq 0.29$};
			\node [scale=2, box, below= -1cm of p, xshift=1.4cm]     (a2)    {income $\geq 0.69$};
			\node [scale=2,below= of a1, xshift=-1.2cm]  (a11)    { $1$};
			\node [scale=2,below= of a1, xshift=1.2cm]   (a12)    { $0 $};
			\node [scale=2,below= of a2, xshift=-1.2cm]  (a21)    { $1$};
			\node [scale=2,below= of a2, xshift=1.2cm]  (a22)    { $0$};
			%
			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
			\end{tikzpicture}}
	\end{minipage}%
	\caption{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}

\end{figure}
\subsubsection{Fair ML.} Statistical discriminations caused by ML algorithms have motivated researchers to develop several frameworks to ensure fairness and several algorithms to mitigate bias.
Existing fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. 
Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of an algorithm to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
Separation metrics, such as equalized odds, define an algorithm to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}.
Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}. %,wu2019counterfactual

In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups. Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same. Sufficiency is satisfied if the eligibility is independent of their age given their attributes are the same.
Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}.
Different algorithms have also been devised to ensure one or multiple of the fairness definitions.
These algorithms try to rectify and mitigate the bias in the data and thus in the prediction-model in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the algorithm~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes~\cite{kamiran2012decision,hardt2016equality}.

\subsubsection{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.

In order to verify fairness as a model property on a dataset, verifiers like \textit{FairSquare}~\cite{albarghouthi2017fairsquare} and \textit{VeriFair}~\cite{bastani2019probabilistic} have been proposed. 
%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
%These papers operate for a specific Boolean sensitive attribute.
These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the attributes in the dataset and a model of a suitable form, and their objective is to verify fairness w.r.t. the distribution and the model.
Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive attributes.
In contrast to the distributional verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}. 
Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.

%\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of an ML algorithm, which is \textit{scalable}, capable of \textit{handling compound protected groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.

\subsubsection{Our Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}.
SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}. In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}. 
SSAT is a conceptual framework that has been employed to capture several fundamental problems in AI such as computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.

%We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning algorithm using a unified scheme.
%Our encodings not only allow us to compute for non-Boolean and compound sensitive attributes but also to scale significantly better than existing formal verifiers. 
%We perform experimental analysis for multiple fairness metrics, datasets, and algorithms to instantiate the efficiency and effectiveness of our approach.

Our contributions are summarised below:
\begin{itemize}
	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of fairness for different datasets and classification algorithms.
	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing algorithms with respect to the data  generating distribution}.
	\item Unlike previously proposed formal distributional verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive attributes.%, and also \red{separation metrics}.
	\item Our experiments validate that our method is more accurate and scalable than the distributional verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
\end{itemize}

It is worth remarking that significant advances in AI bear testimony to the right choice of formulation, for example, formulation of planning as SAT~\cite{kautz1992planning}. In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.

\iffalse
With this growing set of algorithms and definitions, it has become important to measure and verify fairness and bias of different algorithms and datasets. 
One popular approach is to use specific test dataset to compute the related statistical quantities and to certify fairness for that specific test dataset.
AIF360~\cite{aif360-oct-2018} provides a unified framework to implement multiple algorithms and to measure their fairness depending on such test datasets.
Though this method of verification works for a specified datasets, such verifiers do not explain how much a fairness measure depends on which sensitive attribute and is not robust to the selection of test dataset and its size.
\fi

### 2022-01-26 chapters/fairness/justicia/practical.tex
adding two fairness papers

\subsection{Practical Settings}
\label{sec:practical-setting}
We now relax the assumptions on access to Boolean classifiers and Boolean attributes, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous attributes.

\subsubsection{Extending to Decision Trees and Linear Classifiers. }
In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
We extend {\justicia} beyond CNF classifiers to decision trees and linear classifiers, which are widely used in the fairness studies~\cite{zemel2013learning,raff2018fair,zhang2019faht}.
%{\color{red}In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental}.  We leverage these techniques. (not needed)}

\textit{Binary decision trees} are trivially encoded as  CNF formulas.  In the binary decision tree, each node in the tree is a literal. A \textit{path from the root to the leaf} is a conjunction of literals and thus, a \textit{clause}. The \textit{tree} itself is a disjunction of all paths and thus, a \textit{DNF (Disjunctive Normal Form)}. In order to derive a CNF of a decision tree, we first construct a DNF by including all paths terminating at leaves with negative class label ($ \hat{Y} = 0 $) and then complement the DNF to CNF using De Morgan's rule.

\textit{Linear classifiers on Boolean attributes} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}. We consider a linear classifier  $ W^T X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathbb{R}^{|X|} $ and bias $ b \in \mathbb{R} $.  We first normalize $W$ and $b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint.  We then apply  pseudo-Boolean constraints to CNF translation to encode the decision boundary to CNF. This encoding usually introduces additional Boolean variables and results in large CNF. In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider attributes with higher weights only. For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in W $, we can set $ w_i = 0 $. Thus, the attributes with lower weights and thus, less importance do not appear in the encoded CNF.  Moreover, all introduced variables in this CNF translation are given existential ($ \exists $) quantification and they appear in the inner-most position in the prefix of the SSAT formula. Thus, the presented ER-SSAT formulas become effectively ERE-SSAT formulas.

\subsubsection{Extending to Continuous Attributes.}
In practical problems, attributes are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of the input attributes.
In order to perform this Boolean abstraction, each categorical attribute is one-hot encoded and each real-valued attribute is discretised into a set of Boolean attributes~\cite{LKCL2019,GMM20}.

For a binary decision tree, each attribute, including the continuous ones, is compared against a constant at each internal node of the tree. We fix a Boolean variable for each internal node, where the Boolean assignment to the variable decides one of the two branches to choose from the current node.

Linear classifiers are generally trained on continuous attributes, where we apply the following discretization. 
Let us consider a continuous attribute $X_c$, where $w$ is its weight during training. 
We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean attributes and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $. 
For the discretization of $X_c$, we consider the interval-based approach\footnote{Our implementation is agnostic to any discretization technique.}. 
For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned TRUE when the attribute-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned FALSE otherwise. 
Following that, we assign the weight of $ B_i $ to be $ \mu_i\times w $, when $ \mu_i $ is the mean of the $i^{\mathrm{th}}$ interval and  $ B_i $ is TRUE. 
We can show that if we consider infinite number of intervals, $ X_c \approx \sum_i \mu_i B_i $.

### 2022-01-26 chapters/fairness/justicia/prectical.tex
adding two fairness papers

--- 2022-01-26 chapters/fairness/justicia/appendix.tex
@@
-\appendix
-\section{Proofs}
-\begin{lemma}
-	\label{thm:complexity}
-	Solving the ER-SSAT and RE-SSAT problems are $\mathrm{NP}^{\mathrm{PP}}$ hard~\cite{littman2001stochastic}.
-\end{lemma}
-
-\begin{proof}[Proof of Lemma~\ref{thm:complexity}]
-	The decision version of ER-SSAT problem is 
-	\begin{equation}
-		\Phi := \exists a_1,\dots, \exists a_n, \R^{p_{x_1}}x_1, \dots, \R^{p_{x_m}}x_m. \; \Pr[\phi_{\hat{y}}] \geq t,\notag
-	\end{equation}
-	where $t$ is a threshold in $[0,1]$.
-	It is exactly the an E-MAJSAT (or threshold SAT) problem which is $NP^{PP}$ hard~\cite{littman2001stochastic}.
-	If there's no random variable and $t=1$, ER-SSAT reduces to a SAT problem, which is NP-hard. If there's no existential variable, ER-SSAT reduces to a MAJSAT problem, which is PP-hard.
-	Similar arguments also hold for RE-SSAT problem.
-\end{proof}
-
-\begin{equation}
-\Phi_{\mathsf{UR}} := \forall \sensitive_1,\dots, \forall \sensitive_n,
-\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
-\label{eq:ar}
-\end{equation}
-\begin{equation}
-\Phi'_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
-\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \neg \phi_{\hat{Y}}.
-\label{eq:er_complement}
-\end{equation}
-\begin{lemma}\label{thm:dual}
-	Given Eq.~\eqref{eq:ar} and~\eqref{eq:er_complement},	$ \Pr[\Phi_{\mathsf{UR}}] = 1 - \Pr[\Phi'_{\mathsf{ER}}]  $.
-\end{lemma}
-
-\begin{proof}[Proof of Lemma~\ref{thm:dual}]
-	Both $ \Phi_{\mathsf{UR}} $ and $ \Phi'_{\mathsf{ER}} $ have  random quantified variables in the identical order in the prefix. According to the definition of SSAT formulas,
-\begin{equation}
-\Pr[\Phi_{\mathsf{UR}}] = \min\limits_{a_1, \dots,a_n} \Pr[\phi_{\hat{Y}}] \text{ and } \Pr[\Phi'_{\mathsf{ER}}] = \max\limits_{a_1, \dots,a_n} \Pr[\neg\phi_{\hat{Y}}].\notag
-\end{equation}
-We can show the following duality between ER-SSAT and UR-SSAT,
-\begin{equation}
-\begin{split}
-\Pr[\Phi'_{\mathsf{ER}}] &= \max\limits_{a_1, \dots,a_n} \Pr[\neg\phi_{\hat{Y}}]  \\
-&= \min\limits_{a_1, \dots,a_n} (1 - \Pr[\phi_{\hat{Y}}])\\
-&= 1 - \min\limits_{a_1, \dots,a_n}  \Pr[\phi_{\hat{Y}}]\\
-&= 1 - \Pr[\Phi_{\mathsf{UR}}].\notag
-\end{split}
-\end{equation}
-\end{proof}
-
-\begin{lemma}
-	\label{lm:equivalence}
-	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound protected group $ \mathbf{a} \in A $.
-If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
-	$\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   
-	and
-	$\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.   
-\end{lemma}
-
-\begin{proof}[Proof of Lemma~\ref{lm:equivalence}] 
-	\blue{It is trivial} that the PPV of most favored group $ \mathbf{a}_{\mathsf{fav}} $ is the maximum PPV of all compound groups $ \mathbf{a} \in A $.
-Similarly, the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}} $ is the minimum PPV of all compound groups $ \mathbf{a} \in A $.
-
-	By construction of the SSAT formulas, the PPV of $ \mathbf{a}_{\mathsf{fav}} $ and $ \mathbf{a}_{\mathsf{unfav}} $ are $ \Pr[\Phi_{\mathsf{ER}}] $ and $ \Pr[\Phi_{\mathsf{UR}}] $ respectively.
-Since $ \Pr[\Phi_{\mathbf{a}}] $ is the PPV of the compound group $ \mathbf{a} $,
-	\begin{equation}
-	\begin{split}
-		\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]
-	\text{ and }
-	\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}].\notag
-	\end{split}
-	\end{equation}
-\end{proof}
-
 \section{Practical Settings}
+\label{sec:practical-setting}
 In this section, we  relax assumptions of Boolean classifiers and Boolean attributes and extend {\framework} to verify fairness metrics in a more practical setting.
@@
 \subsubsection{Beyond CNF Classifiers.}
-In the presented SSAT approach for verifying fairness, we assume the classifier $ \hat{Y} $ to be represented as a CNF formula.
+In the presented SSAT approach for verifying fairness, we assume the classifiers \red{$ \hat{y} $ (or $ \mathcal{A} $, why not $ \hat{Y} $)} to be represented as a CNF formula.
 In the literature of interpretable machine learning, several studies have been conducted for learning CNF classifiers in the supervised learning setting, which include but are not limited to the work of~\cite{angelino2017learning,malioutov2018mlic,ghosh19incremental,yu2020computing}.
@@
 
-\textbf{Encoding Decision Trees as CNF.} Existing  rule-based classifiers, for example, binary decision trees  can be trivially encoded as  CNF formulas.
+\textbf{Decision Trees as CNF.} Existing  rule-based classifiers, for example, binary decision trees  can be trivially encoded as  CNF formulas.
 In the binary decision tree, each node in the tree is a literal.
@@
 Therefore, for any input that is classified positive by the decision tree satisfies $ \phi $ and vice versa.
-In {\frameworklearn} for learning the least favored group, we can construct a negated CNF classifier in Eq.~\ref{eq:er_complement} by only including paths terminating on positive labeled leaves.
+In {\frameworklearn} for learning the least favored group, we can construct a negated CNF classifier in Eq.~\ref{eq:er_complement} by only including paths terminating on positive labelled leaves.
 
-\textbf{Encoding Linear Classifiers as CNF.} Linear classifiers on Boolean attributes can  be encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}.
-We consider a linear classifier  $ W\cdot X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathcal{R}^{|X}| $ and bias $ b \in \mathcal{R} $.
+\textbf{Linear Classifiers as CNF.} Linear classifiers on Boolean attributes can  be encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}.
+We consider a linear classifier  $ W\cdot X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathbb{R}^{|X}| $ and bias $ b \in \mathbb{R} $.
 We first normalize $ W$ and $b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint, e.g., \textit{at-least} $ k $ constraint.
@@
 We can show trivially that if we consider infinite number of intervals, $ x \approx \sum_i \mu_ib_i $.
-
-\begin{example}[ER-SSAT encoding]
-	\label{example:er_ssat}
-	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple protected groups.
-As the example in Figure~\ref{fig:fair_example} is degenerate for this purpose, we introduce another protected group `sex $ \in $ \{male, female\}'.
-Consider a Boolean variable $ S $ for `sex' where the literal $ S $ denotes `sex = male'.
-With this new protected attribute, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ F,I,J $ have same distributions as discussed in Example~\ref{example:re_ssat}.
-	Hence, we obtain the ER-SSAT formula of $\alg$ to learn the most favored group:
-	\begin{equation}
-	\Phi_{\mathsf{ER}} :=  \exists S,\exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; (\neg F \vee I \vee S) \wedge (F \vee J).\notag
-	\end{equation}
-	As we solve $ \Phi_{\mathsf{ER}} $, we learn that the assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, i.e.
-`male individuals with age $ < 40 $' is the most favored group with PPV computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
-Similarly, to learn the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
-	\begin{equation}
-	\Phi_{\mathsf{ER'}} :=  \exists S, \exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; \neg((\neg F \vee I \vee S) \wedge (F \vee J)).\notag
-	\end{equation}
-	Solving $ \Phi_{\mathsf{ER'}} $, we learn the assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi_{\mathsf{ER'}}] = 0.57 $.
-Thus, `female individuals with age $ < 40 $' constitute the least favored group with PPV:  $ 1-0.57 = 0.43$.
-	Thus, {\frameworklearn} allows us to learn the most and least favored groups and the corresponding discrimination.
-\end{example}
-
-\section{Experiments}
-\subsection{Experimental Setup}
-Since both {\framework}  and FairSquare take a  probability distribution of the attributes as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution.
-
-\subsection{Comparative Evaluation of Two Encodings}
-\iffalse
-\red{Accuracy is removed}
-\textbf{Accuracy.} In Section~\ref{sec:framework},  we have theoretically shown the equivalence between RE and ER encodings  and experimentally validate this in Figure~\ref{fig:equivalence} where both encodings output the same value for different fairness metrics.
-In the RE encoding, we consider two variations: namely `RE' and `RE(cor)' where the former does not consider the correlation between the protected and the non-protected attributes and the later includes correlation.
-\blue{In Figure~\ref{fig:equivalence}, we find that disparate impact usually decreases when the correlation is considered.
-This result is justified by the implicit dependence between the protected and non-protected attributes that the classifiers learn during training.}
-\fi
-
-%\textbf{Execution Time.} 
-While both {\frameworkenum} and {\frameworklearn}  have the same output, the {\frameworklearn} encoding  improves exponentially  in runtime  than {\frameworkenum}  on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fig:runtime_diff_encodings}.
-This analysis justifies that the na\"ive enumeration-based approach cannot verify large-scale fairness problems containing multiple protected attributes.
-\begin{figure}[t!]
-	\begin{center}
-		\includegraphics[scale=.32]{figures/encoding_runtime_Adult_DT.pdf}
-		\includegraphics[scale=.32]{figures/encoding_runtime_Adult_IMLI.pdf}
-		\hfill
-		\caption{Runtime comparison of different encodings for varying total protected groups in the Adult dataset}
-		\label{fig:runtime_diff_encodings}
-	\end{center}
-\end{figure}\textbf{}
-
-\begin{table*}[t!]        
-	\centering
-	\caption{Verification of different fairness enhancing algorithms for multiple datasets and classifiers using {\framework}.
-Numbers in bold refer to fairness improvement  compared against the unprocessed (orig.) dataset.
-RW and OP refer to reweighing and optimized-preprocessing algorithm respectively.}\label{tab:fair_algo_verification_appendix}
-	\vspace*{-.2em}
-	%        \setlength{\tabcolsep}{.3em}
-	\begin{tabular}{ll
-			ccc
-			ccc
-			%            				ccc
-			%            				ccc
-			ccc
-			ccc}
-		\toprule
-		\multirow{3}{*}{Classifier}& Dataset $ \rightarrow $   & 
-		\multicolumn{6}{c}{German} \\
-		\cmidrule(lr){3-8}
-		& Protected  $ \rightarrow $ & 
-                \multicolumn{3}{c}{Age}   & \multicolumn{3}{c}{Sex}  
-		\\ 
-		\cmidrule(lr){3-5}
-		\cmidrule(lr){6-8}
-
-		& Algorithm  $ \rightarrow $ &  
-		orig. & RW & OP &
-		orig. & RW & OP \\ 
-		\midrule
-
-		\multirow{3}{*}{\shortstack{Logistic \\ regression}}
-		& Disparte impact&  $ 0.00 $ &  $ 0.00 $ &  $ \mathbf{0.31} $ &  $ 0.27 $ &  $ \mathbf{0.46} $ &  $ 0.17 $  \\
-		& Stat. parity&  $ 0.45 $ &  $ \mathbf{0.03} $ &  $ \mathbf{0.12} $ &  $ 0.03 $ &  $ \mathbf{0.02} $ &  $ 0.07 $  \\
-		& Equalized odds&  $ 0.65 $ &  $ \mathbf{0.04} $ &  $ \mathbf{0.14} $ &  $ 0.10 $ &  $ \mathbf{0.08} $ &  $ 0.13 $  \\
-		\midrule
-		\multirow{3}{*}{\shortstack{Decision \\ tree}}
-		& Disparte impact&  $ 0.00 $ &  $ \mathbf{0.56} $ &  $ \mathbf{0.12} $ &  $ 0.35 $ &  $ \mathbf{0.37} $ &  $ \mathbf{0.38} $  \\
-		& Stat. parity&  $ 0.35 $ &  $ \mathbf{0.02} $ &  $ \mathbf{0.22} $ &  $ 0.05 $ &  $ 0.10 $ &  $ 0.11 $  \\
-		& Equalized odds&  $ 0.36 $ &  $ \mathbf{0.05} $ &  $ \mathbf{0.28} $ &  $ 0.06 $ &  $ 0.16 $ &  $ 0.17 $  \\
-
-		\bottomrule
-	\end{tabular}
-\end{table*}
-
-\begin{comment}
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_op_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_op_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_op_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_op_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying optimized preprocessing (op) algorithm on Adult dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_op_DT_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_op_DT_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_op_LR_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_op_LR_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying optimized preprocessing (op) algorithm on German-credit dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_op_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_op_DT_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_op_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_op_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying optimized preprocessing (op) algorithm on Compas dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_DT_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_DT_age.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_LR_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_LR_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_op_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_op_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_rw_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_rw_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_rw_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_rw_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Adult_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Adult_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying reweighing (rw) algorithm on Adult dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_rw_DT_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_rw_DT_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_rw_LR_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_rw_LR_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_German_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_German_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying reweighing (rw) algorithm on German-credit dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_rw_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_rw_DT_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_rw_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_rw_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/DI_Compas_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/SPD_Compas_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Result for applying reweighing (rw) algorithm on Compas dataset.
-We show results for three verifiers: RE-SSAT encoding with correlation (RE-cor), RE-SSAT encoding without correlation (RE), and statistical tool AIF360 and for two classifiers: decision tree (DT) and logistic regression (LR).
-If the preprocessing algorithm is successful, we observe an increase in disparate impact and a decrease in statistical parity difference as we apply the fairness algorithm.
-RE-cor and RE finds instances where the preprocessing algorithm fails to mitigate bias in the dataset.}
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Adult_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Adult_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_DT_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_DT_race.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_LR_race.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_LR_race.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_Compas_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_Compas_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_DT_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_DT_age.pdf}
-		} \hfil  
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_LR_age.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_LR_age.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_DT_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_DT_sex.pdf}
-		} \hfil
-
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_before_German_rw_LR_sex.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.5]{figures/sampling_DI_after_German_rw_LR_sex.pdf}
-		} \hfil
-
-	\end{center}
-
-	\caption{Effect of sampling.} 
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race, sex.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_race.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_dt_sex.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race, sex.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_race.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_lr_sex.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race, sex.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_race.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_sex, age.pdf}
-	\includegraphics[width=.23\linewidth]{figures/equivalence_di_adult_mlic_sex.pdf}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race, sex.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_race.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_dt_sex.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race, sex.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_race.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_lr_sex.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race, sex.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_race.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_sex, age.pdf}
-\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_adult_mlic_sex.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_dt_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_lr_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_german_mlic_sex.pdf}
-
-\end{figure*}
-
-\begin{figure*}
-		\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_dt_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_lr_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_age, sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_german_mlic_sex.pdf}
-
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_dt_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_lr_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_di_titanic_mlic_sex.pdf}
-
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_dt_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_lr_sex.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, passenger class, age.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex, passenger class.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/equivalence_spd_titanic_mlic_sex.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_dt.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_lr.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_adult_mlic.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_dt.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_lr.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_german_mlic.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_dt.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_lr.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_ricci_mlic.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_dt.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_lr.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/encoding_runtime_titanic_mlic.pdf}
-\end{figure*}
-
-\begin{figure*}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_mlic_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_mlic_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_mlic_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_mlic_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_mlic_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_dt_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_lr_RE.pdf}
-\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_mlic_RE.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_adult_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_adult_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_di_adult_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_race_spd_adult_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_adult_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_adult_mlic_RE(cor).pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_german_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_german_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_german_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_german_mlic_RE.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_mlic_RE.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_di_titanic_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_age_spd_titanic_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_di_titanic_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_passenger class_spd_titanic_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_di_titanic_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_sex_spd_titanic_mlic_RE(cor).pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_di_ricci_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Position_spd_ricci_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_di_ricci_mlic_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_dt_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_dt_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_lr_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_lr_RE.pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_mlic_RE(cor).pdf}
-	\includegraphics[width=0.23\linewidth]{figures/sensitive_attribute_Race_spd_ricci_mlic_RE.pdf}
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race, sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_race.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE(cor)_sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race, sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_race.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_adult_lr_RE_sex.pdf}
-
-\end{figure*}
-
-\begin{figure*}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_age, sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE(cor)_sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_age, sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_german_lr_RE_sex.pdf}
-\end{figure*}
-
-\begin{figure*}
-
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Position.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Race, Position.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE(cor)_Race.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Position.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Race, Position.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_ricci_lr_RE_Race.pdf}
-\end{figure*}
-
-\begin{figure*}
-
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_passenger class, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_passenger class.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, passenger class, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex, passenger class.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE(cor)_sex.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_passenger class, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_passenger class.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, passenger class, age.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex, passenger class.pdf}
-	\includegraphics[width=.235\linewidth]{figures/threshold_titanic_lr_RE_sex.pdf}
-\end{figure*}
-
-\end{comment}

### 2022-01-26 chapters/fairness/justicia/preliminaries.tex
adding two fairness papers

\section{Background: Fairness and SSAT}
\label{fairness_justicia_sec:preliminaries}
In this section, we define different fairness metrics for a supervised learning problem. Following that, we discuss Stochastic Boolean Satisfiability (SSAT) problem.

\subsection{Fairness Metrics for Machine Learning}\label{fairness_justicia_sec:fairness}
Let us represent a dataset $D$ as a collection of triples $(X, A, Y)$ sampled from an underlying data generating distribution $\mathcal{D} $.
$X \triangleq \lbrace \nonsensitive_1, \ldots, \nonsensitive_m\rbrace \in \mathbb{R}^m $ is the set of non-protected (or non-sensitive) attributes.
$A \triangleq \lbrace A_1, \ldots, A_n\rbrace$ is the set of categorical protected attributes.
$Y$ is the binary label (or class) of $(X, A)$. 
A compound protected attribute $\mathbf{a} = \{a_1, \ldots, a_n\}$ is a valuation to all $A_i$'s and represents a \textit{compound protected} group. For example, $A = \{\textrm{race}, \textrm{sex}\}$, where $\textrm{race} \in \{\textrm{Asian}, \textrm{Colour}, \textrm{White}\}$ and  $\textrm{sex} \in \{\textrm{female},  \textrm{male}\}$. 
Thus, $\mathbf{a} =  \{\textrm{Colour}, \textrm{female}\}$ is a compound protected group. 
We define $\alg \triangleq \Pr(\hat{Y}|X, A)$ to be a binary classifier trained from samples in the distribution $\mathcal{D} $. 
Here, $\hat{Y}$ is the predicted label (or class) of the corresponding data.

As we illustrated in Example~\ref{fairness_justicia_example:intro}, a classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound protected groups over others~\cite{chouldechova2020snapshot}.
Now, we describe two family of fairness metrics that compute bias induced by a classifier and are later verified by {\justicia}.

\begin{comment}
$ X $ is the set of non-protected attributes. Each non-protected attribute $ x_i \in X $ takes the valuation as real numbers or categories. 
We consider a multi-protected attributes setting where $ A $ is the set of protected (or sensitive) attributes. Each protected attribute $ a_i \in A $ is a categorical attribute with at least two categories.  
$ y $ is the binary class label of the data sample.

Based on the valuation of the protected attributes, the samples are  tagged as being in different protected groups, i.e., 
the privileged group(s)  and the unprivileged group(s). For instance, let the dataset contain two protected attributes:  race and gender. The attribute race can be white, black or Asian and the attribute gender can be male or female. For simplicity, we assume that the protected group white-male is the most privileged group and all other protected groups (attribute combinations) refer to unprivileged groups with black-female being the most unprivileged group. Generally the dataset contains bias that results in the discrimination between the privileged and unprivileged groups. In this paper, we do not require the knowledge of privileged and unprivileged groups beforehand. We instead ask for the set $ A $ of protected attributes from which we derive privileged and unprivileged groups. 
\end{comment}

\begin{comment}
We now consider a more specific setting where the samples are divided into privileged and unprivileged groups based on the valuation of $ a $, i.e, when $ a=1 $, the sample is included into a privileged group and vice versa. 
\end{comment}

\subsubsection{Independence Metrics of Fairness.} 
The \textit{independence (or calibration) metrics} of fairness state that the output of the classifier should be independent of the compound protected group.
A notion of independence is referred to \textit{group fairness} that specifies an \textit{equal positive predictive value (PPV) across all compound protected groups} for an algorithm $\alg$, i.e., $\Pr[\hat{Y} = 1 | A = \mathbf{a}, \alg] = \Pr [\hat{Y} =1 | A = \mathbf{b}, \alg], \forall \mathbf{a}, \mathbf{b} \in A$.
Since satisfying group fairness exactly is hard, relaxations of group fairness, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.

\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, a classifier satisfies $(1 - \epsilon)$-disparate impact if, for $\epsilon \in [0,1] $,
\[
	\min_{\mathbf{a} \in A} \Pr[\hat{Y} =1 | \mathbf{a}, \alg]  \ge (1 - \epsilon) \max_{\mathbf{b} \in A} \Pr[\hat{Y} =1 | \mathbf{b}, \alg].
\]
Another popular relaxation of group fairness, \textit{statistical parity} (SP) measures the difference of PPV among the compound groups, and prescribe this to be near zero. Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $, 
\[
\max_{\mathbf{a}, \mathbf{b} \in A}|\Pr[\hat{Y} =1 | \mathbf{a}, \alg] - \Pr [\hat{Y} = 1| \mathbf{b}, \alg]| \le \epsilon.
\]
For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier $\alg$.

\begin{comment}These two metrics of fairness has a disadvantage that a fully accurate classifier with different base rates (i.e., the proportion of actual positive prediction) in different protected groups may be considered unfair. As a result, similar individuals from different protected groups may be treated differently in order to achieve group fairness{\textemdash} such treatment is prohibited by law in some cases. 
\end{comment}

\subsubsection{Separation Metrics of Fairness.}
In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive attributes $A$ given the actual class labels $Y$. In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound protected groups. 
Lower value of equalized odds indicates better fairness.
A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound protected groups $\mathbf{a}, \mathbf{b} \in A$,
\begin{equation}
\begin{split}
|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1| A = \mathbf{b}, Y = 0]| &\le \epsilon, \\ 
|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1| A = \mathbf{b}, Y = 1]| &\le \epsilon.\notag
\end{split}
\end{equation}

%\begin{comment}
%In contrast to disparate impact and statistical parity difference metrics, a fully accurate classifier will necessarily satisfy the two constraints in equalized odds measure.

%\textbf{Group fairness.}  Group fairness requires the probability for an individual to be assigned the favorable outcome to be equal across the privileged  and unprivileged groups. Throughout the manuscript, by favorable outcome we refer to the predictor $ \hat{y} = 1 $ in  the standard binary classification setting. Now group fairness is formally defined in the following. 
%
%\[
%\Pr(\hat{y} =1 | a =1  ) = \Pr (\hat{y} = 1| a = 0)
%\]
%
%By abusing notations, we use the following shorthand. 
%
%\[
%\Pr_1(\hat{y} =1) = \Pr_0(\hat{y} = 1)
%\]
%
%
%
%
%In practice, the above definition can be relaxed. Thus we define the notion of $ \epsilon $-\textit{group fairness} in the following. 
%\[
% |\Pr_1(\hat{y} =1) - \Pr_0(\hat{y} = 1) |< \epsilon
%\]
%
%\end{comment}

In this paper, we formulate verifying the aforementioned independence and separation metrics of fairness as stochastic Boolean satisfiability (SSAT) problem, which we define next.

\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form (CNF)} if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
Let $ \sigma $ be an assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $ where $ 1 $ is logical TRUE and $ 0 $ is logical FALSE. The propositional  \textit{satisfiability} problem (SAT)~\cite{biere2009handbook} finds an assignment $ \sigma $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$. 
In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is concerned with the  probability of the satisfaction of the formula $\phi$. 
An SSAT formula is of the form
\begin{equation}\label{fairness_justicia_eq:ssat}
\Phi = Q_1\bool_1, \dots, Q_m \bool_m,\; \phi, 
\end{equation}
where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers over the Boolean variable $\bool_i$ and $\phi$ is a quantifier-free CNF formula. In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $. In case of randomized quantification $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $. Given an SSAT formula $ \Phi $, let $ \bool $ be the outermost variable in the prefix. The satisfying probability of $ \Phi $ can be computed by the following \textit{rules}:
\begin{enumerate}
	\item $ \Pr[\text{TRUE}] = 1 $,  $ \Pr[\text{FALSE}] = 0 $, 
	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{TRUE}$,
\end{enumerate}
where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the formula $ \phi $ with $ 1 $ and $ 0 $ respectively. In this paper, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT. In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
\begin{lemma}{~\cite{littman2001stochastic}}
	\label{fairness_justicia_thm:complexity}
	Solving the ER-SSAT and RE-SSAT problems are $\mathrm{NP}^{\mathrm{PP}}$ hard.
\end{lemma}

The problem of SSAT and its variants have been pursued by theoreticians and practitioners alike for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}. We refer the reader to~\cite{lee2017solving,lee2018solving} for detailed survey. It is worth remarking that the past decade has witnessed a significant performance improvements thanks to close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.

%In these extreme situations, the reduced problem is NP-hard and PP-hard respectively.
%We can show solving ER- and RE-SSAT are equivalent to solving an E-MAJSAT/e-threshold SAT problem. Thus it is $NP^{PP}$ hard. 
%Equivalent formulation of this is computing MAP in a belief network.

\begin{comment}
	The Stochastic Boolean Satisfiability problem (SSAT) allows existential ($ \exists $) and randomized ($ \R $) quantifiers over the Boolean variables and evaluates the probability of satisfaction of the CNF formula.  Based on the order of quantified variables, an SSAT instance can be either a random-exists (\textbf{RE}) SSAT  formula or an exist-random (\textbf{ER}) SSAT formula.

	Let $ \phi(E,R) $ be a CNF formula over Boolean variables $ E \cup R $ where each $ e_i \in E $ is an exist quantified variable and each $ r_j \in R $ is a randomized quantified variable.  For each $ r_j $, $ p_j = \Pr(\sigma(r_j) = 1) $ denotes the probability of $ r_j $ assigning to TRUE. Now an \textbf{RE}-SSAT formula is defined as

	\[
	\R^{p_1} r_1, \dots, \R^{p_m} r_m, \exists e_1, \dots, \exists e_n \; \phi (E,R)
	\]

	where $ |R| = m,  |E| = n  $. Intuitively,  for the random assignment of $ r_1, \dots, r_m $ there exists assignments of $ e_1, \dots, e_n $, for which, we ask the \red{maximum}  probability of satisfaction of $ \phi $. Similarly, an \textbf{ER}-SSAT formula is defined as 
	\[
	\exists e_1, \dots, \exists e_n, \R^{p_1} r_1, \dots, \R^{p_m} r_m, \; \phi (E,R). 
	\]

	is to evaluate a Boolean formula containing both existential and randomized quantifiers~\cite{littman2001stochastic}.  For example:

	\[
	\exists x_1, \R^{p_2} y_2, \exists x_3, \dots, \R^{p_n} y_n (E[\phi (\mathbf{x})]\ge \theta)
	\]

	In words, this formula asks whether there is a value for $ x_1 $ such that for random values of $ y_2 $ (choose $ y_2 = 1 $ with probability $ p_2 $), there exists a value of $ x_2 $, such that$ \dots $ the expected value of the Boolean formula $ \phi(\mathbf{x}) $ meets or exceeds a threshold $ 0 \le \theta \le 1 $. Here variables $ x_i, y_i \in \mathbf{x} $ are Boolean.

	While the above definition of SSAT has alternating quantifiers, it is not a strict restriction. One could first place random quantifiers followed by exists quantifiers or vice versa. Moreover, the threshold $ \theta $ can be omitted from the definition where the question is to find out the probability of $ \phi(\mathbf{x}) =1 $ given the quantifiers over Boolean variables $ x_i $ and $ y_i $.

\end{comment}

### 2022-01-26 chapters/fairness/justicia/table_synthetic.tex
adding two fairness papers

\begin{table}[t!]
    \centering
%        \vspace*{-.2em}
        \setlength{\tabcolsep}{.1em}
            \begin{tabular}{ccccccc}
                \toprule
                Metric & Exact  & {\justicia} & FairSquare & VeriFair & AIF360\\
                 \midrule
				Disparate impact &  $ 0.26 $  &  $ 0.25 $    &  $ 0.99 $  &  $ 0.99 $  &  $ 0.25 $  \\
				Stat. parity &  $ 0.53 $  &  $ 0.54 $    & \textemdash & \textemdash &  $ 0.54 $  \\
                \bottomrule
    \end{tabular}
 \caption{Results on synthetic benchmark.  `\textemdash'~ refers that the verifier cannot compute the metric. }
\label{fairness_justicia_tab:synthetic}
\end{table}

### 2022-01-26 chapters/introduction.tex
adding two fairness papers

	\vspace{1em}
\section{Introduction}
The tremendous success of machine learning in recent years has found its applications in different high-stake and  safety-critical domains, such as medical, law, education, and transportation. In such domains, the interpretability and fairness of machine learning classifiers are of paramount importance. In this work, we focus on improving the interpretability and fairness in machine learning.

There are two-fold objectives in interpretable machine learning: (i) designing classifiers that are interpretable by design and (ii) leveraging already interpretable models to explain the prediction of black-box classifiers. In this work, we propose an efficient learning framework~\cite{ghosh19incremental} for a rule-based classifier\textemdash a popular interpretable model because of expressing the decision function in terms of rules corresponding to input features.  Our framework allows us to learn classification rules expressible in propositional logic, for example, CNF (Conjunctive Normal Form) and DNF (Disjunctive formal form) formulas.  The benefit of learning CNF classifiers is that this is the fundamental block for learning other popular interpretable representations: decision lists and decision sets. Learning rule-based classifiers is an intractable problem because of solving a combinatorial optimization problem. To efficiently learn these classifiers, we propose a MaxSAT-based formulation and scale it by combining incremental learning. The result is an efficient learning framework that can classify datasets even with 1 million samples while also balancing between high accuracy and small rule size. As an application of incremental learning, we propose to learn more relaxed logical rules~\cite{ghosh20classification} which demonstrate higher succinctness in their representation while achieving better accuracy. Furthermore, as an application of MaxSAT, we solve group testing problems based on compressed sensing~\cite{ciampiconi20maxsat}. Our MaxSAT-based formulation for group testing has impressive scalability than state-of-the-art approaches.

We study the application of interpretable machine learning models in explaining the prediction of black-box models such as neural networks. To this end, we propose a PAC-based (Probably Approximately Correct) local explanations of feed-forward and recurrent neural networks using rule-based classifiers and their variants~\cite{neider2020probably,ghosh2020formal}. To explain feed-forward neural networks, we learn flexible classification rules using SyGuS~\cite{neider2020probably}. Such rules have the theoretical property that they are PAC approximate to the neural network within the locality of a query of interest. Similarly, to explain recurrent neural networks, we propose learning explanations in LTL formulas (Linear Temporal Logic)~\cite{ghosh2020formal}.

Machine learning models may become biased in their decisions towards certain demographic populations. One important question in fairness in machine learning is to verify whether the model is unfair or not with respect to multiple definitions of fairness studied in the literature. To this end, we propose a scalable fairness verification framework based on formal methods, particularly stochastic satisfiability (SSAT)~\cite{ghosh2020justicia} and stochastic subset-sum problem (S3P)~\cite{ghosh2021algorithmic}. The input to the fairness verifier is a machine learning model and the distribution of input features and the output is an assessment of different fairness metrics that the model obtains given the distribution. Our fairness verifier has the following advantages over existing verifiers. We can efficiently verify the fairness of compound sensitive groups\textemdash such as, white-male and black-female\textemdash by combining multiple sensitive features, unlike existing works that only consider a single binary sensitive feature. We also consider feature-correlation represented as a Bayesian network whereas existing methods consider limited correlation among features. As a result, the proposed verifier has higher accuracy than existing verifiers. Finally, our verifier is more scalable than existing verifiers because of the efficient formulation based on SSAT and S3P.

As future work, we aim to combine the interpretability and fairness of machine learning in a single framework. In particular, we explain the unfairness of models by combining explanation methods in machine learning. Our hypothesis is that such a combination would be effective in identifying the bias of a model and also mitigate the bias through imperative actions.

### 2022-01-26 figures/fairness/fvgm/tables/dag_structure_complexity.tex
adding two fairness papers

\begin{figure}[!t]
		\begin{center}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_DI_Adult_LR_sex__age}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_Adult_LR_sex__age}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_notears_Adult_LR_sex__age}}
%
%
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_DI_Compas_LR_age__sex}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_Compas_LR_age__sex}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_notears_Compas_LR_age__sex}}
%			
%			
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_DI_Ricci_LR_Race__Position}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_Ricci_LR_Race__Position}}
%			\subfloat[]{\includegraphics[scale=0.25]{figures/dag_complexity_DAG_time_notears_Ricci_LR_Race__Position}}
%

	\subfloat{\includegraphics[scale=0.24]{figures/cactus_dag_LR_time}}
	\subfloat{\includegraphics[scale=0.24]{figures/cactus_dag_LR_time_notears}}

		\end{center}

		\caption{Effect of number of variables in the learned Bayesian Network on computation time of {\framework}. In both plots, we vary $ \frac{|\mathbf{V}|}{|\nonsensitive \cup \sensitive|} $, that is the ratio between the number of variables in the Bayesian Network to the number of features. We observe that as this ratio increases to $ 1 $, both runtime of {\framework} (left plot) and network learning time (right plot) increase. } 
		\label{fig:results_DAG_complexity}
\end{figure}

### 2022-01-26 figures/fairness/fvgm/tables/ssat_vs_subset_sum.tex
adding two fairness papers

\begin{figure}
		\begin{center}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Adult_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Adult_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Adult_LR}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Adult_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Adult_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Adult_SVM}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Compas_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Compas_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Compas_LR}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Compas_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Compas_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Compas_SVM}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Ricci_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Ricci_LR}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Ricci_LR}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_Independent_Ricci_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_DAG_Ricci_SVM}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/ssat_vs_subsetsum_time_scatter_plot_Ricci_SVM}}\\

			\subfloat[]{\includegraphics[scale=0.25]{figures/cactus_subsetsum_ssat_LR_time_.pdf}}
			\subfloat[]{\includegraphics[scale=0.25]{figures/cactus_subsetsum_ssat_SVM_time_.pdf}}

		\end{center}

		\caption{Scalability between {\framework} and Justicia. \red{Each instance represents the runtime of verifying on a dataset and a set of sensitive attributes.}}
\end{figure}

### 2022-01-26 tables/fairness/fvgm/fairness_attacks.tex
adding two fairness papers

\begin{figure}[!t]
	\begin{center}
		\subfloat{\includegraphics[scale=0.24]{figures/disp_fairness_attack}}
		\subfloat{\includegraphics[scale=0.24]{figures/stat_fairness_attack}}\\
	\end{center}
	\caption{Verifying fairness poisoning attack using {\framework}. The red line denotes safety margin, which being exceeded denotes system-vulnerability by the attack algorithm. As the number of poisoned samples increase, disparate impact (DI) decreases and statistical parity (SP)  increases.}
	\label{fig:attack_extended}
\end{figure}

### 2022-01-26 tables/fairness/fvgm/fairness_influence_functions.tex
adding two fairness papers

\begin{figure}[h]
	\begin{center}

		\subfloat[COMPAS]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_compas_lr_sex_0}}
		\subfloat[COMPAS]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_compas_lr_sex_1}}
		\subfloat[COMPAS]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_compas_lr_sex_DI}}\\

		\subfloat[]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_adult_lr_sex_0}}
		\subfloat[]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_adult_lr_sex_1}}
		\subfloat[]{\includegraphics[scale=0.25]{figures/dependency_exp_Learn-dependency_adult_lr_sex_DI}}\\

	\end{center}
	\caption{Results on computing feature influence functions (FIF) for COMPAS  and Adult dataset.}
\end{figure}

### 2022-01-26 tables/fairness/fvgm/fairness_metrics_on_compound_sensitive_groups.tex
adding two fairness papers

\begin{figure*}[!t]
	\begin{center}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Disparate_impact_Adult_LR_Learn-efficient-dependency}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Statistical_parity_Adult_LR_Learn-efficient-dependency}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute_eqo_Adult_LR_Learn-efficient-dependency}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__PCF_Adult_LR_Learn-efficient-dependency}}	
		\\

		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Disparate_impact_Compas_LR_Learn-efficient-dependency}}			
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Statistical_parity_Compas_LR_Learn-efficient-dependency}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute_eqo_Compas_LR_Learn-efficient-dependency}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__PCF_Compas_LR_Learn-efficient-dependency}}

	\end{center}
		\caption{Verifying compound sensitive groups with respect to multiple fairness metrics. In each plot, the $ X $-axis shows sensitive features with the number of compound groups (within parenthesis) and $ Y $-axis shows computed group and causal fairness metrics. }
		\label{fig:multiple_metrics}

\end{figure*}

### 2022-01-26 tables/fairness/fvgm/sanity_check.tex
adding two fairness papers

\begin{figure}
	\begin{center}
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_probability}\label{fig:synthetic_dist}}
		\subfloat[]{\includegraphics[scale=0.45]{figures/synthetic_example_BN}\label{fig:synthetic_bn}}

%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_LR}}
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_SVM}}
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_DT}}
%		
%		
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_LR}}
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_SVM}}
%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_DT}}
%		
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_LR}\label{fig:synthetic_ppv_max_LR}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_SVM}}\\
%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_DT}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_LR}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_SVM}\label{fig:synthetic_ppv_min_SVM}}
%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_DT}\label{fig:synthetic_ppv_b}}

		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_LR}\label{fig:synthetic_sample_size_LR}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_SVM}}\\
%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_DT}}	
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_LR}}
		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_SVM}\label{fig:synthetic_sample_size_SVM}}
%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_DT}\label{fig:synthetic_sample_size_b}}

	\end{center}

	\caption{
		Measuring accuracy of different fairness verifiers for Example~\ref{example:intro}. LR, SVM, DT refers to Logistic Regression classifier, Support Vector Machine, and Decision Tree, respectively.}
	\label{fig:synthetic_results}
\end{figure}

### 2022-01-27 chapters/interpretability/relaxed-cnf/abstract.tex
relaxed-cnf paper added

\section{abstract}
	 Machine learning algorithms that produce  rule-based predictions   in Conjunctive Normal form (CNF) or in Disjunctive Normal form (DNF) are arguably some of the most interpretable ones. For example, decision set is an interpretable model in practice, that represents the decision function in the form of DNF. In this paper, we consider relaxed definitions of standard OR/AND operators which allow exceptions in the construction of a clause and also in the selection of clauses in a rule. Building on these relaxed definition,  we introduce \textit{relaxed-}CNF rules, which are motivated by the popular usage of checklists in the medical domain and generalizes the widely employed rule representations including CNF, DNF, and decision sets.  While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the naive learning techniques are computationally expensive. 	 To this end, we  propose a novel incremental mini-batch learning procedure, called {\crr}, that employs  advances in the  Mixed-Integer Linear Programming (MILP) solvers to efficiently learn relaxed-CNF rules.  
	 Our  experimental analysis demonstrates that {\crr} can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.

### 2022-01-27 chapters/interpretability/relaxed-cnf/appendix.tex
relaxed-cnf paper added

\begin{table}
	\setlength{\tabcolsep}{3pt}
	%\scriptsize 
	\begin{center}
		\begin{tabular}{l  r  r r r r r r r rrr}
			\toprule
			{{Dataset}} & Size & Features  & RIPPER & BRS & IMLI & CRR\\
			\midrule
			{ Parkinsons}   & $  195 $  & $  51 $  & $   88.75  $    & $   89.74  $    & $   92.3  $    & $   89.47  $   \\ & & 
			& $   3 $    & $   9 $    & $   6 $    & $   4 $   \\ & & 
			& $   5.32 \text{s}  $    & $   16.3 \text{s}  $    & $   1.44 \text{s}  $    & $   3.54 \text{s}  $   \\[.1cm] 
			{ Heart}   & $  303 $  & $  31 $  & $   78.69  $    & $   72.13  $    & $   72.13  $    & $   77.69  $   \\ & & 
			& $   7 $    & $   19 $    & $   13 $    & $   4.5  $   \\ & & 
			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   1.8 \text{s}  $    & $   122.5 \text{s}  $   \\[.1cm] 
			{ Ionosphere}   & $  351 $  & $  144 $  & $   88.65  $    & $   91.43  $    & $   89.29  $    & $   91.43  $   \\ & & 
			& $   8 $    & $   4 $    & $   8.5  $    & $   20 $   \\ & & 
			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   2.09 \text{s}  $    & $   5.59 \text{s}  $   \\[.1cm] 
			{ WDBC}   & $  569 $  & $  88 $  & $   95.22  $    & $   95.65  $    & $   93.91  $    & $   94.69  $   \\ & & 
			& $   7.5  $    & $   12 $    & $   7 $    & $   34.5  $   \\ & & 
			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   1.38 \text{s}  $    & $   316.32 \text{s}  $   \\[.1cm] 
			{ Ilpd}   & $  583 $  & $  14 $  & $   71.55  $    & $   68.38  $    & $   71.55  $    & $   71.55  $   \\ & & 
			& $   1 $    & $   3 $    & $   0 $    & $   2 $   \\ & & 
			& $   5.07 \text{s}  $    & $   19.46 \text{s}  $    & $   1.51 \text{s}  $    & $   2.24 \text{s}  $   \\[.1cm] 
			{ Transfusion}   & $  748 $  & $  6 $  & $   76 $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   76 $    & $   76 $   \\ & & 
			& $   1 $    &       & $   0 $    & $   2 $   \\ & & 
			& $   5.1 \text{s}  $    &       & $   1.38 \text{s}  $    & $   1.99 \text{s}  $   \\[.1cm] 
			{ Pima}   & $  768 $  & $  30 $  & $   76.3  $    & $   73.38  $    & $   70.59  $    & $   74.51  $   \\ & & 
			& $   9.5  $    & $   3 $    & $   7 $    & $   6 $   \\ & & 
			& $   5.45 \text{s}  $    & $   24.58 \text{s}  $    & $   2.02 \text{s}  $    & $   11.06 \text{s}  $   \\[.1cm] 
			{ TicTacToe}   & $  958 $  & $  27 $  & $   96.88  $    & $   100 $    & $   84.33  $    & $   76.96  $   \\ & & 
			& $   25 $    & $   24 $    & $   6 $    & $   27 $   \\ & & 
			& $   5.66 \text{s}  $    & $   15.09 \text{s}  $    & $   194.36 \text{s}  $    & $   1002.04 \text{s}  $   \\[.1cm] 
			{ Titanic}   & $  1309 $  & $  26 $  & $   79.77  $    & $   78.54  $    & $   77.1  $    & $   77.48  $   \\ & & 
			& $   5 $    & $   2 $    & $   7 $    & $   13 $   \\ & & 
			& $   5.72 \text{s}  $    & $   15.33 \text{s}  $    & $   2.78 \text{s}  $    & $   3.72 \text{s}  $   \\[.1cm] 
			{ Banknote}   & $  1372 $  & $  16 $  & $   94.53  $    & $   81.09  $    & $   92.34  $    & $   94.89  $   \\ & & 
			& $   10 $    & $   2 $    & $   10 $    & $   9 $   \\ & & 
			& $   5.6 \text{s}  $    & $   11.02 \text{s}  $    & $   2.07 \text{s}  $    & $   3.31 \text{s}  $   \\[.1cm] 
			{ Compas}   & $  7210 $  & $  19 $  & $   67.65  $    & $   34.26  $    & $   65.19  $    & $   66.02  $   \\ & & 
			& $   19 $    & $   8 $    & $   3 $    & $   11 $   \\ & & 
			& $   7.38 \text{s}  $    & $   20.42 \text{s}  $    & $   4.16 \text{s}  $    & $   577.03 \text{s}  $   \\[.1cm] 
			{ Mushroom}   & $  8124 $  & $  115 $  & $   100 $    & $   99.91  $    & $   99.75  $    & $   100 $   \\ & & 
			& $   12 $    & $   20.5  $    & $   10 $    & $   17 $   \\ & & 
			& $   8.02 \text{s}  $    & $   35.82 \text{s}  $    & $   6.84 \text{s}  $    & $   720.13 \text{s}  $   \\[.1cm] 
			{ Magic}   & $  19020 $  & $  79 $  & $   84.04  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\ & & 
			& $   115 $    & $   3 $    & $   24 $    & $   31 $   \\ & & 
			& $   15.86 \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\[.1cm] 
			{ Tom's HW}   & $  28179 $  & $  910 $  & $   97.4  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\ & & 
			& $   36 $    &       & $   30 $    & $   4 $   \\ & & 
			& $   42.73 \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\[.1cm] 
			{ Credit}   & $  30000 $  & $  110 $  & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $   82.04  $   \\ & & 
			& $   38.5  $    &       & $   10 $    & $   32 $   \\ & & 
			& $   14.52 \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\[.1cm] 
			{ Adult}   & $  32561 $  & $  144 $  & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   84.86  $   \\ & & 
			& $   94 $    &       & $   23 $    & $   18 $   \\ & & 
			& $   27.61 \text{s}  $    &       & $   11.91 \text{s}  $    & $   1016.36 \text{s}  $   \\[.1cm] 
			{ Twitter}   & $  49999 $  & $  1511 $  & $   95.74  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\ & & 
			& $   179.5  $    &       & $   57 $    & $   12 $   \\ & & 
			& $   170.87 \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\[.1cm] 
			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   84.57  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\ & & 
			& $   195 $    &       & $   7 $    & $   2 $   \\ & & 
			& $   121.02 \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\[.1cm] 
			{ Skin}   & $  245057 $  & $  119 $  & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   98.92  $    & $   95.08  $   \\ & & 
			& $   725 $    &       & $   201 $    & $   29 $   \\ & & 
			& $   1313.19 \text{s}  $    &       & $   103.8 \text{s}  $    & $   825.6 \text{s}  $   \\[.1cm] 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Comparisons of test accuracy, validation accuracy and training time among different classifiers. Every cell in the last eight columns contain the test accuracy in percentage (top value), validation accuracy in percentage (middle value within square bracket), and  train time in seconds (bottom value within  parentheses).}
	\label{tab:result}
\end{table}

\begin{table}
	\begin{center}
		\begin{tabular}{l   r r r r r  r }
			\toprule
			{Dataset}  & LR & SVC & RF & $k$-NN & CRR\\\midrule
			\multirow{1}{*}{ Parkinsons}   & $   92.5  $    & $   94.74  $    & $   92.31  $    & $   92.5  $    & $   89.47  $   \\ 
			\multirow{1}{*}{ Heart}   & $   84.29  $    & $   83.33  $    & $   81.97  $    & $   78.69  $    & $   77.69  $   \\ 
			\multirow{1}{*}{ Ionosphere}   & $   94.29  $    & $   91.43  $    & $   92.96  $    & $   91.43  $    & $   91.43  $   \\ 
			\multirow{1}{*}{ WDBC}   & $   98.26  $    & $   96.46  $    & $   96.9  $    & $   95.61  $    & $   94.69  $   \\ 
			\multirow{1}{*}{ Ilpd}   & $   71.55  $    & $   71.55  $    & $   71.55  $    & $   69.65  $    & $   71.55  $   \\ 
			\multirow{1}{*}{ Transfusion}   & $   76.0  $    & $   76.0  $    & $   76.0  $    & $   76.51  $    & $   76.0  $   \\ 
			\multirow{1}{*}{ Pima}   & $   81.37  $    & $   75.82  $    & $   74.03  $    & $   78.43  $    & $   74.51  $   \\ 
			\multirow{1}{*}{ TicTacToe}   & $   98.95  $    & $   98.95  $    & $   98.69  $    & $   92.75  $    & $   76.96  $   \\ 
			\multirow{1}{*}{ Titanic}   & $   79.01  $    & $   77.48  $    & $   79.54  $    & $   77.1  $    & $   77.48  $   \\ 
			\multirow{1}{*}{ Banknote}   & $   93.8  $    & $   94.89  $    & $   94.89  $    & $   93.25  $    & $   94.89  $   \\ 
			\multirow{1}{*}{ Compas}   & $   68.03  $    & $   65.6  $    & $   67.82  $    & $   67.13  $    & $   66.02  $   \\ 
			\multirow{1}{*}{ Mushroom}   & $   100.0  $    & $   100.0  $    & $   100.0  $    & $   100.0  $    & $   100.0  $   \\ 
			\multirow{1}{*}{ Magic}   & $   85.15  $    & $   84.45  $    & $   85.3  $    & $   77.9  $    & $   81.31  $   \\ 
			\multirow{1}{*}{ Tom's HW}   & $   97.62  $    & $   97.66  $    & $   97.52  $    & $   94.59  $    & $   97.34  $   \\ 
			\multirow{1}{*}{ Credit}   & $   82.04  $    & $   82.12  $    & $   81.97  $    & $   80.5  $    & $   82.04  $   \\ 
			\multirow{1}{*}{ Adult}   & $   87.24  $    & $   86.82  $    & $   86.84  $    & $   84.68  $    & $   84.86  $   \\ 
			\multirow{1}{*}{ Twitter}   & $   96.28  $    & $   96.34  $    & $   96.37  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.16  $   \\ 
			\multirow{1}{*}{ Weather-AUS}   & $   85.71  $     &  \multicolumn{1}{c}{\textemdash}      & $   85.63  $     &  \multicolumn{1}{c}{\textemdash}      & $   83.34  $   \\ 
			\multirow{1}{*}{ Skin}   & $   97.21  $     &  \multicolumn{1}{c}{\textemdash}      & $   99.81  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.08  $   \\ 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Size of the rules generated by  interpretable classifiers.}
	\label{tab:rule_size}
\end{table}

\begin{table*}[h]		
	\setlength{\tabcolsep}{3pt}
	%\scriptsize 
	\begin{center}
		\begin{tabular}{l  r  r r r r r r r rrr}
			\toprule	
			{{Dataset}} & Size & Features  & LR & SVC & RF & $k$-NN & RIPPER & BRS & IMLI & CRR\\	
			\midrule
			{ Parkinsons}   & $  195 $  & $  51 $  & $   92.5  $    & $   94.74  $    & $   92.31  $    & $   92.5  $    & $   88.75  $    & $   89.74  $    & $   92.3  $    & $   89.47  $   \\ & & 
			& $  [ 93.75 ] $    & $  [ 93.75 ] $    & $  [ 93.75 ] $    & $  [ 93.33 ] $    & $  [ 88.71 ] $    & $  [ 93.75 ] $    & $  [ 90.62 ] $    & $  [ 93.75 ] $   \\ & & 
			& $  ( 1.23 \text{s} ) $    & $  ( 1.01 \text{s} ) $    & $  ( 2.91 \text{s} ) $    & $  ( 1.03 \text{s} ) $    & $  ( 5.32 \text{s} ) $    & $  ( 16.3 \text{s} ) $    & $  ( 1.44 \text{s} ) $    & $  ( 3.54 \text{s} ) $   \\[.1cm] 
			{ Heart}   & $  303 $  & $  31 $  & $   84.29  $    & $   83.33  $    & $   81.97  $    & $   78.69  $    & $   78.69  $    & $   72.13  $    & $   72.13  $    & $   77.69  $   \\ & & 
			& $  [ 85.71 ] $    & $  [ 85.42 ] $    & $  [ 85.42 ] $    & $  [ 83.67 ] $    & $  [ 81.25 ] $    & $  [ 79.17 ] $    & $  [ 82] $    & $  [ 85.42 ] $   \\ & & 
			& $  ( 1.04 \text{s} ) $    & $  ( 0.98 \text{s} ) $    & $  ( 3.02 \text{s} ) $    & $  ( 1.02 \text{s} ) $    & $  ( 5.27 \text{s} ) $    & $  ( 25.07 \text{s} ) $    & $  ( 1.8 \text{s} ) $    & $  ( 122.5 \text{s} ) $   \\[.1cm] 
			{ Ionosphere}   & $  351 $  & $  144 $  & $   94.29  $    & $   91.43  $    & $   92.96  $    & $   91.43  $    & $   88.65  $    & $   91.43  $    & $   89.29  $    & $   91.43  $   \\ & & 
			& $  [ 94.64 ] $    & $  [ 94.64 ] $    & $  [ 92.92 ] $    & $  [ 91.07 ] $    & $  [ 91.23 ] $    & $  [ 92.86 ] $    & $  [ 89.29 ] $    & $  [ 94.64 ] $   \\ & & 
			& $  ( 2.21 \text{s} ) $    & $  ( 1.01 \text{s} ) $    & $  ( 2.93 \text{s} ) $    & $  ( 1.03 \text{s} ) $    & $  ( 5.87 \text{s} ) $    & $  ( 75.53 \text{s} ) $    & $  ( 2.09 \text{s} ) $    & $  ( 5.59 \text{s} ) $   \\[.1cm] 
			{ WDBC}   & $  569 $  & $  88 $  & $   98.26  $    & $   96.46  $    & $   96.9  $    & $   95.61  $    & $   95.22  $    & $   95.65  $    & $   93.91  $    & $   94.69  $   \\ & & 
			& $  [ 96.7 ] $    & $  [ 97.8 ] $    & $  [ 96.7 ] $    & $  [ 96.7 ] $    & $  [ 95.6 ] $    & $  [ 96.74 ] $    & $  [ 94.51 ] $    & $  [ 97.25 ] $   \\ & & 
			& $  ( 2.18 \text{s} ) $    & $  ( 0.99 \text{s} ) $    & $  ( 3.08 \text{s} ) $    & $  ( 1.03 \text{s} ) $    & $  ( 5.7 \text{s} ) $    & $  ( 630.23 \text{s} ) $    & $  ( 1.38 \text{s} ) $    & $  ( 316.32 \text{s} ) $   \\[.1cm] 
			{ Ilpd}   & $  583 $  & $  14 $  & $   71.55  $    & $   71.55  $    & $   71.55  $    & $   69.65  $    & $   71.55  $    & $   68.38  $    & $   71.55  $    & $   71.55  $   \\ & & 
			& $  [ 71.28 ] $    & $  [ 71.28 ] $    & $  [ 71.28 ] $    & $  [ 71.28 ] $    & $  [ 71.28 ] $    & $  [ 70.21 ] $    & $  [ 71.28 ] $    & $  [ 71.28 ] $   \\ & & 
			& $  ( 1.01 \text{s} ) $    & $  ( 1.02 \text{s} ) $    & $  ( 2.86 \text{s} ) $    & $  ( 1.03 \text{s} ) $    & $  ( 5.07 \text{s} ) $    & $  ( 19.46 \text{s} ) $    & $  ( 1.51 \text{s} ) $    & $  ( 2.24 \text{s} ) $   \\[.1cm] 
			{ Transfusion}   & $  748 $  & $  6 $  & $   76 $    & $   76 $    & $   76 $    & $   76.51  $    & $   76 $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   76 $    & $   76 $   \\ & & 
			& $  [ 76.47 ] $    & $  [ 76.47 ] $    & $  [ 76.47 ] $    & $  [ 76.47 ] $    & $  [ 76.47 ] $    &       & $  [ 76.47 ] $    & $  [ 76.47 ] $   \\ & & 
			& $  ( 1.03 \text{s} ) $    & $  ( 1\text{s} ) $    & $  ( 2.91 \text{s} ) $    & $  ( 1\text{s} ) $    & $  ( 5.1 \text{s} ) $    &       & $  ( 1.38 \text{s} ) $    & $  ( 1.99 \text{s} ) $   \\[.1cm] 
			{ Pima}   & $  768 $  & $  30 $  & $   81.37  $    & $   75.82  $    & $   74.03  $    & $   78.43  $    & $   76.3  $    & $   73.38  $    & $   70.59  $    & $   74.51  $   \\ & & 
			& $  [ 80.45 ] $    & $  [ 78.46 ] $    & $  [ 78.86 ] $    & $  [ 79.27 ] $    & $  [ 80.08 ] $    & $  [ 79.67 ] $    & $  [ 76.23 ] $    & $  [ 78.86 ] $   \\ & & 
			& $  ( 1.3 \text{s} ) $    & $  ( 0.94 \text{s} ) $    & $  ( 3.37 \text{s} ) $    & $  ( 0.98 \text{s} ) $    & $  ( 5.45 \text{s} ) $    & $  ( 24.58 \text{s} ) $    & $  ( 2.02 \text{s} ) $    & $  ( 11.06 \text{s} ) $   \\[.1cm] 
			{ TicTacToe}   & $  958 $  & $  27 $  & $   98.95  $    & $   98.95  $    & $   98.69  $    & $   92.75  $    & $   96.88  $    & $   100 $    & $   84.33  $    & $   76.96  $   \\ & & 
			& $  [ 98.69 ] $    & $  [ 98.69 ] $    & $  [ 97.39 ] $    & $  [ 92.81 ] $    & $  [ 98.69 ] $    & $  [ 100] $    & $  [ 83.34 ] $    & $  [ 84.97 ] $   \\ & & 
			& $  ( 1.52 \text{s} ) $    & $  ( 1.06 \text{s} ) $    & $  ( 3.82 \text{s} ) $    & $  ( 1.01 \text{s} ) $    & $  ( 5.66 \text{s} ) $    & $  ( 15.09 \text{s} ) $    & $  ( 194.36 \text{s} ) $    & $  ( 1002.04 \text{s} ) $   \\[.1cm] 
			{ Titanic}   & $  1309 $  & $  26 $  & $   79.01  $    & $   77.48  $    & $   79.54  $    & $   77.1  $    & $   79.77  $    & $   78.54  $    & $   77.1  $    & $   77.48  $   \\ & & 
			& $  [ 78.1 ] $    & $  [ 78.47 ] $    & $  [ 79] $    & $  [ 78.1 ] $    & $  [ 79.05 ] $    & $  [ 78.95 ] $    & $  [ 79.19 ] $    & $  [ 79.52 ] $   \\ & & 
			& $  ( 1.84 \text{s} ) $    & $  ( 1.06 \text{s} ) $    & $  ( 3.78 \text{s} ) $    & $  ( 1.04 \text{s} ) $    & $  ( 5.72 \text{s} ) $    & $  ( 15.33 \text{s} ) $    & $  ( 2.78 \text{s} ) $    & $  ( 3.72 \text{s} ) $   \\[.1cm] 
			{ Banknote}   & $  1372 $  & $  16 $  & $   93.8  $    & $   94.89  $    & $   94.89  $    & $   93.25  $    & $   94.53  $    & $   81.09  $    & $   92.34  $    & $   94.89  $   \\ & & 
			& $  [ 94.52 ] $    & $  [ 94.52 ] $    & $  [ 94.09 ] $    & $  [ 92.92 ] $    & $  [ 94.09 ] $    & $  [ 81.74 ] $    & $  [ 94.52 ] $    & $  [ 94.52 ] $   \\ & & 
			& $  ( 1.36 \text{s} ) $    & $  ( 1.01 \text{s} ) $    & $  ( 3.66 \text{s} ) $    & $  ( 1.03 \text{s} ) $    & $  ( 5.6 \text{s} ) $    & $  ( 11.02 \text{s} ) $    & $  ( 2.07 \text{s} ) $    & $  ( 3.31 \text{s} ) $   \\[.1cm] 
			{ Compas}   & $  7210 $  & $  19 $  & $   68.03  $    & $   65.6  $    & $   67.82  $    & $   67.13  $    & $   67.65  $    & $   34.26  $    & $   65.19  $    & $   66.02  $   \\ & & 
			& $  [ 67.76 ] $    & $  [ 66.09 ] $    & $  [ 66.58 ] $    & $  [ 67.16 ] $    & $  [ 67.1 ] $    & $  [ 64.61 ] $    & $  [ 66] $    & $  [ 67.65 ] $   \\ & & 
			& $  ( 3.6 \text{s} ) $    & $  ( 4.06 \text{s} ) $    & $  ( 12.76 \text{s} ) $    & $  ( 3.28 \text{s} ) $    & $  ( 7.38 \text{s} ) $    & $  ( 20.42 \text{s} ) $    & $  ( 4.16 \text{s} ) $    & $  ( 577.03 \text{s} ) $   \\[.1cm] 
			{ Mushroom}   & $  8124 $  & $  115 $  & $   100 $    & $   100 $    & $   100 $    & $   100 $    & $   100 $    & $   99.91  $    & $   99.75  $    & $   100 $   \\ & & 
			& $  [ 100] $    & $  [ 100] $    & $  [ 100] $    & $  [ 100] $    & $  [ 100] $    & $  [ 100] $    & $  [ 99.77 ] $    & $  [ 100] $   \\ & & 
			& $  ( 3.33 \text{s} ) $    & $  ( 4.33 \text{s} ) $    & $  ( 13.15 \text{s} ) $    & $  ( 9.92 \text{s} ) $    & $  ( 8.02 \text{s} ) $    & $  ( 35.82 \text{s} ) $    & $  ( 6.84 \text{s} ) $    & $  ( 720.13 \text{s} ) $   \\[.1cm] 
			{ Magic}   & $  19020 $  & $  79 $  & $   85.15  $    & $   84.45  $    & $   85.3  $    & $   77.9  $    & $   84.04  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\ & & 
			& $  [ 84.78 ] $    & $  [ 84.49 ] $    & $  [ 85.56 ] $    & $  [ 77.46 ] $    & $  [ 84.39 ] $    & $  [ 76.11 ] $    & $  [ 77.42 ] $    & $  [ 81.86 ] $   \\ & & 
			& $  ( 4.06 \text{s} ) $    & $  ( 58.06 \text{s} ) $    & $  ( 24.57 \text{s} ) $    & $  ( 49.08 \text{s} ) $    & $  ( 15.86 \text{s} ) $    & $  ( 56.46 \text{s} ) $    & $  ( 141.2 \text{s} ) $    & $  ( 1012.6 \text{s} ) $   \\[.1cm] 
			{ Tom's HW}   & $  28179 $  & $  910 $  & $   97.62  $    & $   97.66  $    & $   97.52  $    & $   94.59  $    & $   97.4  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\ & & 
			& $  [ 97.54 ] $    & $  [ 97.47 ] $    & $  [ 97.47 ] $    & $  [ 94.52 ] $    & $  [ 97.52 ] $    &       & $  [ 96.32 ] $    & $  [ 97.32 ] $   \\ & & 
			& $  ( 5.37 \text{s} ) $    & $  ( 231.87 \text{s} ) $    & $  ( 66.53 \text{s} ) $    & $  ( 1366.39 \text{s} ) $    & $  ( 42.73 \text{s} ) $    &       & $  ( 92.65 \text{s} ) $    & $  ( 1071.58 \text{s} ) $   \\[.1cm] 
			{ Credit}   & $  30000 $  & $  110 $  & $   82.04  $    & $   82.12  $    & $   81.97  $    & $   80.5  $    & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $   82.04  $   \\ & & 
			& $  [ 82.17 ] $    & $  [ 81.94 ] $    & $  [ 81.99 ] $    & $  [ 80.52 ] $    & $  [ 82] $    &       & $  [ 81.91 ] $    & $  [ 81.98 ] $   \\ & & 
			& $  ( 4.78 \text{s} ) $    & $  ( 170.76 \text{s} ) $    & $  ( 41.19 \text{s} ) $    & $  ( 258.2 \text{s} ) $    & $  ( 14.52 \text{s} ) $    &       & $  ( 17.66 \text{s} ) $    & $  ( 1021.35 \text{s} ) $   \\[.1cm] 
			{ Adult}   & $  32561 $  & $  144 $  & $   87.24  $    & $   86.82  $    & $   86.84  $    & $   84.68  $    & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   84.86  $   \\ & & 
			& $  [ 87.38 ] $    & $  [ 87.18 ] $    & $  [ 86.84 ] $    & $  [ 84.53 ] $    & $  [ 84.35 ] $    &       & $  [ 82.65 ] $    & $  [ 85.35 ] $   \\ & & 
			& $  ( 6.46 \text{s} ) $    & $  ( 326.45 \text{s} ) $    & $  ( 47.44 \text{s} ) $    & $  ( 391.66 \text{s} ) $    & $  ( 27.61 \text{s} ) $    &       & $  ( 11.91 \text{s} ) $    & $  ( 1016.36 \text{s} ) $   \\[.1cm] 
			{ Twitter}   & $  49999 $  & $  1511 $  & $   96.28  $    & $   96.34  $    & $   96.37  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.74  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\ & & 
			& $  [ 96.32 ] $    & $  [ 96.36 ] $    & $  [ 96.41 ] $    &       & $  [ 96.23 ] $    &       & $  [ 95] $    & $  [ 95.68 ] $   \\ & & 
			& $  ( 8.28 \text{s} ) $    & $  ( 1419.34 \text{s} ) $    & $  ( 230.56 \text{s} ) $    &       & $  ( 170.87 \text{s} ) $    &       & $  ( 238.29 \text{s} ) $    & $  ( 1144.66 \text{s} ) $   \\[.1cm] 
			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   85.71  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   85.63  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   84.57  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\ & & 
			& $  [ 85.71 ] $    &       & $  [ 85.61 ] $    &       & $  [ 84.78 ] $    &       & $  [ 82.75 ] $    & $  [ 83.52 ] $   \\ & & 
			& $  ( 11.21 \text{s} ) $    &       & $  ( 174.59 \text{s} ) $    &       & $  ( 121.02 \text{s} ) $    &       & $  ( 366.12 \text{s} ) $    & $  ( 1115.27 \text{s} ) $   \\[.1cm] 
			{ Skin}   & $  245057 $  & $  119 $  & $   97.21  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   99.81  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   98.92  $    & $   95.08  $   \\ & & 
			& $  [ 97.23 ] $    &       & $  [ 99.8 ] $    &       & $  [ 98.37 ] $    &       & $  [ 98.88 ] $    & $  [ 95.25 ] $   \\ & & 
			& $  ( 8.36 \text{s} ) $    &       & $  ( 460.25 \text{s} ) $    &       & $  ( 1313.19 \text{s} ) $    &       & $  ( 103.8 \text{s} ) $    & $  ( 825.6 \text{s} ) $   \\[.1cm] 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Comparisons of test accuracy, validation accuracy and training time among different classifiers. Every cell in the last eight columns contain the test accuracy in percentage (top value), validation accuracy in percentage (middle value within square bracket), and  train time in seconds (bottom value within  parentheses).}
	\label{tab:result}
\end{table*}

\begin{table*}[h]		
	\begin{center}
		\begin{tabular}{l   r r r r r  }
			\toprule
			{Dataset}  & RIPPER & BRS & IMLI & CRR\\\midrule
			\multirow{1}{*}{ Parkinsons}   & $   3 $    & $   9 $    & $   6 $    & $   4 $   \\ 
			\multirow{1}{*}{ Heart}   & $   7 $    & $   19 $    & $   13 $    & $   4.5  $   \\ 
			\multirow{1}{*}{ Ionosphere}   & $   8 $    & $   4 $    & $   8.5  $    & $   20 $   \\ 
			\multirow{1}{*}{ WDBC}   & $   7.5  $    & $   12 $    & $   7 $    & $   34.5  $   \\ 
			\multirow{1}{*}{ Ilpd}   & $   1 $    & $   3 $    & $   0 $    & $   2 $   \\ 
			\multirow{1}{*}{ Transfusion}   & $   1 $     &  \multicolumn{1}{c}{\textemdash}      & $   0 $    & $   2 $   \\ 
			\multirow{1}{*}{ Pima}   & $   9.5  $    & $   3 $    & $   7 $    & $   6 $   \\ 
			\multirow{1}{*}{ TicTacToe}   & $   25 $    & $   24 $    & $   6 $    & $   27 $   \\ 
			\multirow{1}{*}{ Titanic}   & $   5 $    & $   2 $    & $   7 $    & $   13 $   \\ 
			\multirow{1}{*}{ Banknote}   & $   10 $    & $   2 $    & $   10 $    & $   9 $   \\ 
			\multirow{1}{*}{ Compas}   & $   19 $    & $   8 $    & $   3 $    & $   11 $   \\ 
			\multirow{1}{*}{ Mushroom}   & $   12 $    & $   20.5  $    & $   10 $    & $   17 $   \\ 
			\multirow{1}{*}{ Magic}   & $   115 $    & $   3 $    & $   24 $    & $   31 $   \\ 
			\multirow{1}{*}{ Tom's HW}   & $   36 $     &  \multicolumn{1}{c}{\textemdash}      & $   30 $    & $   4 $   \\ 
			\multirow{1}{*}{ Credit}   & $   38.5  $     &  \multicolumn{1}{c}{\textemdash}      & $   10 $    & $   32 $   \\ 
			\multirow{1}{*}{ Adult}   & $   94 $     &  \multicolumn{1}{c}{\textemdash}      & $   23 $    & $   18 $   \\ 
			\multirow{1}{*}{ Twitter}   & $   179.5  $     &  \multicolumn{1}{c}{\textemdash}      & $   57 $    & $   12 $   \\ 
			\multirow{1}{*}{ Weather-AUS}   & $   195 $     &  \multicolumn{1}{c}{\textemdash}      & $   7 $    & $   2 $   \\ 
			\multirow{1}{*}{ Skin}   & $   725 $     &  \multicolumn{1}{c}{\textemdash}      & $   201 $    & $   29 $   \\ 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Size of the rules generated by  interpretable classifiers.}
	\label{tab:rule_size}	
\end{table*}

\begin{figure*}[t]
\includegraphics[width=0.18\linewidth]{figures/adult_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_validation_accuracy_vary_lambda.pdf}

\end{figure*}
\begin{figure*}

\includegraphics[width=0.18\linewidth]{figures/skin_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_validation_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_rule_size_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_test_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_time_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_train_accuracy_vary_lambda.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_validation_accuracy_vary_lambda.pdf}
\end{figure*}

\begin{figure*}[t]
\includegraphics[width=0.18\linewidth]{figures/adult_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/adult_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/banknote_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/compas_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/credit_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/heart_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ilpd_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/ionosphere_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/magic_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/mushroom_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/parkinsons_validation_accuracy_vary_clause.pdf}

\end{figure*}

\begin{figure*}
\includegraphics[width=0.18\linewidth]{figures/skin_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/skin_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/tictactoe_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/titanic_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/transfusion_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/wdbc_validation_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_rule_size_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_test_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_time_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_train_accuracy_vary_clause.pdf}
\includegraphics[width=0.18\linewidth]{figures/weatherAUS_validation_accuracy_vary_clause.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.18\linewidth]{figures/adult_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_validation_accuracy_vary_iteration.pdf}

	\end{figure*}

	\begin{figure*}

	\includegraphics[width=0.18\linewidth]{figures/skin_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_validation_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_rule_size_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_test_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_time_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_train_accuracy_vary_iteration.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_validation_accuracy_vary_iteration.pdf}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.18\linewidth]{figures/adult_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/adult_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/banknote_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/credit_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/heart_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ilpd_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/ionosphere_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/magic_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/mushroom_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/parkinsons_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/pima_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/pima_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/pima_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/pima_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/pima_validation_accuracy_vary_subsamplesize.pdf}

	\end{figure*}

	\begin{figure*}
	\includegraphics[width=0.18\linewidth]{figures/skin_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/skin_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/tictactoe_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/titanic_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/transfusion_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/wdbc_validation_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_rule_size_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_test_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_time_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_train_accuracy_vary_subsamplesize.pdf}
	\includegraphics[width=0.18\linewidth]{figures/weatherAUS_validation_accuracy_vary_subsamplesize.pdf}
\end{figure*}

### 2022-01-27 chapters/interpretability/relaxed-cnf/conclusion.tex
relaxed-cnf paper added

\section{Conclusion}
\label{interpretability_crr_sec:conclusion}

In this paper, we proposed an efficient combinatorial framework called {\crr}, for learning relaxed-CNF  classification rules, that are  more expressive  than CNF/DNF rules.  {\crr}  uses a novel integration of mini-batch learning procedure with the MILP framework to learn sparse relaxed-CNF rules. Our experimental results demonstrate that {\crr} is able to learn relaxed-CNF  rules with higher accuracy and more concise representation than CNF rules. Moreover, the generated rules are sparser than decision lists in large datasets.

### 2022-01-27 chapters/interpretability/relaxed-cnf/detail.tex
relaxed-cnf paper added

%\cite{emadlearning}

\section{Description}
The current version of MaxSat based prediction rule facilitates both \textbf{and} and \textbf{or} type prediction rule over the  input features. \textbf{And} (\textbf{or}) type prediction rule is a similar representation of DNF (CNF) clauses. We consider that each clause has $ n $ literals and there are $ c $ such clauses in the prediction rule. \textbf{And} (\textbf{or}) type rule requires that  all  of $ n $ (at least $ 1 $ of $ n $) literals in  at least $ 1 $ of  $ c $ (all of $ c $)  clauses have to be satisfied for the prediction rule to predict \textit{true}. For simplification of understanding we assume binary class data (class label is  \textit{true} or \textit{false}) in this context. A significant benefit of SAT based prediction rule is that we can easily convert an \textbf{and} type rule to \textbf{or} type rule in the model designing phase and underlying SAT solver  works independently.

	A relaxed but more domain specific extension of current MaxSAT based \textbf{or} type prediction rule is designing $ m $ of $ n $ rule where satisfying at least $ m $  out of $ n $ literals  in each of $ c $ clauses predicts \textit{true}. This type of rule has various real life applications such as clinical prediction and checklist etc. In the following part we will formally represent our problem setting and solution method.

		%We can also design similar extension of \textbf{and} type rule where at most $ m $ out of $ n $ literals of any $ 1 $ of $ c $ clauses need to be satisfied for predicting \textit{true}.

	\paragraph{Problem Formulation:} Throughout the paper we consider $true =1 $ and $ false=0 $ interchangeably. Let us assume tuple $ (X_i,y_i)$ is the  $ i^{th} $ sample where $ X_i=[x_1^i,x_2^i,....,x_n^i] $ is a feature vector with  $ n $ features  $( |X_i|=n )$ and $ y_i \in \{0,1\}$ is corresponding class label. We specify there are $ s $ samples i.e., $ 1 \le i\le s $.  For simplicity we assume that $ X_i $ is a binary vector i.e., $\forall j,  x_j^i \in \{0,1\}  $ and $ 1\le j \le n $.  Any type of feature like numerical, real valued or categorical can be converted to binary feature through discretization.

	We take $ m $ as input for this  problem setting. The intuition behind selecting $ m $
	is application specific. For example, suppose a doctor wants to keep record of possible breast cancer patients. He has come up with $ 5 $ different tests and knows that if any patient has been diagnosed as infected in at least $ 3 $ tests, he is predicted as \textit{malignant} cancer otherwise  \textit{benign}. In  this case $ m=3 $.

	We assume the prediction rule has $ c $ clauses and each clause has $ n $ literals. We call each literal as a column variable $ b $ as it  corresponds to a binary column in the feature matrix. For $ c $ clause prediction rule we consider total $ c $ column variables against a single feature column. $ \{b_1^1, b_2^1,...., b_n^1,b_1^2,....,b_n^2,....,b_n^c \} $ is the set of column variables.  Our prediction model returns an assignment (\textit{true} or \textit{false}) of each variable $ b_j^k $ where $ 1\le j \le n $ and $ 1 \le k \le c $. A \textit{true} assignment of $ b_j^k $ signifies that $ j^{th} $ feature is an essential factor of our prediction rule. Whereas a  \textit{false}  assignment of $ b_j^k $ says that $ j^{th} $ feature does not take part in prediction.

	Our goal is to learn the assignment of each column variable from given sample space and each sample contributes in learning this assignment. For ease of representation we define a  vector $ B_k=[b_1^k, b_2^k,...., b_n^k] $  consisting of all column variables of $ k^{th} $ clause $ (1 \le k \le c) $. We define the operator $ \boldsymbol{\vee} $ over two binary vector in a special way for this problem setting. $ X_i \boldsymbol{\vee} B_k= (x_1^i \wedge b_1^k) + (x_2^i \wedge b_2^k) + .... +(x_n^i \wedge b_n^k)$. Here $ \wedge $ is logical \textit{and} operator i.e., $ 0 \wedge b_j^k=0 $ and $  1 \wedge b_j^k=b_j^k $.  Intuitively we can see that for $ i^{th} $ sample if $ j^{th} $ feature is $ 1 $, it ignites all the column variables $   b_j^k  (1\le k \le c) $ associated with this feature so that $ b_j^k $ can possibly be assigned \textit{true} after model returns the final assignment. However after we get final assignment of each column variable, $ X_i \boldsymbol{\vee} B_k $ returns a numerical value between $ 0 $ and $ n $ and we  predict $ i^{th} $ sample as \textit{true} if $ \forall k, (X_i \boldsymbol{\vee} B_k) \ge m $ and \textit{false} otherwise.

	Now we will formally define our prediction function $ \hat{y_i} $  for $ i^{th} $ sample as follows.
	\[
	\hat{y}=
	\begin{cases}
		\text{ 1 } & \quad\text{if }\forall k, ( X_i \vee B_k ) \ge m  \text{ where }  1\le k \le c\\
		\text{ 0 } &\quad\text{otherwise}\\
	\end{cases}		
	\]

	SAT based framework can keep track of accuracy and other performance measures simultaneously in the learning phase. Hence we introduce a noise variable $ \eta_i $ corresponding to $ i^{th} $ sample. A $ true $ assignment of $ \eta_i $ signifies that $ i^{th} $ sample is misclassified by our generated prediction rule.  So we will falsify as many noise variables as possible. Moreover an interpretable machine learning model encourages a shorter prediction rule. So we target at falsifying as many column variable $ b_j^k $ as possible while ensuring higher prediction accuracy. Here we see an optimization problem regarding boolean variables and constraints.

%	MaxSAT facilitates optimizing weighted clauses in CNF forms.

	We aim to represent our model according to MaxSAT setting. MaxSAT is an optimization of satisfiability problem where maximum number of clauses are aimed to be satisfied. MaxSAT is more difficult than SAT problem. However this optimization version is very useful and acts as a bridge between a satisfiability problem and an optimization problem. Weighted MaxSAT considers weight on each clause and it maximizes (minimizes) the sum of weight of satisfiable  (unsatisfiable) clauses. Two types of clauses are considered in MaxSAT setting- hard clause and soft clause. The hard clauses are  considered to be satisfied first and then the soft clauses are considered. This two different types of weight facilitates  putting priority over clauses. However each clause in CNF form is a disjunction of literals. But our problem setting ($ m $ of $ n $ rule) is not a direct form of disjunction of variables or literals rather a   pseudo-boolean  constraints. We will discuss later about how pseudo-boolean constraints can be converted to CNF form efficiently using existing encoder.
	%Our final prediction rule will be following.
	%\[		
	%\]

	\paragraph{Constraints}
	We define the following constraints.

	\begin{equation}
	V_j^k:=\neg b_{j}^k\qquad W(V_j^k)=1
	\label{eq:sparse_constraint}
	\end{equation}
	\begin{equation}
	S_i:=\neg \eta_i\qquad W(S_i)=\lambda
	\label{eq:noise_reduce}
	\end{equation}
	\begin{equation}
	D_i:=\neg \eta_i \rightarrow (y_i \leftrightarrow \hat{y}_i) \qquad W(D_i)=\infty
	\label{eq:classifier_prediction}
	\end{equation}

%	Here ~\ref{eq:classifier_prediction}  is a hard constraint and  ~\ref{eq:sparse_constraint},~\ref{eq:noise_reduce} are soft constraints. 
	Finally the set of constraints $ Q^c $ set by our approach is defined as follows.
	\[
	Q^c:=\bigwedge_{j=1,k=1}^{j=n,k=c} V_j^k \bigwedge_{i=1}^{i=s} S_i \bigwedge_{i=1}^{i=s} D_i
	\]

	\paragraph{Description of constraints:} 
	Here $ W(.) $ is a weight function over clauses. We put a weight of $ 1 $ to the column variables for gaining a sparse representation of prediction rule. Moreover we put $ \lambda $ as weight of noise variables where $ \lambda $ is regularization factor. Higher value of $ \lambda $ ensures less number of prediction error and vice versa. $ D_i $ is considered as hard clause in this setting. $ D_i $ can be interpreted in this way. If  $ i^{th} $ sample is not treated as noise, 
	prediction function $ \hat{y_i} $ should match sample class label $ y_i $.

	\paragraph{CNF Encoding of constraints:} **

### 2022-01-27 chapters/interpretability/relaxed-cnf/example.tex
relaxed-cnf paper added

\section{Illustrative Example}
\label{sec:example}

In this section, we provide a toy example to illustrate  a relaxed-CNF rule which we learn for  Breast Cancer Wisconsin Diagnostic Data Set (WDBC\footnote{\url{https://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+(Diagnostic)}}). The dataset is used to predict whether the cancer is benign or malignant based on the characteristics of an image of  a tumor cell.  {\Framework} generates the following rule for detecting malignant cancer. 
%\vspace{-10pt}

%\begin{comment}
\[
\begin{split}
&[ (  \text{ smoothness }\ge 0.089  + \text{ standard error of area  }\ge 53.78 \\
& + \text{ largest radius }\ge 18.225   )\ge 2  ] +\\
&[   (98.76\le \text{ perimeter } < 114.8  +  \text{ largest smoothness }\ge 0.136 \\
& + 105.95 \le \text{ largest perimeter }  < 117.45     )\ge 2  ]  \ge 1 \end{split}
\]
%\end{comment}

The generated rule has two clauses, each clause consisting of three characteristics of the image of a tumor cell. In this rule, a clause is satisfied if two out of three characteristics match for a test sample. Finally, the rule predicts ``malignant'' cancer if at least one clause is satisfied. The size of this rule is $ 3+3=6 $.  If each clause in this rule is encoded into CNF,  standard encoding technique~\cite{steinke2014pblib} would generate more than one CNF clauses, resulting in higher rule-size. Therefore,   relaxed-CNF rules possesse higher expressibility than  CNF rules. 
We believe that this type of rule is very intuitive and helpful to human experts in decision making.

\begin{comment}
{\scriptsize
	\begin{table}[]
		\setlength{\tabcolsep}{5pt}

		%		%\vspace{-6 pt}
		\begin{center}
			\begin{tabular}{|l | l| }
				%%				\newcommand{\softand}{\ensuremath{\mathsf{soft-AND}}}
				\hline

				\multicolumn{2}{|c|}{ $  \Rule = \ensuremath{\mathsf{softAND}}(\{\text{Test } 1 , \text{Test 
					} 2 \}, 1 $) }  \\
				\hline
				\multicolumn{2}{c}{ }  \\
				\hline
				\multicolumn{1}{|c|}{ Test $ 1 =  \ensuremath{\mathsf{hardOR}} (\cdot ,2 )$} & check \\
				\hline
				$\text{perimeter }> 700.66  $ & \\
				\hline
				$\text{standard error of perimeter }> 91.31 $ & \\
				\hline
				$\text{largest radius }> 23.58   $& \\
				\hline
				$\text{largest texture }> 115.9   $& \\
				\hline
				$\text{largest  concave portions of contour }> 0.15    $& \\
				\hline

				\multicolumn{2}{c}{\multirow{1}{*}{ }} \\
				%				\multicolumn{2}{c}{}\\
				\hline
				\multicolumn{1}{|c|}{ Test $ 2 =  \ensuremath{\mathsf{hardOR}} (\cdot ,2 )$}&  check \\
				\hline
				$ \text{area }> 0.1 $ & \\
				\hline 
				$ \text{largest radius }> 33.65   $ & \\
				\hline
				$ \text{largest perimeter }> 781.18  $&  \\
				\hline

			\end{tabular}
		\end{center}
		%\vspace{-10 pt}
		\caption{Example of relaxed-CNF rules for WDBC dataset.  \ensuremath{\mathsf{hardOR}} takes  a   clause and  $ \eta_l $ as input   and returns $ 1 $  when at least $ \eta_l $ literals are  $ 1 $. \ensuremath{\mathsf{softAND}} takes   a formula and  $ \eta_c $ as input and returns $ 1 $  when at least $ \eta_c $ clauses are  $ 1 $.  Informally,   a test is diagnosed as positive if at least two properties of breast tissue are matched. The patient has malignant cancer if she/he is diagnosed positive in at least one test, otherwise she/he has benign cancer.}
		\label{tab:illustration}
%		\vspace{-40 pt}
	\end{table}
}

\end{comment}

### 2022-01-27 chapters/interpretability/relaxed-cnf/experiment.tex
relaxed-cnf paper added

\section{Experiments}\label{interpretability_crr_sec:experiment}
We implement a prototype  of  the incremental version of $ \crr $\footnote{The source code is available at \url{https://github.com/meelgroup/mlic}} based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances. The objective of our experimental evaluation is to answer the following questions:

\begin{enumerate}
	\item How do the accuracy and training time for  {\crr}  behave vis-a-vis state-of-the-art classifiers on large datasets arising in machine learning problems in practice?
	\item Can {\crr} generate sparse rules compared to that of  other rule-based models? 
	\item How do the training time, accuracy, and rule size vary with model hyper-parameters?
\end{enumerate}

In summary, relaxed-CNF rules generated by {\crr} achieves higher accuracy and more concise representation than CNF rules in most of the datasets. Moreover, relaxed-CNF rules are shown to be sparser than decision lists with competitive accuracy in large datasets.  Finally, we show how to control the trade-off between rule-sparsity and accuracy using the hyper-parameter $ \lambda $; and between accuracy and training time using the hyper-parameter $ k $ and size $ n_p $ of each mini-batch. In the following, we give a detailed description of the experiments.

%\vspace{-3pt}
\subsection{Experiment Methodology}
We perform experiments on a high-performance computer cluster, where each node consists of E$ 5$-$2690\text{ v}3 $ CPU with $ 24 $ cores, $ 96 $ GB of RAM. Each experiment is run on four cores of a node with $ 16 $ GB memory.   We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g. $ \IMLI  $~\cite{ghosh19incremental}, RIPPER~\cite{C1995}, BRS~\cite{WRDLKM2017},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR). Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers. In particular,
{\IMLI} generates classification rules in CNF using a MaxSAT-based formulation and we use Open-WBO \cite{MML2014} as the MaxSAT solver for {\IMLI}. We compare with propositional rule learning algorithm RIPPER, which is implemented in WEKA~\cite{HFHPRW2009} and generates classification rules in the form of decision lists. BRS is a Bayesian framework for generating rule sets expressed as DNF. For  other classifiers, we use the Scikit-learn module of Python~\cite{PVGMTGB2011}. For all  classifiers, we set the training cut-off time to $ 1500 $  {seconds}.

\begin{table}[b]
	\setlength{\tabcolsep}{3pt}
	\scriptsize 
	\begin{center}
		\begin{tabular}{l  r  r r r r r r r rrr}
			\toprule
			{{Dataset}} & Size & Features  & RIPPER & BRS & IMLI & CRR\\
			\midrule
			{ Heart}   & $  303 $  & $  31 $  & $   78.69  $    & $   72.13  $    & $   72.13  $    & $   77.69  $   \\[0.05cm] & & 
			& $   7 $    & $   19 $    & $   13 $    & $   4.5  $   \\[0.05cm] & & 
			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   1.8 \text{s}  $    & $   122.5 \text{s}  $   \\[.3cm] 
			{ Ionosphere}   & $  351 $  & $  144 $  & $   88.65  $    & $   91.43  $    & $   89.29  $    & $   91.43  $   \\[0.05cm] & & 
			& $   8 $    & $   4 $    & $   8.5  $    & $   20 $   \\[0.05cm] & & 
			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   2.09 \text{s}  $    & $   5.59 \text{s}  $   \\[.3cm] 
			{ WDBC}   & $  569 $  & $  88 $  & $   95.22  $    & $   95.65  $    & $   93.91  $    & $   94.69  $   \\[0.05cm] & & 
			& $   7.5  $    & $   12 $    & $   7 $    & $   34.5  $   \\[0.05cm] & & 
			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   1.38 \text{s}  $    & $   316.32 \text{s}  $   \\[.3cm] 
			{ Magic}   & $  19020 $  & $  79 $  & $   84.04  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\[0.05cm] & & 
			& $   115 $    & $   3 $    & $   24 $    & $   31 $   \\[0.05cm] & & 
			& $   15.86 \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\[.3cm] 
			{ Tom's HW}   & $  28179 $  & $  910 $  & $   97.4  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\[0.05cm] & & 
			& $   36 $    &       & $   30 $    & $   4 $   \\[0.05cm] & & 
			& $   42.73 \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\[.3cm] 
			{ Credit}   & $  30000 $  & $  110 $  & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $   82.04  $   \\[0.05cm] & & 
			& $   38.5  $    &       & $   10 $    & $   32 $   \\[0.05cm] & & 
			& $   14.52 \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\[.3cm] 
			{ Adult}   & $  32561 $  & $  144 $  & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   84.86  $   \\[0.05cm] & & 
			& $   94 $    &       & $   23 $    & $   18 $   \\[0.05cm] & & 
			& $   27.61 \text{s}  $    &       & $   11.91 \text{s}  $    & $   1016.36 \text{s}  $   \\[.3cm] 
			{ Twitter}   & $  49999 $  & $  1511 $  & $   95.74  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\[0.05cm] & & 
			& $   179.5  $    &       & $   57 $    & $   12 $   \\[0.05cm] & & 
			& $   170.87 \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\[.3cm] 
			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   84.57  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\[0.05cm] & & 
			& $   195 $    &       & $   7 $    & $   2 $   \\[0.05cm] & & 
			& $   121.02 \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\[.3cm] 
			{ Skin}   & $  245057 $  & $  119 $  & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   98.92  $    & $   95.08  $   \\[0.05cm] & & 
			& $   725 $    &       & $   201 $    & $   29 $   \\[0.05cm] & & 
			& $   1313.19 \text{s}  $    &       & $   103.8 \text{s}  $    & $   825.6 \text{s}  $   \\[.3cm] 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Comparisons of test accuracy, rule-size and training time among different rule-based classifiers. Every cell in the last four columns contains the test accuracy in percentage (top value), rule size  (middle value), and  training time in seconds (bottom value). In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.}
	\label{interpretability_crr_tab:rule_based_classifiers}
\end{table}

We consider a comparable number  of hyper-parameter choices  for each classifier. Specifically for {\crr}, we choose the data-fidelity parameter $ \lambda \in \{0.5, 0.67, 0.84, 0.99\} $, the number of clauses $ k \in \{1,2,3\} $, the relative size of mini-batch $ \frac{n_p}{n} \in \{0.25, 0.50, 0.75\} $, and the number of iterations $ \tau \in \{2,4,8,16\} $. We learn the value of $ \card_c $ and $ \eta_{l} $ from the dataset as described in Eq.~\ref{interpretability_crr_sec:ilp_query}.  In CPLEX, we set the  maximum solving time of the LP solver to $ 1000 $  seconds ($ \frac{ 1000 }{\tau} $ seconds for each iteration) and the remaining $ 500 $ seconds is allotted to construct the MLIP instances, parse the solutions and execute  other auxiliary tasks of the learning algorithm. We present the current best solution of CPLEX when the solver times out while finding the optimal solution.

We control the cut-off of the number of examples in the leaf node in the case of RF and RIPPER. For SVC, $ k $-NN, and LR we discretize the regularization parameter on a logarithmic grid. For BRS, we vary the max clause-length $ \in \{3,4,5\} $, support $ \in  \{5,10,15\} $, and two other parameters $ s \in \{100,1000,10000\} $ and $ \rho \in \{0.9,0.95,0.99\} $.  For  {\IMLI}, we consider  $ \lambda \in \{1,5,10\} $ and  $ k \in \{1,2,3\} $ and  vary the number of batches $ \tau $ such that each batch has at least $ 32 $ samples and at most $ 512 $ samples.

%\vspace{-9pt}
\subsection{Results}
In the following, we first discuss empirical results of rule-based classifiers, then extend analysis to non-rule-based classifiers, and finally discuss the effect of different choices of hyper-parameters. 
%\vspace{-3pt}
\subsubsection*{Performance Evaluation of {\crr} with Rule-based Classifiers:}
We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule-size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. 
Specifically, we show the dataset,  the number of samples and the number of discretized features in the first three columns in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. Inside each cell of column   four to $    11 $,  we present the   test accuracy (top value), rule-size (middle value) and training time (bottom value) of each classifier for each dataset.

We first compare relaxed-CNF rules generated by {\crr} with CNF rules generated by IMLI. In {Table}~\ref{interpretability_crr_tab:rule_based_classifiers}, relaxed-CNF rules exhibit higher prediction accuracy than CNF rules in the majority of the datasets, showing the effectiveness of using a more expressive representation of classification rules in capturing the decision boundary. In addition, the generated relaxed-CNF rules are comparatively smaller than CNF rules in terms of rule-size in most of the datasets. Therefore, relaxed-CNF rules improve upon CNF rules in terms of both prediction accuracy and rule size in the majority of the datasets. In this context, {\crr} provides a trade-off between accuracy and rule-size depending on the choice of hyper-parameters and the experimental results are discussed later in Section~\ref{interpretability_crr_sec:model_parameters}.  We then compare relaxed-CNF rules with DNF rules generated by BRS and find that relaxed-CNF rules outperform DNF rules with respect to prediction accuracy in several datasets. At this point,   BRS fails to scale for larger datasets as shown in Table~\ref{interpretability_crr_tab:rule_based_classifiers}. We finally compare relaxed-CNF rules with decision lists generated by RIPPER.  In the experiments, relaxed-CNF rules achieve comparable prediction accuracy with decision lists in most of the datasets. In contrast,   RIPPER generates very large decision lists compared to relaxed-CNF rules in the majority of the datasets, more precisely in large datasets.  To summarize the performance of {\crr} among different rule-based classifiers,  {\crr} can generate smaller relaxed-CNF rules with better accuracy in numbers of the cases with a couple of exceptions.

Moving focus on the training time,  the non-incremental version of {\crr} times out on larger instances in the experiments, potentially producing sub-optimal rules with reduced accuracy, thereby highlighting the need for the incremental approach. On the other hand, the incremental version of {\crr} can handle most of the datasets within the allotted amount of time. In Table~\ref{interpretability_crr_tab:rule_based_classifiers}, {\crr} takes a comparatively longer time to generate relaxed-CNF rules in comparison with other rule-based classifiers, e.g., RIPPER, and IMLI because of  the flexible combinatorial structure of relaxed-CNF rules. However, the testing time of {\crr} is insignificant ($ < 0.01 $ seconds) and thus can 
 be deployed in practice.

\subsubsection*{Performance Evaluation of {\crr} with Non-rule-based Classifiers:}

We compare  the  test accuracy of {\crr} with  non-rule-based classifiers:  LR, SVC, RF, and $ k $-NN and report the results in Table~\ref{interpretability_crr_tab:all_classifiers}. In the experiments, we  find that {\crr},  in spite of being a rule-based classifier, is able to achieve competitive prediction accuracy with non-rule-based classifiers.  In this context, SVC, and $ k $-NN  can not complete training  within the allotted time particularly in the  datasets with more than $ 10^5 $ samples, while {\crr} can still generate relaxed-CNF rules with competitive accuracy. 
Therefore, {\crr} shows the promise of applying rule-based classifiers in practice with an added benefit of interpretability along with competitive accuracy.

\begin{table}[tb]
	\begin{center}
		\begin{tabular}{l   r r r r r  r }
			\toprule
			{Dataset}  & LR & SVC & RF & $k$-NN & CRR\\\midrule
			\multirow{1}{*}{ Heart}   & $   84.29  $    & $   83.33  $    & $   81.97  $    & $   78.69  $    & $   77.69  $   \\ 
			\multirow{1}{*}{ Ionosphere}   & $   94.29  $    & $   91.43  $    & $   92.96  $    & $   91.43  $    & $   91.43  $   \\ 
			\multirow{1}{*}{ WDBC}   & $   98.26  $    & $   96.46  $    & $   96.9  $    & $   95.61  $    & $   94.69  $   \\ 
			\multirow{1}{*}{ Magic}   & $   85.15  $    & $   84.45  $    & $   85.3  $    & $   77.9  $    & $   81.31  $   \\ 
			\multirow{1}{*}{ Tom's HW}   & $   97.62  $    & $   97.66  $    & $   97.52  $    & $   94.59  $    & $   97.34  $   \\ 
			\multirow{1}{*}{ Credit}   & $   82.04  $    & $   82.12  $    & $   81.97  $    & $   80.5  $    & $   82.04  $   \\ 
			\multirow{1}{*}{ Adult}   & $   87.24  $    & $   86.82  $    & $   86.84  $    & $   84.68  $    & $   84.86  $   \\ 
			\multirow{1}{*}{ Twitter}   & $   96.28  $    & $   96.34  $    & $   96.37  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.16  $   \\ 
			\multirow{1}{*}{ Weather-AUS}   & $   85.71  $     &  \multicolumn{1}{c}{\textemdash}      & $   85.63  $     &  \multicolumn{1}{c}{\textemdash}      & $   83.34  $   \\ 
			\multirow{1}{*}{ Skin}   & $   97.21  $     &  \multicolumn{1}{c}{\textemdash}      & $   99.81  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.08  $   \\ 
			\bottomrule
		\end{tabular}
	\end{center}
	\caption{Comparisons of test accuracy among {\crr} and  non-rule-based classifiers. In the experiments, {\crr} achieves competitive prediction accuracy in spite of being a rule-based classifier.}
	\label{interpretability_crr_tab:all_classifiers}
\end{table}

\begin{figure}[t]
	%\vspace{-25pt}
	\centering
	\subfloat
	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_lambda.pdf}\label{interpretability_crr_fig:d}} \hfill
	\subfloat
	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_clause.pdf}\label{interpretability_crr_fig:f}} \hfill
	\subfloat
	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_subsamplesize.pdf}
		\label{interpretability_crr_fig:a}} \hfill
	\subfloat
	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_iteration.pdf}\label{interpretability_crr_fig:g}} \hfill
	\caption{Effect of model hyper-parameters on test accuracy  for  Heart disease dataset. } 
	\label{interpretability_crr_fig:result_1}
\end{figure}

%\vspace{-3pt}
	\subsubsection*{Varying Model Parameters: }
	\label{interpretability_crr_sec:model_parameters}
	In Figure~\ref{interpretability_crr_fig:result_1},~\ref{interpretability_crr_fig:result_2}, and~\ref{interpretability_crr_fig:result_3}, we demonstrate the effect of varying the hyper-parameters of {\crr}. To understand the effect of a single hyper-parameter, we fix the values of other hyper-parameters to a default choice where the default choice results in the most accurate rule.

	%	%\vspace{-3pt}
	\textbf{Varying data-fidelity parameter ($ \lambda $):}  As we increase data-fidelity parameter $ \lambda $ in the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj} and Eq.~\ref{interpretability_crr_eq:ilp_inc}, more priority is given to the prediction accuracy than the sparsity of the rules.  In most of the datasets, we similarly observe an increase in accuracy  and also an increase in the size of the rules when $ \lambda $ is higher. This suggests that improved interpretability can often come at a  minor cost in accuracy.  
	In addition, we  find an increase in training time for most of the datasets indicating that the MILP query usually takes a longer time to find the solution when more priority is given on the prediction accuracy.

	\begin{figure}[t]
		%\vspace{-25pt}
		\centering
		\subfloat
		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_lambda.pdf}\label{interpretability_crr_fig:c}} \hfill
		\subfloat
		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_clause.pdf}\label{interpretability_crr_fig:h}} \hfill
		\subfloat
		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_subsamplesize.pdf}\label{interpretability_crr_fig:b}} \hfill 
		\subfloat
		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
		%\vspace{-10pt}
		\caption{Effect of model hyper-parameters on rule-size for  Heart disease dataset.}

		\vspace{-10pt}
		\label{interpretability_crr_fig:result_2}
	\end{figure}

	{\textbf{Varying the number of clauses ($ k $): }}
	%	In Figure~\ref{interpretability_crr_fig:clause(a)}, we show the effect of $ k $ on  test accuracy   for Parkinsons {dataset}. 
	As we increase  $ k $, {\crr} allows the generated rules to capture the variance in the given dataset more effectively, that results in higher  accuracy in most of the datasets. The  rule-size also increases as we learn more clauses. In addition, the training time  increases, that can be reasoned by the fact that the number of constraints in the MILP formulation is linear with $ k $. Thereby, the number of clauses $ k $ provides a control over the accuracy of the generated rule and training time.

	 \begin{figure}[btp]
	 	%\vspace{-25pt}
	 	\centering
	 	\subfloat
	 	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_lambda.pdf}\label{interpretability_crr_fig:c}} \hfill
	 	\subfloat
	 	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_clause.pdf}\label{interpretability_crr_fig:h}} \hfill
	 		\subfloat
	 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_subsamplesize.pdf}\label{interpretability_crr_fig:b}} \hfill 
	 		\subfloat
	 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
	 	%	%\vspace{-10pt}
	 	\caption{Effect of model hyper-parameters on training  time for  Heart disease dataset.}

	 	\vspace{-10pt}
	 	\label{interpretability_crr_fig:result_3}
	 \end{figure}

	\textbf{Relative size of the mini-batch:} We vary the relative size of the mini-batch $ \frac{n_p}{n} $ to observe its effect on the accuracy and the size of the rules. In most datasets,  the accuracy increases when more samples are considered in the mini-batch, costing higher   training time.  Moreover, the size of the generated rule increases as $ \frac{n_p}{n} $ increases, that can be supported by the increase  in the variance of the samples in the mini-batch.

	%	%\vspace{-3pt}
	\textbf{Varying the number of iterations ($ \tau $):}   
	As we allow more iterations in the learning process, we find an increase in accuracy in most datasets. The training time also increases with $ \tau $ because {\crr} is required to solve in total $ \tau $ queries. We also observe an increase in rule-size in most datasets. The reason is that the objective function in the incremental mini-batch approach in Eq.~\ref{interpretability_crr_eq:ilp_inc} does not put a restriction on the size of the rules, rather on the change of rules in consecutive iterations.  In addition, the learned values of the thresholds $ \eta_c$ and $ \eta_l $  in one iteration are not carried to the MILP query in the next iteration, that may cause an increase of rule-size.

### 2022-01-27 chapters/interpretability/relaxed-cnf/framework.tex
relaxed-cnf paper added

\section{Description of $ \Framework $}\label{sec:maxsat}

	\subsection{Transformation to MaxSAT Query}\label{subsec:maxsat_query}

% The construction of $Q_i$ takes in four parameters: (i) $ \noclause $, the desired number of clauses in CNF rule, (ii) $ \lambda $, the regularization parameter, (iii) a matrix ${\mathbf{X}^{i}} \in \{0,1\}^{n \times m}$  describing the  binary value of  $m$ features for each of    $n$ samples with $\mathbf{X}^{i}_q$ being a binary valued vector for the $ q $-th sample corresponding to feature vector $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$, (iv) a label vector $\mathbf{y}^{i} \in \{0,1\}^n$ containing a class label $y^{i}_q$ for the  sample $\X^{i}_q$.

   Given an instance space $ \{ \mathbf{X}, \mathbf{y} \} $ and  model parameters $ \noclause,\card$, and  $\lambda $, {{\Framework}} constructs a {MaxSAT} query  $ Q $ and invokes an off the shelf MaxSAT solver to compute the underlying rule $ \Rule $.

	{\Framework} considers two types of propositional variables: (i) feature variables and (ii) noise variables.
%	 Feature variables are defined once for all partitiones. 
 	   In our problem setting, a $ \noclause $-clause CNF rule $ \Rule=\bigwedge_{i=1}^\noclause C_i $ is  represented as the conjunction of  $ \noclause $ cardinality constraints where  each $ C_i $ is defined as   the  arithmetic sum of   feature variables compared against a cardinal constant $ \card $. A  sample $ \X_q $ satisfies  $ C_i $ if $ \X_q $ has at least $ \card $ features whose representative variables are present in $ C_i $. If $ \X_q $ satisfies  $ \forall i, C_i$, then  $ \Rule(\X_q)=1 $ otherwise $ \Rule(\X_q)=0 $. Since  feature $ x_j $ can be present or not present  in each of   $ \noclause $ constraints, {\Framework} considers $ \noclause $  boolean variables, each denoted by $ b_j^i $ ($ i \in [1,\noclause], j \in [1,m] $) for  feature $ x_j $ to denote its participation in each of $ \noclause $ clauses. A sample $ \X_q $, however, can be misclassified by $ \Rule $ i.e. $ \Rule(\X_q)\oplus  y_q =1 $.  {\Framework}  introduces  a noise variable $ \noise_q  $ ($ q \in [1,n] $) corresponding to  sample $ \X_q $ so that the assignment of $ \noise_q $ can be interpreted whether $ \X_q $ is misclassified by $ \Rule $ or not.  Hence the key idea of {{\Framework}} for learning $ \Rule $ is to define a {MaxSAT} query over $k \times m + n$ propositional variables, denoted by $\{ b_{1}^{1}, b_{2}^{1}, \dots ,b_{m}^{1}, \dots ,b_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.   The {\MaxSAT} query of {\Framework} consists of the following three sets of constraints:

	\begin{enumerate}

		\item Since our objective is to find sparser rules,  {\Framework}  adds a constraint to falsify as many $ b_j^i $  as possible by constructing a soft clause $ (\neg b_j^i) $.   The weight  corresponding to this clause is $ 1 $. 
	\begin{equation}
	V^i_j:= 
	\neg b^i_j
	;\quad  W(V^i_j)= 1
	\label{eq:constraint-v}
	\end{equation}
		\item We use noise variables to handle mis-classifications and therefore, {\Framework} tries to falsify as many   noise variables as possible. Since regularization parameter $ \lambda $ is proportionate to accuracy (higher $ \lambda $ ensures higher accuracy), {\Framework} puts $ \lambda $ weight to each of the following soft  clause. 
%		containing the complement of  noise variable.   
		\begin{equation}
		N_q:= (\neg \noise_q ); \qquad \qquad W(N_q) = \lambda
		\label{eq:constraint_two}
		\end{equation}
		%		Here $N_i $ is also a soft clause.

		\item 
		Let $\mathbf{B}_i = \{b^i_{j} \mid j \in [1,m] \}$ be the vector of feature variables corresponding to the $ i $-th cardinality constraint. 
		Here we provide the third set of constraints of $ Q $. 
		\begin{equation}
		D_q:= (\neg \noise_q \rightarrow ( y_q \leftrightarrow \bigwedge_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i \ge \card })));    W(D_q) = \infty
		\label{eq:constraint_three}	
		\end{equation}
		Every hard clause $D_q$  can be interpreted as follows. If $\noise_q$ is assigned to $ 0 $ (i.e. $ q $-th sample is not considered as noise) then ${y}_q = \Rule(\X_q) $.  The operator ``$\circ $'' is defined in Sect. \ref{sec:preliminaries}.

		%		We are not showing in details how $ D_i $ can be encoded into CNF forms using standard procedure. We refer \cite{MM18} for detailed description.

	\end{enumerate}
	Finally, the set of constraints $Q$  constructed by {{\Framework}} is defined as follows:
	\begin{equation*}
	\label{eq:constraint-imli}
	Q_i :=  \bigwedge_{j=1, i = 1}^{j=m,l=k} V^i_j \wedge \bigwedge_{q=1}^{n} N_q \wedge \bigwedge_{q=1}^{n} D_q
	\end{equation*}

	Next, we extract $\Rule$ from the solution of $Q$ as follows. 
	\begin{construction}
		Let $\sigma^* = \MaxSAT(Q,W)$, then $x_j \in {\clause(\Rule,i)}$ iff $\sigma^*(b_{j}^i) = 1$.
	\end{construction}

	\subsection{CNF Encoding of MaxSAT Constraints $ Q $}
	A MaxSAT query consists of CNF clauses with associated weight for each clause. Here we discuss the CNF encodings of $ Q $. The constraints $ V^i_j $ and $ N_q $ are unit clauses and do not require further processing.

	We consider a function $ \CNF(\dots) $ which takes a cardinality constraint as input and outputs a CNF formula. We assume that an off the self PB-Encoder (pseudo-boolean encoder) can implement  $ \CNF $, and we skip the inner knowledge of  $ \CNF $ of converting any pseudo-boolean constraint into a CNF formula since it is beyond the scope of this paper.

	Now we present the CNF encoding of $ D_q $. 
	Recall that, $ y_q $ is input to our framework.  if $ y_q=1 $, $ D_q $ can be encoded as follows.
	\[ \neg \xi_q \rightarrow \bigwedge_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i \ge \card }) \Rightarrow  \xi_q \vee \bigwedge_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i \ge \card }) \]
	 Here we apply the formula $ a \rightarrow b \Rightarrow \neg a \vee b $.   Therefore, we call $ \CNF $ $ k $ times for each  $ {\mathbf{X}_q} \circ {\mathbf{B}_i \ge \card } $ as input and let $ F_1, \dots, F_k $ be the output CNF formulas for each input. Now $ \xi_q \vee (\bigwedge_{i=1}^k  F_i) \Rightarrow \xi_q \vee   \hat{F}  $ by applying associative law. We further apply 
	 distributive law on  ($ \xi_q \vee \hat{F})  $ to get CNF clauses.
	 When $ y_q = 0 $, $ D_q $ is encoded as follows 
	  \[ 
	  \begin{split}
	  \neg \xi_q \rightarrow \neg (\bigwedge_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i \ge \card })) 
	  &\Rightarrow \neg \xi_q \rightarrow \bigvee_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i < \card })\\
	   &\Rightarrow  \xi_q \vee \bigvee_{i=1}^{k} ({\mathbf{X}_q} \circ {\mathbf{B}_i < \card })
	  \end{split}
	   \]
	   We again call $ \CNF $ for each $ {\mathbf{X}_q} \circ {\mathbf{B}_i < \card } $ as input  and  let $ F_1, \dots, F_k $ be the  output CNF formulas. Now $ \xi_q \vee ( \bigvee_{i=1}^k F_i) $ can be encoded into CNF by applying Tseytin transformation. We introduce $ k $ additional boolean variables  $ \xi_{q,1}, \dots, \xi_{q,k}   $  and apply the following transformation. 
	   \[ 
	   \begin{split}
	   &\quad\;\; \xi_q \vee ( \bigvee_{i=1}^k F_i)\\
	   &\Rightarrow (\xi_q \vee (\bigvee_{i=1}^k \xi_{q,i})) \wedge (\bigwedge_{i=1}^k \xi_{q,i} \rightarrow F_i) \\
	   &\Rightarrow (\xi_q \vee (\bigvee_{i=1}^k \xi_{q,i})) \wedge (\bigwedge_{i=1}^k \neg \xi_{q,i} \vee F_i) 
	   \end{split} \]

### 2022-01-27 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
relaxed-cnf paper added

\section{{\crr}: Classification Rules in Relaxed Logical Form}
\label{interpretability_crr_sec:ilp}
In this section, we describe the main contribution of our work, {\crr}, a  framework for learning   relaxed-CNF rules.  {\crr}  converts the learning problem into  an ILP-based formulation,  learns the optimal assignment of variables and constructs rule $ \Rule $ based on the assignment. We organize the rest of this section as follows.
We  discuss the decision variables in Section ~\ref{interpretability_crr_sec:variables},  the constraints and the linear programming relaxation in Section~\ref{interpretability_crr_sec:ilp_query}, the incremental learning in Section~\ref{interpretability_crr_sec:inc_ilp}, feature discretization in Section~\ref{interpretability_crr_sec:discretization}, and adaptation of {\crr} for learning other classification rules in Section~\ref{interpretability_crr_sec:other_rules}.
% We conclude this section by describing how {\crr} can be adjusted to learn interpretable rules in other logical forms in Section~\ref{interpretability_crr_sec:other_format}.

%Given an instance space $ \{ \mathbf{X}, \mathbf{y} \} $ and  model parameters $ \noclause,\card_c,\card_l$, and  $\lambda $, {{\crr}} constructs an ILP query $ Q $  and invokes an off the shelf ILP solver to compute the underlying rule $ \Rule $.

\subsection{Description of Variables}
\label{interpretability_crr_sec:variables}
{\crr} considers two types of  \emph{decision} variables: (i) feature variables and (ii) noise (classification error) variables.
Since  feature $ x_j $ can be present or not present  in each of  the  $ \noclause $ clauses, {\crr} considers $ \noclause $   variables, each denoted by $ b^i_j  $  corresponding to   feature $ x_j $  to denote its participation  in the $ i $-th clause, i.e, $ b^i_j=\mathbb{1}[j\text{- th feature is selected in }i\text{-th clause}]  $.  The $ q $-th sample in the training dataset, however, can be misclassified by $ \Rule $.  Therefore, {\crr}  introduces  a noise variable $ \noise_q \in \{0,1\}  $  corresponding to the $ q $-th sample, so that the assignment of $ \noise_q  $ can be interpreted whether $ \X_q $ is misclassified by $ \Rule $ or not, i.e., $ \noise_q=\mathbb{1}[q\text{-th sample is misclassified}] $.  Hence, the key idea of {{\crr}} for learning $ \Rule $ is to define an ILP query over $k \cdot m + n$  decision variables, denoted by $\{ b_{1}^{1}, b_{2}^{1}, \dots ,b_{m}^{1}, \dots ,b_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.   In this context, we define $\B_i = \{b^i_{j} \mid j= 1,\dots,m\}$ as a \emph{vector of   feature variables} corresponding to the $ i $-th clause.

\subsection{Construction of ILP Query}
\label{interpretability_crr_sec:ilp_query}
We first discuss the objective function of the  ILP query $ Q $ for learning  a $ k $-clause relaxed-CNF rule  $ \Rule $. The objective function takes care of  both the rule-sparsity and the prediction accuracy of $ \Rule $. Since {\crr} prefers a sparser rule  with as few literals as possible, we construct the objective function by preferring  $ b^i_j $ to be  $ 0 $. Moreover, to encourage $ \Rule $ to predict  the training samples accurately, we penalize the number of variables $ \noise_q $ that are different from $ 0 $. In this context, we utilize  the   parameter $ \lambda $ to trade off between sparsity and accuracy.  Therefore,  the objective function of the ILP query $ Q $ is to minimize the normalized sum of all  noise variables $ \noise_q $  weighed  by the data-fidelity parameter $ \lambda $ and feature variables $ b^i_j $ weighed by $ 1-\lambda $ (Eq.~\ref{interpretability_crr_eq:ilp_obj}).

We formulate the constraints of the ILP query $ Q $ as follows. Initially, we define the range of the decision variables and   add constraints accordingly (Eq.~\ref{interpretability_crr_eq:ilp_range} and ~\ref{interpretability_crr_eq:ilp_range_2}).  For each sample, at first, we  add constraints to mimic the behavior of hard-OR of literals in a clause, and then we add constraints to apply  soft-AND of clauses in a formula.

We first consider the case when  the $ q $-th  sample  has positive class label (Eq.~\ref{interpretability_crr_eq:ilp_pos}). 
$ \X_q\circ \B_i \ge \card_l $ resembles the hard-OR operation of literals in a clause.  We introduce   $ k $ \emph{auxiliary} $ 0 $-$ 1 $ variables $ \{\xi_{q,1}, \dots, \xi_{q,k}\} $  to check whether at least $ \card_c $ clauses are satisfied, i.e., $ \noise_{q,i}=\mathbb{1}[i\text{-th clause is dissatisfied for }q\text{-th sample}] $, which let us impose the operation of soft-AND over clauses.    We then add a constraint to make sure that at most $ k-\eta_{c} $ clauses are allowed to be dissatisfied, otherwise the noise variable $ \noise_q $  is assigned to $ 1 $, i.e., the $ q $-th sample is detected as aw noise.

A negative labeled sample has to dissatisfy  more than $ k-\card_c $ clauses in $ \Rule $ so that the sample is predicted as $ 0 $, which is equivalent to satisfying  more than $ k-\card_c$ constraints $ \X_q\circ \B_i < \eta_l  $. As mentioned earlier, we  introduce $ 0 $-$ 1 $ variable $ \noise_{q,i} $ to specify if the constraint $ \X_q\circ \B_i < \eta_l  $ is dissatisfied or not and  restrict the count of dissatisfied clauses $ \sum_{i=1}^k\xi_{q,i} $ to be less than $ \card_c $.

In the following, we present the ILP query $ Q $.

%\vspace{-15pt}
\begin{subequations}
	\label{interpretability_crr_eq:ilp}
 \begin{align} 	
% 	\begin{split}
 	&\min   \frac{\lambda}{n} \sum_{q=1}^n \xi_q + \frac{1-\lambda}{k\cdot m}\sum_{i=1}^k\sum_{j=1}^m b_j^i \label{interpretability_crr_eq:ilp_obj} \\
 	&\text{such that,} \notag \\
 	&b_j^i \in \{0,1\},  i = 1,\dots,k,  j = 1,\dots,m \label{interpretability_crr_eq:ilp_range}\\
 	& \xi_q \in \{0,1\},q = 1,\dots,n  \label{interpretability_crr_eq:ilp_range_2}\\
 	&\text{if }  \forall q \in \{1,\dots,n\},\text{ if }  y_q=1, \label{interpretability_crr_eq:ilp_pos}\\
 	&\qquad \X_q\circ \B_i+ m\xi_{q,i} \ge \eta_{l},  i = 1,\dots,k  \label{interpretability_crr_eq:ilp_ela_l_pos}\\
 	& \qquad   k\xi_q + k-\eta_c \ge \sum_{i=1}^k \xi_{q,i} \notag \\ &\qquad \xi_{q,i} \in \{0,1\},  i = 1,\dots,k \notag\\
 	&\text{if } \forall q \in \{1,\dots,n\}, y_q=0, \label{interpretability_crr_eq:ilp_neg}\\
 	&\qquad \X_q\circ \B_i < \eta_{l} + m  \xi_{q,i},  i = 1,\dots,k \label{interpretability_crr_eq:ilp_ela_l_neg}\\
 	&\qquad  k\xi_q  + \eta_{c} > \sum_{i=1}^k\xi_{q,i} \notag\\ &\qquad \xi_{q,i} \in \{0,1\},  i = 1,\dots,k \notag
% 	\end{split}
\end{align}
\end{subequations}
%\vspace{-15pt}

In the following, we show the complexity of the ILP query in terms of the number of variables and constraints.

\begin{proposition}
	Given a training dataset with $ n $ samples and $ m $ binary features, the ILP query $ Q $ for learning a binary classification rule in relaxed-CNF has $ k\cdot m + n $ decision variables, $ k\cdot n $ auxiliary variables, and  $ k\cdot m + n\cdot (k+3)  $ integer constraints. 
\end{proposition}

An ILP solver takes  query  $ Q $ as input and returns the optimal assignment $ \sigma^* $of the variables.
We extract relaxed-CNF rule $\Rule$ from the solution  as follows.

\begin{construction}
	\label{interpretability_crr_cons:rule}
	Let $\sigma^* = \ILP(Q)$, then $x_j \in {\clause(\Rule,i)}$ iff $\sigma^*(b_{j}^i) = 1$.
\end{construction}

%\textbf{Hardness of the ILP problem:}The ILP query in Eq.~\ref{interpretability_crr_eq:ilp} has integer constraints, and the problem is NP-hard. We prove this by reducing the presented ILP query to learn a sparse single clause CNF rule, that is consistent with the training dataset (does not make any classification error) by specifying $ k=1, \eta_{c}=1, \eta_l=1, \text{ and } \lambda=1$ in Eq.~\ref{interpretability_crr_eq:ilp}.  In this context, Malioutov et al.~\cite{MV2013} has demonstrated that learning a sparse single clause CNF rule consistent on the training dataset  is a combinatorial problem in NP-hard by showing connections between compressed sensing and sparse signal recovery~\cite{natarajan1995sparse}.

\textbf{Learning thresholds $ \eta_l $ and $ \eta_c $:}
Given the training dataset $ (\mathbf{X}, \mathbf{y})$ and data-fidelity parameter $ \lambda $, one could learn the optimum value of the thresholds $ \card_{c} $ and $ \card_l $ of the desired rule $ \Rule $  by specifying their range as constraints in the ILP query $ Q $ in Eq.~\ref{interpretability_crr_eq:ilp}. More precisely, we need to add two integer constraints  $ \card_c  \in \{0,\dots,k\}$ and $ \card_l \in \{0,\dots,m\} $ in the above query and then learn their values from the solution.

A more generalized version to CNF rules would be to learn different thresholds on literals for different clauses, i.e., $ \eta_{l,i} $ for the $ i $-th clause. This facilitates to capture the complex decision boundaries, while still producing rule-based decisions. 
In our ILP formulation, it is straight-forward to consider such generalization where we put constraints $ \eta_{l,i} \in \{0,\dots,m\}, i=1,\dots, k $ and replace
 $ \eta_l $ with $ \eta_{l,i} $ in Eq.~\ref{interpretability_crr_eq:ilp_ela_l_pos} and~\ref{interpretability_crr_eq:ilp_ela_l_neg}.

%\[
%\begin{split}
%	& \eta_c \in [1,k]\\
%	& \eta_l \in [1,m] \\
%	& \sum_{j=1}^m b_i^j \ge \eta_l, \forall i \in [1,k] \\ 
%\end{split}
%\]

\textbf{Relaxing the ILP problem:}
\label{interpretability_crr_sec:relaxation}
The ILP query $ Q $ has binary integer constraints and the solution of this integer program is computationally intractable. 
A common approach that efficiently finds an approximate solution to such a problem extends to relax the integer constraints,  solves the LP-relaxed (linear programming relaxation) problem, and then rounds the non-integer variables to get an integer solution as in~\cite{MV2013}. 
In our case, we cannot relax all integer constraints because it may violate the structure of relaxed-CNF rules. Specifically, $ \eta_c $ and $ \eta_l $ (or $ \eta_{l,i} $) must be integers in the construction of relaxed-CNF rules, and $ \noise_{q,i} $ needs to be an integer to precisely calculate the noise variable $ \noise_q $.  However,  we can relax feature variable $ b^i_j $ and noise variable $ \noise_q $ and solve the corresponding MILP  problem.  To construct the MILP problem, we replace Eq.~\ref{interpretability_crr_eq:ilp_range} with $ 0 \le b^i_j \le 1 $ and Eq.~\ref{interpretability_crr_eq:ilp_range_2} with $ 0 \le \noise_q \le 1 $, and the rest of the constraints in $ Q $ remain the same.  If non-zero $ b^i_j $ is found in the solution, we set it to $ 1 $ and then construct the rule according to Construction~\ref{interpretability_crr_cons:rule}.

%{\color{red} WRite about learning DNF rule}

%\vspace{-5pt}
\subsection{Incremental Mini-batch Learning}
\label{interpretability_crr_sec:inc_ilp}

In Section~\ref{interpretability_crr_sec:ilp_query}, we  present an ILP formulation for learning relaxed-CNF rules and then discussed the relaxation to the integer constraints in the MILP formulation to make the approach computationally tractable. However, not all integer constraints in the formulation can be relaxed, as discussed  in Section~\ref{interpretability_crr_sec:relaxation}. Thus we require an improved learning technique to achieve scalability. We now describe a mini-batch learning approach to {\crr}, that learns relaxed-CNF rule $ \Rule $ incrementally from a set of mini-batches while solving a modified MILP query for each mini-batch.

In incremental mini-batch learning, the learning process repeats for a fixed number of iterations. In each iteration, we randomly select an equal number of samples from the full training set and generate a mini-batch with less number of samples.  We then construct a MILP query on the current mini-batch with a modified objective function. This objective function simultaneously penalizes the prediction errors
on the current mini-batch and the change in the rules learned in consecutive batches.  
In the following, we discuss the modified objective function and the  MILP formulation.

% The number of constraints in the ILP query  is linear with the number of training samples, which results in poor scalability  for larger datasets.  Hence, we propose an incremental learning technique via a mini-batch training methodology \cite{GM19}.  In this context, we define $ \Rule_p $ to be the rule learned for the $ p $-th batch. The key idea of incremental learning is to  divide the training dataset sequentially into a fixed number  of mini-batches $ \tau $, learn rule for each batch one after another such that  $ \Rule_p $  depends not only on the training samples in the $ p $-th batch but also regularizes itself with the  rule $ \Rule_{p-1} $ learned for the  $ (p-1) $-th batch.  To this end, we use notation $ (\X^p,\mathbf{y}^p)$  to refer to the training data for the $ p $-th batch. We assume that $ \forall p \in \{1,\dots,\tau\}, |\X^p|=n_\tau $. We propose  the following modifications in the ILP-based formulation for incrementally learning relaxed-CNF rules.

Let $ (\X,\mathbf{y})$ be a training dataset with $ n $ samples and $ m $ binary features and $ \tau $ be the number of iterations in the learning process. In the $ p $-th iteration, we randomly sample a mini-batch $ (\X_p,\mathbf{y}_p)$ with equal number $ n_p $ of  samples and $ n_p \le n, p=1,\dots, \tau $. Note that all mini-batches have the same  binary features $ \mathbf{x} =\{x_1, \dots, x_m\} $ as in the  training set and thus share the same feature variables $ b_j^i $ in the MILP query. Therefore, we devise an objective function that prefers to keep the  assignment of $ b_j^i $ learned in the $ (p-1) $-th iteration while minimizing the prediction error on  the current mini-batch $ (\X_p,\mathbf{y}_p)$. To this end, each $ b_j^i $ is assigned $ 0 $ initially in the learning process.  This technique  enables us to update the learned rule in terms of the update in  assignment of feature variables $ b_j^i $ over mini-batches.

Let  $ Q_p $ be the MILP query for the $ p $-th mini-batch for learning the $ k $-clause relaxed-CNF rule $ \Rule_p $. Thus, $ \Rule_0 $ is an empty rule.
We consider an indicator function $ I(\cdot): b^i_j \rightarrow \{1,-1\} $, that takes a feature variable $ b^i_j $ as input and outputs $ -1 $  if $ b^i_j $ is assigned $ 1 $ in the solution of $ Q_{p-1} $ (i.e., feature $ x_j $ is selected in the $ i $-th clause of   $ \Rule_{p-1} $), otherwise outputs 1.

%\vspace{-8pt}
\[
I(b^i_j)=
\begin{cases}
-1 \quad \text{if }  x_j \in \clause(\Rule_{p-1},i)
\\
1 \quad \text{otherwise}
\end{cases}
\]
%\vspace{-8pt}

We now discuss the modified objective function where we multiply $ I(b^i_j) $ with $ b^i_j $ differently  from the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj}.

%\vspace{-10pt}
\begin{equation}
\label{interpretability_crr_eq:ilp_inc}
	\begin{split}
	&\min  \frac{\lambda}{n} \sum_{q=1}^{n_p} \xi_q  +  \frac{1-\lambda}{k\cdot m} \sum_{i=1}^k\sum_{j=1}^m b_j^i \cdot I(b^i_j) \\
	\end{split}
\end{equation}
%\vspace{-10pt}
In the objective function, the first term penalizes the prediction error of samples in the current mini-batch and the second term penalizes when $ b_j^i $ is assigned differently than its previous  assignment. Note that the total prediction error is normalized by dividing by $ n $, which is the size of the full training dataset. This normalization is useful as it assists in updating $ \Rule_p $ while also considering the relative size of the mini-batch. Intuitively, if the size $ n_p $ of the mini-batch  is close to the size $ n $
of the training set, more priority is given to the prediction accuracy and vice versa. The constraints in the  query $ Q_p $ are similar to the constraints in Eq.~\ref{interpretability_crr_eq:ilp}.  
Finally, the prediction rule  $ \Rule $ is  $ \Rule_\tau $ that is learned for the last mini-batch.

%The core idea of our incremental mini-batch learning is motivated by~\cite{GM19}. However, our approach is  different from~\cite{GM19} in two ways:  the generation of mini-batches and the design of the objective function.  The incremental approach in~\cite{GM19} adopts a sequential partitioning of the training set to  learn a CNF rule incrementally, and  the generated rule highly depends  on the order/position of samples in the training set. However, in this paper, we randomly sample mini-batches to avoid such a dependency on the order of samples. Additionally, we consider a normalized objective function with an added flexibility in updating rules by considering the size of the mini-batch, which~\cite{GM19} does not consider.

\begin{proposition}
	At each iteration, the MILP query in the incremental mini-batch learning approach has $ k\cdot m + n_p \cdot (k+3) $  constraints, where $ n_p $  is the size of the mini-batch. 
\end{proposition}

\textbf{Learning thresholds $ \eta_l $ and $ \eta_c $:} In the incremental learning, the objective function in Eq~\ref{interpretability_crr_eq:ilp_inc} does not impose any constraint on the thresholds. In fact, the thresholds are learned in each iteration by solving the corresponding MILP query and we  consider their final values in the last iteration.

\subsection{Learning with Non-binary Features}
\label{interpretability_crr_sec:discretization}
Since our problem formulation requires input instances to have binary features, datasets with categorical and continuous features  require a preprocessing stage. Initially, for all continuous features, we apply entropy-based discretization~\cite{fayyad1993multi} to infer the most appropriate number of categories/intervals by recursively splitting the domain of each continuous feature to minimize the class-entropy of the given dataset.\footnote{A simple quantile-based discretization also works, but it requires an extra parameter (i.e., the number of quantiles).}  For example, let $ x_c \in [a,b] $ be a continuous feature, and  entropy-based discretization splits the domain $ [a,b] $ into three intervals with two split points $ \{a',b'\} $, where $ a<a'<b'<b $.  Therefore, the result intervals are $ { x_c < a'} $, $ {a' \le x_c < b'} $, and $ { x_c  \ge b'} $.

After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{ghosh19incremental,LKCL2019}. In this encoding, a Boolean vector is introduced with cardinality equal to the number of distinct categories. Let a categorical feature have three  categories `red',`green', and `yellow'. In one-hot encoding, samples with category-value `red',`green', and `yellow' would be converted into binary features while taking values $ 100 $, $ 010 $, and $ 001 $, respectively.

\subsection{Learning Rules in Other Logical Forms}
\label{interpretability_crr_sec:other_rules}
While {\crr} learns  classification rules in relaxed-CNF form, we can leverage this framework for learning classification rules in other logical forms, for example, CNF,  DNF, and decision sets. To learn a CNF rule one can set  $ \card_l=1 $ and $ \card_c=k $. Moreover,  to learn a DNF rule,  one would first complement the class label of all samples, then learn a CNF rule by setting the parameters as described and finally negate the learned rule. We note that the literature on decision sets also fits into our formulation, because decision sets can be represented by DNF formulas. As a future work, we are investigating whether related formulations could be used to address other learning paradigms such as decision lists.

%We can extend this formulation for learning decision sets, a popular rule-based interpretable representation of the decision function~\cite{ignatiev2018sat,lakkaraju2016interpretable}. A decision set is an unordered set of ``if-then'' clauses where each clause is a conjunction of literals and  is associated with a class label. In the restricted setting, a decision set is a DNF rule because each clause in DNF is a conjunction of literals and predicts class $ 1 $ when the clause becomes true. As a future work, we remove this restriction and leverage the presented formulation for learning decision sets.

%Decision lists are slightly different from decision sets, and one would think of decision lists as ``if-then-elseif-$ \dots $else'' rule~\cite{R1987}. \red{Can we comment on learning decision lists? Is learning decision sets sufficient in our context as a future work?}

%\begin{proof} (Theorem~\ref{interpretability_crr_thm:succinctness})
%	\label{interpretability_crr_proof:succinctness}
%	Let $ F= \bigwedge_i C_i $  be a CNF formula of $ {m \choose{m-\eta+1}} $  clauses where each clause $ C_i $ is distinct and has $ m-\eta+1 $ literals of $ C $. $ F $ is an equivalent CNF formula of clause $ (C,\eta) $ if $ \sigma \models (C,\eta) \Leftrightarrow  \sigma \models F $. 
%	
%	We will first prove $ \sigma \models (C,\eta) \Rightarrow \sigma \models F $.  
%	Since $ \sigma \models (C,\eta) $, at least $ \eta $ literals in $ C $ are  true  and let $ T $ be the set of those $ \eta $ literals. Therefore, according to pigeon hole theorem we cannot construct any $ C_i $ with $ m-\eta+1 $ literals such that $ T \cap C_i = \emptyset $. Hence, at least $ 1 $ literal in $ C_i $ is also in $ T $. Therefore, $ \sigma \models (C_i,\eta) \Rightarrow \sigma \models F$.
%	
%	We prove $ \sigma \models F \Rightarrow \sigma \models (C,\eta) $ using a proof by contradiction. Let assume $  \sigma \models F $ and  $ \sigma \not \models (C,\eta) $. Hence at least $ m-\eta+1 $ literals in $ C $ must be $ false $. Then there exists a clause $ C_{i'}  $ that consists of these $ m-\eta+1 $ literals. But $ F = \bigwedge_i C_i $ must include  $ C_{i'}  $ and $ \sigma \models  C_{i'} $, which is a contradiction. Hence $ \sigma \models F \Rightarrow \sigma \models (C,\eta) $. Therefore, $ F $ is an equivalent CNF formula of $ (C,\eta) $. 
%	
%	We now prove that $ F $ is a compact encoding. Suppose $ F $ excludes clause $ C_{i'} $ but includes all other clauses as well as clauses with greater than $ m-\eta+1 $ variables. We  set the variables in $ C_{i'} $ as  false  and every other variables  true. Then $ F $ is satisfiable. But $ (C_{i'},\eta) $ is not satisfiable because each literal in $ C_{i'} $ is $ false $. Therefore, $ F $ must include all $ {m \choose{m-\eta+1}} $ clauses and the total literal count in $ F $ is $ (m-\eta +1) {m \choose{m-\eta+1}}  $. When $ \eta=1 $, $ (m-\eta+1){m\choose{m-\eta+1}} = m $, and when $ \eta > 1 $, $ (m-\eta+1){m\choose{m-\eta+1}} > m $, thus showing the gap in succinctness. 
%\end{proof}

### 2022-01-27 chapters/interpretability/relaxed-cnf/incremental_maxsat.tex
relaxed-cnf paper added

\section{Incremental Approach for Efficient Learning }
In this section, we describe an efficient incremental approach for learning $ M $ of $ N $ rule. We have described a MaxSAT based optimal rule learning technique in Section ~\ref{sec:maxsat}. The size of the  MaxSAT query  (number of clause) in that technique is polynomial with the number of training samples, which results in poor scalability  for larger dataset.  Hence we propose an incremental learning technique via a partition based training methodology \cite{GM19}. In this context, we define $ \Rule_p $ to be the rule learned for the $ p $-th partition. The key idea of incremental learning is to  divide the training dataset into a fixed number of partitions, learn rule for each partition in a  linear order such that  $ \Rule_p $  depends on both the training samples in the $ p $-th partition and  on the  rule $ \Rule_{p-1} $ learned for the  $ (p-1) $-th partition.   We propose  the following modifications in the MaxSAT constraints $ Q $ for learning $ M $ of $ N $ rule in an incremental fashion.

\subsection{MaxSAT Constraints for Incremental Learning}
Let us consider that we want to learn rule $ \Rule_p $ for the $ p $-th partition and we have access to  $ \Rule_{p-1} $.  We want $ \Rule_p $ to not only generalize for the $ p $-th partition but also be consistent with $ \Rule_{p-1} $. Hence  we construct $ V_i^j $ as follows. If $ x_j \in \clause(\Rule_{p-1},i)  $ , we  deviate from the default behavior of falsifying $ b_j^i $ by adding a constraint to keep the corresponding literal \emph{true} in the optimal assignment. Otherwise we construct a soft clause to falsify $ b_j^i $. 
\[
V^i_j:=
\begin{cases}
b_j^i \quad \text{if }  x_j \in \clause(\Rule_{p-1},i)
\\
\neg b_j^i\quad \text{otherwise}
\end{cases}; \quad W(V_j^i)=1
\]

The second (Eq. ~\ref{eq:constraint_two}) and the third (Eq. ~\ref{eq:constraint_three}) constraint of $ Q $ do not require any modification in the  incremental approach of learning.

### 2022-01-27 chapters/interpretability/relaxed-cnf/introduction.tex
relaxed-cnf paper added

%%\vspace{-5pt}
\section{Introduction}
%%\vspace{-5pt}
\label{interpretability_crr_sec:introduction}

The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.  To enable  safe, robust and trustworthy integration of such systems, the end users require these systems to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
In this context,  rule-based representations are particularly effective for  presenting the decision functions to people~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability (defined in terms of rule-sparsity) with minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}. Although CNF/DNF rules are considered interpretable, they are less expressive compared to the Boolean cardinality constraints in propositional logic, wherein a Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.  
In this work, we introduce relaxed-CNF,  a classification rule in propositional logic that has the benefits of  both worlds: it represents the decision boundary in an interpretable manner similar to CNF/DNF  and it is more expressive  for allowing Boolean cardinality constraints in its representation.

 We find the motivation of designing relaxed-CNF rules from checklists, where a checklist is a list of conditions one needs to check, e.g., a list of items to take on a travel trip.  There are several practical applications of checklists in interpretable decision making~\cite{M1976,gage2001validation}. For example, the  CHADS2 score in medicine~\cite{gage2001validation}, that is a clinical prediction rule for estimating the risk of stroke.
Another example is a psychometric test, known as Myers–Briggs Type Indicator (MBTI)~\cite{M1976}, which indicates differing psychological preferences in how people
perceive the world around them and make decisions.  An influential study on the importance of {checklists}~\cite{G2010} finds that highly complex and specialized problems can be handled smoothly by the development and consistent usage of checklists. In this paper, we aim to learn  classification rules in the form of a checklist, which we call relaxed-CNF rules.

We now provide an introduction to relaxed-CNF rules. The simplest logical rules are single level-rules: ORs or ANDs of a subset of \textit{literals}, where a literal can be assigned  $ 1 $ (true) or  $ 0 $ (false). 
An OR operator requires $ 1 $ out of $ N $ literals to be assigned to  $ 1 $, while an AND operator requires all $N$ out of $N$ literals to be assigned to  $ 1 $.   A \emph{clause} is a collection of literals connected by  OR/AND.  A CNF   formula  is a conjunction (AND) of {clauses} where each clause is a disjunction (OR) of {literals}, while  a DNF  formula is a disjunction of clauses where each clause is a conjunction of literals. Therefore, CNF  and DNF  formulas can be viewed as two-level rules. Some popular interpretable rule-based models can be viewed as  CNF/DNF formulas. For example, a decision set is a DNF rule that refers to a set of ``if-else'' clauses~\cite{ignatiev2018sat,lakkaraju2016interpretable}. 
In this paper, we consider a richer set of logical formulas that capture the structure of checklists. We consider \emph{hard}-OR clauses, where at least $ M > 1$ out of $ N $ literals need to be $ 1 $, and we similarly define {\emph{soft}-AND} clauses which allow some of the literals (at most $ N-M $) to be $ 0 $. 
To be precise,  the definitions of hard-OR and soft-AND overlap. We use the term hard-OR when $M \le N/2$ and soft-AND otherwise.     
To extend the standard definition of CNF (which is ANDs of ORs),  we define \emph{relaxed}-CNF to denote soft-ANDs of hard-ORs. Similarly, \emph{relaxed}-DNF is hard-ORs of soft-ANDs. Since   hard-OR and soft-AND are differentiated based on the value of $ M $ and $ N $, relaxed-CNF and relaxed-DNF have a similar structural representation.
In early work, Craven and Shavlik~\cite{craven1996extracting} considered single level $ M $-of-$ N $ rules to explain black-box neural-network classifiers. Recently, Emad et al.~\cite{EVM2015} have developed a semi-quantitative group testing approach for learning sparse single level $M$-of-$N$ rules, which are quite restrictive in their ability to fit the data. In this work, we study a much richer family of two-level relaxed-CNF rules.

\begin{figure*}
	\begin{center}
		\includegraphics[scale=.7]{figures/interpretability/relaxed-cnf/checklist.pdf}
	\end{center}
	\caption{An illustrative example of a  relaxed CNF classification rule that describes the decision function in the form of a  two-level checklist. This classifier is learned on the WDBC (Wisconsin diagnostic breast cancer) dataset and predicts whether a tumor cell is malignant or not based on the characteristics of the  tumor cell. The first column in each checklist contains Boolean predicate/literals. An entry in the second column is  $ 1 $ if the corresponding predicate is true by an observed tumor cell, and $ 0 $ otherwise. In the first level, checklist A (resp. B) is true if the count of true  predicates is at least $ 2 $ (resp. $ 3 $). In the second level, a tumor cell is predicted as malignant if the count of true checklists is at least $ 1 $. }
	\label{interpretability_crr_exmpl:wdbc}
\end{figure*}

 {Relaxed-CNF rules are  more flexible than pure CNF rules and  can accurately fit more complex classification boundaries.} For example, relaxed-CNF clauses  allow  compact encoding of the majority function, which would require exponentially many clauses in CNF, showing the exponential gap in the succinctness of the two representations.  In addition, relaxed-CNF and CNF rules have the same functional form where a user has to compute the sum of true literals/clauses  and then compare the sum  to  different thresholds {(as in the example in Figure~\ref{interpretability_crr_exmpl:wdbc})}. From the computational perspective, the structural flexibility of relaxed-CNF rules compared to CNF/DNF rules makes them harder to learn, which poses the following question, \textit{can we design a combinatorial framework to efficiently learn relaxed-CNF rules?}

The primary contribution of this paper is an affirmative answer to the above question by proposing an efficient combinatorial framework for learning relaxed-CNF rules which we call {\crr} (\textbf{C}lassification \textbf{R}ules in \textbf{R}elaxed form).  In CRR, we construct an objective function to maximize both prediction accuracy and rule-sparsity and design a  Mixed-Integer Linear Programming (MILP) formulation for learning the optimal relaxed-CNF rules.
%We theoretically prove the succinctness of relaxed-CNF rules and propose a Mixed Integer Linear Programming (MILP)-based  approach for learning optimal rules.
To learn a $ k $-clause relaxed-CNF rule (say $ \Rule $) using the direct (naive) MILP formulation, the size of the MILP query expressed as the number of constraints is $ \mathcal{O}(n\cdot k) $, where the  number of features/attributes in the dataset is fixed and $ n $ is the number of training samples. Consequently, the naive MILP formulation fails to handle large datasets. To address scalability concerns,  we propose an efficient mini-batch training methodology based  incremental approach  for learning $ \Rule $. The proposed incremental approach significantly enhances the scalability of {\crr}. 
%In the incremental approach, we divide the datasets into $ \rho $ partitions and make $ \rho $ queries to ILP solvers with each query of $ \mathcal{O}(\frac{n}{\rho}\cdot k) $ constraints. This approach learns a relaxed-CNF rule for each partition in a  sequential order such that the ILP query constructed for the $ p $-th partition relies on the training samples in the $ p $-th partition and the rule learned for the $ (p-1)  $-th partition.

{Through a comprehensive experimental evaluation over datasets from the UCI and Kaggle repository, we observe that {\crr} with relaxed-CNF rules achieves an improved trade-off between accuracy and rule sparsity and scales to datasets with more than $ 10^5 $ samples. More significantly, {\crr} can generate relaxed-CNF rules which have higher accuracy than CNF rules generated by~\cite{ghosh19incremental}. Also, compared to decision lists generated by~\cite{C1995}, relaxed-CNF rules are sparser in large datasets.}
%can also reach the same level of accuracy compared to plain CNF/DNF rules with smaller relaxed-CNF rules. 
%
%We have conducted a comprehensive experimental study on the datasets from the UCI repository. Our empirical study demonstrates  an increase in the prediction accuracy for choosing relaxed-CNF rules over regular CNF rules. The incremental approach achieves runtime scalability without loss of accuracy.

\begin{comment}
\begin{example}
	\label{interpretability_crr_exmpl:wdbc}
	As an example, we illustrate a relaxed-CNF rule learned 
	for  Breast Cancer Wisconsin Diagnostic Dataset (WDBC) from UCI repository. 
	The dataset is used to predict whether the tumor is  malignant or benign based on the characteristics of  a tumor cell. We can learn the following rule for detecting malignant cancer. 
	%

	%\vspace{-13pt}
	\[
	\boxed{
		\begin{split}
		&\text{Tumor is diagnosed as malignant if, }\\
		&[ \{  (\text{ smoothness }\ge 0.089)   + (\text{standard error of area  }\ge 53.78)  + \\ & (\text{largest radius }\ge  18.225)   \}\ge 2  ] \; + \\&  [   \{(98.76\le \text{ perimeter } < 114.8)  +    (\text{largest smoothness }\ge  0.136) \\& 
		+ (105.95 \le \text{ largest perimeter }  < 117.45)    \}\ge 2  ]  \ge 1 \end{split}}
	\]

	The generated rule has two clauses presented within the square brackets. Each clause consists of three literals  (presented within the parentheses) and a comparator with a \textit{threshold  on literals} (shown outside the curly brace). For example, ``smoothness $ \ge  0.089  $'' is a literal/predicate and is assigned $ 1 $ (or true)  if a tumor cell satisfies this characteristic and $ 0 $ otherwise. Inside a clause, 
	we connect the literals  using ``$ + $'' to sum   the number of true  literals  and then compare the sum with the threshold, which is $ 2 $ in both clauses in the above example. A clause is assigned $ 1 $ (or true)  if the sum of true literals satisfies the threshold.
	Similarly, each clause is connected using ``$ + $'' to sum the number of true clauses in the rule. Finally, the rule is satisfied, i.e, a tumor is diagnosed as malignant, if the number of true clauses is at least equal to the \textit{threshold on clauses}, which is  $ 1 $ in this example. In this context, we choose to measure the size of the rule by the number of literals it contains, which in this case is   $ 3+3=6 $.   In this  example, the decision function is represented  as a two-level checklist, where one would  check items (i.e., literals/predicates) inside a checklist in the first level   and then check among the checklists (i.e., clauses) in the second level. 
\end{example}

\end{comment}

%KSM: Please feel free to comment out the text below if it exceeds page limit
The rest of the paper is organized as follows. We discuss related works in Section~\ref{interpretability_crr_sec:related_work}, introduce the notation and preliminaries in Section~\ref{interpretability_crr_sec:preliminaries} and formulate the problem in Section~\ref{interpretability_crr_sec:problem}. We describe the main contribution of this paper, $ \crr $,  in Section~\ref{interpretability_crr_sec:ilp} and extend to an incremental mini-batch learning in Section~\ref{interpretability_crr_sec:inc_ilp}. We then describe the experimental results in Section ~\ref{interpretability_crr_sec:experiment} and conclude in Section~\ref{interpretability_crr_sec:conclusion}.

### 2022-01-27 chapters/interpretability/relaxed-cnf/preliminaries.tex
relaxed-cnf paper added

%\vspace{-8pt}
\section{Preliminaries}
\label{interpretability_crr_sec:preliminaries}
We use capital  boldface letters such as $\mathbf{X}$ to denote matrices, while lower boldface letters $\mathbf{y}$ are reserved for vectors. For a matrix $\mathbf{X}$, $\mathbf{X}_i$  represents the $ i $-th row of $\mathbf{X}$ while for a vector $\mathbf{y}$, $y_i$ represents the $ i $-th element of $\mathbf{y}$. 
% For a vector/set $ \mathbf{u} $ we  use $ |\mathbf{u}| $ to denote the number of elements in $ \mathbf{u} $.

% introduce cardinality constraint in this part

Let $F$ be a Boolean formula and  $\mathbf{b} = \{b_1,b_2,\dots ,b_m \}$ be the set of Boolean propositional variables appearing in $F$. A literal $ v_i $ is a variable ($b_i$) or its complement ($\neg b_i$).  A \emph{satisfying assignment} or a \emph{witness} $ \sigma $ of $F$ is an assignment of variables in $\mathbf{b}$ that makes $F$ evaluate to $ 1 $ and is denoted as $ \sigma \models F $.  If $\sigma$ is an assignment of variables and $b_i \in \mathbf{b}$, we use $\sigma(b_i)$ to denote the value assigned to $b_i$ in $\sigma$. $F$ is in CNF if $F :=  \bigwedge_{i=1}^k C_i$ is conjunction (AND) of clauses, where each clause $C_i :=  \bigvee_j v_j $  is represented as a disjunction (OR) of literals.  {A hard OR (resp. soft AND) clause $ (C,\eta) $ has an extra parameter $ \eta $, where $ \eta $ is the threshold on the literals in $ C $.} Let $ \mathbb{1}[true]=1 $ and $ \mathbb{1}[false]=0 $. Motivated by checklists, we propose relaxed-CNF where in addition to $F$, we have two more parameters $\eta_c$ and $\eta_l$. We say that $(F,\eta_c, \eta_l)$ is in relaxed-CNF and $ \sigma $ is its witness if $\sigma \models (F,\eta_c, \eta_l)$ whenever $\sum_{i=1}^k \mathbb{1}[\sigma \models (C_i,\eta_l)] \ge \eta_c$ where $\sigma \models (C_i, \eta_l)$ iff $\sum_{v \in C_i} \mathbb{1}[\sigma \models v] \geq \eta_l$. Informally,  $\sigma$ satisfies a clause $(C_i, \eta_l)$ if  at least $\eta_l$ literals in $C_i$ are set to true by $\sigma$ and $\sigma$ satisfies $(F,\eta_c, \eta_l)$ if at least $\eta_c$ of $\{(C_i,\eta_l)\}$ are true. In the  example in Figure~\ref{interpretability_crr_exmpl:wdbc}, $ \eta_c=1 $ and $ \eta_l=2 $.

\begin{theorem}[\cite{benhamou1994two}]
	\label{interpretability_crr_thm:succinctness}	
	Let $ (C,\eta) $ be a clause in relaxed-CNF where  $ C $ has $ m  $ literals and $ \eta \in \{1,\dots,m\} $ is the threshold on literals.  An equivalent compact encoding of $ (C,\eta) $  into a CNF formula $ F= \bigwedge_i C_i $ requires $ {m \choose{m-\eta+1}} $  clauses where each clause is distinct and has $ m-\eta+1 $ literals of $ (C,\eta) $. Therefore,  the total number of literals in $ F $ is $ (m-\eta+1){m\choose{m-\eta+1}} = m $ when $ \eta=1 $, otherwise $ (m-\eta+1){m\choose{m-\eta+1}} > m $.
\end{theorem}

%\red{Relaxed-CNF rules generalize both CNF and DNF rules. }
%Properly define k-DNF, i.e., every clause has k literals. Every $-$DNF rule can be viewed as a relaxed-CNF rule with $\eta_c = 1$ and $\eta_l = k$. Similarly, every $k-$CNF rule can be viewed as a relaxed-CNF rule with $\eta_c=k$ 
% $ F $ is a relaxed-CNF if $ F:=[\sum_{i=1}^k C_i] \ge \eta_c $, where each clause $ C_i := [\sum_j l_j] \ge \eta_l $. Here $ \eta_c $  and $ \eta_l $ are two positive integer thresholds.

%We use $|C_i|$ to denote the number of literals in $C_i$.  
%A \emph{pseudo-boolean constraint} over literals is defined as $ \sum_i a_i\cdot l_i \ge \card; a_i,\card \in \mathbb{R} $.  The special case $ \sum_i l_i \ge \card $ is denoted as a \emph{cardinality constraint}. In the  case of cardinality constraint, $ \card $ is called a  cardinality constant and $ \card \in \mathbb{Z} ^+$. 
%A pseudo-boolean constraint can be encoded into a CNF formula by applying the state of the art encoding technique. 
%Therefore, a cardinality constraint is a variant of general  CNF clause. 
% Specifically,  a cardinality constraint is satisfied if at least $ \card $ literals in that clause are assigned to $ 1 $, which is different from a CNF clause where $ \card=1 $ is applied.

Between two vectors  $\mathbf{u}$ and $\mathbf{v}$ of the same size over Boolean variables/constants (e.g.,  $ 0 $, $ 1 $), we use $\mathbf{u} \circ \mathbf{v} $  to denote the inner product, i.e.,     $\mathbf{u} \circ \mathbf{v} = \sum_{i} (u_{i} \cdot v_{i})$, where $u_{i}$ and $v_{i}$ are a variable/constant  at the $i$-th index of $\mathbf{u}$ and $\mathbf{v}$, respectively. In this context,  the operator ``$\cdot$'' between a variable and a constant follows the standard interpretation, i.e., $0 \cdot b = 0$ and $1 \cdot b = b$.

%KSM: Define CNF and define relaxed-CNF before you start talking about clauses of CNF/relaxed-CNF. Also, define literal

We consider a standard binary classification problem, where we are given a collection of 
training samples $\{ (\mathbf{X}_i, y_i) \}$. For sample $ i $, $\mathbf{X}_i \in \{0,1\}^m$ contains the valuation of the feature vector $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$, and $y_i \in \{0,1\}$ is the binary class label.
A classifier {\Rule} is a mapping that takes in the feature vector $\mathbf{x}$ and returns a class $\hat{y} \in \{0,1\}$, i.e., $\hat{y} = \Rule(\mathbf{x})$. The goal is not 
only to design $\Rule$ to approximate the training set, but also to generalize to unseen samples arising from the same distribution. 
We focus on classifiers that can be expressed  as  relaxed-CNF. We use $\clause(\Rule,i)$ to denote  the  $i$-th clause of $\Rule$ and $ |\clause(\Rule,i)| $ to  denote the size of  $\clause(\Rule,i)$, which is measured as the number of literals in the $ i $-th clause. Furthermore, we use $|\Rule|$ to denote the rule-size of classifier $ \Rule $, which is defined as the number of literals in all the clauses, i.e., $|\Rule| = \Sigma_i |\clause(\Rule,i)|$. In the example in Figure~\ref{interpretability_crr_exmpl:wdbc}, $ |\Rule| = |\clause(\Rule, 1)| + |\clause(\Rule, 2)| = 3+3=6 $.

In this work, we consider the learning problem as an optimization problem, wherein we reduce the construction of $\Rule$ to computing assignments of appropriately designed variables.  An optimization problem consisting of Boolean variables can be solved using Integer Linear Programming (ILP) where each variable takes value either $ 0 $ or $ 1 $.\footnote{The problem can be viewed as either ILP or MaxSAT, but we obtained better performance from ILP solvers.} Given an objective function and a set of constraints comprising of variables with range $ \{0,1\} $, ILP finds an optimal assignment of variables that minimizes the objective function.
% We assume $ \sigma(\cdot) $ is a function over variables which outputs the assignment (either $ 0 $ or $ 1 $) of each variable and the task is to learn $ \sigma $ by solving an ILP query.  
% The inner knowledge of how ILP solves the problem, is not required to understand the paper.

%In our context, we define $ \Rule $ to be a set of important features which we aim to identify and $ \Rule \subseteq \mathbf{x} $. We use the notation $ |\Rule| $ to denote the count of features in $ \Rule $.For example, let $ \mathbf{x}=\{x_1,x_2,x_3,x_4\} $ and we learn  $ \Rule=\{x_1,x_3,x_4\} $. Therefore $ |\Rule|=3 $.

%We define two rules  $ \Rule_1 $ and $ \Rule_2 $ to be  equivalent  if $ \forall i,\Rule_1(\X_i)=\Rule_2(\X_i) $.

%We denote the set of all witnesses of $F$ by {\satisfying{F}}.

%In this work, we focus on the weighted variant of CNF wherein a weight function is defined over clauses. For a clause $C_i$ and weight function {\weight{\cdot}}, we use {\weight{C_i}} to denote the weight of clause $C_i$. We say that a clause $C_i$ is hard if $\weight{C_i} = \infty$, otherwise $C_i$ is called a soft clause.  To avoid notational clutter, we overload {\weight{\cdot}} to denote the weight of an assignment or clause, depending on the context. We define weight of an assignment $\sigma$ as the sum of weight of clauses that $\sigma$ does not satisfy. Formally, $\weight{\sigma} = \Sigma_{i | \sigma \not\models C_i} \weight{C_i}$.

%Given $F$ and weight function $\weight{\cdot}$, the problem of {MaxSAT} is to find an assignment $\sigma^*$ that has the minimum weight, i.e.  $\sigma^* = \MaxSAT(F,W)$ if $\forall \sigma \neq \sigma^*, \weight{\sigma*} \leq \weight{\sigma}$. Our formulation will have positive clause weights, hence MaxSAT corresponds to satisfying as many clauses as possible, and picking the strongest clauses among the unsatisfied ones. 
%Note that the above formulation is different from the typical definition of {\MaxSAT} but the difference is only syntactic. Borrowing terminology of community focused on developing {MaxSAT} solvers, we are solving a partial weighted {MaxSAT} instance wherein we mark all the clauses with $\infty$ weight as hard and clauses with other positive value less than $ \infty $ weight as soft  and ask for a solution that optimizes the partial weighted {MaxSAT} formula.  The knowledge of encoding a  cardinality constraint into CNF clauses, the  inner working of {MaxSAT} solvers,  and  the encoding of our representation into weighted {MaxSAT} is not required for this paper.

### 2022-01-27 chapters/interpretability/relaxed-cnf/preliminaries_prev.tex
relaxed-cnf paper added

\section{Preliminaries}
\label{sec:preliminaries}

 We use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters $\mathbf{y}$ are reserved for vectors/sets. For a matrix $\mathbf{X}$, $\mathbf{X}_i$  represents the $ i $-th row of $\mathbf{X}$ while for a vector/set $\mathbf{y}$, $y_i$ represents the $ i $-th element of $\mathbf{y}$. 
% For a vector/set $ \mathbf{u} $ we  use $ |\mathbf{u}| $ to denote the number of elements in $ \mathbf{u} $.

% introduce cardinality constraint in this part

Let $F$ be a Boolean formula and  $\mathbf{b} = \{b_1,b_2,\dots ,b_m \}$ be the set of variables appearing in $F$. A literal $ l_i $ is a variable ($b_i$) or its complement ($\neg b_i$).  A \emph{satisfying assignment} or a \emph{witness} of $F$ is an assignment of variables in $\mathbf{b}$ that makes $F$ evaluate to $ 1 $.  If $\sigma$ is an assignment of variables and $b_i \in \mathbf{b}$, we use $\sigma(b_i) : b_i \rightarrow \{0,1\}$ to denote the value assigned to $b_i$ in $\sigma$. $F$ is in Conjunctive Normal Form (CNF) if $F :=  \bigwedge_i C_i$, where each clause $C_i :=  \bigvee_j l_j $ is represented as disjunction of literals.

%We use $|C_i|$ to denote the number of literals in $C_i$.  
A \emph{pseudo-boolean constraint} over literals is defined as $ \sum_i a_i\cdot l_i \ge \card; a_i,\card \in \mathbb{R} $.  The special case $ \sum_i l_i \ge \card $ is denoted as a \emph{cardinality constraint}. In the  case of cardinality constraint, $ \card $ is called a  cardinality constant and $ \card \in \mathbb{Z} ^+$. 
%A pseudo-boolean constraint can be encoded into a CNF formula by applying the state of the art encoding technique. 
Therefore, a cardinality constraint is a variant of general  CNF clause. 
 Specifically,  a cardinality constraint is satisfied if at least $ \card $ literals in that clause are assigned to $ 1 $, which is different from a CNF clause where $ \card=1 $ is applied.

We define an operator $ \circ $ between  two vectors $\mathbf{u}$ and $\mathbf{v}$ over propositional variables or constants (e.g.,  $ 0 $, $ 1 $) as follows;    $\mathbf{u} \circ \mathbf{v} = \sum_{i} (u_{i} \wedge v_{i})$, where $u_{i}$ and $v_{i}$ denote a variable/constant  at the $i$-th index of $\mathbf{u}$ and $\mathbf{v}$ respectively. In this context, note that the operation $\wedge$ between a variable and a constant follows the standard interpretation, i.e. $0 \wedge b = 0$ and $1 \wedge b = b$.

We consider a standard binary classification problem, where we are given a collection of 
training samples $\{ \mathbf{X}_i, y_i \}$. Each vector $\mathbf{X}_i \in \mathcal{X}$ contains the valuation of the features $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$ for sample $i$, and $y_i \in \{0,1\}$ is the binary label for sample $i$.
A classifier {\Rule} is a mapping that takes in a feature vector $\mathbf{x}$ and return a class $\hat{y}$, i.e. $\hat{y} = \Rule(\mathbf{x})$. The goal is not 
only to design $\Rule$ to approximate our training set, but also to generalize to unseen samples arising from the same distribution. 
We focus on classifiers that can be expressed compactly in CNF, particularly \emph{conjunction of cardinality constraint}. {We use $\clause(\Rule,i)$ to denote the $i$-th clause of $\Rule$}. {Furthermore, we use $|\Rule|$ to denote the rule-size of classifier $ \Rule $ that is the sum of the count of literals in all the clauses, i.e. $|\Rule| = \Sigma_i |\clause(\Rule,i)|$}.

%In our context, we define $ \Rule $ to be a set of important features which we aim to identify and $ \Rule \subseteq \mathbf{x} $. We use the notation $ |\Rule| $ to denote the count of features in $ \Rule $.For example, let $ \mathbf{x}=\{x_1,x_2,x_3,x_4\} $ and we learn  $ \Rule=\{x_1,x_3,x_4\} $. Therefore $ |\Rule|=3 $.

%We define two rules  $ \Rule_1 $ and $ \Rule_2 $ to be  equivalent  if $ \forall i,\Rule_1(\X_i)=\Rule_2(\X_i) $.

%We denote the set of all witnesses of $F$ by {\satisfying{F}}.

In this work, we focus on the weighted variant of CNF wherein a weight function is defined over clauses. For a clause $C_i$ and 
weight function {\weight{\cdot}}, we use {\weight{C_i}} to denote the weight of clause $C_i$. We say that a clause $C_i$ is hard if $\weight{C_i} = \infty$, otherwise $C_i$ is called a soft clause.  To avoid notational clutter, we overload {\weight{\cdot}} to denote the weight of an assignment or clause, depending on the context. We define weight of an assignment $\sigma$ as the sum of weight of clauses that $\sigma$ does not satisfy. Formally, $\weight{\sigma} = \Sigma_{i | \sigma \not\models C_i} \weight{C_i}$.

Given $F$ and weight function $\weight{\cdot}$, the problem of {MaxSAT} is to find an assignment $\sigma^*$ that has the minimum weight, i.e.  $\sigma^* = \MaxSAT(F,W)$ if $\forall \sigma \neq \sigma^*, \weight{\sigma*} \leq \weight{\sigma}$. Our formulation will have positive clause weights, hence MaxSAT corresponds to satisfying as many clauses as possible, and picking the strongest clauses among the unsatisfied ones. 
%Note that the above formulation is different from the typical definition of {\MaxSAT} but the difference is only syntactic. 
Borrowing terminology of community focused on developing {MaxSAT} solvers, we are solving a partial weighted {MaxSAT} instance wherein we mark all the clauses with $\infty$ weight as hard and clauses with other positive value less than $ \infty $ weight as soft  and ask for a solution that optimizes the partial weighted {MaxSAT} formula.  The knowledge of encoding a  cardinality constraint into CNF clauses, the  inner working of {MaxSAT} solvers,  and  the encoding of our representation into weighted {MaxSAT} is not required for this paper.

### 2022-01-27 chapters/interpretability/relaxed-cnf/problem_formulation.tex
relaxed-cnf paper added

%\vspace{-2pt}
\section{Problem Formulation}
\label{interpretability_crr_sec:problem}
We now present the formal definition of the problem. Given  (i) an instance space   $ (\mathbf{X}, \mathbf{y})$, where $ \mathbf{X} \in \{0,1\}^{n\times m} $ is the feature matrix with $ m $ binary features and $ n $ samples,\footnote{$\mathbf{X}$ contains both the features and their complements as columns as in~\cite{MV2013}.} and $ \mathbf{y} \in \{0,1\}^n $ is the class label vector, (ii) {a positive integer $ \noclause $ indicating the number of clauses in the desired rule}, (iii)  an integer threshold on literals $ \card_{l} \in  \{0,\dots,m\} $ indicating  the  minimum number of literals required to be true  to satisfy a clause, (iv)  an  integer threshold on clauses $ \card_{c} \in  \{0,\dots,k\} $ indicating the minimum number of clauses required to be  true  to satisfy a formula, and  (v) the data-fidelity  parameter $ \lambda \in [0,1]$, we  learn  a classification rule $ \Rule $, that is expressed as a $ k $ clause relaxed-CNF formula.

% \[\hat{y}=\Rule(\mathbf{x})=I[\sum_{x_i\in\Rule, \Rule \subseteq \mathbf{x} } x_i \ge k]\]

Our goal is to find rules that balance two 
goals:   being accurate while also sparse to avoid over-fitting.  
To this end, we seek to minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $. In particular, suppose $ \Rule $ classifies all samples correctly, i.e., $ \forall i, y_i=\Rule(\mathbf{X}_i) $. Among all the rules that classify all samples correctly,  we choose the sparsest  such $ \Rule $:

%\vspace{-10pt}
\[
\min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y_i=\mathcal{R}(\mathbf{X}_i)
\]
%\vspace{-10pt}

% $ |\mathcal{R}|=\sum_{i=1}^k |C_i (\Rule)| $ is the  total number of literals appearing in $ \mathcal{R} $.

In practical classification tasks, perfect classification is very unusual. Hence, we need to balance rule-sparsity with prediction error.  Let $ \mathcal{E}_\mathcal{R} $   be  the set of samples which are misclassified  by $ \mathcal{R} $, i.e., $\mathcal{E}_\mathcal{R}=\{\mathbf{X}_i | y_i \ne \mathcal{R}(\mathbf{X}_i) \}$. Hence we aim to find $ \mathcal{R} $ as follows:\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g. to penalize  false-alarms more than mis-detects), and to allow instance weights (per sample).}

%\vspace{-12pt}
\[
\min\limits_{\mathcal{R}} \;  \frac{\lambda}{n}|\mathcal{E}_\mathcal{R}| + \frac{1-\lambda}{k\cdot m} |\mathcal{R}|  \text{ such that }\forall \mathbf{X}_i \in \mathcal{E}_\mathcal{R} ,y_i \ne \mathcal{R}(\mathbf{X}_i) .
\]
%\vspace{-13pt}

Each term in the  objective function is normalized in $ [0,1] $. 
The data-fidelity parameter $ \lambda \in [0,1] $  serves to balance
the trade-off between prediction accuracy and sparsity. Higher 
values of $ \lambda $ produce lower prediction errors by sacrificing the sparsity 
of $ \Rule $, and vice versa. It can be viewed as an inverse of the regularization 
parameter.

### 2022-01-27 chapters/interpretability/relaxed-cnf/related_work.tex
relaxed-cnf paper added

\section{Related Work}
\label{interpretability_crr_sec:related_work}
In the growing field of interpretable machine learning,  several rule-based systems such as decision trees~\cite{narodytska2018learning},  decision lists~\cite{R1987}, decision sets~\cite{ignatiev2018sat,lakkaraju2016interpretable} and classification rules~\cite{C1995,DGW2018} have been proposed over the years. More recently, a Bayesian framework has been adopted to learn rule sets~\cite{WRDLKM2017}, rule lists \cite{LRMM2015},  and falling rule lists~\cite{WR2015}. Building on the connection between rule learning and Boolean compression sensing,  a rule-based classification system was proposed to trade-off classification accuracy and representation size~\cite{MV2013}. Su et al.~\cite{SWVM2015} proposed an extension that allows two-level Boolean rules. In addition to designing classifiers that produce forms amenable to end-users, there is a large body of work to describe the working of opaque models (model-agnostic interpretability)~\cite{LKCL2019,lundberg2017unified,ribeiro2016should}.  While our work can also be applied to interpret black-box classifiers, we do not pursue this direction in this paper and leave for future work.

In recent work,  Ghosh and Meel~\cite{ghosh19incremental} propose an incremental approach for efficiently learning CNF classification rules using a MaxSAT-based framework. They adopt a sequential partition-based training methodology in incremental learning. Although the said approach achieves the runtime improvement,  the generated rules highly depend on the order/position of samples in the training dataset. In contrast, the incremental framework proposed in this work does not suffer from strong dependencies on the order of samples in the training dataset.

### 2022-01-27 chapters/interpretability/imli/abstract.tex
imli journal paper added

\section{abstract}
Machine learning has become omnipresent with applications in various safety-critical domains such as medical, law, and transportation. In these domains, high-stake decisions provided by machine learning necessitate researchers to design \emph{interpretable} models, where the prediction is understandable to a human. In interpretable machine learning, \emph{rule-based classifiers} are particularly effective in representing the decision boundary through a set of rules comprising input features. Examples of such classifiers include decision trees, decision lists, and decision sets. The interpretability of rule-based classifiers is in general related to the size of the rules, where smaller rules are considered more interpretable. To learn such a classifier, the brute-force direct approach is to consider an optimization problem 
that tries to learn the smallest classification rule that has close to maximum accuracy. This optimization problem is computationally intractable due to its combinatorial nature and thus, the problem is not scalable in large datasets. To this end, in this paper we study the triangular relationship among the accuracy, interpretability, and scalability of learning rule-based classifiers.

The contribution of this paper is an interpretable learning framework {\imli}, that is based on maximum satisfiability (MaxSAT) for synthesizing classification rules expressible in proposition logic. {\imli} considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns an optimal rule by solving an appropriately designed MaxSAT query. Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to practical classification datasets containing thousands to millions of samples. Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning. The resulting framework learns a classifier by iteratively covering the training data, wherein in each iteration, it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch. In our experiments, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. For instance, {\imli} attains a competitive prediction accuracy and interpretability w.r.t.\ existing interpretable classifiers and  demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail. As an application, we deploy {\imli} in learning popular interpretable classifiers such as decision lists and decision sets.

### 2022-01-27 chapters/interpretability/imli/appendix.tex
imli journal paper added

\section*{Appendix}

In the following, we present representative CNF classifiers learned in different datasets. In each dataset, if an input satisfies the CNF formula, it is predicted class $ 1 $ and vice-versa.

\noindent \textbf{ Parkinsons: }\\

\noindent 0.3 $\le$ Average vocal fundamental frequency $<$ 0.4 \textbf{OR} 0.5 $\le$ Average vocal fundamental frequency $<$ 0.6 \textbf{OR} 0.1 $\le$ Maximum vocal fundamental frequency $<$ 0.2 \textbf{OR} 0.5 $\le$ Minimum vocal fundamental frequency $<$ 0.6 \textbf{OR} 0.1 $\le$ Shimmer:APQ5 $<$ 0.2 \textbf{OR} 0.5 $\le$ DFA $<$ 0.6 \textbf{OR} 0.3 $\le$ spread1 $<$ 0.4 \textbf{OR} 0.8 $\le$ spread2 $<$ 0.9 \textbf{OR} 0.3 $\le$ PPE $<$ 0.4 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ MDVP:APQ $<$ 0.1 \blue{\textbf{AND}}\\\\ \textbf{NOT} 0.8 $\le$ Average vocal fundamental frequency $<$ 0.9 \\

\noindent \textbf{ WDBC: }\\

\noindent 0.4 $\le$ perimeter $<$ 0.5 \textbf{OR} 0.8 $\le$ symmetry $<$ 0.9 \textbf{OR} 0.5 $\le$ largest concave points $<$ 0.6 \textbf{OR} 0.6 $\le$ largest concave points $<$ 0.7 \textbf{OR} 0.7 $\le$ largest concave points $<$ 0.8 \textbf{OR} 0.6 $\le$ largest symmetry $<$ 0.7 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ area SE $<$ 0.1 \blue{\textbf{AND}}\\\\0.4 $\le$ texture $<$ 0.5 \textbf{OR} 0.5 $\le$ texture $<$ 0.6 \textbf{OR} 0.5 $\le$ perimeter $<$ 0.6 \textbf{OR} 0.5 $\le$ smoothness $<$ 0.6 \textbf{OR} 0.4 $\le$ concave points $<$ 0.5 \textbf{OR} 0.5 $\le$ concave points $<$ 0.6 \textbf{OR} 0.2 $\le$ largest concavity $<$ 0.3 \textbf{OR} 0.6 $\le$ largest concave points $<$ 0.7 \textbf{OR} 0.7 $\le$ largest concave points $<$ 0.8 \textbf{OR} 0.4 $\le$ largest symmetry $<$ 0.5 \textbf{OR} 0.6 $\le$ largest symmetry $<$ 0.7 \\

\noindent \textbf{ Pima: }\\

\noindent x1 = 11 \textbf{OR} x1 = 14 \textbf{OR} x1 = 15 \textbf{OR} 0.7 $\le$ x2 $<$ 0.8 \textbf{OR} 0.8 $\le$ x2 $<$ 0.9 \textbf{OR} 0.9 $\le$ x2 $<$ $\infty$ \textbf{OR} 0.9 $\le$ x3 $<$ $\infty$ \textbf{OR} 0.4 $\le$ x5 $<$ 0.5 \textbf{OR} 0.7 $\le$ x5 $<$ 0.8 \textbf{OR} 0.7 $\le$ x6 $<$ 0.8 \textbf{OR} 0.4 $\le$ x7 $<$ 0.5 \textbf{OR} 0.5 $\le$ x7 $<$ 0.6 \textbf{OR} 0.9 $\le$ x7 $<$ $\infty$ \blue{\textbf{AND}}\\\\x1 = 7 \textbf{OR} x1 = 8 \textbf{OR} 0.6 $\le$ x2 $<$ 0.7 \textbf{OR} 0.8 $\le$ x2 $<$ 0.9 \textbf{OR} 0.9 $\le$ x2 $<$ $\infty$ \textbf{OR} -$\infty$ $\le$ x3 $<$ 0.1 \textbf{OR} 0.7 $\le$ x3 $<$ 0.8 \textbf{OR} 0.9 $\le$ x3 $<$ $\infty$ \textbf{OR} 0.2 $\le$ x5 $<$ 0.3 \textbf{OR} 0.7 $\le$ x6 $<$ 0.8 \textbf{OR} 0.1 $\le$ x7 $<$ 0.2 \textbf{OR} 0.3 $\le$ x7 $<$ 0.4 \textbf{OR} 0.4 $\le$ x8 $<$ 0.5 \textbf{OR} 0.8 $\le$ x8 $<$ 0.9 \blue{\textbf{AND}}\\\\x1 = 2 \textbf{OR} x1 = 6 \textbf{OR} x1 = 7 \textbf{OR} x1 = 9 \textbf{OR} -$\infty$ $\le$ x3 $<$ 0.1 \textbf{OR} 0.8 $\le$ x3 $<$ 0.9 \textbf{OR} 0.1 $\le$ x5 $<$ 0.2 \textbf{OR} 0.2 $\le$ x5 $<$ 0.3 \textbf{OR} 0.7 $\le$ x5 $<$ 0.8 \textbf{OR} 0.5 $\le$ x6 $<$ 0.6 \textbf{OR} 0.6 $\le$ x6 $<$ 0.7 \textbf{OR} 0.4 $\le$ x7 $<$ 0.5 \textbf{OR} 0.9 $\le$ x7 $<$ $\infty$ \textbf{OR} 0.1 $\le$ x8 $<$ 0.2 \textbf{OR} 0.3 $\le$ x8 $<$ 0.4 \textbf{OR} 0.6 $\le$ x8 $<$ 0.7 \textbf{OR} 0.8 $\le$ x8 $<$ 0.9 \blue{\textbf{AND}}\\\\x1 = 8 \textbf{OR} 0.5 $\le$ x2 $<$ 0.6 \textbf{OR} 0.8 $\le$ x2 $<$ 0.9 \textbf{OR} 0.9 $\le$ x2 $<$ $\infty$ \textbf{OR} 0.7 $\le$ x3 $<$ 0.8 \textbf{OR} -$\infty$ $\le$ x4 $<$ 0.1 \textbf{OR} 0.1 $\le$ x5 $<$ 0.2 \textbf{OR} 0.3 $\le$ x5 $<$ 0.4 \textbf{OR} 0.5 $\le$ x7 $<$ 0.6 \textbf{OR} 0.1 $\le$ x8 $<$ 0.2 \textbf{OR} 0.4 $\le$ x8 $<$ 0.5 \textbf{OR} 0.8 $\le$ x8 $<$ 0.9 \\

\noindent \textbf{ Titanic: }\\

\noindent -$\infty$ $\le$ age $<$ 0.1 \textbf{OR} 0.9 $\le$ age $<$ $\infty$ \textbf{OR} 0.9 $\le$ fare $<$ $\infty$ \textbf{OR}  \textbf{NOT} sex \blue{\textbf{AND}}\\\\passenger-class = 2 \textbf{OR} 0.4 $\le$ age $<$ 0.5 \textbf{OR} 0.6 $\le$ age $<$ 0.7 \textbf{OR} siblings-or-spouces-aboard = 0 \textbf{OR} 0.1 $\le$ fare $<$ 0.2 \textbf{OR} 0.5 $\le$ fare $<$ 0.6 \textbf{OR} embarked = C \blue{\textbf{AND}}\\\\-$\infty$ $\le$ age $<$ 0.1 \textbf{OR} 0.1 $\le$ age $<$ 0.2 \textbf{OR} 0.7 $\le$ age $<$ 0.8 \textbf{OR} siblings-or-spouces-aboard = 1 \textbf{OR} parents-or-childred-aboard = 1 \textbf{OR} embarked = C \textbf{OR}  \textbf{NOT} passenger-class = 3 \blue{\textbf{AND}}\\\\-$\infty$ $\le$ age $<$ 0.1 \textbf{OR} 0.2 $\le$ age $<$ 0.3 \textbf{OR} 0.3 $\le$ age $<$ 0.4 \textbf{OR} parents-or-childred-aboard = 0 \textbf{OR}  \textbf{NOT} passenger-class = 3 \textbf{OR}  \textbf{NOT} siblings-or-spouces-aboard = 0 \blue{\textbf{AND}}\\\\ \textbf{NOT} passenger-class = 2 \textbf{OR}  \textbf{NOT} 0.7 $\le$ age $<$ 0.8 \\

\noindent \textbf{ MAGIC: }\\

\noindent -$\infty$ $\le$ length $<$ 0.1 \textbf{OR} -$\infty$ $\le$ size $<$ 0.1 \textbf{OR} 0.8 $\le$ conc $<$ 0.9 \textbf{OR} -$\infty$ $\le$ alpha $<$ 0.1 \textbf{OR} 0.1 $\le$ alpha $<$ 0.2 \textbf{OR} 0.2 $\le$ dist $<$ 0.3 \\

\noindent \textbf{ Tom's HW: }\\

\noindent 0.4 $\le$ x10 $<$ 0.5 \textbf{OR} 0.1 $\le$ x11 $<$ 0.2 \textbf{OR} 0.4 $\le$ x11 $<$ 0.5 \textbf{OR} 0.4 $\le$ x12 $<$ 0.5 \textbf{OR} 0.3 $\le$ x13 $<$ 0.4 \textbf{OR} 0.6 $\le$ x13 $<$ 0.7 \textbf{OR} 0.3 $\le$ x15 $<$ 0.4 \textbf{OR} 0.6 $\le$ x15 $<$ 0.7 \textbf{OR} 0.4 $\le$ x16 $<$ 0.5 \textbf{OR} 0.1 $\le$ x57 $<$ 0.2 \textbf{OR} 0.5 $\le$ x59 $<$ 0.6 \textbf{OR} 0.3 $\le$ x60 $<$ 0.4 \textbf{OR} 0.2 $\le$ x62 $<$ 0.3 \textbf{OR} 0.2 $\le$ x73 $<$ 0.3 \textbf{OR} 0.1 $\le$ x74 $<$ 0.2 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x74 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x96 $<$ 0.1 \blue{\textbf{AND}}\\\\0.4 $\le$ x9 $<$ 0.5 \textbf{OR} 0.8 $\le$ x9 $<$ 0.9 \textbf{OR} 0.2 $\le$ x12 $<$ 0.3 \textbf{OR} 0.3 $\le$ x12 $<$ 0.4 \textbf{OR} 0.3 $\le$ x15 $<$ 0.4 \textbf{OR} 0.6 $\le$ x15 $<$ 0.7 \textbf{OR} 0.1 $\le$ x60 $<$ 0.2 \textbf{OR} 0.1 $\le$ x61 $<$ 0.2 \textbf{OR} 0.6 $\le$ x78 $<$ 0.7 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x11 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x13 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x46 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x64 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x72 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x77 $<$ 0.1 \blue{\textbf{AND}}\\\\-$\infty$ $\le$ x50 $<$ 0.1 \textbf{OR} 0.1 $\le$ x57 $<$ 0.2 \blue{\textbf{AND}}\\\\-$\infty$ $\le$ x49 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x17 $<$ 0.1 \blue{\textbf{AND}}\\\\0.3 $\le$ x14 $<$ 0.4 \textbf{OR} 0.6 $\le$ x74 $<$ 0.7 \textbf{OR} 0.9 $\le$ x79 $<$ $\infty$ \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x9 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x15 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x73 $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ x80 $<$ 0.1 \\

\noindent \textbf{ Credit: }\\

\noindent Repayment-status-in-September = 2 \textbf{OR} Repayment-status-in-September = 3 \textbf{OR} Repayment-status-in-August = 3 \textbf{OR} Repayment-status-in-May = 3 \textbf{OR} Repayment-status-in-May = 5 \textbf{OR} Repayment-status-in-April = 3 \blue{\textbf{AND}}\\\\0.4 $\le$ Age $<$ 0.5 \textbf{OR} Repayment-status-in-September = 1 \textbf{OR} Repayment-status-in-August = 0 \textbf{OR} Repayment-status-in-July = 2 \textbf{OR} Repayment-status-in-May = 2 \textbf{OR} Repayment-status-in-April = 7 \textbf{OR} 0.3 $\le$ Amount-of-bill-statement-in-April $<$ 0.4 \textbf{OR}  \textbf{NOT} Repayment-status-in-May = -1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ Amount-of-bill-statement-in-May $<$ 0.1 \blue{\textbf{AND}}\\\\Gender \textbf{OR} -$\infty$ $\le$ Age $<$ 0.1 \textbf{OR} 0.3 $\le$ Age $<$ 0.4 \textbf{OR} Repayment-status-in-September = 1 \textbf{OR} 0.1 $\le$ Amount-of-bill-statement-in-August $<$ 0.2 \textbf{OR}  \textbf{NOT} Education = 3 \textbf{OR}  \textbf{NOT} 0.2 $\le$ Amount-of-bill-statement-in-September $<$ 0.3 \\

\noindent \textbf{ Adult: }\\

\noindent education =  Doctorate \textbf{OR} education =  Prof-school \textbf{OR} 0.1 $\le$ capital-gain $<$ 0.2 \textbf{OR} 0.2 $\le$ capital-gain $<$ 0.3 \textbf{OR} 0.9 $\le$ capital-gain $<$ $\infty$ \textbf{OR} 0.4 $\le$ capital-loss $<$ 0.5 \textbf{OR} 0.5 $\le$ capital-loss $<$ 0.6 \blue{\textbf{AND}}\\\\relationship =  Husband \textbf{OR} relationship =  Wife \textbf{OR} 0.5 $\le$ capital-loss $<$ 0.6 \textbf{OR} 0.6 $\le$ capital-loss $<$ 0.7 \textbf{OR} 0.8 $\le$ capital-loss $<$ 0.9 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ capital-gain $<$ 0.1 \blue{\textbf{AND}}\\\\0.1 $\le$ age $<$ 0.2 \textbf{OR} 0.7 $\le$ age $<$ 0.8 \textbf{OR} workclass =  State-gov \textbf{OR} education =  Bachelors \textbf{OR} marital-status =  Separated \textbf{OR} occupation =  Exec-managerial \textbf{OR} occupation =  Farming-fishing \textbf{OR} occupation =  Prof-specialty \textbf{OR} occupation =  Protective-serv \textbf{OR} occupation =  Tech-support \textbf{OR} relationship =  Unmarried \textbf{OR} 0.4 $\le$ capital-loss $<$ 0.5 \textbf{OR} 0.6 $\le$ capital-loss $<$ 0.7 \textbf{OR} 0.2 $\le$ hours-per-week $<$ 0.3 \textbf{OR} 0.4 $\le$ hours-per-week $<$ 0.5 \textbf{OR} 0.7 $\le$ hours-per-week $<$ 0.8 \textbf{OR} 0.9 $\le$ hours-per-week $<$ $\infty$ \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ capital-gain $<$ 0.1 \\

\noindent \textbf{ Bank Marketing: }\\

\noindent 0.7 $\le$ age $<$ 0.8 \textbf{OR} 0.2 $\le$ duration $<$ 0.3 \textbf{OR} 0.3 $\le$ duration $<$ 0.4 \textbf{OR} 0.4 $\le$ duration $<$ 0.5 \textbf{OR} 0.5 $\le$ duration $<$ 0.6 \textbf{OR} poutcome = success \blue{\textbf{AND}}\\\\housing \textbf{OR} 0.8 $\le$ age $<$ 0.9 \textbf{OR} job = admin. \textbf{OR} job = management \textbf{OR} job = self-employed \textbf{OR} job = services \textbf{OR} job = unemployed \textbf{OR} 0.2 $\le$ duration $<$ 0.3 \textbf{OR} 0.1 $\le$ campaign $<$ 0.2 \textbf{OR} poutcome = success \blue{\textbf{AND}}\\\\job = services \textbf{OR} 0.2 $\le$ balance $<$ 0.3 \textbf{OR} contact = cellular \textbf{OR} 0.1 $\le$ duration $<$ 0.2 \textbf{OR} 0.2 $\le$ duration $<$ 0.3 \textbf{OR} 0.2 $\le$ pdays $<$ 0.3 \textbf{OR} poutcome = unknown \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ balance $<$ 0.1 \textbf{OR}  \textbf{NOT} -$\infty$ $\le$ previous $<$ 0.1 \blue{\textbf{AND}}\\\\0.2 $\le$ age $<$ 0.3 \textbf{OR} 0.3 $\le$ age $<$ 0.4 \textbf{OR} 0.8 $\le$ age $<$ 0.9 \textbf{OR} job = management \textbf{OR} job = student \textbf{OR} job = unemployed \textbf{OR} education = secondary \textbf{OR} 0.1 $\le$ balance $<$ 0.2 \textbf{OR} 0.1 $\le$ duration $<$ 0.2 \textbf{OR} -$\infty$ $\le$ pdays $<$ 0.1 \textbf{OR} 0.2 $\le$ pdays $<$ 0.3 \textbf{OR} poutcome = unknown \\

\noindent \textbf{ Connect-4: }\\

\noindent b2 = 1 \textbf{OR} c2 = 1 \textbf{OR} d2 = 1 \textbf{OR} d4 = 1 \textbf{OR} d5 = 0 \textbf{OR} e2 = 1 \textbf{OR} f3 = 0 \textbf{OR}  \textbf{NOT} d1 = 0 \blue{\textbf{AND}}\\\\b2 = 1 \textbf{OR} b3 = 1 \textbf{OR} b4 = 1 \textbf{OR} d4 = 1 \textbf{OR} f2 = 1 \textbf{OR}  \textbf{NOT} d3 = 0 \textbf{OR}  \textbf{NOT} f3 = 2 \blue{\textbf{AND}}\\\\c1 = 2 \textbf{OR} c2 = 2 \textbf{OR} c3 = 1 \textbf{OR} c3 = 2 \textbf{OR} c4 = 1 \textbf{OR} c6 = 0 \textbf{OR} d2 = 1 \textbf{OR} d4 = 1 \textbf{OR} e2 = 1 \textbf{OR} e3 = 1 \textbf{OR} f4 = 1 \textbf{OR}  \textbf{NOT} a3 = 2 \textbf{OR}  \textbf{NOT} b6 = 2 \blue{\textbf{AND}}\\\\a1 = 0 \textbf{OR} a2 = 0 \textbf{OR} a6 = 0 \textbf{OR} b2 = 1 \textbf{OR} b4 = 1 \textbf{OR} b5 = 0 \textbf{OR} c2 = 1 \textbf{OR} c4 = 1 \textbf{OR} c5 = 0 \textbf{OR} d1 = 1 \textbf{OR} d2 = 1 \textbf{OR} e2 = 1 \textbf{OR} g1 = 0 \textbf{OR} g2 = 0 \textbf{OR}  \textbf{NOT} d5 = 2 \textbf{OR}  \textbf{NOT} d6 = 2 \blue{\textbf{AND}}\\\\b2 = 1 \textbf{OR} b4 = 1 \textbf{OR} c3 = 1 \textbf{OR} d3 = 1 \textbf{OR} e2 = 1 \textbf{OR} f2 = 1 \textbf{OR} f3 = 1 \textbf{OR} g3 = 0 \textbf{OR} g5 = 0 \textbf{OR}  \textbf{NOT} c2 = 0 \textbf{OR}  \textbf{NOT} c4 = 2 \textbf{OR}  \textbf{NOT} d2 = 2 \\

\noindent \textbf{ Weather AUS: }\\

\noindent 0.7 $\le$ Rainfall $<$ 0.8 \textbf{OR} 0.8 $\le$ Humidity3pm $<$ 0.9 \textbf{OR} 0.9 $\le$ Humidity3pm $<$ $\infty$ \blue{\textbf{AND}}\\\\RainToday \textbf{OR} 0.5 $\le$ WindGustSpeed $<$ 0.6 \textbf{OR} 0.7 $\le$ Humidity9am $<$ 0.8 \textbf{OR} 0.9 $\le$ Humidity3pm $<$ $\infty$ \textbf{OR} 0.4 $\le$ Pressure3pm $<$ 0.5 \textbf{OR}  \textbf{NOT} 0.9 $\le$ Humidity9am $<$ $\infty$ \blue{\textbf{AND}}\\\\0.1 $\le$ MinTemp $<$ 0.2 \textbf{OR} 0.8 $\le$ MinTemp $<$ 0.9 \textbf{OR} 0.9 $\le$ WindSpeed9am $<$ $\infty$ \textbf{OR} 0.9 $\le$ Humidity9am $<$ $\infty$ \textbf{OR} 0.8 $\le$ Humidity3pm $<$ 0.9 \textbf{OR} 0.9 $\le$ Humidity3pm $<$ $\infty$ \textbf{OR} 0.5 $\le$ Temp3pm $<$ 0.6 \textbf{OR}  \textbf{NOT} 0.7 $\le$ Rainfall $<$ 0.8 \blue{\textbf{AND}}\\\\WindGustDir = NNW \textbf{OR} WindDir3pm = W \textbf{OR}  \textbf{NOT} 0.8 $\le$ Temp3pm $<$ 0.9 \blue{\textbf{AND}}\\\\WindGustDir = NNW \textbf{OR} WindGustDir = NW \textbf{OR} 0.1 $\le$ Pressure9am $<$ 0.2 \textbf{OR}  \textbf{NOT} 0.7 $\le$ Temp3pm $<$ 0.8 \\

\noindent \textbf{ Vote: }\\

\noindent  \textbf{NOT} physician-fee-freeze \blue{\textbf{AND}}\\\\ \textbf{NOT} adoption-of-the-budget-resolution \textbf{OR}  \textbf{NOT} anti-satellite-test-ban \textbf{OR}  \textbf{NOT} synfuels-corporation-cutback \blue{\textbf{AND}}\\\\adoption-of-the-budget-resolution \textbf{OR} el-salvador-aid \textbf{OR}  \textbf{NOT} duty-free-exports \blue{\textbf{AND}}\\\\mx-missile \textbf{OR}  \textbf{NOT} adoption-of-the-budget-resolution \textbf{OR}  \textbf{NOT} el-salvador-aid \textbf{OR}  \textbf{NOT} anti-satellite-test-ban \blue{\textbf{AND}}\\\\adoption-of-the-budget-resolution \textbf{OR} mx-missile \textbf{OR}  \textbf{NOT} synfuels-corporation-cutback \textbf{OR}  \textbf{NOT} education-spending \\

\noindent \textbf{ Skin Seg: }\\

\noindent 0.2 $\le$ Red $<$ 0.3 \textbf{OR} 0.3 $\le$ Red $<$ 0.4 \textbf{OR} 0.4 $\le$ Red $<$ 0.5 \textbf{OR} 0.9 $\le$ Red $<$ $\infty$ \textbf{OR} 0.9 $\le$ Green $<$ $\infty$ \textbf{OR} 0.4 $\le$ Blue $<$ 0.5 \textbf{OR}  \textbf{NOT} 0.9 $\le$ Blue $<$ $\infty$ \blue{\textbf{AND}}\\\\0.2 $\le$ Red $<$ 0.3 \textbf{OR} 0.7 $\le$ Red $<$ 0.8 \textbf{OR} 0.8 $\le$ Red $<$ 0.9 \textbf{OR} 0.9 $\le$ Red $<$ $\infty$ \textbf{OR}  \textbf{NOT} 0.8 $\le$ Blue $<$ 0.9 \blue{\textbf{AND}}\\\\0.7 $\le$ Red $<$ 0.8 \textbf{OR} 0.8 $\le$ Red $<$ 0.9 \textbf{OR} 0.9 $\le$ Red $<$ $\infty$ \textbf{OR} -$\infty$ $\le$ Green $<$ 0.1 \textbf{OR}  \textbf{NOT} 0.7 $\le$ Blue $<$ 0.8 \\

\noindent \textbf{ BNG(labor): }\\

\noindent 0.5 $\le$ wage-increase-first-year $<$ 0.6 \textbf{OR} pension = empl-contr \textbf{OR} contribution-to-dental-plan = full \blue{\textbf{AND}}\\\\0.6 $\le$ wage-increase-first-year $<$ 0.7 \textbf{OR} 0.7 $\le$ wage-increase-second-year $<$ 0.8 \textbf{OR} 0.5 $\le$ shift-differential $<$ 0.6 \textbf{OR}  \textbf{NOT} longterm-disability-assistance \blue{\textbf{AND}}\\\\0.2 $\le$ wage-increase-first-year $<$ 0.3 \textbf{OR} 0.3 $\le$ wage-increase-first-year $<$ 0.4 \textbf{OR} 0.4 $\le$ wage-increase-first-year $<$ 0.5 \textbf{OR} 0.5 $\le$ wage-increase-first-year $<$ 0.6 \textbf{OR} 0.6 $\le$ wage-increase-first-year $<$ 0.7 \textbf{OR} 0.7 $\le$ wage-increase-first-year $<$ 0.8 \textbf{OR} 0.6 $\le$ wage-increase-second-year $<$ 0.7 \textbf{OR} cost-of-living-adjustment = tcf \textbf{OR} 0.3 $\le$ working-hours $<$ 0.4 \textbf{OR} 0.4 $\le$ working-hours $<$ 0.5 \textbf{OR} 0.5 $\le$ working-hours $<$ 0.6 \textbf{OR} pension = ret-allw \textbf{OR} 0.7 $\le$ standby-pay $<$ 0.8 \textbf{OR} contribution-to-dental-plan = full \textbf{OR} contribution-to-health-plan = half \textbf{OR}  \textbf{NOT} education-allowance \blue{\textbf{AND}}\\\\0.5 $\le$ shift-differential $<$ 0.6 \textbf{OR} contribution-to-dental-plan = full \textbf{OR} contribution-to-dental-plan = half \textbf{OR} contribution-to-health-plan = half \textbf{OR} contribution-to-health-plan = none \blue{\textbf{AND}}\\\\0.3 $\le$ wage-increase-first-year $<$ 0.4 \textbf{OR} 0.6 $\le$ wage-increase-first-year $<$ 0.7 \textbf{OR} 0.7 $\le$ wage-increase-second-year $<$ 0.8 \textbf{OR} 0.4 $\le$ working-hours $<$ 0.5 \textbf{OR} pension = empl-contr \textbf{OR} pension = ret-allw \textbf{OR} 0.9 $\le$ statutory-holidays $<$ inf \\

\noindent \textbf{ BNG(credit-g): }\\

\noindent foreign-worker \textbf{OR} checking-status = 'no checking' \textbf{OR} checking-status = >=200 \textbf{OR} -$\infty$ $\le$ duration $<$ 0.1 \textbf{OR} 0.1 $\le$ duration $<$ 0.2 \textbf{OR} credit-history = 'critical/other existing credit' \textbf{OR} purpose = 'used car' \textbf{OR} savings-status = 'no known savings' \textbf{OR} savings-status = >=1000 \textbf{OR} other-parties = guarantor \blue{\textbf{AND}}\\\\foreign-worker \textbf{OR} checking-status = 'no checking' \textbf{OR} checking-status = >=200 \textbf{OR} credit-history = 'delayed previously' \textbf{OR} purpose = 'used car' \textbf{OR} purpose = radio/tv \textbf{OR} other-parties = guarantor \textbf{OR} 0.4 $\le$ age $<$ 0.5 \textbf{OR}  \textbf{NOT} other-payment-plans = bank \blue{\textbf{AND}}\\\\checking-status = 'no checking' \textbf{OR} credit-history = 'critical/other existing credit' \textbf{OR} purpose = 'used car' \textbf{OR} purpose = radio/tv \textbf{OR} purpose = retraining \textbf{OR} employment = 4$\le$X$<$7 \textbf{OR} property-magnitude = 'real estate' \textbf{OR}  \textbf{NOT} checking-status = $<$0 \\

### 2022-01-27 chapters/interpretability/imli/baseline.tex
imli journal paper added

\section{MaxSAT-based Interpretable Classification Rule Learning}
\label{interpretability_imli_sec:baseline}
In this section, we discuss a MaxSAT-based framework for optimal learning of an interpretable rule-based classifier, particularly a CNF classifier $ \Rule $. 
We first describe the decision variables in Section~\ref{interpretability_imli_sec:variables} and present the MaxSAT encoding  in Section~\ref{interpretability_imli_sec:encoding}. Our MaxSAT formulation assumes binary features as input. We
conclude this section by learning $ \Rule $ with non-binary features in Section~\ref{interpretability_imli_sec:non-binary} and discussing more flexible interpretability constraints of $ \Rule $ in Section~\ref{interpretability_imli_sec:complex_interpretability_objectives}.

\subsection{Description of Variables}
\label{interpretability_imli_sec:variables} 
We initially preprocess $ \mathbf{X} $  to account for the negation of  Boolean features while learning a classifier.\footnote{This preprocessing is similarly applied in  \cite{malioutov2013exact}.} In the preprocessing step, we negate each column in $ \mathbf{X} $ to a new column and append it to $ \mathbf{X} $. For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{X} $ by negating the column ``$ \text{age }\ge 25 $''. Hence, in the rest of the paper, we refer $ m $ as the modified number of columns in $ \mathbf{X} $. We next discuss the variables in the MaxSAT problem.

We consider two types of Boolean variables: (i) \emph{feature} variables $ b $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples. We define a Boolean variable $ b^i_j $ that becomes true if feature $ x_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ b^i_j $ is assigned false, otherwise. Moreover, we define an error variable $ \eta_l $ to attribute to whether the sample $ \mathbf{X}_l $ is classified correctly or not. Specifically, $ \eta_l $ becomes true if $ \mathbf{X}_l $ is misclassified, and becomes false otherwise.  We next discuss the MaxSAT encoding to solve the classification problem.

\subsection{MaxSAT Encoding}
\label{interpretability_imli_sec:encoding}
We consider a partial-weighted MaxSAT formula, where we encode the objective function in Eq.~\eqref{interpretability_imli_eq:obj} as soft clauses. For each sample in $ \mathbf{X} $, we construct hard clauses specifying that the sample is either  classified correctly and $ \eta_l $ is false or it is misclassified and $ \eta_l $ is true. We next discuss the MaxSAT encoding in detail.

\begin{itemize}
	\item \textbf{Soft clauses for maximizing training accuracy:} For each training sample, {\imli} constructs a soft unit\footnote{A unit clause has a single literal.} clause $ \neg \eta_l $ to account for a penalty for misclassification. Since the penalty for misclassification of a sample is $ 1 $  in Eq.~\eqref{interpretability_imli_eq:obj},  the weight of this soft clause is also $ 1 $.

	\begin{equation}
		\label{interpretability_imli_soft_clause_error}
		E_l:=\neg \eta_l;  \qquad \mathsf{wt}(E_l)= 1
	\end{equation}
	Intuitively, if a sample is misclassified, the associated error variable becomes true, thereby dissatisfying the soft clause $ E_l $.

	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible. Hence, for each feature variable $ b_j^i $, {\imli} constructs a unit clause as $ \neg b_j^i $. Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
	\begin{equation}
		\label{interpretability_imli_soft_clause_feature}
		S_j^i:= \neg b_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
	\end{equation}

	\item \textbf{Hard clauses for encoding constraints:} In a MaxSAT problem, constraints that must be satisfied are encoded as hard clauses. In {\imli}, we have a constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa. Let  $\mathbf{b}_i= \{b^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $. Then, we define the following hard clause.

	\begin{equation}
			\label{interpretability_imli_hard_clause_rule_consistency}
		H_l:= \neg \eta_l \rightarrow \Big(y_l\leftrightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
	\end{equation}

	In the hard clause, $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ is a CNF formula including variables $ b_j^i $ for which the associated  feature-value is $ 1 $ in  $ \mathbf{X}_l $. Since $ y_l \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement. Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement is true depending on the class-label of the sample.  We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \eta_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.  This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver. Therefore, the single-implication ``$ \rightarrow $'' in the hard clause $ H_l $ acts as a double-implication ``$ \leftrightarrow $''  due to the soft clause $ E_l $.

\end{itemize}

	  We next discuss the translation of soft and hard clauses into a CNF formula, which can be invoked by any MaxSAT solver.

\paragraph{Translating $ E_l, S^i_j, H_l $ to a CNF formula.} 
The soft clauses $ E_l $ and $ S_j^i $ are unit clauses and hence, no translation is required for them.  In the hard clause, when $ y_l=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $. In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF. In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$. Since $ \mathbf{X}_l \circ \mathbf{b}_{i} $ constitutes a disjunction of literals (defined in Section~\ref{interpretability_imli_sec:preliminaries}), we apply Tseytin transformation to encode  $ \neg (\mathbf{X}_l \circ \mathbf{b}_{i}) $ into CNF. More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $. Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})  $ for $ i=\{1,\dots,k\} $. Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF. For either value of $ y_l $, the weight of each translated hard clause in the CNF formula is $ \infty $.

% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ when $ y_l=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$.

% In the hard clause,  $\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ is already in CNF. When $ y_l=1 $, the constraint $ H_l $ can be directly encoded as CNF by using equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $. When $ y_i=0 $, we apply

Once we translate  all soft and hard clauses into CNF, the MaxSAT query $ Q $ is the conjunction of all clauses. 
\[
Q:= \bigwedge_{l=1}^n E_l \wedge  \bigwedge_{i=1,j=1}^{i=k,j=m} S_j^i \wedge \bigwedge_{l=1}^n H_l 
\]

Any off-the-shelf MaxSAT solver can output an optimal assignment $ \sigma^* $ of the MaxSAT query $ (Q, \weight{\cdot}) $. We extract $ \sigma^* $ to construct the classifier $ \Rule $ and compute training errors as follows.
\begin{construction}
	\label{interpretability_imli_construction:rule}
	Let $\sigma^* = \MaxSAT(Q,\weight{\cdot})$. Then $x_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(b_{j}^{i}) = 1$. Additionally, $ \mathbf{X}_l $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
\end{construction}

\paragraph{Complexity of the MaxSAT query.}

We analyze the complexity of the MaxSAT query in terms of the number of Boolean variables and clauses in the CNF formula $ Q $.

\begin{proposition}
	\label{interpretability_imli_prop:maxsat_variables}
	 To learn a $ k $-clause CNF classifier for a dataset of  $ n $ samples over $ m $ boolean features, the MaxSAT query $ Q $ defines $ km +n $ Boolean variables. Let $ n_{neg} $ be the number of negative samples in the training dataset. Then the number of auxiliary variables in $ Q $ is  $ kn_{neg}  $.
\end{proposition}

\begin{proposition}
	\label{interpretability_imli_prop:maxsat_clauses}
	The MaxSAT query $ Q $ has $ k m+n $ unit clauses corresponding to the constraints $ E_l $ and $ S_j^i $. For a positive sample, the hard clause $ H_l $ is translated into $ k $  clauses. For each negative sample, the CNF translation requires at most $ k m+1 $ clauses. Let $ n_{pos} $ and $ n_{neg} $ be the number of positive and negative samples in the training set. Therefore, the number of clauses in the MaxSAT Query $ Q $ is $ k m+n+k n_{pos}+(k m+1)n_{neg} \approx \mathcal{O}(k m  n ) $ when $ n_{pos}= n_{neg} =\frac{n}{2} $. 
\end{proposition}

\subsubsection*{Example of encoding}

We illustrate the MaxSAT encoding for a toy example consisting of four samples and two binary features. Our goal is to learn a two clause CNF classifier that can approximate the training data. 
\[ 
\mathbf{X}_\mathsf{orig} = 
\begin{blockarray}{cc}
x_1 & x_2 \\
\begin{block}{[cc]}
0 & 0 \\
0 & 1 \\
1 & 0 \\
1 & 1 \\
\end{block}
\end{blockarray} \implies 
\mathbf{X} = 
\begin{blockarray}{cccc}
x_1 & \neg x_1 & x_2  & \neg x_2\\
\begin{block}{[cccc]}
0 & 1 & 0 & 1\\
0 & 1 & 1 & 0\\
1 & 0 & 0 & 1\\
1 & 0 & 1 & 0\\
\end{block}
\end{blockarray}; \quad 
\mathbf{y} = 
\begin{bmatrix} 
1 \\ 0 \\ 0 \\ 1
\end{bmatrix}
\]
In the preprocessing step, we negate the columns in $ \mathbf{X}_\mathsf{orig} $ and add complemented features $ \{\neg x_1, \neg x_2\} $ in $ \mathbf{X} $. In the MaxSAT encoding, we define $ 8 $ feature variables ($ 4 $ features $ \times $ $ 2 $ clauses in the classifier) and $ 4 $ error variables. For example, for feature $ x_2 $, introduced feature variables are $ \{b^1_3, b^2_3\} $ and for feature $ \neg x_2 $, introduced variables are $ \{b^1_4, b^2_4\} $. For four samples, error variables are $ \{\eta_1, \eta_2, \eta_3, \eta_4\} $. We next show the soft and hard clauses in the MaxSAT encoding

\[
\begin{split}
&E_1:= (\neg \eta_1); \quad
E_2:= (\neg \eta_2); \quad E_3:= (\neg \eta_3); \quad
E_4:= (\neg \eta_4)\\
& S_1^1 := (\neg b_1^1);\quad 
 S^1_2 := (\neg b^1_2);\quad 
 S^1_3 := (\neg b^1_3);\quad
 S^1_4 := (\neg b^1_4);\quad \\ 
& S_1^2 := (\neg b_1^2);\quad 
S^2_2 := (\neg b^2_2);\quad 
S^2_3 := (\neg b^2_3);\quad 
S^2_4 := (\neg b^2_4);\quad\\ 
& H_1:= (\neg \eta_1 \rightarrow ((b_2^1 \vee b_4^1)\wedge (b_2^2 \vee b_4^2)));\\
& H_2:= (\neg \eta_2 \rightarrow (\neg(b_2^1 \vee b_3^1) \vee \neg(b_2^2 \vee b_3^2)));\\
& H_3:= (\neg \eta_3 \rightarrow (\neg (b_1^1 \vee b_4^1) \vee \neg(b_1^2 \vee b_4^2)));\\
& H_4:= (\neg \eta_4 \rightarrow ((b_1^1 \vee b_3^1) \wedge (b_1^2 \vee b_3^2)));\\
\end{split}
\]

In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \weight{E_i} = 1 $ and rule-sparsity weight as $ \weight{S_j^i} = 0.1 $. Intuitively, the penalty for misclassifying a sample is $ 10 $ times more than the penalty for adding a feature in the classifier. For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{b^1_1, b^1_4, b^2_2, b^2_3\} $ to true. Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (x_1 \vee \neg x_2)  \wedge (\neg x_1 \vee x_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ x_i $ and $ \neg x_i $ in the same clause. The reason is that a clause with both $ x_i $ and $ \neg x_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}

\subsection{Learning with Non-binary Features}
	\label{interpretability_imli_sec:non-binary}
	The presented MaxSAT encoding requires input samples to have binary features. Therefore, we discretize datasets with categorical and continuous features into binary features. For each continuous feature, we apply equal-width discretization that splits the feature into a fixed number of bins. For example, consider a continuous feature $ x_c \in [a,b] $. In discretization, we split the domain $ [a,b] $ into three bins with two split points $ \{a',b'\} $ such that $ a<a'<b'<b $.  Therefore, the resulting three discretized features are $ { a \le x_c < a'} $, $ {a' \le x_c < b'} $, and $ { b' \le x_c  \le b } $. An alternate to this close-interval discretization is open-interval discretization, where we consider six discretized features with each feature being compared with one threshold. In that case, the discretized features are: $ x_c \ge a, x_c \ge a', x_c \ge b', x_c < a', x_c < b', x_c \le b $. Both open-interval and close-interval discretization techniques have their use-cases where one or the other is appropriate. For simplicity, we experiment with close-interval discretization in this paper.

	After applying discretization on continuous features, the dataset contains categorical features only, which we convert to binary features using one-hot encoding~\cite{lakkaraju2019faithful,ghosh19imli}. In one-hot encoding, a Boolean vector of features is introduced with cardinality equal to the number of distinct categories. For example, consider a categorical feature having three categories `red', `green', and `yellow'. In one hot encoding, samples with category-value `red', `green', and `yellow' would be converted into binary features by taking values $ 100 $, $ 010 $, and $ 001 $, respectively.

\subsection{Flexible Interpretability Objectives}
\label{interpretability_imli_sec:complex_interpretability_objectives}
{\imli} can be extended to consider more flexible interpretability objectives than the simplified one in Eq.~\eqref{interpretability_imli_eq:obj}. In Eq.~\eqref{interpretability_imli_eq:obj}, we prioritize all features equally by providing the same weight to the clause $ S_j^i $ for all  values of $ i $ and $ j $. In practice, users may prefer rules containing certain features. In {\imli}, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $. For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg b_j^i $ can be set to $ \infty $. In contrast, to always include a feature, we can define $ S_j^i := b_j^i $ with weight $ \infty $. In both cases, we treat $ S_j^i $ as a hard clause.

Another use case may be to learn rules where clauses have disjoint set of features. To this end, we consider a pseudo-Boolean constraint $ \sum_{i=1}^k b^i_j \le 1 $, which specifies that the feature $ x_j $ appears in at-most one of the $ k $ clauses. This constraint may be soft or hard depending on the priority in the application domain. In either case, we convert this constraint to CNF using pseudo-Boolean to CNF translation~\cite{philipp2015pblib}. Thus, {\imli} allows us to consider varied interpretability constraints by only modifying the MaxSAT query without requiring changes in the MaxSAT solving. Therefore, the \emph{separation between modeling and solving} turns out to be the key strength of {\imli}.

### 2022-01-27 chapters/interpretability/imli/conclusion.tex
imli journal paper added

\section{Conclusion}
\label{interpretability_imli_sec:conclusion}
Interpretable machine learning is gaining more focus with applications in many safety-critical domains. Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.  In this paper, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas. {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.  In our empirical evaluation, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability. In particular, {\imli} demonstrates competitive prediction accuracy and rule-size compared to existing interpretable rule-based classifiers. In addition, {\imli} achieves impressive scalability than both interpretable and non-interpretable classifiers by learning interpretable rules on million-size datasets with higher accuracy.  Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework.

%\section*{Acknowledgments}
%This work was supported in part by National Research Foundation, Singapore under its NRF Fellowship Programme [NRF-NRFFAI1-2019-0004 ] and AI Singapore Programme [AISG-RP-2018-005],  and NUS ODPRT Grant [R-252-000-685-13]. The computational work for this article was performed on resources of Max Planck Institute for Software Systems, Germany and the National Supercomputing Centre, Singapore (\url{https://www.nscc.sg}.

### 2022-01-27 chapters/interpretability/imli/experiment.tex
imli journal paper added

\section{Empirical Performance Analysis}
\label{interpretability_imli_sec:experiments}
In this section, we empirically evaluate the performance of {\imli}. We first present the experimental setup and the objective of the experiments, followed by experimental results.

\subsection{Experimental Setup}
We implement a prototype of {\imli} in Python to evaluate the performance of the proposed MaxSAT-based formulation for learning classification rules. To implement {\imli}, we deploy a state-of-the-art MaxSAT solver Open-WBO~\cite{martins2014open}, which returns the current best solution upon reaching a timeout.

We compare {\imli} with state-of-the-art interpretable and non-interpretable classifiers. Among interpretable classifiers, we compare with RIPPER~\cite{cohen1995fast}, BRL~\cite{letham2015interpretable}, CORELS~\cite{angelino2017learning}, and BRS~\cite{wang2017bayesian}. Among non-interpretable classifiers, we compare with Random Forest (RF), Support Vector Machine with linear kernels (SVM), Logistic Regression classifier (LR), and k-Nearest Neighbors classifier (kNN). We deploy the Scikit-learn library in Python for implementing non-interpretable classifiers.

We experiment with real-world binary classification datasets from  UCI~\cite{Dua:2019}, Open-ML~\cite{OpenML2013}, and Kaggle repository (\url{https://www.kaggle.com/datasets}), as listed in Table~\ref{interpretability_imli_table:interpretable_classifiers}. In these datasets, the number of samples vary from about $ 200 $ to $ 1,000,000 $.  The datasets contain both real-valued and categorical features. We process them to binary features by setting the maximum number of bins as $ 10 $ during discretization. For non-interpretable classifiers such as RF, SVM, LR, and kNN that take real-valued features as inputs, we only convert categorical features to one-hot encoded binary features.

We perform ten-fold cross-validation on each dataset and evaluate the performance of different classifiers based on the median prediction accuracy on the test data. Additionally, we compare the median size of generated rules among rule-based interpretable classifiers. We consider a comparable combination ($ 100 $) of  hyper-parameters choices for all classifiers, that we fine-tune during cross-validation. For {\imli}, we vary the number of clauses $ k \in \{1, 2, \dots, 5\} $ and the regularization parameter $ \lambda $  in a logarithmic grid by choosing $ 5 $ values between $ 10^{-4} $ and $ 10^1 $. For mini-batch learning in {\imli}, we set the number of samples in each mini-batch, $ n' \in \{50, 100, 200, 400\} $. Thus, we consider $ \lceil n / n' \rceil $ mini-batches, where $ n $ denotes the size of training data. To construct mini-batches from a training dataset, we sequentially split the data into $ \lceil n / n' \rceil $  batches with each batch having $ n' $ samples. Furthermore, to ignore the effect of batch-ordering, we perform mini-batch learning in two rounds such that each batch participates twice in the training.

For BRL algorithm, we vary four hyper-parameters: the maximum cardinality of rules in $ \{2, 3, 4\} $, the minimum support of rules in $ \{0.05, 0.175, 0.3\} $, and the prior on the expected length and width of rules in $ \{2, 4, 6, 8\} $ and $ \{2, 5, 8\} $, respectively. For CORELS algorithm, we vary three hyper-parameters:  the maximum cardinality of rules in $ \{2,, 5\} $, the minimum support of rules in $ \{0.01, 0.17, 0.33, 0.5  \} $, and the regularization parameter in $ \{0.005, 0.01 , 0.015, 0.02 , 0.025, 0.03\}  $. For BRS algorithm, we vary three hyper-parameters: the number of initial rules in $ \{500, 1000, 1500, 2000, 2500, 3000\} $, the maximum length of rules in $ \{1,2,3,4\} $, and the minimum support of rules in $ \{1, 4,7, 10\} $. For RF and RIPPER classifiers, we vary the cut-off on the number of samples in the leaf node using a linear grid between $ 3 $ to $ 500 $ and $ 1 $ to $ 300 $, respectively. For SVM and LR classifiers, we discretize the regularization parameter on a logarithmic grid between  $ 10^{-3} $ and $ 10^3 $. For kNN, we vary the number of neighbors in a linear grid between $ 1 $ and $ 500 $. We conduct each experiment on an Intel Xeon E$ 7-8857 $ v$ 2 $ CPU using a single core with $ 16 $ GB of RAM running on a 64bit Linux distribution based on Debian. For all classifiers, we set the training timeout to $ 1000 $ seconds.

\paragraph{Objectives of Experiments.} In the following, we present the objectives of our experimental study.

\begin{enumerate}
	\item How are the accuracy and size of classification rules generated by  {\imli} compared to existing interpretable classifiers?
	\item How is the scalability of {\imli} in solving large-scale classification problems compared to existing interpretable classifiers?
	\item How does {\imli} perform in terms of accuracy and scalability compared to classifiers that are non-interpretable? 
	\item How does the incremental learning in {\imli} perform compared to non-incremental MaxSAT-based learning in terms of accuracy, rule-sparsity, and scalability?
	\item How do different interpretable classification rules learned using {\imli} perform in terms of accuracy and rule-size?
	\item What are the effects of different hyper-parameters in {\imli}?
\end{enumerate}

\paragraph{Summary of Experimental Results.}
To summarize our experimental results, {\imli} achieves the best balance among prediction accuracy, interpretability, and scalability compared to existing interpretable rule-based classifiers. Particularly, compared to the most accurate classifier RIPPER, {\imli} demonstrates on average $ 1\% $ lower prediction accuracy, wherein the accuracy of {\imli} is higher than BRL, CORELS, and BRS in almost all datasets. In contrast, {\imli} generates significantly smaller rules than RIPPER, specifically in large datasets. Moreover, BRL, CORELS, and BRS report comparatively smaller rule-size than {\imli} on average, but with a significant decrease in accuracy.  In terms of scalability, {\imli} achieves the best performance compared to other interpretable and non-interpretable classifiers by classifying datasets with one million samples. While CORELS also scales to such large datasets,  its accuracy is lower than that of {\imli} by at least $ 2\% $ on average.  Therefore, {\imli} is not only scalable but also accurate in practical classification problems while also being interpretable by design.  We additionally analyze the comparative performance of different formulations presented in this paper, where the  incremental approach empirically proves its efficiency than the na\"ive MaxSAT formulation.  Furthermore, we learn and compare the performance of different interpretable representations: decision lists, decision sets, CNF, and DNF formulas using {\imli} and present the efficacy of {\imli} in learning varied interpretable classifiers. Finally, we study the effect of different hyper-parameters in {\imli}, where each hyper-parameter provides a precise control among training time,  prediction accuracy, and rule-sparsity. In the following, we discuss our experimental results in detail.

\begin{table*}[!t]
	\centering

	\caption{Comparison of accuracy and rule-size among interpretable classifiers. Each cell from the fourth to the eighth column contains test accuracy (top) and rule-size (bottom). `\textemdash' represents a timeout. Numbers in bold represent the best performing results among different classifiers.}
	\label{interpretability_imli_table:interpretable_classifiers} 
	%	\setlength{\tabcolsep}{.2em}
	\small
	\begin{tabular}{lrrrrrrrrrrrr}

	\toprule
	Dataset & Size & Features & RIPPER & BRL & CORELS & BRS & IMLI \\

	\midrule
	\multirow{2}{*}{Parkinsons} & \multirow{2}{*}{ $ 195 $ } & \multirow{2}{*}{ $ 202 $ }  &
	$ 94.44 $  &  $ \mathbf{94.74} $  &  $ 89.74 $  &  $ 84.61 $  &  $ \mathbf{94.74} $  \\
	&&& $ 7.0 $  &  $ 11.5 $  &  $ \mathbf{2.0} $  &  $ 5.0 $  &  $ 7.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{WDBC} & \multirow{2}{*}{ $ 569 $ } & \multirow{2}{*}{ $ 278 $ }  &
	$ \mathbf{98.08} $  &  $ 93.81 $  &  $ 92.04 $  &  $ 92.98 $  &  $ 94.74 $  \\
	&&& $ 13.0 $  &  $ 22.0 $  &  $ \mathbf{2.0} $  &  $ 7.0 $  &  $ 11.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Pima} & \multirow{2}{*}{ $ 768 $ } & \multirow{2}{*}{ $ 83 $ }  &
	$ 77.14 $  &  $ 68.18 $  &  $ 75.32 $  &  $ 75.32 $  &  $ \mathbf{78.43} $  \\
	&&& $ 6.0 $  &  $ 13.5 $  &  $ \mathbf{2.0} $  &  $ 3.0 $  &  $ 23.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Titanic} & \multirow{2}{*}{ $ 1,043 $ } & \multirow{2}{*}{ $ 38 $ }  &
	$ 78.72 $  &  $ 62.98 $  &  $ \mathbf{81.9} $  &  $ 80.86 $  &  $ 81.82 $  \\
	&&& $ 6.0 $  &  $ 15.0 $  &  $ \mathbf{4.0} $  &  $ \mathbf{4.0} $  &  $ 5.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{MAGIC} & \multirow{2}{*}{ $ 19,020 $ } & \multirow{2}{*}{ $ 100 $ }  &
	$ \mathbf{82.68} $  &  $ 76.95 $  &  $ 78.05 $  &  $ 77.5 $  &  $ 78.26 $  \\
	&&& $ 102.0 $  &  $ 81.0 $  &  $ 4.0 $  &  $ \mathbf{3.0} $  &  $ 8.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Tom's HW} & \multirow{2}{*}{ $ 28,179 $ } & \multirow{2}{*}{ $ 946 $ }  &
	$ \mathbf{85.91} $  & \textemdash &  $ 83.27 $  &  $ 83.13 $  &  $ 85.24 $  \\
	&&& $ 30.0 $  & \textemdash &  $ \mathbf{4.0} $  &  $ 18.5 $  &  $ 44.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Credit} & \multirow{2}{*}{ $ 30,000 $ } & \multirow{2}{*}{ $ 199 $ }  &
	$ \mathbf{82.39} $  &  $ 46.12 $  &  $ 81.18 $  &  $ 80.45 $  &  $ 82.12 $  \\
	&&& $ 32.5 $  &  $ 26.5 $  &  $ \mathbf{2.0} $  &  $ 7.0 $  &  $ 17.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Adult} & \multirow{2}{*}{ $ 32,561 $ } & \multirow{2}{*}{ $ 94 $ }  &
	$ \mathbf{84.37} $  &  $ 72.08 $  &  $ 79.78 $  &  $ 70.75 $  &  $ 81.2 $  \\
	&&& $ 115.5 $  &  $ 46.5 $  &  $ \mathbf{4.0} $  &  $ \mathbf{4.0} $  &  $ 30.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Bank Marketing} & \multirow{2}{*}{ $ 45,211 $ } & \multirow{2}{*}{ $ 82 $ }  &
	$ \mathbf{90.01} $  &  $ 84.66 $  &  $ 89.62 $  &  $ 86.75 $  &  $ 89.84 $  \\
	&&& $ 36.5 $  &  $ 13.0 $  &  $ \mathbf{2.0} $  &  $ \mathbf{2.0} $  &  $ 24.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Connect-4} & \multirow{2}{*}{ $ 67,557 $ } & \multirow{2}{*}{ $ 126 $ }  &
	$ \mathbf{76.72} $  &  $ 65.83 $  &  $ 68.68 $  &  $ 70.49 $  &  $ 75.36 $  \\
	&&& $ 118.0 $  &  $ 18.5 $  &  $ \mathbf{4.0} $  &  $ 11.0 $  &  $ 50.5 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Weather AUS} & \multirow{2}{*}{ $ 107,696 $ } & \multirow{2}{*}{ $ 169 $ }  &
	$ \mathbf{84.22} $  &  $ 43.26 $  &  $ 83.67 $  & \textemdash &  $ 83.78 $  \\
	&&& $ 26.0 $  &  $ 22.0 $  &  $ \mathbf{2.0} $  & \textemdash &  $ 22.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Vote} & \multirow{2}{*}{ $ 131,072 $ } & \multirow{2}{*}{ $ 16 $ }  &
	$ \mathbf{97.12} $  &  $ 94.78 $  &  $ 95.86 $  &  $ 95.14 $  &  $ 96.69 $  \\
	&&& $ 132.0 $  &  $ 41.5 $  &  $ 3.5 $  &  $ \mathbf{1.0} $  &  $ 15.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{Skin Seg} & \multirow{2}{*}{ $ 245,057 $ } & \multirow{2}{*}{ $ 30 $ }  &
	\textemdash &  $ 79.25 $  &  $ 91.62 $  &  $ 68.48 $  &  $ \mathbf{94.71} $  \\
	&&&\textemdash &  $ 6.0 $  &  $ 9.0 $  &  $ \mathbf{5.0} $  &  $ 30.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{BNG(labor)} & \multirow{2}{*}{ $ 1,000,000 $ } & \multirow{2}{*}{ $ 89 $ }  &
	\textemdash & \textemdash &  $ 88.56 $  & \textemdash &  $ \mathbf{90.91} $  \\
	&&&\textemdash & \textemdash &  $ \mathbf{2.0} $  & \textemdash &  $ 24.0 $  \\
	\addlinespace[0.5em]

	\multirow{2}{*}{BNG(credit-g)} & \multirow{2}{*}{ $ 1,000,000 $ } & \multirow{2}{*}{ $ 97 $ }  &
	\textemdash & \textemdash &  $ 72.08 $  & \textemdash &  $ \mathbf{75.48} $  \\
	&&&\textemdash & \textemdash &  $ \mathbf{2.0} $  & \textemdash &  $ 27.5 $  \\
	\bottomrule

	\end{tabular}

\end{table*}

\subsection{Experimental Results}

\paragraph{Comparing {\imli} with interpretable classifiers.} We compare {\imli} with existing interpretable classifiers in three aspects: test accuracy, rule-size, and scalability.

\textit{Test accuracy and rule-size.}  We present the experimental  results of test accuracy and rule-size among interpretable classifiers in Table~\ref{interpretability_imli_table:interpretable_classifiers}, where the first, second, and third columns represent  the name of the dataset, the number of samples, and the number of features in the dataset, respectively. In each cell from the fourth to the eighth column in the table, the top value represents the median test accuracy and the bottom value represents the median size of rules measured through ten-fold cross-validation.

In Table~\ref{interpretability_imli_table:interpretable_classifiers}, {\imli} and CORELS generate interpretable classification rules in all $ 15 $ datasets in our experiments. In contrast, within a timeout of $ 1000 $ seconds, RIPPER, BRL, and BRS fail to generate any classification rule in three datasets,  specifically in large datasets ($ \ge 200,000 $ samples).

We now compare {\imli} with each interpretable classifier in detail. Compared to RIPPER, {\imli} has lower accuracy in $ 9 $ out  of $ 12 $ datasets. More specifically, the accuracy of {\imli} is $ 1\% $ lower on average than RIPPER. The improved accuracy of RIPPER, however, results in the generation of higher size classification rules than {\imli} in most datasets. In particular, {\imli} generates sparser rules than RIPPER in $ 9 $ out of $ 12 $ datasets, wherein RIPPER times out in $ 3 $ datasets. Interestingly, the difference in rule-size is more significant in larger datasets, such as in `Vote' dataset, where RIPPER learns a classifier with $ 132 $ Boolean predicates compared to $ 15 $ Boolean predicates by {\imli}.  Therefore, {\imli} is better than RIPPER in terms of rule-sparsity, but lags slightly in accuracy.

{\imli} performs better than BRL both in terms of accuracy and rule-sparsity. In particular, {\imli} has  higher accuracy and lower rule-size than BRL in $ 12 $ and $ 8 $ datasets, respectively, in a total of $ 12 $ datasets, wherein BRL times out in $ 3 $ datasets.  While comparing with CORELS, {\imli} achieves higher accuracy in almost all datasets ($ 14 $ out of $ 15 $ datasets). A similar trend is observed in comparison with BRS, where {\imli} achieves higher accuracy in all of $ 12 $ datasets and BRS times out in $ 3 $ datasets. CORELS and BRS, however, generates sparser rules than {\imli} in most datasets, but by costing a significant decrease in accuracy.  For example, in the largest dataset `BNG(credit-g)' with $ 1 $ Million samples, BRS times out and CORELS generates a classifier with  $ 72.08\% $ accuracy with rule-size $ 2 $. {\imli}, in contrast, learns a classifier with $ 27.5 $ Boolean predicates achieving $ 75.48\% $ accuracy, which is $ 3\% $ higher than CORELS. Therefore, {\imli}  makes a good balance between accuracy and rule-size compared to existing interpretable classifiers while also being highly scalable. In the following, we discuss the results on the scalability of all interpretable classifiers in detail.

\textit{Scalability.} We analyze the scalability among interpretable classifiers by comparing their training time. In Figure~\ref{interpretability_imli_fig:interpretable_classifiers}, we use cactus plots\footnote{Cactus plots are often used in (Max)SAT community to present the scalability of different solvers/methods~\cite{argelich2008first,balyo2017sat}.} to represent the training time (in seconds) of all classifiers in $ 1000 $ instances ($ 10 $ folds $ \times  $ $ 100 $ choices of hyper-parameters) derived for each dataset. In the cactus plot, the number of solved instances (within $ 1000 $ seconds) is on the $ X $-axis, whereas the training time is  on the $ Y $-axis. A point $ (x,y) $ on the plot implies that a classifier yields lower than or equal to $ y $ seconds of training in $ x $ many instances.

In Figure~\ref{interpretability_imli_fig:interpretable_classifiers}, we present results in an increasing number of samples in a dataset (from left to right).
In WDBC and Adult datasets presented on the first two plots in Figure~\ref{interpretability_imli_fig:interpretable_classifiers}, CORELS solves lower than $  600 $ instances within a timeout of $ 1000 $ seconds. The scalability performance of BRS is even worse, where it solves around $ 700 $ and $ 200 $ instances in WDBC and Adult datasets, respectively. The other three classifiers: {\imli}, BRL, and RIPPER  solve all $ 1000 $ instances, where BRL takes comparatively higher training time than the other two. The performance of {\imli} and RIPPER is similar, with RIPPER being comparatively better in the two datasets. However, the efficiency of {\imli} compared to other classifiers becomes significant as the number of samples in a dataset increases. In particular, in `Weather AUS' dataset, BRS cannot solve a single instance, BRL and CORELS solves $ ~400 $ instances, and RIPPER solves around $ 600 $ instances. {\imli}, however, solves all $ 1000 $ instances in this dataset. Similarly, in `BNG(labor)' dataset, all other classifiers except {\imli} and CORELS cannot solve any instance. While {\imli} mostly takes the maximum allowable time ($ 1000 $ seconds) in solving all instances in this dataset, CORELS can solve lower than $ 400 $ instances. 
Thus, {\imli} establishes itself as the most scalable classifier compared to other state-of-the-art interpretable classifiers.

\begin{figure}[!t]

	\centering

	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}

	\caption{Comparison of scalability among interpretable classifiers. The plots are arranged in increasing sizes of datasets (from left to right). In each cactus plot, {\imli} solves all $ 1000 $ instances for each dataset, while competitive classifiers often fail to scale, specially in larger datasets.}
	\label{interpretability_imli_fig:interpretable_classifiers}
\end{figure}

\paragraph{Comparison with non-interpretable classifiers.} We compare {\imli} with state-of-the-art non-interpretable classifiers such as LR, SVM, kNN, and RF in terms of their median test accuracy in Table~\ref{interpretability_imli_table:non_interpretable_classifiers}. In the majority of the datasets, {\imli} achieves comparatively lower test accuracy than the best performing non-interpretable classifier. The decrease in the test accuracy of {\imli} is attributed to two factors. Firstly, while we train {\imli} on discretized data, non-interpretable classifiers are trained on non-discretized data and thus {\imli} incurs additional classification errors due to discretization. Secondly, {\imli} learns a rule-based classifier, whereas non-interpretable classifiers can learn more flexible decision boundaries and thus fit data well. In Table~\ref{interpretability_imli_table:non_interpretable_classifiers}, we also observe that {\imli} achieves impressive scalability than competing classifiers by solving datasets with $ 1, 000,000 $  samples where most of the non-interpretable classifiers fail to learn any decision boundary on such large datasets. Thus, {\imli}, being an interpretable classifier, demonstrates lower accuracy than competing non-interpretable classifiers, but higher scalability in practice.

\begin{table*}[!t]        
	\centering
	\caption{Comparison of {\imli} with non-interpretable classifiers in terms of test accuracy. In the table, `\textemdash' represents a timeout. Numbers in bold represent the best performing results among different classifiers.}
	\label{interpretability_imli_table:non_interpretable_classifiers}
	%	\setlength{\tabcolsep}{.2em}
	\small
	\begin{tabular}{lrrrrrrrrrrrrrrr}

		\toprule
		Dataset & Size & LR & SVM & kNN & RF & IMLI \\

		\midrule
		\multirow{1}{*}{Parkinsons} & \multirow{1}{*}{ $ 195 $ }  &
		$ 89.74 $  &  $ 89.74 $  &  $ \mathbf{97.5} $  &  $ 90.0 $  &  $ 94.74 $  \\
		\multirow{1}{*}{WDBC} & \multirow{1}{*}{ $ 569 $ }  &
		$ \mathbf{98.25} $  &  $ \mathbf{98.25} $  &  $ 98.23 $  &  $ 96.49 $  &  $ 94.74 $  \\
		\multirow{1}{*}{Pima} & \multirow{1}{*}{ $ 768 $ }  &
		$ 78.43 $  &  $ 79.08 $  &  $ 74.5 $  &  $ \mathbf{79.22} $  &  $ 78.43 $  \\
		\multirow{1}{*}{Titanic} & \multirow{1}{*}{ $ 1,043 $ }  &
		$ 80.86 $  &  $ 80.38 $  &  $ 81.34 $  &  $ \mathbf{82.69} $  &  $ 81.82 $  \\
		\multirow{1}{*}{MAGIC} & \multirow{1}{*}{ $ 19,020 $ }  &
		$ 79.18 $  &  $ 79.34 $  &  $ 84.6 $  &  $ \mathbf{88.2} $  &  $ 78.26 $  \\
		\multirow{1}{*}{Tom's HW} & \multirow{1}{*}{ $ 28,179 $ }  &
		$ 96.2 $  &  $ 97.13 $  &  $ 88.15 $  &  $ \mathbf{97.78} $  &  $ 85.24 $  \\
		\multirow{1}{*}{Credit} & \multirow{1}{*}{ $ 30,000 $ }  &
		$ \mathbf{82.2} $  &  $ 81.9 $  &  $ 81.83 $  &  $ 82.15 $  &  $ 82.12 $  \\
		\multirow{1}{*}{Adult} & \multirow{1}{*}{ $ 32,561 $ }  &
		$ 85.26 $  &  $ 85.05 $  &  $ 83.8 $  &  $ \mathbf{86.69} $  &  $ 81.2 $  \\
		\multirow{1}{*}{Bank Marketing} & \multirow{1}{*}{ $ 45,211 $ }  &
		$ 90.09 $  &  $ 89.28 $  &  $ 89.43 $  &  $ \mathbf{90.27} $  &  $ 89.84 $  \\
		\multirow{1}{*}{Connect-4} & \multirow{1}{*}{ $ 67,557 $ }  &
		$ 79.39 $  & \textemdash &  $ 85.51 $  &  $ \mathbf{88.11} $  &  $ 75.36 $  \\
		\multirow{1}{*}{Weather AUS} & \multirow{1}{*}{ $ 107,696 $ }  &
		$ 85.64 $  & \textemdash &  $ 78.59 $  &  $ \mathbf{86.26} $  &  $ 83.78 $  \\
		\multirow{1}{*}{Vote} & \multirow{1}{*}{ $ 131,072 $ }  &
		$ 96.43 $  &  $ 96.37 $  &  $ 97.05 $  &  $ \mathbf{97.38} $  &  $ 96.69 $  \\
		\multirow{1}{*}{Skin Seg} & \multirow{1}{*}{ $ 245,057 $ }  &
		$ 91.86 $  & \textemdash &  $ \mathbf{99.96} $  &  $ \mathbf{99.96} $  &  $ 94.71 $  \\
		\multirow{1}{*}{BNG(labor)} & \multirow{1}{*}{ $ 1,000,000 $ }  &
		\textemdash & \textemdash & \textemdash & \textemdash &  $ \mathbf{90.91} $  \\
		\multirow{1}{*}{BNG(credit-g)} & \multirow{1}{*}{ $ 1,000,000 $ }  &
		\textemdash & \textemdash & \textemdash &  $ \mathbf{80.58} $  &  $ 75.48 $  \\
		\bottomrule

	\end{tabular}

\end{table*}

\begin{figure}[!t]

	\centering

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_train_val_fit_time}}	
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_test_accuracy}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_predicate_count}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_train_val_fit_time}}	
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_test_accuracy}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_predicate_count}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_train_val_fit_time}}	
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_test_accuracy}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_predicate_count}}

	\caption{Comparison of training time, test error, and rule-size among different formulations presented in the paper. In each cactus plot, the incremental formulation {\imli} with both mini-batch and iterative learning demonstrates the best performance in training time and test error than compared two formulations: non-incremental MaxSAT formulation {\imli}-$\mathsf{B}$ and incremental formulation with only mini-batch learning {\imli}-$\mathsf{M}$. In terms of rule-size, {\imli} often generates higher size rules than {\imli}-$\mathsf{M}$.}
	\label{interpretability_imli_fig:different_formulations}
\end{figure}

\paragraph{Comparison among different formulations in {\imli}.}  We compare the performance of different formulations for learning classification rules as presented in this paper. In Figure~\ref{interpretability_imli_fig:different_formulations}, we show cactus plots for assessing training time (in seconds), test error (in percentage), and rule-size among different formulations. In the cactus plot, a point $ (x,y) $ denotes that the formulation yields lower than or equal to $ y $ training time (similarly, test error and rule-size) in $ x $ many instances in each dataset.

In Figure~\ref{interpretability_imli_fig:different_formulations}, we denote the baseline non-incremental MaxSAT-based formulation as {\imli}-$\mathsf{B}$\footnote{Since  {\imli}-$\mathsf{B}$ is a non-incremental formulation and does not involve any mini-batch learning, we consider two hyper-parameters for {\imli}-$\mathsf{B}$: the number of clauses in $ \{1, 2, \dots, 10\} $ and the regularization parameter $ \lambda $ as $ 10 $ values chosen from a logarithmic grid between $ 10^{-4} $ and $ 10^1 $.}, incremental MaxSAT-based formulation with only mini-batch learning as {\imli}-$\mathsf{M}$, and incremental MaxSAT-based formulation with both mini-batch and iterative learning as {\imli}.  We first observe the training time of different formulations in the left-most column in Figure~\ref{interpretability_imli_fig:different_formulations}, where {\imli}-$\mathsf{B}$ soon times out and solves lower than $ 300 $ instances out of $ 1000 $ instances in each dataset. This result suggests that the non-incremental formulation cannot scale in practical classification task. Comparing between {\imli} and {\imli}-$\mathsf{M}$, both formulations solve all $ 1000 $ instances in each dataset with {\imli}-$\mathsf{M}$ undertaking significantly higher training time than $ {\imli} $.  Therefore, {\imli} achieves better scalability than {\imli}-$\mathsf{M}$ indicating that an integration of mini-batch and iterative learning achieves a significant progress in terms of scalability than mini-batch learning alone.

We next focus on the test error of different formulations in the middle column in Figure~\ref{interpretability_imli_fig:different_formulations}. Firstly, {\imli}-$\mathsf{B}$ has a higher test error than the other two formulations since {\imli}-$\mathsf{B}$ times out in most instances and learns a sub-optimal classification rule with reduced prediction accuracy. In contrast, {\imli} has the lowest test error compared to two formulations in all datasets. This result indicates the effectiveness of integrating both iterative and mini-batch learning with MaxSAT-based formulation in generating more accurate classification rules.

Moving focus on the rule-size in the rightmost column in Figure~\ref{interpretability_imli_fig:different_formulations}, {\imli}-$\mathsf{B}$ achieves the highest rule-size in WDBC dataset. In contrast, the rule-size of {\imli}-$\mathsf{B}$  is lowest (zero) in Credit and Adult datasets. In the last two datasets, {\imli}-$\mathsf{B}$ times out during training and returns the default rule ``true'' by predicting all samples as class $ 1 $. The other two formulations {\imli} and {\imli}-$\mathsf{M}$  demonstrate a similar trend in rule-size in all datasets with {\imli}-$\mathsf{M}$ generating comparatively smaller size rules in Credit and Adult datasets. In this context,  the improvement of rule-sparsity of {\imli}-$\mathsf{M}$ is due to a comparatively higher test error (or lower accuracy) than {\imli} as observed in all three datasets. Therefore, {\imli} appears to be the best performing formulation w.r.t.\ training time, test error, and rule-size by balancing between accuracy and rule-size while being more scalable.

\begin{table*}[!t]        
	\centering
	\caption{Comparison of test accuracy (top value) and rule-size (bottom value) among different rule-based representations learned using {\imli}. Numbers in bold denote the best performing results among different representations.}
	\label{interpretability_imli_table:different_representations}
	%	\setlength{\tabcolsep}{.2em}
	\small
	\begin{tabular}{lrrrrrrrrrrrrrrr}

\toprule
Dataset & CNF & DNF & Decision Sets & Decision Lists \\

\midrule
\multirow{2}{*}{Parkinsons}  &
$ 94.74 $  &  $ 89.47 $  &  $ \mathbf{94.87} $  &  $ 89.74 $  \\
& $ 7.5 $  &  $ \mathbf{6.0} $  &  $ 15.0 $  &  $ 6.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{WDBC}  &
$ 94.74 $  &  $ \mathbf{96.49} $  &  $ 95.61 $  &  $ 95.61 $  \\
& $ 11.5 $  &  $ 15.0 $  &  $ 15.5 $  &  $ \mathbf{10.0} $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Pima}  &
$ \mathbf{78.43} $  &  $ 77.13 $  &  $ 76.97 $  &  $ 76.97 $  \\
& $ 23.0 $  &  $ \mathbf{9.0} $  &  $ 15.0 $  &  $ 13.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Titanic}  &
$ 81.82 $  &  $ 82.29 $  &  $ 81.82 $  &  $ \mathbf{82.3} $  \\
& $ \mathbf{5.5} $  &  $ 10.5 $  &  $ 8.5 $  &  $ 8.0 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{MAGIC}  &
$ \mathbf{78.26} $  &  $ 77.44 $  &  $ 75.87 $  &  $ 77.79 $  \\
& $ \mathbf{8.5} $  &  $ 41.5 $  &  $ 10.0 $  &  $ 14.0 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Tom's HW}  &
$ 85.24 $  &  $ 85.15 $  &  $ 85.72 $  &  $ \mathbf{85.95} $  \\
& $ 44.5 $  &  $ \mathbf{26.5} $  &  $ 45.0 $  &  $ 59.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Credit}  &
$ 82.12 $  &  $ 82.15 $  &  $ 82.03 $  &  $ \mathbf{82.22} $  \\
& $ 17.5 $  &  $ 14.0 $  &  $ \mathbf{9.5} $  &  $ 21.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Adult}  &
$ 81.2 $  &  $ \mathbf{84.28} $  &  $ 80.07 $  &  $ 80.96 $  \\
& $ 30.0 $  &  $ 34.5 $  &  $ \mathbf{7.0} $  &  $ 24.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Bank Marketing}  &
$ \mathbf{89.84} $  &  $ 89.77 $  &  $ 89.67 $  &  $ 89.79 $  \\
& $ 24.5 $  &  $ 7.5 $  &  $ \mathbf{6.0} $  &  $ 10.5 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Connect-4}  &
$ \mathbf{75.36} $  &  $ 70.63 $  &  $ 68.09 $  &  $ 69.83 $  \\
& $ 50.5 $  &  $ 42.0 $  &  $ \mathbf{4.5} $  &  $ 24.0 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Weather AUS}  &
$ 83.78 $  &  $ \mathbf{84.23} $  &  $ 83.69 $  &  $ 83.85 $  \\
& $ 22.0 $  &  $ 14.0 $  &  $ \mathbf{4.0} $  &  $ 26.0 $  \\
\addlinespace[0.5em]

\multirow{2}{*}{Skin Seg}  &
$ \mathbf{94.71} $  &  $ 93.68 $  &  $ 87.92 $  &  $ 91.17 $  \\
& $ 30.0 $  &  $ 15.0 $  &  $ \mathbf{3.0} $  &  $ 7.0 $  \\
\bottomrule

	\end{tabular}

\end{table*}

\paragraph{Performance evaluation of different interpretable representations in {\imli}.}

We deploy {\imli} to learn different interpretable rule-based representations: CNF and DNF classifiers, decision lists, and decision sets and present their comparative performance w.r.t.\ test accuracy and rule-size in Table~\ref{interpretability_imli_table:different_representations}. In each cell in this table, the top value represents the test accuracy and the bottom value represents the size of generated rules.

We learn all four interpretable representations on twelve datasets, where the CNF classifier appears to be the most accurate representation by achieving the highest accuracy in five datasets. In contrast, both DNF and decision lists achieve the highest accuracy in three datasets each; decision sets demonstrate the least performance in test accuracy by being more accurate in one dataset. To this end, the poor accuracy of decision sets is traded off by its rule-size as decision sets generate the sparsest rules compared to other representations. More precisely, decision sets have the smallest rule-size in six datasets, while CNF, DNF, and decision lists have the smallest rule-size in two, three, and one dataset, respectively. These results suggest that CNF classifiers are more favored in applications where higher accuracy is preferred, while decision sets are preferred in applications where higher interpretability is desired. In both cases, one could deploy {\imli} for learning varied representations of classification rules.

\begin{figure}[!t]

	\centering

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_wdbc}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_titanic}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_adult}}

	\caption{Effect of the number of clauses $ k $ on accuracy (test and train), rule-size, and training time. As $ k $ increases, both train and test accuracy of {\imli} increase while generating rules with higher size by incurring higher training time. }
	\label{interpretability_imli_fig:effect_of_number_of_clauses}
\end{figure}

\paragraph{Ablation study.} We experiment the effect of different hyper-parameters in {\imli} on prediction accuracy, rule-size, and training time in different datasets. In the following, we discuss the impact of the number of clauses, regularization parameter, and size of mini-batches in {\imli}.

\textbf{Effect of the number of clauses $ k $.} In Figure~\ref{interpretability_imli_fig:effect_of_number_of_clauses}, we vary  $ k $ while learning CNF classifiers in {\imli}. As $ k $ increases, both training and test accuracy generally increase in different datasets (plots in the first and second columns). Similarly, the size of rules increases with $ k $ by incurring higher training time (plots in the third and fourth columns). The reason is that a higher value of $ k $ allows more flexibility in fitting the data well by incurring more training time and generating higher size classification rules. Therefore, the number of clauses in {\imli} provides control on training-time vs accuracy and also on accuracy vs rule-sparsity.

\textbf{Effect of regularizer $ \lambda $.} In Figure~\ref{interpretability_imli_fig:effect_of_number_of_regularization}, we vary $ \lambda $ in a logarithmic grid between $ 10^{-4} $ and $ 10^1 $. As stated in Eq.~\eqref{interpretability_imli_eq:obj_incr}, a higher value of $ \lambda $ puts more priority on the minimal changes in rules between consecutive mini-batches in incremental learning while allowing higher mini-batch errors. Thus, in the first and second columns in Figure~\ref{interpretability_imli_fig:effect_of_number_of_regularization}, as $ \lambda $ increases, both training and test accuracy gradually decrease. In addition, the size of rules (plots in the third column) also decreases. Finally, we observe that the training time generally decreases with $ \lambda $. This observation indicates that higher $ \lambda $ puts lower computational load to the MaxSAT solver as a fraction of training examples is allowed to be misclassified. Thus, similar to the number of clauses, regularization parameter $ \lambda $ in {\imli} allows to trade-off between accuracy and rule-size in a precise manner.

\textbf{Effect of the size of mini-batch.} In Figure~\ref{interpretability_imli_fig:effect_of_batch_size}, we present the effect of mini-batch size in {\imli}. As we consider more samples in a batch, both test and training accuracy increase in general as presented in the first and second columns in Figure~\ref{interpretability_imli_fig:effect_of_batch_size}. Similarly, the size of generated rules also increases with the number of samples. Due to solving higher size MaxSAT queries, the training time also increases in general with an increase in mini-batch size. Therefore, by varying the size of mini-batches, {\imli} allows controlling on training time vs the prediction accuracy (and rule-size) of generated rules.

\begin{figure}[!t]

	\centering

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_wdbc}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_titanic}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_adult}}

	\caption{Effect of regularization $ \lambda $ on accuracy (test and train), rule-size, and training time. As $ \lambda $ increases, lower priority is given to accuracy. As a result, both training and test accuracy decrease with $ \lambda $ by generating smaller rules. }
	\label{interpretability_imli_fig:effect_of_number_of_regularization}
\end{figure}

\begin{figure}[!t]

	\centering

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_wdbc}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_wdbc}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_titanic}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_titanic}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_adult}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_adult}}

	\caption{Effect of bath size on accuracy (test and train), rule-size, and training time. As we consider more samples in a mini-batch, {\imli} generates more accurate and larger size classification rules.}
	\label{interpretability_imli_fig:effect_of_batch_size}
\end{figure}

\begin{comment}

\begin{figure}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_parkinsons_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_pima_cactus_train_val_fit_time}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_titanic_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_magic_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_toms_cactus_train_val_fit_time}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_credit_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_bank_cactus_train_val_fit_time}}

	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_connect_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_dota_cactus_train_val_fit_time}}
	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}

	\caption{Comparison among interpretable classifiers.}
	\label{interpretability_imli_fig:interpretable_classifiers_all}
\end{figure}

%
%\begin{figure}
%	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/all_datasets_cactus_predicate_count}}
%	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/all_datasets_cactus_test_accuracy}}
%	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/all_datasets_cactus_train_val_fit_time}}
%	\caption{}
%\end{figure}

\end{comment}

%

### 2022-01-27 chapters/interpretability/imli/incremental_learning.tex
imli journal paper added

\section{Incremental Learning with MaxSAT-based formulation}
\label{interpretability_imli_sec:incremental_learning}
The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to learn in a CNF formula. In this section, we discuss an incremental learning technique that along with MaxSAT formulation can generate classification rules efficiently. Our incremental learning is built on two concepts: (i)
mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.

\subsection{Mini-batch Learning}

Our first improvement is to implement a mini-batch learning technique tailored for rule-based classifiers. Mini-batch learning has two-fold advantages. Firstly, instead of solving a large MaxSAT query for the whole training data, this approach solves a sequence of smaller MaxSAT queries derived from mini-batches of smaller size.  Secondly, this approach extends to life-long learning~\cite{chen2018lifelong}, where the classifier can be updated incrementally with new samples while also generalizing to previously observed samples. In our context of rule-based classifiers, we consider the following heuristic in mini-batch learning.

\paragraph{A Heuristic for Mini-Batch Learning.} Let $ \Rule' $ be a classifier learned on the previous batch. In mini-batch learning, we aim to learn a new classifier $ \Rule $ that can generalize to both the current batch and previously seen samples. For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch. Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution. Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity. We define the Hamming distance between two CNF classifiers $ \Rule $, $ \Rule' $ with the same number of clauses as follows.
\[
	d_{H}(\Rule, \Rule') = \sum_{i=1}^k \Big(\sum_{v\in C_i} \mathds{1}(v \not\in C'_i) +  \sum_{v\in C'_i} \mathds{1}(v \not\in C_i) \Big), 
\]

 where $ C_i $ and $ C'_i $ are the $ i^\text{th} $ clause in $ \Rule $ and $ \Rule' $, respectively and $ \mathds{1} $ is an indicator function that returns $ 1 $ if the input is true and returns $ 0 $ otherwise.  Intuitively, $ d_{H}(\Rule', \Rule) $ calculates the total number of different literals in each (ordered) pair of clauses in $ \Rule $  and $ \Rule' $.  For example, consider $ \Rule = (x_1 \vee x_2) \wedge (\neg x_1) $ and $ \Rule' = (\neg x_1 \vee x_2) \wedge (\neg x_1) $. Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{x_1, \neg x_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $. In the following, we discuss a modified objective function for mini-batch learning.

\paragraph{Objective Function.}
\label{interpretability_imli_sec:obj_incremental}
In mini-batch learning, we design an objective function to penalize both classification errors on the current batch and the Hamming distance between new and previous classifiers. Let $ (\mathbf{X}^b, \mathbf{y}^b) $ be the current mini-batch. The objective function in mini-batch learning is

%\red{Moreover, $ (\mathbf{X}, \mathbf{y}) $ is the training dataset, on which we measure the performance of $ \Rule $ after completing incremental learning\textemdash we will expand shortly on how we measure the performance. 
%}

\begin{align}
\label{interpretability_imli_eq:obj_incr}
\min_{\Rule} |\mathcal{E}_{\mathbf{X}_b}|+\lambda d_{H}(\Rule, \Rule').
\end{align}

In the objective function, $ \mathcal{E}_{\mathbf{X}_b} $ is the misclassified subset of samples in the current batch $ (\mathbf{X}^b, \mathbf{y}^b) $. Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ controls the trade-off between classification errors and the syntactic differences between consecutive classifiers. We next discuss how to encode the modified objective function as a MaxSAT query.

\paragraph{MaxSAT Encoding of Mini-batch Learning.}
\label{interpretability_imli_sec:encoding_incremental}
In order to account for the modified objective function, we only modify the soft clause $ S_j^i $ in the MaxSAT query $ Q $. In particular, we define $ S_j^i $ to penalize for the complemented assignment  of the feature variable $ b_j^i $ in $ \Rule $ compared to $ \Rule' $.
\[
S_j^i:=
\begin{cases}
b_j^i& \text{if }  x_j \in  \clause(\Rule',i)  \\
\neg b_j^i & \text{otherwise}
\end{cases};\qquad \mathsf{wt}(S_j^i)=\lambda
\]

$ S_j^i $ is either a unit clause $ b_j^i $ or $ \neg b_j^i $ depending on whether the associated feature $ x_j $ appears in the previous classifier $ \Rule' $ or not. In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg b_j^i $ for the first batch. We next analyze the complexity of the MaxSAT encoding for mini-batch learning.

\begin{proposition}
	\label{interpretability_imli_prop:maxsat_complexity_incremental}
	Let $ n' \triangleq |\mathbf{X}_b| \ll |\mathbf{X}| $ be the size of a mini-batch, $ n'_{neg} \le n' $ be the number of negative samples in the batch, and $ m $ be the number of Boolean features. According to Proposition~\ref{interpretability_imli_prop:maxsat_variables}, to learn a $ k $-clause CNF classifier in mini-batch learning, the MaxSAT encoding for each batch has $ km + n' $ Boolean variables and $ kn'_{neg} $ auxiliary variables. Let $ n'_{pos} = n' - n'_{neg} $ be the number of positive samples in the batch. Then, according to Proposition~\ref{interpretability_imli_prop:maxsat_clauses}, the number of clauses in the MaxSAT query is $ k m+n'+k n'_{pos}+(k m+1)n'_{neg} \approx \mathcal{O}(k m  n' ) $ when $ n'_{pos}= n'_{neg} =\frac{n'}{2} $.
\end{proposition}

%
%\paragraph{Constructing Mini-batch.}
%\red{Can be put in experiments.}
%We construct mini-batches in several ways: (i) randomly sampling a fixed number of samples from the training data, and (ii) sequentially partitioning the training data into batches of fixed size. In our experiments, we have focused on the later approach. Additionally, to ignore the effect of batch-order, we repeat the mini-batch learning multiple times.

\paragraph{Assessing  the Performance of $ \Rule $.} Since we apply a heuristic objective in mini-batch learning,  $ \Rule $ may be optimized for the current batch while generalizing poorly on the full training data. To tackle this, after learning on each batch, we decide whether to keep $ \Rule $ or not by assessing the performance of $ \Rule $ on the training data and keep $ \Rule $ if it achieves higher performance. We measure the performance of $ \Rule $ on the full training data $ (\mathbf{X}, \mathbf{y}) $ using a weighted combination of classification errors and rule-size. In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}| $ on the training data $ (\mathbf{X}, \mathbf{y}) $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}. Additionally, we  discard the current classifier $ \Rule $ when the loss does not decrease ($ \mathsf{loss}(\Rule) > \mathsf{loss}(\Rule') $).

We present the algorithm for mini-batch learning in Algorithm~\ref{interpretability_imli_algo:incremental}. 
\begin{algorithm}
	\caption{MaxSAT-based Mini-batch Learning}
	\label{interpretability_imli_algo:incremental}
	\begin{algorithmic}[1]
		\Function{$ \mathsf{MiniBatchLearning} $}{$ \mathsf{X},\mathbf{y},\lambda, k $}
		\State $\mathcal{R} = \leftarrow \bigwedge_{i=1}^k\text{false} $ \Comment{Empty CNF formula}
		\State $ \mathsf{loss}_{\max} \leftarrow \infty$ 
		\For{$ i \leftarrow 1, \dots, N $} \Comment{$ N $ is the total batch-count}
		\State $ \mathbf{X}^b, \mathbf{y}^b \leftarrow \mathsf{GetBatch}(\mathsf{X},\mathbf{y}) $
		\State $ Q, \weight{\cdot} \leftarrow \mathsf{MaxSATEncoding}(\Rule, \mathbf{X}^b, \mathbf{y}^b, \lambda, k) $ \Comment{Returns a weighted-partial CNF}
		\State $ \sigma^* \leftarrow \MaxSAT(Q,\weight{\cdot}) $
		\State $ \Rule_{\mathsf{new}} \leftarrow \mathsf{ConstructClassifier(\sigma^*)} $
		\State $ \mathsf{loss} \leftarrow  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}_{\mathsf{new}}| $ \Comment{Compute loss}
		\If{$ \mathsf{loss} < \mathsf{loss}_{\max} $}
		\State $ \mathsf{loss}_{\max} \leftarrow \mathsf{loss} $
		\State $ \Rule \leftarrow \Rule_{\mathsf{new}} $
		\EndIf
		\EndFor
		\State \Return $ \Rule $
		\EndFunction
	\end{algorithmic}
\end{algorithm}

### 2022-01-27 chapters/interpretability/imli/introduction.tex
imli journal paper added

\section{Introduction}

%The last decade has witnessed significant progress in machine learning (ML). 
Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}. In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}. In this study, we focus on learning an interpretable machine learning classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box models.

%Machine learning techniques are now being deployed in critical decision making, for example, hiring employees, giving a loan to people, predicting recidivism, predicting the likelihood of stroke of patients, etc. In the classical setting of classification problems, making an accurate prediction has been the objective of several prediction systems e.g., ad predictions, movie recommendations, etc. at the cost of making such systems essentially black-box. However, there are other safety-critical domains e.g., legal, medical, law, and so on, where interpretability, privacy, and fairness of the prediction system are of paramount importance. In this study, we show a principled approach to generate accurate and interpretable decision functions expressible in propositional logic.

In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}. Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.  At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging. In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean predicates), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.

In this paper, we study decision functions expressible in \emph{propositional logic}. In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean predicates that the formula contains. In the following, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.

%\begin{subequations}
%	\begin{empheq}[box=\widefbox]{align*}
%	& \text{A tumor is malignant if}\\
%	&[(\text{compactness SE} < 0.1) \text{\textbf{ OR }}  \neg (0.1 \le \text{concave points} < 0.2)]
%	\text{\textbf{ \blue{AND} }} \\
%	&[\neg (0.2 \le \text{area} < 0.3) \text{\textbf{ OR }} \neg (0.1 \le \text{largest symmetry }< 0.2)]
%	\text{\textbf{ \blue{AND} }} \\
%	&[(0.6 \le \text{compactness} < 0.7) \text{\textbf{ OR }} (0.4 \le \text{concave points} < 0.5) \text{\textbf{ OR }}  (\text{area SE} \ge 0.1)\\
%	& \text{\textbf{ OR }} (0.3 \le \text{largest concavity} < 0.4) \text{\textbf{ OR }} (0.2 \le \text{largest area} < 0.3)]
%	\end{empheq}
%\end{subequations}

\boxed{
\begin{array}{rl}
	& \text{A tumor is malignant if}\\
	&[(\text{compactness SE} < 0.1) \text{\textbf{ OR }}  \neg (0.1 \le \text{concave points} < 0.2)]
	\text{\textbf{ \blue{AND} }} \\
	&[\neg (0.2 \le \text{area} < 0.3) \text{\textbf{ OR }} \neg (0.1 \le \text{largest symmetry }< 0.2)]
	\text{\textbf{ \blue{AND} }} \\
	&[(0.6 \le \text{compactness} < 0.7) \text{\textbf{ OR }} (0.4 \le \text{concave points} < 0.5) \text{\textbf{ OR }}  (\text{area SE} \ge 0.1)\\
	& \text{\textbf{ OR }} (0.3 \le \text{largest concavity} < 0.4) \text{\textbf{ OR }} (0.2 \le \text{largest area} < 0.3)]
\end{array}
}
\begin{example}
	We learn an interpretable classification rule in conjunctive normal form for predicting the classification of a tumor cell. We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with three \emph{clauses} and nine Boolean \emph{predicates}. In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{predicates} connected by Boolean `OR'. Informally, a tumor cell is malignant if at least one predicate in each clause becomes true.
\end{example}

The problem of learning rule-based classifiers is known to be computationally intractable. The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}. In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc. In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}. Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule-size. In our study, we propose an alternate optimization approach that fits particularly well to rule-learning problems. Particularly,  we propose a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.

The MaxSAT problem is an optimization analog to the satisfiability (SAT) problem, which is complete for the class $ \mathrm{FP}^{\mathrm{NP}} $. Although the MaxSAT problem is NP-hard, dramatic progress has been made in designing solvers that can handle large-scale problems arising in practice. This has encouraged researchers to reduce several optimization problems into MaxSAT such as  optimal planning~\cite{robinson2010partial}, automotive configuration~\cite{walter2013applications}, data analysis in machine learning~\cite{berg2019applications}, and automatic test pattern generation for cancer therapy~\cite{lin2012application}.

The primary contribution of this paper is a MaxSAT-based formulation, called {\imli}, for learning interpretable classification rules expressible in propositional logic. For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF (Conjunctive Normal Form) formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets. In this paper, {\imli} provides precise control to jointly optimize the accuracy and the size of classification rules. {\imli} constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples. Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $. To scale {\imli} to large datasets, we propose an incremental learning technique along with the MaxSAT formulation. Our incremental learning is an integration of mini-batch learning and iterative rule-learning, which are studied separately in classical learning problems. In the presented incremental approach, {\imli} learns a $ k $-clause CNF formula using an iterative separate-and-conquer algorithm, where in each iteration, a single clause is learned by covering a part of the training data. Furthermore, to efficiently learn a single clause in a CNF, {\imli} relies on mini-batch learning, where it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch.

In our experimental evaluations, {\imli} demonstrates the best balance among prediction accuracy, interpretability, and scalability in learning classification rules. In particular, {\imli} achieves competitive prediction accuracy and interpretability w.r.t.\ state-of-the-art interpretable classifiers. Besides, {\imli} achieves impressive scalability by classifying datasets even with $ 1 $ million samples, wherein existing classifiers, including those of the non-interpretable ones, either fail to scale or achieve poor prediction accuracy. Finally, as an application, we deploy {\imli} in learning interpretable classifiers such as decision lists and decision sets.

The presented framework {\imli} extends our prior papers ~\cite{ghosh19imli,MM18} in the following aspects. \cite{MM18} presented a MaxSAT-based formulation for interpretable classification rule-learning, which jointly optimizes between the accuracy and size of rules.  \cite{ghosh19imli} improved the MaxSAT formulation by unifying mini-batch learning. This unification allows learning on moderate-size datasets (about $ 50 $ thousand samples). This paper takes a significant step in terms of scalability by integrating both iterative learning and mini-batch learning inside the incremental technique. The result is a practical scalable learning framework that can classify datasets with $ 1 $ million samples and can potentially scale beyond that.

We organize the rest of the paper as follows. We discuss related literature in Section~\ref{interpretability_imli_sec:related} and preliminaries in Section~\ref{interpretability_imli_sec:preliminaries}. In Section~\ref{interpretability_imli_sec:problem}, we formally define our learning problem and provide a MaxSAT-based solution in Section~\ref{interpretability_imli_sec:baseline}. We improve the MaxSAT-based formulation through an incremental learning technique in Section~\ref{interpretability_imli_sec:incremental_learning}. In Section~\ref{interpretability_imli_sec:application}, we apply {\imli} in learning different interpretable classifiers. We then discuss our experimental results in Section~\ref{interpretability_imli_sec:experiments} and conclude in Section~\ref{interpretability_imli_sec:conclusion}.

### 2022-01-27 chapters/interpretability/imli/iterative_learning.tex
imli journal paper added

\subsection{Iterative Learning}  We now discuss an iterative learning algorithm for rule-based classifiers.  The major advantage of iterative learning is that we solve a smaller MaxSAT query because of learning a \textit{partial} classifier in each iteration. Our iterative approach is motivated by the set-covering algorithm\textemdash also known as separate-and-conquer algorithm\textemdash in symbolic rule learning~\cite{furnkranz1999separate}. In this approach, the core idea is to define the \emph{coverage of a partial classifier} (for example, a clause in a CNF classifier). For a specific definition of coverage, this  algorithm separates samples covered by the partial classifier and recursively conquers remaining samples in the training data by learning another partial classifier until no sample remains. The final classifier is an aggregation of all partial classifiers\textemdash the conjunction of clauses in a CNF formula, for example.

Iterative learning is different from mini-batch learning in several aspects. In mini-batch learning, we learn all clauses in a CNF formula together, while in iterative learning, we learn a single clause of a CNF in each iteration. Additionally, in mini-batch learning, we improve scalability by reducing the number of samples in the training data using mini-batches, while in iterative learning, we improve scalability by reducing the number of clauses to learn at once. Therefore, an efficient integration of iterative learning and mini-batch learning would benefit scalability from both worlds. In the following, we discuss this integration by first stating iterative learning for CNF classifiers.

In iterative learning, we learn one clause of a CNF classifier in each iteration, where the clause refers to a partial classifier. The coverage of a clause in a CNF formula is the set of samples that \emph{do not satisfy} the clause. The reason is that if a sample does not satisfy at least one clause in a CNF formula, the prediction of the sample by the full formula is class $ 0 $ because CNF is a conjunction of clauses. As a result, considering covered samples in the next iteration does not change their prediction regardless of whatever clause we learn in later iterations. To this end, a single clause learning can be performed efficiently by applying mini-batch learning discussed before. In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we provide an algorithm for learning a CNF classifier iteratively by leveraging mini-batch learning. This algorithm is a double-loop algorithm, where in the outer loop we apply iterative learning and in the inner loop, we apply mini-batch learning.

\begin{algorithm}
	\caption{Iterative CNF Classifier Learning}
	\label{interpretability_imli_algo:iterative_CNF_learning}
	\begin{algorithmic}[1]
		\Procedure{IterativeCNFLearning}{$ \mathbf{X},\mathbf{y},\lambda, k $}
		\State $ \mathcal{R} \leftarrow $ true \Comment{Initial formula}
		\For{$ i \leftarrow 1, \dots, k $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
		\State $ C_i \leftarrow \mathsf{MiniBatchLearning}( \mathbf{X},\mathbf{y},\lambda, 1) $ \Comment{Single clause learning, $ k = 1 $}
		\label{interpretability_imli_algo:iterative_CNF_learning_incremental_learning}
		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{Coverage}(\mathbf{X}, C_i) $
		\label{interpretability_imli_algo:iterative_CNF_learning_coverage}
		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} \Comment{Terminating conditions}
		\State break
		\EndIf
		\State $ \mathcal{R} \leftarrow \mathcal{R}  \wedge C_i $
		\State $ \mathbf{X},\mathbf{y}  \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $ \Comment{Removing covered samples}
		\EndFor
		\State \Return $ \Rule $
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

\paragraph{Terminating conditions.} In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we terminate iterative learning based on three conditions: (i) when $ \Rule $ contains all $ k $ clauses , (ii) the training data $ (\mathbf{X},\mathbf{y}) $ is empty (that is, no sample remains uncovered), and (iii) no new sample is covered by the partial classifier. Since the first two conditions are trivial, we elaborate on the third condition. When clause $ C_i $ cannot cover any new sample from the training dataset $ (\mathbf{X},\mathbf{y}) $, the next iteration will result in the same clause $ C_i $ because the training data remains the same. In this case, we do not include clause $ C_i $ to classifier $ \Rule $ because of zero coverage.

\section{Applying {\imli} on Learning Other Interpretable Classifiers.}
\label{interpretability_imli_sec:application}
In earlier sections, we discuss the learning of CNF classifiers using  {\imli}. {\imli} can also be applied to learning other interpretable rule-based representations. In this section, we discuss how {\imli} can be applied in learning DNF classifiers, decision lists, and decision sets.

\subsection{Learning DNF classifiers} 
\label{interpretability_imli_sec:dnf_learning}
For learning DNF classifiers, we leverage De Morgan's law where complementing a CNF formula results in a DNF formula. To learn a DNF classifier, say $ \Rule'(\mathbf{x}) $, we can trivially show that $ y = \Rule'(\mathbf{x}) \leftrightarrow \neg (y = \neg \Rule'(\mathbf{x})) $ for the feature vector $ \mathbf{x} $. Here $ \neg \Rule'(\mathbf{x}) $ is a CNF formula, by definition. Thus, we learn a DNF classifier by first complementing the class-label $ \mathbf{y} $ to $ \neg \mathbf{y}$ in the training dataset, learning a CNF classifier on $ (\mathbf{X}, \neg \mathbf{y}) $ and finally complementing the learned classifier to DNF. For example, the CNF classifier  ``(Male $ \vee $ Age $ < 50 $) $ \wedge $ (Education $ = $ Graduate $ \vee $ Income  $ \ge 1500 $)'' 	is complemented to a DNF classifier as 	 ``(not Male $ \wedge $ Age $ \ge 50 $) $ \vee $ (Education $ \ne $ Graduate $ \wedge $ Income $ \le 1500 $)''.

To learn a DNF classifier incrementally, such as through mini-batch and iterative learning, we adopt the following procedure. For learning a DNF classifier using mini-batch learning, we first learn a CNF classifier on dataset $ (\mathbf{X}, \neg \mathbf{y}) $ and complement the classifier to a DNF classifier at the end of mini-batch learning. To learn a DNF classifier in the iterative approach, we  learn a single clause of the DNF classifier in each iteration, remove covered samples, and continue till no training sample remains. In this context, the coverage of a clause in a DNF formula is the set of samples satisfying the clause.

\subsection{Learning Decision Lists}
In {\imli}, we apply an iterative learning approach for efficiently learning a  decision list. A decision list  $ \Rule_L $ is a list of pairs $ (C_1, v_1), \dots, (C_k, v_k) $, where we propose to learn one pair in each iteration.  We note that the clause $ C_i $ is a conjunction of literals\textemdash equivalently, a single clause DNF formula. Hence, our task is to deploy {\imli} to efficiently learn a single clause DNF formula $ C_i $. In particular, we opt to learn this clause for the majority class, say $ v_i $, in the training dataset by setting the majority samples as class $ 1 $ and all other samples as class $ 0 $. As a result, even if the MaxSAT-based learning, presented in this paper, is targeted for binary classification, we can learn a multi-class decision list in {\imli}.

In Algorithm~\ref{interpretability_imli_algo:iterative_decision_lists_learning}, we present the iterative algorithm for learning decision lists.  In each iteration, the algorithm learns a pair $ (C_i, v_i) $, separates the training set based on the coverage of $ (C_i, v_i) $ and conquers the remaining samples recursively. The coverage of $ (C_i, v_i) $ is the set of samples that satisfies clause $ C_i $. Finally, we add a default rule $ (C_k, v_\mathsf{default}) $ to $ \Rule_L $ where $ C_k \triangleq  $ true denoting that the clause is satisfied by all samples. We select the default class $ v_\mathsf{default} $ in the following order: (i) if any class(s) is not in the predicted classes $ \{v_i\}_{i=1}^{k-1} $ of the decision list, 
 $ v_{\mathsf{default}} $ is the majority class among missing classes, and (ii)  $ v_{\mathsf{default}} $ is the majority class of the original training set $ (\mathbf{X}, \mathbf{y}) $, otherwise.

\begin{algorithm}
	\caption{Iterative learning of decision lists}
	\label{interpretability_imli_algo:iterative_decision_lists_learning}
	\begin{algorithmic}[1]
		\Procedure{DecisionListLearning}{$ \mathbf{X},\mathbf{y},\lambda, k $}
		\State $ \mathcal{R}_L \leftarrow \{\}$
		\For{$ i \leftarrow 1, \dots, k-1 $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
		\State $ v_i \leftarrow \mathsf{MajorityClass(\mathbf{y})} $ \Comment{$ v_i $ specifies the target class}
		\State $ C_i \leftarrow \mathsf{MiniBatchDNFLearning}( \mathbf{X},\mathbf{y},\lambda, 1, v_i) $ \Comment{Ref. Section~\ref{interpretability_imli_sec:dnf_learning}}

%		\label{interpretability_imli_algo:iterative_CNF_learning_incremental_learning}
		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{Coverage}(\mathbf{X}, C_i) $
%		\label{interpretability_imli_algo:iterative_CNF_learning_coverage}
		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} 
		\State break
		\EndIf
		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L \cup \{(C_i, v_i)\} $
		\State $ \mathbf{X},\mathbf{y}  \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $ 
		\EndFor
		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L  \cup \{(\text{true}, v_{\mathsf{default}} )\}$	\Comment{Default rule}
		\State \Return $ \Rule_L $
		\EndProcedure
	\end{algorithmic}

\end{algorithm}

\subsection{Learning Decision Sets}
We now describe an iterative procedure for learning decision sets. A decision set comprises of an individual clause-class  pair $ (C_i, v_i) $ where $ C_i $ denotes a single clause DNF formula similar to decision lists. In a decision set, a sample can satisfy multiple clauses simultaneously, which is attributed as an \textit{overlapping between clauses}~\cite{lakkaraju2016interpretable}. Concretely, the overlap between two clauses $ C_i $ and $ C_j $ with $ i \ne j $ is the set of samples $ \{\mathbf{X}_l | \mathbf{X}_l \models C_i \wedge \mathbf{X}_l \models C_j\} $ satisfying both clauses. One additional objective in learning a decision set is to minimize the overlap between clauses, as studied in~\cite{lakkaraju2016interpretable}. 
Therefore, along with optimizing accuracy and rule-sparsity, we propose an iterative procedure for decision sets that additionally minimizes the overlap between clauses.

\begin{algorithm}
	\caption{Iterative learning of decision sets}
	\label{interpretability_imli_algo:iterative_decision_sets_learning}
	\begin{algorithmic}[1]
		\Procedure{DecisionSetsLearning}{$ \mathbf{X},\mathbf{y},\lambda, k$}
		\State $ \mathcal{R}_S=\{\} $
		\State $ \mathbf{X}^\text{cc} = \{\}$  \Comment{Contains correctly covered samples}
		\For{$ i \leftarrow 1, \dots, k-1 $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
		\State $ v_i \leftarrow \mathsf{MajorityClass(\mathbf{X},\mathbf{y})} $
		\State $ \mathbf{X}^\text{w} \leftarrow \mathbf{X} \cup \mathbf{X}^\text{cc} $ \Comment{Correctly covered samples are included}
		\State $ \mathbf{y}^\text{w} \leftarrow  \mathbf{y} \cup  \{\neg v_i\}^{|\mathbf{X}^\text{cc}|}$ \Comment{$ \mathbf{X}^\text{cc} $ have complemented class $ \neg v_i $ to minimize overlap}
		\State $ C_i \leftarrow \mathsf{MiniBatchDNFLearning}( \mathbf{X}^\text{w},\mathbf{y}^\text{w},\lambda, 1, v_i) $ 
		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{CorrectCoverage}(\mathbf{X}, \mathbf{y}, C_i, v_i) $
		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} 
		\State break
		\EndIf
		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S \cup \{(C_i, v_i)\} $
		\State $ \mathbf{X},\mathbf{y} \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $
		\State $ \mathbf{X}^\text{cc} = \mathbf{X}^\text{cc} \cup \mathbf{X}'$ 
		\EndFor
		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S  \cup \{(\text{true}, v_{\mathsf{default}} )\}$
		\State \Return $ \Rule_S $
		\EndProcedure
	\end{algorithmic}
\end{algorithm}

In Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}, we present an iterative algorithm for learning a decision set. The iterative algorithm is a modification of separate-and-conquer algorithm  by additionally focusing on minimizing overlaps in a decision set.  Given a training data $ (\mathbf{X}, \mathbf{y}) $, a  regularization parameter $ \lambda $, and the number of clauses $ k $, the core idea of Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning} is to learn a pair $ (C_i,v_i) $ in each iteration, separate covered samples from $ (\mathbf{X}, \mathbf{y}) $, and conquer remaining samples recursively. In contrast to learning decision lists, we have following modifications in Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}.

\begin{itemize}
	\item The first modification is with respect to the definition of coverage for decision sets. Unlike decision lists, we separate samples that are \textit{correctly covered} by $ (C_i, v_i) $ in each iteration. Given  $ (\mathbf{X}, \mathbf{y}) $, the correctly covered samples of $ (C_i, v_i) $ is a set $  \{(\mathbf{X}_l, \mathbf{y}_l)| \mathbf{X}_l \models C_i \wedge \mathbf{y}_l = v_i \} \subseteq (\mathbf{X}, \mathbf{y}) $ of samples that satisfy $ C_i $ and have matching class-label as $ v_i $. 
	\item The second modification is related to the  training dataset considered in each iteration. Let $ (\mathbf{X},\mathbf{y}) $ denote the remaining training set in the current iteration and $ v_i $ be the majority class in  $ (\mathbf{X},\mathbf{y}) $. Hence, $ v_i $ is the target class in the current iteration. In Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}, we learn $ C_i $ on a sample set $ (\mathbf{X}^\text{w},\mathbf{y}^\text{w}) $, where $ \mathbf{X}^\text{w} \triangleq \mathbf{X} \cup \mathbf{X}^\text{cc} $ comprises of both remaining and already covered samples. However, the vector of class labels $ \mathbf{y}^\text{w} \triangleq  \mathbf{y} \cup  \{\neg v_i\}^{|\mathbf{X}^\text{cc}|} $  comprises of remaining class-label vector $ \mathbf{y} $ and complemented class label vector $ \{\neg v_i\}^{|\mathbf{X}^\text{cc}|} $ associated with feature-matrix $ \mathbf{X}^\text{cc} $. By explicitly labeling covered samples as class $ \neg v_i $, the new clause $ C_i $ learns to falsify already covered samples. This heuristic allows us to minimize the overlap of $ C_i $ compared to previously learned clauses $ \{C_j\}_{j=1}^{i-1} $.
\end{itemize}

Finally, the default clause for decision sets is learned similarly as in decision lists.

\newcommand{\maxsatquery}{\ensuremath{\mathsf{ConstructQuery}}}
\newcommand{\createclause}{\ensuremath{\mathsf{ConstructClause}}}
\newcommand{\iterativeClauseLearning}{\ensuremath{\mathsf{IterativeClauseLearning}}}
\let\oldReturn\Return
\renewcommand{\Return}{\State\oldReturn}

### 2022-01-27 chapters/interpretability/imli/preliminaries.tex
imli journal paper added

\section{Preliminaries}
\label{interpretability_imli_sec:preliminaries}

In this section, we first discuss propositional satisfiability and MaxSAT, followed by rule-based classification in machine learning. Throughout the paper, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters $\mathbf{x}$ are reserved for vectors/sets.

\subsection{Propositional Satisfiability}

Let $F$ be a Boolean formula defined over Boolean variables $\mathbf{b} = \{b_1,b_2,\dots ,b_m \}$. A literal or predicate $ v $ is a variable $b$ or its complement $\neg b$, and a clause $ C $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals in the literature, we use clause for both disjunction and conjunction.}  $F$ is in Conjunctive Normal Form (CNF) if $F \triangleq  \bigwedge_i C_i$ is a conjunction of clauses where each clause $C_i \triangleq  \bigvee_j v_j $  is a disjunction of literals. In contrast, $ F $ is in Disjunctive Normal Form (DNF) if $F \triangleq  \bigvee_i C_i$ is a disjunction of clauses where each clause $C_i \triangleq  \bigwedge_j v_j $ is a conjunction of literals. We use $\sigma$ to denote an assignment of variables in  $\mathbf{b}$ where $ \sigma(b_i) \in $ \{true, false\}. A \emph{satisfying assignment} $ \sigma^* $ of $F$ is an assignment  that evaluates $F$  to true and is denoted by $ \sigma^* \models F $. The propositional satisfiability (SAT) problem finds a satisfying assignment $ \sigma^* $ to a CNF formula $ F $  such that $ \forall i, \sigma^* \models C_i $, wherein $ \sigma^*\models C_i $ if and only if $ \exists v_j \in C_i, \sigma^*(v_j) = \text{true} $. Informally, $ \sigma^* $ satisfies at least one literal in each clause of a CNF. 
%\red{Moreover, $ F $ is called SAT if there exists a satisfying assignment of $ F $, and $ F $ is called UNSAT otherwise.}

In this paper, we use true as $ 1 $ and false as $ 0 $ interchangeably. Between two vectors  $\mathbf{u}$ and $\mathbf{v}$ of the same length defined over Boolean variables or constants (such as  $ 0 $ and $ 1 $), we define $\mathbf{u} \circ \mathbf{v} $  to refer to  their inner product. Formally, $\mathbf{u} \circ \mathbf{v} \triangleq \bigvee_{i} (u_{i} \wedge v_{i})$ is a disjunction of element-wise conjunction where $u_{i}$ and $v_{i}$ denote the $ i^\text{th} $ variable/constant of $\mathbf{u}$ and $\mathbf{v}$, respectively. In this context, the conjunction ``$\wedge$'' between a variable and a constant follows the standard interpretation: $b \wedge 0 = 0$ and $b \wedge 1 = b$.  For example, if $ \mathbf{u} = [u_1, u_2, u_3] $ and $ \mathbf{v} = [0,1,1] $, then $ \mathbf{u} \circ \mathbf{v} = u_2 \vee u_3 $.

\subsection{MaxSAT}

The MaxSAT problem is an optimization analog to the SAT problem that finds an optimal assignment satisfying the maximum number of clauses in a CNF. In this work, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula. In a weighted-CNF formula, a weight $ {\weight{C_i}} \in \mathbb{R}^+ \cup \{\infty\} $ is defined over each clause $ C_i $, wherein $C_i$ is called a \emph{hard} clause if $\weight{C_i} = \infty$, and  $C_i$ is  a \emph{soft} clause, otherwise.  To avoid notational clutter, we overload {\weight{\cdot}} to denote the weight of an assignment $ \sigma $. In particular, we define $\weight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $\weight{\sigma} = \Sigma_{i | \sigma \not \models C_i} \weight{C_i}$.

Given a weighted-CNF formula $F \triangleq \bigvee_i C_i$ with weight $ \weight{C_i} $, the weighted-partial MaxSAT problems finds an optimal assignment $\sigma^*$ that achieves the minimum weight. Formally,  $\sigma^* = \MaxSAT(F,\weight{\cdot})$ if $\forall \sigma \neq \sigma^*, \weight{\sigma^*} \leq \weight{\sigma}$. The optimal weight of the MaxSAT problem $  \weight{\sigma^*} $ is infinity ($ \infty $) when at least one hard clause becomes unsatisfied.  Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum. In this paper, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem. The knowledge of the inner workings of {MaxSAT} solvers is  not required for this paper.

\subsection{Rule-based Classification}

We consider a standard binary classification problem where $ \mathbf{X} \in \{0,1\}^{n\times m} $ is a binary feature matrix\footnote{Our framework allows learning on real-valued and categorical features, as discussed in Section~\ref{interpretability_imli_sec:non-binary}.} defined over Boolean features $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$ and $ \mathbf{y} \in \{0,1\}^n $ is a vector of binary class labels.  The $ i^\text{th} $ row of $ \mathbf{X} $, denoted by $ \mathbf{X}_i \in \mathcal{X} $, is called a sample, which is generated from a product distribution $ \mathcal{X} $. Thus, $ \mathbf{X}_i $ is the valuation of features in $ \mathbf{x} $ and $ y_i $ is its class label.  $ \mathbf{X}_i $ is called a positive sample if $ y_i = 1 $, otherwise a negative sample.

A classifier $ \Rule: \mathbf{x} \rightarrow \hat{y} \in  \{0,1\} $ is a  function that takes a Boolean vector as input and outputs the predicted class label $\hat{y}$.  %We define two rules  $ \R_1 $ and $ \R_2 $ to be  equivalent  if $ \forall i,\R_1(\X_i)=\R_2(\X_i) $. 
In a rule-based classifier, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{x} $ such that, if $ \Rule $ becomes true for an input (or an assignment), the prediction $ \hat{y} = 1 $ and $ \hat{y} = 0 $, otherwise.  The goal is to design $\Rule$ not only to approximate the training data, but also to generalize to unseen samples arising from the same distribution. Additionally, we prefer to learn a \emph{sparse} $ \Rule $ in order to favor interpretability. We define the structural complexity of $ \Rule $, referred to as \emph{rule-size}, in terms of the number of literals it contains. Let $\clause(\Rule,i)$ denote  the  $i^\text{th}$ clause of $\Rule$ and $ |\clause(\Rule,i)| $ be  the number of literals in $\clause(\Rule,i)$. Thus, the size of the classifier is $|\Rule| =  \Sigma_i |\clause(\Rule,i)| $, which is the sum of literals in all clauses in $ \Rule $.

For example, let $ \Rule \triangleq (x_1 \vee x_3) \wedge (\neg x_2 \vee x_3) $ be a CNF classifier defined over three Boolean features $ \{x_1, x_2, x_3\} $. For an input $ [0, 1, 1] $, the prediction of $ \Rule  $ is $ 1 $ whereas for an input $ [1, 1, 0] $, the prediction is $ 0 $ because $ \Rule $ is true and false in these two cases, respectively. Moreover, $ |\Rule| = 2 + 2 = 4 $ is the rule-size of $ \Rule $.

\paragraph{Decision Lists.} A decision list is a rule-based classifier consisting of ``if-then-else'' rules. Formally, a decision list~\cite{rivest1987learning} is an ordered list $ \Rule_L $ of pairs $ (C_1, v_1), \dots, (C_k, v_k) $ where $ C_i $ is a conjunction of literals (alternately, a single clause in a DNF formula) and $ v_i \in \{0,1\} $ is  a Boolean class label\footnote{In a more practical setting, $ v_i \in \{0, \dots, N\} $  can be multi-class for $ N \ge 1 $.}. Additionally, the last clause $ C_k \triangleq  $ true, thereby $ v_k $ is the default class. A decision list is defined as a classifier with the following interpretation: for a feature vector $ \mathbf{x} $,  $ \Rule_L(\mathbf{x}) $ is equal to $ v_i $ where $ i $ is the least index such that $ \mathbf{x} \models C_i $. Since the last clause is true, $ \Rule_L(\mathbf{x}) $ always exists. Intuitively, whichever clause (starting from the first) in $ \Rule_L $ is satisfied for an input, the associated class-label is considered as its prediction.

\paragraph{Decision sets.} A decision set is a set of \textit{independent} ``if-then'' rules. Formally,  a decision set $ \Rule_S $ is a set of pairs $ \{(C_1, v_1), \dots, (C_{k-1}, v_{k-1})\}  $ and a  default pair $ (C_k, v_k) $, where $ C_i $ is\textemdash similar to a decision list\textemdash a conjunction of literals and $ v_i \in \{0,1\} $ is a Boolean class label. In addition, the last clause $ C_k \triangleq  $ true and $ v_k $ is the default class. For a  decision set, if an input $ \mathbf{x} $ satisfies one clause, say $ C_i $, then the prediction is $ v_i $. If $ \mathbf{x} $ satisfies no clause, then the prediction is the default class $ v_k $. Finally, if $ \mathbf{x} $ satisfies  $ \ge 2 $ clauses, $ \mathbf{x} $ is assigned a class using a tie-breaking~\cite{lakkaraju2016interpretable}.

### 2022-01-27 chapters/interpretability/imli/problem_formulation.tex
imli journal paper added

 \section{Problem Formulation}
 \label{interpretability_imli_sec:problem}
Given (i) a binary feature matrix $ \mathbf{X} \in \{0,1\}^{n \times m} $ with $ n $ samples and $ m $ features, (ii) a class-label vector $ \mathbf{y} \in \{0,1\}^n $, (iii) a positive integer $ k \ge 1 $ denoting the number of clauses, and (iv) a regularization parameter $ \lambda \in \mathbb{R}^+ $, we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.

 Our goal is to learn classifiers that balance two 
 goals: of  being accurate but also interpretable.  
 Various notions of interpretability have been proposed in the context of   classification problems. A common proxy for interpretability in the context of decision rules 
 is the sparsity of rule. For instance, a rule involving fewer literals is highly interpretable.  In this work, we minimize the total number of literals in all clauses, as discussed in Section~\ref{interpretability_imli_sec:preliminaries}, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $. Let $ \Rule $ classify all samples correctly during training. Among all the classification rules that classify all samples correctly,  we choose the sparsest (most interpretable) such $ \Rule $.

 \[
 \min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y_i=\mathcal{R}(\mathbf{X}_i)
 \]

 In practical classification tasks, perfect classification is unlikely. Hence, we need to balance interpretability with classification error.  Let $ \mathcal{E}_\mathbf{X} = \{(\mathbf{X}_i,y_i) | y_i \ne \mathcal{R}(\mathbf{X}_i) \} $   be  the set of samples in $ \mathbf{X} $ that are misclassified  by $ \mathcal{R} $. Therefore, we balance between classification-accuracy and rule-sparsity and optimize the following function.\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g., to penalize  false-alarms more than mis-detects), and to allow instance weights (per sample).}

 \begin{equation}
  \label{interpretability_imli_eq:obj}
 \min\limits_{\mathcal{R}} \;  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}|
 \end{equation}

Higher  values of $ \lambda $ generate a rule with a smaller rule-size but of more training errors, and vice-versa. Thus, $ \lambda $ can be tuned to trade-off between accuracy and interpretability for a rule-based classifier, which we experiment extensively in Section~\ref{interpretability_imli_sec:experiments}.

### 2022-01-27 chapters/interpretability/imli/related_work.tex
imli journal paper added

\section{Related Works}
\label{interpretability_imli_sec:related}

The progress in designing interpretable rule-based classifiers finds its root in the development of decision trees \cite{bessiere2009minimising,quinlan1986induction,quinlan1987simplifying}, decision lists \cite{rivest1987learning}, classification rules \cite{cohen1995fast} etc.  In early works, the focus was to improve the efficiency and scalability of the model rather than designing models that are interpretable. For example,  decision rule approaches such as  C4.5 rules \cite{quinlan2014}, CN2~\cite{ClarkN1989}, RIPPER \cite{cohen1995fast}, and SLIPPER  \cite{CohenS1999} 
rely on heuristic-based branch pruning and ad-hoc local criteria e.g., maximizing information gain, coverage, etc.

Recently, several optimization frameworks have been proposed for interpretable classification, where both accuracy and rule-size are optimized during training. For example,~\cite{malioutov2013exact} proposed exact learning of rule-based classifiers based on Boolean compressed sensing using a linear programming formulation.~\cite{su2016learning} presented two-level Boolean rules, where the trade-off between classification accuracy and interpretability is studied. In their work, the Hamming loss is used to characterize prediction accuracy, and sparsity is used to characterize the interpretability of rules.~\cite{wang2015falling} proposed a Bayesian optimization framework for learning falling rule lists, which is an ordered list of if-then rules. Other similar approaches based on Bayesian analysis for learning classification rules are~\cite{letham2015interpretable,wang2017bayesian}. Building on custom discrete optimization techniques,~\cite{angelino2017learning} proposed an optimal learning technique for decision lists using a branch-and-bound algorithm. In a separate study,~\cite{lakkaraju2016interpretable} highlighted the importance of decision sets over decision lists in terms of interpretability and considered a sub-modular optimization problem for learning a near-optimal solution for decision sets. Our proposed method for interpretable classification, however, relies on the improvement in formal methods over the decades, particularly the efficient CDCL-based solution for satisfiability (SAT) problems~\cite{silva2003grasp}.

Formal methods, particularly SAT and its variants, have been deployed in interpretable classification in recent years. In the context of learning decision trees, SAT and MaxSAT-based solutions are proposed by~\cite{alos2021learning,janota2020sat,narodytska2018learning,shati2021sat}. In addition, researchers have applied SAT for learning explainable decision sets~\cite{ignatiev2018sat,ignatiev2021scalable,schidler2021sat,yu2020computing}. In most cases, SAT/MaxSAT solutions are not sufficient in solving large-scale classification tasks because of the $ \mathrm{NP} $-hardness of the underlying problem. This observation motivates us in combining MaxSAT with more practical algorithms such as incremental learning.

Incremental learning has been studied in improving the scalability of learning problems, where data is processed in parts and results are combined to use lower computation overhead. In case of non-interpretable classifiers such as SVM, several solutions adopting incremental learning are available~\cite{syed1999incremental,ruping2001incremental}. For example,~\cite{cauwenberghs2001incremental} proposed an online recursive algorithm for SVM that learns one support-vector at a time. Based on radial basis kernel function,~\cite{ralaivola2001incremental} proposed a local incremental learning algorithm for SVM. In the context of deep neural networks, stochastic gradient descent is a well-known convex optimization technique\textemdash a variant of which includes computing the gradient on mini-batches~\cite{hinton2012neural,li2014efficient,masters2018revisiting}.  Federated learning, on the other hand, decentralizes training on multiple local nodes based on local data samples with only exchanging learned parameters to construct a global model in a central node~\cite{konevcny2015federated,konevcny2016federated}. Another notable technique is Lagrangian relaxation that decomposes the original problem into several sub-problems, assigns Lagrangian multipliers to make sure that sub-problems agree, and iterates by solving sub-problems and adjusting weights based on disagreements~\cite{fisher1981lagrangian,johnson2007lagrangian,lemarechal2001lagrangian}. To the best of our knowledge, our method is the first method that unifies incremental learning with MaxSAT based formulation to improve the scalability of learning rule-based classifiers.

Classifiers that are interpretable by design can be applied to improve the explainability of complex black-box machine learning classifiers. There is rich literature on extracting decompositional and pedagogical rules from non-linear classifiers such as support vector machines~\cite{barakat2004learning,barakat2005eclectic,diederich2008rule,martens2008rule,nunez2002rule} and neural networks~\cite{augasta2012rule,hailesilassie2016rule,setiono1995understanding,sato2001rule,zilke2016deepred,zhou2004rule}. In recent years, local model-agnostic approaches for explaining black-box classifiers  are  proposed by learning surrogate simpler classifiers such as rule-based classifiers~\cite{guidotti2018local,pastor2019explaining,rajapaksha2020lormika,ribeiro2018anchors}. The core idea in local approaches is to use a rule-learner that can classify synthetically generated neighboring samples with class labels provided by the black-box classifier. To this end, our framework {\imli} can be directly deployed as an efficient rule-learner and can explain the inner-working of black-box classifiers by generating interpretable rules.

### 2022-01-30 chapters/fairness/justicia/preliminaries.tex
preliminaries of justicia updated

@@
 \label{fairness_justicia_sec:preliminaries}
+
 In this section, we define different fairness metrics for a supervised learning problem. Following that, we discuss Stochastic Boolean Satisfiability (SSAT) problem.
@@
 \subsection{Fairness Metrics for Machine Learning}\label{fairness_justicia_sec:fairness}
-Let us represent a dataset $D$ as a collection of triples $(X, A, Y)$ sampled from an underlying data generating distribution $\mathcal{D} $.
-$X \triangleq \lbrace \nonsensitive_1, \ldots, \nonsensitive_m\rbrace \in \mathbb{R}^m $ is the set of non-protected (or non-sensitive) attributes.
-$A \triangleq \lbrace A_1, \ldots, A_n\rbrace$ is the set of categorical protected attributes.
-$Y$ is the binary label (or class) of $(X, A)$. 
-A compound protected attribute $\mathbf{a} = \{a_1, \ldots, a_n\}$ is a valuation to all $A_i$'s and represents a \textit{compound protected} group.
-For example, $A = \{\textrm{race}, \textrm{sex}\}$, where $\textrm{race} \in \{\textrm{Asian}, \textrm{Colour}, \textrm{White}\}$ and  $\textrm{sex} \in \{\textrm{female},  \textrm{male}\}$.
-Thus, $\mathbf{a} =  \{\textrm{Colour}, \textrm{female}\}$ is a compound protected group. 
-We define $\alg \triangleq \Pr(\hat{Y}|X, A)$ to be a binary classifier trained from samples in the distribution $\mathcal{D} $. 
-Here, $\hat{Y}$ is the predicted label (or class) of the corresponding data.
 
-As we illustrated in Example~\ref{fairness_justicia_example:intro}, a classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound protected groups over others~\cite{chouldechova2020snapshot}.
-Now, we describe two family of fairness metrics that compute bias induced by a classifier and are later verified by {\justicia}.
+We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
+We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of triples $ (\nonsensitive, \sensitive, Y) $ generated from an underlying distribution $\mathcal{D}$.
+$ \nonsensitive \triangleq \{X_1, \dots, X_{\numnonsensitive}\} $ are non-sensitive features whereas $ \sensitive \triangleq \{A_1, \dots, A_{\numsensitive}\} $ are categorical sensitive features.
+$Y \in \{0,1\}$ is the binary label (or class) of $(\nonsensitive,\sensitive)$.
+Each non-sensitive feature $ X_i$ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each sensitive feature $ A_j \in \{0, \dots, N_j\}  $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}.
+We use $ (\mathbf{x}, \mathbf{a}) $ to denote the feature-values of  $ (\nonsensitive, \sensitive) $.
+For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{m_2}] $ is called a \textit{compound sensitive group}.
+For example, consider $ \sensitive = $ \{race, sex\} where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}.
+Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group.
+We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \hat{Y} $.
+Here, $\hat{Y} \in \{0,1\}$ is the predicted class of $ (\nonsensitive, \sensitive) $.
 
-\begin{comment}
-$ X $ is the set of non-protected attributes. Each non-protected attribute $ x_i \in X $ takes the valuation as real numbers or categories. 
-We consider a multi-protected attributes setting where $ A $ is the set of protected (or sensitive) attributes.
-Each protected attribute $ a_i \in A $ is a categorical attribute with at least two categories.
-$ y $ is the binary class label of the data sample.
+As we illustrated in Example~\ref{fairness_justicia_example:intro}, a classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
+In the following, we describe two well-known fairness definitions verified by {\justicia}: group fairness and causal fairness.
+\red{In each definition, the fairness metric relies on computing the conditional Positive Predictive Value (PPV) of the classifier, denoted by $ \Pr[\hat{Y} = 1 | \mathsf{C}] $, where different fairness metrics are constituted based on the conditional $ \mathsf{C} $.}
 
-Based on the valuation of the protected attributes, the samples are  tagged as being in different protected groups, i.e., 
-the privileged group(s)  and the unprivileged group(s).
-For instance, let the dataset contain two protected attributes:  race and gender.
-The attribute race can be white, black or Asian and the attribute gender can be male or female.
-For simplicity, we assume that the protected group white-male is the most privileged group and all other protected groups (attribute combinations) refer to unprivileged groups with black-female being the most unprivileged group.
-Generally the dataset contains bias that results in the discrimination between the privileged and unprivileged groups.
-In this paper, we do not require the knowledge of privileged and unprivileged groups beforehand.
-We instead ask for the set $ A $ of protected attributes from which we derive privileged and unprivileged groups.
-\end{comment}
-
-\begin{comment}
-We now consider a more specific setting where the samples are divided into privileged and unprivileged groups based on the valuation of $ a $, i.e, when $ a=1 $, the sample is included into a privileged group and vice versa.
-\end{comment}
-
-\subsubsection{Independence Metrics of Fairness.} 
-The \textit{independence (or calibration) metrics} of fairness state that the output of the classifier should be independent of the compound protected group.
-A notion of independence is referred to \textit{group fairness} that specifies an \textit{equal positive predictive value (PPV) across all compound protected groups} for an algorithm $\alg$, i.e., $\Pr[\hat{Y} = 1 | A = \mathbf{a}, \alg] = \Pr [\hat{Y} =1 | A = \mathbf{b}, \alg], \forall \mathbf{a}, \mathbf{b} \in A$.
-Since satisfying group fairness exactly is hard, relaxations of group fairness, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.
+\subsubsection{Group Fairness.} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\justicia} verifies independence and separation metrics. 
+The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups.
+Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier $\alg$, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in \sensitive$.
+Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.
 
@@
 \[
-	\min_{\mathbf{a} \in A} \Pr[\hat{Y} =1 | \mathbf{a}, \alg]  \ge (1 - \epsilon) \max_{\mathbf{b} \in A} \Pr[\hat{Y} =1 | \mathbf{b}, \alg].
+\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}].
 \]
-Another popular relaxation of group fairness, \textit{statistical parity} (SP) measures the difference of PPV among the compound groups, and prescribe this to be near zero.
+Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the maximum difference of PPVs among sensitive groups, and prescribe this to be near zero.
 Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $,
 \[
-\max_{\mathbf{a}, \mathbf{b} \in A}|\Pr[\hat{Y} =1 | \mathbf{a}, \alg] - \Pr [\hat{Y} = 1| \mathbf{b}, \alg]| \le \epsilon.
+\max_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}] \le \epsilon.
 \]
-For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier $\alg$.
+For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier.
 
-\begin{comment}These two metrics of fairness has a disadvantage that a fully accurate classifier with different base rates (i.e., the proportion of actual positive prediction) in different protected groups may be considered unfair.
-As a result, similar individuals from different protected groups may be treated differently in order to achieve group fairness{\textemdash} such treatment is prohibited by law in some cases.
-\end{comment}
+In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier is independent of the sensitive features $\sensitive$ given the class labels $Y$.
+In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the maximum difference of false positive rates \red{(FPR)} and true positive rates \red{(TPR)} conditioned on sensitive groups.
+Lower value of equalized odds indicates better fairness.
+A classifier satisfies $\epsilon$-equalized odds when the following conditions hold.
+\begin{align*}
+	&\max_{\mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - 	\min_{\mathbf{a}} \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 0] \le \epsilon\\
+	&\max_{\mathbf{a}}\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - 	\min_{\mathbf{a}}  \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 1] \le \epsilon
+\end{align*}
 
-\subsubsection{Separation Metrics of Fairness.}
-In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive attributes $A$ given the actual class labels $Y$.
-In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound protected groups.
-Lower value of equalized odds indicates better fairness.
-A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound protected groups $\mathbf{a}, \mathbf{b} \in A$,
-\begin{equation}
-\begin{split}
-|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1| A = \mathbf{b}, Y = 0]| &\le \epsilon, \\ 
-|\Pr[\hat{Y} =1 | A = \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1| A = \mathbf{b}, Y = 1]| &\le \epsilon.\notag
-\end{split}
-\end{equation}
+\subsubsection{Path-specific Causal Fairness.}
+Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.
+We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $.
+This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
+For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as
+\[
+\max_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] -  \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ] \le \epsilon.
+\]
 
-%\begin{comment}
-%In contrast to disparate impact and statistical parity difference metrics, a fully accurate classifier will necessarily satisfy the two constraints in equalized odds measure.
+Therefore, PCF constrains that $ \hat{Y} $ is not directly dependent of $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through $ \mediator $.
+PCF is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
 
-%\textbf{Group fairness.}  Group fairness requires the probability for an individual to be assigned the favorable outcome to be equal across the privileged  and unprivileged groups.
-Throughout the manuscript, by favorable outcome we refer to the predictor $ \hat{y} = 1 $ in  the standard binary classification setting.
-Now group fairness is formally defined in the following.
-%
-%\[
-%\Pr(\hat{y} =1 | a =1  ) = \Pr (\hat{y} = 1| a = 0)
-%\]
-%
-%By abusing notations, we use the following shorthand. 
-%
-%\[
-%\Pr_1(\hat{y} =1) = \Pr_0(\hat{y} = 1)
-%\]
-%
-%
-%
-%
-%In practice, the above definition can be relaxed. Thus we define the notion of $ \epsilon $-\textit{group fairness} in the following. 
-%\[
-% |\Pr_1(\hat{y} =1) - \Pr_0(\hat{y} = 1) |< \epsilon
-%\]
-%
-%\end{comment}
-
-In this paper, we formulate verifying the aforementioned independence and separation metrics of fairness as stochastic Boolean satisfiability (SSAT) problem, which we define next.
+\begin{example}
+	Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator).
+It is practical to consider that gender $ \in $ \{male, female\} can affect the college-participation of individuals, and all three features are determining factors for the hiring process.
+Let `male' be the most favored group by the classifier, for instance.
+Path-specific causal fairness (PCF) ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
+She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.
+Therefore, PCF measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
+\end{example}
 
@@
 %\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
-Let $ \sigma $ be an assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $ where $ 1 $ is logical TRUE and $ 0 $ is logical FALSE.
-The propositional  \textit{satisfiability} problem (SAT)~\cite{biere2009handbook} finds an assignment $ \sigma $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$.
-In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is concerned with the  probability of the satisfaction of the formula $\phi$.
+Let $ \sigma $ be an assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $.
+The propositional  \textit{satisfiability} problem (SAT) finds an assignment $ \sigma $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true).
+In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is computes the  probability of the satisfaction of the formula $\phi$ defined on \textit{quantified} Boolean variables.
 An SSAT formula is of the form
@@
 \end{equation}
-where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers over the Boolean variable $\bool_i$ and $\phi$ is a quantifier-free CNF formula.
+where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula.
+In case of a randomized quantifier $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
 In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.
-In case of randomized quantification $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
-Given an SSAT formula $ \Phi $, let $ \bool $ be the outermost variable in the prefix.
-The satisfying probability of $ \Phi $ can be computed by the following \textit{rules}:
+Let $ \bool $ be the outermost variable in the prefix.
+The semantics of SSAT formulas are defined recursively in the following.
 \begin{enumerate}
-	\item $ \Pr[\text{TRUE}] = 1 $,  $ \Pr[\text{FALSE}] = 0 $, 
+	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
 	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
 	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
-	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{TRUE}$,
+	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
 \end{enumerate}
-where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the formula $ \phi $ with $ 1 $ and $ 0 $ respectively.
+where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the formula $ \phi $ with $ 1 $ and $ 0 $, respectively.
 In this paper, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
 In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
-\begin{lemma}{~\cite{littman2001stochastic}}
-	\label{fairness_justicia_thm:complexity}
-	Solving the ER-SSAT and RE-SSAT problems are $\mathrm{NP}^{\mathrm{PP}}$ hard.
-\end{lemma}
 
-The problem of SSAT and its variants have been pursued by theoreticians and practitioners alike for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
+\begin{remark}
+	ER-SSAT problem is $\mathrm{NP}^{\mathrm{PP}}$-hard whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
+\end{remark}
+
+The problem of SSAT and its variants have been pursued by theoreticians and practitioners for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
 We refer the reader to~\cite{lee2017solving,lee2018solving} for detailed survey.
-It is worth remarking that the past decade has witnessed a significant performance improvements thanks to close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
+It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
 
@@
 	Let $ \phi(E,R) $ be a CNF formula over Boolean variables $ E \cup R $ where each $ e_i \in E $ is an exist quantified variable and each $ r_j \in R $ is a randomized quantified variable.
-For each $ r_j $, $ p_j = \Pr(\sigma(r_j) = 1) $ denotes the probability of $ r_j $ assigning to TRUE.
+For each $ r_j $, $ p_j = \Pr(\sigma(r_j) = 1) $ denotes the probability of $ r_j $ assigning to true.
 Now an \textbf{RE}-SSAT formula is defined as

### 2022-01-31 chapters/fairness/justicia/framework.tex
working on enumeration approach

@@
-\section{{\justicia}: An SSAT Framework to Verify Fairness Metrics}
+\section{{\justicia}: An SSAT-based Fairness Verifier}
 \label{fairness_justicia_sec:framework}
-In this section, we present the primary contribution of this paper, {\justicia}, which is an SSAT-based framework for verifying independence and separation metrics of fairness.
-
-Given a binary classifier $\alg$ and a probability distribution over dataset $(X,A,Y) \sim \mathcal{D} $, our goal is to verify whether $\alg$ achieves independence and separation metrics with respect to the distribution $\mathcal{D}$.
-We  focus on a classifier that can be translated to a CNF formula of Boolean variables $\mathbf{B} $.
-The probability $ p_i $ of $\bool_i \in \mathbf{B}$ being assigned to $1$ is induced by the data generating distribution $\mathcal{D}$. 
-In order to verify fairness metrics in compound protected groups, we discuss an enumeration-based approach and an equivalent learning-based approach. 
-%We conclude this section by proposing a conditional distribution based enumeration for compound protected groups in Section~\ref{fairness_justicia_sec:cond_ssat}. 
-We then provide a theoretical analysis for a high-probability error bound on the fairness metric and conclude with extension of {\justicia} in practical settings.
+In this section, we present the primary contribution of this paper, {\justicia}, which is an SSAT-based framework for verifying multiple group and causal fairness metrics.
+Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness definition, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.
+Additionally, if a fairness threshold $ \epsilon \in [0,1] $ is provided, {\justicia} verifies whether the classifier is $ \epsilon $-fair by comparing $ f $ with $ \epsilon $.
+In {\justicia}, we focus on classifiers that can be represented as a CNF formula defined over a set of Boolean variables.
+Additionally, for each variable, we query the distribution $ \mathcal{D} $ to derive the marginal probability of assigning to $ 1 $.
+In this section, we discuss two equivalent approaches for fairness verification based on SSAT-based encodings: enumeration approach and inferring approach.
+In both approaches, we verify fairness with the presence of compound sensitive groups.
+We then provide a theoretical analysis for a high-probability error bound on the fairness metric and conclude with an extension of {\justicia} in practical settings.
 
@@
 We first state the problem formally in Section~\ref{fairness_justicia_sec:problem_statement}. 
-To verify fairness metrics in compound protected groups, we discuss an enumeration approach in Section~\ref{fairness_justicia_sec:enumeration_ssat} and an equivalent but more efficient learning approach in Section~\ref{fairness_justicia_sec:learn_ssat}.
-We conclude this section by proposing a conditional distribution based enumeration for compound protected groups in Section~\ref{fairness_justicia_sec:cond_ssat}.
+To verify fairness metrics in compound sensitive groups, we discuss an enumeration approach in Section~\ref{fairness_justicia_sec:enumeration_ssat} and an equivalent but more efficient learning approach in Section~\ref{fairness_justicia_sec:learn_ssat}.
+We conclude this section by proposing a conditional distribution based enumeration for compound sensitive groups in Section~\ref{fairness_justicia_sec:cond_ssat}.
 
@@
 
-\subsection{Evaluating Fairness with RE-SSAT Encoding}
+\subsection{Enumeration Approach  using RE-SSAT encoding}
 \label{fairness_justicia_sec:enumeration_ssat}
-In order to verify independence and separation metrics, the core component of {\justicia} is to compute the positive predictive value $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$ for a compound protected group $\mathbf{a}$.
-For simplicity, we  initially make some assumptions and discuss their practical relaxations later in this section.
-We first assume the classifier $\alg$ is representable as a CNF formula, namely $\phi_{\hat{Y}}$, such that $ \hat{Y} = 1 $ when $ \phi_{\hat{Y}}$ is satisfied and  $\hat{Y} =0$ otherwise.
-Since a Boolean CNF classifier is defined over Boolean variables, we  assume all attributes in $X$ and $A$ to be Boolean.
-Finally, we assume independence of non-protected attributes on protected attributes and $p_i $ is the  probability of the attribute $ \nonsensitive_i $ being assigned to $ 1 $ for any  $\nonsensitive_i \in X  $.
+In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional PPV of the classifier, $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ based on an appropriately designed SSAT formula.
+For simplicity, we  initially make assumptions on the classifier $ \alg $ and discuss practical relaxations later in this section.
+We first assume $\alg$ to be represented as a CNF formula, denoted by $\phi_{\hat{Y}}$, such that the prediction $ \hat{Y} = 1 $ when $ \phi_{\hat{Y}}$ is satisfied and  $\hat{Y} =0$ otherwise.
+Additionally, all features $ \nonsensitive \cup \sensitive $ are assumed to be Boolean variables.
+Finally, we consider independence probability assumption of non-sensitive features on sensitive features such that $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of the non-sensitive feature $ X_i $.
 
-Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$.
-In the prefix of $ \Phi_{\mathbf{a}} $,  all non-protected Boolean attributes in $X$ are assigned randomized quantification and they are followed by the protected Boolean attributes in $ A $ with existential quantification.
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
+
+Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the conditional probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$.
+In the prefix of $ \Phi_{\mathbf{a}} $,  all non-sensitive Boolean features in $X$ are assigned randomized quantification and they are followed by the sensitive Boolean features in $ A $ with existential quantification.
 The CNF formula $ \phi $ in $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\hat{Y} = 1 | A = \mathbf{a}] $.
-In order to encode the conditional $ A = \mathbf{a} $, we take the conjunction of the Boolean variables in $ A $ that symbolically specifies the compound protected group $ \mathbf{a} $.
-For example, we represent two protected attributes: race $ \in $ \{White, Colour\} and sex $ \in $ \{male, female\} by the Boolean variables $ R $ and $ S $  respectively.
+In order to encode the conditional $ A = \mathbf{a} $, we take the conjunction of the Boolean variables in $ A $ that symbolically specifies the compound sensitive group $ \mathbf{a} $.
+For example, we represent two sensitive features: race $ \in $ \{White, Colour\} and sex $ \in $ \{male, female\} by the Boolean variables $ R $ and $ S $  respectively.
 Hence, the compound groups $\{\textrm{White}, \textrm{male}\}$ and $\{\textrm{Colour}, \textrm{female}\}$ are represented by $ R \wedge S $ and $ \neg R \wedge \neg S $, respectively.
@@
 \begin{split}
-	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m}_{\text{non-protected attributes}},  &\underbrace{\exists \sensitive_1,\dots, \exists \sensitive_n}_{\text{protected attributes}},
+	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m}_{\text{non-sensitive features}},  &\underbrace{\exists \sensitive_1,\dots, \exists \sensitive_n}_{\text{sensitive features}},
 \phi_{\hat{Y}} \wedge (A=\mathbf{a}).\notag
@@
 In $ \Phi_{\mathbf{a}} $, the existentially quantified variables $ \sensitive_1, \dots, \sensitive_n $ are assigned values  according  to the constraint $ A=\mathbf{a} $.
-\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Therefore, by solving the SSAT formula $ \Phi_{\mathbf{a}} $,  the SSAT solver finds the probability $ \Pr[\Phi_{\mathbf{a}}] $ for the protected group $ A=\mathbf{a} $ given the random values of $ \nonsensitive_1, \dots, \nonsensitive_m $, which is the PPV of the protected group $\mathbf{a} $ for the distribution $ \mathcal{D} $ and algorithm $\alg$.
+\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Therefore, by solving the SSAT formula $ \Phi_{\mathbf{a}} $,  the SSAT solver finds the probability $ \Pr[\Phi_{\mathbf{a}}] $ for the sensitive group $ A=\mathbf{a} $ given the random values of $ \nonsensitive_1, \dots, \nonsensitive_m $, which is the PPV of the sensitive group $\mathbf{a} $ for the distribution $ \mathcal{D} $ and algorithm $\alg$.
 
-For simplicity, we have described computing the PPV of each compound protected group without considering the correlation between the protected and non-protected attributes.
-In reality, correlation exists between the protected and non-protected attributes.
-Thus, the non-protected attributes may have different conditional distributions for different protected groups.
+For simplicity, we have described computing the PPV of each compound sensitive group without considering the correlation between the sensitive and non-sensitive features.
+In reality, correlation exists between the sensitive and non-sensitive features.
+Thus, the non-sensitive features may have different conditional distributions for different sensitive groups.
 We incorporate these conditional distributions in RE-SSAT encoding by evaluating the conditional probability $ p_i = \Pr[\nonsensitive_i =\text{TRUE}| A=\mathbf{a}] $ instead of the independent probability $\Pr[\nonsensitive_i =\text{TRUE}]$ for any $\nonsensitive_i\in \nonsensitive$.
@@
 	\label{fairness_justicia_example:re_ssat}
-	Here, we illustrate the RE-SSAT formula for calculating the PPV for the protected group `age $ \ge 40 $' in the decision tree of Figure~\ref{fairness_justicia_fig:fair_example}.
+	Here, we illustrate the RE-SSAT formula for calculating the PPV for the sensitive group `age $ \ge 40 $' in the decision tree of Figure~\ref{fairness_justicia_fig:fair_example}.
 We assign three Boolean variables $ F,I,J $ for the three nodes in the tree such that the literal $ F,I,J $ denote `fitness $ \ge 0.61 $', `income $ \ge 0.29 $', and `income $ \ge 0.69 $', respectively.
-We consider another Boolean variable $A$  where the literal $ A $ represents the protected group `age $ \ge 40 $'.
+We consider another Boolean variable $A$  where the literal $ A $ represents the sensitive group `age $ \ge 40 $'.
 Thus, the CNF formula  for the decision tree is $ (\neg F \vee I) \wedge (F \vee J) $.
 From the distribution in Figure~\ref{fairness_justicia_fig:fair_example}, we get $ \Pr[F] = 0.41, \Pr[I] = 0.93 $, and $ \Pr[J] = 0.09 $.
-Given this information, we calculate the PPV for the protected group `age $ \ge 40 $' by solving the RE-SSAT formula:
+Given this information, we calculate the PPV for the sensitive group `age $ \ge 40 $' by solving the RE-SSAT formula:
 	\begin{equation}
@@
 Similarly, to calculate the PPV for the group `age $ < 40 $', we replace the unit (single-literal) clause $ A $ with $ \neg A $ in the CNF in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $ where $ \Pr[\Phi_{\neg A}] = 0.43 $.
-	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of $A$ and $\neg A$, both age groups demonstrate equal PPV as the protected attribute is not explicitly present in the classifier. 
-	However, there is an implicit bias in the data distribution for different protected groups and the classifier unintentionally learns it. 
+	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of $A$ and $\neg A$, both age groups demonstrate equal PPV as the sensitive feature is not explicitly present in the classifier. 
+	However, there is an implicit bias in the data distribution for different sensitive groups and the classifier unintentionally learns it. 
 	To capture this implicit bias, we calculate the conditional probabilities  $ \Pr[F|A] = 0.01, \Pr[I|A] = 0.99 $, and $ \Pr[J|A] = 0.18 $ from the distribution.
@@
 For `age $ < 40 $',  we similarly obtain $ \Pr[F|\neg A] = 0.82, \Pr[I|\neg A] = 0.88 $, and $ \Pr[J|\neg A] = 0.01 $, and thus  $ \Pr[\Phi_{\neg A}] = 0.72 $.
-	Therefore, presented RE-SSAT encoding detects the discrimination of the classifier among different protected groups.
+	Therefore, presented RE-SSAT encoding detects the discrimination of the classifier among different sensitive groups.
 An astute reader would observe that $I$ and $J$ are not independent.
@@
 \label{fairness_justicia_sec:learn_ssat}
-In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the protected attributes. 
+In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the sensitive features. 
 Therefore, the enumeration approach may suffer from scalability issues. 
@@
 \end{equation}
-%In the prefix of the ER-SSAT formula $ \Phi_{\mathsf{ER}} $, the protected variables are  existentially quantified and they are followed by randomized quantified non-protected variables, which is in reverse order  in Eq.~\eqref{fairness_justicia_eq:re}.
-The CNF formula $\phi_{\hat{Y}}$ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound protected group.
+%In the prefix of the ER-SSAT formula $ \Phi_{\mathsf{ER}} $, the sensitive variables are  existentially quantified and they are followed by randomized quantified non-sensitive variables, which is in reverse order  in Eq.~\eqref{fairness_justicia_eq:re}.
+The CNF formula $\phi_{\hat{Y}}$ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound sensitive group.
 Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_n = a^{\max}_n $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum.
 Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_n \}$ achieving the highest PPV. 
-%Since an assignment to $\{\sensitive_1, \dots, \sensitive_n\}$ uniquely denotes a compound protected group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.
+%Since an assignment to $\{\sensitive_1, \dots, \sensitive_n\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.
 
 \subsubsection{Learning the Least Favored Group.}
-In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-protected variables $ \nonsensitive_1, \dots, \nonsensitive_m $.
-In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the protected variables and randomized quantification over the non-protected variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
+In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ \nonsensitive_1, \dots, \nonsensitive_m $.
+In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the sensitive variables and randomized quantification over the non-sensitive variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
 \begin{equation}
@@
 %\begin{equation}
-%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_n}_{\text{protected attributes}},
-%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m}_{\text{non-protected attributes}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
+%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_n}_{\text{sensitive features}},
+%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m}_{\text{non-sensitive features}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
 %\label{fairness_justicia_eq:er_complement}
@@
 \end{lemma}
-As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the protected attributes $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_n\}$ that maximizes $\Phi'_{\mathsf{ER}}$. 
+As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the sensitive features $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_n\}$ that maximizes $\Phi'_{\mathsf{ER}}$. 
 If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$.
@@
 	\label{fairness_justicia_example:er_ssat}
-	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple protected groups.
-As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another protected group `sex $ \in $ \{male, female\}'.
+	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple sensitive groups.
+As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another sensitive group `sex $ \in $ \{male, female\}'.
 Consider a Boolean variable $ S $ for `sex' where the literal $ S $ denotes `sex = male'.
-With this new protected attribute, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ A,F,I,J $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
+With this new sensitive feature, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ A,F,I,J $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
 	Hence, we obtain the ER-SSAT formula of $\alg$ to learn the most favored group:
@@
 	\label{fairness_justicia_lm:equivalence}
-	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound protected group $ \mathbf{a} \in A $.
+	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound sensitive group $ \mathbf{a} \in A $.
 If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
@@
 \begin{comment}
-\red{Let $ \bool_i $ and $ \bool_j $ be two Boolean (non-protected) attributes and we want to encode their pair-wise correlation in {\justicia}.
+\red{Let $ \bool_i $ and $ \bool_j $ be two Boolean (non-sensitive) features and we want to encode their pair-wise correlation in {\justicia}.
 For an assignment to $ \bool_i, \bool_j $, we consider a new Boolean variable $ v $ and add a constraint $ v \leftrightarrow \bool_i \wedge \bool_j $ in the CNF $ \phi $ of the SSAT formula $ \Phi $.
 Additionally, $ v $ is given a randomized quantification in the prefix of $ \Phi $ and the probability of $ v $ is calculated as the probability of both $ \bool_i $ and $ \bool_j $ assigning to $ 1 $ in the distribution $ \mathcal{D} $.
-Note that, to encode the total correlation of all Boolean attributes, the mentioned approach introduces exponentially (with the number of attributes) many new variables and add an exponential number of constraints to $ \phi $ with an aim of computing a more precise satisfying probability of $ \Phi $.
-Optionally, one can encode the correlation among a selected set of attributes of interest in {\justicia}.
+Note that, to encode the total correlation of all Boolean features, the mentioned approach introduces exponentially (with the number of features) many new variables and add an exponential number of constraints to $ \phi $ with an aim of computing a more precise satisfying probability of $ \Phi $.
+Optionally, one can encode the correlation among a selected set of features of interest in {\justicia}.
 }

### 2022-01-31 chapters/fairness/justicia/introduction.tex
working on enumeration approach

@@
 Figure~\ref{fairness_justicia_fig:fair_example}) selects a person above and below $40$ years with probabilities $0.18$ and $0.72$ respectively.
-This simple example demonstrates that even if an ML algorithm does not explicitly learn to differentiate on the basis of a sensitive attribute, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
+This simple example demonstrates that even if an ML algorithm does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
 \end{example}
@@
 Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same.
-Sufficiency is satisfied if the eligibility is independent of their age given their attributes are the same.
+Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
 Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}.
@@
 %VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
-%These papers operate for a specific Boolean sensitive attribute.
-These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the attributes in the dataset and a model of a suitable form, and their objective is to verify fairness w.r.t.
+%These papers operate for a specific Boolean sensitive feature.
+These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a model of a suitable form, and their objective is to verify fairness w.r.t.
 the distribution and the model.
-Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive attributes.
+Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
 In contrast to the distributional verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}.
@@
 %We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning algorithm using a unified scheme.
-%Our encodings not only allow us to compute for non-Boolean and compound sensitive attributes but also to scale significantly better than existing formal verifiers. 
+%Our encodings not only allow us to compute for non-Boolean and compound sensitive features but also to scale significantly better than existing formal verifiers. 
 %We perform experimental analysis for multiple fairness metrics, datasets, and algorithms to instantiate the efficiency and effectiveness of our approach.
@@
 	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing algorithms with respect to the data  generating distribution}.
-	\item Unlike previously proposed formal distributional verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive attributes.%, and also \red{separation metrics}.
+	\item Unlike previously proposed formal distributional verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
 	\item Our experiments validate that our method is more accurate and scalable than the distributional verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
@@
 AIF360~\cite{aif360-oct-2018} provides a unified framework to implement multiple algorithms and to measure their fairness depending on such test datasets.
-Though this method of verification works for a specified datasets, such verifiers do not explain how much a fairness measure depends on which sensitive attribute and is not robust to the selection of test dataset and its size.
+Though this method of verification works for a specified datasets, such verifiers do not explain how much a fairness measure depends on which sensitive feature and is not robust to the selection of test dataset and its size.
 \fi

### 2022-01-31 chapters/fairness/justicia/preliminaries.tex
working on enumeration approach

@@
 We use $ (\mathbf{x}, \mathbf{a}) $ to denote the feature-values of  $ (\nonsensitive, \sensitive) $.
-For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{m_2}] $ is called a \textit{compound sensitive group}.
+For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{\numsensitive}] $ is called a \textit{compound sensitive group}.
 For example, consider $ \sensitive = $ \{race, sex\} where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}.
@@
 In the following, we describe two well-known fairness definitions verified by {\justicia}: group fairness and causal fairness.
-\red{In each definition, the fairness metric relies on computing the conditional Positive Predictive Value (PPV) of the classifier, denoted by $ \Pr[\hat{Y} = 1 | \mathsf{C}] $, where different fairness metrics are constituted based on the conditional $ \mathsf{C} $.}
+To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ provided the distribution of features $ \mathcal{D} $.
+More specifically,  {\justicia} verifies fairness metrics where $ f(\alg, \mathcal{D}) $ depends on the conditional Positive Predictive Value (PPV) of the classifier, denoted by $ \Pr[\hat{Y} = 1 | \mathsf{C}] $.
+Henceforth, each fairness metric is identified by the particular choice of the conditional $ \mathsf{C} $, as discussed next.
 
@@
 The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups.
-Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier $\alg$, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in \sensitive$.
-Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.
+Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in \sensitive$.
+Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}, are proposed in the literature~\cite{dwork2012fairness,feldman2015certifying}.
 
-\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$.
-Formally, a classifier satisfies $(1 - \epsilon)$-disparate impact if, for $\epsilon \in [0,1] $,
+\textit{Disparate impact} (DI) measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$~\cite{feldman2015certifying}.
+Formally, the disparate impact of a classifier is
 \[
-\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}].
+f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]},
 \]
-Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the maximum difference of PPVs among sensitive groups, and prescribe this to be near zero.
-Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $,
-\[
-\max_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}] \le \epsilon.
-\]
-For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier.
+where the subscript of $ f $ denotes the corresponding fairness metric.
+
+Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the maximum difference of PPVs among sensitive groups, and prescribe this quantity to be near zero.
+Thus, the statistical parity of  a classifier is
+\[ f_{\mathsf{SP}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}]. \]
 
@@
 In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the maximum difference of false positive rates \red{(FPR)} and true positive rates \red{(TPR)} conditioned on sensitive groups.
-Lower value of equalized odds indicates better fairness.
-A classifier satisfies $\epsilon$-equalized odds when the following conditions hold.
+Formally,  for $ Y \in \{0,1\} $,  the equalized odds of a classifier is defined as
 \begin{align*}
-	&\max_{\mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - 	\min_{\mathbf{a}} \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 0] \le \epsilon\\
-	&\max_{\mathbf{a}}\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - 	\min_{\mathbf{a}}  \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 1] \le \epsilon
+f_{\mathsf{EO}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \\
+&\max_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1]). 
 \end{align*}
@@
 This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
-For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as
-\[
-\max_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] -  \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ] \le \epsilon.
+To this end, the path-specific causal fairness of a classifier is \[
+ f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
 \]
@@
 \end{example}
+
+\paragraph{Fairness Certification.} We certify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, which quantifies the desired level of fairness.
+In particular, a classifier is $ \epsilon $-fair with respect to statistical parity, equalized odds, or path-specific causal fairness if and only if $ f(\alg, \mathcal{D}) \le \epsilon $.
+In contrast, a classifier satisfies $(1 - \epsilon)$-disparate impact  if and only if $ f(\alg, \mathcal{D}) \ge 1 - \epsilon $.
+In all above fairness metrics, a lower value of $ \epsilon $ refers to higher fairness of the classifier.
 
@@
 It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
-
-%In these extreme situations, the reduced problem is NP-hard and PP-hard respectively.
-%We can show solving ER- and RE-SSAT are equivalent to solving an E-MAJSAT/e-threshold SAT problem. Thus it is $NP^{PP}$ hard. 
-%Equivalent formulation of this is computing MAP in a belief network.
-
-\begin{comment}
-	The Stochastic Boolean Satisfiability problem (SSAT) allows existential ($ \exists $) and randomized ($ \R $) quantifiers over the Boolean variables and evaluates the probability of satisfaction of the CNF formula.
-Based on the order of quantified variables, an SSAT instance can be either a random-exists (\textbf{RE}) SSAT  formula or an exist-random (\textbf{ER}) SSAT formula.
-
-	Let $ \phi(E,R) $ be a CNF formula over Boolean variables $ E \cup R $ where each $ e_i \in E $ is an exist quantified variable and each $ r_j \in R $ is a randomized quantified variable.
-For each $ r_j $, $ p_j = \Pr(\sigma(r_j) = 1) $ denotes the probability of $ r_j $ assigning to true.
-Now an \textbf{RE}-SSAT formula is defined as
-
-	\[
-	\R^{p_1} r_1, \dots, \R^{p_m} r_m, \exists e_1, \dots, \exists e_n \; \phi (E,R)
-	\]
-
-	where $ |R| = m,  |E| = n  $.
-Intuitively,  for the random assignment of $ r_1, \dots, r_m $ there exists assignments of $ e_1, \dots, e_n $, for which, we ask the \red{maximum}  probability of satisfaction of $ \phi $.
-Similarly, an \textbf{ER}-SSAT formula is defined as
-	\[
-	\exists e_1, \dots, \exists e_n, \R^{p_1} r_1, \dots, \R^{p_m} r_m, \; \phi (E,R). 
-	\]
-
-	is to evaluate a Boolean formula containing both existential and randomized quantifiers~\cite{littman2001stochastic}.  For example:
-
-	\[
-	\exists x_1, \R^{p_2} y_2, \exists x_3, \dots, \R^{p_n} y_n (E[\phi (\mathbf{x})]\ge \theta)
-	\]
-
-	In words, this formula asks whether there is a value for $ x_1 $ such that for random values of $ y_2 $ (choose $ y_2 = 1 $ with probability $ p_2 $), there exists a value of $ x_2 $, such that$ \dots $ the expected value of the Boolean formula $ \phi(\mathbf{x}) $ meets or exceeds a threshold $ 0 \le \theta \le 1 $.
-Here variables $ x_i, y_i \in \mathbf{x} $ are Boolean.
-
-	While the above definition of SSAT has alternating quantifiers, it is not a strict restriction.
-One could first place random quantifiers followed by exists quantifiers or vice versa.
-Moreover, the threshold $ \theta $ can be omitted from the definition where the question is to find out the probability of $ \phi(\mathbf{x}) =1 $ given the quantifiers over Boolean variables $ x_i $ and $ y_i $.
-
-\end{comment}

### 2022-04-07 chapters/fairness/justicia/framework.tex
minor update

@@
 \label{fairness_justicia_sec:enumeration_ssat}
-In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional PPV of the classifier, $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ based on an appropriately designed SSAT formula.
+In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional PPV of the classifier, $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ based on an appropriately designed SSAT formula.
 For simplicity, we  initially make assumptions on the classifier $ \alg $ and discuss practical relaxations later in this section.
@@
 Additionally, all features $ \nonsensitive \cup \sensitive $ are assumed to be Boolean variables.
-Finally, we consider independence probability assumption of non-sensitive features on sensitive features such that $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of the non-sensitive feature $ X_i $.
+Finally, we consider independence probability assumption of non-sensitive features on sensitive features, where $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of the non-sensitive feature $ X_i $.
 
@@
 
-Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the conditional probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$.
-In the prefix of $ \Phi_{\mathbf{a}} $,  all non-sensitive Boolean features in $X$ are assigned randomized quantification and they are followed by the sensitive Boolean features in $ A $ with existential quantification.
-The CNF formula $ \phi $ in $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\hat{Y} = 1 | A = \mathbf{a}] $.
-In order to encode the conditional $ A = \mathbf{a} $, we take the conjunction of the Boolean variables in $ A $ that symbolically specifies the compound sensitive group $ \mathbf{a} $.
-For example, we represent two sensitive features: race $ \in $ \{White, Colour\} and sex $ \in $ \{male, female\} by the Boolean variables $ R $ and $ S $  respectively.
-Hence, the compound groups $\{\textrm{White}, \textrm{male}\}$ and $\{\textrm{Colour}, \textrm{female}\}$ are represented by $ R \wedge S $ and $ \neg R \wedge \neg S $, respectively.
-Thus, the RE-SSAT formula for computing the probability  $ \Pr[\hat{Y} = 1 | A = \mathbf{a}] $ is
+Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the conditional probability $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$.
+In the prefix of $ \Phi_{\mathbf{a}} $,  all non-sensitive features $ \nonsensitive $ are assigned randomized quantifiers and they are followed by sensitive features $ \sensitive $ with existential quantifiers.
+In addition, the CNF formula $ \phi $ in the SSAT formula $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $.
+In order to specify the sensitive group $ \sensitive = \mathbf{a} $, we take the conjunction of the Boolean variables in $ \sensitive $ that symbolically specifies the compound sensitive group $ \sensitive = \mathbf{a} $.
+For example, let us consider two sensitive features: race $ \in $ \{White, Colour\} and sex $ \in $ \{male, female\} by Boolean variables $ R $ and $ S $,  respectively.
+Hence, the compound groups $ [\textrm{White}, \textrm{male}] $ and $[\textrm{Colour}, \textrm{female}]$ are represented by $ R \wedge S $ and $ \neg R \wedge \neg S $, respectively.
+Thus, the RE-SSAT formula for computing the probability  $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ is
 \begin{equation}	\label{fairness_justicia_eq:re}
 \begin{split}
-	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m}_{\text{non-sensitive features}},  &\underbrace{\exists \sensitive_1,\dots, \exists \sensitive_n}_{\text{sensitive features}},
-\phi_{\hat{Y}} \wedge (A=\mathbf{a}).\notag
+	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}}_{\text{non-sensitive features}},  &\underbrace{\exists A_1,\dots, \exists A_{\numsensitive}}_{\text{sensitive features}},
+\phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}).\notag
 \end{split}
 \end{equation}
-In $ \Phi_{\mathbf{a}} $, the existentially quantified variables $ \sensitive_1, \dots, \sensitive_n $ are assigned values  according  to the constraint $ A=\mathbf{a} $.
-\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Therefore, by solving the SSAT formula $ \Phi_{\mathbf{a}} $,  the SSAT solver finds the probability $ \Pr[\Phi_{\mathbf{a}}] $ for the sensitive group $ A=\mathbf{a} $ given the random values of $ \nonsensitive_1, \dots, \nonsensitive_m $, which is the PPV of the sensitive group $\mathbf{a} $ for the distribution $ \mathcal{D} $ and algorithm $\alg$.
+In the RE-SSAT formula $ \Phi_{\mathbf{a}} $, existentially quantified variables  $ \{A_1, \dots, A_{\numsensitive}\} $ are assigned values  according  to the constraint $ \sensitive =\mathbf{a} $.
+\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Next,  an SSAT solver computes the probability $ \Pr[\Phi_{\mathbf{a}}] $ by considering the random values of $ \{X_1, \dots, X_{\numnonsensitive}\} $ while fixing the assignment of $ \{A_1, \dots, A_{\numsensitive\}} $.
+Therefore, $ \Pr[\Phi_{\mathbf{a}}] $ equals the conditional PPV of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $, for the sensitive group $ \sensitive = \mathbf{a} $.
 
-For simplicity, we have described computing the PPV of each compound sensitive group without considering the correlation between the sensitive and non-sensitive features.
-In reality, correlation exists between the sensitive and non-sensitive features.
-Thus, the non-sensitive features may have different conditional distributions for different sensitive groups.
-We incorporate these conditional distributions in RE-SSAT encoding by evaluating the conditional probability $ p_i = \Pr[\nonsensitive_i =\text{TRUE}| A=\mathbf{a}] $ instead of the independent probability $\Pr[\nonsensitive_i =\text{TRUE}]$ for any $\nonsensitive_i\in \nonsensitive$.
+For simplicity, we have described the computation of conditional PPVs without considering the correlation among sensitive and non-sensitive features.
+In reality, correlation exists among sensitive and non-sensitive features.
+Thus, non-sensitive features may have different conditional distributions for different sensitive groups.
+We incorporate these conditional distributions in this RE-SSAT encoding by providing the conditional probability $ p_i = \Pr[X_i = 1| \sensitive =\mathbf{a}] $ instead of the independent probability $\Pr[X_i = 1]$ for each $X_i\in \nonsensitive$.
 We illustrate this method in Example~\ref{fairness_justicia_example:re_ssat}.
@@
 We assign three Boolean variables $ F,I,J $ for the three nodes in the tree such that the literal $ F,I,J $ denote `fitness $ \ge 0.61 $', `income $ \ge 0.29 $', and `income $ \ge 0.69 $', respectively.
-We consider another Boolean variable $A$  where the literal $ A $ represents the sensitive group `age $ \ge 40 $'.
+We consider another Boolean variable $A$  where the literal $ \sensitive $ represents the sensitive group `age $ \ge 40 $'.
 Thus, the CNF formula  for the decision tree is $ (\neg F \vee I) \wedge (F \vee J) $.
@@
 	From the solution to this SSAT formula, we get $ \Pr[\Phi_A] = 0.43 $.
-Similarly, to calculate the PPV for the group `age $ < 40 $', we replace the unit (single-literal) clause $ A $ with $ \neg A $ in the CNF in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $ where $ \Pr[\Phi_{\neg A}] = 0.43 $.
+Similarly, to calculate the PPV for the group `age $ < 40 $', we replace the unit (single-literal) clause $ \sensitive $ with $ \neg A $ in the CNF in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $ where $ \Pr[\Phi_{\neg A}] = 0.43 $.
 	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of $A$ and $\neg A$, both age groups demonstrate equal PPV as the sensitive feature is not explicitly present in the classifier. 
@@
 \subsubsection{Measuring Fairness Metrics.}
-As we compute the probability $\Pr[\hat{Y} = 1 | A = \mathbf{a}]$ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. 
+As we compute the probability $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. 
 For that, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ that requires solving exponential (with $n$) number of SSAT instances. 
@@
 		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y} = 1) $
-		%\State $ p_{i} = \mathsf{CalculateProb}(\nonsensitive_i), \forall \nonsensitive_i \in X $
+		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
 		\ForAll{$\mathbf{a} \in A$ }
-		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(\nonsensitive_i | \mathbf{a}), \forall \nonsensitive_i \in X $
-		\State $ \phi := \phi_{\hat{Y}} \wedge (A=\mathbf{a}) $
-		\State $  \Phi_\mathbf{a} := \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \exists \sensitive_1,\dots, \exists \sensitive_n,  \phi $
+		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i | \mathbf{a}), \forall X_i \in X $
+		\State $ \phi := \phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}) $
+		\State $  \Phi_\mathbf{a} := \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive},  \phi $
 		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns a probability}
@@
 		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
-		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(\nonsensitive_i), \forall \nonsensitive_i \in X $
-		\State $  \Phi_\mathbf{ER} := \exists \sensitive_1,\dots, \exists \sensitive_n, \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \phi_{\hat{Y}} $
-		\State $  \Phi'_\mathbf{ER} := \exists \sensitive_1,\dots, \exists \sensitive_n, \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m, \neg \phi_{\hat{Y}} $
+		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i), \forall X_i \in X $
+		\State $  \Phi_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
+		\State $  \Phi'_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
 		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
@@
 Moreover, to measure equalized odds, we compute two SSAT instances for each compound group with modified values of $ p_i $. 
-Specifically, to compute TPR, we use the conditional probability $ p_i = \Pr[\nonsensitive_i|Y=1] $ on samples with class label $ Y = 1 $ and take the difference between the maximum and the minimum probabilities of all compound groups.
-In addition, to compute FPR, we use the conditional probability $ p_i = \Pr[\nonsensitive_i|Y=0] $ on samples with $ Y = 0 $ and take the difference similarly.
+Specifically, to compute TPR, we use the conditional probability $ p_i = \Pr[X_i|Y=1] $ on samples with class label $ Y = 1 $ and take the difference between the maximum and the minimum probabilities of all compound groups.
+In addition, to compute FPR, we use the conditional probability $ p_i = \Pr[X_i|Y=0] $ on samples with $ Y = 0 $ and take the difference similarly.
 Thus, {\justiciaenum} allows us to compute different fairness metrics using a unified algorithmic framework.
@@
 \begin{equation}
-\Phi_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
- \R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
+\Phi_{\mathsf{ER}} := \exists A_1,\dots, \exists A_{\numsensitive},
+ \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}.
  \label{fairness_justicia_eq:er}
@@
 The CNF formula $\phi_{\hat{Y}}$ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound sensitive group.
-Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_n = a^{\max}_n $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum.
-Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_n \}$ achieving the highest PPV. 
-%Since an assignment to $\{\sensitive_1, \dots, \sensitive_n\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.
+Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_{\numsensitive} = a^{\max}_{\numsensitive} $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum.
+Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_{\numsensitive} \}$ achieving the highest PPV. 
+%Since an assignment to $\{A_1, \dots, A_{\numsensitive}\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.
 
 \subsubsection{Learning the Least Favored Group.}
-In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ \nonsensitive_1, \dots, \nonsensitive_m $.
+In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ X_1, \dots, X_{\numnonsensitive} $.
 In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the sensitive variables and randomized quantification over the non-sensitive variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
 \begin{equation}
-\Phi_{\mathsf{UR}} := \forall \sensitive_1,\dots, \forall \sensitive_n,
-\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \phi_{\hat{Y}}.
+\Phi_{\mathsf{UR}} := \forall A_1,\dots, \forall A_{\numsensitive},
+\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}.
 \label{fairness_justicia_eq:ar}
@@
 \begin{equation}
-\Phi'_{\mathsf{ER}} := \exists \sensitive_1,\dots, \exists \sensitive_n,
-\R^{p_{1}}\nonsensitive_1, \dots, \R^{p_{m}}\nonsensitive_m,   \; \neg \phi_{\hat{Y}}.
+\Phi'_{\mathsf{ER}} := \exists A_1,\dots, \exists A_{\numsensitive},
+\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \neg \phi_{\hat{Y}}.
 \label{fairness_justicia_eq:er_complement}
@@
 %\begin{equation}
-%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_n}_{\text{sensitive features}},
-%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m}_{\text{non-sensitive features}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
+%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_{\numsensitive}}_{\text{sensitive features}},
+%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}}_{\text{non-sensitive features}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
 %\label{fairness_justicia_eq:er_complement}
@@
 \end{lemma}
-As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the sensitive features $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_n\}$ that maximizes $\Phi'_{\mathsf{ER}}$. 
+As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the sensitive features $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_{\numsensitive}\}$ that maximizes $\Phi'_{\mathsf{ER}}$.
 If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$.
@@
 		\State $ p_{i} = \mathsf{CalculateProb}(x_i), \forall x_i \in X $
-		\State $  \Phi_\mathbf{ER} = \exists a_1,\dots, \exists a_n, \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m. \; \phi_{\hat{Y}} $
-		\State $  \Phi'_\mathbf{ER} = \exists a_1,\dots, \exists a_n, \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m. \; \neg \phi_{\hat{Y}} $
+		\State $  \Phi_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \phi_{\hat{Y}} $
+		\State $  \Phi'_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \neg \phi_{\hat{Y}} $
 		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
@@
 %		\State $ p_{i} = \mathsf{\blue{calculateCondProb}}(x_i, \mathbf{a}), \forall x_i \in X $
-%		\State $  \Phi_\mathbf{a} = \R^{p_{1}}x_1, \dots, \R^{p_{m}}x_m, \exists \sensitive_1,\dots, \exists \sensitive_n. \; \phi_{\hat{Y}} \wedge \mathsf{CNF}(A=\mathbf{a}) $
+%		\State $  \Phi_\mathbf{a} = \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive}.
+\; \phi_{\hat{Y}} \wedge \mathsf{CNF}(\sensitive =\mathbf{a}) $
 %		\State $ \Pr[\Phi_\mathbf{a}]  = \ma

### 2022-04-07 chapters/fairness/fvgm/abstract.tex
updated files

@@
-\section{abstract}
+\begin{abstract}
 In recent years, machine learning (ML) algorithms have been deployed in safety-critical and high-stake decision-making, where the \textit{fairness} of algorithms is of paramount importance.
@@
 In contrast to existing verifiers, \fvgm~proposes a \textit{stochastic subset-sum} based approach for verifying linear classifiers.
-Experimentally, we show that  \fvgm~leads to an \textit{accurate} and \textit{scalable} assessment for more diverse families of fairness-enhancing algorithms, fairness attacks, and group/causal fairness metrics than the state-of-the-art.
+Experimentally, we show that  \fvgm~leads to an \textit{accurate} and \textit{scalable} assessment for more diverse families of fairness-enhancing algorithms, fairness attacks, and group/causal fairness metrics than the state-of-the-art fairness verifiers.
 We also demonstrate that {\fvgm} facilitates the computation of fairness influence functions as a stepping stone to detect the source of bias induced by subsets of features.
+\end{abstract}
 

### 2022-04-07 chapters/fairness/fvgm/applications.tex
updated files

@@
-\section{Applications of {\fvgm}}\label{fairness_fvgm_sec:applications}
+\section{Applications of {\fvgm}}\label{fvgm_sec:applications}
 In this section, we apply {\fvgm} for verifying fairness-enhancing algorithms and depreciation attacks.
-We also show that {\fvgm} facilitates computation of fairness influence functions enabling detection of bias due to individual features.
+We also demonstrate that {\fvgm} facilitates computation of fairness influence functions by enabling the detection of bias due to individual features.
 
@@
 	%\vspace{-0.5em}
-	\caption{Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
+	\caption{Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
 \end{figure}
@@
 pre-processing algorithms can be validated by applying {\fvgm} on the unprocessed and the processed data separately and comparing different fairness metrics.
-In Table~\ref{fairness_fvgm_tab:fair_algo_verification}, we report the effect of fairness algorithms w.r.t.
+In Table~\ref{fvgm_tab:fair_algo_verification}, we report the effect of fairness algorithms w.r.t.
 four fairness metrics: disparate impact (DI), path-specific causal fairness (PCF), statistical parity (SP), and equalized odds (EO).
 Note that, fairness is improved if DI increases and the rest of the metrics decrease.
-In most instances for Adult dataset, reweighing (RW)~\cite{kamiran2012data} and optimized pre-processing (OP)~\cite{calmon2017optimized} algorithms are successful in improving fairness.
-Except for the unfairness regarding the sensitive feature `sex', where OP algorithm fails.
+For instance, in most instances for Adult dataset, reweighing (RW)~\cite{kamiran2012data} and optimized pre-processing (OP)~\cite{calmon2017optimized} algorithms are successful in improving fairness.
+The exceptional case is the unfairness regarding the sensitive feature `sex', where OP algorithm fails in improving fairness metric EO.
 Thus, \textit{{\fvgm} verifies the enhancement and decrement in fairness by  fairness-enhancing algorithms.
 }
+
+\begin{figure}
+	\centering
+	\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/disp_fairness_attack}}
+	\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/stat_fairness_attack}}
+
+	\captionof{figure}{Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
+\end{figure}
 
@@
 This algorithm injects a small fraction of poisoned samples into the training data such that the classifier becomes relatively unfair~\cite{solans2020poisoning}.
-We apply this attack to add $\{1, 5, \dots, 160\}$ poisoned samples  and measure the corresponding DI's.
-In Figure~\ref{fairness_fvgm_fig:fairness_attacks},  {\fvgm} verifies that DI of the classifier decreases, i.e.\textit{ the classifier becomes more unfair},\textit{ as the number of poisoned samples increases}.
+We apply this attack to add $\{1, 5, \dots, 160\}$ poisoned samples  and measure the corresponding disparate impact and statistical parity.
+In Figure~\ref{fvgm_fig:fairness_attacks},  {\fvgm} verifies that the disparate impact of the classifier decreases and statistical parity increases, i.e.\textit{ the classifier becomes more unfair},\textit{ as the number of poisoned samples increases}.
 Therefore, {\fvgm} shows the potential of being deployed in safety-critical applications to detect fairness attacks.
-For example, if we set $0.9$ as an acceptable threshold of DI, {\fvgm} can raise an alarm once $160$ poisoned samples are added.
+For example, if we set $0.9$ as an acceptable threshold of disparate impact, {\fvgm} can raise an alarm once $160$ poisoned samples are added.
 
@@
 RW and OP refer to reweighing and optimized-preprocessing algorithms.
-}\label{fairness_fvgm_tab:fair_algo_verification}
+}\label{fvgm_tab:fair_algo_verification}
 	\setlength{\tabcolsep}{.4em}
@@
 Another application of \fvgm{} as a fairness verifier is to quantify the effect of a subset of features on fairness.
-Thus, we define fairness influence function (FIF)  that computes the effect of a subset of features $\mathbf{S}$ on the PPV of a classifier given a specific sensitive group $ \sensitive = \mathbf{a}$,
+Thus, we define fairness influence function (FIF)  that computes the effect of a subset of non-sensitive features $\mathbf{S} \subseteq \nonsensitive$ on the PPV of a classifier given a specific sensitive group $ \sensitive = \mathbf{a}$,
 $
@@
 In practice, we compute FIF of $\mathbf{S}$ by replacing the probability distribution of $\mathbf{S}$ with a uniformly random distribution, referred to as  $ \mathcal{D}_{-\mathbf{S}} $, and reporting differences in the conditional PPVs.
-Further details of FIF are in Appendix.
 
-In Figure~\ref{fairness_fvgm_fig:influence function}, we compute FIF for all features in COMPAS dataset, separately for two sensitive groups: Female (`sex' $ = 1 $) and Male.
+In Figure~\ref{fvgm_fig:influence function}, we compute FIF for all features in COMPAS dataset, separately for two sensitive groups: Female (`sex' $ = 1 $) and Male.
 This dataset concerns the likelihood of a person of re-offending crimes within the next two years.
@@
 %	\vspace{-2ex}
-	\caption{\footnotesize FIF computation for COMPAS dataset.
-For Female (left plot) and Male, influence of `age' decreases and the PPV of the classifier increases  by different magnitudes.}\label{fairness_fvgm_fig:influence function}
+	\caption{FIF computation for COMPAS dataset.
+For Female (left plot) and Male, influence of `age' decreases and the PPV of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}
 %	\vspace{-2ex}

### 2022-04-07 chapters/fairness/fvgm/background.tex
updated files

@@
  		\itemsep0em 
- 		\item \textit{Disparate impact} (DI): A classifier  satisfies $(1 - \epsilon)$-disparate impact if for $\epsilon \in [0,1] $,
+ 		\item \textit{Disparate Impact} (DI): A classifier  satisfies $(1 - \epsilon)$-disparate impact if for $\epsilon \in [0,1] $,
  		$
- 		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}'} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'].
+ 		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}].
  		$
@@
  		$
- 		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}']| \le \epsilon.
+ 		\max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}] - \min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}] \le \epsilon.
  		$
  		\item \textit{Equalized Odds} (EO): 	A classifier satisfies $\epsilon$-equalized odds if for $\epsilon \in [0,1] $,
- 		$ \max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 0]| \le \epsilon, $ and $
- 		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 1]| \le \epsilon.
+ 		$ \max_{\mathbf{a}}\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \min_{\mathbf{a}}\Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 0] \le \epsilon, $ and $
+ 		\max_{\mathbf{a}}\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \min_{\mathbf{a}}\Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}, Y = 1] \le \epsilon.
  		$
@@
  		$
- 		\max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}', \mediator ]| \le \epsilon
+ 		\max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ] \le \epsilon
  		$.
@@
 Following the observation of~\cite{ghosh2020justicia},  computing all of the aforementioned fairness metrics is equivalent to computing the maximum and minimum of the \textit{Positive Predictive Value} (PPV) of the classifier, denoted as $\Pr[\hat{Y}=1|\mathbf{A} =\mathbf{a}]$, for all compound sensitive groups $\mathbf{a}$ from $ \sensitive $.
-Thus, in Section~\ref{fairness_fvgm_sec:fvgm}, we focus on computing the maximum and minimum PPVs and then extend it to assess corresponding fairness metrics.
+Thus, in Section~\ref{fvgm_sec:fvgm}, we focus on computing the maximum and minimum PPVs and then extend it to assess corresponding fairness metrics.
 We call the group for which PPV is maximum (minimum) as the \textit{most (least) favored group}.
@@
 	Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ w_i \in \mathbb{Z} $ be the weight of $ B_i $.
-Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} w_i B_i = \tau $, for a constant $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$.
+Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} w_i B_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$.
 Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
@@
 More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} w_iB_i \ge \tau] $.
-Details of {\stochastic} are in Section~\ref{fairness_fvgm_sec:stochastic_sum_set_sum}.
+Details of {\stochastic} are in Section~\ref{fvgm_sec:stochastic_sum_set_sum}.
 
@@
 	In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.
-In this paper, we leverage an access to a Bayesian network on $(\nonsensitive, \sensitive)$ that represents the joint distribution on them.
-	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a Directed Acyclic Graph (DAG), and $\theta$ is a set of parameters that encodes the conditional probabilities induced by the joint distribution under investigation.
+In this paper, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
+	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
 Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable.
@@
 
-	\begin{align}
-	\Pr[V_1=v_1, \dots, V_{|\mathbf{V}|}=v_{|\mathbf{V}|}] = \prod_{i=1}^{|\mathbf{V}|} \Pr[V_i = v_i | \parent(V_i) = \mathbf{u}; \theta].
-	\label{fairness_fvgm_eq:BN}
-	\end{align}
+	\begin{equation}
+		\begin{split}
+				\Pr[V_1=v_1, \dots, &V_{|\mathbf{V}|}=v_{|\mathbf{V}|}] = \\ 
+			&\prod_{i=1}^{|\mathbf{V}|} \Pr[V_i = v_i | \parent(V_i) = \mathbf{u}; \theta].
+		\end{split}
+	\label{fvgm_eq:BN}
+	\end{equation}
 

### 2022-04-07 chapters/fairness/fvgm/conclusion.tex
updated files

@@
 Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
+%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.
+
+\iffalse
+\section*{Acknowledgments}
+This work was supported in part by National Research Foundation, Singapore under its NRF Fellowship Programme [NRF-NRFFAI1-2019-0004 ] and AI Singapore Programme [AISG-RP-2018-005],  and NUS ODPRT Grant [R-252-000-685-13].
+The computational work for this paper was performed on resources of Max Planck Institute for Software Systems, Germany and the National Supercomputing Centre, Singapore (\url{https://www.nscc.sg}).
+\fi

### 2022-04-07 chapters/fairness/fvgm/dp_illustration.tex
updated files

@@
 		}
-\label{fairness_fvgm_fig:example_dp}}
-\hspace*{-0.5cm}
-\subfloat[Probabilities computed with Bayesian network.]
+\label{fvgm_fig:example_dp}}
+%\hspace*{-0.5cm}
+\subfloat[Probabilities computed with a Bayesian network.]
 {  
@@
 	}
-	\label{fairness_fvgm_fig:example_BN}
+	\label{fvgm_fig:example_BN}
 }
 
-\caption{\footnotesize Search tree representation of {\stochastic} for computing the maximum PPV of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
+\caption{Search tree representation of {\stochastic} for computing the maximum PPV of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
 Each node is labeled by $ (i,\tau') $, where $ i $ is the index of $ \mathbf{B} $ and $ \tau' $ is the residual threshold.
@@
 Additionally, a green node denotes a collision, in which case a previously computed $ \mathsf{dp} $ solution is returned.
-Leaf nodes (gray) are computed based on terminating conditions in Eq.~\ref{fairness_fvgm_eq:dp_terminus}.
-In Figure~\ref{fairness_fvgm_fig:example_BN},  nodes with double circles, such as $ \{(1,2), (2,1), (2,2)\} $,  are enumerated exponentially to compute conditional probabilities from the Bayesian network.}
-\label{fairness_fvgm_fig:example_tree_exploration}
+Leaf nodes (gray) are computed based on terminating conditions in Eq.~\ref{fvgm_eq:dp_terminus}.
+In Figure~\ref{fvgm_fig:example_BN},  nodes with double circles, such as $ \{(1,2), (2,1), (2,2)\} $,  are enumerated exponentially to compute conditional probabilities from the Bayesian network.}
+\label{fvgm_fig:example_tree_exploration}
 \end{figure*}

### 2022-04-07 chapters/fairness/fvgm/evaluations.tex
updated files

@@
 \section{Empirical Performance Analysis}
-\label{fairness_fvgm_sec:experiments}
-In this section, we empirically evaluate the performance of {\fvgm}.
-We first present the experimental setup and the objective of our experiments, followed by experimental results.\footnote{Due to page limit, we present additional experimental results, proof of lemmas, and miscellaneous details in Appendix.}
+\label{fvgm_sec:experiments}
+In this section, we empirically evaluate the performance of {\fvgm}. We first present the experimental setup and the objective of our experiments, followed by experimental results.
+%\footnote{Due to page limit, we present additional experimental results, proof of lemmas, and miscellaneous details in Appendix.}
 
@@
 	\begin{center}		
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_LR_time_}}
+		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_LR_time_}}\\
 		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_SVM_time_}}		
@@
 	%	\vspace{-2ex}
-	\caption{\footnotesize A cactus plot to present the scalability of different fairness verifiers.
-The number of solved benchmarks are on the $ X $-axis and the time taken on the $ Y $-axis; a point $ (x,y) $ implies that a verifier takes less than or equal to $ y $ seconds to compute fairness metrics of $ x $ many benchmarks.
+	\caption{A cactus plot to present the scalability of different fairness verifiers.
+The number of solved benchmarks are on the $ X $-axis and the required time is on the $ Y $-axis; a point $ (x,y) $ implies that a verifier takes less than or equal to $ y $ seconds to compute fairness metrics of $ x $ many benchmarks.
 We consider $ 100 $ benchmarks generated from $ 5 $ real-world datasets using $ 5 $-fold cross-validation.
-In each fold, we consider $\{25, 50, 75, 100\} $ percent of non-sensitive features.}\label{fairness_fvgm_fig:scalability_exp}
+In each fold, we consider $\{25, 50, 75, 100\} $ percent of non-sensitive features.}\label{fvgm_fig:scalability_exp}
 	%	\vspace{-2ex}
@@
 We deploy the Scikit-learn library for learning linear classifiers such as Logistic Regression (LR) and Support Vector Machine (SVM) with linear kernels.
-We perform five-fold cross-validation.
+We perform five-fold cross-validation on a dataset.
 While the classifier is trained on continuous features, we discretize them to Boolean features to be invoked by {\fvgm}.
-While discretizing, we apply a gird-search to estimate the best bin-size within a maximum bin of $ 10 $.
+During discretization, we apply a gird-search to estimate the best bin-size within a maximum bin of $ 10 $.
 To convert the coefficients of features into integers, we employ another grid-search to choose the best multiplier within $ \{1,2, \dots, 100\} $.
@@
 
-\textbf{Results.} In Figure~\ref{fairness_fvgm_fig:scalability_exp}, we present the scalability results of different verifiers.
+\textbf{Results.} In Figure~\ref{fvgm_fig:scalability_exp}, we present the scalability results of different verifiers.
 First, we observe that FairSquare often times out ($ =900 $ seconds) and can solve $ \le 5 $ benchmarks.
@@
 	\begin{center}	
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_LR_01}}
+		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_LR_01}}\\
 		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_SVM_01}}		
@@
 	%	\vspace{-2ex}
-	\caption{\footnotesize Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features.
-{\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of DI for LR and SVM classifiers.}\label{fairness_fvgm_fig:sanity_exp}
+	\caption{Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features.
+{\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of DI for LR and SVM classifiers.}\label{fvgm_fig:sanity_exp}
 	%	\vspace{-2ex}
@@
  We  assess the accuracy of the competing verifiers in estimating fairness metrics, specifically DI with LR and SVM classifiers.
-In Figure~\ref{fairness_fvgm_fig:sanity_exp}, {\fvgm} computes DI closest to the \textit{Exact} value for different number of features and both type of classifiers.
+In Figure~\ref{fvgm_fig:sanity_exp}, {\fvgm} computes DI closest to the \textit{Exact} value for different number of features and both type of classifiers.
 In contrast, Justicia, FairSquare, and VeriFair measure DI far from the \textit{Exact} because of ignoring correlations among the features.
-For example, for SVM classifier with  $ n = 5 $ (right plot in Figure~\ref{fairness_fvgm_fig:sanity_exp}), \textit{Exact} DI is $ 0.089 $ (average over 100 random benchmarks).
+For example, for SVM classifier with  $ n = 5 $ (right plot in Figure~\ref{fvgm_fig:sanity_exp}), \textit{Exact} DI is $ 0.089 $ (average over 100 random benchmarks).
 Here, {\fvgm} computes DI as $ 0.094 $, while all other verifiers compute DI as at least $ 0.233 $.
@@
 
-\iffalse
 \begin{table}[!t]
+		\centering
 
-		\centering
-		\small
-		\caption{Verification of fairness algorithms using {\fvgm}.
-Numbers in bold refer to fairness improvement.
-RW and OP refer to reweighing and optimized-preprocessing algorithms respectively.
-}\label{fairness_fvgm_tab:fair_algo_verification}
-		\setlength{\tabcolsep}{.2em}
-
+		\setlength{\tabcolsep}{0.45em}
 		\begin{tabular}{lllrrrrrrrrrrrrr}
@@
 			\toprule
-				Dataset & Sensitive & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta$ EO\\
+			Dataset & $ \sensitive $ & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta $EO\\
 			\midrule
 
-				\multirow{4}{*}{Adult}&\multirow{2}{*}{race}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
+			\multirow{4}{*}{\rotatebox[origin=c]{0}{Adult}}&\multirow{2}{*}{\rotatebox[origin=c]{0}{race}}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
 			&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
 			\cmidrule{2-7}
-				&\multirow{2}{*}{sex}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
+			&\multirow{2}{*}{\rotatebox[origin=c]{0}{sex}}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
 			&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\
@@
 		\end{tabular}
+
+	\caption{Verification of fairness algorithms using {\fvgm}.
+$ \sensitive $ denotes sensitive features.
+RW and OP refer to reweighing and optimized-preprocessing algorithms.
+Numbers in bold refer to fairness improvement.
+}\label{fvgm_tab:fair_algo_verification}
+
 \end{table}
-\fi
 
-\iffalse
-\textbf{Verification: Verifying Fairness Attacks and Fairness-enhancing Algorithms.}
-We now demonstrate the applications of {\fvgm} starting with verifying fairness attacks.
-We instantiate a poisoning-attack algorithm that injects a small fraction of poisoned samples to the training data such that the classifier becomes relatively unfair~\cite{solans2020poisoning}.
-We use this algorithm to add $x, y, z$ number of poisoned samples and to measure the corresponding DI and SP.
-From Figure~\ref{fairness_fvgm_fig:fairness_attacks}, we observe that {\fvgm} verifies that DI of the classifier decreases and SP increases, i.e.\textit{ the classifier becomes more unfair},\textit{ as the number of poisoned samples increases}.
-Therefore, {\fvgm} shows the potential of being deployed in safety-critical applications to detect fairness attacks.
-For example, if we set $DI=0.9$ as an acceptable threshold of DI, {\fvgm} can raise an alarm once $xx$ poisoned samples are added.
-
-We have employed {\fvgm} in verifying fairness-enhancing algorithms that pre-process the dataset to ameliorate bias.
-In Table~\ref{fairness_fvgm_tab:fair_algo_verification}, We report the effect of fairness algorithms w.r.t.
-four fairness metrics: disparate impact (DI), path-specific causal fairness (PCF), statistical parity (SP) and equalized odds (EO).
-Note that, fairness is improved if DI increases and rest of the metrics decrease.
-We observe that in most instances for Adult and COMPAS dataset, reweighing (RW) and optimized pre-processing (OP) algorithms are successful in improving fairness.
-Except the unfairness regarding the sensitive feature `sex', where both of the algorithms fail.
-Thus, {\fvgm} verifies the enhancement and decrement in fairness of a classifier on a dataset as promised.
-%\iffalse
-%More precisely, {\fvgm} guarantees that while these algorithms improve fairness on  a finite dataset, say a training dataset, the improvement also generalizes to the distribution of the dataset.
-%\fi
-
-\textbf{Tracing Sources of Unfairness: Computing Fairness Influence Function (FIF).}
-In Figure~\ref{fairness_fvgm_fig:influence function}, we have computed the fairness influence function for all features in COMPAS dataset, separately for two sensitive groups: Female (`sex' $ = 1 $) and Male.
-This dataset concerns with the likelihood of a person of re-offending crimes within next two years.
-We first observe that the base PPV values are different for the two groups ($ 0.46 $ vs $ 0.61 $ for Female and Male), thereby showing Male as more probable to re-offend crimes than Female.
-Moreover, FIF of the feature `age' is comparatively higher in magnitude for Male than Female.
-This implies while deciding recidivism the algorithm assumes that Female individuals across different ages re-offend crimes with almost same probability and the probability of re-offending for Male individuals highly depends on age.
-Thus, applying {\fvgm} to compute provides us insights about sources of bias and thus, indicators to improve fairness.
-\fi
-
-\iffalse
-\begin{figure}[t!]
-		\centering
-		\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/disp_fairness_attack}}
-		\subfloat{\includegraphics[scale=0.25]{figures/fairness/fvgm/stat_fairness_attack}}
-		\caption{Verifying fairness poisoning attack using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
-\end{figure}
-\fi
-
-\begin{table}[!t]
-	\begin{minipage}{0.6\columnwidth}
-		\centering
-		\small
-
-		\setlength{\tabcolsep}{.25em}
-
-		\begin{tabular}{lllrrrrrrrrrrrrr}
-
-			\toprule
-			$\mathbf{D}$ & $ \sensitive $ & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta $EO\\
-			\midrule
-
-			\multirow{4}{*}{\rotatebox[origin=c]{90}{Adult}}&\multirow{2}{*}{\rotatebox[origin=c]{90}{race}}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
-			&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
-			\cmidrule{2-7}
-			&\multirow{2}{*}{\rotatebox[origin=c]{90}{sex}}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
-			&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\
-
-			\bottomrule
-		\end{tabular}
-
-		\caption{\footnotesize Verification of fairness algorithms using {\fvgm}.
+%\begin{table}[!t]
+%	\begin{minipage}{0.6\columnwidth}
+%		\centering
+%		\small
+%		
+%		\setlength{\tabcolsep}{.25em}
+%		
+%		\begin{tabular}{lllrrrrrrrrrrrrr}
+%			
+%			\toprule
+%			$\mathbf{D}$ & $ \sensitive $ & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta $EO\\
+%			\midrule
+%			
+%			
+%			\multirow{4}{*}{\rotatebox[origin=c]{90}{Adult}}&\multirow{2}{*}{\rotatebox[origin=c]{90}{race}}&RW&$ \textbf{0.53} $&$ \textbf{-0.06} $&$ \textbf{-0.06} $&$ \textbf{-0.02} $\\
+%			&&OP&$ \textbf{0.57} $&$ \textbf{-0.07} $&$ \textbf{-0.07} $&$ \textbf{-0.02} $\\
+%			\cmidrule{2-7}
+%			&\multirow{2}{*}{\rotatebox[origin=c]{90}{sex}}&RW&$ \textbf{0.96} $&$ \textbf{-0.16} $&$ \textbf{-0.15} $&$ \textbf{-0.08} $\\
+%			&&OP&$ \textbf{0.43} $&$ \textbf{-0.08} $&$ \textbf{-0.08} $&$ 0.03 $\\
+%			
+%			\bottomrule
+%		\end{tabular}
+%		
+%		\caption{\footnotesize Verification of fairness algorithms using {\fvgm}.
 $ \mathbf{D} $ and $ \sensitive $ denote dataset and sensitive features.
@@
 Numbers in bold refer to fairness improvement.
-}\label{fairness_fvgm_tab:fair_algo_verification}
-
-	\end{minipage}\hspace*{1em}
-	\begin{minipage}{0.4\columnwidth}
-		\centering
-		\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/disp_fairness_attack}}
-		\captionof{figure}{\footnotesize Verifying poisoning attack against fairness using {\fvgm}.
-The red line denotes the safety margin of the ML model against the attack.}\label{fairness_fvgm_fig:fairness_attacks}
-	\end{minipage}\hspace*{1em}
-%	\vspace{-2ex}
-
-\end{table}
+}\label{fvgm_tab:fair_algo_verification}
+%		
+%	\end{minipage}\hspace*{1em}
+%	\begin{minipage}{0.4\columnwidth}
+%		\centering
+%		\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/disp_fairness_attack}}
+%		\captionof{figure}{\footnotesize Verifying poisoning attack against fairness using {\fvgm}.
+The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
+%	\end{minipage}\hspace*{1em}
+%%	\vspace{-2ex}
+%	
+%\end{table}

### 2022-04-07 chapters/fairness/fvgm/framework.tex
updated files

@@
-\section{{\fvgm}: Fairness Verification with Graphical Models}\label{fairness_fvgm_sec:fvgm}
+\section{{\fvgm}: Fairness Verification with Graphical Models}\label{fvgm_sec:fvgm}
 
@@
 We then extend {\stochastic} to consider correlated features as input.
-We conclude by discussing fairness verification based on the solution of {\stochastic} in practice.
+Finally, we conclude by discussing fairness verification based on the solution of {\stochastic}.
 
@@
 
-Here, $ w $ denotes the weight (or coefficients) of a feature, $ \tau $ denotes the bias of the classifier, and $\mathds{1}$ is an indicator function.
+Here, $ w $ denotes the weight (or coefficients) of a feature, $ \tau $ denotes the bias or the offset parameter of the classifier, and $\mathds{1}$ is an indicator function.
 Hence, the prediction $ \hat{Y} =1 $ if and only if the inner inequality holds.
@@
 To perform this reduction, we assume  weights $ w $ and bias $ \tau $ as integers, and features $\nonsensitive \cup \sensitive $ as Boolean.
-In Sec.~\ref{fairness_fvgm_sec:practical}, we relax these assumptions and extend to the practical settings.
+In Sec.~\ref{fvgm_sec:practical}, we relax these assumptions and extend to the practical settings.
 
-\subsection{$ \stochastic $: Stochastic Subset Sum Problem}\label{fairness_fvgm_sec:stochastic_sum_set_sum}
+\subsection{$ \stochastic $: Stochastic Subset Sum Problem}\label{fvgm_sec:stochastic_sum_set_sum}
 Now, we formally describe the specification and semantics of {\stochastic}.
@@
 To specify the category of variables, we consider a \textit{quantifier} $ q_i \in \{\R^{p_i}, \exists, \forall\} $ for each $ B_i $.
-Elaborately, $ \R^{p} $ is a \textit{random quantifier} corresponding to a chance variable $ B $, where  $ p\triangleq \Pr[B = 1]$.
+Elaborately, $ \R^{p} $ is a \textit{random quantifier} corresponding to a chance variable $ B \in \mathbf{B} $, where  $ p\triangleq \Pr[B = 1]$.
 In contrast, $ \exists $ is an \textit{existential quantifier} corresponding to a choice variable $ B $ such that a Boolean assignment of $ B $  \textit{maximizes}  $ S(\mathbf{B}, \tau) $.
@@
 
-\begin{align}\label{fairness_fvgm_eq:semantics_recurse}
- \hspace*{-1.2em} S(\mathbf{B},\tau) =
+\begin{align}\label{fvgm_eq:semantics_recurse}
+  S(\mathbf{B},\tau) =
  \begin{cases}
@@
 \begin{remark}
-	\label{fairness_fvgm_lm:property_s3p}
+	\label{fvgm_lm:property_s3p}
 	$ S(\mathbf{B},\tau) $ does not depend on the order of $ \mathbf{B} $.
@@
 \begin{lemma}
-	\label{fairness_fvgm_lm:hardness}
+	\label{fvgm_lm:hardness}
 	The decision version of the fairness verification problem for linear classifiers is in $\mathrm{NP^{PP}}$.
@@
  \subsection{A Dynamic Programming Solution for {\stochastic}}
- \label{fairness_fvgm_sec:dp_formulation}
+ \label{fvgm_sec:dp_formulation}
  %\dbcomment{Use $ \mathsf{dp}(i, \tau) $ given $ B $}
@@
 We propose a dynamic programming approach~\cite{pisinger1999linear,woeginger1992equal} to solve {\stochastic} as the problem has overlapping sub-problem properties. 
-For example, $S(\mathbf{B}, \tau)$ can be solved by solving $S(\mathbf{B}[2:n], \tau')$, where the updated threshold $\tau'$, called the \textit{residual threshold}, depends on the original threshold $\tau$ and the assignment of $B_1$ as shown in Eq.~\eqref{fairness_fvgm_eq:semantics_recurse}.
+For example, $S(\mathbf{B}, \tau)$ can be solved by solving $S(\mathbf{B}[2:n], \tau')$, where the updated threshold $\tau'$, called the \textit{residual threshold}, depends on the original threshold $\tau$ and the assignment of $B_1$ as shown in Eq.~\eqref{fvgm_eq:semantics_recurse}.
 %\red{For example, the residual threshold $ \tau $ can be equal for a fixed subset of $ \mathbf{B} $, say $ \mathbf{B}[i:n] \triangleq \{B_j\}_{j=i}^{n} $, derived from different assignments of $ B_1, \dots, B_{i-1} $.}
-Building on this observation, we propose a recursion and terminating condition leading to our dynamic programming algorithm.
+Building on this observation, we propose the recursion and terminating condition leading to our dynamic programming algorithm.
 
 \textit{Recursion.} We consider a function $ \mathsf{dp}(i, \tau) $ that solves the sub-problem $ S(\mathbf{B}[i:n],\tau)  $, for $ i \in \{1,\ldots,n\}  $.
-The semantics of  $ S(\mathbf{B},\tau) $ in Eq.~\eqref{fairness_fvgm_eq:semantics_recurse} induces the recursive definition of $ \mathsf{dp}(i,\tau) $ as:
+The semantics of  $ S(\mathbf{B},\tau) $ in Eq.~\eqref{fvgm_eq:semantics_recurse} induces the recursive definition of $ \mathsf{dp}(i,\tau) $ as:
 
-\begin{align}\label{fairness_fvgm_eq:dp_recurse}
+\begin{align}\label{fvgm_eq:dp_recurse}
  \mathsf{dp}(i,\tau)=
@@
 
-Eq.~\eqref{fairness_fvgm_eq:dp_recurse} shows that $ S(\mathbf{B},\tau) $ can be solved by instantiating $ \mathbf{dp}(1, \tau) $, which includes all the variables in $ \mathbf{B} $.
+Eq.~\eqref{fvgm_eq:dp_recurse} shows that $ S(\mathbf{B},\tau) $ can be solved by instantiating $ \mathbf{dp}(1, \tau) $, which includes all the variables in $ \mathbf{B} $.
 
@@
 Thus, for any $ i $, if the {residual} threshold $ \tau \le w_{neg}$, there is always a subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.
-Conversely, when $ \tau > w_{pos}$, there is no subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.	We leverage this bound and tighten the terminating conditions of $ \mathsf{dp}(i, \tau) $ in Eq.~\eqref{fairness_fvgm_eq:dp_terminus}.
+Conversely, when $ \tau > w_{pos}$, there is no subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.	We leverage this bound and tighten the terminating conditions of $ \mathsf{dp}(i, \tau) $ in Eq.~\eqref{fvgm_eq:dp_terminus}.
 
-\begin{align}\label{fairness_fvgm_eq:dp_terminus}
+\begin{align}\label{fvgm_eq:dp_terminus}
  \mathsf{dp}(i, \tau) =
@@
 
-Eq.~\eqref{fairness_fvgm_eq:dp_recurse} and~\eqref{fairness_fvgm_eq:dp_terminus} together define our dynamic programming algorithm.
+Eq.~\eqref{fvgm_eq:dp_recurse} and~\eqref{fvgm_eq:dp_terminus} together define our dynamic programming algorithm.
 While deploying the algorithm, we store $ \mathsf{dp}(i, \tau) $ in memory to avoid repetitive computations.
-This allows us to achieve a pseudo-polynomial algorithm (Lemma~\ref{fairness_fvgm_lemma:complexity_sss}) instead of a na\"ive exponential algorithm enumerating all possible assignments.
+This allows us to achieve a pseudo-polynomial algorithm (Lemma~\ref{fvgm_lemma:complexity_sss}) instead of a na\"ive exponential algorithm enumerating all possible assignments.
+In particular, the time complexity is pseudo-polynomial for chance (random) variables and linear for choice (existential and universal) variables.
 
-\begin{lemma}\label{fairness_fvgm_lemma:complexity_sss}
+\begin{lemma}\label{fvgm_lemma:complexity_sss}
  	Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $.
@@
 We propose two improvements for a faster computation of the dynamic programming solution.
-Firstly, we observe that in Eq.~\eqref{fairness_fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights.
+Firstly, we observe that in Eq.~\eqref{fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights.
 Hence, we reorder $ \mathbf{B} $ such that existential/universal variables appear earlier in $ \mathbf{B} $ than random variables.
 This allows us to avoid unnecessary repeated exploration of existential/universal variables in $ \mathsf{dp} $.
-Moreover, according to the remark in Section~\ref{fairness_fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}.
+Moreover, according to the remark in Section~\ref{fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}.
 Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables.
@@
 Otherwise,  we sort in ascending order.
-We illustrate our dynamic programming approach in Example~\ref{fairness_fvgm_example:subset-sum}.
+We illustrate our dynamic programming approach in Example~\ref{fvgm_example:subset-sum}.
 
-\begin{example}\label{fairness_fvgm_example:subset-sum}
+\begin{example}\label{fvgm_example:subset-sum}
 We consider a linear classifier $ P + Q + R - S \ge 2$.
@@
 We now require to compute $ \Pr[Q + R - S \ge 1] $, which by dynamic programming, is computed as $ 0.55 $.
-The solution is visualized as a search tree in Figure~\ref{fairness_fvgm_fig:example_dp}, where we observe that storing solution of sub-problems in the memory avoids repetitive computation, such as exploring the node $ (4,0) $.
-Similarly, the minimum PPV  of the classifier is $ 0.14 $ (not shown in Figure~\ref{fairness_fvgm_fig:example_dp}) where we impose a universal quantifier on $P$ to obtain $ P = 0 $ as the optimal assignment.
+The solution is visualized as a search tree in Figure~\ref{fvgm_fig:example_dp}, where we observe that storing the solution of sub-problems in the memory avoids repetitive computation, such as exploring the node $ (4,0) $.
+Similarly, the minimum PPV  of the classifier is $ 0.14 $ (not shown in Figure~\ref{fvgm_fig:example_dp}) where we impose a universal quantifier on $P$ to obtain $ P = 0 $ as the optimal assignment.
 \end{example}

### 2022-04-07 chapters/fairness/fvgm/framework2.tex
updated files

@@
 \subsection{{\stochastic} with Correlated Variables} 
-\label{sec:dp_with_BN}
-In {\stochastic} presented in Section~\ref{fairness_fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent.
+\label{fvgm_sec:dp_with_BN}
+In {\stochastic} presented in Section~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent.
 This independence assumption often leads to an \textit{inaccurate estimate} of PPVs because both sensitive and non-sensitive features can be correlated in practical fairness problems.
@@
 In practice, we achieve this by allowing no incoming edge on choice variables while learning $ \BN $ (ref.
-Section~\ref{fairness_fvgm_sec:experiments}).
+Section~\ref{fvgm_sec:experiments}).
 
 For a chance variable $ B_i \in \mathbf{V} $, let $ \parent(B_i) $ denote its parents.
-According to Eq.~\eqref{fairness_fvgm_eq:BN},  for an assignment $ \mathbf{u} $ of $ \parent(B_i) $, $ \BN $ ensures $ B_i $ to be independent of other non-descendant variables in $ \mathbf{V} $.
-Hence, in the recursion of Eq.~\eqref{fairness_fvgm_eq:dp_recurse}, we substitute  $ p_i $  with  $ \Pr[B_i = 1| \parent(B_i) = \mathbf{u}] $.
+According to Eq.~\eqref{fvgm_eq:BN},  for an assignment $ \mathbf{u} $ of $ \parent(B_i) $, $ \BN $ ensures $ B_i $ to be independent of other non-descendant variables in $ \mathbf{V} $.
+Hence, in the recursion of Eq.~\eqref{fvgm_eq:dp_recurse}, we substitute  $ p_i $  with  $ \Pr[B_i = 1| \parent(B_i) = \mathbf{u}] $.
 In order to explicate the dependence on $ \mathbf{u} $, we denote the expected solution of $ S(\mathbf{B}[i:n], \tau) $ as
-$ \mathsf{dp}(i, \tau, \mathbf{u}) $, which for $ B_i \in \mathbf{V} $ is modified as:
+$ \mathsf{dp}(i, \tau, \mathbf{u}) $, which for $ B_i \in \mathbf{V} $ is modified as follows:
 
@@
 Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to know the assignment of parents before computing $ \mathsf{dp} $ on the child.
-Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}') $ for all $ \mathbf{u}' \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities $ \Pr[B_i = 1|\parent(B_i) = \mathbf{u}'] $ into $ \stochastic $.
+Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}) $ for $ \mathbf{u} \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities into $ \stochastic $.
 For this enumeration, we do not store $ \mathsf{dp}(i, \tau, \mathbf{u}) $ in memory.
-However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Section~\ref{fairness_fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.
-Lemma~\ref{fairness_fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fairness_fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).
+However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Section~\ref{fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.
+Lemma~\ref{fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).
 
 \begin{lemma}
-	\label{fairness_fvgm_lm:complexity_dp_with_bn}
+	\label{fvgm_lm:complexity_dp_with_bn}
 	Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $.
 Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $.
-To exactly compute {\stochastic} with correlated variables in dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
+To exactly compute {\stochastic} with correlated variables in the dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
 \end{lemma}
@@
 	\begin{example}
-		We extend Example~\ref{fairness_fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $.
+		We extend Example~\ref{fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $.
 Parameters $ \theta $ imply conditional probabilities $ \Pr[Q|P] = 0.6 $ and $ \Pr[Q|\neg P] = 0.3 $.
-	In Figure~\ref{fairness_fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $.
-We, however, observe that the dynamic programming solution in Section~\ref{fairness_fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $.
-Hence following the calculation in Figure~\ref{fairness_fvgm_fig:example_BN}, we obtain the maximum PPV  as $ 0.65 $ for $ P = 1 $.
+	In Figure~\ref{fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $.
+We, however, observe that the dynamic programming solution in Section~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $.
+Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum PPV  as $ 0.65 $ for $ P = 1 $.
 The minimum PPV (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $.
@@
 
-	\subsection{Extension to Practical Settings}\label{fairness_fvgm_sec:practical}
+	\subsection{Extension to Practical Settings}\label{fvgm_sec:practical}
 	For verifying linear classifiers with real-valued features and coefficients, we preprocess them so that {\fvgm} can be invoked.
 Let $ X_c \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ w_c \in \mathbb{R} $ in the classifier.
-We discretize $ X_c $ to a set of $ k $ Boolean variables $ \mathbf{B}_{c} $ based on the binning-based discretization and assign a Boolean variable to each bin.
+We discretize $ X_c $ to a set $ \mathbf{B}_{c} $ of $ k $ Boolean variables using binning-based discretization and assign a Boolean variable to each bin.
 Hence, $ B_i \in \mathbf{B}_{c} $ becomes $ 1 $, when $ X_c $ belongs to the $ i^\text{th} $ bin.
@@
 		\centering
-		\subfloat[Exploration tree of stochastic sub-set sum problem.]{	\includegraphics[scale=0.3]{figures/exploration_tree_DP}\label{fig:example_dp}}
-		\subfloat[Bayesian network]{\includegraphics[scale=.4]{figures/example_BN}\label{fig:example_BN}}
-		\subfloat[Exploration tree of stochastic sub-set sum problem with Bayesian network]{\includegraphics[scale=.4]{figures/exploration_tree_DP_BN}	\label{fig:example_dp_BN}
+		\subfloat[Exploration tree of stochastic sub-set sum problem.]{	\includegraphics[scale=0.3]{figures/exploration_tree_DP}\label{fvgm_fig:example_dp}}
+		\subfloat[Bayesian network]{\includegraphics[scale=.4]{figures/example_BN}\label{fvgm_fig:example_BN}}
+		\subfloat[Exploration tree of stochastic sub-set sum problem with Bayesian network]{\includegraphics[scale=.4]{figures/exploration_tree_DP_BN}	\label{fvgm_fig:example_dp_BN}
 		}

### 2022-04-07 chapters/fairness/fvgm/introduction.tex
updated files

@@
 	 The following example illustrates a plausible case of unfairness induced by ML algorithms. 
-	\begin{example}\label{fairness_fvgm_example:intro}
-		Following~\cite[Example 1.]{ghosh2020justicia}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fairness_fvgm_fig:example1}).
+	\begin{example}\label{fvgm_example:intro}
+		Following~\cite[Example 1.]{ghosh2020justicia}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}).
 Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions.
@@
 	\vspace*{-1em}
-	\caption{\footnotesize Age-dependent distributions of income and fitness in Example~\ref{fairness_fvgm_example:intro}.}\label{fairness_fvgm_fig:example1}%\vspace*{-1.6em}
+	\caption{Age-dependent distributions of income and fitness in Example~\ref{fvgm_example:intro}.}\label{fvgm_fig:example1}%\vspace*{-1.6em}
 \end{figure}
@@
 More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.
-For Example~\ref{fairness_fvgm_example:intro}, a fairness verifier takes the SVM classifier and the distribution of features $ I, F, A $ as an input and outputs the probability of positive prediction of the classifier for  different sensitive groups.
+For Example~\ref{fvgm_example:intro}, a fairness verifier takes the SVM classifier and the distribution of features $ I, F, A $ as an input and outputs the probability of positive prediction of the classifier for  different sensitive groups.
 
@@
 
-	In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly, \cite{ghosh2020justicia} and \cite{albarghouthi2017fairsquare} propose formal methods that reduce the problem into a solution of an SSAT or an SMT~ formula respectively.	Secondly, \cite{bastani2019probabilistic} propose a sampling-based approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
+	In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly, \cite{ghosh2020justicia} and \cite{albarghouthi2017fairsquare} propose formal methods that reduce the problem into a solution of an SSAT or an SMT~ formula respectively.	Secondly, \cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
 Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.
-sensitive features.
+sensitive features and ignore correlations among non-sensitive features.
 As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.
@@
 
-	\textbf{Our Contributions.} In this paper, we propose a fairness verification framework, namely {\fvgm}, for accurately and efficiently verifying linear classifiers.
+	\textbf{Our Contributions.} In this paper, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.

### 2022-04-07 chapters/fairness/justicia/abstract.tex
working on justicia

@@
 \section{Abstract}
-	As a technology ML is oblivious to societal good or bad, and thus, the field of fair machine learning has stepped up to propose multiple mathematical definitions, algorithms, and systems to ensure different notions of fairness in ML applications.
-	Given the multitude of propositions, it has become imperative to formally verify the fairness metrics satisfied by different algorithms on different datasets.
-	In this paper, we propose a \textit{stochastic satisfiability} (SSAT) framework, {\justicia}, that formally verifies different fairness measures of supervised learning algorithms with respect to the underlying data distribution.
-	We instantiate {\justicia} on multiple classification and bias mitigation algorithms, and datasets to verify different fairness metrics, such as disparate impact, statistical parity, and equalized odds.
-	{\justicia} is scalable, accurate, and operates on non-Boolean and compound sensitive attributes unlike existing distribution-based verifiers, such as FairSquare and VeriFair.
-	Being distribution-based by design, {\justicia} is more robust than the verifiers, such as AIF360, that operate on specific test samples.
+	As a technology machine learning\improvement{Unnecessary abbreviation} is oblivious to societal good or bad, and thus, the field of fair machine learning has stepped up to propose multiple mathematical definitions and algorithms to ensure different notions of fairness in machine learning classifiers.	Given the multitude of propositions, it has become imperative to formally verify the fairness of different classifiers on different datasets.	In this chapter\change{Paper replaced with chapter}, we propose a \textit{stochastic satisfiability} (SSAT) based probabilistic fairness verification framework, called {\justicia}, that formally verifies different fairness measures of supervised classifiers given the distribution of input features.
+We instantiate {\justicia} on multiple classifiers, bias mitigation algorithms, and fairness datasets to verify different fairness metrics, such as disparate impact, statistical parity, and equalized odds.
+	{\justicia} is scalable, accurate, and operates on non-Boolean and compound sensitive features\change{Attributes replaced with features} unlike existing probabilistic verifiers, such as FairSquare and VeriFair.
+	Being distribution-based by design, {\justicia} is more robust than verifiers such as AIF360, that operate on specific test samples.
 	We also theoretically bound the finite-sample error of the verified fairness measure.

### 2022-04-07 chapters/fairness/justicia/introduction.tex
working on justicia

@@
 \section{Introduction}
-%\paragraph{Motivation: ML + Fairness}
-Machine learning (ML) is becoming the omnipresent technology of our time. 
-ML algorithms are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions.
-Thus, human lives are now pervasively influenced by data, ML, and their inherent bias.
+%\paragraph{Motivation: \red{ML} + Fairness}
+Machine learning is becoming the omnipresent technology of our time.
+Machine learning classifiers are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions.
+Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.
 
 \begin{example}\label{fairness_justicia_example:intro}
-Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding eligibility for health insurance depending on the fitness and income of the individuals of different age groups (20-40 and 40-60).
+Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60).
 Typically, incomes of individuals increase as their ages increase while their fitness deteriorates.
-We assume relation of income and fitness depends on the age as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example}.
-Now, if we train a decision tree~\cite{narodytska2018learning} on these fitness and income indicators to decide the eligibility of an individual to get a health insurance, we observe that the `optimal' decision tree (ref.
-Figure~\ref{fairness_justicia_fig:fair_example}) selects a person above and below $40$ years with probabilities $0.18$ and $0.72$ respectively.
-This simple example demonstrates that even if an ML algorithm does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
+We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example}.
+Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref.
+Figure~\ref{fairness_justicia_fig:fair_example}) does not predict based on the sensitive feature age.
+However, a fairness verifier would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively.
+This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
 \end{example}
 
-\begin{figure}[t!]
+\begin{figure}[t!]%%\vspace*{-1em}
+	\centering\hspace*{-4em}
+	\begin{minipage}{.33\columnwidth}
 		\centering
-	\begin{minipage}{.18\columnwidth}
-		\centering
-		\scalebox{0.5}{	
+		\scalebox{0.8}{	
 			\begin{tikzpicture}[x=1.5cm,y=1.8cm]
@@
 			\node[latent,scale=1.5] (a1) {$\textrm{age}$} ; %
-			\node[obs, scale=1.5, below=of a1, xshift=-.6cm] (h) {$\textrm{fitness}$}; %
-			\node[obs, scale=1.5, below=of a1, xshift=.6cm] (i) {$\textrm{income}$}; %
-			\node[obs, scale=1.5, below=of h, xshift=.6cm] (p) {$\hat{Y}$}; %			
+			\node[obs, scale=1.5, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
+			\node[obs, scale=1.5, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
+			\node[obs, scale=1.5, below=of h, xshift=.8cm] (p) {$\hat{Y}$}; %			
 			%%add edge
@@
 		}
-	\end{minipage}\hfill
-	\begin{minipage}{.2\columnwidth}
+	\end{minipage}\hspace*{-2em}
+	\begin{minipage}{.33\columnwidth}
 		\centering
-		\includegraphics[scale=0.35]{figures/fairness/justicia/example.pdf}
-	\end{minipage}\hfill
-	\begin{minipage}{.5\columnwidth}
-		\scalebox{0.37}{	
+		\includegraphics[scale=0.5]{figures/fairness/justicia/example.pdf}
+	\end{minipage}\hspace*{-2em}
+	\begin{minipage}{.35\columnwidth}
+		\scalebox{0.48}{	
 			\begin{tikzpicture}[x=1cm,y=1.8cm]
-			\node [box, scale=2]                                    (p)      {fitness $\geq 0.61$};
-			\node [scale=2, above= -0.1cm of p]  (a)    {Trained Decision Tree};
-			\node [scale=2, box, below=  -1cm of p, xshift=-1.4cm]    (a1)    {income\\ $\geq 0.29$};
-			\node [scale=2, box, below= -1cm of p, xshift=1.4cm]     (a2)    {income $\geq 0.69$};
-			\node [scale=2,below= of a1, xshift=-1.2cm]  (a11)    { $1$};
-			\node [scale=2,below= of a1, xshift=1.2cm]   (a12)    { $0 $};
-			\node [scale=2,below= of a2, xshift=-1.2cm]  (a21)    { $1$};
-			\node [scale=2,below= of a2, xshift=1.2cm]  (a22)    { $0$};
+			\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
+			\node [scale=1.5, above= 0.1cm of p]  (a)    {Trained Decision Tree};
+			\node [scale=1.5, box, below= of p, xshift=-2.5cm]    (a1)    {income\\ $\geq 0.29$};
+			\node [scale=1.5, box, below= of p, xshift=2.5cm]     (a2)    {income $\geq 0.69$};
+			\node [scale=1.5,below= of a1, xshift=-1.8cm]  (a11)    { $\hat{Y}= 1$};
+			\node [scale=1.5,below= of a1, xshift=1.8cm]   (a12)    { $\hat{Y}=0 $};
+			\node [scale=1.5,below= of a2, xshift=-1.8cm]  (a21)    { $\hat{Y}= 1$};
+			\node [scale=1.5,below= of a2, xshift=1.8cm]  (a22)    { $\hat{Y}= 0$};
 			%
-			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
-			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
-			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
-			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
-			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
-			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
+			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Yes};
+			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {No};
+			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Yes};
+			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {No};
+			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Yes};
+			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {No};
 			\end{tikzpicture}}
 	\end{minipage}%
-	\caption{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}
+	\caption{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}%\vspace*{-2em}
+\end{figure}
 
-\end{figure}
-\subsubsection{Fair ML.} Statistical discriminations caused by ML algorithms have motivated researchers to develop several frameworks to ensure fairness and several algorithms to mitigate bias.
-Existing fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}. 
-Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of an algorithm to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
-Separation metrics, such as equalized odds, define an algorithm to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}.
-Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
-%,wu2019counterfactual
+\subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
+\red{In this chapter, we focus on a popular concept of fairness, known as group fairness.}\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
+Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
+Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}.
+Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
 
@@
 Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same.
-Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
+\red{Sufficiency is satisfied if the eligibility is independent of their age given their features are the same}\unsure{check}.
 Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}.
-Different algorithms have also been devised to ensure one or multiple of the fairness definitions.
-These algorithms try to rectify and mitigate the bias in the data and thus in the prediction-model in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the algorithm~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes~\cite{kamiran2012decision,hardt2016equality}.
+To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions.
+These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
 
-\subsubsection{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
+\subsubsection{Fairness Verifiers} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
 
@@
 %These papers operate for a specific Boolean sensitive feature.
-These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a model of a suitable form, and their objective is to verify fairness w.r.t.
-the distribution and the model.
+These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
 Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
@@
 %\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
-Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of an ML algorithm, which is \textit{scalable}, capable of \textit{handling compound protected groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.
+Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of a classifer, which is \textit{scalable}, capable of \textit{handling compound protected groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.
 
@@
 In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
-SSAT is a conceptual framework that has been employed to capture several fundamental problems in AI such as computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
+SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
 Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.
 
-%We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning algorithm using a unified scheme.
+%We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning \red{algorithm} using a unified scheme.
 %Our encodings not only allow us to compute for non-Boolean and compound sensitive features but also to scale significantly better than existing formal verifiers. 
-%We perform experimental analysis for multiple fairness metrics, datasets, and algorithms to instantiate the efficiency and effectiveness of our approach.
+%We perform experimental analysis for multiple fairness metrics, datasets, and \red{algorithm}s to instantiate the efficiency and effectiveness of our approach.
 
@@
 \begin{itemize}
-	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of fairness for different datasets and classification algorithms.
-	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing algorithms with respect to the data  generating distribution}.
+	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of fairness for different datasets and classifiers.
+	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
 	\item Unlike previously proposed formal distributional verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
@@
 
-It is worth remarking that significant advances in AI bear testimony to the right choice of formulation, for example, formulation of planning as SAT~\cite{kautz1992planning}.
+It is worth remarking that significant advances in artificial intelligence bear testimony to the right choice of formulation, for example, formulation of planning as satisfiability (SAT)~\cite{kautz1992planning}.
 In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.
@@
 \iffalse
-With this growing set of algorithms and definitions, it has become important to measure and verify fairness and bias of different algorithms and datasets. 
+With this growing set of \red{algorithm}s and definitions, it has become important to measure and verify fairness and bias of different \red{algorithm}s and datasets. 
 One popular approach is to use specific test dataset to compute the related statistical quantities and to certify fairness for that specific test dataset.
-AIF360~\cite{aif360-oct-2018} provides a unified framework to implement multiple algorithms and to measure their fairness depending on such test datasets.
+AIF360~\cite{aif360-oct-2018} provides a unified framework to implement multiple \red{algorithm}s and to measure their fairness depending on such test datasets.
 Though this method of verification works for a specified datasets, such verifiers do not explain how much a fairness measure depends on which sensitive feature and is not robust to the selection of test dataset and its size.

### 2022-04-07 chapters/fairness/justicia/preliminaries.tex
working on justicia

@@
 
+\red{\red{PPV}}\change{PPV to PP, better to not use short form}
+
 As we illustrated in Example~\ref{fairness_justicia_example:intro}, a classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
@@
 To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ provided the distribution of features $ \mathcal{D} $.
-More specifically,  {\justicia} verifies fairness metrics where $ f(\alg, \mathcal{D}) $ depends on the conditional Positive Predictive Value (PPV) of the classifier, denoted by $ \Pr[\hat{Y} = 1 | \mathsf{C}] $.
+More specifically,  {\justicia} verifies fairness metrics where $ f(\alg, \mathcal{D}) $ depends on the conditional probability of positive prediction of the classifier, denoted by $ \Pr[\hat{Y} = 1 | \mathsf{C}] $.
 Henceforth, each fairness metric is identified by the particular choice of the conditional $ \mathsf{C} $, as discussed next.
@@
 The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups.
-Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in \sensitive$.
+Formally, independence notion specifies an equal probability of positive prediction across all sensitive groups for a classifier, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in \sensitive$.
 Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}, are proposed in the literature~\cite{dwork2012fairness,feldman2015certifying}.
 
-\textit{Disparate impact} (DI) measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$~\cite{feldman2015certifying}.
+\textit{Disparate impact} (DI) measures the ratio of probability of positive predictions between the most favored group and least favored group, and prescribe it to be close to $1$~\cite{feldman2015certifying}.
 Formally, the disparate impact of a classifier is
@@
 
-Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the maximum difference of PPVs among sensitive groups, and prescribe this quantity to be near zero.
+Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the maximum difference of probability of positive predictions among sensitive groups, and prescribe this quantity to be near zero.
 Thus, the statistical parity of  a classifier is
@@
 In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier is independent of the sensitive features $\sensitive$ given the class labels $Y$.
-In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the maximum difference of false positive rates \red{(FPR)} and true positive rates \red{(TPR)} conditioned on sensitive groups.
+In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the maximum difference of conditional false positive rates \red{(FPR)} and conditional true positive rates \red{(TPR)} with respect to sensitive groups.
 Formally,  for $ Y \in \{0,1\} $,  the equalized odds of a classifier is defined as
@@
 This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
-To this end, the path-specific causal fairness of a classifier is \[
+To this end, the path-specific causal fairness, abbreviated as PCF, of a classifier is \[
  f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
@@
 
-Therefore, PCF constrains that $ \hat{Y} $ is not directly dependent of $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through $ \mediator $.
-PCF is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
+Therefore, path-specific cause fairness constrains that the prediction $ \hat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through mediator features $ \mediator $.
+Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
 
@@
 Let `male' be the most favored group by the classifier, for instance.
-Path-specific causal fairness (PCF) ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
+Path-specific causal fairness ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
 She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.
-Therefore, PCF measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
+Therefore, path-specific causal fairness measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
 \end{example}
 
-\paragraph{Fairness Certification.} We certify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, which quantifies the desired level of fairness.
+\paragraph{Fairness Certification.} We certify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, that quantifies the desired level of fairness.
 In particular, a classifier is $ \epsilon $-fair with respect to statistical parity, equalized odds, or path-specific causal fairness if and only if $ f(\alg, \mathcal{D}) \le \epsilon $.
-In contrast, a classifier satisfies $(1 - \epsilon)$-disparate impact  if and only if $ f(\alg, \mathcal{D}) \ge 1 - \epsilon $.
+In contrast, a classifier achieves $(1 - \epsilon)$-disparate impact  if and only if $ f(\alg, \mathcal{D}) \ge 1 - \epsilon $.
 In all above fairness metrics, a lower value of $ \epsilon $ refers to higher fairness of the classifier.
@@
 \end{enumerate}
-where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the formula $ \phi $ with $ 1 $ and $ 0 $, respectively.
+where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively.
 In this paper, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
@@
 The problem of SSAT and its variants have been pursued by theoreticians and practitioners for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
-We refer the reader to~\cite{lee2017solving,lee2018solving} for detailed survey.
+We refer the reader to~\cite{lee2017solving,lee2018solving} for a detailed survey.
 It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.

### 2022-04-07 chapters/fairness/justicia/framework.tex
working on justicia

@@
 \label{fairness_justicia_sec:framework}
-In this section, we present the primary contribution of this paper, {\justicia}, which is an SSAT-based framework for verifying multiple group and causal fairness metrics.
+In this section, we present the primary contribution of this chapter, {\justicia}, which is an SSAT-based framework for verifying group and causal fairness metrics.
 Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness definition, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.

### 2022-04-08 chapters/fairness/justicia/framework.tex
upto example in re-ssat encoding

@@
 In this section, we present the primary contribution of this chapter, {\justicia}, which is an SSAT-based framework for verifying group and causal fairness metrics.
-Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness definition, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.
+Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness metric $ f(\alg, \mathcal{D}) $, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.
 Additionally, if a fairness threshold $ \epsilon \in [0,1] $ is provided, {\justicia} verifies whether the classifier is $ \epsilon $-fair by comparing $ f $ with $ \epsilon $.
 In {\justicia}, we focus on classifiers that can be represented as a CNF formula defined over a set of Boolean variables.
-Additionally, for each variable, we query the distribution $ \mathcal{D} $ to derive the marginal probability of assigning to $ 1 $.
-In this section, we discuss two equivalent approaches for fairness verification based on SSAT-based encodings: enumeration approach and inferring approach.
+Additionally, for each variable, we query the distribution $ \mathcal{D} $ to derive the marginal probability of the variable to be assigned to $ 1 $.
+In this section, we discuss two equivalent approaches for fairness verification based on SSAT-based encodings: enumeration approach and \red{inferring} approach.
 In both approaches, we verify fairness with the presence of compound sensitive groups.
@@
 \label{fairness_justicia_sec:enumeration_ssat}
-In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional PPV of the classifier, $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ based on an appropriately designed SSAT formula.
+In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional probability of positive prediction of the classifier, $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ by solving an appropriately designed SSAT formula.
 For simplicity, we  initially make assumptions on the classifier $ \alg $ and discuss practical relaxations later in this section.
@@
 Additionally, all features $ \nonsensitive \cup \sensitive $ are assumed to be Boolean variables.
-Finally, we consider independence probability assumption of non-sensitive features on sensitive features, where $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of the non-sensitive feature $ X_i $.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
+Finally, we consider independence probability assumption of non-sensitive features $ \nonsensitive $, where $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of $ X_i $.
 
@@
 In the RE-SSAT formula $ \Phi_{\mathbf{a}} $, existentially quantified variables  $ \{A_1, \dots, A_{\numsensitive}\} $ are assigned values  according  to the constraint $ \sensitive =\mathbf{a} $.
-\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Next,  an SSAT solver computes the probability $ \Pr[\Phi_{\mathbf{a}}] $ by considering the random values of $ \{X_1, \dots, X_{\numnonsensitive}\} $ while fixing the assignment of $ \{A_1, \dots, A_{\numsensitive\}} $.
-Therefore, $ \Pr[\Phi_{\mathbf{a}}] $ equals the conditional PPV of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $, for the sensitive group $ \sensitive = \mathbf{a} $.
+\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Next,  an SSAT solver computes the probability $ \Pr[\Phi_{\mathbf{a}}] $ by considering the random values of $ \{X_1, \dots, X_{\numnonsensitive}\} $ while fixing the assignment of $ \{A_1, \dots, A_{\numsensitive}\} $.
+Therefore, $ \Pr[\Phi_{\mathbf{a}}] $ equals the conditional probability of positive prediction of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $, for the sensitive group $ \sensitive = \mathbf{a} $.
 
-For simplicity, we have described the computation of conditional PPVs without considering the correlation among sensitive and non-sensitive features.
-In reality, correlation exists among sensitive and non-sensitive features.
-Thus, non-sensitive features may have different conditional distributions for different sensitive groups.
-We incorporate these conditional distributions in this RE-SSAT encoding by providing the conditional probability $ p_i = \Pr[X_i = 1| \sensitive =\mathbf{a}] $ instead of the independent probability $\Pr[X_i = 1]$ for each $X_i\in \nonsensitive$.
-We illustrate this method in Example~\ref{fairness_justicia_example:re_ssat}.
+For simplicity, we have described the computation of conditional probability $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ without considering the correlation among sensitive and non-sensitive features.
+In reality, correlation exists among these features.
+As a result, non-sensitive features may have different conditional distributions for different sensitive groups.
+For a non-sensitive feature $ X_i $, we incorporate its conditional probability in the RE-SSAT encoding by setting $ p_i = \Pr[X_i = 1| \sensitive =\mathbf{a}] $ instead of the independent probability $\Pr[X_i = 1]$.
+Next, we illustrate this enumeration approach in Example~\ref{fairness_justicia_example:re_ssat}.
 %This relaxes the independence requirement.
@@
 	\label{fairness_justicia_example:re_ssat}
-	Here, we illustrate the RE-SSAT formula for calculating the PPV for the sensitive group `age $ \ge 40 $' in the decision tree of Figure~\ref{fairness_justicia_fig:fair_example}.
-We assign three Boolean variables $ F,I,J $ for the three nodes in the tree such that the literal $ F,I,J $ denote `fitness $ \ge 0.61 $', `income $ \ge 0.29 $', and `income $ \ge 0.69 $', respectively.
-We consider another Boolean variable $A$  where the literal $ \sensitive $ represents the sensitive group `age $ \ge 40 $'.
+	We illustrate the RE-SSAT encoding for calculating the probability of positive prediction for the sensitive group age $ \ge 40 $ in the decision tree of Figure~\ref{fairness_justicia_fig:fair_example}.
+We assign three Boolean variables $ F,I,J $ for three nodes in the decision tree, where literal $ F,I,J $ denote fitness $ \ge 0.61 $, income $ \ge 0.29 $, and income $ \ge 0.69 $, respectively.
+We consider another Boolean variable $A$  where the literal $ A $ represents the sensitive group age $ \ge 40 $ and $ \neg A $ denotes age $ < 40 $.
 Thus, the CNF formula  for the decision tree is $ (\neg F \vee I) \wedge (F \vee J) $.
 From the distribution in Figure~\ref{fairness_justicia_fig:fair_example}, we get $ \Pr[F] = 0.41, \Pr[I] = 0.93 $, and $ \Pr[J] = 0.09 $.
-Given this information, we calculate the PPV for the sensitive group `age $ \ge 40 $' by solving the RE-SSAT formula:
+Given this information, we calculate the probability of positive prediction for the sensitive group age $ \ge 40 $ by solving the following RE-SSAT formula:
 	\begin{equation}
@@
 	From the solution to this SSAT formula, we get $ \Pr[\Phi_A] = 0.43 $.
-Similarly, to calculate the PPV for the group `age $ < 40 $', we replace the unit (single-literal) clause $ \sensitive $ with $ \neg A $ in the CNF in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $ where $ \Pr[\Phi_{\neg A}] = 0.43 $.
-	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of $A$ and $\neg A$, both age groups demonstrate equal PPV as the sensitive feature is not explicitly present in the classifier. 
+Similarly, to calculate the probability of positive prediction for the group age $ < 40 $, we replace the unit clause\footnote{A unit clause is a clause with a single literal.} $ A $ with $ \neg A $ in the CNF formula in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $.
+	\[ \Phi_{\neg A} := \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge \neg A \]
+	For $ \Phi_{\neg A} $, the solution $ \Pr[\Phi_{\neg A}] = 0.43 $ is similarly derived from an SSAT solver. 
+	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of the sensitive feature $A$, both age groups demonstrate equal probability of positive prediction as the sensitive feature is not explicitly present in the classifier.
+
 	However, there is an implicit bias in the data distribution for different sensitive groups and the classifier unintentionally learns it.
-	To capture this implicit bias, we calculate the conditional probabilities  $ \Pr[F|A] = 0.01, \Pr[I|A] = 0.99 $, and $ \Pr[J|A] = 0.18 $ from the distribution.
-Using the conditional probabilities in  $\Phi_A $, we find that $ \Pr[\Phi_A] = 0.18 $ for `age $ \ge 40 $'.
-For `age $ < 40 $',  we similarly obtain $ \Pr[F|\neg A] = 0.82, \Pr[I|\neg A] = 0.88 $, and $ \Pr[J|\neg A] = 0.01 $, and thus  $ \Pr[\Phi_{\neg A}] = 0.72 $.
-	Therefore, presented RE-SSAT encoding detects the discrimination of the classifier among different sensitive groups.
-An astute reader would observe that $I$ and $J$ are not independent.
+To capture this implicit bias, we calculate conditional probabilities  $ \Pr[F|A] = 0.01, \Pr[I|A] = 0.99 $, and $ \Pr[J|A] = 0.18 $ from the distribution for group age $ \ge 40 $ in Figure~\ref{fairness_justicia_fig:fair_example}.
+Providing the conditional probabilities, we construct a modified SSAT formula $\Phi'_A $ and compute $ \Pr[\Phi'_A] = 0.18 $ for age $ \ge 40 $.
+	\begin{equation}
+	\Phi'_A := \R^{0.01}F, \R^{0.99}I, \R^{0.18}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge A\notag
+	\end{equation}
+	For the sensitive group age $ < 40 $,  we similarly obtain $ \Pr[F|\neg A] = 0.82, \Pr[I|\neg A] = 0.88 $, $ \Pr[J|\neg A] = 0.01 $, construct the modified formula $ \Phi'_{\neg A} $ and get  $ \Pr[\Phi'_{\neg A}] = 0.72 $.
+	\[ \Phi'_{\neg A} := \R^{0.82}F, \R^{0.88}I, \R^{0.01}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge \neg A \]	
+	In the later case with correlations among sensitive and non-sensitive features, the RE-SSAT encoding detects the discrimination of the classifier among different sensitive groups, where the classifier is more biased towards the younger group with age $ < 40 $ than the elderly group with age $ \ge 40 $.
+
+	\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
+
+%	An astute reader would observe that $I$ and $J$ are not independent.
 Following~\cite{chavira2008probabilistic}, we can simply capture relationship between the variables using constraints and if needed, auxiliary variables.
@@
 
-\subsection{Learning Fairness with ER-SSAT Encoding}
+\subsection{\red{Learning Fairness with ER-SSAT Encoding}}
 \label{fairness_justicia_sec:learn_ssat}
@@
 Therefore, the enumeration approach may suffer from scalability issues. 
-Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their PPVs to measure different fairness metrics.
+Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their \red{probability of positive prediction}s to measure different fairness metrics.
 
@@
 In ER-SSAT formula, the probability of satisfying $ \Phi $ is the \textit{maximum} satisfying probability over the existentially quantified variables given the randomized quantified variables (by Rule 2, Sec.~\ref{fairness_justicia_sec:ssat}).
-In this paper, we leverage this property to compute the most favored group with the highest PPV. 
+In this paper, we leverage this property to compute the most favored group with the highest \red{probability of positive prediction}. 
 We consider the following ER-SSAT formula. 
@@
 Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_{\numsensitive} = a^{\max}_{\numsensitive} $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum.
-Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_{\numsensitive} \}$ achieving the highest PPV. 
-%Since an assignment to $\{A_1, \dots, A_{\numsensitive}\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest PPV.
+Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_{\numsensitive} \}$ achieving the highest \red{probability of positive prediction}. 
+%Since an assignment to $\{A_1, \dots, A_{\numsensitive}\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest \red{probability of positive prediction}.
 
 \subsubsection{Learning the Least Favored Group.}
-In order to learn the least favored group in terms of PPV, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ X_1, \dots, X_{\numnonsensitive} $.
+In order to learn the least favored group in terms of \red{probability of positive prediction}, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ X_1, \dots, X_{\numnonsensitive} $.
 In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the sensitive variables and randomized quantification over the non-sensitive variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
@@
 As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the sensitive features $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_{\numsensitive}\}$ that maximizes $\Phi'_{\mathsf{ER}}$.
-If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the PPV of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$.
+If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the \red{probability of positive prediction} of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$.
 We present the algorithm for this learning approach, namely {\justicialearn} in Algorithm~\ref{fairness_justicia_algo:enum} (Line~\ref{fairness_justicia_algo:justicia_learn_begin}--\ref{fairness_justicia_algo:justicia_learn_end}).
@@
 	As we solve $ \Phi_{\mathsf{ER}} $, we learn that the assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, i.e.
-`male individuals with age $ < 40 $' is the most favored group with PPV computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
+`male individuals with age $ < 40 $' is the most favored group with \red{probability of positive prediction} computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
 Similarly, to learn the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
@@
 	Solving $ \Phi_{\mathsf{ER'}} $, we learn the assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi_{\mathsf{ER'}}] = 0.57 $.
-Thus, `female individuals with age $ < 40 $' constitute the least favored group with PPV:  $ 1-0.57 = 0.43$.
+Thus, `female individuals with age $ < 40 $' constitute the least favored group with \red{probability of positive prediction}:  $ 1-0.57 = 0.43$.
 	Thus, {\justicialearn} allows us to learn the most and least favored groups and the corresponding discrimination.
 \end{example}
-We use the PPVs of the most and least favored groups to compute different fairness metrics.
+We use the \red{probability of positive prediction}s of the most and least favored groups to compute different fairness metrics.
 We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.
@@
 	\label{fairness_justicia_lm:equivalence}
-	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the PPV of the compound sensitive group $ \mathbf{a} \in A $.
+	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the \red{probability of positive prediction} of the compound sensitive group $ \mathbf{a} \in A $.
 If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
@@
 These finite sample set introduce errors in the computed probabilities of the randomised quantifiers being $1$.
-These finite-sample errors in computed probabilities induce further errors in the computed positive predictive value (PPV) and fairness metrics.
+These finite-sample errors in computed probabilities induce further errors in the computed positive predictive value (\red{probability of positive prediction}) and fairness metrics.
 We next provide a bound on this finite-sample error.

### 2022-04-08 chapters/fairness/justicia/framework.tex
enumeration approach complete

@@
 
-	\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
-
 %	An astute reader would observe that $I$ and $J$ are not independent.
@@
 \subsubsection{Measuring Fairness Metrics.}
-As we compute the probability $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics. 
-For that, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ that requires solving exponential (with $n$) number of SSAT instances. 
+As we compute $ \Pr[\Phi_\mathbf{a}] = \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics.
+To this end, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ by solving an exponential number (with $ \numsensitive $) of SSAT formulas.
 We elaborate this enumeration approach, namely {\justiciaenum}, in Algorithm~\ref{fairness_justicia_algo:enum}  (Line~\ref{fairness_justicia_algo:justicia_enum_begin}--\ref{fairness_justicia_algo:justicia_enum_end}).
@@
 	\begin{algorithmic}[1]
-		\Function{{\justiciaenum}}{$ X,A,\hat{Y} $}
+		\Function{{\justiciaenum}}{$ \nonsensitive, \sensitive,\hat{Y} $}
 		\label{fairness_justicia_algo:justicia_enum_begin}
@@
 		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
-		\ForAll{$\mathbf{a} \in A$ }
-		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i | \mathbf{a}), \forall X_i \in X $
+		\ForAll{$\mathbf{a} \in \sensitive$ }
+		\State $ p_{i} \leftarrow \Pr[X_i = 1|\sensitive = \mathbf{a}], \forall X_i \in \nonsensitive $
 		\State $ \phi := \phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}) $
@@
 		\EndFor
-		\State \Return $ \max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] $
+		\State \Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
 		\label{fairness_justicia_algo:justicia_enum_end}
@@
 
-		\Function{{\justicialearn}}{$ X,A,\hat{Y} $}
+		\Function{{\justicialearn}}{$ \nonsensitive, \sensitive,\hat{Y} $}
 		\label{fairness_justicia_algo:justicia_learn_begin}
 		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
-		\State $ p_{i} \leftarrow \mathsf{CalculateProb}(X_i), \forall X_i \in X $
+		\State $ p_{i} \leftarrow \Pr[X_i=1], \forall X_i \in \nonsensitive $
 		\State $  \Phi_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
@@
 
-We calculate the ratio of the minimum and the maximum probabilities according to the definition of disparate impact. 
-We compute statistical parity by taking the difference between the maximum and the minimum probabilities of all $ \Pr[\Phi_{\mathbf{a}}] $.
-Moreover, to measure equalized odds, we compute two SSAT instances for each compound group with modified values of $ p_i $. 
-Specifically, to compute TPR, we use the conditional probability $ p_i = \Pr[X_i|Y=1] $ on samples with class label $ Y = 1 $ and take the difference between the maximum and the minimum probabilities of all compound groups.
-In addition, to compute FPR, we use the conditional probability $ p_i = \Pr[X_i|Y=0] $ on samples with $ Y = 0 $ and take the difference similarly.
-Thus, {\justiciaenum} allows us to compute different fairness metrics using a unified algorithmic framework.
+To measure the disparate impact of a classifier, we calculate the ratio between the minimum and the maximum conditional probability of positive prediction of the classifier, which are  $ \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $ and $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $, respectively.
+We compute statistical parity by taking the difference between $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $ and $ \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $.
 
-\subsection{\red{Learning Fairness with ER-SSAT Encoding}}
+Moreover, to compute equalized odds, we call {\justicia} twice, one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $.
+In both calls, we compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of equalized odds.
+For measuring path-specific causal fairness, we compute  $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ by conditioning the distribution $ \mathcal{D} $ by mediator features $ \mathbf{Z} $ and take their difference.
+Thus, {\justiciaenum} allows us to compute different group and causal fairness metrics using a unified algorithmic framework.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
+
+\subsection{\red{Inference Approach with ER-SSAT Encoding}}
 \label{fairness_justicia_sec:learn_ssat}

### 2022-04-09 chapters/fairness/justicia/framework.tex
inference appraoch

@@
 
-\subsection{\red{Inference Approach with ER-SSAT Encoding}}
+\subsection{\red{Inference Approach using ER-SSAT Encoding}}
 \label{fairness_justicia_sec:learn_ssat}
-In most practical problems, there can be exponentially many compound groups based on the different combinations of valuation to the sensitive features. 
-Therefore, the enumeration approach may suffer from scalability issues. 
-Hence, we propose efficient SSAT encodings to \textit{learn} the most favored group and the least favored group for given  $\alg$ and $ \mathcal{D} $, and to compute their \red{probability of positive prediction}s to measure different fairness metrics.
+In most practical problems, there can be exponentially many compound groups due to the combinatorial combination of categorical sensitive features.
+As a result, the enumeration approach may suffer from scalability issues due to an exponential number of calls to the SSAT solver.
+Hence, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the the most favored sensitive group with the maximum conditional probability of positive prediction of the classifier and another for inferring the least favored sensitive group with the minimum conditional probability of positive prediction of the same classifier.
+As shown above, these two probabilities are sufficient to measure different group and causal fairness metrics.
 
-\subsubsection{Learning the Most Favored Group.}
+\subsubsection{Inferring the Most Favored Sensitive Group.}
 In an SSAT formula $ \Phi $, the order of quantification of the Boolean variables in the prefix  carries distinct interpretation of the satisfying probability of $ \Phi $.  

### 2022-04-10 chapters/fairness/justicia/framework.tex
upto inference approach

@@
 
-\subsubsection{Measuring Fairness Metrics.}
+\subsubsection{Measuring Fairness Metrics}
 As we compute $ \Pr[\Phi_\mathbf{a}] = \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics.
@@
 		\State $ p_{i} \leftarrow \Pr[X_i=1], \forall X_i \in \nonsensitive $
-		\State $  \Phi_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
-		\State $  \Phi'_\mathbf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
-		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
+		\State $  \Phi_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
+		\State $  \Phi'_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
+		\State \Return $ \mathsf{SSAT}(\Phi_\mathsf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathsf{ER}) $
 		\label{fairness_justicia_algo:justicia_learn_end}
@@
 
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
-
 \subsection{\red{Inference Approach using ER-SSAT Encoding}}
 \label{fairness_justicia_sec:learn_ssat}
-In most practical problems, there can be exponentially many compound groups due to the combinatorial combination of categorical sensitive features.
+In most practical problems, there can be exponentially many compound sensitive groups due to the combinatorial combination of categorical sensitive features.
 As a result, the enumeration approach may suffer from scalability issues due to an exponential number of calls to the SSAT solver.
-Hence, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the the most favored sensitive group with the maximum conditional probability of positive prediction of the classifier and another for inferring the least favored sensitive group with the minimum conditional probability of positive prediction of the same classifier.
+To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the the most favored sensitive group with the maximum conditional probability of positive prediction of the classifier and another for inferring the least favored sensitive group with the minimum conditional probability of positive prediction of the same classifier.
 As shown above, these two probabilities are sufficient to measure different group and causal fairness metrics.
 
-\subsubsection{Inferring the Most Favored Sensitive Group.}
-In an SSAT formula $ \Phi $, the order of quantification of the Boolean variables in the prefix  carries distinct interpretation of the satisfying probability of $ \Phi $.  
-In ER-SSAT formula, the probability of satisfying $ \Phi $ is the \textit{maximum} satisfying probability over the existentially quantified variables given the randomized quantified variables (by Rule 2, Sec.~\ref{fairness_justicia_sec:ssat}).
-In this paper, we leverage this property to compute the most favored group with the highest \red{probability of positive prediction}. 
-We consider the following ER-SSAT formula. 
+\subsubsection{Inferring the Most Favored Sensitive Group}
+In the prefix of an SSAT formula $ \Phi $, the order of quantified variables carries distinct interpretation of  $ \Pr[\Phi] $.
+In an ER-SSAT formula, $ \Pr[\Phi] $ is the \textit{maximum} satisfying probability of $ \Phi $ over the optimal assignment of existentially quantified variables given the randomized quantified variables (by Semantic 2, Sec.~\ref{fairness_justicia_sec:ssat}).
+In this paper, we leverage this property to compute the most favored sensitive group with the highest probability of positive prediction of a classifier.
+In particular, we consider the following ER-SSAT formula:
+
 \begin{equation}
@@
 \end{equation}
-%In the prefix of the ER-SSAT formula $ \Phi_{\mathsf{ER}} $, the sensitive variables are  existentially quantified and they are followed by randomized quantified non-sensitive variables, which is in reverse order  in Eq.~\eqref{fairness_justicia_eq:re}.
-The CNF formula $\phi_{\hat{Y}}$ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound sensitive group.
-Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_{\numsensitive} = a^{\max}_{\numsensitive} $ for which the satisfying probability $ \Pr[\Phi_{\mathsf{ER}}] $ is maximum.
-Thus, we compute  the most favored group $ \mathbf{a}_{\mathsf{fav}} \triangleq \{ a^{\max}_1, \dots, a^{\max}_{\numsensitive} \}$ achieving the highest \red{probability of positive prediction}. 
-%Since an assignment to $\{A_1, \dots, A_{\numsensitive}\}$ uniquely denotes a compound sensitive group, say $ \mathbf{a}_{\mathsf{fav}} \in A $, we find the most favored group $ \mathbf{a}_{\mathsf{fav}} $ for which $ \Pr[\Phi_{\mathsf{ER}}] $ achieves the highest \red{probability of positive prediction}.
 
-\subsubsection{Learning the Least Favored Group.}
-In order to learn the least favored group in terms of \red{probability of positive prediction}, we  compute the \textit{minimum} satisfying probability of the classifier $ \phi_{\hat{Y}} $ given the random values of the non-sensitive variables $ X_1, \dots, X_{\numnonsensitive} $.
-In order to do so, we have to solve a `universal-random' (UR) SSAT formula (Eq.~\eqref{fairness_justicia_eq:ar}) with  universal quantification over the sensitive variables and randomized quantification over the non-sensitive variables (by Rule 3, Sec.~\ref{fairness_justicia_sec:ssat}).
+The CNF formula $\phi_{\hat{Y}}$ in $ \Phi_{\mathsf{ER}} $ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound sensitive group.
+Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the optimal assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_{\numsensitive} = a^{\max}_{\numsensitive} $ for which the probability of satisfaction of the ER-SSAT formula $ \Pr[\Phi_{\mathsf{ER}}] $ becomes maximum.
+Thus, we compute  the most favored group $ \red{\mathbf{a}_{\max}} \triangleq [ a^{\max}_1, \dots, a^{\max}_{\numsensitive} ] $ achieving the highest probability of positive prediction of the classifier.
+
+\subsubsection{Inferring the Least Favored Sensitive Group}
+In order to infer the least favored sensitive group by a classifier, we  compute the \textit{minimum} conditional probability of positive prediction of the classifier with respect to all sensitive groups given the random values of the non-sensitive features.
+To this end, we  solve a `universal-random' (UR) SSAT formula with  universal quantifiers over sensitive features and randomized quantification over non-sensitive features (by Semantic 3, Sec.~\ref{fairness_justicia_sec:ssat}).
 \begin{equation}
 \Phi_{\mathsf{UR}} := \forall A_1,\dots, \forall A_{\numsensitive},
-\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}.
+\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}
 \label{fairness_justicia_eq:ar}
 \end{equation}
-A UR-SSAT formula returns the minimum satisfying probability of $ \phi $ over the universally quantified variables in contrast to the ER-SSAT formula that returns the maximum satisfying  probability over the existentially quantified variables.
-Due to practical issues to solve UR-SSAT formula, in this paper, we leverage the \textit{duality} between UR-SSAT (Eq.~\eqref{fairness_justicia_eq:ar}) and ER-SSAT formulas (Eq.~\eqref{fairness_justicia_eq:er_complement})
+%By solving $ \Phi_{\mathsf{UR}} $, we find the optimal assignment to the universally quantified variables $ A_1 = a^{\min}_1, \dots,A_{\numsensitive} = a^{\min}_{\numsensitive} $ for which the probability of satisfaction of the UR-SSAT formula $ \Pr[\Phi_{\mathsf{UR}}] $ becomes minimum.
+Thus, $ \red{\mathbf{a}_{\min}} \triangleq [ a^{\min}_1, \dots, a^{\min}_{\numsensitive} ] $ becomes the least favored sensitive group achieving the lowest probability of positive prediction of the classifier.
+
+\red{Solving an UR-SSAT formula raises several practical issues.} As a result, there is an unavailability of solvers for UR-SSAT formulas.
+To resolve this problem, we leverage the \textit{duality} between UR-SSAT  and ER-SSAT formulas, where we solve an UR-SSAT formula on the CNF $ \phi $ using the solution of an ER-SSAT formula on the complemented CNF $ \neg \phi $~\cite{littman2001stochastic}.
+More specifically, we solve the following ER-SSAT formula for finding the least favored sensitive group.
+
 \begin{equation}
 	\Phi'_{\mathsf{ER}} := \exists A_1,\dots, \exists A_{\numsensitive},
-\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \neg \phi_{\hat{Y}}.
+	\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \neg \phi_{\hat{Y}}
 	\label{fairness_justicia_eq:er_complement}
 \end{equation}
-%\begin{equation}
-%\Phi'_{\mathsf{ER}} = \underbrace{\exists a_1,\dots, \exists a_{\numsensitive}}_{\text{sensitive features}},
-%\underbrace{\R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}}_{\text{non-sensitive features}},   \; \underbrace{\neg \phi_{\hat{Y}} }_{\phi}.
-%\label{fairness_justicia_eq:er_complement}
-%\end{equation}
-and solve the UR-SSAT formula on the CNF $ \phi $ using the ER-SSAT formula on the complemented CNF $ \neg \phi $~\cite{littman2001stochastic}.
-Lemma~\ref{fairness_justicia_thm:dual} encodes this duality.
-%We apply such a duality because of the practical unavailability of an UR-SSAT solver (to the best of our knowledge).
-Formally, we solve the following ER-SSAT formula to find the least favored group.
+
+In Lemma~\ref{fairness_justicia_thm:dual}, we discuss the duality between UR-SSAT and ER-SSAT formulas.
+
 \begin{lemma}\label{fairness_justicia_thm:dual}
@@
 \end{lemma}
-As we solve $\Phi'_{\mathsf{ER}}$, we obtain the assignment to the sensitive features $\mathbf{a}_{\mathsf{unfav}} \triangleq \{a^{min}_1, \dots, a^{min}_{\numsensitive}\}$ that maximizes $\Phi'_{\mathsf{ER}}$.
-If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $,  which is the \red{probability of positive prediction} of the least favored group $ \mathbf{a}_{\mathsf{unfav}}$.
-We present the algorithm for this learning approach, namely {\justicialearn} in Algorithm~\ref{fairness_justicia_algo:enum} (Line~\ref{fairness_justicia_algo:justicia_learn_begin}--\ref{fairness_justicia_algo:justicia_learn_end}).
+
+As we solve $\Phi'_{\mathsf{ER}}$, we obtain the optimal assignment to the sensitive features $\mathbf{a}_{\min} \triangleq [a^{min}_1, \dots, a^{min}_{\numsensitive}]$ that maximizes $\Phi'_{\mathsf{ER}}$.
+If $ p $ is the maximum satisfying probability of $ \Phi'_{\mathsf{ER}} $, then according to Lemma~\ref{fairness_justicia_thm:dual}, $ 1 - p $ is the minimum satisfying probability of $ \Phi_{\mathsf{UR}} $;  which is also the minimum probability of positive prediction of the classifier.
+We present the algorithm for the inference approach, namely {\justicialearn} in Algorithm~\ref{fairness_justicia_algo:enum} (Line~\ref{fairness_justicia_algo:justicia_learn_begin}--\ref{fairness_justicia_algo:justicia_learn_end}).
 
@@
 
-In ER-SSAT formula of Eq.~\eqref{fairness_justicia_eq:er_complement}, we need to negate the classifier $ \phi_{\hat{Y}} $ to another CNF formula $ \neg \phi_{\hat{Y}} $.
-The na\"ive approach of negating a CNF to another CNF generates exponential number of new clauses.
-Here, we can apply Tseitin transformation that increases the clauses linearly while introducing linear number of new variables~\cite{tseitin1983complexity}.
-As an alternative, we also directly encode the classifier $\alg$ for the negative class label $\hat{Y} = 0$ as a CNF formula and pass it to $\Phi'_{\mathsf{ER}} $, if possible.
+In the ER-SSAT formula in Eq.~\eqref{fairness_justicia_eq:er_complement}, we need to negate the classifier $ \phi_{\hat{Y}} $ to another CNF formula $ \neg \phi_{\hat{Y}} $.
+The na\"ive approach of negating one CNF to another CNF generates an exponential number of new clauses.
+Here, we apply Tseitin transformation for the negation, which increases the number of clauses linearly while introducing a linear number of new variables~\cite{tseitin1983complexity}.
+As an alternative approach, we also directly encode the binary classifier $\alg$ for the negative class label $\hat{Y} = 0$ as a CNF formula and pass it to $\Phi'_{\mathsf{ER}} $, whenever it is possible.
 The last approach is generally more efficient than the other approaches as the resulting CNF is often smaller.
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
 

### 2022-04-10 chapters/fairness/justicia/preliminaries.tex
upto inference approach

@@
 
-\subsubsection{Group Fairness.} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\justicia} verifies independence and separation metrics. 
+\subsubsection{Group Fairness} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\justicia} verifies independence and separation metrics. 
 The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups.
@@
 
-\subsubsection{Path-specific Causal Fairness.}
+\subsubsection{Path-specific Causal Fairness}
 Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.

### 2022-04-11 chapters/fairness/justicia/framework.tex
example for RE-SSAT encoding

@@
 The last approach is generally more efficient than the other approaches as the resulting CNF is often smaller.
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
 
@@
 	\label{fairness_justicia_example:er_ssat}
-	Here, we illustrate the ER-SSAT encodings for learning the most favored and the least favored group in presence of multiple sensitive groups.
-As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another sensitive group `sex $ \in $ \{male, female\}'.
-Consider a Boolean variable $ S $ for `sex' where the literal $ S $ denotes `sex = male'.
-With this new sensitive feature, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where $ A,F,I,J $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
-	Hence, we obtain the ER-SSAT formula of $\alg$ to learn the most favored group:
-	$ \Phi_{\mathsf{ER}} =  \exists S,\exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; (\neg F \vee I \vee S) \wedge (F \vee J).
-	$
+	Here, we illustrate the ER-SSAT encoding for inferring the most favored and the least favored sensitive group of a classifier in the presence of multiple compound sensitive groups.
+As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another Boolean sensitive feature `sex' $ \in $ \{male, female\}.
+Consider a Boolean variable $ S $ for sex where the literal $ S $ denotes sex = male.
+With this new sensitive feature, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where variables $ F,I,J $, and $ A $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
+Hence, we obtain the following ER-SSAT formula of $\alg$ to infer the most favored sensitive group:
 
-	As we solve $ \Phi_{\mathsf{ER}} $, we learn that the assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, i.e.
-`male individuals with age $ < 40 $' is the most favored group with \red{probability of positive prediction} computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
-Similarly, to learn the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
-	$	\Phi_{\mathsf{ER'}} =  \exists S, \exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; \neg((\neg F \vee I \vee S) \wedge (F \vee J)).
-	$
+	\[ \Phi_{\mathsf{ER}} =  \exists S,\exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; (\neg F \vee I \vee S) \wedge (F \vee J).
+	\]
 
-	Solving $ \Phi_{\mathsf{ER'}} $, we learn the assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi_{\mathsf{ER'}}] = 0.57 $.
-Thus, `female individuals with age $ < 40 $' constitute the least favored group with \red{probability of positive prediction}:  $ 1-0.57 = 0.43$.
-	Thus, {\justicialearn} allows us to learn the most and least favored groups and the corresponding discrimination.
+	As we solve $ \Phi_{\mathsf{ER}} $, we infer that the optimal assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, which implies that `male individuals with age $ < 40 $' is the most favored group with probability of positive prediction computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
+Similarly, to infer the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
+
+	\[\Phi'_{\mathsf{ER}} =  \exists S, \exists A, \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \; \neg((\neg F \vee I \vee S) \wedge (F \vee J)).
+	\]
+
+	Solving $ \Phi'_{\mathsf{ER}} $, we learn the optimal assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi'_{\mathsf{ER}}] = 0.57 $.
+Thus, `female individuals with age $ < 40 $' constitute the least favored group with probability of positive prediction:  $ 1-0.57 = 0.43$.
+	Thus, {\justicialearn} allows us to infer the most and least favored sensitive groups and the corresponding discrimination.
 \end{example}
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
+
 We use the \red{probability of positive prediction}s of the most and least favored groups to compute different fairness metrics.
 We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.
+
 \begin{lemma}

### 2022-04-11 chapters/fairness/justicia/framework.tex
revised upto practical setting

@@
 
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2.5pt}}
-
-We use the \red{probability of positive prediction}s of the most and least favored groups to compute different fairness metrics.
 We next prove the equivalence of {\justiciaenum} and {\justicialearn} in Lemma~\ref{fairness_justicia_lm:equivalence}.
@@
 	\label{fairness_justicia_lm:equivalence}
-	Let $ \Phi_{\mathbf{a}} $ be the RE-SSAT formula for computing the \red{probability of positive prediction} of the compound sensitive group $ \mathbf{a} \in A $.
-If $ \Phi_{\mathsf{ER}} $ is the ER-SSAT formula for learning the most favored group and $ \Phi_{\mathsf{UR}} $ is the UR-SSAT formula for learning the least favored group, then
-	$\max_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   
-	and
-	$\min_{\mathbf{a}} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.   
+	Let $ \Phi_{\mathbf{a}} $ be an RE-SSAT formula for computing the probability of positive prediction of a classifier corresponding to the sensitive group $  \sensitive = \mathbf{a} $.
+Additionally, for the same classifer, let $ \Phi_{\mathsf{ER}} $ be an ER-SSAT formula for inferring the most favored sensitive group and $ \Phi_{\mathsf{UR}} $ be a UR-SSAT formula for inferring the least favored sensitive group.
+Therefore,	$\max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   	and	$\min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.\improvement{Rewrite this lemma in Appendix.}
 \end{lemma}
@@
 \subsection{Theoretical Analysis: Error Bounds}\label{fairness_justicia_sec:theory}
-We access the data generating distribution through  finite number of samples observed from it.
-These finite sample set introduce errors in the computed probabilities of the randomised quantifiers being $1$.
-These finite-sample errors in computed probabilities induce further errors in the computed positive predictive value (\red{probability of positive prediction}) and fairness metrics.
+For practical fairness problems, we do not have access to the distribution of features rather to a finite-sample dataset.
+Since the input of {\justicia} is the distribution of features, we compute the distribution from the dataset.
+The finite-sample dataset introduces errors in the computed probabilities of randomized quantified features being assigned to $1$.
+These finite-sample errors in computed probabilities induce further errors in the computed probability of positive prediction of the classifier and corresponding fairness metrics.
 We next provide a bound on this finite-sample error.
@@
 %If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.
-Thus, the true satisfying probability $p$ of $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{\bool_i \in \sigma}p_i$.
-This probability is estimated as $\hat{p}$ using $k$-samples from the data generating distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$. 
+\red{Thus, the true satisfying probability $p$ of an SSAT formula $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{l_i \in \sigma}l_i * p_i + (1 - l_i) * (1 - p_i)$, where the literal $ l_i \in \{0,1\} $ denotes either the variable $ B_i $ (when $ l_i = 1 $) or its negation (when $ l_i = 0 $).
+}This probability $ p $ is estimated as $\hat{p}$ using $k$-samples from the distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$.
 %However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
@@
 \begin{corollary}
-	If $k$ samples are considered from the data-generating distribution in {\justicia} such that 
+	If $k$ samples are considered from the  distribution in {\justicia} such that 
 	$
@@
 \end{corollary}
-This implies that given a classifier $ \alg \triangleq \Pr(\hat{Y}|X, A) $ represented as a CNF formula and a data-generating distribution $ (X,A,Y) \sim \mathcal{D} $, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
+This implies that given a binary classifier represented as a CNF formula and a distribution of sensitive and non-sensitive features, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
 Thus, {\justicia} is a sound framework of fairness verification with high probability.

### 2022-04-11 chapters/fairness/justicia/practical.tex
revised upto practical setting

@@
 \label{sec:practical-setting}
-We now relax the assumptions on access to Boolean classifiers and Boolean attributes, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous attributes.
+We now relax the assumptions of {\justicia} on access to Boolean classifiers and Boolean features, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous features.
 
-\subsubsection{Extending to Decision Trees and Linear Classifiers. }
+\subsubsection{Extending {\justicia} to Decision Trees and Linear Classifiers. }
 In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
@@
 \textit{Binary decision trees} are trivially encoded as  CNF formulas.
-In the binary decision tree, each node in the tree is a literal.
+In the binary decision tree, each node in the tree is considered as a literal.
 A \textit{path from the root to the leaf} is a conjunction of literals and thus, a \textit{clause}.
@@
 
-\textit{Linear classifiers on Boolean attributes} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}.
-We consider a linear classifier  $ W^T X + b \ge 0 $ on Boolean attributes $ X $ with weights $ W \in \mathbb{R}^{|X|} $ and bias $ b \in \mathbb{R} $.
-We first normalize $W$ and $b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint.
-We then apply  pseudo-Boolean constraints to CNF translation to encode the decision boundary to CNF.
+\textit{Linear classifiers on Boolean features} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}.
+We consider a linear classifier  $ \mathbf{W}^T \mathbf{X} + b \ge 0 $ on Boolean features $ \mathbf{X} $ with weights $ \mathbf{W} \in \mathbb{R}^{|\mathbf{X}|} $ and bias $ b \in \mathbb{R} $.
+We first normalize $ \mathbf{W} $ and $ b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint~\cite{roussel2009pseudo}.
+We then apply  pseudo-Boolean constraints to CNF translation~\cite{philipp2015pblib} to encode the decision boundary to CNF.
 This encoding usually introduces additional Boolean variables and results in large CNF.
-In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider attributes with higher weights only.
-For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in W $, we can set $ w_i = 0 $.
-Thus, the attributes with lower weights and thus, less importance do not appear in the encoded CNF.
+In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider features with higher weights only.
+For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in \mathbf{W} $, we can set $ w_i = 0 $.
+Thus, features with lower weights (less important) do not appear in the encoded CNF.
 Moreover, all introduced variables in this CNF translation are given existential ($ \exists $) quantification and they appear in the inner-most position in the prefix of the SSAT formula.
@@
 
-\subsubsection{Extending to Continuous Attributes.}
-In practical problems, attributes are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of the input attributes.
-In order to perform this Boolean abstraction, each categorical attribute is one-hot encoded and each real-valued attribute is discretised into a set of Boolean attributes~\cite{LKCL2019,GMM20}.
+\subsubsection{Extending to Continuous Features.}
+In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features.
+In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.
 
-For a binary decision tree, each attribute, including the continuous ones, is compared against a constant at each internal node of the tree.
-We fix a Boolean variable for each internal node, where the Boolean assignment to the variable decides one of the two branches to choose from the current node.
+For a binary decision tree, each feature, including the continuous ones, is compared against a constant at each node (except leaves) of the tree.
+We assign a Boolean variable to each node (except leaves), where the $ \{0,1\} $ assignment to the variable decides one of the two branches to choose from the current node.
 
-Linear classifiers are generally trained on continuous attributes, where we apply the following discretization. 
-Let us consider a continuous attribute $X_c$, where $w$ is its weight during training. 
-We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean attributes and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $. 
+Linear classifiers are generally trained on continuous features, where we apply discretization in the following way.
+Let us consider a continuous feature $X_c$, where $w$ is its weight during training.
+We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean features and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $.
 For the discretization of $X_c$, we consider the interval-based approach\footnote{Our implementation is agnostic to any discretization technique.}.
-For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned TRUE when the attribute-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned FALSE otherwise.
-Following that, we assign the weight of $ B_i $ to be $ \mu_i\times w $, when $ \mu_i $ is the mean of the $i^{\mathrm{th}}$ interval and  $ B_i $ is TRUE. 
+For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned $ 1 $ when the feature-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned $ 0 $ otherwise.
+Following that, we assign the weight of $ B_i $ to be $ \mu_iw $, where $ \mu_i $ is the mean of feature values in the $i^{\mathrm{th}}$ interval.
 We can show that if we consider infinite number of intervals, $ X_c \approx \sum_i \mu_i B_i $.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-04-11 chapters/fairness/justicia/experiments.tex
working on experiments

@@
 For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art distributional verifiers FairSquare and VeriFair, and also a sample-based fairness measuring tool: AIF360.
-In the experiments, we have studied three type of classifiers: CNF learner, decision trees and logistic regression classifier.
-Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner IMLI of~\cite{ghosh19incremental}.
+In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner.
+Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{ghosh19incremental}.
 We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula.
-We have also verified two fairness-enhancing algorithms: reweighing algorithm~\cite{kamiran2012data} and the optimized pre-processing  algorithm~\cite{calmon2017optimized}. 
-We have experimented on multiple datasets containing multiple protected attributes: the UCI\footnote{\url{ http://archive.ics.uci.edu/ml}} Adult and German-credit dataset,  ProPublica’s COMPAS recidivism dataset~\cite{angwin2016machine}, Ricci dataset~\cite{mcginley2010ricci}, and Titanic dataset\footnote{\url{https://www.kaggle.com/c/titanic}}.
+In our experiments, we have verified two fairness-enhancing algorithms: reweighing algorithm~\cite{kamiran2012data} and optimized pre-processing  algorithm~\cite{calmon2017optimized}.
+We have experimented on multiple datasets containing multiple sensitive features: the UCI\footnote{\url{ http://archive.ics.uci.edu/ml}} Adult and German-credit dataset,  ProPublica’s COMPAS recidivism dataset~\cite{angwin2016machine}, Ricci dataset~\cite{mcginley2010ricci}, and Titanic dataset\footnote{\url{https://www.kaggle.com/c/titanic}}.
 \iffalse 
-Since both {\justicia}  and FairSquare take a  probability distribution of the attributes as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution.
+Since both {\justicia}  and FairSquare take a  probability distribution of the features as input, we perform five-fold cross validation, use the train set for learning the classifier, compute distribution on the test set and finally verify fairness metrics such as disparate impact and statistical parity difference on the distribution.
 \fi
@@
 
-Our empirical studies have the following objectives:
+Our empirical studies have following objectives:
 
 \begin{enumerate}
-	\item How accurate and scalable {\justicia} is with respect to existing fairness verifiers, FairSquare and VeriFair?
+	\item How accurate and scalable {\justicia} is with respect to existing fairness verifiers: FairSquare and VeriFair?
 	\item Can {\justicia} verify the effectiveness of different fairness-enhancing algorithms on different datasets?
@@
 
-Our experimental studies validate that {\justicia} is more accurate and scalable than the state-of-the-art verifiers FairSquare and VeriFair.
-{\justicia} is able to verify the effectiveness of different fairness-enhancing algorithms for multiple fairness metrics, and datasets.
+Our experimental studies validate that {\justicia} is more accurate and scalable than the state-of-the-art verifiers: FairSquare and VeriFair.
+{\justicia} is able to verify the effectiveness of different fairness-enhancing algorithms for multiple fairness metrics and datasets.
 {\justicia} achieves scalable performance in the presence of compound sensitive groups that the existing verifiers cannot handle.
@@
 
-\input{chapters/fairness/justicia/table_synthetic.tex}
-\input{chapters/fairness/justicia/table_scalability.tex}
-\input{chapters/fairness/justicia/table_fairness_algorithm.tex}
-
-%\red{Add lines related to research question 5.}
-
-%\subsection{Experimental Results}
-%We now discuss our empirical findings.
-\begin{figure*}[t!]
-		\begin{minipage}{0.48\textwidth}
-		\centering
-		\begin{minipage}{0.48\columnwidth}
-			\includegraphics[scale=.15]{figures/fairness/justicia/sensitive_attribute_race_di_Adult_DT_RE.pdf}
-		\end{minipage}
-		\begin{minipage}{0.48\columnwidth}
-			\includegraphics[scale=.15]{figures/fairness/justicia/sensitive_attribute_race_spd_Adult_DT_RE.pdf}
-		\end{minipage}
-%		\begin{minipage}{0.32\columnwidth}
-%			\includegraphics[scale=.14]{figures/fairness/justicia/sensitive_attribute_race_eqo_Adult_DT_RE.pdf}
-%		\end{minipage}
-		\caption{Fairness metrics measured by {\justicia} for different protected groups in the Adult dataset. The number within parenthesis in the xticks denotes total compound groups.}
-		\label{fairness_justicia_fig:protected_groups}
-	\end{minipage}
-	\hfill
-	\begin{minipage}{0.48\textwidth}
-		\includegraphics[scale=.15]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}
-		\hfill
-		\includegraphics[scale=.15]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}
-		\caption{Standard deviation in estimation of disparate impact (DI) and stat. parity (SP)  for different sample sizes. {\justicia} is more robust with variation of sample size than  AIF360. }
-		\label{fairness_justicia_fig:sample-size}
-	\end{minipage}
-\end{figure*}
-
 \subsection{Experimental Analysis}
 
-\begin{comment}
-\begin{figure}[t]
-\centering
-\includegraphics[scale=.32]{figures/fairness/justicia/sanity_DI_DT.pdf}
-\caption{Results of different verifiers on  a synthetic problem.
-}
-\label{fairness_justicia_fig:synthetic}
-\end{figure}
-\end{comment}
+\input{chapters/fairness/justicia/table_synthetic.tex}
 
 %\subsubsection{Performance of Different Verifiers.}
-\subsubsection{Accuracy: Less Than $  1\%$-error.} 
+\subsubsection{Accuracy: Less Than $  1\%$-error} 
 In order to assess the accuracy of different verifiers, we have considered the decision tree in Figure~\ref{fairness_justicia_fig:fair_example} for which the fairness metrics  are analytically computable.
@@
 
-\iffalse
-\begin{figure}[t]
-	\begin{center}
-		\subfloat{
-			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_DT.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_LR.pdf}
-		} \hfill
-		%		\subfloat{
-		%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_DT.pdf}
-		%		}
-		%		\subfloat{
-		%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_LR.pdf}
-		%		} \hfill
-		\subfloat{
-			\includegraphics[scale=.32]{figures/fairness/justicia/time_ratio_FS_DT.pdf}
-		}
-		\subfloat{
-			\includegraphics[scale=.32]{figures/fairness/justicia/time_ratio_FS_LR.pdf}
-		} \hfill
+\input{chapters/fairness/justicia/table_scalability.tex}
 
-		\caption{Comparison of {\justicia} (RE(cor) encoding) with FairSquare (FS).
-The top two figures show measured disparate impact and the bottom two figures show ratio of runtime of the two verifiers for decision tree (DT) and logistic regression (LR).}
-		\label{fairness_justicia_fig:fairsquare}
-	\end{center}
-\end{figure}
-\fi
-
-\subsubsection{Scalability: $ 1 $ to $ 3 $ Orders of Magnitude Speed-up.} 
+\subsubsection{Scalability: $ 1 $ to $ 3 $ Orders of Magnitude Speed-up} 
 %Since {\justicia} appears to be more accurate than FairSquare and VeriFair in the synthetic benchmark, 
@@
 
-\begin{comment}
-We now discuss the performance of  {\justicia} compared against FairSquare in practical benchmarks.
-In general, we have observed that FairSquare cannot  scale to benchmarks with large number of attributes.
-Hence we show result on smaller benchmarks e.g.,  Ricci and Titanic dataset.
-In this context, we emphasize that FairSquare only handles single binary protected attribute, whereas {\justicia} does not have such a restriction.
-In Figure~\ref{fairness_justicia_fig:fairsquare}, we see that Both {\justicia} (RE(cor) encoding) and FairSquare (FS) certify higher disparate impact for Ricci dataset and lower disparate impact for  Titanic dataset on decision tree classifier (DT).
-When we consider logistic regression (LR), both {\justicia} and FairSquare certify that disparate impact decreases for Ricci dataset.
-In this context, FairSquare cannot scale to Titanic dataset for the logistic regression classifier within the cut off time of 400 seconds.
-Moving focus on the  runtime, we have observed that FairSquare takes one to two magnitude higher time for decision tree  and two to three magnitude higher time for logistic regression, thereby demonstrating the poor scalability of FairSquare in verifying fairness.
-\end{comment}
+\begin{figure*}
+	\centering
+	\subfloat{
+		\includegraphics[scale=.2]{figures/fairness/justicia/sensitive_attribute_race_di_Adult_DT_RE.pdf}
+	}
+	\subfloat{
+		\includegraphics[scale=.2]{figures/fairness/justicia/sensitive_attribute_race_spd_Adult_DT_RE.pdf}
+	}
+	\caption{Fairness metrics measured by {\justicia} for different sensitive groups in the Adult dataset. The number within parenthesis in the xticks denotes total compound groups.}
+	\label{fairness_justicia_fig:sensitive_groups}
+\end{figure*}
 
 \subsubsection{Verification: Detecting Compounded Discrimination in Protected Groups.}
-We have tested {\justicia} for datasets consisting of multiple protected attributes and reported the results in Figure~\ref{fairness_justicia_fig:protected_groups}.
-{\justicia} operates on datasets with even 40 compound protected groups and can potentially scale more than that while the state-of-the-art fairness verifiers (e.g., FairSquare and VeriFair) consider a single protected attribute.
+We have tested {\justicia} for datasets consisting of multiple sensitive features and reported the results in Figure~\ref{fairness_justicia_fig:sensitive_groups}.
+{\justicia} operates on datasets with even 40 compound sensitive groups and can potentially scale more than that while the state-of-the-art fairness verifiers (e.g., FairSquare and VeriFair) consider a single sensitive feature.
 Thus, {\justicia} removes an important limitation in practical fairness verification. 
-Additionally, we observe in most datasets the disparate impact decreases and thus, discrimination increases as more compound protected groups are considered.
+Additionally, we observe in most datasets the disparate impact decreases and thus, discrimination increases as more compound sensitive groups are considered.
 For instance, when we increase the total  groups from $ 5 $ to $ 40 $ in the Adult dataset, disparate impact decreases from around $ 0.9 $ to $ 0.3 $, thereby detecting higher discrimination.
 Thus, {\justicia} detects that the marginalized individuals of a specific type (e.g., `race')  are even more discriminated and marginalized when they also belong to a marginalized group of another type (e.g., `sex').
+
+\input{chapters/fairness/justicia/table_fairness_algorithm.tex}
 
@@
 Both of them pre-process to remove statistical bias from the dataset. 
-We study the effectiveness of these algorithms using {\justicia} on three datasets each with two different protected attributes.  
+We study the effectiveness of these algorithms using {\justicia} on three datasets each with two different sensitive features.  
 In Table~\ref{fairness_justicia_tab:fair_algo_verification}, we report different fairness metrics on logistic regression and decision tree.
 We observe that {\justicia} verifies fairness improvement as the bias mitigating algorithms are applied.
-For example, for the Adult dataset with `race' as the protected attribute, disparate impact increases from $ 0.23 $ to $ 0.85 $ for applying the reweighing algorithm on logistic regression classifier.
+For example, for the Adult dataset with `race' as the sensitive feature, disparate impact increases from $ 0.23 $ to $ 0.85 $ for applying the reweighing algorithm on logistic regression classifier.
 In addition, statistical parity decreases from $ 0.09 $ to $ 0.01 $, and equalized odds decreases from $ 0.13 $ to $ 0.03 $, thereby showing the effectiveness of reweighing algorithm in all three fairness metrics.
@@
 
-\iffalse
-\begin{figure}[t]
-	\begin{center}
-		\begin{minipage}{.32\columnwidth}
-			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_di_Adult_DT_race__sex.pdf}
-		\end{minipage}
-		\begin{minipage}{.32\columnwidth}
-			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_spd_Adult_DT_race__sex.pdf}
-		\end{minipage}
-		\begin{minipage}{.32\columnwidth}
-			\includegraphics[scale=.22]{figures/fairness/justicia/equivalence_eqo_Adult_DT_race__sex.pdf}
-		\end{minipage}
-		%		\subfloat{
-		%			\includegraphics[scale=.32]{figures/fairness/justicia/equivalence_di_adult_lr_sex.pdf}
-		%		}
-		%		\subfloat{
-		%			\includegraphics[scale=.32]{figures/fairness/justicia/equivalence_spd_adult_lr_sex.pdf}
-		%		} \hfil
-		\caption{Fairness metrics measured by different encodings.
-Both RE and ER encodings result in same fairness value while RE(cor) encoding usually shows lower fairness value for considering correlation of the protected attributes.\red{Can be removed}}
-		\label{fairness_justicia_fig:equivalence}
-	\end{center}
-\end{figure}
-\fi
-
-\iffalse
-\begin{figure}
-	\subfloat{
-		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Adult_op_LR_sex.pdf}
+\begin{figure*}
+	\centering
+	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}}
+	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}		\caption{Standard deviation in estimation of disparate impact (DI) and stat.
+parity (SP)  for different sample sizes.
+{\justicia} is more robust with variation of sample size than  AIF360.
 }
-	\subfloat{
-		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Adult_rw_LR_sex.pdf}
-	}\\
-	\subfloat{
-		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Compas_op_LR_sex.pdf}
-	}
-	\subfloat{
-		\includegraphics[scale=.32]{figures/fairness/justicia/DI_Compas_rw_LR_sex.pdf}
-	}
-	\caption{Fairness metrics for applying optimized pre-processing (op) and reweighing (rw) algorithm on the Adult dataset for the protected attribute sex.
-As we apply these algorithms, fairness is usually enhanced certified by SSAT based verifiers {\justicia} (e.g.,  RE(cor) and RE encoding) and the statistical tool AIF360 (AIF).
-In each pair of columns, the first and second column refer to before and after applying the algorithm, respectively.}
-	\label{fairness_justicia_fig:fairness_algorithm_aif}
-\end{figure}
-\fi
-
-%\begin{figure}
-%	\begin{center}
-%		\subfloat{
-%			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_DT.pdf}
-%		} 
-%		\subfloat{
-%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_DT.pdf}
-%		} \hfil
-%		\subfloat{
-%			\includegraphics[scale=.32]{figures/fairness/justicia/DI_FS_LR.pdf}
-%		} 
-%		\subfloat{
-%			\includegraphics[scale=.32]{figures/fairness/justicia/time_FS_LR.pdf}
-%		} \hfil
-%			
-%		\caption{Comparison with FairSquare.}
-%	\end{center}
-%\end{figure}
-
-%\begin{figure}
-%	\begin{center}
-%		\subfloat{
-%			\includegraphics[scale=.32]{figures/fairness/justicia/threshold_adult_lr_RE(cor)_age.pdf}
-%		} 
-%			\subfloat{
-%		\includegraphics[scale=.32]{figures/fairness/justicia/threshold_ricci_lr_RE(cor)_Race.pdf}
-%		} \hfil
-%	
-%		\caption{Effect of thresholding.}
-%	\end{center}
-%\end{figure}
+	\label{fairness_justicia_fig:sample-size}
+\end{figure*}
 
@@
 	\begin{center}
-		\includegraphics[scale=.32]{figures/fairness/justicia/encoding_runtime_Adult_DT.pdf}
-		\includegraphics[scale=.32]{figures/fairness/justicia/encoding_runtime_Adult_IMLI.pdf}
+		\includegraphics[scale=.4]{figures/fairness/justicia/encoding_runtime_Adult_DT.pdf}
+		\includegraphics[scale=.4]{figures/fairness/justicia/encoding_runtime_Adult_IMLI.pdf}
 		\hfill
-		\caption{Runtime comparison of different encodings while varying total protected groups in the Adult dataset.}
+		\caption{Runtime comparison of different encodings while varying total sensitive groups in the Adult dataset. \red{Label `learn' to be `infer'???}}
 		\label{fairness_justicia_fig:runtime_diff_encodings}
@@
 {\justiciacond} (conditional probabilities w.r.t.
-protected groups) also has an exponential trend in runtime similar to {\justiciaenum}.
-This analysis justifies that the na\"ive enumeration-based approach cannot verify large-scale fairness problems containing multiple protected attributes, and {\justicialearn} is a more efficient approach for practical use.
+sensitive groups) also has an exponential trend in runtime similar to {\justiciaenum}.
+This analysis justifies that the na\"ive enumeration-based approach cannot verify large-scale fairness problems containing multiple sensitive features, and {\justicialearn} is a more efficient approach for practical use.
 

### 2022-04-11 chapters/fairness/justicia/discussion.tex
justicia revised

@@
 \section{Discussion and Future Work}
-Though formal verification of different fairness metrics of an ML algorithm for different datasets is an important question, existing verifiers are not scalable, accurate, and extendable to non-Boolean protected attributes.
-We propose a stochastic SAT-based approach, {\justicia}, that formally verifies independence and separation metrics of fairness for different classifiers and distributions for compound protected groups.
+Though formal verification of different fairness metrics of an ML algorithm for different datasets is an important question, existing verifiers are not scalable, accurate, and extendable to non-Boolean sensitive features.
+We propose a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups.
 Experimental evaluations demonstrate that {\justicia} achieves \textit{higher accuracy} and \textit{scalability} in comparison to the state-of-the-art verifiers, FairSquare and VeriFair, while yielding \textit{higher robustness} than the sample-based tools, such as AIF360.
 
-Our work opens up several new directions of research.
+\red{Our work opens up several new directions of research.
 One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs.
-Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
+Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.}
 %Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2022-04-11 chapters/fairness/justicia/experiments.tex
justicia revised

@@
 In Table~\ref{fairness_justicia_tab:synthetic}, we show the computed fairness metrics by {\justicia}, FairSquare, VeriFair, and AIF360.
-We observe that {\justicia} and AIF360  yield more accurate estimates of DI and SP compared against the ground truth with less than $1\%$ error.
-FairSquare and VeriFair  estimate the disparate impact to be $0.99$ and thus, being unable to verify the fairness violation. 
-Thus, {\justicia} is significantly accurate than the existing formal verifiers: FairSquare and VeriFair. 
+We observe that {\justicia} and AIF360  yield more accurate estimates of DI and SP compared against the ground truth values of fairness metrics with less than $1\%$ error.
+In contrast, FairSquare and VeriFair  estimate DI to be $0.99$ and thus, being unable to verify the fairness violation.
+Therefore, {\justicia} is significantly more accurate than the existing formal verifiers: FairSquare and VeriFair.
 %First we observe that both RE and ER encoding result in the same disparate impact, thereby showing the equivalence between the two encodings.
@@
 \subsubsection{Verification: Detecting Compounded Discrimination in Protected Groups.}
-We have tested {\justicia} for datasets consisting of multiple sensitive features and reported the results in Figure~\ref{fairness_justicia_fig:sensitive_groups}.
-{\justicia} operates on datasets with even 40 compound sensitive groups and can potentially scale more than that while the state-of-the-art fairness verifiers (e.g., FairSquare and VeriFair) consider a single sensitive feature.
-Thus, {\justicia} removes an important limitation in practical fairness verification. 
-Additionally, we observe in most datasets the disparate impact decreases and thus, discrimination increases as more compound sensitive groups are considered.
+We have tested {\justicia} for datasets consisting of multiple sensitive features and reported results in Figure~\ref{fairness_justicia_fig:sensitive_groups}.
+{\justicia} operates on datasets with even 40 compound sensitive groups and can potentially scale more than that while the state-of-the-art fairness verifiers (e.g., FairSquare and VeriFair) consider a single sensitive feature with two sensitive groups.
+Thus, {\justicia} removes an important limitation in practical fairness verification, which was previously restricted to Boolean sensitive groups.
+Additionally, in most datasets, we observe that disparate impact decreases and thus, discrimination increases as more compound sensitive groups are considered.
 For instance, when we increase the total  groups from $ 5 $ to $ 40 $ in the Adult dataset, disparate impact decreases from around $ 0.9 $ to $ 0.3 $, thereby detecting higher discrimination.
@@
 
-\subsubsection{Verification: Fairness of Algorithms on Datasets.}
-We have experimented with two fairness-enhancing algorithms: the reweighing (RW) algorithm and the optimized-preprocessing (OP) algorithm.
+\subsubsection{Verification: Fairness of Algorithms on Datasets}
+We have experimented with two fairness-enhancing algorithms: reweighing (RW) algorithm and optimized-preprocessing (OP) algorithm.
 Both of them pre-process to remove statistical bias from the dataset.
@@
 {\justicia} also finds instances where the fairness algorithms fail, specially when considering the decision tree classifier. 
-Thus, {\justicia} enables verification of different fairness enhancing algorithms in literature.
+Thus, {\justicia} verifies the effectiveness of different fairness enhancing algorithms.
 
@@
 	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}		\caption{Standard deviation in estimation of disparate impact (DI) and stat.
-parity (SP)  for different sample sizes.
+parity (SP)  for different sample sizes (sample size $ = 1 $ refers to the entire dataset).
 {\justicia} is more robust with variation of sample size than  AIF360.
@@
 While both {\justiciaenum} and {\justicialearn}  have the same output according to Lemma~\ref{fairness_justicia_lm:equivalence},  {\justicialearn} encoding  improves exponentially  in runtime  than {\justiciaenum} encoding on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fairness_justicia_fig:runtime_diff_encodings}.
-{\justiciacond} (conditional probabilities w.r.t.
+{\justiciacond} ({\justiciaenum} encoding where we consider conditional probabilities of non-sensitive features w.r.t.
 sensitive groups) also has an exponential trend in runtime similar to {\justiciaenum}.

### 2022-04-12 chapters/interpretability/relaxed-cnf/abstract.tex
workig on relaxed-cnf

@@
 \section{abstract}
-	 Machine learning algorithms that produce  rule-based predictions   in Conjunctive Normal form (CNF) or in Disjunctive Normal form (DNF) are arguably some of the most interpretable ones.
-For example, decision set is an interpretable model in practice, that represents the decision function in the form of DNF.
-In this paper, we consider relaxed definitions of standard OR/AND operators which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
-Building on these relaxed definition,  we introduce \textit{relaxed-}CNF rules, which are motivated by the popular usage of checklists in the medical domain and generalizes the widely employed rule representations including CNF, DNF, and decision sets.
-While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the naive learning techniques are computationally expensive.
+	 Machine learning classifiers that produce  rule-based predictions in Boolean representations such as in Conjunctive Normal form (CNF) or in Disjunctive Normal form (DNF) are arguably some of the most interpretable ones.
+For example, decision set is a popular interpretable model, which represents the decision function in the form of DNF.
+In this paper, we consider relaxed definitions of standard OR/AND operators in CNF/DNF, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
+Building on these relaxed definition,  we introduce \textit{relaxed-}CNF rules, which are motivated by the popular usage of checklists in the medical domain and generalize the widely employed rule representations including CNF, DNF, and decision sets.
+While the combinatorial structure of relaxed-CNF rules offers an exponential succinctness, the na\"ive learning techniques are computationally expensive.
 	 To this end, we  propose a novel incremental mini-batch learning procedure, called {\crr}, that employs  advances in the  Mixed-Integer Linear Programming (MILP) solvers to efficiently learn relaxed-CNF rules.
-	 Our  experimental analysis demonstrates that {\crr} can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
+Our  experimental analysis demonstrates that {\crr} can generate more accurate and sparser relaxed-CNF rules compared to the alternative rule-based classifiers.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-04-21 chapters/interpretability/relaxed-cnf/abstract.tex
CRR intro revised

@@
 For example, decision set is a popular interpretable model, which represents the decision function in the form of DNF.
-In this paper, we consider relaxed definitions of standard OR/AND operators in CNF/DNF, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
-Building on these relaxed definition,  we introduce \textit{relaxed-}CNF rules, which are motivated by the popular usage of checklists in the medical domain and generalize the widely employed rule representations including CNF, DNF, and decision sets.
+In this chapter, we consider relaxed definitions of standard OR/AND operators in CNF/DNF, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
+Building on these relaxed definitions,  we introduce \textit{relaxed-}CNF rules, which are motivated by the popular usage of checklists in the medical domain and generalize the widely employed rule representations including CNF, DNF, and decision sets.
 While the combinatorial structure of relaxed-CNF rules offers an exponential succinctness, the na\"ive learning techniques are computationally expensive.
@@
 Our  experimental analysis demonstrates that {\crr} can generate more accurate and sparser relaxed-CNF rules compared to the alternative rule-based classifiers.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-04-21 chapters/interpretability/relaxed-cnf/introduction.tex
CRR intro revised

@@
 The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
-To enable  safe, robust and trustworthy integration of such systems, the end users require these systems to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
-In this context,  rule-based representations are particularly effective for  presenting the decision functions to people~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
-A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability (defined in terms of rule-sparsity) with minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}.
-Although CNF/DNF rules are considered interpretable, they are less expressive compared to the Boolean cardinality constraints in propositional logic, wherein a Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
-In this work, we introduce relaxed-CNF,  a classification rule in propositional logic that has the benefits of  both worlds: it represents the decision boundary in an interpretable manner similar to CNF/DNF  and it is more expressive  for allowing Boolean cardinality constraints in its representation.
+To enable  safe, robust, and trustworthy integration of such systems, the end users require them to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
+In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
+A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}.
+Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
+A Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
+In this work, we introduce a novel formulation of interpretable classification rules, namely \emph{relaxed-CNF}, which achieve benefits of  both worlds: it is interpretable similar to CNF/DNF  but more expressible by allowing cardinality constraints in the representation.
 
- We find the motivation of designing relaxed-CNF rules from checklists, where a checklist is a list of conditions one needs to check, e.g., a list of items to take on a travel trip.
-There are several practical applications of checklists in interpretable decision making~\cite{M1976,gage2001validation}.
-For example, the  CHADS2 score in medicine~\cite{gage2001validation}, that is a clinical prediction rule for estimating the risk of stroke.
-Another example is a psychometric test, known as Myers–Briggs Type Indicator (MBTI)~\cite{M1976}, which indicates differing psychological preferences in how people
-perceive the world around them and make decisions.
-An influential study on the importance of {checklists}~\cite{G2010} finds that highly complex and specialized problems can be handled smoothly by the development and consistent usage of checklists.
-In this paper, we aim to learn  classification rules in the form of a checklist, which we call relaxed-CNF rules.
+We find the motivation of relaxed-CNF rules from checklists.
+A checklist is a list of conditions that one needs to check, e.g., a list of items to take on a travel trip.
+Checklists have several applications in interpretable decision making~\cite{M1976,gage2001validation}, particularly in the medical domain.
+For example, the  CHADS2 score in medicine is a clinical prediction rule for estimating the risk of stroke~\cite{gage2001validation}.
+Another example of checklists is a psychometric test, known as Myers–Briggs Type Indicator (MBTI)~\cite{M1976}, which indicates differing psychological preferences in how people perceive the world around them and make decisions.
+An influential study on the importance of {checklists}~\cite{G2010} finds that highly complex and specialized problems can be handled smoothly by the development and consistent usage of checklists, which we formally call as relaxed-CNF rules.
 
 We now provide an introduction to relaxed-CNF rules.
-The simplest logical rules are single level-rules: ORs or ANDs of a subset of \textit{literals}, where a literal can be assigned  $ 1 $ (true) or  $ 0 $ (false).
-An OR operator requires $ 1 $ out of $ N $ literals to be assigned to  $ 1 $, while an AND operator requires all $N$ out of $N$ literals to be assigned to  $ 1 $.
-A \emph{clause} is a collection of literals connected by  OR/AND.
-A CNF   formula  is a conjunction (AND) of {clauses} where each clause is a disjunction (OR) of {literals}, while  a DNF  formula is a disjunction of clauses where each clause is a conjunction of literals.
-Therefore, CNF  and DNF  formulas can be viewed as two-level rules.
-Some popular interpretable rule-based models can be viewed as  CNF/DNF formulas.
-For example, a decision set is a DNF rule that refers to a set of ``if-else'' clauses~\cite{ignatiev2018sat,lakkaraju2016interpretable}.
-In this paper, we consider a richer set of logical formulas that capture the structure of checklists.
-We consider \emph{hard}-OR clauses, where at least $ M > 1$ out of $ N $ literals need to be $ 1 $, and we similarly define {\emph{soft}-AND} clauses which allow some of the literals (at most $ N-M $) to be $ 0 $.
-To be precise,  the definitions of hard-OR and soft-AND overlap. We use the term hard-OR when $M \le N/2$ and soft-AND otherwise.     
+In interpretable rule-based classification, the simplest logical rules are single level-rules: ORs or ANDs of Boolean \textit{literals}, where each literal denotes either a Boolean input feature or its negation.
+A \emph{clause} is a collection of $ N $ literals connected by  OR/AND.
+To satisfy a clause, an OR operator requires $ 1 $ out of $ N $ literals to be assigned to  $ 1 $ in the clause, while an AND operator requires all $N$ out of $N$ literals to be assigned to  $ 1 $.
+A CNF   formula  is a conjunction (AND) of {clauses} where each clause is a disjunction (OR) of {literals}, and  a DNF  formula is a disjunction of clauses where each clause is a conjunction of literals.
+Therefore, CNF  and DNF  formulas can be viewed as two-level rules with several applications in interpretable decision-making.
+For example, a decision set is a DNF rule referring to a set of ``if-else'' conditions~\cite{ignatiev2018sat,lakkaraju2016interpretable}.
+In this chapter, we consider a richer set of logical formulas that capture the structure of checklists.
+To this end, we consider \emph{hard}-OR clauses, where at least $ M > 1$ out of $ N $ literals are assigned to $ 1 $, and we similarly define {\emph{soft}-AND} clauses which allow some of the literals (at most $ N-M $) to be $ 0 $.
+To be precise,  the definitions of hard-OR and soft-AND overlap.
+Consequently, we use hard-OR when $M \le N/2$ and soft-AND otherwise.
 To extend the standard definition of CNF (which is ANDs of ORs),  we define \emph{relaxed}-CNF to denote soft-ANDs of hard-ORs.
 Similarly, \emph{relaxed}-DNF is hard-ORs of soft-ANDs.
-Since   hard-OR and soft-AND are differentiated based on the value of $ M $ and $ N $, relaxed-CNF and relaxed-DNF have a similar structural representation.
+Since   hard-OR and soft-AND are differentiated based on the value of $ M $ and $ N $, relaxed-CNF and relaxed-DNF have the same structural representation.
 In early work, Craven and Shavlik~\cite{craven1996extracting} considered single level $ M $-of-$ N $ rules to explain black-box neural-network classifiers.
 Recently, Emad et al.~\cite{EVM2015} have developed a semi-quantitative group testing approach for learning sparse single level $M$-of-$N$ rules, which are quite restrictive in their ability to fit the data.
-In this work, we study a much richer family of two-level relaxed-CNF rules.
+In contrast, in this work, we study a much richer family of two-level relaxed-CNF rules.
 
@@
 	\end{center}
-	\caption{An illustrative example of a  relaxed CNF classification rule that describes the decision function in the form of a  two-level checklist.
-This classifier is learned on the WDBC (Wisconsin diagnostic breast cancer) dataset and predicts whether a tumor cell is malignant or not based on the characteristics of the  tumor cell.
-The first column in each checklist contains Boolean predicate/literals.
-An entry in the second column is  $ 1 $ if the corresponding predicate is true by an observed tumor cell, and $ 0 $ otherwise.
-In the first level, checklist A (resp.
-B) is true if the count of true  predicates is at least $ 2 $ (resp.
-$ 3 $).
+	\caption{An illustrative example of a  relaxed CNF classification rule, which describes the decision function in the form of a  two-level checklist.
+This classifier is learned on the WDBC (Wisconsin diagnostic breast cancer) dataset and it predicts whether a tumor cell is malignant or not based on the characteristics of the  tumor cell.
+The first column in each checklist contains Boolean literals.
+An entry in the second column is  $ 1 $ if the corresponding literal is true by an observed tumor cell, and $ 0 $ otherwise.
+In the first level, checklist A (resp.\ B) is true if the count of true  predicates is at least $ 2 $ (resp.\ $ 3 $).
 In the second level, a tumor cell is predicted as malignant if the count of true checklists is at least $ 1 $.
@@
 
- {Relaxed-CNF rules are  more flexible than pure CNF rules and  can accurately fit more complex classification boundaries.} For example, relaxed-CNF clauses  allow  compact encoding of the majority function, which would require exponentially many clauses in CNF, showing the exponential gap in the succinctness of the two representations.
+ {Relaxed-CNF rules are  more flexible than pure CNF rules, and they can accurately fit more complex classification boundaries.} For example, relaxed-CNF clauses  allow  a compact encoding of the majority function\footnote{The majority function is a Boolean function that evaluates to $ 0 $ when half or more arguments are false, and $ 1 $ otherwise (\url{https://en.wikipedia.org/wiki/Majority_function}).} in Boolean logic, which would require exponentially many clauses in CNF, showing the exponential gap in the succinctness of the two representations.
 In addition, relaxed-CNF and CNF rules have the same functional form where a user has to compute the sum of true literals/clauses  and then compare the sum  to  different thresholds {(as in the example in Figure~\ref{interpretability_crr_exmpl:wdbc})}.
-From the computational perspective, the structural flexibility of relaxed-CNF rules compared to CNF/DNF rules makes them harder to learn, which poses the following question, \textit{can we design a combinatorial framework to efficiently learn relaxed-CNF rules?}
+From the computational perspective, the structural flexibility of relaxed-CNF compared to CNF/DNF makes it harder to learn.
+Therefore, in this chapter, we ask the following research question: \textit{can we design a combinatorial framework to efficiently learn relaxed-CNF rules?}
 
-The primary contribution of this paper is an affirmative answer to the above question by proposing an efficient combinatorial framework for learning relaxed-CNF rules which we call {\crr} (\textbf{C}lassification \textbf{R}ules in \textbf{R}elaxed form).
-In CRR, we construct an objective function to maximize both prediction accuracy and rule-sparsity and design a  Mixed-Integer Linear Programming (MILP) formulation for learning the optimal relaxed-CNF rules.
-%We theoretically prove the succinctness of relaxed-CNF rules and propose a Mixed Integer Linear Programming (MILP)-based  approach for learning optimal rules.
-To learn a $ k $-clause relaxed-CNF rule (say $ \Rule $) using the direct (naive) MILP formulation, the size of the MILP query expressed as the number of constraints is $ \mathcal{O}(n\cdot k) $, where the  number of features/attributes in the dataset is fixed and $ n $ is the number of training samples.
-Consequently, the naive MILP formulation fails to handle large datasets.
-To address scalability concerns,  we propose an efficient mini-batch training methodology based  incremental approach  for learning $ \Rule $.
-The proposed incremental approach significantly enhances the scalability of {\crr}.
-%In the incremental approach, we divide the datasets into $ \rho $ partitions and make $ \rho $ queries to ILP solvers with each query of $ \mathcal{O}(\frac{n}{\rho}\cdot k) $ constraints.
-This approach learns a relaxed-CNF rule for each partition in a  sequential order such that the ILP query constructed for the $ p $-th partition relies on the training samples in the $ p $-th partition and the rule learned for the $ (p-1)  $-th partition.
+The primary contribution of this chapter is an affirmative answer to the above question by proposing an efficient combinatorial learning framework for relaxed-CNF rules, namely {\crr} (\textbf{C}lassification \textbf{R}ules in \textbf{R}elaxed form).
+In CRR, we construct a learning objective to maximize both the prediction accuracy and the rule-sparsity of the generated classification rule.
+To this end, we design a  Mixed-Integer Linear Programming (MILP) formulation for learning the optimal relaxed-CNF rule from data.
+To learn a $ k $-clause relaxed-CNF rule (say $ \Rule $) using the na\"ive MILP formulation, the size of the MILP query expressed as the number of constraints is $ \mathcal{O}(nk) $, where $ n $ is the number of training samples in the dataset.
+Consequently, this formulation fails to handle large datasets.
+To address the scalability of {\crr},  we propose an efficient mini-batch training methodology, where we incrementally learn $ \Rule $ from data by iteratively solving smaller MILP queries corresponding to batches.
 
 {Through a comprehensive experimental evaluation over datasets from the UCI and Kaggle repository, we observe that {\crr} with relaxed-CNF rules achieves an improved trade-off between accuracy and rule sparsity and scales to datasets with more than $ 10^5 $ samples.
-More significantly, {\crr} can generate relaxed-CNF rules which have higher accuracy than CNF rules generated by~\cite{ghosh19incremental}.
-Also, compared to decision lists generated by~\cite{C1995}, relaxed-CNF rules are sparser in large datasets.}
+More significantly, {\crr} generates relaxed-CNF rules with higher accuracy than CNF rules generated by~\cite{ghosh19incremental}.
+Furthermore, compared to decision lists generated by~\cite{C1995}, relaxed-CNF rules are sparser in large datasets.}
 %can also reach the same level of accuracy compared to plain CNF/DNF rules with smaller relaxed-CNF rules. 
@@
 %KSM: Please feel free to comment out the text below if it exceeds page limit
-The rest of the paper is organized as follows.
+The rest of the chapter is organized as follows.
 We discuss related works in Section~\ref{interpretability_crr_sec:related_work}, introduce the notation and preliminaries in Section~\ref{interpretability_crr_sec:preliminaries} and formulate the problem in Section~\ref{interpretability_crr_sec:problem}.
-We describe the main contribution of this paper, $ \crr $,  in Section~\ref{interpretability_crr_sec:ilp} and extend to an incremental mini-batch learning in Section~\ref{interpretability_crr_sec:inc_ilp}.
+We describe the main contribution of this chapter, $ \crr $,  in Section~\ref{interpretability_crr_sec:ilp} and extend to an incremental mini-batch learning in Section~\ref{interpretability_crr_sec:inc_ilp}.
 We then describe the experimental results in Section ~\ref{interpretability_crr_sec:experiment} and conclude in Section~\ref{interpretability_crr_sec:conclusion}.
+
+\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-05-27 chapters/abstract.tex
updated abstract

\begin{abstract}
	\begin{center}
		Interpretability and Fairness in Machine Learning: A Formal Method Approach\\
		\vspace{1em}
		by\\
		\vspace{1em}
		Bishwamittra Ghosh\\
		\vspace{1em}
	\end{center}

	The significant success of machine learning in past decades has witnessed a host of applications of algorithmic decision-making in different safety-critical domains. The high-stake predictions of machine learning in medical, law, education, transportation, etc.\ have far-reaching consequences on the end-users. Consequently, researchers call for the regulation of machine learning by defining and improving different safety-critical properties of predictions, such as interpretability, fairness, robustness, and privacy.  In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable classification}, \emph{verifying fairness}, and \emph{identifying the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both. To alleviate these limitations, we apply formal methods in interpretable and fair machine learning and provide scalable and accurate solutions to the underlying problems.

	In interpretable machine learning, rule-based classifiers are particularly effective in representing the decision boundary using a set of rules. The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice. As such, interpretable learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets. To this end, we propose an incremental learning framework, called {\imli}, which extends the classification to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning. Built on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning classification rules represented as relaxed logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.

	Fairness in machine learning centers on quantifying and mitigating the bias/unfairness of machine learning. In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of certifying whether the classifier achieves the desired level of fairness given the distribution of features. We also extend {\justicia} by considering feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness. Finally, we propose a framework for fairness influence functions (FIF) with an aim of quantifying the influence of features as their contribution on the bias of a classifier. FIF analysis interprets fairness by revealing potential features or subset of features attributing highly to the bias. Building on global sensitivity analysis, we propose an algorithm, called {\fairxplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.

\end{abstract}

### 2022-05-27 chapters/interpretability/relaxed-cnf/related_work.tex
updated abstract

@@
 More recently, a Bayesian framework has been adopted to learn rule sets~\cite{WRDLKM2017}, rule lists \cite{LRMM2015},  and falling rule lists~\cite{WR2015}.
-Building on the connection between rule learning and Boolean compression sensing,  a rule-based classification system was proposed to trade-off classification accuracy and representation size~\cite{MV2013}.
-Su et al.~\cite{SWVM2015} proposed an extension that allows two-level Boolean rules.
-In addition to designing classifiers that produce forms amenable to end-users, there is a large body of work to describe the working of opaque models (model-agnostic interpretability)~\cite{LKCL2019,lundberg2017unified,ribeiro2016should}.
+Building on the connection between rule learning and Boolean compression sensing, Malioutov and Varshney proposed a rule-based classification framework to trade-off classification accuracy and representation size~\cite{MV2013}.
+Furthermore, Su et al.~\cite{SWVM2015} proposed an extension that allows learning two-level Boolean rules.
+In addition to designing classifiers that produce representations amenable to the end users, there is a large body of work to describe the prediction of opaque black-box models~\cite{LKCL2019,lundberg2017unified,ribeiro2016should}.
 While our work can also be applied to interpret black-box classifiers, we do not pursue this direction in this paper and leave for future work.
 
-In recent work,  Ghosh and Meel~\cite{ghosh19incremental} propose an incremental approach for efficiently learning CNF classification rules using a MaxSAT-based framework.
-They adopt a sequential partition-based training methodology in incremental learning.
-Although the said approach achieves the runtime improvement,  the generated rules highly depend on the order/position of samples in the training dataset.
-In contrast, the incremental framework proposed in this work does not suffer from strong dependencies on the order of samples in the training dataset.
+Recently,  Ghosh and Meel~\cite{ghosh19incremental} propose an incremental approach for efficiently learning CNF classification rules using a MaxSAT-based framework.
+They adopt a sequential partition-based training methodology in incremental learning to achieve scalability in learning.
+Although the said approach achieves an improvement in runtime,  the generated rules highly depend on the order of samples in the training dataset.
+In contrast, the incremental framework proposed in this work does not suffer from the strong dependencies on the order of samples in training.\change{Do we keep this related work?}

### 2022-05-27 chapters/introduction.tex
updated files
updated abstract

@@
-	\vspace{1em}
-\section{Introduction}
+\chapter{Introduction}
 The tremendous success of machine learning in recent years has found its applications in different high-stake and  safety-critical domains, such as medical, law, education, and transportation.

### 2022-05-27 chapters/list_of_papers.tex
updated abstract

%\clearpage
%\begin{center}
%	\large\textbf{List of Publications}
%	\normalsize
%\end{center}

\section*{List of Publications}
This thesis is based on the following publications.

\begin{enumerate}
	\item \href{https://arxiv.org/pdf/2205.06936.pdf}{\blue{Efficient Learning of Interpretable Classification Rules}} \\
	\underline{Bishwamittra Ghosh}, Dmitry Malioutov, Kuldeep S. Meel\\
	In Proceedings of JAIR, 2022

	\item \href{https://arxiv.org/pdf/2109.09447.pdf}{\blue{Algorithmic Fairness Verification with Graphical Models}} \\
	\underline{Bishwamittra Ghosh}, Debabrota Basu, Kuldeep S. Meel\\
	In Proceedings of AAAI, 2022

	\item \href{https://arxiv.org/pdf/2009.06516.pdf}{\blue{Justicia: A Stochastic SAT Approach to Formally Verify Fairness}} \\
	\underline{Bishwamittra Ghosh}, Debabrota Basu, Kuldeep S. Meel\\
	In Proceedings of AAAI, 2021

	\item \href{https://bishwamittra.github.io/publication/ecai_2020/paper.pdf}{\blue{Classification Rules in Relaxed Logical Form
	}} \\
	\underline{Bishwamittra Ghosh}, Dmitry Malioutov, Kuldeep S. Meel\\
	In Proceedings of ECAI, 2020

	\item \href{https://bishwamittra.github.io/publication/imli-ghosh.pdf}{\blue{IMLI: An Incremental Framework for MaxSAT-Based Learning of Interpretable Classification Rules}}\\
	\underline{Bishwamittra Ghosh}, Kuldeep S. Meel\\
	In Proceedings of AIES, 2019

\end{enumerate}
\clearpage

### 2022-05-30 chapters/abstract.tex
intro started

@@
 The high-stake predictions of machine learning in medical, law, education, transportation, etc.\ have far-reaching consequences on the end-users.
-Consequently, researchers call for the regulation of machine learning by defining and improving different safety-critical properties of predictions, such as interpretability, fairness, robustness, and privacy.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable classification}, \emph{verifying fairness}, and \emph{identifying the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
+Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{identifying the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
 To alleviate these limitations, we apply formal methods in interpretable and fair machine learning and provide scalable and accurate solutions to the underlying problems.
@@
 To this end, we propose an incremental learning framework, called {\imli}, which extends the classification to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
-Built on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning classification rules represented as relaxed logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.
+Built on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning a more expressible classification rule by relaxing logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.
 

### 2022-05-30 chapters/introduction.tex
intro started

@@
 \chapter{Introduction}
+
+The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical, law, education, and transportation.
+In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users.
+With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability, fairness, robustness, and privacy.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning.
+The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness.
+
+Towards responsible machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers.
+\emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features.
+Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
+Thus, we propose efficient algorithms for learning interpretable rule-based classifiers.
+In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
+Consequently, fairness research centers on quantifying fairness using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
+In this thesis, we study fairness verification problems to assess the fairness claim of different classifiers under the lens of multiple fairness definitions and algorithms.
+In addition, we study the sources of the unfairness of classifiers by identifying a subset of features attributing highly to bias.
+In both interpretable and fair machine learning, we prioritize improving both the \emph{scalability} and the \emph{accuracy} of solutions, either or both of which prior approaches fail to achieve.
+
+The contributions of the thesis are discussed in the following.
+
+\section{Interpretable Machine Learning}
+The problem in interpretable machine learning is to learn a classifier making predictions interpretable to the human.
+To achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
+At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging.
+In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
+
+In this thesis, we study two interpretable rule-based classifiers, characterized by their expressiveness.
+At first, we study classifiers represented as formulas in propositional logic.
+In propositional logic, Conjunctive Formal Form (CNF) and Disjunctive Normal Form (DNF) are useful representations of Boolean formulas.
+Popular interpretable rule-based classifiers such as decision tree, decision lists, and decision sets share the logical structure of CNF/DNF in their representation of the decision function.
+Thereby, we propose to learn a CNF classifier, wherein its interpretability is defined by the number of Boolean literals that the formula contains.
+Compared to CNF, Boolean cardinality constraints are more expressive as they allow numerical bounds on Boolean literals~\cite{sinz2005towards}.
+Relying on the concept of cardinality constraint to increase expressiveness, our second interpretable rule-based classifier is a logical relaxation of CNF/DNF classifiers, namely \emph{relaxed-}CNF.
+
+The problem of learning rule-based classifiers is known to be computationally intractable.
+The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}.
+In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc.
+In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}.
+Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
+\red{Now comes incremental learning}
+
+In our study, we propose an alternate optimization approach that fits particularly well to rule-learning problems.
+Particularly,  we propose a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.
+
+In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean predicates that the formula contains.
+
+The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
+To enable  safe, robust, and trustworthy integration of such systems, the end users require them to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
+In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
+A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}.
+Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
+A Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
+In this work, we introduce a novel formulation of interpretable classification rules, namely \emph{relaxed-CNF}, which achieve benefits of  both worlds: it is interpretable similar to CNF/DNF  but more expressible by allowing cardinality constraints in the representation.
+\section{Fairness in Machine Learning}
+
+\section{Contributions}
+
+\section{Outline}
+
 The tremendous success of machine learning in recent years has found its applications in different high-stake and  safety-critical domains, such as medical, law, education, and transportation.

### 2022-05-31 chapters/fairness/fvgm/abstract.tex
intro: fairness part start

@@
 
-In this paper, we propose an efficient fairness verifier, called \fvgm, that encodes the correlations among features as a Bayesian network.
-In contrast to existing verifiers, \fvgm~proposes a \textit{stochastic subset-sum} based approach for verifying linear classifiers.
-Experimentally, we show that  \fvgm~leads to an \textit{accurate} and \textit{scalable} assessment for more diverse families of fairness-enhancing algorithms, fairness attacks, and group/causal fairness metrics than the state-of-the-art fairness verifiers.
+In this paper, we propose an efficient fairness verifier, called {\fvgm}, that encodes the correlations among features as a Bayesian network.
+In contrast to existing verifiers, {\fvgm}~proposes a \textit{stochastic subset-sum} based approach for verifying linear classifiers.
+Experimentally, we show that  {\fvgm}~leads to an \textit{accurate} and \textit{scalable} assessment for more diverse families of fairness-enhancing algorithms, fairness attacks, and group/causal fairness metrics than the state-of-the-art fairness verifiers.
 We also demonstrate that {\fvgm} facilitates the computation of fairness influence functions as a stepping stone to detect the source of bias induced by subsets of features.

### 2022-05-31 chapters/fairness/fvgm/applications.tex
intro: fairness part start

@@
 Another application of \fvgm{} as a fairness verifier is to quantify the effect of a subset of features on fairness.
-Thus, we define fairness influence function (FIF)  that computes the effect of a subset of non-sensitive features $\mathbf{S} \subseteq \nonsensitive$ on the PPV of a classifier given a specific sensitive group $ \sensitive = \mathbf{a}$,
+Thus, we define fairness influence function (FIF)  that computes the effect of a subset of non-sensitive features $\mathbf{S} \subseteq \nonsensitive$ on the probability of positive prediction of a classifier given a specific sensitive group $ \sensitive = \mathbf{a}$,
 $
@@
 FIF allows us to explain the sources of unfairness in the classifier.
-In practice, we compute FIF of $\mathbf{S}$ by replacing the probability distribution of $\mathbf{S}$ with a uniformly random distribution, referred to as  $ \mathcal{D}_{-\mathbf{S}} $, and reporting differences in the conditional PPVs.
+In practice, we compute FIF of $\mathbf{S}$ by replacing the probability distribution of $\mathbf{S}$ with a uniformly random distribution, referred to as  $ \mathcal{D}_{-\mathbf{S}} $, and reporting differences in the conditional probability of positive prediction of the classifier.
 
@@
 This dataset concerns the likelihood of a person of re-offending crimes within the next two years.
-We first observe that the base PPV values are different for the two groups ($ 0.46 $ vs $ 0.61 $ for Female and Male), thereby showing Male as more probable to re-offend crimes than Female.
+We first observe that the base values of the probability of positive prediction are different for the two groups ($ 0.46 $ vs $ 0.61 $ for Female and Male), thereby showing Male as more probable to re-offend crimes than Female.
 Moreover, FIF of the feature `age' is comparatively higher in magnitude for Male than Female.
@@
 Thus, applying {\fvgm} to compute FIF provides us insights about sources of bias and thus, indicators to improve fairness.
+$\Pr[\hat{Y} = 1|$age\_0 $\ge$ 0.5]
 
@@
 	\caption{FIF computation for COMPAS dataset.
-For Female (left plot) and Male, influence of `age' decreases and the PPV of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}
+For Female (left plot) and Male, influence of `age' decreases and the probability of positive prediction of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}
 %	\vspace{-2ex}

### 2022-05-31 chapters/fairness/fvgm/background.tex
intro: fairness part start

@@
 Let $ \alg : (\nonsensitive, \sensitive) \rightarrow \hat{Y} $  be a binary classier trained on $ D $, where $ \hat{Y} \in \{0,1\} $ is the predicted class.
-The positive predictive value (PPV) of $ \alg $, denoted as $ \Pr[\hat{Y} = 1] $, is the probability of positive outcome.
+The positive predictive value (\red{PPV}) of $ \alg $, denoted as $ \Pr[\hat{Y} = 1] $, is the probability of positive outcome.
  	We use $ (\mathbf{x}, \mathbf{a}) $ to denote the feature-values of  $ (\nonsensitive, \sensitive) $.
@@
  	  For all of the above metrics, lower value of $\epsilon$ indicates higher fairness demonstrated by the classifier $\mathcal{M}$.
-Following the observation of~\cite{ghosh2020justicia},  computing all of the aforementioned fairness metrics is equivalent to computing the maximum and minimum of the \textit{Positive Predictive Value} (PPV) of the classifier, denoted as $\Pr[\hat{Y}=1|\mathbf{A} =\mathbf{a}]$, for all compound sensitive groups $\mathbf{a}$ from $ \sensitive $.
-Thus, in Section~\ref{fvgm_sec:fvgm}, we focus on computing the maximum and minimum PPVs and then extend it to assess corresponding fairness metrics.
-We call the group for which PPV is maximum (minimum) as the \textit{most (least) favored group}.
+Following the observation of~\cite{ghosh2020justicia},  computing all of the aforementioned fairness metrics is equivalent to computing the maximum and minimum of the \textit{probability of positive prediction} of the classifier, denoted as $\Pr[\hat{Y}=1|\mathbf{A} =\mathbf{a}]$, for all compound sensitive groups $\mathbf{a}$ from $ \sensitive $.
+Thus, in Section~\ref{fvgm_sec:fvgm}, we focus on computing the maximum and minimum probability of positive prediction of the classifier and then extend it to assess corresponding fairness metrics.
+We call the group for which the probability of positive prediction is maximum (minimum) as the \textit{most (least) favored group} of the classifier.
 
-% 	\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, 
-% 	Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of PPV among sensitive groups, and prescribe this to be near zero.
+% 	\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of \red{PPV}s between the most favored group and least favored group, and prescribe it to be close to $1$. Formally, 
+% 	Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of \red{PPV} among sensitive groups, and prescribe this to be near zero.
 Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $,
@@
 % 	
-% 	\blue{It may be space consuming to discuss different fairness metrics, in which case we can mention that computing PPV is the core component of estimating different fairness metrics.}
+% 	\blue{It may be space consuming to discuss different fairness metrics, in which case we can mention that computing \red{PPV} is the core component of estimating different fairness metrics.}
 %

### 2022-05-31 chapters/fairness/fvgm/dp_illustration.tex
intro: fairness part start

@@
 
-\caption{Search tree representation of {\stochastic} for computing the maximum PPV of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
+\caption{Search tree representation of {\stochastic} for computing the maximum probability of positive prediction of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
 Each node is labeled by $ (i,\tau') $, where $ i $ is the index of $ \mathbf{B} $ and $ \tau' $ is the residual threshold.
@@
 Additionally, a green node denotes a collision, in which case a previously computed $ \mathsf{dp} $ solution is returned.
-Leaf nodes (gray) are computed based on terminating conditions in Eq.~\ref{fvgm_eq:dp_terminus}.
+Leaf nodes (gray) are computed based on terminating conditions in Eq.~\eqref{fvgm_eq:dp_terminus}.
 In Figure~\ref{fvgm_fig:example_BN},  nodes with double circles, such as $ \{(1,2), (2,1), (2,2)\} $,  are enumerated exponentially to compute conditional probabilities from the Bayesian network.}

### 2022-05-31 chapters/fairness/fvgm/evaluations.tex
intro: fairness part start

@@
 Since all non-sensitive features are from Gaussian distributions, we compute the probability of the predicted class $ \Pr[\hat{Y} | A = 1]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i, \sigma_{\hat{Y}}^2) $ and $ \Pr[\hat{Y} | A = 0]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i', \sigma_{\hat{Y}}^2) $ with $ \sigma_{\hat{Y}}^2 =   (\sum_{i=1}^{n-1}w_i^2)\sigma^2 $.
-Hence, the PPV of the classifier  is $  1 - \mathsf{CDF}_{\hat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\hat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function.
-Finally, we compute DI by taking the ratio of the minimum and the maximum of the two PPVs.}
+Hence, the probability of positive prediction of the classifier  is $  1 - \mathsf{CDF}_{\hat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\hat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function.
+Finally, we compute DI by taking the ratio of the minimum and the maximum of the probability of positive prediction of the classifier.}
 using different verifiers.

### 2022-05-31 chapters/fairness/fvgm/framework.tex
intro: fairness part start

@@
 In this section, we present {\fvgm}, a fairness verification framework for linear classifiers that accounts for correlated features represented as a graphical model.
-The core idea of verifying fairness of a classifier is to compute the PPV of the classifier with respect to all compound sensitive groups.
-To this end, {\fvgm} solves a stochastic subset sum problem, {\stochastic}, that is equivalent to computing the PPV of the classifier for the most (and the least) favored sensitive group.
+The core idea of verifying fairness of a classifier is to compute the probability of positive prediction of the classifier with respect to all compound sensitive groups.
+To this end, {\fvgm} solves a stochastic subset sum problem, {\stochastic}, that is equivalent to computing the probability of positive prediction of the classifier for the most (and the least) favored sensitive group.
 In this section, we first define {\stochastic} and present an efficient dynamic programming solution for {\stochastic}.
@@
 Hence, the prediction $ \hat{Y} =1 $ if and only if the inner inequality holds.
-Thus, computing the maximum (resp.\ minimum) PPV is equivalent to finding out the assignment of $A_j$'s for which the probability of satisfying the inner inequality is highest (resp.\ lowest).
+Thus, computing the maximum (resp.\ minimum) probability of positive prediction is equivalent to finding out the assignment of $A_j$'s for which the probability of satisfying the inner inequality is highest (resp.\ lowest).
 We reduce this computation into an instance of {\stochastic}.
@@
 
-\textbf{Computing Minimum and Maximum PPVs of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
+\textbf{Computing Minimum and Maximum probability of positive prediction of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
 The coefficients $ w_{A_j} $ and $ w_{X_i} $ of the classifier become weights of $ \mathbf{B} $.
@@
 For non-sensitive variables $ X_i $, which are chance variables, we derive their marginal probability $ p_i = \Pr[X_i = 1] $ from the distribution $ \mathcal{D} $.
-According to semantic of {\stochastic}, setting $ \exists $ quantifiers on $ \sensitive $ computes the maximum value of $ S(\mathbf{B}, \tau) $ that equalizes the maximum PPV of the classifier.
+According to semantic of {\stochastic}, setting $ \exists $ quantifiers on $ \sensitive $ computes the maximum value of $ S(\mathbf{B}, \tau) $ that equalizes the maximum probability of positive prediction of the classifier.
 In this case, the \textit{inferred} assignment of $ \sensitive $ implies the most favored group $ \mathbf{a}_{{\max}} =  \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
-In contrast, to compute the minimum PPV, we instead assign each variable  $ A_j $ a universal  quantifier  while keeping random quantifiers over $ X_i $, and infer the least favored group $ \mathbf{a}_{{\min}} =  \argmin_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
+In contrast, to compute the minimum probability of positive prediction, we instead assign each variable  $ A_j $ a universal  quantifier  while keeping random quantifiers over $ X_i $, and infer the least favored group $ \mathbf{a}_{{\min}} =  \argmin_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
 
 \iffalse
-\red{The decision version of computing the maximum (minimum) PPV is to decide whether there is an assignment of \textit{sensitive} or \textit{choice variables}, for which the \textit{non-sensitive} or \textit{chance variables} yield a PPV greater or less than $\alpha \in [0,1]$.
+\red{The decision version of computing the maximum (minimum) \red{\red{probability of positive prediction}} is to decide whether there is an assignment of \textit{sensitive} or \textit{choice variables}, for which the \textit{non-sensitive} or \textit{chance variables} yield a \red{\red{probability of positive prediction}} greater or less than $\alpha \in [0,1]$.
 Now, we formally state the hardness of verifying linear classifiers followed by an efficient dynamic programming solution.
@@
 Herein, $ P $ is a Boolean sensitive feature, and $ Q, R, S $ are Boolean non-sensitive features with $ \Pr[Q] = 0.4,  \Pr[R] = 0.5 $, and $ \Pr[S] = 0.3 $.
-To compute the maximum PPV of the classifier,  we impose an existential quantifier on $P$ and randomized quantifiers on others.
+To compute the maximum probability of positive prediction of the classifier,  we impose an existential quantifier on $P$ and randomized quantifiers on others.
 This leads us to the observation that $ P = 1 $ is the optimal assignment as $ w_P = 1 > 0 $.
@@
 The solution is visualized as a search tree in Figure~\ref{fvgm_fig:example_dp}, where we observe that storing the solution of sub-problems in the memory avoids repetitive computation, such as exploring the node $ (4,0) $.
-Similarly, the minimum PPV  of the classifier is $ 0.14 $ (not shown in Figure~\ref{fvgm_fig:example_dp}) where we impose a universal quantifier on $P$ to obtain $ P = 0 $ as the optimal assignment.
+Similarly, the minimum probability of positive prediction  of the classifier is $ 0.14 $ (not shown in Figure~\ref{fvgm_fig:example_dp}) where we impose a universal quantifier on $P$ to obtain $ P = 0 $ as the optimal assignment.
 \end{example}

### 2022-05-31 chapters/fairness/fvgm/framework2.tex
intro: fairness part start

@@
 In {\stochastic} presented in Section~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent.
-This independence assumption often leads to an \textit{inaccurate estimate} of PPVs because both sensitive and non-sensitive features can be correlated in practical fairness problems.
+This independence assumption often leads to an \textit{inaccurate estimate} of the probability of positive prediction of the classifier because both sensitive and non-sensitive features can be correlated in practical fairness problems.
 Therefore, we extend {\stochastic} to include correlations among variables.
@@
 \textbf{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $.
-For computing PPVs of a linear classifier with correlated features, we consider a heuristic to sort variables in $ \mathbf{B} = \sensitive \cup \nonsensitive $.
+For computing the probability of positive prediction of a linear classifier with correlated features, we consider a heuristic to sort variables in $ \mathbf{B} = \sensitive \cup \nonsensitive $.
 Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the network and $ \mathbf{V}^c = \mathbf{B} \setminus \mathbf{V} $.
@@
 We, however, observe that the dynamic programming solution in Section~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $.
-Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum PPV  as $ 0.65 $ for $ P = 1 $.
-The minimum PPV (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $.
+Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum probability of positive prediction of the classifier as $ 0.65 $ for $ P = 1 $.
+The minimum probability of positive prediction (not shown)  is similarly calculated as $ 0.11 $ for $ P = 0 $.
 	\end{example}
 
-	\subsection{Fairness Verification with Computed PPVs} 
+	\subsection{Fairness Verification with Computed Probability of Positive Prediction} 
 	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair ($\epsilon \in [0,1]$) is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$.
-We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum PPV $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum PPV $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.
+We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum probability of positive prediction $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum probability of positive prediction $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.
 

### 2022-05-31 chapters/introduction.tex
intro: fairness part start

@@
 Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
-Thus, we propose efficient algorithms for learning interpretable rule-based classifiers.
+Thus, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
 In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
-Consequently, fairness research centers on quantifying fairness using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
+Consequently, fairness research centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
 In this thesis, we study fairness verification problems to assess the fairness claim of different classifiers under the lens of multiple fairness definitions and algorithms.
@@
 
-The contributions of the thesis are discussed in the following.
-
 \section{Interpretable Machine Learning}
-The problem in interpretable machine learning is to learn a classifier making predictions interpretable to the human.
+The problem in interpretable machine learning is to learn a classifier making interpretable predictions to the end-users.
 To achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
@@
 Popular interpretable rule-based classifiers such as decision tree, decision lists, and decision sets share the logical structure of CNF/DNF in their representation of the decision function.
-Thereby, we propose to learn a CNF classifier, wherein its interpretability is defined by the number of Boolean literals that the formula contains.
+Thereby, we propose to learn an interpretable CNF classifier, wherein its interpretability is defined by the number of Boolean literals that the formula contains.
 Compared to CNF, Boolean cardinality constraints are more expressive as they allow numerical bounds on Boolean literals~\cite{sinz2005towards}.
-Relying on the concept of cardinality constraint to increase expressiveness, our second interpretable rule-based classifier is a logical relaxation of CNF/DNF classifiers, namely \emph{relaxed-}CNF.
+Relying on the concept of cardinality constraints to increase expressiveness, our second interpretable rule-based classifier is a logical relaxation of CNF/DNF classifiers, namely \emph{relaxed-}CNF.
 
@@
 Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
-\red{Now comes incremental learning}
+The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets.
+Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP (mixed integer linear programming) to efficiently learn rule-based classifiers in a mini-batch learning setting.
 
-In our study, we propose an alternate optimization approach that fits particularly well to rule-learning problems.
-Particularly,  we propose a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.
+The contributions of this thesis on interpretable machine learning are summarized in the following.
 
-In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean predicates that the formula contains.
+\subsection*{Scalability via Incremental Learning}
+We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for synthesizing interpretable classification rules expressible in proposition logic.
+$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns an optimal rule by solving an appropriately designed MaxSAT query.
+Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to practical classification datasets containing thousands to millions of samples.
+Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning.
+The resulting framework learns a classifier by iteratively covering the training data, wherein in each iteration, it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch.
+In our experiments, $ \mathsf{IMLI} $ achieves the best balance among prediction accuracy, interpretability, and scalability.
+For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability w.r.t.\ existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
+As an application, we deploy $ \mathsf{IMLI} $ in learning popular interpretable classifiers such as decision lists and decision sets.
 
-The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
-To enable  safe, robust, and trustworthy integration of such systems, the end users require them to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
-In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
-A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}.
-Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
-A Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
-In this work, we introduce a novel formulation of interpretable classification rules, namely \emph{relaxed-CNF}, which achieve benefits of  both worlds: it is interpretable similar to CNF/DNF  but more expressible by allowing cardinality constraints in the representation.
+\subsection*{Expressiveness via Logical Relaxation}
+We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}.
+Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
+Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain.
+Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets.
+While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive.
+To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
+Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
+
 \section{Fairness in Machine Learning}
+As a technology machine learning is oblivious to societal good or bad.
+The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
+In such applications, the deployed  classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}.
+For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.
+Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier.
+For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.
+To this end, different fairness enhancing algorithms have been devised to improve fairness with respect to one or multiple fairness metrics.
+These algorithms try to rectify and mitigate bias in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
+Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we focus on two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.
 
-\section{Contributions}
+\subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features.
+
+\subsection*{Identification of Sources of Unfairness}
 

### 2022-06-01 chapters/abstract.tex
adding fifs and introduction

@@
 FIF analysis interprets fairness by revealing potential features or subset of features attributing highly to the bias.
-Building on global sensitivity analysis, we propose an algorithm, called {\fairxplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.
+Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.
 

### 2022-06-01 chapters/fairness/fif/abstract.tex
adding fifs and introduction

\begin{abstract}

	Fairness in machine learning has attained significant focus due to the widespread application of machine learning in high-stake decision-making tasks. Unless regulated with a fairness objective, machine learning classifiers might demonstrate unfairness/bias towards certain demographic populations in the data. Thus, the quantification and mitigation of the bias induced by classifiers have become a central concern. In this paper, \textit{we aim to quantify the influence of different features on the bias of a classifier}. To this end, we propose a framework of \emph{Fairness Influence Function} (FIF), and compute it as a scaled difference of conditional variances in the classifier's prediction. We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features. Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.

\end{abstract}

### 2022-06-01 chapters/fairness/fif/additional_exp.tex
adding fifs and introduction

\begin{comment}
TODO
\begin{itemize}
	\item all group fairness metrics
	\item vary sample size  and dimension of features (add statistics for each dataset in the figure)
	\item runtime vs max-order $ \lambda $
	\item draw $ 1/n $, $ 1/n^{1/2} $, $ 1/n^{1/3} $, std.
	\item \red{Experiment of fairness enhancing and attack algo on the same dataset.} 
\end{itemize}
\end{comment}

\clearpage

\begin{figure}
	\centering
	\subfloat[]{\includegraphics[scale=0.3]{figures/sp_train_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/eo_train_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_train_time}}\\
	\subfloat[]{\includegraphics[scale=0.3]{figures/sp_test_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/eo_test_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_test_time}}\\
	\caption{Comparison between {\framework} and SHAP on runtime. Lower values on the $ Y $-axis are better. \red{[Should be discussed in the Supplementary material.]}}
\end{figure}

\begin{figure}
	\centering
	\subfloat[Unsorted weights]{\includegraphics[scale=0.45]{figures/feature_weight_sobol_unsorted}}\\
	\subfloat[Sorted weights including first and second order effect]{\includegraphics[scale=0.45]{figures/feature_weight_sobol}}\\
	\caption{Explaining statistical parity in COMPAS dataset for the sensitive feature `race' using Sobol method. This approach takes the (independent) distribution of features as input instead of a dataset. We apply KDE (Kernel Density Estimation) to learn the PDF of each feature.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_shap}}\\
	\caption{Explaining statistical parity in COMPAS dataset for the sensitive feature `race' using Shap method. This method is based on local explanations of classifiers. We observe higher discrepancy of this explanation as the computed statistical parity is often far from the exact value. Shap method can explain effect of individual features, which is a crucial limitation.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_eo_y_0}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_eo_y_1}}\\
	\caption{Explaining equalized odds in COMPAS dataset for the sensitive feature `race' using HDMR method.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_sufficiency_y_0}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_sufficiency_y_1}}\\
	\caption{Explaining sufficiency fairness metrics in COMPAS dataset for the sensitive feature `race' using HDMR method.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/german_rw_dt_sex_before}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/german_rw_dt_sex_after}}
	\caption{Fairness enhancing preprocessing algorithm, reweighing..}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/german_op_dt_sex_before}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/german_op_dt_sex_after}}
	\caption{Fairness enhancing preprocessing algorithm, optimized preprocessing. f\red{X axis should have the same X limit}}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_0}}
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_1}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_0_and_feature_1}}
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_higher-order_influence}}\\
	\caption{Fairness attack. Fairness influence is presented  in bar while statistical parity is shown in a line.}
\end{figure}

### 2022-06-01 chapters/fairness/fif/appendix.tex
adding fifs and introduction

\appendix
\section*{Appendix}
\section{Societal Impact}
In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset. This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias. This also allows us to understand the effect of the fairness enhancing algorithms in removing bias and their achievement/failure to do so. To the best of our understanding, this paper does not have any negative societal impact.

\section{FIF as Difference of Conditional Variances: Proof of Theorem 1}\label{sec:proof_1}

\begin{theoremrep}[FIF as Difference of Conditional Variances]
	\label{lemma:fif_rep}
	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied. Let $\mathbf{a}_{\max}$ and $\mathbf{a}_{\min}$ be the most and the least favored groups, respectively. Then, FIF of $ \nonsensitive_{\mathbf{S}} $ corresponding to statistical parity is
	\begin{equation}\tag{\ref{eq:fif_decompose}}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])},
	\end{equation}
	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
\end{theoremrep}

\begin{proof}

Let $ B_{\mathbf{a}} \in \{0, 1\} $ be a Bernoulli random variable such that $ B_{\mathbf{a}} = 1 $ denotes the classifier predicting positive class $ \hat{Y} = 1 $ for an individual belonging to the sensitive group $ \sensitive = \mathbf{a} $, and $ B_{\mathbf{a}} = 0 $ denotes $ \hat{Y} = 0 $. Hence, if $ p_\mathbf{a} \triangleq \Pr[B_\mathbf{a} = 1] $, then $ p_\mathbf{a} $ is also the conditional probability of positive prediction of the classifier,  $ p_\mathbf{a} = \Pr[\hat{Y} = 1 |  \sensitive = \mathbf{a}] $. The variance of $ B_{\mathbf{a}} $ is computed as
\begin{align*}
	\mathsf{Var}[B_\mathbf{a}] = p_{\mathbf{a}}(1 - p_{\mathbf{a}}). 
\end{align*}

Now, we consider two Bernoulli random variables $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ associated with two different sensitive groups $ \sensitive = \mathbf{a} $ and $ \sensitive = \mathbf{a}' $, respectively. The difference in variances between $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ implies that
\begin{align*}
	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}] &= p_{\mathbf{a}}(1 - p_{\mathbf{a}}) - p_{\mathbf{a'}}(1 - p_{\mathbf{a'}}) \\
	&= p_{\mathbf{a}} - p_{\mathbf{a}}^2 - p_{\mathbf{a'}} + p_{\mathbf{a'}}^2 \\ 
	&= (p_\mathbf{a} -  p_\mathbf{a'}) (1 - (p_\mathbf{a} + p_\mathbf{a'}))
\end{align*}

After reorganization, we express the difference in probabilities $ p_\mathbf{a} -  p_\mathbf{a'} $ as
\begin{align}
	\label{eq:sp_to_var}
	p_\mathbf{a} -  p_\mathbf{a'}  = \frac{	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}]}{1 - (p_\mathbf{a} + p_\mathbf{a'})}
\end{align}

By setting $ \mathbf{a} = \mathbf{a}_{\max} $ and $ \mathbf{a'} = \mathbf{a}_{\min} $ in Equation~\eqref{eq:sp_to_var}, we express the statistical parity of the classifier in terms of the scaled difference in the conditional variance of the positive prediction of the classifier. 
\begin{align}
	f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq p_{\mathbf{a}_{\max}} -  ~~p_{\mathbf{a}_{\min}}  &= \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}\label{eq:sp_var_diff}\\
	&= \frac{	\mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\max}] - \mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\min}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \notag.
\end{align}

%The last inequality is obtained by the definition of $B_{\mathbf{a}}$.

Next, we apply variance decomposition (Equation~\eqref{eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $. From Equation~\eqref{eq:variance_decomposition_set_notation} in Section~\ref{sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as 
\[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
where $ V_{\mathbf{a}, \mathbf{S}} $ denotes the decomposed variance of $ B_\mathbf{a} $ w.r.t.\ the features $ \nonsensitive_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.

Now, we apply variance decomposition on both $ \mathsf{Var}[B_{\mathbf{a}_{\max}}] $ and $ \mathsf{Var}[B_{\mathbf{a}_{\min}}] $ in Equation~\ref{eq:sp_var_diff} to obtain

\begin{align}
f_{\mathsf{SP}}(\alg, \mathbf{D}) & = \frac{\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\max}})}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \label{eq:influence_value}
\end{align}
Finally, we estimate  the FIF of the subset of features $ \nonsensitive_{\mathbf{S}}  $ as 
\begin{align}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}
\end{align} 
by separating the terms corresponding to $ \nonsensitive_{\mathbf{S}} $.

Following Equation~\eqref{eq:influence_value}, $	w_{\mathbf{S}} $ satisfies Axiom~\ref{axm:additivity} as $f_{\mathsf{SP}}(\alg, \mathbf{D}) = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} $.
\end{proof}

\section{A Smoothing Operator $\textsc{Smooth}$: Cubic Splines}
\label{sec:smoothing} 
In the $\textsc{LocalRegression}$ module of \framework{} (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}, Algorithm~\ref{algo:framework}), we use a smoothing operator $\textsc{Smooth}$ (Line~\ref{alg_line:backfitting_step}). In our experiments, \textit{we use cubic splines as the smoothing operator}. Here, we elucidate the technical details of cubic splines.

In interpolation problems, a B-spline of order $ n $ is traditionally used to smoothen the intersection of piecewise interpolators~\cite{schumaker2007spline}. A B-spline of degree $n$ is a piecewise polynomial of degree $ n - 1 $ defined over a variable $ X $. Each piecewise term is computed on local points and is aggregated as a global curve smoothly fitting the data. The values of $ X $ where the polynomial pieces meet together are called knots, and are denoted by $\{ \dots, t_0, t_1, t_2, \dots\} $.

Let $ B_{r, n}(X) $ denote the basis function for a B-spline of order $ n $, and $ r $ is the index of the knot vector. According to Carl de Boor~\cite{de1971subroutine}, $ B_{r,1}(X) $, for $ n = 1 $, is defined as
\begin{align*}
B_{r,1}(X) = 
\begin{cases}
0 &\text{ if } X < t_r \text{ or } X \ge t_{r+1},\\
1 &\text{ otherwise}
\end{cases}
\end{align*}

This definition satisfies $ \sum_i B_{r, 1}(X) = 1 $. The higher order basis functions are defined recursively as
\begin{align*}
	B_{r, n + 1}(X) = p_{r, n}(X)B_{r, n}(X) + (1 - p_{r + 1, n}(X))B_{r + 1, n}(X),
\end{align*}
where 
\begin{align*}
p_{r,n}(X) = 
\begin{cases}
\frac{X - t_r}{t_{r + n} - t_r} &\text{ if } t_{r + n} \ne t_r,\\
0 &\text{ otherwise.}
\end{cases}
\end{align*}

In this paper, we consider cubic splines with the basis function $ B_{r,4}(X) $ that constitutes a B-spline of degree $ 3 $. This polynomial has $ C^2 $ continuity, i.e. for each piecewise term, derivatives up to the second order are zero at the endpoints of each interval in the knot vector. We estimate component functions $ \function_{\mathbf{a}, \mathbf{S}} $'s with the basis function $ B_{r,4}(\mathbf{X})  $ of cubic splines~\cite{li2010global}, as shown in Equation~\eqref{eq:cubic_splines}.  
\begin{align}\label{eq:cubic_splines}
\begin{split}
&\function_{\mathbf{a}, \{i\}} (\mathbf{X}_{\{i\}}) \approx \sum_{r=-1}^{m+1}\alpha_r^iB_{r,n}(\mathbf{X}_{\{i\}})\\
&\function_{\mathbf{a}, \{i, j\}} (\mathbf{X}_{\{i, j\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\beta_{pq}^{ij}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})\\
&\function_{\mathbf{a}, \{i, j, k\}} (\mathbf{X}_{\{i, j, k\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\sum_{r=-1}^{m+1}\gamma_{pqr}^{ijk}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})B_{r,n}(\mathbf{X}_{\{j\}})
\end{split}
\end{align}

Here, $ m $ is the number of knots. We learn the coefficients $ \alpha, \beta, \gamma $ using the backfitting algorithm (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}, Algorithm~\ref{algo:framework}).

\begin{comment}
\paragraph{Kernel Smoothing.}
Kernel smoother estimates a real-valued function $ \function : \mathbb{R}^k \rightarrow \mathbb{R} $ as the weighted average of neighboring observed data. The weight is defined by the kernel such that closer points are given higher weights. The estimated function is smooth, where the level of smoothness is specified by a single parameter. Let $ k_{h_\lambda}(\mathbf{X}_0, \mathbf{X}) $ be a kernel defined by
\[
k_{h_\lambda}(\mathbf{X}_0, \mathbf{X}) = D\Big(\frac{||\mathbf{X}_0, \mathbf{X}||}{h_\lambda(\mathbf{X}_0)}\Big)
\]

where: 
\begin{itemize}
	\item $ \mathbf{X}_0, \mathbf{X} \in \mathbb{R}^k $
	\item $ ||\cdot|| $ is Euclidean norm
	\item $ h_\lambda(\mathbf{X}_0) $ is a parameter known as kernel radius
	\item $ D(t) $ is a positive real-valued function, whose value is decreasing (or not increasing) for the increasing distance $ t $
\end{itemize}
Popular kernels used for smoothing include parabolic, tricube, and Gaussian kernels. Popular kernel smoothers are Gaussian kernel smoother, nearest neighbor smoother, kernel average smoother, local linear or polynomial regression.

\end{comment}

%\clearpage
\section{Experimental Evaluations}\label{sec:additional_experiments}

\subsection{Experimental Setup}
We conduct experiments on Red Hat Enterprise Linux Server release 6.10 (Santiago) with $ \text{E}5-2690 \text{ v}3 $ CPU and $ 8 $GB of RAM. Since our objective is to compute FIFs for any classifier, we do not perform any tuning of hyper-parameters during training. We use the default hyper-parameter choices available in Scikit-learn~\cite{scikit-learn}. The only parameter that we choose in {\framework} is the maximum order of intersectionality, namely $ \lambda $. We present the ablation study w.r.t.\ $ \lambda $ in Appendix~\ref{sec:ablation}.

\subsection{Runtime of {\framework}} 
In Figure~\ref{fig:runtime}, we report the runtime of {\framework} in computing FIFs (with $ \lambda = 2 $) for different group fairness metrics on four classifiers and five datasets. The dimension of the dataset, measured by the number of samples and the number of features, is a key factor for runtime. For example, Adult dataset has the maximum number of samples ($ = 26048 $) compared to other datasets. In contrast, German dataset has the maximum features ($ = 26 $). Consequently, {\framework} requires more runtime in both Adult and German dataset compared to others. However, \emph{in all datasets, FIF computation takes at most $ 25 $ seconds, that demonstrates the efficiency of {\framework} in practical fairness problems.}

\begin{figure}[h!]
	\centering
	\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/sp_train_time_detailed_fairXplainer_only}}
	\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/eo_train_time_detailed_fairXplainer_only}}\\
	\subfloat[Predictive parity]{\includegraphics[scale=0.39]{figures/suff_train_time_detailed_fairXplainer_only}}
	\caption{Runtime of {\framework} (in seconds) for computing FIFs in different datasets. For each dataset, we mention its dimension as a tuple (\# sample, \# features) in the xticks. As the dimension increases, we observe higher runtime of {\framework}.}\label{fig:runtime}\vspace*{-1em}
\end{figure}

\subsection{Ablation Study: Effect of Maximum Order of Intersectionality $ \lambda $}\label{sec:ablation} 
We test the effect of $ \lambda \in \{1, 2, 3\} $ on the runtime of {\framework}. For each $ \lambda $, we consider $ 540 $ fairness instances ($ 5 $ datasets with $ 27 $ different combinations of sensitive features $ \times $ $ 5 $-fold cross validation $ \times $ $ 4 $ classifiers). We plot the corresponding results in the cumulative distribution plot of  Figure~\ref{fig:varying_max_order}. Here, $ X $-axis denotes the runtime (in seconds) and $ Y $-axis denotes the total number of solved instances, i.e. a point $ (x,y) $ denotes that $ y $ number of instances are solved within $ x $ seconds.

As we increase from $ \lambda = 1 $ to $ \lambda = 2 $ and then to $ \lambda = 3 $, {\framework} requires around $ 0.5 $ and $ 2 $ orders of magnitude more runtime, respectively, to solve the equal number of instances. This is due to the fact that with increase in $ \lambda $, there is a combinatorial explosion in the number of component functions in the backfitting algorithm. Thus, \emph{computing higher-order FIFs requires a higher runtime of {\framework}}.

\begin{figure}[t!]
	\centering
	{\includegraphics[scale=0.38]{figures/vary_max_order_train}}
	%	\subfloat[]{\includegraphics[scale=0.4]{figures/vary_max_order_test}}	
	\caption{Effect on runtime while varying max-order $ \lambda $ in {\framework}. Higher $ \lambda $ incurs more a computational effort, hereby solving less fairness instances within the same time limit.}\label{fig:varying_max_order}
\end{figure}

\subsection{Accuracy in Estimating Different Fairness Metrics}
We compare and evaluate the accuracies of {\framework} and SHAP in estimating a fairness metric, computed as the sum of FIFs (Axiom~\ref{axm:additivity}). We illustrate the results with a scatter-plot in Figure~\ref{fig:accuracy_scatter_plot}. In the plot, each dot represents the estimation of a fairness metric on a dataset and a classifier.

The majority of dots for {\framework} are on the line connecting $ (0, 0) $ to $ (1, 1) $  in Figure~\ref{fig:accuracy_scatter_plot}. This implies that the estimation of {\framework} is significantly accurate. Interestingly, error rarely occurs in {\framework} due to the degenerate cases mentioned in Section~\ref{sec:fifs}. 
In contrast, SHAP often computes a metric inaccurately, thereby showing the drawback of applying a local explanation approach for FIF computation. \emph{Thus, {\framework} is significantly more accurate than SHAP in computing FIFs and the corresponding fairness metrics.}

\begin{figure}[t!]\vspace*{-1em}
	\centering
	\subfloat[Statistical parity]{\includegraphics[scale=0.35]{figures/scatter_plot_sp_train_accuracy}}
	\subfloat[Equalized odds]{\includegraphics[scale=0.35]{figures/scatter_plot_eo_train_accuracy}}
	\subfloat[Predictive parity]{\includegraphics[scale=0.35]{figures/scatter_plot_suff_train_accuracy}}
	\caption{Comparison between the estimated and exact value of a fairness metric, computed as the sum of FIFs according to Axiom~\ref{axm:additivity}. Results are presented separately for {\framework} and SHAP. Each dot represents the result for a fairness metric, carried over multiple datasets and classifiers. Ideally, if a dot lies on the line connecting $ (0,0) $ to $ (1,1) $, it denotes a correct fairness estimate. {\framework} demonstrates significantly better accuracy of estimation than SHAP. For predictive parity metric, only {\framework} allows associated FIFs computation.}
	\label{fig:accuracy_scatter_plot}
\end{figure}

\subsection{FIF of Different Datasets}
We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult, Ricci, and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fig:individual_vs_intersectional_influence_adult},~\ref{fig:individual_vs_intersectional_influence_ricci}, and \ref{fig:individual_vs_intersectional_influence_titanic}, respectively. In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Section~\ref{sec:experiments}.

\textbf{In Adult dataset,} the classifier predicts whether an individual earns more than $ \$50 $k per year or not, where `race' and `sex' are sensitive features. We observe that the trained network is unfair and it demonstrates statistical parity as $ 0.23 $. As we analyze FIFs,  \textit{education number}, \textit{age}, and \textit{capital gain/loss} are key features responsible for the bias. Unlike COMPAS dataset (Figure~\ref{fig:individual_and_intersectional_fifs}), there does not exist any feature or subset of features in Adult dataset that potentially reduces the bias.

\begin{figure}[!t]
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/adult_feature_weight_unsorted}\label{fig:individual_fifs_adult}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/adult_feature_weight}\label{fig:individual_and_intersectional_fifs_adult}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{FIFs for Adult dataset on explaining statistical parity. }\label{fig:individual_vs_intersectional_influence_adult}

\end{figure}

\textbf{In Ricci dataset,} the classification task  is to predict whether a firefighter obtains promotion or not in New Haven, Connecticut administered exams. We consider `race' as the sensitive feature for this experiment. We observed that the classifier has statistical parity of $ 0.3 $ based on the race of a person. In the computed FIFs, the \textit{combined score} and \textit{desired position} of a person increases bias, whereas written exam score reduces bias while interacting with other features. In addition, the higher order influences demonstrate a higher value (FIF $ = 0.26 $), meaning that the features are highly correlated and act in favor of bias.

\begin{figure}[h!]
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/ricci_feature_weight_unsorted}\label{fig:individual_fifs_ricci}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/ricci_feature_weight}\label{fig:individual_and_intersectional_fifs_ricci}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{FIFs for Ricci dataset on explaining statistical parity.}\label{fig:individual_vs_intersectional_influence_ricci}

\end{figure}

\textbf{In Titanic dataset,} the neural network predicts whether a person survives the Titanic shipwreck or not. In this experiment, we consider the `sex' of a person as a sensitive feature and observe that the classifier is highly unfair (statistical parity as $ 0.83 $) on the basis of `sex'. In the computed FIFs, most of the features except the \textit{passenger class} and the \textit{age} of a person increases the bias of the classifier.

\begin{figure}[h!]
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/titanic_feature_weight_unsorted}\label{fig:individual_fifs_titanic}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/titanic_feature_weight}\label{fig:individual_and_intersectional_fifs_titanic}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{FIFs for Titanic dataset on explaining statistical parity.}\label{fig:individual_vs_intersectional_influence_titanic}

\end{figure}

### 2022-06-01 chapters/fairness/fif/conclusion.tex
adding fifs and introduction

%\vspace{-0.5em}
\section{Conclusion}
%\vspace{-1em}
We propose the Fairness Influence Function (FIF) to quantify the contribution of input features on the resulting bias of a classifier for a given dataset. Relying on an additive axiom, we express group fairness metrics computed for sensitive groups as the sum of FIFs of all the subsets of non-sensitive features. To compute FIFs, we first prove existing group fairness metrics as the scaled difference of the conditional variances in the predictions of the classifier and then apply variance decomposition based on global sensitivity analysis. Finally, we propose {\framework}, an algorithm for efficiently and accurately computing FIFs by deploying a local regression to learn a set-additive decomposition of the classifier. The experimental results show that {\framework} estimates bias with significantly higher accuracy than the local explanation based approach SHAP. Also, {\framework} computes both individual and intersectional FIFs unlike SHAP to yield a better understanding of the sources of bias.
In future,  we aim to develop algorithms that leverage FIFs to yield unbiased decisions.

### 2022-06-01 chapters/fairness/fif/experiments.tex
adding fifs and introduction

%\clearpage
\section{Empirical Performance Analysis}\label{sec:experiments}
In this section, we perform an empirical evaluation of {\fairXplainer}. In the following, we discuss the experimental setup, the objectives of experiments, and experimental results. { We adjoin the source code and additional experimental results on the effects of maximum order $\lambda$, runtime, accuracy on different datasets and corresponding explanations (Appendix~\ref{sec:additional_experiments}) to the supplementary material.}

\noindent\textbf{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $). In {\fairXplainer}, we deploy SAlib library~\cite{Herman2017} to compute FIFs leveraging techniques of global sensitivity analysis. In our experiments, we consider five widely studied datasets from fairness literature, namely COMPAS~\cite{angwin2016machine}, Adult~\cite{DK2017uci}, German-credit~\cite{DK2017}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}). We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: Logistic Regression, Support Vector Machine, Decision Tree, and Neural Networks. In experiments, we specify {\fairXplainer} to compute intersectional influences up to second order ($ \lambda = 2 $). We compare {\fairXplainer} with Shapley-valued based FIF computational framework, referred as SHAP~\cite{lundberg2020explaining}. In addition, we deploy {\fairXplainer} along with different fairness-enhancing~\cite{kamiran2012data} and fairness attack~\cite{solans2020poisoning} algorithms, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objective of our empirical study. 
\begin{enumerate}[nosep,leftmargin=*]
%\begin{enumerate}
	\item How do \textbf{accuracies} of {\fairXplainer} and SHAP compare for estimating FIFs and resulting bias?
	\item What is the impact of\textbf{ intersectional vs.\ individual FIFs} in tracing the sources of bias?
    \item How do FIFs change while \textbf{applying} different fairness enhancing algorithms, i.e. \textbf{affirmative actions}, and fairness attacks, i.e. \textbf{punitive actions}?
	%\item \red{What is the effect of $ \lambda $ on the runtime of {\fairXplainer}?}
\end{enumerate}
In summary, \textit{{\fairXplainer} achieves significantly less estimation error than SHAP in all the datasets.} This shows the importance of adopting global sensitivity analysis to compute FIFs for group fairness metrics than local explanation approaches. Moreover, \textit{{\fairXplainer} effectively traces the sources of bias by computing intersectional FIFs, which earlier method like SHAP {did not capture}.} {\fairXplainer} also detects the effects of the affirmative and punitive actions on the bias of a classifier and the corresponding tensions between different subsets of features. %Due to brevity of space, we defer additional experimental results to the Supplementary material. %In the following, we elaborate results on estimation error and individual vs.\ intersectional FIFs.

\begin{figure}[!t]
	\begin{minipage}{0.49\textwidth}
		\centering
		\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/fairness/fif/sp_train_accuracy}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/fairness/fif/eo_train_accuracy}}
	\end{minipage}
	\caption{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics. Lower values on the $ Y $-axis denote a better result. {\fairXplainer} has significantly less error than SHAP.}\label{fig:estimation_error}
%	\vspace{-0.5em}
\end{figure}
\noindent\textbf{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}). We consider five datasets, each trained on different classifiers by applying five-fold cross-validation. We compute estimation error by taking the absolute difference between the exact and  estimated value of a metric, and present results in Figure~\ref{fig:estimation_error}. In both metrics, {\fairXplainer} demonstrates significantly less error than SHAP. For example, the mean error of {\fairXplainer} in estimating $ \mathsf{SP} $ on Titanic dataset is $ 0.1 $ vs.\ $ 0.42 $ for SHAP. In this context, the error of {\fairXplainer} is often zero in most datasets and only insignificant ($ \le 0.13 $) due to the degenerate cases (Section~\ref{sec:fifs}) with $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{0, 1\} $.  Therefore, \emph{global sensitivity analysis based approach {\fairXplainer} is significantly more accurate in computing FIFs and corresponding fairness metric than local explanation based approach SHAP}.

%\clearpage

\noindent\textbf{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs. We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity as $ 0.17 $, which means a non-Caucasian  has $ 0.17 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the source of bias and present individual FIFs in Figure~\ref{fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fig:individual_and_intersectional_fifs}. In both figures, we show top seven influential FIFs sorted by absolute values,  followed by residual higher-order FIFs. In Figure~\ref{fig:individual_fifs}, `priors count' of an individual is dominating in increasing statistical parity (FIF $ = 0.12 $). Other non-sensitive features appear to have almost zero FIFs. However, the sum of second-order influences of all features increases statistical parity by $ 0.06 $, denoting that the data is highly correlated and presenting only individual FIFs does not trace the true sources of bias. For example, while both `sex' and `age' of persons have zero influence on bias (Figure~\ref{fig:individual_fifs}), these features together with `c charge degree', `priors count', and `juvenile other count' contribute highly on statistical parity (sum of FIFs $ = 0.03 + 0.02 + 0.01 + 0.01 - 0.01 - 0.01 = 0.05$). \emph{Therefore, intersectional influences together with individual influences lead to a clearer understanding on the source of bias of a classifier. Note that, unlike SHAP, {\fairXplainer} is the only framework that computes beyond individual FIFs.}
\begin{figure}[t!]
	\begin{minipage}{0.43\textwidth}
		\centering
		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fig:individual_fifs}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fig:individual_and_intersectional_fifs}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{FIFs for COMPAS dataset on explaining statistical parity. Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}

\end{figure}
\begin{figure}[t!]
	\vspace{-0.7em}
	\begin{minipage}{0.49\textwidth}
		\centering
		\subfloat[Fairness attack (punitive actions)]{\includegraphics[scale=0.345]{figures/fairness/fif/fairness_attack}\label{fig:fairness_attack}}
	\end{minipage}
	\begin{minipage}{0.5\textwidth}
		\centering	
		\subfloat[Fairness enchancing algorithm (affirmative actions)]{\includegraphics[scale=0.345]{figures/fairness/fif/fairness_repair}\label{fig:fairness_repair}}
	\end{minipage}
	\vspace{-0.2em}
	\caption{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing algorithm~\cite{kamiran2012data} on FIFs.}\label{fig:affirmative_punitive_actions}
%	\vspace{-0.5em}
\end{figure}

\noindent\textbf{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.  In Figure~\ref{fig:individual_vs_intersectional_influence}, statistical parity of the classifier is $ 0.17 $. Applying a data poisoning fairness attack~\cite{solans2020poisoning} increases statistical parity to $ 0.5 $ and the data reweighing-based fairness-enhancing algorithm~\cite{kamiran2012data} decreases it to $ 0.15$. We observe that in both cases, there are a subset of features decreasing bias while another subset of features increasing it, e.g. `age' vs.\ `priors count'. Figure~\ref{fig:fairness_attack} suggests that the attack algorithm would be more successful if it could hide the influence of `age' of a person in receiving discriminating prediction for re-offending crimes. Figure~\ref{fig:fairness_repair} suggests that the fairness enhancing algorithm can improve by ameliorating the effect of `priors count' further. Thus, \textit{{\fairXplainer} provides a dissecting tool to undertake necessary steps to improve or worsen fairness of the classifier.}

### 2022-06-01 chapters/fairness/fif/experiments_conf.tex
adding fifs and introduction

%\clearpage
\section{Empirical Performance Analysis}\label{sec:experiments}
In this section, we perform an empirical evaluation of {\framework}. In the following, we discuss experimental setup, objective of experiments, and experimental results\footnote{Source code, details on preliminaries, and additional experimental results are in the supplementary material.}.

\noindent\textbf{Experimental Setup.} We implement a prototype of {\framework} in Python (version $ 3.8 $). In {\framework}, we deploy SAlib library~\cite{Herman2017} to compute FIFs based on global sensitivity analysis. In our experiments, we consider five widely studied datasets in fairness literature, namely COMPAS, Adult, German-credit, Ricci, and Titanic. We deploy Scikit-learn library to learn different classifiers: Logistic Regression, Support Vector Machine, Decision Tree, and \red{Neural Networks}. \red{We set $ \lambda = 2 $ meaning that we allow {\framework} to compute inter-sectional influences up to order two.}

We compare {\framework} with Shapley-valued based FIF computational framework, referred as SHAP. In addition, we deploy {\framework} on different fairness enhancing and fairness attack algorithms and analyze the effect of these algorithms on the FIFs and the resultant fairness metric. In the following, we discuss the objective of our empirical study.

1. How does {\framework} compare with SHAP in terms of accuracy in computing FIFs and the resulting bias?

2. What is the impact of inter-sectional vs.\ individual FIFs in understanding the sources of bias of a classifer?

3. How do FIFs change while applying different fairness enhancing and fairness attack algorithms?

4. What is the effect of $ \lambda $ on the runtime of {\framework}?

In summary...

%\subsection{Experimental Results}
%In the following we discuss our experimental results in detail.

%\clearpage
\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.4]{figures/sp_train_accuracy}}
	\subfloat{\includegraphics[scale=0.4]{figures/eo_train_accuracy}}
	%	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_train_accuracy}}\\
	%	\subfloat[]{\includegraphics[scale=0.3]{figures/sp_test_accuracy}}
	%	\subfloat[]{\includegraphics[scale=0.3]{figures/eo_test_accuracy}}
	%	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_test_accuracy}}\\

	\caption{Comparison between {\framework} and SHAP on estimation error of fairness metrics. Lower values on the $ Y $-axis denote a better result. \red{Add results for EO and CUAE}}
	\label{fig:estimation_error}
\end{figure}

\paragraph{Estimation Error of Bias.} We compare {\framework} with SHAP in estimating statistical parity $ \mathsf{SP} $ and equalized odds $ \mathsf{EO} $, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}). We consider \red{five} datasets, each trained on \red{three} different classifiers by applying five fold cross-validation. We compute estimation error by taking the absolute difference between the exact and  estimated value of a metric and present results in Figure~\ref{fig:estimation_error}. In both metrics, {\framework} demonstrates significantly less estimation error than SHAP. For example, the mean error of {\framework} in estimating $ \mathsf{SP} $ on Titanic dataset is $ 0.1 $ vs $ 0.42 $ for SHAP. In this context, the estimation error of {\framework} is due to the degenerate cases in Section~\ref{sec:fifs}, where $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ is either $ 1 $ or $ 0 $.  Therefore, local explanation based approach SHAP is significantly inaccurate in computing FIFs and corresponding fairness metric than global sensitivity analysis based approach {\framework}.

\begin{figure}
	\centering
	\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/feature_weight_unsorted}\label{fig:individual_fifs}}
	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/feature_weight}\label{fig:individual_and_intersectional_fifs}}\\
	\caption{Explaining statistical parity in COMPAS dataset for the sensitive feature `race' using HDMR method. \red{Effect to be influence. Change 'Feature' from Y axis. SP in different color and in the title. X axis label = statistical parity w.r.t.\ race. An experiment comparing LR and NN (showing accuracy, sp and FIFs side by side.)}}
	\label{fig:individual_vs_intersectional_influence}
\end{figure}

\paragraph{Individual vs.\ Inter-sectional FIFs.} Our first experiment is to understand the importance of inter-sectional FIFs over individual FIFs. We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years. Since the classifier only optimizes training error, it demonstrates statistical parity as $ 0.17 $, denoting that a non-Caucasian  has $ 0.17 $ higher probability of re-offending crimes than a Caucasian. Next, we investigate the source of bias and present individual FIFs in Figure~\ref{fig:individual_fifs} and both individual and intersectional FIFs in Figure~\ref{fig:individual_and_intersectional_fifs}. In both figures, we show top seven influential FIFs sorted by absolute values, followed by residual high-order FIFs. In Figure~\ref{fig:individual_fifs}, `priors count' of an individual is dominating in increasing statistical parity (FIF $ = 0.12 $). Other non-sensitive features appear to have almost zero FIFs. However, the sum of second-order influences of all features increases statistical parity by $ 0.06 $, denoting that the data is highly correlated and presenting only individual FIFs does not demonstrate the true sources of bias. For example, while both `sex' and `age' of persons have zero influence on bias in an individual level (Figure~\ref{fig:individual_fifs}), these features together with `c charge degree', `priors count', and `juvenile other count' contribute highly on statistical parity (sum of FIF $ = 0.03 + 0.02 + 0.01 + 0.01 - 0.01 - 0.01 = 0.05$). Therefore, inter-sectional influences together with individual influences demonstrate a clear understanding on the source of bias of a classifier. Note that, unlike SHAP, {\framework} is the only framework to compute beyond individual FIFs.

%\clearpage

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.4]{figures/vary_max_order_train}}
%	\subfloat[]{\includegraphics[scale=0.4]{figures/vary_max_order_test}}	
	\caption{Varying max-order $ \lambda $.}
	\label{fig:varying_max_order}
\end{figure}

\paragraph{Effect of Max-order $ \lambda $.} We experiment the effect of \red{$ \lambda \in \{1, 2, 3\} $} on the runtime of {\framework}. For each $ \lambda $, \blue{we consider $ 1086 $ fairness instances ($ 5 $ datasets totaling to $ 27 $ different combinations of sensitive features $ \times $ $ 5 $-fold cross validation $ \times $ $ 8 $ classifiers)}. In the \red{cumulative plot} in  Figure~\ref{fig:varying_max_order}, $ X $-axis denotes runtime and $ Y $-axis denotes total solved instances; a point $ (x,y) $ denotes that $ y $ many instances are solved within $ x $ seconds. As we increase $ \lambda $, {\framework} requires more time (around $ 0.5 $ order of magnitude) to solve equal number of instances. Because, for higher $ \lambda $, there is a combinatorial explosion in the number of component functions in the backfitting algorithm. Thus, an computing higher-order FIFs results in a higher runtime of {\framework}.

\begin{comment}
TODO
\begin{itemize}
	\item all group fairness metrics
	\item vary sample size  and dimension of features (add statistics for each dataset in the figure)
	\item runtime vs max-order $ \lambda $
	\item draw $ 1/n $, $ 1/n^{1/2} $, $ 1/n^{1/3} $, std.
	\item \red{Experiment of fairness enhancing and attack algo on the same dataset.} 
\end{itemize}
\end{comment}

\input{sections/conclusion.tex}

\clearpage

\begin{figure}
	\centering
	\subfloat[]{\includegraphics[scale=0.3]{figures/sp_train_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/eo_train_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_train_time}}\\
	\subfloat[]{\includegraphics[scale=0.3]{figures/sp_test_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/eo_test_time}}
	\subfloat[]{\includegraphics[scale=0.3]{figures/suff_test_time}}\\
	\caption{Comparison between {\framework} and SHAP on runtime. Lower values on the $ Y $-axis are better. \red{[Should be discussed in the Supplementary material.]}}
\end{figure}

\begin{figure}
	\centering
	\subfloat[Unsorted weights]{\includegraphics[scale=0.45]{figures/feature_weight_sobol_unsorted}}\\
	\subfloat[Sorted weights including first and second order effect]{\includegraphics[scale=0.45]{figures/feature_weight_sobol}}\\
	\caption{Explaining statistical parity in COMPAS dataset for the sensitive feature `race' using Sobol method. This approach takes the (independent) distribution of features as input instead of a dataset. We apply KDE (Kernel Density Estimation) to learn the PDF of each feature.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_shap}}\\
	\caption{Explaining statistical parity in COMPAS dataset for the sensitive feature `race' using Shap method. This method is based on local explanations of classifiers. We observe higher discrepancy of this explanation as the computed statistical parity is often far from the exact value. Shap method can explain effect of individual features, which is a crucial limitation.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_eo_y_0}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_eo_y_1}}\\
	\caption{Explaining equalized odds in COMPAS dataset for the sensitive feature `race' using HDMR method.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_sufficiency_y_0}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/feature_weight_sufficiency_y_1}}\\
	\caption{Explaining sufficiency fairness metrics in COMPAS dataset for the sensitive feature `race' using HDMR method.}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/german_rw_dt_sex_before}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/german_rw_dt_sex_after}}
	\caption{Fairness enhancing preprocessing algorithm, reweighing..}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/german_op_dt_sex_before}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/german_op_dt_sex_after}}
	\caption{Fairness enhancing preprocessing algorithm, optimized preprocessing. f\red{X axis should have the same X limit}}
\end{figure}

\begin{figure}
	\centering
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_0}}
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_1}}\\
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_feature_0_and_feature_1}}
	\subfloat{\includegraphics[scale=0.45]{figures/fairness_attack_higher-order_influence}}\\
	\caption{Fairness attack. Fairness influence is presented  in bar while statistical parity is shown in a line.}
\end{figure}

### 2022-06-01 chapters/fairness/fif/extra.tex
adding fifs and introduction

\clearpage
\red{Extra}

\section{Introduction}
The last decade has witnessed a tremendous progress in machine learning with a host of applications of algorithmic decision making. Presently, machine learning classifiers make high-stake decisions in various safety-critical domains such as college admission, recidivism prediction, and job applications. In these domains, the fairness and explainability of the deployed classifiers are of paramount importance. Fairness in machine learning detects and improves the unfairness/bias of the classifier towards certain demographic population in the dataset, such as offering more college admission to White-male candidates than to Black-female candidates. Moreover, the black-box nature of classifiers, such as neural networks, encourages researchers in developing explainable methods for black-box predictions. The explainability methods identify important features for tabular data or highlight regions in an image that the black-box classifer may focus on during prediction. In this paper, we combine both the fairness and explainability aspects in machine learning and propose an explanation framework for the fairness of a classifer.

There is a rich literature on algorithmic fairness and explainability in machine learning. To quantify the bias of an unfair classifier, multiple fairness definitions have been proposed. In this paper, we focus on \red{\emph{group fairness}}, which quantifies the bias in the prediction of the classifier among \emph{sensitive} groups in the data. Group-based fairness metrics mostly belong to three categories: independence, separation, and sufficiency. In binary classification, independence group fairness metrics, such as statistical parity, quantify whether the probability of positive prediction of the classifier is equal across all sensitive groups. Separation metrics, such as equalized odds, apply conditions on the ground-truth (actual) class in the data and constrain that for population belonging to each ground-truth class, the probability of positive prediction is equal across all sensitive groups. In contrast, sufficiency metrics, such as conditional use accuracy equality, apply conditions on the predicted class and emphasize that for each predicted class, the probability of individuals originally belonging to a particular ground truth class is equal across all sensitive groups. From the measurement aspect of these metrics, one  computes the difference in the (conditional)  probability of positive prediction of the classifier between each pair of sensitive groups and outputs the maximum difference as a measure of the fairness metric.

There is a growing interest in explaining the prediction of opaque black-box classifiers. \red{When the input to the classifer is a feature vector}, the most popular explanation methods are based on additive feature attribution~\cite{lundberg2017unified,ribeiro2016should}. In this method, the prediction of the classifier (almost) equals the sum of weights/effects attributable to each individual feature~\cite{lundberg2017unified} or the higher-order interactions among features (or a subset of features)~\cite{sundararajan2020shapley}. Moreover, these weights are computed locally for a single sample, thereby tagging such methods as local explainers. To explain a classifier globally on a dataset, local explanation is performed on each sample separately and later each individual local weight is aggregated (for example, by averaging) for all samples as a global weight of a feature (or a subset of features).

While fairness and explainability in machine learning are studied separately in the literature, there has been an insignificant effort in applying explainability methods to explain the fairness of a classifer. To the best of our knowledge, there are two studies that explain fairness using Shapley values~\cite{lundberg2020explaining,benesse2021fairness}. The key idea in both studies is to compute the global weight of individual features on the prediction of the classifier separately for each sensitive group and take the difference in individual weights between specific groups. Consequently, the sum of differences in weights corresponds to group fairness metrics such as statistical parity. This explainability approach for fairness suffers from two-fold limitations. Firstly, fairness, particularly group fairness, is a global property of a classifier. Hence, the state-of-the-art local to global explainability procedure leads to an \emph{inaccurate explanation} of fairness. Secondly, existing approaches are limited to explaining fairness in terms of individual features. However, features are often correlated in real-world fairness problems. To this end, practical algorithms are required to explain fairness by considering higher-order interactions among features, where existing Shapley-based computation suffers from \emph{poor scalability}. Therefore, in this paper, we ask the following research question: \emph{Can we design an explainability framework that can accurately and scalably explain the fairness of a classifer by attributing to individual features and their higher-order interactions?}

The primary contribution of this paper is an explanation framework for fairness, namely {\framework}, which explains the fairness of a classifier based on additive feature (or subset of features) attribution. The key idea is to compute the \emph{\red{fairness influence function}}\unsure{Do we introduce this terminology?} of a subset of non-sensitive features, such that the sum of influences of all possible subsets of non-sensitive features equals the fairness value achieved by the classifer. {\framework} relies on relating the fairness of a classifier with the difference in the conditional variances of the classifier between specific sensitive groups. To this end, we apply techniques from global sensitive analysis to decompose the variance of the classifier to compute the influence of features. Furthermore, to account for correlated features, we apply covariance decomposition while computing influences.

In our experiments....

\noindent\makebox[\textwidth]{\rule{\textwidth}{2pt}}\info{Up to this}

In order to formulate fairness influence functions, we next apply variance decomposition on $ \mathsf{Var}[B_{\mathbf{a}}] $
\begin{align}
\mathsf{Var}[B_\mathbf{a}] = \mathsf{Var}[Y=1|\sensitive = \mathbf{a}] =  \sum_{i=1}^{n}V_{\{i\}}^{(\mathbf{a})} +  \sum_{i<j}^{n} V_{\{i,j\}}^{(\mathbf{a})}  + \cdots  + V_{\{1,2,\dots,n\}}^{(\mathbf{a})}
\end{align}

where the superscript of $ V_{\mathbf{S}}^{(\mathbf{a})} $ denotes the decomposed variance $ V_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.  Thus, we rewrite Eq.~\ref{eq:sp_to_var} in terms of differences in the decomposed conditional \red{variances}.

\begin{align}
p_\mathbf{a} -  p_\mathbf{a'}  &= \frac{	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}]}{1 - (p_\mathbf{a} + p_\mathbf{a'})}\notag \\ 
&= \frac{\sum_{i=1}^{n} (V_{\{i\}}^{(\mathbf{a})} - V_{\{i\}}^{(\mathbf{a'})})+  \sum_{i<j}^{n} (V_{\{i,j\}}^{(\mathbf{a})} - V_{\{i,j\}}^{(\mathbf{a'})})  + \cdots  + (V_{\{1,2\dots,n\}}^{(\mathbf{a})} - V_{\{1,2\dots,n\}}^{(\mathbf{a'})})}{1 - (p_\mathbf{a} + p_\mathbf{a'})} \notag \\
&= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{S}}^{(\mathbf{a})} - V_{\mathbf{S}}^{(\mathbf{a'})}}{1 - (p_\mathbf{a} + p_\mathbf{a'})} \label{eq:influence_value}
\end{align}

Next, we compute $ w_{\mathbf{S}} $ by taking out terms associated with the subset of features $ \mathbf{S} $ in Eq.~\ref{eq:influence_value}.

\begin{align*}
p_\mathbf{a} -  p_\mathbf{a'} =  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{ V_{\mathbf{S}}^{(\mathbf{a})} - V_{\mathbf{S}}^{(\mathbf{a'})}}{1 - (p_\mathbf{a} + p_\mathbf{a'})} \\
\Rightarrow w_{\mathbf{S}} = \frac{V_{\mathbf{S}}^{(\mathbf{a})} - V_{\mathbf{S}}^{(\mathbf{a'})}}{1 - (p_\mathbf{a} + p_\mathbf{a'})}
\end{align*}

Intuitively, $ w_{\mathbf{S}} $ is the difference in the decomposed conditional \red{variances} of the classifier corresponding to the most and least favored sensitive groups, where the variance of the classifier is computed with respect to features $ \mathbf{X}_{\mathbf{S}} $.

\paragraph{Important points regarding Approach 1}
\begin{itemize}
	\item \textbf{Numerical instability:} The denominator of $ w_{\mathbf{S}} $ is $ 1 - (p_\mathbf{a} + p_{\mathbf{a}}') \ne 0 $.  We may require to add a small number to the denominator to avoid the cases when the value is $ 0 $.

	\item \textbf{Computation of variance decomposition:} There is rich literature on computing the decomposed variance of a function, such as Fourier-based algorithm FAST and Sobol sensitivity analysis. FAST only computes first order \red{variances}, when $ |\mathbf{S}| = 1 $, whereas Sobol sensitivity approaches can compute both first and second order \red{variances}, $ 1 \le |\mathbf{S}| \le 2 $. From a practitioner's perspective, first and second order effects should be sufficient to inspect the cause of unfairness of a classifier.

	\item \textbf{Sign of $ w_{\mathbf{S}} $:} $ w_{\mathbf{S}} $ can be  positive, negative or zero. If $ w_{\mathbf{S}} > 0 $, then the interaction $ \mathbf{X}_{\mathbf{S}} $ is responsible for increasing the statistical parity of the classifier, thereby leading to higher unfairness. When $ w_{\mathbf{S}} < 0 $, then the interaction $ \mathbf{X}_{\mathbf{S}} $ favors fairness. Finally, the interaction $ \mathbf{X}_{\mathbf{S}} $ is neutral when $ w_{\mathbf{S}} = 0 $. 
\end{itemize}

\begin{comment}
Next, we relate SP with the difference in \red{variances} of the classifier by substituting $ \mathbf{a} $ as $ \mathbf{a}_{\max} $ and $ \mathbf{a'} $ as $ \mathbf{a}_{\min} $.

\begin{align*}
\text{Statistical parity }= p_{\mathbf{a}_{\max}} -  p_{\mathbf{a}_{\min}}  = \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}
\end{align*}

Thus, statistical parity can be distributed to all subset of features in $ \nonsensitive \cup \sensitive $ based on variance decomposition. We observe that computing the variance for all possible subsets of features is computationally expensive and in practice one may look into only first and second-order \red{variances}.

\end{comment}

\section{Variance Decomposition}
There is a rich literature on variance decomposition for the sensitivity analysis of a function in terms of the cumulative effect of all possible subset of variables in the function. In our case, the non-sensitive features are correlated. As such, we discuss a meta-learning approach, where we apply functional decomposition on the function $ f(\mathbf{X}) $ with $ \delta $ being the implicit error in this decomposition.

\begin{align*}
f(\mathbf{X}) = f_{\emptyset} +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} f_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \delta
\end{align*}

Here, $ f_{\mathbf{S}} $ is known as a component function, where $ \lambda $ denotes its maximum order. In functional decomposition, lower order component functions impacts more on $ f(\mathbf{X}) $ than higher order functions. Once each $ f_{\mathbf{S}} $ is learned, we compute its variance $ V_{\mathbf{S}}  = \mathsf{Var}_{X_{\mathbf{S}}}[f_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}})]$. Furthermore, features in $ \mathbf{X} $ may be correlated with each other. Therefore, the component function $ f_{\mathbf{S}} $ is in correlation with other component function $ f_{\mathbf{S}'} $, for $ \mathbf{S} \ne \mathbf{S}' $. To account for this correlation,~\cite{li2010global} propose to decompose the variance of $ f(\mathbf{X}) $ as the sum of the variance of $ f_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) $'s along with its covariance with other component functions $ \sum_{\mathbf{S}' \ne \mathbf{S}} f_{\mathbf{S'}}(\mathbf{X}_{\mathbf{S}'}) $.

\[
\mathsf{Var}[f(\mathbf{X})] = \sum_{\mathbf{S} \subseteq [k] \setminus \emptyset} \Big(V_{\mathbf{S}}  + \mathbf{Cov}[f_{\mathbf{S}}, \sum_{\mathbf{S}' \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}'| \le \lambda, \mathbf{S} \ne \mathbf{S}'} f_{\mathbf{S}'}]\Big)
\]

Now, we focus on learning the component function $ f_{\mathbf{S}} $. In this paper, we apply backfitting algorithm, which is an iterative procedure to learn component functions in additive models. In each iteration, all but one component function, say $ f_{\mathbf{S}} $ is learned, while other components are fixed constant. Also, $ f_{\mathbf{S}} $ is learned as a smoothed function of $ f(\mathbf{X}) $ and $ f_{\mathbf{S}'} $, for $ \mathbf{S} \ne \mathbf{S}' $. In the following, we present the pseudo-code of backfitting algorithm.

\begin{algorithm}
	\caption{Backfitting algorithm}
	\label{algo:iterative_decision_sets_learning}
	\begin{algorithmic}[1]
		\State \textbf{Initialize}: $ \hat{f}_{\emptyset} \leftarrow \frac{1}{n}\sum_{i=1}^ny_i, \hat{f}_{\mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [k]\setminus\emptyset, |\mathbf{S}|\le \lambda$ 
		\While{$ \hat{f}_{\mathbf{S}} $ does not converge}
		\For{each $\mathbf{S}$}
		\State $ \hat{f}_{\mathbf{S}} \leftarrow \mathsf{Smooth}[\{y_i - \hat{f}_{\emptyset} - \sum_{\mathbf{S} \ne \mathbf{S}'} \hat{f}_{\mathbf{S}'}(x_{\mathbf{S}}^{(i)})\}_{i=1}^n] $ \Comment{Backfitting step}
		\State $\hat{f}_{\mathbf{S}}  \leftarrow \hat{f}_{\mathbf{S}} - \frac{1}{n}\sum_{i=1}^n \hat{f}_{\mathbf{S}}(x_{\mathbf{S}}^{(i)})$ \Comment{Mean centering}
		\EndFor
		\EndWhile
	\end{algorithmic}
\end{algorithm}

Here, $ \mathsf{Smooth} $ is a smoothing operator, which is typically chosen to be cubic spline smoother or Kernel smoother. In this paper, we have experimented with both approaches. First, we discuss cubic spline methods, as discussed in~\cite{li2010global}.

\paragraph{Cubic Splines as Smoothing Function.} Spline is a piecewise polynomial function studied in interpolation problems. Each piecewise term is computed on local points and is aggregated as a global curve with smoothly fitting the data. Cubic splines are of polynomials with degree $ 3 $ with continuity $ C^2 $ meaning that the second derivatives of each piecewise term is zero at the endpoints of each interval in the interpolation. ~\cite{li2010global} propose to estimate component function $ f_{\mathbf{S}} $'s in terms of cubic B spline function $ B_r(\mathbf{X}) $, where $ r $ is the index of a specified knot vector. More specifically, each component function $ f_{\mathbf{S}} $ is defined in terms of these basis functions.

\begin{align*}
&f_{\{i\}} (\mathbf{X}_{\{i\}}) \approx \sum_{r=-1}^{m+1}\alpha_r^iB_r(\mathbf{X}_{\{i\}})\\
&f_{\{i, j\}} (\mathbf{X}_{\{i, j\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\beta_{pq}^{ij}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})\\
&f_{\{i, j, k\}} (\mathbf{X}_{\{i, j, k\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\sum_{r=-1}^{m+1}\gamma_{pqr}^{ijk}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})B_r(\mathbf{X}_{\{j\}})\\
\end{align*}

Here, $ m $ is the number of knots, with $ \alpha, \beta, \gamma $ being constant coefficients that need to be determined.

\paragraph{Kernel Smoothing.}
Kernel smoother estimates a real-valued function $ f:\mathbb{R}^k \rightarrow \mathbb{R} $ as the weighted average of neighboring observed data. The weight is defined by the kernel such that closer points are given higher weights. The estimated function is smooth, where the level of smoothness is specified by a single parameter.

Let $ k_{h_\lambda}(X_0, X) $ be a kernel defined by
\[
k_{h_\lambda}(X_0, X) = D\Big(\frac{||X_0 - X||}{h_\lambda(X_0)}\Big)
\]

where: 
\begin{itemize}
	\item $ X, X_0 \in \mathbb{R}^k $
	\item $ ||\cdot|| $ is Euclidean norm
	\item $ h_\lambda(X_0) $ is a parameter, known as kernel radius
	\item $ D(t) $ is a positive real-valued function, whose value is decreasing (or not increasing) for the increasing distance $ t $
\end{itemize}
Popular kernels used for smoothing include parabolic, Tricube, and Gaussian kernels. Popular smoother based kernels are Gaussian kernel smoother, nearest neighbor smoother, kernel average smoother, local linear or polynomial regression.

\clearpage
\section{Methodology [\red{Backup}]}
In this section, we discuss the computational aspects of fairness influence $ w_{\mathbf{S}} $ of a subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $. The key idea in computing $ w_{\mathbf{S}} $ is to decompose the conditional variance $ \mathsf{Var}[B_{\mathbf{a}}] $ such that $ \mathsf{Var}[B_{\mathbf{a}}] = \sum_{\mathbf{S} \in [k] \setminus \emptyset} V_{\mathbf{S}}^{(\mathbf{a})} $. To compute $ V_{\mathbf{S}}^{(\mathbf{a})} $, assumptions are made on the distribution of input features. In the following, we discuss two approaches, one where features are independent and another where feature correlations are taken into account.

\subsection{Variance Decomposition on Independent Features}

Sobol sensitivity analysis is a popular approach for computing $ V_{\mathbf{S}} $ where input non-sensitive features are probabilistically independent of others. Sobol approach relies on Monte Carlo estimation where quasi Monte Carlo samples are generated by uniformly covering the sample space of features. These are also called Sobol sequences. Since the distribution of features is not necessarily uniform, each random sample is rescaled based on each feature's independent distribution. Let $ x_i \in [0,1] $ be the uniform random value of feature $ X_i $ during the Monte Carlo sampling. Let $ \mathsf{ICDF}_{i}: \{0,1\} \rightarrow \mathsf{supp}(X_i) $ is the inverse cumulative distribution function of $ X_i $, where $ \mathsf{supp}(X_i) $ denotes the possible feature values of $ X_i $. Then, $ x_i $ is rescaled to $ \mathsf{ICDF}_{i}(x_i) $.  In practice, the distribution of individual features are not readily available. We apply univariate KDE (kernel density estimation) to learn a distribution of features, which can be queried as $ \mathsf{ICDF} $.

Once a sufficient number of samples are generated (for example, $ 2^n $, for $ n \ge 10 $), Sobol approach computes $ V_{\mathbf{S}} $, for $ |\mathbf{S}| \in \{1,2\} $.\footnote{While Sobol approach computes the total effect of a subset of feature, we do not consider them in our setting. Because the sum of total Sobol indices does to add up to $ 1 $, which is a  requirement in our case.}

\textbf{Equations for Monte Carlo Estimation.}
Let sample matrices $ \mathbf{P} $, $ \mathbf{Q} $, and $ \mathbf{P}_{\mathbf{Q}}^i $, for $ i = 1, \dots, \numnonsensitive $.

Computing the first order ($ |\mathbf{S}| = 1 $) variance: 
\[
V_{\mathbf{S}}^{(\mathbf{a})} = 
\mathsf{Var}_{\mathbf{X}_{\mathbf{S}}}[\mathsf{E}_{\mathbf{X}_{\sim \mathbf{S}}}[\hat{Y} = 1 | \mathbf{X}_{\mathbf{S}}, \sensitive = \mathbf{a}]] = \frac{1}{n}\sum_{j=1}^{n}\alg(\mathbf{Q}, \sensitive)_j\Big(\alg(\mathbf{P}_{\mathbf{Q}}^i, \sensitive)_j - \alg(\mathbf{P}, \sensitive)_j\Big)
\]

\subsection{Variance Decomposition on Correlated Features}
The earlier Monte Carlo estimation approach for variance decomposition does not consider any correlation among features. For including feature-correlation we apply following techniques.

\paragraph{High Dimensional Model Representation (HDMR)~\cite{li2010global}} The key idea is to decompose the variance of $ \hat{Y} $ using the covariance of individual term $ \mathbf{X}_{\mathbf{S}} $, for $ \mathbf{S} \subseteq [k] \setminus \emptyset $.

\red{Goes to preliminaries.}
Let $ Y =  f(\mathbf{X}) $ be decomposed into a limited number of subsets of features $ \mathbf{X}_{\mathbf{S}} $.

\[
f(\mathbf{X})  =  f_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} f_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \epsilon
\]

Here parameter $ \lambda $ sets the upper bound of the cardinality of $ \mathbf{S} $. $ \epsilon $ is the error incurred due to limiting $ |\mathbf{S}| $ up to $ \numnonsensitive $.  Then, the variance of $ Y $ can be decomposed as the covariance of subset of features.
\[
\mathsf{Var}[Y] = \sum_{\mathbf{S} \subseteq [k] \setminus \emptyset} (V_{\mathbf{S}}  + \mathbf{Cov}[\mathbf{X}_{\mathbf{S}}, \sum_{\mathbf{S}' \subseteq [k] \setminus (\emptyset \cup \mathbf{S}), |\mathbf{S}'| \le \lambda} V_{\mathbf{S}'}])
\]

Here, $ V_{\mathbf{S}} $ is the uncorrelated effect on variance, whereas $ \mathbf{Cov}[\mathbf{X}_{\mathbf{S}}, \sum_{\mathbf{S}' \subseteq [k] \setminus (\emptyset \cup \mathbf{S}), |\mathbf{S}'| \le \lambda} V_{\mathbf{S}'}] $ is the correlated effect of variance of $ \mathbf{X}_{\mathbf{S}} $ on the total variance.

\paragraph{HDMR: Estimation Aspects~\cite{li2010global}}

\begin{enumerate}
	\item The first step is to compute component functions $ f_{\mathbf{S}} $'s using a regression method from a set of input-output examples. The cited paper applies RS-HDMR combined with a $ F $-test.

	\item RS-HDMR is a general approach to optimally construct component functions sequentially from lower to higher order, such that the lower order contributions are maximized and the higher order contributions are minimized. In this process, the higher order component functions (if they exist) are decomposed and portions are included in the low order ones.

	\item RS-HDMR computes $ f_{\mathbf{S}} $ with sampling of $ \mathbf{X} $ following any given PDF (Probability Density Function).

	\item To reduce the sampling effort, the RS-HDMR component functions are approximated by expansions in terms of some suitable basis functions. The cited paper used \red{cubic B spline functions} $ B_k(\mathbf{X}) $.

	\item The first, second, and third order RS-HDMR component functions are approximately expanded as 
	\begin{align*}
	&f_{\{i\}} (\mathbf{X}_{\{i\}}) \approx \sum_{r=-1}^{m+1}\alpha_r^iB_r(\mathbf{X}_{\{i\}})\\
	&f_{\{i, j\}} (\mathbf{X}_{\{i, j\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\beta_{pq}^{ij}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})\\
	&f_{\{i, j, k\}} (\mathbf{X}_{\{i, j, k\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\sum_{r=-1}^{m+1}\gamma_{pqr}^{ijk}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})B_r(\mathbf{X}_{\{j\}})\\
	\end{align*}

	Here, $ m $ is the number of knots, with $ \alpha, \beta, \gamma $ being constant coefficients that need to be determined.

	\item \textbf{Determination of component functions by backfitting}

	\begin{enumerate}
		\item After the component functions being approximated by suitable basis functions, a statistical $ F $-test is carried out to provide confidence bands for the predicted functions.

		\item Component functions can be determined sequentially or simultaneously by least-squares regression. If the system has independent inputs, the sequential determination is more efficient because the process does not depend on the order of the inputs and in each step only one component function needs to be determined. For correlated inputs, this approach is improper because the process depends on the order of inputs. For this case, either simultaneous or backfitting determinations can be used. The cited paper prefers backfitting approach in high-dimension systems because each iteration of backfitting only treats one component function whose dimension is always low (low to high order construction?)

		\item A new estimate of component function is estimated through backfitting.

		\[
		Y^{(i)} - f_{0}  - \sum_{\mathbf{S'} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S'}| \le \lambda, \mathbf{S} \ne \mathbf{S}'} f_{\mathbf{S}'}(\mathbf{X}_{\mathbf{S'}}^{(i)}) = f_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}}^{(i)}),  
		\]

		where $ i = 1, 2, \dots, n $ denotes the $ i $-th sample, and $ n $ is the total number of samples. This procedure is performed for each component function $ f_{\mathbf{S}} $ in term, using the current estimates of the $ f_{\mathbf{S}'} $'s. This process continues until the estimated $ f_{\mathbf{S}} $'s converge.

		\item \textbf{$ F $-testing}: Let $ RSS_1 $ represent the residual sum of squares for the least-squares regression of the large model $ f_0 + \sum_{\mathbf{S}' \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}'| \le \lambda} f_{\mathbf{S}'}(\mathbf{X}_{\mathbf{S}'})  $ with $ p_1 $ unknown parameters (e.g., the coefficients $ \alpha, \beta, \gamma $ in the spline function expansions). Also, let $ RSS_0 $ be the same quantity for a small model $ f_0 + \sum_{\mathbf{S}' \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}'| \le \lambda, \mathbf{S} \ne \mathbf{S}'} f_{\mathbf{S}'}(\mathbf{X}_{\mathbf{S}'})  $ nested in the large model, but with $ p_0 $ unknown parameters.

		\[
		F = \frac{(RSS_0 - RSS_1)/(p1 - p_0)}{RSS_1 / (n - p_1)}
		\]

		The $ F $ statistic has an $ F $ distribution with $ (p_1 - p_0) $ and $ (n - p_1) $ degrees of freedom. If the observed $ F $ is larger than the tabulated value of the $ F $ distribution with $ (p_1 - p_0) $ and $ (n - p_1) $ degrees of freedom at the $ 99 \% $ confidence level (or with other desired confidence level), the $ f_{\mathbf{S}} $ is significant  and should be included in the approximated. Otherwise, $ f_{\mathbf{S}} $ can be excluded. 
	\end{enumerate}
\end{enumerate}

\section{Approach $ 2 $ [\red{Revision required}]: Variance of PPV over Sensitive Features}

In the second approach, we express $ p_\mathbf{a} - p_{\mathbf{a}'} $ with variance of $ p_\mathbf{a} $ and  $ p_{\mathbf{a}'} $.  Let assume $ p_\mathbf{a} \ge  p_{\mathbf{a}'} $.

\begin{align*}
	\mathsf{Var}[\{p_\mathbf{a}, p_{\mathbf{a}'} \}] &= \frac{1}{2}((p_\mathbf{a} - \frac{p_\mathbf{a} + p_{\mathbf{a}'}}{2})^2 + (p_\mathbf{a}' - \frac{p_\mathbf{a} + p_{\mathbf{a}'}}{2})^2)\\
	&= \frac{1}{2^2}(p_\mathbf{a} -  p_{\mathbf{a}'})^2 = \frac{1}{2^2}\mathsf{sp}^2
\end{align*}

Now, we express $ \mathsf{Var}[\{p_\mathbf{a}, p_{\mathbf{a}'} \}] =  \mathsf{Var}_\sensitive[\mathsf{E}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]] $, where $ \sensitive $ takes two values $ \mathbf{a} $ and $ \mathbf{a}' $.  Reorganizing the above equation and applying the law of total variance result in the following derivation for $ \mathsf{sp} $.

\begin{align*}
	\frac{1}{2^2}\mathsf{sp}^2 &= \mathsf{Var}_\sensitive[\mathsf{E}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]\\
	&= \mathsf{Var}[\hat{Y}] - \mathsf{E}_\sensitive[\mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]\\
	&= \mathsf{Var}[\hat{Y}] - \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}] \mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive=\mathbf{a}]\\
	&= \mathsf{Var}[\hat{Y}] - \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Big(\Pr[\sensitive=\mathbf{a}] \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}}^{(\mathbf{a})} \Big)\\
	&= \mathsf{Var}[\hat{Y}] -  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \underbrace{\sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}}_{---}\\
	&\red{= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \left(\frac{1}{2^\numnonsensitive - 1} \mathsf{Var}[\hat{Y}] -   \underbrace{\sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}}_{w_{\mathbf{S}}}\right)}\\
\end{align*}

We compute influence of $ \mathbf{X}_{\mathbf{S}} $ as follows:

\begin{align}
	w_{\mathbf{S}} =  C\sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}
\end{align}

where the constant $ C $ is defined as:
\begin{align*}
	C &\triangleq \frac{\mathsf{Var}[\hat{Y}] - \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}}{\sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}} \\
	&= \frac{\mathsf{Var}[\hat{Y}] - \mathsf{E}_\sensitive[\mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]}{\mathsf{E}_\sensitive[\mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]}
\end{align*}

\red{
	From the above equation, we derive the influence of $ \mathbf{X}_{\mathbf{S}} $ in the following.
	\[
	w_{\mathbf{S}} =  \frac{1}{2^\numnonsensitive - 1} \mathsf{Var}[\hat{Y}] - \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}
	\]
	Here, $ \frac{1}{2^\numnonsensitive - 1} \mathsf{Var}[\hat{Y}] $ is the variance of $ \hat{Y} $ normalized by the total combination of subsets of features in $ \nonsensitive $. \red{Ideally, each $ \mathbf{X}_{\mathbf{S}} $ should have equal variance imposed by $ \hat{Y} $.} On the other hand, $ \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})} $ is the decomposition of the expected variance of $ \hat{Y} $ conditioned on different sensitive groups evaluated on the subset $ \mathbf{X}_{\mathbf{S}} $. Therefore, $ w_{\mathbf{S}} $ quantifies the difference between the ideal variance of the prediction $ \hat{Y} $ on $ \mathbf{X}_{\mathbf{S}} $ and  the expected conditional variance of $ \mathbf{X}_{\mathbf{S}} $ due to considering different sensitive groups.  
}

\begin{comment}
	\textit{ An alternate} is to further decompose $ \mathsf{Var}[\hat{Y}] $ and include in $ w_{\mathbf{S}} $.

	\begin{align*}
		\frac{1}{2^2}\mathsf{sp}^2 &= \mathsf{Var}[\hat{Y}] -  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})} \\ 
		&= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \Big(V_{\mathbf{S}}  - 
		\sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})} \Big) \\ 
	\end{align*}

	In that case, we compute $ w_{\mathbf{S}} $ as follows:

	\[
	w_{\mathbf{S}} = V_{\mathbf{S}}  - 
	\sum_{\mathbf{a} \in \{\mathbf{a}, \mathbf{a}'\}} \Pr[\sensitive=\mathbf{a}]V_{\mathbf{S}}^{(\mathbf{a})}
	\]

	\red{Update:}
	\begin{itemize}
		\item In current implementation, $ w_\mathbf{S} $ becomes $ 0 $ for each \textit{individual} features (when $ |\mathbf{S}| $ = 1) .  
	\end{itemize}
\end{comment}

\paragraph{Difference between approach $ 1 $ and $ 2 $}
\begin{itemize}
	\item In approach 1, $ w_{\mathbf{S}} $ is the difference in conditional variances while in approach 2, it is the expected conditional variance with inverse sign.
	\item In approach 2, we explain $ \mathsf{sp}^2 $, while in approach 1, we explain $ \mathsf{sp} $. 
	\item The interpretation of $ w_{\mathsf{S}} $ is opposite in the two approaches, because of the sign of $ w_{\mathbf{S}} $. 
\end{itemize}

\red{Why do we consider more than two sensitive groups? Especially, because we are explaining the difference of conditional PPVs between two sensitive groups.}

\section{Approach $ 2 $: General case.}
Let consider $ k $ different sensitive groups $ \mathbf{a}_1, \dots, \mathbf{a}_k $ each denoting an assignment of sensitive features $ \sensitive $. Also, let the conditional PPV of the classifier be $ p_i = \Pr[\hat{Y}=1|\sensitive=\mathbf{a}_i] $. The variance of the PPVs are derived as follows.

\begin{align*}
	\mathsf{Var}[\{p_1, \dots, p_k\}] 
	&= \frac{1}{k} \sum_i (p_i - \frac{1}{k}\sum_j p_j)^2 \\
	&= \frac{1}{k^3}\sum_i (\sum_j (p_i - p_j))^2 \\ 
	%		&\ge \frac{1}{k^3}\sum_i (\sum_j |p_i - p_j|)^2 \\ 
	&\ge \frac{1}{k^3} \sum_i (\max |p_i - p_j|)^2 \\
	&= \frac{1}{k^3} k (\max |p_i - p_j|)^2 \\
	&= \frac{1}{k^2} (\max |p_i - p_j|)^2
\end{align*}

Therefore,

\begin{align*}
	\max|p_i - p_j| \le k \sqrt{\mathsf{Var}[\{p_1, \dots, p_k\}] }
\end{align*}

Equality happens when $ k = 2 $. \red{and SP = 0 (why?)}

For example, when $ k  = 2 $, statistical parity, $ \max|p_i - p_j| = 2 \sqrt{\mathsf{Var}[\{p_1, \dots, p_k\}] } $.

\[
\mathsf{Var}[\{p_1, \dots, p_k\}]  \triangleq \mathsf{Var}_\sensitive[\mathsf{E}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]
\]

We recall the law of total variance:

\begin{align*}
	\mathsf{Var}[\hat{Y}] &= \mathsf{E}_\sensitive[\mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]] + \mathsf{Var}_\sensitive[\mathsf{E}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]] \\
	\Rightarrow \mathsf{Var}_\sensitive[\mathsf{E}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]] &= \mathsf{Var}[\hat{Y}] - \mathsf{E}_\sensitive[\mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive]]\\
	&=\mathsf{Var}[\hat{Y}] - \sum_{i=1}^k \Pr[\sensitive=\mathbf{a}_i] \mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive=\mathbf{a}_i]
\end{align*}

Therefore, $ \text{Statistical parity } = \max|p_i - p_j| \le k \sqrt{\mathsf{Var}[\hat{Y}] - \sum_{i=1}^k \Pr[\sensitive=\mathbf{a}_i] \mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive=\mathbf{a}_i]}$

Finally, we decompose $ \mathsf{Var}_{\mathbf{X}}[\hat{Y}|\mathbf{X}, \sensitive = \mathbf{a}_i] $ to variances of $1, 2, \dots, n $ order.

\paragraph{\red{Todo}}:

\begin{itemize}
	\item  Can we prove the central limit theorem of $ w_{\mathbf{S}} $ that relates to the theorem 2 in GSA based fairness paper?

	\item Can we prove error bound on finite sample?

	\item Approach 1 takes the subtraction of variance decomposition of two random variables. Our error analysis should consider this case. 
\end{itemize}

### 2022-06-01 chapters/fairness/fif/fairxplainer.tex
adding fifs and introduction

\section{\fairXplainer: An Algorithm to Compute Fairness Influence Functions}\label{sec:fairxplainer}
\begin{comment}
Represent the algorithm as a sequence of two blocks: (1) Component function learning (2) Covariance computation of component functions. 
\end{comment}
We propose an algorithm, {\fairXplainer}, that leverages the variance decomposition of Eq.~\eqref{eq:fif_decompose} to compute the Fairness Influence Functions (FIFs) of all the subset of features. {\fairXplainer} has two algorithmic blocks: (i) local regression to decompose the classifier into component functions and (ii) computing the variance (or covariance) of each component. We describe the schematic of \fairXplainer{} in Algorithm~\ref{algo:framework}.
\setlength{\textfloatsep}{12pt}% Remove \textfloatsep

\noindent\textbf{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}). Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \function_{\mathbf{a}}(\nonsensitive) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $. We express $ \function_{\mathbf{a}} $ as a set-additive model:
\begin{align}
\label{eq:set_additive_classifier}
\function_{\mathbf{a}}(\mathbf{X}) = \function_{\mathbf{a}, 0} +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} \function_{\mathbf{a},\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \delta
\end{align}
\begin{algorithm}
	\caption{{\fairXplainer}}\label{algo:framework}
	\hspace*{\algorithmicindent} \textbf{Input:} $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y},  \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n, f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}, \lambda $  \\
	\hspace*{\algorithmicindent} \textbf{Output:} $ w_{\mathbf{S}}$ 
	\begin{algorithmic}[1]
		\State $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], k \gets    |\nonsensitive| $
		\label{algo_line:fif_computation_start}

		\For{$ \mathbf{a} \in \{\mathbf{a}_{\max}, \mathbf{a}_{\min}\} $} \Comment{Enumerate for specific sensitive groups}
		\State $ \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0} \gets \textsc{LocalRegression}(\alg(\nonsensitive, \sensitive = \mathbf{a}), \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k) $

		\State $ V_{\mathbf{a}, \mathbf{S}} \gets \textsc{Covariance}(\function_{\mathbf{a}},  \{\mathbf{x}^{(i)}\}_{i=1}^n, \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0}) $
		\EndFor
		\State Compute $ w_{\mathbf{S}}$ using  $V_{\mathbf{a}_{\max},\mathbf{S}}$ and $V_{\mathbf{a}_{\min},\mathbf{S}}$ as in Equation~\eqref{eq:fif_decompose}
		%\State $ w_{\mathbf{S}}  = (V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}})/(1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])) $
		\label{algo_line:fif_computation_end}
		\Comment{Theorem~\ref{lemma:fif}}

%		\Statex
		\Function{LocalRegression}{$ g_\mathbf{a}, \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k $}
		\label{algo_line:local_regression_start}
		\State \textbf{Initialize}: $ \function_{\mathbf{a}, 0} \leftarrow \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n), \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [k]\setminus\emptyset, |\mathbf{S}|\le \lambda$ \label{alg_line:initialization}
		\While{each $ \widehat{\function}_{\mathbf{a},\mathbf{S}} $ does not converge}
		\For{each $\mathbf{S}$}
		\State $ \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow \textsc{Smooth}[\{\function_{\mathbf{a}}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\function}_{\mathbf{a}, \mathbf{S}'}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n] 
		$\label{alg_line:backfitting_step} \Comment{Backfitting}
		\State $\widehat{\function}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\function}_{\mathbf{a}, \mathbf{S}} - \textsc{Mean}(\{\widehat{\function}_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n)$ \label{alg_line:mean_centered} \Comment{Mean centering}
		\EndFor
		\EndWhile
		\State \Return $ \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0} $
		\label{algo_line:local_regression_end} 
		\EndFunction

		\Function{Covariance}{$\function_{\mathbf{a}},  \{\mathbf{x}^{(i)}\}_{i=1}^n, \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0}$}
		\label{algo_line:covariance_computation_start}
		\State \Return $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}) $
		\label{algo_line:covariance_computation_end}
		%		 \Comment{Eq.~\eqref{eq:covariance_computation}}
		\EndFunction
	\end{algorithmic}
\end{algorithm}
Here, $ \function_{\mathbf{a}, 0} $ is a constant and $ \function_{\mathbf{a},\mathbf{S}} $ is a \emph{component function} of $ \function_{\mathbf{a}} $ taking  $ \nonsensitive_{\mathbf{S}} $ as input, and $ \delta $ is the approximation error. For computational tractability, we consider only components of \emph{maximum order} $ \lambda $. {\fairXplainer} deploys backfitting algorithm for learning component functions in Eq.~\eqref{eq:set_additive_classifier}.

\noindent\textbf{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}). Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \function_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed. Specifically, $ \function_{\mathbf{a}, \mathbf{S}} $ is learned as a smoothed function of $ g $ and rest of the components $ \function_{\mathbf{a}, \mathbf{S}'} $, where $ \mathbf{S}' \ne \mathbf{S} $ is a non-empty subset of $ [k]\setminus \emptyset $. To keep every component function mean centered, backfitting requires to impose the constraints $ \function_{\mathbf{a}, 0} = \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n) $ (Line~\ref{alg_line:initialization}), which is the mean of $ \function_{\mathbf{a}} $ evaluated on samples belonging to the sensitive group $ \sensitive = \mathbf{a} $;  and $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}) = 0$ (Line~\ref{alg_line:mean_centered}), where $ \mathbf{x}_{\mathbf{S}}^{(i)} $ be the subset of feature values associated with feature indices $ \mathbf{S} $ for the $ i $-th sample $ \mathbf{x}^{(i)} $. These constraints assign the expectation of $ \function_{\mathbf{a}} $ on the constant term $ \function_{\mathbf{a}, 0} $ and the variance of $ \function_{\mathbf{a}} $ to the component functions.

While performing local regression, backfitting uses a smoothing operator~\cite{loader2012smoothing} over the set of samples (Line~\ref{alg_line:backfitting_step}). A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating the local curves obtained by local regression~\cite{loader2012smoothing}. In this paper, we apply cubic spline smoothing~\cite{li2010global} to learn each component function. Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity. Hence, up to the second derivatives of each piecewise term are zero at the endpoints of intervals. We refer to Appendix~\ref{sec:smoothing} for further details.%\todo{Add pseudo-code of {\fairXplainer}}

\begin{comment}
\begin{algorithm}[!t]
	\caption{Local Regression with Backfitting Algorithm}
	\label{algo:backfitting_algorithm}
%	\hspace*{\algorithmicindent} \textbf{Input:} $ g_\mathbf{a}(\mathbf{X}), \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda $ \\
%	\hspace*{\algorithmicindent} \textbf{Output:} $ \function_{\mathbf{a}, 0}, \function_{\mathbf{a}, \mathbf{S}} $ 
	\begin{algorithmic}[1]
		\Procedure{LocalRegression}{$ g_\mathbf{a}, \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k $}
		\State \textbf{Initialize}: $ \function_{\mathbf{a}, 0} \leftarrow \frac{1}{n}\sum_{i=1}^n\function_{\mathbf{a}}(\mathbf{x}^{(i)}), \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [k]\setminus\emptyset, |\mathbf{S}|\le \lambda$ \label{alg_line:initialization}
		\While{each $ \widehat{\function}_{\mathbf{a},\mathbf{S}} $ does not converge}
		\For{each $\mathbf{S}$}
		\State $ \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow \mathsf{Smooth}[\{\function_{\mathbf{a}}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\function}_{\mathbf{a}, \mathbf{S}'}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1}^n] $\label{alg_line:backfitting_step} \Comment{Backfitting step}
		\State $\widehat{\function}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\function}_{\mathbf{a}, \mathbf{S}} - \frac{1}{n}\sum_{i=1}^n \widehat{\function}_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})$ \label{alg_line:mean_centered} \Comment{Mean centering}
		\EndFor
		\EndWhile
		\State \Return $ \function_{\mathbf{a}, 0}, \function_{\mathbf{a}, \mathbf{S}} $ 
		\EndProcedure
	\end{algorithmic}
\end{algorithm}
\end{comment}

\noindent\textbf{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $. Since each component function is mean centered (Line~\ref{alg_line:mean_centered}), we compute the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ on the dataset as
$  \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n (\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}))^2 $. Hence, variance captures the independent effect of $ \function_{\mathbf{a}, \mathbf{S}} $.
%$ V_{\mathbf{a}, \mathbf{S}} $ is the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ evaluated on each $ \mathbf{x}_{\mathbf{S}}^{(i)} $ in the dataset. 
Covariance is computed to account for the correlation among features $ \nonsensitive $. We compute the covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ with $ \function_{\mathbf{a}} $ on the dataset as
$
\label{eq:covariance_computation}
	 \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}). 
$
Here, $ g_\mathbf{a}(\cdot) - \function_{\mathbf{a}, 0} $ is the mean centered form of $ \function_{\mathbf{a}} $. Covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ can be both positive and negative depending on whether the features $ \nonsensitive_{\mathbf{S}} $ are positively or negatively correlated with $ \function_{\mathbf{a}} $. Specifically, under the set additive model, we obtain $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $. Now, we use $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] $ as the effective variance of $ \nonsensitive_{\mathbf{S}} $ for a given sensitive group $ \sensitive = \mathbf{a} $ (Line~\ref{algo_line:covariance_computation_start}--\ref{algo_line:covariance_computation_end}). In Line~\ref{algo_line:fif_computation_start}--\ref{algo_line:fif_computation_end}, we compute $ V_{\mathbf{a}, \mathbf{S}} $ for the most and the least favored groups, and plug them in Theorem~\ref{lemma:fif} to compute FIF of $ \nonsensitive_{\mathbf{S}} $.

%In both variance and covariance decomposition, $ \mathsf{Var}[\function_{\mathbf{a}}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} V_{\mathbf{a}, \mathbf{S}}$. The difference between variance and covariance decomposition is that $ V_{\mathbf{a}, \mathbf{S}} \in \mathbb{R}^{\ge 0} $ is positive in  variance decomposition, while $ V_{\mathbf{a}, \mathbf{S}} \in \mathbb{R} $ may also be negative in covariance decomposition. In covariance decomposition, we can separate between the structural and correlative effects of $ \function_{\mathbf{a}, \mathbf{S}} $. Formally, $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S}' \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}'| \le \lambda, \mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $, where  to the right side of equality, first term refers to the structural effect and the second term refers to correlative effect of $ \function_{\mathbf{a}, \mathbf{S}} $.

\noindent\textbf{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
We extend {\fairXplainer} in computing FIF of group fairness metrics beyond statistical parity, namely equalized odds and predictive parity. For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of samples in the dataset where $ Y = 1 $ and another on samples with $ Y = 0 $. Then, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ quantifies the equalized odds of the classifer.

To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $. For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label. This contrasts the computation for  statistical parity and equalized odds, where the predicted class $ \widehat{Y} $ is considered as label. Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ quantifies the predictive parity of the classifier.

### 2022-06-01 chapters/fairness/fif/figure_intro.tex
adding fifs and introduction

\begin{figure}[t!]
	\begin{minipage}[t]{0.13\textwidth}			
		\scalebox{1}{	
		\begin{tikzpicture}[x=1cm,y=0.3cm]
		\node[] (a1) {\textbf{Data:}};			
		\end{tikzpicture}
		}
	\end{minipage}%
	\begin{minipage}{0.35\textwidth}
		\centering
		\subfloat[Dependency among features and prediction]{
			\scalebox{0.4}{	
				\begin{tikzpicture}[x=1.5cm,y=0.3cm]
				% Define nodes
				\node[latent,scale=2] (a1) {$\textrm{age}$} ; %
				\node[obs, scale=2, below=of a1, xshift=-2cm] (h) {$\textrm{income}$}; %
				\node[obs, scale=2, below=of a1, xshift=2cm] (i) {$\textrm{fitness}$}; %
				\node[obs, scale=2, below=of h, xshift=2cm] (p) {$\widehat{Y}$}; %			
				%%add edge
				\edge[] {a1} {h,i} ;
				\edge[] {h,i} {p} ;
				\end{tikzpicture}
			}	
			\label{fig:dag_age_income_fitness}}
	\end{minipage}%
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Age-dependent distributions of non-sensitive features]{\includegraphics[scale=0.4]{figures/fairness/fif/sanity_distribution}
		\label{fig:distribution_example}}
	\end{minipage}

	\begin{minipage}[t]{0.13\textwidth}
		\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.3cm]
			\node[] (a1) {\textbf{Classifier:}};			
			\end{tikzpicture}
		}
	\end{minipage}%
	\begin{minipage}{0.43\textwidth}
		\centering
		\subfloat[Decision tree (DT$ 1 $)]{
			\scalebox{0.45}{	
				\begin{tikzpicture}[x=1cm,y=1.8cm]
				\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
				\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
				\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
				\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
				%
				\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
				\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
				\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
				\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
				\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
				\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
				\end{tikzpicture}}
			\label{fig:dt_original}}
	\end{minipage}%
	\begin{minipage}{0.45\textwidth}
		\centering
		\subfloat[Decision tree with affirmative action (DT$ 2 $)]{
			\scalebox{0.45}{	
				\begin{tikzpicture}[x=1cm,y=1.8cm]
				\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
				\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm, fill=affirmative]     (a2)    {income $\geq 0.555$};
				\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
				\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
				\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
				%
				\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
				\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
				\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
				\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
				\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
				\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
				\end{tikzpicture}}

			\label{fig:dt_affirmative}}
	\end{minipage}

	\begin{minipage}[t]{0.07\textwidth}
		\vspace{-2em}
		\scalebox{1}{	
			\begin{tikzpicture}[x=1cm,y=0.3cm]
			\node[] (a1) {\textbf{FIF:}};			
			\end{tikzpicture}
		}
	\end{minipage}%
	\begin{minipage}{0.46\textwidth}
		\centering
		\subfloat[Fairness influence functions (FIF) for DT$ 1 $]{\includegraphics[scale=0.45]{figures/fairness/fif/fif_example}
		\label{fig:fif_original}}	
	\end{minipage}%
	\begin{minipage}{0.5\textwidth}
		\centering
		\subfloat[Modified FIFs for DT$ 2 $]{\includegraphics[scale=0.45]{figures/fairness/fif/fif_example_affirmative_action}
		\label{fig:fif_affirmative}}
	\end{minipage}

	\caption{FIFs of input features to investigate the bias (i.e., statistical parity) of a decision tree classifier predicting the eligibility for health insurance using age-dependent features `fitness' and `income'. An affirmative action reduces bias, and corresponding FIFs present the evidence.}
	\label{fig:fair_example_fif}
	\vspace{-1em}
\end{figure}

### 2022-06-01 chapters/fairness/fif/framework.tex
adding fifs and introduction

%\clearpage
\section{Fairness Influence Functions: Definition and Computation}\label{sec:fifs}
In this section, we discuss our contribution of formalizing and computing Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier on a dataset. The key idea in the formalization of FIFs is their additive formulation. The key idea in the computation of FIFs is to relate bias with the scaled difference of the conditional variance of positive prediction of the classifier over specific sensitive groups, and to apply variance decomposition to compute individual and intersectional FIFs. %\red{In this section, we first formalize fairness influence functions, and then discuss a variance decomposition based approach to compute FIFs for group fairness metrics.}\todo{Structure: 1. Background (measuring bias using fairness metrics) 2.FIFs as a measure of feature bias 3. Reducing FIF estimation to variance decomposition 4. An algorithm to do the variance decomposition}

%in Section~\ref{??}, and conclude by extending influence computation to other group fairness metrics in Section~\ref{??}
%\subsection{An Axiomatic Formulation}
%\label{sec:problem_formulation}
\noindent\textbf{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $. Our objective is to compute the influences of non-sensitive features on $ f $. We compute influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.

\begin{definition}[Fairness Influence Function]
Fairness Influence Function (FIF) $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$. We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $\textemdash particularly, a second-order influence when $ |\mathbf{S}| = 2 $ and a higher-order influence when $ 3 \le |\mathbf{S}| \le \numnonsensitive $. %\unsure{What do we mean by function here? Isn't FIF scalar?}
\end{definition}	
Since FIF is meant to attribute the bias of the classifier to the subset of features, it is natural to choose a function such that the FIFs of all the subsets of features exactly adds up to the total bias. This leads to the following additivity axiom.
%{\fairXplainer} is an axiomatic approach to relate individual and inter-sectional influences with associated fairness metrics, which we formalize in the following.
\begin{axiom}[Additivity of influence]
	\label{axm:additivity}
	Let $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$ be the bias of the classifer and $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ be the FIF of features $ \mathbf{X}_{\mathbf{S}} $ on $ f $. The additivity axiom states that the total influences of all possible subset of non-sensitive features is equal to the bias of the classifier.
	\begin{align}\label{eq:constraint_framework}
		f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
	\end{align}
\end{axiom}
\noindent\textbf{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}. In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.

%\begin{comment}
%	\[
%	\mathsf{Var}_v[v] = \mathsf{Var}[v]
%	\]
%\end{comment}

\noindent\textbf{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction. Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $. Then, the probability of the random variable is $ p_\mathbf{a} \triangleq \Pr[B_{\mathbf{a}} = 1] = \Pr[\widehat{Y} = 1| \sensitive = \mathbf{a}] $, and variance is $ V_{\mathbf{a}} = p_{\mathbf{a}}(1-p_{\mathbf{a}}) $. Therefore, considering two random variables $ B_{\mathbf{a}_{\max}} $ and $ B_{\mathbf{a}_{\min}} $, we state the statistical parity of the classifier as $ p_{a_{\max}} - p_{a_{\min}} = (V_{a_{\max}} - V_{a_{\min}})/(1  - (p_{a_{\max}} + p_{a_{\min}})) $, where $ p_{a_{\max}} $ and $ p_{a_{\min}} $ is the probability of positive prediction of the classifier for the group $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively. Intuitively, the statistical parity of the classifier is the scaled (by $ 1/(1  - (p_{a_{\max}} + p_{a_{\min}})) $) difference of the conditional variance of the positive prediction of the classifier over the most and the least favored sensitive groups.

\begin{example}
	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively. Thus, the statistical parity of the classifier is $ p_{a_{\max}} - p_{a_{\min}} = 0.704 - 0.172 =  0.532 $. Now, we compute conditional variances of positive prediction as $ V_{a_{\max}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } < 40] = 0.208 $ and $ V_{a_{\min}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } \ge 40] = 0.142 $. Thus, following our argument, we compute the scaled difference in conditional variances as $ (0.208 - 0.142)/(1 - (0.702 + 0.172)) = 0.532 $, which coincides with the statistical parity.
\end{example}
Now, we apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}) for both the conditional variances $ V_{\mathbf{a}_{\max}} $ and $ V_{\mathbf{a}_{\min}} $ to compute the FIF of features, as stated in Theorem~\ref{lemma:fif}.
\begin{theorem}[FIF as Difference of Conditional Variances]
	\label{lemma:fif}
	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied. Then, we can express FIF of $ \nonsensitive_{\mathbf{S}} $ as
	\begin{equation}\label{eq:fif_decompose}
	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])}
	\end{equation}
	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
\end{theorem}
For brevity of space. we defer the detailed proof to Appendix~\ref{sec:proof_1}.

\noindent\textbf{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.

1. \textit{Algorithm Design.} Expressing FIF in terms of the variance decomposition over subset of features allows us to import and extend well-studied techniques of global sensitivity analysis to perform FIF estimation accurately and at scale.
%To prove Lemma~\ref{lemma:fif}, we present that the statistical parity of a classifier is equal to the scaled difference in the conditional variance of positive prediction of the classifier over specific sensitive groups. Later, we apply variance decomposition to derive the expression for $ w_{\mathbf{S}} $.

2. \textit{Bias Amplifying and Eliminating Features.} The sign of a FIF $ w_{\mathbf{S}} $ denotes whether the features $ \nonsensitive_{\mathbf{S}} $ are amplifying the bias of the classifier or eliminating it. When $ w_{\mathbf{S}} > 0 $, $ \nonsensitive_{\mathbf{S}} $ increases bias. When $ w_{\mathbf{S}} < 0 $, $ \nonsensitive_{\mathbf{S}} $ eliminates bias, and thus improves fairness. The features $ \nonsensitive_{\mathbf{S}} $ are neutral when $ w_{\mathbf{S}} = 0 $.

3. \textit{Degenerate Cases:} If the classifier always makes positive prediction or never makes positive prediction, i.e. $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{1, 0\} $ for any sensitive group $ \sensitive = \mathbf{a} $, the conditional variance $ \mathsf{Var}[\widehat{Y} = 1 | \sensitive = \mathbf{a}]  = 0 $. When total variance is zero, the decomposed variances $V_{\mathbf{a},\mathbf{S}}$ may become zero for each feature combination $\nonsensitive_\mathbf{S}$, and Equation~\eqref{eq:fif_decompose} cannot trivially compute fairness influence functions. Though such degenerate cases rarely occur for real-life data, we can avoid them by randomly perturbing at least one sample in the dataset to yield the opposite predicted class.

4. \textit{Numerical Instability.} In Equation~\eqref{eq:fif_decompose}, the conditional probabilities of positive prediction of the classifier $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}]  $ and $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]  $ are constant. When $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]   = 1 $, the denominator in $ w_{\mathbf{S}} $ becomes $ 0 $ and hence, $ w_{\mathbf{S}} $ is undefined. In this case, we add/subtract a small number in the denominator to avoid numerical instability of division though we have not encountered any such case in our experiments.%\red{However, the interpretation of influence functions such as whether $ \nonsensitive_{\mathbf{S}} $ improve or worsen fairness becomes inconsistent due to the choice of addition/subtraction.}

### 2022-06-01 chapters/fairness/fif/introduction.tex
adding fifs and introduction

\section{Introduction}
%\begin{itemize}
%	\item Bias -> fairness metric 
%	\item Fairness influence function (FIF)
%	\item Example
%	\item Contribution (1) framework (2) algorithm (3) experimental validation
%	\item Related works.
%	\item Notation paragraph
%\end{itemize}
The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc. In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}. For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}. Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset. For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.

Although fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}. In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}. To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier. Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias. FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias.

\noindent\textbf{Our Contributions.} The contribution of this paper is three-fold.

\textit{1. Formalism:} We propose to compute individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier (Sec~\ref{sec:fifs}). The \textit{intersectionality}~\cite{buolamwini2018gender} allows us to detect the higher order interactions among the features. In the formalism, we first axiomatize that to be a proper attribution of bias, the sum of FIFs of all subsets of the features is equal to the bias of the classifier (Axiom~\ref{axm:additivity}). Following that, we show that computing FIFs over subsets of features is equivalent to computing a scaled difference in the \emph{conditional variances} (Theorem~\ref{lemma:fif}) of the classifier between the sensitive groups of interest. This allows us to connect global sensitivity analysis, a standard technique recommended by regulators to asses numerical models~\cite{eu,usepa}, with computing feature influences leading to bias in the classifier's output.%\improvement{Application of GSA in EU and modeling. }

\textit{2. Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for computing FIFs for subsets of features for a given classifier, a dataset, and any group fairness metrics, namely statistical parity, equalized odds, and predictive parity (Sec.~\ref{sec:fairxplainer}).  The key ideas are to import techniques from Global Sensitivity Analysis (GSA)~\cite{saltelli2008global} to decompose the variance of the prediction of the classifier among the subset of features and to use a local regression method~\cite{loader2006local} to compute FIFs.

\textit{3. Experimental:} We experimentally validate that {\fairXplainer} is significantly more accurate and can capture the intersectional or joint effect of the features on the bias induced by a classifier (Sec.~\ref{sec:experiments}). Improvement in accuracy and extension to intersectional FIFs are both absent in the existing works aimed to this problem~\cite{begley2020explainability,lundberg2020explaining}. Also, {\fairXplainer} enables us to detect the change in FIFs due to different fairness enhancing algorithms and fairness reducing attacks, which opens up new avenues to analyze the effects of these algorithms.

We illustrate the usefulness of our contributions via an example scenario proposed in~\cite{ghosh2020justicia}.

%\dbcomment{In recent times, we have observed growing interest to explain sources of unfairness in ML algorithms. Specifically, [a,b] tried to transfer the existing tools from interpretability literature to quantify and explain fairness. But these works mostly demonstrate empirical evaluations to justify the choices of interpretability tools and does not consider the global nature of fairness metrics. In this paper, we develop a formal framework to explain sources of unfairness in an ML algorithm and also a novel methodology to address it. To the best of our knowledge, this is the first work to do so.}

 %\red{This allows practitioners to deploy various affirmative or punitive actions.}\improvement{Examples?}
\begin{example}
	\label{ex:motivating_example}
 Following~\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features `fitness' and `income'.
	`fitness' and `income' depend on a sensitive feature `age' leading to two sensitive groups ``young" (age $ < 40 $) and ``elderly" (age $ \ge 40 $), as highlighted in (Figure~\ref{fig:dag_age_income_fitness}--\ref{fig:distribution_example}).

	\textbf{Case study 1:} For each sensitive group, we generate $ 500 $ samples of (income, fitness) and train a decision tree (DT$ 1 $), which predicts without explicitly using the sensitive feature `age' (Figure~\ref{fig:dt_original}). %This classifier predicts the eligibility for health insurance without any direct effect of the sensitive feature: age. 	
	Using standard off-the-shelf techniques, we can compute statistical parity as $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.7 - 0.17 = 0.53 $, and therefore DT$ 1 $ is unfair towards ``elderly".

	Applying techniques developed in this paper, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}. In Figure~\ref{fig:fif_original}, fitness $(\mathrm{FIF} = 0.74)$, and the joint effect of fitness and income $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e. higher bias. This observation is invoked as DT$ 1 $ predicts positively for the higher values of fitness (root node in DT$ 1 $) and elderly individuals often possess lower fitness (Figure~\ref{fig:distribution_example}). In contrast, the income of individuals  $(\mathrm{FIF}  = -0.33)$ decreases statistical parity, as elderly individuals have a higher income but lower fitness.

	\textbf{Case study 2:} Since DT$ 1 $ is unfair to the ``elderly", we learn another decision tree (DT$ 2 $) by applying an affirmative action, where we decrease the threshold on income from $ 0.69 $ to $ 0.555 $ when the fitness is lower ({\color{affirmative}green} node in Figure~\ref{fig:dt_affirmative}). This action allows more elderly individuals to receive insurance by including more people with lower fitness, and thus the statistical parity becomes $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.71 - 0.7 = 0.01$. This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this paper, we compute FIF of features in Figure~\ref{fig:fif_affirmative}. Here, the FIF of income and fitness reflects the reduction in statistical parity as their influences almost nullify each other. Thus, FIF depicts the effects of different features and their combinations on the resultant bias incurred by the classifier. %As we analyse the source of unfairness, the combined influence of income and fitness changes its direction ($ - 0.01 $ vs.\ $ 0.05 $) towards improving fairness.
\end{example}

\input{chapters/fairness/fif/figure_intro.tex}

\section{Related Work} Recently, several studies apply \emph{local explanation methods} of black-box prediction to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}. Since our work adopts feature-attribution approach, it reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in computing FIFs and (ii) \emph{failing to compute intersectional} FIFs. Elaborately, FIF computation in~\cite{begley2020explainability,lundberg2020explaining} is inaccurate because of applying local explainers such as SHAP~\cite{lundberg2017unified} to compute global properties of the classifier such as group fairness. In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the unfairness of the classifer. Also, these works provide empirical evaluations to justify the choices of SHAP-based tools for explaining fairness and does not consider the global nature of group fairness metrics. In this paper, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it. To the best of our knowledge, this is the first work to do the both}. %\todo{1. GSA with fairness, 2. Citation not correct, 3. Influence of Yair Zick.}
Among other related works, \cite{benesse2021fairness} link GSA measures such as Sobol and Cram{\'e}r-von-Mises indices to different fairness metrics. While their approach relates the GSA of sensitive features on the resulting bias, we focus on applying GSA to all features to compute FIFs. \textit{Their approach only detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all features}. In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} compute feature-influence as the shift of bias from its original value by randomly intervening features. Their work is different from our axiomatic approach, where the sum of FIFs equals the bias.

### 2022-06-01 chapters/fairness/fif/preliminaries.tex
adding fifs and introduction

%\clearpage
\section{Background: Fairness and Global Sensitivity Analysis}\label{sec:preliminaries}\vspace*{-.5em}
Before proceeding to the details of our contribution, we present the fundamentals of group fairness metrics as quantifiers of bias and global sensitivity analysis as a classical method of feature attribution.\vspace*{-.5em}
\subsection{Fairness in Machine Learning: Fairness Metrics}
We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ 
 generated from an underlying distribution $\mathcal{D}$. Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $\{\mathbf{x}^{(i)}_1, \dots, \mathbf{x}^{(i)}_{\numnonsensitive}\} $. Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $\{\mathbf{a}^{(i)}_1, \dots, \mathbf{a}^{(i)}_{\numsensitive}\} $.  $y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$. %Each non-sensitive feature $ X_i$ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each sensitive feature $ A_j \in \{0, \dots, N_j\}  $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}. 
 We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.  
%\blue{For sensitive features, a vector $ \mathbf{a} = [a_1, .., a_{\numsensitive}] $ of categorical values is called a \textit{compound sensitive group}. For example, consider $ \sensitive = $ \{race, sex\}, where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}. Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group.}\unsure{blue part is not needed} 
 We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $. Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
Given this setup, we discuss different fairness metrics to compute bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.  %\red{We use $ f(\alg, \mathbf{D}) $ to denote a fairness metric computed for a classifier $ \alg $ on the dataset $ \mathbf{D} $}. %\blue{We certify a classifier $ \epsilon $-fair, for a constant $ \epsilon \in [0,1] $, by computing $ \mathds{1}(f(\alg, \mathbf{D}) \le \epsilon) $. Here, $ \mathds{1} $ is an indicator function evaluating to $ 1 $ whenever the input is true, and $ 0 $ otherwise.}\info{Not necessary!} 
\begin{enumerate}[leftmargin=*]
	%		\itemsep0em 
	\item \textit{Statistical Parity} ($ \mathsf{SP} $)~\cite{feldman2015certifying}: Statistical parity belongs to \textit{independence} measuring group fairness metrics, where the prediction $ \widehat{Y} $ is statistically independent of sensitive features $ \sensitive $.  The statistical parity of  a classifier is measured as $ f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] $, which is the difference between the maximum and minimum conditional probability of positive prediction the classifier for different sensitive groups.

	\item \textit{Equalized Odds} ($ \mathsf{EO} $)~\cite{hardt2016equality}: \textit{Separation} measuring group fairness metrics such as equalized odds constrain that $ \widehat{Y} $ is independent of $ \sensitive $ given the ground class $ Y $.  Formally, for $ Y \in \{0,1\} $, equalized odds is $ f_{\mathsf{EO}}(\alg, \mathbf{D})  \triangleq \max(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1])$.

	\item \textit{Predictive Parity} ($ \mathsf{PP} $)~\cite{verma2018fairness}: \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $. Formally, $ f_{\mathsf{PP}}(\alg, \mathbf{D})  \triangleq \max(\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1])$. % \textit{Conditional Use Accuracy Equality (CUAE):}
%	\item \blue{\textit{Path-specific Causal Fairness} (PCF): 
%		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 |\mathbf{A}=  \mathbf{a}] $ be the sensitive group with maximum conditional probability of positive prediction of the classifier. We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $. This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.  To this end, the path-specific causal fairness of a classifier is $ f(\alg, \mathbf{D}) = \max_{\mathbf{a}}\Pr[\widehat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ] $.}\change{How to sample $ \mathbf{Z} $?}
\end{enumerate}
All of the aforementioned group fairness metrics depend on the difference between different conditional probabilities of positive prediction of a classifier.
For all of these metrics, lower value of $ f(\alg, \mathbf{D}) $ indicates higher fairness demonstrated by the classifier $\mathcal{M}$. \textit{We deploy these fairness metrics as the measures of bias of a classifier w.r.t\ a given dataset.}%\info{Not necessary!}

\subsection{Global Sensitivity Analysis: Variance Decomposition}
Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
%\paragraph{Sensitivity Analysis:}
\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}. To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
% and $ Y $ is a univariate output. 
%\blue{In sensitivity analysis, assumptions are made on the independence and uniform distribution of $ X_i $'s within the unit hypercube}, that is, $ X_i \in [0,1] $ for $ i = \{ 1, \dots, k\} $.\footnote{In our experiments, we have not considered this yet.} Let $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $. Then, 
Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
\begin{align}
	\function(\mathbf{Z}) &= \function_0 + \sum_{i=1}^{\numnonsensitive} \function_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \function_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \function_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
	\\    
	&= \function_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
\end{align}
%	$S_> = \{{S_0}, \ldots, {S_{|s|}}\}$ i.e. a monotonically increasing ordering of $|S|$ features for a given starting feature $Z_{S_0}$ such that $S_0 <S_1 <\ldots <S_{|S|}$.
In this decomposition, $ \function_0 $ is a constant, $ \function_{\{i\}} $ is a function of $ Z_i $, $ \function_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on. Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.  We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \function_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.  The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}. $ \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously. %\unsure{Is this sentence necessary?}  
For $|\mathbf{S}|=1$, it quantifies an individual variable's effect. For $|\mathbf{S}|>1$, it quantifies the higher-order interactive effect of variables.
\iffalse
\begin{align}
	\label{eq:orthogonality_in_decomposition}
	\int_{0}^{1} \function_{\mathbf{S}}(\mathbf{X}_{\mathbf{S}})dX_i = 0, \text{ for } i \in \mathbf{S} = \{S_1, S_2, \dots, S_{|\mathbf{S}|}\}
\end{align}\info{let's discuss this}

This orthogonality constraint leads to the definitions of each term expressed as integrals (equivalently, the conditional expected values) of $ Y = f(\mathbf{X}) $.

\begin{align*}
	&\int f(\mathbf{X})\prod_l dX_l =  \mathsf{E}[Y] = \function_0\\
	&\int f(\mathbf{X})\prod_{l \ne i} dX_l =  \mathsf{E}[Y|X_i] = \function_{\{i\}}(X_i) + \function_0\\
	&\int f(\mathbf{X})\prod_{l \ne i, j} dX_l = \mathsf{E}[Y|X_i, X_j] = \function_{\{i,j\}}(X_i, X_j)  +  \function_{\{i\}}(X_i) +  \function_{\{j\}}(X_j) + \function_0 \\
	&\vdots
\end{align*}

Thus, $ \function_{\{i\}} $ is the effect of varying $ X_i $ alone and it is known as the \emph{main effect} of $ X_i $. Similarly, $ \function_{\{i,j\}} $ is the effect of varying both $ X_i $ and $ X_j $ simultaneously, additional to their effect of individual variations. $ \function_{\{i,j\}} $ is thus known as the \emph{second-order interaction} between $ X_i $ and $ X_j $. Higher order interactions have analogous interpretations.

Now, we assume $ f $ to be square integrable. Then each term $ \function_{\mathbf{S}} $ in Eq.~\eqref{eq:functional_decomposition_set_notation} are also square integrable. Squaring Eq.~\eqref{eq:functional_decomposition_set_notation}, integrating over $  [0,1]^\numnonsensitive$, and applying orthogonality constraint in Eq.~\ref{eq:orthogonality_in_decomposition}, we get
\begin{align*}
	\int f^2(\mathbf{X})\prod_l dX_l - \function_0^2 = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset}  \function_{\mathbf{S}}^2(\mathbf{X}_{\mathbf{S}}) \prod_{S_i \in \mathbf{S}} d Z_{S_i}
\end{align*}

Here, the left hand side is equal to the variance of $ Y = f(\mathbf{X}) $, and the terms in the right hand side are variance terms decomposed with respect to  $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X} $. 
\fi
Now, if we assume $ \function $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
\begin{align}\label{eq:variance_decomposition_set_notation}
	\mathsf{Var}[g(\mathbf{Z})] &= \sum_{i=1}^{\numnonsensitive}V_{\{i\}} +  \sum_{i<j}^{\numnonsensitive} V_{\{i,j\}}  + \cdots  + V_{\{1, 2\dots, \numnonsensitive\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}} 
\end{align}
where $ V_{\{i\}} $ is the variance of $ \function_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \function_{\{i,j\}} $ and so on. Formally,  $V_{\mathbf{S}} \triangleq \mathsf{Var} _{\mathbf{Z}_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}/\mathbf{Z}_{\mathbf{S}}}[g(\mathbf{Z})\mid \mathbf{Z}_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}/\emptyset}{V} _{\mathbf{S}'}$. Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$. %\improvement{Check notations.}
This variance decomposition shows how the variance of $\function(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them. Together all terms sum to the total variance of the model output. 
\textit{We reduce the problem of computing FIFs of subsets of features into a variance decomposition problem.}

\iffalse

Formally, $V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y | X_i]]$ and $V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y | X_i, X_j]] - V_{\{i\}} - V_{\{j\}}$
\begin{align*}
	& V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y | X_i]]\\
	& V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y | X_i, X_j]] - V_{\{i\}} - V_{\{j\}}%\\
	%& \vdots
\end{align*}
$ \mathbf{X}_{\sim \mathbf{S}} \triangleq \mathbf{X} \setminus \mathbf{X}_{\mathbf{S}} $ notation indicates the set of all variables except that are in  $ \mathbf{X}_{\mathbf{S}} $. Thus, in each definition of $ V_{\mathbf{S}} $, the expectation of $ Y $ is computed on variables in $ \mathbf{X}_{\sim \mathbf{S}} $ and then, the variance of expectations is computed on variables in $ \mathbf{X}_{\mathbf{S}} $.

This variance decomposition shows how the variance of $\function(\mathbf{X}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them. Together all terms sum to the total variance of the model output. 
\textit{We reduce the problem of computing FIFs of subsets of features into a variance decomposition problem.}
%	In the rest of the manuscript, we refer $ V_i $ as the first order variance of $ X_i $, $ V_{ij} $ as the second order variance of $ X_i $ and $ X_j $, and so on.

\paragraph{Law of total variance:}

if $ X $ and $ Y $ are random variables in the same probability space, and the variance of $ Y $ is finite, then law of total variance states the following.

\begin{align*}
	\mathsf{Var}[Y] = \mathsf{E}_X[\mathsf{Var}_Y[Y|X]] + \mathsf{Var}_X[\mathsf{E}_Y[Y|X]]
\end{align*}

Law of total variance is useful when we can compute the conditional variance/expectation of $ Y $ given $ X $ instead of the variance of $ Y $. 
\fi

### 2022-06-01 chapters/introduction.tex
adding fifs and introduction

@@
 The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-In such applications, the deployed  classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}.
+In such applications, the deployed  classifier often demonstrate bias towards certain sensitive demographic groups involved in the data~\cite{dwork2012fairness}.
 For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.
@@
 \subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features.
+The early works on fairness verification focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}.
+Naturally, such techniques were limited in enhancing confidence of users for wide deployment.
+Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare, bastani2019probabilistic}.
+More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.
+
+In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula.
+Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
+Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.
+sensitive features and ignore correlations among non-sensitive features.
+As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.
+
+In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers.
+Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
+{\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.
+Because SSAT encoding allows separate quantification to each sensitive feature without any restriction on the number of features.
+In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}.
+We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.
+
+Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+We extend probabilistic fairness verification to linear classifiers by proposing a tractable verification framework in contrast to the existing $ \mathrm{NP}^{\mathrm{PP}} $-hardness of SSAT-based verification in {\justicia}.
+We also improve the accuracy of verification by considering feature correlations.
+Specifically, we propose a fairness verification framework for linear classifiers, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurate and scalable fairness verification.
+{\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
+To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
+Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
+We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) \red{computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.}
 
 \subsection*{Identification of Sources of Unfairness}
+While fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}.
+To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
+Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias.
+FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias.
+We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features.
+Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.
 
 \section{Outline}
+
+\begin{comment}
 
@@
 Our hypothesis is that such a combination would be effective in identifying the bias of a model and also mitigate the bias through imperative actions.
+
+\end{comment}

### 2022-06-01 chapters/fairness_background.tex
upto preliminaries of part 1

\part{Fairness in Machine Learning}
\chapter{Preliminaries}

### 2022-06-01 chapters/interpretability_background.tex
upto preliminaries of part 1

--- 2022-01-27 chapters/interpretability/imli/preliminaries.tex
@@
-\section{Preliminaries}
-\label{interpretability_imli_sec:preliminaries}
+\part{Interpretable Machine Learning}
 
-In this section, we first discuss propositional satisfiability and MaxSAT, followed by rule-based classification in machine learning.
-Throughout the paper, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters $\mathbf{x}$ are reserved for vectors/sets.
+\chapter{Preliminaries}
 
-\subsection{Propositional Satisfiability}
+%\section{Preliminaries}
+%\label{interpretability_imli_sec:preliminaries}
+
+In this chapter, we discuss preliminaries for learning interpretable rule-based classifiers.
+Throughout this part of the thesis, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters $\mathbf{x}$ are reserved for vectors/sets.
+
+\subsection*{Propositional Satisfiability}
 
@@
 Informally, $ \sigma^* $ satisfies at least one literal in each clause of a CNF.
-%\red{Moreover, $ F $ is called SAT if there exists a satisfying assignment of $ F $, and $ F $ is called UNSAT otherwise.}
 
-In this paper, we use true as $ 1 $ and false as $ 0 $ interchangeably.
+\subsection*{Relaxation of Logical Formulas}
+
+A hard-OR or a soft-AND clause is a tuple $ (C,\eta) $ with an extra parameter $ \eta $, where $ \eta $ is the threshold on the literals in $ C $.
+Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $.
+Motivated by \red{Boolean cardinality constraints}, we propose relaxed-CNF formulas, where in addition to $F$, we have two more parameters $\eta_c$ and $\eta_l$.
+We say that $(F,\eta_c, \eta_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (F,\eta_c, \eta_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (C_i,\eta_l)] \ge \eta_c$, where $\sigma^* \models (C_i, \eta_l)$ if and only if $\sum_{v \in C_i} \mathds{1}[\sigma^* \models v] \geq \eta_l$.
+Informally,  $\sigma^*$ satisfies a clause $(C_i, \eta_l)$ if  at least $\eta_l$ literals in $C_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(F,\eta_c, \eta_l)$ if at least $\eta_c$  clauses out of all $\{(C_i,\eta_l)\}$ clauses are true.
+
+\begin{theorem}[\cite{benhamou1994two}]
+	\label{interpretability_crr_thm:succinctness}	
+	Let $ (C,\eta) $ be a clause in relaxed-CNF where  $ C $ has $ m  $ literals and $ \eta \in \{1,\dots,m\} $ is the threshold on literals.
+An equivalent compact encoding of $ (C,\eta) $  into a CNF formula $ F= \bigwedge_i C_i $ requires $ {m \choose{m-\eta+1}} $  clauses where each clause is distinct and has $ m-\eta+1 $ literals of $ (C,\eta) $.
+Therefore,  the total number of literals in $ F $ is $ (m-\eta+1){m\choose{m-\eta+1}} = m $ when $ \eta=1 $, otherwise $ (m-\eta+1){m\choose{m-\eta+1}} > m $.
+\end{theorem}
+
+\subsection*{Inner Product}
+In this part of the thesis, we use true as $ 1 $ and false as $ 0 $ interchangeably.
 Between two vectors  $\mathbf{u}$ and $\mathbf{v}$ of the same length defined over Boolean variables or constants (such as  $ 0 $ and $ 1 $), we define $\mathbf{u} \circ \mathbf{v} $  to refer to  their inner product.
@@
 For example, if $ \mathbf{u} = [u_1, u_2, u_3] $ and $ \mathbf{v} = [0,1,1] $, then $ \mathbf{u} \circ \mathbf{v} = u_2 \vee u_3 $.
+Applying numerical interpretation, the inner product can also be expressed as $\mathbf{u} \circ \mathbf{v} \triangleq \sum_{i} (u_{i} \cdot v_{i}) $.
+Hence, for the running example $ \mathbf{u} = [u_1, u_2, u_3] $ and $ \mathbf{v} = [0,1,1] $, we derive $ \mathbf{u} \circ \mathbf{v} = u_2  + u_3 $.
+In this part of the thesis, we use Boolean interpretation of the inner product for learning CNF classification rules and numerical interpretation for learning relaxed-CNF classification rules.
 
-\subsection{MaxSAT}
+\subsection*{MaxSAT}
 
 The MaxSAT problem is an optimization analog to the SAT problem that finds an optimal assignment satisfying the maximum number of clauses in a CNF.
-In this work, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula.
+In interpretable machine learning, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula.
 In a weighted-CNF formula, a weight $ {\weight{C_i}} \in \mathbb{R}^+ \cup \{\infty\} $ is defined over each clause $ C_i $, wherein $C_i$ is called a \emph{hard} clause if $\weight{C_i} = \infty$, and  $C_i$ is  a \emph{soft} clause, otherwise.
@@
 
-\subsection{Rule-based Classification}
+\subsection*{\red{Mixed Integer Linear Programming (MILP)}}
+\red{[Not required]}The MILP problem is an optimization problem operating on both integer and real-valued variables.
+Given an objective function and a set of constraints, a (minimization) MILP finds an optimal assignment of variables that minimizes the objective function.
+
+\subsection*{Rule-based Classification}
 
@@
 Finally, if $ \mathbf{x} $ satisfies  $ \ge 2 $ clauses, $ \mathbf{x} $ is assigned a class using a tie-breaking~\cite{lakkaraju2016interpretable}.
+
+%\todo{End}

### 2022-06-02 chapters/fairness/fif/conclusion.tex
all papers added

@@
 %\vspace{-0.5em}
-\section{Conclusion}
+\section{Chapter Summary}
 %\vspace{-1em}
@@
 To compute FIFs, we first prove existing group fairness metrics as the scaled difference of the conditional variances in the predictions of the classifier and then apply variance decomposition based on global sensitivity analysis.
-Finally, we propose {\framework}, an algorithm for efficiently and accurately computing FIFs by deploying a local regression to learn a set-additive decomposition of the classifier.
-The experimental results show that {\framework} estimates bias with significantly higher accuracy than the local explanation based approach SHAP.
-Also, {\framework} computes both individual and intersectional FIFs unlike SHAP to yield a better understanding of the sources of bias.
+Finally, we propose {\fairXplainer}, an algorithm for efficiently and accurately computing FIFs by deploying a local regression to learn a set-additive decomposition of the classifier.
+The experimental results show that {\fairXplainer} estimates bias with significantly higher accuracy than the local explanation based approach SHAP.
+Also, {\fairXplainer} computes both individual and intersectional FIFs unlike SHAP to yield a better understanding of the sources of bias.
 In future,  we aim to develop algorithms that leverage FIFs to yield unbiased decisions.

### 2022-06-02 chapters/fairness/fvgm/conclusion.tex
intro: fairness part start
all papers added

--- 2022-01-26 chapters/fairness/fvgm/conclusion.tex
@@
-\section{Conclusion}
+\section{Chapter Summary}
 In this paper, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem.
@@
 Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
+%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.

### 2022-06-02 chapters/fairness/fvgm/introduction.tex
intro: fairness part start
adding fifs and introduction
all papers added

@@
-\section{Introduction}
+%\section{Introduction}
 	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}.
@@
 	\end{center}
-	\vspace*{-1em}
+%	\vspace*{-1em}
 	\caption{Age-dependent distributions of income and fitness in Example~\ref{fvgm_example:intro}.}\label{fvgm_fig:example1}%\vspace*{-1.6em}

### 2022-06-02 chapters/fairness/justicia/discussion.tex
all papers added

@@
-\section{Discussion and Future Work}
+\section{Chapter Summary}
 Though formal verification of different fairness metrics of an ML algorithm for different datasets is an important question, existing verifiers are not scalable, accurate, and extendable to non-Boolean sensitive features.

### 2022-06-02 chapters/fairness_background.tex
all papers added

@@
 \chapter{Preliminaries}
+We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.
+
+\subsection*{Dataset and Distribution}
+We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
+Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $\{\mathbf{x}^{(i)}_1, \dots, \mathbf{x}^{(i)}_{\numnonsensitive}\} $.
+Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $\{\mathbf{a}^{(i)}_1, \dots, \mathbf{a}^{(i)}_{\numsensitive}\} $.
+$y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.
+
+We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.
+Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}.
+Thus, $ \mathcal{D} $ is the product distribution, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.
+
+For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{\numsensitive}] $ is called a \textit{compound sensitive group}.
+For example, consider $ \sensitive = $ [race, sex] where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}.
+Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group.
+
+We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
+Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
+Given this setup, we discuss different fairness metrics to compute bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.
+
+\subsection*{Fairness Metrics}
+
+A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
+In the following, we describe two well-known fairness definitions: group fairness and causal fairness.
+To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $.
+Alternatively, a fairness metric can be computed on a finite sampled dataset instead of the distribution.
+In that case, we use the notation $ f(\alg, \mathbf{D}) $, which is the bias of the classifier $ \alg $ on a dataset $ \mathbf{D} $.
+In the following, we discuss different fairness metrics.
+
+\paragraph{Statistical Parity ($ \mathsf{SP} $).} Statistical parity belongs to \textit{independence} measuring group fairness metrics, where the prediction $ \widehat{Y} $ is statistically independent of sensitive features $ \sensitive $~\cite{feldman2015certifying}.
+The statistical parity of  a classifier is measured as
+\[ f_{\mathsf{SP}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}], \] which is the difference between the maximum and minimum conditional probability of positive prediction the classifier for different sensitive groups.
+
+\paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.
+Formally, the disparate impact of a classifier is
+\[
+f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}.
+\]
+
+\paragraph{Equalized Odds ($ \mathsf{EO} $).} \textit{Separation} measuring group fairness metrics such as equalized odds constrain that $ \widehat{Y} $ is independent of $ \sensitive $ given the ground class $ Y $~\cite{hardt2016equality}.
+Formally, for $ Y \in \{0,1\} $, equalized odds is
+
+\begin{align*}
+	f_{\mathsf{EO}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \\ &\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1]).
+\end{align*}
+
+\paragraph{Predictive Parity ($ \mathsf{PP} $).} \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $~\cite{verma2018fairness}.
+Formally,
+
+\begin{align*}
+	f_{\mathsf{PP}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \\&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1]).
+\end{align*}
+
+\paragraph{Path-specific Causal Fairness ($ \mathsf{PCF} $).}
+Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.
+We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $.
+This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
+To this end, the path-specific causal fairness, abbreviated as $ \mathsf{PCF} $, of a classifier is \[
+f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
+\]
+
+Therefore, path-specific cause fairness constrains that the prediction $ \hat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through mediator features $ \mediator $.
+Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
+
+\begin{example}
+	Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator).
+It is practical to consider that gender $ \in $ \{male, female\} can affect the college-participation of individuals, and all three features are determining factors for the hiring process.
+Let `male' be the most favored group by the classifier, for instance.
+Path-specific causal fairness ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
+She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.
+Therefore, path-specific causal fairness measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
+\end{example}
+
+\paragraph{Fairness Certification.} We certify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, that quantifies the desired level of fairness.
+In particular, a classifier is $ \epsilon $-fair with respect to statistical parity, equalized odds, predictive parity, and path-specific causal fairness if and only if $ f(\alg, \mathcal{D}) \le \epsilon $.
+In contrast, a classifier achieves $(1 - \epsilon)$-disparate impact  if and only if $ f(\alg, \mathcal{D}) \ge 1 - \epsilon $.
+In all above fairness metrics, a lower value of $ \epsilon $ refers to higher fairness of the classifier.
+
+\subsection*{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
+Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
+A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form (CNF)} if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
+%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
+Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $.
+The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true).
+In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is computes the  probability of the satisfaction of the formula $\phi$ defined on \textit{quantified} Boolean variables.
+An SSAT formula is of the form
+\begin{equation}\label{fairness_justicia_eq:ssat}
+\Phi = Q_1\bool_1, \dots, Q_m \bool_m,\; \phi, 
+\end{equation}
+where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula.
+In case of a randomized quantifier $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
+In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.
+Let $ \bool $ be the outermost variable in the prefix.
+The semantics of SSAT formulas are defined recursively in the following.
+\begin{enumerate}
+	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
+	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
+	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
+	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
+\end{enumerate}
+where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively.
+In this paper, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
+In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
+
+\begin{remark}
+	ER-SSAT problem is $\mathrm{NP}^{\mathrm{PP}}$-hard whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
+\end{remark}
+
+The problem of SSAT and its variants have been pursued by theoreticians and practitioners for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
+We refer the reader to~\cite{lee2017solving,lee2018solving} for a detailed survey.
+It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
+
+\subsection*{Stochastic Subset Sum Problem ({\stochastic})} 
+Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ W_i \in \mathbb{Z} $ be the weight\change{Weight notation $ w $ to $ W $} of $ B_i $.
+Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$.
+Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
+The \textit{counting} version of the subset-sum problem counts all $ \mathbf{b} $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
+In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i \ge \tau $ where variables $ B_i $'s are stochastic.
+More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} W_iB_i \ge \tau] $.
+Details of {\stochastic} are in Section~\ref{fvgm_sec:stochastic_sum_set_sum}.
+
+\subsection*{Bayesian Network}
+In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.
+In this paper, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
+	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
+Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable.
+Edges $ \mathbf{E} \in \mathbf{V} \times \mathbf{V} $ imply conditional dependencies among variables.
+For each variable $ V_i \in \mathbf{V} $, let $ \parent(V_i) \subseteq \mathbf{V} \setminus \{V_i\} $ denote the set of parents of $ V_i $.
+Given $\parent(V_i)$ and parameters $\theta$, $ V_i $ is independent of its other non-descendant variables in $\graph$.
+Thus, for the assignment $ v_i $ of $ V_i $ and $ \mathbf{u} $ of $ \parent(V_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\mathbf{V}$ as:
+
+\begin{equation}
+\begin{split}
+\Pr[V_1=v_1, \dots, &V_{|\mathbf{V}|}=v_{|\mathbf{V}|}] = \prod_{i=1}^{|\mathbf{V}|} \Pr[V_i = v_i | \parent(V_i) = \mathbf{u}; \theta].
+\end{split}
+\label{fvgm_eq:BN}
+\end{equation}
+
+\subsection*{Global Sensitivity Analysis: Variance Decomposition}
+Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
+Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
+%\paragraph{Sensitivity Analysis:}
+\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}.
+To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
+
+\begin{align}
+\function(\mathbf{Z}) &= \function_0 + \sum_{i=1}^{\numnonsensitive} \function_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \function_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \function_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
+\\    
+&= \function_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
+\end{align}
+
+In this decomposition, $ \function_0 $ is a constant, $ \function_{\{i\}} $ is a function of $ Z_i $, $ \function_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
+Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.
+We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \function_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.
+The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}.
+$ \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously.
+%\unsure{Is this sentence necessary?}
+For $|\mathbf{S}|=1$, it quantifies an individual variable's effect. For $|\mathbf{S}|>1$, it quantifies the higher-order interactive effect of variables.
+
+Now, if we assume $ \function $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
+
+\begin{align}\label{eq:variance_decomposition_set_notation}
+\mathsf{Var}[g(\mathbf{Z})] &= \sum_{i=1}^{\numnonsensitive}V_{\{i\}} +  \sum_{i<j}^{\numnonsensitive} V_{\{i,j\}}  + \cdots  + V_{\{1, 2\dots, \numnonsensitive\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}}
+\end{align}
+
+where $ V_{\{i\}} $ is the variance of $ \function_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \function_{\{i,j\}} $ and so on. Formally,
+
+\[ V_{\mathbf{S}} \triangleq \mathsf{Var} _{\mathbf{Z}_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}/\mathbf{Z}_{\mathbf{S}}}[g(\mathbf{Z})\mid \mathbf{Z}_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}/\emptyset}{V} _{\mathbf{S}'}.\]
+
+Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$. %\improvement{Check notations.}
+This variance decomposition shows how the variance of $\function(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them.
+Together all terms sum to the total variance of the model output.
+
+%\todo{End}

### 2022-06-02 chapters/list_of_papers.tex
intro: fairness part start
all papers added

@@
 \section*{List of Publications}
+\addcontentsline{toc}{chapter}{List of Publications}
 This thesis is based on the following publications.
@@
 \begin{enumerate}
-	\item \href{https://arxiv.org/pdf/2205.06936.pdf}{\blue{Efficient Learning of Interpretable Classification Rules}} \\
+
+	\item ``How Biased is Your Feature?": Computing Fairness Influence Functions with Global Sensitivity Analysis\\
+	\underline{Bishwamittra Ghosh}, Debabrota Basu, Kuldeep S. Meel\\
+	Under review, 2022
+
+	\item \href{https://arxiv.org/pdf/2205.06936.pdf}{Efficient Learning of Interpretable Classification Rules} \\
 	\underline{Bishwamittra Ghosh}, Dmitry Malioutov, Kuldeep S. Meel\\
@@
 
-	\item \href{https://arxiv.org/pdf/2109.09447.pdf}{\blue{Algorithmic Fairness Verification with Graphical Models}} \\
+	\item \href{https://arxiv.org/pdf/2109.09447.pdf}{Algorithmic Fairness Verification with Graphical Models} \\
 	\underline{Bishwamittra Ghosh}, Debabrota Basu, Kuldeep S. Meel\\
@@
 
-	\item \href{https://arxiv.org/pdf/2009.06516.pdf}{\blue{Justicia: A Stochastic SAT Approach to Formally Verify Fairness}} \\
+	\item \href{https://arxiv.org/pdf/2009.06516.pdf}{Justicia: A Stochastic SAT Approach to Formally Verify Fairness} \\
 	\underline{Bishwamittra Ghosh}, Debabrota Basu, Kuldeep S. Meel\\
@@
 
-	\item \href{https://bishwamittra.github.io/publication/ecai_2020/paper.pdf}{\blue{Classification Rules in Relaxed Logical Form
-	}} \\
+	\item \href{https://bishwamittra.github.io/publication/ecai_2020/paper.pdf}{Classification Rules in Relaxed Logical Form
+	} \\
 	\underline{Bishwamittra Ghosh}, Dmitry Malioutov, Kuldeep S. Meel\\
@@
 
-	\item \href{https://bishwamittra.github.io/publication/imli-ghosh.pdf}{\blue{IMLI: An Incremental Framework for MaxSAT-Based Learning of Interpretable Classification Rules}}\\
+	\item \href{https://bishwamittra.github.io/publication/imli-ghosh.pdf}{IMLI: An Incremental Framework for MaxSAT-Based Learning of Interpretable Classification Rules}\\
 	\underline{Bishwamittra Ghosh}, Kuldeep S. Meel\\

### 2022-06-02 chapters/fairness/fif/experiments.tex
tables and figures updated with short caption

@@
 	\end{minipage}
-	\caption{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics.
+	\caption[Accuracy of {\fairXplainer}]{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics.
 Lower values on the $ Y $-axis denote a better result.
@@
 	\vspace{-0.2em}
-	\caption{FIFs for COMPAS dataset on explaining statistical parity. Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}
+	\caption[Individual vs.\ intersectional FIFs]{FIFs for COMPAS dataset on explaining statistical parity.
+Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}
 
@@
 	\vspace{-0.2em}
-	\caption{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing algorithm~\cite{kamiran2012data} on FIFs.}\label{fig:affirmative_punitive_actions}
+	\caption[FIFs under fairness attack and fairness enhancing algorithms]{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing algorithm~\cite{kamiran2012data} on FIFs.}\label{fig:affirmative_punitive_actions}
 %	\vspace{-0.5em}

### 2022-06-02 chapters/fairness/fif/figure_intro.tex
tables and figures updated with short caption

@@
 
-	\caption{FIFs of input features to investigate the bias (i.e., statistical parity) of a decision tree classifier predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
+	\caption[Illustration of fairness influence functions]{FIFs of input features to investigate the bias (i.e., statistical parity) of a decision tree classifier predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
 An affirmative action reduces bias, and corresponding FIFs present the evidence.}

### 2022-06-02 chapters/fairness/fvgm/applications.tex
tables and figures updated with short caption

@@
 	\centering
-	\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/disp_fairness_attack}}
-	\subfloat{\includegraphics[scale=0.2]{figures/fairness/fvgm/stat_fairness_attack}}
+	\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/disp_fairness_attack}}
+	\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/stat_fairness_attack}}
 
-	\captionof{figure}{Verifying poisoning attack against fairness using {\fvgm}. The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
+	\caption[Verifying fairness attacks]{Verifying poisoning attack against fairness using {\fvgm}.
+The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
 \end{figure}
@@
 	\vspace{-2ex}
-	\subfloat{\includegraphics[scale=0.23]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_1}}
-	\subfloat{\includegraphics[scale=0.23]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_0}}
+	\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_1}}
+	\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_0}}
 	%\vspace{-0.5em}
 %	\vspace{-2ex}
-	\caption{FIF computation for COMPAS dataset.
+	\caption[FIF computation using {\fvgm}]{FIF computation for COMPAS dataset.
 For Female (left plot) and Male, influence of `age' decreases and the probability of positive prediction of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}

### 2022-06-02 chapters/fairness/fvgm/dp_illustration.tex
tables and figures updated with short caption

@@
 
-\caption{Search tree representation of {\stochastic} for computing the maximum probability of positive prediction of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
+\caption[Simulation of stochastic subset-sum problem]{Search tree representation of {\stochastic} for computing the maximum probability of positive prediction of the classifier on variables $ \mathbf{B} =  \{P,Q,R,S\} $ with weights $ \{1,1,1,-1\} $ and threshold $ \tau = 2 $ .
 Each node is labeled by $ (i,\tau') $, where $ i $ is the index of $ \mathbf{B} $ and $ \tau' $ is the residual threshold.

### 2022-06-02 chapters/fairness/fvgm/evaluations.tex
tables and figures updated with short caption

@@
 	\begin{center}		
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_LR_time_}}\\
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/cactus_all_verifiers_SVM_time_}}		
+		\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/cactus_all_verifiers_LR_time_}}
+		\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/cactus_all_verifiers_SVM_time_}}		
 	\end{center}
 	%	\vspace{-2ex}
-	\caption{A cactus plot to present the scalability of different fairness verifiers.
+	\caption[Scalability of {\fvgm}]{A cactus plot to present the scalability of different fairness verifiers.
 The number of solved benchmarks are on the $ X $-axis and the required time is on the $ Y $-axis; a point $ (x,y) $ implies that a verifier takes less than or equal to $ y $ seconds to compute fairness metrics of $ x $ many benchmarks.
@@
 	\begin{center}	
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_LR_01}}\\
-		\subfloat{\includegraphics[scale=0.24]{figures/fairness/fvgm/sanity_DI_SVM_01}}		
+		\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/sanity_DI_LR_01}}
+		\subfloat{\includegraphics[scale=0.3]{figures/fairness/fvgm/sanity_DI_SVM_01}}		
 		%		\subfloat[]{\includegraphics[scale=0.25]{figures/fairness/fvgm/sanity_DI_LR_005}}
@@
 	%	\vspace{-2ex}
-	\caption{Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features.
+	\caption[Accuracy of {\fvgm}]{Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features.
 {\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of DI for LR and SVM classifiers.}\label{fvgm_fig:sanity_exp}
@@
 \begin{table}[!t]
+
+	\caption[Verification of fairness metrics and algorithms using {\fvgm}]{Verification of fairness algorithms using {\fvgm}.
+$ \sensitive $ denotes sensitive features.
+RW and OP refer to reweighing and optimized-preprocessing algorithms.
+Numbers in bold refer to fairness improvement.
+}\label{fvgm_tab:fair_algo_verification}
+
 		\centering
 
-		\setlength{\tabcolsep}{0.45em}
+%		\setlength{\tabcolsep}{0.45em}
 		\begin{tabular}{lllrrrrrrrrrrrrr}
@@
 		\end{tabular}
-
-	\caption{Verification of fairness algorithms using {\fvgm}.
-$ \sensitive $ denotes sensitive features.
-RW and OP refer to reweighing and optimized-preprocessing algorithms.
-Numbers in bold refer to fairness improvement.
-}\label{fvgm_tab:fair_algo_verification}
 

### 2022-06-02 chapters/fairness/fvgm/introduction.tex
tables and figures updated with short caption

@@
 \begin{figure}[h!]
-	%		{r}{0.25\textwidth}%\vspace*{-2.4em}
 	\begin{center}
-		\includegraphics[scale = 0.4]{figures/fairness/fvgm/sanity_probability}
+		\includegraphics[scale = 0.5]{figures/fairness/fvgm/sanity_probability}
 	\end{center}
-%	\vspace*{-1em}
-	\caption{Age-dependent distributions of income and fitness in Example~\ref{fvgm_example:intro}.}\label{fvgm_fig:example1}%\vspace*{-1.6em}
+	\caption[Illustrative distribution of features]{Age-dependent distributions of income and fitness in Example~\ref{fvgm_example:intro}.}\label{fvgm_fig:example1}
 \end{figure}

### 2022-06-02 chapters/fairness/justicia/experiments.tex
tables and figures updated with short caption

@@
 	}
-	\caption{Fairness metrics measured by {\justicia} for different sensitive groups in the Adult dataset. The number within parenthesis in the xticks denotes total compound groups.}
+	\caption[Fairness verification on compound sensitive groups]{Fairness metrics measured by {\justicia} for different sensitive groups in the Adult dataset.
+The number within parenthesis in the xticks denotes total compound groups.}
 	\label{fairness_justicia_fig:sensitive_groups}
@@
 	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}}
-	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}		\caption{Standard deviation in estimation of disparate impact (DI) and stat.
+	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}
+
+	\caption[Robustness of fairness verification]{Standard deviation in estimation of disparate impact (DI) and stat.
 parity (SP)  for different sample sizes (sample size $ = 1 $ refers to the entire dataset).
@@
 		\hfill
-		\caption{Runtime comparison of different encodings while varying total sensitive groups in the Adult dataset. \red{Label `learn' to be `infer'???}}
+		\caption[Runtime of different encodings in {\justicia}]{Runtime comparison of different encodings while varying total sensitive groups in the Adult dataset. \red{Label `learn' to be `infer'???}}
 		\label{fairness_justicia_fig:runtime_diff_encodings}

### 2022-06-02 chapters/fairness/justicia/introduction.tex
all papers added
tables and figures updated with short caption

@@
-\section{Introduction}
+%\section{Introduction}
 %\paragraph{Motivation: \red{ML} + Fairness}
@@
 
-\begin{figure}[t!]%%\vspace*{-1em}
+\begin{figure}
 	\centering\hspace*{-4em}
-	\begin{minipage}{.33\columnwidth}
+	\begin{minipage}{.25\columnwidth}
 		\centering
-		\scalebox{0.8}{	
+		\scalebox{0.6}{	
 			\begin{tikzpicture}[x=1.5cm,y=1.8cm]
@@
 	\end{minipage}%
-	\caption{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}%\vspace*{-2em}
+	\caption[A decision tree classifier on sensitive and non-sensitive features]{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}%\vspace*{-2em}
 \end{figure}

### 2022-06-02 chapters/interpretability/imli/experiment.tex
all papers added
tables and figures updated with short caption

@@
 
-	\caption{Comparison of accuracy and rule-size among interpretable classifiers.
+	\caption[Accuracy and rule-size of interpretable classifiers]{Comparison of accuracy and rule-size among interpretable classifiers.
 Each cell from the fourth to the eighth column contains test accuracy (top) and rule-size (bottom).
@@
 
-	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
-	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}
-	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
-	\subfloat{\includegraphics[scale=0.21]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}\\
+	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}
 
-	\caption{Comparison of scalability among interpretable classifiers.
+	\caption[Scalability of Interpretable Classifiers]{Comparison of scalability among interpretable classifiers.
 The plots are arranged in increasing sizes of datasets (from left to right).
@@
 	\centering
-	\caption{Comparison of {\imli} with non-interpretable classifiers in terms of test accuracy.
+	\caption[Accuracy of {\imli} and non-interpretable classifiers]{Comparison of {\imli} with non-interpretable classifiers in terms of test accuracy.
 In the table, `\textemdash' represents a timeout.
@@
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_wdbc_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_predicate_count}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_credit_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_predicate_count}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/imli_exp_adult_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_predicate_count}}
 
-	\caption{Comparison of training time, test error, and rule-size among different formulations presented in the paper.
+	\caption[Training time, test error, and rule-size of different formulations in {\imli}]{Comparison of training time, test error, and rule-size among different formulations presented in the paper.
 In each cactus plot, the incremental formulation {\imli} with both mini-batch and iterative learning demonstrates the best performance in training time and test error than compared two formulations: non-incremental MaxSAT formulation {\imli}-$\mathsf{B}$ and incremental formulation with only mini-batch learning {\imli}-$\mathsf{M}$.
@@
 	\centering
-	\caption{Comparison of test accuracy (top value) and rule-size (bottom value) among different rule-based representations learned using {\imli}.
+	\caption[Accuracy and rule-size of different classification rules learned using {\imli}]{Comparison of test accuracy (top value) and rule-size (bottom value) among different rule-based representations learned using {\imli}.
 Numbers in bold denote the best performing results among different representations.}
@@
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_test_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_predicate_count_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_fit_time_wdbc}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_test_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_predicate_count_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_fit_time_titanic}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_test_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_predicate_count_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/clauses_train_val_fit_time_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_test_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_predicate_count_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/clauses_train_val_fit_time_adult}}
 
-	\caption{Effect of the number of clauses $ k $ on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of the number of clauses in {\imli}]{Effect of the number of clauses $ k $ on accuracy (test and train), rule-size, and training time.
 As $ k $ increases, both train and test accuracy of {\imli} increase while generating rules with higher size by incurring higher training time.
@@
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__test_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__predicate_count_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_fit_time_wdbc}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__test_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__predicate_count_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_fit_time_titanic}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__test_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__predicate_count_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/lambda__train_val_fit_time_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__test_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__predicate_count_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/lambda__train_val_fit_time_adult}}
 
-	\caption{Effect of regularization $ \lambda $ on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of regularization $ \lambda $ in {\imli}]{Effect of regularization $ \lambda $ on accuracy (test and train), rule-size, and training time.
 As $ \lambda $ increases, lower priority is given to accuracy.
@@
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_wdbc}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_test_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_accuracy_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_predicate_count_wdbc}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_fit_time_wdbc}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_titanic}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_test_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_accuracy_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_predicate_count_titanic}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_fit_time_titanic}}
 
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_test_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_accuracy_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_predicate_count_adult}}
-	\subfloat{\includegraphics[scale=0.25]{figures/interpretability/imli/batch_size_train_val_fit_time_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_test_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_accuracy_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_predicate_count_adult}}
+	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/batch_size_train_val_fit_time_adult}}
 
-	\caption{Effect of bath size on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of batch-size in {\imli}]{Effect of bath size on accuracy (test and train), rule-size, and training time.
 As we consider more samples in a mini-batch, {\imli} generates more accurate and larger size classification rules.}

### 2022-06-02 chapters/interpretability/relaxed-cnf/experiment.tex
tables and figures updated with short caption

@@
 
-%\vspace{-3pt}
+%%\vspace{-3pt}
 \subsection{Experiment Methodology}
@@
 
-\begin{table}[b]
+\begin{table}
+	\caption[Accuracy, rule-size, and training time of rule-based classifiers]{Comparisons of test accuracy, rule-size and training time among different rule-based classifiers.
+Every cell in the last four columns contains the test accuracy in percentage (top value), rule size  (middle value), and  training time in seconds (bottom value).
+In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.}
+	\label{interpretability_crr_tab:rule_based_classifiers}
 	\setlength{\tabcolsep}{3pt} 
-	\scriptsize 
 	\begin{center}
@@
 	\end{center}
-	\caption{Comparisons of test accuracy, rule-size and training time among different rule-based classifiers.
-Every cell in the last four columns contains the test accuracy in percentage (top value), rule size  (middle value), and  training time in seconds (bottom value).
-In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.}
-	\label{interpretability_crr_tab:rule_based_classifiers}
+
 \end{table}
@@
 
-%\vspace{-9pt}
+%%\vspace{-9pt}
 \subsection{Results}
 In the following, we first discuss empirical results of rule-based classifiers, then extend analysis to non-rule-based classifiers, and finally discuss the effect of different choices of hyper-parameters.
-%\vspace{-3pt}
+%%\vspace{-3pt}
 \subsubsection*{Performance Evaluation of {\crr} with Rule-based Classifiers:}
@@
 
-\begin{table}[tb]
+\begin{table}
+	\caption[Accuracy of {\crr} and non-rule-based classifiers]{Comparisons of test accuracy among {\crr} and  non-rule-based classifiers.
+In the experiments, {\crr} achieves competitive prediction accuracy in spite of being a rule-based classifier.}
+	\label{interpretability_crr_tab:all_classifiers}
 	\begin{center}
@@
 	\end{center}
-	\caption{Comparisons of test accuracy among {\crr} and  non-rule-based classifiers. In the experiments, {\crr} achieves competitive prediction accuracy in spite of being a rule-based classifier.}
-	\label{interpretability_crr_tab:all_classifiers}
 \end{table}
@@
 \begin{figure}[t]
-	%\vspace{-25pt}
+	%%\vspace{-25pt}
 	\centering
@@
 	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_iteration.pdf}\label{interpretability_crr_fig:g}} \hfill
-	\caption{Effect of model hyper-parameters on test accuracy  for  Heart disease dataset. } 
+	\caption[Effect of hyper-parameters on accuracy in {\crr}]{Effect of model hyper-parameters on test accuracy  for  Heart disease dataset. } 
 	\label{interpretability_crr_fig:result_1}
@@
 
-%\vspace{-3pt}
+%%\vspace{-3pt}
 	\subsubsection*{Varying Model Parameters: }
@@
 
-	%	%\vspace{-3pt}
+	%	%%\vspace{-3pt}
 	\textbf{Varying data-fidelity parameter ($ \lambda $):}  As we increase data-fidelity parameter $ \lambda $ in the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj} and Eq.~\ref{interpretability_crr_eq:ilp_inc}, more priority is given to the prediction accuracy than the sparsity of the rules.
@@
 	\begin{figure}[t]
-		%\vspace{-25pt}
+		%%\vspace{-25pt}
 		\centering
@@
 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
+		%%\vspace{-10pt}
+		\caption[Effect of hyper-parameters on rule-size in {\crr}]{Effect of model hyper-parameters on rule-size for  Heart disease dataset.}
+
 		%\vspace{-10pt}
-		\caption{Effect of model hyper-parameters on rule-size for  Heart disease dataset.}
-
-		\vspace{-10pt}
 		\label{interpretability_crr_fig:result_2}
@@
 	 \begin{figure}[btp]
-	 	%\vspace{-25pt}
+	 	%%\vspace{-25pt}
 	 	\centering
@@
 	 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
-	 	%	%\vspace{-10pt}
-	 	\caption{Effect of model hyper-parameters on training  time for  Heart disease dataset.}
+	 	%	%%\vspace{-10pt}
+	 	\caption[Effect of hyper-parameters on training time in {\crr}]{Effect of model hyper-parameters on training  time for  Heart disease dataset.}
 
-	 	\vspace{-10pt}
+	 	%\vspace{-10pt}
 	 	\label{interpretability_crr_fig:result_3}
@@
 
-	%	%\vspace{-3pt}
+	%	%%\vspace{-3pt}
 	\textbf{Varying the number of iterations ($ \tau $):}   

### 2022-06-02 chapters/interpretability/relaxed-cnf/introduction.tex
intro started
all papers added
tables and figures updated with short caption

@@
 %%\vspace{-5pt}
-\section{Introduction}
+%\section{Introduction}
 %%\vspace{-5pt}
-\label{interpretability_crr_sec:introduction}
+%\label{interpretability_crr_sec:introduction}
 
@@
 	\begin{center}
-		\includegraphics[scale=.7]{figures/interpretability/relaxed-cnf/checklist.pdf}
+		\includegraphics[scale=.65]{figures/interpretability/relaxed-cnf/checklist.pdf}
 	\end{center}
-	\caption{An illustrative example of a  relaxed CNF classification rule, which describes the decision function in the form of a  two-level checklist.
+	\caption[Illustration of a relaxed-CNF classification rule]{An illustrative example of a  relaxed CNF classification rule, which describes the decision function in the form of a  two-level checklist.
 This classifier is learned on the WDBC (Wisconsin diagnostic breast cancer) dataset and it predicts whether a tumor cell is malignant or not based on the characteristics of the  tumor cell.
@@
 We then describe the experimental results in Section ~\ref{interpretability_crr_sec:experiment} and conclude in Section~\ref{interpretability_crr_sec:conclusion}.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-06-02 chapters/introduction.tex
tables and figures updated with short caption

@@
 Consequently, fairness research centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
-In this thesis, we study fairness verification problems to assess the fairness claim of different classifiers under the lens of multiple fairness definitions and algorithms.
-In addition, we study the sources of the unfairness of classifiers by identifying a subset of features attributing highly to bias.
+In this thesis, we study fairness verification problems to assess the fairness of different classifiers under the lens of multiple fairness definitions and algorithms.
+In addition, we study the sources of the unfairness of classifiers by identifying  features attributing highly to bias.
 In both interpretable and fair machine learning, we prioritize improving both the \emph{scalability} and the \emph{accuracy} of solutions, either or both of which prior approaches fail to achieve.
@@
 Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
-Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain.
+Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and Boolean cardinality constraints in logic.
 Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets.
@@
 Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
-We extend probabilistic fairness verification to linear classifiers by proposing a tractable verification framework in contrast to the existing $ \mathrm{NP}^{\mathrm{PP}} $-hardness of SSAT-based verification in {\justicia}.
-We also improve the accuracy of verification by considering feature correlations.
-Specifically, we propose a fairness verification framework for linear classifiers, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurate and scalable fairness verification.
+Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
+Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurate and scalable fairness verification.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.

### 2022-06-02 chapters/fairness/fif/experiments.tex
minor edits in many parts of the thesis

@@
 
-\noindent\textbf{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $).
+\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $).
 In {\fairXplainer}, we deploy SAlib library~\cite{Herman2017} to compute FIFs leveraging techniques of global sensitivity analysis.
@@
 In the following, we discuss the objective of our empirical study.
-\begin{enumerate}[nosep,leftmargin=*]
+
+\begin{enumerate}
 %\begin{enumerate}
@@
 \end{enumerate}
+
 In summary, \textit{{\fairXplainer} achieves significantly less estimation error than SHAP in all the datasets.} This shows the importance of adopting global sensitivity analysis to compute FIFs for group fairness metrics than local explanation approaches.
@@
 \end{figure}
-\noindent\textbf{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}).
+\paragraph{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}).
 We consider five datasets, each trained on different classifiers by applying five-fold cross-validation.
@@
 
-\noindent\textbf{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs.
+\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs.
 We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years.
@@
 
-\noindent\textbf{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.
+\paragraph{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.
 In Figure~\ref{fig:individual_vs_intersectional_influence}, statistical parity of the classifier is $ 0.17 $.

### 2022-06-02 chapters/fairness/fif/fairxplainer.tex
minor edits in many parts of the thesis

@@
 
-\noindent\textbf{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}).
+\paragraph{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}).
 Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \function_{\mathbf{a}}(\nonsensitive) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $.
 We express $ \function_{\mathbf{a}} $ as a set-additive model:
+
 \begin{align}
@@
 \end{align}
+
 \begin{algorithm}
@@
 
-\noindent\textbf{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}).
+\paragraph{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}).
 Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \function_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed.
@@
 A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating the local curves obtained by local regression~\cite{loader2012smoothing}.
-In this paper, we apply cubic spline smoothing~\cite{li2010global} to learn each component function.
+In this chapter, we apply cubic spline smoothing~\cite{li2010global} to learn each component function.
 Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity.
@@
 
-\noindent\textbf{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $.
+\paragraph{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $.
 Since each component function is mean centered (Line~\ref{alg_line:mean_centered}), we compute the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ on the dataset as
@@
 
-\noindent\textbf{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
+\paragraph{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
 We extend {\fairXplainer} in computing FIF of group fairness metrics beyond statistical parity, namely equalized odds and predictive parity.

### 2022-06-02 chapters/fairness/fif/framework.tex
minor edits in many parts of the thesis

@@
 %\label{sec:problem_formulation}
-\noindent\textbf{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $.
+\paragraph{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $.
 Our objective is to compute the influences of non-sensitive features on $ f $.
@@
 \end{axiom}
-\noindent\textbf{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}.
+\paragraph{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}.
 In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.
@@
 
-\noindent\textbf{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction.
+\paragraph{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction.
 Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $.
@@
 \begin{example}
+	\normalfont
 	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively.
@@
 \end{theorem}
-For brevity of space. we defer the detailed proof to Appendix~\ref{sec:proof_1}.
 
-\noindent\textbf{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.
+\begin{proof}
+
+	Let $ B_{\mathbf{a}} \in \{0, 1\} $ be a Bernoulli random variable such that $ B_{\mathbf{a}} = 1 $ denotes the classifier predicting positive class $ \hat{Y} = 1 $ for an individual belonging to the sensitive group $ \sensitive = \mathbf{a} $, and $ B_{\mathbf{a}} = 0 $ denotes $ \hat{Y} = 0 $.
+Hence, if $ p_\mathbf{a} \triangleq \Pr[B_\mathbf{a} = 1] $, then $ p_\mathbf{a} $ is also the conditional probability of positive prediction of the classifier,  $ p_\mathbf{a} = \Pr[\hat{Y} = 1 |  \sensitive = \mathbf{a}] $.
+The variance of $ B_{\mathbf{a}} $ is computed as
+	\begin{align*}
+	\mathsf{Var}[B_\mathbf{a}] = p_{\mathbf{a}}(1 - p_{\mathbf{a}}). 
+	\end{align*}
+
+	Now, we consider two Bernoulli random variables $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ associated with two different sensitive groups $ \sensitive = \mathbf{a} $ and $ \sensitive = \mathbf{a}' $, respectively.
+The difference in variances between $ B_\mathbf{a} $ and $ B_\mathbf{a'} $ implies that
+	\begin{align*}
+	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}] &= p_{\mathbf{a}}(1 - p_{\mathbf{a}}) - p_{\mathbf{a'}}(1 - p_{\mathbf{a'}}) \\
+	&= p_{\mathbf{a}} - p_{\mathbf{a}}^2 - p_{\mathbf{a'}} + p_{\mathbf{a'}}^2 \\ 
+	&= (p_\mathbf{a} -  p_\mathbf{a'}) (1 - (p_\mathbf{a} + p_\mathbf{a'}))
+	\end{align*}
+
+	After reorganization, we express the difference in probabilities $ p_\mathbf{a} -  p_\mathbf{a'} $ as
+	\begin{align}
+	\label{eq:sp_to_var}
+	p_\mathbf{a} -  p_\mathbf{a'}  = \frac{	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}]}{1 - (p_\mathbf{a} + p_\mathbf{a'})}
+	\end{align}
+
+	By setting $ \mathbf{a} = \mathbf{a}_{\max} $ and $ \mathbf{a'} = \mathbf{a}_{\min} $ in Equation~\eqref{eq:sp_to_var}, we express the statistical parity of the classifier in terms of the scaled difference in the conditional variance of the positive prediction of the classifier.
+	\begin{align}
+	f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq p_{\mathbf{a}_{\max}} -  ~~p_{\mathbf{a}_{\min}}  &= \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}\label{eq:sp_var_diff}\\
+	&= \frac{	\mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\max}] - \mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\min}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \notag.
+	\end{align}
+
+	%The last inequality is obtained by the definition of $B_{\mathbf{a}}$.
+
+	Next, we apply variance decomposition (Equation~\eqref{eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $.
+From Equation~\eqref{eq:variance_decomposition_set_notation}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as
+	\[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
+	where $ V_{\mathbf{a}, \mathbf{S}} $ denotes the decomposed variance of $ B_\mathbf{a} $ w.r.t.\ the features $ \nonsensitive_{\mathbf{S}} $ conditioned on the sensitive group $ \sensitive = \mathbf{a} $.
+
+	Now, we apply variance decomposition on both $ \mathsf{Var}[B_{\mathbf{a}_{\max}}] $ and $ \mathsf{Var}[B_{\mathbf{a}_{\min}}] $ in Equation~\ref{eq:sp_var_diff} to obtain
+
+	\begin{align}
+	f_{\mathsf{SP}}(\alg, \mathbf{D}) & = \frac{\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\max}})}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \label{eq:influence_value}
+	\end{align}
+	Finally, we estimate  the FIF of the subset of features $ \nonsensitive_{\mathbf{S}}  $ as 
+	\begin{align}
+	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}
+	\end{align} 
+	by separating the terms corresponding to $ \nonsensitive_{\mathbf{S}} $.
+
+	Following Equation~\eqref{eq:influence_value}, $	w_{\mathbf{S}} $ satisfies Axiom~\ref{axm:additivity} as $f_{\mathsf{SP}}(\alg, \mathbf{D}) = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} $.
+\end{proof}
+
+\paragraph{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.
 

### 2022-06-02 chapters/fairness/fif/introduction.tex
all papers added
minor edits in many parts of the thesis

@@
-\section{Introduction}
+%\section{Introduction}
 %\begin{itemize}
@@
 
-\noindent\textbf{Our Contributions.} The contribution of this paper is three-fold.
+\paragraph{Our Contributions.} The contribution of this chapter is three-fold.
 
@@
 
-We illustrate the usefulness of our contributions via an example scenario proposed in~\cite{ghosh2020justicia}.
+We illustrate the usefulness of our contributions via an example scenario in Example~\ref{fairness_justicia_example:intro}.
 
@@
 \begin{example}
+	\normalfont
 	\label{ex:motivating_example}
@@
 
-	Applying techniques developed in this paper, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}.
+	Applying techniques developed in this chapter, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}.
 In Figure~\ref{fig:fif_original}, fitness $(\mathrm{FIF} = 0.74)$, and the joint effect of fitness and income $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e.
@@
 This action allows more elderly individuals to receive insurance by including more people with lower fitness, and thus the statistical parity becomes $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.71 - 0.7 = 0.01$.
-This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this paper, we compute FIF of features in Figure~\ref{fig:fif_affirmative}.
+This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this chapter, we compute FIF of features in Figure~\ref{fig:fif_affirmative}.
 Here, the FIF of income and fitness reflects the reduction in statistical parity as their influences almost nullify each other.
@@
 Also, these works provide empirical evaluations to justify the choices of SHAP-based tools for explaining fairness and does not consider the global nature of group fairness metrics.
-In this paper, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it.
+In this chapter, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it.
 To the best of our knowledge, this is the first work to do the both}.

### 2022-06-02 chapters/fairness/fvgm/conclusion.tex
minor edits in many parts of the thesis

@@
 \section{Chapter Summary}
-In this paper, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem.
+In this chapter, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem.
 {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately.

### 2022-06-02 chapters/fairness/fvgm/evaluations.tex
minor edits in many parts of the thesis

@@
 Finally, we create label $ Y = \mathds{1}[ \sum_{i=1}^{n-1} X_i \ge 0.5 \sum_{i=1}^{n-1} (\mu_i + \mu_i')] $ such that $ Y $ does not directly depend on the sensitive feature.
-For each $ n $, we generate $ 100 $ random benchmarks, learn LR and SVM classifiers on them, and compute DI\footnote{
+For each $ n $, we generate $ 100 $ random benchmarks, learn LR and SVM classifiers on them, and compute DI using different verifiers.
+
 To analytically compute DI, let the coefficients of the classifier be $ w_i $ for $ X_i $ and $ w_A $ for $ A $, and bias be $ \tau $.
@@
 Hence, the probability of positive prediction of the classifier  is $  1 - \mathsf{CDF}_{\hat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\hat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function.
-Finally, we compute DI by taking the ratio of the minimum and the maximum of the probability of positive prediction of the classifier.}
-using different verifiers.
+Finally, we compute DI by taking the ratio of the minimum and the maximum of the probability of positive prediction of the classifier.
 

### 2022-06-02 chapters/fairness/fvgm/framework2.tex
minor edits in many parts of the thesis

@@
 	\begin{example}
+		\normalfont
 		We extend Example~\ref{fvgm_example:subset-sum} with a Bayesian Network $ (\graph, \theta) $ with $ \mathbf{V} = \{P, Q\} $ and $ \mathbf{E} = \{(P,Q)\} $.

### 2022-06-02 chapters/fairness/fvgm/introduction.tex
minor edits in many parts of the thesis

@@
 	\begin{example}\label{fvgm_example:intro}
-		Following~\cite[Example 1.]{ghosh2020justicia}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}).
+		\normalfont
+		Following Example~\ref{fairness_justicia_example:intro}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}).
 Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions.
@@
 
-	In order to identify and mitigate the bias of ML classifiers, different fairness definitions and fairness algorithms have been proposed~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.	In this paper, we focus on two families of fairness definitions: \textit{group} and \textit{causal} fairness.
+	In order to identify and mitigate the bias of ML classifiers, different fairness definitions and fairness algorithms have been proposed~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.	In this chapter, we focus on two families of fairness definitions: \textit{group} and \textit{causal} fairness.
 Group fairness metrics, such as disparate impact and equalized odds constrain the probability of the positive prediction of the classifier to be (almost) equal among different sensitive groups~\cite{dwork2012fairness,feldman2015certifying}.	On the other hand, causal fairness metrics assess the difference in positive predictions  if every feature in the causal relation  remains identical except the sensitive feature~\cite{nabi2018fair,zhang2018fairness}.
@@
 
-	\textbf{Our Contributions.} In this paper, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
+	\textbf{Our Contributions.} In this chapter, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.

### 2022-06-02 chapters/fairness/justicia/experiments.tex
minor edits in many parts of the thesis

@@
 In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner.
-Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{ghosh19incremental}.
+Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}.
 We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula.

### 2022-06-02 chapters/fairness/justicia/framework.tex
minor edits in many parts of the thesis

@@
 \begin{example}[RE-SSAT encoding]
+	\normalfont
 	\label{fairness_justicia_example:re_ssat}
@@
 Given this information, we calculate the probability of positive prediction for the sensitive group age $ \ge 40 $ by solving the following RE-SSAT formula:
+
 	\begin{equation}
@@
 	\end{equation}
+
 	From the solution to this SSAT formula, we get $ \Pr[\Phi_A] = 0.43 $.
 Similarly, to calculate the probability of positive prediction for the group age $ < 40 $, we replace the unit clause\footnote{A unit clause is a clause with a single literal.} $ A $ with $ \neg A $ in the CNF formula in $ \Phi_A $ and construct another SSAT formula $ \Phi_{\neg A} $.
+
 	\[ \Phi_{\neg A} := \R^{0.41}F, \R^{0.93}I, \R^{0.09}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge \neg A \]
+
 	For $ \Phi_{\neg A} $, the solution $ \Pr[\Phi_{\neg A}] = 0.43 $ is similarly derived from an SSAT solver. 
@@
 Providing the conditional probabilities, we construct a modified SSAT formula $\Phi'_A $ and compute $ \Pr[\Phi'_A] = 0.18 $ for age $ \ge 40 $.
+
 	\begin{equation}
@@
 	\end{equation}
+
 	For the sensitive group age $ < 40 $,  we similarly obtain $ \Pr[F|\neg A] = 0.82, \Pr[I|\neg A] = 0.88 $, $ \Pr[J|\neg A] = 0.01 $, construct the modified formula $ \Phi'_{\neg A} $ and get  $ \Pr[\Phi'_{\neg A}] = 0.72 $.
+
 	\[ \Phi'_{\neg A} := \R^{0.82}F, \R^{0.88}I, \R^{0.01}J, \exists A, \; (\neg F \vee I) \wedge (F \vee J) \wedge \neg A \]
+
 	In the later case with correlations among sensitive and non-sensitive features, the RE-SSAT encoding detects the discrimination of the classifier among different sensitive groups, where the classifier is more biased towards the younger group with age $ < 40 $ than the elderly group with age $ \ge 40 $.
@@
 In an ER-SSAT formula, $ \Pr[\Phi] $ is the \textit{maximum} satisfying probability of $ \Phi $ over the optimal assignment of existentially quantified variables given the randomized quantified variables (by Semantic 2, Sec.~\ref{fairness_justicia_sec:ssat}).
-In this paper, we leverage this property to compute the most favored sensitive group with the highest probability of positive prediction of a classifier.
+In this chapter, we leverage this property to compute the most favored sensitive group with the highest probability of positive prediction of a classifier.
 In particular, we consider the following ER-SSAT formula:
@@
 \begin{example}[ER-SSAT encoding]
+	\normalfont
 	\label{fairness_justicia_example:er_ssat}

### 2022-06-02 chapters/fairness/justicia/introduction.tex
minor edits in many parts of the thesis

@@
 \begin{example}\label{fairness_justicia_example:intro}
+\normalfont
 Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60).

### 2022-06-02 chapters/fairness_background.tex
minor edits in many parts of the thesis

@@
 \chapter{Preliminaries}
-We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters. We express random variables in uppercase, and an assignment of a random variable in lowercase.
+In this chapter, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness.
+We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
+We express random variables in uppercase, and an assignment of a random variable in lowercase.
 
@@
 \begin{example}
+	\normalfont
 	Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator).
@@
 where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively.
-In this paper, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
+In this thesis, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
 In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
@@
 In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.
-In this paper, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
+In this chapter, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
 	A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.

### 2022-06-02 chapters/interpretability/imli/baseline.tex
minor edits in many parts of the thesis

@@
 For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{X} $ by negating the column ``$ \text{age }\ge 25 $''.
-Hence, in the rest of the paper, we refer $ m $ as the modified number of columns in $ \mathbf{X} $.
+Hence, in the rest of the chapter, we refer $ m $ as the modified number of columns in $ \mathbf{X} $.
 We next discuss the variables in the MaxSAT problem.
@@
 In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$.
-Since $ \mathbf{X}_l \circ \mathbf{b}_{i} $ constitutes a disjunction of literals (defined in Section~\ref{interpretability_imli_sec:preliminaries}), we apply Tseytin transformation to encode  $ \neg (\mathbf{X}_l \circ \mathbf{b}_{i}) $ into CNF.
+Since $ \mathbf{X}_l \circ \mathbf{b}_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{X}_l \circ \mathbf{b}_{i}) $ into CNF.
 More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $.
@@
 Both open-interval and close-interval discretization techniques have their use-cases where one or the other is appropriate.
-For simplicity, we experiment with close-interval discretization in this paper.
+For simplicity, we experiment with close-interval discretization in this chapter.
 
-	After applying discretization on continuous features, the dataset contains categorical features only, which we convert to binary features using one-hot encoding~\cite{lakkaraju2019faithful,ghosh19imli}.
+	After applying discretization on continuous features, the dataset contains categorical features only, which we convert to binary features using one-hot encoding~\cite{lakkaraju2019faithful}.
 In one-hot encoding, a Boolean vector of features is introduced with cardinality equal to the number of distinct categories.

### 2022-06-02 chapters/interpretability/imli/experiment.tex
minor edits in many parts of the thesis

@@
 Therefore, {\imli} is not only scalable but also accurate in practical classification problems while also being interpretable by design.
-We additionally analyze the comparative performance of different formulations presented in this paper, where the  incremental approach empirically proves its efficiency than the na\"ive MaxSAT formulation.
+We additionally analyze the comparative performance of different formulations presented in this chapter, where the  incremental approach empirically proves its efficiency than the na\"ive MaxSAT formulation.
 Furthermore, we learn and compare the performance of different interpretable representations: decision lists, decision sets, CNF, and DNF formulas using {\imli} and present the efficacy of {\imli} in learning varied interpretable classifiers.
@@
 
-	\caption[Training time, test error, and rule-size of different formulations in {\imli}]{Comparison of training time, test error, and rule-size among different formulations presented in the paper.
+	\caption[Training time, test error, and rule-size of different formulations in {\imli}]{Comparison of training time, test error, and rule-size among different formulations presented in the chapter.
 In each cactus plot, the incremental formulation {\imli} with both mini-batch and iterative learning demonstrates the best performance in training time and test error than compared two formulations: non-incremental MaxSAT formulation {\imli}-$\mathsf{B}$ and incremental formulation with only mini-batch learning {\imli}-$\mathsf{M}$.
@@
 
-\paragraph{Comparison among different formulations in {\imli}.}  We compare the performance of different formulations for learning classification rules as presented in this paper.
+\paragraph{Comparison among different formulations in {\imli}.}  We compare the performance of different formulations for learning classification rules as presented in this chapter.
 In Figure~\ref{interpretability_imli_fig:different_formulations}, we show cactus plots for assessing training time (in seconds), test error (in percentage), and rule-size among different formulations.

### 2022-06-02 chapters/interpretability/imli/introduction.tex
all papers added
minor edits in many parts of the thesis

@@
-\section{Introduction}
+%\section{Introduction}
 
@@
 
-In this paper, we study decision functions expressible in \emph{propositional logic}.
+In this chapter, we study decision functions expressible in \emph{propositional logic}.
 In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean predicates that the formula contains.
 In the following, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.
-
-%\begin{subequations}
-%	\begin{empheq}[box=\widefbox]{align*}
-%	& \text{A tumor is malignant if}\\
-%	&[(\text{compactness SE} < 0.1) \text{\textbf{ OR }}  \neg (0.1 \le \text{concave points} < 0.2)]
-%	\text{\textbf{ \blue{AND} }} \\
-%	&[\neg (0.2 \le \text{area} < 0.3) \text{\textbf{ OR }} \neg (0.1 \le \text{largest symmetry }< 0.2)]
-%	\text{\textbf{ \blue{AND} }} \\
-%	&[(0.6 \le \text{compactness} < 0.7) \text{\textbf{ OR }} (0.4 \le \text{concave points} < 0.5) \text{\textbf{ OR }}  (\text{area SE} \ge 0.1)\\
-%	& \text{\textbf{ OR }} (0.3 \le \text{largest concavity} < 0.4) \text{\textbf{ OR }} (0.2 \le \text{largest area} < 0.3)]
-%	\end{empheq}
-%\end{subequations}
 
@@
 }
+
 \begin{example}
+	\normalfont
 	We learn an interpretable classification rule in conjunctive normal form for predicting the classification of a tumor cell.
@@
 Although the MaxSAT problem is NP-hard, dramatic progress has been made in designing solvers that can handle large-scale problems arising in practice.
-This has encouraged researchers to reduce several optimization problems into MaxSAT such as  optimal planning~\cite{robinson2010partial}, automotive configuration~\cite{walter2013applications}, data analysis in machine learning~\cite{berg2019applications}, and automatic test pattern generation for cancer therapy~\cite{lin2012application}.
+This has encouraged researchers to reduce several optimization problems into MaxSAT such as  optimal planning~\cite{robinson2010partial}, automotive configuration~\cite{walter2013applications}, group-testing~\cite{CGSM2020}, data analysis in machine learning~\cite{berg2019applications}, and automatic test pattern generation for cancer therapy~\cite{lin2012application}.
 
-The primary contribution of this paper is a MaxSAT-based formulation, called {\imli}, for learning interpretable classification rules expressible in propositional logic.
+The primary contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
 For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF (Conjunctive Normal Form) formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.
-In this paper, {\imli} provides precise control to jointly optimize the accuracy and the size of classification rules.
+In this chapter, {\imli} provides precise control to jointly optimize the accuracy and the size of classification rules.
 {\imli} constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples.
@@
 
-The presented framework {\imli} extends our prior papers ~\cite{ghosh19imli,MM18} in the following aspects.
+%The presented framework {\imli} extends our prior chapters ~\cite{ghosh19imli,MM18} in the following aspects.
 \cite{MM18} presented a MaxSAT-based formulation for interpretable classification rule-learning, which jointly optimizes between the accuracy and size of rules.
@@
 This unification allows learning on moderate-size datasets (about $ 50 $ thousand samples).
-This paper takes a significant step in terms of scalability by integrating both iterative learning and mini-batch learning inside the incremental technique.
+This chapter takes a significant step in terms of scalability by integrating both iterative learning and mini-batch learning inside the incremental technique.
 The result is a practical scalable learning framework that can classify datasets with $ 1 $ million samples and can potentially scale beyond that.
 
-We organize the rest of the paper as follows.
+%We organize the rest of the chapter as follows.
 We discuss related literature in Section~\ref{interpretability_imli_sec:related} and preliminaries in Section~\ref{interpretability_imli_sec:preliminaries}.

### 2022-06-02 chapters/interpretability/imli/iterative_learning.tex
minor edits in many parts of the thesis

@@
 In particular, we opt to learn this clause for the majority class, say $ v_i $, in the training dataset by setting the majority samples as class $ 1 $ and all other samples as class $ 0 $.
-As a result, even if the MaxSAT-based learning, presented in this paper, is targeted for binary classification, we can learn a multi-class decision list in {\imli}.
+As a result, even if the MaxSAT-based learning, presented in this chapter, is targeted for binary classification, we can learn a multi-class decision list in {\imli}.
 

### 2022-06-02 chapters/interpretability/imli/problem_formulation.tex
minor edits in many parts of the thesis

@@
 For instance, a rule involving fewer literals is highly interpretable.
-In this work, we minimize the total number of literals in all clauses, as discussed in Section~\ref{interpretability_imli_sec:preliminaries}, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $.
+In this work, we minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $.
 Let $ \Rule $ classify all samples correctly during training.

### 2022-06-02 chapters/interpretability/relaxed-cnf/conclusion.tex
all papers added
minor edits in many parts of the thesis

@@
-\section{Conclusion}
+\section{Chapter Summary}
 \label{interpretability_crr_sec:conclusion}
 
-In this paper, we proposed an efficient combinatorial framework called {\crr}, for learning relaxed-CNF  classification rules, that are  more expressive  than CNF/DNF rules.
+In this chapter, we proposed an efficient combinatorial framework called {\crr}, for learning relaxed-CNF  classification rules, that are  more expressive  than CNF/DNF rules.
 {\crr}  uses a novel integration of mini-batch learning procedure with the MILP framework to learn sparse relaxed-CNF rules.

### 2022-06-02 chapters/interpretability/relaxed-cnf/experiment.tex
minor edits in many parts of the thesis

@@
 We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g.
-$ \IMLI  $~\cite{ghosh19incremental}, RIPPER~\cite{C1995}, BRS~\cite{WRDLKM2017},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
+$ \IMLI  $~\cite{GM2019}, RIPPER~\cite{C1995}, BRS~\cite{WRDLKM2017},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
 Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers.

### 2022-06-02 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
minor edits in many parts of the thesis

@@
 
-After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{ghosh19incremental,LKCL2019}.
+After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{LKCL2019}.
 In this encoding, a Boolean vector is introduced with cardinality equal to the number of distinct categories.

### 2022-06-02 chapters/interpretability/relaxed-cnf/introduction.tex
minor edits in many parts of the thesis

@@
 To enable  safe, robust, and trustworthy integration of such systems, the end users require them to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
-In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,MM18,WR2015,WRDLKM2017}.
-A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{ghosh19incremental,LKCL2019,MM18}.
+In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,WR2015,WRDLKM2017}.
+A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{GM2019,LKCL2019,MM18}.
 Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
@@
 {Through a comprehensive experimental evaluation over datasets from the UCI and Kaggle repository, we observe that {\crr} with relaxed-CNF rules achieves an improved trade-off between accuracy and rule sparsity and scales to datasets with more than $ 10^5 $ samples.
-More significantly, {\crr} generates relaxed-CNF rules with higher accuracy than CNF rules generated by~\cite{ghosh19incremental}.
+More significantly, {\crr} generates relaxed-CNF rules with higher accuracy than CNF rules generated by~\cite{GM2019}.
 Furthermore, compared to decision lists generated by~\cite{C1995}, relaxed-CNF rules are sparser in large datasets.}
@@
 %KSM: Please feel free to comment out the text below if it exceeds page limit
-The rest of the chapter is organized as follows.
+%The rest of the chapter is organized as follows.
 We discuss related works in Section~\ref{interpretability_crr_sec:related_work}, introduce the notation and preliminaries in Section~\ref{interpretability_crr_sec:preliminaries} and formulate the problem in Section~\ref{interpretability_crr_sec:problem}.

### 2022-06-02 chapters/interpretability_background.tex
all papers added
minor edits in many parts of the thesis

@@
 In this chapter, we discuss preliminaries for learning interpretable rule-based classifiers.
-Throughout this part of the thesis, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters $\mathbf{x}$ are reserved for vectors/sets.
+Throughout this part of the thesis, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters such as $\mathbf{x}$ are reserved for vectors/sets.
 
@@
 Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum.
-In this paper, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem.
-The knowledge of the inner workings of {MaxSAT} solvers is  not required for this paper.
+In this chapter, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem.
+The knowledge of the inner workings of {MaxSAT} solvers is  not required for this chapter.
 
-\subsection*{\red{Mixed Integer Linear Programming (MILP)}}
-\red{[Not required]}The MILP problem is an optimization problem operating on both integer and real-valued variables.
+%
+%\subsection*{\red{Mixed Integer Linear Programming (MILP)}}
+%\red{[Not required]}The MILP problem is an optimization problem operating on both integer and real-valued variables.
 Given an objective function and a set of constraints, a (minimization) MILP finds an optimal assignment of variables that minimizes the objective function.

### 2022-06-02 chapters/fairness/fif/experiments.tex
minor edits

@@
 In the following, we discuss the experimental setup, the objectives of experiments, and experimental results.
-{ We adjoin the source code and additional experimental results on the effects of maximum order $\lambda$, runtime, accuracy on different datasets and corresponding explanations (Appendix~\ref{sec:additional_experiments}) to the supplementary material.}
+{ We adjoin the source code and additional experimental results on the effects of maximum order $\lambda$, runtime, accuracy on different datasets and corresponding explanations (\red{Appendix(ref{sec:additionalexperiments})}) to the supplementary material.}
 

### 2022-06-03 chapters/fairness/fif/introduction.tex
upto preliminaries in fairness

@@
 %\end{itemize}
+\label{chapter:fif}
 The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}.
+In such applications, the deployed machine learning classifiers often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}.
 For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.

### 2022-06-03 chapters/fairness/fvgm/introduction.tex
upto preliminaries in fairness

@@
 %\section{Introduction}
+
+\label{chapter:fvgm}
 	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}.

### 2022-06-03 chapters/fairness/justicia/introduction.tex
upto preliminaries in fairness

@@
 %\paragraph{Motivation: \red{ML} + Fairness}
+\label{chapter:justicia}
 Machine learning is becoming the omnipresent technology of our time.
@@
 \subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
-\red{In this chapter, we focus on a popular concept of fairness, known as group fairness.}\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
+In this chapter, we focus on a popular concept of fairness, known as group fairness.\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
 Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.

### 2022-06-03 chapters/fairness_background.tex
upto preliminaries in fairness

@@
 \part{Fairness in Machine Learning}
+
+In this part of the thesis, we discuss preliminaries on fairness in machine learning in Chapter~\ref{chapter_fairness_preliminaries}.
+Next, we discuss fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
+Finally, we discuss identification of sources of unfairness in Chapter~\ref{chapter:fif}.
+
 \chapter{Preliminaries}
+\label{chapter_fairness_preliminaries}
 In this chapter, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness.
@@
 
-\subsection*{Dataset and Distribution}
+\subsection{Dataset and Distribution}
 We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
-Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $\{\mathbf{x}^{(i)}_1, \dots, \mathbf{x}^{(i)}_{\numnonsensitive}\} $.
-Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $\{\mathbf{a}^{(i)}_1, \dots, \mathbf{a}^{(i)}_{\numsensitive}\} $.
+Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $.
+Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.
 $y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.
@@
 Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}.
-Thus, $ \mathcal{D} $ is the product distribution, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.
+Thus, $ \mathcal{D} $ is the product distribution of all features, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.
 
@@
 Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
-Given this setup, we discuss different fairness metrics to compute bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.
+Given this setup, we discuss different fairness metrics to compute the bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.
 
-\subsection*{Fairness Metrics}
+\subsection{Fairness Metrics}
 
-A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\hat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
+A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\widehat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
 In the following, we describe two well-known fairness definitions: group fairness and causal fairness.
 To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $.
-Alternatively, a fairness metric can be computed on a finite sampled dataset instead of the distribution.
+Alternatively, a fairness metric can be computed on a finite dataset instead of on the distribution.
 In that case, we use the notation $ f(\alg, \mathbf{D}) $, which is the bias of the classifier $ \alg $ on a dataset $ \mathbf{D} $.
-In the following, we discuss different fairness metrics.
 
@@
 
-\paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.
+\paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact also belongs to independence measuring group fairness metrics.
+Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.
 Formally, the disparate impact of a classifier is
 \[
-f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]}.
+f_{\mathsf{DI}}(\alg, \mathcal{D}) \triangleq \frac{\min_{\mathbf{a}} \Pr[\widehat{Y} =1 | \mathbf{A} =  \mathbf{a}]}{\max_{\mathbf{a}} \Pr[\widehat{Y} =1 | \mathbf{A} =  \mathbf{a}]}.
 \]
@@
 \paragraph{Path-specific Causal Fairness ($ \mathsf{PCF} $).}
-Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.
+Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.
 We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $.
@@
 To this end, the path-specific causal fairness, abbreviated as $ \mathsf{PCF} $, of a classifier is \[
-f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
+f_{\mathsf{PCF}}(\alg, \mathcal{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ].
 \]
 
-Therefore, path-specific cause fairness constrains that the prediction $ \hat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through mediator features $ \mediator $.
+Therefore, path-specific cause fairness constrains that the prediction $ \widehat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \widehat{Y} $ only through mediator features $ \mediator $.
 Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
@@
 
-\subsection*{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
+\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
 Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
-A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form (CNF)} if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
+A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form} (CNF) if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
 %\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
@@
 The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true).
-In contrast to the SAT problem, the \textit{Stochastic Boolean Satisfiability} (SSAT) problem~\cite{littman2001stochastic} is computes the  probability of the satisfaction of the formula $\phi$ defined on \textit{quantified} Boolean variables.
+In contrast to the SAT problem, the \textit{stochastic Boolean satisfiability} (SSAT) problem~\cite{littman2001stochastic} computes the  probability of the satisfaction of the formula $\phi$ defined on an ordered set of \textit{quantified} Boolean variables.
 An SSAT formula is of the form
@@
 where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula.
-In case of a randomized quantifier $ \R^{p_i} $, $ p_i \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
+In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[B_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
 In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.
@@
 The semantics of SSAT formulas are defined recursively in the following.
+
 \begin{enumerate}
@@
 \end{enumerate}
+
 where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively.
@@
 
-\subsection*{Stochastic Subset Sum Problem ({\stochastic})} 
+\subsection{Stochastic Subset Sum Problem ({\stochastic})} 
 Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ W_i \in \mathbb{Z} $ be the weight\change{Weight notation $ w $ to $ W $} of $ B_i $.
@@
 More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} W_iB_i \ge \tau] $.
-Details of {\stochastic} are in Section~\ref{fvgm_sec:stochastic_sum_set_sum}.
+Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.
 
-\subsection*{Bayesian Network}
+\subsection{Bayesian Network}
 In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.
-In this chapter, we leverage an access to a Bayesian network on $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
+In fairness verification with correlated features, we leverage an access to a Bayesian network on features $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
 A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
@@
 
-\subsection*{Global Sensitivity Analysis: Variance Decomposition}
+\subsection{Global Sensitivity Analysis: Variance Decomposition}
 Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.

### 2022-06-03 chapters/interpretability/imli/experiment.tex
upto preliminaries in fairness

@@
 In particular, {\imli} generates sparser rules than RIPPER in $ 9 $ out of $ 12 $ datasets, wherein RIPPER times out in $ 3 $ datasets.
-Interestingly, the difference in rule-size is more significant in larger datasets, such as in `Vote' dataset, where RIPPER learns a classifier with $ 132 $ Boolean predicates compared to $ 15 $ Boolean predicates by {\imli}.
+Interestingly, the difference in rule-size is more significant in larger datasets, such as in `Vote' dataset, where RIPPER learns a classifier with $ 132 $ Boolean literals compared to $ 15 $ Boolean literals by {\imli}.
 Therefore, {\imli} is better than RIPPER in terms of rule-sparsity, but lags slightly in accuracy.
@@
 For example, in the largest dataset `BNG(credit-g)' with $ 1 $ Million samples, BRS times out and CORELS generates a classifier with  $ 72.08\% $ accuracy with rule-size $ 2 $.
-{\imli}, in contrast, learns a classifier with $ 27.5 $ Boolean predicates achieving $ 75.48\% $ accuracy, which is $ 3\% $ higher than CORELS.
+{\imli}, in contrast, learns a classifier with $ 27.5 $ Boolean literals achieving $ 75.48\% $ accuracy, which is $ 3\% $ higher than CORELS.
 Therefore, {\imli}  makes a good balance between accuracy and rule-size compared to existing interpretable classifiers while also being highly scalable.

### 2022-06-03 chapters/interpretability/imli/introduction.tex
upto preliminaries in fairness

@@
 %The last decade has witnessed significant progress in machine learning (ML). 
+\label{chapter:imli}
 Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}.
@@
 At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging.
-In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean predicates), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
+In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
 
 In this chapter, we study decision functions expressible in \emph{propositional logic}.
-In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean predicates that the formula contains.
+In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean literals that the formula contains.
 In the following, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.
 
+\vspace{1em}
 \boxed{
@@
 }
+\vspace{1em}
 
@@
 	We learn an interpretable classification rule in conjunctive normal form for predicting the classification of a tumor cell.
-We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with three \emph{clauses} and nine Boolean \emph{predicates}.
-In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{predicates} connected by Boolean `OR'.
-Informally, a tumor cell is malignant if at least one predicate in each clause becomes true.
+We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with three \emph{clauses} and nine Boolean \emph{literals}.
+In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{literals} connected by Boolean `OR'.
+Informally, a tumor cell is malignant if at least one literal in each clause becomes true.
 \end{example}

### 2022-06-03 chapters/interpretability/relaxed-cnf/introduction.tex
upto preliminaries in fairness

@@
 
+\label{chapter:crr}
 The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
@@
 An entry in the second column is  $ 1 $ if the corresponding literal is true by an observed tumor cell, and $ 0 $ otherwise.
-In the first level, checklist A (resp.\ B) is true if the count of true  predicates is at least $ 2 $ (resp.\ $ 3 $).
+In the first level, checklist A (resp.\ B) is true if the count of true  literals is at least $ 2 $ (resp.\ $ 3 $).
 In the second level, a tumor cell is predicted as malignant if the count of true checklists is at least $ 1 $.

### 2022-06-03 chapters/interpretability_background.tex
upto preliminaries in fairness

@@
 
+In this part of the thesis, we discuss preliminaries on interpretable machine learning in Chapter~\ref{chapter_interpretability_preliminaries}, an incremental learning framework for interpretable classification rules in Chapter~\ref{chapter:imli}, and extend incremental learning to more expressible classification rules in Chapter~\ref{chapter:crr}.
+
 \chapter{Preliminaries}
-
+\label{chapter_interpretability_preliminaries}
 %\section{Preliminaries}
@@
 
-\subsection*{Propositional Satisfiability}
+\subsection{Propositional Satisfiability}
 
 Let $F$ be a Boolean formula defined over Boolean variables $\mathbf{b} = \{b_1,b_2,\dots ,b_m \}$.
-A literal or predicate $ v $ is a variable $b$ or its complement $\neg b$, and a clause $ C $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals in the literature, we use clause for both disjunction and conjunction.}  $F$ is in Conjunctive Normal Form (CNF) if $F \triangleq  \bigwedge_i C_i$ is a conjunction of clauses where each clause $C_i \triangleq  \bigvee_j v_j $  is a disjunction of literals.
+A literal $ v $ is a variable $b$ or its complement $\neg b$, and a clause $ C $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals in the literature, we use clause for both disjunction and conjunction.}  $F$ is in Conjunctive Normal Form (CNF) if $F \triangleq  \bigwedge_i C_i$ is a conjunction of clauses where each clause $C_i \triangleq  \bigvee_j v_j $  is a disjunction of literals.
 In contrast, $ F $ is in Disjunctive Normal Form (DNF) if $F \triangleq  \bigvee_i C_i$ is a disjunction of clauses where each clause $C_i \triangleq  \bigwedge_j v_j $ is a conjunction of literals.
@@
 
-\subsection*{Relaxation of Logical Formulas}
+\subsection{Relaxation of Logical Formulas}
 
@@
 
-\subsection*{Inner Product}
+\subsection{Inner Product}
 In this part of the thesis, we use true as $ 1 $ and false as $ 0 $ interchangeably.
@@
 
-\subsection*{MaxSAT}
+\subsection{MaxSAT}
 
@@
 %
-%\subsection*{\red{Mixed Integer Linear Programming (MILP)}}
+%\subsection{\red{Mixed Integer Linear Programming (MILP)}}
 %\red{[Not required]}The MILP problem is an optimization problem operating on both integer and real-valued variables.
@@
 
-\subsection*{Rule-based Classification}
+\subsection{Rule-based Classification}
 

### 2022-06-03 chapters/introduction.tex
upto preliminaries in fairness

@@
 In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users.
-With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability, fairness, robustness, and privacy.
+With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability, fairness, robustness, and privacy in predictions.
 In this thesis, we focus on the interpretability and fairness aspects of machine learning.
@@
 Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
-Thus, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
+In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
 In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
@@
 At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging.
-In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
+In our context of rule-based classifiers, we use \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
 
@@
 The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-In such applications, the deployed  classifier often demonstrate bias towards certain sensitive demographic groups involved in the data~\cite{dwork2012fairness}.
+In such applications, the deployed  classifier often demonstrates bias towards certain \emph{sensitive demographic groups} involved in the data~\cite{dwork2012fairness}.
 For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.
@@
 Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we focus on two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.
 
@@
 
-In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula.
+In order to solve the fairness verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula.
 Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
@@
 Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
-Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurate and scalable fairness verification.
+Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for accurate and scalable fairness verification.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
@@
 Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
-We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) \red{computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.}
+We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias, which we study in detail in the following contribution of the thesis.
 
@@
 
+\section{Tools}
 \section{Outline}

### 2022-06-03 chapters/fairness/justicia/introduction.tex
upto intro of justicia

@@
 Machine learning is becoming the omnipresent technology of our time.
-Machine learning classifiers are being used for high-stake decisions like college admissions, crime recidivism, insurance, and loan decisions.
+Machine learning classifiers are being used for high-stake decisions such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
 Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.
@@
 
-\begin{figure}
-	\centering\hspace*{-4em}
-	\begin{minipage}{.25\columnwidth}
+\begin{figure}[!t]
 	\centering
-		\scalebox{0.6}{	
-			\begin{tikzpicture}[x=1.5cm,y=1.8cm]
+	\subfloat[Dependency among features and prediction]{\scalebox{1}{	
+			\begin{tikzpicture}[x=1cm,y=0.6cm]
 			% Define nodes
-			\node[latent,scale=1.5] (a1) {$\textrm{age}$} ; %
-			\node[obs, scale=1.5, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
-			\node[obs, scale=1.5, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
-			\node[obs, scale=1.5, below=of h, xshift=.8cm] (p) {$\hat{Y}$}; %			
+			\node[latent,scale=1] (a1) {$\textrm{age}$} ; %
+			\node[obs, scale=1, below=of a1, xshift=-.8cm] (h) {$\textrm{fitness}$}; %
+			\node[obs, scale=1, below=of a1, xshift=.8cm] (i) {$\textrm{income}$}; %
+			\node[obs, scale=1, below=of h, xshift=.8cm] (p) {$\widehat{Y}$}; %			
 			%%add edge
@@
 			\end{tikzpicture}
+	}}\hfil
+	\subfloat[Age-dependent distributions of non-sensitive features]{
+		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
 	}
-	\end{minipage}\hspace*{-2em}
-	\begin{minipage}{.33\columnwidth}
-		\centering
-		\includegraphics[scale=0.5]{figures/fairness/justicia/example.pdf}
-	\end{minipage}\hspace*{-2em}
-	\begin{minipage}{.35\columnwidth}
-		\scalebox{0.48}{	
+
+	\subfloat[Trained decision tree]{
+		\scalebox{0.55}{	
 			\begin{tikzpicture}[x=1cm,y=1.8cm]
 			\node [box, scale=1.5]                                    (p)      {fitness $\geq 0.61$};
-			\node [scale=1.5, above= 0.1cm of p]  (a)    {Trained Decision Tree};
-			\node [scale=1.5, box, below= of p, xshift=-2.5cm]    (a1)    {income\\ $\geq 0.29$};
-			\node [scale=1.5, box, below= of p, xshift=2.5cm]     (a2)    {income $\geq 0.69$};
-			\node [scale=1.5,below= of a1, xshift=-1.8cm]  (a11)    { $\hat{Y}= 1$};
-			\node [scale=1.5,below= of a1, xshift=1.8cm]   (a12)    { $\hat{Y}=0 $};
-			\node [scale=1.5,below= of a2, xshift=-1.8cm]  (a21)    { $\hat{Y}= 1$};
-			\node [scale=1.5,below= of a2, xshift=1.8cm]  (a22)    { $\hat{Y}= 0$};
+			\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
+			\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm]     (a2)    {income $\geq 0.69$};
+			\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
+			\node [scale=1.5,below= of a1, xshift=1.5cm, yshift=0.8cm]   (a12)    { $\widehat{Y}=0 $};
+			\node [scale=1.5,below= of a2, xshift=-1.5cm, yshift=0.8cm]  (a21)    { $\widehat{Y}= 1$};
+			\node [scale=1.5,below= of a2, xshift=1.5cm, yshift=0.8cm]  (a22)    { $\widehat{Y}= 0$};
 			%
-			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Yes};
-			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {No};
-			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Yes};
-			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {No};
-			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Yes};
-			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {No};
+			\path [line] (p) -|         (a1) node [scale=1.5,midway, above]  {Y};
+			\path [line] (p) -|         (a2) node [scale=1.5,midway, above]  {N};
+			\path [line] (a1) -|       (a11) node [scale=1.5,midway, above]  {Y};
+			\path [line] (a1) -|       (a12) node [scale=1.5,midway, above]  {N};
+			\path [line] (a2) -|       (a21) node [scale=1.5,midway, above]  {Y};
+			\path [line] (a2) -|       (a22) node [scale=1.5,midway, above]  {N};
 			\end{tikzpicture}}
-	\end{minipage}%
-	\caption[A decision tree classifier on sensitive and non-sensitive features]{A trained decision tree to learn eligibility for health insurance using age-dependent fitness and income indicators.}\label{fairness_justicia_fig:fair_example}%\vspace*{-2em}
+	}\hfil
+	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
+		\scalebox{1}{
+			\begin{tikzpicture}
+			\begin{axis}[
+			enlarge x limits=0.2,
+			symbolic x coords={$ \text{age} \ge 40 $, $ \text{age} < 40 $},
+			xtick={$ \text{age} \ge 40 $, $ \text{age} < 40 $},  
+%			xticklabel style={text height=2ex}, 
+			ylabel={$ \Pr[\hat{Y} =  1] $},
+			xlabel={Sensitive groups},
+			ymin=0,
+			width= 0.3\textwidth,]
+
+			\addplot[ybar,fill] coordinates {
+				($ \text{age} \ge 40 $, 0.18)
+				($ \text{age} < 40 $, 0.72)};
+
+			\end{axis}
+
+			\end{tikzpicture}
+		}
+	}
+	\caption[A decision tree classifier on sensitive and non-sensitive features]{A trained decision tree to learn the eligibility for health insurance using age-dependent fitness and income indicators.
+This classifier makes unfair prediction to individuals with age above $ 40 $.}\label{fairness_justicia_fig:fair_example}%\vspace*{-2em}
 \end{figure}
@@
 Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
-Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same~\cite{hardt2016equality}.
+Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth label~\cite{hardt2016equality}.
 Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
@@
 %These papers operate for a specific Boolean sensitive feature.
-These verifiers are referred to as {\em distributional verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
+These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
 Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
-In contrast to the distributional verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}.
+In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}.
 Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.
@@
 	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
-	\item Unlike previously proposed formal distributional verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
-	\item Our experiments validate that our method is more accurate and scalable than the distributional verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
+	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
+	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
 	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.

### 2022-06-06 chapters/fairness/justicia/framework.tex
upto theoretical analysis of justicia

@@
 Additionally, for each variable, we query the distribution $ \mathcal{D} $ to derive the marginal probability of the variable to be assigned to $ 1 $.
-In this section, we discuss two equivalent approaches for fairness verification based on SSAT-based encodings: enumeration approach and \red{inferring} approach.
+In this section, we discuss two equivalent approaches for fairness verification based on SSAT-based encodings: \emph{enumeration} approach and \emph{inference} approach.
 In both approaches, we verify fairness with the presence of compound sensitive groups.
@@
 \label{fairness_justicia_sec:enumeration_ssat}
-In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional probability of positive prediction of the classifier, $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ by solving an appropriately designed SSAT formula.
+In order to estimate $ f(\alg, \mathcal{D}) $ in the enumeration approach, the key idea  is to compute the conditional probability of positive prediction of the classifier, $\Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}]$, for the compound sensitive group $\sensitive = \mathbf{a}$ by solving an appropriately designed SSAT formula.
 For simplicity, we  initially make assumptions on the classifier $ \alg $ and discuss practical relaxations later in this section.
-We first assume $\alg$ to be represented as a CNF formula, denoted by $\phi_{\hat{Y}}$, such that the prediction $ \hat{Y} = 1 $ when $ \phi_{\hat{Y}}$ is satisfied and  $\hat{Y} =0$ otherwise.
+We first assume $\alg$ to be represented as a CNF formula, denoted by $\phi_{\widehat{Y}}$, such that the prediction $ \widehat{Y} = 1 $ when $ \phi_{\widehat{Y}}$ is satisfied and  $\widehat{Y} =0$ otherwise.
 Additionally, all features $ \nonsensitive \cup \sensitive $ are assumed to be Boolean variables.
-Finally, we consider independence probability assumption of non-sensitive features $ \nonsensitive $, where $p_i \triangleq \Pr[X_i = 1] $ is the independent probability of $ X_i $.
+Finally, we consider independence probability assumption of non-sensitive features $ \nonsensitive $, where $p_i \triangleq \Pr[X_i = 1] $ is the marginal probability of $ X_i $.
 
-Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the conditional probability $\Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}]$.
+Now, we define an RE-SSAT formula $\Phi_{\mathbf{a}}$ to compute the conditional probability $\Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}]$ using the probability of satisfaction of $\Phi_{\mathbf{a}}$.
 In the prefix of $ \Phi_{\mathbf{a}} $,  all non-sensitive features $ \nonsensitive $ are assigned randomized quantifiers and they are followed by sensitive features $ \sensitive $ with existential quantifiers.
-In addition, the CNF formula $ \phi $ in the SSAT formula $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $.
+In addition, the CNF formula $ \phi $ in the SSAT formula $ \Phi_{\mathbf{a}} $ is constructed such that $ \phi $ encodes the event inside the target probability $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $.
 In order to specify the sensitive group $ \sensitive = \mathbf{a} $, we take the conjunction of the Boolean variables in $ \sensitive $ that symbolically specifies the compound sensitive group $ \sensitive = \mathbf{a} $.
@@
 Hence, the compound groups $ [\textrm{White}, \textrm{male}] $ and $[\textrm{Colour}, \textrm{female}]$ are represented by $ R \wedge S $ and $ \neg R \wedge \neg S $, respectively.
-Thus, the RE-SSAT formula for computing the probability  $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ is
+Thus, the RE-SSAT formula for computing the probability  $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ is
 \begin{equation}	\label{fairness_justicia_eq:re}
@@
 	\Phi_{\mathbf{a}} := \underbrace{\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}}_{\text{non-sensitive features}},  &\underbrace{\exists A_1,\dots, \exists A_{\numsensitive}}_{\text{sensitive features}},
-\phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}).\notag
+\phi_{\widehat{Y}} \wedge (\sensitive =\mathbf{a}).\notag
 \end{split}
 \end{equation}
-In the RE-SSAT formula $ \Phi_{\mathbf{a}} $, existentially quantified variables  $ \{A_1, \dots, A_{\numsensitive}\} $ are assigned values  according  to the constraint $ \sensitive =\mathbf{a} $.
-\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Next,  an SSAT solver computes the probability $ \Pr[\Phi_{\mathbf{a}}] $ by considering the random values of $ \{X_1, \dots, X_{\numnonsensitive}\} $ while fixing the assignment of $ \{A_1, \dots, A_{\numsensitive}\} $.
-Therefore, $ \Pr[\Phi_{\mathbf{a}}] $ equals the conditional probability of positive prediction of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $, for the sensitive group $ \sensitive = \mathbf{a} $.
+In the RE-SSAT formula $ \Phi_{\mathbf{a}} $, existentially quantified variables  $ \{A_1, \dots, A_{\numsensitive}\} $ are assigned Boolean values  according  to the constraint $ \sensitive =\mathbf{a} $.\footnote{An RE-SSAT formula becomes an R-SSAT formula when the assignment to the existential variables are fixed.} Next,  an SSAT solver computes the probability $ \Pr[\Phi_{\mathbf{a}}] $ by considering the random values of $ \{X_1, \dots, X_{\numnonsensitive}\} $ while fixing the assignment of $ \{A_1, \dots, A_{\numsensitive}\} $.
+Therefore, $ \Pr[\Phi_{\mathbf{a}}] $ equals the conditional probability of positive prediction of the classifier, $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $, for the sensitive group $ \sensitive = \mathbf{a} $.
 
-For simplicity, we have described the computation of conditional probability $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ without considering the correlation among sensitive and non-sensitive features.
-In reality, correlation exists among these features.
+For simplicity, we have described the computation of conditional probability $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ without considering the correlation among sensitive and non-sensitive features.
+In reality, correlation exists among these features (for a detailed study on feature correlations in fairness verification, we refer to Chapter~\ref{chapter:fvgm}).
 As a result, non-sensitive features may have different conditional distributions for different sensitive groups.
@@
 \subsubsection{Measuring Fairness Metrics}
-As we compute $ \Pr[\Phi_\mathbf{a}] = \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}] $ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics.
-To this end, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in A$ by solving an exponential number (with $ \numsensitive $) of SSAT formulas.
+As we compute $ \Pr[\Phi_\mathbf{a}] = \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ by solving the SSAT formula $ \Phi_\mathbf{a} $, we  use $ \Pr[\Phi_\mathbf{a}] $ to measure different fairness metrics.
+To this end, we compute $ \Pr[\Phi_\mathbf{a}] $ for all compound groups $\mathbf{a} \in \sensitive $ by solving an exponential number (with $ \numsensitive $) of SSAT formulas.
 We elaborate this enumeration approach, namely {\justiciaenum}, in Algorithm~\ref{fairness_justicia_algo:enum}  (Line~\ref{fairness_justicia_algo:justicia_enum_begin}--\ref{fairness_justicia_algo:justicia_enum_end}).
@@
 \begin{algorithm}[t!]
-	\caption{\justicia: SSAT-based Fairness Verifier}
+	\caption{\justicia: An SSAT-based Fairness Verifier}
 	\label{fairness_justicia_algo:enum}
-	\footnotesize
 	\begin{algorithmic}[1]
-		\Function{{\justiciaenum}}{$ \nonsensitive, \sensitive,\hat{Y} $}
+		\Function{{\justiciaenum}}{$ \nonsensitive, \sensitive,\widehat{Y} $}
 		\label{fairness_justicia_algo:justicia_enum_begin}
-		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y} = 1) $
+		\State $ \phi_{\widehat{Y}} := \mathsf{CNF}(\widehat{Y} = 1) $
 		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
@@
 		\State $ p_{i} \leftarrow \Pr[X_i = 1|\sensitive = \mathbf{a}], \forall X_i \in \nonsensitive $
-		\State $ \phi := \phi_{\hat{Y}} \wedge (\sensitive =\mathbf{a}) $
+		\State $ \phi := \phi_{\widehat{Y}} \wedge (\sensitive =\mathbf{a}) $
 		\State $  \Phi_\mathbf{a} := \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive},  \phi $
@@
 		\EndFor
-		\State \Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
+		\State\Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
 		\label{fairness_justicia_algo:justicia_enum_end}
 		\EndFunction
+		\Statex
 
-		\Function{{\justicialearn}}{$ \nonsensitive, \sensitive,\hat{Y} $}
+		\Function{{\justicialearn}}{$ \nonsensitive, \sensitive,\widehat{Y} $}
 		\label{fairness_justicia_algo:justicia_learn_begin}
-		\State $ \phi_{\hat{Y}} := \mathsf{CNF}(\hat{Y}  = 1) $
+		\State $ \phi_{\widehat{Y}} := \mathsf{CNF}(\widehat{Y}  = 1) $
 		\State $ p_{i} \leftarrow \Pr[X_i=1], \forall X_i \in \nonsensitive $
-		\State $  \Phi_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\hat{Y}} $
-		\State $  \Phi'_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\hat{Y}} $
-		\State \Return $ \mathsf{SSAT}(\Phi_\mathsf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathsf{ER}) $
+		\State $  \Phi_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \phi_{\widehat{Y}} $
+		\State $  \Phi'_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\widehat{Y}} $
+		\State\Return $ \mathsf{SSAT}(\Phi_\mathsf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathsf{ER}) $
 		\label{fairness_justicia_algo:justicia_learn_end}
@@
 Moreover, to compute equalized odds, we call {\justicia} twice, one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $.
-In both calls, we compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of equalized odds.
-For measuring path-specific causal fairness, we compute  $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ by conditioning the distribution $ \mathcal{D} $ by mediator features $ \mathbf{Z} $ and take their difference.
+In both calls, we compute $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of equalized odds.
+For measuring path-specific causal fairness, we compute  $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ by conditioning the distribution $ \mathcal{D} $ by mediator features $ \mathbf{Z} $ and take their difference.
 Thus, {\justiciaenum} allows us to compute different group and causal fairness metrics using a unified algorithmic framework.
 
-\subsection{\red{Inference Approach using ER-SSAT Encoding}}
+\subsection{Inference Approach using ER-SSAT Encoding}
 \label{fairness_justicia_sec:learn_ssat}
-In most practical problems, there can be exponentially many compound sensitive groups due to the combinatorial combination of categorical sensitive features.
+In most practical problems, there can be exponentially many compound sensitive groups due to the combination of categorical sensitive features.
 As a result, the enumeration approach may suffer from scalability issues due to an exponential number of calls to the SSAT solver.
-To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the the most favored sensitive group with the maximum conditional probability of positive prediction of the classifier and another for inferring the least favored sensitive group with the minimum conditional probability of positive prediction of the same classifier.
-As shown above, these two probabilities are sufficient to measure different group and causal fairness metrics.
+To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier.
+As discussed above, these two probabilities are sufficient to measure different group and causal fairness metrics.
+
+\todo{End of revision}
 
@@
 In an ER-SSAT formula, $ \Pr[\Phi] $ is the \textit{maximum} satisfying probability of $ \Phi $ over the optimal assignment of existentially quantified variables given the randomized quantified variables (by Semantic 2, Sec.~\ref{fairness_justicia_sec:ssat}).
-In this chapter, we leverage this property to compute the most favored sensitive group with the highest probability of positive prediction of a classifier.
+In this chapter, we leverage this property to compute the most favored sensitive group with the highest probability of positive prediction of the classifier.
 In particular, we consider the following ER-SSAT formula:
@@
 \Phi_{\mathsf{ER}} := \exists A_1,\dots, \exists A_{\numsensitive},
- \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}.
+ \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\widehat{Y}}.
  \label{fairness_justicia_eq:er}
@@
 
-The CNF formula $\phi_{\hat{Y}}$ in $ \Phi_{\mathsf{ER}} $ is the CNF translation  of the classifier $ \hat{Y} = 1 $ without any specification of the compound sensitive group.
+The CNF formula $\phi_{\widehat{Y}}$ in $ \Phi_{\mathsf{ER}} $ is the CNF translation  of the classifier $ \widehat{Y} = 1 $ without any specification of the compound sensitive group.
 Therefore, as we solve $ \Phi_{\mathsf{ER}} $, we find the optimal assignment to the existentially quantified variables $ A_1 = a^{\max}_1, \dots,A_{\numsensitive} = a^{\max}_{\numsensitive} $ for which the probability of satisfaction of the ER-SSAT formula $ \Pr[\Phi_{\mathsf{ER}}] $ becomes maximum.
-Thus, we compute  the most favored group $ \red{\mathbf{a}_{\max}} \triangleq [ a^{\max}_1, \dots, a^{\max}_{\numsensitive} ] $ achieving the highest probability of positive prediction of the classifier.
+Thus, we compute  the most favored group $ \mathbf{a}_{\max} \triangleq [ a^{\max}_1, \dots, a^{\max}_{\numsensitive} ] $ achieving the highest probability of positive prediction of the classifier.
 
 \subsubsection{Inferring the Least Favored Sensitive Group}
-In order to infer the least favored sensitive group by a classifier, we  compute the \textit{minimum} conditional probability of positive prediction of the classifier with respect to all sensitive groups given the random values of the non-sensitive features.
+In order to infer the least favored sensitive group of the classifier, we  compute the \textit{minimum} conditional probability of positive prediction of the classifier with respect to all sensitive groups given the random values of the non-sensitive features.
 To this end, we  solve a `universal-random' (UR) SSAT formula with  universal quantifiers over sensitive features and randomized quantification over non-sensitive features (by Semantic 3, Sec.~\ref{fairness_justicia_sec:ssat}).
@@
 \Phi_{\mathsf{UR}} := \forall A_1,\dots, \forall A_{\numsensitive},
-\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\hat{Y}}
+\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \phi_{\widehat{Y}}
 \label{fairness_justicia_eq:ar}
@@
 
-\red{Solving an UR-SSAT formula raises several practical issues.} As a result, there is an unavailability of solvers for UR-SSAT formulas.
+Solving an UR-SSAT formula raises several practical issues and thus, there is an unavailability of a UR-SSAT solver.
 To resolve this problem, we leverage the \textit{duality} between UR-SSAT  and ER-SSAT formulas, where we solve an UR-SSAT formula on the CNF $ \phi $ using the solution of an ER-SSAT formula on the complemented CNF $ \neg \phi $~\cite{littman2001stochastic}.
@@
 	\Phi'_{\mathsf{ER}} := \exists A_1,\dots, \exists A_{\numsensitive},
-	\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \neg \phi_{\hat{Y}}
+	\R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive},   \; \neg \phi_{\widehat{Y}}
 	\label{fairness_justicia_eq:er_complement}
@@
 \end{lemma}
+
+\begin{proof}[Proof of Lemma~\ref{fairness_justicia_thm:dual}]
+	Both $ \Phi_{\mathsf{UR}} $ and $ \Phi'_{\mathsf{ER}} $ have  random quantified variables in the identical order in the prefix. According to the definition of SSAT formulas,
+	\begin{equation*}
+	\Pr[\Phi_{\mathsf{UR}}] = \min\limits_{a_1, \dots,a_{\numsensitive}} \Pr[\phi_{\widehat{Y}}] \text{ and } \Pr[\Phi'_{\mathsf{ER}}] = \max\limits_{a_1, \dots,a_{\numsensitive}} \Pr[\neg\phi_{\widehat{Y}}],
+	\end{equation*}
+
+	where $ \Pr[\phi_{\widehat{Y}}] $ and $ \Pr[\neg \phi_{\widehat{Y}}] $ are both computed for the random values of non-sensitive features $ \nonsensitive $.
+
+	Therefore, we derive the following duality between ER-SSAT and UR-SSAT,
+	\begin{equation}
+	\begin{split}
+	\Pr[\Phi'_{\mathsf{ER}}] &= \max\limits_{a_1, \dots,a_{\numsensitive}} \Pr[\neg\phi_{\widehat{Y}}]  \\
+	&= \min\limits_{a_1, \dots,a_{\numsensitive}} (1 - \Pr[\phi_{\widehat{Y}}])\\
+	&= 1 - \min\limits_{a_1, \dots,a_{\numsensitive}}  \Pr[\phi_{\widehat{Y}}]\\
+	&= 1 - \Pr[\Phi_{\mathsf{UR}}].\notag
+	\end{split}
+	\end{equation}
+\end{proof}
 
@@
 	\begin{algorithmic}[1]
-		\Function{{\justicialearn}}{$ X,A,\hat{Y} $}
-		\State $ \phi_{\hat{Y}} = \mathsf{CNF}(\hat{Y}  = 1) $
+		\Function{{\justicialearn}}{$ X,A,\widehat{Y} $}
+		\State $ \phi_{\widehat{Y}} = \mathsf{CNF}(\widehat{Y}  = 1) $
 		\State $ p_{i} = \mathsf{CalculateProb}(x_i), \forall x_i \in X $
-		\State $  \Phi_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \phi_{\hat{Y}} $
-		\State $  \Phi'_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \neg \phi_{\hat{Y}} $
+		\State $  \Phi_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \phi_{\widehat{Y}} $
+		\State $  \Phi'_\mathbf{ER} = \exists a_1,\dots, \exists a_{\numsensitive}, \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}. \; \neg \phi_{\widehat{Y}} $
 		\State \Return $ \mathsf{SSAT}(\Phi_\mathbf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathbf{ER}) $
@@
 
-In the ER-SSAT formula in Eq.~\eqref{fairness_justicia_eq:er_complement}, we need to negate the classifier $ \phi_{\hat{Y}} $ to another CNF formula $ \neg \phi_{\hat{Y}} $.
+In the ER-SSAT formula in Eq.~\eqref{fairness_justicia_eq:er_complement}, we need to negate the classifier $ \phi_{\widehat{Y}} $ to another CNF formula $ \neg \phi_{\widehat{Y}} $.
 The na\"ive approach of negating one CNF to another CNF generates an exponential number of new clauses.
 Here, we apply Tseitin transformation for the negation, which increases the number of clauses linearly while introducing a linear number of new variables~\cite{tseitin1983complexity}.
-As an alternative approach, we also directly encode the binary classifier $\alg$ for the negative class label $\hat{Y} = 0$ as a CNF formula and pass it to $\Phi'_{\mathsf{ER}} $, whenever it is possible.
-The last approach is generally more efficient than the other approaches as the resulting CNF is often smaller.
+As an alternative approach, we directly encode the binary classifier $\alg$ for the negative class label $\widehat{Y} = 0$ as a CNF formula and pass it to $\Phi'_{\mathsf{ER}} $, whenever it is possible.
+The latter approach is generally more efficient than the former approach as the resulting CNF is often smaller.
 
@@
 	\label{fairness_justicia_example:er_ssat}
-	Here, we illustrate the ER-SSAT encoding for inferring the most favored and the least favored sensitive group of a classifier in the presence of multiple compound sensitive groups.
+	Here, we illustrate the ER-SSAT encoding for inferring the most favored and the least favored sensitive group of a classifier in the presence of compound sensitive groups.
 As the example in Figure~\ref{fairness_justicia_fig:fair_example} is degenerate for this purpose, we introduce another Boolean sensitive feature `sex' $ \in $ \{male, female\}.
-Consider a Boolean variable $ S $ for sex where the literal $ S $ denotes sex = male.
-With this new sensitive feature, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where variables $ F,I,J $, and $ A $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
+We consider a Boolean variable $ S $ for sex where the literal $ S $ denotes sex = male.
+With this new sensitive feature, let the classifier be  $\alg \triangleq (\neg F \vee I \vee S) \wedge (F \vee J)$, where variables corresponding to sensitive features $ F,I$, and  $J $ have same distributions as discussed in Example~\ref{fairness_justicia_example:re_ssat}.
 Hence, we obtain the following ER-SSAT formula of $\alg$ to infer the most favored sensitive group:
@@
 	As we solve $ \Phi_{\mathsf{ER}} $, we infer that the optimal assignment to the existential variables $ \sigma(S) = 1, \sigma(A) = 0$, which implies that `male individuals with age $ < 40 $' is the most favored group with probability of positive prediction computed as $ \Pr[\Phi_{\mathsf{ER}}] = 0.46$.
-Similarly, to infer the least favored group, we negate the CNF of the classifier $\alg$ to obtain the following ER-SSAT formula:
+Similarly, to infer the least favored group, we negate the CNF translation of the classifier $\alg$ to obtain the following ER-SSAT formula:
 
@@
 	Solving $ \Phi'_{\mathsf{ER}} $, we learn the optimal assignment $ \sigma(S) = 0, \sigma(A) = 0  $ and $  \Pr[\Phi'_{\mathsf{ER}}] = 0.57 $.
-Thus, `female individuals with age $ < 40 $' constitute the least favored group with probability of positive prediction:  $ 1-0.57 = 0.43$.
+Thus, `female individuals with age $ < 40 $' constitute the least favored group with probability of positive prediction as  $ 1-0.57 = 0.43$.
 	Thus, {\justicialearn} allows us to infer the most and least favored sensitive groups and the corresponding discrimination.
@@
 Additionally, for the same classifer, let $ \Phi_{\mathsf{ER}} $ be an ER-SSAT formula for inferring the most favored sensitive group and $ \Phi_{\mathsf{UR}} $ be a UR-SSAT formula for inferring the least favored sensitive group.
-Therefore,	$\max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   	and	$\min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.\improvement{Rewrite this lemma in Appendix.}
+Therefore,	$\max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]$   	and	$\min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]$.
 \end{lemma}
+
+\begin{proof}[Proof of Lemma~\ref{fairness_justicia_lm:equivalence}] 
+	It is trivial that the probability of positive prediction of the classifier for the most favored group $ \mathbf{a}_{\max} $ is the maximum computed probability of all compound groups $ \mathbf{a} \in \sensitive $.
+Similar argument holds for the least favored group $ \mathbf{a}_{\min} $, which obtains the minimum probability of positive prediction of the classifier among all compound groups $ \mathbf{a} \in A $.
+
+	By construction of the SSAT formulas, $ \Pr[\Phi_{\mathsf{ER}}] $ and $ \Pr[\Phi_{\mathsf{UR}}] $ are the probabilities corresponding to the groups $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively.
+Now, since $ \Pr[\Phi_{\mathbf{a}}] $ is the probability for the group $ \sensitive = \mathbf{a} $, we derive the following.
+
+	\begin{equation*}
+	\begin{split}
+	\max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{ER}}]
+	\text{ and }
+	\min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}] = \Pr[\Phi_{\mathsf{UR}}]
+	\end{split}
+	\end{equation*}
+\end{proof}
+
 \iffalse
@@
 	\[ k = O\left(\frac{1}{\epsilon^2} (n + \ln(1/\delta))  \right)\]
-	where $\hat{p} - p \leq \epsilon$ with probability $1-\delta$.
+	where $\widehat{p} - p \leq \epsilon$ with probability $1-\delta$.
 \end{theorem}
@@
 %	\begin{algorithmic}[1]
-%		\Function{{\justiciacond}}{$ X,A,\hat{Y} $}
-%		\State $ \phi_{\hat{Y}} = \mathsf{CNF}(\hat{Y}  = 1) $
+%		\Function{{\justiciacond}}{$ X,A,\widehat{Y} $}
+%		\State $ \phi_{\widehat{Y}} = \mathsf{CNF}(\widehat{Y}  = 1) $
 %		\ForAll{$\mathbf{a} \in A$ }
@@
 %		\State $  \Phi_\mathbf{a} = \R^{p_{1}}x_1, \dots, \R^{p_{\numnonsensitive}}x_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive}.
-\; \phi_{\hat{Y}} \wedge \mathsf{CNF}(\sensitive =\mathbf{a}) $
+\; \phi_{\widehat{Y}} \wedge \mathsf{CNF}(\sensitive =\mathbf{a}) $
 %		\State $ \Pr[\Phi_\mathbf{a}]  = \ma
@@
 
-Let us consider that $\hat{p_i}$ is the estimated probability of a Boolean variable $ \bool_i $ being assigned to $ 1 $ from $k$-samples and $p_i$ is the true probability according to $ \mathcal{D} $.
-%If $ | p_i - \hat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.
-\red{Thus, the true satisfying probability $p$ of an SSAT formula $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $: $p = \sum_\sigma \prod_{l_i \in \sigma}l_i * p_i + (1 - l_i) * (1 - p_i)$, where the literal $ l_i \in \{0,1\} $ denotes either the variable $ B_i $ (when $ l_i = 1 $) or its negation (when $ l_i = 0 $).
-}This probability $ p $ is estimated as $\hat{p}$ using $k$-samples from the distribution $\mathcal{D}$ such that $\hat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$.
-%However, the true satisfying probability $ \hat{p} $  of $ \Phi $ is 
+Let us consider that $\widehat{p_i}$ is the estimated probability of a Boolean variable $ \bool_i $ being assigned to $ 1 $ from $ n $ samples, and $p_i$ is the true probability of $ \bool_i = 1 $ according to $ \mathcal{D} $.
+%If $ | p_i - \widehat{p_i}| \le \epsilon $, i.e., $ \epsilon $ additive error for small $ \epsilon \approx 0 $, we want to compute the error of probability $ p $ of the satisfaction of the SSAT formula $ \Phi $.
+Thus, the true satisfying probability $p$ of an SSAT formula $ \Phi $ is the weighted sum of all satisfying assignments of the CNF $ \phi $, $p = \sum_{\sigma^*} \prod_{L_i \in {\sigma^*}}L_i * p_i + (1 - L_i) * (1 - p_i)$, where the literal $ L_i \in \{0,1\} $ denotes either the variable $ B_i $ when $ L_i = 1 $ or its negation when $ L_i = 0 $.
+This probability $ p $ is estimated as $\widehat{p}$ using $n$ samples from the distribution $\mathcal{D}$ such that $\widehat{p} \leq \epsilon_0 p$ for $\epsilon_0 \geq 1$.
+%However, the true satisfying probability $ \widehat{p} $  of $ \Phi $ is 
 %\[
-%\hat{p} = \sum_\sigma \prod_{\bool_i \in \sigma}\hat{p_i}  \leq \sum_\sigma \prod_{\bool_i \in \sigma}(1 \pm \epsilon_i) p_i = \prod_{i =1}^m (1 \pm \epsilon_i) p = \epsilon_0 p
+%\widehat{p} = \sum_\sigma \prod_{\bool_i \in \sigma}\widehat{p_i}  \leq \sum_\sigma \prod_{\bool_i \in \sigma}(1 \pm \epsilon_i) p_i = \prod_{i =1}^m (1 \pm \epsilon_i) p = \epsilon_0 p
 %\]
@@
 	For an ER-SSAT problem, the sample complexity is given by 
-	$ k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0} \right),$
-	where $\frac{\hat{p}}{p} \leq \epsilon_0$ with probability $1-\delta$ such that $\epsilon_0 \geq 1$.
+	$ n = O\left((\numsensitive+ \ln(1/\delta))\frac{\ln \numnonsensitive}{\ln \epsilon_0} \right),$
+	where $\frac{\widehat{p}}{p} \leq \epsilon_0$ with probability $1-\delta$ such that $\epsilon_0 \geq 1$.
 \end{theorem}
-%Theorem~\ref{fairness_justicia_thm:sample} states that for the prescribed number of samples the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ would also satisfy $\hat{DI} \leq  \epsilon_0 DI, $ and $\hat{SP} \leq 2\epsilon_0 SP$.
+\todo{Check correctness of the theorem}
+%Theorem~\ref{fairness_justicia_thm:sample} states that for the prescribed number of samples the estimated disparate impact $\widehat{DI}$ and statistical parity $\widehat{SP}$ would also satisfy $\widehat{DI} \leq  \epsilon_0 DI, $ and $\widehat{SP} \leq 2\epsilon_0 SP$.
 %Apply Hoeffding inequality for each term and then use Union bound for $2^n$ existential variables.
 \begin{corollary}
-	If $k$ samples are considered from the  distribution in {\justicia} such that 
-	$
-	k = O\left((n+ \ln(1/\delta))\frac{\ln m}{\ln \epsilon_0}\right),
-	$
-	the estimated disparate impact $\hat{DI}$ and statistical parity $\hat{SP}$ satisfy, with probability $1-\delta$,
-	$
-	\hat{DI} \leq  \epsilon_0 DI, \quad \text{and} \quad \hat{SP} \leq 2\epsilon_0 SP.
-	$
+	\label{fairness_justicia_cor:error}
+	If $n$ samples are considered from the  distribution in {\justicia} such that 
+	\[
+	n = O\left((\numsensitive + \ln(1/\delta))\frac{\ln \numnonsensitive}{\ln \epsilon_0}\right),
+	\]
+
+	the estimated disparate impact    $\widehat{\mathsf{DI}}$ and statistical parity  $\widehat{\mathsf{SP}}$ satisfy, with probability $1-\delta$,
+	\[
+	\widehat{\mathsf{DI}} \leq  \epsilon_0 \mathsf{DI},\text{ and }  \widehat{\mathsf{SP}} \leq 2\epsilon_0 \mathsf{SP}.
+	\]
 \end{corollary}
-This implies that given a binary classifier represented as a CNF formula and a distribution of sensitive and non-sensitive features, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
-Thus, {\justicia} is a sound framework of fairness verification with high probability.
+\todo{Proof is wrong}
+
+\begin{proof}[Proof of Corollary~\ref{fairness_justicia_cor:error}] 
+By Theorem~\ref{fairness_justicia_thm:sample}, we get that for $n$ samples obtained from the distribution, where
+
+$$ n \geq (\numsensitive + \ln(1/\delta))\frac{\ln \numnonsensitive}{\ln \epsilon_0},$$
+
+the estimated probability of satisfaction of SSAT formulas for the most and least favored sensitive groups, referred as $\widehat{p}_{max}$ and  $\widehat{p}_{min}$, respectively satisfy
+
+\begin{equation}
+\widehat{p}_{max} \leq \epsilon_0 \max_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}]
+\text{ and }
+\widehat{p}_{min} \leq \epsilon_0 \min_{\mathbf{a} \in \sensitive} \; \Pr[\Phi_{\mathbf{a}}].\notag
+\end{equation}
+
+with probability $ 1-\delta $.
+Thus, the estimated value of disparate impact will satisfy
+
+$$\widehat{\mathsf{DI}} \triangleq \frac{\widehat{p}_{max}}{\widehat{p}_{min}} \leq \epsilon_0  \frac{p_{max}}{p_{min}} \leq \epsilon_0 \mathsf{DI},$$
+
+and statistical parity will satisfy
+
+$$\widehat{SP} \triangleq \abs{\widehat{p}_{max} - \widehat{p}_{min}} \leq  \epsilon_0  \abs{p_{max} - p_{min}} \leq \epsilon_0 SP,$$
+
+with probability $1-\delta$.
+\end{proof}
+
+\red{This implies that given a binary classifier represented as a CNF formula and a distribution of sensitive and non-sensitive features, {\justicia} can verify independence and separation notion of fairness up to an error level $\epsilon_0$ and $2\epsilon_0$ with probability $1-\delta$.
+Thus, {\justicia} is a sound framework of fairness verification with high probability.}
 %\red{Proof of all theorems for arxiv version}
@@
 \begin{corollary}[Hypothesis]
-	If $k$ samples are considered in \justicia and the estimated group fairness value $\hat{GF}$ satisfies
+	If $k$ samples are considered in \justicia and the estimated group fairness value $\widehat{GF}$ satisfies
 	\begin{equation*}
-	1- e^{\epsilon_0} \leq \frac{\hat{GF}}{GF} \leq 1 + e^{\epsilon_0}
+	1- e^{\epsilon_0} \leq \frac{\widehat{GF}}{GF} \leq 1 + e^{\epsilon_0}
 	\end{equation*}

### 2022-06-06 chapters/fairness/justicia/introduction.tex
upto theoretical analysis of justicia

@@
 \subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
-In this chapter, we focus on a popular concept of fairness, known as group fairness.\change{Add causal fairness here}  Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
+In this chapter, we focus on a popular concept of fairness, known as group fairness.
+Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
 Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
-Separation metrics\unsure{Incorrect definition}, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth label~\cite{hardt2016equality}.
+Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth label~\cite{hardt2016equality}.
 Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
@@
 \begin{itemize}
-	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of fairness for different datasets and classifiers.
+	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group fairness and causal fairness metrics for different datasets and classifiers.
 	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.

### 2022-06-07 chapters/fairness/justicia/discussion.tex
revising justicia

@@
 \section{Chapter Summary}
-Though formal verification of different fairness metrics of an ML algorithm for different datasets is an important question, existing verifiers are not scalable, accurate, and extendable to non-Boolean sensitive features.
+Formal verification of different fairness metrics of machine learning for different datasets is an important question.
+Existing fairness verifiers, however, are not scalable, accurate, and extendable to non-Boolean sensitive features.
 We propose a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups.
@@
 
-\red{Our work opens up several new directions of research.
+Our work opens up several new directions of research.
 One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs.
-Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.}
+Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
 %Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2022-06-07 chapters/fairness/justicia/experiments.tex
revising justicia

@@
 \label{fairness_justicia_sec:experiments}
-In this section, we discuss the empirical studies to evaluate the performance of {\justicia} in verifying different fairness metrics.
+In this section, we discuss the empirical studies to evaluate the performance of {\justicia} in verifying different fairness metrics and algorithms.
 We first discuss the experimental setup and the objective of the experiments and then evaluate the experimental results.
@@
 
-For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art distributional verifiers FairSquare and VeriFair, and also a sample-based fairness measuring tool: AIF360.
+For comparative evaluation of {\justicia}, we have experimented with two state-of-the-art probabilistic fairness verifiers FairSquare and VeriFair, and a sample-based fairness measuring tool: AIF360.
 In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner.
@@
 	\item Can {\justicia} verify fairness in the presence of compound sensitive groups?
-	\item How robust is {\justicia} in comparison to sample-based tools like AIF360 for varying sample sizes?
+	\item How robust is {\justicia} in comparison to sample-based tools such as AIF360 for varying sample sizes?
 	\item How do the computational efficiencies of {\justicialearn} and {\justiciaenum} compare?
@@
 In Table~\ref{fairness_justicia_tab:synthetic}, we show the computed fairness metrics by {\justicia}, FairSquare, VeriFair, and AIF360.
-We observe that {\justicia} and AIF360  yield more accurate estimates of DI and SP compared against the ground truth values of fairness metrics with less than $1\%$ error.
-In contrast, FairSquare and VeriFair  estimate DI to be $0.99$ and thus, being unable to verify the fairness violation.
+We observe that {\justicia} and AIF360  yield more accurate estimates of disparate impact and statistical compared against the ground truth values of fairness metrics with less than $1\%$ error.
+In contrast, FairSquare and VeriFair  estimate disparate impact to be $0.99$ and thus, being unable to verify the fairness violation.
 Therefore, {\justicia} is significantly more accurate than the existing formal verifiers: FairSquare and VeriFair.
@@
 We observe that {\justicia} shows impressive scalability than the competing verifiers.
-Particularly, {\justicia} is $ 1 $ to $ 2 $ orders of magnitude faster than FairSquare and  $ 1 $ to $ 3 $ orders of magnitude faster than VeriFair.
+Particularly, among benchmarks where all three verifiers output results, {\justicia} is $ 1 $ to $ 2 $ orders of magnitude faster than FairSquare and  $ 1 $ to $ 3 $ orders of magnitude faster than VeriFair.
 Additionally, FairSquare times out in most  benchmarks.
@@
 	\subfloat{
-		\includegraphics[scale=.2]{figures/fairness/justicia/sensitive_attribute_race_di_Adult_DT_RE.pdf}
+		\includegraphics[scale=.35]{figures/fairness/justicia/sensitive_attribute_race_di_Adult_DT_RE.pdf}
 	}
 	\subfloat{
-		\includegraphics[scale=.2]{figures/fairness/justicia/sensitive_attribute_race_spd_Adult_DT_RE.pdf}
+		\includegraphics[scale=.35]{figures/fairness/justicia/sensitive_attribute_race_spd_Adult_DT_RE.pdf}
 	}
 	\caption[Fairness verification on compound sensitive groups]{Fairness metrics measured by {\justicia} for different sensitive groups in the Adult dataset.
-The number within parenthesis in the xticks denotes total compound groups.}
+The number within parenthesis in the xticks denotes total compound sensitive groups, which increases due to the increasing combination of sensitive features.
+For higher sensitive groups, fairness becomes worse.}
 	\label{fairness_justicia_fig:sensitive_groups}
@@
 
-\subsubsection{Verification: Detecting Compounded Discrimination in Protected Groups.}
+\subsubsection{Verification: Detecting Compounded Discrimination in Sensitive Groups.}
 We have tested {\justicia} for datasets consisting of multiple sensitive features and reported results in Figure~\ref{fairness_justicia_fig:sensitive_groups}.
@@
 Both of them pre-process to remove statistical bias from the dataset.
-We study the effectiveness of these algorithms using {\justicia} on three datasets each with two different sensitive features.
+We study the effectiveness of these algorithms using {\justicia} on different5 datasets each with two different sensitive features.
 In Table~\ref{fairness_justicia_tab:fair_algo_verification}, we report different fairness metrics on logistic regression and decision tree.
@@
 \subsubsection{Robustness: Stability to Sample Size.} 
-We have compared the robustness of {\justicia} with AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics. 
+We have compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics.
 In Figure~\ref{fairness_justicia_fig:sample-size}, AIF360 shows higher standard deviation for lower sample-size and the value decreases as  the sample-size increases. 
 In contrast, {\justicia} shows significantly lower ($\sim10\times$ to $100\times$) standard deviation for different sample-sizes. 
-The reason is that AIF360 empirically measures on a fixed test dataset whereas {\justicia} provides estimates over the data generating distribution.
+The reason is that AIF360 empirically measures on a fixed dataset whereas {\justicia} provides estimates over the distribution.
 Thus, {\justicia} is more robust than the sample-based verifier AIF360.
@@
 	\centering
-	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}}
-	\subfloat{\includegraphics[scale=.2]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}
+	\subfloat{\includegraphics[scale=.35]{figures/fairness/justicia/sampling_DI_after_Adult_rw_LR_race.pdf}}
+	\subfloat{\includegraphics[scale=.35]{figures/fairness/justicia/sampling_SPD_after_Adult_rw_LR_race.pdf}}
 
@@
 {\justicia} is more robust with variation of sample size than  AIF360.
-}
+Sample size $ = 1 $ denotes the full dataset considering all samples.}
 	\label{fairness_justicia_fig:sample-size}
@@
 	\begin{center}
-		\includegraphics[scale=.4]{figures/fairness/justicia/encoding_runtime_Adult_DT.pdf}
-		\includegraphics[scale=.4]{figures/fairness/justicia/encoding_runtime_Adult_IMLI.pdf}
+		\includegraphics[scale=.35]{figures/fairness/justicia/encoding_runtime_Adult_DT.pdf}
+		\includegraphics[scale=.35]{figures/fairness/justicia/encoding_runtime_Adult_IMLI.pdf}
 		\hfill
-		\caption[Runtime of different encodings in {\justicia}]{Runtime comparison of different encodings while varying total sensitive groups in the Adult dataset. \red{Label `learn' to be `infer'???}}
+		\caption[Runtime of different encodings in {\justicia}]{Runtime comparison of different encodings while varying total sensitive groups in the Adult dataset.}
 		\label{fairness_justicia_fig:runtime_diff_encodings}
@@
 \subsubsection{Comparative Evaluation of Different Encodings.}
-While both {\justiciaenum} and {\justicialearn}  have the same output according to Lemma~\ref{fairness_justicia_lm:equivalence},  {\justicialearn} encoding  improves exponentially  in runtime  than {\justiciaenum} encoding on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fairness_justicia_fig:runtime_diff_encodings}.
+Both {\justiciaenum} and {\justicialearn}  have the same output according to Lemma~\ref{fairness_justicia_lm:equivalence}.
+However, {\justicialearn}  improves exponentially  in runtime  than {\justiciaenum} on both decision tree and Boolean CNF classifiers as we vary the total compound groups in Figure~\ref{fairness_justicia_fig:runtime_diff_encodings}.
 {\justiciacond} ({\justiciaenum} encoding where we consider conditional probabilities of non-sensitive features w.r.t.

### 2022-06-07 chapters/fairness/justicia/practical.tex
workig on relaxed-cnf
minor edits in many parts of the thesis
revising justicia

@@
 The \textit{tree} itself is a disjunction of all paths and thus, a \textit{DNF (Disjunctive Normal Form)}.
-In order to derive a CNF of a decision tree, we first construct a DNF by including all paths terminating at leaves with negative class label ($ \hat{Y} = 0 $) and then complement the DNF to CNF using De Morgan's rule.
+In order to derive a CNF of a decision tree, we first construct a DNF by including all paths terminating at leaves with negative class label ($ \widehat{Y} = 0 $) and then complement the DNF to CNF using De Morgan's rule.
 
 \textit{Linear classifiers on Boolean features} are encoded into CNF formulas using pseudo-Boolean encoding~\cite{philipp2015pblib}.
-We consider a linear classifier  $ \mathbf{W}^T \mathbf{X} + b \ge 0 $ on Boolean features $ \mathbf{X} $ with weights $ \mathbf{W} \in \mathbb{R}^{|\mathbf{X}|} $ and bias $ b \in \mathbb{R} $.
+We consider a linear classifier  $ \mathbf{W} \mathbf{X} + b \ge 0 $ on Boolean features $ \mathbf{X} $ with weights $ \mathbf{W} \in \mathbb{R}^{|\mathbf{X}|} $ and bias $ b \in \mathbb{R} $.
 We first normalize $ \mathbf{W} $ and $ b $ in $ [-1,1] $ and then round to integers so that the decision boundary becomes a pseudo-Boolean constraint~\cite{roussel2009pseudo}.
-We then apply  pseudo-Boolean constraints to CNF translation~\cite{philipp2015pblib} to encode the decision boundary to CNF.
-This encoding usually introduces additional Boolean variables and results in large CNF.
+Then we apply  pseudo-Boolean constraints to CNF translation~\cite{philipp2015pblib} to encode the decision boundary to CNF.
+This encoding usually introduces additional Boolean variables and results in a large CNF.
 In order to generate a smaller CNF, we can trivially apply thresholding  on the weights to consider features with higher weights only.
-For instance, if the weight $  |w_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ w_i \in \mathbf{W} $, we can set $ w_i = 0 $.
+For instance, if the weight $  |W_i| \le \lambda $ for a threshold $ \lambda  \in \mathbb{R}^+$ and $ W_i \in \mathbf{W} $, we can set $ W_i = 0 $.
 Thus, features with lower weights (less important) do not appear in the encoded CNF.
@@
 For a binary decision tree, each feature, including the continuous ones, is compared against a constant at each node (except leaves) of the tree.
-We assign a Boolean variable to each node (except leaves), where the $ \{0,1\} $ assignment to the variable decides one of the two branches to choose from the current node.
+We assign a Boolean variable to each internal node of the tree (except leaves), where the $ \{0,1\} $ assignment to the variable decides one of the two branches to choose from the current node.
 
 Linear classifiers are generally trained on continuous features, where we apply discretization in the following way.
-Let us consider a continuous feature $X_c$, where $w$ is its weight during training.
-We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean features and recalculate the weight of each variable in $ \mathbf{B} $ based on $ w $.
-For the discretization of $X_c$, we consider the interval-based approach\footnote{Our implementation is agnostic to any discretization technique.}.
+Let us consider a continuous feature $X_c$, where $W$ is its weight during training.
+We discretize $ X_c $ to a set $ \mathbf{B} $ of Boolean features and recalculate the weight of each variable in $ \mathbf{B} $ based on $ W $.
+We consider the an interval-based approach for discretizing $ X_c $.
 For each interval in the continuous space of $X_c$, we consider a Boolean variable $B_i \in \mathbf{B}$, such that $ B_i $ is assigned $ 1 $ when the feature-value of $X_c$ lies within the $i^{\mathrm{th}}$ interval and $ B_i $ is assigned $ 0 $ otherwise.
-Following that, we assign the weight of $ B_i $ to be $ \mu_iw $, where $ \mu_i $ is the mean of feature values in the $i^{\mathrm{th}}$ interval.
+Following that, we assign the weight of $ B_i $ to be $ \mu_iW $, where $ \mu_i $ is the mean of feature values in the $i^{\mathrm{th}}$ interval.
 We can show that if we consider infinite number of intervals, $ X_c \approx \sum_i \mu_i B_i $.
-
-\noindent\makebox[\linewidth]{\rule{\paperwidth}{2pt}}

### 2022-06-07 chapters/fairness/fvgm/framework.tex
minor edits in many parts of the thesis
revision on fvgm

@@
 The core idea of verifying fairness of a classifier is to compute the probability of positive prediction of the classifier with respect to all compound sensitive groups.
-To this end, {\fvgm} solves a stochastic subset sum problem, {\stochastic}, that is equivalent to computing the probability of positive prediction of the classifier for the most (and the least) favored sensitive group.
+To this end, {\fvgm} solves a stochastic subset sum problem, {\stochastic}, that is equivalent to computing the probability of positive prediction of the classifier for the most and the least favored sensitive group\footnote{The most (resp.\ least) favored sensitive group obtains the maximum (resp.\ minimum) probability of positive prediction of the classifier.}.
 In this section, we first define {\stochastic} and present an efficient dynamic programming solution for {\stochastic}.
@@
 \textbf{Problem Formulation.}	
-Given a linear classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \hat{Y} $ and a probability distribution $ \mathcal{D} $ of $ \nonsensitive \cup \sensitive $, our objective is to compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ with respect to $ \mathcal{D} $.
+Given a linear classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $ and a probability distribution $ \mathcal{D} $ of $ \nonsensitive \cup \sensitive $, our objective is to compute $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ and $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ with respect to $ \mathcal{D} $.
 In this study, we express a linear classifier $\mathcal{M}$ as
 
-\[	\hat{Y} = \mathds{1}\Big[\sum_{i} w_{X_i}X_i + \sum_{j} w_{A_j}A_j \ge \tau\Big].\]
+\[	\widehat{Y} = \mathds{1}\Big[\sum_{i} W_{X_i}X_i + \sum_{j} W_{A_j}A_j \ge \tau\Big].\]
 
-Here, $ w $ denotes the weight (or coefficients) of a feature, $ \tau $ denotes the bias or the offset parameter of the classifier, and $\mathds{1}$ is an indicator function.
-Hence, the prediction $ \hat{Y} =1 $ if and only if the inner inequality holds.
+Here, $ W $ denotes the weight (or coefficients) of a feature, $ \tau $ denotes the bias or the offset parameter of the classifier, and $\mathds{1}$ is an indicator function.
+Hence, the prediction $ \widehat{Y} =1 $ if and only if the inner inequality holds.
 Thus, computing the maximum (resp.\ minimum) probability of positive prediction is equivalent to finding out the assignment of $A_j$'s for which the probability of satisfying the inner inequality is highest (resp.\ lowest).
 We reduce this computation into an instance of {\stochastic}.
-To perform this reduction, we assume  weights $ w $ and bias $ \tau $ as integers, and features $\nonsensitive \cup \sensitive $ as Boolean.
+To perform this reduction, we assume  weight $ W $ and bias $ \tau $ as integers, and features $\nonsensitive \cup \sensitive $ as Boolean.
 In Sec.~\ref{fvgm_sec:practical}, we relax these assumptions and extend to the practical settings.
@@
 Now, we formally describe the specification and semantics of {\stochastic}.
-{\stochastic} operates on  a set of Boolean variables $ \mathbf{B} = \{B_i\}_{i=1}^{n} \in \{0,1\}^{n} $, where $ w_i \in \mathbb{Z} $ is the weight of $ B_i $, and $n \triangleq |\mathbf{B}|$.
+{\stochastic} operates on  a set of Boolean variables $ \mathbf{B} = \{B_i\}_{i=1}^{n} \in \{0,1\}^{n} $, where $ W_i \in \mathbb{Z} $ is the weight of $ B_i $, and $n \triangleq |\mathbf{B}|$.
 Given a constant threshold $ \tau \in \mathbb{Z} $, {\stochastic} computes the \textit{probability} of a subset of $ \mathbf{B} $ with sum of weights of non-zero variables to be at least $ \tau $.
@@
 
-\[S(\mathbf{B}, \tau) \triangleq \Pr\Big[\sum_{i} w_iB_i \ge \tau \Big] \in [0,1].\]
+\[S(\mathbf{B}, \tau) \triangleq \Pr\Big[\sum_{i} W_iB_i \ge \tau \Big].\]
 
 Aligning with terminologies in stochastic satisfiability (SSAT)~\cite{littman2001stochastic}, we categorize the variables $ \mathbf{B} $ into two types: (i) \textit{chance variables} that are stochastic and have an associated probability of being assigned to $ 1 $ and (ii) \textit{choice variables} that we  optimize while computing $ S(\mathbf{B}, \tau) $.
-To specify the category of variables, we consider a \textit{quantifier} $ q_i \in \{\R^{p_i}, \exists, \forall\} $ for each $ B_i $.
+To specify the category of variables, we consider a \textit{quantifier} $ Q_i \in \{\R^{p_i}, \exists, \forall\} $ for each $ B_i $.
 Elaborately, $ \R^{p} $ is a \textit{random quantifier} corresponding to a chance variable $ B \in \mathbf{B} $, where  $ p\triangleq \Pr[B = 1]$.
@@
 
-Now, we formally present the semantics of $ S(\mathbf{B}, \tau) $ provided that each variable $ B_i $ has weight $ w_i $ and quantifier $ q_i $.
+Now, we formally present the semantics of $ S(\mathbf{B}, \tau) $ provided that each variable $ B_i $ has weight $ W_i $ and quantifier $ Q_i $.
 Let  $ \mathbf{B}[2:n] \triangleq \{B_j\}_{j=2}^{n} $ be the subset of $\mathbf{B}$ without the first variable $ B_1 $.
@@
  \mathds{1}[ \tau \le 0 ], \; \text{if } \textbf{B} = \emptyset \\
- S(\mathbf{B}[2:n],\tau - \max\{w_1, 0\}), \; \text{if } q_1 = \exists\\
- S(\mathbf{B}[2:n],\tau - \min\{w_1, 0\}), \; \text{if } q_1 = \forall\\
- p_1\times S(\mathbf{B}[2:n],\tau - w_1) +\\ (1-p_1)\times S(\mathbf{B}[2:n],\tau),\; \text{if } q_1 = \R^{p_1}
+ S(\mathbf{B}[2:n],\tau - \max\{W_1, 0\}), \; \text{if } Q_1 = \exists\\
+ S(\mathbf{B}[2:n],\tau - \min\{W_1, 0\}), \; \text{if } Q_1 = \forall\\
+ p_1\times S(\mathbf{B}[2:n],\tau - W_1) +\\ (1-p_1)\times S(\mathbf{B}[2:n],\tau),\; \text{if } Q_1 = \R^{p_1}
  \end{cases}
@@
 % 	\item $ S(\mathbf{B},\tau) = \mathds{1}[ \tau \le 0 ] $, if $ \textbf{B} = \emptyset $, %Here $ \mathds{1}[\text{true}] = 1 $ and $ \mathds{1}[\text{false}] = 0 $,
-% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \max\{w_1, 0\}) $, if $ q_1 =  \exists $,
-% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \min\{w_1, 0\}) $, if $ q_1 =  \forall $,
-% 	\item $ S(\mathbf{B},\tau) = p_1S(\mathbf{B}_{-1},\tau - w_1) + (1-p_1) S(\mathbf{B}_{-1},\tau) $, if $ q_1 = \R^{p_1}  $.
+% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \max\{W_1, 0\}) $, if $ Q_1 =  \exists $,
+% 	\item $ S(\mathbf{B},\tau) = S(\mathbf{B}_{-1},\tau - \min\{W_1, 0\}) $, if $ Q_1 =  \forall $,
+% 	\item $ S(\mathbf{B},\tau) = p_1S(\mathbf{B}_{-1},\tau - W_1) + (1-p_1) S(\mathbf{B}_{-1},\tau) $, if $ Q_1 = \R^{p_1}  $.
 %\end{enumerate} 
@@
 For  existential and universal quantifiers, we compute $ S $ based on the weight.
-Specifically, if $ q_1 = \exists $, we decrement the threshold $ \tau $ by the maximum between $ w_1 $ and $ 0 $.
-For example, if $ w_1 > 0 $, $ B_1 $ is assigned $ 1 $, and assigned $ 0 $ otherwise.
+Specifically, if $ Q_1 = \exists $, we decrement the threshold $ \tau $ by the maximum between $ W_1 $ and $ 0 $.
+For example, if $ W_1 > 0 $, $ B_1 $ is assigned $ 1 $, and assigned $ 0 $ otherwise.
 Therefore, by solving for an existential variable, we  maximize $ S $.
-In contrast, when if $ q_1 = \forall $, we fix an assignment of $ B_1 $ that minimizes $ S $ by choosing between the minimum of $ w_1 $ and $ 0 $.
-Finally, for random quantifiers, we decompose the computation of $S$ into two sub-problems: one sub-problem where $ B_1 = 1 $ and the updated threshold becomes $ \tau - w_1 $ and another sub-problem where $ B_1 = 0 $ and the updated threshold remains the same.
+In contrast, when if $ Q_1 = \forall $, we fix an assignment of $ B_1 $ that minimizes $ S $ by choosing between the minimum of $ W_1 $ and $ 0 $.
+Finally, for random quantifiers, we decompose the computation of $S$ into two sub-problems: one sub-problem where $ B_1 = 1 $ and the updated threshold becomes $ \tau - W_1 $ and another sub-problem where $ B_1 = 0 $ and the updated threshold remains the same.
 Herein, we compute $ S $ as the expected output of both sub-problems.
@@
 
-\textbf{Computing Minimum and Maximum probability of positive prediction of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
-The coefficients $ w_{A_j} $ and $ w_{X_i} $ of the classifier become weights of $ \mathbf{B} $.
+\textbf{Computing Minimum and Maximum probability of positive prediction of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
+The coefficients $ W_{A_j} $ and $ W_{X_i} $ of the classifier become weights of $ \mathbf{B} $.
 Also, we get $n=m_1 +m_2$.
@@
 According to semantic of {\stochastic}, setting $ \exists $ quantifiers on $ \sensitive $ computes the maximum value of $ S(\mathbf{B}, \tau) $ that equalizes the maximum probability of positive prediction of the classifier.
-In this case, the \textit{inferred} assignment of $ \sensitive $ implies the most favored group $ \mathbf{a}_{{\max}} =  \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
-In contrast, to compute the minimum probability of positive prediction, we instead assign each variable  $ A_j $ a universal  quantifier  while keeping random quantifiers over $ X_i $, and infer the least favored group $ \mathbf{a}_{{\min}} =  \argmin_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
+In this case, the \textit{inferred} assignment of $ \sensitive $ implies the most favored group $ \mathbf{a}_{{\max}} =  \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $.
+In contrast, to compute the minimum probability of positive prediction, we instead assign each variable  $ A_j $ a universal  quantifier  while keeping random quantifiers over $ X_i $, and infer the least favored group $ \mathbf{a}_{{\min}} =  \argmin_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $.
 
@@
  \begin{cases}
- \mathsf{dp}(i+1, \tau-\max\{w_i, 0\}), \; \text{if } q_i = \exists\\
- \mathsf{dp}(i+1, \tau-\min\{w_i, 0\}), \; \text{if } q_i = \forall\\
- p_i \times \mathsf{dp}(i+1, \tau-w_i) + \\ (1- p_i) \times \mathsf{dp}(i+1, \tau) ,\; \text{if } q_i = \R^{p_i}
+ \mathsf{dp}(i+1, \tau-\max\{W_i, 0\}), \; \text{if } Q_i = \exists\\
+ \mathsf{dp}(i+1, \tau-\min\{W_i, 0\}), \; \text{if } Q_i = \forall\\
+ p_i \times \mathsf{dp}(i+1, \tau-W_i) + \\ (1- p_i) \times \mathsf{dp}(i+1, \tau) ,\; \text{if } Q_i = \R^{p_i}
  \end{cases}
@@
 \textit{Terminating Condition.} %We now define the \textit{terminating conditions} of the recursion. 
-Let $ w_{neg} $, $ w_{pos} $, and $ w_{all} $ be the sum of negative, positive, and all weights of $ \mathbf{B} $, respectively.
-We observe that $ w_{neg} \le w_{all} \le w_{pos}$.
-Thus, for any $ i $, if the {residual} threshold $ \tau \le w_{neg}$, there is always a subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.
-Conversely, when $ \tau > w_{pos}$, there is no subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.	We leverage this bound and tighten the terminating conditions of $ \mathsf{dp}(i, \tau) $ in Eq.~\eqref{fvgm_eq:dp_terminus}.
+Let $ W_{neg} $, $ W_{pos} $, and $ W_{all} $ be the sum of negative, positive, and all weights of $ \mathbf{B} $, respectively.
+We observe that $ W_{neg} \le W_{all} \le W_{pos}$.
+Thus, for any $ i $, if the {residual} threshold $ \tau \le W_{neg}$, there is always a subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.
+Conversely, when $ \tau > W_{pos}$, there is no subset of $ \mathbf{B}[i:n] $ with sum of weights at least $ \tau $.	We leverage this bound and tighten the terminating conditions of $ \mathsf{dp}(i, \tau) $ in Eq.~\eqref{fvgm_eq:dp_terminus}.
 
@@
  \begin{cases}
- 1\text{ if }\tau \le w_{neg}\\
- 0\text{ if } \tau > w_{pos}\\
+ 1\text{ if }\tau \le W_{neg}\\
+ 0\text{ if } \tau > W_{pos}\\
  \mathds{1}[\tau \le 0] \text{ if } i=n + 1
@@
  	Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $.
-Let $ w_{\exists} = \sum_{B_i \in \mathbf{B} | q_i = \exists} \max\{w_i, 0\}$  and $ w_{\forall} = \sum_{B_i \in \mathbf{B} | q_i = \forall} \min\{w_i, 0\}$ be the considered sum of weights of existential and universal variables, respectively.
-We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $.
-The space complexity is  $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
+Let $$ W_{\exists} = \sum_{B_i \in \mathbf{B} | Q_i = \exists} \max\{W_i, 0\}\text{ and } W_{\forall} = \sum_{B_i \in \mathbf{B} | Q_i = \forall} \min\{W_i, 0\} $$ be the considered sum of weights of existential and universal variables, respectively.
+We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |W_{neg}| - W_{\exists} - W_{\forall}) + n') $.
+The space complexity is  $ \mathcal{O}((n - n')(\tau + |W_{neg}| - W_{\exists} - W_{\forall})) $.
 \end{lemma}
+
+\begin{proof}
+	\textbf{Case 1: All $n$ variables in $\mathbf{B}$ have randomized quantifiers.}
+
+	At step $i$ of the dynamic programming (Eq.~\eqref{fvgm_eq:dp_recurse}), we modify the residual threshold of that step, namely $\tau_i$, either by subtracting $W_i$ or by retaining it.
+	Now, we observe that the residual threshold $\tau_i$ for any $i \in \lbrace1,\ldots,n\rbrace$ will be in $[0, \tau+|W_{neg}|]$.
+This holds because if $\tau_i$ crosses these bounds, the dynamic programming is terminated as shown in Equation~\eqref{fvgm_eq:dp_terminus}.
+	Since all weights of $\{W_i\}_{i=1}^{n}$ are integers, the maximum number of values that the residual threshold can take, is $ \tau + |W_{neg}| $.
+	Thus, we need to store at most $ n(\tau + |W_{neg}|)$ values in the memory for performing dynamic programming with $n$ variables and $ (\tau + |W_{neg}|) $ number of possible weights.
+	Thus, the space complexity  is $ \mathcal{O}(n(\tau + |W_{neg}|)) $.
+
+	In order to construct the dynamic programming table, we have to call the $\mathsf{dp}$ function $ \mathcal{O}(n(\tau + |W_{neg}|)) $ times, in the worst-case.
+	Thus, the time complexity of our method is $ \mathcal{O}(n(\tau + |W_{neg}|) $.
+
+	\textbf{Case 2: $n'$ variables have existential and universal quantifiers and $n-n'$ variables have randomized quantifiers in $\mathbf{B}$.}
+
+	According to Eq.~\eqref{fvgm_eq:dp_recurse}, $ W_\exists $ and $ W_\forall $ are the fixed weights of all existential and universal variables, respectively.
+Therefore, we need to consider at most $ \tau + |W_{neg}| -  W_\exists - W_\forall $ values of residual weights for random variables.
+By applying analysis in Case 1, the space and time complexity is derived as $ \mathcal{O}((n - n')(\tau + |W_{neg}| - W_{\exists} - W_{\forall})) $.
+
+	We note that there is an additional time complexity of $ \mathcal{O}(n') $ for existential and universal variables in Eq.~\eqref{fvgm_eq:dp_recurse}.
+Thus the time complexity becomes $ \mathcal{O}((n - n')(\tau + |W_{neg}| - W_{\exists} - W_{\forall}) + n') $.
+We, however, do not require to store any entry for existential and universal variables in $ \mathsf{dp} $  function and thus, the space complexity remains the same as $ \mathcal{O}((n - n')(\tau + |W_{neg}| - W_{\exists} - W_{\forall})) $.
+\end{proof}
 
@@
 Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables.
-In particular, if $ \tau \le 0.5(w_{pos} - w_{neg}) $, $ \tau $ is closer to $ w_{pos} $ than $ w_{neg} $.
+In particular, if $ \tau \le 0.5(W_{pos} - W_{neg}) $, $ \tau $ is closer to $ W_{pos} $ than $ W_{neg} $.
 Hence, we sort each cluster in descending order of weights.
@@
 \begin{example}\label{fvgm_example:subset-sum}
+\normalfont
 We consider a linear classifier $ P + Q + R - S \ge 2$.
@@
 To compute the maximum probability of positive prediction of the classifier,  we impose an existential quantifier on $P$ and randomized quantifiers on others.
-This leads us to the observation that $ P = 1 $ is the optimal assignment as $ w_P = 1 > 0 $.
+This leads us to the observation that $ P = 1 $ is the optimal assignment as $ W_P = 1 > 0 $.
 We now require to compute $ \Pr[Q + R - S \ge 1] $, which by dynamic programming, is computed as $ 0.55 $.

### 2022-06-07 chapters/fairness/fvgm/framework2.tex
revision on fvgm

@@
 \begin{align*}
-	\mathsf{dp}(i,  &\tau, \mathbf{u}) = \Pr[B_i = 0| \parent(B_i) = \mathbf{u}] \mathsf{dp}(i+1, \tau, \mathbf{u} \cup \{0\})  \\
-	& + \Pr[B_i = 1| \parent(B_i) = \mathbf{u}]  \mathsf{dp}(i+1, \tau-w_i, \mathbf{u} \cup \{1\}).
+	\mathsf{dp}(i,  \tau, \mathbf{u}) = & \Pr[B_i = 0| \parent(B_i) = \mathbf{u}] \mathsf{dp}(i+1, \tau, \mathbf{u} \cup \{0\})  \\
+	 + & \Pr[B_i = 1| \parent(B_i) = \mathbf{u}]  \mathsf{dp}(i+1, \tau-W_i, \mathbf{u} \cup \{1\}).
 \end{align*}
 
-Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to know the assignment of parents before computing $ \mathsf{dp} $ on the child.
+Since $ \mathsf{dp}(i, \tau, \mathbf{u}) $ involves  $ \mathbf{u} $, we initially perform a topological sort of $ \mathbf{V} $ to enumerate the assignment of parents before computing $ \mathsf{dp} $ on the child.
 Moreover, there are $ 2^{|\parent(B_i)|} $ assignments of $ \parent(B_i) $, and we compute $ \mathsf{dp}(i, \tau, \mathbf{u}) $ for $ \mathbf{u} \in \{0,1\}^{|\parent(B_i)|} $ to incorporate all conditional probabilities into $ \stochastic $.
@@
 	Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $.
-Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $.
-To exactly compute {\stochastic} with correlated variables in the dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
+Let $$ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | Q_i = \exists} \max\{W_i, 0\} \text{ and } w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | Q_i = \forall} \min\{W_i, 0\}$$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $.
+To exactly compute {\stochastic} with correlated variables in the dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall})) $.
 \end{lemma}
 
-\textbf{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $.
+\begin{proof}
+	We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $.
+For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{fvgm_sec:dp_with_BN}.
+
+	For variables in $\mathbf{B}\setminus \mathbf{V} $, we apply analysis from  Lemma~\ref{fvgm_lemma:complexity_sss}, where we consider $ (n - n'' - |\mathbf{V}|) $ random variables, $ n'' $ existential/universal variables, and residual weights can take at most $ (\tau + |W_{neg}| - w'_{\exists} - w'_{\forall}) $ values.
+Hence, time complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $, and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall})) $
+
+	Combining two cases, overall time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |W_{neg}| - w'_{\exists} - w'_{\forall})) $.
+\end{proof}
+
+\textbf{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $ that are in the Bayesian network.
 For computing the probability of positive prediction of a linear classifier with correlated features, we consider a heuristic to sort variables in $ \mathbf{B} = \sensitive \cup \nonsensitive $.
@@
 In this heuristic, we sort sensitive variables $ \sensitive $ by positioning $ \sensitive \cap \mathbf{V} $ in the beginning followed by $ \sensitive \cap \mathbf{V}^c $.
-Then we order the variables $ \mathbf{B} $ such that variables in $ \sensitive $ precedes those in $ \nonsensitive \cap \mathbf{V} $, and the variables in $ \nonsensitive \cap \mathbf{V}^c $ follows the ones in $ \nonsensitive \cap \mathbf{V} $.
+Then we order the variables $ \mathbf{B} $ such that variables in $ \nonsensitive $ precedes those in $ \nonsensitive \cap \mathbf{V} $, and the variables in $ \nonsensitive \cap \mathbf{V}^c $ follows the ones in $ \nonsensitive \cap \mathbf{V} $.
 This sorting allows us to avoid repetitive enumeration of variables in $ \mathbf{V} \subseteq \mathbf{B} $ as they are placed earlier in $ \mathbf{B} $.
@@
 	\subsection{Fairness Verification with Computed Probability of Positive Prediction} 
-	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair ($\epsilon \in [0,1]$) is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$.
-We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum probability of positive prediction $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum probability of positive prediction $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.
+	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair for $\epsilon \in [0,1]$ is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$.
+We now compute $f(\mathcal{M}|\mathcal{D})$ based on the maximum probability of positive prediction $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ and the  minimum probability of positive prediction $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a classifier.
 
-	For measuring fairness metric SP, we compute the difference $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
-We, however, deploy {\fvgm} twice while measuring EO: one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $.
-	In each case, we compute $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of EO.
-	For measuring causal metric PCF, we compute  $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\hat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ conditioned on mediator features $ \mathbf{Z} $ and take their difference.
-	To measure DI, we compute the ratio $ \max_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] / \min_{ \mathbf{a}} \Pr[\hat{Y} =1 | \sensitive = \mathbf{a}] $.
-In contrast to other fairness metrics, DI closer to 1 indicates higher fairness level.
-Thus, we verify whether a classifier achieves $(1 - \epsilon)$-DI by checking $ \mathds{1}[DI(\alg| \mathcal{D}) \ge 1 - \epsilon] $.
+	For measuring fairness metric $ \mathsf{SP} $, we compute the difference $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}]  - \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $.
+We, however, deploy {\fvgm} twice while measuring $ \mathsf{EO} $, one for the distribution $ \mathcal{D} $ conditioned on $ Y = 1  $ and another for $ Y = 0 $.
+In each case, we compute $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, Y = y ]  - \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, Y = y ] $ for $ y \in \{0,1\} $ and take the  maximum difference as the value of $ \mathsf{EO} $.
+For measuring causal metric $ \mathsf{PCF} $, we compute  $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}, \mathbf{Z}] $ and  $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1, \mathbf{Z}| \sensitive = \mathbf{a} , \mathbf{Z}] $ conditioned on mediator features $ \mathbf{Z} $ and take their difference.
+	To measure disparate impact $ \mathsf{DI} $, we compute the ratio $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] / \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $.
+In contrast to other fairness metrics, $ \mathsf{DI} $ closer to 1 indicates higher fairness level.
+Thus, we verify whether a classifier achieves $(1 - \epsilon)$-$ \mathsf{DI} $ by checking $ \mathds{1}[f_{\mathsf{DI}}(\alg| \mathcal{D}) \ge 1 - \epsilon] $.
+
+	\todo{End of revision}
 
@@
 	For verifying linear classifiers with real-valued features and coefficients, we preprocess them so that {\fvgm} can be invoked.
-Let $ X_c \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ w_c \in \mathbb{R} $ in the classifier.
+Let $ X_c \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ W_c \in \mathbb{R} $ in the classifier.
 We discretize $ X_c $ to a set $ \mathbf{B}_{c} $ of $ k $ Boolean variables using binning-based discretization and assign a Boolean variable to each bin.
@@
 Let $ \mu_i $ denote the mean of feature-values within $ i^\text{th} $ bin.
-We then set the coefficient of $ B_i $ as $ w_c\mu_i $.
+We then set the coefficient of $ B_i $ as $ W_c\mu_i $.
 By law of large numbers, $ X_c \approx \sum_i \mu_iB_i $ for infinitely many bins~\cite{grimmett2020probability}.

### 2022-06-07 chapters/fairness/fvgm/introduction.tex
revision on fvgm

--- 2022-06-02 chapters/fairness/fvgm/introduction.tex
@@
 %\section{Introduction}
-	The significant improvement of machine learning (ML) over the decades has led to a host of applications of ML in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}.
-ML algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
+
+\label{chapter:fvgm}
+	The significant improvement of machine learning over the decades has led to a host of applications of machine learning in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}.
+Machine learning algorithms often have an accuracy-centric learning objective, which may cause them to be biased towards certain part of the dataset belonging to a certain economically or socially sensitive groups~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
 %	Fairness in Machine Learning (ML) has gained a significant focus  in recent years as ML algorithms are deployed in high-stake decision-making, such as college admission, hiring of employees, and recidivism prediction.
 Thus, if the ML algorithm is biased towards certain part of the dataset belonging to certain economically or socially sensitive groups, it can cause adverse effects in real-life.
-	 The following example illustrates a plausible case of unfairness induced by ML algorithms. 
+	 The following example illustrates a plausible case of unfairness of machine learning. 
 	\begin{example}\label{fvgm_example:intro}
 		\normalfont
-		Following Example~\ref{fairness_justicia_example:intro}, let us consider an ML problem where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}).
+		Following Example~\ref{fairness_justicia_example:intro}, let us consider a machine learning classification where the classifier decides the eligibility of an individual for health insurance given their income and fitness (Figure~\ref{fvgm_fig:example1}).
 Here, the sensitive feature `age' ($ A $) follows a Bernoulli distribution, and income ($ I $) and fitness ($ F $) follow Gaussian distributions.
@@
 
-	In order to identify and mitigate the bias of ML classifiers, different fairness definitions and fairness algorithms have been proposed~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.	In this chapter, we focus on two families of fairness definitions: \textit{group} and \textit{causal} fairness.
+	In order to identify and mitigate the bias of classifiers, different fairness definitions and fairness algorithms have been proposed~\cite{hardt2016equality,kusner2017counterfactual,mehrabi2019survey}.	In this chapter, we focus on two families of fairness definitions: \textit{group} and \textit{causal} fairness.
 Group fairness metrics, such as disparate impact and equalized odds constrain the probability of the positive prediction of the classifier to be (almost) equal among different sensitive groups~\cite{dwork2012fairness,feldman2015certifying}.	On the other hand, causal fairness metrics assess the difference in positive predictions  if every feature in the causal relation  remains identical except the sensitive feature~\cite{nabi2018fair,zhang2018fairness}.
@@
 
-	\textbf{Our Contributions.} In this chapter, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
-{\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
-To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
-	%Finally, {\fvgm} encodes the Bayesian Network in a SSAT-based formulation to verify classifiers in CNF and in the stochastic subset-sum based formulation to verify linear classifiers.  
-	Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
-We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.
+	\textbf{Our Contributions.} The contributions of this chapter are summarized below.
+
+	\begin{itemize}
+		\item \textit{Framework:} In this chapter, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
+		\item \textit{Scalability:} {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
+		\item \textit{Accuracy:} To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
+		\item \textit{Experimental Results:}	Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
+		\item \textit{Applications:} We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.
+	\end{itemize}

### 2022-06-07 chapters/fairness/fvgm/applications.tex
updated upto fvgm

@@
 In this section, we apply {\fvgm} for verifying fairness-enhancing algorithms and depreciation attacks.
-We also demonstrate that {\fvgm} facilitates computation of fairness influence functions by enabling the detection of bias due to individual features.
+We also demonstrate that {\fvgm} facilitates the computation of fairness influence functions by enabling the shift of bias from the original value due to individual features.
 
@@
 In Table~\ref{fvgm_tab:fair_algo_verification}, we report the effect of fairness algorithms w.r.t.
-four fairness metrics: disparate impact (DI), path-specific causal fairness (PCF), statistical parity (SP), and equalized odds (EO).
-Note that, fairness is improved if DI increases and the rest of the metrics decrease.
+four fairness metrics: disparate impact ($ \mathsf{DI} $), path-specific causal fairness ($ \mathsf{PCF} $), statistical parity ($ \mathsf{SP} $), and equalized odds ($ \mathsf{EO} $).
+Note that, fairness is improved if $ \mathsf{DI} $ increases and the rest of the metrics decrease.
 For instance, in most instances for Adult dataset, reweighing (RW)~\cite{kamiran2012data} and optimized pre-processing (OP)~\cite{calmon2017optimized} algorithms are successful in improving fairness.
-The exceptional case is the unfairness regarding the sensitive feature `sex', where OP algorithm fails in improving fairness metric EO.
+The exceptional case is the unfairness regarding the sensitive feature `sex', where OP algorithm fails in improving fairness metric $ \mathsf{EO} $.
 Thus, \textit{{\fvgm} verifies the enhancement and decrement in fairness by  fairness-enhancing algorithms.
@@
 	\caption[Verifying fairness attacks]{Verifying poisoning attack against fairness using {\fvgm}.
-The red line denotes the safety margin of the ML model against the attack.}\label{fvgm_fig:fairness_attacks}
+The red line denotes the safety margin of the machine learning model against the attack.}\label{fvgm_fig:fairness_attacks}
 \end{figure}
@@
 Thus, applying {\fvgm} to compute FIF provides us insights about sources of bias and thus, indicators to improve fairness.
-$\Pr[\hat{Y} = 1|$age\_0 $\ge$ 0.5]
 
@@
 	\caption[FIF computation using {\fvgm}]{FIF computation for COMPAS dataset.
-For Female (left plot) and Male, influence of `age' decreases and the probability of positive prediction of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}
+For Female (left plot) and Male (right plot) individuals,  the influence of `age' decreases and the probability of positive prediction of the classifier increases  by different magnitudes.}\label{fvgm_fig:influence function}
 %	\vspace{-2ex}

### 2022-06-07 chapters/fairness/fvgm/conclusion.tex
updated upto fvgm

@@
 %We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF).
-Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
+Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other machine learning models.
 %Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.

### 2022-06-07 chapters/fairness/fvgm/evaluations.tex
updated upto fvgm

@@
 \label{fvgm_sec:experiments}
-In this section, we empirically evaluate the performance of {\fvgm}. We first present the experimental setup and the objective of our experiments, followed by experimental results.
+In this section, we empirically evaluate the performance of {\fvgm}. We first present the experimental setup  followed by experimental results.
 %\footnote{Due to page limit, we present additional experimental results, proof of lemmas, and miscellaneous details in Appendix.}
@@
 
-\textbf{Experimental Setup.}
+\paragraph{Experimental Setup.}
 We  implement a  prototype of {\fvgm} in Python (version 3.8).  
@@
 
-We compare {\fvgm} with three existing fairness verifiers: Justicia~\cite{ghosh2020justicia}, FairSquare~\cite{albarghouthi2017fairsquare}, and VeriFair~\cite{bastani2019probabilistic}. 
+We compare {\fvgm} with three existing fairness verifiers: Justicia~\cite{ghosh2020justicia}, FairSquare~\cite{albarghouthi2017fairsquare}, and VeriFair~\cite{bastani2019probabilistic}.In the following, we discuss a comparative analysis among all verifiers based on scalability and accuracy, where {\fvgm} yields superior performance than others.
+
 %Additionally, we have verified two fairness algorithms: reweighing algorithm (RW)~\cite{kamiran2012data} and the optimized pre-processing algorithm (OP)~\cite{calmon2017optimized}, and a fairness poisoning-attack algorithm~\cite{solans2020poisoning}.
@@
 \subsection{Scalability Analysis}
-\textbf{Benchmarks.} We perform the scalability analysis on five real-world datasets studied in fair ML literature: UCI Adult, German-credit~\cite{DK2017}, COMPAS~\cite{angwin2016machine}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}).
-We consider $ 100 $ benchmarks generated from 5 real-world datasets and report the computation times (for DI and SP) of different verifiers.
+\paragraph{Benchmarks.} We perform the scalability analysis on five real-world datasets studied in the literature of fairness in machine learning: UCI Adult, German-credit~\cite{DK2017}, COMPAS~\cite{angwin2016machine}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}).
+We consider $ 100 $ benchmarks generated from 5 real-world datasets and report the computation times for disparate impact and statistical parity metrics of different verifiers.
 
-\textbf{Results.} In Figure~\ref{fvgm_fig:scalability_exp}, we present the scalability results of different verifiers.
+\paragraph{Results.} In Figure~\ref{fvgm_fig:scalability_exp}, we present the scalability results of different verifiers.
 First, we observe that FairSquare often times out ($ =900 $ seconds) and can solve $ \le 5 $ benchmarks.
@@
 Similarly, SSAT-based verifier Justicia that performs pseudo-Boolean to CNF translation for linear classifiers, times out for around $  20 $ out of $ 100 $ benchmarks.
-Sampling-based framework, VeriFair, has comparatively better scalability than SMT/SSAT based frameworks and can solve more than $ 90 $ benchmarks.
+Sampling-based framework, VeriFair, has comparatively better scalability than SMT/SSAT-based frameworks and can solve more than $ 90 $ benchmarks.
 Finally, {\fvgm} achieves impressive scalability by solving all $ 100 $ benchmarks with $ 1 $ to $ 2 $ orders of magnitude runtime improvements than existing verifiers.
@@
 	\caption[Accuracy of {\fvgm}]{Comparing the average accuracy of different verifiers over 100 synthetic benchmarks while varying the number of features.
-{\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of DI for LR and SVM classifiers.}\label{fvgm_fig:sanity_exp}
+{\fvgm} yields the closest estimation of the analytically calculated \textit{Exact} values of disparate impact for LR and SVM classifiers.}\label{fvgm_fig:sanity_exp}
 	%	\vspace{-2ex}
@@
 \subsection{Accuracy Analysis}
-\noindent\textbf{Benchmark Generation.} To perform accuracy analysis, we require the ground truth, which is not available for real-world instances.
-Therefore, we focus on generating synthetic benchmarks for analytically computing the ground truth of different fairness metrics, such as DI, from the known distribution of features.
+\paragraph{Benchmark Generation.} To perform accuracy analysis, we require the ground truth, which is not available for real-world instances.
+Therefore, we focus on generating synthetic benchmarks for analytically computing the ground truth of different fairness metrics, such as disparate impact, from the known distribution of features.
 In each benchmark, we consider $ n \in \{2, 3, 4, 5\} $ features including one Boolean sensitive feature, say $ A $, generated from a Bernoulli distribution with mean $ 0.5 $.
@@
 Finally, we create label $ Y = \mathds{1}[ \sum_{i=1}^{n-1} X_i \ge 0.5 \sum_{i=1}^{n-1} (\mu_i + \mu_i')] $ such that $ Y $ does not directly depend on the sensitive feature.
-For each $ n $, we generate $ 100 $ random benchmarks, learn LR and SVM classifiers on them, and compute DI using different verifiers.
+For each $ n $, we generate $ 100 $ random benchmarks, learn LR and SVM classifiers on them, and compute disparate impact using different verifiers.
 
-To analytically compute DI, let the coefficients of the classifier be $ w_i $ for $ X_i $ and $ w_A $ for $ A $, and bias be $ \tau $.
-Since all non-sensitive features are from Gaussian distributions, we compute the probability of the predicted class $ \Pr[\hat{Y} | A = 1]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i, \sigma_{\hat{Y}}^2) $ and $ \Pr[\hat{Y} | A = 0]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i', \sigma_{\hat{Y}}^2) $ with $ \sigma_{\hat{Y}}^2 =   (\sum_{i=1}^{n-1}w_i^2)\sigma^2 $.
-Hence, the probability of positive prediction of the classifier  is $  1 - \mathsf{CDF}_{\hat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\hat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function.
-Finally, we compute DI by taking the ratio of the minimum and the maximum of the probability of positive prediction of the classifier.
+\paragraph{Analytic Computation of  Disparate Impact.} Let the coefficients of the classifier be $ w_i $ for $ X_i $ and $ w_A $ for $ A $, and bias be $ \tau $.
+Since all non-sensitive features are from Gaussian distributions, we compute the probability of the predicted class $ \Pr[\widehat{Y} | A = 1]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i, \sigma_{\widehat{Y}}^2) $ and $ \Pr[\widehat{Y} | A = 0]  \sim \mathcal{N}(\sum_{i=1}^{n-1}w_i\mu_i', \sigma_{\widehat{Y}}^2) $ with $ \sigma_{\widehat{Y}}^2 =   (\sum_{i=1}^{n-1}w_i^2)\sigma^2 $.
+Hence, the probability of positive prediction of the classifier  is $  1 - \mathsf{CDF}_{\widehat{Y}| A =1}(\tau - w_A) $ for $ A = 1 $ and $  1 - \mathsf{CDF}_{\widehat{Y}|A=0}(\tau) $ for $ A = 0 $, where $ \mathsf{CDF} $ is the cumulative distribution function.
+Finally, we compute disparate impact by taking the ratio of the minimum and the maximum of the probability of positive prediction of the classifier for $ A = 1 $ and $ A = 0 $.
 
-\textbf{Results.} 
+\paragraph{Results.} 
 %\begin{enumerate}[leftmargin=*]
@@
 %\red{Due to limited space, we discuss additional experiments including verification of CNF-based classifiers with feature correlations and performance analysis of encoding Bayesian networks in Appendix.}
- We  assess the accuracy of the competing verifiers in estimating fairness metrics, specifically DI with LR and SVM classifiers.
-In Figure~\ref{fvgm_fig:sanity_exp}, {\fvgm} computes DI closest to the \textit{Exact} value for different number of features and both type of classifiers.
-In contrast, Justicia, FairSquare, and VeriFair measure DI far from the \textit{Exact} because of ignoring correlations among the features.
-For example, for SVM classifier with  $ n = 5 $ (right plot in Figure~\ref{fvgm_fig:sanity_exp}), \textit{Exact} DI is $ 0.089 $ (average over 100 random benchmarks).
-Here, {\fvgm} computes DI as $ 0.094 $, while all other verifiers compute DI as at least $ 0.233 $.
+ We  assess the accuracy of the competing verifiers in estimating fairness metrics, specifically disparate impact with LR and SVM classifiers.
+In Figure~\ref{fvgm_fig:sanity_exp}, {\fvgm} computes disparate impact closest to the \textit{Exact} value for different number of features and both type of classifiers.
+In contrast, Justicia, FairSquare, and VeriFair measure disparate impact far from the \textit{Exact} because of ignoring correlations among the features.
+For example, for SVM classifier with  $ n = 5 $ (right plot in Figure~\ref{fvgm_fig:sanity_exp}), \textit{Exact} disparate impact is $ 0.089 $ (average over 100 random benchmarks).
+Here, {\fvgm} computes disparate impact as $ 0.094 $, while all other verifiers compute disparate impact as at least $ 0.233 $.
 Therefore, \textit{{\fvgm} is more accurate than existing verifiers as it explicitly considers correlations among features}.
@@
 			\toprule
-			Dataset & $ \sensitive $ & Algo. & $ \Delta $DI &  $ \Delta $PCF & $ \Delta $SP & $ \Delta $EO\\
+			Dataset & $ \sensitive $ & Algo. & $ \Delta \mathsf{DI}$ &  $ \Delta \mathsf{PCF}$ & $ \Delta $ $\mathsf{SP}$ $$ & $ \Delta \mathsf{EO}$\\
 			\midrule

### 2022-06-07 chapters/fairness/fvgm/framework.tex
updated upto fvgm

@@
 
-\textbf{Problem Formulation.}	
+\paragraph{Problem Formulation.}	
 Given a linear classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $ and a probability distribution $ \mathcal{D} $ of $ \nonsensitive \cup \sensitive $, our objective is to compute $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ and $ \min_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ with respect to $ \mathcal{D} $.
@@
 
-\textbf{Computing Minimum and Maximum probability of positive prediction of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
+\paragraph{Computing Minimum and Maximum probability of positive prediction of Linear Classifiers Using  {\stochastic}.} For computing $ \max_{ \mathbf{a}} \Pr[\widehat{Y} =1 | \sensitive = \mathbf{a}] $ of a linear classifier, we set existential quantifiers $ \exists $ to sensitive features $ A_j $, randomized quantifiers $ \R $ to non-sensitive features $ X_i $ and construct a set $ \mathbf{B} = \sensitive \cup \nonsensitive $.
 The coefficients $ W_{A_j} $ and $ W_{X_i} $ of the classifier become weights of $ \mathbf{B} $.
@@
 
-\textbf{A Heuristic for Faster Computation.} 
+\paragraph{A Heuristic for Faster Computation.} 
 We propose two improvements for a faster computation of the dynamic programming solution.

### 2022-06-07 chapters/fairness/fvgm/framework2.tex
updated upto fvgm

@@
 
-\textbf{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $ that are in the Bayesian network.
+\paragraph{A Heuristic for Faster Computation.} We observe that to encode conditional probabilities, we enumerate all assignments of variables in $ \mathbf{V} $ that are in the Bayesian network.
 For computing the probability of positive prediction of a linear classifier with correlated features, we consider a heuristic to sort variables in $ \mathbf{B} = \sensitive \cup \nonsensitive $.
@@
 
-	\todo{End of revision}
-
 	\subsection{Extension to Practical Settings}\label{fvgm_sec:practical}
 	For verifying linear classifiers with real-valued features and coefficients, we preprocess them so that {\fvgm} can be invoked.
-Let $ X_c \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ W_c \in \mathbb{R} $ in the classifier.
-We discretize $ X_c $ to a set $ \mathbf{B}_{c} $ of $ k $ Boolean variables using binning-based discretization and assign a Boolean variable to each bin.
-Hence, $ B_i \in \mathbf{B}_{c} $ becomes $ 1 $, when $ X_c $ belongs to the $ i^\text{th} $ bin.
+Let $ X \in \mathbb{R} $ be a continuous real-valued feature with coefficient $ W \in \mathbb{R} $ in the classifier.
+We discretize $ X $ to a set $ \mathbf{B} $ of $ k $ Boolean variables using binning-based discretization and assign a Boolean variable to each bin.
+Hence, $ B_i \in \mathbf{B} $ becomes $ 1 $, when $ X $ belongs to the $ i^\text{th} $ bin.
 Let $ \mu_i $ denote the mean of feature-values within $ i^\text{th} $ bin.
-We then set the coefficient of $ B_i $ as $ W_c\mu_i $.
-By law of large numbers, $ X_c \approx \sum_i \mu_iB_i $ for infinitely many bins~\cite{grimmett2020probability}.
+We then set the coefficient of $ B_i $ as $ \mu_iW $.
+By the law of large numbers, $ X \approx \sum_i \mu_iB_i $ for infinitely many bins~\cite{grimmett2020probability}.
 Finally, we multiply the coefficients of discretized variables by $ l \in \mathbb{N} \setminus \{0\} $ and round to an integer.
-Accuracy of the preprocessing step relies on the number of bins $ k $ and the multiplier $ l $.
+The accuracy of the preprocessing step relies on the number of bins $ k $ and the multiplier $ l $.
 Therefore, we empirically fine-tune both $ k $ and $ l $ by comparing  the processed classifier with the initial classifier on a validation dataset.

### 2022-06-08 chapters/interpretability/imli/introduction.tex
minor edits

@@
 In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}.
-In this study, we focus on learning an interpretable machine learning classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box models.
+In this chapter, we focus on learning an interpretable machine learning classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box models.
 
@@
 
-The primary contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
+\paragraph{Contributions.} The primary contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
 For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF (Conjunctive Normal Form) formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.

### 2022-06-08 chapters/interpretability_background.tex
minor edits

@@
 
-In this part of the thesis, we discuss preliminaries on interpretable machine learning in Chapter~\ref{chapter_interpretability_preliminaries}, an incremental learning framework for interpretable classification rules in Chapter~\ref{chapter:imli}, and extend incremental learning to more expressible classification rules in Chapter~\ref{chapter:crr}.
+In this part of the thesis, we discuss preliminaries in interpretable machine learning in Chapter~\ref{chapter_interpretability_preliminaries}, an incremental learning framework for interpretable classification rules in Chapter~\ref{chapter:imli}, and extend incremental learning to more expressible classification rules in Chapter~\ref{chapter:crr}.
 
@@
 Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $.
-Motivated by \red{Boolean cardinality constraints}, we propose relaxed-CNF formulas, where in addition to $F$, we have two more parameters $\eta_c$ and $\eta_l$.
+Motivated by Boolean cardinality constraints, we propose relaxed-CNF formulas, where in addition to $F$, we have two more parameters $\eta_c$ and $\eta_l$.
 We say that $(F,\eta_c, \eta_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (F,\eta_c, \eta_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (C_i,\eta_l)] \ge \eta_c$, where $\sigma^* \models (C_i, \eta_l)$ if and only if $\sum_{v \in C_i} \mathds{1}[\sigma^* \models v] \geq \eta_l$.
@@
 Hence, for the running example $ \mathbf{u} = [u_1, u_2, u_3] $ and $ \mathbf{v} = [0,1,1] $, we derive $ \mathbf{u} \circ \mathbf{v} = u_2  + u_3 $.
-In this part of the thesis, we use Boolean interpretation of the inner product for learning CNF classification rules and numerical interpretation for learning relaxed-CNF classification rules.
+In this part of the thesis, we use Boolean interpretation of the inner product for learning CNF classification rules in Chapter~\ref{chapter:imli} and numerical interpretation for learning relaxed-CNF classification rules in Chapter~\ref{chapter:crr}.
 
@@
 To avoid notational clutter, we overload {\weight{\cdot}} to denote the weight of an assignment $ \sigma $.
-In particular, we define $\weight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $\weight{\sigma} = \Sigma_{i | \sigma \not \models C_i} \weight{C_i}$.
+In particular, we define $\weight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $$\weight{\sigma} = \sum_{i | \sigma \not \models C_i} \weight{C_i}.$$
 
@@
 The optimal weight of the MaxSAT problem $  \weight{\sigma^*} $ is infinity ($ \infty $) when at least one hard clause becomes unsatisfied.
-Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum.
-In this chapter, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem.
+Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses\footnote{In our formulation, we assume that there is a satisfying assignment to a CNF formula containing all hard clauses.} and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum.
+In Chapter~\ref{chapter:imli}, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem.
 The knowledge of the inner workings of {MaxSAT} solvers is  not required for this chapter.
@@
 We consider a standard binary classification problem where $ \mathbf{X} \in \{0,1\}^{n\times m} $ is a binary feature matrix\footnote{Our framework allows learning on real-valued and categorical features, as discussed in Section~\ref{interpretability_imli_sec:non-binary}.} defined over Boolean features $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$ and $ \mathbf{y} \in \{0,1\}^n $ is a vector of binary class labels.
-The $ i^\text{th} $ row of $ \mathbf{X} $, denoted by $ \mathbf{X}_i \in \mathcal{X} $, is called a sample, which is generated from a product distribution $ \mathcal{X} $.
+The $ i^\text{th} $ row of $ \mathbf{X} $, denoted by $ \mathbf{X}_i \in \mathcal{X} $, is called a sample generated from a product distribution of all features $ \mathcal{X} $.
 Thus, $ \mathbf{X}_i $ is the valuation of features in $ \mathbf{x} $ and $ y_i $ is its class label.
-$ \mathbf{X}_i $ is called a positive sample if $ y_i = 1 $, otherwise a negative sample.
+$ \mathbf{X}_i $ is called a positive sample if $ y_i = 1 $, and a negative sample otherwise.
 
-A classifier $ \Rule: \mathbf{x} \rightarrow \hat{y} \in  \{0,1\} $ is a  function that takes a Boolean vector as input and outputs the predicted class label $\hat{y}$.
+A classifier $ \Rule: \mathbf{x} \rightarrow \widehat{y} \in  \{0,1\} $ is a  function that takes a Boolean vector as input and outputs the predicted class label $\widehat{y}$.
 %We define two rules  $ \R_1 $ and $ \R_2 $ to be  equivalent  if $ \forall i,\R_1(\X_i)=\R_2(\X_i) $.
-In a rule-based classifier, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{x} $ such that, if $ \Rule $ becomes true for an input (or an assignment), the prediction $ \hat{y} = 1 $ and $ \hat{y} = 0 $, otherwise.
+In a rule-based classifier, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{x} $ such that, if $ \Rule $ becomes true for an input (or an assignment), the prediction $ \widehat{y} = 1 $ and $ \widehat{y} = 0 $, otherwise.
 The goal is to design $\Rule$ not only to approximate the training data, but also to generalize to unseen samples arising from the same distribution.
@@
 
-For example, let $ \Rule \triangleq (x_1 \vee x_3) \wedge (\neg x_2 \vee x_3) $ be a CNF classifier defined over three Boolean features $ \{x_1, x_2, x_3\} $.
-For an input $ [0, 1, 1] $, the prediction of $ \Rule  $ is $ 1 $ whereas for an input $ [1, 1, 0] $, the prediction is $ 0 $ because $ \Rule $ is true and false in these two cases, respectively.
-Moreover, $ |\Rule| = 2 + 2 = 4 $ is the rule-size of $ \Rule $.
+\begin{example}
+	\normalfont
+	Let $ \Rule \triangleq (x_1 \vee x_3) \wedge (\neg x_2 \vee x_3) $ be a CNF classifier defined over three Boolean features $ \{x_1, x_2, x_3\} $.
+For an input $ [0, 1, 1] $, the prediction of $ \Rule  $ is $ 1 $, whereas for an input $ [1, 1, 0] $, the prediction is $ 0 $ because $ \Rule $ is true and false in these two cases, respectively.
+Moreover, $ |\Rule| = 2 + 2 = 4 $ is the size of $ \Rule $.
+\end{example}
 

### 2022-06-08 chapters/introduction.tex
upto theoretical analysis of justicia
minor edits

@@
 \section{Tools}
+	We develop following open-source tools for interpretable and fair machine learning.
+	\begin{itemize}
+		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
+		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification
+	\end{itemize}
 \section{Outline}
+We organize the thesis in the following way.
+In the first part of the thesis, we discuss interpretable machine learning, starting with preliminaries in Chapter~\ref{chapter_interpretability_preliminaries}.
+In Chapter~\ref{chapter:imli}, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable CNF classifiers.
+We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
+In the second part of this thesis, we discuss fairness in machine learning, starting with preliminaries in Chapter~\ref{chapter_fairness_preliminaries}.
+Next, we discuss fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
+We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}.
+Finally, we conclude our thesis in Chapter~\ref{}.
 

### 2022-06-14 chapters/abstract.tex
all papers added
upto preliminaries in fairness
preliminaries notations updated

@@
 \begin{abstract}
+	\addcontentsline{toc}{chapter}{Abstract}
 	\begin{center}
-		Interpretability and Fairness in Machine Learning: A Formal Method Approach\\
+		Interpretability and Fairness in Machine Learning: A Formal Methods Approach\\
 		\vspace{1em}
@@
 The interpretability of rule-based classifiers is generally related to the size of the rules, where smaller rules with higher accuracy are preferable in practice.
-As such, interpretable learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
+As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
 To this end, we propose an incremental learning framework, called {\imli}, which extends the classification to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.

### 2022-06-14 chapters/background.tex
preliminaries notations updated

--- 2022-06-03 chapters/fairness_background.tex
@@
-\part{Fairness in Machine Learning}
+\chapter{Preliminaries}
 
-In this part of the thesis, we discuss preliminaries on fairness in machine learning in Chapter~\ref{chapter_fairness_preliminaries}.
-Next, we discuss fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
-Finally, we discuss identification of sources of unfairness in Chapter~\ref{chapter:fif}.
+We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
+We express random variables in uppercase, and the valuation or an assignment of a random variable in lowercase.
 
-\chapter{Preliminaries}
+%In this chapter, we discuss preliminaries for learning interpretable rule-based classifiers.
+Throughout this part of the thesis, we use capital  boldface letters such as $\mathbf{X}$ to denote matrices while lower boldface letters such as $\mathbf{x}$ are reserved for vectors/sets.
+
+\section{Formal Methods}
+
+\subsection{Propositional Satisfiability}
+
+Let $\propformula$ be a Boolean formula defined over a set of Boolean variables $\boolset = \{\bool_1,\bool_2,\dots ,\bool_m \}$.
+A literal $ \literal $ is a variable $ \bool $ or its complement $\neg \bool$, and a clause $ \propclause $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals in the literature, we use clause for both disjunction and conjunction.}  $\propformula$ is in Conjunctive Normal Form (CNF) if $\propformula \triangleq  \bigwedge_i \propclause_i$ is a conjunction of clauses where each clause $\propclause_i \triangleq  \bigvee_j \literal_j $  is a disjunction of literals.
+In contrast, $ \propformula $ is in Disjunctive Normal Form (DNF) if $\propformula \triangleq  \bigvee_i \propclause_i$ is a disjunction of clauses where each clause $\propclause_i \triangleq  \bigwedge_j \literal_j $ is a conjunction of literals.
+We use $\sigma$ to denote an assignment of variables in  $\boolset$ where $ \sigma(\bool_i) \in $ \{true, false\}.
+A \emph{satisfying assignment} $ \sigma^* $ of $\propformula$ is an assignment  that evaluates $\propformula$  to true and is denoted by $ \sigma^* \models \propformula $.
+The propositional satisfiability (SAT) problem finds a satisfying assignment $ \sigma^* $ to a CNF formula $ \propformula $  such that $ \forall i, \sigma^* \models \propclause_i $, wherein $ \sigma^*\models \propclause_i $ if and only if $ \exists \literal_j \in \propclause_i, \sigma^*(\literal_j) = \text{true} $.
+Informally, $ \sigma^* $ satisfies at least one literal in each clause of a CNF.
+
+\subsection{Relaxation of Logical Formulas}
+
+A hard-OR or a soft-AND clause is a tuple $ (\propclause,\card) $ with an extra parameter $ \card $, where $ \card $ is the threshold on the literals in $ C $.
+Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $.
+Motivated by Boolean cardinality constraints, we propose relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$.
+We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{v \in \propclause_i} \mathds{1}[\sigma^* \models v] \geq \card_l$.
+Informally,  $\sigma^*$ satisfies a clause $(\propclause_i, \card_l)$ if  at least $\card_l$ literals in $\propclause_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(\propformula,\card_c, \card_l)$ if at least $\card_c$  clauses out of all $\{(\propclause_i,\card_l)\}$ clauses are true.
+
+\begin{theorem}[\cite{benhamou1994two}]
+	\label{interpretability_crr_thm:succinctness}	
+	Let $ (\propclause,\card) $ be a clause in relaxed-CNF where  $ C $ has $ m  $ literals and $ \card \in \{1,\dots,m\} $ is the threshold on literals.
+An equivalent compact encoding of $ (\propclause,\card) $  into a CNF formula $ \propformula= \bigwedge_i \propclause_i $ requires $ {m \choose{m-\card+1}} $  clauses where each clause is distinct and has $ m-\card+1 $ literals of $ (\propclause,\card) $.
+Therefore,  the total number of literals in $ \propformula $ is $ (m-\card+1){m\choose{m-\card+1}} = m $ when $ \card=1 $, otherwise $ (m-\card+1){m\choose{m-\card+1}} > m $.
+\end{theorem}
+
+\subsection{Inner Product}
+In the thesis, we use true as $ 1 $ and false as $ 0 $ interchangeably.
+Between two vectors  $\mathbf{U}$ and $\mathbf{V}$ of the same length defined over Boolean variables or constants (such as  $ 0 $ and $ 1 $), we define $\mathbf{U} \circ \mathbf{V} $  to refer to  their inner product.
+Formally, $\mathbf{U} \circ \mathbf{V} \triangleq \bigvee_{i} (U_{i} \wedge V_{i})$ is a disjunction of element-wise conjunction where $U_{i}$ and $V_{i}$ denote the $ i^\text{th} $ variable/constant of $\mathbf{U}$ and $\mathbf{V}$, respectively.
+In this context, the conjunction ``$\wedge$'' between a variable and a constant follows the standard interpretation: $\bool \wedge 0 = 0$ and $\bool \wedge 1 = \bool$.
+
+For example, let us consider a vector of variables $ \mathbf{U} = [U_1, U_2, U_3] $ and a vector of constants of the same length $ \mathbf{v} = [0,1,1] $.
+Then, $ \mathbf{U} \circ \mathbf{v} = U_2 \vee U_3 $.
+Applying numerical interpretation, the inner product can also be expressed as $\mathbf{U} \circ \mathbf{V} \triangleq \sum_{i} (U_{i} \cdot V_{i}) $.
+Hence, for the running example $ \mathbf{U} = [U_1, U_2, U_3] $ and $ \mathbf{v} = [0,1,1] $, we derive $ \mathbf{U} \circ \mathbf{v} = u_2  + u_3 $.
+In the thesis, we use Boolean interpretation of the inner product for learning CNF classification rules in Chapter~\ref{chapter:imli} and numerical interpretation for learning relaxed-CNF classification rules in Chapter~\ref{chapter:crr}.
+
+\subsection{MaxSAT}
+
+The MaxSAT problem is an optimization analog to the SAT problem that finds an optimal assignment satisfying the maximum number of clauses in a CNF.
+In interpretable machine learning, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula.
+In a weighted-CNF formula, a weight $ {\clauseweight{\propclause_i}} \in \mathbb{R}^+ \cup \{\infty\} $ is defined over each clause $ \propclause_i $, wherein $\propclause_i$ is called a \emph{hard} clause if $\clauseweight{\propclause_i} = \infty$, and  $\propclause_i$ is  a \emph{soft} clause, otherwise.
+To avoid notational clutter, we overload {\clauseweight{\cdot}} to denote the weight of an assignment $ \sigma $.
+In particular, we define $\clauseweight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $$\clauseweight{\sigma} = \sum_{i | \sigma \not \models \propclause_i} \clauseweight{\propclause_i}.$$
+
+Given a weighted-CNF formula $\propformula \triangleq \bigvee_i \propclause_i$ with weight $ \clauseweight{\propclause_i} $, the weighted-partial MaxSAT problems finds an optimal assignment $\sigma^*$ that achieves the minimum weight.
+Formally,  $\sigma^* = \MaxSAT(\propformula,\clauseweight{\cdot})$ if $\forall \sigma \neq \sigma^*, \clauseweight{\sigma^*} \leq \clauseweight{\sigma}$.
+The optimal weight of the MaxSAT problem $  \clauseweight{\sigma^*} $ is infinity ($ \infty $) when at least one hard clause becomes unsatisfied.
+Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses\footnote{In our formulation, we assume that there is a satisfying assignment to a CNF formula containing all hard clauses.} and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum.
+In Chapter~\ref{chapter:imli}, we reduce the classification problem  to the solution of a weighted-partial MaxSAT problem.
+The knowledge of the inner workings of {MaxSAT} solvers is  not required for this chapter.
+
+\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
+%Let $\boolset  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
+%A propositional formula $\propformula$ defined over $\boolset$ is in \textit{Conjunctive Normal Form} (CNF) if $\propformula$   is  a conjunction of clauses and each clause is a disjunction of literals.
+%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
+%Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \boolset $  such that $ \sigma(\bool_i) \in \{1, 0\} $.
+The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \boolset $ such that the formula $ \propformula $ is evaluated to be $1$ (equivalently, true).
+%In contrast to the SAT problem,
+
+The stochastic Boolean satisfiability (SSAT) problem~\cite{littman2001stochastic} is a counting analog of the SAT problem concerning with the probability of satisfaction of the formula.
+The SSAT problem computes the  probability of satisfaction of a CNF formula $\propformula$ defined on the ordered set of \textit{quantified} Boolean variables $ \boolset $.
+An SSAT formula is of the form
+\begin{equation}\label{fairness_justicia_eq:ssat}
+	\ssatformula = \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m,\; \propformula, 
+\end{equation}
+where $ \varquantifier_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\propformula$ is a quantifier-free CNF formula.
+In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[\bool_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
+In the SSAT formula $ \ssatformula $, the quantifier part $ \varquantifier_1\bool_1, \dots, \varquantifier_m \bool_m $ is known as the \textit{prefix} of the formula $ \propformula $.
+Let $ \bool $ be the outermost variable in the prefix.
+The semantics of SSAT formulas are defined recursively in the following.
+
+\begin{enumerate}
+	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
+	\item $ \Pr [\ssatformula] = \max_{\bool} \{\Pr[\ssatformula|_{\bool}], \Pr[\ssatformula|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
+	\item $ \Pr [\ssatformula] = \min_{\bool} \{\Pr[\ssatformula|_{\bool}], \Pr[\ssatformula|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
+	\item $ \Pr [\ssatformula] = p\Pr[\ssatformula|_{\bool}] + (1-p) \Pr[\ssatformula|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
+\end{enumerate}
+
+where $ \ssatformula|_{\bool} $ and $ \ssatformula|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \propformula $ with $ 1 $ and $ 0 $, respectively.
+In this thesis, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
+In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
+
+\begin{remark}
+	ER-SSAT problem is $\mathrm{NP}^{\mathrm{PP}}$-hard whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
+\end{remark}
+
+The problem of SSAT and its variants have been pursued by theoreticians and practitioners for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
+We refer the reader to~\cite{lee2017solving,lee2018solving} for a detailed survey.
+It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
+
+\subsection{Stochastic Subset Sum Problem ({\stochastic})} 
+Let $ \boolset \triangleq \{\bool_i\}_{i=1}^{|\boolset|}$ be a set of Boolean variables and $ \varweight_i \in \mathbb{Z} $ be the weight of $ \bool_i $.
+Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $\boolset$.
+Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
+The \textit{counting} version of the subset-sum problem counts all $ \boolset $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
+In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic.
+More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $.
+Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.
+
+\section{Interpretable Machine Learning}
+\label{chapter_interpretability_preliminaries}
+
+We consider a standard binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
+Here, the feature valuation vector $ \mathbf{x} = [x_1, \dots, x_{m}] \in \{0, 1\}^m $ is a vector of $ m $ Boolean features with $ y \in \{0,1\} $ being the binary class label.
+Thus, $ (\mathbf{x}, y) $ is called a positive sample if $ y = 1 $, and a negative sample otherwise.
+We use $ (\nonsensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, y) $.
+Hence, each feature $ X_j \in \nonsensitive $ is sampled from a Bernoulli distribution $ \mathcal{X}_j $, and $ \mathcal{D} = \prod_{j=1}^m \mathcal{X}_j  $ is the product distribution of all features.
+
+%$ \mathbf{X} \in \{0,1\}^{n\times m} $ is a binary feature matrix\footnote{Our framework allows learning on real-valued and categorical features, as discussed in Section~\ref{interpretability_imli_sec:non-binary}.} defined over Boolean features $\mathbf{x} = \{x_1, x_2, \dots, x_m\}$ and $ \mathbf{y} \in \{0,1\}^n $ is a vector of binary class labels.
+The $ i^\text{th} $ row of $ \mathbf{X} $, denoted by $ \mathbf{X}_i \in \mathcal{X} $, is called a sample generated from a product distribution of all features $ \mathcal{X} $.
+Thus, $ \mathbf{X}_i $ is the valuation of features in $ \mathbf{x} $ and $ y_i $ is its class label.
+$ \mathbf{X}_i $ is called a positive sample if $ y_i = 1 $, and a negative sample otherwise.
+
+\subsection{Rule-based Classification}
+A classifier $ \Rule: \mathbf{X} \rightarrow \widehat{Y} \in  \{0,1\} $ is a  function that takes a feature vector as input and outputs the predicted class label $\widehat{Y}$.
+%We define two rules  $ \R_1 $ and $ \R_2 $ to be  equivalent  if $ \forall i,\R_1(\X_i)=\R_2(\X_i) $.
+In a rule-based classifier, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{x} $ such that, if $ \Rule $ becomes true for an input (or an assignment), the prediction $ \widehat{Y} = 1 $ and $ \widehat{Y} = 0 $, otherwise.
+The goal is to design $\Rule$ not only to approximate the training data, but also to generalize to unseen samples arising from the same distribution.
+Additionally, we prefer to learn a \emph{sparse} $ \Rule $ in order to favor interpretability.
+We define the structural complexity of $ \Rule $, referred to as \emph{rule-size}, in terms of the number of literals it contains.
+Let $\clause(\Rule,i)$ denote  the  $i^\text{th}$ clause of $\Rule$ and $ |\clause(\Rule,i)| $ be  the number of literals in $\clause(\Rule,i)$.
+Thus, the size of the classifier is $|\Rule| =  \Sigma_i |\clause(\Rule,i)| $, which is the sum of literals in all clauses in $ \Rule $.
+
+\begin{example}
+	\normalfont
+	Let $ \Rule \triangleq (X_1 \vee X_3) \wedge (\neg X_2 \vee X_3) $ be a CNF classifier defined over three Boolean features $ [X_1, X_2, X_3] $.
+For an input $ [0, 1, 1] $, the prediction of $ \Rule  $ is $ 1 $, whereas for an input $ [1, 1, 0] $, the prediction is $ 0 $ because $ \Rule $ is true and false in these two cases, respectively.
+Moreover, $ |\Rule| = 2 + 2 = 4 $ is the size of $ \Rule $.
+\end{example}
+
+\subsection{Decision Lists.} A decision list is a rule-based classifier consisting of ``if-then-else'' rules.
+Formally, a decision list~\cite{rivest1987learning} is an ordered list $ \Rule_L $ pairs $ (C_1, V_1), \dots, (C_k, V_k) $ where the  rule $ C_i $ is a conjunction of literals (alternately, a single clause in a DNF formula) and $ V_i \in \{0,1\} $ is  a Boolean class label\footnote{In a more practical setting, $ V_i \in \{0, \dots, N\} $  can be multi-class for $ N \ge 1 $.}.
+Additionally, the last clause $ C_k \triangleq  $ true, thereby $ V_k $ is the default class.
+A decision list is defined as a classifier with the following interpretation: for an input feature vector $ \mathbf{x} $,  $ \Rule_L(\mathbf{x}) $ is equal to $ V_i $ where $ i $ is the least index such that the feature vector satisfies the rule, $ \mathbf{x} \models C_i $.
+Since the last clause is true, $ \Rule_L(\mathbf{x}) $ always exists.
+Intuitively, whichever clause (starting from the first) in $ \Rule_L $ is satisfied for an input, the associated class-label is considered as its prediction.
+
+\subsection{Decision sets.} A decision set is a set of \textit{independent} ``if-then'' rules.
+Formally,  a decision set $ \Rule_S $ is a set of pairs $ \{(C_1, V_1), \dots, (C_{k-1}, V_{k-1})\}  $ and a  default pair $ (C_k, V_k) $, where $ C_i $ is\textemdash similar to a decision list\textemdash a conjunction of literals and $ V_i \in \{0,1\} $ is a Boolean class label.
+In addition, the last clause $ C_k \triangleq  $ true and $ V_k $ is the default class.
+For a  decision set, if an input $ \mathbf{x} $ satisfies one clause, say $ C_i $, then the prediction is $ V_i $.
+If $ \mathbf{x} $ satisfies no clause, then the prediction is the default class $ V_k $.
+Finally, if $ \mathbf{x} $ satisfies  $ \ge 2 $ clauses, $ \mathbf{x} $ is assigned a class using a tie-breaking~\cite{lakkaraju2016interpretable}.
+
+\section{Fairness in Machine Learning}
 \label{chapter_fairness_preliminaries}
-In this chapter, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness.
-We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
-We express random variables in uppercase, and an assignment of a random variable in lowercase.
+In this section, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness.
 
 \subsection{Dataset and Distribution}
-We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
+We consider a dataset $ \dataset $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
 Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $.
@@
 
-We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
+We represent a binary classifier trained on the dataset $\dataset$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
 Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
@@
 Alternatively, a fairness metric can be computed on a finite dataset instead of on the distribution.
-In that case, we use the notation $ f(\alg, \mathbf{D}) $, which is the bias of the classifier $ \alg $ on a dataset $ \mathbf{D} $.
+In that case, we use the notation $ f(\alg, \dataset) $, which is the bias of the classifier $ \alg $ on a dataset $ \dataset $.
 
@@
 
-\subsection{Stochastic Boolean Satisfiability (SSAT)}\label{fairness_justicia_sec:ssat}
-Let $\mathbf{B}  = \{\bool_1, \dots, \bool_m\}  $ be a set of Boolean variables. A \textit{literal} is a variable $ \bool_i $ or its complement $ \neg \bool_i $. 
-A propositional formula $\phi$ defined over $\mathbf{B}$ is in \textit{Conjunctive Normal Form} (CNF) if $\phi$   is  a conjunction of clauses and each clause is a disjunction of literals. 
-%\red{DNF (disjunctive normal form) is the  complement of CNF where  the formula is a disjunction of clauses and each clause is  a conjunction of literals.} 
-Let $ \sigma $ be a assignment to the  variables $ \bool_i \in \mathbf{B} $  such that $ \sigma(\bool_i) \in \{1, 0\} $.
-The propositional  \textit{satisfiability} problem (SAT) finds a satisfying assignment $ \sigma^* $ to all $ \bool_i \in  \mathbf{B} $ such that the formula $ \phi $ is evaluated to be $1$ (equivalently, true).
-In contrast to the SAT problem, the \textit{stochastic Boolean satisfiability} (SSAT) problem~\cite{littman2001stochastic} computes the  probability of the satisfaction of the formula $\phi$ defined on an ordered set of \textit{quantified} Boolean variables.
-An SSAT formula is of the form
-\begin{equation}\label{fairness_justicia_eq:ssat}
-\Phi = Q_1\bool_1, \dots, Q_m \bool_m,\; \phi, 
-\end{equation}
-where $ Q_i \in \{\exists, \forall, \R^{p_i}\} $ is either of the existential ($\exists$), universal ($\forall$), or randomized ($\R^{p_i}$) quantifiers on $\bool_i$ and $\phi$ is a quantifier-free CNF formula.
-In case of a randomized quantifier $ \R^{p_i} $, $ p_i \triangleq \Pr[B_i = 1] \in [0,1] $ is the probability of $ \bool_i $ being assigned to $ 1 $.
-In the SSAT formula $ \Phi $, the quantifier part $ Q_1\bool_1, \dots, Q_m \bool_m $ is known as the \textit{prefix} of the formula $ \phi $.
-Let $ \bool $ be the outermost variable in the prefix.
-The semantics of SSAT formulas are defined recursively in the following.
-
-\begin{enumerate}
-	\item $ \Pr[\text{true}] = 1 $,  $ \Pr[\text{false}] = 0 $, 
-	\item $ \Pr [\Phi] = \max_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\}$ if $ \bool $ is existentially quantified ($ \exists $), 
-	\item $ \Pr [\Phi] = \min_{\bool} \{\Pr[\Phi|_{\bool}], \Pr[\Phi|_{\neg \bool}]\} $ if $ \bool $ is universally quantified ($ \forall $), 
-	\item $ \Pr [\Phi] = p\Pr[\Phi|_{\bool}] + (1-p) \Pr[\Phi|_{\neg \bool}] $ if $ \bool $ is randomized quantified ($\R^{p}$) with probability $p$ of being $\text{true}$,
-\end{enumerate}
-
-where $ \Phi|_{\bool} $ and $ \Phi|_{\neg \bool} $ denote the SSAT formulas derived by eliminating the outermost quantifier of $ \bool $  by substituting the value of $ \bool $ in the CNF $ \phi $ with $ 1 $ and $ 0 $, respectively.
-In this thesis, we focus on two specific types of SSAT formulas:  \textit{random-exist} (RE) SSAT and \textit{exist-random} (ER) SSAT.
-In the ER-SSAT (resp.\ RE-SSAT) formula, all existentially (resp.\ randomized) quantified variables are followed by randomized (resp.\ existentially) quantified variables in the prefix.
-
-\begin{remark}
-	ER-SSAT problem is $\mathrm{NP}^{\mathrm{PP}}$-hard whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
-\end{remark}
-
-The problem of SSAT and its variants have been pursued by theoreticians and practitioners for over three decades~\cite{majercik2005dc,fremont2017maximum,huang2006combining}.
-We refer the reader to~\cite{lee2017solving,lee2018solving} for a detailed survey.
-It is worth remarking that the past decade has witnessed a significant performance improvements of SSAT solving, thanks to the close integration of techniques from SAT solving with advances in weighted model counting~\cite{sang2004combining,chakraborty2013scalable,chakraborty2014distribution}.
-
-\subsection{Stochastic Subset Sum Problem ({\stochastic})} 
-Let $ \mathbf{B} \triangleq \{B_i\}_{i=1}^{|\mathbf{B}|}$ be a set of Boolean variables and $ W_i \in \mathbb{Z} $ be the weight\change{Weight notation $ w $ to $ W $} of $ B_i $.
-Given a constraint of the form  $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\mathbf{b} \in \{0,1\}^{|\mathbf{B}|}$ such that the constraint evaluates to true when $\mathbf{B}$ is substituted with $\mathbf{b}$.
-Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
-The \textit{counting} version of the subset-sum problem counts all $ \mathbf{b} $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
-In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\mathbf{B}|} W_i B_i \ge \tau $ where variables $ B_i $'s are stochastic.
-More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\mathbf{B}|} W_iB_i \ge \tau] $.
-Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.
-
 \subsection{Bayesian Network}
@@
 In fairness verification with correlated features, we leverage an access to a Bayesian network on features $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
-A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\mathbf{V}, \mathbf{E}) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
-Each vertex $V_i \in \mathbf{V}$ corresponds to a random variable.
-Edges $ \mathbf{E} \in \mathbf{V} \times \mathbf{V} $ imply conditional dependencies among variables.
-For each variable $ V_i \in \mathbf{V} $, let $ \parent(V_i) \subseteq \mathbf{V} \setminus \{V_i\} $ denote the set of parents of $ V_i $.
-Given $\parent(V_i)$ and parameters $\theta$, $ V_i $ is independent of its other non-descendant variables in $\graph$.
-Thus, for the assignment $ v_i $ of $ V_i $ and $ \mathbf{u} $ of $ \parent(V_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\mathbf{V}$ as:
+A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\graphvertices, \graphedges) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
+Each vertex $\graphvertex_i \in \graphvertices$ corresponds to a random variable.
+Edges $ \graphedges \in \graphvertices \times \graphvertices $ imply conditional dependencies among variables.
+For each variable $ \graphvertex_i \in \graphvertices $, let $ \parent(\graphvertex_i) \subseteq \graphvertices \setminus \{\graphvertex_i\} $ denote the set of parents of $ \graphvertex_i $.
+Given $\parent(\graphvertex_i)$ and parameters $\theta$, $ \graphvertex_i $ is independent of its other non-descendant variables in $\graph$.
+Thus, for the assignment $ v_i $ of $ \graphvertex_i $ and $ \mathbf{u} $ of $ \parent(\graphvertex_i) $, the aforementioned semantics of a Bayesian network encodes the joint distribution of $\graphvertices$ as:
 
@@
 	\begin{split}
-\Pr[V_1=v_1, \dots, &V_{|\mathbf{V}|}=v_{|\mathbf{V}|}] = \prod_{i=1}^{|\mathbf{V}|} \Pr[V_i = v_i | \parent(V_i) = \mathbf{u}; \theta].
+		\Pr[\graphvertex_1=v_1, \dots, &\graphvertex_{|\graphvertices|}=v_{|\graphvertices|}] = \prod_{i=1}^{|\graphvertices|} \Pr[\graphvertex_i = v_i | \parent(\graphvertex_i) = \mathbf{u}; \theta].
 	\end{split}
@@
 Together all terms sum to the total variance of the model output.
-
-%\todo{End}

### 2022-06-14 chapters/interpretability/imli/baseline.tex
minor edits
preliminaries notations updated

@@
 \label{interpretability_imli_sec:variables} 
-We initially preprocess $ \mathbf{X} $  to account for the negation of  Boolean features while learning a classifier.\footnote{This preprocessing is similarly applied in  \cite{malioutov2013exact}.} In the preprocessing step, we negate each column in $ \mathbf{X} $ to a new column and append it to $ \mathbf{X} $.
-For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{X} $ by negating the column ``$ \text{age }\ge 25 $''.
-Hence, in the rest of the chapter, we refer $ m $ as the modified number of columns in $ \mathbf{X} $.
+We initially preprocess feature vector $ \mathbf{x} $  to account for the negation of  Boolean features while learning a classifier.\footnote{This preprocessing is similarly applied in  \cite{malioutov2013exact}.} In the preprocessing step, we negate each feature in $ \mathbf{x} $ to a new feature and append it to $ \mathbf{x} $.
+For example, if ``$ \text{age }\ge 25 $'' is a Boolean feature, we add another feature ``$ \text{age }< 25 $'' in  $ \mathbf{x} $ by negating the feature ``$ \text{age }\ge 25 $''.
+Hence, in the rest of the chapter, we refer $ m $ as the modified number of features in $ \mathbf{x} $.
 We next discuss the variables in the MaxSAT problem.
 
-We consider two types of Boolean variables: (i) \emph{feature} variables $ b $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples.
-We define a Boolean variable $ b^i_j $ that becomes true if feature $ x_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ b^i_j $ is assigned false, otherwise.
-Moreover, we define an error variable $ \eta_l $ to attribute to whether the sample $ \mathbf{X}_l $ is classified correctly or not.
-Specifically, $ \eta_l $ becomes true if $ \mathbf{X}_l $ is misclassified, and becomes false otherwise.
+We consider two types of Boolean variables: (i) \emph{feature} variables $ \bool $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples.
+We define a Boolean variable $ \bool^i_j $ that becomes true if feature $ X_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ \bool^i_j $ is assigned false, otherwise.
+Moreover, we define an error variable $ \eta_l $ to attribute to whether the $ l^\text{th} $ sample $ (\mathbf{x}^{(l)}, y^{(l)}) $ is classified correctly or not.
+Specifically, $ \eta_l $ becomes true if $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified, and becomes false otherwise.
 We next discuss the MaxSAT encoding to solve the classification problem.
@@
 \label{interpretability_imli_sec:encoding}
-We consider a partial-weighted MaxSAT formula, where we encode the objective function in Eq.~\eqref{interpretability_imli_eq:obj} as soft clauses.
-For each sample in $ \mathbf{X} $, we construct hard clauses specifying that the sample is either  classified correctly and $ \eta_l $ is false or it is misclassified and $ \eta_l $ is true.
+We consider a partial-weighted MaxSAT formula, where we encode the objective function in Eq.~\eqref{interpretability_imli_eq:obj} as soft clauses and the learning constraints as hard clauses.
 We next discuss the MaxSAT encoding in detail.
@@
 	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible.
-Hence, for each feature variable $ b_j^i $, {\imli} constructs a unit clause as $ \neg b_j^i $.
+Hence, for each feature variable $ \bool_j^i $, {\imli} constructs a unit clause as $ \neg \bool_j^i $.
 Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
@@
 		\label{interpretability_imli_soft_clause_feature}
-		S_j^i:= \neg b_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
+		S_j^i:= \neg \bool_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
 	\end{equation}
@@
 In {\imli}, we have a constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa.
-Let  $\mathbf{b}_i= \{b^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $.
+Let  $\boolset_i= \{\bool^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $.
 Then, we define the following hard clause.
@@
 			\label{interpretability_imli_hard_clause_rule_consistency}
-		H_l:= \neg \eta_l \rightarrow \Big(y_l\leftrightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
+		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
 	\end{equation}
 
-	In the hard clause, $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ is a CNF formula including variables $ b_j^i $ for which the associated  feature-value is $ 1 $ in  $ \mathbf{X}_l $.
-Since $ y_l \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement.
-Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}\ $ or its complement is true depending on the class-label of the sample.
+	In the hard clause, $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ is a CNF formula including variables $ \bool_j^i $ for which the associated  feature-value is $ 1 $ in  $ \mathbf{x}^{(l)} $.
+Since $ y^{(l)} \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement.
+Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement is true depending on the class-label of the sample.
 We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \eta_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
@@
 The soft clauses $ E_l $ and $ S_j^i $ are unit clauses and hence, no translation is required for them.
-In the hard clause, when $ y_l=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $.
+In the hard clause, when $ y^{(l)}=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $.
 In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF.
-In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$.
-Since $ \mathbf{X}_l \circ \mathbf{b}_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{X}_l \circ \mathbf{b}_{i}) $ into CNF.
-More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $.
-Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{X}_l} \circ {\mathbf{b}_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})  $ for $ i=\{1,\dots,k\} $.
+In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
+Since $ \mathbf{x}^{(l)} \circ \boolset_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{x}^{(l)} \circ \boolset_{i}) $ into CNF.
+More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $.
+Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})  $ for $ i=\{1,\dots,k\} $.
 Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF.
-For either value of $ y_l $, the weight of each translated hard clause in the CNF formula is $ \infty $.
+For either value of $ y^{(l)} $, the weight of each translated hard clause in the CNF formula is $ \infty $.
 
-% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ when $ y_l=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{X}_l} \circ {\mathbf{b}_{i}})$.
+% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ when $ y^{(l)}=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
 
-% In the hard clause,  $\bigwedge_{i=1}^k {\mathbf{X}_l} \circ {\mathbf{b}_{i}}  $ is already in CNF.
-When $ y_l=1 $, the constraint $ H_l $ can be directly encoded as CNF by using equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $.
+% In the hard clause,  $\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ is already in CNF.
+When $ y^{(l)}=1 $, the constraint $ H_l $ can be directly encoded as CNF by using equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $.
 When $ y_i=0 $, we apply
@@
 
-Any off-the-shelf MaxSAT solver can output an optimal assignment $ \sigma^* $ of the MaxSAT query $ (Q, \weight{\cdot}) $.
+Any off-the-shelf MaxSAT solver can output an optimal assignment $ \sigma^* $ of the MaxSAT query $ (Q, \clauseweight{\cdot}) $.
 We extract $ \sigma^* $ to construct the classifier $ \Rule $ and compute training errors as follows.
@@
 	\label{interpretability_imli_construction:rule}
-	Let $\sigma^* = \MaxSAT(Q,\weight{\cdot})$.
-Then $x_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(b_{j}^{i}) = 1$.
-Additionally, $ \mathbf{X}_l $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
+	Let $\sigma^* = \MaxSAT(Q,\clauseweight{\cdot})$.
+Then $X_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(\bool_{j}^{i}) = 1$.
+Additionally, $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
 \end{construction}
@@
 
-\subsubsection*{Example of encoding}
+\begin{example}{MaxSAT Encoding.}
+\normalfont
 
@@
 \[ 
-\mathbf{X}_\mathsf{orig} = 
-\begin{blockarray}{cc}
-x_1 & x_2 \\
-\begin{block}{[cc]}
-0 & 0 \\
-0 & 1 \\
-1 & 0 \\
-1 & 1 \\
+\dataset_\mathsf{orig} = 
+\begin{blockarray}{ccc}
+X_1 & X_2 & Y\\
+\begin{block}{[ccc]}
+0 & 0 & 1\\
+0 & 1 & 0\\
+1 & 0 & 0\\
+1 & 1 & 1\\
 \end{block}
 \end{blockarray} \implies 
-\mathbf{X} = 
-\begin{blockarray}{cccc}
-x_1 & \neg x_1 & x_2  & \neg x_2\\
-\begin{block}{[cccc]}
-0 & 1 & 0 & 1\\
-0 & 1 & 1 & 0\\
-1 & 0 & 0 & 1\\
-1 & 0 & 1 & 0\\
+\dataset = 
+\begin{blockarray}{ccccc}
+X_1 & \neg X_1 & X_2  & \neg X_2 & Y\\
+\begin{block}{[ccccc]}
+0 & 1 & 0 & 1 & 1\\
+0 & 1 & 1 & 0 & 0\\
+1 & 0 & 0 & 1 & 0\\
+1 & 0 & 1 & 0 & 1\\
 \end{block}
-\end{blockarray}; \quad 
-\mathbf{y} = 
-\begin{bmatrix} 
-1 \\ 0 \\ 0 \\ 1
-\end{bmatrix}
+\end{blockarray}
 \]
-In the preprocessing step, we negate the columns in $ \mathbf{X}_\mathsf{orig} $ and add complemented features $ \{\neg x_1, \neg x_2\} $ in $ \mathbf{X} $.
+In the preprocessing step, we negate the features in $ \dataset_\mathsf{orig} $ and add complemented features $ \{\neg X_1, \neg X_2\} $ in $ \dataset $.
 In the MaxSAT encoding, we define $ 8 $ feature variables ($ 4 $ features $ \times $ $ 2 $ clauses in the classifier) and $ 4 $ error variables.
-For example, for feature $ x_2 $, introduced feature variables are $ \{b^1_3, b^2_3\} $ and for feature $ \neg x_2 $, introduced variables are $ \{b^1_4, b^2_4\} $.
+For example, for feature $ X_2 $, introduced feature variables are $ \{\bool^1_3, \bool^2_3\} $ and for feature $ \neg X_2 $, introduced variables are $ \{\bool^1_4, \bool^2_4\} $.
 For four samples, error variables are $ \{\eta_1, \eta_2, \eta_3, \eta_4\} $.
@@
 E_4:= (\neg \eta_4)\\
-& S_1^1 := (\neg b_1^1);\quad 
- S^1_2 := (\neg b^1_2);\quad 
- S^1_3 := (\neg b^1_3);\quad
- S^1_4 := (\neg b^1_4);\quad \\ 
-& S_1^2 := (\neg b_1^2);\quad 
-S^2_2 := (\neg b^2_2);\quad 
-S^2_3 := (\neg b^2_3);\quad 
-S^2_4 := (\neg b^2_4);\quad\\ 
-& H_1:= (\neg \eta_1 \rightarrow ((b_2^1 \vee b_4^1)\wedge (b_2^2 \vee b_4^2)));\\
-& H_2:= (\neg \eta_2 \rightarrow (\neg(b_2^1 \vee b_3^1) \vee \neg(b_2^2 \vee b_3^2)));\\
-& H_3:= (\neg \eta_3 \rightarrow (\neg (b_1^1 \vee b_4^1) \vee \neg(b_1^2 \vee b_4^2)));\\
-& H_4:= (\neg \eta_4 \rightarrow ((b_1^1 \vee b_3^1) \wedge (b_1^2 \vee b_3^2)));\\
+& S_1^1 := (\neg \bool_1^1);\quad 
+ S^1_2 := (\neg \bool^1_2);\quad 
+ S^1_3 := (\neg \bool^1_3);\quad
+ S^1_4 := (\neg \bool^1_4);\quad \\ 
+& S_1^2 := (\neg \bool_1^2);\quad 
+S^2_2 := (\neg \bool^2_2);\quad 
+S^2_3 := (\neg \bool^2_3);\quad 
+S^2_4 := (\neg \bool^2_4);\quad\\ 
+& H_1:= (\neg \eta_1 \rightarrow ((\bool_2^1 \vee \bool_4^1)\wedge (\bool_2^2 \vee \bool_4^2)));\\
+& H_2:= (\neg \eta_2 \rightarrow (\neg(\bool_2^1 \vee \bool_3^1) \vee \neg(\bool_2^2 \vee \bool_3^2)));\\
+& H_3:= (\neg \eta_3 \rightarrow (\neg (\bool_1^1 \vee \bool_4^1) \vee \neg(\bool_1^2 \vee \bool_4^2)));\\
+& H_4:= (\neg \eta_4 \rightarrow ((\bool_1^1 \vee \bool_3^1) \wedge (\bool_1^2 \vee \bool_3^2)));\\
 \end{split}
@@
 
-In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \weight{E_i} = 1 $ and rule-sparsity weight as $ \weight{S_j^i} = 0.1 $.
+In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \clauseweight{E_i} = 1 $ and rule-sparsity weight as $ \clauseweight{S_j^i} = 0.1 $.
 Intuitively, the penalty for misclassifying a sample is $ 10 $ times more than the penalty for adding a feature in the classifier.
-For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{b^1_1, b^1_4, b^2_2, b^2_3\} $ to true.
-Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (x_1 \vee \neg x_2)  \wedge (\neg x_1 \vee x_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ x_i $ and $ \neg x_i $ in the same clause.
-The reason is that a clause with both $ x_i $ and $ \neg x_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}
+For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{\bool^1_1, \bool^1_4, \bool^2_2, \bool^2_3\} $ to true.
+Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (X_1 \vee \neg X_2)  \wedge (\neg X_1 \vee X_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ X_i $ and $ \neg X_i $ in the same clause.
+The reason is that a clause with both $ X_i $ and $ \neg X_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}
+
+\end{example}
 
@@
 For each continuous feature, we apply equal-width discretization that splits the feature into a fixed number of bins.
-For example, consider a continuous feature $ x_c \in [a,b] $.
+For example, consider a continuous feature $ X_c \in [a,b] $.
 In discretization, we split the domain $ [a,b] $ into three bins with two split points $ \{a',b'\} $ such that $ a<a'<b'<b $.
-Therefore, the resulting three discretized features are $ { a \le x_c < a'} $, $ {a' \le x_c < b'} $, and $ { b' \le x_c  \le b } $.
+Therefore, the resulting three discretized features are $ { a \le X_c < a'} $, $ {a' \le X_c < b'} $, and $ { b' \le X_c  \le b } $.
 An alternate to this close-interval discretization is open-interval discretization, where we consider six discretized features with each feature being compared with one threshold.
-In that case, the discretized features are: $ x_c \ge a, x_c \ge a', x_c \ge b', x_c < a', x_c < b', x_c \le b $.
+In that case, the discretized features are: $ X_c \ge a, X_c \ge a', X_c \ge b', X_c < a', X_c < b', X_c \le b $.
 Both open-interval and close-interval discretization techniques have their use-cases where one or the other is appropriate.
@@
 In {\imli}, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $.
-For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg b_j^i $ can be set to $ \infty $.
-In contrast, to always include a feature, we can define $ S_j^i := b_j^i $ with weight $ \infty $.
+For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg \bool_j^i $ can be set to $ \infty $.
+In contrast, to always include a feature, we can define $ S_j^i := \bool_j^i $ with weight $ \infty $.
 In both cases, we treat $ S_j^i $ as a hard clause.
@@
 Another use case may be to learn rules where clauses have disjoint set of features.
-To this end, we consider a pseudo-Boolean constraint $ \sum_{i=1}^k b^i_j \le 1 $, which specifies that the feature $ x_j $ appears in at-most one of the $ k $ clauses.
+To this end, we consider a pseudo-Boolean constraint $ \sum_{i=1}^k \bool^i_j \le 1 $, which specifies that the feature $ X_j $ appears in at-most one of the $ k $ clauses.
 This constraint may be soft or hard depending on the priority in the application domain.

### 2022-06-14 chapters/interpretability/imli/problem_formulation.tex
preliminaries notations updated

@@
  \label{interpretability_imli_sec:problem}
-Given (i) a binary feature matrix $ \mathbf{X} \in \{0,1\}^{n \times m} $ with $ n $ samples and $ m $ features, (ii) a class-label vector $ \mathbf{y} \in \{0,1\}^n $, (iii) a positive integer $ k \ge 1 $ denoting the number of clauses, and (iv) a regularization parameter $ \lambda \in \mathbb{R}^+ $, we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.
+Given 
+\begin{enumerate}
+	\item a dataset $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ of $ n $ samples, where feature vector $ \mathbf{x}^{(i)} \in \{0, 1\}^m $ contains $ m $ features and class label $ y^{(i)} \in \{0,1\} $,
+	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, and
+	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $,
+\end{enumerate}
+we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.
 
@@
  \[
- \min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y_i=\mathcal{R}(\mathbf{X}_i)
+ \min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y^{(i)}=\mathcal{R}(\mathbf{x}^{(i)})
  \]
@@
 Hence, we need to balance interpretability with classification error.
-Let $ \mathcal{E}_\mathbf{X} = \{(\mathbf{X}_i,y_i) | y_i \ne \mathcal{R}(\mathbf{X}_i) \} $   be  the set of samples in $ \mathbf{X} $ that are misclassified  by $ \mathcal{R} $.
+Let $ \mathcal{E}_\dataset = \{(\mathbf{x}^{(i)},y^{(i)}) | y^{(i)} \ne \mathcal{R}(\mathbf{x}^{(i)}) \} $   be  the set of samples in $ \dataset $ that are misclassified  by $ \mathcal{R} $.
 Therefore, we balance between classification-accuracy and rule-sparsity and optimize the following function.\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g., to penalize  false-alarms more than mis-detects), and to allow instance weights (per sample).}
@@
   \label{interpretability_imli_eq:obj}
- \min\limits_{\mathcal{R}} \;  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}|
+ \min\limits_{\mathcal{R}} \;  |\mathcal{E}_\dataset| +\lambda |\mathcal{R}|
  \end{equation}

### 2022-06-14 chapters/interpretability_background.tex
preliminaries notations updated

@@
 In interpretable machine learning, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula.
-In a weighted-CNF formula, a weight $ {\weight{C_i}} \in \mathbb{R}^+ \cup \{\infty\} $ is defined over each clause $ C_i $, wherein $C_i$ is called a \emph{hard} clause if $\weight{C_i} = \infty$, and  $C_i$ is  a \emph{soft} clause, otherwise.
-To avoid notational clutter, we overload {\weight{\cdot}} to denote the weight of an assignment $ \sigma $.
-In particular, we define $\weight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $$\weight{\sigma} = \sum_{i | \sigma \not \models C_i} \weight{C_i}.$$
+In a weighted-CNF formula, a weight $ {\clauseweight{C_i}} \in \mathbb{R}^+ \cup \{\infty\} $ is defined over each clause $ C_i $, wherein $C_i$ is called a \emph{hard} clause if $\clauseweight{C_i} = \infty$, and  $C_i$ is  a \emph{soft} clause, otherwise.
+To avoid notational clutter, we overload {\clauseweight{\cdot}} to denote the weight of an assignment $ \sigma $.
+In particular, we define $\clauseweight{\sigma}$ as the sum of the weights of \emph{unsatisfied clauses} in a CNF, formally, $$\clauseweight{\sigma} = \sum_{i | \sigma \not \models C_i} \clauseweight{C_i}.$$
 
-Given a weighted-CNF formula $F \triangleq \bigvee_i C_i$ with weight $ \weight{C_i} $, the weighted-partial MaxSAT problems finds an optimal assignment $\sigma^*$ that achieves the minimum weight.
-Formally,  $\sigma^* = \MaxSAT(F,\weight{\cdot})$ if $\forall \sigma \neq \sigma^*, \weight{\sigma^*} \leq \weight{\sigma}$.
-The optimal weight of the MaxSAT problem $  \weight{\sigma^*} $ is infinity ($ \infty $) when at least one hard clause becomes unsatisfied.
+Given a weighted-CNF formula $F \triangleq \bigvee_i C_i$ with weight $ \clauseweight{C_i} $, the weighted-partial MaxSAT problems finds an optimal assignment $\sigma^*$ that achieves the minimum weight.
+Formally,  $\sigma^* = \MaxSAT(F,\clauseweight{\cdot})$ if $\forall \sigma \neq \sigma^*, \clauseweight{\sigma^*} \leq \clauseweight{\sigma}$.
+The optimal weight of the MaxSAT problem $  \clauseweight{\sigma^*} $ is infinity ($ \infty $) when at least one hard clause becomes unsatisfied.
 Therefore,  the weighted-partial MaxSAT problem finds $ \sigma^* $ that satisfies all hard clauses\footnote{In our formulation, we assume that there is a satisfying assignment to a CNF formula containing all hard clauses.} and as many soft clauses as possible such that the total weight of unsatisfied soft clauses is minimum.

### 2022-06-15 chapters/interpretability/imli/baseline.tex
imli paper updated with notations

@@
 		\label{interpretability_imli_soft_clause_error}
-		E_l:=\neg \eta_l;  \qquad \mathsf{wt}(E_l)= 1
+		E_l:=\neg \eta_l;  \qquad \clauseweight{E_l}= 1
 	\end{equation}
@@
 		\label{interpretability_imli_soft_clause_feature}
-		S_j^i:= \neg \bool_j^i; \qquad \mathsf{wt}(S_j^i)= \lambda
+		S_j^i:= \neg \bool_j^i; \qquad \clauseweight{S_j^i}= \lambda
 	\end{equation}
@@
 			\label{interpretability_imli_hard_clause_rule_consistency}
-		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \mathsf{wt}(H_l)=\infty
+		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \clauseweight{H_l}=\infty
 	\end{equation}

### 2022-06-15 chapters/interpretability/imli/experiment.tex
imli paper updated with notations

@@
 
-	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
-	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}\\
-	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
-	\subfloat{\includegraphics[scale=0.35]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_wdbc_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_adult_cactus_train_val_fit_time}}\\
+	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_weatherAUS_cactus_train_val_fit_time}}
+	\subfloat{\includegraphics[scale=0.4]{figures/interpretability/imli/dataset_labor_cactus_train_val_fit_time}}
 

### 2022-06-15 chapters/interpretability/imli/incremental_learning.tex
tables and figures updated with short caption
preliminaries notations updated
imli paper updated with notations

@@
 The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to learn in a CNF formula.
-In this section, we discuss an incremental learning technique that along with MaxSAT formulation can generate classification rules efficiently.
+In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently.
 Our incremental learning is built on two concepts: (i)
@@
 Intuitively, $ d_{H}(\Rule', \Rule) $ calculates the total number of different literals in each (ordered) pair of clauses in $ \Rule $  and $ \Rule' $.
-For example, consider $ \Rule = (x_1 \vee x_2) \wedge (\neg x_1) $ and $ \Rule' = (\neg x_1 \vee x_2) \wedge (\neg x_1) $.
-Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{x_1, \neg x_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $.
+For example, consider $ \Rule = (X_1 \vee X_2) \wedge (\neg X_1) $ and $ \Rule' = (\neg X_1 \vee X_2) \wedge (\neg X_1) $.
+Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{X_1, \neg X_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $.
 In the following, we discuss a modified objective function for mini-batch learning.
@@
 In mini-batch learning, we design an objective function to penalize both classification errors on the current batch and the Hamming distance between new and previous classifiers.
-Let $ (\mathbf{X}^b, \mathbf{y}^b) $ be the current mini-batch.
+Let $ \dataset_b \subset \dataset $ be the current mini-batch, where $ |\dataset_b| \ll |\dataset| $.
 The objective function in mini-batch learning is
 
-%\red{Moreover, $ (\mathbf{X}, \mathbf{y}) $ is the training dataset, on which we measure the performance of $ \Rule $ after completing incremental learning\textemdash we will expand shortly on how we measure the performance.
+%\red{Moreover, $ (\red{\dataset}, \mathbf{y}) $ is the training dataset, on which we measure the performance of $ \Rule $ after completing incremental learning\textemdash we will expand shortly on how we measure the performance.
 %}
@@
 \label{interpretability_imli_eq:obj_incr}
-\min_{\Rule} |\mathcal{E}_{\mathbf{X}_b}|+\lambda d_{H}(\Rule, \Rule').
+\min_{\Rule} |\mathcal{E}_{\dataset_b}|+\lambda d_{H}(\Rule, \Rule').
 \end{align}
 
-In the objective function, $ \mathcal{E}_{\mathbf{X}_b} $ is the misclassified subset of samples in the current batch $ (\mathbf{X}^b, \mathbf{y}^b) $.
-Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ controls the trade-off between classification errors and the syntactic differences between consecutive classifiers.
+In the objective function, $ \mathcal{E}_{\dataset_b} $ is the misclassified subset of samples in the current batch $ \dataset_b $.
+Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ \emph{controls the trade-off between classification errors and the syntactic differences between consecutive classifiers}.
 We next discuss how to encode the modified objective function as a MaxSAT query.
@@
 In order to account for the modified objective function, we only modify the soft clause $ S_j^i $ in the MaxSAT query $ Q $.
-In particular, we define $ S_j^i $ to penalize for the complemented assignment  of the feature variable $ b_j^i $ in $ \Rule $ compared to $ \Rule' $.
+In particular, we define $ S_j^i $ to penalize for the complemented assignment  of the feature variable $ \bool_j^i $ in $ \Rule $ compared to $ \Rule' $.
 \[
@@
 \begin{cases}
-b_j^i& \text{if }  x_j \in  \clause(\Rule',i)  \\
-\neg b_j^i & \text{otherwise}
-\end{cases};\qquad \mathsf{wt}(S_j^i)=\lambda
+\bool_j^i& \text{if }  X_j \in  \clause(\Rule',i)  \\
+\neg \bool_j^i & \text{otherwise}
+\end{cases};\qquad \clauseweight{S_j^i}=\lambda
 \]
 
-$ S_j^i $ is either a unit clause $ b_j^i $ or $ \neg b_j^i $ depending on whether the associated feature $ x_j $ appears in the previous classifier $ \Rule' $ or not.
-In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg b_j^i $ for the first batch.
+$ S_j^i $ is either a unit clause $ \bool_j^i $ or $ \neg \bool_j^i $ depending on whether the associated feature $ X_j $ appears in the previous classifier $ \Rule' $ or not.
+In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg \bool_j^i $ for the first batch.
 We next analyze the complexity of the MaxSAT encoding for mini-batch learning.
@@
 	\label{interpretability_imli_prop:maxsat_complexity_incremental}
-	Let $ n' \triangleq |\mathbf{X}_b| \ll |\mathbf{X}| $ be the size of a mini-batch, $ n'_{neg} \le n' $ be the number of negative samples in the batch, and $ m $ be the number of Boolean features.
+	Let $ n' \triangleq |\dataset_b| \ll |\dataset| $ be the size of a mini-batch, $ n'_{neg} \le n' $ be the number of negative samples in the batch, and $ m $ be the number of Boolean features.
 According to Proposition~\ref{interpretability_imli_prop:maxsat_variables}, to learn a $ k $-clause CNF classifier in mini-batch learning, the MaxSAT encoding for each batch has $ km + n' $ Boolean variables and $ kn'_{neg} $ auxiliary variables.
@@
 To tackle this, after learning on each batch, we decide whether to keep $ \Rule $ or not by assessing the performance of $ \Rule $ on the training data and keep $ \Rule $ if it achieves higher performance.
-We measure the performance of $ \Rule $ on the full training data $ (\mathbf{X}, \mathbf{y}) $ using a weighted combination of classification errors and rule-size.
-In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}| $ on the training data $ (\mathbf{X}, \mathbf{y}) $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}.
+We measure the performance of $ \Rule $ on the full training data $ \dataset $ using a weighted combination of classification errors and rule-size.
+In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\dataset| +\lambda |\mathcal{R}| $ on the training data $ \dataset $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}.
 Additionally, we  discard the current classifier $ \Rule $ when the loss does not decrease ($ \mathsf{loss}(\Rule) > \mathsf{loss}(\Rule') $).
@@
 	\begin{algorithmic}[1]
-		\Function{$ \mathsf{MiniBatchLearning} $}{$ \mathsf{X},\mathbf{y},\lambda, k $}
+		\Function{MiniBatchLearning}{$ \dataset, \lambda, k $}
 		\State $\mathcal{R} = \leftarrow \bigwedge_{i=1}^k\text{false} $ \Comment{Empty CNF formula}
@@
 		\For{$ i \leftarrow 1, \dots, N $} \Comment{$ N $ is the total batch-count}
-		\State $ \mathbf{X}^b, \mathbf{y}^b \leftarrow \mathsf{GetBatch}(\mathsf{X},\mathbf{y}) $
-		\State $ Q, \weight{\cdot} \leftarrow \mathsf{MaxSATEncoding}(\Rule, \mathbf{X}^b, \mathbf{y}^b, \lambda, k) $ \Comment{Returns a weighted-partial CNF}
-		\State $ \sigma^* \leftarrow \MaxSAT(Q,\weight{\cdot}) $
-		\State $ \Rule_{\mathsf{new}} \leftarrow \mathsf{ConstructClassifier(\sigma^*)} $
-		\State $ \mathsf{loss} \leftarrow  |\mathcal{E}_\mathbf{X}| +\lambda |\mathcal{R}_{\mathsf{new}}| $ \Comment{Compute loss}
+		\State $ \dataset_b \leftarrow \textsc{GetBatch}(\dataset) $
+		\State $ Q, \clauseweight{\cdot} \leftarrow \textsc{MaxSATEncoding}(\Rule, \dataset_b, \lambda, k) $ \Comment{Returns a weighted-partial CNF}
+		\State $ \sigma^* \leftarrow \MaxSAT(Q,\clauseweight{\cdot}) $
+		\State $ \Rule_{\mathsf{new}} \leftarrow \textsc{ConstructClassifier}(\sigma^*) $
+		\State $ \mathsf{loss} \leftarrow  |\mathcal{E}_\dataset| +\lambda |\mathcal{R}_{\mathsf{new}}| $ \Comment{Compute loss}
 		\If{$ \mathsf{loss} < \mathsf{loss}_{\max} $}

### 2022-06-15 chapters/interpretability/imli/iterative_learning.tex
imli paper updated with notations

@@
 		\State $ \mathcal{R} \leftarrow $ true \Comment{Initial formula}
-		\For{$ i \leftarrow 1, \dots, k $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
-		\State $ C_i \leftarrow \mathsf{MiniBatchLearning}( \mathbf{X},\mathbf{y},\lambda, 1) $ \Comment{Single clause learning, $ k = 1 $}
+		\For{$ i \leftarrow 1, \dots, k $ and $ \dataset \ne \emptyset $}
+		\State $ C_i \leftarrow \textsc{MiniBatchLearning}( \dataset,\lambda, 1) $ \Comment{Single clause learning, $ k = 1 $}
 		\label{interpretability_imli_algo:iterative_CNF_learning_incremental_learning}
-		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{Coverage}(\mathbf{X}, C_i) $
+		\State $ \dataset' \leftarrow \textsc{Coverage}(\dataset, C_i) $
 		\label{interpretability_imli_algo:iterative_CNF_learning_coverage}
-		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} \Comment{Terminating conditions}
+		\If{$ \dataset' = \emptyset $} \Comment{Terminating conditions}
 		\State break
@@
 		\State $ \mathcal{R} \leftarrow \mathcal{R}  \wedge C_i $
-		\State $ \mathbf{X},\mathbf{y}  \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $ \Comment{Removing covered samples}
+		\State $ \dataset  \leftarrow \dataset  \setminus \dataset' $ \Comment{Removing covered samples}
 		\EndFor
@@
 
-\paragraph{Terminating conditions.} In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we terminate iterative learning based on three conditions: (i) when $ \Rule $ contains all $ k $ clauses , (ii) the training data $ (\mathbf{X},\mathbf{y}) $ is empty (that is, no sample remains uncovered), and (iii) no new sample is covered by the partial classifier.
+\paragraph{Terminating conditions.} In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we terminate iterative learning based on three conditions: (i) when $ \Rule $ contains all $ k $ clauses , (ii) the training data $ \dataset $ is empty (that is, no sample remains uncovered), and (iii) no new sample is covered by the partial classifier.
 Since the first two conditions are trivial, we elaborate on the third condition.
-When clause $ C_i $ cannot cover any new sample from the training dataset $ (\mathbf{X},\mathbf{y}) $, the next iteration will result in the same clause $ C_i $ because the training data remains the same.
+When clause $ C_i $ cannot cover any new sample from the training dataset $ \dataset $, the next iteration will result in the same clause $ C_i $ because the training data remains the same.
 In this case, we do not include clause $ C_i $ to classifier $ \Rule $ because of zero coverage.
@@
 For learning DNF classifiers, we leverage De Morgan's law where complementing a CNF formula results in a DNF formula.
-To learn a DNF classifier, say $ \Rule'(\mathbf{x}) $, we can trivially show that $ y = \Rule'(\mathbf{x}) \leftrightarrow \neg (y = \neg \Rule'(\mathbf{x})) $ for the feature vector $ \mathbf{x} $.
-Here $ \neg \Rule'(\mathbf{x}) $ is a CNF formula, by definition.
-Thus, we learn a DNF classifier by first complementing the class-label $ \mathbf{y} $ to $ \neg \mathbf{y}$ in the training dataset, learning a CNF classifier on $ (\mathbf{X}, \neg \mathbf{y}) $ and finally complementing the learned classifier to DNF.
+To learn a DNF classifier, say $ \Rule'(\nonsensitive) $, we can trivially show that $ Y = \Rule'(\nonsensitive) \leftrightarrow \neg (Y = \neg \Rule'(\nonsensitive)) $ for the feature vector $ \nonsensitive $.
+Here $ \neg \Rule'(\nonsensitive) $ is a CNF formula, by definition.
+Thus, we learn a DNF classifier by first complementing the class-label $ y $ to $ \neg y $ for each sample in the training dataset $ \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $, learning a CNF classifier on $ \{(\mathbf{x}^{(i)}, \neg y^{(i)})\}_{i=1}^n $ and finally complementing the learned classifier to DNF.
 For example, the CNF classifier  ``(Male $ \vee $ Age $ < 50 $) $ \wedge $ (Education $ = $ Graduate $ \vee $ Income  $ \ge 1500 $)'' 	is complemented to a DNF classifier as 	 ``(not Male $ \wedge $ Age $ \ge 50 $) $ \vee $ (Education $ \ne $ Graduate $ \wedge $ Income $ \le 1500 $)''.
@@
 To learn a DNF classifier incrementally, such as through mini-batch and iterative learning, we adopt the following procedure.
-For learning a DNF classifier using mini-batch learning, we first learn a CNF classifier on dataset $ (\mathbf{X}, \neg \mathbf{y}) $ and complement the classifier to a DNF classifier at the end of mini-batch learning.
+For learning a DNF classifier using mini-batch learning, we first learn a CNF classifier on dataset $ \{(\mathbf{x}^{(i)}, \neg y^{(i)})\}_{i=1}^n $ and complement the classifier to a DNF classifier at the end of mini-batch learning.
 To learn a DNF classifier in the iterative approach, we  learn a single clause of the DNF classifier in each iteration, remove covered samples, and continue till no training sample remains.
@@
 In {\imli}, we apply an iterative learning approach for efficiently learning a  decision list.
-A decision list  $ \Rule_L $ is a list of pairs $ (C_1, v_1), \dots, (C_k, v_k) $, where we propose to learn one pair in each iteration.
+A decision list  $ \Rule_L $ is a list of pairs $ (C_1, V_1), \dots, (C_k, V_k) $, where we propose to learn one pair in each iteration.
 We note that the clause $ C_i $ is a conjunction of literals\textemdash equivalently, a single clause DNF formula.
 Hence, our task is to deploy {\imli} to efficiently learn a single clause DNF formula $ C_i $.
-In particular, we opt to learn this clause for the majority class, say $ v_i $, in the training dataset by setting the majority samples as class $ 1 $ and all other samples as class $ 0 $.
+In particular, we opt to learn this clause for the majority class, say $ V_i $, in the training dataset by setting the majority samples as class $ 1 $ and all other samples as class $ 0 $.
 As a result, even if the MaxSAT-based learning, presented in this chapter, is targeted for binary classification, we can learn a multi-class decision list in {\imli}.
@@
 In Algorithm~\ref{interpretability_imli_algo:iterative_decision_lists_learning}, we present the iterative algorithm for learning decision lists.
-In each iteration, the algorithm learns a pair $ (C_i, v_i) $, separates the training set based on the coverage of $ (C_i, v_i) $ and conquers the remaining samples recursively.
-The coverage of $ (C_i, v_i) $ is the set of samples that satisfies clause $ C_i $.
-Finally, we add a default rule $ (C_k, v_\mathsf{default}) $ to $ \Rule_L $ where $ C_k \triangleq  $ true denoting that the clause is satisfied by all samples.
-We select the default class $ v_\mathsf{default} $ in the following order: (i) if any class(s) is not in the predicted classes $ \{v_i\}_{i=1}^{k-1} $ of the decision list,
- $ v_{\mathsf{default}} $ is the majority class among missing classes, and (ii)  $ v_{\mathsf{default}} $ is the majority class of the original training set $ (\mathbf{X}, \mathbf{y}) $, otherwise.
+In each iteration, the algorithm learns a pair $ (C_i, V_i) $, separates the training set based on the coverage of $ (C_i, V_i) $ and conquers the remaining samples recursively.
+The coverage of $ (C_i, V_i) $ is the set of samples that satisfies clause $ C_i $.
+Finally, we add a default rule $ (C_k, V_\mathsf{default}) $ to $ \Rule_L $ where $ C_k \triangleq  $ true denoting that the clause is satisfied by all samples.
+We select the default class $ V_\mathsf{default} $ in the following order: (i) if any class(s) is not in the predicted classes $ \{V_i\}_{i=1}^{k-1} $ of the decision list,
+ $ V_{\mathsf{default}} $ is the majority class among missing classes, and (ii)  $ V_{\mathsf{default}} $ is the majority class of the original training set $ \dataset $, otherwise.
 
@@
 	\begin{algorithmic}[1]
-		\Procedure{DecisionListLearning}{$ \mathbf{X},\mathbf{y},\lambda, k $}
+		\Procedure{DecisionListLearning}{$ \dataset, \lambda, k $}
 		\State $ \mathcal{R}_L \leftarrow \{\}$
-		\For{$ i \leftarrow 1, \dots, k-1 $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
-		\State $ v_i \leftarrow \mathsf{MajorityClass(\mathbf{y})} $ \Comment{$ v_i $ specifies the target class}
-		\State $ C_i \leftarrow \mathsf{MiniBatchDNFLearning}( \mathbf{X},\mathbf{y},\lambda, 1, v_i) $ \Comment{Ref. Section~\ref{interpretability_imli_sec:dnf_learning}}
+		\For{$ i \leftarrow 1, \dots, k-1 $ and $ \dataset \ne \emptyset $}
+		\State $ V_i \leftarrow \textsc{MajorityClass}(\{y^{(i)}\}_{i=1}^{|\dataset|}) $ \Comment{$ V_i $ specifies the target class given class label}
+		\State $ C_i \leftarrow \textsc{MiniBatchDNFLearning}( \dataset, \lambda, 1, V_i) $ \Comment{Ref. Section~\ref{interpretability_imli_sec:dnf_learning}}
 
 %		\label{interpretability_imli_algo:iterative_CNF_learning_incremental_learning}
-		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{Coverage}(\mathbf{X}, C_i) $
+		\State $ \dataset' \leftarrow \textsc{Coverage}(\dataset, C_i) $
 %		\label{interpretability_imli_algo:iterative_CNF_learning_coverage}
-		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} 
+		\If{$ \dataset' = \emptyset $} 
 		\State break
 		\EndIf
-		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L \cup \{(C_i, v_i)\} $
-		\State $ \mathbf{X},\mathbf{y}  \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $ 
+		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L \cup \{(C_i, V_i)\} $
+		\State $ \dataset  \leftarrow \dataset  \setminus \dataset' $ 
 		\EndFor
-		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L  \cup \{(\text{true}, v_{\mathsf{default}} )\}$	\Comment{Default rule}
+		\State $ \mathcal{R}_L \leftarrow \mathcal{R}_L  \cup \{(\text{true}, V_{\mathsf{default}} )\}$	\Comment{Default rule}
 		\State \Return $ \Rule_L $
@@
 We now describe an iterative procedure for learning decision sets.
-A decision set comprises of an individual clause-class  pair $ (C_i, v_i) $ where $ C_i $ denotes a single clause DNF formula similar to decision lists.
+A decision set comprises of an individual clause-class  pair $ (C_i, V_i) $ where $ C_i $ denotes a single clause DNF formula similar to decision lists.
 In a decision set, a sample can satisfy multiple clauses simultaneously, which is attributed as an \textit{overlapping between clauses}~\cite{lakkaraju2016interpretable}.
-Concretely, the overlap between two clauses $ C_i $ and $ C_j $ with $ i \ne j $ is the set of samples $ \{\mathbf{X}_l | \mathbf{X}_l \models C_i \wedge \mathbf{X}_l \models C_j\} $ satisfying both clauses.
+Concretely, the overlap between two clauses $ C_i $ and $ C_j $ with $ i \ne j $ is the set of samples $ \{\mathbf{x} | \mathbf{x} \models C_i \wedge \mathbf{x} \models C_j\} $ satisfying both clauses.
 One additional objective in learning a decision set is to minimize the overlap between clauses, as studied in~\cite{lakkaraju2016interpretable}.
@@
 	\begin{algorithmic}[1]
-		\Procedure{DecisionSetsLearning}{$ \mathbf{X},\mathbf{y},\lambda, k$}
+		\Procedure{DecisionSetsLearning}{$\dataset,\lambda, k$}
 		\State $ \mathcal{R}_S=\{\} $
-		\State $ \mathbf{X}^\text{cc} = \{\}$  \Comment{Contains correctly covered samples}
-		\For{$ i \leftarrow 1, \dots, k-1 $ and $ (\mathbf{X},\mathbf{y}) \ne \emptyset $}
-		\State $ v_i \leftarrow \mathsf{MajorityClass(\mathbf{X},\mathbf{y})} $
-		\State $ \mathbf{X}^\text{w} \leftarrow \mathbf{X} \cup \mathbf{X}^\text{cc} $ \Comment{Correctly covered samples are included}
-		\State $ \mathbf{y}^\text{w} \leftarrow  \mathbf{y} \cup  \{\neg v_i\}^{|\mathbf{X}^\text{cc}|}$ \Comment{$ \mathbf{X}^\text{cc} $ have complemented class $ \neg v_i $ to minimize overlap}
-		\State $ C_i \leftarrow \mathsf{MiniBatchDNFLearning}( \mathbf{X}^\text{w},\mathbf{y}^\text{w},\lambda, 1, v_i) $ 
-		\State $ \mathbf{X}',\mathbf{y}' \leftarrow \mathsf{CorrectCoverage}(\mathbf{X}, \mathbf{y}, C_i, v_i) $
-		\If{$ (\mathbf{X}',\mathbf{y}') = \emptyset $} 
+		\State $ \dataset_{cc} = \{\}$  \Comment{Contains correctly covered samples}
+		\For{$ i \leftarrow 1, \dots, k-1 $ and $ \dataset \ne \emptyset $}
+		\State $ V_i \leftarrow \textsc{MajorityClass}(\{y^{(i)}\}_{i=1}^{|\dataset|}) $
+		\State $ \dataset_{w} = \dataset  \cup \{(\mathbf{x}, \neg V_i) | (\mathbf{x}, y) \in \dataset_{cc}\} $ \Comment{Appending covered samples with complemented class }
+%		\State $ \mathbf{X}^\text{w} \leftarrow \mathbf{X} \cup \mathbf{X}^\text{cc} $ \Comment{Correctly covered samples are included}
+%		\State $ \mathbf{y}^\text{w} \leftarrow  \mathbf{y} \cup  \{\neg V_i\}^{|\mathbf{X}^\text{cc}|}$ \Comment{$ \mathbf{X}^\text{cc} $ have complemented class $ \neg V_i $ to minimize overlap}
+		\State $ C_i \leftarrow \textsc{MiniBatchDNFLearning}( \dataset_w,\lambda, 1, V_i) $ 
+		\State $ \dataset' \leftarrow \textsc{CorrectCoverage}(\dataset, C_i, V_i) $
+		\If{$ (\dataset' = \emptyset $} 
 		\State break
 		\EndIf
-		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S \cup \{(C_i, v_i)\} $
-		\State $ \mathbf{X},\mathbf{y} \leftarrow (\mathbf{X},\mathbf{y})  \setminus (\mathbf{X}',\mathbf{y}') $
-		\State $ \mathbf{X}^\text{cc} = \mathbf{X}^\text{cc} \cup \mathbf{X}'$ 
+		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S \cup \{(C_i, V_i)\} $
+		\State $ \dataset \leftarrow \dataset  \setminus \dataset' $
+		\State $ \dataset_{cc} = \dataset_{cc} \cup \dataset' $
+%		\State $ \mathbf{X}^\text{cc} = \mathbf{X}^\text{cc} \cup \mathbf{X}'$ 
 		\EndFor
-		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S  \cup \{(\text{true}, v_{\mathsf{default}} )\}$
+		\State $ \mathcal{R}_S \leftarrow \mathcal{R}_S  \cup \{(\text{true}, V_{\mathsf{default}} )\}$
 		\State \Return $ \Rule_S $
@@
 The iterative algorithm is a modification of separate-and-conquer algorithm  by additionally focusing on minimizing overlaps in a decision set.
-Given a training data $ (\mathbf{X}, \mathbf{y}) $, a  regularization parameter $ \lambda $, and the number of clauses $ k $, the core idea of Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning} is to learn a pair $ (C_i,v_i) $ in each iteration, separate covered samples from $ (\mathbf{X}, \mathbf{y}) $, and conquer remaining samples recursively.
+Given a training data $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $, a  regularization parameter $ \lambda $, and the number of clauses $ k $, the core idea of Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning} is to learn a pair $ (C_i,V_i) $ in each iteration, separate covered samples from $ \dataset $, and conquer remaining samples recursively.
 In contrast to learning decision lists, we have following modifications in Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}.
@@
 	\item The first modification is with respect to the definition of coverage for decision sets.
-Unlike decision lists, we separate samples that are \textit{correctly covered} by $ (C_i, v_i) $ in each iteration.
-Given  $ (\mathbf{X}, \mathbf{y}) $, the correctly covered samples of $ (C_i, v_i) $ is a set $  \{(\mathbf{X}_l, \mathbf{y}_l)| \mathbf{X}_l \models C_i \wedge \mathbf{y}_l = v_i \} \subseteq (\mathbf{X}, \mathbf{y}) $ of samples that satisfy $ C_i $ and have matching class-label as $ v_i $.
+Unlike decision lists, we separate samples that are \textit{correctly covered} by $ (C_i, V_i) $ in each iteration.
+Given  $ (\mathbf{X}, \mathbf{y}) $, the correctly covered samples of $ (C_i, V_i) $ is a dataset $ \dataset_{cc}  =  \{(\mathbf{x}^{(l)}, y^{(l)}) | \mathbf{x}^{(l)} \models C_i \wedge y^{(l)} = V_i \}_{l=1}^{|\dataset|} \subseteq \dataset $ of samples that satisfy $ C_i $ and have matching class-label as $ V_i $.
 	\item The second modification is related to the  training dataset considered in each iteration.
-Let $ (\mathbf{X},\mathbf{y}) $ denote the remaining training set in the current iteration and $ v_i $ be the majority class in  $ (\mathbf{X},\mathbf{y}) $.
-Hence, $ v_i $ is the target class in the current iteration.
-In Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}, we learn $ C_i $ on a sample set $ (\mathbf{X}^\text{w},\mathbf{y}^\text{w}) $, where $ \mathbf{X}^\text{w} \triangleq \mathbf{X} \cup \mathbf{X}^\text{cc} $ comprises of both remaining and already covered samples.
-However, the vector of class labels $ \mathbf{y}^\text{w} \triangleq  \mathbf{y} \cup  \{\neg v_i\}^{|\mathbf{X}^\text{cc}|} $  comprises of remaining class-label vector $ \mathbf{y} $ and complemented class label vector $ \{\neg v_i\}^{|\mathbf{X}^\text{cc}|} $ associated with feature-matrix $ \mathbf{X}^\text{cc} $.
-By explicitly labeling covered samples as class $ \neg v_i $, the new clause $ C_i $ learns to falsify already covered samples.
+Let $ \dataset $ denote \emph{the remaining training dataset} in the current iteration and $ V_i $ be the majority class in  $ \dataset $.
+Hence, $ V_i $ is the target class in the current iteration.
+Also, let $ \dataset_{cc} $ denotes the set of samples \emph{correctly covered in all previous iterations}.
+In Algorithm~\ref{interpretability_imli_algo:iterative_decision_sets_learning}, we learn the current rule $ C_i $ on the working dataset $ \dataset_{w} = \dataset  \cup \{(\mathbf{x}, \neg V_i) | (\mathbf{x}, y) \in \dataset_{cc}\} $.
+Thus, $ \dataset_{w} $ is constructed by joining the remaining training samples with correctly covered samples in $ \dataset_{cc} $ with complemented class label to the target class $ V_i $.
+By explicitly labeling covered samples as class $ \neg V_i $, the new clause $ C_i $ learns to falsify already covered samples.
 This heuristic allows us to minimize the overlap of $ C_i $ compared to previously learned clauses $ \{C_j\}_{j=1}^{i-1} $.

### 2022-06-15 chapters/interpretability/imli/baseline.tex
notations updated

@@
 
-We consider two types of Boolean variables: (i) \emph{feature} variables $ \bool $ corresponding to input features and (ii) \emph{error} variables $ \eta $ corresponding to the classification error of samples.
+We consider two types of Boolean variables: (i) \emph{feature} variables $ \bool $ corresponding to input features and (ii) \emph{error} variables $ \noise $ corresponding to the classification error of samples.
 We define a Boolean variable $ \bool^i_j $ that becomes true if feature $ X_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ \bool^i_j $ is assigned false, otherwise.
-Moreover, we define an error variable $ \eta_l $ to attribute to whether the $ l^\text{th} $ sample $ (\mathbf{x}^{(l)}, y^{(l)}) $ is classified correctly or not.
-Specifically, $ \eta_l $ becomes true if $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified, and becomes false otherwise.
+Moreover, we define an error variable $ \noise_l $ to attribute to whether the $ l^\text{th} $ sample $ (\mathbf{x}^{(l)}, y^{(l)}) $ is classified correctly or not.
+Specifically, $ \noise_l $ becomes true if $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified, and becomes false otherwise.
 We next discuss the MaxSAT encoding to solve the classification problem.
@@
 \begin{itemize}
-	\item \textbf{Soft clauses for maximizing training accuracy:} For each training sample, {\imli} constructs a soft unit\footnote{A unit clause has a single literal.} clause $ \neg \eta_l $ to account for a penalty for misclassification.
+	\item \textbf{Soft clauses for maximizing training accuracy:} For each training sample, {\imli} constructs a soft unit\footnote{A unit clause has a single literal.} clause $ \neg \noise_l $ to account for a penalty for misclassification.
 Since the penalty for misclassification of a sample is $ 1 $  in Eq.~\eqref{interpretability_imli_eq:obj},  the weight of this soft clause is also $ 1 $.
@@
 		\label{interpretability_imli_soft_clause_error}
-		E_l:=\neg \eta_l;  \qquad \clauseweight{E_l}= 1
+		E_l:=\neg \noise_l;  \qquad \clauseweight{E_l}= 1
 	\end{equation}
@@
 			\label{interpretability_imli_hard_clause_rule_consistency}
-		H_l:= \neg \eta_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \clauseweight{H_l}=\infty
+		H_l:= \neg \noise_l \rightarrow \Big(y^{(l)}\leftrightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\Big); \qquad  \clauseweight{H_l}=\infty
 	\end{equation}
@@
 Since $ y^{(l)} \in \{0,1\} $ is a constant, the constraint to the right of the implication ``$ \rightarrow $'' is either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement.
-Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \eta_l $), either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement is true depending on the class-label of the sample.
-We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \eta_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
+Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \noise_l $), either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement is true depending on the class-label of the sample.
+We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \noise_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
 This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver.
@@
 The soft clauses $ E_l $ and $ S_j^i $ are unit clauses and hence, no translation is required for them.
-In the hard clause, when $ y^{(l)}=1 $, the simplification is $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $.
+In the hard clause, when $ y^{(l)}=1 $, the simplification is $ H_l:= \neg \noise_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $.
 In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF.
-In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
+In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \noise_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \noise_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
 Since $ \mathbf{x}^{(l)} \circ \boolset_{i} $ constitutes a disjunction of literals, we apply Tseytin transformation to encode  $ \neg (\mathbf{x}^{(l)} \circ \boolset_{i}) $ into CNF.
 More specifically, we introduce an auxiliary variable $ z_{l,i} $ corresponding to the clause $ \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $.
-Formally, we replace $ H_l := \neg \eta_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \eta_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})  $ for $ i=\{1,\dots,k\} $.
+Formally, we replace $ H_l := \neg \noise_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \noise_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})  $ for $ i=\{1,\dots,k\} $.
 Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF.
@@
 
-% Now,  the hard clause can be simplified as $ H_l:= \neg \eta_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ when $ y^{(l)}=1 $, otherwise  $ H_l:= \neg \eta_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \eta_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
+% Now,  the hard clause can be simplified as $ H_l:= \neg \noise_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $ when $ y^{(l)}=1 $, otherwise  $ H_l:= \neg \noise_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \noise_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
 
@@
 Then $X_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(\bool_{j}^{i}) = 1$.
-Additionally, $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified if and only if $ \sigma^*(\eta_l) = 1 $.
+Additionally, $ (\mathbf{x}^{(l)}, y^{(l)}) $ is misclassified if and only if $ \sigma^*(\noise_l) = 1 $.
 \end{construction}
@@
 For example, for feature $ X_2 $, introduced feature variables are $ \{\bool^1_3, \bool^2_3\} $ and for feature $ \neg X_2 $, introduced variables are $ \{\bool^1_4, \bool^2_4\} $.
-For four samples, error variables are $ \{\eta_1, \eta_2, \eta_3, \eta_4\} $.
+For four samples, error variables are $ \{\noise_1, \noise_2, \noise_3, \noise_4\} $.
 We next show the soft and hard clauses in the MaxSAT encoding
@@
 \begin{split}
-&E_1:= (\neg \eta_1); \quad
-E_2:= (\neg \eta_2); \quad E_3:= (\neg \eta_3); \quad
-E_4:= (\neg \eta_4)\\
+&E_1:= (\neg \noise_1); \quad
+E_2:= (\neg \noise_2); \quad E_3:= (\neg \noise_3); \quad
+E_4:= (\neg \noise_4)\\
 & S_1^1 := (\neg \bool_1^1);\quad 
@@
 S^2_4 := (\neg \bool^2_4);\quad\\ 
-& H_1:= (\neg \eta_1 \rightarrow ((\bool_2^1 \vee \bool_4^1)\wedge (\bool_2^2 \vee \bool_4^2)));\\
-& H_2:= (\neg \eta_2 \rightarrow (\neg(\bool_2^1 \vee \bool_3^1) \vee \neg(\bool_2^2 \vee \bool_3^2)));\\
-& H_3:= (\neg \eta_3 \rightarrow (\neg (\bool_1^1 \vee \bool_4^1) \vee \neg(\bool_1^2 \vee \bool_4^2)));\\
-& H_4:= (\neg \eta_4 \rightarrow ((\bool_1^1 \vee \bool_3^1) \wedge (\bool_1^2 \vee \bool_3^2)));\\
+& H_1:= (\neg \noise_1 \rightarrow ((\bool_2^1 \vee \bool_4^1)\wedge (\bool_2^2 \vee \bool_4^2)));\\
+& H_2:= (\neg \noise_2 \rightarrow (\neg(\bool_2^1 \vee \bool_3^1) \vee \neg(\bool_2^2 \vee \bool_3^2)));\\
+& H_3:= (\neg \noise_3 \rightarrow (\neg (\bool_1^1 \vee \bool_4^1) \vee \neg(\bool_1^2 \vee \bool_4^2)));\\
+& H_4:= (\neg \noise_4 \rightarrow ((\bool_1^1 \vee \bool_3^1) \wedge (\bool_1^2 \vee \bool_3^2)));\\
 \end{split}

### 2022-06-15 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
notations updated

@@
 {\crr} considers two types of  \emph{decision} variables: (i) feature variables and (ii) noise (classification error) variables.
-Since  feature $ x_j $ can be present or not present  in each of  the  $ \noclause $ clauses, {\crr} considers $ \noclause $   variables, each denoted by $ b^i_j  $  corresponding to   feature $ x_j $  to denote its participation  in the $ i $-th clause, i.e, $ b^i_j=\mathbb{1}[j\text{- th feature is selected in }i\text{-th clause}]  $.
-The $ q $-th sample in the training dataset, however, can be misclassified by $ \Rule $.
-Therefore, {\crr}  introduces  a noise variable $ \noise_q \in \{0,1\}  $  corresponding to the $ q $-th sample, so that the assignment of $ \noise_q  $ can be interpreted whether $ \X_q $ is misclassified by $ \Rule $ or not, i.e., $ \noise_q=\mathbb{1}[q\text{-th sample is misclassified}] $.
-Hence, the key idea of {{\crr}} for learning $ \Rule $ is to define an ILP query over $k \cdot m + n$  decision variables, denoted by $\{ b_{1}^{1}, b_{2}^{1}, \dots ,b_{m}^{1}, \dots ,b_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.
-In this context, we define $\B_i = \{b^i_{j} \mid j= 1,\dots,m\}$ as a \emph{vector of   feature variables} corresponding to the $ i $-th clause.
+Since  feature $ X_j $ can be present or not present  in each of  the  $ \noclause $ clauses, {\crr} considers $ \noclause $   variables, each denoted by $ \bool^i_j  $  corresponding to   feature $ X_j $  to denote its participation  in the $ i^\text{th} $ clause, i.e, $ \bool^i_j=\mathds{1}[j^\text{th}\text{ feature is selected in }i^\text{th}\text{ clause}]  $.
+The $ q^\text{th} $ sample in the training dataset, however, can be misclassified by $ \Rule $.
+Therefore, {\crr}  introduces  a noise variable $ \noise_q \in \{0,1\}  $  corresponding to the $ q^\text{th} $ sample, so that the assignment of $ \noise_q  $ can be interpreted whether $ (\mathbf{x}^{(q)}, y^{(q)}) $ is misclassified by $ \Rule $ or not, i.e., $ \noise_q=\mathds{1}[q^\text{th}\text{ sample is misclassified}] $.
+Hence, the key idea of {{\crr}} for learning $ \Rule $ is to define an ILP query over $k \cdot m + n$  decision variables, denoted by $\{ \bool_{1}^{1}, \bool_{2}^{1}, \dots ,\bool_{m}^{1}, \dots ,\bool_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.
+In this context, we define $\boolset_i = \{\bool^i_{j} \mid j= 1,\dots,m\}$ as a \emph{vector of   feature variables} corresponding to the $ i^\text{th} $ clause.
 
@@
 The objective function takes care of  both the rule-sparsity and the prediction accuracy of $ \Rule $.
-Since {\crr} prefers a sparser rule  with as few literals as possible, we construct the objective function by preferring  $ b^i_j $ to be  $ 0 $.
+Since {\crr} prefers a sparser rule  with as few literals as possible, we construct the objective function by preferring  $ \bool^i_j $ to be  $ 0 $.
 Moreover, to encourage $ \Rule $ to predict  the training samples accurately, we penalize the number of variables $ \noise_q $ that are different from $ 0 $.
 In this context, we utilize  the   parameter $ \lambda $ to trade off between sparsity and accuracy.
-Therefore,  the objective function of the ILP query $ Q $ is to minimize the normalized sum of all  noise variables $ \noise_q $  weighed  by the data-fidelity parameter $ \lambda $ and feature variables $ b^i_j $ weighed by $ 1-\lambda $ (Eq.~\ref{interpretability_crr_eq:ilp_obj}).
+Therefore,  the objective function of the ILP query $ Q $ is to minimize the normalized sum of all  noise variables $ \noise_q $  weighed  by the data-fidelity parameter $ \lambda $ and feature variables $ \bool^i_j $ weighed by $ 1-\lambda $ (Eq.~\ref{interpretability_crr_eq:ilp_obj}).
 
@@
 
-We first consider the case when  the $ q $-th  sample  has positive class label (Eq.~\ref{interpretability_crr_eq:ilp_pos}). 
-$ \X_q\circ \B_i \ge \card_l $ resembles the hard-OR operation of literals in a clause.
-We introduce   $ k $ \emph{auxiliary} $ 0 $-$ 1 $ variables $ \{\xi_{q,1}, \dots, \xi_{q,k}\} $  to check whether at least $ \card_c $ clauses are satisfied, i.e., $ \noise_{q,i}=\mathbb{1}[i\text{-th clause is dissatisfied for }q\text{-th sample}] $, which let us impose the operation of soft-AND over clauses.
-We then add a constraint to make sure that at most $ k-\eta_{c} $ clauses are allowed to be dissatisfied, otherwise the noise variable $ \noise_q $  is assigned to $ 1 $, i.e., the $ q $-th sample is detected as aw noise.
+We first consider the case when  the $ q^\text{th} $  sample  has positive class label (Eq.~\ref{interpretability_crr_eq:ilp_pos}). 
+$ \mathbf{x}^{(q)}\circ \boolset_i \ge \card_l $ resembles the hard-OR operation of literals in a clause.
+We introduce   $ k $ \emph{auxiliary} $ \{0, 1\} $ variables $ \{\noise_{q,1}, \dots, \noise_{q,k}\} $  to check whether at least $ \card_c $ clauses are satisfied, i.e., $ \noise_{q,i}=\mathds{1}[i^\text{th}\text{ clause is dissatisfied for }q^\text{th}\text{ sample}] $, which let us impose the operation of soft-AND over clauses.
+We then add a constraint to make sure that at most $ k-\eta_{c} $ clauses are allowed to be dissatisfied, otherwise the noise variable $ \noise_q $  is assigned to $ 1 $, i.e., the $ q^\text{th} $ sample is detected as aw noise.
 
-A negative labeled sample has to dissatisfy  more than $ k-\card_c $ clauses in $ \Rule $ so that the sample is predicted as $ 0 $, which is equivalent to satisfying  more than $ k-\card_c$ constraints $ \X_q\circ \B_i < \eta_l  $.
-As mentioned earlier, we  introduce $ 0 $-$ 1 $ variable $ \noise_{q,i} $ to specify if the constraint $ \X_q\circ \B_i < \eta_l  $ is dissatisfied or not and  restrict the count of dissatisfied clauses $ \sum_{i=1}^k\xi_{q,i} $ to be less than $ \card_c $.
+A negative labeled sample has to dissatisfy  more than $ k-\card_c $ clauses in $ \Rule $ so that the sample is predicted as $ 0 $, which is equivalent to satisfying  more than $ k-\card_c$ constraints $ \mathbf{x}^{(q)}\circ \boolset_i < \eta_l  $.
+As mentioned earlier, we  introduce $ \{0, 1\} $ variable $ \noise_{q,i} $ to specify if the constraint $ \mathbf{x}^{(q)}\circ \boolset_i < \eta_l  $ is dissatisfied or not and  restrict the count of dissatisfied clauses $ \sum_{i=1}^k\noise_{q,i} $ to be less than $ \card_c $.
 
@@
 % 	\begin{split}
- 	&\min   \frac{\lambda}{n} \sum_{q=1}^n \xi_q + \frac{1-\lambda}{k\cdot m}\sum_{i=1}^k\sum_{j=1}^m b_j^i \label{interpretability_crr_eq:ilp_obj} \\
+ 	&\min   \frac{\lambda}{n} \sum_{q=1}^n \noise_q + \frac{1-\lambda}{k\cdot m}\sum_{i=1}^k\sum_{j=1}^m \bool_j^i \label{interpretability_crr_eq:ilp_obj} \\
  	&\text{such that,} \notag \\
- 	&b_j^i \in \{0,1\},  i = 1,\dots,k,  j = 1,\dots,m \label{interpretability_crr_eq:ilp_range}\\
- 	& \xi_q \in \{0,1\},q = 1,\dots,n  \label{interpretability_crr_eq:ilp_range_2}\\
- 	&\text{if }  \forall q \in \{1,\dots,n\},\text{ if }  y_q=1, \label{interpretability_crr_eq:ilp_pos}\\
- 	&\qquad \X_q\circ \B_i+ m\xi_{q,i} \ge \eta_{l},  i = 1,\dots,k  \label{interpretability_crr_eq:ilp_ela_l_pos}\\
- 	& \qquad   k\xi_q + k-\eta_c \ge \sum_{i=1}^k \xi_{q,i} \notag \\ &\qquad \xi_{q,i} \in \{0,1\},  i = 1,\dots,k \notag\\
- 	&\text{if } \forall q \in \{1,\dots,n\}, y_q=0, \label{interpretability_crr_eq:ilp_neg}\\
- 	&\qquad \X_q\circ \B_i < \eta_{l} + m  \xi_{q,i},  i = 1,\dots,k \label{interpretability_crr_eq:ilp_ela_l_neg}\\
- 	&\qquad  k\xi_q  + \eta_{c} > \sum_{i=1}^k\xi_{q,i} \notag\\ &\qquad \xi_{q,i} \in \{0,1\},  i = 1,\dots,k \notag
+ 	&\bool_j^i \in \{0,1\},  i = 1,\dots,k,  j = 1,\dots,m \label{interpretability_crr_eq:ilp_range}\\
+ 	& \noise_q \in \{0,1\},q = 1,\dots,n  \label{interpretability_crr_eq:ilp_range_2}\\
+ 	&\text{if }  \forall q \in \{1,\dots,n\},\text{ if }  y^{(q)}=1, \label{interpretability_crr_eq:ilp_pos}\\
+ 	&\qquad \mathbf{x}^{(q)}\circ \boolset_i+ m\noise_{q,i} \ge \eta_{l},  i = 1,\dots,k  \label{interpretability_crr_eq:ilp_ela_l_pos}\\
+ 	& \qquad   k\noise_q + k-\eta_c \ge \sum_{i=1}^k \noise_{q,i} \notag \\ &\qquad \noise_{q,i} \in \{0,1\},  i = 1,\dots,k \notag\\
+ 	&\text{if } \forall q \in \{1,\dots,n\}, y^{(q)}=0, \label{interpretability_crr_eq:ilp_neg}\\
+ 	&\qquad \mathbf{x}^{(q)}\circ \boolset_i < \eta_{l} + m  \noise_{q,i},  i = 1,\dots,k \label{interpretability_crr_eq:ilp_ela_l_neg}\\
+ 	&\qquad  k\noise_q  + \eta_{c} > \sum_{i=1}^k\noise_{q,i} \notag\\ &\qquad \noise_{q,i} \in \{0,1\},  i = 1,\dots,k \notag
 % 	\end{split}
@@
 	\label{interpretability_crr_cons:rule}
-	Let $\sigma^* = \ILP(Q)$, then $x_j \in {\clause(\Rule,i)}$ iff $\sigma^*(b_{j}^i) = 1$.
+	Let $\sigma^* = \ILP(Q)$, then $X_j \in {\clause(\Rule,i)}$ iff $\sigma^*(\bool_{j}^i) = 1$.
 \end{construction}
@@
 
-A more generalized version to CNF rules would be to learn different thresholds on literals for different clauses, i.e., $ \eta_{l,i} $ for the $ i $-th clause.
+A more generalized version to CNF rules would be to learn different thresholds on literals for different clauses, i.e., $ \eta_{l,i} $ for the $ i^\text{th} $ clause.
 This facilitates to capture the complex decision boundaries, while still producing rule-based decisions.
@@
 %	& \eta_l \in [1,m] \\
-%	& \sum_{j=1}^m b_i^j \ge \eta_l, \forall i \in [1,k] \\ 
+%	& \sum_{j=1}^m \bool_i^j \ge \eta_l, \forall i \in [1,k] \\ 
 %\end{split}
@@
 Specifically, $ \eta_c $ and $ \eta_l $ (or $ \eta_{l,i} $) must be integers in the construction of relaxed-CNF rules, and $ \noise_{q,i} $ needs to be an integer to precisely calculate the noise variable $ \noise_q $.
-However,  we can relax feature variable $ b^i_j $ and noise variable $ \noise_q $ and solve the corresponding MILP  problem.
-To construct the MILP problem, we replace Eq.~\ref{interpretability_crr_eq:ilp_range} with $ 0 \le b^i_j \le 1 $ and Eq.~\ref{interpretability_crr_eq:ilp_range_2} with $ 0 \le \noise_q \le 1 $, and the rest of the constraints in $ Q $ remain the same.
-If non-zero $ b^i_j $ is found in the solution, we set it to $ 1 $ and then construct the rule according to Construction~\ref{interpretability_crr_cons:rule}.
+However,  we can relax feature variable $ \bool^i_j $ and noise variable $ \noise_q $ and solve the corresponding MILP  problem.
+To construct the MILP problem, we replace Eq.~\ref{interpretability_crr_eq:ilp_range} with $ 0 \le \bool^i_j \le 1 $ and Eq.~\ref{interpretability_crr_eq:ilp_range_2} with $ 0 \le \noise_q \le 1 $, and the rest of the constraints in $ Q $ remain the same.
+If non-zero $ \bool^i_j $ is found in the solution, we set it to $ 1 $ and then construct the rule according to Construction~\ref{interpretability_crr_cons:rule}.
 
@@
 Hence, we propose an incremental learning technique via a mini-batch training methodology \cite{GM19}.
-In this context, we define $ \Rule_p $ to be the rule learned for the $ p $-th batch.
-The key idea of incremental learning is to  divide the training dataset sequentially into a fixed number  of mini-batches $ \tau $, learn rule for each batch one after another such that  $ \Rule_p $  depends not only on the training samples in the $ p $-th batch but also regularizes itself with the  rule $ \Rule_{p-1} $ learned for the  $ (p-1) $-th batch.
-To this end, we use notation $ (\X^p,\mathbf{y}^p)$  to refer to the training data for the $ p $-th batch.
-We assume that $ \forall p \in \{1,\dots,\tau\}, |\X^p|=n_\tau $.
+In this context, we define $ \Rule_p $ to be the rule learned for the $ p^\text{th} $ batch.
+The key idea of incremental learning is to  divide the training dataset sequentially into a fixed number  of mini-batches $ \tau $, learn rule for each batch one after another such that  $ \Rule_p $  depends not only on the training samples in the $ p^\text{th} $ batch but also regularizes itself with the  rule $ \Rule_{p-1} $ learned for the  $ (p-1)^\text{th} $ batch.
+To this end, we use notation $ (\mathbf{x}^p,\mathbf{y}^p)$  to refer to the training data for the $ p^\text{th} $ batch.
+We assume that $ \forall p \in \{1,\dots,\tau\}, |\mathbf{x}^p|=n_\tau $.
 We propose  the following modifications in the ILP-based formulation for incrementally learning relaxed-CNF rules.
 
-Let $ (\X,\mathbf{y})$ be a training dataset with $ n $ samples and $ m $ binary features and $ \tau $ be the number of iterations in the learning process.
-In the $ p $-th iteration, we randomly sample a mini-batch $ (\X_p,\mathbf{y}_p)$ with equal number $ n_p $ of  samples and $ n_p \le n, p=1,\dots, \tau $.
-Note that all mini-batches have the same  binary features $ \mathbf{x} =\{x_1, \dots, x_m\} $ as in the  training set and thus share the same feature variables $ b_j^i $ in the MILP query.
-Therefore, we devise an objective function that prefers to keep the  assignment of $ b_j^i $ learned in the $ (p-1) $-th iteration while minimizing the prediction error on  the current mini-batch $ (\X_p,\mathbf{y}_p)$.
-To this end, each $ b_j^i $ is assigned $ 0 $ initially in the learning process.
-This technique  enables us to update the learned rule in terms of the update in  assignment of feature variables $ b_j^i $ over mini-batches.
+Let $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $ be a training dataset with $ n $ samples and $ m $ binary features and $ \tau $ be the number of iterations in the learning process.
+In the $ p^\text{th} $ iteration, we randomly construct a mini-batch from $ \dataset $ with equal number $ n_p  $ of  samples and $ n_p \ll n, p=1,\dots, \tau $.
+Note that all mini-batches have the same  binary features $ \nonsensitive =\{X_1, \dots, X_m\} $ as in the  training set and thus share the same feature variables $ \bool_j^i $ in the MILP query.
+Therefore, we devise an objective function that prefers to keep the  assignment of $ \bool_j^i $ learned in the $ (p-1)^\text{th} $ iteration while minimizing the prediction error on  the current mini-batch.
+To this end, each $ \bool_j^i $ is assigned $ 0 $ initially in the learning process.
+This technique  enables us to update the learned rule in terms of the update in  assignment of feature variables $ \bool_j^i $ over mini-batches.
 
-Let  $ Q_p $ be the MILP query for the $ p $-th mini-batch for learning the $ k $-clause relaxed-CNF rule $ \Rule_p $. Thus, $ \Rule_0 $ is an empty rule.
-We consider an indicator function $ I(\cdot): b^i_j \rightarrow \{1,-1\} $, that takes a feature variable $ b^i_j $ as input and outputs $ -1 $  if $ b^i_j $ is assigned $ 1 $ in the solution of $ Q_{p-1} $ (i.e., feature $ x_j $ is selected in the $ i $-th clause of   $ \Rule_{p-1} $), otherwise outputs 1.
+Let  $ Q_p $ be the MILP query for the $ p^\text{th} $ mini-batch for learning the $ k $-clause relaxed-CNF rule $ \Rule_p $. Thus, $ \Rule_0 $ is an empty rule.
+We consider an indicator function $ I(\cdot): \bool^i_j \rightarrow \{1,-1\} $, that takes a feature variable $ \bool^i_j $ as input and outputs $ -1 $  if $ \bool^i_j $ is assigned $ 1 $ in the solution of $ Q_{p-1} $ (i.e., feature $ X_j $ is selected in the $ i^\text{th} $ clause of   $ \Rule_{p-1} $), otherwise outputs 1.
 
@@
 \[
-I(b^i_j)=
+I(\bool^i_j)=
 \begin{cases}
--1 \quad \text{if }  x_j \in \clause(\Rule_{p-1},i)
+-1 \quad \text{if }  X_j \in \clause(\Rule_{p-1},i)
 \\
@@
 
-We now discuss the modified objective function where we multiply $ I(b^i_j) $ with $ b^i_j $ differently  from the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj}.
+We now discuss the modified objective function where we multiply $ I(\bool^i_j) $ with $ \bool^i_j $ differently  from the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj}.
 
@@
 	\begin{split}
-	&\min  \frac{\lambda}{n} \sum_{q=1}^{n_p} \xi_q  +  \frac{1-\lambda}{k\cdot m} \sum_{i=1}^k\sum_{j=1}^m b_j^i \cdot I(b^i_j) \\
+	&\min  \frac{\lambda}{n} \sum_{q=1}^{n_p} \noise_q  +  \frac{1-\lambda}{k\cdot m} \sum_{i=1}^k\sum_{j=1}^m \bool_j^i \cdot I(\bool^i_j) \\
 	\end{split}
@@
 %\vspace{-10pt}
-In the objective function, the first term penalizes the prediction error of samples in the current mini-batch and the second term penalizes when $ b_j^i $ is assigned differently than its previous  assignment.
+In the objective function, the first term penalizes the prediction error of samples in the current mini-batch and the second term penalizes when $ \bool_j^i $ is assigned differently than its previous  assignment.
 Note that the total prediction error is normalized by dividing by $ n $, which is the size of the full training dataset.
@@
 Since our problem formulation requires input instances to have binary features, datasets with categorical and continuous features  require a preprocessing stage.
-Initially, for all continuous features, we apply entropy-based discretization~\cite{fayyad1993multi} to infer the most appropriate number of categories/intervals by recursively splitting the domain of each continuous feature to minimize the class-entropy of the given dataset.\footnote{A simple quantile-based discretization also works, but it requires an extra parameter (i.e., the number of quantiles).}  For example, let $ x_c \in [a,b] $ be a continuous feature, and  entropy-based discretization splits the domain $ [a,b] $ into three intervals with two split points $ \{a',b'\} $, where $ a<a'<b'<b $.
-Therefore, the result intervals are $ { x_c < a'} $, $ {a' \le x_c < b'} $, and $ { x_c  \ge b'} $.
+Initially, for all continuous features, we apply entropy-based discretization~\cite{fayyad1993multi} to infer the most appropriate number of categories/intervals by recursively splitting the domain of each continuous feature to minimize the class-entropy of the given dataset.\footnote{A simple quantile-based discretization also works, but it requires an extra parameter (i.e., the number of quantiles).}  For example, let $ X_c \in [a,b] $ be a continuous feature, and  entropy-based discretization splits the domain $ [a,b] $ into three intervals with two split points $ \{a',b'\} $, where $ a<a'<b'<b $.
+Therefore, the result intervals are $ { X_c < a'} $, $ {a' \le X_c < b'} $, and $ { X_c  \ge b'} $.
 

### 2022-06-15 chapters/interpretability/relaxed-cnf/problem_formulation.tex
notations updated

@@
 \label{interpretability_crr_sec:problem}
-We now present the formal definition of the problem.
+
+Given 
+\begin{enumerate}
+	\item a dataset $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ of $ n $ samples, where feature vector $ \mathbf{x}^{(i)} \in \{0, 1\}^m $ contains $ m $ features and class label $ y^{(i)} \in \{0,1\} $,
+	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, 
+	\item an integer threshold on literals $ \card_{l} \in  \{0,\dots,m\} $ indicating  the  minimum number of literals required to be true  to satisfy a clause,
+	\item an  integer threshold on clauses $ \card_{c} \in  \{0,\dots,k\} $ indicating the minimum number of clauses required to be  true  to satisfy a formula, and
+	\item a data-fidelity  parameter $ \lambda \in \mathbb{R}^+ $,
+\end{enumerate}
+we  learn  a classification rule $ \Rule $ expressed as a $ k $ clause relaxed-CNF formula separating samples of class $ 1 $ from class $ 0 $.
+
+%We now present the formal definition of the problem.
 Given  (i) an instance space   $ (\mathbf{X}, \mathbf{y})$, where $ \mathbf{X} \in \{0,1\}^{n\times m} $ is the feature matrix with $ m $ binary features and $ n $ samples,\footnote{$\mathbf{X}$ contains both the features and their complements as columns as in~\cite{MV2013}.} and $ \mathbf{y} \in \{0,1\}^n $ is the class label vector, (ii) {a positive integer $ \noclause $ indicating the number of clauses in the desired rule}, (iii)  an integer threshold on literals $ \card_{l} \in  \{0,\dots,m\} $ indicating  the  minimum number of literals required to be true  to satisfy a clause, (iv)  an  integer threshold on clauses $ \card_{c} \in  \{0,\dots,k\} $ indicating the minimum number of clauses required to be  true  to satisfy a formula, and  (v) the data-fidelity  parameter $ \lambda \in [0,1]$, we  learn  a classification rule $ \Rule $, that is expressed as a $ k $ clause relaxed-CNF formula.
@@
 To this end, we seek to minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $.
-In particular, suppose $ \Rule $ classifies all samples correctly, i.e., $ \forall i, y_i=\Rule(\mathbf{X}_i) $.
+In particular, suppose $ \Rule $ classifies all samples correctly, i.e., $ \forall i, y^{(i)}=\Rule(\mathbf{x}^{(i)}) $.
 Among all the rules that classify all samples correctly,  we choose the sparsest  such $ \Rule $:
@@
 \[
-\min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y_i=\mathcal{R}(\mathbf{X}_i)
+\min\limits_{\mathcal{R}} |\mathcal{R}|\text{ such that }\forall i, y^{(i)}=\Rule(\mathbf{x}^{(i)})
 \]
@@
 Hence, we need to balance rule-sparsity with prediction error.
-Let $ \mathcal{E}_\mathcal{R} $   be  the set of samples which are misclassified  by $ \mathcal{R} $, i.e., $\mathcal{E}_\mathcal{R}=\{\mathbf{X}_i | y_i \ne \mathcal{R}(\mathbf{X}_i) \}$.
+Let $ \mathcal{E}_{\dataset} $   be  the set of samples which are misclassified  by $ \mathcal{R} $ on the dataset $ \dataset $, i.e., $ \mathcal{E}_\dataset = \{(\mathbf{x}^{(i)},y^{(i)}) | y^{(i)} \ne \mathcal{R}(\mathbf{x}^{(i)}) \} $.
 Hence we aim to find $ \mathcal{R} $ as follows:\footnote{In our formulation, it is  straightforward to add class-conditional weights  (e.g.
@@
 \[
-\min\limits_{\mathcal{R}} \;  \frac{\lambda}{n}|\mathcal{E}_\mathcal{R}| + \frac{1-\lambda}{k\cdot m} |\mathcal{R}|  \text{ such that }\forall \mathbf{X}_i \in \mathcal{E}_\mathcal{R} ,y_i \ne \mathcal{R}(\mathbf{X}_i) .
+\min\limits_{\mathcal{R}} \;  \frac{\lambda}{n}|\mathcal{E}_{\dataset}| + \frac{1-\lambda}{k\cdot m} |\mathcal{R}|  
 \]

### 2022-07-15 chapters/abstract.tex
considering all minor comments

@@
 	The significant success of machine learning in past decades has witnessed a host of applications of algorithmic decision-making in different safety-critical domains.
-The high-stake predictions of machine learning in medical, law, education, transportation, etc.\ have far-reaching consequences on the end-users.
+The high-stake predictions of machine learning in medical, law, education, transportation and so on have far-reaching consequences on the end-users.
 Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
@@
 As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
-To this end, we propose an incremental learning framework, called {\imli}, which extends the classification to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
-Built on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning a more expressible classification rule by relaxing logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.
+To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
+Building on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning a more expressible classification rule by relaxing logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.
 
 	Fairness in machine learning centers on quantifying and mitigating the bias/unfairness of machine learning.
-In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of certifying whether the classifier achieves the desired level of fairness given the distribution of features.
-We also extend {\justicia} by considering feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness.
+In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of verifying whether the classifier achieves the desired level of fairness given the distribution of features.
+We also extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness.
 Finally, we propose a framework for fairness influence functions (FIF) with an aim of quantifying the influence of features as their contribution on the bias of a classifier.

### 2022-07-15 chapters/background.tex
considering all minor comments

@@
 \chapter{Preliminaries}
-
+\label{chapter:preliminaries}
 We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
@@
 Let $\propformula$ be a Boolean formula defined over a set of Boolean variables $\boolset = \{\bool_1,\bool_2,\dots ,\bool_m \}$.
-A literal $ \literal $ is a variable $ \bool $ or its complement $\neg \bool$, and a clause $ \propclause $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals in the literature, we use clause for both disjunction and conjunction.}  $\propformula$ is in Conjunctive Normal Form (CNF) if $\propformula \triangleq  \bigwedge_i \propclause_i$ is a conjunction of clauses where each clause $\propclause_i \triangleq  \bigvee_j \literal_j $  is a disjunction of literals.
+A literal $ \literal $ is a variable $ \bool $ or its complement $\neg \bool$, and a clause $ \propclause $ is a disjunction ($ \vee $) or a conjunction ($ \wedge $) of literals.\footnote{While \emph{term} is reserved to denote disjunction of literals, we use clause for both disjunction and conjunction.}  $\propformula$ is in Conjunctive Normal Form (CNF) if $\propformula \triangleq  \bigwedge_i \propclause_i$ is a conjunction of clauses where each clause $\propclause_i \triangleq  \bigvee_j \literal_j $  is a disjunction of literals.
 In contrast, $ \propformula $ is in Disjunctive Normal Form (DNF) if $\propformula \triangleq  \bigvee_i \propclause_i$ is a disjunction of clauses where each clause $\propclause_i \triangleq  \bigwedge_j \literal_j $ is a conjunction of literals.
@@
 Motivated by Boolean cardinality constraints, we propose relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$.
-We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{v \in \propclause_i} \mathds{1}[\sigma^* \models v] \geq \card_l$.
-Informally,  $\sigma^*$ satisfies a clause $(\propclause_i, \card_l)$ if  at least $\card_l$ literals in $\propclause_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(\propformula,\card_c, \card_l)$ if at least $\card_c$  clauses out of all $\{(\propclause_i,\card_l)\}$ clauses are true.
+We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{V \in \propclause_i} \mathds{1}[\sigma^* \models V] \geq \card_l$.
+Informally,  $\sigma^*$ satisfies a clause $(\propclause_i, \card_l)$ if  at least $\card_l$ literals in $\propclause_i$ are set to true by $\sigma^*$ and $\sigma^*$ satisfies $(\propformula,\card_c, \card_l)$ if at least $\card_c$  clauses out of all $\{(\propclause_i,\card_l)\}_{i=1}^k$ clauses are true.
 
@@
 	\label{interpretability_crr_thm:succinctness}	
-	Let $ (\propclause,\card) $ be a clause in relaxed-CNF where  $ C $ has $ m  $ literals and $ \card \in \{1,\dots,m\} $ is the threshold on literals.
+	Let $ (\propclause,\card) $ be a clause in a relaxed-CNF where  $ C $ has $ m  $ literals and $ \card \in \{1,\dots,m\} $ is the threshold on literals.
 An equivalent compact encoding of $ (\propclause,\card) $  into a CNF formula $ \propformula= \bigwedge_i \propclause_i $ requires $ {m \choose{m-\card+1}} $  clauses where each clause is distinct and has $ m-\card+1 $ literals of $ (\propclause,\card) $.
-Therefore,  the total number of literals in $ \propformula $ is $ (m-\card+1){m\choose{m-\card+1}} = m $ when $ \card=1 $, otherwise $ (m-\card+1){m\choose{m-\card+1}} > m $.
+Therefore,  the total number of literals in $ \propformula $ is $ (m-\card+1){m\choose{m-\card+1}} = m $ (equal succinctness) when $ \card=1 $, otherwise $ (m-\card+1){m\choose{m-\card+1}} > m $ (exponential succinctness).
 \end{theorem}
@@
 Let $ \boolset \triangleq \{\bool_i\}_{i=1}^{|\boolset|}$ be a set of Boolean variables and $ \varweight_i \in \mathbb{Z} $ be the weight of $ \bool_i $.
-Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $\boolset$.
+Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $ \mathbf{b} $.
 Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
-The \textit{counting} version of the subset-sum problem counts all $ \boolset $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
+The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
 In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic.
@@
 %We define two rules  $ \R_1 $ and $ \R_2 $ to be  equivalent  if $ \forall i,\R_1(\X_i)=\R_2(\X_i) $.
-In a rule-based classifier, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{x} $ such that, if $ \Rule $ becomes true for an input (or an assignment), the prediction $ \widehat{Y} = 1 $ and $ \widehat{Y} = 0 $, otherwise.
+In a rule-based classifier such as a CNF formula, we view $ \mathcal{R} $ as a propositional formula defined over Boolean features $ \mathbf{X} $ such that, if $ \Rule $ becomes true for an input, the prediction $ \widehat{Y} = 1 $ and $ \widehat{Y} = 0 $, otherwise.
 The goal is to design $\Rule$ not only to approximate the training data, but also to generalize to unseen samples arising from the same distribution.
@@
 
-Therefore, path-specific cause fairness constrains that the prediction $ \widehat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \widehat{Y} $ only through mediator features $ \mediator $.
+Therefore, path-specific causal fairness constrains that the prediction $ \widehat{Y} $ is not directly dependent of sensitive features $ \sensitive $ while $ \sensitive $ may indirectly affects $ \widehat{Y} $ only through mediator features $ \mediator $.
 Hence, path-specific causal fairness is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
@@
 Let `male' be the most favored group by the classifier, for instance.
-Path-specific causal fairness ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
+Path-specific causal fairness ensures that a female candidate should be given a job offer with similar probability as a male candidate.
 She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.
@@
 In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.
-In fairness verification with correlated features, we leverage an access to a Bayesian network on features $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
+In fairness verification with correlated features, we leverage an access to a Bayesian network on sensitive and non-sensitive features $ \nonsensitive \cup \sensitive $ that represents the joint distribution on them.
 A Bayesian network is denoted by a pair $ (\graph, \theta)$, where $ \graph \triangleq (\graphvertices, \graphedges) $ is a DAG (Directed Acyclic Graph), and $\theta$ is a set of parameters encoding the conditional probabilities induced by the joint distribution under investigation.
@@
 \emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}.
-To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
+To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
+Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
 

### 2022-07-15 chapters/fairness/justicia/framework.tex
revision on fvgm
considering all minor comments

@@
 Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness metric $ f(\alg, \mathcal{D}) $, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.
-Additionally, if a fairness threshold $ \epsilon \in [0,1] $ is provided, {\justicia} verifies whether the classifier is $ \epsilon $-fair by comparing $ f $ with $ \epsilon $.
+Additionally, if a fairness threshold $ \epsilon \in [0,1] $ is provided, {\justicia} verifies whether the classifier is $ \epsilon $-fair by comparing $ f $ with $ \epsilon $ (refer to Chapter~\ref{chapter:preliminaries}).
 In {\justicia}, we focus on classifiers that can be represented as a CNF formula defined over a set of Boolean variables.
@@
 	For $ \Phi_{\neg A} $, the solution $ \Pr[\Phi_{\neg A}] = 0.43 $ is similarly derived from an SSAT solver. 
-	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of the sensitive feature $A$, both age groups demonstrate equal probability of positive prediction as the sensitive feature is not explicitly present in the classifier.
+	Therefore, if $\Pr[F], \Pr[I], \Pr[J]$ are computed independently of the sensitive feature $A$, both age groups achieve an equal probability of positive prediction as the sensitive feature is not explicitly present in the classifier.
 
@@
 		\label{fairness_justicia_algo:justicia_enum_begin}
-		\State $ \phi_{\widehat{Y}} := \mathsf{CNF}(\widehat{Y} = 1) $
+		\State $ \phi_{\widehat{Y}} := \textsc{CNF}(\widehat{Y} = 1) $
 		%\State $ p_{i} = \mathsf{CalculateProb}(X_i), \forall X_i \in X $
@@
 		\label{fairness_justicia_algo:justicia_learn_begin}
-		\State $ \phi_{\widehat{Y}} := \mathsf{CNF}(\widehat{Y}  = 1) $
+		\State $ \phi_{\widehat{Y}} := \textsc{CNF}(\widehat{Y}  = 1) $
 		\State $ p_{i} \leftarrow \Pr[X_i=1], \forall X_i \in \nonsensitive $
@@
 In most practical problems, there can be exponentially many compound sensitive groups due to the combination of categorical sensitive features.
-As a result, the enumeration approach may suffer from scalability issues due to an exponential number of calls to the SSAT solver.
+As a result, the enumeration approach may suffer from scalability issues due to the exponential number of calls to the SSAT solver.
 To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier.
 As discussed above, these two probabilities are sufficient to measure different group and causal fairness metrics.
-
-\todo{End of revision}
 
@@
 
-Solving an UR-SSAT formula raises several practical issues and thus, there is an unavailability of a UR-SSAT solver.
+Solving an UR-SSAT formula raises several practical issues and thus, there is an unavailability of an UR-SSAT solver.
 To resolve this problem, we leverage the \textit{duality} between UR-SSAT  and ER-SSAT formulas, where we solve an UR-SSAT formula on the CNF $ \phi $ using the solution of an ER-SSAT formula on the complemented CNF $ \neg \phi $~\cite{littman2001stochastic}.
@@
 
+\begin{comment}
 \subsection{Theoretical Analysis: Error Bounds}\label{fairness_justicia_sec:theory}
@@
 \fi
+
+\end{comment}

### 2022-07-15 chapters/fairness/justicia/introduction.tex
considering all minor comments

@@
 Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60).
-Typically, incomes of individuals increase as their ages increase while their fitness deteriorates.
-We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example}.
+Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}).
+We assume that the relation of income and fitness depends on ages as per the Normal distributions in Figure~\ref{fairness_justicia_fig:fair_example_distribution}.
 Now, if we train a decision tree~\cite{narodytska2018learning} to decide the eligibility of an individual to get a health insurance given three features: fitness, income and age, we observe that the `optimal' decision tree (ref.
-Figure~\ref{fairness_justicia_fig:fair_example}) does not predict based on the sensitive feature age.
-However, a fairness verifier would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively.
+Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age.
+However, a fairness verifier would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}).
 This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
@@
 			\end{tikzpicture}
-	}}\hfil
+	}\label{fairness_justicia_fig:fair_example_dependency}}\hfil
 	\subfloat[Age-dependent distributions of non-sensitive features]{
 		\includegraphics[scale=0.45]{figures/fairness/fif/sanity_distribution}
-	}
+	\label{fairness_justicia_fig:fair_example_distribution}}
 
@@
 			\end{tikzpicture}}
-	}\hfil
+	\label{fairness_justicia_fig:fair_example_dt}}\hfil
 	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
@@
 		}
-	}
+	\label{fairness_justicia_fig:fair_example_probability}}
 	\caption[A decision tree classifier on sensitive and non-sensitive features]{A trained decision tree to learn the eligibility for health insurance using age-dependent fitness and income indicators.
@@
 Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
-Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth label~\cite{hardt2016equality}.
-Sufficiency metrics\unsure{Check definition}, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
+Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}.
+Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
 
@@
 Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same.
-\red{Sufficiency is satisfied if the eligibility is independent of their age given their features are the same}\unsure{check}.
+Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
 Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}.
@@
 In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
-SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
+SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
 Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.
@@
 \begin{itemize}
-	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group fairness and causal fairness metrics for different datasets and classifiers.
+	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group and causal fairness metrics for different datasets and classifiers.
 	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
@@
 	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
-	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
+%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
 \end{itemize}

### 2022-07-15 chapters/fairness/justicia/practical.tex
considering all minor comments

@@
 \label{sec:practical-setting}
-We now relax the assumptions of {\justicia} on access to Boolean classifiers and Boolean features, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous features.
+We now relax the assumptions of {\justicia} on an access to Boolean classifiers and Boolean features, and extend {\justicia} to verify fairness metrics for more practical settings of decision trees, linear classifiers, and continuous features.
 

### 2022-07-15 chapters/interpretability/imli/baseline.tex
considering all minor comments

@@
 	\item \textbf{Hard clauses for encoding constraints:} In a MaxSAT problem, constraints that must be satisfied are encoded as hard clauses.
-In {\imli}, we have a constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa.
+In {\imli}, we have a learning constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa.
 Let  $\boolset_i= \{\bool^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $.
@@
 Therefore, the hard clause enforces that if the sample is correctly classified (using $ \neg \noise_l $), either  $ \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}\ $ or its complement is true depending on the class-label of the sample.
-We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \noise_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
+We highlight that  the single-implication ``$ \rightarrow $'' in the hard clause $ H_l $ acts as a double-implication ``$ \leftrightarrow $''  due to the soft clause $ E_l $.
+Because, according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \noise_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
+This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver.
+
+%	We highlight that according to the definition of ``$ \rightarrow $'', the left constraint $ \neg \noise_l $ can  be false while the right constraint of 	``$ \rightarrow $'' is true.
 This, however, incurs unnecessarily dissatisfying the soft clause $ E_l $, which is a sub-optimal solution and hence this solution is not returned by the MaxSAT solver.
@@
 In the hard clause, when $ y^{(l)}=1 $, the simplification is $ H_l:= \neg \noise_l \rightarrow \bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}  $.
-In this case, we apply the equivalence rule in propositional logic $ (a \rightarrow b) \equiv (\neg a \vee b) $  to encode $ H_l $ into  CNF.
+In this case, we apply the equivalence rule in propositional logic $ (A \rightarrow B) \equiv (\neg A \vee B) $  to encode $ H_l $ into  CNF.
 In contrast, when $ y_i=0 $, we simplify the hard clause as $ H_l:= \neg \noise_l \rightarrow \neg (\bigwedge_{i=1}^k {\mathbf{x}^{(l)}} \circ {\boolset_{i}}) \Rightarrow  \neg \noise_l \rightarrow \bigvee_{i=1}^k \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})$.
@@
 Formally, we replace $ H_l := \neg \noise_l \rightarrow  \bigvee_{i=1}^k \neg ({\mathbf{x}^{(l)}} \circ {\boolset_{i}}) $ with   $ \bigwedge_{i=0}^k H_{l,i} $,  where $ H_{l,0}:= (\neg \noise_l \rightarrow  \bigvee_{i=1}^k  z_{l,i}) $  and  $  H_{l,i}:= z_{l,i} \rightarrow  \neg({\mathbf{x}^{(l)}} \circ {\boolset_{i}})  $ for $ i=\{1,\dots,k\} $.
-Finally, we  apply the equivalence of $ (a \rightarrow b) \equiv (\neg a \vee b) $ on $ H_{l,i} $ to translate them into CNF.
+Finally, we  apply the equivalence of $ (A \rightarrow B) \equiv (\neg A \vee B) $ on $ H_{l,i} $ to translate them into CNF.
 For either value of $ y^{(l)} $, the weight of each translated hard clause in the CNF formula is $ \infty $.
@@
 	The MaxSAT query $ Q $ has $ k m+n $ unit clauses corresponding to the constraints $ E_l $ and $ S_j^i $.
-For a positive sample, the hard clause $ H_l $ is translated into $ k $  clauses.
+For each positive sample, the hard clause $ H_l $ is translated into $ k $  clauses.
 For each negative sample, the CNF translation requires at most $ k m+1 $ clauses.
@@
 \]
-In the preprocessing step, we negate the features in $ \dataset_\mathsf{orig} $ and add complemented features $ \{\neg X_1, \neg X_2\} $ in $ \dataset $.
+In the preprocessing step, we negate the features $ \{X_1, X_2\} $ in $ \dataset_\mathsf{orig} $ and add complemented features $ \{\neg X_1, \neg X_2\} $ in $ \dataset $.
 In the MaxSAT encoding, we define $ 8 $ feature variables ($ 4 $ features $ \times $ $ 2 $ clauses in the classifier) and $ 4 $ error variables.
@@
 In this example, we consider regularizer $ \lambda = 0.1 $, thereby setting the weight on accuracy as $ \clauseweight{E_i} = 1 $ and rule-sparsity weight as $ \clauseweight{S_j^i} = 0.1 $.
-Intuitively, the penalty for misclassifying a sample is $ 10 $ times more than the penalty for adding a feature in the classifier.
+Intuitively, the penalty for misclassifying a sample is $ 10 $ times than the penalty for adding a feature in the classifier.
 For this MaxSAT query, the optimal solution classifies all samples correctly by assigning four feature variables $ \{\bool^1_1, \bool^1_4, \bool^2_2, \bool^2_3\} $ to true.

### 2022-07-15 chapters/interpretability/imli/experiment.tex
considering all minor comments

@@
 
-\paragraph{Comparison with non-interpretable classifiers.} We compare {\imli} with state-of-the-art non-interpretable classifiers such as LR, SVM, kNN, and RF in terms of their median test accuracy in Table~\ref{interpretability_imli_table:non_interpretable_classifiers}.
+\paragraph{Comparison with non-interpretable classifiers.} We compare {\imli} with state of the art non-interpretable classifiers such as LR, SVM, kNN, and RF in terms of their median test accuracy in Table~\ref{interpretability_imli_table:non_interpretable_classifiers}.
 In the majority of the datasets, {\imli} achieves comparatively lower test accuracy than the best performing non-interpretable classifier.

### 2022-07-15 chapters/interpretability/imli/incremental_learning.tex
considering all minor comments

@@
 \label{interpretability_imli_sec:incremental_learning}
-The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to learn in a CNF formula.
+The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to be learned in a CNF formula.
 In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently.
@@
 Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity.
-We define the Hamming distance between two CNF classifiers $ \Rule $, $ \Rule' $ with the same number of clauses as follows.
+In the following, we define the Hamming distance between two CNF classifiers $ \Rule $, $ \Rule' $ with the same number of clauses as follows.
 \[
@@
 For example, consider $ \Rule = (X_1 \vee X_2) \wedge (\neg X_1) $ and $ \Rule' = (\neg X_1 \vee X_2) \wedge (\neg X_1) $.
-Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $ because in the first clause, the  literals in $ \{X_1, \neg X_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $.
+Then $ 	d_{H}(\Rule, \Rule') = 2 + 0 = 2 $, because in the first clause, the  literals in $ \{X_1, \neg X_1\} $ are absent in either formulas, and the second clause is identical for both $ \Rule $ and $ \Rule' $.
 In the following, we discuss a modified objective function for mini-batch learning.
@@
 Unlike controlling the rule-sparsity in the non-incremental approach in the earlier section, in Eq.~\eqref{interpretability_imli_eq:obj_incr}, $ \lambda $ \emph{controls the trade-off between classification errors and the syntactic differences between consecutive classifiers}.
-We next discuss how to encode the modified objective function as a MaxSAT query.
+Next, we discuss how to encode the modified objective function as a MaxSAT query.
 
@@
 $ S_j^i $ is either a unit clause $ \bool_j^i $ or $ \neg \bool_j^i $ depending on whether the associated feature $ X_j $ appears in the previous classifier $ \Rule' $ or not.
-In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ with no feature, and thus $ S_j^i := \neg \bool_j^i $ for the first batch.
+In this encoding, the Hamming distance of $ \Rule $ and $ \Rule' $ is minimized while attaining minimum classification errors on the current batch (using soft clause $ E_l $ in Eq.~\eqref{interpretability_imli_soft_clause_error}) To this end,  mini-batch learning starts with an empty CNF formula as $ \Rule' $ without any feature, and thus $ S_j^i := \neg \bool_j^i $ for the first batch.
 We next analyze the complexity of the MaxSAT encoding for mini-batch learning.
@@
 Let $ n'_{pos} = n' - n'_{neg} $ be the number of positive samples in the batch.
-Then, according to Proposition~\ref{interpretability_imli_prop:maxsat_clauses}, the number of clauses in the MaxSAT query is $ k m+n'+k n'_{pos}+(k m+1)n'_{neg} \approx \mathcal{O}(k m  n' ) $ when $ n'_{pos}= n'_{neg} =\frac{n'}{2} $.
+Then, according to Proposition~\ref{interpretability_imli_prop:maxsat_clauses}, the number of clauses in the MaxSAT query for each batch is $ k m+n'+k n'_{pos}+(k m+1)n'_{neg} \approx \mathcal{O}(k m  n' ) $ when $ n'_{pos}= n'_{neg} =\frac{n'}{2} $.
 \end{proposition}

### 2022-07-15 chapters/interpretability/imli/introduction.tex
considering all minor comments

@@
 Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.
-At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging.
-In our context of rule-based classifiers, we will use the \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
+In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
 
@@
 	\normalfont
-	We learn an interpretable classification rule in conjunctive normal form for predicting the classification of a tumor cell.
+	We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
 We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with three \emph{clauses} and nine Boolean \emph{literals}.
@@
 \paragraph{Contributions.} The primary contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
-For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF (Conjunctive Normal Form) formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.
+For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.
 In this chapter, {\imli} provides precise control to jointly optimize the accuracy and the size of classification rules.

### 2022-07-15 chapters/interpretability/imli/iterative_learning.tex
considering all minor comments

@@
 The coverage of a clause in a CNF formula is the set of samples that \emph{do not satisfy} the clause.
-The reason is that if a sample does not satisfy at least one clause in a CNF formula, the prediction of the sample by the full formula is class $ 0 $ because CNF is a conjunction of clauses.
+The reason is that if a sample does not satisfy at least one clause in a CNF formula, the prediction of the sample by the full formula is class $ 0 $, because CNF is a conjunction of clauses.
 As a result, considering covered samples in the next iteration does not change their prediction regardless of whatever clause we learn in later iterations.
@@
 
-\paragraph{Terminating conditions.} In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we terminate iterative learning based on three conditions: (i) when $ \Rule $ contains all $ k $ clauses , (ii) the training data $ \dataset $ is empty (that is, no sample remains uncovered), and (iii) no new sample is covered by the partial classifier.
+\paragraph{Terminating conditions.} In Algorithm~\ref{interpretability_imli_algo:iterative_CNF_learning}, we terminate iterative learning based on three conditions: (i) when $ \Rule $ contains all $ k $ clauses , (ii) the training data $ \dataset $ is empty (that is, no sample remains uncovered), and (iii) no new sample is covered by the current partial classifier.
 Since the first two conditions are trivial, we elaborate on the third condition.
@@
 Here $ \neg \Rule'(\nonsensitive) $ is a CNF formula, by definition.
-Thus, we learn a DNF classifier by first complementing the class-label $ y $ to $ \neg y $ for each sample in the training dataset $ \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $, learning a CNF classifier on $ \{(\mathbf{x}^{(i)}, \neg y^{(i)})\}_{i=1}^n $ and finally complementing the learned classifier to DNF.
+Thus, we learn a DNF classifier by first complementing the class-label $ y^{(i)} $ to $ \neg y^{(i)} $ for each sample in the training dataset $ \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $, learning a CNF classifier on $ \{(\mathbf{x}^{(i)}, \neg y^{(i)})\}_{i=1}^n $, and finally complementing the learned classifier to DNF.
 For example, the CNF classifier  ``(Male $ \vee $ Age $ < 50 $) $ \wedge $ (Education $ = $ Graduate $ \vee $ Income  $ \ge 1500 $)'' 	is complemented to a DNF classifier as 	 ``(not Male $ \wedge $ Age $ \ge 50 $) $ \vee $ (Education $ \ne $ Graduate $ \wedge $ Income $ \le 1500 $)''.
@@
 \begin{algorithm}
-	\caption{Iterative learning of decision lists}
+	\caption{Iterative Learning of Decision Lists}
 	\label{interpretability_imli_algo:iterative_decision_lists_learning}
@@
 \begin{algorithm}
-	\caption{Iterative learning of decision sets}
+	\caption{Iterative Learning of Decision Sets}
 	\label{interpretability_imli_algo:iterative_decision_sets_learning}
@@
 Unlike decision lists, we separate samples that are \textit{correctly covered} by $ (C_i, V_i) $ in each iteration.
-Given  $ (\mathbf{X}, \mathbf{y}) $, the correctly covered samples of $ (C_i, V_i) $ is a dataset $ \dataset_{cc}  =  \{(\mathbf{x}^{(l)}, y^{(l)}) | \mathbf{x}^{(l)} \models C_i \wedge y^{(l)} = V_i \}_{l=1}^{|\dataset|} \subseteq \dataset $ of samples that satisfy $ C_i $ and have matching class-label as $ V_i $.
+Given  a dataset $ \dataset = \{(\mathbf{x}^{(l)}, y^{(l)})\}$, the correctly covered samples of $ (C_i, V_i) $ is a dataset $ \dataset_{cc}  =  \{(\mathbf{x}^{(l)}, y^{(l)}) | \mathbf{x}^{(l)} \models C_i \wedge y^{(l)} = V_i \}_{l=1}^{|\dataset|} \subseteq \dataset $ of samples that satisfy $ C_i $ and have matching class-label as $ V_i $.
 	\item The second modification is related to the  training dataset considered in each iteration.
@@
 Thus, $ \dataset_{w} $ is constructed by joining the remaining training samples with correctly covered samples in $ \dataset_{cc} $ with complemented class label to the target class $ V_i $.
-By explicitly labeling covered samples as class $ \neg V_i $, the new clause $ C_i $ learns to falsify already covered samples.
+Thus, by explicitly labeling covered samples as class $ \neg V_i $, the new clause $ C_i $ learns to falsify already covered samples.
 This heuristic allows us to minimize the overlap of $ C_i $ compared to previously learned clauses $ \{C_j\}_{j=1}^{i-1} $.

### 2022-07-15 chapters/interpretability/relaxed-cnf/conclusion.tex
considering all minor comments

@@
 
-In this chapter, we proposed an efficient combinatorial framework called {\crr}, for learning relaxed-CNF  classification rules, that are  more expressive  than CNF/DNF rules.
+In this chapter, we proposed an efficient combinatorial framework, called {\crr}, for learning relaxed-CNF  classification rules.
+Relaxed-CNF rules are  more expressive  than CNF/DNF rules.
 {\crr}  uses a novel integration of mini-batch learning procedure with the MILP framework to learn sparse relaxed-CNF rules.

### 2022-07-15 chapters/interpretability/relaxed-cnf/experiment.tex
notations updated
considering all minor comments

@@
 \section{Experiments}\label{interpretability_crr_sec:experiment}
-We implement a prototype  of  the incremental version of $ \crr $\footnote{The source code is available at \url{https://github.com/meelgroup/mlic}} based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances.
+We implement a prototype  of  $ \crr $\footnote{The source code is available at \url{https://github.com/meelgroup/mlic}} based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances.
 The objective of our experimental evaluation is to answer the following questions:
@@
 Every cell in the last four columns contains the test accuracy in percentage (top value), rule size  (middle value), and  training time in seconds (bottom value).
-In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.}
+In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.
+Number in bold denotes the best result, such as maximum test accuracy, minimum rule-size, and minimum training time among competitive classifiers.}
 	\label{interpretability_crr_tab:rule_based_classifiers}
-	\setlength{\tabcolsep}{3pt} 
+%	\setlength{\tabcolsep}{3pt} 
 	\begin{center}
@@
 			\midrule
-			{ Heart}   & $  303 $  & $  31 $  & $   78.69  $    & $   72.13  $    & $   72.13  $    & $   77.69  $   \\[0.05cm] & & 
-			& $   7 $    & $   19 $    & $   13 $    & $   4.5  $   \\[0.05cm] & & 
-			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   1.8 \text{s}  $    & $   122.5 \text{s}  $   \\[.3cm] 
-			{ Ionosphere}   & $  351 $  & $  144 $  & $   88.65  $    & $   91.43  $    & $   89.29  $    & $   91.43  $   \\[0.05cm] & & 
-			& $   8 $    & $   4 $    & $   8.5  $    & $   20 $   \\[0.05cm] & & 
-			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   2.09 \text{s}  $    & $   5.59 \text{s}  $   \\[.3cm] 
-			{ WDBC}   & $  569 $  & $  88 $  & $   95.22  $    & $   95.65  $    & $   93.91  $    & $   94.69  $   \\[0.05cm] & & 
-			& $   7.5  $    & $   12 $    & $   7 $    & $   34.5  $   \\[0.05cm] & & 
-			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   1.38 \text{s}  $    & $   316.32 \text{s}  $   \\[.3cm] 
-			{ Magic}   & $  19020 $  & $  79 $  & $   84.04  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\[0.05cm] & & 
-			& $   115 $    & $   3 $    & $   24 $    & $   31 $   \\[0.05cm] & & 
-			& $   15.86 \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\[.3cm] 
-			{ Tom's HW}   & $  28179 $  & $  910 $  & $   97.4  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\[0.05cm] & & 
-			& $   36 $    &       & $   30 $    & $   4 $   \\[0.05cm] & & 
-			& $   42.73 \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\[.3cm] 
-			{ Credit}   & $  30000 $  & $  110 $  & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $   82.04  $   \\[0.05cm] & & 
-			& $   38.5  $    &       & $   10 $    & $   32 $   \\[0.05cm] & & 
-			& $   14.52 \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\[.3cm] 
-			{ Adult}   & $  32561 $  & $  144 $  & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   84.86  $   \\[0.05cm] & & 
-			& $   94 $    &       & $   23 $    & $   18 $   \\[0.05cm] & & 
-			& $   27.61 \text{s}  $    &       & $   11.91 \text{s}  $    & $   1016.36 \text{s}  $   \\[.3cm] 
-			{ Twitter}   & $  49999 $  & $  1511 $  & $   95.74  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\[0.05cm] & & 
-			& $   179.5  $    &       & $   57 $    & $   12 $   \\[0.05cm] & & 
-			& $   170.87 \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\[.3cm] 
-			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   84.57  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\[0.05cm] & & 
-			& $   195 $    &       & $   7 $    & $   2 $   \\[0.05cm] & & 
-			& $   121.02 \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\[.3cm] 
-			{ Skin}   & $  245057 $  & $  119 $  & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   98.92  $    & $   95.08  $   \\[0.05cm] & & 
-			& $   725 $    &       & $   201 $    & $   29 $   \\[0.05cm] & & 
-			& $   1313.19 \text{s}  $    &       & $   103.8 \text{s}  $    & $   825.6 \text{s}  $   \\[.3cm] 
+			{ Heart}   & $  303 $  & $  31 $  & $   \textbf{78.69}  $    & $   72.13  $    & $   72.13  $    & $   77.69  $   \\ & & 
+			& $   7 $    & $   19 $    & $   13 $    & $   \textbf{4.5}  $   \\ & & 
+			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   \textbf{1.8} \text{s}  $    & $   122.5 \text{s}  $   \\\\
+			{ Ionosphere}   & $  351 $  & $  144 $  & $   88.65  $    & $   \textbf{91.43}  $    & $   89.29  $    & $   \textbf{91.43}  $   \\ & & 
+			& $   8 $    & $   \textbf{4} $    & $   8.5  $    & $   20 $   \\ & & 
+			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   \textbf{2.09} \text{s}  $    & $   5.59 \text{s}  $   \\\\ 
+			{ WDBC}   & $  569 $  & $  88 $  & $   95.22  $    & $   \textbf{95.65}  $    & $   93.91  $    & $   94.69  $   \\ & & 
+			& $   7.5  $    & $   12 $    & $   \textbf{7} $    & $   34.5  $   \\ & & 
+			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   \textbf{1.38} \text{s}  $    & $   316.32 \text{s}  $   \\\\ 
+			{ Magic}   & $  19020 $  & $  79 $  & $   \textbf{84.04}  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\ & & 
+			& $   115 $    & $   \textbf{3} $    & $   24 $    & $   31 $   \\ & & 
+			& $   \textbf{15.86} \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\\\ 
+			{ Tom's HW}   & $  28179 $  & $  910 $  & $   \textbf{97.4}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\ & & 
+			& $   36 $    &       & $   30 $    & $   \textbf{4} $   \\ & & 
+			& $   \textbf{42.73} \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\\\ 
+			{ Credit}   & $  30000 $  & $  110 $  & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $  \textbf{ 82.04}  $   \\ & & 
+			& $   38.5  $    &       & $   \textbf{10} $    & $   32 $   \\ & & 
+			& $   \textbf{14.52} \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\\\ 
+			{ Adult}   & $  32561 $  & $  144 $  & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   \textbf{84.86}  $   \\ & & 
+			& $   94 $    &       & $   23 $    & $   \textbf{18} $   \\ & & 
+			& $   27.61 \text{s}  $    &       & $   \textbf{11.91} \text{s}  $    & $   1016.36 \text{s}  $   \\\\ 
+			{ Twitter}   & $  49999 $  & $  1511 $  & $   \textbf{95.74}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\ & & 
+			& $   179.5  $    &       & $   57 $    & $   \textbf{12} $   \\ & & 
+			& $   \textbf{170.87} \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\\\ 
+			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   \textbf{84.57}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\ & & 
+			& $   195 $    &       & $   7 $    & $   \textbf{2} $   \\ & & 
+			& $   \textbf{121.02} \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\\\ 
+			{ Skin}   & $  245057 $  & $  119 $  & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   \textbf{98.92}  $    & $   95.08  $   \\ & & 
+			& $   725 $    &       & $   201 $    & $   \textbf{29} $   \\ & & 
+			& $   1313.19 \text{s}  $    &       & $   \textbf{103.8} \text{s}  $    & $   825.6 \text{s}  $   \\\\ 
 			\bottomrule
@@
 Specifically for {\crr}, we choose the data-fidelity parameter $ \lambda \in \{0.5, 0.67, 0.84, 0.99\} $, the number of clauses $ k \in \{1,2,3\} $, the relative size of mini-batch $ \frac{n_p}{n} \in \{0.25, 0.50, 0.75\} $, and the number of iterations $ \tau \in \{2,4,8,16\} $.
-We learn the value of $ \card_c $ and $ \eta_{l} $ from the dataset as described in Eq.~\ref{interpretability_crr_sec:ilp_query}.
+We learn the value of $ \card_c $ and $ \eta_{l} $ from the dataset as described in Chapter.~\ref{interpretability_crr_sec:ilp_query}.
 In CPLEX, we set the  maximum solving time of the LP solver to $ 1000 $  seconds ($ \frac{ 1000 }{\tau} $ seconds for each iteration) and the remaining $ 500 $ seconds is allotted to construct the MLIP instances, parse the solutions and execute  other auxiliary tasks of the learning algorithm.
@@
 We then compare relaxed-CNF rules with DNF rules generated by BRS and find that relaxed-CNF rules outperform DNF rules with respect to prediction accuracy in several datasets.
-At this point,   BRS fails to scale for larger datasets as shown in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
+At this point,   BRS fails to scale on larger datasets as shown in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
 We finally compare relaxed-CNF rules with decision lists generated by RIPPER.
@@
 			{Dataset}  & LR & SVC & RF & $k$-NN & CRR\\\midrule
-			\multirow{1}{*}{ Heart}   & $   84.29  $    & $   83.33  $    & $   81.97  $    & $   78.69  $    & $   77.69  $   \\ 
-			\multirow{1}{*}{ Ionosphere}   & $   94.29  $    & $   91.43  $    & $   92.96  $    & $   91.43  $    & $   91.43  $   \\ 
-			\multirow{1}{*}{ WDBC}   & $   98.26  $    & $   96.46  $    & $   96.9  $    & $   95.61  $    & $   94.69  $   \\ 
-			\multirow{1}{*}{ Magic}   & $   85.15  $    & $   84.45  $    & $   85.3  $    & $   77.9  $    & $   81.31  $   \\ 
-			\multirow{1}{*}{ Tom's HW}   & $   97.62  $    & $   97.66  $    & $   97.52  $    & $   94.59  $    & $   97.34  $   \\ 
-			\multirow{1}{*}{ Credit}   & $   82.04  $    & $   82.12  $    & $   81.97  $    & $   80.5  $    & $   82.04  $   \\ 
-			\multirow{1}{*}{ Adult}   & $   87.24  $    & $   86.82  $    & $   86.84  $    & $   84.68  $    & $   84.86  $   \\ 
-			\multirow{1}{*}{ Twitter}   & $   96.28  $    & $   96.34  $    & $   96.37  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.16  $   \\ 
-			\multirow{1}{*}{ Weather-AUS}   & $   85.71  $     &  \multicolumn{1}{c}{\textemdash}      & $   85.63  $     &  \multicolumn{1}{c}{\textemdash}      & $   83.34  $   \\ 
-			\multirow{1}{*}{ Skin}   & $   97.21  $     &  \multicolumn{1}{c}{\textemdash}      & $   99.81  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.08  $   \\ 
+			\multirow{1}{*}{ Heart}   & $   \textbf{84.29}  $    & $   83.33  $    & $   81.97  $    & $   78.69  $    & $   77.69  $   \\ 
+			\multirow{1}{*}{ Ionosphere}   & $   \textbf{94.29}  $    & $   91.43  $    & $   92.96  $    & $   91.43  $    & $   91.43  $   \\ 
+			\multirow{1}{*}{ WDBC}   & $   \textbf{98.26}  $    & $   96.46  $    & $   96.90  $    & $   95.61  $    & $   94.69  $   \\ 
+			\multirow{1}{*}{ Magic}   & $   85.15  $    & $   84.45  $    & $   \textbf{85.30}  $    & $   77.9  $    & $   81.31  $   \\ 
+			\multirow{1}{*}{ Tom's HW}   & $   97.62  $    & $   \textbf{97.66}  $    & $   97.52  $    & $   94.59  $    & $   97.34  $   \\ 
+			\multirow{1}{*}{ Credit}   & $   82.04  $    & $   \textbf{82.12}  $    & $   81.97  $    & $   80.5  $    & $   82.04  $   \\ 
+			\multirow{1}{*}{ Adult}   & $   \textbf{87.24}  $    & $   86.82  $    & $   86.84  $    & $   84.68  $    & $   84.86  $   \\ 
+			\multirow{1}{*}{ Twitter}   & $   96.28  $    & $   96.34  $    & $   \textbf{96.37}  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.16  $   \\ 
+			\multirow{1}{*}{ Weather-AUS}   & $   \textbf{85.71}  $     &  \multicolumn{1}{c}{\textemdash}      & $   85.63  $     &  \multicolumn{1}{c}{\textemdash}      & $   83.34  $   \\ 
+			\multirow{1}{*}{ Skin}   & $   97.21  $     &  \multicolumn{1}{c}{\textemdash}      & $   \textbf{99.81}  $     &  \multicolumn{1}{c}{\textemdash}      & $   95.08  $   \\ 
 			\bottomrule
@@
 \begin{figure}[t]
-	%%\vspace{-25pt}
 	\centering
+
 	\subfloat
-	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_lambda.pdf}\label{interpretability_crr_fig:d}} \hfill
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_lambda.pdf}}
 	\subfloat
-	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_clause.pdf}\label{interpretability_crr_fig:f}} \hfill
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_lambda.pdf}}
 	\subfloat
-	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_subsamplesize.pdf}
-		\label{interpretability_crr_fig:a}} \hfill
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_lambda.pdf}} 
+	\\
+
 	\subfloat
-	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_iteration.pdf}\label{interpretability_crr_fig:g}} \hfill
-	\caption[Effect of hyper-parameters on accuracy in {\crr}]{Effect of model hyper-parameters on test accuracy  for  Heart disease dataset. } 
-	\label{interpretability_crr_fig:result_1}
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_test_accuracy_vary_lambda.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_rule_size_vary_lambda.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_time_vary_lambda.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_test_accuracy_vary_lambda.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_rule_size_vary_lambda.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_time_vary_lambda.pdf}} 
+	\\
+
+	\caption[Effect of data-fidelity $ \lambda $ in {\crr}]{Effect of data-fidelity $ \lambda $ on test accuracy, rule-size, and training time in {\crr}. } 
+	\label{interpretability_crr_fig:result_lambda}
 \end{figure}
 
-%%\vspace{-3pt}
+\begin{figure}[t]
+	\centering
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_clause.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_test_accuracy_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_rule_size_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_time_vary_clause.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_test_accuracy_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_rule_size_vary_clause.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_time_vary_clause.pdf}} 
+	\\
+
+	\caption[Effect of the number of clause $ k $ in {\crr}]{Effect of the number of clause $ k $ on test accuracy, rule-size, and training time in {\crr}. } 
+	\label{interpretability_crr_fig:result_clause}
+\end{figure}
+
+\begin{figure}[t]
+	\centering
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_subsamplesize.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_test_accuracy_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_rule_size_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_time_vary_subsamplesize.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_test_accuracy_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_rule_size_vary_subsamplesize.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_time_vary_subsamplesize.pdf}} 
+	\\
+
+	\caption[Effect of mini-batch size in {\crr}]{Effect of mini-batch size on test accuracy, rule-size, and training time in {\crr}. } 
+	\label{interpretability_crr_fig:result_minibatch}
+\end{figure}
+
+\begin{figure}[t]
+	\centering
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_test_accuracy_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_iteration.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_test_accuracy_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_rule_size_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/wdbc_time_vary_iteration.pdf}} 
+	\\
+
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_test_accuracy_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_rule_size_vary_iteration.pdf}}
+	\subfloat
+	{\includegraphics[width=0.27\textwidth]{figures/interpretability/relaxed-cnf/adult_time_vary_iteration.pdf}} 
+	\\
+
+	\caption[Effect of the number of iterations in {\crr}]{Effect of the number of iterations on test accuracy, rule-size, and training time in {\crr}. } 
+	\label{interpretability_crr_fig:result_iteration}
+\end{figure}
+
 	\subsubsection*{Varying Model Parameters: }
 	\label{interpretability_crr_sec:model_parameters}
-	In Figure~\ref{interpretability_crr_fig:result_1},~\ref{interpretability_crr_fig:result_2}, and~\ref{interpretability_crr_fig:result_3}, we demonstrate the effect of varying the hyper-parameters of {\crr}.
+	In Figure~\ref{interpretability_crr_fig:result_lambda},~\ref{interpretability_crr_fig:result_clause}, ,~\ref{interpretability_crr_fig:result_minibatch}, and~\ref{interpretability_crr_fig:result_iteration}, we demonstrate the effect of varying the hyper-parameters of {\crr}.
 To understand the effect of a single hyper-parameter, we fix the values of other hyper-parameters to a default choice where the default choice results in the most accurate rule.
 
-	%	%%\vspace{-3pt}
-	\textbf{Varying data-fidelity parameter ($ \lambda $):}  As we increase data-fidelity parameter $ \lambda $ in the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj} and Eq.~\ref{interpretability_crr_eq:ilp_inc}, more priority is given to the prediction accuracy than the sparsity of the rules.
+	\paragraph{Varying data-fidelity parameter ($ \lambda $):} In Figure~\ref{interpretability_crr_fig:result_lambda}, as we increase data-fidelity parameter $ \lambda $ in the objective function in Eq.~\ref{interpretability_crr_eq:ilp_obj} and Eq.~\ref{interpretability_crr_eq:ilp_inc}, more priority is given to the prediction accuracy than the sparsity of the rules.
 In most of the datasets, we similarly observe an increase in accuracy  and also an increase in the size of the rules when $ \lambda $ is higher.
@@
 
-	\begin{figure}[t]
-		%%\vspace{-25pt}
-		\centering
-		\subfloat
-		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_lambda.pdf}\label{interpretability_crr_fig:c}} \hfill
-		\subfloat
-		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_clause.pdf}\label{interpretability_crr_fig:h}} \hfill
-		\subfloat
-		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_subsamplesize.pdf}\label{interpretability_crr_fig:b}} \hfill 
-		\subfloat
-		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_rule_size_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
-		%%\vspace{-10pt}
-		\caption[Effect of hyper-parameters on rule-size in {\crr}]{Effect of model hyper-parameters on rule-size for  Heart disease dataset.}
-
-		%\vspace{-10pt}
-		\label{interpretability_crr_fig:result_2}
-	\end{figure}
-
-	{\textbf{Varying the number of clauses ($ k $): }}
-	%	In Figure~\ref{interpretability_crr_fig:clause(a)}, we show the effect of $ k $ on  test accuracy   for Parkinsons {dataset}. 
-	As we increase  $ k $, {\crr} allows the generated rules to capture the variance in the given dataset more effectively, that results in higher  accuracy in most of the datasets.
+	\paragraph{Varying the number of clauses ($ k $): }
+	In Figure~\ref{interpretability_crr_fig:result_clause}, as we increase  $ k $, {\crr} allows the generated rules to capture the variance in the given dataset more effectively, that results in higher  accuracy in most of the datasets.
 The  rule-size also increases as we learn more clauses.
@@
 
-	 \begin{figure}[btp]
-	 	%%\vspace{-25pt}
-	 	\centering
-	 	\subfloat
-	 	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_lambda.pdf}\label{interpretability_crr_fig:c}} \hfill
-	 	\subfloat
-	 	{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_clause.pdf}\label{interpretability_crr_fig:h}} \hfill
-	 		\subfloat
-	 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_subsamplesize.pdf}\label{interpretability_crr_fig:b}} \hfill 
-	 		\subfloat
-	 		{\includegraphics[width=0.23\textwidth]{figures/interpretability/relaxed-cnf/heart_time_vary_iteration.pdf}\label{interpretability_crr_fig:e}} \hfill
-	 	%	%%\vspace{-10pt}
-	 	\caption[Effect of hyper-parameters on training time in {\crr}]{Effect of model hyper-parameters on training  time for  Heart disease dataset.}
-
-	 	%\vspace{-10pt}
-	 	\label{interpretability_crr_fig:result_3}
-	 \end{figure}
-
-	\textbf{Relative size of the mini-batch:} We vary the relative size of the mini-batch $ \frac{n_p}{n} $ to observe its effect on the accuracy and the size of the rules.
+	\paragraph{Relative size of the mini-batch:} In Figure~\ref{interpretability_crr_fig:result_minibatch}, we vary the relative size of the mini-batch $ \frac{n_p}{n} $ to observe its effect on the accuracy and the size of the rules.
 In most datasets,  the accuracy increases when more samples are considered in the mini-batch, costing higher   training time.
@@
 	%	%%\vspace{-3pt}
-	\textbf{Varying the number of iterations ($ \tau $):}   
-	As we allow more iterations in the learning process, we find an increase in accuracy in most datasets.
+	\paragraph{Varying the number of iterations ($ \tau $):}   
+	In Figure~\ref{interpretability_crr_fig:result_iteration}, as we allow more iterations in the learning process, we find an increase in accuracy in most datasets.
 The training time also increases with $ \tau $ because {\crr} is required to solve in total $ \tau $ queries.

### 2022-07-15 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
considering all minor comments

@@
 \label{interpretability_crr_sec:variables}
-{\crr} considers two types of  \emph{decision} variables: (i) feature variables and (ii) noise (classification error) variables.
+{\crr} considers two types of  \emph{decision} variables: (i) feature variables and (ii) noise or classification error variables.
 Since  feature $ X_j $ can be present or not present  in each of  the  $ \noclause $ clauses, {\crr} considers $ \noclause $   variables, each denoted by $ \bool^i_j  $  corresponding to   feature $ X_j $  to denote its participation  in the $ i^\text{th} $ clause, i.e, $ \bool^i_j=\mathds{1}[j^\text{th}\text{ feature is selected in }i^\text{th}\text{ clause}]  $.
@@
 Therefore, {\crr}  introduces  a noise variable $ \noise_q \in \{0,1\}  $  corresponding to the $ q^\text{th} $ sample, so that the assignment of $ \noise_q  $ can be interpreted whether $ (\mathbf{x}^{(q)}, y^{(q)}) $ is misclassified by $ \Rule $ or not, i.e., $ \noise_q=\mathds{1}[q^\text{th}\text{ sample is misclassified}] $.
-Hence, the key idea of {{\crr}} for learning $ \Rule $ is to define an ILP query over $k \cdot m + n$  decision variables, denoted by $\{ \bool_{1}^{1}, \bool_{2}^{1}, \dots ,\bool_{m}^{1}, \dots ,\bool_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.
+Hence, the key idea of {{\crr}} for learning $ \Rule $ is to define an ILP (Integer Linear Program) query over $k \cdot m + n$  decision variables, denoted by $\{ \bool_{1}^{1}, \bool_{2}^{1}, \dots ,\bool_{m}^{1}, \dots ,\bool_{m}^{\noclause}, \noise_1, \dots ,\noise_n\}$.
 In this context, we define $\boolset_i = \{\bool^i_{j} \mid j= 1,\dots,m\}$ as a \emph{vector of   feature variables} corresponding to the $ i^\text{th} $ clause.
 
-\subsection{Construction of ILP Query}
+\subsection{Construction of the ILP Query}
 \label{interpretability_crr_sec:ilp_query}
-We first discuss the objective function of the  ILP query $ Q $ for learning  a $ k $-clause relaxed-CNF rule  $ \Rule $.
-The objective function takes care of  both the rule-sparsity and the prediction accuracy of $ \Rule $.
+In Eq.~\eqref{interpretability_crr_eq:ilp}, we discuss the ILP query $ Q $  for learning  a $ k $-clause relaxed-CNF rule  $ \Rule $.
+The objective function in Eq.~\eqref{interpretability_crr_eq:ilp_obj} takes care of  both the rule-sparsity and the prediction accuracy of $ \Rule $.
 Since {\crr} prefers a sparser rule  with as few literals as possible, we construct the objective function by preferring  $ \bool^i_j $ to be  $ 0 $.
@@
 In this context, we utilize  the   parameter $ \lambda $ to trade off between sparsity and accuracy.
-Therefore,  the objective function of the ILP query $ Q $ is to minimize the normalized sum of all  noise variables $ \noise_q $  weighed  by the data-fidelity parameter $ \lambda $ and feature variables $ \bool^i_j $ weighed by $ 1-\lambda $ (Eq.~\ref{interpretability_crr_eq:ilp_obj}).
+Therefore,  the objective function of the ILP query $ Q $ is to minimize the normalized sum of all  noise variables $ \noise_q $  weighed  by the data-fidelity parameter $ \lambda $ and feature variables $ \bool^i_j $ weighed by $ 1-\lambda $.
 
@@
 Initially, we define the range of the decision variables and   add constraints accordingly (Eq.~\ref{interpretability_crr_eq:ilp_range} and ~\ref{interpretability_crr_eq:ilp_range_2}).
-For each sample, at first, we  add constraints to mimic the behavior of hard-OR of literals in a clause, and then we add constraints to apply  soft-AND of clauses in a formula.
+For each sample, at first, we  add constraints to mimic the behavior of hard-OR of literals in a clause, and then we add constraints to apply  soft-AND of clauses in a formula (refer to Preliminaries in Chapter~\ref{chapter:preliminaries}).
 
@@
 
-In the following, we present the ILP query $ Q $.
+%In the following, we present the ILP query $ Q $.
 
@@
 	\label{interpretability_crr_cons:rule}
-	Let $\sigma^* = \ILP(Q)$, then $X_j \in {\clause(\Rule,i)}$ iff $\sigma^*(\bool_{j}^i) = 1$.
+	Let $\sigma^* = \ILP(Q)$, then $X_j \in {\clause(\Rule,i)}$ if and only if $\sigma^*(\bool_{j}^i) = 1$.
 \end{construction}
@@
 \textbf{Learning thresholds $ \eta_l $ and $ \eta_c $:}
-Given the training dataset $ (\mathbf{X}, \mathbf{y})$ and data-fidelity parameter $ \lambda $, one could learn the optimum value of the thresholds $ \card_{c} $ and $ \card_l $ of the desired rule $ \Rule $  by specifying their range as constraints in the ILP query $ Q $ in Eq.~\ref{interpretability_crr_eq:ilp}.
+Given the training dataset $ \dataset $ and data-fidelity parameter $ \lambda $, one could learn the optimum value of the thresholds $ \card_{c} $ and $ \card_l $ of the desired rule $ \Rule $  by specifying their range as constraints in the ILP query $ Q $ in Eq.~\ref{interpretability_crr_eq:ilp}.
 More precisely, we need to add two integer constraints  $ \card_c  \in \{0,\dots,k\}$ and $ \card_l \in \{0,\dots,m\} $ in the above query and then learn their values from the solution.
@@
 In Section~\ref{interpretability_crr_sec:ilp_query}, we  present an ILP formulation for learning relaxed-CNF rules and then discussed the relaxation to the integer constraints in the MILP formulation to make the approach computationally tractable.
-However, not all integer constraints in the formulation can be relaxed, as discussed  in Section~\ref{interpretability_crr_sec:relaxation}.
+However, each integer constraints in the formulation cannot be relaxed, as discussed  in Section~\ref{interpretability_crr_sec:relaxation}.
 Thus we require an improved learning technique to achieve scalability.
@@
 In incremental mini-batch learning, the learning process repeats for a fixed number of iterations.
-In each iteration, we randomly select an equal number of samples from the full training set and generate a mini-batch with less number of samples.
-We then construct a MILP query on the current mini-batch with a modified objective function.
+In each iteration, we randomly select an equal number of samples from the full training set and generate a mini-batch with samples.
+We then construct an MILP query on the current mini-batch with a modified objective function.
 This objective function simultaneously penalizes the prediction errors
@@
 Let $ \dataset = \{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n $ be a training dataset with $ n $ samples and $ m $ binary features and $ \tau $ be the number of iterations in the learning process.
-In the $ p^\text{th} $ iteration, we randomly construct a mini-batch from $ \dataset $ with equal number $ n_p  $ of  samples and $ n_p \ll n, p=1,\dots, \tau $.
+In the $ p^\text{th} $ iteration, we randomly construct a mini-batch from $ \dataset $ with equal number $ n_p  $ of  samples and $ n_p \ll n, $  for $ p=1,\dots, \tau $.
 Note that all mini-batches have the same  binary features $ \nonsensitive =\{X_1, \dots, X_m\} $ as in the  training set and thus share the same feature variables $ \bool_j^i $ in the MILP query.
@@
 To this end, each $ \bool_j^i $ is assigned $ 0 $ initially in the learning process.
-This technique  enables us to update the learned rule in terms of the update in  assignment of feature variables $ \bool_j^i $ over mini-batches.
+This technique  enables us to update the learned rule in terms of the update in the assignment of feature variables $ \bool_j^i $ over mini-batches.
 
-Let  $ Q_p $ be the MILP query for the $ p^\text{th} $ mini-batch for learning the $ k $-clause relaxed-CNF rule $ \Rule_p $. Thus, $ \Rule_0 $ is an empty rule.
+Let  $ Q_p $ be the MILP query for the $ p^\text{th} $ mini-batch for learning a $ k $-clause relaxed-CNF rule $ \Rule_p $. Thus, $ \Rule_0 $ is an empty rule.
 We consider an indicator function $ I(\cdot): \bool^i_j \rightarrow \{1,-1\} $, that takes a feature variable $ \bool^i_j $ as input and outputs $ -1 $  if $ \bool^i_j $ is assigned $ 1 $ in the solution of $ Q_{p-1} $ (i.e., feature $ X_j $ is selected in the $ i^\text{th} $ clause of   $ \Rule_{p-1} $), otherwise outputs 1.
@@
 Intuitively, if the size $ n_p $ of the mini-batch  is close to the size $ n $
-of the training set, more priority is given to the prediction accuracy and vice versa. The constraints in the  query $ Q_p $ are similar to the constraints in Eq.~\ref{interpretability_crr_eq:ilp}.  
+of the training set, more priority is given to the prediction accuracy on the current batch and vice versa.
+The constraints in the  query $ Q_p $ are similar to the constraints in Eq.~\ref{interpretability_crr_eq:ilp}.
 Finally, the prediction rule  $ \Rule $ is  $ \Rule_\tau $ that is learned for the last mini-batch.
@@
 \label{interpretability_crr_sec:other_rules}
-While {\crr} learns  classification rules in relaxed-CNF form, we can leverage this framework for learning classification rules in other logical forms, for example, CNF,  DNF, and decision sets.
-To learn a CNF rule one can set  $ \card_l=1 $ and $ \card_c=k $.
-Moreover,  to learn a DNF rule,  one would first complement the class label of all samples, then learn a CNF rule by setting the parameters as described and finally negate the learned rule.
-We note that the literature on decision sets also fits into our formulation, because decision sets can be represented by DNF formulas.
+While {\crr} learns  classification rules in relaxed-CNF form, we can leverage this framework for learning classification rules in other logical forms, for example, CNF and  DNF.
+To learn a CNF rule, we set  $ \card_l=1 $ and $ \card_c=k $, which converts a relaxed-CNF to a CNF formula.
+Moreover,  to learn a DNF rule,  we first complement the class label of all samples, learn a CNF rule by setting the parameters as described and finally negate the learned rule, which we have discussed elaborately in Chapter~\ref{chapter:imli}.
+%We note that the literature on decision sets also fits into our formulation, because decision sets can be represented by DNF formulas.
 As a future work, we are investigating whether related formulations could be used to address other learning paradigms such as decision lists.

### 2022-07-15 chapters/interpretability/relaxed-cnf/introduction.tex
considering all minor comments

@@
 In this context,  rule-based representations are considered interpretable in presenting the decision functions to users~\cite{LKCL2019,WR2015,WRDLKM2017}.
-A recent body of work has studied sparsity-inducing objectives for classification rules in CNF ({Conjunctive Normal Form}) or DNF ({Disjunctive Normal Form}) and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{GM2019,LKCL2019,MM18}.
+A recent body of work has studied sparsity-inducing objectives for classification rules in CNF or DNF and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{GM2019,LKCL2019,MM18}.
 Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.

### 2022-07-15 chapters/introduction.tex
considering all minor comments

--- 2022-06-03 chapters/introduction.tex
@@
 
-The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical, law, education, and transportation.
-In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users.
-With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability, fairness, robustness, and privacy in predictions.
+The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, education~\cite{luckin2018machine}, and transportation~\cite{peled2019model,zantalis2019review}.
+In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users~\cite{eshete2021making}.
+With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox}, and privacy~\cite{papernot2016towards} in predictions.
 In this thesis, we focus on the interpretability and fairness aspects of machine learning.
-The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness.
+The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
 
@@
 In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
-Consequently, fairness research centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
+Consequently, the research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
 In this thesis, we study fairness verification problems to assess the fairness of different classifiers under the lens of multiple fairness definitions and algorithms.
 In addition, we study the sources of the unfairness of classifiers by identifying  features attributing highly to bias.
-In both interpretable and fair machine learning, we prioritize improving both the \emph{scalability} and the \emph{accuracy} of solutions, either or both of which prior approaches fail to achieve.
+To summarize our thesis, in interpretable and fair machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which prior approaches fail to achieve.
 
@@
 We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}.
-Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic, which allow exceptions in the construction of a clause and also in the selection of clauses in a rule.
+Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule.
 Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and Boolean cardinality constraints in logic.
@@
 sensitive features and ignore correlations among non-sensitive features.
-As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.
+As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remain major challenges.
 
@@
 In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}.
-We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.
+%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.
 
@@
 Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
-Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for accurate and scalable fairness verification.
+Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
-To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
+To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
 Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
-We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias, which we study in detail in the following contribution of the thesis.
+We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
+%, which we study in detail in the following contribution of the thesis.
 
@@
 \section{Tools}
+	We develop following open-source tools for interpretable and fair machine learning.
+	\begin{itemize}
+		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
+		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification and FIF computation
+	\end{itemize}
 \section{Outline}
+We organize the thesis in the following way.
+We discuss preliminaries in Chapter~\ref{chapter:preliminaries}.
+In the first part of the thesis, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable classifiers.
+We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
+In the second part of this thesis, we discuss  fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
+We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}.
+Finally, we conclude our thesis in Chapter~\ref{}.
 

### 2022-07-17 chapters/conclusion.tex
thesis conclusion add

\chapter{Conclusion And Future Work} In the past decade, the advancement of machine learning has witnessed a host of  applications in different safety-critical domains. In such domains, the interpretability, fairness, robustness, and privacy etc.\ of classifiers are considered important in trustworthy and responsible AI (artificial intelligence). In this thesis, we focus on the interpretability and fairness aspects of machine learning, where we prioritize on improving the scalability and accuracy of the underlying problems. Relying on formal methods, our contributions are summarized as follows: (i) in interpretable machine learning, we design incremental learning technique for interpretable rule-based classifiers, (ii) in fairness in machine learning, we develop a formal fairness verification framework to verify multiple fairness definitions, algorithms, and classifiers; and we develop techniques to identify the sources of unfairness of classifiers by computing fairness influence functions.

To validate the efficiency of our methods, we develop open-source tools and conduct experiments on real-world interpretability and fairness datasets in machine learning. In interpretable machine learning, our framework {\imli} scales classification to million-size datasets while achieving competitive prediction accuracy and rule-size compared to existing rule-based classifiers. In our quest for learning more expressible interpretable classifiers, we develop another learning framework {\crr} based on incremental learning. In experiments, {\crr} learns more concise and accurate rule-based classifiers while demonstrating scalability owing to incremental learning.

In fairness in machine learning, our probabilistic fairness verifier {\justicia} and {\fvgm} demonstrates a superior performance in accuracy and scalability than the state of the art verifiers. Being a stochastic-SAT-based verifier, {\justicia} verifies fairness of compound sensitive groups of Boolean classifiers such as decision trees with higher scalability. {\fvgm}, in addition, considers correlated features and verifies fairness of linear classifiers with better scalability and accuracy than existing verifiers. Finally, given a classifier and a dataset, our framework {\fairXplainer} computes fairness influences of each subset of non-sensitive features as their contribution to the incurred bias of the classifier. In experiments, {\fairXplainer} estimates bias with significantly higher accuracy than the existing local explanation approach, while enabling computation of intersectional influences as a mean of better understanding the sources fo bias.

### 2022-07-18 chapters/conclusion.tex
preparing for a submission

@@
-\chapter{Conclusion And Future Work} In the past decade, the advancement of machine learning has witnessed a host of  applications in different safety-critical domains.
+\chapter{Conclusion And Future Work}
+\label{chapter:conclusion} 
+In the past decade, the advancement of machine learning has witnessed a host of  applications in different safety-critical domains.
 In such domains, the interpretability, fairness, robustness, and privacy etc.\ of classifiers are considered important in trustworthy and responsible AI (artificial intelligence).
 In this thesis, we focus on the interpretability and fairness aspects of machine learning, where we prioritize on improving the scalability and accuracy of the underlying problems.
-Relying on formal methods, our contributions are summarized as follows: (i) in interpretable machine learning, we design incremental learning technique for interpretable rule-based classifiers, (ii) in fairness in machine learning, we develop a formal fairness verification framework to verify multiple fairness definitions, algorithms, and classifiers; and we develop techniques to identify the sources of unfairness of classifiers by computing fairness influence functions.
+Based on formal methods, the contributions of our thesis are summarized as follows.
+(i) In interpretable machine learning, we design incremental learning technique for interpretable rule-based classifiers.
+(ii) In fairness in machine learning, we develop a formal fairness verification framework to verify multiple fairness definitions, algorithms, and classifiers; and we develop techniques to identify the sources of unfairness of classifiers by computing fairness influence functions.
 
@@
 In interpretable machine learning, our framework {\imli} scales classification to million-size datasets while achieving competitive prediction accuracy and rule-size compared to existing rule-based classifiers.
-In our quest for learning more expressible interpretable classifiers, we develop another learning framework {\crr} based on incremental learning.
-In experiments, {\crr} learns more concise and accurate rule-based classifiers while demonstrating scalability owing to incremental learning.
+In our quest for learning more expressible interpretable classifiers, we develop another learning framework {\crr} for logical relaxed classification rules based on incremental learning.
+In experiments, {\crr} learns more concise and accurate rule-based classifiers while demonstrating scalability owing to the incremental learning.
 
@@
 Being a stochastic-SAT-based verifier, {\justicia} verifies fairness of compound sensitive groups of Boolean classifiers such as decision trees with higher scalability.
-{\fvgm}, in addition, considers correlated features and verifies fairness of linear classifiers with better scalability and accuracy than existing verifiers.
+{\fvgm}, in addition, considers correlated features and verifies the fairness of linear classifiers with better scalability and accuracy than existing verifiers.
 Finally, given a classifier and a dataset, our framework {\fairXplainer} computes fairness influences of each subset of non-sensitive features as their contribution to the incurred bias of the classifier.
 In experiments, {\fairXplainer} estimates bias with significantly higher accuracy than the existing local explanation approach, while enabling computation of intersectional influences as a mean of better understanding the sources fo bias.
+
+As a future work, we continue our research on fair and interpretable machine learning.
+An immediate research question is to design a fairness improvement or repair algorithm of machine learning classifiers, which benefits from fairness verifier such as {\justicia} and bias identification framework such as {\fairXplainer}.
+Another research question is to extend fairness verification to more complex classifiers, for example, neural networks.
+In addition, in future, we plan to deploy {\fairXplainer} on image classifiers, language models, and computer vision tasks.
+In a separate line of work, we aim to apply incremental solving, as presented in the context of interpretable classification learning, on other optimization problems, even beyond machine learning.
+Therefore, the long term vision of this thesis is to improve machine learning in interpretability, fairness and beyond, with the assistance of formal methods and automated reasoning.
+
+\begin{comment}
+	\begin{itemize}
+		\item Fairness repair
+		\item Incremental solving
+		\item Fairness and interpretability as a service to more complex models.
+	\end{itemize}
+\end{comment}

### 2022-10-31 chapters/background.tex
minor edits

@@
 
+\blue{The complexity $ \bncomplexity(\graph) $ of a Bayesian network $ (\graph, \theta) $ with discrete random variables $ \graphvertices $ is defined as the number of independent parameters used to define the probability distribution of $ \graphvertices $~\cite{murakonda2021quantifying}.
+Let $ \bnnumdiscrete(\graphvertex) $ denote the number of distinct values that a random variable $ \graphvertex $ can take.
+As such, for each conditional probability $ \Pr[\graphvertex_i = v_i | \parent(\graphvertex_i) = \mathbf{u}; \theta] $, we need $ \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1) $ independent parameters.
+Thus, the total complexity of a Bayesian network is:}
+
+\blue{
+\begin{align}
+	\bncomplexity(\graph) = \sum_{i=1}^{|\graphvertices|} \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1)
+\end{align}
+}
+
 \subsection{Global Sensitivity Analysis: Variance Decomposition}

### 2022-10-31 chapters/interpretability/imli/introduction.tex
minor edits

@@
 	&[\neg (0.2 \le \text{area} < 0.3) \text{\textbf{ OR }} \neg (0.1 \le \text{largest symmetry }< 0.2)]
-	\text{\textbf{ \blue{AND} }} \\
-	&[(0.6 \le \text{compactness} < 0.7) \text{\textbf{ OR }} (0.4 \le \text{concave points} < 0.5) \text{\textbf{ OR }}  (\text{area SE} \ge 0.1)\\
-	& \text{\textbf{ OR }} (0.3 \le \text{largest concavity} < 0.4) \text{\textbf{ OR }} (0.2 \le \text{largest area} < 0.3)]
 \end{array}
@@
 	\normalfont
-	We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
-We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with three \emph{clauses} and nine Boolean \emph{literals}.
+	\blue{We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
+We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with two \emph{clauses} and four Boolean \emph{literals}.
 In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{literals} connected by Boolean `OR'.
-Informally, a tumor cell is malignant if at least one literal in each clause becomes true.
+Informally, a tumor cell is malignant if at least one literal in each clause becomes true.}
 \end{example}

### 2022-11-01 chapters/background.tex
minor edits

@@
 \begin{remark}
-	ER-SSAT problem is $\mathrm{NP}^{\mathrm{PP}}$-hard whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
+	The decision problem of ER-SSAT is $\mathrm{NP}^{\mathrm{PP}}$ whereas RE-SSAT problem is $\mathrm{PP}^{\mathrm{NP}}$-complete~\cite{littman2001stochastic}.
 \end{remark}

### 2022-11-15 chapters/interpretability/imli/experiment.tex
addressing reviewers comments

@@
 While {\imli} mostly takes the maximum allowable time ($ 1000 $ seconds) in solving all instances in this dataset, CORELS can solve lower than $ 400 $ instances.
-Thus, {\imli} establishes itself as the most scalable classifier compared to other state-of-the-art interpretable classifiers.
+\blue{The improved performance of {\imli} is due to incremental learning based on the novel integration of iterative learning and mini-batch learning.
+Contrary to our incremental learning, earlier approaches are based on heuristic rule-pruning (e.g., RIPPER) and rule-mining followed by Bayesian optimization (e.g., BRL, BRS) and branch and bounds algorithms (e.g., CORELS).
+Thus, owing to incremental learning, {\imli} establishes itself as the most scalable classifier compared to other state-of-the-art interpretable classifiers.}
 

### 2022-11-15 chapters/interpretability/imli/incremental_learning.tex
addressing reviewers comments

@@
 Firstly, instead of solving a large MaxSAT query for the whole training data, this approach solves a sequence of smaller MaxSAT queries derived from mini-batches of smaller size.
-Secondly, this approach extends to life-long learning~\cite{chen2018lifelong}, where the classifier can be updated incrementally with new samples while also generalizing to previously observed samples.
-In our context of rule-based classifiers, we consider the following heuristic in mini-batch learning.
+\blue{Secondly, this approach extends to online learning, where the classifier can be updated incrementally with new samples while also generalizing to previously observed samples.} In our context of rule-based classifiers, we consider the following heuristic in mini-batch learning.
 

### 2022-11-18 response.tex
preparing response letter

\begin{center}
	\large
	Response Letter
\end{center}
We thank the reviewers for their useful comments and feedback on improving the manuscript and the presentation. In the following, we summarize changes reflected in the revised manuscript.

\paragraph{Distribution shifts in life-long learning.} We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same. One way to account for distribution shift is to consider last $ p $ batches instead of the (single) previous batch in the objective function in mini-batch learning. For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch. Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.
%When distribution changes, the classifier $ \mathcal{R}_{i} $ for the $ i^ $ 
\red{[Add page number]}

\paragraph{Discussions of related works in experiments.}{\imli} scales well than existing interpretable rule-based classifiers because of the incremental solving approach. Most existing works rely on rule-mining of potential classification rules followed by an optimization algorithm such as Bayesian optimization and branch and bound algorithm. We observe incrementality of learning to be more effective. We elaborate this in Section~\ref{interpretability_imli_sec:experiments_scalability}, page 39.

\paragraph{How does {\justicia} outperform a direct approach by learning a Bayesian Network on limited samples?} We agree with the reviewer that learning a Bayesian Network on limited samples does not always result in a better robustness (e.g., less standard deviation of fairness metrics estimation) of {\justicia} than the direct approach estimating fairness on a dataset.

Elaborately, in Chapter~\ref{chapter:justicia} Figure~\ref{fairness_justicia_fig:sample-size}, {\justicia} demonstrates higher robustness than the direct approach, where we consider the distribution in two ways: (i) non-sensitive features are independent and (ii) they are dependent on sensitive features only. Thus, Figure~\ref{fairness_justicia_fig:sample-size} does not involve experiments with Bayesian Network, which we introduce later in Chapter~\ref{chapter:fvgm}. In our revised experiment, we observe that {\justicia} with learning Bayesian network on a dataset, called {\fvgm}, exhibits less robustness due to network learning.

\paragraph{Clarification on Fairness Influence Function (FIF).} We consider an axiom that FIFs are additive where the sum of FIFs of all subsets of non-sensitive features is equal \textit{to the resultant unfairness of the classifier}. Here the unfairness of the classifier is a real value in $ [0,1] $, such as statistical parity, equalized odds, and predictive parity. While one may choose a different axiom, such as multiplicative FIFs, our choice is motivated by the explainability literature for machine learning, such as in SHAP~\cite{begley2020explainability,lundberg2020explaining}. In particular, additivity of FIFs is based on the idea of decomposing the bias/unfairness as the sum of the contribution of different subsets of features on the unfairness of the classifier.

### 2022-12-02 response.tex
adding appendix

@@
 
-\paragraph{Distribution shifts in life-long learning.} We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same.
-One way to account for distribution shift is to consider last $ p $ batches instead of the (single) previous batch in the objective function in mini-batch learning.
+\paragraph{Distribution shifts in life-long learning.} We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same across batches.
+One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning.
 For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch.
 Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.
+We mention this in Section~\ref{interpretability_imli_sec:incremental_learning}, page 28.
 %When distribution changes, the classifier $ \mathcal{R}_{i} $ for the $ i^ $
-\red{[Add page number]}
 
@@
 Most existing works rely on rule-mining of potential classification rules followed by an optimization algorithm such as Bayesian optimization and branch and bound algorithm.
-We observe incrementality of learning to be more effective.
+In experiments, we observe the incremental learning of {\imli} to be more scalable  than the state-of-the-art methods.
 We elaborate this in Section~\ref{interpretability_imli_sec:experiments_scalability}, page 39.
 
-\paragraph{How does {\justicia} outperform a direct approach by learning a Bayesian Network on limited samples?} We agree with the reviewer that learning a Bayesian Network on limited samples does not always result in a better robustness (e.g., less standard deviation of fairness metrics estimation) of {\justicia} than the direct approach estimating fairness on a dataset.
+\paragraph{How does {\justicia} outperform a direct approach by learning a Bayesian Network on limited samples?} We agree with the reviewer that learning a Bayesian Network on limited samples does not always result in a better robustness (e.g., less standard deviation of fairness metrics estimation) of {\justicia} than the direct approach of estimating fairness on a dataset.
 
-Elaborately, in Chapter~\ref{chapter:justicia} Figure~\ref{fairness_justicia_fig:sample-size}, {\justicia} demonstrates higher robustness than the direct approach, where we consider the distribution in two ways: (i) non-sensitive features are independent and (ii) they are dependent on sensitive features only.
-Thus, Figure~\ref{fairness_justicia_fig:sample-size} does not involve experiments with Bayesian Network, which we introduce later in Chapter~\ref{chapter:fvgm}.
-In our revised experiment, we observe that {\justicia} with learning Bayesian network on a dataset, called {\fvgm}, exhibits less robustness due to network learning.
+Elaborately, in Chapter~\ref{chapter:justicia} Figure~\ref{fairness_justicia_fig:sample-size}, {\justicia} demonstrates higher robustness than the direct approach, where we consider a specific distribution of non-sensitive features conditioning only on sensitive features.
+Thus, Figure~\ref{fairness_justicia_fig:sample-size} does not involve experiments with Bayesian Network capturing correlations of all features, which we indeed introduce later in Chapter~\ref{chapter:fvgm}.
+In our revised experiment, we observe that {\justicia} with Bayesian network, called {\fvgm}, exhibits less robustness due to Bayesian network learning.
 
-\paragraph{Clarification on Fairness Influence Function (FIF).} We consider an axiom that FIFs are additive where the sum of FIFs of all subsets of non-sensitive features is equal \textit{to the resultant unfairness of the classifier}.
-Here the unfairness of the classifier is a real value in $ [0,1] $, such as statistical parity, equalized odds, and predictive parity.
-While one may choose a different axiom, such as multiplicative FIFs, our choice is motivated by the explainability literature for machine learning, such as in SHAP~\cite{begley2020explainability,lundberg2020explaining}.
-In particular, additivity of FIFs is based on the idea of decomposing the bias/unfairness as the sum of the contribution of different subsets of features on the unfairness of the classifier.
+\paragraph{Clarification on Fairness Influence Function (FIF).} Our additive axiom for FIFs is based on the idea of decomposing the total unfairness of the classifiers among different subsets of features~\cite{begley2020explainability,lundberg2020explaining}.
+We consider that the sum of FIFs of all subsets of non-sensitive features is equal \textit{to the resultant unfairness of the classifier}, where unfairness is a real number in $ [0,1] $, such as statistical parity, equalized odds, and predictive parity.

### 2022-12-02 chapters/appendix.tex
adding appendix

\appendix

\chapter{Feature Correlations in SSAT-based Fairness Verifier: {\justicia}}
\label{sec:CNF_feature_correlation}
In Chapter~\ref{chapter:justicia}, {\justicia} verifies the fairness of CNF classifiers without considering correlation among features. In this chapter, we address fairness verification of CNF classifiers with correlated features. In particular, we present the encoding of conditional probabilities from a Bayesian network into the SSAT formulation in {\justicia}.

\subsection*{Methodology}
For CNF classifiers, SSAT is a natural choice as it computes the probability of satisfaction of a CNF formula  given quantified Boolean variables, where quantifiers distinguish between (random) non-sensitive variables and (existential or universal) sensitive variables. Let $ \phi_{\widehat{Y}} $ be a CNF classifier such that a satisfying assignment of $ \phi_{\widehat{Y}} $ denotes the positive prediction of the classifier $ \widehat{Y} = 1 $. We consider another CNF formula $ \phi_\BN $ to encode the conditional dependencies among variables in the Bayesian Network $ \BN $, which is given as the input distribution. $ \phi_\BN $ contains auxiliary variables to encode the conditional dependencies, which we discuss shortly. The outline of our methodology is to construct a conjoined CNF formula $ \phi_{\widehat{Y}} \wedge \phi_\BN $, assign appropriate quantifiers to the variables, and solve an SSAT problem on quantified formula $ \phi_{\widehat{Y}} \wedge \phi_\BN $ to answer queries such as $ \max_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ and $ \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $\textemdash the maximum and minimum conditional positive prediction of the classifiers with correlated features, respectively.

\paragraph{Encoding a Bayesian Network as a CNF Formula.}\label{sec:BN_to_CNF}
Our goal is to encode the Bayesian network $ \BN = (\graph, \factors) $ into a CNF formula $ \phi_\BN $ such that \textit{the weighted model count} of $ \phi_\BN $ exactly computes the joint probability distribution in $ \BN $~\cite{chavira2008probabilistic}.  In this context, an SSAT formula trivially does not allow conditional probabilities of randomized quantified variables. Hence, $ \phi_\BN $ contains additional \textit{auxiliary variables} to capture the conditional probabilities, as discussed next.

Let  $ G = (\mathbf{V}, \mathbf{E}) $ where $  $  $ \mathbf{V} \subseteq \nonsensitive \cup \sensitive $, $ \mathbf{E} \subseteq \mathbf{V} \times \mathbf{V} $, and each variable $ V_i \in \mathbf{V} $ is Boolean. For each network variable $ V_i \in \mathbf{V} $, we define a Boolean \textit{indicator}  variable $ \lambda_{V_i} $ such that $ \Pr[\lambda_{V_i}] \triangleq \Pr[V_i] $. We add following constraint in $ \phi_\BN $ to establish the relation between $ \lambda_{V_i} $ and $ V_i $. 
\begin{align}
\lambda_{V_i} \leftrightarrow V_i,
\label{eq:indicator_constraint}
\end{align}
Intuitively, both $ \lambda_{V_i} $ and $ V_i $ are either true or false. This constraint can be trivially translated to clauses in CNF using the equivalence rule $ A \leftrightarrow B \equiv (\neg A \vee B)  \wedge (A \vee \neg B) $ for Boolean variables $ A, B$.

We now present the encoding of conditional probabilities induced by parameters in the network $ \theta $. Let $ V_i \in  \mathbf{V}  $  be a vertex in $ G $ where $ \parent(V_i) \ne \emptyset $ be $ V_i $'s parents and $ |\parent(V_i)| = k $. Additionally, let $ v $ and $ \mathbf{u} \triangleq [u_1,.., u_k] $ be an assignment of  $ V_i $ and $ \parent(V_i)  $, respectively.  To encode $ \Pr[V_i = v| \parent(V_i) = \mathbf{u}]$, we introduce auxiliary variable $ \lambda_{v,\mathbf{u}} $ and add following constraints in $ \phi_{\BN} $.
\begin{align}
\lambda_{v,\mathbf{u}}  \wedge \bigwedge_{j=1}^{k} \lambda_{u_j} \rightarrow \lambda_v
\label{eq:factor_pos}
\end{align}
\begin{align}
\neg \lambda_{v,\mathbf{u}}  \wedge \bigwedge_{j=1}^{k} \lambda_{u_j} \rightarrow \neg \lambda_v
\label{eq:factor_neg}
\end{align}

where $ \lambda_v \equiv \lambda_{V_i} $. Moreover, $ \lambda_{u_j} $ is the indicator variable corresponding to the $ j^\text{th} $ parent in $ \parent(V_i) $. In the above two constraints,	for a fixed assignment $ \mathbf{u} $ of parents $ \parent(V_i) $, both $ \lambda_v $ and $ \lambda_{v,\mathbf{u}} $ are either true or false.  Hence, these two constraints encode the conditional probability of $ V_i = v $ given $ \parent(V_i) = \mathbf{u} $ using $ \Pr[\lambda_{v,\mathbf{u}}] = \Pr[V_i = v| \parent(V_i) = \mathbf{u}]$. Both constraints can be translated to CNF clauses trivially. For example, Eq.~\ref{eq:factor_pos} is translated as $ \neg \lambda_{v,\mathbf{u}}  \vee \bigvee_{j=1}^{k} \neg \lambda_{u_j} \vee \lambda_v $. We next analyze the complexity of $ \phi_\BN $ in terms of the number of variables and clauses.

\begin{lemma}
	For a Bayesian network $ \BN = (\graph, \factors) $ defined over $ n $ Boolean variables and $ \bncomplexity(\graph) $ network complexity, the encoded CNF formula $ \phi_\BN $ has $ n + \bncomplexity(\graph) $  variables and $ 2(n + \bncomplexity(\graph)) $ clauses. 
\end{lemma}
\begin{proof}
	Since the DAG in the Bayesian network has $ n $ vertices, we consider $ n $ indicator variables. Moreover, for encoding conditional probabilities, we consider $ \bncomplexity(\graph) $ auxiliary variables where $ \bncomplexity(\graph) $ denotes the the number of independent parameters in the network (ref. Chapter~\ref{chapter_fairness_preliminaries_BN}). Hence, total variables in $ \phi_\BN $ is $ n + \bncomplexity(\graph)  $.

	According to Eq.~\eqref{eq:indicator_constraint},~\eqref{eq:factor_pos},~\eqref{eq:factor_neg}, there are $2( n + \bncomplexity(\graph)) $ clauses in $ \phi_\BN $.

\end{proof}

\paragraph{Quantifiers in $ \phi_{\widehat{Y}} \wedge \phi_\BN $.}
We now discuss the quantifiers in $ \phi_{\widehat{Y}} \wedge \phi_\BN $, the SSAT solution of which  constitutes the maximum (minimum) probability of positive prediction of a CNF classifier. $ \phi_{\widehat{Y}} \wedge \phi_\BN $ contains four  categories of variables : (i) sensitive variables $ \sensitive $, (ii) non-sensitive variables $ \nonsensitive $, (iii) indicator variables $  \lambda_{V_i} $, and (iv) auxiliary variables $ \lambda_{v,\mathbf{u}} $. Among them, (iii) and (iv) are associated with $ \phi_\BN $ and the rest for $ \phi_{\widehat{Y}} $. For computing the maximum probability of positive prediction of the classifier, we construct  an exists-random-exists (ERE) SSAT formula with following quantifiers: we set sensitive features $ \sensitive $ with existential quantifiers in the beginning of the prefix of the SSAT formula followed by $  \lambda_{V_i}, \lambda_{v,\mathbf{u}} $, and $ X_j \in \nonsensitive \setminus \mathbf{V} $ with randomized quantifiers. The remaining variables  $ X_i \in \mathbf{V} $ are existentially quantified as their assignment is fixed by indicator variables $ \lambda_{V_i} $. In contrast, for computing the minimum probability of positive prediction of the classifier, we consider an universal-random-exists (URE) SSAT formula  where we set sensitive features $ \sensitive $ as universal quantifiers with all other quantifiers remaining same.

### 2022-12-02 chapters/background.tex
adding appendix

@@
 \subsection{Bayesian Network}
+\label{chapter_fairness_preliminaries_BN}
 In general, a Probabilistic Graphical Model~\cite{koller2009probabilistic}, and specifically a \textit{Bayesian network}~\cite{pearl1985bayesian,chavira2008probabilistic}, encodes the dependencies and conditional independence between a set of random variables.

### 2022-12-02 chapters/interpretability/imli/incremental_learning.tex
adding appendix

@@
 For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch.
-Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution.
+Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{\blue{We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same across batches.
+One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning.
+For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch.
+Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.}}.
 Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity.

### 2022-12-27 response.tex
updated response

@@
 We consider that the sum of FIFs of all subsets of non-sensitive features is equal \textit{to the resultant unfairness of the classifier}, where unfairness is a real number in $ [0,1] $, such as statistical parity, equalized odds, and predictive parity.
+
+\paragraph{Clarification on Bayesian Network.} We consider a Bayesian network as an input distribution to express the conditional dependencies and independencies among features.
+In Chapter~\ref{chapter:CNF_feature_correlation}, we demonstrate the encoding of probabilistic inference into SSAT via additional Boolean variables and clauses~\cite{chavira2008probabilistic}.
+We also discuss the complexity of the SSAT encoding in terms of the complexity of the Bayesian network (ref.~\ref{chapter_fairness_preliminaries_BN}, page 15).

### 2022-12-27 chapters/appendix.tex
updated response

@@
 \chapter{Feature Correlations in SSAT-based Fairness Verifier: {\justicia}}
-\label{sec:CNF_feature_correlation}
+\label{chapter:CNF_feature_correlation}
 In Chapter~\ref{chapter:justicia}, {\justicia} verifies the fairness of CNF classifiers without considering correlation among features.

### 2022-12-27 tp_report/main.tex
added tp report

\documentclass[10pt]{article}
% Define margins
\setlength{\topmargin}{-1.0cm}
\setlength{\oddsidemargin}{0.2cm}
\setlength{\textwidth}{16cm}
\setlength{\textheight}{23.0cm}

\input{macros}

\begin{document}
	We thank the reviewers for their constructive feedback during the doctoral seminar. In this report, we discuss our ongoing research to be added to the thesis along with explaining some questions during the presentation.

	\section*{Fairness Influence Functions}  The goal of this project is to interpret group fairness of classifiers by identifying (subset of) non-sensitive features responsible for the bias/unfairness of the classifier with respect to a dataset. Since both sensitive and non-sensitive features may be correlated in a dataset containing historical bias, such an identification of malicious features will guide a future endeavor to design a fairness improvement algorithm. Therefore, we propose a method to compute the fairness influence function (FIF) of features. Intuitively, FIF of non-sensitive features implies how much those features contribute to the bias of the classifier\textemdash a well known measure of bias is the maximum difference in the positive prediction of the classifier conditioned on all sensitive groups.

	\paragraph{Problem Statement.} Let $ \nonsensitive \in \mathbb{R}^k $ and $ \sensitive \in \mathbb{N}^m $ denote non-sensitive and sensitive features, respectively. We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric\footnote{Statistical parity belongs to \textit{independence} measuring group fairness metrics, where the prediction $ \widehat{Y} $ is statistically independent of sensitive features $ \sensitive $.  The statistical parity of  a classifier is measured as $ f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] $, which is the difference between the maximum and minimum conditional probability of positive prediction the classifier for different sensitive groups.} $ f(\alg, \mathbf{D}) \in [0, 1] $.  Our objective is to compute the influences of non-sensitive features on $ f $. Particularly, we compute the influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.

	\begin{definition}[Fairness Influence Function]
		Fairness Influence Function (FIF) $ w_{\mathbf{S}}: (\alg, \mathbf{X}_{\mathbf{S}}) \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$. 
		\begin{equation}\label{eq:fif_decompose}
			w_{\mathbf{S}}  \triangleq \frac{\mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}_{\max}]}{1 - \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}_{\min}] }{1 - \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}]}
		\end{equation}

	\end{definition}

	\paragraph{Interpretation of FIFs.} In the definition of $ w_{\mathbf{S}} $, $ \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ is the decomposed variance  of the conditional variance of classifier's prediction $ \mathsf{Var}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ with respect to features $ \nonsensitive_{\mathbf{S}} $; and  sensitive group $ \mathbf{a}_{\max} $ (resp.\ $ \mathbf{a}_{\min} $) obtains the maximum (resp.\ minimum) conditional probability of positive prediction by the classifier, e.g., $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $. Intuitively, between two specific sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $, the FIF $ w_{\mathbf{S}} $ implies the difference in the normalized\footnote{The numerator in each term of $ w_{\mathbf{S}} $ is a constant, which is in proportion to classifier's positive prediction conditioned on the sensitive group.} conditional variance of classifier's prediction attributed for features  $ \nonsensitive_{\mathbf{S}} $. Thus, FIF of features is non-zero when the variance-difference of prediction between sensitive groups is non-zero for those features, and vice versa. Our definition of $ w_{\mathbf{S}} $ results in the following property of FIFs.

	\paragraph{Additivity of FIFs.} FIF provides a non-trivial way to decompose the bias $ f(\alg, \mathbf{D}) $ of the classifier on dataset $ \mathbf{D} $ among all possible subsets of non-sensitive features.

	\begin{theorem}
		\label{thm:fif}
		Let $ f(\alg, \mathbf{D}) $ be the bias/unfairness of the classifier $ \alg $ on dataset $ \mathbf{D} $ according to linear fairness metrics such as statistical parity. Let $ w_{\mathbf{S}}  $ be the FIF of a subset of non-sensitive features $ \nonsensitive_{\mathbf{S}} $ as defined in Eq.~\eqref{eq:fif_decompose}. Therefore, the sum of FIFs of all subset of non-sensitive features is equal to the bias of the classifier. 
		\begin{align}\label{eq:constraint_framework}
			f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
		\end{align}
	\end{theorem}

	\begin{proof}
	The proof-sketch of Theorem~\ref{thm:fif} is to first view fairness metrics in terms of the variance of classifier's prediction and then apply variance decomposition to derive FIFs of features. For statistical parity fairness metric, the derivation is the following.

	\begin{align*}
		f_{\mathsf{SP}}(\alg, \mathbf{D}) &\triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] \\ 
		&= \Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}_{\max}] - \Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}_{\min}] \\
		& = \frac{\mathsf{Var}[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}_{\max}]}{1 - \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{\mathsf{Var}[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}_{\min}]}}{1 - \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}]}
	\end{align*}

	Next, we decompose the total variance with respect to all non-sensitive  into the sum of variances of  all subsets of non-sensitive features based on global sensitivity analysis (GSA).

	\[ \mathsf{Var}[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] = \sum_{\mathbf{S} \subseteq [k] \setminus \emptyset} \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} =1 | \mathbf{A}= \mathbf{a}]
	\]

	From this decomposition, term associated with features $ \nonsensitive_{\mathbf{S}} $ constitutes the FIF of those features in Eq.~\eqref{eq:fif_decompose}.

	\end{proof}

	\section*{Revised Experiments on Fairness Quantification} In the earlier experimental setup for comparing different fairness quantifiers, our method Justicia considers Bernoulli distributions of features. Competitive methods FairSquare and VeriFair, however, consider Gaussian distributions along with Bernoulli distributions. To make a fair comparison, we are designing an experimental setup where all three methods take the same Bernoulli distributions of features as input. We add the experimental findings in the thesis.

	\section*{Accuracy of  Interpretable Rule-based Classifiers} Our motivation for proposing an incremental MaxSAT encoding for interpretable rule-based classification learning is to improve \emph{the scalability of learning} than the existing non-incremental and direct MaxSAT encoding. Since the priority is only on scalability, we do not claim to improve prediction accuracy using the incremental learning method. As pointed out by the reviewers that the experimental result on accuracy is not statistically significant, we are working on a detailed experimental comparison between the incremental and non-incremental encodings and add the results to the thesis.
\end{document}

### 2023-02-07 chapters/acknowledgements.tex
with new formatting

\begin{acknowledgments}

\end{acknowledgments}

### 2023-02-07 chapters/background.tex
with new formatting

@@
 	Let $ (\propclause,\card) $ be a clause in a relaxed-CNF where  $ C $ has $ m  $ literals and $ \card \in \{1,\dots,m\} $ is the threshold on literals.
-An equivalent compact encoding of $ (\propclause,\card) $  into a CNF formula $ \propformula= \bigwedge_i \propclause_i $ requires $ {m \choose{m-\card+1}} $  clauses where each clause is distinct and has $ m-\card+1 $ literals of $ (\propclause,\card) $.
-Therefore,  the total number of literals in $ \propformula $ is $ (m-\card+1){m\choose{m-\card+1}} = m $ (equal succinctness) when $ \card=1 $, otherwise $ (m-\card+1){m\choose{m-\card+1}} > m $ (exponential succinctness).
+An equivalent compact encoding of $ (\propclause,\card) $  into a CNF formula $ \propformula= \bigwedge_i \propclause_i $ requires $ {\binom{m}{m-\card+1}} $  clauses where each clause is distinct and has $ m-\card+1 $ literals of $ (\propclause,\card) $.
+Therefore,  the total number of literals in $ \propformula $ is $ (m-\card+1){\binom{m}{m-\card+1}} = m $ (equal succinctness) when $ \card=1 $, otherwise $ (m-\card+1){\binom{m}{m-\card+1}} > m $ (exponential succinctness).
 \end{theorem}
@@
 
-\subsection{Stochastic Subset Sum Problem ({\stochastic})} 
+\subsection{Stochastic Subset Sum Problem} 
 Let $ \boolset \triangleq \{\bool_i\}_{i=1}^{|\boolset|}$ be a set of Boolean variables and $ \varweight_i \in \mathbb{Z} $ be the weight of $ \bool_i $.
 Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $ \mathbf{b} $.
-Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
+Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical
+computer science~\cite{kleinberg2006algorithm}.
 The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
@@
 \label{chapter_interpretability_preliminaries}
-
 We consider a standard binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
@@
 \emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}.
-To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
-Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
+To illustrate, let us consider a real-valued function $  \fiffunc(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
+Now, we decompose $ \fiffunc(\mathbf{Z}) $ among the subsets of inputs, such that:
 
 \begin{align}
-	\function(\mathbf{Z}) &= \function_0 + \sum_{i=1}^{\numnonsensitive} \function_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \function_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \function_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
+	\fiffunc(\mathbf{Z}) &= \fiffunc_0 + \sum_{i=1}^{\numnonsensitive} \fiffunc_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \fiffunc_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \fiffunc_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
 	\\    
-	&= \function_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
+	&= \fiffunc_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
 \end{align}
 
-In this decomposition, $ \function_0 $ is a constant, $ \function_{\{i\}} $ is a function of $ Z_i $, $ \function_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
+In this decomposition, $ \fiffunc_0 $ is a constant, $ \fiffunc_{\{i\}} $ is a function of $ Z_i $, $ \fiffunc_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
 Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.
-We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \function_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.
+We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \fiffunc_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.
 The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}.
-$ \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously.
+$ \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously.
 %\unsure{Is this sentence necessary?}
@@
 
-Now, if we assume $ \function $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
+Now, if we assume $ \fiffunc $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
 
@@
 
-where $ V_{\{i\}} $ is the variance of $ \function_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \function_{\{i,j\}} $ and so on. Formally,
+where $ V_{\{i\}} $ is the variance of $ \fiffunc_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \fiffunc_{\{i,j\}} $ and so on. Formally,
 
@@
 Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$. %\improvement{Check notations.}
-This variance decomposition shows how the variance of $\function(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them.
+This variance decomposition shows how the variance of $\fiffunc(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them.
 Together all terms sum to the total variance of the model output.

### 2023-02-07 chapters/fairness/fvgm/applications.tex
with new formatting

@@
-\section{Applications of {\fvgm}}\label{fvgm_sec:applications}
+\section{Applications}\label{fvgm_sec:applications}
 In this section, we apply {\fvgm} for verifying fairness-enhancing algorithms and depreciation attacks.

### 2023-02-07 chapters/fairness/fvgm/framework.tex
with new formatting

@@
-\section{{\fvgm}: Fairness Verification with Graphical Models}\label{fvgm_sec:fvgm}
+\section{Fairness Verification with Graphical Models}\label{fvgm_sec:fvgm}
 
@@
 
-\subsection{$ \stochastic $: Stochastic Subset Sum Problem}\label{fvgm_sec:stochastic_sum_set_sum}
+\subsection{Stochastic Subset Sum Problem}\label{fvgm_sec:stochastic_sum_set_sum}
 Now, we formally describe the specification and semantics of {\stochastic}.
@@
 
- \subsection{A Dynamic Programming Solution for {\stochastic}}
+ \subsection{A Dynamic Programming Solution}
  \label{fvgm_sec:dp_formulation}

### 2023-02-07 chapters/background.tex
adding fairXplainer

@@
 
-\subsection{Global Sensitivity Analysis: Variance Decomposition}
-Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
+\subsection{Global Sensitivity Analysis (GSA): Variance Decomposition}
+Global sensitivity analysis is an active field of research that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input variables~\cite{saltelli2008global}.
 Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
@@
 \emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}.
-To illustrate, let us consider a real-valued function $  \fiffunc(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
-Now, we decompose $ \fiffunc(\mathbf{Z}) $ among the subsets of inputs, such that:
+To illustrate, let us consider a real-valued function $  \fiffunc(\feature) $, where $ \feature $ is a vector of $ \numfeatures $ input variables $ \{Z_1, \dots, Z_\numfeatures\} $.
+% and $ Y $ is a univariate output. 
+%\blue{In sensitivity analysis, assumptions are made on the independence and uniform distribution of $ X_i $'s within the unit hypercube}, that is, $ X_i \in [0,1] $ for $ i = \{ 1, \dots, k\} $.\footnote{In our experiments, we have not considered this yet.} Let $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $.
+Then,
+Now, we decompose $ \fiffunc(\feature) $ among the subsets of inputs, such that:
+\begin{align}
+	\fiffunc(\feature) &= \fiffunc_0 + \sum_{i=1}^{\numfeatures} \fiffunc_{\{i\}}(Z_i) +  \sum_{i < j}^{\numfeatures} \fiffunc_{\{i,j\}}(Z_i, Z_j)  + \cdots\notag + \fiffunc_{\{1, 2, \dots, \numfeatures\}} (Z_1, Z_2, \dots, Z_{\numfeatures})\\
+	&= \fiffunc_0 +  \sum_{\mathbf{S} \subseteq [\numfeatures]} \fiffunc_{\mathbf{S}}(\feature_{\mathbf{S}})\label{fairness_fairXplainer_eq:functional_decomposition_set_notation}
+\end{align}
+%	$S_> = \{{S_0}, \ldots, {S_{|s|}}\}$ i.e. a monotonically increasing ordering of $|S|$ features for a given starting feature $Z_{S_0}$ such that $S_0 <S_1 <\ldots <S_{|S|}$.
+The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{fairness_fairXplainer_eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}.
+In this decomposition, $ \fiffunc_0 $ is a constant, $ \fiffunc_{\{i\}} $ is a function of $ Z_i $, $ \fiffunc_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
+Adhering to the set-based notations, we denote by $ g_{\mathbf{S}} $ a function of a non-empty subset of variables $ \feature_{\mathbf{S}}  \triangleq \{Z_i \mid i \in \mathbf{S}\} \subseteq \feature $, where $ \mathbf{S} = \{S_i \mid 1 \le |S_i| \le \numfeatures\} \subseteq [\numfeatures] $ is a non-empty subset of indices with $ [\numfeatures] \triangleq \{1,2,\dots, \numfeatures\}  $.
+Here, $|\mathbf{S}|=1$ quantifies an individual variable's effect while $|\mathbf{S}|>1$ quantifies the higher-order intersectional effect of variables.
 
-\begin{align}
-	\fiffunc(\mathbf{Z}) &= \fiffunc_0 + \sum_{i=1}^{\numnonsensitive} \fiffunc_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \fiffunc_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \fiffunc_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
-	\\    
-	&= \fiffunc_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
+Considering $ \fiffunc $ as square integrable, we obtain the decomposition of the variance of $ \fiffunc(\feature) $ expressed as the sum of variances of $\fiffunc_{\mathbf{S}}$'s~\cite{sobol1990sensitivity}.
+\begin{align}\label{fairness_fairXplainer_eq:variance_decomposition_set_notation}
+	\mathsf{Var}[g(\feature)] &= \sum_{i=1}^{\numfeatures}V_{\{i\}} +  \sum_{i<j}^{\numfeatures} V_{\{i,j\}}  + \dots +  V_{\{1, 2\dots, \numfeatures\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive]} V_{\mathbf{S}}
 \end{align}
-
-In this decomposition, $ \fiffunc_0 $ is a constant, $ \fiffunc_{\{i\}} $ is a function of $ Z_i $, $ \fiffunc_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
-Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.
-We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \fiffunc_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.
-The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}.
-$ \fiffunc_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously.
-%\unsure{Is this sentence necessary?}
-For $|\mathbf{S}|=1$, it quantifies an individual variable's effect. For $|\mathbf{S}|>1$, it quantifies the higher-order interactive effect of variables.
-
-Now, if we assume $ \fiffunc $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
-
-\begin{align}\label{eq:variance_decomposition_set_notation}
-	\mathsf{Var}[g(\mathbf{Z})] &= \sum_{i=1}^{\numnonsensitive}V_{\{i\}} +  \sum_{i<j}^{\numnonsensitive} V_{\{i,j\}}  + \cdots  + V_{\{1, 2\dots, \numnonsensitive\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}}
-\end{align}
-
 where $ V_{\{i\}} $ is the variance of $ \fiffunc_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \fiffunc_{\{i,j\}} $ and so on. Formally,   
-
-\[ V_{\mathbf{S}} \triangleq \mathsf{Var} _{\mathbf{Z}_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}/\mathbf{Z}_{\mathbf{S}}}[g(\mathbf{Z})\mid \mathbf{Z}_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}/\emptyset}{V} _{\mathbf{S}'}.\]
-
-Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$. %\improvement{Check notations.}
-This variance decomposition shows how the variance of $\fiffunc(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them.
-Together all terms sum to the total variance of the model output.
+\begin{equation}
+	\label{eq:decomposed_variance}
+	V_{\mathbf{S}} \triangleq \mathsf{Var} _{\feature_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}\setminus\feature_{\mathbf{S}}}[g(\feature)\mid \feature_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}}{V} _{\mathbf{S}'}.
+\end{equation} 
+Here, $\mathbf{S}'$ denotes all the non-empty and ordered proper subsets of $\mathbf{S}$.
+Thus, $ V_{\mathbf{S}} $ is the variance w.r.t.\ $ \feature_{\mathbf{S}} $ by subtracting the variances of all non-empty proper subsets of $ \feature_{\mathbf{S}} $.
+As a result, Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} demonstrates how the variance of $\fiffunc(\feature) $ can be decomposed into terms attributable to each input feature, as well as the intersectional effects among them.
+Conversely, together all terms sum to the total variance of the model output.

### 2023-02-07 chapters/fairness/fif/abstract.tex
adding fairXplainer

@@
 \begin{abstract}
-
-	Fairness in machine learning has attained significant focus due to the widespread application of machine learning in high-stake decision-making tasks.
-Unless regulated with a fairness objective, machine learning classifiers might demonstrate unfairness/bias towards certain demographic populations in the data.
-Thus, the quantification and mitigation of the bias induced by classifiers have become a central concern.
-In this paper, \textit{we aim to quantify the influence of different features on the bias of a classifier}.
-To this end, we propose a framework of \emph{Fairness Influence Function} (FIF), and compute it as a scaled difference of conditional variances in the classifier's prediction.
-We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features.
-Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.
-
+	Fairness in machine learning has attained significant focus due to the widespread application in high-stake decision-making tasks.
+Unregulated machine learning classifiers can exhibit bias towards certain demographic groups in data, thus the quantification and mitigation of classifier bias is a central concern in fairness in machine learning.
+In this paper, \textit{we aim to quantify the influence of different features in a dataset on the bias of a classifier}.
+To do this, we introduce the \textit{Fairness Influence Function} (FIF).
+This function breaks down bias into its components among individual features and the intersection of multiple features.
+The key idea is to represent existing group fairness metrics as the difference of the scaled conditional variances in the classifier's prediction and apply a decomposition of variance according to global sensitivity  analysis.
+To estimate FIFs, we instantiate an algorithm {\framework} that applies variance decomposition of classifier's prediction following local regression.
+	Experiments demonstrate that {\framework} captures FIFs of individual feature and intersectional features, provides a better approximation of bias based on FIFs, demonstrates higher correlation of FIFs with fairness interventions, and detects changes in bias due to fairness affirmative/punitive actions in the classifier.
 \end{abstract}

### 2023-02-07 chapters/fairness/fif/appendix.tex
adding fairXplainer

@@
 \appendix
+\onecolumn
 \section*{Appendix}
@@
 
-\section{FIF as Difference of Conditional Variances: Proof of Theorem 1}\label{sec:proof_1}
+\begin{comment}
+\section{FIF as Difference of Conditional Variances: Proof of Theorem 1}\label{fairness_fairXplainer_sec:proof_1}
 
-\begin{theoremrep}[FIF as Difference of Conditional Variances]
-	\label{lemma:fif_rep}
+\begin{reptheorem}{lemma:fif_rep}[FIF as Difference of Conditional Variances]
 	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied.
@@
 Then, FIF of $ \nonsensitive_{\mathbf{S}} $ corresponding to statistical parity is
-	\begin{equation}\tag{\ref{eq:fif_decompose}}
+	\begin{equation}\tag{\ref{fairness_fairXplainer_eq:fif_decompose}}
 	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])},
 	\end{equation}
-	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
-\end{theoremrep}
+	and FIFs defined by Equation~\eqref{fairness_fairXplainer_eq:fif_decompose} satisfy Axiom~\ref{fairness_fairXplainer_axm:additivity}.
+\end{reptheorem}
 
@@
 \begin{align}
-	\label{eq:sp_to_var}
+	\label{fairness_fairXplainer_eq:sp_to_var}
 	p_\mathbf{a} -  p_\mathbf{a'}  = \frac{	\mathsf{Var}[B_\mathbf{a}] - \mathsf{Var}[B_\mathbf{a'}]}{1 - (p_\mathbf{a} + p_\mathbf{a'})}
@@
 
-By setting $ \mathbf{a} = \mathbf{a}_{\max} $ and $ \mathbf{a'} = \mathbf{a}_{\min} $ in Equation~\eqref{eq:sp_to_var}, we express the statistical parity of the classifier in terms of the scaled difference in the conditional variance of the positive prediction of the classifier.
+By setting $ \mathbf{a} = \mathbf{a}_{\max} $ and $ \mathbf{a'} = \mathbf{a}_{\min} $ in Equation~\eqref{fairness_fairXplainer_eq:sp_to_var}, we express the statistical parity of the classifier in terms of the scaled difference in the conditional variance of the positive prediction of the classifier.
 \begin{align}
-	f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq p_{\mathbf{a}_{\max}} -  ~~p_{\mathbf{a}_{\min}}  &= \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}\label{eq:sp_var_diff}\\
+	f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq p_{\mathbf{a}_{\max}} -  ~~p_{\mathbf{a}_{\min}}  &= \frac{	\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})}\label{fairness_fairXplainer_eq:sp_var_diff}\\
 	&= \frac{	\mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\max}] - \mathsf{Var}[\hat{Y} = 1| \sensitive = \mathbf{a}_{\min}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \notag.
@@
 
-Next, we apply variance decomposition (Equation~\eqref{eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $.
-From Equation~\eqref{eq:variance_decomposition_set_notation} in Section~\ref{sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as
+Next, we apply variance decomposition (Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $.
+From Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} in Section~\ref{fairness_fairXplainer_sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as
 \[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
@@
 
-Now, we apply variance decomposition on both $ \mathsf{Var}[B_{\mathbf{a}_{\max}}] $ and $ \mathsf{Var}[B_{\mathbf{a}_{\min}}] $ in Equation~\ref{eq:sp_var_diff} to obtain
+Now, we apply variance decomposition on both $ \mathsf{Var}[B_{\mathbf{a}_{\max}}] $ and $ \mathsf{Var}[B_{\mathbf{a}_{\min}}] $ in Equation~\ref{fairness_fairXplainer_eq:sp_var_diff} to obtain
 
 \begin{align}
-f_{\mathsf{SP}}(\alg, \mathbf{D}) & = \frac{\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\max}})}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \label{eq:influence_value}
+f_{\mathsf{SP}}(\alg, \mathbf{D}) & = \frac{\mathsf{Var}[B_{\mathbf{a}_{\max}}] - \mathsf{Var}[B_{\mathbf{a}_{\min}}]}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\max}})}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}} - V_{\mathbf{a}_{\min}, \mathbf{S}}}{1 - (p_{\mathbf{a}_{\max}} + p_{\mathbf{a}_{\min}})} \label{fairness_fairXplainer_eq:influence_value}
 \end{align}
@@
 
-Following Equation~\eqref{eq:influence_value}, $	w_{\mathbf{S}} $ satisfies Axiom~\ref{axm:additivity} as $f_{\mathsf{SP}}(\alg, \mathbf{D}) = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} $.
+Following Equation~\eqref{fairness_fairXplainer_eq:influence_value}, $	w_{\mathbf{S}} $ satisfies Axiom~\ref{fairness_fairXplainer_axm:additivity} as $f_{\mathsf{SP}}(\alg, \mathbf{D}) = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}} $.
 \end{proof}
 
+\end{comment}
+
+\section{Proofs of Properties and Implications of FIF}
+\begin{reptheorem}{fairness_fairXplainer_thm:fif_property}
+	Let $ f(\alg, \mathbf{D}) $ be the bias/unfairness of the classifier $ \alg $ on dataset $ \mathbf{D} $ according to linear group fairness metrics such as statistical parity.
+Let $ w_{\mathbf{S}}  $ be the FIF of a subset of  features $ \feature_{\mathbf{S}} $ as defined in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose}.
+	\begin{enumerate}
+		\item[(a)] \textit{The decomposability property} of FIF states that the sum of FIFs of all subset of  features is equal to the bias of the classifier. 
+		\begin{align}
+		\sum_{\mathbf{S} \subseteq [\numfeatures] } w_{\mathbf{S}} = f(\alg, \mathbf{D})
+		\end{align}
+		\item[(b)] \textit{The symmetry property} states that  two features $ Z_i $ and $ Z_j $ are equivalent based on FIF if the sum of corresponding individual influences and the intersectional influences with all other features are the same.
+Mathematically,
+		\begin{align}
+		\sum_{\mathbf{S}'' \subseteq [m]\setminus \{i,j\}}w_{\mathbf{S}''\cup \{i\}} = \sum_{\mathbf{S}'' \subseteq [m]\setminus\{i,j\}}w_{\mathbf{S}''\cup \{j\}}  
+		\end{align}
+		if $\sum_{\mathbf{S}' \subseteq \mathbf{S}\cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{j\}} w_{\mathbf{S}'}$ for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ containing neither $ i $ nor $ j $.
+		\item[(c)] \textit{The null property} of FIF states that feature $ X_i $ is a dummy or neutral feature if sum of its individual influence and the intersectional influences with all other features is zero.
+Mathematically,
+		\begin{align}
+		\sum_{\mathbf{S}'' \subseteq [m]\setminus\{i\}}w_{\mathbf{S}''\cup \{i\}} = 0 	 
+		\end{align}
+		if	$\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S}} w_{\mathbf{S}'}$ for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ that does not contain $ i $.
+	\end{enumerate}
+\end{reptheorem}
+
+\begin{proof}
+	(a) The decomposability property of FIF is based on GSA, where the total variance is decomposed to the variances of individual and intersectional inputs. 
+	\begin{align*}
+	\sum_{\mathbf{S} \subseteq [\numfeatures]} w_{\mathbf{S}} &= \sum_{\mathbf{S} \subseteq [\numfeatures]} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+	&= \frac{\sum_{\mathbf{S} \subseteq [\numfeatures]} V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\sum_{\mathbf{S} \subseteq [\numfeatures]} V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+	&= \frac{V_{\mathbf{a}_{\max}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+	&=\frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\min}] }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+	&= f_{\mathsf{SP}}(\alg, \mathbf{D})
+	\end{align*}
+	Thus, applying Lemma~\ref{fairness_fairXplainer_lm:sp_var_relation}, we prove the decomposability property of FIF for statistical parity.
+
+	(c) We observe that 
+	\begin{align}\label{fairness_fairXplainer_eq:sum_decomp}
+		\sum_{\mathbf{S}' \subseteq \mathbf{S}\cup \{i\}} w_{\mathbf{S}'} = \sum_{\mathbf{S}' \subseteq \mathbf{S}} w_{\mathbf{S}'} + \sum_{\mathbf{S}'' \subseteq \mathbf{S}} w_{\mathbf{S}''\cup \{i\}}.
+	\end{align}
+	This means that we can decompose the sum of FIFs of all the subsets of $\mathbf{S}\cup \{i\}$ into two non-overlapping sums: the subsets that include $i$ and the subsets that do not include.
+
+	Since we assume that for $i$,  $\sum_{\mathbf{S}' \subseteq \mathbf{S}\cup \{i\}} w_{\mathbf{S}'} = \sum_{\mathbf{S}' \subseteq \mathbf{S}} w_{\mathbf{S}'}$ holds true, it implies
+		\begin{align*}
+	\sum_{\mathbf{S}'' \subseteq \mathbf{S}} w_{\mathbf{S}''\cup \{i\}} = 0.
+	\end{align*}
+	Now, considering $\mathbf{S} = [m]\setminus \{i\}$, i.e. the set of all features  except $i$, concludes the proof.
+
+	(b) Proof of the symmetry property follows similar decomposition of the sum of FIFs as Equation~\eqref{fairness_fairXplainer_eq:sum_decomp}.
+\end{proof}
+
+\begin{repproposition}{fairness_fairXplainer_prop:neg_fif}
+	When $ w_{\mathbf{S}} < 0 $, i.e.
+features $ \feature_{\mathbf{S}} $ decrease bias, the decomposed variance of CPPs w.r.t.\ $ \feature_{\mathbf{S}} $ follows $ V_{\mathbf{a}_{\max}, \mathbf{S}} < V_{\mathbf{a}_{\min}, \mathbf{S}}  $.
+\end{repproposition}
+
+\begin{proof}
+	When $ w_{\mathbf{S}} < 0 $,
+	\begin{align*}
+	&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} < 0 \\
+	\implies&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} < \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \\
+	\implies&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} < \frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \le 1 \\
+	\implies&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} < 1 \\ 
+	\implies&V_{\mathbf{a}_{\max}, \mathbf{S}} < V_{\mathbf{a}_{\min}, \mathbf{S}}
+	\end{align*}
+\end{proof}
+
+\begin{repproposition}{fairness_fairXplainer_prop:pos_fif}
+	{If the decomposed variance of CPPs w.r.t.\ $ \feature_{\mathbf{S}} $ satisfies $ V_{\mathbf{a}_{\max}, \mathbf{S}} > {V_{\mathbf{a}_{\min}, \mathbf{S}}} $, the corresponding FIF $ w_{\mathbf{S}} > 0  $, i.e.
+features $ \feature_{\mathbf{S}} $ increase bias.}
+\end{repproposition}
+\begin{proof}\
+	Since $	V_{\mathbf{a}_{\max}, \mathbf{S}} > V_{\mathbf{a}_{\min}, \mathbf{S}}$, we obtain $
+	\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} > 1.$
+Since $\mathbf{a}_{\max}$ and $\mathbf{a}_{\min}$ are the most and least favored groups respectively, the probability of yielding a positive prediction is greater or equal for $\mathbf{a}_{\max}$ than $\mathbf{a}_{\min}$.
+Thus, ${\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} \leq {\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}$, which implies that $\frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \leq 1$.
+
+Combining both the observations, we obtain
+\begin{align*}
+&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} > \frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+\implies &\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} > \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
+\implies &\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} > 0\\
+\implies &w_{\mathbf{S}} > 0.
+\end{align*}
+\end{proof}
+
+\begin{comment}
+\begin{repproposition}{prop:pos_fif}
+Let $ n_{\max} $ be the number of data points in the dataset $ \mathbf{D} $ belonging to the most favored sensitive group $ \mathbf{a}_{\max} $ and $ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}] \ne 0 $.
+Therefore, when $ w_{\mathbf{S}} > 0  $, i.e.
+features $ \feature_{\mathbf{S}} $ increase bias, the decomposed variance of CPPs w.r.t.\ $ \feature_{\mathbf{S}} $ follows $ V_{\mathbf{a}_{\max}, \mathbf{S}} \ge \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{n_{\max}} $.
+\end{repproposition}
+
+\begin{proof}
+Let $ n_{\min} $ be the number of data points in the dataset $ \mathbf{D} $ belonging to the least favored sensitive group $ \mathbf{a}_{\min} $. Then,	when $ w_{\mathbf{S}} > 0 $,
+\begin{align*}
+& \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} > \frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \ge \frac{\frac{1}{n_{\max}}}{\frac{n_{\min}}{n_{\min}}}  = \frac{1}{n_{\max}} \\
+& \Rightarrow V_{\mathbf{a}_{\max}, \mathbf{S}} \ge \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{n_{\max}}
+\end{align*}
+
+\end{proof}
+
+\end{comment}
+
 \section{A Smoothing Operator $\textsc{Smooth}$: Cubic Splines}
-\label{sec:smoothing} 
-In the $\textsc{LocalRegression}$ module of \framework{} (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}, Algorithm~\ref{algo:framework}), we use a smoothing operator $\textsc{Smooth}$ (Line~\ref{alg_line:backfitting_step}).
+\label{fairness_fairXplainer_sec:smoothing} 
+In the $\textsc{LocalRegression}$ module of {\fairXplainer} (Line~\ref{fairness_fairXplainer_algo_line:local_regression_start}--\ref{fairness_fairXplainer_algo_line:local_regression_end}, Algorithm~\ref{fairness_fairXplainer_algo:framework}), we use a smoothing operator $\textsc{Smooth}$ (Line~\ref{fairness_fairXplainer_alg_line:backfitting_step}).
 In our experiments, \textit{we use cubic splines as the smoothing operator}.
@@
 In interpolation problems, a B-spline of order $ n $ is traditionally used to smoothen the intersection of piecewise interpolators~\cite{schumaker2007spline}.
-A B-spline of degree $n$ is a piecewise polynomial of degree $ n - 1 $ defined over a variable $ X $.
-Each piecewise term is computed on local points and is aggregated as a global curve smoothly fitting the data.
-The values of $ X $ where the polynomial pieces meet together are called knots, and are denoted by $\{ \dots, t_0, t_1, t_2, \dots\} $.
+A B-spline of degree $n$ is a piecewise polynomial of degree $ n - 1 $ defined over a variable $ Z $.
+Each piece-wise term is computed on local points and is aggregated as a global curve smoothly fitting the data.
+The values of $ Z $ where the polynomial pieces meet together are called knots, and are denoted by $\{ \dots, t_0, t_1, t_2, \dots\} $.
 
-Let $ B_{r, n}(X) $ denote the basis function for a B-spline of order $ n $, and $ r $ is the index of the knot vector.
-According to Carl de Boor~\cite{de1971subroutine}, $ B_{r,1}(X) $, for $ n = 1 $, is defined as
+Let $ B_{r, n}(Z) $ denote the basis function for a B-spline of order $ n $, and $ r $ is the index of the knot vector.
+According to Carl de Boor~\cite{de1971subroutine}, $ B_{r,1}(Z) $, for $ n = 1 $, is defined as
 \begin{align*}
-B_{r,1}(X) = 
+B_{r,1}(Z) = 
 \begin{cases}
-0 &\text{ if } X < t_r \text{ or } X \ge t_{r+1},\\
+0 &\text{ if } Z < t_r \text{ or } Z \ge t_{r+1},\\
 1 &\text{ otherwise}
@@
 
-This definition satisfies $ \sum_i B_{r, 1}(X) = 1 $. The higher order basis functions are defined recursively as
+This definition satisfies $ \sum_i B_{r, 1}(Z) = 1 $. The higher order basis functions are defined recursively as
 \begin{align*}
-	B_{r, n + 1}(X) = p_{r, n}(X)B_{r, n}(X) + (1 - p_{r + 1, n}(X))B_{r + 1, n}(X),
+	B_{r, n + 1}(Z) = p_{r, n}(Z)B_{r, n}(Z) + (1 - p_{r + 1, n}(Z))B_{r + 1, n}(Z),
 \end{align*}
@@
 \begin{align*}
-p_{r,n}(X) = 
+p_{r,n}(Z) = 
 \begin{cases}
-\frac{X - t_r}{t_{r + n} - t_r} &\text{ if } t_{r + n} \ne t_r,\\
+\frac{Z - t_r}{t_{r + n} - t_r} &\text{ if } t_{r + n} \ne t_r,\\
 0 &\text{ otherwise.}
@@
 
-In this paper, we consider cubic splines with the basis function $ B_{r,4}(X) $ that constitutes a B-spline of degree $ 3 $.
+In this paper, we consider cubic splines with the basis function $ B_{r,4}(Z) $ that constitutes a B-spline of degree $ 3 $.
 This polynomial has $ C^2 $ continuity, i.e.
 for each piecewise term, derivatives up to the second order are zero at the endpoints of each interval in the knot vector.
-We estimate component functions $ \function_{\mathbf{a}, \mathbf{S}} $'s with the basis function $ B_{r,4}(\mathbf{X})  $ of cubic splines~\cite{li2010global}, as shown in Equation~\eqref{eq:cubic_splines}.
-\begin{align}\label{eq:cubic_splines}
+We estimate component functions $ \fiffunc_{\mathbf{a}, \mathbf{S}} $'s with the basis function $ B_{r,4}(\mathbf{Z})  $ of cubic splines~\cite{li2010global}, as shown in Equation~\eqref{fairness_fairXplainer_eq:cubic_splines}.
+\begin{align}\label{fairness_fairXplainer_eq:cubic_splines}
 \begin{split}
-&\function_{\mathbf{a}, \{i\}} (\mathbf{X}_{\{i\}}) \approx \sum_{r=-1}^{m+1}\alpha_r^iB_{r,n}(\mathbf{X}_{\{i\}})\\
-&\function_{\mathbf{a}, \{i, j\}} (\mathbf{X}_{\{i, j\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\beta_{pq}^{ij}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})\\
-&\function_{\mathbf{a}, \{i, j, k\}} (\mathbf{X}_{\{i, j, k\}}) \approx \sum_{p=-1}^{m+1}\sum_{q=-1}^{m+1}\sum_{r=-1}^{m+1}\gamma_{pqr}^{ijk}B_p(\mathbf{X}_{\{i\}})B_q(\mathbf{X}_{\{j\}})B_{r,n}(\mathbf{X}_{\{j\}})
+&\fiffunc_{\mathbf{a}, \{i\}} (\mathbf{Z}_{\{i\}}) \approx \sum_{r=-1}^{\tau+1}\alpha_r^iB_{r,n}(\mathbf{Z}_{\{i\}})\\
+&\fiffunc_{\mathbf{a}, \{i, j\}} (\mathbf{Z}_{\{i, j\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\beta_{pq}^{ij}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})\\
+&\fiffunc_{\mathbf{a}, \{i, j, k\}} (\mathbf{Z}_{\{i, j, k\}}) \approx \sum_{p=-1}^{\tau+1}\sum_{q=-1}^{\tau+1}\sum_{r=-1}^{\tau+1}\gamma_{pqr}^{ijk}B_p(\mathbf{Z}_{\{i\}})B_q(\mathbf{Z}_{\{j\}})B_{r,n}(\mathbf{Z}_{\{j\}})
 \end{split}
@@
 
-Here, $ m $ is the number of knots.
-We learn the coefficients $ \alpha, \beta, \gamma $ using the backfitting algorithm (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}, Algorithm~\ref{algo:framework}).
+Here, $ \tau $ is the number of knots, also called spline intervals.
+We learn the coefficients $ \alpha, \beta, \gamma $ using the backfitting algorithm (Line~\ref{fairness_fairXplainer_algo_line:local_regression_start}--\ref{fairness_fairXplainer_algo_line:local_regression_end}, Algorithm~\ref{fairness_fairXplainer_algo:framework}).
+
+$\tau$ is a hyper-parameter that influences the accuracy of the local regression and thus, the FIFs.
+We perform an ablation study to explicate the impact of $\tau$ on the performance of {\fairXplainer} in Appendix~\ref{fairness_fairXplainer_sec:additional_experiments}.
 
@@
 \paragraph{Kernel Smoothing.}
-Kernel smoother estimates a real-valued function $ \function : \mathbb{R}^k \rightarrow \mathbb{R} $ as the weighted average of neighboring observed data.
+Kernel smoother estimates a real-valued function $ \fiffunc : \mathbb{R}^k \rightarrow \mathbb{R} $ as the weighted average of neighboring observed data.
 The weight is defined by the kernel such that closer points are given higher weights.
 The estimated function is smooth, where the level of smoothness is specified by a single parameter.
-Let $ k_{h_\lambda}(\mathbf{X}_0, \mathbf{X}) $ be a kernel defined by
+Let $ k_{h_\lambda}(\mathbf{Z}_0, \mathbf{Z}) $ be a kernel defined by
 \[
-k_{h_\lambda}(\mathbf{X}_0, \mathbf{X}) = D\Big(\frac{||\mathbf{X}_0, \mathbf{X}||}{h_\lambda(\mathbf{X}_0)}\Big)
+k_{h_\lambda}(\mathbf{Z}_0, \mathbf{Z}) = D\Big(\frac{||\mathbf{Z}_0, \mathbf{Z}||}{h_\lambda(\mathbf{Z}_0)}\Big)
 \]
@@
 \begin{itemize}
-	\item $ \mathbf{X}_0, \mathbf{X} \in \mathbb{R}^k $
+	\item $ \mathbf{Z}_0, \mathbf{Z} \in \mathbb{R}^k $
 	\item $ ||\cdot|| $ is Euclidean norm
-	\item $ h_\lambda(\mathbf{X}_0) $ is a parameter known as kernel radius
+	\item $ h_\lambda(\mathbf{Z}_0) $ is a parameter known as kernel radius
 	\item $ D(t) $ is a positive real-valued function, whose value is decreasing (or not increasing) for the increasing distance $ t $
@@
 
+\begin{comment}
+\section{Runtime Complexity of {\fairXplainer}}\label{fairness_fairXplainer_sec:runtime}
+\red{The runtime complexity of {\fairXplainer} is $ \mathcal{O}\Big(ts\sum_{i=1}^{\lambda} {k \choose i}\Big),$
+where $ t $ is the number of iterations of the backfitting algorithm, $ k $ is the number of non-sensitive features, $ \lambda $ is the maximum order of interactions among features, and $ s $ is the runtime complexity of the smoothing oracle\footnote{Typically, the runtime complexity of smoothing oracles, particularly of cubic splines, is linear with respect to $ n $, i.e.
+the number of samples~\cite{toraichi1987computational}.}.
+For example, if we are interested in first and second order interactions among features, the runtime complexities of FairXplainer are $\mathcal{O}(tsk)$ and $\mathcal{O}(tsk^2)$, respectively.
+}
+
+We have experimentally investigate the runtime of {\fairXplainer} (Appendix~\ref{fairness_fairXplainer_sec:runtime_exp}).
+The complexity of {\fairXplainer} relies on both the number of samples and features, and increases with both.
+In all these datasets ($\#$feature upto 26, $\#$samples upto 26048), we compute FIFs of up to order 2 within 25
+seconds. 
+\end{comment}
+
+\section{Computing FIFs for Equalized Odds and Predictive Parity}\label{fairness_fairXplainer_app:pp}
+Due to brevity of space, we elaborate definition of statistical parity and corresponding methodology to compute FIFs in the main text.
+Here, we provide definition of other group-based fairness metrics~\cite{verma2018fairness}: equalized odds and predictive parity.
+We also explain the methodology to use {\fairXplainer} in order to compute FIFs corresponding to these metrics.
+
+Following the classification setting and notations described in Section~\ref{chapter:preliminaries}, here we consider a binary classifier trained on dataset $\mathbf{D}\triangleq \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
+$\widehat{Y} \in \{0,1\}$ and ${Y} \in \{0,1\}$ represents the predicted class and the true class for a data point $ (\nonsensitive, \sensitive) $ with sensitive features $\sensitive$ and non-sensitive features $\nonsensitive$.
+
+\subsection{Equalized Odds}
+
+\textit{Equalized Odds} ($ \mathsf{EO} $)~\cite{hardt2016equality}: \textit{Separation} measuring group fairness metrics such as equalized odds constrain that the predicted class $ \widehat{Y} $ is independent of $ \sensitive $ given the true class $ Y $.
+Formally, for $ Y \in \{0,1\} $, equalized odds is defined as
+\[ f_{\mathsf{EO}}(\alg, \mathbf{D}) \triangleq  \max_{y} \Big(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y]\Big).
+\]
+
+Equalized odds intuitively implies the maximum statistical parity conditioned on $ Y $.
+
+\paragraph{Extension of {\fairXplainer} to Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $.
+Since, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ is the equalized odds of the classifier, we finally report FIFs of features corresponding to the maximum sum of FIFs between $ Y \in \{0, 1\} $.
+
+\subsection{Predictive Parity}
+
+\textit{Sufficiency} measuring group fairness metrics such as Predictive Parity ($ \mathsf{PP} $) constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $.
+Formally,
+\begin{align*}
+	f_{\mathsf{PP}}(\alg, \mathbf{D})  
+	\triangleq \max_{y} \Big(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y]\Big).
+\end{align*}
+
+\paragraph{Extension of {\fairXplainer} to Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
+For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label.
+This contrasts the computation for  statistical parity and equalized odds, where the predicted class $ \widehat{Y} $ is considered as label.
+Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ measures the predictive parity.
+Similar to equalized odds, FIFs achieving the greatest sum of FIFs for $ \widehat{Y} \in \{0, 1\} $ are the reported FIFs for the predictive parity of the classifier.
+
+%In Appendix~\ref{fairness_fairXplainer_sec:additional_experiments}, we provide runtime analysis and accuracy of {\fairXplainer} to compute FIFs corresponding to predictive parity for different datasets and classifiers.
+
+\begin{table}
+	\caption{Median estimation error (over 5-fold cross validation and all combinations of sensitive features) of equalized odds (columns $ 5 $ to $ 7 $) and predictive parity (columns $ 8 $ and $ 9 $) in terms of estimated FIFs by different methods.
+Best results (lowest error) are in bold color.
+`{\textemdash}' denotes timeout.
+SHAP cannot estimate FIFs for predictive parity due to limited methodology.}
+	\label{fairness_fairXplainer_tab:accuracy_eo_pp}
+	\centering
+	%        \setlength{\tabcolsep}{.3em}
+	\begin{tabular}{lrclrrr|rrr}
+		\toprule
+		\multirow{3}{*}{Dataset} & \multirow{3}{*}{\shortstack[1]{Dimension\\$ (n, \numfeatures) $}} & \multirow{3}{*}{\shortstack[1]{Max Sensitive\\Features, $ |\sensitive| $}} &\multirow{3}{*}{Classifier} & \multicolumn{3}{c|}{Equalized Odds} & \multicolumn{3}{c}{Predictive Parity} \\
+		& & & & \multirow{2}{*}{SHAP} & \multicolumn{2}{c|}{\fairXplainer} &  \multicolumn{2}{c}{\fairXplainer}\\
+		& & & & & $ \lambda = 1 $ & $ \lambda = 2 $ & $ \lambda = 1 $ & $ \lambda = 2 $ \\
+		\midrule
+
+		\multirow{4}{*}{Titanic} & \multirow{4}{*}{$ (834, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & $ 1.697 $  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ 0.251 $  & $ \mathbf{0.148} $  \\ 
+		& & & SVM & $ 1.000 $  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ 0.092 $  & $ \mathbf{0.045} $  \\ 
+		& & & Neural Network & \textemdash  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ 0.349 $  & $ \mathbf{0.171} $  \\ 
+		& & & Decision Tree & $ 0.074 $  & $ 0.185 $  & $ \mathbf{0.059} $  & $ \mathbf{0.097} $  & $ \mathbf{0.097} $  \\ 
+		\midrule
+
+		\multirow{4}{*}{German} & \multirow{4}{*}{$ (417, 23) $} & \multirow{4}{*}{$ 2$} 
+		& Logistic Regression & $ 0.382 $  & $ 0.109 $  & $ \mathbf{0.001} $  & $ 0.075 $  & $ \mathbf{0.001} $  \\ 
+		& & & SVM & $ 0.435 $  & $ \mathbf{0.082} $  & \textemdash  & $ 0.060 $  & $ \mathbf{0.001} $  \\ 
+		& & & Neural Network & \textemdash  & $ 0.149 $  & $ \mathbf{0.000} $  & $ 0.184 $  & $ \mathbf{0.000} $  \\ 
+		& & & Decision Tree & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  & $ \mathbf{0.000} $  \\ 
+		\midrule
+
+		\multirow{4}{*}{COMPAS} & \multirow{4}{*}{$ (5771, 8) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & $ 0.380 $  & $ 0.167 $  & $ \mathbf{0.071} $  & $ \mathbf{0.201} $  & $ 0.214 $  \\ 
+		& & & SVM & $ 0.481 $  & $ 0.043 $  & $ \mathbf{0.024} $  & $ 0.124 $  & $ \mathbf{0.117} $  \\ 
+		& & & Neural Network & \textemdash  & $ 0.143 $  & $ \mathbf{0.055} $  & $ \mathbf{0.078} $  & $ 0.129 $  \\ 
+		& & & Decision Tree & $ 0.071 $  & $ 0.069 $  & $ \mathbf{0.031} $  & $ 0.348 $  & $ \mathbf{0.340} $  \\ 
+		\midrule
+
+		\multirow{4}{*}{Adult} & \multirow{4}{*}{$ (26048, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & $ 1.647 $  & $ 0.186 $  & $ \mathbf{0.013} $  & $ 0.090 $  & $ \mathbf{0.002} $  \\ 
+		& & & SVM & $ 0.703 $  & $ 0.081 $  & $ \mathbf{0.001} $  & $ 0.109 $  & $ \mathbf{0.002} $  \\ 
+		& & & Neural Network & \textemdash  & $ 0.077 $  & $ \mathbf{0.000} $  & $ 0.091 $  & $ \mathbf{0.002} $  \\ 
+		& & & Decision Tree & $ \mathbf{0.062} $  & $ 0.263 $  & $ 0.190 $  & $ 0.216 $  & $ \mathbf{0.203} $  \\
+
+		\bottomrule
+	\end{tabular}
+\end{table}
+
 %\clearpage
-\section{Experimental Evaluations}\label{sec:additional_experiments}
+\section{Experimental Evaluations}\label{fairness_fairXplainer_sec:additional_experiments}
+\subsection{Experimental Setup}
+We perform experiments on a Red Hat Enterprise Linux Server release 6.10 (Santiago) that has an $\text{E}5-2690 \text{ v}3$ CPU and 16GB of RAM.
+With the aim of computing FIFs for any classifier, we do not adjust the classifier's hyper-parameters during training.
+Instead, we utilize the default hyper-parameters provided by Scikit-learn~\cite{scikit-learn}.
+For equalized odds and predictive parity, {\fairXplainer}  (similarly SHAP) is deployed twice.
+Hence, we double the time limit, $ 2*300 = 600 $ seconds.
 
-\subsection{Experimental Setup}
-We conduct experiments on Red Hat Enterprise Linux Server release 6.10 (Santiago) with $ \text{E}5-2690 \text{ v}3 $ CPU and $ 8 $GB of RAM.
-Since our objective is to compute FIFs for any classifier, we do not perform any tuning of hyper-parameters during training.
-We use the default hyper-parameter choices available in Scikit-learn~\cite{scikit-learn}.
-The only parameter that we choose in {\framework} is the maximum order of intersectionality, namely $ \lambda $.
-We present the ablation study w.r.t.\ $ \lambda $ in Appendix~\ref{sec:ablation}.
+\subsection{Accurate Approximation of Equalized Odds and Predictive Parity using FIFs.}
+The accuracy of approximating equalized odds and predictive parity for {\fairXplainer} and SHAP is compared in Table~\ref{fairness_fairXplainer_tab:accuracy_eo_pp}.
+{\fairXplainer} shows lower estimation error compared to SHAP, especially when $\lambda=2$.
+SHAP is unable to explain predictive parity as predictive parity relies on the ground label $Y$, which is not available for randomly generated data points by SHAP for estimating local explanations.
 
-\subsection{Runtime of {\framework}} 
-In Figure~\ref{fig:runtime}, we report the runtime of {\framework} in computing FIFs (with $ \lambda = 2 $) for different group fairness metrics on four classifiers and five datasets.
+\begin{figure}
+	\subfloat[Equalized Odds]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_eo}} 
+	\subfloat[Predictive Parity]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_suff}}
+	\caption{Execution time of different methods for estimating FIFs for equalized odds and predictive parity.
+{\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
+SHAP cannot explain predictive parity.
+	}
+	\label{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}
+\end{figure}
+
+\subsection{Execution Time for FIF Estimation of Equalized Odds and Predictive Parity}
+In Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}, we demonstrate the execution time of different methods in estimating FIFs of equalized odds and predictive parity in cactus plots.
+{\fairXplainer} with $ \lambda = 1 $ is more efficient than $ \lambda = 2 $ and solves all $ 480 $ fairness instances with at least one order of magnitude less execution time.
+Compared to {\fairXplainer} ($ \lambda = 1 $), SHAP demonstrates less computational efficiency.
+
+\begin{figure}
+	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_error.pdf}}
+	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_mean_rmse.pdf}}
+	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_time.pdf}}
+
+	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_error.pdf}}
+	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_mean_rmse.pdf}}
+	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_time.pdf}}
+
+	\caption{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
+	\label{fairness_fairXplainer_fig:effect_spline_intervals_appendix}
+\end{figure}
+
+\subsection{Ablation Study: Effect of Spline Intervals} 
+To understand the impact of spline intervals $\tau$ on {\fairXplainer}, we conduct an experiment.
+$\tau$ determines the number of local points to include in the cubic-spline based smoothing, with higher values providing better approximation of the component functions in the set-additive decomposition of the classifier (ref.\ Eq.~\eqref{fairness_fairXplainer_eq:set_additive_classifier}).
+As shown in Figure~\ref{fairness_fairXplainer_fig:effect_spline_intervals_appendix}, as $\tau$ increases, the approximation error of statistical parity based on FIFs decreases as well as the root mean square error of the set-additive approximation of the classifier.
+On the other hand, with higher $ \tau $, the execution time of {\fairXplainer} increases.
+Therefore, \emph{$ \tau $ exhibits a trade-off between the estimation accuracy and execution time of {\fairXplainer}.}
+
+\begin{comment}
+\subsection{Runtime of {\fairXplainer}} \label{fairness_fairXplainer_sec:runtime_exp}
+In Figure~\ref{fairness_fairXplainer_fig:runtime}, we report the runtime of {\fairXplainer} in computing FIFs (with $ \lambda = 2 $) for different group fairness metrics on four classifiers and five datasets.
 The dimension of the dataset, measured by the number of samples and the number of features, is a key factor for runtime.
@@
 In contrast, German dataset has the maximum features ($ = 26 $).
-Consequently, {\framework} requires more runtime in both Adult and German dataset compared to others.
-However, \emph{in all datasets, FIF computation takes at most $ 25 $ seconds, that demonstrates the efficiency of {\framework} in practical fairness problems.}
+Consequently, {\fairXplainer} requires more runtime in both Adult and German dataset compared to others.
+However, \emph{in all datasets, FIF computation takes at most $ 25 $ seconds, that demonstrates the efficiency of {\fairXplainer} in practical fairness problems.}
+\end{comment}
 
-\begin{figure}[h!]
+\begin{comment}
+\begin{figure}
 	\centering
-	\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/sp_train_time_detailed_fairXplainer_only}}
-	\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/eo_train_time_detailed_fairXplainer_only}}\\
-	\subfloat[Predictive parity]{\includegraphics[scale=0.39]{figures/suff_train_time_detailed_fairXplainer_only}}
-	\caption{Runtime of {\framework} (in seconds) for computing FIFs in different datasets.
+	\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/fairness/fif/sp_train_time_detailed_fairXplainer_only}}
+	\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/fairness/fif/eo_train_time_detailed_fairXplainer_only}}\\
+	\subfloat[Predictive parity]{\includegraphics[scale=0.39]{figures/fairness/fif/suff_train_time_detailed_fairXplainer_only}}
+	\caption{Runtime of {\fairXplainer} (in seconds) for computing FIFs in different datasets.
 For each dataset, we mention its dimension as a tuple (\# sample, \# features) in the xticks.
-As the dimension increases, we observe higher runtime of {\framework}.}\label{fig:runtime}\vspace*{-1em}
+As the dimension increases, we observe higher runtime of {\fairXplainer}.}\label{fairness_fairXplainer_fig:runtime}\vspace*{-1em}
+\end{figure}
+\end{comment}
+
+\begin{figure}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_error.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_mean_rmse.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_time.pdf}}
+
+	\caption{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
+	\label{fairness_fairXplainer_fig:effect_maxorder_appendix}
 \end{figure}
 
-\subsection{Ablation Study: Effect of Maximum Order of Intersectionality $ \lambda $}\label{sec:ablation} 
-We test the effect of $ \lambda \in \{1, 2, 3\} $ on the runtime of {\framework}.
-For each $ \lambda $, we consider $ 540 $ fairness instances ($ 5 $ datasets with $ 27 $ different combinations of sensitive features $ \times $ $ 5 $-fold cross validation $ \times $ $ 4 $ classifiers).
-We plot the corresponding results in the cumulative distribution plot of  Figure~\ref{fig:varying_max_order}.
-Here, $ X $-axis denotes the runtime (in seconds) and $ Y $-axis denotes the total number of solved instances, i.e.
-a point $ (x,y) $ denotes that $ y $ number of instances are solved within $ x $ seconds.
+\subsection{Ablation Study: Effect of Maximum Order of Intersectionality}\label{fairness_fairXplainer_sec:ablation} 
+Figure~\ref{fairness_fairXplainer_fig:effect_maxorder_appendix} examines the impact of the maximum order of intersectionality ($\lambda$) on {\fairXplainer} in terms of accuracy and execution time on Ricci~\cite{mcginley2010ricci}, Titanic, and COMPAS datasets.
+As $\lambda$ increases, we see a decrease in approximation error for statistical parity based on FIFs, a decrease in the root mean squared error of the classifier's set additive decomposition, and an increase in execution time across different datasets.
+This means $\lambda$ provides a trade-off between accuracy and efficiency in {\fairXplainer}.
 
-As we increase from $ \lambda = 1 $ to $ \lambda = 2 $ and then to $ \lambda = 3 $, {\framework} requires around $ 0.5 $ and $ 2 $ orders of magnitude more runtime, respectively, to solve the equal number of instances.
-This is due to the fact that with increase in $ \lambda $, there is a combinatorial explosion in the number of component functions in the backfitting algorithm.
-Thus, \emph{computing higher-order FIFs requires a higher runtime of {\framework}}.
-
+\begin{comment}
 \begin{figure}[t!]
 	\centering
-	{\includegraphics[scale=0.38]{figures/vary_max_order_train}}
-	%	\subfloat[]{\includegraphics[scale=0.4]{figures/vary_max_order_test}}	
-	\caption{Effect on runtime while varying max-order $ \lambda $ in {\framework}.
-Higher $ \lambda $ incurs more a computational effort, hereby solving less fairness instances within the same time limit.}\label{fig:varying_max_order}
+	{\includegraphics[scale=0.38]{figures/fairness/fif/vary_max_order_train}}
+	%	\subfloat[]{\includegraphics[scale=0.4]{figures/fairness/fif/vary_max_order_test}}	
+	\vspace{-1em}
+	\caption{Effect on runtime while varying max-order $ \lambda $ in {\fairXplainer}.
+Higher $ \lambda $ incurs more a computational effort, hereby solving less fairness instances within the same time limit.}\label{fairness_fairXplainer_fig:varying_max_order}
 \end{figure}
+\end{comment}
 
+\begin{comment}
+
+\begin{table*}    
+	\caption{Median estimation error of equalized odds, $ |\mathsf{EO} - \widehat{\mathsf{EO}}| $, in terms of estimated FIFs by different methods (columns $ 5 $ to $ 7 $, less is better).
+Best results are in bold color.
+{\textemdash} denotes timeout in estimating FIFs.}
+%	\label{fairness_fairXplainer_tab:fair_algo_verification}   
+	\centering
+	%        \setlength{\tabcolsep}{.3em}
+	\begin{tabular}{lrclrrrr}
+		\toprule
+		\multirow{2}{*}{Dataset} & \multirow{2}{*}{\shortstack[1]{Dimension\\$ (n, \numfeatures) $}} & \multirow{2}{*}{\shortstack[1]{Maximum\\Sensitive Features}} &\multirow{2}{*}{Classifier} & \multirow{2}{*}{SHAP} & \multicolumn{2}{c}{\fairXplainer}\\
+		& & & & & $ \lambda = 1 $ & $ \lambda = 2 $\\
+		\midrule
+
+		\multirow{4}{*}{Ricci} & \multirow{4}{*}{$ (94, 5) $} & \multirow{4}{*}{$ 2$} 
+		& Logistic Regression &  $ 0.568 $ &  $ 0.002 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & SVM &  $ 0.388 $ &  $ 0.003 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & Neural Network &  $ \mathbf{0.0} $ &  $ 1.0 $ &  $ 1.0 $ &  \\ 
+		& & & Decision Tree &  $ \mathbf{0.0} $ &  $ 1.0 $ &  $ 1.0 $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{Titanic} & \multirow{4}{*}{$ (834, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression &  $ 1.697 $ &  $ \mathbf{0.0} $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & SVM &  $ 1.0 $ &  $ \mathbf{0.0} $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ \mathbf{0.0} $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Decision Tree &  $ 0.074 $ &  $ 0.185 $ &  $ \mathbf{0.059} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{German} & \multirow{4}{*}{$ (417, 23) $} & \multirow{4}{*}{$ 2$} 
+		& Logistic Regression &  $ 0.382 $ &  $ 0.109 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & SVM &  $ 0.435 $ &  $ \mathbf{0.082} $ & \textemdash &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.149 $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Decision Tree &  $ \mathbf{0.0} $ &  $ 1.0 $ &  $ 1.0 $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{COMPAS} & \multirow{4}{*}{$ (5771, 8) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression &  $ 0.38 $ &  $ 0.167 $ &  $ \mathbf{0.071} $ &  \\ 
+		& & & SVM &  $ 0.481 $ &  $ 0.043 $ &  $ \mathbf{0.024} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.143 $ &  $ \mathbf{0.055} $ &  \\ 
+		& & & Decision Tree &  $ 0.071 $ &  $ 0.069 $ &  $ \mathbf{0.031} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{Adult} & \multirow{4}{*}{$ (26048, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression &  $ 1.647 $ &  $ 0.186 $ &  $ \mathbf{0.013} $ &  \\ 
+		& & & SVM &  $ 0.703 $ &  $ 0.081 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.077 $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Decision Tree &  $ \mathbf{0.062} $ &  $ 0.263 $ &  $ 0.2 $ &  \\
+
+		\bottomrule
+	\end{tabular}
+\end{table*}
+
+\begin{table*}    
+	\caption{Median estimation error of predictive parity, $ |\mathsf{PP} - \widehat{\mathsf{PP}}| $, in terms of estimated FIFs by different methods (columns $ 5 $ to $ 7 $, less is better).
+Best results are in bold color.
+{\textemdash} denotes timeout in estimating FIFs.}
+	%	\label{fairness_fairXplainer_tab:fair_algo_verification}   
+	\centering
+	%        \setlength{\tabcolsep}{.3em}
+	\begin{tabular}{lrclrrrr}
+		\toprule
+		\multirow{2}{*}{Dataset} & \multirow{2}{*}{\shortstack[1]{Dimension\\$ (n, \numfeatures) $}} & \multirow{2}{*}{\shortstack[1]{Maximum\\Sensitive Features}} &\multirow{2}{*}{Classifier} & \multirow{2}{*}{SHAP} & \multicolumn{2}{c}{\fairXplainer}\\
+		& & & & & $ \lambda = 1 $ & $ \lambda = 2 $\\
+		\midrule
+
+		\multirow{4}{*}{Ricci} & \multirow{4}{*}{$ (94, 5) $} & \multirow{4}{*}{$ 2$} 
+		& Logistic Regression & \textemdash &  $ 0.006 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & SVM & \textemdash &  $ 0.018 $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ \mathbf{1.0} $ &  $ \mathbf{1.0} $ &  \\ 
+		& & & Decision Tree & \textemdash &  $ \mathbf{1.0} $ &  $ \mathbf{1.0} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{Titanic} & \multirow{4}{*}{$ (834, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & \textemdash &  $ 0.251 $ &  $ \mathbf{0.148} $ &  \\ 
+		& & & SVM & \textemdash &  $ 0.092 $ &  $ \mathbf{0.049} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.349 $ &  $ \mathbf{0.171} $ &  \\ 
+		& & & Decision Tree & \textemdash &  $ \mathbf{0.097} $ &  $ \mathbf{0.097} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{German} & \multirow{4}{*}{$ (417, 23) $} & \multirow{4}{*}{$ 2$} 
+		& Logistic Regression & \textemdash &  $ 0.075 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & SVM & \textemdash &  $ 0.06 $ &  $ \mathbf{0.001} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.184 $ &  $ \mathbf{0.0} $ &  \\ 
+		& & & Decision Tree & \textemdash &  $ \mathbf{1.0} $ &  $ \mathbf{1.0} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{COMPAS} & \multirow{4}{*}{$ (5771, 8) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & \textemdash &  $ \mathbf{0.201} $ &  $ 0.214 $ &  \\ 
+		& & & SVM & \textemdash &  $ 0.124 $ &  $ \mathbf{0.117} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ \mathbf{0.078} $ &  $ 0.129 $ &  \\ 
+		& & & Decision Tree & \textemdash &  $ 0.348 $ &  $ \mathbf{0.34} $ &  \\ 
+		\midrule
+
+		\multirow{4}{*}{Adult} & \multirow{4}{*}{$ (26048, 11) $} & \multirow{4}{*}{$ 3$} 
+		& Logistic Regression & \textemdash &  $ 0.09 $ &  $ \mathbf{0.002} $ &  \\ 
+		& & & SVM & \textemdash &  $ 0.109 $ &  $ \mathbf{0.002} $ &  \\ 
+		& & & Neural Network & \textemdash &  $ 0.091 $ &  $ \mathbf{0.002} $ &  \\ 
+		& & & Decision Tree & \textemdash &  $ 0.216 $ &  $ \mathbf{0.154} $ &  \\
+
+		\bottomrule
+	\end{tabular}
+\end{table*}
+
+\end{comment}
+
+\begin{comment}
 \subsection{Accuracy in Estimating Different Fairness Metrics}
-We compare and evaluate the accuracies of {\framework} and SHAP in estimating a fairness metric, computed as the sum of FIFs (Axiom~\ref{axm:additivity}).
-We illustrate the results with a scatter-plot in Figure~\ref{fig:accuracy_scatter_plot}.
+We compare and evaluate the accuracies of {\fairXplainer} and SHAP in estimating a fairness metric, computed as the sum of FIFs.
+We illustrate the results with a scatter-plot in Figure~\ref{fairness_fairXplainer_fig:accuracy_scatter_plot}.
 In the plot, each dot represents the estimation of a fairness metric on a dataset and a classifier.
 
-The majority of dots for {\framework} are on the line connecting $ (0, 0) $ to $ (1, 1) $  in Figure~\ref{fig:accuracy_scatter_plot}.
-This implies that the estimation of {\framework} is significantly accurate.
-Interestingly, error rarely occurs in {\framework} due to the degenerate cases mentioned in Section~\ref{sec:fifs}.
+The majority of dots for {\fairXplainer} are on the line connecting $ (0, 0) $ to $ (1, 1) $  in Figure~\ref{fairness_fairXplainer_fig:accuracy_scatter_plot}.
+This implies that the estimation of {\fairXplainer} is significantly accurate.
+Interestingly, error rarely occurs in {\fairXplainer} due to the degenerate cases mentioned in Section~\ref{fairness_fairXplainer_sec:fifs}.
 In contrast, SHAP often computes a metric inaccurately, thereby showing the drawback of applying a local explanation approach for FIF computation.
-\emph{Thus, {\framework} is significantly more accurate than SHAP in computing FIFs and the corresponding fairness metrics.}
+\emph{Thus, {\fairXplainer} is significantly more accurate than SHAP in computing FIFs and the corresponding fairness metrics.}
 
@@
 	\centering
-	\subfloat[Statistical parity]{\includegraphics[scale=0.35]{figures/scatter_plot_sp_train_accuracy}}
-	\subfloat[Equalized odds]{\includegraphics[scale=0.35]{figures/scatter_plot_eo_train_accuracy}}
-	\subfloat[Predictive parity]{\includegraphics[scale=0.35]{figures/scatter_plot_suff_train_accuracy}}
-	\caption{Comparison between the estimated and exact value of a fairness metric, computed as the sum of FIFs according to Axiom~\ref{axm:additivity}.
-Results are presented separately for {\framework} and SHAP.
+	\subfloat[Statistical parity]{\includegraphics[scale=0.35]{figures/fairness/fif/scatter_plot_sp_train_accuracy}}
+	\subfloat[Equalized odds]{\includegraphics[scale=0.35]{figures/fairness/fif/scatter_plot_eo_train_accuracy}}
+	\subfloat[Predictive parity]{\includegraphics[scale=0.35]{figures/fairness/fif/scatter_plot_suff_train_accuracy}}
+	\caption{Comparison between the estimated and exact value of a fairness metric, computed as the sum of FIFs according to Axiom~\ref{fairness_fairXplainer_axm:additivity}.
+Results are presented separately for {\fairXplainer} and SHAP.
 Each dot represents the result for a fairness metric, carried over multiple datasets and classifiers.
 Ideally, if a dot lies on the line connecting $ (0,0) $ to $ (1,1) $, it denotes a correct fairness estimate.
-{\framework} demonstrates significantly better accuracy of estimation than SHAP.
-For predictive parity metric, only {\framework} allows associated FIFs computation.}
-	\label{fig:accuracy_scatter_plot}
+{\fairXplainer} demonstrates significantly better accuracy of estimation than SHAP.
+For predictive parity metric, only {\fairXplainer} allows associated FIFs computation.}
+	\label{fairness_fairXplainer_fig:accuracy_scatter_plot}
 \end{figure}
 
+\end{comment}
+
 \subsection{FIF of Different Datasets}
-We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult, Ricci, and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fig:individual_vs_intersectional_influence_adult},~\ref{fig:individual_vs_intersectional_influence_ricci}, and \ref{fig:individual_vs_intersectional_influence_titanic}, respectively.
-In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Section~\ref{sec:experiments}.
+We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult} and \ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}, respectively.
+In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Section~\ref{fairness_fairXplainer_sec:experiments}.
 
-\textbf{In Adult dataset,} the classifier predicts whether an individual earns more than $ \$50 $k per year or not, where `race' and `sex' are sensitive features.
+\textbf{In Adult dataset,} the classifier predicts whether an individual earns more than $ \$50 $k per year or not, where race and sex are sensitive features.
 We observe that the trained network is unfair and it demonstrates statistical parity as $ 0.23 $.
-As we analyze FIFs,  \textit{education number}, \textit{age}, and \textit{capital gain/loss} are key features responsible for the bias.
-Unlike COMPAS dataset (Figure~\ref{fig:individual_and_intersectional_fifs}), there does not exist any feature or subset of features in Adult dataset that potentially reduces the bias.
+As we analyze FIFs,  education number, age, and capital gain/loss are key features responsible for the bias.
 
@@
 		\centering
-		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/adult_feature_weight_unsorted}\label{fig:individual_fifs_adult}}
+		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_adult}}
 	\end{minipage}
@@
 		\centering
-		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/adult_feature_weight}\label{fig:individual_and_intersectional_fifs_adult}}
+		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_adult}}
 	\end{minipage}
 	\vspace{-0.2em}
-	\caption{FIFs for Adult dataset on explaining statistical parity. }\label{fig:individual_vs_intersectional_influence_adult}
-
+	\caption{FIFs for Adult dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult}
 \end{figure}
 
+\begin{comment}
 \textbf{In Ricci dataset,} the classification task  is to predict whether a firefighter obtains promotion or not in New Haven, Connecticut administered exams.
@@
 In addition, the higher order influences demonstrate a higher value (FIF $ = 0.26 $), meaning that the features are highly correlated and act in favor of bias.
-
-\begin{figure}[h!]
+\begin{figure}
 	\begin{minipage}{0.45\textwidth}
 		\centering
-		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/ricci_feature_weight_unsorted}\label{fig:individual_fifs_ricci}}
+		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/ricci_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_ricci}}
 	\end{minipage}
@@
 		\centering
-		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/ricci_feature_weight}\label{fig:individual_and_intersectional_fifs_ricci}}
+		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/ricci_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_ricci}}
 	\end{minipage}
 	\vspace{-0.2em}
-	\caption{FIFs for Ricci dataset on explaining statistical parity.}\label{fig:individual_vs_intersectional_influence_ricci}
-
+	\caption{FIFs for Ricci dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_ricci}
 \end{figure}
 
+\end{comment}
+
 \textbf{In Titanic dataset,} the neural network predicts whether a person survives the Titanic shipwreck or not.
-In this experiment, we consider the `sex' of a person as a sensitive feature and observe that the classifier is highly unfair (statistical parity as $ 0.83 $) on the basis of `sex'.
-In the computed FIFs, most of the features except the \textit{passenger class} and the \textit{age} of a person increases the bias of the classifier.
+In this experiment, we consider the sex of a person as a sensitive feature and observe that the classifier is highly unfair achieving statistical parity as $ 0.83 $.
+Our FIF analysis reveals high correlation in Titanic, where individual FIFs are mostly zero while intersectional FIFs achieve high absolute values.
 
-\begin{figure}[h!]
-	\begin{minipage}{0.45\textwidth}
+\begin{figure}
 	\centering
-		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/titanic_feature_weight_unsorted}\label{fig:individual_fifs_titanic}}
-	\end{minipage}
-	\begin{minipage}{0.5\textwidth}
-		\centering
-		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/titanic_feature_weight}\label{fig:individual_and_intersectional_fifs_titanic}}
-	\end{minipage}
-	\vspace{-0.2em}
-	\caption{FIFs for Titanic dataset on explaining statistical parity.}\label{fig:individual_vs_intersectional_influence_titanic}
+	\subfloat[Individual FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_titanic}}\hfil
+	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_titanic}}
+	\caption{FIFs for Titanic dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}
+\end{figure}
 
+\begin{comment}
+
+\begin{figure}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_1000_0_0_2_0_5_pearson_correlation_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_1000_0_0_2_0_5_p-value_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_1000_0_0_2_0_5_scatter_accuracy_.pdf}}
+
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_5000_0_0_2_0_5_pearson_correlation_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_5000_0_0_2_0_5_p-value_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_5000_0_0_2_0_5_scatter_accuracy_.pdf}}
+
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_10000_0_0_2_0_5_pearson_correlation_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_10000_0_0_2_0_5_p-value_.pdf}}
+	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/fairXplainer_sanity_check_10000_0_0_2_0_5_scatter_accuracy_.pdf}}
 \end{figure}
+
+\end{comment}
+
+\begin{comment}
+\begin{figure}[t!]
+	\subfloat[COMPAS]{\includegraphics[scale=0.35]{figures/fairness/fif/intervention_compas_intersectional}}\hfil
+	\subfloat[Adult]{\includegraphics[scale=0.35]{figures/fairness/fif/intervention_adult_intersectional}}
+	\caption{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
+Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}
+	\label{fairness_fairXplainer_fig:fairness_intervention_with_intersectional}
+\end{figure}
+
+\end{comment}
+
+\begin{comment}
+
+\begin{figure}
+	\subfloat[{\fairXplainer} ($ \lambda = 1 $)]{\includegraphics[scale=0.38]{figures/fairness/fif/intervention_individual}\label{fairness_fairXplainer_fig:intervention_individual}}
+	\subfloat[{\fairXplainer} ($ \lambda = 2 $)]{\includegraphics[scale=0.38]{figures/fairness/fif/intervention_intersectional}\label{fairness_fairXplainer_fig:intervention_intersectional}}
+	\caption{\red{[In Appendix]} Intervention results. \red{Diff with QII, Yair ZIck's paper: data vs classifier intervention.}}
+\end{figure}
+
+\end{comment}

### 2023-02-07 chapters/fairness/fif/conclusion.tex
adding fairXplainer

--- 2022-06-01 chapters/fairness/fif/conclusion.tex
@@
 %\vspace{-1em}
-We propose the Fairness Influence Function (FIF) to quantify the contribution of input features on the resulting bias of a classifier for a given dataset.
-Relying on an additive axiom, we express group fairness metrics computed for sensitive groups as the sum of FIFs of all the subsets of non-sensitive features.
-To compute FIFs, we first prove existing group fairness metrics as the scaled difference of the conditional variances in the predictions of the classifier and then apply variance decomposition based on global sensitivity analysis.
-Finally, we propose {\framework}, an algorithm for efficiently and accurately computing FIFs by deploying a local regression to learn a set-additive decomposition of the classifier.
-The experimental results show that {\framework} estimates bias with significantly higher accuracy than the local explanation based approach SHAP.
-Also, {\framework} computes both individual and intersectional FIFs unlike SHAP to yield a better understanding of the sources of bias.
-In future,  we aim to develop algorithms that leverage FIFs to yield unbiased decisions.
+
+In this paper, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset.
+Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning.
+Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity.
+We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression.
+In experiments, {\fairXplainer} estimates individual and intersectional FIFs in real-world datasets and classifiers, approximates bias using FIFs with less estimation error than earlier methods, demonstrates a high correlation between FIFs and fairness interventions, and analyzes the impact of fairness enhancing and attack algorithms on FIFs.
+The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias.
+Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier.
+Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.
+%

### 2023-02-07 chapters/fairness/fif/experiments.tex
minor edits
adding fairXplainer

--- 2022-06-02 chapters/fairness/fif/experiments.tex
@@
-%\clearpage
-\section{Empirical Performance Analysis}\label{sec:experiments}
-In this section, we perform an empirical evaluation of {\fairXplainer}.
-In the following, we discuss the experimental setup, the objectives of experiments, and experimental results.
-{ We adjoin the source code and additional experimental results on the effects of maximum order $\lambda$, runtime, accuracy on different datasets and corresponding explanations (Appendix~\ref{sec:additional_experiments}) to the supplementary material.}
+\section{Empirical Performance Analysis}\label{fairness_fairXplainer_sec:experiments}
+In this section, we perform an empirical evaluation of {\fairXplainer}. Particularly, we discuss the experimental setup, the objectives of experiments, and experimental results.
 
-\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.8 $).
-In {\fairXplainer}, we deploy SAlib library~\cite{Herman2017} to compute FIFs leveraging techniques of global sensitivity analysis.
-In our experiments, we consider five widely studied datasets from fairness literature, namely COMPAS~\cite{angwin2016machine}, Adult~\cite{DK2017uci}, German-credit~\cite{DK2017}, Ricci~\cite{mcginley2010ricci}, and Titanic (\url{https://www.kaggle.com/c/titanic}).
-We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: Logistic Regression, Support Vector Machine, Decision Tree, and Neural Networks.
-In experiments, we specify {\fairXplainer} to compute intersectional influences up to second order ($ \lambda = 2 $).
-We compare {\fairXplainer} with Shapley-valued based FIF computational framework, referred as SHAP~\cite{lundberg2020explaining}.
-In addition, we deploy {\fairXplainer} along with different fairness-enhancing~\cite{kamiran2012data} and fairness attack~\cite{solans2020poisoning} algorithms, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric.
-In the following, we discuss the objective of our empirical study.
+\paragraph{Experimental Setup.} We implement a prototype of {\fairXplainer} in Python (version $ 3.7.6 $).
+To estimate FIFs, we leverage and modify the `HDMR' module in SALib library~\cite{Herman2017} based on global sensitivity analysis.
+In experiments, we consider four widely studied datasets from fairness literature, namely German-credit~\cite{DK2017},
+% Ricci~\cite{mcginley2010ricci}, 
+Titanic (\url{https://www.kaggle.com/c/titanic}), COMPAS~\cite{angwin2016machine}, and Adult~\cite{DK2017uci}.
+We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: logistic regression classifier, support vector machine (SVM), neural network, and decision tree with $ 5 $-fold cross-validation.
+In experiments, we specify {\fairXplainer} to compute intersectional influences up to the second order ($ \lambda = 2 $).
+While applying cubic-spline based local regression in {\fairXplainer}, we set $ \tau $, the number of spline intervals to $ 6 $.
+We compare {\fairXplainer} with the existing Shapley-valued based FIF computational framework (\url{https://shorturl.at/iqtuX}), referred as SHAP~\cite{lundberg2020explaining}.
+For both {\fairXplainer} and SHAP, we set a timeout of $ 300 $ seconds for estimating FIFs.
+In addition, we deploy {\fairXplainer} along with a fairness-enhancing algorithm~\cite{kamiran2012data} and a fairness attack~\cite{solans2020poisoning} algorithm, and analyze the effect of these algorithms on the FIFs and the resultant fairness metric.
+In the following, we discuss the objectives of our empirical study.
 
+\setlength{\textfloatsep}{4pt}% Remove \textfloatsep
 \begin{enumerate}
 %\begin{enumerate}
-	\item How do \textbf{accuracies} of {\fairXplainer} and SHAP compare for estimating FIFs and resulting bias?
-	\item What is the impact of\textbf{ intersectional vs.\ individual FIFs} in tracing the sources of bias?
-    \item How do FIFs change while \textbf{applying} different fairness enhancing algorithms, i.e. \textbf{affirmative actions}, and fairness attacks, i.e. \textbf{punitive actions}?
+	\item \textit{Performance:} How \textbf{accurate and computationally efficient} {\fairXplainer} and SHAP are in approximating the bias of a classifier based on estimated FIFs?
+	\item \textit{Functionality:} How do FIFs estimated by {\fairXplainer} and SHAP \textbf{correlate with the impact a fairness intervention} strategy on features?
+%    \item \textit{Ablation study:} What are the effects of different \textbf{hyper-parameters} on accuracy and efficiency of {\fairXplainer}?
+	\item \textit{Granularity of explanation:} How effective are the \textbf{ intersectional FIFs} in comparison with the \textbf{individual FIFs} while tracing the sources of bias?
+	\item \textit{Application:} How do FIFs quantify the impact of \textbf{applying} different fairness enhancing algorithms, i.e.
+\textbf{affirmative actions}, and fairness attacks, i.e.
+\textbf{punitive actions}?
  %\item \red{What is the effect of $ \lambda $ on the runtime of {\fairXplainer}?}
@@
 
-In summary, \textit{{\fairXplainer} achieves significantly less estimation error than SHAP in all the datasets.} This shows the importance of adopting global sensitivity analysis to compute FIFs for group fairness metrics than local explanation approaches.
-Moreover, \textit{{\fairXplainer} effectively traces the sources of bias by computing intersectional FIFs, which earlier method like SHAP {did not capture}.} {\fairXplainer} also detects the effects of the affirmative and punitive actions on the bias of a classifier and the corresponding tensions between different subsets of features.
-%Due to brevity of space, we defer additional experimental results to the Supplementary material.
-%In the following, we elaborate results on estimation error and individual vs.\ intersectional FIFs.
+In summary, (1) we observe that {\fairXplainer} yields \textit{less estimation error} than SHAP while approximating statistical parity using FIFs.
+{\fairXplainer} incurs lower execution time, i.e.
+\textit{better efficiency in computing individual FIFs} than SHAP while also enabling computation of intersectional FIFs for real-world datasets and classifiers.
+(2) While considering a fairness intervention, i.e.
+change in bias due to omission of a feature, FIFs estimated by {\fairXplainer} have \textit{higher correlation with increase/decrease in bias due to  the intervention than SHAP}.
+%importantly by top most features with highest absolute FIFs.
+Thus, {\fairXplainer} demonstrates to be \textit{a better choice for identifying the features influencing the group fairness metrics} than SHAP. 
+%(3) In the ablation study, {\fairXplainer} demonstrates the trade-off between estimation accuracy and execution time through hyper-parameters $ \tau $ and $ \lambda $. 
+(3) \textit{By quantifying both the individual and intersectional influences of features, \fairXplainer{} leads to a more accurate and granular interpretation of the sources of bias}, which is absent in earlier bias explaining methods like SHAP.
+(4) Finally, as an application of the FIF formulation, {\fairXplainer} also detects the effects of the affirmative and punitive actions on the bias of a classifier and the corresponding tensions between different subsets of features.
+Here, we elaborate on experimental results, and defer additional experiments such as applying {\fairXplainer} on other fairness metrics: equalized odds and predictive parity, and an ablation study of hyper-parameters: maximum order of intersectionality  $ \lambda $ and spline intervals $ \tau $ to Appendix~\ref{fairness_fairXplainer_sec:additional_experiments}.
 
-\begin{figure}[!t]
-	\begin{minipage}{0.49\textwidth}
+\input{chapters/fairness/fif/table_performance.tex}
+
+\subsection{Performance and Functionality in Estimating FIFs}
+\paragraph{Accurate Approximation of Bias with FIFs.} We compare {\fairXplainer} with SHAP in estimating statistical parity by summing all FIFs, as dictated by the decomposability property (Theorem~\ref{fairness_fairXplainer_thm:fif_property}).
+{To our best knowledge, the ground truth of FIF is not known for real-world datasets and classifiers.
+As such, we cannot compare accuracy of {\fairXplainer} and SHAP directly on the estimated FIFs.
+Since both methods follow the decomposability property, one way to compare them is to test accuracy of the sum of estimated FIFs yielding bias/unfairness, as the ground truth of bias of a classifier can be exactly calculated for a given dataset~\cite{aif360-oct-2018}.}
+We compute estimation error by taking the absolute difference between the exact and  estimated values of statistical parity, and present median results in Table~\ref{fairness_fairXplainer_tab:fair_algo_verification}.
+
+In general, {\fairXplainer} achieves less estimation with $ \lambda = 2 $ than with $ \lambda = 1 $ in all datasets and classifiers.
+This implies that combining intersectional FIFs with individual FIFs captures bias more accurately than the individual FIFs alone.
+In each dataset, {\fairXplainer} ($\lambda = 2$) demonstrates less estimation error than SHAP in all classifiers except in decision trees, denoting that GSA based approach {\fairXplainer} is more accurate in approximating group fairness metrics through FIFs than the local explainer SHAP.
+In decision trees, {\fairXplainer}\textemdash which is model-agnostic in methodology\textemdash with $ \lambda = 2 $ often demonstrates a comparable accuracy with SHAP, especially the optimized tree-based explanation counterpart of SHAP~\cite{lundberg2020local2global}.
+In the context of neural networks, SHAP, particularly Kernel-SHAP, often fails to estimate FIFs within the provided time-limit, while {\fairXplainer} with $ \lambda = 2 $ yields highly accurate estimates (median estimation error between $ 0$ to $0.053 $).
+Therefore, we conclude that \emph{{\fairXplainer} is more accurate in estimating statistical parity using FIFs than SHAP.}
+
+\begin{figure}[t!]
 	\centering
-		\subfloat[Statistical parity]{\includegraphics[scale=0.39]{figures/fairness/fif/sp_train_accuracy}}
+		\begin{minipage}{0.4\columnwidth}
+		\centering
+		\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}\vspace*{-.7em}
+		\caption{Execution time of different methods for estimating FIFs.
+{\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
+		}
+		\label{fairness_fairXplainer_fig:execution_time_cactus_plot}
+	\end{minipage}\hfil
+	\begin{minipage}{0.55\columnwidth}
+		\centering\vspace*{-1em}
+		\subfloat[COMPAS]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_compas}}
+		\subfloat[Adult]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
+		\caption{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
+Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}
+		\label{fairness_fairXplainer_fig:fairness_intervention}
+%		\centering
+%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_error.pdf}}
+%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_time.pdf}}
+%		
+%		\caption{Effect of spline intervals on the estimation error of statistical parity and  execution time of {\fairXplainer} on COMPAS.}
+%		\label{fairness_fairXplainer_fig:effect_spline_intervals}
 	\end{minipage}
-	\begin{minipage}{0.5\textwidth}
-		\centering
-		\subfloat[Equalized odds]{\includegraphics[scale=0.39]{figures/fairness/fif/eo_train_accuracy}}
-	\end{minipage}
-	\caption[Accuracy of {\fairXplainer}]{Comparing {\fairXplainer} and SHAP on the estimation error of fairness metrics.
-Lower values on the $ Y $-axis denote a better result.
-{\fairXplainer} has significantly less error than SHAP.}\label{fig:estimation_error}
-%	\vspace{-0.5em}
-\end{figure}
-\paragraph{Accurate Estimation of Bias.} We compare {\fairXplainer} with SHAP in estimating Statistical Parity $(\mathsf{SP})$ and Equalized Odds $(\mathsf{EO})$, where each metric is calculated by summing all FIFs (Axiom~\ref{axm:additivity}).
-We consider five datasets, each trained on different classifiers by applying five-fold cross-validation.
-We compute estimation error by taking the absolute difference between the exact and  estimated value of a metric, and present results in Figure~\ref{fig:estimation_error}.
-In both metrics, {\fairXplainer} demonstrates significantly less error than SHAP.
-For example, the mean error of {\fairXplainer} in estimating $ \mathsf{SP} $ on Titanic dataset is $ 0.1 $ vs.\ $ 0.42 $ for SHAP.
-In this context, the error of {\fairXplainer} is often zero in most datasets and only insignificant ($ \le 0.13 $) due to the degenerate cases (Section~\ref{sec:fifs}) with $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{0, 1\} $.
-Therefore, \emph{global sensitivity analysis based approach {\fairXplainer} is significantly more accurate in computing FIFs and corresponding fairness metric than local explanation based approach SHAP}.
-
-%\clearpage
-
-\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs.
-We consider COMPAS dataset with race = \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years.
-Since the classifier only optimizes training error, it demonstrates statistical parity as $ 0.17 $, which means a non-Caucasian  has $ 0.17 $ higher probability of re-offending crimes than a Caucasian.
-Next, we investigate the source of bias and present individual FIFs in Figure~\ref{fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fig:individual_and_intersectional_fifs}.
-In both figures, we show top seven influential FIFs sorted by absolute values,  followed by residual higher-order FIFs.
-In Figure~\ref{fig:individual_fifs}, `priors count' of an individual is dominating in increasing statistical parity (FIF $ = 0.12 $).
-Other non-sensitive features appear to have almost zero FIFs.
-However, the sum of second-order influences of all features increases statistical parity by $ 0.06 $, denoting that the data is highly correlated and presenting only individual FIFs does not trace the true sources of bias.
-For example, while both `sex' and `age' of persons have zero influence on bias (Figure~\ref{fig:individual_fifs}), these features together with `c charge degree', `priors count', and `juvenile other count' contribute highly on statistical parity (sum of FIFs $ = 0.03 + 0.02 + 0.01 + 0.01 - 0.01 - 0.01 = 0.05$).
-\emph{Therefore, intersectional influences together with individual influences lead to a clearer understanding on the source of bias of a classifier.
-Note that, unlike SHAP, {\fairXplainer} is the only framework that computes beyond individual FIFs.}
-\begin{figure}[t!]
-	\begin{minipage}{0.43\textwidth}
-		\centering
-		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fig:individual_fifs}}
-	\end{minipage}
-	\begin{minipage}{0.5\textwidth}
-		\centering
-		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fig:individual_and_intersectional_fifs}}
-	\end{minipage}
-	\vspace{-0.2em}
-	\caption[Individual vs.\ intersectional FIFs]{FIFs for COMPAS dataset on explaining statistical parity.
-Intersectional FIFs depict sources of bias in detail than individual FIFs.}\label{fig:individual_vs_intersectional_influence}
-
-\end{figure}
-\begin{figure}[t!]
-	\vspace{-0.7em}
-	\begin{minipage}{0.49\textwidth}
-		\centering
-		\subfloat[Fairness attack (punitive actions)]{\includegraphics[scale=0.345]{figures/fairness/fif/fairness_attack}\label{fig:fairness_attack}}
-	\end{minipage}
-	\begin{minipage}{0.5\textwidth}
-		\centering	
-		\subfloat[Fairness enchancing algorithm (affirmative actions)]{\includegraphics[scale=0.345]{figures/fairness/fif/fairness_repair}\label{fig:fairness_repair}}
-	\end{minipage}
-	\vspace{-0.2em}
-	\caption[FIFs under fairness attack and fairness enhancing algorithms]{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing algorithm~\cite{kamiran2012data} on FIFs.}\label{fig:affirmative_punitive_actions}
-%	\vspace{-0.5em}
 \end{figure}
 
-\paragraph{Fairness affirmative/punitive actions.} Continuing on the experiment in Figure~\ref{fig:individual_vs_intersectional_influence}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs for COMPAS dataset in Figure~\ref{fig:affirmative_punitive_actions}.
-In Figure~\ref{fig:individual_vs_intersectional_influence}, statistical parity of the classifier is $ 0.17 $.
-Applying a data poisoning fairness attack~\cite{solans2020poisoning} increases statistical parity to $ 0.5 $ and the data reweighing-based fairness-enhancing algorithm~\cite{kamiran2012data} decreases it to $ 0.15$.
-We observe that in both cases, there are a subset of features decreasing bias while another subset of features increasing it, e.g.
-`age' vs.\ `priors count'.
-Figure~\ref{fig:fairness_attack} suggests that the attack algorithm would be more successful if it could hide the influence of `age' of a person in receiving discriminating prediction for re-offending crimes.
-Figure~\ref{fig:fairness_repair} suggests that the fairness enhancing algorithm can improve by ameliorating the effect of `priors count' further.
-Thus, \textit{{\fairXplainer} provides a dissecting tool to undertake necessary steps to improve or worsen fairness of the classifier.}
+\paragraph{Execution Time: {\fairXplainer} vs.\ SHAP} We compare the execution time of {\fairXplainer} vs.\ SHAP in a cactus plot in Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot}, where a point $ (x, y) $ denotes that a method computes FIFs of $ y $ many fairness instances within $ x $ seconds.
+We consider $ 480 $ fairness instances from $ 4 $ datasets constituting $ 24 $ distinct combinations of sensitive features, $ 4 $ classifiers, and $ 5 $ cross-validation folds.
+In Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot},  {\fairXplainer} with $ \lambda = 1 $ is faster than  $ \lambda = 2 $.
+For example, within $ 10 $ seconds, {\fairXplainer} with $ \lambda = 1 $ solves $ 443 $ instances vs.\ $ 41 $ instances with $ \lambda = 2 $.
+In addition, {\fairXplainer} with $ \lambda = 1 $ solves all $ 480 $ instances compared to $ 360 $ instances solved by SHAP.
+Thus, {\fairXplainer} with $ \lambda = 1 $ demonstrates its higher efficiency in estimating individual FIFs compared to SHAP.
+While estimating intersectional FIFs, {\fairXplainer} also demonstrates its practical applicability by solving $ 350 $ instances within  $ 160 $ seconds.
+\textit{Therefore, {\fairXplainer} demonstrates computational efficiency in explaining group fairness metrics of real-world datasets and classifiers.}
+
+\paragraph{FIFs under Fairness Intervention} We consider a fairness intervention strategy to hide the impact of an individual feature on a classifier and record the correlation between fairness improvement/reduction of the intervened classifier with the FIF of the feature.
+Our intervention strategy of modifying the classifier is different than~\cite{datta2016algorithmic}, where the dataset is modified by replacing features with random values.
+In particular, we intervene a logistic regression classifier by setting the coefficient to zero for the corresponding feature.
+Intuitively, when the coefficient becomes zero for a feature, the prediction of the classifier may become independent on the feature; thereby, the bias of the classifier may also be independent on the conditional variances of the feature for different sensitive groups.
+As a result, a feature with a positive FIF value (i.e.
+increases bias) is likely to decrease bias under the intervention, and vice versa.
+In Figure~\ref{fairness_fairXplainer_fig:fairness_intervention}, we report Pearson's correlation coefficient between the difference in bias (statistical parity) due to intervention and the FIFs of features in COMPAS and Adult datasets, where features are sorted in descending order of their absolute FIFs estimated by {\fairXplainer} and SHAP.
+In {\fairXplainer}, the correlation coefficient generally decreases with an increase of top influential features, denoting that features with higher absolute FIFs highly correlate with bias-differences.
+SHAP, in contrast, demonstrates less correlation, specially for the top most influential features.
+\emph{Therefore, FIFs estimated by {\fairXplainer} shows the potential of being deployed to design improved fairness algorithms in future.}
+
+\begin{figure}
+	\centering
+	\subfloat[{\fairXplainer} ($ \lambda = 1 $)]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs}}\hfil
+	\subfloat[{\fairXplainer} ($ \lambda = 2 $)]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fairness_fairXplainer_fig:intersectional_fifs}}\hfil
+	\subfloat[SHAP]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_shap}\label{fairness_fairXplainer_fig:fif_shap}}
+
+	\caption{FIFs for COMPAS dataset on explaining statistical parity.
+Both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs} depict sources of bias in detail than individual FIFs alone in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}.
+In Figure~\ref{fairness_fairXplainer_fig:fif_shap}, individual FIFs estimated by SHAP are far from correctly approximating statistical parity (exact value $0.174 $).}
+	\label{fairness_fairXplainer_fig:fif_illustration_compas}
+\end{figure}
+
+\subsection{Explainability and Applicability of FIFs}
+\paragraph{Individual vs.\ Intersectional FIFs.} Now, we aim to understand the importance of intersectional FIFs over individual FIFs in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}.
+We consider COMPAS dataset with race $ \in $ \{Caucasian, non-Caucasian\} as the sensitive feature, and a logistic regression classifier to predict whether a person will re-offend crimes within the next two years.
+Since the classifier only optimizes training error, it demonstrates statistical parity of $ 0.174 $, i.e.
+it suggests that a non-Caucasian  has $ 0.174 $ higher probability of re-offending crimes than a Caucasian.
+Next, we investigate the sources of bias and present individual FIFs in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, and both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs}.
+In both figures, we present influential features and their FIFs in the descending order of absolute values.
+According to {\fairXplainer}, the priors count (FIF $ = 0.1 $) dominates in increasing statistical parity\textemdash between Caucasian and non-Caucasian, their prior count demonstrates the maximum difference in the scaled variance of positive prediction.
+Other non-sensitive features have almost zero FIFs.
+However, in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}, higher-order FIFs ($ \lambda > 1 $) increases statistical parity by $ 0.031 $, denoting that the data is correlated and presenting only individual FIFs is not sufficient for understanding the sources of bias.
+For example, while both sex and age individually demonstrate almost zero influence on bias (Figure~\ref{fairness_fairXplainer_fig:individual_fifs}), their combined effect and intersectional effects with c-charge degree, priors count, and juvenile miscellaneous count contribute highly on statistical parity.
+In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.
+Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19}.
+From Figure 3 in~\cite{rudin19}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively.
+While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features.
+\emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.
+%\todo{For a linear classifier with independent features, see if coefficient and FIFs are related.}
+
+\begin{figure}
+	\subfloat[Fairness attack (punitive actions)]{\includegraphics[scale=0.35]{figures/fairness/fif/fairness_attack}\label{fairness_fairXplainer_fig:fairness_attack}}\hfil
+	\subfloat[Fairness enchancing (affirmative actions)]{\includegraphics[scale=0.35]{figures/fairness/fif/fairness_repair}\label{fairness_fairXplainer_fig:fairness_repair}}
+
+	\caption{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing~\cite{kamiran2012data} algorithms on FIFs.}\label{fairness_fairXplainer_fig:affirmative_punitive_actions}
+\end{figure}
+
+\paragraph{Quantifying Impacts of Affirmative/Punitive Actions.} Continuing on the experiment in Figure~\ref{fairness_fairXplainer_fig:fif_illustration_compas}, we evaluate the effect of fairness attack and enhancing algorithms on FIFs on COMPAS dataset in Figure~\ref{fairness_fairXplainer_fig:affirmative_punitive_actions}.
+Without applying any fairness algorithm, the statistical parity of the classifier is $ 0.174 $.
+Applying a data poisoning fairness attack~\cite{solans2020poisoning} increases statistical parity to $ 0.502 $ (approximated as $ 0.418 $ in Figure~\ref{fairness_fairXplainer_fig:fairness_attack}), whereas  a fairness-enhancing algorithm based on data reweighing~\cite{kamiran2012data} decreases statistical parity to $ 0.151 $ (approximated as $ 0.135 $ in Figure~\ref{fairness_fairXplainer_fig:fairness_repair}).
+In Figure~\ref{fairness_fairXplainer_fig:fairness_attack}, the attack algorithm would be more successful if it could hide the influence of features with positive FIFs, such as priors count and the intersectional effects of age and c-charge degree with sex.
+In contrast, in Figure~\ref{fairness_fairXplainer_fig:fairness_repair}, the fairness enhancing algorithm can improve by  further ameliorating the effect features with  negative FIFs, such as priors count.
+Thus, \textit{{\fairXplainer} demonstrates the potential as a dissecting tool to undertake necessary steps to improve or worsen fairness of a classifier.}

### 2023-02-07 chapters/fairness/fif/fairxplainer.tex
minor edits
considering all minor comments
adding fairXplainer

--- 2022-06-01 chapters/fairness/fif/fairxplainer.tex
@@
-\section{\fairXplainer: An Algorithm to Compute Fairness Influence Functions}\label{sec:fairxplainer}
+\section{ An Algorithm to Estimate Fairness Influence Functions}\label{fairness_fairXplainer_sec:fairxplainer}
 \begin{comment}
@@
 \end{comment}
-We propose an algorithm, {\fairXplainer}, that leverages the variance decomposition of Eq.~\eqref{eq:fif_decompose} to compute the Fairness Influence Functions (FIFs) of all the subset of features.
-{\fairXplainer} has two algorithmic blocks: (i) local regression to decompose the classifier into component functions and (ii) computing the variance (or covariance) of each component.
-We describe the schematic of \fairXplainer{} in Algorithm~\ref{algo:framework}.
+We propose an algorithm, {\fairXplainer}, that leverages the variance decomposition of CPPs to estimate the FIFs of all subsets of features.
+{\fairXplainer} has two algorithmic blocks: (i) local regression to decompose the classifier into component functions taking distinct subsets of features as input and (ii) computing the variance (or covariance) of each component function.
+We describe the schematic of {\fairXplainer} in Algorithm~\ref{fairness_fairXplainer_algo:framework}.
 \setlength{\textfloatsep}{12pt}% Remove \textfloatsep
 
-\noindent\textbf{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{eq:functional_decomposition_set_notation}).
-Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \function_{\mathbf{a}}(\nonsensitive) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $.
-We express $ \function_{\mathbf{a}} $ as a set-additive model:
+\paragraph{A Set-additive Representation of the Classifier.} To apply variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}), we learn a set-additive representation of the classifier (Eq.~\eqref{fairness_fairXplainer_eq:functional_decomposition_set_notation}) with input $ \feature \equiv (\nonsensitive, \sensitive) $.
+Let us denote the classifier $ \alg $ conditioned on a sensitive group $ \mathbf{a} $ as $ \fiffunc_{\mathbf{a}}(\feature) \triangleq \alg(\nonsensitive, \sensitive = \mathbf{a}) $.
+We express $ \fiffunc_{\mathbf{a}} $ as a set-additive model:
 \begin{align}
-\label{eq:set_additive_classifier}
-\function_{\mathbf{a}}(\mathbf{X}) = \function_{\mathbf{a}, 0} +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} \function_{\mathbf{a},\mathbf{S}}(\mathbf{X}_{\mathbf{S}}) + \delta
+\label{fairness_fairXplainer_eq:set_additive_classifier}
+\fiffunc_{\mathbf{a}}(\feature) = \fiffunc_{\mathbf{a}, 0} +  \sum_{\mathbf{S} \subseteq [\numfeatures] , |\mathbf{S}| \le \lambda} \fiffunc_{\mathbf{a},\mathbf{S}}(\feature_{\mathbf{S}}) + \delta
 \end{align}
-\begin{algorithm}
-	\caption{{\fairXplainer}}\label{algo:framework}
-	\hspace*{\algorithmicindent} \textbf{Input:} $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y},  \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n, f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}, \lambda $  \\
-	\hspace*{\algorithmicindent} \textbf{Output:} $ w_{\mathbf{S}}$ 
+
+%\iffalse
+\begin{algorithm}[t!]
+	\caption{{\fairXplainer}}\label{fairness_fairXplainer_algo:framework}
+	\begin{flushleft}
+	\hspace*{\algorithmicindent}\textbf{Input:} Classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y}$, Dataset $\mathbf{D} = \{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$, Fairness metric $f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$, Maximum order of intersectional influence $\lambda $\\
+	\hspace*{\algorithmicindent}\textbf{Output:} FIF $ w_{\mathbf{S}}$ for the subsets of features $\{\feature_{\mathbf{S}}\}$
+	\end{flushleft}
+
 	\begin{algorithmic}[1]
-		\State $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], k \gets    |\nonsensitive| $
-		\label{algo_line:fif_computation_start}
+		\State $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}], \numfeatures \gets    |\feature|, \feature \equiv (\nonsensitive, \sensitive) $
+		\label{fairness_fairXplainer_algo_line:fif_computation_start}
 
 		\For{$ \mathbf{a} \in \{\mathbf{a}_{\max}, \mathbf{a}_{\min}\} $} \Comment{Enumerate for specific sensitive groups}
-		\State $ \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0} \gets \textsc{LocalRegression}(\alg(\nonsensitive, \sensitive = \mathbf{a}), \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k) $
+		\State $ \fiffunc_{\mathbf{a}, \mathbf{S}},\fiffunc_{\mathbf{a}, 0} \gets \textsc{LocalRegression}(\alg(\nonsensitive, \sensitive = \mathbf{a}), \{\mathbf{z}^{(i)}\}_{i=1}^n, \lambda, m) $
 
-		\State $ V_{\mathbf{a}, \mathbf{S}} \gets \textsc{Covariance}(\function_{\mathbf{a}},  \{\mathbf{x}^{(i)}\}_{i=1}^n, \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0}) $
+		\State $ V_{\mathbf{a}, \mathbf{S}} \gets \textsc{Covariance}(\fiffunc_{\mathbf{a}},  \{\mathbf{z}^{(i)}\}_{i=1}^n, \fiffunc_{\mathbf{a}, \mathbf{S}},\fiffunc_{\mathbf{a}, 0}) $
 		\EndFor
-		\State Compute $ w_{\mathbf{S}}$ using  $V_{\mathbf{a}_{\max},\mathbf{S}}$ and $V_{\mathbf{a}_{\min},\mathbf{S}}$ as in Equation~\eqref{eq:fif_decompose}
+		\State Compute $ w_{\mathbf{S}}$ using  $V_{\mathbf{a}_{\max},\mathbf{S}}$ and $V_{\mathbf{a}_{\min},\mathbf{S}}$ as in Equation~\eqref{fairness_fairXplainer_eq:fif_decompose}
 		%\State $ w_{\mathbf{S}}  = (V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}})/(1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])) $
-		\label{algo_line:fif_computation_end}
-		\Comment{Theorem~\ref{lemma:fif}}
+		\label{fairness_fairXplainer_algo_line:fif_computation_end}
 
-%		\Statex
-		\Function{LocalRegression}{$ g_\mathbf{a}, \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k $}
-		\label{algo_line:local_regression_start}
-		\State \textbf{Initialize}: $ \function_{\mathbf{a}, 0} \leftarrow \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n), \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [k]\setminus\emptyset, |\mathbf{S}|\le \lambda$ \label{alg_line:initialization}
-		\While{each $ \widehat{\function}_{\mathbf{a},\mathbf{S}} $ does not converge}
+		\Statex
+		\Function{LocalRegression}{$ g_\mathbf{a}, \{\mathbf{z}^{(i)}\}_{i=1}^n, \lambda, m $}
+		\label{fairness_fairXplainer_algo_line:local_regression_start}
+		\State \textbf{Initialize}: $ \fiffunc_{\mathbf{a}, 0} \leftarrow \textsc{Mean}(\{g(\mathbf{z}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n), \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [m], \mathbf{S} \ne \emptyset, |\mathbf{S}|\le \lambda$ \label{fairness_fairXplainer_alg_line:initialization}
+		\While{each $ \widehat{\fiffunc}_{\mathbf{a},\mathbf{S}} $ does not converge}
 		\For{each $\mathbf{S}$}
-		\State $ \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow \textsc{Smooth}[\{\function_{\mathbf{a}}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\function}_{\mathbf{a}, \mathbf{S}'}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n]
-		$\label{alg_line:backfitting_step} \Comment{Backfitting}
-		\State $\widehat{\function}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\function}_{\mathbf{a}, \mathbf{S}} - \textsc{Mean}(\{\widehat{\function}_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n)$ \label{alg_line:mean_centered} \Comment{Mean centering}
+		\State $ \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}} \leftarrow \textsc{Smooth}\big(\{\fiffunc_{\mathbf{a}}(\mathbf{z}^{(i)}) - \fiffunc_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}'}(\mathbf{z}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\big)
+		$\label{fairness_fairXplainer_alg_line:backfitting_step} \Comment{Backfitting}
+		\State $\widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}} - \textsc{Mean}\big(\{\widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\big)$ \label{fairness_fairXplainer_alg_line:mean_centered} \Comment{Mean centering}
 		\EndFor
 		\EndWhile
-		\State \Return $ \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0} $
-		\label{algo_line:local_regression_end} 
+		\State \Return $ \fiffunc_{\mathbf{a}, \mathbf{S}},\fiffunc_{\mathbf{a}, 0} $
+		\label{fairness_fairXplainer_algo_line:local_regression_end} 
 		\EndFunction
 
-		\Function{Covariance}{$\function_{\mathbf{a}},  \{\mathbf{x}^{(i)}\}_{i=1}^n, \function_{\mathbf{a}, \mathbf{S}},\function_{\mathbf{a}, 0}$}
-		\label{algo_line:covariance_computation_start}
-		\State \Return $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}) $
-		\label{algo_line:covariance_computation_end}
-		%		 \Comment{Eq.~\eqref{eq:covariance_computation}}
+		\Function{Covariance}{$\fiffunc_{\mathbf{a}},  \{\mathbf{z}^{(i)}\}_{i=1}^n, \fiffunc_{\mathbf{a}, \mathbf{S}},\fiffunc_{\mathbf{a}, 0}$}
+		\label{fairness_fairXplainer_algo_line:covariance_computation_start}
+		\State \Return $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \fiffunc_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{z}^{(i)}) - \fiffunc_{\mathbf{a}, 0}) $
+		\label{fairness_fairXplainer_algo_line:covariance_computation_end}
+		%		 \Comment{Eq.~\eqref{fairness_fairXplainer_eq:covariance_computation}}
 		\EndFunction
@@
 \end{algorithm}
-Here, $ \function_{\mathbf{a}, 0} $ is a constant and $ \function_{\mathbf{a},\mathbf{S}} $ is a \emph{component function} of $ \function_{\mathbf{a}} $ taking  $ \nonsensitive_{\mathbf{S}} $ as input, and $ \delta $ is the approximation error.
-For computational tractability, we consider only components of \emph{maximum order} $ \lambda $.
-{\fairXplainer} deploys backfitting algorithm for learning component functions in Eq.~\eqref{eq:set_additive_classifier}.
+%\fi
+Here, $ \fiffunc_{\mathbf{a}, 0} $ is a constant, $ \fiffunc_{\mathbf{a},\mathbf{S}} $ is a \emph{component function} of $ \fiffunc_{\mathbf{a}} $ taking  a non-empty subset of features $ \feature_{\mathbf{S}} $ as input, and $ \delta $ is the approximation error.
+For computational tractability, we consider only components of \emph{maximum order} $ \lambda $, denoting the maximum order of intersectionality.
+{\fairXplainer} deploys backfitting algorithm for learning component functions in Eq.~\eqref{fairness_fairXplainer_eq:set_additive_classifier}, as discussed in the following.
 
-\noindent\textbf{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to order $ \lambda $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}).
-Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \function_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed.
-Specifically, $ \function_{\mathbf{a}, \mathbf{S}} $ is learned as a smoothed function of $ g $ and rest of the components $ \function_{\mathbf{a}, \mathbf{S}'} $, where $ \mathbf{S}' \ne \mathbf{S} $ is a non-empty subset of $ [k]\setminus \emptyset $.
-To keep every component function mean centered, backfitting requires to impose the constraints $ \function_{\mathbf{a}, 0} = \textsc{Mean}(\{g(\mathbf{x}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n) $ (Line~\ref{alg_line:initialization}), which is the mean of $ \function_{\mathbf{a}} $ evaluated on samples belonging to the sensitive group $ \sensitive = \mathbf{a} $;  and $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}) = 0$ (Line~\ref{alg_line:mean_centered}), where $ \mathbf{x}_{\mathbf{S}}^{(i)} $ be the subset of feature values associated with feature indices $ \mathbf{S} $ for the $ i $-th sample $ \mathbf{x}^{(i)} $.
-These constraints assign the expectation of $ \function_{\mathbf{a}} $ on the constant term $ \function_{\mathbf{a}, 0} $ and the variance of $ \function_{\mathbf{a}} $ to the component functions.
+\paragraph{Local Regression with Backfitting.} We perform local regression with backfitting algorithm to learn the component functions up to a given order $ \lambda $ (Line~\ref{fairness_fairXplainer_algo_line:local_regression_start}--\ref{fairness_fairXplainer_algo_line:local_regression_end}).
+Backfitting algorithm is an iterative algorithm, where in each iteration one component function, say $ \fiffunc_{\mathbf{a}, \mathbf{S}} $, is learned while keeping other component functions fixed.
+Specifically, $ \fiffunc_{\mathbf{a}, \mathbf{S}} $ is learned as a smoothed function of $ g $ and rest of the components $ \fiffunc_{\mathbf{a}, \mathbf{S}'} $, where $ \mathbf{S}' \ne \mathbf{S} $ is a non-empty subset of $ [\numfeatures] $.
+To keep every component function mean centered, backfitting requires to impose two constraints: (i) $ \fiffunc_{\mathbf{a}, 0} = \textsc{Mean}\big(\{g(\mathbf{z}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\big) $ (Line~\ref{fairness_fairXplainer_alg_line:initialization}), which is the mean of $ \fiffunc_{\mathbf{a}} $ evaluated on data points belonging to the sensitive group $ \mathbf{a} $;  and (ii) $ \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\fiffunc_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)}) = 0$ (Line~\ref{fairness_fairXplainer_alg_line:mean_centered}), where $ \mathbf{z}_{\mathbf{S}}^{(i)} $ is the subset of feature values associated with feature indices $ \mathbf{S} $ for the $ i $-th data point $  \mathbf{z}^{(i)} $.
+These constraints assign the expectation of $ \fiffunc_{\mathbf{a}} $ on the constant term $ \fiffunc_{\mathbf{a}, 0} $ and the variance of $ \fiffunc_{\mathbf{a}} $ to the component functions.
 
-While performing local regression, backfitting uses a smoothing operator~\cite{loader2012smoothing} over the set of samples (Line~\ref{alg_line:backfitting_step}).
-A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating the local curves obtained by local regression~\cite{loader2012smoothing}.
+While performing local regression, backfitting uses a smoothing operator~\cite{loader2012smoothing} over the set of data points (Line~\ref{fairness_fairXplainer_alg_line:backfitting_step}).
+A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating $\tau + 2$ local points obtained by local regression~\cite{loader2012smoothing}.
 In this paper, we apply cubic spline smoothing~\cite{li2010global} to learn each component function.
-Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity.
-Hence, up to the second derivatives of each piecewise term are zero at the endpoints of intervals.
-We refer to Appendix~\ref{sec:smoothing} for further details.%\todo{Add pseudo-code of {\fairXplainer}}
+Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity interpolating local points in $ \tau $ intervals.
+Hence, the first and second derivatives of each piecewise term are zero at the endpoints of intervals.
+We refer to Appendix~\ref{fairness_fairXplainer_sec:smoothing} for details of implementation.
+An ablation study demonstrating the impacts of the hyperparameters $\tau$ and $\lambda$ on the performance of {\fairXplainer} is in Appendix~\ref{fairness_fairXplainer_sec:additional_experiments}.
+%\todo{Mention $ \tau $ as the number of spline intervals.
+mentioned in appendix.
+makes sense to include there if the ablation study is not in the main.}
 
-\begin{comment}
-\begin{algorithm}[!t]
-	\caption{Local Regression with Backfitting Algorithm}
-	\label{algo:backfitting_algorithm}
-%	\hspace*{\algorithmicindent} \textbf{Input:} $ g_\mathbf{a}(\mathbf{X}), \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda $ \\
-%	\hspace*{\algorithmicindent} \textbf{Output:} $ \function_{\mathbf{a}, 0}, \function_{\mathbf{a}, \mathbf{S}} $ 
-	\begin{algorithmic}[1]
-		\Procedure{LocalRegression}{$ g_\mathbf{a}, \{\mathbf{x}^{(i)}\}_{i=1}^n, \lambda, k $}
-		\State \textbf{Initialize}: $ \function_{\mathbf{a}, 0} \leftarrow \frac{1}{n}\sum_{i=1}^n\function_{\mathbf{a}}(\mathbf{x}^{(i)}), \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow 0, \forall \mathbf{S} \in [k]\setminus\emptyset, |\mathbf{S}|\le \lambda$ \label{alg_line:initialization}
-		\While{each $ \widehat{\function}_{\mathbf{a},\mathbf{S}} $ does not converge}
-		\For{each $\mathbf{S}$}
-		\State $ \widehat{\function}_{\mathbf{a}, \mathbf{S}} \leftarrow \mathsf{Smooth}[\{\function_{\mathbf{a}}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\function}_{\mathbf{a}, \mathbf{S}'}(\mathbf{x}_{\mathbf{S}}^{(i)})\}_{i=1}^n] $\label{alg_line:backfitting_step} \Comment{Backfitting step}
-		\State $\widehat{\function}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\function}_{\mathbf{a}, \mathbf{S}} - \frac{1}{n}\sum_{i=1}^n \widehat{\function}_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})$ \label{alg_line:mean_centered} \Comment{Mean centering}
-		\EndFor
-		\EndWhile
-		\State \Return $ \function_{\mathbf{a}, 0}, \function_{\mathbf{a}, \mathbf{S}} $ 
-		\EndProcedure
-	\end{algorithmic}
-\end{algorithm}
-\end{comment}
+\paragraph{Variance and Covariance Computation.} Once each component function $ \fiffunc_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{fairness_fairXplainer_algo_line:local_regression_start}--\ref{fairness_fairXplainer_algo_line:local_regression_end}), we compute variances of the component functions and their covariances using $g_\mathbf{a}(\cdot)$.
+Since each component function is mean centered (Line~\ref{fairness_fairXplainer_alg_line:mean_centered}), we compute the variance of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $ for the dataset $\mathbf{D}$ as
+$  \mathsf{Var}[\fiffunc_{\mathbf{a}, \mathbf{S}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n (\fiffunc_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)}))^2 $.
+Hence, variance captures the independent effect of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $.
+Covariance is computed to account for the correlation among features $ \feature $.
+We compute the covariance of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $ with $ \fiffunc_{\mathbf{a}} $ on the dataset as
+\[
+\label{fairness_fairXplainer_eq:covariance_computation}
+	 \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \fiffunc_{\mathbf{a}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \fiffunc_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{z}^{(i)}) - \fiffunc_{\mathbf{a}, 0}).
+\]
+Here, $ g_\mathbf{a}(\cdot) - \fiffunc_{\mathbf{a}, 0} $ is the mean centered form of $ \fiffunc_{\mathbf{a}} $.
+Covariance of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $ can be both positive and negative depending on whether the features $ \feature_{\mathbf{S}} $ are positively or negatively correlated with $ \fiffunc_{\mathbf{a}} $.
+Specifically, under the set additive model, we obtain $ \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \fiffunc_{\mathbf{a}}] = \mathsf{Var}[\fiffunc_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S} \ne \mathbf{S}'} \fiffunc_{\mathbf{a}, \mathbf{S}'}] $.
+Now, we use $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \fiffunc_{\mathbf{a}}] $ as the effective variance of $ \feature_{\mathbf{S}} $ for a given sensitive group $ \mathbf{a} $ (Line~\ref{fairness_fairXplainer_algo_line:covariance_computation_start}--\ref{fairness_fairXplainer_algo_line:covariance_computation_end}).
+In Line~\ref{fairness_fairXplainer_algo_line:fif_computation_start}--\ref{fairness_fairXplainer_algo_line:fif_computation_end}, we compute $ V_{\mathbf{a}, \mathbf{S}} $ for the most and the least favored groups, and plug them in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose} to yield an estimate of FIF of $ \feature_{\mathbf{S}} $.
+%\todo{Any error bound?}
 
-\noindent\textbf{Variance and Covariance Computation.} Once each component function $ \function_{\mathbf{a}, \mathbf{S}} $ is learned with $ \textsc{LocalRegression} $ (Line~\ref{algo_line:local_regression_start}--\ref{algo_line:local_regression_end}), we compute variances of the component functions and their covariances with $ \function_{\mathbf{a}} $.
-Since each component function is mean centered (Line~\ref{alg_line:mean_centered}), we compute the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ on the dataset as
-$  \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n (\function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)}))^2 $.
-Hence, variance captures the independent effect of $ \function_{\mathbf{a}, \mathbf{S}} $.
-%$ V_{\mathbf{a}, \mathbf{S}} $ is the variance of $ \function_{\mathbf{a}, \mathbf{S}} $ evaluated on each $ \mathbf{x}_{\mathbf{S}}^{(i)} $ in the dataset. 
-Covariance is computed to account for the correlation among features $ \nonsensitive $.
-We compute the covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ with $ \function_{\mathbf{a}} $ on the dataset as
-$
-\label{eq:covariance_computation}
-	 \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \sum_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n \function_{\mathbf{a}, \mathbf{S}}(\mathbf{x}_{\mathbf{S}}^{(i)})(g_\mathbf{a}(\mathbf{x}^{(i)}) - \function_{\mathbf{a}, 0}).
-$
-Here, $ g_\mathbf{a}(\cdot) - \function_{\mathbf{a}, 0} $ is the mean centered form of $ \function_{\mathbf{a}} $.
-Covariance of $ \function_{\mathbf{a}, \mathbf{S}} $ can be both positive and negative depending on whether the features $ \nonsensitive_{\mathbf{S}} $ are positively or negatively correlated with $ \function_{\mathbf{a}} $.
-Specifically, under the set additive model, we obtain $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $.
-Now, we use $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] $ as the effective variance of $ \nonsensitive_{\mathbf{S}} $ for a given sensitive group $ \sensitive = \mathbf{a} $ (Line~\ref{algo_line:covariance_computation_start}--\ref{algo_line:covariance_computation_end}).
-In Line~\ref{algo_line:fif_computation_start}--\ref{algo_line:fif_computation_end}, we compute $ V_{\mathbf{a}, \mathbf{S}} $ for the most and the least favored groups, and plug them in Theorem~\ref{lemma:fif} to compute FIF of $ \nonsensitive_{\mathbf{S}} $.
+%Here, we provide the runtime complexity of \framework.
+\begin{proposition}[Time Complexity of {\fairXplainer}] Let $ t $ be the number of iterations of the backfitting algorithm, $ \numfeatures $ be the number of features, $ \lambda $ be the maximum order of intersectional features, and $ s $ be the runtime complexity of the smoothing oracle\footnote{Typically, the runtime complexity of smoothing oracles, particularly of cubic splines, is linear with respect to $ n $, i.e.
+the number of samples~\cite{toraichi1987computational}.}.	Then, the runtime complexity of Algorithm~\ref{fairness_fairXplainer_algo:framework} is $ \mathcal{O}\Big(ts\sum_{i=1}^{\lambda} {\numfeatures \choose i}\Big)$.
+For example, if we are interested in up to first and second order intersectional features, the runtime complexities are $\mathcal{O}(ts\numfeatures)$ and $\mathcal{O}(ts\numfeatures^2)$, respectively.
+\end{proposition}
 
-%In both variance and covariance decomposition, $ \mathsf{Var}[\function_{\mathbf{a}}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}| \le \lambda} V_{\mathbf{a}, \mathbf{S}}$.
+%In both variance and covariance decomposition, $ \mathsf{Var}[\fiffunc_{\mathbf{a}}] = \sum_{\mathbf{S} \subseteq [\numfeatures] , |\mathbf{S}| \le \lambda} V_{\mathbf{a}, \mathbf{S}}$.
 The difference between variance and covariance decomposition is that $ V_{\mathbf{a}, \mathbf{S}} \in \mathbb{R}^{\ge 0} $ is positive in  variance decomposition, while $ V_{\mathbf{a}, \mathbf{S}} \in \mathbb{R} $ may also be negative in covariance decomposition.
-In covariance decomposition, we can separate between the structural and correlative effects of $ \function_{\mathbf{a}, \mathbf{S}} $.
-Formally, $ \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \function_{\mathbf{a}}] = \mathsf{Var}[\function_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\function_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S}' \subseteq [\numnonsensitive] \setminus \emptyset, |\mathbf{S}'| \le \lambda, \mathbf{S} \ne \mathbf{S}'} \function_{\mathbf{a}, \mathbf{S}'}] $, where  to the right side of equality, first term refers to the structural effect and the second term refers to correlative effect of $ \function_{\mathbf{a}, \mathbf{S}} $.
+In covariance decomposition, we can separate between the structural and correlative effects of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $.
+Formally, $ \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \fiffunc_{\mathbf{a}}] = \mathsf{Var}[\fiffunc_{\mathbf{a}, \mathbf{S}}] + \mathsf{Cov}[\fiffunc_{\mathbf{a}, \mathbf{S}}, \sum_{\mathbf{S}' \subseteq [\numfeatures] , |\mathbf{S}'| \le \lambda, \mathbf{S} \ne \mathbf{S}'} \fiffunc_{\mathbf{a}, \mathbf{S}'}] $, where  to the right side of equality, first term refers to the structural effect and the second term refers to correlative effect of $ \fiffunc_{\mathbf{a}, \mathbf{S}} $.
 
-\noindent\textbf{Extension of {\fairXplainer} to Equalized Odds and Predictive Parity.} 
-We extend {\fairXplainer} in computing FIF of group fairness metrics beyond statistical parity, namely equalized odds and predictive parity.
-For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of samples in the dataset where $ Y = 1 $ and another on samples with $ Y = 0 $.
-Then, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ quantifies the equalized odds of the classifer.
-
-To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
-For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label.
+%To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
+For each sub-dataset, we deploy $ \framework $ by setting the ground-truth class $ Y $ as label.
 This contrasts the computation for  statistical parity and equalized odds, where the predicted class $ \widehat{Y} $ is considered as label.
-Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ quantifies the predictive parity of the classifier.
+Finally, the maximum of the sum of FIFs between two sub-datasets for $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $ measures the predictive parity.
+% of the classifier.

### 2023-02-07 chapters/fairness/fif/figure_intro.tex
adding fairXplainer

--- 2022-06-01 chapters/fairness/fif/figure_intro.tex
@@
-\begin{figure}[t!]
+\begin{figure}
 	\begin{minipage}[t]{0.13\textwidth}			
@@
 				\node[latent,scale=2] (a1) {$\textrm{age}$} ; %
-				\node[obs, scale=2, below=of a1, xshift=-2cm] (h) {$\textrm{income}$}; %
-				\node[obs, scale=2, below=of a1, xshift=2cm] (i) {$\textrm{fitness}$}; %
+				\node[obs, scale=2, below=of a1, xshift=-2cm] (h) {$\textrm{fitness}$}; %
+				\node[obs, scale=2, below=of a1, xshift=2cm] (i) {$\textrm{income}$}; %
 				\node[obs, scale=2, below=of h, xshift=2cm] (p) {$\widehat{Y}$}; %			
@@
 			}	
-			\label{fig:dag_age_income_fitness}}
+			\label{fairness_fairXplainer_fig:dag_age_income_fitness}}
 	\end{minipage}%
-	\begin{minipage}{0.5\textwidth}
+	\begin{minipage}{0.4\textwidth}
 		\centering
 		\subfloat[Age-dependent distributions of non-sensitive features]{\includegraphics[scale=0.4]{figures/fairness/fif/sanity_distribution}
-		\label{fig:distribution_example}}
+		\label{fairness_fairXplainer_fig:distribution_example}}
 	\end{minipage}
@@
 				\end{tikzpicture}}
-			\label{fig:dt_original}}
+			\label{fairness_fairXplainer_fig:dt_original}}
 	\end{minipage}%
@@
 		\centering
-		\subfloat[Decision tree with affirmative action (DT$ 2 $)]{
+		\subfloat[Decision tree with an affirmative action (DT$ 2 $)]{
 			\scalebox{0.45}{	
@@
 				\node [scale=1.5, box, below= of p, xshift=-2.1cm, yshift=1.2cm]    (a1)    {income\\ $\geq 0.29$};
-				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm, fill=affirmative]     (a2)    {income $\geq 0.555$};
+				\node [scale=1.5, box, below= of p, xshift=2.1cm, yshift=1.2cm, fill=affirmative]     (a2)    {income $\geq 0.55$};
 				\node [scale=1.5,below= of a1, xshift=-1.5cm, yshift=0.8cm]  (a11)    { $\widehat{Y}= 1$};
@@
 
-			\label{fig:dt_affirmative}}
+			\label{fairness_fairXplainer_fig:dt_affirmative}}
 	\end{minipage}
@@
 	\begin{minipage}[t]{0.07\textwidth}
-		\vspace{-2em}
+		%\vspace{-2em}
 		\scalebox{1}{	
@@
 		\centering
-		\subfloat[Fairness influence functions (FIF) for DT$ 1 $]{\includegraphics[scale=0.45]{figures/fairness/fif/fif_example}
-		\label{fig:fif_original}}	
+		\subfloat[Fairness influence functions (FIF) for DT$ 1 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example}
+		\label{fairness_fairXplainer_fig:fif_original}}	
 	\end{minipage}%
@@
 		\centering
-		\subfloat[Modified FIFs for DT$ 2 $]{\includegraphics[scale=0.45]{figures/fairness/fif/fif_example_affirmative_action}
-		\label{fig:fif_affirmative}}
+		\subfloat[Modified FIFs for DT$ 2 $]{\includegraphics[scale=0.42]{figures/fairness/fif/fif_example_affirmative_action}
+		\label{fairness_fairXplainer_fig:fif_affirmative}}
 	\end{minipage}
-
-	\caption{FIFs of input features to investigate the bias (i.e., statistical parity) of a decision tree classifier predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
-An affirmative action reduces bias, and corresponding FIFs present the evidence.}
-	\label{fig:fair_example_fif}
-	\vspace{-1em}
+%\vspace*{-.5em}
+	\caption{FIFs of input features to investigate the bias (statistical parity) of a decision tree predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
+An affirmative action reduces bias as corresponding FIFs reflect it.}
+	\label{fairness_fairXplainer_fig:fair_example_fif}
+	%\vspace{-1.2em}
 \end{figure}

### 2023-02-07 chapters/fairness/fif/framework.tex
adding fairXplainer

--- 2022-06-01 chapters/fairness/fif/framework.tex
@@
 %\clearpage
-\section{Fairness Influence Functions: Definition and Computation}\label{sec:fifs}
-In this section, we discuss our contribution of formalizing and computing Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier on a dataset.
-The key idea in the formalization of FIFs is their additive formulation.
-The key idea in the computation of FIFs is to relate bias with the scaled difference of the conditional variance of positive prediction of the classifier over specific sensitive groups, and to apply variance decomposition to compute individual and intersectional FIFs.
-%\red{In this section, we first formalize fairness influence functions, and then discuss a variance decomposition based approach to compute FIFs for group fairness metrics.}\todo{Structure: 1.
-Background (measuring bias using fairness metrics) 2.FIFs as a measure of feature bias 3.
-Reducing FIF estimation to variance decomposition 4.
-An algorithm to do the variance decomposition}
+\section{Fairness Influence Functions: Formulation and Properties}\label{fairness_fairXplainer_sec:fifs}
+We formalize Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier applied on a dataset.
+In GSA, we observe that the variance of the output of a function can be attributed to the corresponding subset of input variables through variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}).
+To leverage the power of GSA in fairness in machine learning, we first express the existing fairness metrics in terms of the variance of classifier's prediction in Section~\ref{fairness_fairXplainer_sec:metric_as_variance}.
+This allows us to formulate FIF by leveraging variance decomposition (Section~\ref{fairness_fairXplainer_sec:fif_formulation}).
 
-%in Section~\ref{??}, and conclude by extending influence computation to other group fairness metrics in Section~\ref{??}
-%\subsection{An Axiomatic Formulation}
-%\label{sec:problem_formulation}
-\noindent\textbf{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $.
-Our objective is to compute the influences of non-sensitive features on $ f $.
-We compute influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.
+\subsection{Fairness Metrics as the Variance of Prediction}
+\label{fairness_fairXplainer_sec:metric_as_variance}
 
-\begin{definition}[Fairness Influence Function]
-Fairness Influence Function (FIF) $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$.
-We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $\textemdash particularly, a second-order influence when $ |\mathbf{S}| = 2 $ and a higher-order influence when $ 3 \le |\mathbf{S}| \le \numnonsensitive $.
-%\unsure{What do we mean by function here?
-Isn't FIF scalar?}
+First, we observe that the random variable central to computing statistical parity is $ \mathds{1}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $.
+We refer to this indicator function as \textit{Conditional Positive Prediction (CPP)} of a classifier.
+Now, we express statistical parity as a functional of the probability of CPPs for different sensitive groups~\cite{ghosh2020justicia,benesse2021fairness}.
+For brevity, we defer similar formulations for the other fairness metrics, i.e.
+equalized odds and predictive parity, to Appendix~\ref{fairness_fairXplainer_app:pp}.
+%CPP has an additional conditional $ Y = y $.}
+%For indicator functions, probability and expectation are equal.
+
+%To this end, existing fairness metrics such as statistical parity is the first moment (probability or expectation) of prediction distribution while variance is the second moment of distribution.  
+Our key idea for computing FIFs of features is to represent fairness metrics using the variance of CPPs.
+Formally, we express statistical parity using the variance of CPPs in Lemma~\ref{fairness_fairXplainer_lm:sp_var_relation}.
+
+\begin{lemma}[Statistical Parity as Difference of Variances of CPPs]
+	\label{fairness_fairXplainer_lm:sp_var_relation}
+	Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 \mid  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $ be the most and the least favored sensitive groups, respectively.
+The statistical parity of a binary\footnote{For a multi-class classifier, statistical parity of the target class $ y \in \mathbb{N} $ is $ \frac{\mathsf{Var}[\widehat{Y} = y \mid \sensitive = \mathbf{a}_{\max}]}{1 - \Pr[\widehat{Y} = y \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = y \mid \sensitive = \mathbf{a}_{\min}] }{1 - \Pr[\widehat{Y} = y \mid  \sensitive = \mathbf{a}_{\min}]}.
+$} classifier is the difference in the scaled variance of CPPs.
+Formally, if $\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}] \neq 0$ and $\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}] \neq 0$,
+	\[
+	 f_{\mathsf{SP}}(\alg, \mathbf{D}) = \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\min}] }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}.
+	\]
+\end{lemma}
+\begin{proof}
+	For a sensitive group $ \mathbf{a} $, CPP  is a Bernoulli random variable, where probability\footnote{For any binary event $ E $, expectation and probability are identical, $ \mathbb{E}[\mathds{1}(E)] = \Pr[E = 1] $.} $ p_{\mathbf{a}}  = \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $ and variance $ V_{\mathbf{a}} = \mathsf{Var}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] = p_{\mathbf{a}} (1 - p_{\mathbf{a}}) $.
+Thus, for sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $, the statistical parity of the classifier is $ p_{\mathbf{a}_{\max}}  - p_{\mathbf{a}_{\min}} =  \frac{V_{\mathbf{a}_{\max}}}{1 - p_{\mathbf{a}_{\max}}}  - \frac{V_{\mathbf{a}_{\min}}}{1 - p_{\mathbf{a}_{\min}}} $.
+Replacing $ 1 - p_\mathbf{a} = \Pr[\widehat{Y} = 0 \mid \sensitive = \mathbf{a}] $  proves the lemma.
+\end{proof}
+
+\begin{repexample}{fairness_fairXplainer_ex:motivating_example}[Revisited]
+	From Figure~\ref{fairness_fairXplainer_fig:dt_original}, the probability of CPPs of the decision tree for the most and least favored groups are $  \Pr[\widehat{Y} = 1 \mid \text{age = young}] = 0.695 $ and $ \Pr[\widehat{Y} = 1 \mid \text{age = elderly}] = 0.165 $, respectively.
+Thus, the statistical parity is $ 0.695 - 0.165 =  0.53 $.
+Next, we compute variance of CPPs as $  \mathsf{Var}[\widehat{Y} = 1\mid \text{age = young}] = 0.212 $ and $  \mathsf{Var}[\widehat{Y} = 1\mid \text{age = elderly}] = 0.138 $.
+Thus, following Lemma~\ref{fairness_fairXplainer_lm:sp_var_relation}, we compute the difference in the scaled  variance of CPPs as $ \frac{0.212}{1 - 0.695} - \frac{0.138}{1 - 0.165} =  0.529 \approx 0.53 $, which coincides with the statistical parity reported earlier.
+\end{repexample}
+
+\subsection{Formulation of FIF}
+\label{fairness_fairXplainer_sec:fif_formulation}
+We are given a binary classifier  $\alg $, a dataset $ \mathbf{D} $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $.
+Our objective is to compute the influences of  features on $ f $.
+Particularly, we compute the influence of each subset of  features $ \feature_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i \mid 1 \le |S_i| \le \numfeatures\} \subseteq [\numfeatures]  $ is a non-empty subset of indices of $ \feature $.
+
+\begin{definition} \textbf{Fairness Influence Function (FIF)}\footnote{For equalized odds, a precise definition of FIF is $ w_{\mathbf{S}}: (\alg, Y, \feature_{\mathbf{S}}) \rightarrow \mathbb{R} $ taking the ground class $ Y $ as an additional input.
+For predictive parity, FIF is defined as $ w_{\mathbf{S}}: (\alg, \widehat{Y}, \feature_{\mathbf{S}}) \rightarrow \mathbb{R} $ taking the predicted class $ \widehat{Y} $ as an additional input.}, denoted by $ w_{\mathbf{S}}: (\alg, \feature_{\mathbf{S}}) \rightarrow \mathbb{R} $, measures the quantitative contribution of features $ \feature_{\mathbf{S}} \subseteq \feature $ on the incurred bias $ f(\alg, \mathbf{D}) $.
+Leveraging the variance-difference representation of $ f(\alg, \mathbf{D}) $ (Lemma~\ref{fairness_fairXplainer_lm:sp_var_relation}) and variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}), we particularly define $ w_{\mathbf{S}} $ as
+\begin{equation}\label{fairness_fairXplainer_eq:fif_decompose}
+	w_{\mathbf{S}}  \triangleq \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}.
+\end{equation}
+Here, given a sensitive group $ \mathbf{a} $, $ V_{\mathbf{a}, \mathbf{S}} \triangleq  \mathsf{Var} _{\feature_{\mathbf{S}}}[\mathbb{E}_{\feature\setminus\feature_{\mathbf{S}}}[\widehat{Y} = 1\mid \sensitive = \mathbf{a}, \feature_{\mathbf{S}}]]- \sum_{\mathbf{S}' \subset \mathbf{S}}{V} _{\mathbf{a}, \mathbf{S}'} $ is the contribution (decomposed variance) of features $ \feature_{\mathbf{S}} $ in $ \mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}] $.
 \end{definition}
-Since FIF is meant to attribute the bias of the classifier to the subset of features, it is natural to choose a function such that the FIFs of all the subsets of features exactly adds up to the total bias.
-This leads to the following additivity axiom.
-%{\fairXplainer} is an axiomatic approach to relate individual and inter-sectional influences with associated fairness metrics, which we formalize in the following.
-\begin{axiom}[Additivity of influence]
-	\label{axm:additivity}
-	Let $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$ be the bias of the classifer and $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ be the FIF of features $ \mathbf{X}_{\mathbf{S}} $ on $ f $.
-The additivity axiom states that the total influences of all possible subset of non-sensitive features is equal to the bias of the classifier.
-	\begin{align}\label{eq:constraint_framework}
-		f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
+
+Informally, the FIF $ w_{\mathbf{S}} $ of $ \feature_{\mathbf{S}} $ is the difference in the scaled decomposed variance of CPPs between sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $ as induced by  $ \feature_{\mathbf{S}} $.
+Thus, FIF of features is non-zero when the scaled decomposed variance-difference of CPPs is non-zero for those features, and vice versa.
+We refer to $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $, and as an \emph{intersectional influence} when $ |\mathbf{S}| > 1 $.
+Being able to naturally quantify the higher-order influences allow FIFs to explain the sources of bias in a more fine-grained manner.
+We experimentally validate this in Section~\ref{fairness_fairXplainer_sec:experiments}.
+
+\paragraph{Properties of FIF} FIF as defined in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose} yields interesting properties, such as decomposability\footnote{Decomposability property is also known as the efficiency property in the context of Shapley values~\cite{roth1988shapley}.}, symmetry, and null properties, which we formally state in Theorem~\ref{fairness_fairXplainer_thm:fif_property}.
+
+\begin{theorem}[Properties of FIF]
+	\label{fairness_fairXplainer_thm:fif_property}
+	Let $ f(\alg, \mathbf{D}) $ be the bias/unfairness of the classifier $ \alg $ on dataset $ \mathbf{D} $ according to linear group fairness metrics such as statistical parity.
+Let $ w_{\mathbf{S}}  $ be the FIF of a subset of  features $ \feature_{\mathbf{S}} $ as defined in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose}.
+	\begin{enumerate}
+		\item[(a)] \textit{The decomposability property} of FIF states that the sum of FIFs of all subset of  features is equal to the bias of the classifier. 
+		\begin{align}\label{fairness_fairXplainer_eq:constraint_framework}
+		\sum_{\mathbf{S} \subseteq [\numfeatures] } w_{\mathbf{S}} = f(\alg, \mathbf{D})
 		\end{align}
-\end{axiom}
-\noindent\textbf{Computing FIF: Reduction to Variance Decomposition.} Henceforth, our objective is to design an estimator of FIF $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}.
-In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.
+		\item[(b)] \textit{The symmetry property} states that  two features $ Z_i $ and $ Z_j $ are equivalent based on FIF if the sum of corresponding individual influences and the intersectional influences with all other features are the same.
+Mathematically,
+		\begin{align}
+		\sum_{\mathbf{S}'' \subseteq [m]\setminus\{i,j\}}w_{\mathbf{S}''\cup \{i\}} = \sum_{\mathbf{S}'' \subseteq [m]\setminus\{i,j\}}w_{\mathbf{S}''\cup \{j\}}  
+		\end{align}
+		if $\sum_{\mathbf{S}' \subseteq \mathbf{S}\cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{j\}} w_{\mathbf{S}'}$ for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ containing neither $ i $ nor $ j $.
+ 		\item[(c)] \textit{The null property} of FIF states that feature $ X_i $ s a dummy or neutral feature if sum of its individual influence and the intersectional influences with all other features is zero.
+Mathematically,
+	\begin{align}
+	 \sum_{\mathbf{S}'' \subseteq [m]\setminus \{i\}}w_{\mathbf{S}''\cup \{i\}} = 0 	 
+	\end{align}
+	if	$\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S}} w_{\mathbf{S}'}$ for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ that does not contain $ i $.
+\end{enumerate}
+\end{theorem}
+We emphasize that the decomposability property proposed here is global, i.e.
+it holds for a whole dataset, but the one used for Shapley-value based explanations~(ref.
+Def.
+1, \cite{lundberg2017unified}), or any local explanation method~\cite{han2022explanation}, is specific to a given data point~\cite{sliwinski2019axiomatic}.
+Thus, they are fundamentally different.
 
-%\begin{comment}
-%	\[
-%	\mathsf{Var}_v[v] = \mathsf{Var}[v]
-%	\]
-%\end{comment}
+The symmetry and null properties of FIFs also distinguish the FIF quantification proposed in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose} from the attribution methods considering only individual features, like SHAP.
+For example, a feature $i$ has zero impact on bias if not only its individual influence but the sum of influences of all the subsets of features that includes $i$ is zero.
+Thus, the symmetry and null properties stated here are by default global and intersectional, and these two aspects are absent in the existing bias explainers.
 
-\noindent\textbf{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction.
-Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $.
-Then, the probability of the random variable is $ p_\mathbf{a} \triangleq \Pr[B_{\mathbf{a}} = 1] = \Pr[\widehat{Y} = 1| \sensitive = \mathbf{a}] $, and variance is $ V_{\mathbf{a}} = p_{\mathbf{a}}(1-p_{\mathbf{a}}) $.
-Therefore, considering two random variables $ B_{\mathbf{a}_{\max}} $ and $ B_{\mathbf{a}_{\min}} $, we state the statistical parity of the classifier as $ p_{a_{\max}} - p_{a_{\min}} = (V_{a_{\max}} - V_{a_{\min}})/(1  - (p_{a_{\max}} + p_{a_{\min}})) $, where $ p_{a_{\max}} $ and $ p_{a_{\min}} $ is the probability of positive prediction of the classifier for the group $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively.
-Intuitively, the statistical parity of the classifier is the scaled (by $ 1/(1  - (p_{a_{\max}} + p_{a_{\min}})) $) difference of the conditional variance of the positive prediction of the classifier over the most and the least favored sensitive groups.
+\paragraph{Bias Amplifying and Eliminating Features.} The sign of $ w_{\mathbf{S}} $ indicates whether features $ \feature_{\mathbf{S}} $ amplify the bias of the classifier or eliminate it.
+When the scaled decomposed variance of CPPs w.r.t.\ features $ \feature_{\mathbf{S}} $ for the sensitive group $ \mathbf{a}_{\max} $ is higher than the group $ \mathbf{a}_{\min} $, $ w_{\mathbf{S}} > 0 $.
+As such, $ \feature_{\mathbf{S}} $ increase bias.
+Conversely, when $ w_{\mathbf{S}} < 0 $, $ \feature_{\mathbf{S}} $ eliminates bias, and improves fairness.
+Finally, features $ \feature_{\mathbf{S}} $ are neutral in bias when $ w_{\mathbf{S}} = 0 $.
 
-\begin{example}
-	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively.
-Thus, the statistical parity of the classifier is $ p_{a_{\max}} - p_{a_{\min}} = 0.704 - 0.172 =  0.532 $.
-Now, we compute conditional variances of positive prediction as $ V_{a_{\max}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } < 40] = 0.208 $ and $ V_{a_{\min}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } \ge 40] = 0.142 $.
-Thus, following our argument, we compute the scaled difference in conditional variances as $ (0.208 - 0.142)/(1 - (0.702 + 0.172)) = 0.532 $, which coincides with the statistical parity.
-\end{example}
-Now, we apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}) for both the conditional variances $ V_{\mathbf{a}_{\max}} $ and $ V_{\mathbf{a}_{\min}} $ to compute the FIF of features, as stated in Theorem~\ref{lemma:fif}.
-\begin{theorem}[FIF as Difference of Conditional Variances]
-	\label{lemma:fif}
-	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied.
-Then, we can express FIF of $ \nonsensitive_{\mathbf{S}} $ as
-	\begin{equation}\label{eq:fif_decompose}
-	w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])}
-	\end{equation}
-	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
-\end{theorem}
-For brevity of space. we defer the detailed proof to Appendix~\ref{sec:proof_1}.
+Now, we present the following two propositions that relate the sign of $ w_{\mathbf{S}} $ with the decomposed variance of CPPs.
 
-\noindent\textbf{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.
+\begin{proposition}\label{fairness_fairXplainer_prop:neg_fif}
+	When $ w_{\mathbf{S}} < 0 $, i.e.
+features $ \feature_{\mathbf{S}} $ decrease bias, the decomposed variance of CPPs w.r.t.\ $ \feature_{\mathbf{S}} $ follows $ V_{\mathbf{a}_{\max}, \mathbf{S}} < V_{\mathbf{a}_{\min}, \mathbf{S}}  $.
+\end{proposition}
+Proposition~\ref{fairness_fairXplainer_prop:neg_fif} implies that if a subset of features $ \feature_{\mathbf{S}} $ is bias-eliminating, the conditional variance in positive outcomes induced by $ \feature_{\mathbf{S}} $ is smaller for the most favoured group than that of the least favoured group.
+\begin{proposition}\label{fairness_fairXplainer_prop:pos_fif}
+{If the decomposed variance of CPPs w.r.t.\ $ \feature_{\mathbf{S}} $ satisfies $ V_{\mathbf{a}_{\max}, \mathbf{S}} > {V_{\mathbf{a}_{\min}, \mathbf{S}}} $, the corresponding FIF $ w_{\mathbf{S}} > 0  $, i.e.
+features $ \feature_{\mathbf{S}} $ increase bias.}
+\end{proposition}
+Proposition~\ref{fairness_fairXplainer_prop:pos_fif} implies that if the conditional variance in positive outcomes induced by $ \feature_{\mathbf{S}} $ is larger for the most favoured group than that of the least favoured group, then this subset of features $ \feature_{\mathbf{S}} $ is bias-inducing.
 
-1.
-\textit{Algorithm Design.} Expressing FIF in terms of the variance decomposition over subset of features allows us to import and extend well-studied techniques of global sensitivity analysis to perform FIF estimation accurately and at scale.
-%To prove Lemma~\ref{lemma:fif}, we present that the statistical parity of a classifier is equal to the scaled difference in the conditional variance of positive prediction of the classifier over specific sensitive groups.
-Later, we apply variance decomposition to derive the expression for $ w_{\mathbf{S}} $.
+\paragraph{Special Cases.} Since our FIF formulation is based on the variance of prediction, for (degenerate) cases when the conditional prediction of the classifier is always positive or always negative for any sensitive group, the variance of prediction becomes zero.
+This observation results in following two propositions.
 
-2.
-\textit{Bias Amplifying and Eliminating Features.} The sign of a FIF $ w_{\mathbf{S}} $ denotes whether the features $ \nonsensitive_{\mathbf{S}} $ are amplifying the bias of the classifier or eliminating it.
-When $ w_{\mathbf{S}} > 0 $, $ \nonsensitive_{\mathbf{S}} $ increases bias.
-When $ w_{\mathbf{S}} < 0 $, $ \nonsensitive_{\mathbf{S}} $ eliminates bias, and thus improves fairness.
-The features $ \nonsensitive_{\mathbf{S}} $ are neutral when $ w_{\mathbf{S}} = 0 $.
+\begin{proposition}[Perfectly Unbiased Classifiers]
+	When $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] $ and both conditional probabilities are either $ 0 $ or $ 1 $, the FIF $w_{\mathbf{S}}  = 0 $ for all subsets of features $ \feature_{\mathbf{S}} $.
+\end{proposition}
+When $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] $ for a classifier, it means that the classifier equally yields positive predictions for each of the sensitive groups.
+Thus, there is no bias (in terms of statistical parity) in the classifier outcome.
+In that case, our formulation of FIF yields $w_{\mathbf{S}}  = 0 $ for all the subsets of features, and leads to a degenerate conclusion that all subsets of features are neutral or zero-bias inducing.
+\begin{proposition}[Perfectly Biased Classifier]
+	{When statistical parity is $ 1 $, i.e.
+$ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = 1 $ and $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] = 0 $, the sensitive features $ \sensitive $ are solely responsible for bias.
+In this case, the FIF $ w_{\mathbf{S}} = 0 $ for all features.}
+\end{proposition}
+When a classifier always yields positive predictions for the most favoured group and only negative predictions for the least favoured group, we obtain $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = 1 $ and $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] = 0 $.
+This is a perfectly biased classifier, which can be expressed by the binary rule $\widehat{Y} = 1$ if $\sensitive = \mathbf{a}_{\max}$ and $0$ otherwise.
+As the discrimination is only due to the sensitive features, our FIF formulation cannot attribute any bias to the any other subset of  features, which leads to $ w_{\mathbf{S}} = 0 $ for all $\mathbf{S}$.
 
-3.
-\textit{Degenerate Cases:} If the classifier always makes positive prediction or never makes positive prediction, i.e.
-$ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{1, 0\} $ for any sensitive group $ \sensitive = \mathbf{a} $, the conditional variance $ \mathsf{Var}[\widehat{Y} = 1 | \sensitive = \mathbf{a}]  = 0 $.
-When total variance is zero, the decomposed variances $V_{\mathbf{a},\mathbf{S}}$ may become zero for each feature combination $\nonsensitive_\mathbf{S}$, and Equation~\eqref{eq:fif_decompose} cannot trivially compute fairness influence functions.
-Though such degenerate cases rarely occur for real-life data, we can avoid them by randomly perturbing at least one sample in the dataset to yield the opposite predicted class.
-
-4.
-\textit{Numerical Instability.} In Equation~\eqref{eq:fif_decompose}, the conditional probabilities of positive prediction of the classifier $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}]  $ and $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]  $ are constant.
-When $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]   = 1 $, the denominator in $ w_{\mathbf{S}} $ becomes $ 0 $ and hence, $ w_{\mathbf{S}} $ is undefined.
-In this case, we add/subtract a small number in the denominator to avoid numerical instability of division though we have not encountered any such case in our experiments.%\red{However, the interpretation of influence functions such as whether $ \nonsensitive_{\mathbf{S}} $ improve or worsen fairness becomes inconsistent due to the choice of addition/subtraction.}
+\textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Section~\ref{fairness_fairXplainer_sec:fairxplainer}.

### 2023-02-07 chapters/fairness/fif/framework_prev.tex
adding fairXplainer

%\clearpage
\section{Fairness Influence Functions: Definition and Computation}\label{sec:fifs}
In this section, we discuss our contribution of formalizing and computing Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier on a dataset. The key idea in the formalization of FIFs is their additive formulation. The key idea in the computation of FIFs is to relate bias with the scaled difference of the conditional variance of positive prediction of the classifier over specific sensitive groups, and to apply variance decomposition to compute individual and intersectional FIFs. %\red{In this section, we first formalize fairness influence functions, and then discuss a variance decomposition based approach to compute FIFs for group fairness metrics.}\todo{Structure: 1. Background (measuring bias using fairness metrics) 2.FIFs as a measure of feature bias 3. Reducing FIF estimation to variance decomposition 4. An algorithm to do the variance decomposition}

%in Section~\ref{??}, and conclude by extending influence computation to other group fairness metrics in Section~\ref{??}
%\subsection{An Axiomatic Formulation}
%\label{sec:problem_formulation}
\textbf{FIF: An Axiomatic Formulation.} We are given a binary classifier  $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $, a dataset $ \mathbf{D} = \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $. Our objective is to compute the influences of non-sensitive features on $ f $. We compute influence of each subset of non-sensitive features $ \mathbf{X}_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i | 1 \le S_i \le \numnonsensitive\} \subseteq [\numnonsensitive] \setminus \emptyset $ is a non-empty subset of indices.

\begin{definition}[Fairness Influence Function]
	Fairness Influence Function (FIF) $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ measures the quantitative contribution of the subset of features $ \mathbf{X}_{\mathbf{S}} \subseteq \mathbf{X}_{[k]}$ on the incurred bias $ f(\alg, \mathbf{D}) $ of the classifier $ \alg $ for dataset $\mathbf{D}$. We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $\textemdash particularly, a second-order influence when $ |\mathbf{S}| = 2 $ and a higher-order influence when $ 3 \le |\mathbf{S}| \le \numnonsensitive $. %\unsure{What do we mean by function here? Isn't FIF scalar?}
\end{definition}	
Since FIF is meant to attribute the bias of the classifier to the subset of features, it is natural to choose a function such that the FIFs of all the subsets of features exactly adds up to the total bias. This leads to the following additivity axiom.
%{\framework} is an axiomatic approach to relate individual and inter-sectional influences with associated fairness metrics, which we formalize in the following.
\begin{axiom}[Additivity of influence]
	\label{axm:additivity}
	Let $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$ be the bias of the classifer and $ w_{\mathbf{S}}:\mathbf{X}_{\mathbf{S}} \rightarrow \mathbb{R} $ be the FIF of features $ \mathbf{X}_{\mathbf{S}} $ on $ f $. The additivity axiom states that the total influences of all possible subset of non-sensitive features is equal to the bias of the classifier.
	\begin{align}\label{eq:constraint_framework}
		f(\alg, \mathbf{D}) = \sum\nolimits_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} w_{\mathbf{S}}
	\end{align}
\end{axiom}
We emphasize that the additivity axiom proposed here is global, i.e. it holds for a whole dataset, but the one used for Shapley-value based explanations~(ref. Def. 1, \citep{lundberg2017unified}), or any local explanation, is for a given data point. Thus, they are fundamentally different.
\textbf{Computing FIF: Reduction to Variance Decomposition.} The additivity axiom itself does not imply an unique solution of FIFs. The axiom instead constrains that the sum of FIFs is equal to the bias of the classifier. Henceforth, the objective of our paper is to design a unique estimator of FIFs $ w_\mathbf{S} $ that satisfies Axiom~\ref{axm:additivity}. In the following, we discuss the closed form representation of $ w_{\mathbf{S}} $ for statistical parity fairness metric and extend to other metrics in Section~\ref{sec:fairxplainer}.

%\begin{comment}
%	\[
%	\mathsf{Var}_v[v] = \mathsf{Var}[v]
%	\]
%\end{comment}

\textbf{Connecting Statistical Parity and Conditional Variance.} Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive group of the classifier, respectively, according to their conditional probability of positive prediction. Next, we consider a Bernoulli random variable $ B_{\mathbf{a}} $ to denote the positive prediction of the classifier for the sensitive group $ \sensitive = \mathbf{a} $. Then, the probability of the random variable is $ p_\mathbf{a} \triangleq \Pr[B_{\mathbf{a}} = 1] = \Pr[\widehat{Y} = 1| \sensitive = \mathbf{a}] $, and variance is $ V_{\mathbf{a}} = p_{\mathbf{a}}(1-p_{\mathbf{a}}) $. Therefore, considering two random variables $ B_{\mathbf{a}_{\max}} $ and $ B_{\mathbf{a}_{\min}} $, we state statistical parity of the classifier as $ p_{a_{\max}} - p_{a_{\min}} = (V_{a_{\max}} - V_{a_{\min}})/(1  - (p_{a_{\max}} + p_{a_{\min}})) $, where $ p_{a_{\max}} $ and $ p_{a_{\min}} $ is the probability of positive prediction of the classifier for the group $ \sensitive = \mathbf{a}_{\max} $ and $ \sensitive = \mathbf{a}_{\min} $, respectively. Intuitively, the statistical parity of the classifier is the scaled (by $ 1/(1  - (p_{a_{\max}} + p_{a_{\min}})) $) difference of the conditional variance of the positive prediction of the classifier over the most and the least favored sensitive groups.

\begin{example}
	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the conditional probabilities of positive prediction of the decision tree for the most and least favored groups are $  p_{a_{\max}} = \Pr[\widehat{Y} = 1 | \text{age } < 40] = 0.704 $ and $ p_{a_{\min}} = \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.172 $, respectively. Thus, the statistical parity of the classifier is $ p_{a_{\max}} - p_{a_{\min}} = 0.704 - 0.172 =  0.532 $. Now, we compute conditional variances of positive prediction as $ V_{a_{\max}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } < 40] = 0.208 $ and $ V_{a_{\min}} = \mathsf{Var}[\widehat{Y} = 1| \text{age } \ge 40] = 0.142 $. Thus, following our argument, we compute the scaled difference in conditional variances as $ (0.208 - 0.142)/(1 - (0.702 + 0.172)) = 0.532 $, which coincides with the statistical parity.
\end{example}
Now, we apply variance decomposition (Eq.~\eqref{eq:variance_decomposition_set_notation}) for both the conditional variances $ V_{\mathbf{a}_{\max}} $ and $ V_{\mathbf{a}_{\min}} $ to compute the FIF of features, as stated in Theorem~\ref{lemma:fif}.
\begin{theorem}[FIF as Difference of Conditional Variances]
	\label{lemma:fif}
	Let $ V_{\mathbf{a}, \mathbf{S}} \triangleq \mathsf{Var}_{\nonsensitive_{\mathbf{S}}}[\widehat{Y} = 1|\sensitive = \mathbf{a}] $ be the decomposed conditional variance of the classifier's positive prediction for the sensitive group $ \sensitive = \mathbf{a} $, where features $ \nonsensitive_{\mathbf{S}} $ are jointly varied. Then, we can express FIF of $ \nonsensitive_{\mathbf{S}} $ as
	\begin{equation}\label{eq:fif_decompose}
		w_{\mathbf{S}}  = \frac{V_{\mathbf{a}_{\max},\mathbf{S}} - V_{\mathbf{a}_{\min},\mathbf{S}}}{1 - (\Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 |  \sensitive = \mathbf{a}_{\min}])}
	\end{equation}
	and FIFs defined by Equation~\eqref{eq:fif_decompose} satisfy Axiom~\ref{axm:additivity}.
\end{theorem}
For brevity of space. we defer the detailed proof to App.~\ref{sec:proof_1}.

\textbf{Consequences of Theorem~\ref{lemma:fif}.} Here, we discuss the algorithmic consequences of Theorem~\ref{lemma:fif} and probable issues that might appear in computation.

1. \textit{Algorithm Design.} Expressing FIF in terms of the variance decomposition over subset of features allows us to import and extend well-studied techniques of global sensitivity analysis to perform FIF estimation accurately and at scale.
%To prove Lemma~\ref{lemma:fif}, we present that the statistical parity of a classifier is equal to the scaled difference in the conditional variance of positive prediction of the classifier over specific sensitive groups. Later, we apply variance decomposition to derive the expression for $ w_{\mathbf{S}} $.

2. \textit{Bias Amplifying and Eliminating Features.} The sign of a FIF $ w_{\mathbf{S}} $ denotes whether the features $ \nonsensitive_{\mathbf{S}} $ are amplifying the bias of the classifier or eliminating it. When $ w_{\mathbf{S}} > 0 $, $ \nonsensitive_{\mathbf{S}} $ increases bias. When $ w_{\mathbf{S}} < 0 $, $ \nonsensitive_{\mathbf{S}} $ eliminates bias, and improves fairness. The features $ \nonsensitive_{\mathbf{S}} $ are neutral when $ w_{\mathbf{S}} = 0 $.

3. \textit{Degenerate Cases:} If the classifier always makes positive prediction or never makes positive prediction, i.e. $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}] \in \{1, 0\} $ for any sensitive group $ \sensitive = \mathbf{a} $, the conditional variance $ \mathsf{Var}[\widehat{Y} = 1 | \sensitive = \mathbf{a}]  = 0 $. When total variance is zero, the decomposed variances $V_{\mathbf{a},\mathbf{S}}$ may become zero for each feature combination $\nonsensitive_\mathbf{S}$, and Equation~\eqref{eq:fif_decompose} cannot trivially compute fairness influence functions. Though such degenerate cases rarely occur for real-life data, we can avoid them by randomly perturbing at least one sample in the dataset to yield the opposite predicted class.

\[
w_s = \frac{V_{a, s}}{1 - p_a} - \frac{V_{b, s}}{1 - p_b}
\]
\red{[Corollary:]}
Flags.
\begin{itemize}
	\item When $ p_a = p_b = 0 \text{ or } 1,  w_s = 0 $
	\item When $ p_{a} = 1,  p_b = 0  $, sensitive feature is responsible for bias. A checker for the code.
	\item $ V_{a, s} = V_{b, s} $, $ w_s  > 0$ if $ p_a > p_b $ and vice versa.
\end{itemize}

\todo{Shapely proof.}

4. \textit{Numerical Instability.} In Equation~\eqref{eq:fif_decompose}, the conditional probabilities of positive prediction of the classifier $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}]  $ and $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]  $ are constant. When $ \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\max}] + \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}_{\min}]   = 1 $, the denominator in $ w_{\mathbf{S}} $ becomes $ 0 $ and hence, $ w_{\mathbf{S}} $ is undefined. In this case, we add/subtract a small number in the denominator to avoid numerical instability of division though we have not encountered any such case in our experiments.%\red{However, the interpretation of influence functions such as whether $ \nonsensitive_{\mathbf{S}} $ improve or worsen fairness becomes inconsistent due to the choice of addition/subtraction.}

### 2023-02-07 chapters/fairness/fif/framework_temp.tex
adding fairXplainer

%\clearpage
\section{Fairness Influence Functions: Formulation and Properties}\label{sec:fifs}
We formalize Fairness Influence Functions (FIF) as a quantifier of the contribution of a subset of features to the resultant bias of a classifier applied on a dataset.  In GSA, we observe that the variance of the output of a function can be attributed to the corresponding subset of input variables through decomposing variance (Eq.~\eqref{eq:variance_decomposition_set_notation}). To leverage the power of GSA in fairness in machine learning, we first represent existing fairness metric in terms of the variance of classifier's prediction in Section~\ref{sec:metric_as_variance}. This allows us to formulate FIF based on variance decomposition in Section~\ref{sec:fif_formulation}.

\subsection{Fairness Metrics as the Variance of Prediction}
\label{sec:metric_as_variance}
\blue{We refer the indicator function $ \mathds{1}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $ as CPP (Conditional Positive Prediction) of a classifier, which is the central function in existing fairness metrics.} Henceforth, fairness metrics such as statistical parity\footnote{For equalized odds, CPP has an additional conditional $ Y = y $.} are a functional of the probability of CPPs for different sensitive groups~\cite{ghosh2020justicia,benesse2021fairness}.
%For indicator functions, probability and expectation are equal.
Our key idea of computing FIFs of features is to represent fairness metrics using the variance of CPPs. 
%To this end, existing fairness metrics such as statistical parity is the first moment (probability or expectation) of prediction distribution while variance is the second moment of distribution.  
Formally, we express statistical parity using the variance of CPPs in Lemma~\ref{lm:sp_var_relation}.\footnote{\blue{We demonstrate similar derivation for equalized odds and predictive parity in the Appendix.}}

\begin{lemma}
	\label{lm:sp_var_relation}
	Let $ \mathbf{a}_{\max} = \argmax_{\mathbf{a}} \Pr[\widehat{Y} = 1 \mid  \sensitive = \mathbf{a}] $ and $ \mathbf{a}_{\min} = \argmin_{\mathbf{a}} \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $ be the most favored and the least favored sensitive groups, respectively. The statistical parity of a binary\footnote{For a multi-class classifier, statistical parity of the target class $ y \in \mathbb{N} $ is $ \frac{\mathsf{Var}[\widehat{Y} = y \mid \sensitive = \mathbf{a}_{\max}]}{1 - \Pr[\widehat{Y} = y \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = y \mid \sensitive = \mathbf{a}_{\min}] }{1 - \Pr[\widehat{Y} = y \mid  \sensitive = \mathbf{a}_{\min}]}. $} classifier is the difference in the scaled variance of CPPs. Formally, 
	\[
	 f_{\mathsf{SP}}(\alg, \mathbf{D}) = \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\min}] }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}.
	\]
\end{lemma}
\begin{proof}
	For a sensitive group $ \mathbf{a} $, CPP  is a Bernoulli random variable, where probability\footnote{\red{For any binary event $ E $, expectation and probability are identical, $ \mathbb{E}[\mathds{1}(E)] = \Pr[E = 1] $.}} $ p_{\mathbf{a}}  = \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $ and variance $ V_{\mathbf{a}} = \mathsf{Var}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] = p_{\mathbf{a}} (1 - p_{\mathbf{a}}) $. Thus, for sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $, the statistical parity of the classifier is $ p_{\mathbf{a}_{\max}}  - p_{\mathbf{a}_{\min}} =  \frac{V_{\mathbf{a}_{\max}}}{1 - p_{\mathbf{a}_{\max}}}  - \frac{V_{\mathbf{a}_{\min}}}{1 - p_{\mathbf{a}_{\min}}} $. Replacing $ 1 - p_\mathbf{a} = \Pr[\widehat{Y} = 0 \mid \sensitive = \mathbf{a}] $  proves the lemma.
\end{proof}

\begin{example}
	In Example~\ref{ex:motivating_example} (Figure~\ref{fig:dt_original}), the probability of CPPs of the decision tree for the most and least favored groups are $  \Pr[\widehat{Y} = 1 \mid \text{age = young}] = 0.695 $ and $ \Pr[\widehat{Y} = 1 \mid \text{age = elderly}] = 0.165 $, respectively. Thus, the statistical parity is $ 0.695 - 0.165 =  0.53 $. Next, we compute variance of CPPs as $  \mathsf{Var}[\widehat{Y} = 1\mid \text{age = young}] = 0.212 $ and $  \mathsf{Var}[\widehat{Y} = 1\mid \text{age = elderly}] = 0.138 $. Thus, following Lemma~\ref{lm:sp_var_relation}, we compute the difference in the scaled  variance of CPPs as $ \frac{0.212}{1 - 0.695} - \frac{0.138}{1 - 0.165} =  0.529 \approx 0.53 $, which coincides with the statistical parity.
\end{example}

\subsection{Formulation of FIF}
\label{sec:fif_formulation}
We are given a binary classifier  $\alg $, a dataset $ \mathbf{D} $, and a  fairness metric $ f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0} $. Our objective is to compute the influences of  features on $ f $. Particularly, we compute the influence of each subset of  features $ \feature_{\mathbf{S}} $, where $ \mathbf{S} = \{S_i \mid 1 \le |S_i| \le \numfeatures\} \subseteq [\numfeatures]  $ is a non-empty subset of indices of $ \feature $.

\begin{definition} Fairness Influence Function (FIF)\footnote{For equalized odds, a precise definition of FIF is $ w_{\mathbf{S}}: (\alg, Y, \feature_{\mathbf{S}}) \rightarrow \mathbb{R} $ taking the ground class $ Y $ as an additional input.}, denoted by $ w_{\mathbf{S}}: (\alg, \feature_{\mathbf{S}}) \rightarrow \mathbb{R} $, measures the quantitative contribution of features $ \feature_{\mathbf{S}} \subseteq \feature $ on the incurred bias $ f(\alg, \mathbf{D}) $. Representing $ f(\alg, \mathbf{D}) $ as variance-difference (Lemma~\ref{lm:sp_var_relation}), we particularly define $ w_{\mathbf{S}} $ based on variance decomposition.
\begin{equation}\label{eq:fif_decompose}
	w_{\mathbf{S}}  \triangleq \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}
\end{equation}

where, for a sensitive group $ \mathbf{a} $, $ V_{\mathbf{a}, \mathbf{S}} =  \mathsf{Var} _{\feature_{\mathbf{S}}}[\mathbb{E}_{\feature\setminus\feature_{\mathbf{S}}}[\widehat{Y} = 1\mid \sensitive = \mathbf{a}, \feature_{\mathbf{S}}]]- \sum_{\mathbf{S}' \subset \mathbf{S}}{V} _{\mathbf{a}, \mathbf{S}'} $ is the contribution (decomposed variance) of features $ \feature_{\mathbf{S}} $ in $ \mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}] $. Informally, the FIF $ w_{\mathbf{S}} $ of $ \feature_{\mathbf{S}} $ is the difference in the scaled decomposed variance of CPPs between sensitive groups $ \mathbf{a}_{\max} $ and $ \mathbf{a}_{\min} $ as induced by  $ \feature_{\mathbf{S}} $. Thus, FIF of features is non-zero when the scaled decomposed variance-difference of CPPs is non-zero for those features, and vice versa. We refer $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $ and an \emph{inter-sectional influence} when $ |\mathbf{S}| > 1 $.

\end{definition}

FIF in Eq.~\eqref{eq:fif_decompose} results in different desired properties, such as decomposability\footnote{ Decomposability property is also known as efficiency property in the context of Shapley values~\cite{roth1988shapley}.}, symmetry and null FIF properties, which we formally state in Theorem~\ref{thm:fif_property}.

\begin{theorem}
	\label{thm:fif_property}
	Let $ f(\alg, \mathbf{D}) $ be the bias/unfairness of the classifier $ \alg $ on dataset $ \mathbf{D} $ according to linear group fairness metrics such as statistical parity. Let $ w_{\mathbf{S}}  $ be the FIF of a subset of  features $ \feature_{\mathbf{S}} $ as defined in Eq.~\eqref{eq:fif_decompose}. 
	\begin{enumerate}
		\item[(a)] \textit{The decomposability property} of FIF states that the sum of FIFs of all subset of  features is equal to the bias of the classifier. 
		\begin{align}\label{eq:constraint_framework}
		\sum_{\mathbf{S} \subseteq [\numfeatures] } w_{\mathbf{S}} = f(\alg, \mathbf{D})
		\end{align}
		\item[(b)] \textit{The symmetry property} of FIF states that  two features $ X_i $ and $ X_j $ are \red{equivalent} based on FIF, when
		\begin{align}
		w_{\{i\}} = w_{\{j\}} \text{ if } \sum_{\mathbf{S}' \subseteq \mathbf{S}\cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{j\}} w_{\mathbf{S}'} 
		\end{align}
		for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ containing neither $ i $ nor $ j $. 
		\item[(c)] 	\textit{The null property} of FIF states that feature $ X_i $ \blue{becomes a null or neutral feature feature based on FIF}, when
		\begin{align}
		w_{\{i\}} = 0 	\text{ if }	\sum_{\mathbf{S}' \subseteq \mathbf{S} \cup \{i\}} w_{\mathbf{S}'} = 		\sum_{\mathbf{S}' \subseteq \mathbf{S}} w_{\mathbf{S}'} 
		\end{align}
		for every non-empty subset $ \mathbf{S} $ of $ [\numfeatures] $ that does not contain $ i $.

	\end{enumerate}
\end{theorem}

\begin{proof}
	The decomposability property of FIF is based on GSA, where the total variance is decomposed to the variances of individual and intersectional inputs. 
	\begin{align*}
		\sum_{\mathbf{S} \subseteq [\numfeatures]} w_{\mathbf{S}} &= \sum_{\mathbf{S} \subseteq [\numfeatures]} \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
		&= \frac{\sum_{\mathbf{S} \subseteq [\numfeatures]} V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\sum_{\mathbf{S} \subseteq [\numfeatures]} V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
		&= \frac{V_{\mathbf{a}_{\max}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
		&=\frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{\mathsf{Var}[\widehat{Y} = 1\mid\sensitive = \mathbf{a}_{\min}] }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]}\\
		&= f_{\mathsf{SP}}(\alg, \mathbf{D})
	\end{align*}
	Thus, applying Lemma~\ref{lm:sp_var_relation}, we prove the decomposability property of FIF for statistical parity.
	\todo{Proof of symmetry and null property.}

\end{proof}

We emphasize that the decomposability property proposed here is global, i.e. it holds for a whole dataset, but the one used for Shapley-value based explanations~(ref. Def. 1, \citep{lundberg2017unified}), or any local explanation method, is for a given data point. Thus, they are fundamentally different.

\paragraph{Bias Amplifying and Eliminating Features.} The sign of $ w_{\mathbf{S}} $ denotes whether features $ \feature_{\mathbf{S}} $ amplify the bias of the classifier or eliminate it. When the scaled decomposed variance of CPPs w.r.t.\ features $ \feature_{\mathbf{S}} $ for the sensitive group $ \mathbf{a}_{\max} $ is higher than the group $ \mathbf{a}_{\min} $, $ w_{\mathbf{S}} > 0 $. As such, $ \feature_{\mathbf{S}} $ increase bias. Conversely, when $ w_{\mathbf{S}} < 0 $, $ \feature_{\mathbf{S}} $ eliminates bias, and improves fairness. Finally, features $ \feature_{\mathbf{S}} $ are neutral in bias when $ w_{\mathbf{S}} = 0 $. The following two propositions relate the sign of $ w_{\mathbf{S}} $ with the decomposed variance of CPPs.

\begin{proposition}
	When $ w_{\mathbf{S}} < 0 $, $ V_{\mathbf{a}_{\max}, \mathbf{S}} \le V_{\mathbf{a}_{\min}, \mathbf{S}}  $
\end{proposition}

\begin{proof}
	When $ w_{\mathbf{S}} < 0 $,
	\begin{align*}
		&\frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} - \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} < 0 \\
		& \Rightarrow \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]} < \frac{V_{\mathbf{a}_{\min}, \mathbf{S}} }{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \\
		& \Rightarrow  \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} < \frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \le 1 \\
		& \Rightarrow \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} \le 1 \\ 
		& \Rightarrow V_{\mathbf{a}_{\max}, \mathbf{S}} \le V_{\mathbf{a}_{\min}, \mathbf{S}}
	\end{align*}
\end{proof}

\begin{proposition}
	Let $ n_{\max} $ be the number of data points in the dataset $ \mathbf{D} $ belonging to the most favored sensitive group $ \mathbf{a}_{\max} $ and $ \Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}] \ne 0 $. Therefore, when $ w_{\mathbf{S}} > 0  $, $ V_{\mathbf{a}_{\max}, \mathbf{S}} \ge \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{n_{\max}} $.
\end{proposition}

\begin{proof}
Let $ n_{\min} $ be the number of data points in the dataset $ \mathbf{D} $ belonging to the least favored sensitive group $ \mathbf{a}_{\min} $. Then,	when $ w_{\mathbf{S}} > 0 $,
\begin{align*}
	& \frac{V_{\mathbf{a}_{\max}, \mathbf{S}}}{V_{\mathbf{a}_{\min}, \mathbf{S}}} > \frac{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\max}]}{\Pr[\widehat{Y} = 0 \mid  \sensitive = \mathbf{a}_{\min}]} \ge \frac{\frac{1}{n_{\max}}}{\frac{n_{\min}}{n_{\min}}}  = \frac{1}{n_{\max}} \\
	& \Rightarrow V_{\mathbf{a}_{\max}, \mathbf{S}} \ge \frac{V_{\mathbf{a}_{\min}, \mathbf{S}}}{n_{\max}}
\end{align*}

\end{proof}

\paragraph{Special Cases.} Since our FIF formulation is based on the variance of prediction, for (degenerate) cases when the conditional prediction of the classifier is always positive or always negative for any sensitive group, the variance of prediction becomes zero. This observation results in following two propositions.

\begin{proposition}
	When statistical parity is $ 0 $, i.e. $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] $, the FIF $w_{\mathbf{S}}  = 0 $ for all subsets of features $ \feature_{\mathbf{S}} $.
\end{proposition}
\begin{proposition}
	When statistical parity is $ 1 $, i.e. $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ] = 1 $ and $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ] = 0 $, the sensitive features $ \sensitive $ are solely responsible for bias. In this case, the FIF $ w_{\mathbf{S}} = 0 $ for \red{all subsets of features except sensitive features. Hence, the FIF of sensitive features is $ 1 $.} 
\end{proposition}

Expressing FIF in terms of the variance decomposition over subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation, which we elaborate on Section~\ref{sec:fairxplainer}.

### 2023-02-07 chapters/fairness/fif/introduction.tex
adding fairXplainer

--- 2022-06-01 chapters/fairness/fif/introduction.tex
@@
-\section{Introduction}
+\label{chapter:fif}
 %\begin{itemize}
@@
 %\end{itemize}
-The last decades have witnessed a significant progress in machine learning with applications in high-stake decision making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-In such applications, the deployed machine learning classifier often demonstrate bias towards certain demographic groups involved in the data~\cite{dwork2012fairness}.
-For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.
-Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier on a dataset.
-For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.
 
-Although fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
-In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}.
-To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
-Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias.
-FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias.
+The past few decades have witnessed a remarkable advancement in the field of machine learning, with its applications encompassing crucial decision-making processes such as college admissions~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring}, and more.
+In such applications, the deployed machine learning classifier often demonstrates bias towards certain demographic groups present in the data~\cite{dwork2012fairness}.
+For instance, a machine learning classifier employed in college admissions may show a disproportionate preference for White-male candidates over Black-female candidates.
+This can occur due to historical biases present in the admission data, the classifier's accuracy-focused learning objective, or a combination of the two~\cite{landy1978correlates,zliobaite2015relation,berk2019accuracy}.
+Following such phenomena, multiple group-based fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity}, etc.\ have been proposed to quantify the bias of the classifier on a dataset.
+For example, a statistical parity of $ 0.6 $ for the college admission classifier would indicate that White-male candidates are offered admission $ 60\% $ more frequently than Black-female candidates~\cite{garg2020fairness,besse2021survey}.
 
-\noindent\textbf{Our Contributions.} The contribution of this paper is three-fold.
+Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
+Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
+To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
+In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier.
+Thus, \textit{we aim to design a bias attribution method that is global by construction}.
+In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features.
+Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature\cite{buolamwini2018gender,wang2022towards}.
+%\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
+In this paper, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}
 
-\textit{1.
-Formalism:} We propose to compute individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier (Sec~\ref{sec:fifs}).
-The \textit{intersectionality}~\cite{buolamwini2018gender} allows us to detect the higher order interactions among the features.
-In the formalism, we first axiomatize that to be a proper attribution of bias, the sum of FIFs of all subsets of the features is equal to the bias of the classifier (Axiom~\ref{axm:additivity}).
-Following that, we show that computing FIFs over subsets of features is equivalent to computing a scaled difference in the \emph{conditional variances} (Theorem~\ref{lemma:fif}) of the classifier between the sensitive groups of interest.
-This allows us to connect global sensitivity analysis, a standard technique recommended by regulators to asses numerical models~\cite{eu,usepa}, with computing feature influences leading to bias in the classifier's output.%\improvement{Application of GSA in EU and modeling.
-}
+\paragraph{Contributions.}  Our contributions are three-fold.
 
-\textit{2.
-Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for computing FIFs for subsets of features for a given classifier, a dataset, and any group fairness metrics, namely statistical parity, equalized odds, and predictive parity (Sec.~\ref{sec:fairxplainer}).
-The key ideas are to import techniques from Global Sensitivity Analysis (GSA)~\cite{saltelli2008global} to decompose the variance of the prediction of the classifier among the subset of features and to use a local regression method~\cite{loader2006local} to compute FIFs.
+\begin{enumerate}[leftmargin=*]
 
-\textit{3.
-Experimental:} We experimentally validate that {\fairXplainer} is significantly more accurate and can capture the intersectional or joint effect of the features on the bias induced by a classifier (Sec.~\ref{sec:experiments}).
-Improvement in accuracy and extension to intersectional FIFs are both absent in the existing works aimed to this problem~\cite{begley2020explainability,lundberg2020explaining}.
-Also, {\fairXplainer} enables us to detect the change in FIFs due to different fairness enhancing algorithms and fairness reducing attacks, which opens up new avenues to analyze the effects of these algorithms.
+	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Section~\ref{fairness_fairXplainer_sec:fifs}).
+Our method is based on transforming existing fairness metrics into the difference of scaled conditional variances of classifier's prediction, which we then decompose using Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
+FIFs have several desirable properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}), including the decomposability property~\cite{begley2020explainability,lundberg2020explaining}.
+This property states that the sum of FIFs of all individual and intersectional features equals the bias of the classifier.
+With this formulation of FIFs, we can identify which features have the greatest influence on bias by looking at the disparity in their scaled decomposed variance between sensitive groups.
 
-We illustrate the usefulness of our contributions via an example scenario proposed in~\cite{ghosh2020justicia}.
+%	\item \textbf{Formalism:} Given a classifier, a dataset, and a group fairness metric, we propose to estimate individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier induced by the features (Section~\ref{fairness_fairXplainer_sec:fifs}).	Our FIF formulation is based on representing existing fairness metrics as the difference of the scaled conditional variances of classifier's prediction, followed by a decomposition of variance according to Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
+The presented FIF formulation results in different desired properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}): the efficiency property~\cite{begley2020explainability,lundberg2020explaining} states that the sum of FIFs of all subsets of features is equal to the bias of the classifier.
+This allows us to view FIFs as a way to decompose the bias of the classifier among features.
+In this decomposition, features that demonstrate a higher disparity in the scaled decomposed variance of prediction between sensitive groups are attributed to be highly influential on the bias in prediction.
+
+	\item \textbf{Algorithmic:} We propose a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric.
+The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}).
+Building on GSA\cite{saltelli2008global} techniques, {\fairXplainer} solves a local regression problem~\cite{loader2006local} based on cubic splines~\cite{li2010global} to decompose the variance of the classifier's prediction among all the subsets of features.
+
+%	\item \textbf{Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for estimating FIFs of  features for a given classifier, a dataset, and any group fairness metric, namely statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}).
+The key idea is to import techniques from GSA\cite{saltelli2008global} to decompose the variance of the classifier's prediction among all  subsets of features by applying a local regression technique~\cite{loader2006local} based on cubic splines~\cite{li2010global}.
+
+	\item \textbf{Experimental:} We evaluate  {\fairXplainer}  on a variety of real-world datasets and machine learning classifiers to demonstrate its efficiency in estimating the individual and intersectional FIFs of features.
+Our results show that {\fairXplainer} has a higher accuracy in approximating bias using estimated FIFs compared to existing methods (Section~\ref{fairness_fairXplainer_sec:experiments}).
+Our estimation of FIFs also shows a strong correlation with fairness interventions.
+Furthermore, {\fairXplainer} yields more granular explanation of the sources of bias by combining both individual and intersectional FIFs, and also detects patterns that existing fairness explainers cannot.
+Finally, {\fairXplainer} enables us to observe changes in FIFs as a result of different fairness enhancing algorithms~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} and fairness reducing attacks~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning}.
+This creates opportunities to further examine the impact of these algorithms.
+\end{enumerate}
+
+\input{chapters/fairness/fif/figure_intro}
+We illustrate the usefulness of our contributions via an example scenario proposed in\cite{ghosh2020justicia}.
 
@@
 \begin{example}
-	\label{ex:motivating_example}
- Following~\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features `fitness' and `income'.
-	`fitness' and `income' depend on a sensitive feature `age' leading to two sensitive groups ``young" (age $ < 40 $) and ``elderly" (age $ \ge 40 $), as highlighted in (Figure~\ref{fig:dag_age_income_fitness}--\ref{fig:distribution_example}).
+	\label{fairness_fairXplainer_ex:motivating_example}  Following\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features: fitness  and income.
+Fitness and income depend on a sensitive feature age $ \in $ \{young, elderly\} leading to two sensitive groups, as highlighted in (Figure~\ref{fairness_fairXplainer_fig:dag_age_income_fitness}--\ref{fairness_fairXplainer_fig:distribution_example}).
 
-	\textbf{Case study 1:} For each sensitive group, we generate $ 500 $ samples of (income, fitness) and train a decision tree (DT$ 1 $), which predicts without explicitly using the sensitive feature `age' (Figure~\ref{fig:dt_original}).
+	\textbf{Case study 1:} For each sensitive group, we generate $ 1000 $ samples of (income, fitness) and train a decision tree (DT$ 1 $), which predicts without explicitly using the sensitive feature (Figure~\ref{fairness_fairXplainer_fig:dt_original}).
 %This classifier predicts the eligibility for health insurance without any direct effect of the sensitive feature: age.
-	Using standard off-the-shelf techniques, we can compute statistical parity as $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.7 - 0.17 = 0.53 $, and therefore DT$ 1 $ is unfair towards ``elderly".
+	Using the $1000$ samples and corresponding predictions, we estimate the statistical parity of DT1 as $ \Pr[\widehat{Y} = 1  \mid  \text{age = young} ] - \Pr[\widehat{Y} = 1  \mid  \text{age = elderly}] = 0.695 - 0.165 = 0.53 $.
+Therefore, DT$ 1 $ is unfair towards the elderly group.
 
-	Applying techniques developed in this paper, we  investigate the source of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers denote a reduction in fairness and negative numbers denotes fairness improvement}.
-In Figure~\ref{fig:fif_original}, fitness $(\mathrm{FIF} = 0.74)$, and the joint effect of fitness and income $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e.
+	Using the methods described in this paper, we examine the sources of unfairness in DT$ 1 $ by calculating the FIFs of each feature subset.
+Positive values represent a reduction in fairness due to increased statistical parity, while negative values indicate an improvement in fairness.
+The results, shown in the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, indicate that fitness $(\mathrm{FIF} = 0.355)$, income $(\mathrm{FIF} = 0.098)$, and the combination of income and fitness $(\mathrm{FIF} = 0.055)$ contribute to higher statistical parity and bias.
+Fitness, being the root of DT$ 1 $, has a greater impact on statistical parity than income, which is at the second level of DT$ 1 $.
+Our method also reveals the joint effect of income and fitness on statistical parity, which prior methods do not account for (Section~\ref{fairness_fairXplainer_sec:related_work}).
+The total of the FIFs of all features, $ 0.355 + 0.098 + 0.055 = 0.507 \approx 0.53 $, approximately matches the statistical parity of the classifier, providing a way to break down the bias among all feature subsets.
+Note that we discuss the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.
+
+%	Now, applying the techniques developed in this paper, we investigate the sources of unfairness of DT$ 1 $ by computing FIFs of features, where \emph{positive numbers indicate reduction in fairness by increasing statistical parity, while negative numbers indicate improvement in fairness}.
+In the waterfall diagram in Figure~\ref{fairness_fairXplainer_fig:fif_original}, fitness $(\mathrm{FIF} = 0.35)$, income $(\mathrm{FIF}  = 0.1)$, and the joint effect of income and fitness $(\mathrm{FIF}  = 0.05)$ cause higher statistical parity, i.e.
 higher bias.
-This observation is invoked as DT$ 1 $ predicts positively for the higher values of fitness (root node in DT$ 1 $) and elderly individuals often possess lower fitness (Figure~\ref{fig:distribution_example}).
-In contrast, the income of individuals  $(\mathrm{FIF}  = -0.33)$ decreases statistical parity, as elderly individuals have a higher income but lower fitness.
+Intuitively, fitness being at the root of DT$ 1 $ is more influential on statistical parity than income, which is situated at the second level of DT$ 1 $.
+Our method also reveals the joint (or intersectional) effect of income and fitness on statistical parity, which prior approaches do not consider (Section~\ref{fairness_fairXplainer_sec:related_work}).
+Finally, the sum of FIFs of features, $ 0.35 + 0.1 + 0.05 = 0.5 \approx 0.53 $, approximates\footnote{\red{We elaborate on the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.}} the statistical parity of the classifier, thereby demonstrating a way to decompose the bias of the classifier among all the subsets of features.
 
-	\textbf{Case study 2:} Since DT$ 1 $ is unfair to the ``elderly", we learn another decision tree (DT$ 2 $) by applying an affirmative action, where we decrease the threshold on income from $ 0.69 $ to $ 0.555 $ when the fitness is lower ({\color{affirmative}green} node in Figure~\ref{fig:dt_affirmative}).
-This action allows more elderly individuals to receive insurance by including more people with lower fitness, and thus the statistical parity becomes $ \Pr[\widehat{Y} = 1 | \text{age } < 40] - \Pr[\widehat{Y} = 1 | \text{age } \ge 40] = 0.71 - 0.7 = 0.01$.
-This is significantly less than the earlier statistical parity of $ 0.53 $.	Again, applying techniques developed in this paper, we compute FIF of features in Figure~\ref{fig:fif_affirmative}.
-Here, the FIF of income and fitness reflects the reduction in statistical parity as their influences almost nullify each other.
-Thus, FIF depicts the effects of different features and their combinations on the resultant bias incurred by the classifier.
-%As we analyse the source of unfairness, the combined influence of income and fitness changes its direction ($ - 0.01 $ vs.\ $ 0.05 $) towards improving fairness.
+%	\clearpage
+	\textbf{Case study 2:} 	
+	To address the unfairness of DT$ 1 $ towards the elderly, we introduce DT$ 2 $ which applies an affirmative action.
+Specifically, for lower fitness, which is typical for the elderly group (Figure~\ref{fairness_fairXplainer_fig:distribution_example}), we decrease the threshold on income from $ 0.69 $ to $ 0.55 $ ({\color{affirmative}green} node in Figure~\ref{fairness_fairXplainer_fig:dt_affirmative}).
+This allows more elderly individuals to receive insurance as they tend to have higher income, and the lower threshold accommodates their eligibility.
+The statistical parity of DT$ 2 $ is calculated to be $ \Pr[\widehat{Y} = 1 \mid \text{age = young}] - \Pr[\widehat{Y} = 1 \mid \text{age = elderly}] = 0.707 - 0.706 = 0.001 $, which is negligible compared to the earlier statistical parity of $ 0.53 $ in DT$ 1 $.
+We estimate the FIFs of features, with $ - 0.388 $ for income, $ 0.322 $ for fitness, and $ 0.07 $ for both features combined.
+Hence, the negative influence of income confirms the affirmative action, and nullifies the disparity induced by fitness.
+Additionally, the sum of FIFs $ - 0.388 + 0.322 + 0.07 = 0.004 $ coincides with the statistical parity of DT$ 2 $.
+
+%	Since DT$ 1 $ is unfair to the elderly, we learn another decision tree (DT$ 2 $) by applying an affirmative action.
+Specifically, for lower fitness, which is typical for the elderly group (Figure~\ref{fairness_fairXplainer_fig:distribution_example}), we decrease the threshold on income from $ 0.69 $ to $ 0.55 $ ({\color{affirmative}green} node in Figure~\ref{fairness_fairXplainer_fig:dt_affirmative}).
+As {\color{red}the elderly have higher income and} the threshold on income becomes lower for the elderly group, the affirmative action allows more elderly individuals to receive insurance.
+Thus, the statistical parity becomes $ \Pr[\widehat{Y} = 1  \mid  \text{age = young}] - \Pr[\widehat{Y} = 1  \mid  \text{age = elderly}] = 0.71 - 0.71 = 0 $, which is negligible than the earlier statistical parity of $ 0.53 $.	Next, we estimate FIFs of features, which are $ - 0.39 $ for income,  $ 0.32 $ for fitness, and $ 0.07 $ for income and fitness together.
+Hence, the negative influence of income confirms the affirmative action, and nullifies the disparity induced by fitness.
+We also observe that the sum of FIFs $ - 0.39 + 0.32 + 0.07 = 0 $ coincides with the statistical parity of DT$ 2 $.
 \end{example}
 
-\input{chapters/fairness/fif/figure_intro.tex}
+\section{Related Work}
+\label{fairness_fairXplainer_sec:related_work}
+Recently, local explanation methods have been applied to black-box classifiers to explain sources of bias through feature-attribution\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition\cite{pan2021explaining}.
+Our work uses the feature-attribution approach and highlights three limitations of existing methods: (i) a failure to estimate intersectional FIFs, (ii) inaccuracies in approximating bias based on FIFs, and (ii) less correlation of FIFs to fairness intervention.
+Elaborately, to explain group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}.
+They apply a global aggregation (i.e.
+expectation) of all local explanations corresponding to all data points in a dataset.
+Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}).
+In addition, existing methods ignore the joint contribution of correlated features on bias.
+To address these limitations, \textit{we develop a formal framework to explain sources of group unfairness in a classifier and also a novel methodology to estimate FIFs.
+To the best of our knowledge, this is the first work to do both}.
 
-\section{Related Work} Recently, several studies apply \emph{local explanation methods} of black-box prediction to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}.
-Since our work adopts feature-attribution approach, it reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in computing FIFs and (ii) \emph{failing to compute intersectional} FIFs.
-Elaborately, FIF computation in~\cite{begley2020explainability,lundberg2020explaining} is inaccurate because of applying local explainers such as SHAP~\cite{lundberg2017unified} to compute global properties of the classifier such as group fairness.
-In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the unfairness of the classifer.
-Also, these works provide empirical evaluations to justify the choices of SHAP-based tools for explaining fairness and does not consider the global nature of group fairness metrics.
-In this paper, \textit{we develop a formal framework to explain sources of unfairness in a classifier and also a novel methodology to address it.
-To the best of our knowledge, this is the first work to do the both}.
+%Recently, several studies apply \emph{local explanation methods} for black-box classifiers to explain sources of bias by feature-attribution~\cite{begley2020explainability,lundberg2020explaining} and causal path decomposition~\cite{pan2021explaining}.
+Our work adopts feature-attribution approach and reveals two-fold limitations of existing methods: (i) \emph{inaccuracy} in approximating bias in terms of estimated FIFs and (ii) \emph{failing to estimate intersectional} FIFs.
+Elaborately, to explain group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}.
+They apply a global aggregation (such as expectation) of all local explanations corresponding to all data points in a dataset.
+Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}).
+In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the bias of the classifier.
+To this end,  in this paper, \textit{we develop a formal framework to explain sources of group unfairness in a classifier and also a novel methodology to address it.
+To the best of our knowledge, this is the first work to do both}.
 %\todo{1.
@@
 Influence of Yair Zick.}
-Among other related works, \cite{benesse2021fairness} link GSA measures such as Sobol and Cram{\'e}r-von-Mises indices to different fairness metrics.
-While their approach relates the GSA of sensitive features on the resulting bias, we focus on applying GSA to all features to compute FIFs.
-\textit{Their approach only detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all features}.
-In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} compute feature-influence as the shift of bias from its original value by randomly intervening features.
-Their work is different from our axiomatic approach, where the sum of FIFs equals the bias.
+
+Among other related works, \cite{benesse2021fairness} links GSA measures such as Sobol and Cram{\'e}r-von-Mises indices to different fairness metrics.
+While their approach relates the GSA of sensitive features on bias, we focus on applying GSA to all features to estimate FIFs.
+\textit{Their approach detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all feature subsets}.
+In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} estimate feature-influence as the shifting of bias from its original value by randomly intervening features.
+Their work is different from the decomposability property of FIFs, where the sum of FIFs is equal to the bias.
+A separate line of work estimates the fairness influence of data points in a dataset~\cite{li2022achieving,wang2022understanding}, while \textit{our focus is on quantifying influences of input features}.

### 2023-02-07 chapters/fairness/fif/preliminaries.tex
all papers added
adding fairXplainer

@@
-%\clearpage
-\section{Background: Fairness and Global Sensitivity Analysis}\label{sec:preliminaries}\vspace*{-.5em}
-Before proceeding to the details of our contribution, we present the fundamentals of group fairness metrics as quantifiers of bias and global sensitivity analysis as a classical method of feature attribution.\vspace*{-.5em}
+\section{Background: Fairness and Global Sensitivity Analysis}\label{sec:preliminaries}%\vspace*{-.5em}
+Before proceeding to the details of our contribution, we present the fundamentals of group fairness metrics as quantifiers of bias and global sensitivity analysis as a classical method of feature attribution.%\vspace*{-.5em}
 \subsection{Fairness in Machine Learning: Fairness Metrics}
+%We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
+We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $.
+Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.
+$y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.
+We denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$ with $ (\nonsensitive, \sensitive, Y) $.
+We represent a binary classifier trained on dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
+$\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
+Given the setup, we discuss two fairness metrics quantifying the bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,verma2018fairness}.
+
+%\blue{We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
+We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of $n$ samples $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+The feature vector $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $.
+$ \mathbf{x}^{(i)} $ additionally contains sensitive information represented as a vector $ \mathbf{a}^{(i)} \subset \mathbf{x}^{(i)} $.
+Specifically, the sensitive feature vector $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.
+$y^{(i)} \in \{0,1\}$ is the binary class corresponding to $ \mathbf{x}^{(i)} $.
+We denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$ with $ (\nonsensitive, \sensitive, Y) $.
+We represent a binary classifier trained on dataset $\mathbf{D}$ as $\alg: \nonsensitive \rightarrow \widehat{Y} $.
+$\widehat{Y} \in \{0,1\}$ is the class predicted for $ \nonsensitive $.
+Given the setup, we discuss two fairness metrics quantifying the bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,verma2018fairness}.
+}
+
 We consider\footnote{{We represent sets/vectors by bold letters, and the corresponding distributions by calligraphic letters.
-We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$
- generated from an underlying distribution $\mathcal{D}$.
-Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $\{\mathbf{x}^{(i)}_1, \dots, \mathbf{x}^{(i)}_{\numnonsensitive}\} $.
-Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $\{\mathbf{a}^{(i)}_1, \dots, \mathbf{a}^{(i)}_{\numsensitive}\} $.
-$y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.
-%Each non-sensitive feature $ X_i$ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each sensitive feature $ A_j \in \{0, \dots, N_j\}  $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}.
- We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.  
-%\blue{For sensitive features, a vector $ \mathbf{a} = [a_1, .., a_{\numsensitive}] $ of categorical values is called a \textit{compound sensitive group}.
-For example, consider $ \sensitive = $ \{race, sex\}, where race $ \in $ \{Asian, Color, White\} and sex $ \in $ \{female, male\}.
-Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group.}\unsure{blue part is not needed}
- We represent a binary classifier trained on the dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
-Here, $\widehat{Y} \in \{0,1\}$ is the class predicted for $ (\nonsensitive, \sensitive) $.
-Given this setup, we discuss different fairness metrics to compute bias in the prediction of a classifier~\cite{feldman2015certifying,hardt2016equality,nabi2018fair}.
-%\red{We use $ f(\alg, \mathbf{D}) $ to denote a fairness metric computed for a classifier $ \alg $ on the dataset $ \mathbf{D} $}.
-%\blue{We certify a classifier $ \epsilon $-fair, for a constant $ \epsilon \in [0,1] $, by computing $ \mathds{1}(f(\alg, \mathbf{D}) \le \epsilon) $.
-Here, $ \mathds{1} $ is an indicator function evaluating to $ 1 $ whenever the input is true, and $ 0 $ otherwise.}\info{Not necessary!}
-\begin{enumerate}[leftmargin=*]
-	%		\itemsep0em 
-	\item \textit{Statistical Parity} ($ \mathsf{SP} $)~\cite{feldman2015certifying}: Statistical parity belongs to \textit{independence} measuring group fairness metrics, where the prediction $ \widehat{Y} $ is statistically independent of sensitive features $ \sensitive $.
-The statistical parity of  a classifier is measured as $ f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] $, which is the difference between the maximum and minimum conditional probability of positive prediction the classifier for different sensitive groups.
+We express random variables in uppercase, and an assignment of a random variable in lowercase.}} a dataset $ \mathbf{D} $ as a collection of $n$ data points  $\{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+The feature vector $\mathbf{z}^{(i)} \triangleq (\mathbf{x}^{(i)}, \mathbf{a}^{(i)}) $ is a concatenation of non-sensitive features $ \mathbf{x}^{(i)} $ and sensitive features $ \mathbf{a}^{(i)} $.
+Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] \in \mathbb{R}^{\numnonsensitive} $.
+Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] \in \mathbb{N}^{\numsensitive} $.
+Thus, the cardinality of feature vector $ \mathbf{z}^{(i)} $ is denoted by $ m $, formally $ |\mathbf{z}^{(i)}| \triangleq \numfeatures = \numnonsensitive + \numsensitive $.
+The binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$ is $y^{(i)} \in \{0,1\}$.
+We refer to $y^{(i)}$ as the true class.
+We denote the random variables corresponding to $ (\mathbf{z}, \mathbf{x}, \mathbf{a},  y) $ with $ (\feature, \nonsensitive, \sensitive, Y) $.
+Henceforth, we represent a binary classifier trained on dataset $\mathbf{D}$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} \in \{0,1\} $, where $\widehat{Y} $ is the predicted class.
+Given the setup, we discuss a fairness metric $ f: (\alg, \mathbf{D}) \rightarrow \mathbb{R}^{\ge 0} $ quantifying the bias in the prediction of a classifier on a dataset~\cite{feldman2015certifying,hardt2016equality,verma2018fairness}.
 
-	\item \textit{Equalized Odds} ($ \mathsf{EO} $)~\cite{hardt2016equality}: \textit{Separation} measuring group fairness metrics such as equalized odds constrain that $ \widehat{Y} $ is independent of $ \sensitive $ given the ground class $ Y $.
-Formally, for $ Y \in \{0,1\} $, equalized odds is $ f_{\mathsf{EO}}(\alg, \mathbf{D})  \triangleq \max(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1])$.
+	\textit{Statistical Parity} ($ \mathsf{SP} $): Statistical parity belongs to the \textit{independence} measuring group fairness metrics that measures whether the prediction $ \widehat{Y} $ is statistically independent of the sensitive features $ \sensitive $.
+The statistical parity of  a classifier is measured as $ f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}] $, which is the difference between the maximum and minimum conditional probability of positive prediction the classifier for different sensitive groups.
+Lower value of $ f(\alg, \mathbf{D}) $ indicates higher fairness demonstrated by the classifier $\mathcal{M}$ on $ \mathbf{D} $.
+\textit{Henceforth, we deploy statistical parity as the measure of bias of a classifier w.r.t\ a given dataset.}\footnote{For brevity, we define other  fairness metrics equalized odds and predictive parity, and corresponding FIF computation technique in Appendix~\ref{app:pp}.}
 
-	\item \textit{Predictive Parity} ($ \mathsf{PP} $)~\cite{verma2018fairness}: \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $.
-Formally, $ f_{\mathsf{PP}}(\alg, \mathbf{D})  \triangleq \max(\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1])$.
+%	\item \textit{Predictive Parity} ($ \mathsf{PP} $)~\citep{verma2018fairness}: \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $.
+Formally, $ f_{\mathsf{PP}}(\alg, \mathbf{D})  \triangleq \max(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = 1])$.
 % \textit{Conditional Use Accuracy Equality (CUAE):}
 %	\item \blue{\textit{Path-specific Causal Fairness} (PCF): 
-%		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 |\mathbf{A}=  \mathbf{a}] $ be the sensitive group with maximum conditional probability of positive prediction of the classifier.
-We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $.
+%		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\widehat{Y} =1 \mid\mathbf{A}=  \mathbf{a}] $ be the sensitive group with maximum conditional probability of positive prediction of the classifier.
+We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{\mid\mathbf{A} = \mathbf{a}_{\max}}} $.
 This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
-To this end, the path-specific causal fairness of a classifier is $ f(\alg, \mathbf{D}) = \max_{\mathbf{a}}\Pr[\widehat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 | \sensitive = \mathbf{a}, \mediator ] $.}\change{How to sample $ \mathbf{Z} $?}
-\end{enumerate}
-All of the aforementioned group fairness metrics depend on the difference between different conditional probabilities of positive prediction of a classifier.
-For all of these metrics, lower value of $ f(\alg, \mathbf{D}) $ indicates higher fairness demonstrated by the classifier $\mathcal{M}$.
-\textit{We deploy these fairness metrics as the measures of bias of a classifier w.r.t\ a given dataset.}%\info{Not necessary!}
+To this end, the path-specific causal fairness of a classifier is $ f(\alg, \mathbf{D}) = \max_{\mathbf{a}}\Pr[\widehat{Y} = 1 \mid \sensitive =  \mathbf{a}, \mediator] - \min_{\mathbf{a}} \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}, \mediator ] $.}\change{How to sample $ \feature $?}
 
-\subsection{Global Sensitivity Analysis: Variance Decomposition}
-Global sensitivity analysis is a field that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input while considering the whole input domain~\cite{saltelli2008global}.
-Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\cite{eu}, USA~\cite{usepa}, and research communities~\cite{saltelli2020five}.
+%\red{We observe that to yield a good estimator of statistical parity, the central function to estimate is $ \mathds{1}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $.
+Hereafter, we refer to it as  CPP (Conditional Positive Prediction) function.}
+
+%\red{The group fairness metrics depend on the difference between different conditional probabilities of classifier's prediction.
+For all of these metrics, lower value of $ f(\alg, \mathbf{D}) $ indicates higher fairness demonstrated by the classifier $\mathcal{M}$ on $ \mathbf{D} $.
+\textit{We deploy these fairness metrics as the measures of bias of a classifier w.r.t\ a given dataset.}}\footnote{For brevity, we define other  fairness metrics equalized odds and predictive parity, and corresponding FIF computation technique in Appendix~\ref{app:pp}.}%\info{Not necessary!}
+
+\subsection{Global Sensitivity Analysis (GSA): Variance Decomposition}
+Global sensitivity analysis is an active field of research that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input variables~\citep{saltelli2008global}.
+Sensitivity analysis is an essential component for quality assurance and impact assessment of models in EU~\citep{eu}, USA~\citep{usepa}, and research communities~\citep{saltelli2020five}.
 %\paragraph{Sensitivity Analysis:}
-\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\cite{sobol1990sensitivity,sobol2001global}.
-To illustrate, let us consider a real-valued function $  \function(\mathbf{Z}) $, where $ \mathbf{Z} $ is a vector of $ \numnonsensitive $ input variables $ \{Z_1, \dots, Z_k\} $.
+\emph{Variance-based sensitivity analysis} is a form of global sensitivity analysis, where variance is considered as the measure of uncertainty~\citep{sobol1990sensitivity,sobol2001global}.
+To illustrate, let us consider a real-valued function $  \function(\feature) $, where $ \feature $ is a vector of $ \numfeatures $ input variables $ \{Z_1, \dots, Z_\numfeatures\} $.
 % and $ Y $ is a univariate output. 
@@
 Then,
-Now, we decompose $ \function(\mathbf{Z}) $ among the subsets of inputs, such that:
+Now, we decompose $ \function(\feature) $ among the subsets of inputs, such that:
 \begin{align}
-	\function(\mathbf{Z}) &= \function_0 + \sum_{i=1}^{\numnonsensitive} \function_{\{i\}}(Z_i) +  \sum_{i < j}^{\numnonsensitive} \function_{\{i,j\}}(Z_i, Z_j)  + \cdots  + \function_{\{1, 2, \dots, \numnonsensitive\}} (Z_1, Z_2, \dots, Z_{\numnonsensitive})\notag%\label{eq:functional_decomposition}
-	\\    
-	&= \function_0 +  \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
+	\function(\feature) &= \function_0 + \sum_{i=1}^{\numfeatures} \function_{\{i\}}(Z_i) +  \sum_{i < j}^{\numfeatures} \function_{\{i,j\}}(Z_i, Z_j)  + \cdots\notag + \function_{\{1, 2, \dots, \numfeatures\}} (Z_1, Z_2, \dots, Z_{\numfeatures})\\
+	&= \function_0 +  \sum_{\mathbf{S} \subseteq [\numfeatures]} \function_{\mathbf{S}}(\feature_{\mathbf{S}})\label{eq:functional_decomposition_set_notation}
 \end{align}
 %	$S_> = \{{S_0}, \ldots, {S_{|s|}}\}$ i.e. a monotonically increasing ordering of $|S|$ features for a given starting feature $Z_{S_0}$ such that $S_0 <S_1 <\ldots <S_{|S|}$.
+The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\citep{sobol1990sensitivity}.
 In this decomposition, $ \function_0 $ is a constant, $ \function_{\{i\}} $ is a function of $ Z_i $, $ \function_{\{i,j\}} $ is a function of $ Z_i $ and $ Z_j $, and so on.
-Here, $ [\numnonsensitive] \triangleq \{1,2,\dots, \numnonsensitive\} $ and $ \mathbf{S}$ is an ordered subset of $[\numnonsensitive] \setminus \emptyset $.
-We denote $ \mathbf{Z}_{\mathbf{S}}  \triangleq \{Z_i | i \in \mathbf{S}\}$ as the input of $ \function_{\mathbf{S}} $, where $ \mathbf{Z}_{\mathbf{S}}$ is a set of variables with indices belonging to $ \mathbf{S} $.
-The standard condition of this decomposition is the orthogonality of each term in the right-hand side of Eq.~\eqref{eq:functional_decomposition_set_notation}~\cite{sobol1990sensitivity}.
-$ \function_{\mathbf{S}}(\mathbf{Z}_{\mathbf{S}}) $ is the effect of varying all the features in $\mathbf{Z}_{\mathbf{S}}$ simultaneously.
-%\unsure{Is this sentence necessary?}
-For $|\mathbf{S}|=1$, it quantifies an individual variable's effect. For $|\mathbf{S}|>1$, it quantifies the higher-order interactive effect of variables.
+Adhering to the set-based notations, we denote by $ g_{\mathbf{S}} $ a function of a non-empty subset of variables $ \feature_{\mathbf{S}}  \triangleq \{Z_i \mid i \in \mathbf{S}\} \subseteq \feature $, where $ \mathbf{S} = \{S_i \mid 1 \le |S_i| \le \numfeatures\} \subseteq [\numfeatures] $ is a non-empty subset of indices with $ [\numfeatures] \triangleq \{1,2,\dots, \numfeatures\}  $.
+Here, $|\mathbf{S}|=1$ quantifies an individual variable's effect while $|\mathbf{S}|>1$ quantifies the higher-order intersectional effect of variables.
 \iffalse
@@
 	&\int f(\mathbf{X})\prod_l dX_l =  \mathsf{E}[Y] = \function_0\\
-	&\int f(\mathbf{X})\prod_{l \ne i} dX_l =  \mathsf{E}[Y|X_i] = \function_{\{i\}}(X_i) + \function_0\\
-	&\int f(\mathbf{X})\prod_{l \ne i, j} dX_l = \mathsf{E}[Y|X_i, X_j] = \function_{\{i,j\}}(X_i, X_j)  +  \function_{\{i\}}(X_i) +  \function_{\{j\}}(X_j) + \function_0 \\
+	&\int f(\mathbf{X})\prod_{l \ne i} dX_l =  \mathsf{E}[Y\midX_i] = \function_{\{i\}}(X_i) + \function_0\\
+	&\int f(\mathbf{X})\prod_{l \ne i, j} dX_l = \mathsf{E}[Y\midX_i, X_j] = \function_{\{i,j\}}(X_i, X_j)  +  \function_{\{i\}}(X_i) +  \function_{\{j\}}(X_j) + \function_0 \\
 	&\vdots
@@
 Then each term $ \function_{\mathbf{S}} $ in Eq.~\eqref{eq:functional_decomposition_set_notation} are also square integrable.
-Squaring Eq.~\eqref{eq:functional_decomposition_set_notation}, integrating over $  [0,1]^\numnonsensitive$, and applying orthogonality constraint in Eq.~\ref{eq:orthogonality_in_decomposition}, we get
+Squaring Eq.~\eqref{eq:functional_decomposition_set_notation}, integrating over $  [0,1]^\numfeatures$, and applying orthogonality constraint in Eq.~\ref{eq:orthogonality_in_decomposition}, we get
 \begin{align*}
-	\int f^2(\mathbf{X})\prod_l dX_l - \function_0^2 = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset}  \function_{\mathbf{S}}^2(\mathbf{X}_{\mathbf{S}}) \prod_{S_i \in \mathbf{S}} d Z_{S_i}
+	\int f^2(\mathbf{X})\prod_l dX_l - \function_0^2 = \sum_{\mathbf{S} \subseteq [\numfeatures], \mathbf{S}\ne \emptyset}  \function_{\mathbf{S}}^2(\mathbf{X}_{\mathbf{S}}) \prod_{S_i \in \mathbf{S}} d Z_{S_i}
 \end{align*}
@@
 \fi
-Now, if we assume $ \function $ to be square integrable, we obtain the decomposition of the variance of the output~\cite{sobol1990sensitivity}.
+Considering $ \function $ as square integrable, we obtain the decomposition of the variance of $ \function(\feature) $ expressed as the sum of variances of $\function_{\mathbf{S}}$'s~\citep{sobol1990sensitivity}.
 \begin{align}\label{eq:variance_decomposition_set_notation}
-	\mathsf{Var}[g(\mathbf{Z})] &= \sum_{i=1}^{\numnonsensitive}V_{\{i\}} +  \sum_{i<j}^{\numnonsensitive} V_{\{i,j\}}  + \cdots  + V_{\{1, 2\dots, \numnonsensitive\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{S}}
+	\mathsf{Var}[g(\feature)] &= \sum_{i=1}^{\numfeatures}V_{\{i\}} +  \sum_{i<j}^{\numfeatures} V_{\{i,j\}}  + \dots +  V_{\{1, 2\dots, \numfeatures\}}= \sum_{\mathbf{S} \subseteq [\numnonsensitive]} V_{\mathbf{S}}
 \end{align}
-where $ V_{\{i\}} $ is the variance of $ \function_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \function_{\{i,j\}} $ and so on.
-Formally,  $V_{\mathbf{S}} \triangleq \mathsf{Var} _{\mathbf{Z}_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}/\mathbf{Z}_{\mathbf{S}}}[g(\mathbf{Z})\mid \mathbf{Z}_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}/\emptyset}{V} _{\mathbf{S}'}$.
-Here, $\mathbf{S}'$ denotes all the ordered proper subsets of $\mathbf{S}$.
-%\improvement{Check notations.}
-This variance decomposition shows how the variance of $\function(\mathbf{Z}) $ can be decomposed into terms attributable to each input, as well as the interactive effects among them.
-Together all terms sum to the total variance of the model output.
-\textit{We reduce the problem of computing FIFs of subsets of features into a variance decomposition problem.}
+where $ V_{\{i\}} $ is the variance of $ \function_{\{i\}} $, $ V_{\{i,j\}} $ is the variance of $ \function_{\{i,j\}} $ and so on. Formally,  
+\begin{equation}
+\label{eq:decomposed_variance}
+V_{\mathbf{S}} \triangleq \mathsf{Var} _{\feature_{\mathbf{S}}}\left[\mathbb{E}_{{\textbf {Z}}\setminus\feature_{\mathbf{S}}}[g(\feature)\mid \feature_{\mathbf{S}}]\right]- \sum_{\mathbf{S}' \subset \mathbf{S}}{V} _{\mathbf{S}'}.
+\end{equation} 
+Here, $\mathbf{S}'$ denotes all the non-empty and ordered proper subsets of $\mathbf{S}$.
+Thus, $ V_{\mathbf{S}} $ is the variance w.r.t.\ $ \feature_{\mathbf{S}} $ by subtracting the variances of all non-empty proper subsets of $ \feature_{\mathbf{S}} $.
+As a result, Eq.~\eqref{eq:variance_decomposition_set_notation} demonstrates how the variance of $\function(\feature) $ can be decomposed into terms attributable to each input feature, as well as the intersectional effects among them.
+Conversely, together all terms sum to the total variance of the model output.
+\textit{In the next section, we reduce the problem of computing FIFs of subsets of features into a variance decomposition problem.}
 
@@
 
-Formally, $V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y | X_i]]$ and $V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y | X_i, X_j]] - V_{\{i\}} - V_{\{j\}}$
+Formally, $V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y \mid X_i]]$ and $V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y \mid X_i, X_j]] - V_{\{i\}} - V_{\{j\}}$
 \begin{align*}
-	& V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y | X_i]]\\
-	& V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y | X_i, X_j]] - V_{\{i\}} - V_{\{j\}}%\\
+	& V_{\{i\}} = \mathsf{Var}_{\mathbf{X}_{\{i\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i\}}}[Y \mid X_i]]\\
+	& V_{\{i,j\}} = \mathsf{Var}_{\mathbf{X}_{\{i,j\}}}[\mathsf{E}_{\mathbf{X}_{\sim \{i,j\}}}[Y \mid X_i, X_j]] - V_{\{i\}} - V_{\{j\}}%\\
 	%& \vdots
@@
 \begin{align*}
-	\mathsf{Var}[Y] = \mathsf{E}_X[\mathsf{Var}_Y[Y|X]] + \mathsf{Var}_X[\mathsf{E}_Y[Y|X]]
+	\mathsf{Var}[Y] = \mathsf{E}_X[\mathsf{Var}_Y[Y\midX]] + \mathsf{Var}_X[\mathsf{E}_Y[Y\midX]]
 \end{align*}

### 2023-02-08 chapters/acknowledgements.tex
acknowledgement first draft added

@@
 
+With utmost humility, I would like to take this opportunity to express my sincere gratitude to everyone who has supported me throughout my PhD journey.
+
+I come from a humble background in Bangladesh and have always had a deep passion for math, science, and computer science.
+My PhD journey has been a process of self-reflection and growth, from being an inattentive undergraduate student to becoming a more dedicated PhD graduate.
+I have learned to accept challenges and setbacks, which has taught me valuable lessons and contributed to my eventual success.
+
+I am humbly grateful to my advisor, Kuldeep S.
+Meel, who has played a vital role in my PhD.
+He has taught me how to be a more independent researcher and his leadership, passion, and dedication have been a true source of inspiration.
+I feel privileged to have been a part of his research lab and to have benefited from its thriving research culture.
+
+I have also been blessed to work with exceptional collaborators, including Debabrota Basu, Dmitry Malioutov, Daniel Neider, Naheed Anjum Araafat, Jonathan Scarlet, and Lorenzo.
+They have generously shared their knowledge and experience with me and have helped me broaden my perspectives.
+I have come to understand that finding a meaningful research problem is a crucial step and I am grateful to have had the opportunity to collaborate and brainstorm with such insightful colleagues and friends.
+
+I would also like to extend my heartfelt thanks to my family, especially my mother, sister, brother-in-law, and uncle for their constant support and encouragement.
+While I regret that my father could not be here to see this day, I am comforted by the thought that his blessings will always be with me.
+Finally, I would like to thank my wife, who has been my supportive partner and companion throughout this journey.
+Without her unwavering support, it would have been a much more difficult road to completion.
+
+Thank you all for your support and encouragement. I am truly grateful.
+
+\vspace{2em}
+\noindent Best regards,\\
+Bishwamittra Ghosh
+
 \end{acknowledgments}

### 2023-02-08 doctoral_consortium/main.tex
updated templates for doctoral consortium

--- 2022-07-15 chapters/introduction.tex
@@
-\chapter{Introduction}
+%%%% ijcai23.tex
+
+\typeout{IJCAI--23 Instructions for Authors}
+
+% These are the instructions for authors for IJCAI-23.
+
+\documentclass{article}
+\pdfpagewidth=8.5in
+\pdfpageheight=11in
+
+% The file ijcai23.sty is a copy from ijcai22.sty
+% The file ijcai22.sty is NOT the same as previous years'
+\usepackage{ijcai23}
+
+% Use the postscript times font!
+\usepackage{times}
+\usepackage{soul}
+\usepackage{url}
+\usepackage[hidelinks]{hyperref}
+\usepackage[utf8]{inputenc}
+\usepackage[small]{caption}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{amsthm}
+\usepackage{booktabs}
+\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage[switch]{lineno}
+
+% Comment out this line in the camera-ready submission
+\linenumbers
+
+\urlstyle{same}
+
+% the following package is optional:
+%\usepackage{latexsym}
+
+% See https://www.overleaf.com/learn/latex/theorems_and_proofs
+% for a nice explanation of how to define new theorems, but keep
+% in mind that the amsthm package is already included in this
+% template and that you must *not* alter the styling.
+\newtheorem{example}{Example}
+\newtheorem{theorem}{Theorem}
+
+% Following comment is from ijcai97-submit.tex:
+% The preparation of these files was supported by Schlumberger Palo Alto
+% Research, AT\&T Bell Laboratories, and Morgan Kaufmann Publishers.
+% Shirley Jowell, of Morgan Kaufmann Publishers, and Peter F.
+% Patel-Schneider, of AT\&T Bell Laboratories collaborated on their
+% preparation.
+
+% These instructions can be modified and used in other conferences as long
+% as credit to the authors and supporting agencies is retained, this notice
+% is not changed, and further modification or reuse is not restricted.
+% Neither Shirley Jowell nor Peter F. Patel-Schneider can be listed as
+% contacts for providing assistance without their prior permission.
+
+% To use for other conferences, change references to files and the
+% conference appropriate and use other authors, contacts, publishers, and
+% organizations.
+% Also change the deadline and address for returning papers and the length and
+% page charge instructions.
+% Put where the files are available in the appropriate places.
+
+% PDF Info Is REQUIRED.
+% Please **do not** include Title and Author information
+\pdfinfo{
+/TemplateVersion (IJCAI.2023.0)
+}
+
+\title{Interpretability and Fairness in Machine Learning: A Formal Methods Approach}
+
+% Single author syntax
+\author{
+    Bishwamittra Ghosh
+    \affiliations
+    National University of Singapore
+    \emails
+   bghosh@u.nus.edu
+}
+
+% Multiple author syntax (remove the single-author syntax above and the \iffalse ... \fi here)
+\iffalse
+\author{
+First Author$^1$
+\and
+Second Author$^2$\and
+Third Author$^{2,3}$\And
+Fourth Author$^4$
+\affiliations
+$^1$First Affiliation\\
+$^2$Second Affiliation\\
+$^3$Third Affiliation\\
+$^4$Fourth Affiliation
+\emails
+\{first, second\}@example.com,
+third@other.example.com,
+fourth@example.com
+}
+\fi
+
+% Macros
+\newcommand{\imli}{\ensuremath{\mathsf{IMLI}}}
+\newcommand{\crr}{\ensuremath{\mathsf{CRR}}}
+\newcommand{\fairXplainer}{\ensuremath{\mathsf{FairXplainer}}}
+\newcommand{\justicia}{\ensuremath{\mathsf{Justicia}}}
+\newcommand{\fvgm}{\ensuremath{\mathsf{FVGM}}}
+
+\begin{document}
+
+\maketitle
 
@@
 
-\section{Tools}
-	We develop following open-source tools for interpretable and fair machine learning.
-	\begin{itemize}
-		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
-		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification and FIF computation
-	\end{itemize}
-\section{Outline}
-We organize the thesis in the following way.
-We discuss preliminaries in Chapter~\ref{chapter:preliminaries}.
-In the first part of the thesis, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable classifiers.
-We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
-In the second part of this thesis, we discuss  fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
-We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}.
-Finally, we conclude our thesis in Chapter~\ref{}.
+%% The file named.bst is a bibliography style file for BibTeX 0.99c
+\bibliographystyle{named}
+\bibliography{ref}
 
-\begin{comment}
-
-The tremendous success of machine learning in recent years has found its applications in different high-stake and  safety-critical domains, such as medical, law, education, and transportation.
-In such domains, the interpretability and fairness of machine learning classifiers are of paramount importance.
-In this work, we focus on improving the interpretability and fairness in machine learning.
-
-There are two-fold objectives in interpretable machine learning: (i) designing classifiers that are interpretable by design and (ii) leveraging already interpretable models to explain the prediction of black-box classifiers.
-In this work, we propose an efficient learning framework~\cite{ghosh19incremental} for a rule-based classifier\textemdash a popular interpretable model because of expressing the decision function in terms of rules corresponding to input features.
-Our framework allows us to learn classification rules expressible in propositional logic, for example, CNF (Conjunctive Normal Form) and DNF (Disjunctive formal form) formulas.
-The benefit of learning CNF classifiers is that this is the fundamental block for learning other popular interpretable representations: decision lists and decision sets.
-Learning rule-based classifiers is an intractable problem because of solving a combinatorial optimization problem.
-To efficiently learn these classifiers, we propose a MaxSAT-based formulation and scale it by combining incremental learning.
-The result is an efficient learning framework that can classify datasets even with 1 million samples while also balancing between high accuracy and small rule size.
-As an application of incremental learning, we propose to learn more relaxed logical rules~\cite{ghosh20classification} which demonstrate higher succinctness in their representation while achieving better accuracy.
-Furthermore, as an application of MaxSAT, we solve group testing problems based on compressed sensing~\cite{ciampiconi20maxsat}.
-Our MaxSAT-based formulation for group testing has impressive scalability than state-of-the-art approaches.
-
-We study the application of interpretable machine learning models in explaining the prediction of black-box models such as neural networks.
-To this end, we propose a PAC-based (Probably Approximately Correct) local explanations of feed-forward and recurrent neural networks using rule-based classifiers and their variants~\cite{neider2020probably,ghosh2020formal}.
-To explain feed-forward neural networks, we learn flexible classification rules using SyGuS~\cite{neider2020probably}.
-Such rules have the theoretical property that they are PAC approximate to the neural network within the locality of a query of interest.
-Similarly, to explain recurrent neural networks, we propose learning explanations in LTL formulas (Linear Temporal Logic)~\cite{ghosh2020formal}.
-
-Machine learning models may become biased in their decisions towards certain demographic populations.
-One important question in fairness in machine learning is to verify whether the model is unfair or not with respect to multiple definitions of fairness studied in the literature.
-To this end, we propose a scalable fairness verification framework based on formal methods, particularly stochastic satisfiability (SSAT)~\cite{ghosh2020justicia} and stochastic subset-sum problem (S3P)~\cite{ghosh2021algorithmic}.
-The input to the fairness verifier is a machine learning model and the distribution of input features and the output is an assessment of different fairness metrics that the model obtains given the distribution.
-Our fairness verifier has the following advantages over existing verifiers.
-We can efficiently verify the fairness of compound sensitive groups\textemdash such as, white-male and black-female\textemdash by combining multiple sensitive features, unlike existing works that only consider a single binary sensitive feature.
-We also consider feature-correlation represented as a Bayesian network whereas existing methods consider limited correlation among features.
-As a result, the proposed verifier has higher accuracy than existing verifiers.
-Finally, our verifier is more scalable than existing verifiers because of the efficient formulation based on SSAT and S3P.
-
-As future work, we aim to combine the interpretability and fairness of machine learning in a single framework.
-In particular, we explain the unfairness of models by combining explanation methods in machine learning.
-Our hypothesis is that such a combination would be effective in identifying the bias of a model and also mitigate the bias through imperative actions.
-
-\end{comment}
+\end{document}

### 2023-02-10 chapters/fairness/fif/conclusion.tex
fix bib

@@
 The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias.
-Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier.
+Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier.
 Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.

### 2023-02-10 chapters/fairness/fif/experiments.tex
fix bib

@@
 In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.
-Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19}.
-From Figure 3 in~\cite{rudin19}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively.
-While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features.
+Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19stop}.
+From Figure 3 in~\cite{rudin19stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively.
+While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features.
 \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.

### 2023-02-10 doctoral_consortium/main.tex
fix bib

@@
 Naturally, such techniques were limited in enhancing confidence of users for wide deployment.
-Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare, bastani2019probabilistic}.
+Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}.
 More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.
@@
 
-Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy,john2020verifying}.
 Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.

### 2023-02-10 chapters/abstract.tex
submitted version to the committee
with new formatting
fif edits fixed

@@
 \begin{abstract}
-	\addcontentsline{toc}{chapter}{Abstract}
-	\begin{center}
-		Interpretability and Fairness in Machine Learning: A Formal Methods Approach\\
-		\vspace{1em}
-		by\\
-		\vspace{1em}
-		Bishwamittra Ghosh\\
-		\vspace{1em}
-	\end{center}
 
@@
 
-	Fairness in machine learning centers on quantifying and mitigating the bias/unfairness of machine learning.
+	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning.
 In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of verifying whether the classifier achieves the desired level of fairness given the distribution of features.

### 2023-02-10 chapters/background.tex
fif edits fixed

@@
 \subsection{Dataset and Distribution}
-We consider a dataset $ \dataset $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
-Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] $.
-Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] $.
-$y^{(i)} \in \{0,1\}$ is the binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$.
+We consider a dataset $ \mathbf{D} $ as a collection of $n$ data points  $\{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+The feature vector $\mathbf{z}^{(i)} \triangleq (\mathbf{x}^{(i)}, \mathbf{a}^{(i)}) $ is a concatenation of non-sensitive features $ \mathbf{x}^{(i)} $ and sensitive features $ \mathbf{a}^{(i)} $.
+Each non-sensitive data point $\mathbf{x}^{(i)}$ consists of $\numnonsensitive$ features $[x^{(i)}_1, \dots, x^{(i)}_{\numnonsensitive}] \in \mathbb{R}^{\numnonsensitive} $.
+Each sensitive data point $\mathbf{a}^{(i)}$ consists of $\numsensitive$ categorical features $[a^{(i)}_1, \dots, a^{(i)}_{\numsensitive}] \in \mathbb{N}^{\numsensitive} $.
+Thus, the cardinality of feature vector $ \mathbf{z}^{(i)} $ is denoted by $ m $, formally $ |\mathbf{z}^{(i)}| \triangleq \numfeatures = \numnonsensitive + \numsensitive $.
+The binary class corresponding to $(\mathbf{x}^{(i)}, \mathbf{a}^{(i)})$ is $y^{(i)} \in \{0,1\}$.
+We refer to $y^{(i)}$ as the true class.
 
-We use $ (\nonsensitive, \sensitive, Y) $ to denote the random variables corresponding to $ (\mathbf{x}, \mathbf{a}, y)$.
+We denote the random variables corresponding to $ (\mathbf{z}, \mathbf{x}, \mathbf{a},  y) $ with $ (\feature, \nonsensitive, \sensitive, Y) $.
 Each non-sensitive feature $ X_i \in \nonsensitive $ is sampled from a continuous probability distribution {$ \mathcal{X}_i $}, and each categorical sensitive feature $ A_j \in \sensitive $ is sampled from a discrete probability distribution {$ \mathcal{A}_j $}.
@@
 
-\paragraph{Equalized Odds ($ \mathsf{EO} $).} \textit{Separation} measuring group fairness metrics such as equalized odds constrain that $ \widehat{Y} $ is independent of $ \sensitive $ given the ground class $ Y $~\cite{hardt2016equality}.
-Formally, for $ Y \in \{0,1\} $, equalized odds is
+\paragraph{Equalized Odds ($ \mathsf{EO} $).} \textit{Separation} measuring group fairness metrics such as equalized odds~\cite{hardt2016equality} constrain that the predicted class $ \widehat{Y} $ is independent of $ \sensitive $ given the true class $ Y $.
+Formally, for $ Y \in \{0,1\} $, equalized odds is defined as
+\[ f_{\mathsf{EO}}(\alg, \mathbf{D}) \triangleq  \max_{y} \Big(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y]\Big).
+\]
 
+Equalized odds intuitively implies the maximum statistical parity conditioned on $ Y $.
+
+\paragraph{Predictive Parity ($ \mathsf{PP} $).}
+\textit{Sufficiency} measuring group fairness metrics such as Predictive Parity ($ \mathsf{PP} $) constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $.
+Formally,
 \begin{align*}
-	f_{\mathsf{EO}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 0], \\ &\max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}, Y = 1]).
-\end{align*}
-
-\paragraph{Predictive Parity ($ \mathsf{PP} $).} \textit{Sufficiency} measuring group fairness metrics such as predictive parity constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $~\cite{verma2018fairness}.
-Formally,
-
-\begin{align*}
-	f_{\mathsf{PP}}(\alg, \mathcal{D})  \triangleq \max(&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 0], \\&\max_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1] - \min_{\mathbf{a}}\Pr[Y =1 | \mathbf{A} = \mathbf{a}, \widehat{Y} = 1]).
+	f_{\mathsf{PP}}(\alg, \mathbf{D})  
+	\triangleq \max_{y} \Big(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y]\Big).
 \end{align*}
@@
 
-\blue{The complexity $ \bncomplexity(\graph) $ of a Bayesian network $ (\graph, \theta) $ with discrete random variables $ \graphvertices $ is defined as the number of independent parameters used to define the probability distribution of $ \graphvertices $~\cite{murakonda2021quantifying}.
+The complexity $ \bncomplexity(\graph) $ of a Bayesian network $ (\graph, \theta) $ with discrete random variables $ \graphvertices $ is defined as the number of independent parameters used to define the probability distribution of $ \graphvertices $~\cite{murakonda2021quantifying}.
 Let $ \bnnumdiscrete(\graphvertex) $ denote the number of distinct values that a random variable $ \graphvertex $ can take.
 As such, for each conditional probability $ \Pr[\graphvertex_i = v_i | \parent(\graphvertex_i) = \mathbf{u}; \theta] $, we need $ \bnnumdiscrete(\parent(\graphvertex_i))(\bnnumdiscrete(\graphvertex_i) - 1) $ independent parameters.
-Thus, the total complexity of a Bayesian network is:}
+Thus, the total complexity of a Bayesian network is:
 
-\blue{
 \begin{align}
@@
 \end{align}
-}
 

### 2023-02-10 chapters/fairness/fif/appendix.tex
fif edits fixed

@@
 \appendix
-\onecolumn
-\section*{Appendix}
+\chapter{Fairness Influence Functions}
 \section{Societal Impact}
@@
 
-\subsection{Equalized Odds}
-
-\textit{Equalized Odds} ($ \mathsf{EO} $)~\cite{hardt2016equality}: \textit{Separation} measuring group fairness metrics such as equalized odds constrain that the predicted class $ \widehat{Y} $ is independent of $ \sensitive $ given the true class $ Y $.
-Formally, for $ Y \in \{0,1\} $, equalized odds is defined as
-\[ f_{\mathsf{EO}}(\alg, \mathbf{D}) \triangleq  \max_{y} \Big(\max_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 \mid \mathbf{A} = \mathbf{a}, Y = y]\Big).
-\]
-
-Equalized odds intuitively implies the maximum statistical parity conditioned on $ Y $.
-
-\paragraph{Extension of {\fairXplainer} to Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $.
+\subsection{FIFs of Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $.
 Since, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ is the equalized odds of the classifier, we finally report FIFs of features corresponding to the maximum sum of FIFs between $ Y \in \{0, 1\} $.
 
-\subsection{Predictive Parity}
-
-\textit{Sufficiency} measuring group fairness metrics such as Predictive Parity ($ \mathsf{PP} $) constrain that the ground class $ Y $ is independent of $ \sensitive $ given the prediction $ \widehat{Y} $.
-Formally,
-\begin{align*}
-	f_{\mathsf{PP}}(\alg, \mathbf{D})  
-	\triangleq \max_{y} \Big(\max_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y] - \min_{\mathbf{a}}\Pr[Y =1 \mid \mathbf{A} = \mathbf{a}, \widehat{Y} = y]\Big).
-\end{align*}
-
-\paragraph{Extension of {\fairXplainer} to Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
+\subsection{FIFs of Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
 For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label.
@@
 
-\begin{table}
-	\caption{Median estimation error (over 5-fold cross validation and all combinations of sensitive features) of equalized odds (columns $ 5 $ to $ 7 $) and predictive parity (columns $ 8 $ and $ 9 $) in terms of estimated FIFs by different methods.
+\begin{sidewaystable}
+	\caption[Approximation error of $ \mathsf{EO} $ and $ \mathsf{PP} $ using FIFs]{Median estimation error (over 5-fold cross validation and all combinations of sensitive features) of equalized odds (columns $ 5 $ to $ 7 $) and predictive parity (columns $ 8 $ and $ 9 $) in terms of estimated FIFs by different methods.
 Best results (lowest error) are in bold color.
@@
 	\end{tabular}
-\end{table}
+\end{sidewaystable}
 
@@
 \begin{figure}
-	\subfloat[Equalized Odds]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_eo}} 
-	\subfloat[Predictive Parity]{\includegraphics[scale=0.35]{figures/fairness/fif/runtime_diff_methods_train_suff}}
-	\caption{Execution time of different methods for estimating FIFs for equalized odds and predictive parity.
+	\centering
+	\subfloat[Equalized Odds]{\includegraphics[scale=0.45]{figures/fairness/fif/runtime_diff_methods_train_eo}} 
+	\subfloat[Predictive Parity]{\includegraphics[scale=0.45]{figures/fairness/fif/runtime_diff_methods_train_suff}}
+	\caption[Execution time of FIFs]{Execution time of different methods for estimating FIFs for equalized odds and predictive parity.
 {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
@@
 \begin{figure}
-	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_error.pdf}}
-	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_mean_rmse.pdf}}
-	\subfloat[Adult]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_adult_train_time.pdf}}
+	\centering
+	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_error.pdf}}
+	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_mean_rmse.pdf}}
+	\subfloat[Adult]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_adult_train_time.pdf}}
 
-	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_error.pdf}}
-	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_mean_rmse.pdf}}
-	\subfloat[COMPAS]{\includegraphics[scale=0.4]{figures/fairness/fif/sp_compas_train_time.pdf}}
+	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_error.pdf}}
+	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_mean_rmse.pdf}}
+	\subfloat[COMPAS]{\includegraphics[scale=0.38]{figures/fairness/fif/sp_compas_train_time.pdf}}
 
-	\caption{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
+	\caption[Ablation study on FIFs: $ \tau $]{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
 	\label{fairness_fairXplainer_fig:effect_spline_intervals_appendix}
@@
 \begin{figure}
+	\centering
 	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fif/ablation_study_max_order_sp_train_error.pdf}}
@@
 
-	\caption{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
+	\caption[Ablation study on FIFs: $ \lambda $]{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
 	\label{fairness_fairXplainer_fig:effect_maxorder_appendix}
@@
 		\centering
-		\subfloat[Individual FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_adult}}
+		\subfloat[Individual FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/adult_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_adult}}
 	\end{minipage}
@@
 		\centering
-		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.35]{figures/fairness/fif/adult_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_adult}}
+		\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/adult_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_adult}}
 	\end{minipage}
-	\vspace{-0.2em}
-	\caption{FIFs for Adult dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult}
+	\caption[FIFs for Adult dataset]{FIFs for Adult dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult}
 \end{figure}
@@
 	\centering
-	\subfloat[Individual FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_titanic}}\hfil
-	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.4]{figures/fairness/fif/titanic_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_titanic}}
-	\caption{FIFs for Titanic dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}
+	\subfloat[Individual FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/titanic_feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs_titanic}}\hfil
+	\subfloat[Individual and intersectional FIFs]{\includegraphics[scale=0.45]{figures/fairness/fif/titanic_feature_weight}\label{fairness_fairXplainer_fig:individual_and_intersectional_fifs_titanic}}
+	\caption[FIFs for Titanic dataset]{FIFs for Titanic dataset on explaining statistical parity.}\label{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}
 \end{figure}

### 2023-02-10 chapters/fairness/fif/experiments.tex
fif edits fixed

@@
 
-\begin{figure}[t!]
+\begin{figure}
 	\centering
-		\begin{minipage}{0.4\columnwidth}
-		\centering
-		\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}\vspace*{-.7em}
-		\caption{Execution time of different methods for estimating FIFs.
+	\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}
+	\caption[Execution time of FIFs in $ \mathsf{SP} $]{Execution time of different methods for estimating FIFs.
 {\fairXplainer} with $ \lambda = 1 $ is more efficient than SHAP, while {\fairXplainer} ($ \lambda = 2 $) requires more computational effort.
@@
 	\label{fairness_fairXplainer_fig:execution_time_cactus_plot}
-	\end{minipage}\hfil
-	\begin{minipage}{0.55\columnwidth}
+\end{figure}
+\begin{figure}
 	\centering\vspace*{-1em}
@@
 	\subfloat[Adult]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
-		\caption{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
+	\caption[FIFs under fairness intervention]{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
 Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}
 	\label{fairness_fairXplainer_fig:fairness_intervention}
-%		\centering
-%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_error.pdf}}
-%		\subfloat{\includegraphics[scale=0.35]{figures/fairness/fif/sp_compas_train_time.pdf}}
-%		
-%		\caption{Effect of spline intervals on the estimation error of statistical parity and  execution time of {\fairXplainer} on COMPAS.}
-%		\label{fairness_fairXplainer_fig:effect_spline_intervals}
-	\end{minipage}
 \end{figure}
@@
 
-	\caption{FIFs for COMPAS dataset on explaining statistical parity.
+	\caption[FIFs for COMPAS dataset]{FIFs for COMPAS dataset on explaining statistical parity.
 Both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs} depict sources of bias in detail than individual FIFs alone in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}.
@@
 
-	\caption{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing~\cite{kamiran2012data} algorithms on FIFs.}\label{fairness_fairXplainer_fig:affirmative_punitive_actions}
+	\caption[FIFs under fairness affirmative/punitive actions]{Effects of a fairness attack~\cite{solans2020poisoning} and a fairness enhancing~\cite{kamiran2012data} algorithms on FIFs.}\label{fairness_fairXplainer_fig:affirmative_punitive_actions}
 \end{figure}

### 2023-02-10 chapters/fairness/fif/figure_intro.tex
fif edits fixed

@@
 %\vspace*{-.5em}
-	\caption{FIFs of input features to investigate the bias (statistical parity) of a decision tree predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
+	\caption[Demonstration of FIF  in health insurance]{FIFs of input features to investigate the bias (statistical parity) of a decision tree predicting the eligibility for health insurance using age-dependent features `fitness' and `income'.
 An affirmative action reduces bias as corresponding FIFs reflect it.}

### 2023-02-10 chapters/background.tex
formatting

@@
 
-\subsection{Decision Lists.} A decision list is a rule-based classifier consisting of ``if-then-else'' rules.
+\subsection{Decision Lists} A decision list is a rule-based classifier consisting of ``if-then-else'' rules.
 Formally, a decision list~\cite{rivest1987learning} is an ordered list $ \Rule_L $ pairs $ (C_1, V_1), \dots, (C_k, V_k) $ where the  rule $ C_i $ is a conjunction of literals (alternately, a single clause in a DNF formula) and $ V_i \in \{0,1\} $ is  a Boolean class label\footnote{In a more practical setting, $ V_i \in \{0, \dots, N\} $  can be multi-class for $ N \ge 1 $.}.
@@
 
-\subsection{Decision sets.} A decision set is a set of \textit{independent} ``if-then'' rules.
+\subsection{Decision sets} A decision set is a set of \textit{independent} ``if-then'' rules.
 Formally,  a decision set $ \Rule_S $ is a set of pairs $ \{(C_1, V_1), \dots, (C_{k-1}, V_{k-1})\}  $ and a  default pair $ (C_k, V_k) $, where $ C_i $ is\textemdash similar to a decision list\textemdash a conjunction of literals and $ V_i \in \{0,1\} $ is a Boolean class label.

### 2023-02-10 chapters/fairness/fif/experiments.tex
formatting

@@
 
-\setlength{\textfloatsep}{4pt}% Remove \textfloatsep
 \begin{enumerate}
@@
 	\centering
-	\subfloat{\includegraphics[scale=0.38]{figures/fairness/fif/runtime_diff_methods_train_sp}}
+	\subfloat{\includegraphics[scale=0.5]{figures/fairness/fif/runtime_diff_methods_train_sp}}
 	\caption[Execution time of FIFs in $ \mathsf{SP} $]{Execution time of different methods for estimating FIFs.
@@
 	\label{fairness_fairXplainer_fig:execution_time_cactus_plot}
-\end{figure}
-\begin{figure}
-	\centering\vspace*{-1em}
-	\subfloat[COMPAS]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_compas}}
-	\subfloat[Adult]{\includegraphics[scale=0.33]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
-	\caption[FIFs under fairness intervention]{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
-Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}
-	\label{fairness_fairXplainer_fig:fairness_intervention}
 \end{figure}
@@
 \textit{Therefore, {\fairXplainer} demonstrates computational efficiency in explaining group fairness metrics of real-world datasets and classifiers.}
+
+\begin{figure}
+	\centering
+	\subfloat[COMPAS]{\includegraphics[scale=0.5]{figures/fairness/fif/intervention_compas}}
+	\subfloat[Adult]{\includegraphics[scale=0.5]{figures/fairness/fif/intervention_adult}}\vspace*{-1em}
+	\caption[FIFs under fairness intervention]{Results on fairness intervention on logistic regression classifiers for COMPAS and Adult datasets.
+Pearson's correlation coefficient between bias-difference due to the intervention and FIFs is higher for top ranked influential features by {\fairXplainer} compared to SHAP.}
+	\label{fairness_fairXplainer_fig:fairness_intervention}
+\end{figure}
 
@@
 	\centering
-	\subfloat[{\fairXplainer} ($ \lambda = 1 $)]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs}}\hfil
-	\subfloat[{\fairXplainer} ($ \lambda = 2 $)]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight}\label{fairness_fairXplainer_fig:intersectional_fifs}}\hfil
-	\subfloat[SHAP]{\includegraphics[scale=0.35]{figures/fairness/fif/feature_weight_shap}\label{fairness_fairXplainer_fig:fif_shap}}
+	\subfloat[{\fairXplainer} ($ \lambda = 1 $)]{\includegraphics[scale=0.45]{figures/fairness/fif/feature_weight_unsorted}\label{fairness_fairXplainer_fig:individual_fifs}}\hfil
+	\subfloat[{\fairXplainer} ($ \lambda = 2 $)]{\includegraphics[scale=0.45]{figures/fairness/fif/feature_weight}\label{fairness_fairXplainer_fig:intersectional_fifs}}\hfil
+	\subfloat[SHAP]{\includegraphics[scale=0.45]{figures/fairness/fif/feature_weight_shap}\label{fairness_fairXplainer_fig:fif_shap}}
 
@@
 \begin{figure}
-	\subfloat[Fairness attack (punitive actions)]{\includegraphics[scale=0.35]{figures/fairness/fif/fairness_attack}\label{fairness_fairXplainer_fig:fairness_attack}}\hfil
-	\subfloat[Fairness enchancing (affirmative actions)]{\includegraphics[scale=0.35]{figures/fairness/fif/fairness_repair}\label{fairness_fairXplainer_fig:fairness_repair}}
+	\subfloat[Fairness attack (punitive actions)]{\includegraphics[scale=0.4]{figures/fairness/fif/fairness_attack}\label{fairness_fairXplainer_fig:fairness_attack}}\hfil
+	\subfloat[Fairness enchancing (affirmative actions)]{\includegraphics[scale=0.4]{figures/fairness/fif/fairness_repair}\label{fairness_fairXplainer_fig:fairness_repair}}
 

### 2023-02-10 chapters/interpretability/imli/experiment.tex
preparing response letter
formatting

@@
 For CORELS algorithm, we vary three hyper-parameters:  the maximum cardinality of rules in $ \{2,, 5\} $, the minimum support of rules in $ \{0.01, 0.17, 0.33, 0.5  \} $, and the regularization parameter in $ \{0.005, 0.01 , 0.015, 0.02 , 0.025, 0.03\}  $.
-For BRS algorithm, we vary three hyper-parameters: the number of initial rules in $ \{500, 1000, 1500, 2000, 2500, 3000\} $, the maximum length of rules in $ \{1,2,3,4\} $, and the minimum support of rules in $ \{1, 4,7, 10\} $.
+For BRS algorithm, we vary three hyper-parameters: the maximum length of rules in $ \{1,2,3,4\} $, the number of initial rules in $ \{500, 1000, 1500, 2000, 2500, 3000\} $, and the minimum support of rules in $ \{1, 4,7, 10\} $.
 For RF and RIPPER classifiers, we vary the cut-off on the number of samples in the leaf node using a linear grid between $ 3 $ to $ 500 $ and $ 1 $ to $ 300 $, respectively.
@@
 \subsection{Experimental Results}
-
+\label{interpretability_imli_sec:experiments_scalability}
 \paragraph{Comparing {\imli} with interpretable classifiers.} We compare {\imli} with existing interpretable classifiers in three aspects: test accuracy, rule-size, and scalability.
@@
 While {\imli} mostly takes the maximum allowable time ($ 1000 $ seconds) in solving all instances in this dataset, CORELS can solve lower than $ 400 $ instances.
-\blue{The improved performance of {\imli} is due to incremental learning based on the novel integration of iterative learning and mini-batch learning.
+The improved performance of {\imli} is due to incremental learning based on the novel integration of iterative learning and mini-batch learning.
 Contrary to our incremental learning, earlier approaches are based on heuristic rule-pruning (e.g., RIPPER) and rule-mining followed by Bayesian optimization (e.g., BRL, BRS) and branch and bounds algorithms (e.g., CORELS).
-Thus, owing to incremental learning, {\imli} establishes itself as the most scalable classifier compared to other state-of-the-art interpretable classifiers.}
+Thus, owing to incremental learning, {\imli} establishes itself as the most scalable classifier compared to other state-of-the-art interpretable classifiers.
 
@@
 
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_wdbc_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_wdbc_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_wdbc_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_wdbc_cactus_predicate_count}}
 
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_credit_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_credit_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_credit_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_credit_cactus_predicate_count}}
 
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_train_val_fit_time}}	
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_test_accuracy}}
-	\subfloat{\includegraphics[scale=0.3]{figures/interpretability/imli/imli_exp_adult_cactus_predicate_count}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_adult_cactus_train_val_fit_time}}	
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_adult_cactus_test_accuracy}}
+	\subfloat{\includegraphics[scale=0.28]{figures/interpretability/imli/imli_exp_adult_cactus_predicate_count}}
 

### 2023-02-10 chapters/interpretability/imli/incremental_learning.tex
formatting

--- 2022-07-15 chapters/interpretability/imli/incremental_learning.tex
@@
 Firstly, instead of solving a large MaxSAT query for the whole training data, this approach solves a sequence of smaller MaxSAT queries derived from mini-batches of smaller size.
-Secondly, this approach extends to life-long learning~\cite{chen2018lifelong}, where the classifier can be updated incrementally with new samples while also generalizing to previously observed samples.
+Secondly, this approach extends to online learning, where the classifier can be updated incrementally with new samples while also generalizing to previously observed samples.
 In our context of rule-based classifiers, we consider the following heuristic in mini-batch learning.
@@
 For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch.
-Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution.
+Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same across batches.
+One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning.
+For a feature variable $ B^i_j $, we consider its majority assignment in last $ p $ classification rules and encode as a soft clause to retain the majority assignment in the current batch.
+Moreover, we can reweigh the soft clause by prioritizing assignments of $ B^i_j $ in recent batches.}.
 Since our study focuses on rule-based classifiers, we consider the Hamming distance between two classifiers as a notion of their syntactic dissimilarity.

### 2023-02-10 chapters/interpretability/imli/introduction.tex
formatting

@@
 	\normalfont
-	\blue{We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
+	We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
 We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with two \emph{clauses} and four Boolean \emph{literals}.
 In the formula, clauses are connected by Boolean `AND', where each clause contains \emph{literals} connected by Boolean `OR'.
-Informally, a tumor cell is malignant if at least one literal in each clause becomes true.}
+Informally, a tumor cell is malignant if at least one literal in each clause becomes true.
 \end{example}

### 2023-02-10 chapters/introduction.tex
preparing for a submission
formatting

@@
 \chapter{Introduction}
-
 The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, education~\cite{luckin2018machine}, and transportation~\cite{peled2019model,zantalis2019review}.
@@
 We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}.
-Finally, we conclude our thesis in Chapter~\ref{}.
+Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.
 

### 2023-02-10 chapters/abstract.tex
updated abstract

@@
 Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{identifying the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{explaining the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
 To alleviate these limitations, we apply formal methods in interpretable and fair machine learning and provide scalable and accurate solutions to the underlying problems.
@@
 To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
-Building on incremental learning and MILP (mixed integer linear programming) solving, we propose another interpretable learning framework, called {\crr}, for learning a more expressible classification rule by relaxing logical formulas, obtaining higher accuracy and less rule-size than existing interpretable classifiers.
+\blue{Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic.
+To learn more expressible yet interpretable classification rules, we propose a relaxation of rule-based classifiers based on logical formulas and  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming).
+{\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.}
 
-	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning.
-In the presence of multiple fairness definitions and fairness algorithms, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of verifying whether the classifier achieves the desired level of fairness given the distribution of features.
-We also extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness.
-Finally, we propose a framework for fairness influence functions (FIF) with an aim of quantifying the influence of features as their contribution on the bias of a classifier.
-FIF analysis interprets fairness by revealing potential features or subset of features attributing highly to the bias.
-Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for computing FIFs of features, resulting in an accurate estimate of the bias of the classifier decomposed to input features.
+	\blue{Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers.
+In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of formally verifying the bias of a classifier given the probability distribution of features.
+Additionally, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness.
+Fairness metrics globally quantify bias, but do not detect or explain its sources.
+To explain group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier.
+FIF  explains fairness by revealing potential individual or intersectional features attributing highly to the bias.
+Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.}
 

### 2023-02-10 chapters/introduction.tex
introduction updated

@@
 
-Towards responsible machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers.
+Towards responsible and trustworthy machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers.
 \emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features.
@@
 In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
-In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
-Consequently, the research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
-In this thesis, we study fairness verification problems to assess the fairness of different classifiers under the lens of multiple fairness definitions and algorithms.
-In addition, we study the sources of the unfairness of classifiers by identifying  features attributing highly to bias.
-To summarize our thesis, in interpretable and fair machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which prior approaches fail to achieve.
+In another research theme of \emph{fairness in machine learning}, unregulated classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless classifiers are trained with a fairness objective.
+Consequently, research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
+In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features.
+Finally, we explain the sources of bias based on fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
+To summarize our thesis on the interpretability and fairness in machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.
 
@@
 
-In this thesis, we study two interpretable rule-based classifiers, characterized by their expressiveness.
+In the thesis, we study two interpretable rule-based classifiers, characterized by their expressiveness.
 At first, we study classifiers represented as formulas in propositional logic.
@@
 Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP (mixed integer linear programming) to efficiently learn rule-based classifiers in a mini-batch learning setting.
-
-The contributions of this thesis on interpretable machine learning are summarized in the following.
+Here we summarize the contributions of the thesis on interpretable rule-based machine learning.
 
 \subsection*{Scalability via Incremental Learning}
-We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for synthesizing interpretable classification rules expressible in proposition logic.
-$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns an optimal rule by solving an appropriately designed MaxSAT query.
-Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to practical classification datasets containing thousands to millions of samples.
+We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for learning interpretable classification rules expressed in CNF.
+$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query.
+Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to real-world classification datasets of thousands to millions of samples.
 Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning.
@@
 In our experiments, $ \mathsf{IMLI} $ achieves the best balance among prediction accuracy, interpretability, and scalability.
-For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability w.r.t.\ existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
-As an application, we deploy $ \mathsf{IMLI} $ in learning popular interpretable classifiers such as decision lists and decision sets.
+For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability with respect to existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
+As an application, we deploy $ \mathsf{IMLI} $ in learning other interpretable representations: DNF classifiers, decision lists, and decision sets.
 
@@
 We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}.
-Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule.
-Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and Boolean cardinality constraints in logic.
+Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule.
+Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and the higher expressiveness of Boolean cardinality constraints.
 Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets.
@@
 To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
-Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
+Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.\todo{Up to this.}
 

### 2023-02-12 chapters/abstract.tex
introduction rewritten

@@
 To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
-\blue{Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic.
+Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic.
 To learn more expressible yet interpretable classification rules, we propose a relaxation of rule-based classifiers based on logical formulas and  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming).
-{\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.}
+{\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.
 
-	\blue{Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers.
+	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers.
 In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of formally verifying the bias of a classifier given the probability distribution of features.
@@
 FIF  explains fairness by revealing potential individual or intersectional features attributing highly to the bias.
-Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.}
+Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.
 

### 2023-02-12 chapters/fairness/fif/introduction.tex
introduction rewritten

@@
 In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features.
-Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature\cite{buolamwini2018gender,wang2022towards}.
+Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
 %\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 

### 2023-02-12 chapters/introduction.tex
introduction rewritten

@@
 To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
-Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.\todo{Up to this.}
+Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
 
@@
 Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier.
-For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.
+For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more frequently than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.
 To this end, different fairness enhancing algorithms have been devised to improve fairness with respect to one or multiple fairness metrics.
@@
 Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting or explaining of sources of unfairness.
 
@@
 Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
-Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.
-sensitive features and ignore correlations among non-sensitive features.
+Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.\ sensitive features and ignore correlations among non-sensitive features.
 As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remain major challenges.
@@
 {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.
-Because SSAT encoding allows separate quantification to each sensitive feature without any restriction on the number of features.
-In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}.
+Because SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby extending the verification problem to multiple sensitive features unlike earlier methods.
+In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic} because of the efficient SSAT encoding.
+%, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}. 
 %We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.
 
-Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
-Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
-Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
+Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+In the context of verifying the bias of linear classifiers, existing fairness verifiers suffer from two-fold limitations: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
+To alleviate these limitations, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
 {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
 To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
-Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
+Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics.
 We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
@@
 
-\subsection*{Identification of Sources of Unfairness}
-While fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
-In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}.
-To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
-Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias.
-FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias.
-We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features.
-Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.
+\subsection*{Explaining Fairness metrics: Identifying the Sources of Bias}
+Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
+Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
+To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
+In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier.
+Thus, \textit{we aim to design a bias attribution method that is global by construction}.
+In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features.
+Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
+In this thesis, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.} Building on global variance decomposition, we propose an algorithm {\fairXplainer} to estimate the fairness influence function (FIF) of features.
+In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods.
+We also demonstrate higher correlation of FIFs with fairness interventions, thereby demonstrating the importance of FIFs in designing improved fairness algorithms in future.
 
@@
 	\begin{itemize}
-		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for interpretable machine learning
-		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for fairness verification and FIF computation
+		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
+		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
+		\item \red{FairXplainer for explaining group-based fairness metrics in machine learning.}
 	\end{itemize}
@@
 We discuss preliminaries in Chapter~\ref{chapter:preliminaries}.
-In the first part of the thesis, we discuss an incremental learning framework based on MaxSAT solving for learning interpretable classifiers.
-We conclude this part by applying incremental learning on a more expressible but interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
-In the second part of this thesis, we discuss  fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
-We conclude this part by discussing the identification of sources of unfairness of machine learning classifiers in Chapter~\ref{chapter:fif}.
+In the first part of the thesis on interpretable machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers.
+We conclude this part by applying incremental learning on a more expressible yet interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
+In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
+We conclude this part by explaining group fairness metrics in Chapter~\ref{chapter:fif}.
 Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.

### 2023-02-13 chapters/background.tex
highlevel pass

@@
 
-A classifier $\alg$ that solely optimizes accuracy, i.e., the average number of times $\widehat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
+A classifier $\alg$ that solely optimizes accuracy, i.e.
+the average number of times $\widehat{Y} = Y$, may discriminate certain compound sensitive groups over others~\cite{chouldechova2020snapshot}.
 In the following, we describe two well-known fairness definitions: group fairness and causal fairness.
-To this end, we use $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $.
+To this end, we denote $ f(\alg, \mathcal{D}) $ to quantify the fairness of the classifier $ \alg $ given the distribution of features $ \mathcal{D} $.
 Alternatively, a fairness metric can be computed on a finite dataset instead of on the distribution.
@@
 \paragraph{Disparate impact ($ \mathsf{DI} $).} Disparate impact also belongs to independence measuring group fairness metrics.
-Disparate impact measures the ratio between the minimum and the maximum probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.
+Disparate impact measures the ratio between the minimum and the maximum conditional probability of positive prediction of the classifier over all sensitive groups, and prescribe the ratio to be close to $1$~\cite{feldman2015certifying}.
 Formally, the disparate impact of a classifier is
@@
 
-\paragraph{Fairness Certification.} We certify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, that quantifies the desired level of fairness.
+\paragraph{Fairness Verification.} We verify the fairness of a classifier by comparing $ f(\alg, \mathcal{D}) $ with a fairness threshold, denoted by $ \epsilon \in [0,1] $, that quantifies the desired level of fairness.
 In particular, a classifier is $ \epsilon $-fair with respect to statistical parity, equalized odds, predictive parity, and path-specific causal fairness if and only if $ f(\alg, \mathcal{D}) \le \epsilon $.

### 2023-02-13 chapters/fairness/fif/appendix.tex
highlevel pass

@@
 \chapter{Fairness Influence Functions}
-\section{Societal Impact}
-In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset.
+%\section{Societal Impact}
+%In this paper, we quantify the influence of input features on the incurred bias/unfairness of the classifier on a dataset.
 This quantification facilitates our understanding of potential features or subsets of features attributing highly to the bias.
@@
 
-\section{A Smoothing Operator $\textsc{Smooth}$: Cubic Splines}
+\section{A Smoothing Operator: Cubic Splines}
 \label{fairness_fairXplainer_sec:smoothing} 
@@
 
-\subsection{FIFs of Equalized Odds.}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $.
+\subsection{FIFs of Equalized Odds}  For equalized odds, we deploy {\fairXplainer} twice, one for computing FIFs on a subset of data points in the dataset where $ Y = 1 $ and another on data points with $ Y = 0 $.
 Since, the maximum of the sum of FIFs between $ Y = 1 $ and $ Y = 0 $ is the equalized odds of the classifier, we finally report FIFs of features corresponding to the maximum sum of FIFs between $ Y \in \{0, 1\} $.
 
-\subsection{FIFs of Predictive Parity.}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
+\subsection{FIFs of Predictive Parity}  To compute FIFs for predictive parity, we condition the dataset by the predicted class $ \widehat{Y} $ and separate into two sub-datasets: $ \widehat{Y} = 1 $ and  $ \widehat{Y} = 0 $.
 For each sub-dataset, we deploy $ \fairXplainer $ by setting the ground-truth class $ Y $ as label.
@@
 
-	\caption[Ablation study on FIFs: $ \tau $]{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
+	\caption[Ablation study on FIFs: effect of spline intervals]{Effect of spline intervals on the approximation error of statistical parity, root-mean square error (RMSE), and  execution time of {\fairXplainer}.}
 	\label{fairness_fairXplainer_fig:effect_spline_intervals_appendix}
@@
 
-	\caption[Ablation study on FIFs: $ \lambda $]{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
+	\caption[Ablation study on FIFs: effect of maximum order of intersectionality]{Effect of maximum order $ \lambda $ on the approximation error of statistical parity, root-mean square error (RMSE) and  execution time of {\fairXplainer}.}
 	\label{fairness_fairXplainer_fig:effect_maxorder_appendix}
@@
 
-\begin{figure}[!t]
+\begin{figure}
 	\begin{minipage}{0.45\textwidth}

### 2023-02-13 chapters/fairness/fif/fairxplainer.tex
introduction updated
highlevel pass

@@
 \begin{algorithm}[t!]
-	\caption{{\fairXplainer}}\label{fairness_fairXplainer_algo:framework}
+	\small{
+	\caption{{\fairXplainer}: An algorithm for estimating FIFs}\label{fairness_fairXplainer_algo:framework}
 	\begin{flushleft}
-	\hspace*{\algorithmicindent}\textbf{Input:} Classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y}$, Dataset $\mathbf{D} = \{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$, Fairness metric $f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$, Maximum order of intersectional influence $\lambda $\\
+	\hspace*{\algorithmicindent}\textbf{Input:} Classifier $ \alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y}$, dataset $\mathbf{D} = \{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$, fairness metric $f(\alg, \mathbf{D}) \in \mathbb{R}^{\geq 0}$, and maximum order of intersectional influence $\lambda $\\
 	\hspace*{\algorithmicindent}\textbf{Output:} FIF $ w_{\mathbf{S}}$ for the subsets of features $\{\feature_{\mathbf{S}}\}$
@@
 		\State $ \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}} \leftarrow \textsc{Smooth}\big(\{\fiffunc_{\mathbf{a}}(\mathbf{z}^{(i)}) - \fiffunc_{\mathbf{a}, 0} - \sum_{\mathbf{S} \ne \mathbf{S}'} \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}'}(\mathbf{z}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\big)
-		$\label{fairness_fairXplainer_alg_line:backfitting_step} \Comment{Backfitting}
+		$\label{fairness_fairXplainer_alg_line:backfitting_step}
 		\State $\widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}}  \leftarrow \widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}} - \textsc{Mean}\big(\{\widehat{\fiffunc}_{\mathbf{a}, \mathbf{S}}(\mathbf{z}_{\mathbf{S}}^{(i)})\}_{i=1, \mathbf{a}^{(i)} = \mathbf{a}}^n\big)$ \label{fairness_fairXplainer_alg_line:mean_centered} \Comment{Mean centering}
@@
 	\end{algorithmic}
+	}
 \end{algorithm}

### 2023-02-13 chapters/fairness/justicia/experiments.tex
highlevel pass

@@
 
-\subsubsection{Verification: Detecting Compounded Discrimination in Sensitive Groups.}
+\subsubsection{Verification: Detecting Compounded Discrimination in Sensitive Groups}
 We have tested {\justicia} for datasets consisting of multiple sensitive features and reported results in Figure~\ref{fairness_justicia_fig:sensitive_groups}.
@@
 
-\subsubsection{Robustness: Stability to Sample Size.} 
+\subsubsection{Robustness: Stability to Sample Size} 
 We have compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics.
@@
 
-\subsubsection{Comparative Evaluation of Different Encodings.}
+\subsubsection{Comparative Evaluation of Different Encodings}
 Both {\justiciaenum} and {\justicialearn}  have the same output according to Lemma~\ref{fairness_justicia_lm:equivalence}.

### 2023-02-13 chapters/fairness/justicia/introduction.tex
addressing reviewers comments
highlevel pass

@@
 			\end{tikzpicture}
-	}\label{fairness_justicia_fig:fair_example_dependency}}\hfil
+	}\label{fairness_justicia_fig:fair_example_dependency}}\hfill
 	\subfloat[Age-dependent distributions of non-sensitive features]{
@@
 			\end{tikzpicture}}
-	\label{fairness_justicia_fig:fair_example_dt}}\hfil
+	\label{fairness_justicia_fig:fair_example_dt}}\hfill
 	\subfloat[Group specific conditional probability of the classifier's positive prediction]{
@@
 
-\subsubsection{Fair Machine Learning} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
+\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
 In this chapter, we focus on a popular concept of fairness, known as group fairness.
@@
 Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
-Thus, we see that the metrics of fairness can be contradictory and complimentary depending on the application and the data~\cite{corbett2018measure}.
+Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}.
 To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions.
@@
 
-\subsubsection{Fairness Verifiers} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
+\paragraph{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
 
@@
 
-\subsubsection{Our Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}.
+\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}.
 SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}.

### 2023-02-13 chapters/fairness/justicia/practical.tex
with new formatting
highlevel pass

@@
 
-\subsubsection{Extending {\justicia} to Decision Trees and Linear Classifiers. }
+\subsubsection{Extension to Decision Trees and Linear Classifiers}
 In the SSAT approach, we assume that the classifier $\alg$ is represented as a CNF formula.  
@@
 
-\subsubsection{Extending to Continuous Features.}
+\subsubsection{Extension to Continuous Features}
 In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features.

### 2023-02-13 chapters/interpretability/imli/incremental_learning.tex
highlevel pass

@@
-\section{Incremental Learning with MaxSAT-based formulation}
+\section{Incremental Learning with MaxSAT}
 \label{interpretability_imli_sec:incremental_learning}
@@
 For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch.
-Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that distribution remains same across batches.
+Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that the probability distribution of features remains same across batches.
 One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning.

### 2023-02-13 chapters/interpretability/imli/problem_formulation.tex
notations updated
highlevel pass

@@
  \label{interpretability_imli_sec:problem}
-Given 
+We are given 
 \begin{enumerate}
@@
 	\item a positive integer $ k \ge 1 $ denoting the number of clauses to be learned in the classification rule, and
-	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $,
+	\item a regularization parameter $ \lambda \in \mathbb{R}^+ $.
 \end{enumerate}
-we learn a classifier $ \Rule $ represented as a $ k $-clause CNF formula that can separate samples of class $ 1 $ from class $ 0 $.
+Our goal is to learn a rule-based classifier $ \Rule $ represented as a $ k $-clause CNF formula separating samples of class $ 1 $ from class $ 0 $.
 
- Our goal is to learn classifiers that balance two 
+ We learn classifiers that balance two 
  goals: of  being accurate but also interpretable.  

### 2023-02-13 chapters/fairness/fif/conclusion.tex
fif edits fixed
added appendix

@@
 %\vspace{-0.5em}
-\section{Conclusion}
+\section{Chapter Summary}
 %\vspace{-1em}
 
-In this paper, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset.
+In this chapter, we propose the Fairness Influence Function (FIF) to measure the effect of input features on the bias of classifiers on a given dataset.
 Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning.

### 2023-02-13 chapters/fairness/fif/introduction.tex
added appendix

@@
 %\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
-In this paper, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}
+In this chapter, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.}
 
@@
 
-\begin{enumerate}[leftmargin=*]
+\begin{enumerate}
 
@@
 
-	Using the methods described in this paper, we examine the sources of unfairness in DT$ 1 $ by calculating the FIFs of each feature subset.
+	Using the methods described in this chapter, we examine the sources of unfairness in DT$ 1 $ by calculating the FIFs of each feature subset.
 Positive values represent a reduction in fairness due to increased statistical parity, while negative values indicate an improvement in fairness.

### 2023-02-13 chapters/appendix.tex
fvgm appendix

--- 2022-12-02 chapters/appendix.tex
@@
-\appendix
-
-\chapter{Feature Correlations in SSAT-based Fairness Verifier: {\justicia}}
-\label{sec:CNF_feature_correlation}
+\chapter{Feature Correlations in SSAT-based Fairness Verifier}
+%\label{chapter:CNF_feature_correlation}
 In Chapter~\ref{chapter:justicia}, {\justicia} verifies the fairness of CNF classifiers without considering correlation among features.
-In this chapter, we address fairness verification of CNF classifiers with correlated features.
+Now, we address fairness verification of CNF classifiers with correlated features.
 In particular, we present the encoding of conditional probabilities from a Bayesian network into the SSAT formulation in {\justicia}.
@@
 We now discuss the quantifiers in $ \phi_{\widehat{Y}} \wedge \phi_\BN $, the SSAT solution of which  constitutes the maximum (minimum) probability of positive prediction of a CNF classifier.
-$ \phi_{\widehat{Y}} \wedge \phi_\BN $ contains four  categories of variables : (i) sensitive variables $ \sensitive $, (ii) non-sensitive variables $ \nonsensitive $, (iii) indicator variables $  \lambda_{V_i} $, and (iv) auxiliary variables $ \lambda_{v,\mathbf{u}} $.
+$ \phi_{\widehat{Y}} \wedge \phi_\BN $ contains four  categories of variables: (i) sensitive variables $ \sensitive $, (ii) non-sensitive variables $ \nonsensitive $, (iii) indicator variables $  \lambda_{V_i} $, and (iv) auxiliary variables $ \lambda_{v,\mathbf{u}} $.
 Among them, (iii) and (iv) are associated with $ \phi_\BN $ and the rest for $ \phi_{\widehat{Y}} $.

### 2023-02-13 chapters/fairness/fvgm/appendix.tex
fvgm appendix

@@
-\clearpage
-\appendix
+\chapter{Fairness Verification with Feature Correlation}
+	In this chapter, we discuss the extended experimental results as a continuation of Chapter~\ref{chapter:fvgm}.
 
-\clearpage
+	\section{Extended Experimental Results}
+	\label{fairness_fvgm_appendix:experiments}
+	Each experiment is performed on Intel Xeon E$ 7-8857 $ v$2 $ CPUs with $ 16 $GB memory, $ 64 $bit Linux distribution based on Debian OS and clock speed $ 3 $ GHz.
 
-		\section{Broader Impact}
-		We have presented an efficient framework for verifying fairness of linear classifiers.
-This work does not bear any negative societal impact.
-The sole purpose of our framework is to detect bias of machine learning models that are deployed in high-stake decision making.
-
-		\section{Background}
-
-		\subsection{Fairness Metrics}
-		In this section, we state the fairness metrics verified in this paper in further details.
-		\subsubsection{Group Fairness.} Group fairness is categorized into three families: independence, separation and sufficiency, of which {\framework} verifies independence and separation metrics. 
-		The independence metrics state that the predicition of the classifier should be independent of compound sensitive groups.
-Formally, independence notion specifies an equal PPV across all sensitive groups for a classifier $\alg$, i.e., $\Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  =  \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'] , \forall \mathbf{a}, \mathbf{a}' \in A$.
-		Since satisfying independence exactly is hard, relaxations of independence fairness metrics, such as \textit{disparate impact} and \textit{statistical parity}~\cite{dwork2012fairness,feldman2015certifying}, are proposed.
-
-		\textit{Disparate impact} (DI)~\cite{feldman2015certifying} measures the ratio of PPVs between the most favored group and least favored group, and prescribe it to be close to $1$.
-Formally, a classifier satisfies $(1 - \epsilon)$-disparate impact if, for $\epsilon \in [0,1] $,
-		\[
-		\min_{\mathbf{a}} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}]  \ge (1 - \epsilon) \max_{\mathbf{a}'} \Pr[\hat{Y} =1 | \mathbf{A} =  \mathbf{a}'].
-		\]
-		Another popular relaxation of independence metrics  is \textit{statistical parity} (SP) that measures the difference of PPV among sensitive groups, and prescribe this to be near zero.
-Formally, an algorithm satisfies $\epsilon$-statistical parity if, for $\epsilon \in [0,1] $,
-		\[
-		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 | \mathbf{A} = \mathbf{a}] - \Pr [\hat{Y} = 1| \mathbf{A} = \mathbf{a}']| \le \epsilon.
-		\]
-		For both disparate impact and statistical parity, lower value of $\epsilon$ indicates higher group fairness of the classifier $\alg$.
-
-		In the \textit{separation (or classification parity)} notion of fairness, the predicted label $\hat{Y}$ of a classifier $\alg$ is independent of the sensitive features $\sensitive$ given the class labels $Y$.
-In case of binary classifiers, a popular separation metric is \textit{equalized odds} (EO)~\cite{hardt2016equality} that computes the difference of false positive rates (FPR) and the difference of true positive rates (TPR) among all compound sensitive groups.
-		Lower value of equalized odds indicates better fairness.
-		A classifier $\alg$ satisfies $\epsilon$-equalized odds if, for all compound sensitive groups $\mathbf{a}, \mathbf{a}' \in A$,
-		$ \max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 0  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 0]| \le \epsilon, $ and $
-		\max_{\mathbf{a}, \mathbf{a}'}|\Pr[\hat{Y} =1 |\mathbf{A}= \mathbf{a}, Y= 1  ] - \Pr [\hat{Y} = 1|\mathbf{A}= \mathbf{a}', Y = 1]| \le \epsilon.
-		$
-
-		\subsubsection{Path-specific Causal Fairness.}
-		Let $ \mathbf{a}_{\max}  \triangleq \argmax_{ \mathbf{a}} \Pr[\hat{Y} =1 |\mathbf{A}=  \mathbf{a}] $.
-We consider mediator features $ \mediator \subseteq \nonsensitive $ sampled from the conditional distribution $ {\mathcal{Z}_{|\mathbf{A} = \mathbf{a}_{\max}}} $.
-This emulates the fact that mediator variables have the same sensitive features $ \mathbf{a}_{\max} $.
-For $ \epsilon \in [0,1] $,  path-specific causal fairness is defined as
-		$
-		\max_{\mathbf{a}, \mathbf{a}'} |\Pr[\hat{Y} = 1 | \sensitive =  \mathbf{a}, \mediator] - \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}', \mediator ]| \le \epsilon
-		$.
-
-		Therefore, PCF constrains that $ \hat{Y} $ is not directly dependent of $ \sensitive $ while $ \sensitive $ may indirectly affects $ \hat{Y} $ only through $ \mediator $.
-PCF is a variation of counterfactual fairness and causal fairness without mediator features~\cite{bastani2019probabilistic}.
-
-		\begin{example}
-			Following~\cite{bastani2019probabilistic}, we consider a classifier that decides the hiring of employees based on three features: gender (sensitive), years of experience (non-sensitive), and college-participation (mediator).
-It is practical to consider that gender $ \in $ \{male, female\} can affect the college-participation of individuals, and all three features are determining factors for the hiring process.
-Let `male' be the most favored group by the classifier, for instance.
-Path-specific causal fairness (PCF) ensures that a female candidate should be given a job offer with similar probability as a male candidate (by constraining $ \epsilon \approx 0 $).
-She,  however,  went to (participated in) college as if she were a male candidate while other non-mediator features such as  `years of experience' are the same.
-Therefore, PCF measures the effect of gender on job offer, but ignores the effect of gender on whether candidates went to college.
-		\end{example}
-
-%		\newpage
-		\section{Proofs of Theoretical Results}
-		\iffalse
-		\begin{lemmarep}
-			$ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.\footnote{\blue{This is the rectified version of Lemma~\ref{lm:property-subset-sum}.
-We note that this correction does not change any claim in Section~\ref{sec:fvgm} except the proposed heuristic in Section~\ref{sec:dp_formulation}, which we address in final version of the paper.}}
-		\end{lemmarep}
-		\begin{proof}
-		The quantifiers in $\mathbf{B}$ can be either universal, existential, or random.
-The sum of weights for existential and universal variables in $\mathbf{B}$ are calculated as $ w_{\exists} =  \sum_{i|q_i = \exists} \max\{w_i, 0\}$ and  $ w_{\forall} =  \sum_{i|q_i = \forall} \min\{w_i, 0\}  $, respectively by following Eq.~\eqref{eq:dp_recurse}.
-Since the contributions of $ w_\exists $ and $ w_\forall $ in $S(B,\tau)$ are fixed, it suffices to prove that $ S(B,\tau) $ does not rely on the order of \textit{random variables} in $\mathbf{B}$.
-
-		Let  $ \tau' =  \tau - w_\exists - w_\forall$ be the revised threshold for random variables $ B' = \{B_i | q_i = \R\} \subseteq\mathbf{B}$.
-Then $ S(B,\tau) $ is equivalent to solving $ S(B', \tau') $, and we show that $ S(B', \tau') $ does not rely on the order of $ B' $.
-
-		We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ based on its assignment $\mathbf{B}$. 
-\begin{align*}
-		f_w(i, b) = \begin{cases}
-	w_i &\text{ if }\mathbf{B}= 1, \\
-	0 &\text{ if }\mathbf{B}= 0.
-	\end{cases}
-\end{align*}
-		Let $ \sigma \in \{0,1\}^{|B'|} $ be a vector of  assignments of variables in $ B' $. Then, $ S(B', \tau') $ is computed as
-		\[ S(B', \tau') = \sum_{\sigma \in \{0,1\}^{|B'|}}  \mathds{1}[\sum_{i=1}^{|B'|} f_w(i, \sigma_i) \ge \tau'] \prod_{i=1}^{|B'|}\Pr[B_i = \sigma_i]. \]
-		 The aforementioned equation represents a stochastic subset-sum problem, where for each assignment $ \sigma $, we compute the decision version of the subset-sum problem corresponding to $ \sigma $ and multiply with probabilities of random-variables.
-We observe that $ S(B',\tau') $ does not rely on the order of $ B' $, because both multiplication and addition operations are commutative for real numbers.
-			\end{proof}
-		\iffalse
-		\begin{lemmarep}
-			Let $ Q = \{q_i\} $ be the set of quantifiers of $\mathbf{B}$.
-If $ |Q| = 1$, then $ Q $ is called homogeneous, and otherwise $ Q $ is called heterogeneous.
-If $ Q $ is homogeneous, then $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.
-Otherwise, $ S(B, \tau) $ is order-specific to $\mathbf{B}$.
-		\end{lemmarep}
-
-		\begin{proof}
-			When $ |Q|  = 1 $, all variables in $\mathbf{B}$ are either existential, universal or random.
-When all variables are existential, $ S(B,\tau) $ is solved linearly in $ |B| $ as $ S(B, \tau) = \mathds{1}[\sum_i \max\{w_i, 0\} \ge \tau] $.
-Similarly, when all variables are universal, $ S(B, \tau) = \mathds{1}[\sum_i \min\{w_i, 0\} \ge \tau] $.
-In both cases,  the sum of $ |B| $ integers does not depend on the order of integers.
-
-			We consider an indicator function $ f_w(i, b) $ that outputs the weight of $ B_i $ depending on its assignment $\mathbf{B}$. 
-			\[
-			f_w(i, b) = \begin{cases}
-			w_i \text{ if }\mathbf{B}= 1 \\
-			0 \text{ if }\mathbf{B}= 0
-			\end{cases}
-			\]
-			Let $ \sigma \in \{0,1\}^{|B|} $ be a vector of  assignments of $\mathbf{B}$. When all variables are random, $ S(B, \tau) $ is computed as
-			\[ S(B, \tau) = \sum_{\sigma \in \{0,1\}^{|B|}}  \mathds{1}[\sum_{i=1}^{|B|} f_w(i, \sigma_i) \ge \tau] \prod_{i=1}^{|B|}\Pr[B_i = \sigma_i], \]
-			which also does not rely on the order of $\mathbf{B}$. Therefore, if $ Q $ is homogeneous, $ S(B,\tau) $ does not rely on the order of $\mathbf{B}$.
-
-			\red{Lemma is not correct}
-
-		\end{proof}
-	\fi
-
-		\begin{lemmarep}
-			The decision version of the fairness verification problem for linear classifiers is in $\mathrm{NP^{PP}}$.
-		\end{lemmarep}
-
-		\begin{proof}
-
-			\textbf{Case 1: Subset-sum problem is $ \mathrm{NP} $-complete by a reduction to $ 3 $-SAT.}
-
-			We show that subset-sum problem is a $ \mathrm{NP} $-complete problem.
-In particular, we show that computing $ \sum_{i} w_iB_i = \tau $ for $ B_i \in\mathbf{B}$ is $ \mathrm{NP} $-complete through a reduction to $ 3 $-SAT~\cite{kleinberg2006algorithm}.
-$ 3 $-SAT problem computes a satisfying assignment of  a CNF formula, say $ \phi $, defining over $ n $ variables $ X_1, \dots, X_n $ and $ m $ clauses $ C_1, \dots, C_m $ where each clause has three literals.
-We start by defining two integer weights $ a_i $ and $ b_i $ associated with each variable $ X_i $, where including $ a_i $ corresponds to assigning $ X_i $ to true and including $ b_i $ corresponds to assigning $ X_i $ to false.
-We compute $ a_i $ and $ b_i $ in the following.
-			\begin{align*}
-				&a_i = 10^{m+i} + \sum_{j|X_i \in C_j} 10^j\\
-				&b_i = 10^{m+i} + \sum_{j|\neg X_i \in C_j} 10^j
-			\end{align*}
-
-			We construct a subset-sum problem of $ 2n $ variables with weights $ W =  \cup_{i=1}^n\{a_i, b_i\} $.
-Concretely, let $\mathbf{B}= \{B_{1,1}, B_{1,2}, \dots, B_{n,1}, B_{n,2}\} $ be the set of variables such that $ w(B_{i,1}) = a_i $ and $ w(B_{i,2}) = b_i $ and our goal is to compute an assignment of $\mathbf{B}$ such that $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $.
-
-			Let $ \sigma $ be a satisfying assignment of $ 3 $-SAT formula $ \phi $.
-Given the assignment $\sigma$, we consider a \textit{solution} of the subset-sum problem by assigning $ B_{i, 1} = 1 $ and $ B_{i, 2} = 0 $ whenever $ \sigma(X_i) = 1 $, and $ B_{i, 1} = 0 $ and $ B_{i, 2} = 1 $ whenever $ \sigma(X_i) = 10 $.
-Intuitively,
-			we consider a subset of weights from $ W $ by including 
-			 $ a_i $ if $ X_i $ is true and otherwise we include $ b_i $.  
-			 The sum of \textit{included} weights has following properties. 
-			\begin{itemize}
-				\item  the sum has $ 1 $ in leftmost $ n $ digits, corresponding to $ 10^{m+i} $ for $ i = 1, \dots, n $ as we include one of $ a_i $ or $ b_i $ for each $ i $. 
-				\item the sum has $ 1, 2 $, or $ 3 $ in the next $ m $ digits.
-The digit $ 10^j $ is the number of true literals in clause $ C_j $ and it is non-zero since $ \sigma $ satisfies each clause in  $ \phi $.
-				\item $ 0 $ in the rightmost digit.
-			\end{itemize}
-
-			In the subset-sum problem, we set the threshold $ \tau = \sum_{i=1}^n 10^{m+i} + 3 \sum_{j=1}^m 10^j $.
-We additionally consider integer weights $ c_j =  d_j = 10^j $, which will be added to $ W $ as discussed next.
-For brevity of presentation, we are not assigning Boolean variables in $\mathbf{B}$ for additional weights $ c_j $ and $ d_j $.
-			 We then prove that  $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ if and only if $ \phi $ is satisfiable.
-
-			\begin{itemize}
-				\item If $ \phi $ is satisfiable, then we select weights $  a _i $ if $ \sigma(X_i) = 1 $, else $ \mathbf{B}_i $.
-As discussed above, the sum of selected weights has leftmost $ n $ digits of $ \tau $ correct.
-To make remaining digits correct, we may need to add both of $ c_j $ and $ d_j $, one of $ c_j $ and $ d_j $, or none of $ c_j $ and $ d_j $ depending  if the digit $ 10^j $ in the sum is $ 1,2, $ or $ 3 $.
-We emphasis that this addition is deterministic as it relies on the number of true literals in $ C_j $ for the assignment $ \sigma $.
-Thus, for a satisfying assignment of $ 3 $-SAT, there exists a solution of the subset-sum problem.
-
-				\item Finally, we prove that a solution of the subset-sum problem corresponds to a satisfying assignment of $ \phi $.
-First we observe that in any digit $ 10^k $, for $ k \le m $, there are at most five weights with $ 1 $ in digit $ 10^k $: the three literals corresponding to clause $ k $ and $ c_k $ and $ d_k $ (otherwise when $ k > m $, there are at most $ 2 $  weights with $ 1 $ in digit $ 10^k $).
-Since there are at most $ 5 $ ones in any position, a subset of these weights will not cause any carries in addition.
-Hence, we can only get a total of $ \tau $ by including the right number of $ 1 $s in any digit.
-To do this, one needs to include exactly one of $ a_i $ and $ b_i $, and the corresponding truth assignment $ \sigma $ needs to satisfy $ \phi $ so that in each position $ 1 \le j \le m $, we get at least one $ 1 $ in digit $ 10^j $.
-We can then add $ c_j $ and $ d_j $ to increase the sum of weights to $ \tau $, as discussed previously.
-This, for a solution of the subset-sum problem, there is a corresponding satisfying assignment of $ 3 $-SAT
-			\end{itemize}
-
-			In the above analysis, the construction of $ a_i $ and $ b_i $ has polynomial complexity with respect to $ n $ and $ m $.
-Therefore, the subset-sum problem $ \sum_{i} w(B_{i,1})B_{i,1} + w(B_{i,2})B_{i,2} = \tau $ and equivalently, $ \sum_{i} w_iB_i = \tau $ is $ \mathrm{NP} $-complete because $ 3 $-SAT is also $ \mathrm{NP} $-complete.
-We further claim that computing $ \sum_{i} w_iB_i \ge \tau $ is at least as hard as computing $ \sum_{i} w_iB_i = \tau $, which is required for analyzing the complexity of stochastic subset-sum problem, as discussed next.
-
-			\textbf{Case 2: Stochastic subset-sum problem is $ \mathrm{PSPACE} $-complete by a reduction to $ 3 $-SSAT.}
-
-			We observe that a $ 3 $-SSAT formula that is based on a $ 3 $-SAT formula $ \phi $ and an additional specification of quantifiers for each variable in $ \phi $.
-Concretely, we define a $ 3 $-SSAT formula $$ \Phi \triangleq (\{q(X_i)\}_{i=1}^{n}, \phi) \text{ , where } q(X_i) \in \{\exists, \forall, \R^{p}\} .$$ To analyze the complexity of a stochastic subset-sum problem, we discuss a procedure in the following that is based on a polynomial reduction to $ 3 $-SSAT.
-
-			We apply the aforementioned construction procedure ($ 3 $-SAT to subset-sum problem) to translate  a $ 3 $-SSAT formula to a stochastic sub-set sum problem.
-In particular, for each quantified variable in $ \phi $, we now discuss the quantifier of variables in the stochastic subset-sum problem.
-If $ X_i $ is $ \exists $ (resp.\ $ \forall $) quantified, associated $ B_{i,1} $ and $ B_{i,2} $ are also $ \exists $ (resp.\ $ \forall $) quantified.
-Moreover, if $ X_i $ is randomized ($ \R^{p_i} $) quantified with $ p_i = \Pr[X_i = 1] $, then both $ B_{i,1} $ and $ B_{i,2} $ are randomized quantified with probability $ \Pr[B_{i,1}] = p_i $ and $ \Pr[B_{i,2}] = 1 - p_i $.
-Since the decision version of $ 3 $-SSAT is $ \mathrm{PSPACE} $-complete~\cite{littman2001stochastic}  and the presented reduction procedure has polynomial complexity, the decision version of stochastic subset-sum if also $ \mathrm{PSPACE} $-complete.
-
-			\textbf{Case 3: Fairness verification for linear classifiers is in
-			$\mathrm{NP^{PP}}$ by a reduction to E-MAJSAT.}
-
-			E-MAJSAT is a specific quantifier arrangement of SSAT problems where existential (or universal) variables appear before random variables. Formally, a E-MAJSAT formula 
-			$$ \Phi =  (\{q(X_1) = \exists,\cdots,q(X_i) =\exists, q(X_{i+1}) =\R^{p_{i+1}}, \cdots, q(X_n) = \R^{p_{n}} \}, \phi).$$ The decision version of E-MAJSAT problem has complexity in $\mathrm{NP^{PP}}$~\cite{littman2001stochastic}.
-Furthermore, the decision version of SSAT problems with universal quantifiers followed by random-quantifiers are also in $\mathrm{NP^{PP}}$ as this a dual problem of E-MAJSAT~\cite{littman2001stochastic}.
-			In Section~\ref{sec:fvgm}, we present fairness verification of linear classifiers defined over Boolean variables using a formulation of stochastic subset-sum problem.
-Similar to E-MAJSAT,  existential (or universal) variables appear before random variables in our formulation.
-Thus, the decision version of fairness verification of linear classifiers is in $\mathrm{NP^{PP}}$.
-			\iffalse 
-			\red{We further highlight that verification of linear classifiers on continuous real-valued variables is at least as hard as verifying classifiers on Boolean variables.}\fi
-		\end{proof}
-		\fi
-
-		\begin{lemmarep}
-			Let $ n' $ be the number of existential and universal variables in $ \mathbf{B} $.
-Let $ w_{\exists} = \sum_{B_i \in \mathbf{B} | q_i = \exists} \max\{w_i, 0\}$  and $ w_{\forall} = \sum_{B_i \in \mathbf{B} | q_i = \forall} \min\{w_i, 0\}$ be the considered sum of weights of existential and universal variables, respectively.
-We can exactly solve {\stochastic} using dynamic programming with time complexity $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $.
-The space complexity is  $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
-		\end{lemmarep}
-
-		\begin{proof}
-			\textbf{Case 1: All $n$ variables in $\mathbf{B}$ have randomized quantifiers.}
-
-			At step-$i$ of the dynamic programming (Eq.~\eqref{eq:dp_recurse}), we modify the residual threshold of that step, namely $\tau_i$, either by subtracting $w_i$ or by retaining it.
-			Now, we observe that the residual threshold $\tau_i$ for any $i \in \lbrace1,\ldots,n\rbrace$ will be in $[0, \tau+|w_{neg}|]$.
-This holds because if $\tau_i$ crosses these bounds, the dynamic programming is terminated as shown in Equation~\eqref{eq:dp_terminus}.
-			Since all weights of $\{w_i\}_{i=1}^{n}$ are integers, the maximum number of values that the residual threshold can take, is $ (\tau + |w_{neg}|) $.
-			Thus, we need to store at most $ n(\tau + |w_{neg}|)$ values in the memory for performing dynamic programming with $n$ variables and $ (\tau + |w_{neg}|) $ number of possible weights.
-			Thus, the space complexity  is $ \mathcal{O}(n(\tau + |w_{neg}|)) $.
-
-			In order to construct the dynamic programming table, we have to call the $\mathsf{dp}$ function $ \mathcal{O}(n(\tau + |w_{neg}|)) $ times, in the worst-case.
-			Thus, the time complexity of our method is $ \mathcal{O}(n(\tau + |w_{neg}|) $.
-
-			\textbf{Case 2: $n'$ variables have existential or universal quantifiers and $n-n'$ variables have randomized quantifiers in $\mathbf{B}$.}
-
-			According to Eq.~\eqref{eq:dp_recurse}, $ w_\exists $ and $ w_\forall $ are the fixed weights of all existential and universal variables, respectively.
-Therefore, we need to consider at most $ \tau + |w_{neg}| -  w_\exists - w_\forall $ values of residual weights for random variables.
-By applying analysis in Case 1, the space and time complexity is derived as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
-
-			We note that there is an additional time complexity of $ \mathcal{O}(n') $ for existential and universal variables in Eq.~\eqref{eq:dp_recurse}.
-Thus the time complexity becomes $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall}) + n') $.
-We, however, do not require to store any entry for existential and universal variables in $ \mathsf{dp} $  function and thus, the space complexity remains the same as $ \mathcal{O}((n - n')(\tau + |w_{neg}| - w_{\exists} - w_{\forall})) $.
-		\end{proof}
-
-		\begin{lemmarep}
-			Let $ \mathbf{V} \subseteq \mathbf{B} $ be the set of vertices in the Bayesian network and $ n'' $ be the number of existential and universal variables in $ \mathbf{B} \setminus \mathbf{V} $.
-Let $ w'_{\exists} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \exists} \max\{w_i, 0\}$  and $ w'_{\forall} = \sum_{B_i \in \mathbf{B} \setminus \mathbf{V} | q_i = \forall} \min\{w_i, 0\}$ be the sum of considered weights of existential and universal variables, respectively that only appear in $ \mathbf{B} \setminus \mathbf{V} $.
-To exactly compute {\stochastic} with correlated variables in dynamic programming approach,  time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
-		\end{lemmarep}
-
-		\input{tables/sanity_check.tex}
-
-		\begin{proof}
-			We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $.
-For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{sec:dp_with_BN}.
-
-			For variables in $\mathbf{B}\setminus \mathbf{V} $, we apply analysis from  Lemma~\ref{lemma:complexity_sss}, where we consider $ (n - n'' - |\mathbf{V}|) $ random variables, $ n'' $ existential/universal variables, and residual weights can take at most $ (\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) $ values.
-Hence, time complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $, and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $
-
-			Combining two cases, overall time complexity is $ \mathcal{O}(2^{|\mathbf{V}|} + (n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall}) + n'') $ and space complexity is $ \mathcal{O}((n - n'' - |\mathbf{V}|)(\tau + |w_{neg}| - w'_{\exists} - w'_{\forall})) $.
-		\end{proof}
-
-%	\newpage	
-	\section{Extended Experimental Evaluations}
-	\label{appendix:experiments}
-	Each experiment is performed on Intel Xeon E$ 7-8857 $ v$2 $ CPUs with $ 16 $GB memory, $ 64 $bit Linux distribution based on Debian OS and clock speed $ 3 $ GHz.
-In the following, we discuss extended experimental results.
+	\input{tables/fairness/fvgm/sanity_check.tex}
 
@@
 	We have considered a synthetic problem for comparing accuracy among different verifiers.
-For Example~\ref{example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $.
+For Example~\ref{fvgm_example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $.
 For `income' feature ($ I $), we consider two Gaussian distributions $ \Pr[I | \text{age} \ge 40] \sim \mathcal{N}(0.6, 0.1) $ and $ \Pr[I | \text{age} < 40] \sim \mathcal{N}(0.4, 0.1) $ separated by two age groups.
@@
 
-	In Figure~\ref{fig:synthetic_bn} we show the Bayesian Network on discretized features, in particular for income and fitness features.
-In Figure~\ref{fig:synthetic_ppv_max_LR} \ to Figure~\ref{fig:synthetic_ppv_min_SVM}, we show PPVs of different classifiers computed by different verifiers, where  {\framework} outputs closest to exactly computed values, in comparison with Justicia.
-In Figure~\ref{fig:synthetic_sample_size_LR} to Figure~\ref{fig:synthetic_sample_size_SVM}, we show the effect of sample size on {\framework} in measuring fairness metrics: disparate impact and statistical parity, where with increasing sample size, the estimate becomes more accurate.
+	In Figure~\ref{fairness_fvgm_fig:synthetic_bn} we show the Bayesian Network on discretized features, in particular for income and fitness features.
+In Figure~\ref{fairness_fvgm_fig:synthetic_ppv_max_LR} and Figure~\ref{fairness_fvgm_fig:synthetic_ppv_min_LR}, we show PPVs of logistic regression classifier computed by different verifiers, where  {\fvgm} outputs closest to exactly computed values, in comparison with Justicia.
+In Figure~\ref{fairness_fvgm_fig:synthetic_sample_size_LR_DI} and Figure~\ref{fairness_fvgm_fig:synthetic_sample_size_LR_SPD}, we show the effect of sample size on {\fvgm} in measuring fairness metrics: disparate impact and statistical parity, where with increasing sample size, the estimate becomes more accurate.
+Similar observations are recorded for SVM classifier in Figure~\ref{fairness_fvgm_fig:synthetic_results_SVM}.
 
-		\begin{figure}[!t]
+		\begin{figure}
 		\begin{center}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_LR}}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Adult_SVM}}	\\		
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_LR}}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Compas_SVM}}\\
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Adult_LR}}
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Adult_SVM}}	\\		
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Compas_LR}}
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Compas_SVM}}\\
 
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_LR}}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_German_SVM}}		\\
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_LR}}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Titanic_SVM}}\\
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_German_LR}}
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_German_SVM}}		\\
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Titanic_LR}}
+			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Titanic_SVM}}\\
 
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_LR}}
-			\subfloat{\includegraphics[scale=0.24]{figures/time_vary_features_Ricci_SVM}}
-
+%			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Ricci_LR}}
+%			\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/time_vary_features_Ricci_SVM}}
+%			
 			\caption{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
-			\label{fig:time_vary_features}
+			\label{fairness_fvgm_fig:time_vary_features}
 
@@
 
-	In Figure~\ref{fig:time_vary_features}, we present the runtime of different fairness verifiers while varying the number of features in different datasets.
+	In Figure~\ref{fairness_fvgm_fig:time_vary_features}, we present the runtime of different fairness verifiers while varying the number of features in different datasets.
 We observe that with an increase of features, the runtime increases in general.
 
-	\input{tables/fairness_attacks.tex}
-	%\input{tables/verifier_scalability.tex}
-	%\input{tables/ssat_vs_subset_sum.tex}
-	%\input{tables/fairness_algorithms.tex}
-	\input{tables/fairness_metrics_on_compound_sensitive_groups.tex}	
-	\input{tables/dag_structure_complexity.tex}
+	\input{tables/fairness/fvgm/fairness_attacks.tex}
+	%\input{tables/fairness/fvgm/verifier_scalability.tex}
+	%\input{tables/fairness/fvgm/ssat_vs_subset_sum.tex}
+	%\input{tables/fairness/fvgm/fairness_algorithms.tex}
+	\input{tables/fairness/fvgm/fairness_metrics_on_compound_sensitive_groups.tex}	
+	\input{tables/fairness/fvgm/dag_structure_complexity.tex}
 
@@
 
-	We show extended results on verifying fairness attack in Figure~\ref{fig:attack_extended} for two fairness metrics: disparate impact (DI) and statistical parity (SP).
-We observe that {\framework} can detect poisoning attack for both metrics.
-
-	In Figure~\ref{fig:multiple_metrics} we show verification results on compound sensitive groups with respect to multiple fairness metrics.
+	We show extended results on verifying fairness attack in Figure~\ref{fairness_fvgm_fig:attack_extended} for two fairness metrics: disparate impact (DI) and statistical parity (SP).
+We observe that {\fvgm} can detect poisoning attack for both metrics.
+	In Figure~\ref{fairness_fvgm_fig:multiple_metrics} we show verification results on compound sensitive groups with respect to multiple fairness metrics.
 In this figure, we observe that with an increase in the number of groups, fairness metrics worsens\textemdash disparate impact decreases and other three metrics increases.
-Additionally, in Table~\ref{fig:extended_verifying_fairness_algorithms}, we show extended results for German and COMPAS dataset on verifying fairness algorithms: reweighing and optimized pre-processing algorithms.
 
 	\subsection{Performance Analysis of Bayesian Network}
-	In Figure~\ref{fig:results_DAG_complexity}, we analyze the performance of encoding Bayesian Networks of differing complexity.
+	In Figure~\ref{fairness_fvgm_fig:results_DAG_complexity}, we analyze the performance of encoding Bayesian Networks of differing complexity.
 We define the complexity of the network as  $ \frac{|V|}{|\nonsensitive \cup \sensitive|} $, which is the ratio between the number of features appearing in the network and total features.
-In this figure, as the ratio increases, both computation time of {\framework} and learning time of Bayesian Network increase.
+In this figure, as the ratio increases, both computation time of {\fvgm} and learning time of Bayesian Network increase.
 
-	\begin{table}[!t]
+	\begin{comment}
+	\begin{table}
 
@@
 	\small
-	\caption{Extended results for verification of different fairness enhancing algorithms using {\framework}.
+	\caption{Extended results for verification of different fairness enhancing algorithms using {\fvgm}.
 Numbers in bold refer to fairness improvement w.r.t.
@@
 }
-	\label{fig:extended_verifying_fairness_algorithms}
+	\label{fairness_fvgm_fig:extended_verifying_fairness_algorithms}
 	\setlength{\tabcolsep}{.5em}
@@
 
+	\end{comment}
+
 \iffalse
@@
 We additionally show results for decision tree classifiers (DT).
-{\framework} takes comparatively more time for decision tree classifiers because of encoding feature-correlations represented as a Bayesian network.}
-			\label{tab:scalability_extended}	
+{\fvgm} takes comparatively more time for decision tree classifiers because of encoding feature-correlations represented as a Bayesian network.}
+			\label{fairness_fvgm_tab:scalability_extended}	
 		\begin{tabular}{llrrrrrrrrrrrrrr}
@@
 			\toprule
-			Dataset & Classifier & FairSquare & VeriFair & Justicia & {\framework} \\
+			Dataset & Classifier & FairSquare & VeriFair & Justicia & {\fvgm} \\
 			\midrule
@@
 %\newpage
-\input{sections/fairness_influence_functions.tex}
+\input{chapters/fairness/fvgm/fairness_influence_functions.tex}
 
@@
 Since there is no state of the art SSAT solver that can handle universal quantification, we cannot consider conditional probabilities of universal variables while encoding the Bayesian network.
-In {\framework}, universal quantification appears while learning the least favored group in the learning encoding.
+In {\fvgm}, universal quantification appears while learning the least favored group in the learning encoding.
 When we do not consider conditional probabilities on universal variables, earlier approach could solve this using an existential SSAT solver by first negating the CNF and then subtracting computed probability from $ 1 $ as a solution to the original universal SSAT problem.
@@
 
-\iffalse
-\clearpage
-
-\section{Progress}
-
-\begin{enumerate}
-	\item \textbf{Encoding conditional probabilities:} In our experiments, we have studied two algorithms for learning Bayesian network from data: Notears and pgmpy.
-After learning a DAG structure, we post-process it before feeding to {\framework}.
-
-	\begin{itemize}
-		\item \textbf{Modifying in-degree edges for sensitive variables.} In the SSAT-based encoding, sensitive variables have existential (or universal) quantification because we are interested in knowing which assignment to these variables result in the most (or least) favored group.
-
-		In a Bayesian network, a directed edge $ (a,b) $ between variable $ a $ and $\mathbf{B}$ refers to the conditional dependence of $\mathbf{B}$ on $ a $.
-We now consider a case where variable $\mathbf{B}$ is a sensitive variable.
-Since an existential (or universal) quantification on variable $\mathbf{B}$ fixes its assignment, we first reverse the direction of the edge to $ (b,a) $ in order to capture the dependence between two variables.
-We highlight that {\framework} does not have a strict restriction for DAG structure as input.
-Therefore, even if reversing an edge may violate DAG structure, {\framework} is not affected by this modification.
-
-		If both $ a $ and $\mathbf{B}$ refers to different categories for a sensitive attribute, more specifically they belong to the one-hot encoded variables for a sensitive attribute, we remove the edge $ (a, b) $ from the learned DAG.
-
-	\end{itemize}
-
-	\item \textbf{Verifying Path-specific Causal Fairness.}
-
-	\item \textbf{Fairness Attack Verification:} We apply {\framework} to verify fairness poisoning attack \cite{?}.
-This attack algorithm generates poisoned data, which together with initial training data results in a ML classifier with decaying fairness results.
-With {\framework} we verify the effectiveness of fairness poisoning attacks.
-
-	\item \textbf{Verifying pre-processing and post-processing fairness algorithms:} We apply {\framework} to verify the  effectiveness of fairness enhancing algorithms, in particular pre-processing and post-processing algorithms.
-
-	\begin{itemize}
-		\item \textbf{Pre-processing fairness algorithms.} Pre-processing algorithms process a  dataset such that any model learned on the dataset has better fairness results, computed in terms of disparate impact, statistical parity and so on.
-To verify the effectiveness of a pre-processing algorithm, we first train a classifier on unprocessed data and later another classifier on processed data.
-Finally, we compute the difference in different fairness metrics using {\framework}.
-For an effective fairness algorithm, disparate impact increases and statistical parity decreases for the processed data.
-
-		\item \textbf{Post-processing fairness algorithms.} 
-		Post processing fairness algorithms works on the output of any already trained classifier so that the final output by the fairness algorithm improves fairness of the classifier.
-\red{In order to verify a post-processing fairness algorithm, we learn a new classifier on the processed data\footnote{Learning a new classifier is not a requirement for approaches that can compute fairness on a dataset (includes class-label).
-However, our framework requires a classifier, which learns the mapping between input features and class-label.
-} and report difference in fairness metrics.
-The new classifier should have high accuracy on the processed data, otherwise the verification result by {\framework} is not reliable.
-}
-
-	\end{itemize}
-
-	\item \textbf{Feature Influence.} We employ {\framework} to compute feature-influence of each feature in the data.
-For a distribution $ \mathcal{D} $, a finite sample dataset $ D \sim \mathcal{D} $, and feature $ x $, we define its influence in terms of the positive predictive value (PPV) of the classifier.
-	\[
-	q(x) \triangleq  \Pr[\hat{Y}_{D_{-x}} = 1 | \mathcal{D}_{-x}] - \Pr[\hat{Y}_D = 1 | \mathcal{D}]
-	\]
-
-	In this equation, $ \mathcal{D}_{-x} $ (resp.\ $ D_{-x} $) refers to the distribution (resp.\ dataset) where all but feature $ x $ are present.
-	Intuitively, the influence of feature $ x $ is the difference in positive predictive value of the classifier as if $ x $ does and does not participate in the task of classification.
-Note that, the classifier is learned on dataset $ D $ whereas $ q(x) $ is computed for the distribution $ \mathcal{D} $.
-
-	\textbf{Outline:}
-	\begin{enumerate}
-		\item Train a classifier $ M $ on a dataset $ D $
-		\item From the dataset, compute product distribution of features $ \mathcal{D} $. This is handled by {\framework} \red{Can we implement an interface where the distribution is passed as inputs? }
-		\item Compute base values of different fairness metrics. In our experiments, we have computed disparate impact and statistical parity. '
-		\item To compute the influence of an attribute, we replace that attribute with random value on the dataset $ D' $.
-For a continuous attribute, we randomly generate attribute value within the maximum and minimum of original attribute-values.
-For a categorical attribute, we randomly select a category.
-		\item We again calculate distribution from $ D' $ and run {\framework} on the new distribution.
-		\item Since random-value replacement involves randomness, we have repeated the process for $ 5 $ iterations. 
-	\end{enumerate}
-
-	\textbf{Result Analysis}
-	We first learn dependency among features using Bayesian structure learning algorithm.
-For the dependency structure, we compute feature-influence and report in Figure~\ref{fig:feature_influence_compas}.
-In COMPAS dataset, we consider `race' as a sensitive attribute where race = 1 denotes Caucasian.
-In all three reported classifiers, prior-counts and juvenile-count are top influential features for both sensitive group.
-Other non-crime features such as race, age are also found to be influential for logistic regression classifier and CNF classifier.
-Moreover, depending on sensitive group, age and race have differentiating influence.
-	\red{Incomplete analysis}
-
-\end{enumerate}
-
-For the learning encoding, sensitive variables $ \sensitive $ are existentially quantified in $ \mathsf{prefix_{ERE}} $ and universally quantified in $ \mathsf{prefix_{URE}} $.
-These variables are followed by randomized quantified variables $ \lambda_{X_i}, \lambda_{x,\mathbf{u}} $ and $ X_j \in \nonsensitive \setminus V $.
-Finally, remaining variables such as $ \lambda_{\hat{Y}} $ and $ X_i $ for each $ X_i \in V(G) $ are existentially quantified.
-
-\paragraph{Extra}
-To this end, we consider all features $ \nonsensitive \cup \sensitive $ to be Boolean.
-We additionally constrain the classifier $ \hat{Y} $  to be a Boolean formula $ \phi_{\hat{Y}} $ in CNF defined over $ \nonsensitive \cup \sensitive $ such that, for an assignment of $ \nonsensitive \cup \sensitive $, $ \hat{Y} = 1 $ whenever $ \phi_{\hat{Y}} $ is true, and $ \hat{Y} = 0 $ otherwise.
-Additionally, we construct another CNF formula $ \phi_\BN $ to encode the joint distribution of $ \BN $, which we discuss shortly.
-
-One approach to verify different fairness metrics such as disparate impact, statistical parity, equalized odds, and path-specific causal fairness is to compute the most favored group and the least favored group based on the group specific PPV of the classifier~\cite{ghosh2020justicia}.
-As such, we can either (i) enumerate computation of group specific PPV for all sensitive groups or (ii) learn the most favored group and the least favored group, to compute these metrics.
-	Provided that we have  $ \phi_{\hat{Y}} $ and $ \phi_\BN $ representing the classifier and the Bayesian network, we now discuss two encodings in the following.
-
-\begin{enumerate}
-	\item \textbf{Enumeration.} We consider an RE-SSAT (Randomized Exponential Stochastic SAT) instance with random quantifier over Boolean variables to compute $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \BN] $.
-	\begin{align}
-	\Phi_{\mathbf{a}} = \mathsf{prefix_{Enum}}, \phi_{\hat{Y}} \wedge \phi_\BN \wedge (\sensitive = \mathbf{a})
-	\end{align}
-	where $ \mathsf{prefix_{Enum}} $ is an ordered sequence of randomized quantified Boolean variables and $ \sensitive = \mathbf{a} $ encodes the sensitive group $ \mathbf{a} \in \sensitive $ as a CNF.
-We discuss quantifiers in detail in Section~\ref{?}.
-An RE-SSAT formula can be solved by a projected model counter where we project model-counting (number of satisfying assignments in a CNF) on randomized quantified variables.
-Solving $ \Phi_{\mathbf{a}} $,  the probability of satisfaction $ \Pr[\Phi_{\mathbf{a}}] $ exactly computes group-specific PPV of the classifier, $ \Pr[\hat{Y} = 1 | \sensitive = \mathbf{a}, \BN] $.
-Finally, we enumerate this computation for all sensitive groups $ \forall \mathsf{a} \in A $ and use  $ \max_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $  and $ \min_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $\textemdash the PPV of the most and the least favored group, respectively\textemdash  to compute different fairness metrics.
-
-	\item \textbf{Learning.} The second encoding efficiently learns the most favored group and the least favored group by instantiating two SSAT formulas: ERE (Existential-Random-Existential) and URE (Universal-Random-Existential) SSAT formulas, respectively.
-Both SSAT formulas have the same CNF formula and only difference is in the prefix.
-	\begin{align}
-	&\Phi_\mathsf{ERE} = \mathsf{prefix_{ERE}}, \phi_{\hat{Y}} \wedge \phi_\BN\\
-	&\Phi_\mathsf{URE} = \mathsf{prefix_{URE}}, \phi_{\hat{Y}} \wedge \phi_\BN
-	\end{align}
-	$ \mathsf{prefix_{ERE}} $ (resp.\ $ \mathsf{prefix_{URE}} $) is an ordered sequence of existentially (resp.\ universally) quantified variables followed by randomized quantified variables.
-	We apply existential quantifier ($ \exists $) on sensitive features $ A $ in $ \Phi_\mathsf{ERE} $ using $ \mathsf{prefix_{ERE}} $.
-Therefore, according to rule(??) in Section\ref{??}, we learn an assignment to $ A $, say the most favored sensitive group,  resulting in the maximum satisfying probability of $ \Phi_\mathsf{ERE} $.
-Formally, $ \Pr[\Phi_\mathsf{ER}] $ is equal to $ \max_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $ in the enumeration encoding.
-We similarly apply universal quantifiers ($ \forall $) on $ A $ to learn the least favored sensitive group with $ \Pr[\Phi_\mathsf{URE}] $ equalizing to $ \min_{\mathbf{a}} \Pr[\Phi_{\mathbf{a}}] $.
-\end{enumerate}
-
-\red{Complexity analysis has to be verified.}
-Both enumeration and learning encoding output same results and their worst case complexity is same.
-In particular, solving an  RE-SSAT formula has complexity $ \text{PP}^{\text{NP}} $ and we enumerate for all sensitive groups, which is exponentially many in $ A $.
-Similarly, solving both ER-SSAT and URE-SSAT is $ \text{NP}^{\text{PP}} $ hard.
-
-We now discuss the construction of the CNF formula, particularly $ \phi_{\hat{Y}} $ and $ \phi_\BN $.
-Ghosh et al.~\cite{ghosh2020justicia} have proposed techniques to encode decision trees and linear classifiers as a CNF formula $ \phi_{\hat{Y}} $.
-For linear classifiers, the decision function is viewed as a pseudo-Boolean constraint and can be efficiently encoded to CNF by introducing auxiliary variables, denoted by $ \lambda_{\hat{Y}} $.
-Support-vector classifiers with linear kernels can also be encoded to CNF in the same approach to linear classifiers.
-Hence, we now focus on the construction of $ \phi_\BN $ to encode the Bayesian network to CNF.
-
-\red{Additional}
-\begin{enumerate}
-	\item Advance terminate when $ S(W,\tau, Q) \le w_{neg} $, where $ w_{neg} $ is the sum of negative weights in $ W $.
-	\item Within similar quantifiers, sort based on descending weights to leverage early termination.
-	\item We can learn maximum and minimum of $ S(W,\tau,Q) $ in one go (should be explained in dynamic programming).
-It is not required to convert weights to positive.
-One benefit of assuming all positive weights is that we then have one sub-problem with selection of weights (and this sub-problem has maximum value).
-However, to compute minimum of $ S $, we again need to consider the sub-problem with not selection of weights.
-But we can propose a dynamic approach where both max and min can be solved in one go.
-\end{enumerate}
-
-\section{Experimental Evaluations}
-
-\begin{figure}
-	\begin{center}
-		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Edges}}
-		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Time_(DAG)}}\\
-		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Disparate_impact}}
-		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_Statistical_parity}}\\
-		\subfloat[]{\includegraphics[scale=0.4]{figures/BN_ppv}}\\
-	\end{center}
-	\caption{Experimenting the effect of regularizer in DAG learner inside {\framework}.
-With increasing values of regularizer, learned DAGs become sparser by undertaking less learning time.
-For different DAGs, we evaluate positive predictive value (PPV) for a fixed classification algorithm.
-We occasionally observe that disparate impact  increases and statistical parity decreases for sparser DAG.
-We show break down result for two different sensitive groups where the group `Age $ = 1 $' is often (but not always) more discriminated than `Age $ = 0 $'  in majority values of regularizer.
-Therefore, {\framework} can verify bias induced by data (in form of Bayesian Network)  for the fixed classification algorithm.}
-\end{figure}
-\fi
-
-\input{sections/cnf_correlations}
-
-%\clearpage
-%	\bibliographystyle{plainnat.bst}
-%	\bibliography{ref.bib}
+\input{chapters/appendix.tex}

### 2023-02-13 chapters/fairness/fvgm/fairness_influence_functions.tex
fvgm appendix

@@
 In both cases, we denote the sensitive groups `male' and `female' as `sex=0' and `sex=1', respectively. 
-FIFs computed for these sensitive groups show influence of different features and their relevant disparities. We illustrate the results in Figure~\ref{fig:fif}.
+FIFs computed for these sensitive groups show influence of different features and their relevant disparities. We illustrate the results in Figure~\ref{fairness_fvgm_fig:fif}.
 
-Figure~\ref{fig:fif_a} illustrates that the PPV value for the `male' population in the dataset is $0.61$.
+Figure~\ref{fairness_fvgm_fig:fif_a} illustrates that the PPV value for the `male' population in the dataset is $0.61$.
 If we now replace the feature `age' with uniformly random values, the PPV for the same group becomes $0.46$, Thus, the FIF of the feature `age' for `male' group is $0.15$.
-Similarly, all the red bars in Figures~\ref{fig:fif_a}-\ref{fig:fif_b}, ~\ref{fig:fif_c}-\ref{fig:fif_d} indicate a positive influence of that feature and the green bars indicate a negative influence of the feature on the individual getting classified to $\hat{Y}=1$.
+Similarly, all the red bars in Figures~\ref{fairness_fvgm_fig:fif_a}-\ref{fairness_fvgm_fig:fif_b}, ~\ref{fairness_fvgm_fig:fif_c}-\ref{fairness_fvgm_fig:fif_d} indicate a positive influence of that feature and the green bars indicate a negative influence of the feature on the individual getting classified to $\hat{Y}=1$.
 
-In Figures~\ref{fig:fif_c} and~\ref{fig:fif_f}, we illustrate the influence of different features on the disparate impact between the two groups `male' and `female'. 
+In Figures~\ref{fairness_fvgm_fig:fif_c} and~\ref{fairness_fvgm_fig:fif_f}, we illustrate the influence of different features on the disparate impact between the two groups `male' and `female'. 
 The green bar indicates that removing a feature causes an increase in DI, i.e.
@@
 
-Thus, from Figure~\ref{fig:fif_c}, we conclude that `age' is the most bias-inducing feature among the two groups `male' and `female'.
-From Fig~\ref{fig:fif_a} and~\ref{fig:fif_b}, we observe that `age' is a decisive feature for `male' while it is comparatively insignificant for `female'.
+Thus, from Figure~\ref{fairness_fvgm_fig:fif_c}, we conclude that `age' is the most bias-inducing feature among the two groups `male' and `female'.
+From Fig~\ref{fairness_fvgm_fig:fif_a} and~\ref{fairness_fvgm_fig:fif_b}, we observe that `age' is a decisive feature for `male' while it is comparatively insignificant for `female'.
 
@@
 It is also the most biased-inducing feature as it varies most disparately among the two groups as the `capital-gain' differs significantly between two groups.
-\begin{figure}[t!]
+\begin{figure}
 	\begin{center}
 
-		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_0}\label{fig:fif_a}}
-		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_1}\label{fig:fif_b}}\\
-		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_0}\label{fig:fif_d}}
-		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_1}\label{fig:fif_e}}\\
-		\subfloat[COMPAS]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_compas_lr_sex_DI}\label{fig:fif_c}}
-		\subfloat[Adult]{\includegraphics[scale=0.24]{figures/dependency_exp_Learn-dependency_adult_lr_sex_DI}\label{fig:fif_f}}\\
+		\subfloat[COMPAS]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_0}\label{fairness_fvgm_fig:fif_a}}
+		\subfloat[COMPAS]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_1}\label{fairness_fvgm_fig:fif_b}}\\
+		\subfloat[Adult]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_adult_lr_sex_0}\label{fairness_fvgm_fig:fif_d}}
+		\subfloat[Adult]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_adult_lr_sex_1}\label{fairness_fvgm_fig:fif_e}}\\
+		\subfloat[COMPAS]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_compas_lr_sex_DI}\label{fairness_fvgm_fig:fif_c}}
+		\subfloat[Adult]{\includegraphics[scale=0.35]{figures/fairness/fvgm/dependency_exp_Learn-dependency_adult_lr_sex_DI}\label{fairness_fvgm_fig:fif_f}}\\
 
 	\end{center}
-	\caption{Extended results on computing feature influence functions (FIF) for COMPAS  and Adult dataset.}\label{fig:fif}
+	\caption{Extended results on computing feature influence functions (FIF) for COMPAS  and Adult dataset.}\label{fairness_fvgm_fig:fif}
 \end{figure}

### 2023-02-13 tables/fairness/fvgm/dag_structure_complexity.tex
fvgm appendix

\begin{figure}
		\begin{center}

	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/cactus_dag_LR_time}}
	\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/cactus_dag_LR_time_notears}}

		\end{center}

		\caption{Effect of number of variables in the learned Bayesian Network on computation time of {\fvgm}. In both plots, we vary $ \frac{|\mathbf{V}|}{|\nonsensitive \cup \sensitive|} $, that is the ratio between the number of variables in the Bayesian Network to the number of features. We observe that as this ratio increases to $ 1 $, both runtime of {\fvgm} (left plot) and network learning time (right plot) increase. } 
		\label{fairness_fvgm_fig:results_DAG_complexity}
\end{figure}

### 2023-02-13 tables/fairness/fvgm/fairness_metrics_on_compound_sensitive_groups.tex
fvgm appendix

@@
-\begin{figure*}[!t]
+\begin{figure}
 	\begin{center}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Disparate_impact_Adult_LR_Learn-efficient-dependency}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Statistical_parity_Adult_LR_Learn-efficient-dependency}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute_eqo_Adult_LR_Learn-efficient-dependency}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__PCF_Adult_LR_Learn-efficient-dependency}}	
-		\\
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__Disparate_impact_Adult_LR_Learn-efficient-dependency}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__Statistical_parity_Adult_LR_Learn-efficient-dependency}}
+		\vspace{-1em}
 
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Disparate_impact_Compas_LR_Learn-efficient-dependency}}			
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__Statistical_parity_Compas_LR_Learn-efficient-dependency}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute_eqo_Compas_LR_Learn-efficient-dependency}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sensitive_attribute__PCF_Compas_LR_Learn-efficient-dependency}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute_eqo_Adult_LR_Learn-efficient-dependency}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__PCF_Adult_LR_Learn-efficient-dependency}}	
+		\vspace{-1em}
+
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__Disparate_impact_Compas_LR_Learn-efficient-dependency}}			
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__Statistical_parity_Compas_LR_Learn-efficient-dependency}}
+		\vspace{-1em}
+
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute_eqo_Compas_LR_Learn-efficient-dependency}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sensitive_attribute__PCF_Compas_LR_Learn-efficient-dependency}}
 
@@
 }
-		\label{fig:multiple_metrics}
+		\label{fairness_fvgm_fig:multiple_metrics}
 
-\end{figure*}
+\end{figure}

### 2023-02-13 tables/fairness/fvgm/sanity_check.tex
fvgm appendix

@@
 	\begin{center}
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_probability}\label{fig:synthetic_dist}}
-		\subfloat[]{\includegraphics[scale=0.45]{figures/synthetic_example_BN}\label{fig:synthetic_bn}}
+		\subfloat[]{\includegraphics[scale=0.5]{figures/fairness/fvgm/synthetic_example_BN}\label{fairness_fvgm_fig:synthetic_bn}}
 
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_LR}}
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_SVM}}
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_DI_DT}}
-%		
-%		
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_LR}}
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_SVM}}
-%		\subfloat[]{\includegraphics[scale=0.25]{figures/sanity_SPD_DT}}
-%		
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_LR}\label{fig:synthetic_ppv_max_LR}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_SVM}}\\
-%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_max_PPV_DT}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_LR}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_SVM}\label{fig:synthetic_ppv_min_SVM}}
-%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_ppv_min_PPV_DT}\label{fig:synthetic_ppv_b}}
+		\subfloat[]{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_ppv_max_PPV_LR}\label{fairness_fvgm_fig:synthetic_ppv_max_LR}}
+		\subfloat[]{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_ppv_min_PPV_LR}\label{fairness_fvgm_fig:synthetic_ppv_min_LR}}
 
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_LR}\label{fig:synthetic_sample_size_LR}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_SVM}}\\
-%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_DI_DT}}	
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_LR}}
-		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_SVM}\label{fig:synthetic_sample_size_SVM}}
-%		\subfloat[]{\includegraphics[scale=0.33]{figures/sanity_sample_size_SPD_DT}\label{fig:synthetic_sample_size_b}}
+		\subfloat[]{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_sample_size_DI_LR}\label{fairness_fvgm_fig:synthetic_sample_size_LR_DI}}
+		\subfloat[]{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_sample_size_SPD_LR}\label{fairness_fvgm_fig:synthetic_sample_size_LR_SPD}}
+	\end{center}
+
+	\caption{
+		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on logistic regression classifier.}
+	\label{fairness_fvgm_fig:synthetic_results}
+\end{figure}
+
+\begin{figure}
+	\begin{center}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_ppv_max_PPV_SVM}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_ppv_min_PPV_SVM}}\\
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_sample_size_DI_SVM}}
+		\subfloat{\includegraphics[scale=0.4]{figures/fairness/fvgm/sanity_sample_size_SPD_SVM}}
 
@@
 	\caption{
-		Measuring accuracy of different fairness verifiers for Example~\ref{example:intro}. LR, SVM, DT refers to Logistic Regression classifier, Support Vector Machine, and Decision Tree, respectively.}
-	\label{fig:synthetic_results}
+		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on SVM classifier.}
+	\label{fairness_fvgm_fig:synthetic_results_SVM}
 \end{figure}

### 2023-02-13 chapters/interpretability/imli/appendix.tex
added appendix
imli new results

@@
-\section*{Appendix}
+\chapter{Interpretable Classification Rules}
 
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.38]{figures/interpretability/imli/thesis_exp/thesis_exp_cactus_train_val_fit_time}
+	\includegraphics[scale=0.38]{figures/interpretability/imli/thesis_exp/thesis_exp_cactus_predicate_count}
+	\includegraphics[scale=0.38]{figures/interpretability/imli/thesis_exp/thesis_exp_cactus_test_accuracy}
+	\includegraphics[scale=0.38]{figures/interpretability/imli/thesis_exp/thesis_exp_cactus_train_val_accuracy}
+\end{figure}
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Parkinsons_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_WDBC_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Pima_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Titanic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Magic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Tom's_HW_cactus_train_val_fit_time}		\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Credit_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Adult_cactus_train_val_fit_time}
+\end{figure}
+
+\section{Representative Interpretable Classification Rules in Datasets}
 In the following, we present representative CNF classifiers learned in different datasets. In each dataset, if an input satisfies the CNF formula, it is predicted class $ 1 $ and vice-versa.

### 2023-02-13 chapters/acknowledgements.tex
ack todo

@@
 
-With utmost humility, I would like to take this opportunity to express my sincere gratitude to everyone who has supported me throughout my PhD journey.
+\todo{Incomplete}With utmost humility, I would like to take this opportunity to express my sincere gratitude to everyone who has supported me throughout my PhD journey.
 
@@
 
-I have also been blessed to work with exceptional collaborators, including Debabrota Basu, Dmitry Malioutov, Daniel Neider, Naheed Anjum Araafat, Jonathan Scarlet, and Lorenzo.
+I have also been blessed to work with exceptional collaborators, including Debabrota Basu, Dmitry Malioutov, Daniel Neider, Naheed Anjum Arafat, Jonathan Scarlet, and Lorenzo.
 They have generously shared their knowledge and experience with me and have helped me broaden my perspectives.

### 2023-02-14 tables/fairness/fvgm/sanity_check.tex
short captions

@@
 
-	\caption{
+	\caption[Accuracy of {\fvgm} on logistic regression classifier]{
 		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on logistic regression classifier.}
@@
 
-	\caption{
+	\caption[Accuracy of {\fvgm} on SVM classifier]{
 		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on SVM classifier.}

### 2023-02-15 chapters/conclusion.tex
conclusion modified

--- 2022-07-17 chapters/conclusion.tex
@@
-\chapter{Conclusion And Future Work} In the past decade, the advancement of machine learning has witnessed a host of  applications in different safety-critical domains.
-In such domains, the interpretability, fairness, robustness, and privacy etc.\ of classifiers are considered important in trustworthy and responsible AI (artificial intelligence).
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, where we prioritize on improving the scalability and accuracy of the underlying problems.
-Relying on formal methods, our contributions are summarized as follows: (i) in interpretable machine learning, we design incremental learning technique for interpretable rule-based classifiers, (ii) in fairness in machine learning, we develop a formal fairness verification framework to verify multiple fairness definitions, algorithms, and classifiers; and we develop techniques to identify the sources of unfairness of classifiers by computing fairness influence functions.
+\chapter{Conclusion And Future Work}
+\label{chapter:conclusion} 
+Over the past decade, machine learning has been applied to various safety-critical domains, and it's crucial for classifiers to be interpretable, fair, robust, and private to ensure trustworthy and responsible AI.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning and aim to improve the scalability and accuracy of the underlying problems.
+We utilize formal methods to make the following contributions: (i) In interpretable machine learning, we design an incremental learning technique for interpretable rule-based classifiers of varied expressiveness.
+(ii) In fairness in machine learning, we develop a formal probabilistic fairness verification framework that can verify multiple fairness definitions of classifiers.
+Additionally, we develop techniques to explain fairness metrics by identifying feature combinations responsible for the bias of the classifier.
 
-To validate the efficiency of our methods, we develop open-source tools and conduct experiments on real-world interpretability and fairness datasets in machine learning.
-In interpretable machine learning, our framework {\imli} scales classification to million-size datasets while achieving competitive prediction accuracy and rule-size compared to existing rule-based classifiers.
-In our quest for learning more expressible interpretable classifiers, we develop another learning framework {\crr} based on incremental learning.
-In experiments, {\crr} learns more concise and accurate rule-based classifiers while demonstrating scalability owing to incremental learning.
+To demonstrate the efficacy of our methods, we have constructed open-source tools and conducted experiments on real-world datasets in machine learning to evaluate interpretability and fairness.
+In the context of interpretable rule-based machine learning, we have developed an incremental learning framework, known as {\imli}, that scales classification to million-size datasets while maintaining competitive prediction accuracy and rule-size compared to existing rule-based classifiers.
+Additionally, in our pursuit of more expressive yet interpretable classifiers, we have introduced another learning framework, called {\crr}, for logical relaxed classification rules that are based on incremental learning.
+Experimental results show that {\crr} is capable of learning more concise and accurate rule-based classifiers.
 
-In fairness in machine learning, our probabilistic fairness verifier {\justicia} and {\fvgm} demonstrates a superior performance in accuracy and scalability than the state of the art verifiers.
-Being a stochastic-SAT-based verifier, {\justicia} verifies fairness of compound sensitive groups of Boolean classifiers such as decision trees with higher scalability.
-{\fvgm}, in addition, considers correlated features and verifies fairness of linear classifiers with better scalability and accuracy than existing verifiers.
-Finally, given a classifier and a dataset, our framework {\fairXplainer} computes fairness influences of each subset of non-sensitive features as their contribution to the incurred bias of the classifier.
-In experiments, {\fairXplainer} estimates bias with significantly higher accuracy than the existing local explanation approach, while enabling computation of intersectional influences as a mean of better understanding the sources fo bias.
+Our work on fairness in machine learning introduces two novel probabilistic fairness verifiers, {\justicia} and {\fvgm}, which exhibit superior performance in accuracy and scalability compared to state-of-the-art verifiers.
+{\justicia} is a stochastic-SAT-based verifier that enables scalable verification of fairness for compound sensitive groups of Boolean classifiers, such as decision trees, which was previously infeasible with existing methods.
+On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers.
+Additionally, we propose a global sensitivity analysis-based method, {\fairXplainer}, that explains group fairness metrics by computing fairness influences of individual and intersectional features.
+Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.
+
+In future, our research on fair and interpretable machine learning will focus on several areas.
+One immediate research question is to design a fairness improvement algorithm for machine learning classifiers using the fairness verifiers, {\justicia} and {\fvgm}, and the bias explanation framework, {\fairXplainer}.
+We also aim to extend formal fairness verification to more complex classifiers, such as random forests and neural networks.
+In addition, our future plans include deploying {\fairXplainer} on image classifiers, language models, and computer vision tasks.
+In a separate line of work, we will explore the application of incremental solving, as demonstrated in the context of interpretable classification learning by {\imli} and {\crr}, to other optimization problems, even beyond machine learning.
+To conclude the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the support of formal methods and automated reasoning.
+
+\begin{comment}
+	\begin{itemize}
+		\item Fairness repair
+		\item Incremental solving
+		\item Fairness and interpretability as a service to more complex models.
+	\end{itemize}
+\end{comment}

### 2023-02-16 chapters/interpretability/imli/appendix.tex
imli appendix

@@
 \chapter{Interpretable Classification Rules}
+
+\section{Experimental Comparison between Incremental vs.\ Non-incremental  Encoding}
+We compare the performance of two encoding techniques, na\"ive non-incremental encoding and efficient incremental encoding in {\imli}, for learning interpretable CNF classification rules in datasets of various sizes.
+We demonstrate the scalability results using a cactus plot in Figure~\ref{fig:interpretability_imli_additional_fig:scalability_non_incremental_vs_incremental}.
+Each point $ (x, y) $ represents the ability to solve $ x $ classification instances within $ y $ seconds.
+We test both incremental and non-incremental encoding using $ 360 $ instances for each dataset with varying hyper-parameters.
+As the dataset size increases, the non-incremental encoding times out, failing to solve all $ 360 $ instances within $ 1000 $ seconds in the Credit and Adult datasets.
+In contrast, the incremental encoding in {\imli} demonstrates higher scalability, solving all $ 360 $ instances in less than $ 10 $ seconds in the Titanic dataset, and efficiently solving all $ 360 $ instances within the timeout in the Credit and Adult datasets.
+Hence, the incremental encoding is superior in terms of scalability compared to the non-incremental encoding.
+
+In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental},
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Parkinsons_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_WDBC_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Pima_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Titanic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Magic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Tom's_HW_cactus_train_val_fit_time}		\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Credit_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Adult_cactus_train_val_fit_time}
+	\caption[Scalability: incremental vs.\ non-incremental encoding]{Detailed scalability results of incremental vs.\ non-incremental encoding in {\imli}, presented as a cactus plot.
+As datasets become large, the non-incremental encoding suffers from poor scalability by witnessing time-outs more frequently than the incremental encoding {\imli}.}
+	\label{fig:interpretability_imli_additional_fig:scalability_non_incremental_vs_incremental}
+\end{figure}
 
@@
 	\includegraphics[scale=0.38]{figures/interpretability/imli/thesis_exp/thesis_exp_cactus_train_val_accuracy}
+
+	\caption[Performance comparison between non-incremental vs.\ incremental encoding]{Comparison of training time, rule-size, and test and train accuracy between non-incremental vs.\ incremental encoding in {\imli}.
+In $ X $-axis, the dimension of the dataset (\#samples $ \mid $  \#features) is shown below dataset name.
+In large datasets, non-incremental encoding often times out and learns zero-size classification rules yielding less test and train accuracy than the incremental encoding.}
+	\label{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental}
 \end{figure}
 
-\begin{figure}
-	\centering
-	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Parkinsons_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_WDBC_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Pima_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Titanic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Magic_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Tom's_HW_cactus_train_val_fit_time}		\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Credit_cactus_train_val_fit_time}	\includegraphics[scale=0.27]{figures/interpretability/imli/thesis_exp/thesis_exp_Adult_cactus_train_val_fit_time}
-\end{figure}
-
-\section{Representative Interpretable Classification Rules in Datasets}
+\section{Representative Interpretable Classifiers}
 In the following, we present representative CNF classifiers learned in different datasets. In each dataset, if an input satisfies the CNF formula, it is predicted class $ 1 $ and vice-versa.

### 2023-02-17 chapters/abstract.tex
abstract pass

@@
 Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{explaining the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{interpreting the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
 To alleviate these limitations, we apply formal methods in interpretable and fair machine learning and provide scalable and accurate solutions to the underlying problems.
@@
 As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
-To this end, we propose an incremental learning framework, called {\imli}, which extends the classification problem to million-size datasets through an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning.
-Furthermore, rule-based classifiers suffer from limited expressiveness due to being confined to propositional logic.
-To learn more expressible yet interpretable classification rules, we propose a relaxation of rule-based classifiers based on logical formulas and  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming).
+To this end, we propose an incremental learning framework, called {\imli}, which applies an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning and enables classification to million-size datasets.
+Although being interpretable, rule-based classifiers often suffer from limited expressiveness, for example,  classifiers based on propositional logic.
+To learn more expressible yet interpretable classification rules, we propose a relaxation of classifiers based on logical formulas.
+For learning relaxed rule-based classifiers, we propose  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming).
 {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.
@@
 	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers.
-In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, based on SSAT (stochastic satisfiability) with the goal of formally verifying the bias of a classifier given the probability distribution of features.
-Additionally, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate and scalable verification of fairness.
+In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features.
+Building on SSAT (stochastic satisfiability), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features.
+For a more accurate fairness verification, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, which results in an accurate  verification of fairness.
 Fairness metrics globally quantify bias, but do not detect or explain its sources.

### 2023-02-17 chapters/abstract.tex
abstruct: another pass

@@
 Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{interpreting the sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
-To alleviate these limitations, we apply formal methods in interpretable and fair machine learning and provide scalable and accurate solutions to the underlying problems.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{explaining sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
+To alleviate these limitations, we integrate formal methods and automated reasoning with interpretability and fairness in machine learning and provide scalable and accurate solutions to the underlying problems.
 
@@
 As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
-To this end, we propose an incremental learning framework, called {\imli}, which applies an iterative solving of MaxSAT (maximum satisfiability) queries in mini-batch learning and enables classification to million-size datasets.
+To this end, we propose an incremental learning framework, called {\imli}, which applies an iterative solving of  maximum satisfiability (MaxSAT) queries in mini-batch learning and enables classification on million-size datasets.
 Although being interpretable, rule-based classifiers often suffer from limited expressiveness, for example,  classifiers based on propositional logic.
 To learn more expressible yet interpretable classification rules, we propose a relaxation of classifiers based on logical formulas.
-For learning relaxed rule-based classifiers, we propose  an efficient learning framework, called {\crr},  building on incremental learning and MILP (mixed integer linear programming).
+For learning relaxed rule-based classifiers, we propose  an efficient learning framework, called {\crr},  building on incremental learning and mixed integer linear programming (MILP).
 {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.
@@
 In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features.
-Building on SSAT (stochastic satisfiability), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features.
-For a more accurate fairness verification, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, which results in an accurate  verification of fairness.
+Building on  stochastic satisfiability (SSAT), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features.
+For a more accurate fairness verification, we extend {\justicia} to consider feature correlations represented as a Bayesian Network, resulting in an accurate  verification of fairness.
+
 	Fairness metrics globally quantify bias, but do not detect or explain its sources.

### 2023-02-17 chapters/fairness/fif/experiments.tex
connecting chapters

@@
 To estimate FIFs, we leverage and modify the `HDMR' module in SALib library~\cite{Herman2017} based on global sensitivity analysis.
-In experiments, we consider four widely studied datasets from fairness literature, namely German-credit~\cite{DK2017},
+In experiments, we consider four widely studied datasets from fairness literature, namely German-credit~\cite{Dua:2019},
 % Ricci~\cite{mcginley2010ricci}, 
-Titanic (\url{https://www.kaggle.com/c/titanic}), COMPAS~\cite{angwin2016machine}, and Adult~\cite{DK2017uci}.
+Titanic (\url{https://www.kaggle.com/c/titanic}), COMPAS~\cite{angwin2016machine}, and Adult dataset~\cite{Dua:2019}.
 We deploy Scikit-learn~\cite{scikit-learn} to learn different classifiers: logistic regression classifier, support vector machine (SVM), neural network, and decision tree with $ 5 $-fold cross-validation.

### 2023-02-17 chapters/fairness/fvgm/introduction.tex
updated upto fvgm
connecting chapters

@@
 \label{chapter:fvgm}
+\begin{comment}
 	The significant improvement of machine learning over the decades has led to a host of applications of machine learning in  high-stake decision-making such as college admission~\cite{martinez2021using}, hiring of employees~\cite{ajunwa2016hiring}, and recidivism prediction~\cite{tollenaar2013method,dressel2018accuracy}.
@@
 
-	In order to solve the fairness-verification problem, existing works have proposed two principled approaches.	Firstly, \cite{ghosh2020justicia} and \cite{albarghouthi2017fairsquare} propose formal methods that reduce the problem into a solution of an SSAT or an SMT~ formula respectively.	Secondly, \cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
-Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.
-sensitive features and ignore correlations among non-sensitive features.
-As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remains a major challenge.
+\end{comment}
 
-	In this work, we seek to remedy the aforementioned situation.
-As a first step, we focus on  \textit{linear classifiers}, which has attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
-At this point, it is worth highlighting that our empirical evaluation demonstrates that the existing techniques fail to scale beyond small examples or provide highly inaccurate estimates for comparatively {\em small} linear classifiers.
+	In this chapter, we extend the probabilistic fairness verification problem to a more practical scenario by accounting for the correlation of features, which is prevalent in most classification datasets.
+Existing fairness verifiers \cite{albarghouthi2017fairsquare,bastani2019probabilistic}, including Justicia \cite{ghosh2020justicia} in Chapter~\ref{chapter:justicia}, suffer from poor accuracy due to considering specific input distribution.
+For example, they assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features with respect to sensitive features, and ignore correlations among non-sensitive features.
+Thus, it is desirable to design a fairness verification framework where the input is a probability distribution containing feature correlations.
 
-%	In this paper, we argue that the above solution is not practical in verifying linear classifiers for two reasons: (i) the encoding of a linear classifier into an SSAT or an SMT formula or using a sampling approach suffer from scalability and (ii) ignoring correlations among all features results in an inaccurate estimate of fairness metrics.
+	Towards this goal, we particularly focus on the verification of \textit{linear classifiers} as our first step because there has been significant attention on linear classifiers in the context of fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+In this context, existing approaches suffer from limited scalability while verifying linear classifiers.
+This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}.
+For example, Justicia applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.
 
-%	Given multiple fairness definitions, fairness algorithms either \textit{preprocess} the training data~\cite{calmon2017optimized,kamiran2012data,zemel2013learning}, \textit{in-process} the classifier during training~\cite{zhang2018mitigating}, or \textit{post-process} the prediction of the trained classifiers~\cite{hardt2016equality,kamiran2012decision} to improve fairness.
-In contrast, fairness-attack algorithms try to manipulate the data or the classifier to worsen its fairness~\cite{solans2020poisoning}.
+	Considering these two aspects, our goal in this chapter is to design a fairness verifier particularly tailored for linear classifiers that addresses both the scalability and accuracy challenges of existing verifiers.
 
-%	\red{Among different ML classifiers, researcher This observation motivates us in asking the following research question: \textit{Can we design an efficient fairness verification framework for linear classifiers that can verify different fairness definitions and algorithms accurately?}
-
-%	
-%	Considering the surge of research in ML fairness based on linear classifiers, we focus on designing a fairness verifier particularly tailored for linear classifiers.
-We further show novel applications of the verifier in practical fairness tasks, such as detecting fairness attacks and computing the source of unfairness in the level of input features.
-%
-
-	\textbf{Our Contributions.} The contributions of this chapter are summarized below.
+	\paragraph{Our Contributions.} The contributions of this chapter are summarized below.
 

### 2023-02-17 chapters/fairness/justicia/experiments.tex
connecting chapters

@@
 In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner.
-Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{PVGMTGB2011} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}.
+Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{scikit-learn} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}.
 We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula.
@@
 \subsubsection{Robustness: Stability to Sample Size} 
-We have compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics.
+We have empirically compared the robustness of our probabilistic fairness verifier {\justicia} with dataset-centric verifier AIF360 by varying the sample-size and reporting the standard deviation of different fairness metrics.
 In Figure~\ref{fairness_justicia_fig:sample-size}, AIF360 shows higher standard deviation for lower sample-size and the value decreases as  the sample-size increases. 

### 2023-02-17 chapters/fairness/justicia/introduction.tex
connecting chapters

@@
 SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}.
-In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
+In this chapter, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
 SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.

### 2023-02-17 chapters/fairness/justicia/practical.tex
connecting chapters

@@
 \subsubsection{Extension to Continuous Features}
-In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features.
-In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.
+In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM2020}, are generally trained on a Boolean abstraction of input features.
+In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM2020}.
 

### 2023-02-17 chapters/interpretability/imli/introduction.tex
connecting chapters

@@
-%\section{Introduction}
-
-%The last decade has witnessed significant progress in machine learning (ML). 
 \label{chapter:imli}
-Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}.
+%Machine learning models are presently deployed in safety-critical and high-stake decision-making arising in medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, and transportation~\cite{peled2019model,zantalis2019review}.
 In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}.
-In this chapter, we focus on learning an interpretable machine learning classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box models.
-
-%Machine learning techniques are now being deployed in critical decision making, for example, hiring employees, giving a loan to people, predicting recidivism, predicting the likelihood of stroke of patients, etc.
-In the classical setting of classification problems, making an accurate prediction has been the objective of several prediction systems e.g., ad predictions, movie recommendations, etc.
-at the cost of making such systems essentially black-box.
-However, there are other safety-critical domains e.g., legal, medical, law, and so on, where interpretability, privacy, and fairness of the prediction system are of paramount importance.
-In this study, we show a principled approach to generate accurate and interpretable decision functions expressible in propositional logic.
-
-In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
+In this chapter, we learn an interpretable rule-based classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box classifiers. 
+%In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
 Such models are used to either learn an interpretable model from the start or as proxies to provide post-hoc explanations of pre-trained black-box models~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}.
 In our context of rule-based classifiers, we use the sparsity of rules as a measure of interpretability, which is adopted in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
-
-In this chapter, we study decision functions expressible in \emph{propositional logic}.
-In particular, we focus on learning logical formulas from data as classifiers and define interpretability in terms of the number of Boolean literals that the formula contains.
-In the following, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.
+In particular, we study classifiers expressed as Boolean formulas, wherein we define interpretability in terms of the number of Boolean literals that the formula contains.
+Here, we illustrate an interpretable classifier that decides if a tumor cell is malignant or benign based on different features of tumors.
 
@@
 \begin{example}
-	\normalfont
-	We learn an interpretable classification rule in CNF for predicting the classification of a tumor cell.
+	We illustrate an interpretable classification rule in CNF for classifying a tumor cell.
 We consider WDBC (Wisconsin Diagnostic Breast Cancer) dataset~\cite{agarap2018breast} to learn a CNF formula with two \emph{clauses} and four Boolean \emph{literals}.

### 2023-02-17 chapters/interpretability/relaxed-cnf/experiment.tex
formatting
highlevel pass
connecting chapters

@@
 We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g.
-$ \IMLI  $~\cite{GM2019}, RIPPER~\cite{C1995}, BRS~\cite{WRDLKM2017},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
+$ \IMLI  $~\cite{GM2019}, RIPPER~\cite{cohen1995fast}, BRS~\cite{wang2017bayesian},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
 Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers.
 In particular,
-{\IMLI} generates classification rules in CNF using a MaxSAT-based formulation and we use Open-WBO \cite{MML2014} as the MaxSAT solver for {\IMLI}.
+{\IMLI} generates classification rules in CNF using a MaxSAT-based formulation and we use Open-WBO \cite{martins2014open} as the MaxSAT solver for {\IMLI}.
 We compare with propositional rule learning algorithm RIPPER, which is implemented in WEKA~\cite{HFHPRW2009} and generates classification rules in the form of decision lists.
 BRS is a Bayesian framework for generating rule sets expressed as DNF.
-For  other classifiers, we use the Scikit-learn module of Python~\cite{PVGMTGB2011}.
+For  other classifiers, we use the Scikit-learn module of Python~\cite{scikit-learn}.
 For all  classifiers, we set the training cut-off time to $ 1500 $  {seconds}.
@@
 	\label{interpretability_crr_tab:rule_based_classifiers}
+
 %	\setlength{\tabcolsep}{3pt} 
@@
 			& $   7 $    & $   19 $    & $   13 $    & $   \textbf{4.5}  $   \\ & & 
-			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   \textbf{1.8} \text{s}  $    & $   122.5 \text{s}  $   \\\\
+			& $   5.27 \text{s}  $    & $   25.07 \text{s}  $    & $   \textbf{1.8} \text{s}  $    & $   122.5 \text{s}  $   \\
+			\midrule 
 			{ Ionosphere}   & $  351 $  & $  144 $  & $   88.65  $    & $   \textbf{91.43}  $    & $   89.29  $    & $   \textbf{91.43}  $   \\ & & 
 			& $   8 $    & $   \textbf{4} $    & $   8.5  $    & $   20 $   \\ & & 
-			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   \textbf{2.09} \text{s}  $    & $   5.59 \text{s}  $   \\\\ 
+			& $   5.87 \text{s}  $    & $   75.53 \text{s}  $    & $   \textbf{2.09} \text{s}  $    & $   5.59 \text{s}  $   \\
+			\midrule 
 			{ WDBC}   & $  569 $  & $  88 $  & $   95.22  $    & $   \textbf{95.65}  $    & $   93.91  $    & $   94.69  $   \\ & & 
 			& $   7.5  $    & $   12 $    & $   \textbf{7} $    & $   34.5  $   \\ & & 
-			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   \textbf{1.38} \text{s}  $    & $   316.32 \text{s}  $   \\\\ 
+			& $   5.7 \text{s}  $    & $   630.23 \text{s}  $    & $   \textbf{1.38} \text{s}  $    & $   316.32 \text{s}  $   \\
+			\midrule 
 			{ Magic}   & $  19020 $  & $  79 $  & $   \textbf{84.04}  $    & $   74.15  $    & $   71.97  $    & $   81.31  $   \\ & & 
 			& $   115 $    & $   \textbf{3} $    & $   24 $    & $   31 $   \\ & & 
-			& $   \textbf{15.86} \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\\\ 
+			& $   \textbf{15.86} \text{s}  $    & $   56.46 \text{s}  $    & $   141.2 \text{s}  $    & $   1012.6 \text{s}  $   \\
+			\midrule 
 			{ Tom's HW}   & $  28179 $  & $  910 $  & $   \textbf{97.4}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   95.88  $    & $   97.34  $   \\ & & 
 			& $   36 $    &       & $   30 $    & $   \textbf{4} $   \\ & & 
-			& $   \textbf{42.73} \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\\\ 
+			& $   \textbf{42.73} \text{s}  $    &       & $   92.65 \text{s}  $    & $   1071.58 \text{s}  $   \\
+			\midrule 
 			{ Credit}   & $  30000 $  & $  110 $  & $   81.68  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   81.42  $    & $  \textbf{ 82.04}  $   \\ & & 
 			& $   38.5  $    &       & $   \textbf{10} $    & $   32 $   \\ & & 
-			& $   \textbf{14.52} \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\\\ 
+			& $   \textbf{14.52} \text{s}  $    &       & $   17.66 \text{s}  $    & $   1021.35 \text{s}  $   \\
+			\midrule 
 			{ Adult}   & $  32561 $  & $  144 $  & $   84.31  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.08  $    & $   \textbf{84.86}  $   \\ & & 
 			& $   94 $    &       & $   23 $    & $   \textbf{18} $   \\ & & 
-			& $   27.61 \text{s}  $    &       & $   \textbf{11.91} \text{s}  $    & $   1016.36 \text{s}  $   \\\\ 
+			& $   27.61 \text{s}  $    &       & $   \textbf{11.91} \text{s}  $    & $   1016.36 \text{s}  $   \\
+			\midrule 
 			{ Twitter}   & $  49999 $  & $  1511 $  & $   \textbf{95.74}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   94.24  $    & $   95.16  $   \\ & & 
 			& $   179.5  $    &       & $   57 $    & $   \textbf{12} $   \\ & & 
-			& $   \textbf{170.87} \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\\\ 
+			& $   \textbf{170.87} \text{s}  $    &       & $   238.29 \text{s}  $    & $   1144.66 \text{s}  $   \\
+			\midrule 
 			{ Weather-AUS}   & $  107696 $  & $  141 $  & $   \textbf{84.57}  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   82.83  $    & $   83.34  $   \\ & & 
 			& $   195 $    &       & $   7 $    & $   \textbf{2} $   \\ & & 
-			& $   \textbf{121.02} \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\\\ 
+			& $   \textbf{121.02} \text{s}  $    &       & $   366.12 \text{s}  $    & $   1115.27 \text{s}  $   \\
+			\midrule 
 			{ Skin}   & $  245057 $  & $  119 $  & $   98.32  $    &   \multicolumn{1}{c}{\multirow{3}{*}{\textemdash}}       & $   \textbf{98.92}  $    & $   95.08  $   \\ & & 
 			& $   725 $    &       & $   201 $    & $   \textbf{29} $   \\ & & 
-			& $   1313.19 \text{s}  $    &       & $   \textbf{103.8} \text{s}  $    & $   825.6 \text{s}  $   \\\\ 
+			& $   1313.19 \text{s}  $    &       & $   \textbf{103.8} \text{s}  $    & $   825.6 \text{s}  $   \\
 			\bottomrule
@@
 
-\begin{figure}[t]
+\begin{figure}
 	\centering

### 2023-02-17 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
with new formatting
connecting chapters

@@
-\section{{\crr}: Classification Rules in Relaxed Logical Form}
+\section{Classification Rules in Relaxed Logical Form}
 \label{interpretability_crr_sec:ilp}
-In this section, we describe the main contribution of our work, {\crr}, a  framework for learning   relaxed-CNF rules.
+In this section, we describe the main contribution of this chapter, {\crr}, a  framework for learning   relaxed-CNF rules.
 {\crr}  converts the learning problem into  an ILP-based formulation,  learns the optimal assignment of variables and constructs rule $ \Rule $ based on the assignment.

### 2023-02-17 chapters/interpretability/relaxed-cnf/introduction.tex
formatting
connecting chapters

@@
 \label{chapter:crr}
-The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
+%The widespread adoption of prediction systems in various safety-critical domains such as medical diagnosis, law, education, and many others has led to the increased importance of presenting the decision functions in interpretable representations~\cite{K2001,MVBB2005,SAD2015,S2014,TV2013}.
 To enable  safe, robust, and trustworthy integration of such systems, the end users require them to support interpretability, privacy, and fairness in decision-making~\cite{DF2018,VML2012,WRLKM2015,ZUR2017}.
@@
 A recent body of work has studied sparsity-inducing objectives for classification rules in CNF or DNF and demonstrated that they often achieve high interpretability\textemdash defined in terms of rule-sparsity\textemdash with a minimal sacrifice in classification accuracy ~\cite{GM2019,LKCL2019,MM18}.
-Although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
+
+In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing from chapter~\ref{chapter:imli},
+although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
 A Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
-In this work, we introduce a novel formulation of interpretable classification rules, namely \emph{relaxed-CNF}, which achieve benefits of  both worlds: it is interpretable similar to CNF/DNF  but more expressible by allowing cardinality constraints in the representation.
+In this chapter, we introduce a novel formulation of interpretable classification rules, namely \emph{relaxed-CNF}, which achieve benefits of  both worlds: it is interpretable similar to CNF/DNF  but more expressible by allowing cardinality constraints in the representation.
 
@@
 Recently, Emad et al.~\cite{EVM2015} have developed a semi-quantitative group testing approach for learning sparse single level $M$-of-$N$ rules, which are quite restrictive in their ability to fit the data.
-In contrast, in this work, we study a much richer family of two-level relaxed-CNF rules.
+In contrast, in this chapter, we study a much richer family of two-level relaxed-CNF rules.
 
@@
 	\begin{center}
-		\includegraphics[scale=.65]{figures/interpretability/relaxed-cnf/checklist.pdf}
+		\includegraphics[scale=.62]{figures/interpretability/relaxed-cnf/checklist.pdf}
 	\end{center}
@@
 More significantly, {\crr} generates relaxed-CNF rules with higher accuracy than CNF rules generated by~\cite{GM2019}.
-Furthermore, compared to decision lists generated by~\cite{C1995}, relaxed-CNF rules are sparser in large datasets.}
+Furthermore, compared to decision lists generated by~\cite{cohen1995fast}, relaxed-CNF rules are sparser in large datasets.}
 %can also reach the same level of accuracy compared to plain CNF/DNF rules with smaller relaxed-CNF rules. 

### 2023-02-17 chapters/introduction.tex
connecting chapters

@@
 In this thesis, we focus on the interpretability and fairness aspects of machine learning.
-The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
+The thesis establishes a close integration of formal methods and automated reasoning with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
 
@@
 \emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features.
-Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
+Interpretable rule-based classifiers not only interpret the decision function but also are applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
 In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
@@
 In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features.
-Finally, we explain the sources of bias based on fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
-To summarize our thesis on the interpretability and fairness in machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.
+Finally, we explain the sources of bias by formalizing and computing fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
+To summarize our thesis on interpretability and fairness in machine learning, we prioritize on improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.
 
@@
 In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc.
-In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}.
+In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}.
 Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
 The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets.
-Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP (mixed integer linear programming) to efficiently learn rule-based classifiers in a mini-batch learning setting.
+Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP to efficiently learn rule-based classifiers in a mini-batch learning setting.
 Here we summarize the contributions of the thesis on interpretable rule-based machine learning.
@@
 We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for learning interpretable classification rules expressed in CNF.
-$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query.
+$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and sparsity of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query.
 Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to real-world classification datasets of thousands to millions of samples.
@@
 \subsection*{Expressiveness via Logical Relaxation}
-We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}.
+We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness.
 Elaborately, we consider relaxed definitions of standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule.
@@
 
-In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers.
+In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
 Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
 {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.
-Because SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby extending the verification problem to multiple sensitive features unlike earlier methods.
+SSAT encoding, by construction, allows separate quantification to each sensitive feature without any restriction on the number of features, thereby it is natural in SSAT-based formulation to extend the verification problem to multiple sensitive features unlike earlier methods.
 In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic} because of the efficient SSAT encoding.
@@
 
-\subsection*{Explaining Fairness metrics: Identifying the Sources of Bias}
+\subsection*{Explaining Fairness Metrics: Identifying Sources of Bias}
 Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
@@
 Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
-In this thesis, \textit{we aim to design a global bias attribution framework and a corresponding algorithm that can quantify both the individual and intersectional influences of features leading to granular and functional explanation of the sources of bias.} Building on global variance decomposition, we propose an algorithm {\fairXplainer} to estimate the fairness influence function (FIF) of features.
+In our contribution, we propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features.
 In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods.
@@
 		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
-		\item \red{FairXplainer for explaining group-based fairness metrics in machine learning.}
+		\item \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for explaining group-based fairness metrics in machine learning.
 	\end{itemize}

### 2023-02-19 chapters/fairness/fif/conclusion.tex
connecting different chapters

@@
 The results instantiate {\fairXplainer} as a global, granular, and more accurate explanation method to understand the sources of bias.
-Additionally, the resonance between the rules extracted by CORELS~\cite{rudin19stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier.
+Additionally, the resonance between the rules extracted by CORELS~\cite{rudin2019stop} and the most influential features detected by {\fairXplainer} indicates that FIFs can be exploited in future to create an explainable proxy of a biased/unbiased classifier.
 Also, we aim to extend {\fairXplainer} to compute FIFs for complex data, such as image and text, and design algorithms leveraging FIFs to yield unbiased decisions.

### 2023-02-19 chapters/fairness/fif/experiments.tex
connecting different chapters

@@
 In contrast to {\fairXplainer}, SHAP only estimates individual FIFs  (Figure~\ref{fairness_fairXplainer_fig:fif_shap}) and approximates statistical parity with higher error than {\fairXplainer}.
-Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin19stop}.
-From Figure 3 in~\cite{rudin19stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively.
-While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin19stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features.
+Interestingly, {\fairXplainer} yields FIF estimates of the classifier trained on COMPAS dataset that significantly matches with the rule-based classifier extracted from the COMPAS dataset using Certifiably Optimal Rule Lists (CORELS) algorithm~\cite[Figure 3]{rudin2019stop}.
+From Figure 3 in~\cite{rudin2019stop}, we observe that prior count is the only feature used individually to predict arrests, while age is paired with sex and prior counts respectively.
+While considering second-order intersectionality, {\fairXplainer} ($ \lambda = 2 $) yields priors count, (sex, age), and (age, priors count) as \emph{three of the top four features} that explains the observation of~\cite{rudin2019stop} better than SHAP and {\fairXplainer} ($ \lambda = 1 $) considering only individual features.
 \emph{Therefore, {\fairXplainer} demonstrates a clearer understanding on the sources of bias of a classifier by simultaneously quantifying on intersectional influences and individual influences }.

### 2023-02-19 chapters/fairness/fif/introduction.tex
connecting different chapters

@@
 
-The past few decades have witnessed a remarkable advancement in the field of machine learning, with its applications encompassing crucial decision-making processes such as college admissions~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring}, and more.
+%The past few decades have witnessed a remarkable advancement in the field of machine learning, with its applications encompassing crucial decision-making processes such as college admissions~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring}, and more.
 In such applications, the deployed machine learning classifier often demonstrates bias towards certain demographic groups present in the data~\cite{dwork2012fairness}.
@@
 
-Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+In this chapter, we envision towards combining two research themes in the thesis: interpretability and fairness, and discuss an algorithmic framework to interpret or explain fairness in machine learning.
+As demonstrated in Chapter~\ref{chapter:justicia} and~\ref{chapter:fvgm}, fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
 In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
-Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+Motivated by the GDPR's ``right to explanation'', research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
 In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
@@
 \input{chapters/fairness/fif/figure_intro}
-We illustrate the usefulness of our contributions via an example scenario proposed in\cite{ghosh2020justicia}.
+We illustrate the usefulness of our contributions via an example scenario proposed in Example~\ref{fairness_justicia_example:intro}.
 
@@
 \begin{example}
-	\label{fairness_fairXplainer_ex:motivating_example}  Following\cite{ghosh2020justicia}, we consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features: fitness  and income.
+	\label{fairness_fairXplainer_ex:motivating_example} We consider a classifier that decides an individual's eligibility for health insurance based on non-sensitive features: fitness  and income.
 Fitness and income depend on a sensitive feature age $ \in $ \{young, elderly\} leading to two sensitive groups, as highlighted in (Figure~\ref{fairness_fairXplainer_fig:dag_age_income_fitness}--\ref{fairness_fairXplainer_fig:distribution_example}).

### 2023-02-19 chapters/fairness/fvgm/conclusion.tex
connecting different chapters

--- 2022-06-02 chapters/fairness/fvgm/conclusion.tex
@@
 We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions.
-As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}.
-%We also investigate a principled approach to detect the sources of bias by computing fairness influence functions (FIF).
-Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other ML models.
-%Though the proposed solution with a Bayesian network has exponential worst-case complexity, it would be interesting to develop a pseudo-polynomial approximation in the future.
+In the next chapter, we elaborate on fairness influence functions as a framework to explaining group fairness metrics in machine learning.
+\begin{comment}
+	%As a future work, we aim to design fairness-enhancing algorithms certified by fairness verifiers, such as {\fvgm}.
+Since {\fvgm} serves as an accurate and scalable fairness verifier for linear classifiers, it will be interesting to design such verifiers for other machine learning models.
+\end{comment}

### 2023-02-19 chapters/fairness/fvgm/introduction.tex
connecting different chapters

@@
 
-	Towards this goal, we particularly focus on the verification of \textit{linear classifiers} as our first step because there has been significant attention on linear classifiers in the context of fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
 In this context, existing approaches suffer from limited scalability while verifying linear classifiers.
 This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}.
-For example, Justicia applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.
+For example, {\justicia} applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.
 

### 2023-02-19 chapters/fairness/justicia/discussion.tex
highlevel pass
connecting different chapters

@@
 Our work opens up several new directions of research.
-One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs.
+One direction is to develop SSAT models and verifiers for popular classifiers like deep networks and SVMs.
 Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
+Finally, an interesting direction is to enable {\justicia} to consider correlated feature distribution, which we pursue in the next chapter.
 %Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2023-02-19 chapters/fairness/justicia/practical.tex
connecting different chapters

@@
 In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM2020}, are generally trained on a Boolean abstraction of input features.
-In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM2020}.
+In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{lakkaraju2019faithful,GMM2020}.
 

### 2023-02-19 chapters/interpretability/imli/appendix.tex
connecting different chapters

@@
 
-In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental},
+In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental}, we compare both non-incremental and incremental encoding in terms of median training time, rule-size, test accuracy, and train accuracy over ten cross validation folds.
+In all datasets, the incremental encoding demonstrates higher efficiency than the non-incremental encoding by requiring less training time.
+In large datasets, such as Magic, Credit, and Adult, the non-incremental encoding always times out, thereby producing zero-size rules because of failing to solve the MaxSAT query.
+As a result, the accuracy of the non-incremental encoding is also less than the incremental encoding in large datasets.
+In small datasets, the incremental encoding demonstrates a competitive accuracy and rule-size with the non-incremental encoding.
+Therefore, the incremental encoding in {\imli} demonstrates its effectiveness in training time, accuracy, and rule-size over the non-incremental encoding.
 

### 2023-02-19 chapters/interpretability/imli/conclusion.tex
all papers added
minor edits in many parts of the thesis
connecting different chapters

@@
-\section{Conclusion}
+\section{Chapter Summary}
 \label{interpretability_imli_sec:conclusion}
@@
 Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.
-In this paper, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas.
+In this chapter, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas.
 {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.
@@
 Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework.
+An interesting direction is to leverage the incremental learning in {\imli} to learn more expressible yet interpretable classification rules, which we explore in the next chapter.
 

### 2023-02-19 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
connecting different chapters

@@
 The ILP query $ Q $ has binary integer constraints and the solution of this integer program is computationally intractable. 
-A common approach that efficiently finds an approximate solution to such a problem extends to relax the integer constraints,  solves the LP-relaxed (linear programming relaxation) problem, and then rounds the non-integer variables to get an integer solution as in~\cite{MV2013}.
+A common approach that efficiently finds an approximate solution to such a problem extends to relax the integer constraints,  solves the LP-relaxed (linear programming relaxation) problem, and then rounds the non-integer variables to get an integer solution as in~\cite{malioutov2013exact}.
 In our case, we cannot relax all integer constraints because it may violate the structure of relaxed-CNF rules.
@@
 
-After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{LKCL2019}.
+After applying entropy-based discretization on continuous features,  the dataset  contains only categorical features, that can be converted to binary features using one-hot encoding as in \cite{lakkaraju2019faithful}.
 In this encoding, a Boolean vector is introduced with cardinality equal to the number of distinct categories.

### 2023-02-19 chapters/introduction.tex
connecting different chapters

@@
 In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
-Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+Motivated by the GDPR's ``right to explanation'', research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
 In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.

### 2023-02-19 chapters/fairness/fvgm/appendix.tex
short captions
reference fixed

@@
 	We have considered a synthetic problem for comparing accuracy among different verifiers.
-For Example~\ref{fvgm_example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $.
+For Example~\ref{fairness_justicia_example:intro}, we consider `age $ \ge 40 $' as a Bernoulli random  variable with probability $ 0.5 $.
 For `income' feature ($ I $), we consider two Gaussian distributions $ \Pr[I | \text{age} \ge 40] \sim \mathcal{N}(0.6, 0.1) $ and $ \Pr[I | \text{age} < 40] \sim \mathcal{N}(0.4, 0.1) $ separated by two age groups.
@@
 %			
-			\caption{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
+			\caption[Ablation study: effect of the number of features]{Effect of number of features on the runtime of different datasets for LR and SVM classifiers.}
 			\label{fairness_fvgm_fig:time_vary_features}

### 2023-02-19 tables/fairness/fvgm/sanity_check.tex
reference fixed

--- 2023-02-13 tables/fairness/fvgm/sanity_check.tex
@@
 
-	\caption{
-		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on logistic regression classifier.}
+	\caption[Accuracy of {\fvgm} on logistic regression classifier]{
+		Measuring accuracy of different fairness verifiers for Example~\ref{fairness_justicia_example:intro} on logistic regression classifier.}
 	\label{fairness_fvgm_fig:synthetic_results}
@@
 
-	\caption{
-		Measuring accuracy of different fairness verifiers for Example~\ref{fvgm_example:intro} on SVM classifier.}
+	\caption[Accuracy of {\fvgm} on SVM classifier]{
+		Measuring accuracy of different fairness verifiers for Example~\ref{fairness_justicia_example:intro} on SVM classifier.}
 	\label{fairness_fvgm_fig:synthetic_results_SVM}

### 2023-02-20 chapters/fairness/fif/introduction.tex
extended abstract

@@
 \input{chapters/fairness/fif/figure_intro}
-We illustrate the usefulness of our contributions via an example scenario proposed in Example~\ref{fairness_justicia_example:intro}.
+We illustrate the usefulness of our contributions via an example scenario presented in Example~\ref{fairness_justicia_example:intro}.
 

### 2023-02-20 doctoral_consortium/main.tex
extended abstract

@@
 \usepackage{soul}
+\usepackage{comment}
 \usepackage{url}
@@
 
-The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical~\cite{erickson2017machine,kaissis2020secure,kononenko2001machine}, law~\cite{kumar2018law,surden2014machine}, education~\cite{luckin2018machine}, and transportation~\cite{peled2019model,zantalis2019review}.
-In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users~\cite{eshete2021making}.
-With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox}, and privacy~\cite{papernot2016towards} in predictions.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning.
-The thesis establishes a close integration of formal methods with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
+The last decades have witnessed significant progress in machine learning with a host of applications of algorithmic decision-making in different safety-critical domains, such as medical, law, education, and transportation.
+In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users.
+With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox},  etc.\ in predictions.
+Towards responsible and trustworthy machine learning, we propose two research themes in our dissertation research: interpretability and fairness of machine learning classifiers.
+In particular, we design algorithms to \textit{learn interpretable rule-based classifiers}~\cite{ghosh22efficient,ghosh2019incremental,ghosh2020classification}, \textit{formally verify fairness}~\cite{ghosh2021justicia,ghosh2022algorithmic}, and \textit{explain the sources of unfairness}~\cite{ghosh2022how}.
+Prior approaches to these problems are often limited by scalability, accuracy, or both.
+To overcome these limitations, we closely integrate automated reasoning, formal methods, and statistics with fairness and interpretability to develop scalable and accurate solutions.
 
-Towards responsible machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers.
-\emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features.
-Interpretable rule-based classifiers not only interpret the decision function but also be applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
-In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
-In another research theme of \emph{fairness in machine learning}, classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless the classifiers are trained with a fairness objective.
-Consequently, the research on fairness centers on quantifying bias using multiple fairness definitions and mitigating bias based on multiple fairness algorithms.
-In this thesis, we study fairness verification problems to assess the fairness of different classifiers under the lens of multiple fairness definitions and algorithms.
-In addition, we study the sources of the unfairness of classifiers by identifying  features attributing highly to bias.
-To summarize our thesis, in interpretable and fair machine learning, we prioritize improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which prior approaches fail to achieve.
+\section{Interpretable Rule-based Machine Learning}
 
-\section{Interpretable Machine Learning}
-The problem in interpretable machine learning is to learn a classifier making interpretable predictions to the end-users.
-To achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
-At this point, it is important to acknowledge that interpretability is an informal notion, and formalizing it in full generality is challenging.
-In our context of rule-based classifiers, we use \emph{sparsity of rules} (that is, fewer rules each having fewer Boolean literals), which has been considered a proxy of interpretability in various domains, specifically in the medical domain~\cite{gage2001validation,lakkaraju2019faithful,letham2015interpretable,malioutov2013exact,myers1962myers}.
-
-In this thesis, we study two interpretable rule-based classifiers, characterized by their expressiveness.
-At first, we study classifiers represented as formulas in propositional logic.
-In propositional logic, Conjunctive Formal Form (CNF) and Disjunctive Normal Form (DNF) are useful representations of Boolean formulas.
-Popular interpretable rule-based classifiers such as decision tree, decision lists, and decision sets share the logical structure of CNF/DNF in their representation of the decision function.
-Thereby, we propose to learn an interpretable CNF classifier, wherein its interpretability is defined by the number of Boolean literals that the formula contains.
-Compared to CNF, Boolean cardinality constraints are more expressive as they allow numerical bounds on Boolean literals~\cite{sinz2005towards}.
-Relying on the concept of cardinality constraints to increase expressiveness, our second interpretable rule-based classifier is a logical relaxation of CNF/DNF classifiers, namely \emph{relaxed-}CNF.
-
-The problem of learning rule-based classifiers is known to be computationally intractable.
-The earliest tractable approaches for classifiers such as decision trees and decision lists relied on heuristically chosen objective functions and greedy algorithmic techniques~\cite{ClarkN1989,CohenS1999,quinlan2014}.
-In these approaches, the size of rules is controlled by early stopping,  ad-hoc rule pruning, etc.
-In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}.
-Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, Bayesian optimizations~\cite{letham2015interpretable}, and MaxSAT~\cite{malioutov2018mlic} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
-The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets.
-Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP (mixed integer linear programming) to efficiently learn rule-based classifiers in a mini-batch learning setting.
-
-The contributions of this thesis on interpretable machine learning are summarized in the following.
+Interpretable machine learning often employs rule-based classifiers, which use a set of rules to represent the decision boundary.
+The interpretability of such classifiers depends on the size of the rules: smaller rules with higher accuracy are preferred in practice.
+However, this presents a challenge when dealing with large datasets, as interpretable classification learning becomes a combinatorial optimization problem that suffers from poor scalability.
+To address this issue, we propose an incremental learning framework for interpretable rule-based classification on large datasets.
+Our framework combines maximum satisfiability (MaxSAT) and mixed integer linear programming (MILP) with mini-batch learning.
 
 \subsection*{Scalability via Incremental Learning}
-We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for synthesizing interpretable classification rules expressible in proposition logic.
-$ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and the interpretability of classification rules and learns an optimal rule by solving an appropriately designed MaxSAT query.
-Despite the progress of MaxSAT solving in the last decade, the straightforward MaxSAT-based solution cannot scale to practical classification datasets containing thousands to millions of samples.
-Therefore, we incorporate an efficient incremental learning technique inside the MaxSAT formulation by integrating mini-batch learning and iterative rule-learning.
-The resulting framework learns a classifier by iteratively covering the training data, wherein in each iteration, it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch.
-In our experiments, $ \mathsf{IMLI} $ achieves the best balance among prediction accuracy, interpretability, and scalability.
-For instance, $ \mathsf{IMLI} $ attains a competitive prediction accuracy and interpretability w.r.t.\ existing interpretable classifiers and demonstrates impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
-As an application, we deploy $ \mathsf{IMLI} $ in learning popular interpretable classifiers such as decision lists and decision sets.
+We introduce a new incremental learning framework, referred to as $\mathsf{IMLI}$, which is based on MaxSAT for learning interpretable classification rules in propositional logic.
+The framework aims to optimize both the accuracy and interpretability of the classification rules through a joint objective function, and an optimal rule is learned by solving a specially designed MaxSAT query.
+However, while MaxSAT has made considerable progress in the last decade, it is not scalable to practical classification datasets with thousands to millions of samples.
+To address this, we incorporate an efficient incremental learning technique that integrates mini-batch learning and iterative rule-learning within the MaxSAT formulation.
+This results in a framework that learns a classifier by iteratively covering the training data, solving a sequence of smaller MaxSAT queries corresponding to each mini-batch in each iteration.
+Our experiments demonstrate that $\mathsf{IMLI}$ achieves the best balance among prediction accuracy, interpretability, and scalability, with competitive accuracy and interpretability compared to existing interpretable classifiers, and impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
+Finally, we apply $\mathsf{IMLI}$ to learn popular interpretable classifiers such as decision lists and decision sets.
 
 \subsection*{Expressiveness via Logical Relaxation}
-We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness~\cite{GMM2020}.
-Elaborately, we consider relaxed definitions of standard OR/AND operators in Boolean logic by allowing exceptions in the construction of a clause and also in the selection of clauses in a rule.
-Building on these relaxed definitions, we introduce relaxed-CNF classification rules motivated by the popular usage of checklists in the medical domain and Boolean cardinality constraints in logic.
-Relaxed-CNF generalizes widely employed rule representations including CNF, DNF, and decision sets.
-While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive.
-To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
-Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
+We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness, as described in~\cite{ghosh2020classification}.
+Specifically, we consider relaxed definitions of the standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and in the selection of clauses in a rule.
+Based on these relaxed definitions, we introduce relaxed logical classification rules, which are motivated by the use of checklists in the medical domain and Boolean cardinality constraints.
+These rules generalize widely used rule representations, such as CNF, DNF, and decision sets.
+However, the combinatorial structure of these rules results in exponential succinctness, and na\"ive learning techniques are computationally expensive.
+To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules.
+Our experimental analysis shows that $\mathsf{CRR}$ can generate more accurate and sparser classification rules compared to alternative rule-based classifiers.
 
 \section{Fairness in Machine Learning}
-As a technology machine learning is oblivious to societal good or bad.
-The success of machine learning as an accurate predictor, however, finds applications in high-stake decision-making, such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-In such applications, the deployed  classifier often demonstrates bias towards certain \emph{sensitive demographic groups} involved in the data~\cite{dwork2012fairness}.
-For example, a classifier deciding the eligibility of college admission may offer more admission to White-male candidates than to Black-female candidates\textemdash possibly because of the historical bias in the admission data, or the accuracy-centric learning objective of the classifier, or a combination of both~\cite{berk2019accuracy,landy1978correlates,zliobaite2015relation}.
-Following such phenomena, multiple fairness metrics, such as \textit{statistical parity}, \textit{equalized odds}, \textit{predictive parity} etc, have been proposed to quantify the bias of the classifier.
-For example, if the classifier in college admission demonstrates a {statistical parity} of $ 0.6 $, it means that White-male candidates are offered admission $ 60\% $ more than Black-female candidates~\cite{besse2021survey,feldman2015certifying,garg2020fairness}.
-To this end, different fairness enhancing algorithms have been devised to improve fairness with respect to one or multiple fairness metrics.
-These algorithms try to rectify and mitigate bias in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
-Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) identification of sources of unfairness.
 
-\subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features.
-The early works on fairness verification focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}.
-Naturally, such techniques were limited in enhancing confidence of users for wide deployment.
-Consequently, recent verifiers seek to achieve verification beyond  finite dataset and in turn focus on the  probability distribution of features~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}.
-More specifically, the input to the verifier is a classifier and  the probability distribution of features, and the output is an estimate of fairness metrics that the classifier obtains given the distribution.
+Fairness in machine learning involves quantifying and mitigating bias towards different sensitive groups in the data that may be introduced by the classifier.
+Over the past decade, multiple fairness definitions and metrics have been proposed to quantify bias.
+However, there has been little progress in formally verifying these fairness metrics.
+Furthermore, fairness metrics only measure the overall bias of a classifier and are unable to detect or explain the sources of bias.
+Therefore, our research focuses on two key aspects: formally verifying the bias of a classifier and explaining its sources by breaking down bias into individual features and the intersection of multiple features.
 
-In order to solve the fairness verification problem, existing works have proposed two principled approaches.	Firstly,~\cite{albarghouthi2017fairsquare} propose a formal method approach to reduce the verification problem into the weighted volume computation of an SMT formula.
-Secondly,~\cite{bastani2019probabilistic} propose a sampling approach that relies on extensively enumerating the conditional probabilities of prediction given different sensitive features and thus, incurs high computational cost.
-Additionally, existing works assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features w.r.t.
-sensitive features and ignore correlations among non-sensitive features.
-As a result, the \textit{scalability} and \textit{accuracy} of existing  verifiers remain major challenges.
+\subsection*{Probabilistic Fairness Verification} The problem of probabilistic fairness verification is to verify the resulting bias of a classifier given the distribution of input features.
+Early work on fairness verification focused on quantifying the bias of a classifier for a specific dataset.
+However, such techniques were limited in terms of increasing confidence for wide deployment.
+Consequently, recent verifiers aim to achieve verification beyond a finite dataset and instead focus on the probability distribution of features.
+Specifically, the input to the probabilistic fairness verifier is a classifier and the distribution of features, and the output is a quantification of fairness metrics that the classifier obtains given the distribution.
 
-In this thesis, we propose an efficient fairness verification framework for two classes of machine learning classifiers, classifiers represented as Boolean formulas and linear classifiers.
-Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
-{\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.
-Because SSAT encoding allows separate quantification to each sensitive feature without any restriction on the number of features.
-In experiments, {\justicia} is more scalable than the existing probabilistic verifiers~\cite{albarghouthi2017fairsquare,bastani2019probabilistic}, and more robust than the sample-based empirical verifiers~\cite{aif360-oct-2018}.
-%We also prove a finite-sample error bound on estimated fairness metrics, which is stronger than the existing asymptotic guarantees.
+\paragraph{Formal Fairness Verification.} We propose two approaches to probabilistic fairness verification: a general approach that verifies a finite classifier by encoding it into a Boolean formula~\cite{ghosh2021justicia} and a more tailored approach for linear classifiers~\cite{ghosh2022algorithmic}.
+Based on stochastic satisfiability (SSAT), our proposed verifier, called $\mathsf{Justicia}$, verifies the fairness of classifiers such as decision trees by solving appropriately designed SSAT formulas.
+In contrast to prior methods, $\mathsf{Justicia}$ extends verification to compound sensitive groups by combining multiple categorical sensitive features.
+In experiments, $\mathsf{Justicia}$ is more scalable than existing SMT and sampling-based probabilistic verifiers and more robust than dataset-centric empirical verifiers.
 
-Linear classifiers have attracted significant attention from researchers in the context of fair algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy,john2020verifying}.
-Existing fairness verifier suffers from two-fold limitations for verifying linear classifiers: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
-Consequently, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
-{\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
-To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
-Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
-We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing fairness influences of a subset of features on shifting the incurred bias of the classifiers from the original bias.
-%, which we study in detail in the following contribution of the thesis.
+\paragraph{Tractable Fairness Verification With Feature Correlation.} Linear classifiers have received significant attention from researchers in the context of fair algorithms.
+Existing fairness verifiers suffer from two-fold limitations while verifying linear classifiers: (i) poor scalability due to the use of SSAT, SMT, or sampling-based techniques, and (ii) limited accuracy due to ignoring feature correlations.
+To alleviate both limitations, we extend $\mathsf{Justicia}$ with a novel stochastic subset-sum problem-based encoding that verifies linear classifiers by dynamic programming, obtaining pseudo-polynomial complexity.
+To incorporate feature correlations, we consider a probabilistic graphical model, specifically a Bayesian Network, to represent the conditional dependence and independence among features using directed acyclic graphs.
+Experimentally, $\mathsf{Justicia}$ is more accurate and scalable than existing fairness verifiers for linear classifiers while verifying multiple group and causal fairness metrics.
+We also demonstrate two novel applications of $\mathsf{Justicia}$ as a fairness verifier: (a) detecting fairness attacks and fairness improvement algorithms, and (b) computing the impact of feature subsets on shifting the incurred bias of the classifiers from the original bias.
 
-\subsection*{Identification of Sources of Unfairness}
-While fairness metrics globally quantify bias, they cannot detect or explain the sources of bias~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
-In order to identify the sources of bias and also the effect of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which factors contribute how much to the bias of a classifier on a dataset}.
-To this end, we follow a feature-attribution approach to understand the sources of bias~\cite{begley2020explainability,lundberg2020explaining}, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
-Particularly, we define and compute \textit{Fairness Influence Function} (FIF) that quantifies the contribution of individual and subset of features to the resulting bias.
-FIFs do not only allow practitioners to identify the features to act up on but also to quantify the effect of various affirmative~\cite{calmon2017optimized,hardt2016equality,kamiran2012decision,zemel2013learning,zhang2018mitigating,zhang2018fairness,zhang2019faht} or punitive actions~\cite{hua2021human,mehrabi2020exacerbating,solans2020poisoning} on the resulting bias.
-We also instantiate an algorithm, {\fairXplainer}, that uses variance decomposition among the subset of features and a local regressor to compute FIFs accurately, while also capturing the intersectional effects of the features.
-Our experimental analysis validates that {\fairXplainer} captures the influences of both individual features and higher-order feature interactions, estimates the bias more accurately than existing local explanation methods, and detects the increase/decrease in bias due to affirmative/punitive actions in the classifier.
+\subsubsection*{Explaining Fairness Metrics: Identifying the Sources of Bias}
+Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias.
+To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}.
+We use a feature-attribution approach to explain the sources of bias, which relates the \emph{influences} of input features to the resulting bias of the classifier.
+We formalize the \textit{Fairness Influence Function} (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}.
+We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis.
+In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions.
 
-%% The file named.bst is a bibliography style file for BibTeX 0.99c
 \bibliographystyle{named}
-\bibliography{ref}
+\bibliography{main}
 

### 2023-02-22 doctoral_consortium/main.tex
extended abstract

@@
     \affiliations
-    National University of Singapore
+    School of Computing, National University of Singapore
     \emails
@@
 Our experiments demonstrate that $\mathsf{IMLI}$ achieves the best balance among prediction accuracy, interpretability, and scalability, with competitive accuracy and interpretability compared to existing interpretable classifiers, and impressive scalability on large datasets where both interpretable and non-interpretable classifiers fail.
-Finally, we apply $\mathsf{IMLI}$ to learn popular interpretable classifiers such as decision lists and decision sets.
+We also apply $\mathsf{IMLI}$ to learn popular interpretable classifiers: decision lists and decision sets.
 
@@
 
-\subsubsection*{Explaining Fairness Metrics: Identifying the Sources of Bias}
+\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness.
 Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias.
 To understand the sources of bias, it's necessary to determine \textit{which factors contribute how much to the bias of a classifier on a dataset}.
-We use a feature-attribution approach to explain the sources of bias, which relates the \emph{influences} of input features to the resulting bias of the classifier.
-We formalize the \textit{Fairness Influence Function} (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}.
+We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier.
+We formalize the Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}.
 We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis.
 In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions.
+In future, we aim to develop improved fairness enhancing and attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.
 

### 2023-02-26 doctoral_consortium/personal_statement/bghosh_cover_letter.tex
personal statement

\documentclass[a4paper,10pt]{article}
\usepackage[top=2cm,left=3.1cm,right=3.1cm,bottom=2cm]{geometry}% margins
\usepackage{graphicx}
\usepackage{url}
\usepackage[hidelinks]{hyperref}
\usepackage{comment}
\usepackage{natbib}

\usepackage{xcolor}
\newcommand{\magenta}[1]{\textcolor{magenta}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}

%noindent
\setlength\parindent{0pt}

\begin{document}

%	logo
	\begin{figure}[t]
		\includegraphics[scale=0.03]{logo}
	\end{figure}
	\noindent\makebox[\linewidth]{\rule{\textwidth}{1pt}}

	\begin{flushright}
		\begin{tabular}{l@{}}
			\today \\
			\newline \\
			\textbf{Bishwamittra Ghosh}\\
			205 Bukit Batok Street 21, \#02-46\\ 
			Singapore, 650205\\
			Singapore\\
			\blue{\url{https://bishwamittra.github.io}}\\
			\textbf{Email:} bghosh@u.nus.edu\\
			\textbf{Phone:} (+65)85990160\\
		\end{tabular}%
	\end{flushright}

	\textbf{Subject:} Application for attending IJCAI 2023 Doctoral Consortium Program.

	\vspace{1em}
	Dear Organizing Committee,

	\vspace{1em}
	I am writing to apply for the IJCAI 2023 Doctoral Consortium Program. Please find attached my application materials.

	\vspace{1em}

	My research interests primarily lie in the areas of fairness and interpretability in machine learning. Specifically, in my dissertation research, I focus on formally verifying fairness in machine learning, interpreting fairness by identifying its sources, and designing interpretable rule-based classifiers. To achieve scalable and accurate solutions, I closely integrate automated reasoning, formal methods, and statistics with fairness and interpretability.

	\vspace{1em}
	My research is motivated by prior works on fairness and interpretability. For example, \cite{rudin2019stop} emphasizes the design of interpretable classifiers, such as rule-based classifiers~\citep{lakkaraju2016interpretable}, for high-stake and safety-critical decision-making, instead of interpreting opaque black-box classifiers. We improve the scalability of interpretable classifiers based on maximum satisfiability (MaxSAT) and incremental learning by demonstrating interpretable classification on million-size datasets. In fairness, \cite{albarghouthi2017fairsquare} and \cite{bastani2019probabilistic} study formal fairness verification to verify multiple fairness metrics of classifiers given the probability distribution of features. In our research, we improve the scalability and accuracy of formal fairness verification by integrating stochastic satisfiability (SSAT) and Bayesian network capturing feature-correlations. Furthermore, motivated by the GDPR's ``right to explanation'', research on interpreting model predictions has surged~\citep{lundberg2017unified}. However, interpreting prediction bias has received less attention~\citep{lundberg2020explaining}. In this context, we apply global sensitivity analysis (GSA) to interpret fairness metrics by decomposing model bias into individual features and the intersection of multiple features.

	\vspace{1em}

   I have not previously attended a doctoral consortium. I expect to graduate in Summer 2023 and plan to pursue a research career in either academia or industry, with a research focus centered on solving novel and practical problems in fairness and interpretability in machine learning for the next five years. To this end, I would be delighted to have mentors working in fairness and interpretability, such as Xiting Wang, Arthur Choi, Alexey Ignatiev, Dominik Peters, and Mohamed Siala.

	\vspace{1em}
	My research has thrived through multiple collaborations and internships in industry and academia. We have published our works at premier conferences and journals in artificial intelligence and machine learning (AAAI-2022, 2021, 2020, JAIR-2022, ECAI-2020, AIES-2019) and databases (VLDB-2018, TSAS-2022).

%	\vspace{1em}
%	My application materials are also available from my website \blue{\url{https://bishwamittra.github.io}}.

	\vspace{1em}
	Thank you for considering my application. I look forward to hearing from you.

	\vspace{1em}
	Sincerely,\\
	Bishwamittra Ghosh

	\bibliographystyle{plainnat}
	\bibliography{ref.bib}

\begin{comment}
\clearpage
\section*{Reference}
\textbf{Kuldeep S. Meel}\\
Sung Kah Kay Assistant Professor\\ 
School of Computing, National University of Singapore, Singapore\\
Email: meel@comp.nus.edu.sg\\

\textbf{Daniel Neider}\\
Professor\\
Carl von Ossietzky Universit\"at Oldenburg, Germany\\
Email: daniel.neider@uol.de\\

\textbf{Debabrota Basu}\\
Faculty\\
Scool, Inria, Lille-Nord, France\\
Email: debabrota.basu@inria.fr\\

\end{comment}

\end{document}

### 2023-02-28 chapters/abstract.tex
edits according to Kuldeep

@@
 
-	The significant success of machine learning in past decades has witnessed a host of applications of algorithmic decision-making in different safety-critical domains.
+	The significant success of machine learning in past decades has led to a host of applications of algorithmic decision-making in different safety-critical domains.
 The high-stake predictions of machine learning in medical, law, education, transportation and so on have far-reaching consequences on the end-users.
-Consequently, researchers call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
-In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{explaining sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
+Consequently, there has been a call for the regulation of machine learning by defining and improving the interpretability, fairness, robustness, and privacy of predictions.
+In this thesis, we focus on the interpretability and fairness aspects of machine learning, particularly on \emph{learning interpretable rule-based classifiers}, \emph{verifying fairness}, and \emph{interpreting sources of unfairness.} Prior studies aimed for these problems are limited by either scalability or  accuracy or both.
 To alleviate these limitations, we integrate formal methods and automated reasoning with interpretability and fairness in machine learning and provide scalable and accurate solutions to the underlying problems.
@@
 
-	Fairness metrics globally quantify bias, but do not detect or explain its sources.
-To explain group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier.
-FIF  explains fairness by revealing potential individual or intersectional features attributing highly to the bias.
+	Fairness metrics globally quantify bias, but do not detect or interpret its sources.
+To interpret group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier.
+FIF  interprets fairness by revealing potential individual or intersectional features attributing highly to the bias.
 Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.

### 2023-02-28 chapters/background.tex
edits according to Kuldeep

@@
 
-The MaxSAT problem is an optimization analog to the SAT problem that finds an optimal assignment satisfying the maximum number of clauses in a CNF.
+The MaxSAT problem is an optimization analog to the satisfiability (SAT) problem, which is complete for the class $ \mathrm{FP}^{\mathrm{NP}} $.
+Although the MaxSAT problem is NP-hard, dramatic progress has been made in designing solvers that can handle large-scale problems arising in practice.
+This has encouraged researchers to reduce several optimization problems into MaxSAT such as  optimal planning~\cite{robinson2010partial}, automotive configuration~\cite{walter2013applications}, group-testing~\cite{CGSM2020}, data analysis in machine learning~\cite{berg2019applications}, and automatic test pattern generation for cancer therapy~\cite{lin2012application}.
+
+The MaxSAT problem finds an optimal assignment satisfying the maximum number of clauses in a CNF.
 In interpretable machine learning, we consider a \emph{weighted} variant of the MaxSAT problem\textemdash more specifically, a weighted-partial MaxSAT problem\textemdash that optimizes over a set of hard and soft constraints in the form of a weighted-CNF formula.
@@
 Given a constraint of the form  $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i = \tau $, for a constant threshold $ \tau \in \mathbb{Z} $, the subset-sum problem seeks to compute an assignment $\boolset \in \{0,1\}^{|\boolset|}$ such that the constraint evaluates to true when $\boolset$ is substituted with $ \mathbf{b} $.
-Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical
-computer science~\cite{kleinberg2006algorithm}.
-The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds; and this problem belongs to the complexity class $ \mathrm{\#P} $.
-In this thesis, we consider the counting problem for the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic.
-More precisely, we define a \textit{stochastic subset-sum} problem, namely {\stochastic}, that computes $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $.
+Subset sum problem is known to be a $ \mathrm{NP} $-complete problem and well-studied in theoretical computer science~\cite{kleinberg2006algorithm}.
+The \textit{counting} version of the subset-sum problem counts all $  \mathbf{b}  $'s for which the above constraint holds.
+In this thesis, we consider the constraint $\sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge \tau $ where variables $ \bool_i $'s are stochastic.
+While the counting problem of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i =  \tau $ is NP-hard to approximate, there is a FPRAS solution for approximately counting the solutions of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge  \tau $~\cite{dyer2003approximate}.
+In this thesis, we are interested in computing the probability  $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $, referred to as \textit{stochastic subset-sum}  problem ({\stochastic}).
 Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.
@@
 \label{chapter_fairness_preliminaries}
-In this section, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and identify the sources of unfairness.
+In this section, we discuss preliminaries in fairness in machine learning, followed by methods to verify fairness and interpret unfairness.
 
 \subsection{Dataset and Distribution}
-We consider a dataset $ \mathbf{D} $ as a collection of $n$ data points  $\{(\mathbf{z}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
 The feature vector $\mathbf{z}^{(i)} \triangleq (\mathbf{x}^{(i)}, \mathbf{a}^{(i)}) $ is a concatenation of non-sensitive features $ \mathbf{x}^{(i)} $ and sensitive features $ \mathbf{a}^{(i)} $.
@@
 Thus, $ \mathcal{D} $ is the product distribution of all features, $ \mathcal{D} \triangleq \prod_{i=1}^{\numnonsensitive}\mathcal{X}_i \prod_{j=1}^{\numsensitive} \mathcal{A}_j $.
-
 For sensitive features, a valuation vector $ \mathbf{a} = [a_1, .., a_{\numsensitive}] $ is called a \textit{compound sensitive group}.
@@
 Thus $ \mathbf{a} = $ [Asian, female]  is a compound sensitive group.
-
 We represent a binary classifier trained on the dataset $\dataset$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.

### 2023-02-28 chapters/interpretability/imli/introduction.tex
edits according to Kuldeep

@@
 In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}.
-In this chapter, we learn an interpretable rule-based classifier~\cite{rudin2019stop}, where the rules governing the prediction are explicit (by design) in contrast to black-box classifiers. 
+In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers.
+In interpretable rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}.
 %In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
@@
 
-The MaxSAT problem is an optimization analog to the satisfiability (SAT) problem, which is complete for the class $ \mathrm{FP}^{\mathrm{NP}} $.
-Although the MaxSAT problem is NP-hard, dramatic progress has been made in designing solvers that can handle large-scale problems arising in practice.
-This has encouraged researchers to reduce several optimization problems into MaxSAT such as  optimal planning~\cite{robinson2010partial}, automotive configuration~\cite{walter2013applications}, group-testing~\cite{CGSM2020}, data analysis in machine learning~\cite{berg2019applications}, and automatic test pattern generation for cancer therapy~\cite{lin2012application}.
-
-\paragraph{Contributions.} The primary contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
+\paragraph{Contributions.} The contribution of this chapter is a MaxSAT-based formulation, called {\imli} (\textbf{I}ncremental \textbf{M}axSAT-based \textbf{L}earning of \textbf{I}nterpretable classification rules), for learning interpretable classification rules expressible in propositional logic.
 For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.

### 2023-02-28 chapters/background.tex
addressing kuldeep's comments

@@
 
-\subsection{Dataset and Distribution}
-We consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
+\subsection{Dataset and Distribution} In the fairness literature, features are categorized into two types: non-sensitive and sensitive features. Hence,
+we consider a dataset $ \mathbf{D} $ as a collection of $n$ triples  $\{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$  generated from an underlying distribution $\mathcal{D}$.
 The feature vector $\mathbf{z}^{(i)} \triangleq (\mathbf{x}^{(i)}, \mathbf{a}^{(i)}) $ is a concatenation of non-sensitive features $ \mathbf{x}^{(i)} $ and sensitive features $ \mathbf{a}^{(i)} $.

### 2023-02-28 chapters/conclusion.tex
addressing kuldeep's comments

@@
 (ii) In fairness in machine learning, we develop a formal probabilistic fairness verification framework that can verify multiple fairness definitions of classifiers.
-Additionally, we develop techniques to explain fairness metrics by identifying feature combinations responsible for the bias of the classifier.
+Additionally, we develop techniques to interpret fairness metrics by identifying feature combinations responsible for the bias of the classifier.
 
@@
 On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers.
-Additionally, we propose a global sensitivity analysis-based method, {\fairXplainer}, that explains group fairness metrics by computing fairness influences of individual and intersectional features.
+Additionally, we propose a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features.
 Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.

### 2023-02-28 chapters/fairness/fif/conclusion.tex
addressing kuldeep's comments

@@
 Our approach combines global sensitivity analysis and group-based fairness metrics in machine learning.
-Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs explains bias with higher granularity.
+Thereby, it is natural in our approach to formulate FIF of intersectional features, which together with individual FIFs interprets bias with higher granularity.
 We theoretically analyze the properties of FIFs and provide an algorithm, {\fairXplainer}, for estimating FIFs using global variance decomposition and local regression.

### 2023-02-28 chapters/fairness/fif/experiments.tex
addressing kuldeep's comments

@@
 While estimating intersectional FIFs, {\fairXplainer} also demonstrates its practical applicability by solving $ 350 $ instances within  $ 160 $ seconds.
-\textit{Therefore, {\fairXplainer} demonstrates computational efficiency in explaining group fairness metrics of real-world datasets and classifiers.}
+\textit{Therefore, {\fairXplainer} demonstrates computational efficiency in interpreting group fairness metrics of real-world datasets and classifiers.}
 
@@
 
-	\caption[FIFs for COMPAS dataset]{FIFs for COMPAS dataset on explaining statistical parity.
+	\caption[FIFs for COMPAS dataset]{FIFs for COMPAS dataset on interpreting statistical parity.
 Both individual and intersectional FIFs in Figure~\ref{fairness_fairXplainer_fig:intersectional_fifs} depict sources of bias in detail than individual FIFs alone in Figure~\ref{fairness_fairXplainer_fig:individual_fifs}.

### 2023-02-28 chapters/fairness/fif/framework.tex
addressing kuldeep's comments

@@
 We refer to this indicator function as \textit{Conditional Positive Prediction (CPP)} of a classifier.
-Now, we express statistical parity as a functional of the probability of CPPs for different sensitive groups~\cite{ghosh2020justicia,benesse2021fairness}.
+Now, we express statistical parity as a functional of the probability of CPPs for different sensitive groups~\cite{benesse2021fairness}.
 For brevity, we defer similar formulations for the other fairness metrics, i.e.
@@
 We refer to $ w_{\mathbf{S}} $ as an \emph{individual influence} when $ |\mathbf{S}| = 1 $, and as an \emph{intersectional influence} when $ |\mathbf{S}| > 1 $.
-Being able to naturally quantify the higher-order influences allow FIFs to explain the sources of bias in a more fine-grained manner.
+Being able to naturally quantify the higher-order influences allow FIFs to interpret the sources of bias in a more fine-grained manner.
 We experimentally validate this in Section~\ref{fairness_fairXplainer_sec:experiments}.

### 2023-02-28 chapters/fairness/fif/introduction.tex
addressing kuldeep's comments

@@
 
-In this chapter, we envision towards combining two research themes in the thesis: interpretability and fairness, and discuss an algorithmic framework to interpret or explain fairness in machine learning.
-As demonstrated in Chapter~\ref{chapter:justicia} and~\ref{chapter:fvgm}, fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+In this chapter, we envision towards combining two research themes in the thesis: interpretability and fairness, and discuss an algorithmic framework to interpret  fairness in machine learning.
+As demonstrated in Chapter~\ref{chapter:justicia} and~\ref{chapter:fvgm}, fairness metrics measure global bias, but do not detect or interpret its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
 In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
-Motivated by the GDPR's ``right to explanation'', research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
-In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
-To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
+Motivated by the GDPR's ``right to explanation'', research on interpreting model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but interpreting prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+In order to identify and interpret the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
+To this end, we follow a global feature-attribution approach to interpret the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
 In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier.
@@
 In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features.
-Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
+Quantifying intersectionality allows us to interpret bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
 %\textit{By attributing bias to both individual and intersecting features, we can gain a more detailed understanding of bias} and pave the way for designing more effective fairness algorithms. 
@@
 Our work uses the feature-attribution approach and highlights three limitations of existing methods: (i) a failure to estimate intersectional FIFs, (ii) inaccuracies in approximating bias based on FIFs, and (ii) less correlation of FIFs to fairness intervention.
-Elaborately, to explain group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}.
+Elaborately, to interpret group fairness\textemdash a global property of a classifier\textemdash existing local explanation approaches such as ~\cite{begley2020explainability,lundberg2020explaining} estimate FIFs based on a local black-box explainer SHAP~\cite{lundberg2017unified}.
 They apply a global aggregation (i.e.
@@
 In addition, existing methods ignore the joint contribution of correlated features on bias.
-To address these limitations, \textit{we develop a formal framework to explain sources of group unfairness in a classifier and also a novel methodology to estimate FIFs.
+To address these limitations, \textit{we develop a formal framework to interpret sources of group unfairness in a classifier and also a novel methodology to estimate FIFs.
 To the best of our knowledge, this is the first work to do both}.
@@
 \textit{Their approach detects the presence or absence of bias, while we focus on decomposing bias as the sum of FIFs of all feature subsets}.
-In another line of work, \cite{datta2016algorithmic} and \cite{ghosh2022algorithmic} estimate feature-influence as the shifting of bias from its original value by randomly intervening features.
+In another line of work, \cite{datta2016algorithmic}  estimate feature-influence as the shifting of bias from its original value by randomly intervening features.
 Their work is different from the decomposability property of FIFs, where the sum of FIFs is equal to the bias.

### 2023-02-28 chapters/fairness/fvgm/conclusion.tex
addressing kuldeep's comments

@@
 {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately.
-We experimentally demonstrate that {\fvgm} is not only more accurate and scalable than the existing verifiers but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions.
-In the next chapter, we elaborate on fairness influence functions as a framework to explaining group fairness metrics in machine learning.
+We experimentally demonstrate that {\fvgm} is  more accurate and scalable than the existing verifiers.
+% but also applicable in practical fairness tasks, such as verifying fairness attacks and enhancing algorithms, and computing the fairness influence functions. 
+% In the next chapter, we elaborate on fairness influence functions as a framework to explaining group fairness metrics in machine learning.
 \begin{comment}

### 2023-02-28 chapters/fairness/fvgm/evaluations.tex
highlevel pass
addressing kuldeep's comments

@@
 
-\begin{figure}[t!]
+\begin{figure}
 	\begin{center}		
@@
 
-We compare {\fvgm} with three existing fairness verifiers: Justicia~\cite{ghosh2020justicia}, FairSquare~\cite{albarghouthi2017fairsquare}, and VeriFair~\cite{bastani2019probabilistic}.In the following, we discuss a comparative analysis among all verifiers based on scalability and accuracy, where {\fvgm} yields superior performance than others.
+We compare {\fvgm} with three existing fairness verifiers: Justicia (Chapter~\ref{chapter:justicia}), FairSquare~\cite{albarghouthi2017fairsquare}, and VeriFair~\cite{bastani2019probabilistic}.
+In the following, we discuss a comparative analysis among all verifiers based on scalability and accuracy, where {\fvgm} yields superior performance than others.
 
@@
 
-\begin{figure}[t!]
+\begin{figure}
 	\begin{center}	
@@
 
+\begin{comment}
+
 \begin{table}[!t]
@@
 
+\end{comment}
+
 %\begin{table}[!t]

### 2023-02-28 chapters/fairness/fvgm/introduction.tex
addressing kuldeep's comments

@@
 
-	In this chapter, we extend the probabilistic fairness verification problem to a more practical scenario by accounting for the correlation of features, which is prevalent in most classification datasets.
-Existing fairness verifiers \cite{albarghouthi2017fairsquare,bastani2019probabilistic}, including Justicia \cite{ghosh2020justicia} in Chapter~\ref{chapter:justicia}, suffer from poor accuracy due to considering specific input distribution.
+	In this chapter, we extend formal fairness verification problem to a more practical scenario by accounting for the correlation of features, which is prevalent in most classification datasets.
+Existing fairness verifiers \cite{albarghouthi2017fairsquare,bastani2019probabilistic}, including Justicia in Chapter~\ref{chapter:justicia}, suffer from limited accuracy due to considering specific input distribution.
 For example, they assume feature independence of non-sensitive features and consider correlated features within a limited scope, such as conditional probabilities of non-sensitive features with respect to sensitive features, and ignore correlations among non-sensitive features.
@@
 
-	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step, because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
 In this context, existing approaches suffer from limited scalability while verifying linear classifiers.
-This is due to the encoding of linear classifiers into SSAT/SMT formulas \cite{ghosh2020justicia,albarghouthi2017fairsquare}.
+This is due to the encoding of linear classifiers into SSAT (Chapter~\ref{chapter:justicia}) or SMT formulas \cite{albarghouthi2017fairsquare}.
 For example, {\justicia} applies pseudo-Boolean to CNF translation of linear classifiers as a pre-processing step, and the encoding becomes large depending on the number of features and the precision of real-valued coefficients in linear classifiers.
@@
 
-	\paragraph{Our Contributions.} The contributions of this chapter are summarized below.
+	\paragraph{Contributions.} The contributions of this chapter are summarized below.
 
@@
 		\item \textit{Experimental Results:}	Experimentally,  {\fvgm} is more accurate and scalable than existing fairness verifiers; {\fvgm} can verify group and causal fairness metrics for multiple fairness algorithms.
-		\item \textit{Applications:} We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.
+%		\item \textit{Applications:} We also demonstrate two novel applications of {\fvgm} as a fairness verifier: (a) detecting fairness attacks, and (b) computing Fairness Influence Functions (FIF) of features as a mean of identifying (un)fairness contribution of a subset of features.
 	\end{itemize}

### 2023-02-28 chapters/fairness/justicia/experiments.tex
addressing kuldeep's comments

@@
 In the experiments, we have studied three type of classifiers: decision tree, logistic regression classifier, and CNF learner.
-Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{scikit-learn} and we use the MaxSAT-based CNF learner, namely IMLI~\cite{GM2019}.
+Decision tree and logistic regression are implemented using scikit-learn module of Python~\cite{scikit-learn} and we use the MaxSAT-based CNF learner IMLI (Chapter~\ref{chapter:imli}).
 We have used the PySAT library~\cite{imms-sat18} for encoding the decision function of the logistic regression classifier into a CNF formula.

### 2023-02-28 chapters/fairness/justicia/framework.tex
with new formatting
highlevel pass
addressing kuldeep's comments

@@
-\section{{\justicia}: An SSAT-based Fairness Verifier}
+\section{An SSAT-based Fairness Verifier}
 \label{fairness_justicia_sec:framework}
-In this section, we present the primary contribution of this chapter, {\justicia}, which is an SSAT-based framework for verifying group and causal fairness metrics.
+In this section, we present {\justicia}, which is an SSAT-based framework for verifying group and causal fairness metrics.
 Given a binary classifier $\alg$, a probability distribution over features $(\nonsensitive,\sensitive,Y) \sim \mathcal{D} $, and a target fairness metric $ f(\alg, \mathcal{D}) $, our goal is to estimate the fairness $ f(\alg, \mathcal{D}) $ of the classifier $ \alg $ given the distribution $ \mathcal{D} $ according to the fairness definition.
@@
 		\State $  \Phi_\mathbf{a} := \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \exists A_1,\dots, \exists A_{\numsensitive},  \phi $
-		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns a probability}
+		\State $ \Pr[\Phi_\mathbf{a}]  \leftarrow \mathsf{SSAT}(\Phi_\mathbf{a}) $ \Comment{returns probability}
 		\EndFor
-		\State\Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
+		\Return $ \max_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}], \min_{\mathbf{a} \in \sensitive} \Pr[\Phi_{\mathbf{a}}] $
 		\label{fairness_justicia_algo:justicia_enum_end}
@@
 		\State $  \Phi'_\mathsf{ER} := \exists A_1,\dots, \exists A_{\numsensitive}, \R^{p_{1}}X_1, \dots, \R^{p_{\numnonsensitive}}X_{\numnonsensitive}, \neg \phi_{\widehat{Y}} $
-		\State\Return $ \mathsf{SSAT}(\Phi_\mathsf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathsf{ER}) $
+		\Return $ \mathsf{SSAT}(\Phi_\mathsf{ER}), 1 - \mathsf{SSAT}(\Phi'_\mathsf{ER}) $
 		\label{fairness_justicia_algo:justicia_learn_end}

### 2023-02-28 chapters/fairness/justicia/introduction.tex
addressing kuldeep's comments

@@
 \label{chapter:justicia}
-Machine learning is becoming the omnipresent technology of our time.
+In this chapter, we study the formal verification of fairness in machine learning, where we verify the bias of a classifier given the probability distribution of features.
+We start with an example scenario to depict the fairness problem.
+
+%Machine learning is becoming the omnipresent technology of our time.
 Machine learning classifiers are being used for high-stake decisions such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
@@
 
-\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
+%\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
 In this chapter, we focus on a popular concept of fairness, known as group fairness.
 Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
+Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
+Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}.
+Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
+
+%In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups.
+Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same.
+Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
+Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}.
+To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions.
+These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
+
+%\paragraph{Fairness Verifiers.} 
+Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
+Our focus in this chapter is on group fairness, which constitutes three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
 Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
@@
 Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}.
-To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions.
-These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
+Due to the abundance of fairness metrics, it has become necessary to verify them.
 
-\paragraph{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
-
-In order to verify fairness as a model property on a dataset, verifiers like \textit{FairSquare}~\cite{albarghouthi2017fairsquare} and \textit{VeriFair}~\cite{bastani2019probabilistic} have been proposed.
+In order to verify fairness as a model property, verifiers like \textit{FairSquare}~\cite{albarghouthi2017fairsquare} and \textit{VeriFair}~\cite{bastani2019probabilistic} have been proposed. 
 %FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.

### 2023-02-28 chapters/fairness/justicia/practical.tex
addressing kuldeep's comments

--- 2023-02-13 chapters/fairness/justicia/practical.tex
@@
 \subsubsection{Extension to Continuous Features}
-In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF such as~\cite{GMM20}, are generally trained on a Boolean abstraction of input features.
-In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features~\cite{LKCL2019,GMM20}.
+In practical problems, features are generally real-valued or categorical but classifiers, which are naturally expressed as CNF (Chapter~\ref{chapter:imli}), are generally trained on a Boolean abstraction of input features.
+In order to perform the Boolean abstraction, each categorical feature is one-hot encoded and each real-valued feature is discretized into a set of Boolean features (Chapter~\ref{chapter:imli}).
 

### 2023-02-28 chapters/interpretability/imli/baseline.tex
highlevel pass
addressing kuldeep's comments

@@
-\section{MaxSAT-based Interpretable Classification Rule Learning}
+\section{Interpretable Classification Rule Learning via MaxSAT}
 \label{interpretability_imli_sec:baseline}
-In this section, we discuss a MaxSAT-based framework for optimal learning of an interpretable rule-based classifier, particularly a CNF classifier $ \Rule $. 
+In this section, we revisit~\cite{malioutov2018mlic} by discussing a MaxSAT-based learning framework for an interpretable rule-based classifier, particularly a CNF classifier $ \Rule $. 
 We first describe the decision variables in Section~\ref{interpretability_imli_sec:variables} and present the MaxSAT encoding  in Section~\ref{interpretability_imli_sec:encoding}.
-Our MaxSAT formulation assumes binary features as input.
+The MaxSAT formulation assumes binary features as input.
 We

### 2023-02-28 chapters/interpretability/imli/introduction.tex
addressing kuldeep's comments

@@
 In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers.
-In interpretable rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}.
+In rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}.
 %In order to achieve the interpretability of predictions, decision functions in the form of classification rules such as decision trees,  decision lists, decision sets, etc.\ are particularly effective~\cite{bessiere2009minimising,dash2021lprules,ignatiev2021reasoning,izza2020explaining,lakkaraju2017interpretable,lakkaraju2016interpretable,letham2015interpretable,narodytska2018learning,rivest1987learning,wang2015falling,yu2020optimal}.
@@
 Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
-In our study, we propose an alternate optimization approach that fits particularly well to rule-learning problems.
-Particularly,  we propose a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.
+In our study, we study an alternate optimization approach that fits particularly well to rule-learning problems.
+Particularly,  we study and improve a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.
+Built on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules.
+The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples.
+Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $.
+To scale learning to large datasets, we propose an incremental learning technique along with the MaxSAT formulation.
 
@@
 For the simplicity of exposition, our initial focus is on learning classifiers expressible in CNF formulas; we later discuss how the CNF learning formulation can be extended to popular interpretable classifiers such as decision lists and decision sets.
-In this chapter, {\imli} provides precise control to jointly optimize the accuracy and the size of classification rules.
-{\imli} constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples.
-Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $.
-To scale {\imli} to large datasets, we propose an incremental learning technique along with the MaxSAT formulation.
 Our incremental learning is an integration of mini-batch learning and iterative rule-learning, which are studied separately in classical learning problems.

### 2023-02-28 chapters/interpretability/relaxed-cnf/experiment.tex
addressing kuldeep's comments

@@
 \section{Experiments}\label{interpretability_crr_sec:experiment}
-We implement a prototype  of  $ \crr $\footnote{The source code is available at \url{https://github.com/meelgroup/mlic}} based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances.
+We implement a prototype  of  $ \crr $ based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances.
 The objective of our experimental evaluation is to answer the following questions:
@@
 We compare the performance of  $ \crr $   with  state-of-the-art classifiers, e.g.
-$ \IMLI  $~\cite{GM2019}, RIPPER~\cite{cohen1995fast}, BRS~\cite{wang2017bayesian},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
+$ \IMLI  $ (Chapter~\ref{chapter:imli}), RIPPER~\cite{cohen1995fast}, BRS~\cite{wang2017bayesian},  random forest (RF), support vector classifier (SVC), nearest neighbors classifiers ($ k $-NN), and $ l_1 $ penalized logistic regression (LR).
 Among them, {\IMLI}, BRS, and RIPPER are rule-based classifiers.

### 2023-02-28 chapters/interpretability/relaxed-cnf/introduction.tex
addressing kuldeep's comments

@@
 
-In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing from chapter~\ref{chapter:imli},
+In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing on chapter~\ref{chapter:imli},
 although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
@@
 
-The primary contribution of this chapter is an affirmative answer to the above question by proposing an efficient combinatorial learning framework for relaxed-CNF rules, namely {\crr} (\textbf{C}lassification \textbf{R}ules in \textbf{R}elaxed form).
+\paragraph{Contribution.} The contribution of this chapter is an affirmative answer to the above question by proposing an efficient combinatorial learning framework for relaxed-CNF rules, namely {\crr} (\textbf{C}lassification \textbf{R}ules in \textbf{R}elaxed form).
 In CRR, we construct a learning objective to maximize both the prediction accuracy and the rule-sparsity of the generated classification rule.

### 2023-02-28 chapters/introduction.tex
addressing kuldeep's comments

--- 2023-02-17 chapters/introduction.tex
@@
 In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features.
-Finally, we explain the sources of bias by formalizing and computing fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
+Finally, we interpret the sources of bias by formalizing and computing fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
 To summarize our thesis on interpretability and fairness in machine learning, we prioritize on improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.
 
-\section{Interpretable Machine Learning}
+\section{Interpretable Rule-based Machine Learning}
 The problem in interpretable machine learning is to learn a classifier making interpretable predictions to the end-users.
@@
 
-\subsection*{Scalability via Incremental Learning}
-We propose an incremental learning framework, called $ \imli $~\cite{GMM2022,GM2019},  based on MaxSAT for learning interpretable classification rules expressed in CNF.
+\subsection{Scalability via Incremental Learning}
+We propose an incremental learning framework, called $ \imli $,  based on MaxSAT for learning interpretable classification rules expressed in CNF.
 $ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and sparsity of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query.
@@
 
-\subsection*{Expressiveness via Logical Relaxation}
+\subsection{Expressiveness via Logical Relaxation}
 We extend our incremental learning framework to learn a more relaxed representation of classification rules with higher expressiveness.
@@
 Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
+
+\textbf{Tools:} We publish a python library \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
 
@@
 Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting or explaining of sources of unfairness.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting sources of unfairness.
 
-\subsection*{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features.
+\subsection{Probabilistic Fairness Verification} The problem in probabilistic fairness verification is to verify the bias of a classifier given the distribution of input features.
 The early works on fairness verification focused on measuring fairness metrics of a classifier for a given dataset~\cite{aif360-oct-2018}.
@@
 
-In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
+\paragraph{Contributions.} In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
 Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
@@
 
-\subsection*{Explaining Fairness Metrics: Identifying Sources of Bias}
-Fairness metrics measure global bias, but do not detect or explain its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
+\textbf{Tools:} We publish a python library \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
+
+\subsection{Interpreting Fairness: Identifying Sources of Bias}
+Fairness metrics measure global bias, but do not detect or interpret its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
 In order to diagnose the emergence of bias in the predictions of classifier, it is important to compute explanations, such as how different features attribute to the global bias.
-Motivated by the GDPR's ``right to explanation", research on explaining model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but explaining prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
-In order to identify and explain the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
-To this end, we follow a global feature-attribution approach to explain the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
+Motivated by the GDPR's ``right to explanation'', research on interpreting model predictions~\cite{ribeiro2016should,lundberg2017unified,lundberg2020local2global} has surged, but interpreting prediction bias has received less attention~\cite{begley2020explainability,lundberg2020explaining}.
+In order to identify and interpret the sources of bias and also the impact of affirmative/punitive actions to alleviate/deteriorate bias, it is important to understand \textit{which features contribute how much to the bias of a classifier} applied on a dataset.
+To this end, we follow a global feature-attribution approach to interpret the sources of bias, where we relate the \emph{influences} of input features towards the resulting bias of the classifier.
 In this context, existing bias attributing methods~\cite{begley2020explainability,lundberg2020explaining} are variants of local function approximation~\cite{sliwinski2019axiomatic}, whereas bias is a global statistical property of a classifier.
@@
 In addition, existing methods only attribute the individual influence of features on bias while neglecting the \textit{intersectionality} among features.
-Quantifying intersectionality allows us to explain bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
-In our contribution, we propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features.
+Quantifying intersectionality allows us to interpret bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
+\paragraph{Contribution.} We propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features.
 In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods.
@@
 
-\section{Tools}
-	We develop following open-source tools for interpretable and fair machine learning.
-	\begin{itemize}
-		\item \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
-		\item  \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
-		\item \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for explaining group-based fairness metrics in machine learning.
-	\end{itemize}
-\section{Outline}
+\textbf{Tools:} We publish a python library \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for interpreting group-based fairness metrics in machine learning.
+
+\section{Thesis Outline}
 We organize the thesis in the following way.
@@
 In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
-We conclude this part by explaining group fairness metrics in Chapter~\ref{chapter:fif}.
+In the last part, we combine both interpretability and fairness in machine learning by interpreting group fairness metrics in Chapter~\ref{chapter:fif}.
 Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.

### 2023-02-28 chapters/interpretability/imli/incremental_learning.tex
addressing kuldeep's comments
minor edit

@@
 \label{interpretability_imli_sec:incremental_learning} 
-The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples in the training dataset and the number of clauses to be learned in a CNF formula.
-In this section, we discuss an incremental learning technique that along with the MaxSAT formulation can generate classification rules efficiently.
-Our incremental learning is built on two concepts: (i)
-mini-batch learning and (ii) iterative learning. We discuss both in detail in the following.
+To facilitate a scalable learning, we discuss the key technical contribution of this chapter: an incremental learning framework of interpretable classification rules based on MaxSAT.
+The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples $ n $ in the training dataset and the number of clauses $ k $  in the CNF classifier.
+To scale on large $ n $ and $ k $, our incremental learning is built on two concepts: (i) mini-batch learning and (ii) iterative learning.
 

### 2023-02-28 chapters/conclusion.tex
conclusion

@@
 
-To demonstrate the efficacy of our methods, we have constructed open-source tools and conducted experiments on real-world datasets in machine learning to evaluate interpretability and fairness.
+To demonstrate the efficacy of our methods, we have constructed open-source tools and conducted experiments on real-world datasets in machine learning.
 In the context of interpretable rule-based machine learning, we have developed an incremental learning framework, known as {\imli}, that scales classification to million-size datasets while maintaining competitive prediction accuracy and rule-size compared to existing rule-based classifiers.
@@
 
-In future, our research on fair and interpretable machine learning will focus on several areas.
-One immediate research question is to design a fairness improvement algorithm for machine learning classifiers using the fairness verifiers, {\justicia} and {\fvgm}, and the bias explanation framework, {\fairXplainer}.
-We also aim to extend formal fairness verification to more complex classifiers, such as random forests and neural networks.
-In addition, our future plans include deploying {\fairXplainer} on image classifiers, language models, and computer vision tasks.
-In a separate line of work, we will explore the application of incremental solving, as demonstrated in the context of interpretable classification learning by {\imli} and {\crr}, to other optimization problems, even beyond machine learning.
-To conclude the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the support of formal methods and automated reasoning.
+In future, our research on fair and interpretable machine learning will focus on several areas, which we elaborate in the following.
+
+\begin{itemize}
+	\item 
+	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains.
+We aim to design interpretable classifiers that are explainable by design in this thesis.
+We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers.
+Logic-based explanations, such as abductive explanations, Anchor, LORE, CORELS etc., are effective for this purpose.
+We plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can handle more expressive explanations.
+
+	\item \textbf{Fairness Auditing.} Machine learning has grown as a predictive system in automated decision making.
+However, quantifying and mitigating bias in these systems is an active area of research.
+There has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on.
+Thus, we need a fairness auditing framework with formal guarantees.
+One task of this framework is to categorize different definitions of fairness metrics in the literature.
+This would help stakeholders pick the right definition of fairness for their application.
+We plan to extend formal fairness verification beyond group fairness, such as individual fairness, causal fairness, and counterfactual fairness.
+We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks.
+Additionally, we aim to extend fairness verification beyond tabular data to more complex data such as images and texts.
+
+	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We want to make machine learning models fair and interpretable.
+We are interested in how these two goals relate to each other.
+We have some research questions for the future: (i) are interpretable models fair?
+(ii) how can we improve the fairness of an interpretable model?
+(iii) how can we use interpretability to enhance fairness?
+(iv) how can we optimize a classifier for both fairness and interpretability?
+Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability.
+We plan to explore more directions in the future.
+
+	\item \textbf{Efficiency through Incremental Solving.} We will also apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning.
+This can help us scale up without losing much accuracy.
+For example, we want to use incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets.
+To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.
+\end{itemize}
 

### 2023-02-28 chapters/conclusion.tex
future work

@@
 	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains.
-We aim to design interpretable classifiers that are explainable by design in this thesis.
+In this thesis, we have designed interpretable classifiers that are explainable by design.
 We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers.
-Logic-based explanations, such as abductive explanations, Anchor, LORE, CORELS etc., are effective for this purpose.
-We plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can handle more expressive explanations.
+In this context, logic-based explanations, such as abductive explanations, Anchor, LORE, CORELS etc., are effective for this purpose.
+Hence, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can handle more expressive explanations.
 

### 2023-03-02 chapters/background.tex
shortening

@@
 We express random variables in uppercase, and the valuation or an assignment of a random variable in lowercase.
+For example, $ \mathbf{X} = \{X_i\} $ is a set of random variables.
+The distribution of a random variable is $ X \in \mathcal{X} $ and its valuation  is $ X = x $.
 
@@
 %In contrast to the SAT problem, 
+Stochastic Boolean satisfiability (SSAT) was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}.
+SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
 
-The stochastic Boolean satisfiability (SSAT) problem~\cite{littman2001stochastic} is a counting analog of the SAT problem concerning with the probability of satisfaction of the formula.
-The SSAT problem computes the  probability of satisfaction of a CNF formula $\propformula$ defined on the ordered set of \textit{quantified} Boolean variables $ \boolset $.
+SSAT problem~\cite{littman2001stochastic} is a counting analog of the SAT problem concerning with the probability of satisfaction of the formula.
+SSAT problem computes the  probability of satisfaction of a CNF formula $\propformula$ defined on the ordered set of \textit{quantified} Boolean variables $ \boolset $.
 An SSAT formula is of the form
@@
 While the counting problem of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i =  \tau $ is NP-hard to approximate, there is a FPRAS solution for approximately counting the solutions of $ \sum_{i = 1}^ {|\boolset|} \varweight_i \bool_i \ge  \tau $~\cite{dyer2003approximate}.
-In this thesis, we are interested in computing the probability  $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $, referred to as \textit{stochastic subset-sum}  problem ({\stochastic}).
+We, in particular, are interested in computing the probability  $ \Pr[\sum_{i = 1}^ {|\boolset|} \varweight_i\bool_i \ge \tau] $, referred to as \textit{stochastic subset-sum}  problem ({\stochastic}).
 Details of {\stochastic} are in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}.

### 2023-03-02 chapters/fairness/fvgm/introduction.tex
shortening

@@
 
-	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} as our first step, because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+	Towards the goal of considering feature correlations in fairness verification, we particularly focus on the verification of \textit{linear classifiers} because of the significant attention on linear classifiers in fair algorithms \cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
 In this context, existing approaches suffer from limited scalability while verifying linear classifiers.

### 2023-03-02 chapters/fairness/justicia/introduction.tex
shortening

--- 2023-02-13 chapters/fairness/justicia/introduction.tex
@@
 \label{chapter:justicia}
-Machine learning is becoming the omnipresent technology of our time.
-Machine learning classifiers are being used for high-stake decisions such as college admission~\cite{martinez2021using}, recidivism prediction~\cite{tollenaar2013method}, job applications~\cite{ajunwa2016hiring} etc.
-Thus, human lives are now pervasively influenced by data, classifiers, and their inherent bias.
+In this chapter, we study formal verification of fairness in machine learning, where we verify the bias of a classifier given the probability distribution of features.
+To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed.
+%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
+%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
+%These papers operate for a specific Boolean sensitive feature.
+%These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
+Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
+In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}.
+Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.
+
+%\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
+Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of a classifer, which is \textit{scalable}, capable of \textit{handling compound sensitive groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.
+
+\paragraph{Contribution.} We propose to model verifying different fairness metrics as a SSAT problem.
+We primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
+Our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.
+
+%We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning \red{algorithm} using a unified scheme.
+%Our encodings not only allow us to compute for non-Boolean and compound sensitive features but also to scale significantly better than existing formal verifiers. 
+%We perform experimental analysis for multiple fairness metrics, datasets, and \red{algorithm}s to instantiate the efficiency and effectiveness of our approach.
+
+Our contributions are summarised below:
+\begin{itemize}
+	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
+	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
+	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
+	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
+%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
+\end{itemize}
+
+%It is worth remarking that significant advances in artificial intelligence bear testimony to the right choice of formulation, for example, formulation of planning as satisfiability (SAT)~\cite{kautz1992planning}.
+In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.
+
+We illustrate the contribution of this chapter using an example scenario.
 
@@
 Figure~\ref{fairness_justicia_fig:fair_example_dt}) does not predict based on the sensitive feature age.
-However, a fairness verifier would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}).
+However, a fairness verifier, such as {\justicia}, would verify that the decision tree outputs positive prediction to an individual above and below $40$ years with probabilities $0.18$ and $0.72$ respectively (Figure~\ref{fairness_justicia_fig:fair_example_probability}).
 This simple example demonstrates that even if a classifier does not explicitly learn to differentiate on the basis of a sensitive feature, it discriminates different age groups due to the utilitarian sense of accuracy that it tries to optimize.
@@
 
-\paragraph{Fairness in Machine Learning.} Statistical discrimination caused by classifiers has motivated researchers to formulate several definitions of fairness and develop associated algorithms to mitigate bias.
-In this chapter, we focus on a popular concept of fairness, known as group fairness.
-Existing group fairness metrics mostly belong to three categories: \textit{independence}, \textit{separation}, and \textit{sufficiency}~\cite{mehrabi2019survey}.
-Independence metrics, such as demographic parity, statistical parity, and group parity, try and ensure the outcomes of a classifier to be independent of the groups that the individuals belong to~\cite{feldman2015certifying,dwork2012fairness}.
-Separation metrics, such as equalized odds, define a classifer to be fair if the probability of getting the same outcomes for different groups are same, separately for data from each possible ground-truth class label~\cite{hardt2016equality}.
-Sufficiency metrics, such as counterfactual fairness, constrain the probability of outcomes to be independent of individual's sensitive data given their identical non-sensitive data~\cite{kusner2017counterfactual}.
-
-In Figure~\ref{fairness_justicia_fig:fair_example}, independence is satisfied if the probability of getting insurance is same for both the age groups.
-Separation is satisfied if the number of `actually' (ground-truth) ineligible and eligible people getting the insurance are same.
-Sufficiency is satisfied if the eligibility is independent of their age given their features are the same.
-Thus, we see that the metrics of fairness can be contradictory and complementary depending on the application and the data~\cite{corbett2018measure}.
-To this end, different algorithms have also been devised to ensure one or multiple of the fairness definitions.
-These algorithms try to rectify and mitigate the bias in the data and thus in the classifier in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
-
-\paragraph{Fairness Verifiers.} Due to the abundance of fairness metrics and difference in algorithms to achieve them, it has become necessary to verify different fairness metrics over datasets and algorithms.
-
-In order to verify fairness as a model property on a dataset, verifiers like \textit{FairSquare}~\cite{albarghouthi2017fairsquare} and \textit{VeriFair}~\cite{bastani2019probabilistic} have been proposed.
-%FairSquare verifies demographic parity and individual fairness as a numerical integration problem for a specific program semantics.
-%VeriFair translates fairness metrics to an enumeration problem of a specified Boolean syntax.
-%These papers operate for a specific Boolean sensitive feature.
-These verifiers are referred to as {\em probabilistic verifiers} owing to the fact that their inputs are a probability  distribution of the features in the dataset and a classifier of a suitable form, and their objective is to verify fairness with respect to the distribution and the classifier.
-Though FairSquare and VeriFair are robust and have asymptotic convergence guarantees, we observe that they scale up poorly with the size of inputs and also do not generalize to non-Boolean and compound sensitive features.
-In contrast to the probabilistic verifiers, another line of work, referred to as sample-based verifiers, has focused on the design of testing methodologies  on a given fixed data sample~\cite{galhotra2017fairness,aif360-oct-2018}.
-Since sample-based verifiers are dataset-specific, they generally do not provide robustness over the distribution.
-
-%\blue{Other papers: Probabilistic Verification of Fairness Properties via	Concentration, Verifying Individual Fairness in Machine Learning Models~\cite{john2020verifying}}
-Thus, a \textit{unified formal framework} to verify \textit{different fairness metrics} of a classifer, which is \textit{scalable}, capable of \textit{handling compound protected groups}, \textit{robust} with respect to the test data, and \textit{operational on real-life} datasets and fairness-enhancing algorithms, is missing in the literature.
-
-\paragraph{Contribution.} From this vantage point, \textit{we propose to model verifying different fairness metrics as a Stochastic Boolean Satisfiability (SSAT) problem}~\cite{littman2001stochastic}.
-SSAT was originally introduced by ~\cite{papadimitriou1985games} to model {\em games against nature}.
-In this work, we primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
-SSAT is a conceptual framework that has been employed to capture several fundamental problems in artificial intelligence such as the computation of maximum a posteriori (MAP) hypothesis~\cite{fremont2017maximum},  propositional probabilistic planning~\cite{majercik2007appssat},  and circuit verification~\cite{lee2018towards}.
-Furthermore, our choice of SSAT as a target formulation is motivated by the recent algorithmic progress that has yielded efficient SSAT tools~\cite{lee2017solving,lee2018solving}.
-
-%We formulate SSAT encodings of the fairness verification problems and two methods to evaluate them in order to verify independence and separation metrics for any supervised learning \red{algorithm} using a unified scheme.
-%Our encodings not only allow us to compute for non-Boolean and compound sensitive features but also to scale significantly better than existing formal verifiers. 
-%We perform experimental analysis for multiple fairness metrics, datasets, and \red{algorithm}s to instantiate the efficiency and effectiveness of our approach.
-
-Our contributions are summarised below:
-\begin{itemize}
-	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group and causal fairness metrics for different datasets and classifiers.
-	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
-	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
-	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
-%	\item We prove a finite-sample error bound on our estimated fairness metrics which is stronger than the existing asymptotic guarantees.
-\end{itemize}
-
-It is worth remarking that significant advances in artificial intelligence bear testimony to the right choice of formulation, for example, formulation of planning as satisfiability (SAT)~\cite{kautz1992planning}.
-In this context, we view that formulation of fairness as SSAT has potential to spur future work from both the modeling and encoding perspective as well as core algorithmic improvements in the underlying SSAT solvers.
-
 \iffalse

### 2023-03-02 chapters/interpretability/imli/introduction.tex
minor edit
shortening

@@
 In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}.
-In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers.
+In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers with a particular emphasis on the scalability in learning.
 In rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}.
@@
 Particularly,  we study and improve a \textit{maximum satisfiability} (MaxSAT)  solution for learning interpretable rule-based classifiers.
-Built on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules.
+Building on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules.
 The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples.

### 2023-03-02 chapters/interpretability/relaxed-cnf/experiment.tex
shortening

@@
-\section{Experiments}\label{interpretability_crr_sec:experiment}
+\section{Empirical Performance Analysis}\label{interpretability_crr_sec:experiment}
 We implement a prototype  of  $ \crr $ based on the  Python API for CPLEX  and conduct an extensive  empirical analysis to understand the behavior of $\crr$ on real-world instances.
@@
 %%\vspace{-3pt}
-\subsection{Experiment Methodology}
+\subsection{Experimental Setup}
 We perform experiments on a high-performance computer cluster, where each node consists of E$ 5$-$2690\text{ v}3 $ CPU with $ 24 $ cores, $ 96 $ GB of RAM.
@@
 %%\vspace{-9pt}
-\subsection{Results}
+\subsection{Experimental Results}
 In the following, we first discuss empirical results of rule-based classifiers, then extend analysis to non-rule-based classifiers, and finally discuss the effect of different choices of hyper-parameters.
 %%\vspace{-3pt}
-\subsubsection*{Performance Evaluation of {\crr} with Rule-based Classifiers:}
+\subsubsection{Performance Evaluation of {\crr} with Rule-based Classifiers}
 We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule-size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
@@
 
-\subsubsection*{Performance Evaluation of {\crr} with Non-rule-based Classifiers:}
+\subsubsection{Performance Evaluation of {\crr} with Non-rule-based Classifiers}
 
@@
 
-	\subsubsection*{Varying Model Parameters: }
+	\subsubsection{Varying Model Parameters}
 	\label{interpretability_crr_sec:model_parameters}

### 2023-03-02 chapters/interpretability/relaxed-cnf/introduction.tex
shortening

@@
 
-In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing on chapter~\ref{chapter:imli},
+In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing from chapter~\ref{chapter:imli},
 although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.

### 2023-03-02 chapters/introduction.tex
shortening

@@
 In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features.
-Finally, we interpret the sources of bias by formalizing and computing fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
+Finally, we combine both research themes: interpretability and fairness and propose a framework to interpret the sources of bias.
+In particular, we formalize and compute fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
 To summarize our thesis on interpretability and fairness in machine learning, we prioritize on improving the \emph{scalability} and the \emph{accuracy} of solutions\textemdash either or both of which are absent in prior works.
@@
 
-\textbf{Tools:} We publish a python library \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
+\textbf{Tool:} We publish a python library \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
 
@@
 These algorithms try to rectify and mitigate bias in three ways: \textit{pre-processing} the data~\cite{kamiran2012data,zemel2013learning,calmon2017optimized}, \textit{in-processing} the classifier~\cite{zhang2018mitigating}, and \textit{post-processing} the outcomes of a classifier~\cite{kamiran2012decision,hardt2016equality}.
-Researchers also study fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
+There is also study on fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
 In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting sources of unfairness.
@@
 
-\paragraph{Contributions.} In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
+\subsubsection{Formal Fairness Verification} In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
 Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
@@
 
-Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+\subsubsection{Tractable Fairness Verification With Feature Correlation} Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
 In the context of verifying the bias of linear classifiers, existing fairness verifiers suffer from two-fold limitations: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
@@
 
-\textbf{Tools:} We publish a python library \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
+\textbf{Tool:} We publish a python library \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
 
@@
 
-\textbf{Tools:} We publish a python library \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for interpreting group-based fairness metrics in machine learning.
+\textbf{Tool:} We publish a python library \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for interpreting group-based fairness metrics in machine learning.
 

### 2023-03-02 chapters/conclusion.tex
minor fix: section -> chapter

@@
 We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers.
-In this context, logic-based explanations, such as abductive explanations, Anchor, LORE, CORELS etc., are effective for this purpose.
-Hence, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can handle more expressive explanations.
+In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective for this purpose.
+Therefore, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.
 
-	\item \textbf{Fairness Auditing.} Machine learning has grown as a predictive system in automated decision making.
-However, quantifying and mitigating bias in these systems is an active area of research.
-There has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on.
-Thus, we need a fairness auditing framework with formal guarantees.
+	\item \textbf{Fairness Auditing.} Machine learning has demonstrated significant success as a predictive system in automated decision-making.
+However, quantifying and mitigating bias in these systems is still challenging, e.g.
+bias in generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}.
+As a result, there has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on.
+Thus, we need a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees.
 One task of this framework is to categorize different definitions of fairness metrics in the literature.
-This would help stakeholders pick the right definition of fairness for their application.
-We plan to extend formal fairness verification beyond group fairness, such as individual fairness, causal fairness, and counterfactual fairness.
+This would help stakeholders pick the right definition of fairness for their application~\cite{ruf2021towards}.
+We plan to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}.
 We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks.
@@
 
-	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We want to make machine learning models fair and interpretable.
+	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We aim to design fair and interpretable algorithms for  machine learning.
 We are interested in how these two goals relate to each other.
 We have some research questions for the future: (i) are interpretable models fair?
-(ii) how can we improve the fairness of an interpretable model?
+(ii) how can we improve the fairness of interpretable models?
 (iii) how can we use interpretability to enhance fairness?
@@
 
-	\item \textbf{Efficiency through Incremental Solving.} We will also apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning.
+	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning.
 This can help us scale up without losing much accuracy.
-For example, we want to use incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets.
+For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets.
 To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.

### 2023-03-02 chapters/fairness/fif/appendix.tex
added appendix
minor fix: section -> chapter

@@
-\appendix
 \chapter{Fairness Influence Functions}
@@
 Next, we apply variance decomposition (Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}) to estimate the influence of each $ \nonsensitive_{\mathbf{S}} $.
-From Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} in Section~\ref{fairness_fairXplainer_sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as
+From Equation~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation} in Chapter~\ref{fairness_fairXplainer_sec:preliminaries}, we obtain that the variance of $ B_{\mathbf{a}} $ can be decomposed as
 \[ \mathsf{Var}[B_\mathbf{a}] = \sum_{\mathbf{S} \subseteq [\numnonsensitive] \setminus \emptyset} V_{\mathbf{a}, \mathbf{S}},\] 
@@
 
-In this paper, we consider cubic splines with the basis function $ B_{r,4}(Z) $ that constitutes a B-spline of degree $ 3 $.
+In this chapter, we consider cubic splines with the basis function $ B_{r,4}(Z) $ that constitutes a B-spline of degree $ 3 $.
 This polynomial has $ C^2 $ continuity, i.e.
@@
 
-Following the classification setting and notations described in Section~\ref{chapter:preliminaries}, here we consider a binary classifier trained on dataset $\mathbf{D}\triangleq \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
+Following the classification setting and notations described in Chapter~\ref{chapter:preliminaries}, here we consider a binary classifier trained on dataset $\mathbf{D}\triangleq \{(\mathbf{x}^{(i)}, \mathbf{a}^{(i)}, y^{(i)})\}_{i=1}^n$ as $\alg: (\nonsensitive, \sensitive) \rightarrow \widehat{Y} $.
 $\widehat{Y} \in \{0,1\}$ and ${Y} \in \{0,1\}$ represents the predicted class and the true class for a data point $ (\nonsensitive, \sensitive) $ with sensitive features $\sensitive$ and non-sensitive features $\nonsensitive$.
@@
 This implies that the estimation of {\fairXplainer} is significantly accurate.
-Interestingly, error rarely occurs in {\fairXplainer} due to the degenerate cases mentioned in Section~\ref{fairness_fairXplainer_sec:fifs}.
+Interestingly, error rarely occurs in {\fairXplainer} due to the degenerate cases mentioned in Chapter~\ref{fairness_fairXplainer_sec:fifs}.
 In contrast, SHAP often computes a metric inaccurately, thereby showing the drawback of applying a local explanation approach for FIF computation.
@@
 We deploy a neural network ($ 3 $ hidden layers, each with $ 2 $ neurons, L$ 2 $ penalty regularization term as $ 10^{-5} $, a constant learning rate as $ 0.001 $) on different datasets, namely Adult and Titanic, and demonstrate the corresponding FIFs in Figures~\ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_adult} and \ref{fairness_fairXplainer_fig:individual_vs_intersectional_influence_titanic}, respectively.
-In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Section~\ref{fairness_fairXplainer_sec:experiments}.
+In all the figures, both individual and intersectional FIFs depict the sources of bias more clearly than individual FIFs alone, as argued in Chapter~\ref{fairness_fairXplainer_sec:experiments}.
 

### 2023-03-02 chapters/fairness/fif/framework.tex
minor fix: section -> chapter

@@
 In GSA, we observe that the variance of the output of a function can be attributed to the corresponding subset of input variables through variance decomposition (Eq.~\eqref{fairness_fairXplainer_eq:variance_decomposition_set_notation}).
-To leverage the power of GSA in fairness in machine learning, we first express the existing fairness metrics in terms of the variance of classifier's prediction in Section~\ref{fairness_fairXplainer_sec:metric_as_variance}.
-This allows us to formulate FIF by leveraging variance decomposition (Section~\ref{fairness_fairXplainer_sec:fif_formulation}).
+To leverage the power of GSA in fairness in machine learning, we first express the existing fairness metrics in terms of the variance of classifier's prediction in Chapter~\ref{fairness_fairXplainer_sec:metric_as_variance}.
+This allows us to formulate FIF by leveraging variance decomposition (Chapter~\ref{fairness_fairXplainer_sec:fif_formulation}).
 
@@
 Being able to naturally quantify the higher-order influences allow FIFs to interpret the sources of bias in a more fine-grained manner.
-We experimentally validate this in Section~\ref{fairness_fairXplainer_sec:experiments}.
+We experimentally validate this in Chapter~\ref{fairness_fairXplainer_sec:experiments}.
 
@@
 
-\textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Section~\ref{fairness_fairXplainer_sec:fairxplainer}.
+\textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.

### 2023-03-02 chapters/fairness/fif/introduction.tex
minor fix: section -> chapter

@@
 
-	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Section~\ref{fairness_fairXplainer_sec:fifs}).
+	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Chapter~\ref{fairness_fairXplainer_sec:fifs}).
 Our method is based on transforming existing fairness metrics into the difference of scaled conditional variances of classifier's prediction, which we then decompose using Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
@@
 
-%	\item \textbf{Formalism:} Given a classifier, a dataset, and a group fairness metric, we propose to estimate individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier induced by the features (Section~\ref{fairness_fairXplainer_sec:fifs}).	Our FIF formulation is based on representing existing fairness metrics as the difference of the scaled conditional variances of classifier's prediction, followed by a decomposition of variance according to Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
+%	\item \textbf{Formalism:} Given a classifier, a dataset, and a group fairness metric, we propose to estimate individual and intersectional \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) of features as a measure of contribution towards the bias of the classifier induced by the features (Chapter~\ref{fairness_fairXplainer_sec:fifs}).	Our FIF formulation is based on representing existing fairness metrics as the difference of the scaled conditional variances of classifier's prediction, followed by a decomposition of variance according to Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
 The presented FIF formulation results in different desired properties (Theorem~\ref{fairness_fairXplainer_thm:fif_property}): the efficiency property~\cite{begley2020explainability,lundberg2020explaining} states that the sum of FIFs of all subsets of features is equal to the bias of the classifier.
@@
 	\item \textbf{Algorithmic:} We propose a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric.
-The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}).
+The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}).
 Building on GSA\cite{saltelli2008global} techniques, {\fairXplainer} solves a local regression problem~\cite{loader2006local} based on cubic splines~\cite{li2010global} to decompose the variance of the classifier's prediction among all the subsets of features.
 
-%	\item \textbf{Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for estimating FIFs of  features for a given classifier, a dataset, and any group fairness metric, namely statistical parity, equalized odds, or predictive parity (Section~\ref{fairness_fairXplainer_sec:fairxplainer}).
+%	\item \textbf{Algorithmic:} We instantiate an algorithm, {\fairXplainer}, for estimating FIFs of  features for a given classifier, a dataset, and any group fairness metric, namely statistical parity, equalized odds, or predictive parity (Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}).
 The key idea is to import techniques from GSA\cite{saltelli2008global} to decompose the variance of the classifier's prediction among all  subsets of features by applying a local regression technique~\cite{loader2006local} based on cubic splines~\cite{li2010global}.
@@
 	\item \textbf{Experimental:} We evaluate  {\fairXplainer}  on a variety of real-world datasets and machine learning classifiers to demonstrate its efficiency in estimating the individual and intersectional FIFs of features.
-Our results show that {\fairXplainer} has a higher accuracy in approximating bias using estimated FIFs compared to existing methods (Section~\ref{fairness_fairXplainer_sec:experiments}).
+Our results show that {\fairXplainer} has a higher accuracy in approximating bias using estimated FIFs compared to existing methods (Chapter~\ref{fairness_fairXplainer_sec:experiments}).
 Our estimation of FIFs also shows a strong correlation with fairness interventions.
@@
 Fitness, being the root of DT$ 1 $, has a greater impact on statistical parity than income, which is at the second level of DT$ 1 $.
-Our method also reveals the joint effect of income and fitness on statistical parity, which prior methods do not account for (Section~\ref{fairness_fairXplainer_sec:related_work}).
+Our method also reveals the joint effect of income and fitness on statistical parity, which prior methods do not account for (Chapter~\ref{fairness_fairXplainer_sec:related_work}).
 The total of the FIFs of all features, $ 0.355 + 0.098 + 0.055 = 0.507 \approx 0.53 $, approximately matches the statistical parity of the classifier, providing a way to break down the bias among all feature subsets.
-Note that we discuss the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.
+Note that we discuss the estimation error of FIFs in Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.
 
@@
 Intuitively, fitness being at the root of DT$ 1 $ is more influential on statistical parity than income, which is situated at the second level of DT$ 1 $.
-Our method also reveals the joint (or intersectional) effect of income and fitness on statistical parity, which prior approaches do not consider (Section~\ref{fairness_fairXplainer_sec:related_work}).
-Finally, the sum of FIFs of features, $ 0.35 + 0.1 + 0.05 = 0.5 \approx 0.53 $, approximates\footnote{\red{We elaborate on the estimation error of FIFs in Section~\ref{fairness_fairXplainer_sec:fairxplainer}.}} the statistical parity of the classifier, thereby demonstrating a way to decompose the bias of the classifier among all the subsets of features.
+Our method also reveals the joint (or intersectional) effect of income and fitness on statistical parity, which prior approaches do not consider (Chapter~\ref{fairness_fairXplainer_sec:related_work}).
+Finally, the sum of FIFs of features, $ 0.35 + 0.1 + 0.05 = 0.5 \approx 0.53 $, approximates\footnote{\red{We elaborate on the estimation error of FIFs in Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.}} the statistical parity of the classifier, thereby demonstrating a way to decompose the bias of the classifier among all the subsets of features.
 
@@
 expectation) of all local explanations corresponding to all data points in a dataset.
-Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}).
+Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Chapter~\ref{fairness_fairXplainer_sec:experiments}).
 In addition, existing methods ignore the joint contribution of correlated features on bias.
@@
 They apply a global aggregation (such as expectation) of all local explanations corresponding to all data points in a dataset.
-Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Section~\ref{fairness_fairXplainer_sec:experiments}).
+Such a global aggregation of local explanations is often empirically justified and does not approximate bias accurately (Chapter~\ref{fairness_fairXplainer_sec:experiments}).
 In addition, features are often correlated in practical fairness tasks, and computing only individual FIFs ignores the joint contribution of multiple features on the bias of the classifier.

### 2023-03-02 chapters/fairness/fvgm/framework.tex
minor fix: section -> chapter

@@
 This allows us to avoid unnecessary repeated exploration of existential/universal variables in $ \mathsf{dp} $.
-Moreover, according to the remark in Section~\ref{fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}.
+Moreover, according to the remark in Chapter~\ref{fvgm_lm:property_s3p}, reordering $ \mathbf{B} $ still produces the same exact solution of {\stochastic}.
 Secondly, to reach the terminating condition of $ \mathsf{dp}(i, \tau) $ more frequently, we sort $ \mathbf{B} $ based on their weights\textemdash more specifically, within each cluster of random, existential, and universal variables.

### 2023-03-02 chapters/fairness/fvgm/framework2.tex
with new formatting
minor fix: section -> chapter

@@
 \input{chapters/fairness/fvgm/dp_illustration}	
-\subsection{{\stochastic} with Correlated Variables} 
+\subsection{Stochastic Subset Sum Problem with Correlated Variables} 
 \label{fvgm_sec:dp_with_BN}
-In {\stochastic} presented in Section~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent.
+In {\stochastic} presented in Chapter~\ref{fvgm_sec:stochastic_sum_set_sum}, we consider all  Boolean variables to be probabilistically independent.
 This independence assumption often leads to an \textit{inaccurate estimate} of the probability of positive prediction of the classifier because both sensitive and non-sensitive features can be correlated in practical fairness problems.
@@
 In practice, we achieve this by allowing no incoming edge on choice variables while learning $ \BN $ (ref.
-Section~\ref{fvgm_sec:experiments}).
+Chapter~\ref{fvgm_sec:experiments}).
 
@@
 For this enumeration, we do not store $ \mathsf{dp}(i, \tau, \mathbf{u}) $ in memory.
-However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Section~\ref{fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.
+However, for $ B_i \not \in \mathbf{V} $ that does not appear in the network, we instead compute $ \mathsf{dp}(i, \tau) $ and store it in memory as in Chapter~\ref{fvgm_sec:dp_formulation}, because $ B_i $ is not correlated with other variables.
 Lemma~\ref{fvgm_lm:complexity_dp_with_bn} presents the complexity of solving {\stochastic} with correlated variables, wherein unlike Lemma~\ref{fvgm_lemma:complexity_sss}, the  complexity differentiates based on variables in $ \mathbf{V} $ (exponential) and $ \mathbf{B}\setminus \mathbf{V} $ (pseudo-polynomial).
@@
 	We first separate analysis of space and time complexity for variables in $ \mathbf{V} $ and in $\mathbf{B}\setminus \mathbf{V} $.
-For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Section~\ref{fvgm_sec:dp_with_BN}.
+For each Boolean variable in $ \mathbf{V} $, we enumerate all assignments, which has time complexity of $ 2^{|\mathbf{V}|} $ and there is no space complexity as discussed in Chapter~\ref{fvgm_sec:dp_with_BN}.
 
@@
 	In Figure~\ref{fvgm_fig:example_BN}, we enumerate all  assignment of $ P $ and $ Q $ to  incorporate all conditional probabilities of $ Q $ given $ P $.
-We, however, observe that the dynamic programming solution in Section~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $.
+We, however, observe that the dynamic programming solution in Chapter~\ref{fvgm_sec:dp_formulation} still prunes search space for variables that do not appear in $ \mathbf{V} $, such as $ \{R, S\} $.
 Hence following the calculation in Figure~\ref{fvgm_fig:example_BN}, we obtain the maximum probability of positive prediction of the classifier as $ 0.65 $ for $ P = 1 $.

### 2023-03-02 chapters/fairness/justicia/introduction.tex
minor fix: section -> chapter

@@
 \label{chapter:justicia}
-In this chapter, we study formal verification of fairness in machine learning, where we verify the bias of a classifier given the probability distribution of features.
+In this chapter, we study formal fairness verification problem in machine learning, where we verify the bias of a classifier given the probability distribution of features.
 To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed.
@@
 	\normalfont
-	Let us consider an example (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60).
+	Let us consider a classification problem (Figure~\ref{fairness_justicia_fig:fair_example}) of deciding the eligibility for health insurance depending on the fitness and income of individuals of different age groups (20-40 and 40-60).
 Typically, incomes of individuals increase as their ages increase while their fitness deteriorates (Figure~\ref{fairness_justicia_fig:fair_example_dependency}).

### 2023-03-02 chapters/interpretability/imli/baseline.tex
minor fix: section -> chapter

@@
 In this section, we revisit~\cite{malioutov2018mlic} by discussing a MaxSAT-based learning framework for an interpretable rule-based classifier, particularly a CNF classifier $ \Rule $. 
-We first describe the decision variables in Section~\ref{interpretability_imli_sec:variables} and present the MaxSAT encoding  in Section~\ref{interpretability_imli_sec:encoding}.
+We first describe the decision variables in Chapter~\ref{interpretability_imli_sec:variables} and present the MaxSAT encoding  in Chapter~\ref{interpretability_imli_sec:encoding}.
 The MaxSAT formulation assumes binary features as input.
 We
-conclude this section by learning $ \Rule $ with non-binary features in Section~\ref{interpretability_imli_sec:non-binary} and discussing more flexible interpretability constraints of $ \Rule $ in Section~\ref{interpretability_imli_sec:complex_interpretability_objectives}.
+conclude this section by learning $ \Rule $ with non-binary features in Chapter~\ref{interpretability_imli_sec:non-binary} and discussing more flexible interpretability constraints of $ \Rule $ in Chapter~\ref{interpretability_imli_sec:complex_interpretability_objectives}.
 
@@
 \begin{itemize}
-	\item \textbf{Soft clauses for maximizing training accuracy:} For each training sample, {\imli} constructs a soft unit\footnote{A unit clause has a single literal.} clause $ \neg \noise_l $ to account for a penalty for misclassification.
+	\item \textbf{Soft clauses for maximizing training accuracy:} For each training sample, we construct a soft unit\footnote{A unit clause has a single literal.} clause $ \neg \noise_l $ to account for a penalty for misclassification.
 Since the penalty for misclassification of a sample is $ 1 $  in Eq.~\eqref{interpretability_imli_eq:obj},  the weight of this soft clause is also $ 1 $.
@@
 
-	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, {\imli} tries to learn a classifier with as few literals as possible.
-Hence, for each feature variable $ \bool_j^i $, {\imli} constructs a unit clause as $ \neg \bool_j^i $.
+	\item \textbf{Soft clauses for minimizing rule-sparsity:} To favor rule-sparsity, we learn a classifier with as few literals as possible.
+Hence, for each feature variable $ \bool_j^i $, we construct a unit clause as $ \neg \bool_j^i $.
 Similar to training accuracy, the weight for this clause is derived as $ \lambda $ from Eq.~\eqref{interpretability_imli_eq:obj}.
@@
 	\item \textbf{Hard clauses for encoding constraints:} In a MaxSAT problem, constraints that must be satisfied are encoded as hard clauses.
-In {\imli}, we have a learning constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa.
+In rule-based classification, we have a learning constraint that, if the error variable is false, the associated sample must be correctly classified, and vice-versa.
 Let  $\boolset_i= \{\bool^i_{j} \mid j \in \{1,\dots,m \}\}$ be a vector of feature variables corresponding to the $ i^\text{th} $  clause in $ \mathcal{R} $.
@@
 \label{interpretability_imli_sec:complex_interpretability_objectives}
-{\imli} can be extended to consider more flexible interpretability objectives than the simplified one in Eq.~\eqref{interpretability_imli_eq:obj}.
+In interpretable classification rules, we can consider more flexible interpretability objectives than the simplified one in Eq.~\eqref{interpretability_imli_eq:obj}.
 In Eq.~\eqref{interpretability_imli_eq:obj}, we prioritize all features equally by providing the same weight to the clause $ S_j^i $ for all  values of $ i $ and $ j $.
 In practice, users may prefer rules containing certain features.
-In {\imli}, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $.
+In the MaxSAT formulation, such an extension can be achieved by modifying the weight function and/or the definition of  $ S_j^i $.
 For example, to constrain the classifier to never include a feature, the weight of the clause $ S_j^i := \neg \bool_j^i $ can be set to $ \infty $.
@@
 In either case, we convert this constraint to CNF using pseudo-Boolean to CNF translation~\cite{philipp2015pblib}.
-Thus, {\imli} allows us to consider varied interpretability constraints by only modifying the MaxSAT query without requiring changes in the MaxSAT solving.
-Therefore, the \emph{separation between modeling and solving} turns out to be the key strength of {\imli}.
+Thus, the MaxSAT formulation allows us to consider varied interpretability constraints by only modifying the MaxSAT query without requiring changes in the MaxSAT solving.
+Therefore, the \emph{separation between modeling and solving} turns out to be the key strength of the presented MaxSAT formulation.

### 2023-03-02 chapters/interpretability/imli/incremental_learning.tex
minor fix: section -> chapter

@@
-\section{Incremental Learning with MaxSAT}
+\section{Incremental Learning of Interpretable Classification Rules}
 \label{interpretability_imli_sec:incremental_learning} 
-To facilitate a scalable learning, we discuss the key technical contribution of this chapter: an incremental learning framework of interpretable classification rules based on MaxSAT.
-The complexity of the MaxSAT query in Section~\ref{interpretability_imli_sec:encoding} increases with the number of samples $ n $ in the training dataset and the number of clauses $ k $  in the CNF classifier.
+To facilitate a scalable learning, we discuss the key technical contribution of this chapter, {\imli}, a an incremental learning framework of interpretable classification rules based on MaxSAT.
+The complexity of the MaxSAT query in Chapter~\ref{interpretability_imli_sec:encoding} increases with the number of samples $ n $ in the training dataset and the number of clauses $ k $  in the CNF classifier.
 To scale on large $ n $ and $ k $, our incremental learning is built on two concepts: (i) mini-batch learning and (ii) iterative learning.
@@
 We measure the performance of $ \Rule $ on the full training data $ \dataset $ using a weighted combination of classification errors and rule-size.
-In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\dataset| +\lambda |\mathcal{R}| $ on the training data $ \dataset $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Section~\ref{interpretability_imli_sec:problem}.
+In particular, we compute a combined loss function $ \mathsf{loss}(\Rule) \triangleq |\mathcal{E}_\dataset| +\lambda |\mathcal{R}| $ on the training data $ \dataset $, which is indeed the value of the objective function that we minimize in the non-incremental learning in Chapter~\ref{interpretability_imli_sec:problem}.
 Additionally, we  discard the current classifier $ \Rule $ when the loss does not decrease ($ \mathsf{loss}(\Rule) > \mathsf{loss}(\Rule') $).

### 2023-03-02 chapters/interpretability/imli/introduction.tex
minor fix: section -> chapter

@@
 Building on MaxSAT, ~\cite{malioutov2018mlic} considers a learning framework to jointly optimize the accuracy and the size of classification rules.
-The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples.
+The said approach constructs and solves a MaxSAT query of $ \mathcal{O}(kn) $  size (number of clauses) to learn a $ k $-clause CNF formula on a dataset containing $ n $ samples (Chapter ~\ref{interpretability_imli_sec:encoding}).
 Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $.
@@
 Our incremental learning is an integration of mini-batch learning and iterative rule-learning, which are studied separately in classical learning problems.
-In the presented incremental approach, {\imli} learns a $ k $-clause CNF formula using an iterative separate-and-conquer algorithm, where in each iteration, a single clause is learned by covering a part of the training data.
+In the presented incremental approach, {\imli} learns a $ k $-clause CNF formula using an iterative separate-and-conquer algorithm, where in each iteration, a single clause is learned by covering a part of the training data (Chapter~\ref{interpretability_imli_sec:incremental_learning}).
 Furthermore, to efficiently learn a single clause in a CNF, {\imli} relies on mini-batch learning, where it solves a sequence of smaller MaxSAT queries corresponding to each mini-batch.

### 2023-03-02 chapters/interpretability/imli/problem_formulation.tex
connecting chapters
minor fix: section -> chapter

@@
 For instance, a rule involving fewer literals is highly interpretable.
-In this work, we minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $.
+In this chapter, we minimize the total number of literals in all clauses, which motivates us to  find $ \Rule  $ with minimum  $ |\Rule| $.
 Let $ \Rule $ classify all samples correctly during training.
@@
 Higher  values of $ \lambda $ generate a rule with a smaller rule-size but of more training errors, and vice-versa.
-Thus, $ \lambda $ can be tuned to trade-off between accuracy and interpretability for a rule-based classifier, which we experiment extensively in Section~\ref{interpretability_imli_sec:experiments}.
+Thus, $ \lambda $ can be tuned to trade-off between accuracy and interpretability for a rule-based classifier, which we experiment extensively in Chapter~\ref{interpretability_imli_sec:experiments}.

### 2023-03-02 chapters/interpretability/relaxed-cnf/experiment.tex
minor fix: section -> chapter

@@
 Therefore, relaxed-CNF rules improve upon CNF rules in terms of both prediction accuracy and rule size in the majority of the datasets.
-In this context, {\crr} provides a trade-off between accuracy and rule-size depending on the choice of hyper-parameters and the experimental results are discussed later in Section~\ref{interpretability_crr_sec:model_parameters}.
+In this context, {\crr} provides a trade-off between accuracy and rule-size depending on the choice of hyper-parameters and the experimental results are discussed later in Chapter~\ref{interpretability_crr_sec:model_parameters}.
 We then compare relaxed-CNF rules with DNF rules generated by BRS and find that relaxed-CNF rules outperform DNF rules with respect to prediction accuracy in several datasets.

### 2023-03-02 chapters/interpretability/relaxed-cnf/ilp_formulation.tex
minor fix: section -> chapter

@@
 We organize the rest of this section as follows.
-We  discuss the decision variables in Section ~\ref{interpretability_crr_sec:variables},  the constraints and the linear programming relaxation in Section~\ref{interpretability_crr_sec:ilp_query}, the incremental learning in Section~\ref{interpretability_crr_sec:inc_ilp}, feature discretization in Section~\ref{interpretability_crr_sec:discretization}, and adaptation of {\crr} for learning other classification rules in Section~\ref{interpretability_crr_sec:other_rules}.
-% We conclude this section by describing how {\crr} can be adjusted to learn interpretable rules in other logical forms in Section~\ref{interpretability_crr_sec:other_format}.
+We  discuss the decision variables in Chapter~\ref{interpretability_crr_sec:variables},  the constraints and the linear programming relaxation in Chapter~\ref{interpretability_crr_sec:ilp_query}, the incremental learning in Chapter~\ref{interpretability_crr_sec:inc_ilp}, feature discretization in Chapter~\ref{interpretability_crr_sec:discretization}, and adaptation of {\crr} for learning other classification rules in Chapter~\ref{interpretability_crr_sec:other_rules}.
+% We conclude this section by describing how {\crr} can be adjusted to learn interpretable rules in other logical forms in Chapter~\ref{interpretability_crr_sec:other_format}.
 
@@
 
-In Section~\ref{interpretability_crr_sec:ilp_query}, we  present an ILP formulation for learning relaxed-CNF rules and then discussed the relaxation to the integer constraints in the MILP formulation to make the approach computationally tractable.
-However, each integer constraints in the formulation cannot be relaxed, as discussed  in Section~\ref{interpretability_crr_sec:relaxation}.
+In Chapter~\ref{interpretability_crr_sec:ilp_query}, we  present an ILP formulation for learning relaxed-CNF rules and then discussed the relaxation to the integer constraints in the MILP formulation to make the approach computationally tractable.
+However, each integer constraints in the formulation cannot be relaxed, as discussed  in Chapter~\ref{interpretability_crr_sec:relaxation}.
 Thus we require an improved learning technique to achieve scalability.

### 2023-03-02 chapters/interpretability/relaxed-cnf/introduction.tex
minor fix: section -> chapter

@@
 
-We now provide an introduction to relaxed-CNF rules.
-In interpretable rule-based classification, the simplest logical rules are single level-rules: ORs or ANDs of Boolean \textit{literals}, where each literal denotes either a Boolean input feature or its negation.
+\paragraph{Relaxed-CNF: An Informal Introduction.} In interpretable rule-based classification, the simplest logical rules are single level-rules: ORs or ANDs of Boolean \textit{literals}, where each literal denotes either a Boolean input feature or its negation.
 A \emph{clause} is a collection of $ N $ literals connected by  OR/AND.

### 2023-03-02 chapters/introduction.tex
minor fix: section -> chapter

@@
 \subsubsection{Formal Fairness Verification} In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
-Based on stochastic satisfiability (SSAT)\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
+Based on stochastic satisfiability (SSAT)~\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
 {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.

### 2023-03-02 chapters/conclusion.tex
outline in imli

@@
 We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers.
-In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective for this purpose.
+In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective in generating post-hoc explanations of black-box classifiers.
 Therefore, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.
@@
 This would help stakeholders pick the right definition of fairness for their application~\cite{ruf2021towards}.
-We plan to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}.
+In addition, we aim to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}.
 We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks.

### 2023-03-02 chapters/interpretability/imli/introduction.tex
outline in imli

@@
 
-%We organize the rest of the chapter as follows.
-We discuss related literature in Section~\ref{interpretability_imli_sec:related} and preliminaries in Section~\ref{interpretability_imli_sec:preliminaries}.
-In Section~\ref{interpretability_imli_sec:problem}, we formally define our learning problem and provide a MaxSAT-based solution in Section~\ref{interpretability_imli_sec:baseline}.
-We improve the MaxSAT-based formulation through an incremental learning technique in Section~\ref{interpretability_imli_sec:incremental_learning}.
-In Section~\ref{interpretability_imli_sec:application}, we apply {\imli} in learning different interpretable classifiers.
-We then discuss our experimental results in Section~\ref{interpretability_imli_sec:experiments} and conclude in Section~\ref{interpretability_imli_sec:conclusion}.
+We organize the rest of the chapter as follows.
+We discuss related literature in Chapter~\ref{interpretability_imli_sec:related}.
+In Chapter~\ref{interpretability_imli_sec:problem}, we formally define interpretable rule-based classification problem and discuss a MaxSAT-based solution proposed by~\cite{malioutov2018mlic} in Chapter~\ref{interpretability_imli_sec:baseline}.
+In Chapter~\ref{interpretability_imli_sec:incremental_learning}, we discuss the key technical contribution of this chapter, {\imli}, an improved MaxSAT-based learning formulation based on incremental solving.
+In Chapter~\ref{interpretability_imli_sec:application}, we apply {\imli} in learning different interpretable classifiers.
+We then discuss our experimental results in Chapter~\ref{interpretability_imli_sec:experiments} and conclude in Chapter~\ref{interpretability_imli_sec:conclusion}.

### 2023-03-02 chapters/introduction.tex
minor

@@
 
-\subsubsection{Tractable Fairness Verification With Feature Correlation} Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
+\subsubsection{Tractable Fairness Verification with Feature Correlation} Linear classifiers have attracted significant attention from researchers in the context of designing and testing prototype fairness enhancing and attack algorithms~\cite{pleiss2017fairness,zafar2017fairness,dressel2018accuracy, john2020verifying}.
 In the context of verifying the bias of linear classifiers, existing fairness verifiers suffer from two-fold limitations: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.

### 2023-03-02 doctoral_consortium/main.tex
minor

@@
 
-\paragraph{Tractable Fairness Verification With Feature Correlation.} Linear classifiers have received significant attention from researchers in the context of fair algorithms.
+\paragraph{Tractable Fairness Verification with Feature Correlation.} Linear classifiers have received significant attention from researchers in the context of fair algorithms.
 Existing fairness verifiers suffer from two-fold limitations while verifying linear classifiers: (i) poor scalability due to the use of SSAT, SMT, or sampling-based techniques, and (ii) limited accuracy due to ignoring feature correlations.

### 2023-03-03 chapters/abstract.tex
addressing Kuldeep's comments

@@
 As such, interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability in large datasets.
-To this end, we propose an incremental learning framework, called {\imli}, which applies an iterative solving of  maximum satisfiability (MaxSAT) queries in mini-batch learning and enables classification on million-size datasets.
+To this end, we discuss an incremental learning framework, called {\imli}, which applies an iterative solving of  maximum satisfiability (MaxSAT) queries in mini-batch learning and enables classification on million-size datasets.
 Although being interpretable, rule-based classifiers often suffer from limited expressiveness, for example,  classifiers based on propositional logic.
-To learn more expressible yet interpretable classification rules, we propose a relaxation of classifiers based on logical formulas.
-For learning relaxed rule-based classifiers, we propose  an efficient learning framework, called {\crr},  building on incremental learning and mixed integer linear programming (MILP).
+To learn more expressible yet interpretable classification rules, we discuss a relaxation of classifiers based on logical formulas.
+For learning relaxed rule-based classifiers, we discuss an efficient learning framework, called {\crr},  building on incremental learning and mixed integer linear programming (MILP).
 {\crr} obtains higher accuracy yet less rule-size than existing interpretable classifiers.
@@
 	Fairness in machine learning centers on quantifying and mitigating the bias or unfairness of machine learning classifiers.
-In the presence of multiple fairness metrics for quantifying bias, we propose a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features.
+In the presence of multiple fairness metrics for quantifying bias, we discuss a probabilistic fairness verifier, called {\justicia}, with the goal of formally verifying the bias of a classifier given the probability distribution of features.
 Building on  stochastic satisfiability (SSAT), {\justicia} improves the scalability of verification; and unlike prior approaches, {\justicia} verifies compound sensitive groups combining multiple sensitive features.
@@
 	Fairness metrics globally quantify bias, but do not detect or interpret its sources.
-To interpret group-based fairness metrics, we propose fairness influence functions (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier.
+To interpret group-based fairness metrics, we discuss fairness influence function (FIF) with an aim of quantifying the influence of individual features and the intersection of multiple features  on the bias of a classifier.
 FIF  interprets fairness by revealing potential individual or intersectional features attributing highly to the bias.
-Building on global sensitivity analysis, we propose an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.
+Building on global sensitivity analysis, we discuss an algorithm, called {\fairXplainer}, for estimating the FIFs of features, resulting in a better approximation of bias based on FIFs and a higher correlation of FIFs with fairness interventions.
 

### 2023-03-03 chapters/background.tex
addressing Kuldeep's comments

@@
 
-\subsection{Propositional Satisfiability}
+\subsection{Propositional Satisfiability (SAT)}
 
@@
 Let $ \mathds{1}[\text{true}]=1 $ and $ \mathds{1}[\text{false}]=0 $.
-Motivated by Boolean cardinality constraints, we propose relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$.
+Motivated by Boolean cardinality constraints, we introduce relaxed-CNF formulas, where in addition to $\propformula$, we have two more parameters $\card_c$ and $\card_l$.
 We say that $(\propformula,\card_c, \card_l)$ is in relaxed-CNF and $ \sigma^* $ is its satisfying assignment if and only if $\sigma^* \models (\propformula,\card_c, \card_l)$ whenever $\sum_{i=1}^k \mathds{1}[\sigma^* \models (\propclause_i,\card_l)] \ge \card_c$, where $\sigma^* \models (\propclause_i, \card_l)$ if and only if $\sum_{V \in \propclause_i} \mathds{1}[\sigma^* \models V] \geq \card_l$.
@@
 
-\subsection{MaxSAT}
+\subsection{Maximum Satisfiability (MaxSAT)}
 
@@
 \label{chapter_interpretability_preliminaries}
-We consider a standard binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
+We consider a  binary classification problem where the dataset $ \dataset $ is a collection of $ n $ samples  $\{(\mathbf{x}^{(i)}, y^{(i)})\}_{i=1}^n$ generated from an underlying distribution $\mathcal{D}$.
 Here, the feature valuation vector $ \mathbf{x} = [x_1, \dots, x_{m}] \in \{0, 1\}^m $ is a vector of $ m $ Boolean features with $ y \in \{0,1\} $ being the binary class label.

### 2023-03-03 chapters/conclusion.tex
addressing Kuldeep's comments

@@
 On the other hand, {\fvgm} takes correlated features into account and is capable of verifying the fairness of linear classifiers with higher scalability and accuracy than previous verifiers.
-Additionally, we propose a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features.
+Additionally, we discuss a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features.
 Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.

### 2023-03-03 chapters/fairness/fif/appendix.tex
addressing Kuldeep's comments

@@
 
-\subsection{Accurate Approximation of Equalized Odds and Predictive Parity using FIFs.}
+\subsection{Accuracy: Equalized Odds \& Predictive Parity via FIFs.}
 The accuracy of approximating equalized odds and predictive parity for {\fairXplainer} and SHAP is compared in Table~\ref{fairness_fairXplainer_tab:accuracy_eo_pp}.
@@
 
-\subsection{Execution Time for FIF Estimation of Equalized Odds and Predictive Parity}
+\subsection{Execution Time: Equalized Odds \& Predictive Parity via FIFs}
 In Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot_eo_pp}, we demonstrate the execution time of different methods in estimating FIFs of equalized odds and predictive parity in cactus plots.

### 2023-03-03 chapters/fairness/fvgm/framework2.tex
addressing Kuldeep's comments

@@
 
-	\subsection{Fairness Verification with Computed Probability of Positive Prediction} 
+	\subsection{Fairness Verification using Probability of Positive Prediction} 
 	Given a classifier $\mathcal{M}$, a  distribution $\mathcal{D}$, and a fairness metric $f$, verifying whether a classifier is $\epsilon$-fair for $\epsilon \in [0,1]$ is equivalent to computing $\mathds{1}[f(\mathcal{M}|\mathcal{D})\leq \epsilon]$.

### 2023-03-03 chapters/fairness/justicia/discussion.tex
addressing kuldeep's comments
shortening
addressing Kuldeep's comments

--- 2022-06-07 chapters/fairness/justicia/discussion.tex
@@
 Existing fairness verifiers, however, are not scalable, accurate, and extendable to non-Boolean sensitive features.
-We propose a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups.
+We discuss a stochastic SAT-based approach, {\justicia}, that formally verifies multiple group and causal fairness metrics for different classifiers and distributions for compound sensitive groups.
 Experimental evaluations demonstrate that {\justicia} achieves \textit{higher accuracy} and \textit{scalability} in comparison to the state-of-the-art verifiers, FairSquare and VeriFair, while yielding \textit{higher robustness} than the sample-based tools, such as AIF360.
+A limitation of {\justicia} is the lack of consideration of feature correlations in the SSAT-based verifier, which we alleviate in the next chapter.
 
-Our work opens up several new directions of research.
-One direction is to develop SSAT models and verifiers for popular classifiers like Deep networks and SVMs.
+%Our work opens up several new directions of research.
+One direction is to develop SSAT models and verifiers for popular classifiers like deep networks and SVMs.
 Other direction is to develop SSAT solvers that can accommodate continuous variables and conditional probabilities by design.
+%Finally, an interesting direction is to enable {\justicia} to consider correlated feature distribution, which we pursue in the next chapter.
 %Of particular interest is the highlighted need for design of efficient SSAT solvers given that SSAT can be viewed as a natural target problem in the context of development of formal verification techniques for fairness.

### 2023-03-03 chapters/fairness/justicia/framework.tex
addressing Kuldeep's comments

@@
 As a result, the enumeration approach may suffer from scalability issues due to the exponential number of calls to the SSAT solver.
-To this end, we propose an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier.
+To this end, we discuss an efficient SSAT encoding, where we make two SSAT calls, one for inferring the \emph{the most favored sensitive group} with the maximum conditional probability of positive prediction of the classifier and another for inferring \emph{the least favored sensitive group} with the minimum conditional probability of positive prediction of the same classifier.
 As discussed above, these two probabilities are sufficient to measure different group and causal fairness metrics.

### 2023-03-03 chapters/fairness/justicia/introduction.tex
outline in imli
addressing Kuldeep's comments

@@
 \label{chapter:justicia}
-In this chapter, we study formal fairness verification problem in machine learning, where we verify the bias of a classifier given the probability distribution of features.
+We discuss formal fairness verification problem in machine learning where we verify the bias of a classifier given the probability distribution of features.
 To verify fairness as a model property, several probabilistic fairness verifiers such as FairSquare~\cite{albarghouthi2017fairsquare} and VeriFair~\cite{bastani2019probabilistic} have been proposed.
@@
 
-\paragraph{Contribution.} We propose to model verifying different fairness metrics as a SSAT problem.
+\paragraph{Contribution.} We discuss model verifying different fairness metrics as a SSAT problem.
 We primarily focus on reductions to the exist-random quantified fragment of SSAT, which is also known as E-MAJSAT~\cite{littman2001stochastic}.
@@
 \begin{itemize}
-	\item We propose a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
+	\item We discuss a unified SSAT-based approach, {\justicia}, to verify independence and separation metrics of group  fairness metrics for different datasets and classifiers.
 	%\item \blue{{\justicia} measures fairness metrics for pre-, in-, and post-processing \red{algorithm}s with respect to the data  generating distribution}.
-	\item Unlike previously proposed formal probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
+	\item Unlike earlier probabilistic verifiers, namely FairSquare and VeriFair, {\justicia} verifies fairness for compound and non-Boolean sensitive features.%, and also \red{separation metrics}.
 	\item Our experiments validate that our method is more accurate and scalable than the probabilistic verifiers, such as FairSquare and VeriFair, and more robust than the sample-based empirical verifiers, such as AIF360.
@@
 			\end{tikzpicture}
-		}\label{fairness_justicia_fig:fair_example_dependency}}\hfill
+		}\label{fairness_justicia_fig:fair_example_dependency}}\hfil
 	\subfloat[Age-dependent distributions of non-sensitive features]{
@@
 		\end{tikzpicture}}
-		\label{fairness_justicia_fig:fair_example_dt}}\hfill
+		\label{fairness_justicia_fig:fair_example_dt}}\hfil
 	\subfloat[Group specific conditional probability of the classifier's positive prediction]{

### 2023-03-03 chapters/interpretability/imli/appendix.tex
addressing Kuldeep's comments

@@
 
-\section{Experimental Comparison between Incremental vs.\ Non-incremental  Encoding}
+\section{Performance Comparison: Incremental vs.\ Non-incremental  Encoding}
 We compare the performance of two encoding techniques, na\"ive non-incremental encoding and efficient incremental encoding in {\imli}, for learning interpretable CNF classification rules in datasets of various sizes.

### 2023-03-03 chapters/interpretability/imli/conclusion.tex
shortening
addressing Kuldeep's comments

--- 2022-01-27 chapters/interpretability/imli/conclusion.tex
@@
 Considering the growing demand for interpretable models, it is challenging to design learning frameworks that satisfy all aspects: being accurate, interpretable, and scalable in practical classification tasks.
-In this paper, we have proposed a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas.
+In this chapter, we discuss a MaxSAT-based framework {\imli} for learning interpretable rule-based classifiers expressible in CNF formulas.
 {\imli} is built on efficient integration of incremental learning, specifically mini-batch and iterative learning, with MaxSAT-based formulation.
@@
 Finally, {\imli}  generates other popular interpretable classifiers such as decision lists and decision sets using the same framework.
+In the next chapter, we leverage the incremental learning in {\imli} to learn more expressible yet interpretable classification rules.
 

### 2023-03-03 chapters/interpretability/imli/experiment.tex
highlevel pass
addressing Kuldeep's comments

@@
 \subsection{Experimental Setup}
-We implement a prototype of {\imli} in Python to evaluate the performance of the proposed MaxSAT-based formulation for learning classification rules.
+We implement a prototype of {\imli} in Python to evaluate the performance of the  MaxSAT-based formulation for learning classification rules.
 To implement {\imli}, we deploy a state-of-the-art MaxSAT solver Open-WBO~\cite{martins2014open}, which returns the current best solution upon reaching a timeout.
@@
 
-\begin{table*}[!t]
+\begin{table*}
 	\centering
@@
 	%	\setlength{\tabcolsep}{.2em}
-	\small
+	\small{
 	\begin{tabular}{lrrrrrrrrrrrr}
@@
 	\end{tabular}
-
+	}
 \end{table*}
@@
 
-\begin{figure}[!t]
+\begin{figure}
 
@@
 
-\begin{table*}[!t]        
+\begin{table*}     
 	\centering
@@
 
-\begin{table*}[!t]        
+\begin{table*}
 	\centering

### 2023-03-03 chapters/interpretability/imli/incremental_learning.tex
addressing Kuldeep's comments

@@
 For that, we consider a soft constraint such that $ \Rule $ does not \emph{differ much} from $ \Rule' $ while training on the current batch.
-Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We propose Hamming distance heuristics in {\imli} with the assumption that the probability distribution of features remains same across batches.
+Thus, we hypothesize that by constraining $ \Rule $ to be syntactically similar to $ \Rule' $, it is possible to generalize well; because samples in all batches originate from the same distribution\footnote{We apply Hamming distance heuristics in {\imli} with the assumption that the probability distribution of features remains same across batches.
 One way to account for distribution shifts is to consider last $ p $  $ (> 1) $ batches instead of the (single) previous batch in the objective function in mini-batch learning.

### 2023-03-03 chapters/interpretability/imli/introduction.tex
addressing Kuldeep's comments

@@
 In safety-critical domains, a surge of interest has been observed in designing interpretable, robust, and fair models instead of merely focusing on an accuracy-centric learning objective~\cite{bhagoji2018enhancing,doshi2017towards,du2019techniques,hancox2020robustness,holstein2019improving,mehrabi2021survey,molnar2020interpretable,murdoch2019interpretable}.
-In this chapter, we propose an efficient learning framework for interpretable rule-based classifiers with a particular emphasis on the scalability in learning.
+We discuss an efficient learning framework for interpretable rule-based classifiers with a particular emphasis on the scalability in learning.
 In rule-based classifiers, rules governing the prediction are explicit (by design) in contrast to black-box classifiers~\cite{rudin2019stop}.
@@
 Naturally, the na\"ive formulation cannot scale to large values of $ n $ and $ k $.
-To scale learning to large datasets, we propose an incremental learning technique along with the MaxSAT formulation.
+To scale learning to large datasets, we discuss an incremental learning technique along with the MaxSAT formulation.
 

### 2023-03-03 chapters/interpretability/imli/iterative_learning.tex
with new formatting
minor fix: section -> chapter
addressing Kuldeep's comments

@@
 
-\section{Applying {\imli} on Learning Other Interpretable Classifiers.}
+\section{Learning Other Interpretable Classifiers}
 \label{interpretability_imli_sec:application}
@@
 In {\imli}, we apply an iterative learning approach for efficiently learning a  decision list.
-A decision list  $ \Rule_L $ is a list of pairs $ (C_1, V_1), \dots, (C_k, V_k) $, where we propose to learn one pair in each iteration.
+A decision list  $ \Rule_L $ is a list of pairs $ (C_1, V_1), \dots, (C_k, V_k) $, where we learn one pair in each iteration.
 We note that the clause $ C_i $ is a conjunction of literals\textemdash equivalently, a single clause DNF formula.
@@
 		\State $ V_i \leftarrow \textsc{MajorityClass}(\{y^{(i)}\}_{i=1}^{|\dataset|}) $ \Comment{$ V_i $ specifies the target class given class label}
-		\State $ C_i \leftarrow \textsc{MiniBatchDNFLearning}( \dataset, \lambda, 1, V_i) $ \Comment{Ref. Section~\ref{interpretability_imli_sec:dnf_learning}}
+		\State $ C_i \leftarrow \textsc{MiniBatchDNFLearning}( \dataset, \lambda, 1, V_i) $ \Comment{Ref. Chapter~\ref{interpretability_imli_sec:dnf_learning}}
 
@@
 One additional objective in learning a decision set is to minimize the overlap between clauses, as studied in~\cite{lakkaraju2016interpretable}.
-Therefore, along with optimizing accuracy and rule-sparsity, we propose an iterative procedure for decision sets that additionally minimizes the overlap between clauses.
+Therefore, along with optimizing accuracy and rule-sparsity, we discuss an iterative procedure for decision sets that additionally minimizes the overlap between clauses.
 

### 2023-03-03 chapters/interpretability/imli/related_work.tex
all papers added
addressing Kuldeep's comments

@@
-\section{Related Works}
+\section{Related Work}
 \label{interpretability_imli_sec:related}
@@
 In a separate study,~\cite{lakkaraju2016interpretable} highlighted the importance of decision sets over decision lists in terms of interpretability and considered a sub-modular optimization problem for learning a near-optimal solution for decision sets.
-Our proposed method for interpretable classification, however, relies on the improvement in formal methods over the decades, particularly the efficient CDCL-based solution for satisfiability (SAT) problems~\cite{silva2003grasp}.
+Our approach for interpretable classification, however, relies on the improvement in formal methods over the decades, particularly the efficient CDCL-based solution for satisfiability (SAT) problems~\cite{silva2003grasp}.
 

### 2023-03-03 chapters/interpretability/relaxed-cnf/introduction.tex
addressing Kuldeep's comments

@@
 
-In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers. Continuing from chapter~\ref{chapter:imli},
-although CNF/DNF rules are considered interpretable, they are less expressive compared to Boolean cardinality constraints in propositional logic.
+In this chapter, we focus on improving the expressiveness of interpretable rule-based classifiers.
+As demonstrated in Chapter~\ref{chapter:imli}, CNF/DNF rules are considered interpretable, but they are less expressive compared to Boolean cardinality constraints.
 A Boolean cardinality constraint allows one to express numerical bounds on Boolean variables~\cite{sinz2005towards}.
@@
 Consequently, this formulation fails to handle large datasets.
-To address the scalability of {\crr},  we propose an efficient mini-batch training methodology, where we incrementally learn $ \Rule $ from data by iteratively solving smaller MILP queries corresponding to batches.
+To address the scalability of {\crr},  we discuss an efficient mini-batch training methodology as studied in Chapter~\ref{chapter:imli}, where we incrementally learn $ \Rule $ from data by iteratively solving smaller MILP queries corresponding to batches.
 

### 2023-03-03 chapters/introduction.tex
addressing Kuldeep's comments

@@
 In this thesis, we focus on the interpretability and fairness aspects of machine learning.
-The thesis establishes a close integration of formal methods and automated reasoning with machine learning and proposes efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
+We establish a close integration of formal methods and automated reasoning with machine learning and discuss efficient algorithmic solutions for problems arising in interpretability and fairness in machine learning.
 
-Towards responsible and trustworthy machine learning, we propose two research themes in this thesis: interpretability and fairness of machine learning classifiers.
+Towards responsible and trustworthy machine learning, we discuss two research themes in this thesis: interpretability and fairness of machine learning classifiers.
 \emph{In interpretable machine learning}, rule-based classifiers effectively represent the decision boundary using a set of rules comprising input features.
 Interpretable rule-based classifiers not only interpret the decision function but also are applied to explain the prediction of black-box classifiers~\cite{gill2020responsible,lundberg2017unified,moradi2021post,ribeiro2016should,slack2020fooling}, a fundamental research question in explainable artificial intelligence (XAI).
-In this thesis, we propose efficient algorithms based on incremental learning for interpretable rule-based classifiers.
+In this thesis, we discuss efficient algorithms based on incremental learning for interpretable rule-based classifiers.
 In another research theme of \emph{fairness in machine learning}, unregulated classifiers tend to exhibit bias/unfairness to certain demographic groups in the data unless classifiers are trained with a fairness objective.
@@
 In this thesis, we study probabilistic fairness verification problem, where we formally verify the bias of a classifier given the probability distribution of input features.
-Finally, we combine both research themes: interpretability and fairness and propose a framework to interpret the sources of bias.
+Finally, we combine both research themes: interpretability and fairness and discuss a framework to interpret the sources of bias.
 In particular, we formalize and compute fairness influence functions, a way to quantify the influence of individual and the intersection of multiple features on the bias of a classifier.
@@
 Popular interpretable rule-based classifiers such as decision tree, decision lists, and decision sets share the logical structure of CNF/DNF in their representation of the decision function.
-Thereby, we propose to learn an interpretable CNF classifier, wherein its interpretability is defined by the number of Boolean literals that the formula contains.
+Thereby, we discuss a learning framework for interpretable CNF classifiers, wherein the interpretability of the classification rule is defined by the number of Boolean literals in the CNF formula.
 Compared to CNF, Boolean cardinality constraints are more expressive as they allow numerical bounds on Boolean literals~\cite{sinz2005towards}.
@@
 The discrete combinatorial nature of learning rule-based classifiers leads to the intractability of the problem and suffers from scalability issues in large datasets.
-Therefore, we propose an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP to efficiently learn rule-based classifiers in a mini-batch learning setting.
-Here we summarize the contributions of the thesis on interpretable rule-based machine learning.
+Therefore, we discuss an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP to efficiently learn rule-based classifiers in a mini-batch learning setting.
+Here we summarize the contributions of the thesis on interpretable rule-based machine learning\footnote{We publish a python library (\url{https://github.com/meelgroup/mlic}) for learning interpretable rule-based classification rules in a supervised learning setting.}.
 
 \subsection{Scalability via Incremental Learning}
-We propose an incremental learning framework, called $ \imli $,  based on MaxSAT for learning interpretable classification rules expressed in CNF.
+We discuss an incremental learning framework, called $ \imli $,  based on MaxSAT for learning interpretable classification rules expressed in CNF.
 $ \mathsf{IMLI} $ considers a joint objective function to optimize the accuracy and sparsity of classification rules and learns a rule-based classifier by solving an appropriately designed MaxSAT query.
@@
 While the combinatorial structure of relaxed-CNF rules offers exponential succinctness, the na\"ive learning techniques are computationally expensive.
-To this end, we propose an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
+To this end, we discuss an incremental mini-batch learning procedure, called $ \mathsf{CRR} $, that employs advances in MILP solvers to efficiently learn relaxed-CNF rules.
 Our experimental analysis demonstrates that $ \mathsf{CRR} $ can generate relaxed-CNF rules, which are more accurate and sparser compared to the alternative rule-based models.
-
-\textbf{Tool:} We publish a python library \href{https://github.com/meelgroup/MLIC}{{\imli}} for learning interpretable rule-based classification rules in a supervised learning setting.
 
@@
 There is also study on fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness and (ii) interpreting sources of unfairness.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness\footnote{We publish a python library (\url{https://github.com/meelgroup/justicia}) for probabilistic fairness verification.
+} and (ii) interpreting sources of unfairness\footnote{We publish a python library (\url{https://github.com/ReAILe/bias-explainer}) for interpreting group-based fairness metrics in machine learning.}.
 
@@
 
-\subsubsection{Formal Fairness Verification} In this thesis, we propose an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
-Based on stochastic satisfiability (SSAT)~\cite{littman2001stochastic}, our proposed verifier {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
+\subsubsection{Formal Fairness Verification} We discuss an efficient fairness verification framework, starting with a general approach for finite classifiers by encoding them as Boolean formulas, and later a special case of linear classifiers.
+Based on stochastic satisfiability (SSAT)~\cite{littman2001stochastic}, {\justicia} verifies the fairness of Boolean classifiers such as decision tree by solving appropriately designed SSAT formulas.
 {\justicia} also extends verification to compound sensitive groups, which are a combination of multiple categorical sensitive features such as race $ \in $ \{White, Black\} and gender $ \in $ \{male, female\}.
@@
 In the context of verifying the bias of linear classifiers, existing fairness verifiers suffer from two-fold limitations: (i) poor scalability due to applying SSAT/SMT or sampling based techniques and (ii) inaccuracy due to ignoring feature correlations.
-To alleviate these limitations, we propose a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
-{\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
+To alleviate these limitations, we discuss a fairness verification framework for linear classifiers, namely {\fvgm}, for an accurate and scalable fairness verification.
+{\fvgm} relies on a novel \textit{stochastic subset-sum} encoding for linear classifiers obtaining an efficient pseudo-polynomial solution using dynamic programming.
 To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network to represent the conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).
@@
 %, which we study in detail in the following contribution of the thesis.
-
-\textbf{Tool:} We publish a python library \href{https://github.com/meelgroup/justicia}{\justicia} for probabilistic fairness verification.
 
@@
 Quantifying intersectionality allows us to interpret bias induced by the higher-order interactions among features; hence accounting for intersectionality is important to understand bias as suggested by recent literature~\cite{buolamwini2018gender,wang2022towards}.
-\paragraph{Contribution.} We propose an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features.
+
+We discuss an algorithm, called {\fairXplainer}, based on global sensitivity analysis to estimate the fairness influence function (FIF) of individual and intersectional features.
 In experiments, {\fairXplainer} approximates bias based on FIFs more accurately than existing methods.
@@
 
-\textbf{Tool:} We publish a python library \href{https://github.com/ReAILe/bias-explainer}{\fairXplainer} for interpreting group-based fairness metrics in machine learning.
-
 \section{Thesis Outline}
-We organize the thesis in the following way.
-We discuss preliminaries in Chapter~\ref{chapter:preliminaries}.
-In the first part of the thesis on interpretable machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers.
+We organize the thesis as follows.
+In \textbf{part I}, we discuss preliminaries in Chapter~\ref{chapter:preliminaries}.
+In \textbf{part II} on interpretable rule-based machine learning, we discuss an incremental learning framework based on MaxSAT for learning interpretable rule-based classifiers with higher scalability.
 We conclude this part by applying incremental learning on a more expressible yet interpretable rule-based classifier in Chapter~\ref{chapter:crr}.
-In the second part of the thesis on fairness in machine learning, we discuss probabilistic fairness verification for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia} and linear classifiers in Chapter~\ref{chapter:fvgm}.
-In the last part, we combine both interpretability and fairness in machine learning by interpreting group fairness metrics in Chapter~\ref{chapter:fif}.
-Finally, we conclude our thesis in Chapter~\ref{chapter:conclusion}.
+In \textbf{part III} on fairness in machine learning, we discuss probabilistic fairness verification based on SSAT for classifiers represented as Boolean formulas in Chapter~\ref{chapter:justicia}.
+In Chapter~\ref{chapter:fvgm}, we accommodate feature correlations in fairness verification and discuss a tractable verification algorithm for linear classifiers.
+In \textbf{part IV}, we combine both interpretability and fairness in machine learning by interpreting group fairness metrics in Chapter~\ref{chapter:fif}.
+We conclude the thesis in Chapter~\ref{chapter:conclusion}.
 

### 2023-03-03 chapters/fairness/fif/fairxplainer.tex
added appendix
minor edits

@@
 \end{comment}
-We propose an algorithm, {\fairXplainer}, that leverages the variance decomposition of CPPs to estimate the FIFs of all subsets of features.
+We discuss an algorithm, {\fairXplainer}, that leverages the variance decomposition of CPPs to estimate the FIFs of all subsets of features.
 {\fairXplainer} has two algorithmic blocks: (i) local regression to decompose the classifier into component functions taking distinct subsets of features as input and (ii) computing the variance (or covariance) of each component function.
@@
 A smoothing operator, referred as $ \textsc{Smooth} $, allows us to learn a global representation of a component function by smoothly interpolating $\tau + 2$ local points obtained by local regression~\cite{loader2012smoothing}.
-In this paper, we apply cubic spline smoothing~\cite{li2010global} to learn each component function.
+In this chapter, we apply cubic spline smoothing~\cite{li2010global} to learn each component function.
 Cubic spline is a piecewise polynomial of degree $ 3 $ with $ C^2 $ continuity interpolating local points in $ \tau $ intervals.

### 2023-03-03 chapters/fairness/fif/framework.tex
minor edits

@@
 \end{theorem}
-We emphasize that the decomposability property proposed here is global, i.e.
+We emphasize that the decomposability property discussed here is global, i.e.
 it holds for a whole dataset, but the one used for Shapley-value based explanations~(ref.
@@
 
-The symmetry and null properties of FIFs also distinguish the FIF quantification proposed in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose} from the attribution methods considering only individual features, like SHAP.
+The symmetry and null properties of FIFs also distinguish the FIF quantification discussed in Eq.~\eqref{fairness_fairXplainer_eq:fif_decompose} from the attribution methods considering only individual features, like SHAP.
 For example, a feature $i$ has zero impact on bias if not only its individual influence but the sum of influences of all the subsets of features that includes $i$ is zero.

### 2023-03-03 chapters/fairness/fif/introduction.tex
addressing Kuldeep's comments
minor edits

@@
 
-In this chapter, we envision towards combining two research themes in the thesis: interpretability and fairness, and discuss an algorithmic framework to interpret  fairness in machine learning.
 As demonstrated in Chapter~\ref{chapter:justicia} and~\ref{chapter:fvgm}, fairness metrics measure global bias, but do not detect or interpret its sources~\cite{begley2020explainability,lundberg2020explaining,pan2021explaining}.
@@
 
-	\item \textbf{Formalism:} We propose to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Chapter~\ref{fairness_fairXplainer_sec:fifs}).
+	\item \textbf{Formalism:} We discuss how to measure the contribution of individual and intersectional features towards the bias of a classifier operating on a dataset by estimating their \textbf{F}airness \textbf{I}nfluence \textbf{F}unctions (FIFs) (Chapter~\ref{fairness_fairXplainer_sec:fifs}).
 Our method is based on transforming existing fairness metrics into the difference of scaled conditional variances of classifier's prediction, which we then decompose using Global Sensitivity Analysis (GSA)\textemdash a standard technique recommended by regulators to assess numerical models\cite{eu,usepa}.
@@
 
-	\item \textbf{Algorithmic:} We propose a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric.
+	\item \textbf{Algorithmic:} We discuss a new algorithm, called {\fairXplainer}, to estimate individual and intersectional FIFs of features given a dataset, a classifier, and a fairness metric.
 The algorithm is capable of working with any linear group fairness metric, including statistical parity, equalized odds, or predictive parity (Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}).

### 2023-03-03 chapters/fairness/fvgm/conclusion.tex
minor edits

@@
 \section{Chapter Summary}
-In this chapter, we propose {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem.
+We discuss {\fvgm}, an efficient fairness verification framework for linear classifiers based on a novel stochastic subset-sum problem.
 {\fvgm} encodes a graphical model of feature-correlations, represented as a Bayesian Network, and computes multiple group and causal fairness metrics accurately.

### 2023-03-03 chapters/fairness/fvgm/framework.tex
minor edits

@@
 
-We propose a dynamic programming approach~\cite{pisinger1999linear,woeginger1992equal} to solve {\stochastic} as the problem has overlapping sub-problem properties. 
+We discuss a dynamic programming approach~\cite{pisinger1999linear,woeginger1992equal} to solve {\stochastic} as the problem has overlapping sub-problem properties. 
 For example, $S(\mathbf{B}, \tau)$ can be solved by solving $S(\mathbf{B}[2:n], \tau')$, where the updated threshold $\tau'$, called the \textit{residual threshold}, depends on the original threshold $\tau$ and the assignment of $B_1$ as shown in Eq.~\eqref{fvgm_eq:semantics_recurse}.
 %\red{For example, the residual threshold $ \tau $ can be equal for a fixed subset of $ \mathbf{B} $, say $ \mathbf{B}[i:n] \triangleq \{B_j\}_{j=i}^{n} $, derived from different assignments of $ B_1, \dots, B_{i-1} $.}
-Building on this observation, we propose the recursion and terminating condition leading to our dynamic programming algorithm.
+Building on this observation, we discuss the recursion and terminating condition leading to our dynamic programming algorithm.
 
@@
 \paragraph{A Heuristic for Faster Computation.} 
-We propose two improvements for a faster computation of the dynamic programming solution.
+We discuss two improvements for a faster computation of the dynamic programming solution.
 Firstly, we observe that in Eq.~\eqref{fvgm_eq:dp_recurse}, existential/universal variables are deterministically  assigned based on their weights.

### 2023-03-03 chapters/fairness/fvgm/introduction.tex
minor edits

@@
 	\begin{itemize}
-		\item \textit{Framework:} In this chapter, we propose a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
-		\item \textit{Scalability:} {\fvgm} proposes a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
+		\item \textit{Framework:} We discuss a fairness verification framework, namely {\fvgm} (\textbf{F}airness \textbf{V}erification with \textbf{G}raphical \textbf{M}odels), for accurately and efficiently verifying linear classifiers.
+		\item \textit{Scalability:} {\fvgm} relies a novel \textit{stochastic subset-sum} encoding for linear classifiers with an efficient pseudo-polynomial solution using dynamic programming.
 		\item \textit{Accuracy:} To address feature-correlations, {\fvgm} considers a graphical model, particularly a Bayesian Network that represents conditional dependence (and independence) among features in the form of a Directed Acyclic Graph (DAG).

### 2023-03-07 chapters/conclusion.tex
conclusion

@@
 In this thesis, we have designed interpretable classifiers that are explainable by design.
-We can use these classifiers, such as {\imli} and {\crr}, to generate post-hoc explanations for black-box classifiers.
-In this context, logic-based explanations, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local} have demonstrated to be effective in generating post-hoc explanations of black-box classifiers.
-Therefore, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.
+A natural application of interpretable models is to apply them to explain the working of black-box classifiers.
+For example, rule-based classifiers can be leveraged to provide explanations in the form of classification rules, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local}.
+As such, we can use interpretable classifiers {\imli} and {\crr} to generate post-hoc explanations for black-box classifiers.
+In future, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.
 
-	\item \textbf{Fairness Auditing.} Machine learning has demonstrated significant success as a predictive system in automated decision-making.
-However, quantifying and mitigating bias in these systems is still challenging, e.g.
-bias in generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}.
+	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, even beyond machine learning.
+This can help us scale up solutions without losing much accuracy.
+For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets, where the classifier can be routinely updated with new samples.
+Furthermore, we aim to try different objective functions in the incremental solving, apart from the hamming distance-based heuristic objective, which  works well in the incremental learning of CNF classification rules.
+One idea is to learn different rule-based classifiers, possibly in parallel, on different mini-batches or on different subsets of features, and then construct a rule merging procedure to combine different classification rules into one.
+
+	\item \textbf{Fairness Auditing.} Any technology that is publicly used presently undergoes an audit mechanism, where we understand the impacts and limitations of using that technology, and why are they caused.
+Machine learning is becoming the pervasive technology of our time and the discourse on bias induced by machine learning systems is attracting attention, e.g.
+the demonstration of bias in popular generative machine learning and large language models~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}.
 As a result, there has been significant research interest in auditing classifiers for bias, from standard supervised learning to deep neural networks, computer vision, large language models and so on.
-Thus, we need a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees.
-One task of this framework is to categorize different definitions of fairness metrics in the literature.
-This would help stakeholders pick the right definition of fairness for their application~\cite{ruf2021towards}.
-In addition, we aim to extend formal fairness verification beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}.
-We also plan to cover more classes of machine learning classifiers that are hard or impossible to encode as CNF, such as deep neural networks.
-Additionally, we aim to extend fairness verification beyond tabular data to more complex data such as images and texts.
+Thus, we aim to design a fairness auditing framework~\cite{ruf2021towards,yan2022active} with formal guarantees.
+There are three key questions that we aim to investigate in fairness auditing.
 
-	\item \textbf{Fairness and Interpretability: Bridging the Gap.} We aim to design fair and interpretable algorithms for  machine learning.
+	\begin{enumerate}
+		\item \textbf{Which fairness metrics to choose?} Fairness in machine learning is bestowed with multiple notions of fairness.
+Our first line of investigation is to categorize different fairness metrics  and suggest the best metric based on application, data, and prevailing policy, similar to ``Fairness Compass''~\cite{ruf2021towards}.
+This would help stakeholders pick the right definition of fairness for their application.
+
+		\item \textbf{How to quantify bias?} Accurate quantification of bias is an important step towards designing algorithms to mitigate bias.
+As discussed in the thesis, fairness verification allows us to formally quantify the bias of a classifier.
+To this end, we aim to extend formal fairness verification to broader classes of fairness metrics, classifiers, and data.
+
+		\textbf{Fairness Metrics.} We aim to extend fairness verification of beyond group fairness, such as individual fairness~\cite{john2020verifying}, causal fairness~\cite{pan2021explaining,zhang2018fairness}, and counterfactual fairness~\cite{wu2019counterfactual,chiappa2019path}.
+Each category of fairness metrics poses distinct challenges in formal verification.
+For example, verifying individual fairness is related to verifying robustness of a model.
+As such, statistical methods from robustness verification has been applied to individual fairness~\cite{john2020verifying}.
+Formal methods, such as SMT-based encoding is also proposed in this regard~\cite{biswas2022fairify}.
+In our research endeavor, we aim to leverage  SAT or quantified Boolean formula (QBF) based verification, which may improve the scalability of verification.
+
+		\textbf{Broader Classifiers.} We aim to extend fairness verification to broader machine learning classifiers, such as random forests and deep neural networks.
+For random forests, we can leverage CNF-based translation of the ensemble of trees by converting each tree as a CNF and a cardinality constraint to implement the ranking function of the prediction of multiple trees.
+For a special case of binarized neural networks (BNNs)~\cite{hubara2016binarized}, we can leverage existing CNF encoding and deploy SSAT-based verifier.
+However, for general neural networks with continuous parameters, MILP-based encoding can possibly be explored~\cite{mistry2022milp}.
+In addition, for neural networks, other surrogate representation beyond CNF or MILP can also be studied in future.
+In all cases, fairness verification, particularly for group-based metrics, relies on an access to efficient counter of CNF/MILP encoding; thus a dedicated effort to design better counter is a challenging, yet important research direction to explore.
+
+		\textbf{Complex data.} We have explored fairness verification of tabular data; an important research question is to formally verify the fairness of classifiers with complex data such as images~\cite{nuriel2021permuted} and languages~\cite{abid2021persistent,nadeem2020stereoset,vig2020investigating}.
+For such data, one way to adapt existing verification methods, such as {\justicia}, is to apply them on the learned feature representation by the neural network and propagate verification results back to the input layer with images or text.
+In future, we aim to explore this possibility.
+
+		\item \textbf{How to explain bias?} We aim to design fair and interpretable algorithms for  machine learning.
 We are interested in how these two goals relate to each other.
-We have some research questions for the future: (i) are interpretable models fair?
-(ii) how can we improve the fairness of interpretable models?
-(iii) how can we use interpretability to enhance fairness?
-(iv) how can we optimize a classifier for both fairness and interpretability?
-Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability.
-We plan to explore more directions in the future.
+This is inline with GDPR's emphasis on making models transparent and trustworthy, which is deployed in public.	Towards bridging the gap between fairness and interpretability in predictive systems, we explore following research questions:
 
-	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, not only in machine learning.
-This can help us scale up without losing much accuracy.
-For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets.
-To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond, with the help of formal methods and automated reasoning.
+		\begin{enumerate}
+			\item How fair are interpretable machine learning models? 
+			\item How can we improve the fairness of interpretable models?
+			\item How can we apply interpretability to enhance fairness?
+			\item How can we jointly optimize a classifier for  fairness and interpretability?
+		\end{enumerate}
+
+		Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.
+
+	\end{enumerate}
+
 \end{itemize}
+
+To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond.
+To achieve this goal, we aim to integrate formal methods and automated reasoning with statistics and algorithm design.
 

### 2023-05-03 doctoral_consortium/main.tex
updated dc

@@
 % Comment out this line in the camera-ready submission
-\linenumbers
+%\linenumbers
 
@@
 In high-stake domains,  machine learning predictions have far-reaching consequences on the end-users.
-With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2019stop}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox},  etc.\ in predictions.
+With the aim of applying machine learning for societal goods, there have been increasing efforts to regulate machine learning by imposing interpretability~\cite{rudin2022interpretable}, fairness~\cite{barocas2017fairness}, robustness~\cite{rauber2017foolbox},  etc.\ in predictions.
 Towards responsible and trustworthy machine learning, we propose two research themes in our dissertation research: interpretability and fairness of machine learning classifiers.
@@
 
-Interpretable machine learning often employs rule-based classifiers, which use a set of rules to represent the decision boundary.
-The interpretability of such classifiers depends on the size of the rules: smaller rules with higher accuracy are preferred in practice.
-However, this presents a challenge when dealing with large datasets, as interpretable classification learning becomes a combinatorial optimization problem that suffers from poor scalability.
-To address this issue, we propose an incremental learning framework for interpretable rule-based classification on large datasets.
+Interpretable machine learning often employs \textit{rule-based classifiers}, which use a set of rules to represent the decision boundary.
+One particular notion of interpretability of classification rules is their rule-size: smaller rules with higher accuracy are preferred in medical domains.
+However, this presents a challenge when dealing with large datasets, as interpretable classification learning becomes a combinatorial optimization problem suffering from poor scalability.
+To address this issue, we propose an incremental learning framework for rule-based classification on large datasets.
 Our framework combines maximum satisfiability (MaxSAT) and mixed integer linear programming (MILP) with mini-batch learning.
@@
 \subsection*{Expressiveness via Logical Relaxation}
-We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness, as described in~\cite{ghosh2020classification}.
-Specifically, we consider relaxed definitions of the standard OR/AND operators in propositional logic by allowing exceptions in the construction of a clause and in the selection of clauses in a rule.
-Based on these relaxed definitions, we introduce relaxed logical classification rules, which are motivated by the use of checklists in the medical domain and Boolean cardinality constraints.
+We extend our incremental learning framework to enable the learning of a more relaxed representation of classification rules with higher expressiveness~\cite{ghosh2020classification}.
+Motivated by satisfying $ M $ out of $ N $ literals, we consider relaxed definitions of OR/AND operators in logic, which allows exceptions in the construction of a clause and in the selection of clauses in a rule.
+Based on these relaxed definitions, we introduce relaxed logical classification rules, which are inspired by the use of checklists in the medical domain and Boolean cardinality constraints.
 These rules generalize widely used rule representations, such as CNF, DNF, and decision sets.
@@
 
-\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness.
+\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness \textit{essentially beyond sensitive features}.
 Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias.
@@
 In experiments, $\mathsf{FairXplainer}$ captures the influences of both individual and intersectional features across various datasets and classifiers, approximates bias more accurately using FIFs than existing local explanation methods, and demonstrates a higher correlation of FIFs with fairness interventions.
-In future, we aim to develop improved fairness enhancing and attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.
+In future, we aim to develop improved fairness enhancing and fairness attack algorithms based on fairness verifier $ \mathsf{Justicia} $ and fairness explainer $ \mathsf{FairXplainer} $.
 

### 2023-05-03 doctoral_consortium/main.tex
cr attempt 1

@@
 These rules generalize widely used rule representations, such as CNF, DNF, and decision sets.
-However, the combinatorial structure of these rules results in exponential succinctness, and na\"ive learning techniques are computationally expensive.
+While the combinatorial structure of these rules results in exponential succinctness, na\"ive learning techniques are computationally expensive.
 To overcome this issue, we propose an incremental mini-batch learning procedure, called $\mathsf{CRR}$, which leverages advances in MILP solvers to efficiently learn such rules.
@@
 
-\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness \textit{essentially beyond sensitive features}.
+\subsubsection*{Ongoing and Future Work: Explaining Sources of Bias} In our ongoing work, we combine both research themes: interpretability and fairness and propose a framework to explain the sources of unfairness, \textit{essentially beyond sensitive features}.
 Fairness metrics can quantify bias in a global sense, but they cannot identify or explain the sources of bias.
@@
 We use a feature-attribution approach to explain the sources of bias, which relates the influences of input features to the resulting bias of the classifier.
-We formalize the Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}.
+We formalize Fairness Influence Function (FIF) to quantify the contribution of an individual feature and the intersection of multiple features to the resulting bias~\cite{ghosh2022how}.
 We build an algorithm called $\mathsf{FairXplainer}$, which estimates FIFs by decomposing the variance of the classifier's prediction among all subsets of features, using global sensitivity analysis.

### 2023-07-05 chapters/background.tex
addressing comments

@@
 \subsection{Rule-based Classification}
+\label{chapter_interpretability_preliminaries_rule_based_classification}
 A classifier $ \Rule: \mathbf{X} \rightarrow \widehat{Y} \in  \{0,1\} $ is a  function that takes a feature vector as input and outputs the predicted class label $\widehat{Y}$.
@@
 Additionally, we prefer to learn a \emph{sparse} $ \Rule $ in order to favor interpretability.
-We define the structural complexity of $ \Rule $, referred to as \emph{rule-size}, in terms of the number of literals it contains.
+We define the structural complexity of $ \Rule $, referred to as \emph{rule size}, in terms of the number of literals it contains.
 Let $\clause(\Rule,i)$ denote  the  $i^\text{th}$ clause of $\Rule$ and $ |\clause(\Rule,i)| $ be  the number of literals in $\clause(\Rule,i)$.
@@
 \subsection{Fairness Metrics}
-
+\label{chapter_fairness_preliminaries_fairness_metrics}
 A classifier $\alg$ that solely optimizes accuracy, i.e.
@@
 
-\subsection{Bayesian Network}
+\section{Bayesian Network}
 \label{chapter_fairness_preliminaries_BN}
@@
 
-\subsection{Global Sensitivity Analysis (GSA): Variance Decomposition}
+\section{Global Sensitivity Analysis (GSA): Variance Decomposition}
 Global sensitivity analysis is an active field of research that studies how the global uncertainty in the output of a function can be attributed to the different sources of uncertainties in the input variables~\cite{saltelli2008global}.

### 2023-07-05 chapters/conclusion.tex
addressing comments

@@
 To demonstrate the efficacy of our methods, we have constructed open-source tools and conducted experiments on real-world datasets in machine learning.
-In the context of interpretable rule-based machine learning, we have developed an incremental learning framework, known as {\imli}, that scales classification to million-size datasets while maintaining competitive prediction accuracy and rule-size compared to existing rule-based classifiers.
+In the context of interpretable rule-based machine learning, we have developed an incremental learning framework, known as {\imli}, that scales classification to million-size datasets while maintaining competitive prediction accuracy and rule size compared to existing rule-based classifiers.
 Additionally, in our pursuit of more expressive yet interpretable classifiers, we have introduced another learning framework, called {\crr}, for logical relaxed classification rules that are based on incremental learning.

### 2023-07-05 chapters/fairness/fif/experiments.tex
addressing comments

@@
 
-\paragraph{Execution Time: {\fairXplainer} vs.\ SHAP} We compare the execution time of {\fairXplainer} vs.\ SHAP in a cactus plot in Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot}, where a point $ (x, y) $ denotes that a method computes FIFs of $ y $ many fairness instances within $ x $ seconds.
+\paragraph{Execution Time: {\fairXplainer} vs.\ SHAP} We compare the execution time of {\fairXplainer} vs.\ SHAP in a cactus plot in Figure~\ref{fairness_fairXplainer_fig:execution_time_cactus_plot}, where a point $ (x, y) $ denotes that a method computes FIFs of $ x $ many fairness instances within $ y $ seconds.
 We consider $ 480 $ fairness instances from $ 4 $ datasets constituting $ 24 $ distinct combinations of sensitive features, $ 4 $ classifiers, and $ 5 $ cross-validation folds.

### 2023-07-05 chapters/fairness/fif/framework.tex
addressing comments

@@
 \label{fairness_fairXplainer_sec:metric_as_variance}
+First we recall the definition of statistical parity in Chapter~\ref{chapter_fairness_preliminaries_fairness_metrics}.
+\[ f_{\mathsf{SP}}(\alg, \mathbf{D}) \triangleq \max_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] - \min_{\mathbf{a}}\Pr[\widehat{Y} =1 | \mathbf{A} = \mathbf{a}] \]
 
-First, we observe that the random variable central to computing statistical parity is $ \mathds{1}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $.
+Therefore, the random variable central to computing statistical parity is $ \mathds{1}[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}] $.
 We refer to this indicator function as \textit{Conditional Positive Prediction (CPP)} of a classifier.
@@
 
+\begin{remark}
+	For detecting degenerate cases, we require to compute $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\max} ]  $ and $ \Pr[\widehat{Y} = 1 \mid \sensitive = \mathbf{a}_{\min} ]  $ on the dataset $ \mathbf{D} $.
+Hence, the detection can be performed in a straightforward manner.
+\end{remark}
+
 \textit{Expressing FIF in terms of the variance decomposition over a subset of features allows us to import and extend well-studied techniques of GSA to perform FIF estimation}, which we elaborate on Chapter~\ref{fairness_fairXplainer_sec:fairxplainer}.

### 2023-07-05 chapters/interpretability/imli/appendix.tex
addressing comments

@@
 
-In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental}, we compare both non-incremental and incremental encoding in terms of median training time, rule-size, test accuracy, and train accuracy over ten cross validation folds.
+In Figure~\ref{fig:interpretability_imli_additional_fig:all_preformances_non_incremental_vs_incremental}, we compare both non-incremental and incremental encoding in terms of median training time, rule size, test accuracy, and train accuracy over ten cross validation folds.
 In all datasets, the incremental encoding demonstrates higher efficiency than the non-incremental encoding by requiring less training time.
@@
 As a result, the accuracy of the non-incremental encoding is also less than the incremental encoding in large datasets.
-In small datasets, the incremental encoding demonstrates a competitive accuracy and rule-size with the non-incremental encoding.
-Therefore, the incremental encoding in {\imli} demonstrates its effectiveness in training time, accuracy, and rule-size over the non-incremental encoding.
+In small datasets, the incremental encoding demonstrates a competitive accuracy and rule size with the non-incremental encoding.
+Therefore, the incremental encoding in {\imli} demonstrates its effectiveness in training time, accuracy, and rule size over the non-incremental encoding.
 
@@
 
-	\caption[Performance comparison between non-incremental vs.\ incremental encoding]{Comparison of training time, rule-size, and test and train accuracy between non-incremental vs.\ incremental encoding in {\imli}.
+	\caption[Performance comparison between non-incremental vs.\ incremental encoding]{Comparison of training time, rule size, and test and train accuracy between non-incremental vs.\ incremental encoding in {\imli}.
 In $ X $-axis, the dimension of the dataset (\#samples $ \mid $  \#features) is shown below dataset name.

### 2023-07-05 chapters/interpretability/imli/baseline.tex
addressing comments

@@
 We consider two types of Boolean variables: (i) \emph{feature} variables $ \bool $ corresponding to input features and (ii) \emph{error} variables $ \noise $ corresponding to the classification error of samples.
-We define a Boolean variable $ \bool^i_j $ that becomes true if feature $ X_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule-size of $ \Rule $, and $ \bool^i_j $ is assigned false, otherwise.
+We define a Boolean variable $ \bool^i_j $ that becomes true if feature $ X_j $ appears in the $ i^\text{th} $ clause of $ \Rule $, thereby contributing to an increase in the rule size of $ \Rule $, and $ \bool^i_j $ is assigned false, otherwise.
 Moreover, we define an error variable $ \noise_l $ to attribute to whether the $ l^\text{th} $ sample $ (\mathbf{x}^{(l)}, y^{(l)}) $ is classified correctly or not.
@@
 Therefore, by applying Construction~\ref{interpretability_imli_construction:rule}, the classifier is $ (X_1 \vee \neg X_2)  \wedge (\neg X_1 \vee X_2) $.\footnote{The presented MaxSAT-based formulation does not learn a CNF classifier with both $ X_i $ and $ \neg X_i $ in the same clause.
-The reason is that a clause with both $ X_i $ and $ \neg X_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule-size and hence, this is not an optimal classifier.}
+The reason is that a clause with both $ X_i $ and $ \neg X_i $ connected by OR $ (\vee) $ does not increase accuracy but increases rule size and hence, this is not an optimal classifier.}
 

### 2023-07-05 chapters/interpretability/imli/experiment.tex
addressing comments

@@
 	\item How does the incremental learning in {\imli} perform compared to non-incremental MaxSAT-based learning in terms of accuracy, rule-sparsity, and scalability?
-	\item How do different interpretable classification rules learned using {\imli} perform in terms of accuracy and rule-size?
+	\item How do different interpretable classification rules learned using {\imli} perform in terms of accuracy and rule size?
 	\item What are the effects of different hyper-parameters in {\imli}?
@@
 In contrast, {\imli} generates significantly smaller rules than RIPPER, specifically in large datasets.
-Moreover, BRL, CORELS, and BRS report comparatively smaller rule-size than {\imli} on average, but with a significant decrease in accuracy.
+Moreover, BRL, CORELS, and BRS report comparatively smaller rule size than {\imli} on average, but with a significant decrease in accuracy.
 In terms of scalability, {\imli} achieves the best performance compared to other interpretable and non-interpretable classifiers by classifying datasets with one million samples.
@@
 
-	\caption[Accuracy and rule-size of interpretable classifiers]{Comparison of accuracy and rule-size among interpretable classifiers.
-Each cell from the fourth to the eighth column contains test accuracy (top) and rule-size (bottom).
+	\caption[Accuracy and rule-size of interpretable classifiers]{Comparison of accuracy and rule size among interpretable classifiers.
+Each cell from the fourth to the eighth column contains test accuracy (top) and rule size (bottom).
 `\textemdash' represents a timeout.
@@
 \label{interpretability_imli_sec:experiments_scalability}
-\paragraph{Comparing {\imli} with interpretable classifiers.} We compare {\imli} with existing interpretable classifiers in three aspects: test accuracy, rule-size, and scalability.
+\paragraph{Comparing {\imli} with interpretable classifiers.} We compare {\imli} with existing interpretable classifiers in three aspects: test accuracy, rule size, and scalability.
 
-\textit{Test accuracy and rule-size.}  We present the experimental  results of test accuracy and rule-size among interpretable classifiers in Table~\ref{interpretability_imli_table:interpretable_classifiers}, where the first, second, and third columns represent  the name of the dataset, the number of samples, and the number of features in the dataset, respectively.
+\textit{Test accuracy and rule size.}  We present the experimental  results of test accuracy and rule size among interpretable classifiers in Table~\ref{interpretability_imli_table:interpretable_classifiers}, where the first, second, and third columns represent  the name of the dataset, the number of samples, and the number of features in the dataset, respectively.
 In each cell from the fourth to the eighth column in the table, the top value represents the median test accuracy and the bottom value represents the median size of rules measured through ten-fold cross-validation.
@@
 In particular, {\imli} generates sparser rules than RIPPER in $ 9 $ out of $ 12 $ datasets, wherein RIPPER times out in $ 3 $ datasets.
-Interestingly, the difference in rule-size is more significant in larger datasets, such as in `Vote' dataset, where RIPPER learns a classifier with $ 132 $ Boolean literals compared to $ 15 $ Boolean literals by {\imli}.
+Interestingly, the difference in rule size is more significant in larger datasets, such as in `Vote' dataset, where RIPPER learns a classifier with $ 132 $ Boolean literals compared to $ 15 $ Boolean literals by {\imli}.
 Therefore, {\imli} is better than RIPPER in terms of rule-sparsity, but lags slightly in accuracy.
@@
 {\imli} performs better than BRL both in terms of accuracy and rule-sparsity.
-In particular, {\imli} has  higher accuracy and lower rule-size than BRL in $ 12 $ and $ 8 $ datasets, respectively, in a total of $ 12 $ datasets, wherein BRL times out in $ 3 $ datasets.
+In particular, {\imli} has  higher accuracy and lower rule size than BRL in $ 12 $ and $ 8 $ datasets, respectively, in a total of $ 12 $ datasets, wherein BRL times out in $ 3 $ datasets.
 While comparing with CORELS, {\imli} achieves higher accuracy in almost all datasets ($ 14 $ out of $ 15 $ datasets).
@@
 CORELS and BRS, however, generates sparser rules than {\imli} in most datasets, but by costing a significant decrease in accuracy.
-For example, in the largest dataset `BNG(credit-g)' with $ 1 $ Million samples, BRS times out and CORELS generates a classifier with  $ 72.08\% $ accuracy with rule-size $ 2 $.
+For example, in the largest dataset `BNG(credit-g)' with $ 1 $ Million samples, BRS times out and CORELS generates a classifier with  $ 72.08\% $ accuracy with rule size $ 2 $.
 {\imli}, in contrast, learns a classifier with $ 27.5 $ Boolean literals achieving $ 75.48\% $ accuracy, which is $ 3\% $ higher than CORELS.
-Therefore, {\imli}  makes a good balance between accuracy and rule-size compared to existing interpretable classifiers while also being highly scalable.
+Therefore, {\imli}  makes a good balance between accuracy and rule size compared to existing interpretable classifiers while also being highly scalable.
 In the following, we discuss the results on the scalability of all interpretable classifiers in detail.
@@
 
-	\caption[Training time, test error, and rule-size of different formulations in {\imli}]{Comparison of training time, test error, and rule-size among different formulations presented in the chapter.
+	\caption[Training time, test error, and rule size of different formulations in {\imli}]{Comparison of training time, test error, and rule size among different formulations presented in the chapter.
 In each cactus plot, the incremental formulation {\imli} with both mini-batch and iterative learning demonstrates the best performance in training time and test error than compared two formulations: non-incremental MaxSAT formulation {\imli}-$\mathsf{B}$ and incremental formulation with only mini-batch learning {\imli}-$\mathsf{M}$.
-In terms of rule-size, {\imli} often generates higher size rules than {\imli}-$\mathsf{M}$.}
+In terms of rule size, {\imli} often generates higher size rules than {\imli}-$\mathsf{M}$.}
 	\label{interpretability_imli_fig:different_formulations}
@@
 \paragraph{Comparison among different formulations in {\imli}.}  We compare the performance of different formulations for learning classification rules as presented in this chapter.
-In Figure~\ref{interpretability_imli_fig:different_formulations}, we show cactus plots for assessing training time (in seconds), test error (in percentage), and rule-size among different formulations.
-In the cactus plot, a point $ (x,y) $ denotes that the formulation yields lower than or equal to $ y $ training time (similarly, test error and rule-size) in $ x $ many instances in each dataset.
+In Figure~\ref{interpretability_imli_fig:different_formulations}, we show cactus plots for assessing training time (in seconds), test error (in percentage), and rule size among different formulations.
+In the cactus plot, a point $ (x,y) $ denotes that the formulation yields lower than or equal to $ y $ training time (similarly, test error and rule size) in $ x $ many instances in each dataset.
 
@@
 
-Moving focus on the rule-size in the rightmost column in Figure~\ref{interpretability_imli_fig:different_formulations}, {\imli}-$\mathsf{B}$ achieves the highest rule-size in WDBC dataset.
-In contrast, the rule-size of {\imli}-$\mathsf{B}$  is lowest (zero) in Credit and Adult datasets.
+Moving focus on the rule size in the rightmost column in Figure~\ref{interpretability_imli_fig:different_formulations}, {\imli}-$\mathsf{B}$ achieves the highest rule size in WDBC dataset.
+In contrast, the rule size of {\imli}-$\mathsf{B}$  is lowest (zero) in Credit and Adult datasets.
 In the last two datasets, {\imli}-$\mathsf{B}$ times out during training and returns the default rule ``true'' by predicting all samples as class $ 1 $.
-The other two formulations {\imli} and {\imli}-$\mathsf{M}$  demonstrate a similar trend in rule-size in all datasets with {\imli}-$\mathsf{M}$ generating comparatively smaller size rules in Credit and Adult datasets.
+The other two formulations {\imli} and {\imli}-$\mathsf{M}$  demonstrate a similar trend in rule size in all datasets with {\imli}-$\mathsf{M}$ generating comparatively smaller size rules in Credit and Adult datasets.
 In this context,  the improvement of rule-sparsity of {\imli}-$\mathsf{M}$ is due to a comparatively higher test error (or lower accuracy) than {\imli} as observed in all three datasets.
-Therefore, {\imli} appears to be the best performing formulation w.r.t.\ training time, test error, and rule-size by balancing between accuracy and rule-size while being more scalable.
+Therefore, {\imli} appears to be the best performing formulation w.r.t.\ training time, test error, and rule size by balancing between accuracy and rule size while being more scalable.
 
@@
 	\centering
-	\caption[Accuracy and rule-size of different classification rules learned using {\imli}]{Comparison of test accuracy (top value) and rule-size (bottom value) among different rule-based representations learned using {\imli}.
+	\caption[Accuracy and rule-size of different classification rules learned using {\imli}]{Comparison of test accuracy (top value) and rule size (bottom value) among different rule-based representations learned using {\imli}.
 Numbers in bold denote the best performing results among different representations.}
@@
 
-We deploy {\imli} to learn different interpretable rule-based representations: CNF and DNF classifiers, decision lists, and decision sets and present their comparative performance w.r.t.\ test accuracy and rule-size in Table~\ref{interpretability_imli_table:different_representations}.
+We deploy {\imli} to learn different interpretable rule-based representations: CNF and DNF classifiers, decision lists, and decision sets and present their comparative performance w.r.t.\ test accuracy and rule size in Table~\ref{interpretability_imli_table:different_representations}.
 In each cell in this table, the top value represents the test accuracy and the bottom value represents the size of generated rules.
@@
 In contrast, both DNF and decision lists achieve the highest accuracy in three datasets each; decision sets demonstrate the least performance in test accuracy by being more accurate in one dataset.
-To this end, the poor accuracy of decision sets is traded off by its rule-size as decision sets generate the sparsest rules compared to other representations.
-More precisely, decision sets have the smallest rule-size in six datasets, while CNF, DNF, and decision lists have the smallest rule-size in two, three, and one dataset, respectively.
+To this end, the poor accuracy of decision sets is traded off by its rule size as decision sets generate the sparsest rules compared to other representations.
+More precisely, decision sets have the smallest rule size in six datasets, while CNF, DNF, and decision lists have the smallest rule size in two, three, and one dataset, respectively.
 These results suggest that CNF classifiers are more favored in applications where higher accuracy is preferred, while decision sets are preferred in applications where higher interpretability is desired.
@@
 
-	\caption[Effect of the number of clauses in {\imli}]{Effect of the number of clauses $ k $ on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of the number of clauses in {\imli}]{Effect of the number of clauses $ k $ on accuracy (test and train), rule size, and training time.
 As $ k $ increases, both train and test accuracy of {\imli} increase while generating rules with higher size by incurring higher training time.
@@
 
-\paragraph{Ablation study.} We experiment the effect of different hyper-parameters in {\imli} on prediction accuracy, rule-size, and training time in different datasets.
+\paragraph{Ablation study.} We experiment the effect of different hyper-parameters in {\imli} on prediction accuracy, rule size, and training time in different datasets.
 In the following, we discuss the impact of the number of clauses, regularization parameter, and size of mini-batches in {\imli}.
@@
 This observation indicates that higher $ \lambda $ puts lower computational load to the MaxSAT solver as a fraction of training examples is allowed to be misclassified.
-Thus, similar to the number of clauses, regularization parameter $ \lambda $ in {\imli} allows to trade-off between accuracy and rule-size in a precise manner.
+Thus, similar to the number of clauses, regularization parameter $ \lambda $ in {\imli} allows to trade-off between accuracy and rule size in a precise manner.
 
@@
 Due to solving higher size MaxSAT queries, the training time also increases in general with an increase in mini-batch size.
-Therefore, by varying the size of mini-batches, {\imli} allows controlling on training time vs the prediction accuracy (and rule-size) of generated rules.
+Therefore, by varying the size of mini-batches, {\imli} allows controlling on training time vs the prediction accuracy (and rule size) of generated rules.
 
@@
 
-	\caption[Effect of regularization $ \lambda $ in {\imli}]{Effect of regularization $ \lambda $ on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of regularization $ \lambda $ in {\imli}]{Effect of regularization $ \lambda $ on accuracy (test and train), rule size, and training time.
 As $ \lambda $ increases, lower priority is given to accuracy.
@@
 
-	\caption[Effect of batch-size in {\imli}]{Effect of bath size on accuracy (test and train), rule-size, and training time.
+	\caption[Effect of batch-size in {\imli}]{Effect of bath size on accuracy (test and train), rule size, and training time.
 As we consider more samples in a mini-batch, {\imli} generates more accurate and larger size classification rules.}

### 2023-07-05 chapters/interpretability/imli/introduction.tex
addressing comments

@@
 In recent approaches, the interpretable classification problem is reduced to an optimization problem, where the accuracy and the sparsity of rules are optimized jointly~\cite{lakkaraju2016interpretable,narodytska2018learning}.
-Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule-size.
+Different optimization solvers such as linear programming~\cite{malioutov2013exact}, sub-modular optimizations~\cite{lakkaraju2016interpretable}, and Bayesian optimizations~\cite{letham2015interpretable} are then deployed to find the best classifier with maximum accuracy and minimum rule size.
 In our study, we study an alternate optimization approach that fits particularly well to rule-learning problems.

### 2023-07-05 chapters/interpretability/imli/problem_formulation.tex
addressing Kuldeep's comments
addressing comments

@@
  goals: of  being accurate but also interpretable.  
- Various notions of interpretability have been proposed in the context of   classification problems. A common proxy for interpretability in the context of decision rules 
+ Various notions of interpretability have been discussed in the context of   classification problems. A common proxy for interpretability in the context of decision rules 
  is the sparsity of rule.
@@
 
-Higher  values of $ \lambda $ generate a rule with a smaller rule-size but of more training errors, and vice-versa.
+Higher  values of $ \lambda $ generate a rule with a smaller rule size but of more training errors, and vice-versa.
 Thus, $ \lambda $ can be tuned to trade-off between accuracy and interpretability for a rule-based classifier, which we experiment extensively in Chapter~\ref{interpretability_imli_sec:experiments}.

### 2023-07-05 chapters/interpretability/imli/related_work.tex
addressing comments

@@
 
-Recently, several optimization frameworks have been proposed for interpretable classification, where both accuracy and rule-size are optimized during training.
+Recently, several optimization frameworks have been proposed for interpretable classification, where both accuracy and rule size are optimized during training.
 For example,~\cite{malioutov2013exact} proposed exact learning of rule-based classifiers based on Boolean compressed sensing using a linear programming formulation.~\cite{su2016learning} presented two-level Boolean rules, where the trade-off between classification accuracy and interpretability is studied.

### 2023-07-05 chapters/interpretability/relaxed-cnf/experiment.tex
addressing comments

@@
 \begin{table}
-	\caption[Accuracy, rule-size, and training time of rule-based classifiers]{Comparisons of test accuracy, rule-size and training time among different rule-based classifiers.
+	\caption[Accuracy, rule-size, and training time of rule-based classifiers]{Comparisons of test accuracy, rule size and training time among different rule-based classifiers.
 Every cell in the last four columns contains the test accuracy in percentage (top value), rule size  (middle value), and  training time in seconds (bottom value).
 In the experiments, {\crr} shows higher test accuracy than IMLI and generates sparser rules than RIPPER.
-Number in bold denotes the best result, such as maximum test accuracy, minimum rule-size, and minimum training time among competitive classifiers.}
+Number in bold denotes the best result, such as maximum test accuracy, minimum rule size, and minimum training time among competitive classifiers.}
 	\label{interpretability_crr_tab:rule_based_classifiers}
@@
 \subsubsection{Performance Evaluation of {\crr} with Rule-based Classifiers}
-We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule-size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
+We conduct an assessment of performance  using five-fold nested cross-validation as in~\cite{DGW2018}  and report the median of test accuracy, rule size and training time of all rule-based classifiers in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
 Specifically, we show the dataset,  the number of samples and the number of discretized features in the first three columns in Table~\ref{interpretability_crr_tab:rule_based_classifiers}.
-Inside each cell of column   four to $    11 $,  we present the   test accuracy (top value), rule-size (middle value) and training time (bottom value) of each classifier for each dataset.
+Inside each cell of column   four to $    11 $,  we present the   test accuracy (top value), rule size (middle value) and training time (bottom value) of each classifier for each dataset.
 
@@
 In {Table}~\ref{interpretability_crr_tab:rule_based_classifiers}, relaxed-CNF rules exhibit higher prediction accuracy than CNF rules in the majority of the datasets, showing the effectiveness of using a more expressive representation of classification rules in capturing the decision boundary.
-In addition, the generated relaxed-CNF rules are comparatively smaller than CNF rules in terms of rule-size in most of the datasets.
+In addition, the generated relaxed-CNF rules are comparatively smaller than CNF rules in terms of rule size in most of the datasets.
 Therefore, relaxed-CNF rules improve upon CNF rules in terms of both prediction accuracy and rule size in the majority of the datasets.
-In this context, {\crr} provides a trade-off between accuracy and rule-size depending on the choice of hyper-parameters and the experimental results are discussed later in Chapter~\ref{interpretability_crr_sec:model_parameters}.
+In this context, {\crr} provides a trade-off between accuracy and rule size depending on the choice of hyper-parameters and the experimental results are discussed later in Chapter~\ref{interpretability_crr_sec:model_parameters}.
 We then compare relaxed-CNF rules with DNF rules generated by BRS and find that relaxed-CNF rules outperform DNF rules with respect to prediction accuracy in several datasets.
@@
 
-	\caption[Effect of data-fidelity $ \lambda $ in {\crr}]{Effect of data-fidelity $ \lambda $ on test accuracy, rule-size, and training time in {\crr}. } 
+	\caption[Effect of data-fidelity $ \lambda $ in {\crr}]{Effect of data-fidelity $ \lambda $ on test accuracy, rule size, and training time in {\crr}. } 
 	\label{interpretability_crr_fig:result_lambda}
@@
 
-	\caption[Effect of the number of clause $ k $ in {\crr}]{Effect of the number of clause $ k $ on test accuracy, rule-size, and training time in {\crr}. } 
+	\caption[Effect of the number of clause $ k $ in {\crr}]{Effect of the number of clause $ k $ on test accuracy, rule size, and training time in {\crr}. } 
 	\label{interpretability_crr_fig:result_clause}
@@
 
-	\caption[Effect of mini-batch size in {\crr}]{Effect of mini-batch size on test accuracy, rule-size, and training time in {\crr}. } 
+	\caption[Effect of mini-batch size in {\crr}]{Effect of mini-batch size on test accuracy, rule size, and training time in {\crr}. } 
 	\label{interpretability_crr_fig:result_minibatch}
@@
 
-	\caption[Effect of the number of iterations in {\crr}]{Effect of the number of iterations on test accuracy, rule-size, and training time in {\crr}. } 
+	\caption[Effect of the number of iterations in {\crr}]{Effect of the number of iterations on test accuracy, rule size, and training time in {\crr}. } 
 	\label{interpretability_crr_fig:result_iteration}
@@
 	In Figure~\ref{interpretability_crr_fig:result_clause}, as we increase  $ k $, {\crr} allows the generated rules to capture the variance in the given dataset more effectively, that results in higher  accuracy in most of the datasets.
-The  rule-size also increases as we learn more clauses.
+The  rule size also increases as we learn more clauses.
 In addition, the training time  increases, that can be reasoned by the fact that the number of constraints in the MILP formulation is linear with $ k $.
@@
 The training time also increases with $ \tau $ because {\crr} is required to solve in total $ \tau $ queries.
-We also observe an increase in rule-size in most datasets.
+We also observe an increase in rule size in most datasets.
 The reason is that the objective function in the incremental mini-batch approach in Eq.~\ref{interpretability_crr_eq:ilp_inc} does not put a restriction on the size of the rules, rather on the change of rules in consecutive iterations.
-In addition, the learned values of the thresholds $ \eta_c$ and $ \eta_l $  in one iteration are not carried to the MILP query in the next iteration, that may cause an increase of rule-size.
+In addition, the learned values of the thresholds $ \eta_c$ and $ \eta_l $  in one iteration are not carried to the MILP query in the next iteration, that may cause an increase of rule size.

### 2023-07-05 chapters/interpretability/relaxed-cnf/introduction.tex
addressing comments

@@
 
- {Relaxed-CNF rules are  more flexible than pure CNF rules, and they can accurately fit more complex classification boundaries.} For example, relaxed-CNF clauses  allow  a compact encoding of the majority function\footnote{The majority function is a Boolean function that evaluates to $ 0 $ when half or more arguments are false, and $ 1 $ otherwise (\url{https://en.wikipedia.org/wiki/Majority_function}).} in Boolean logic, which would require exponentially many clauses in CNF, showing the exponential gap in the succinctness of the two representations.
+ {Relaxed-CNF rules are  more flexible than pure CNF rules, and they can accurately fit more complex classification boundaries.} For example, relaxed-CNF clauses  allow  a compact encoding of the majority function\footnote{The majority function is a Boolean function that evaluates to $ 0 $ when half or more arguments are false, and $ 1 $ otherwise~\cite{peterson1972error}.} in Boolean logic, which would require exponentially many clauses in CNF, showing the exponential gap in the succinctness of the two representations.
 In addition, relaxed-CNF and CNF rules have the same functional form where a user has to compute the sum of true literals/clauses  and then compare the sum  to  different thresholds {(as in the example in Figure~\ref{interpretability_crr_exmpl:wdbc})}.

### 2023-07-18 chapters/acknowledgements.tex
acknowledgement added

--- 2023-02-07 chapters/acknowledgements.tex
@@
 
+I would like to express my sincere gratitude to everyone who has supported me throughout my PhD.
+This has been a journey of self-discovery, patience, and continuous effort to surpass myself day by day.
+I am fortunate to have met amazing researchers around the world, communicating ideas and learning from them.
+Reflecting on Socrates' famous words, ``Know Thyself'', my PhD has been a journey of understanding my limitations and fostering a mindset of overcoming shortcomings while enjoying the pursuit of knowledge.
+At this moment, I truly understand that the outcome of my PhD goes beyond the thesis itself; it encompasses my personal growth and development.
+
+I come from a humble background in Bangladesh, where my passion for math, science, and computer science has always been a driving force.
+Today, I am immensely grateful to announce that my dream of pursuing a PhD in computer science has become a reality.
+I am deeply thankful to the brilliant minds I have had the privilege to encounter throughout my doctoral journey.
+
+I am immensely grateful to my advisor, Dr.
+Kuldeep S.
+Meel, for playing a crucial role in my PhD journey.
+Under his guidance, I have learned the importance of being an independent researcher and have been provided with numerous opportunities to achieve this goal.
+I owe my confidence as a researcher to Kuldeep's unwavering support in connecting me with exceptional researchers through conferences, workshops, internships, and research visits.
+His leadership, passion, and dedication have truly inspired me.
+I consider myself privileged to have been a part of his research lab and to have experienced the vibrant research culture it offers.
+I vividly recall the early semesters in NUS, where Kuldeep would spend hours discussing with the research group, creating an environment that fostered growth and learning.
+As a new PhD student, his supervision boosted my confidence and enabled me to flourish in the doctoral program.
+Kuldeep's sharp intellect and innovative thinking have not only made me enjoy graduate school but have also facilitated my personal growth.
+I am grateful for his relentless guidance, which has not only enhanced my skills as a researcher but also shaped me into a better person.
+
+I feel incredibly fortunate to have had the opportunity to collaborate with exceptional individuals throughout my research journey.
+Debabrota Basu, Dmitry Malioutov, Krisnha P.
+Gummadi, Daniel Neider, Arijit Khan, Jonathan Scarlet, Chandrika Bhardwaj, Vijay Saraswat, Naheed Anjum Arafat, and Lorenzo Ciampiconi are among the remarkable collaborators who have generously shared their knowledge and experience with me.
+I come to understand that finding a meaningful research problem is a crucial step and I am grateful to have had the opportunity to collaborate and brainstorm with such insightful collaborators.
+
+I would like to express my sincere appreciation to my friends in Singapore who have provided invaluable support throughout my journey.
+Neamul Kabir, Sanjay Saha, Amlan Saha, Rafiul Islam, Zubair Faruqui, Mohimenul Mahi, Ramzan Miah, Muhim Muktadir Zim, and Shayok Ghosh have been pillars of support and encouragement.
+I am also grateful for the exceptional colleagues who have made a significant impact on my professional growth.
+Priyanka Golia, Teodora Baluta, Yash Pote, Arijit Shaw, Suwei Yang, Jiong Yang, Mate Soos, Tim van Bremen, Anna Latour, Yacine Izza, Gunjan Kumar, Paulius Dilkas, Shubham Sharma, Rahul Gupta, and Zhanzhong Pang have been sources of inspiration and collaboration.
+I truly believe that it takes an entire village to achieve something incredible, and I am fortunate to have such a supportive community.
+Additionally, I extend my thanks to all the staff members in SOC, NUS for their exceptional efficiency in addressing students' needs.
+
+I would like to express my heartfelt gratitude to my family for their support and encouragement.
+My parents, sister, brother-in-law, and uncle have been constant pillars of strength throughout my journey.
+While I deeply regret that my father is not here with me today, I find solace in the belief that his blessings will always guide me.
+I am also immensely grateful to my wife, Dola Ghosh, who has been my unwavering source of support and companionship throughout this challenging journey.
+Her presence has made the road to completion much smoother, and I am grateful for her love and encouragement.
+Additionally, I would like to extend a special thank you to my childhood friends, Kalyan Roy, Sukanto Kundu, Nayem Ul Haque, and Niranjan Chandra Roy.
+Their friendship has been a constant source of strength and support throughout my life.
+
+\begin{flushright}
+	\vspace{1em}
+	Bishwamittra Ghosh
+\end{flushright}
+
 \end{acknowledgments}

### 2023-08-10 chapters/acknowledgements.tex
acknowledgement

@@
 I come to understand that finding a meaningful research problem is a crucial step and I am grateful to have had the opportunity to collaborate and brainstorm with such insightful collaborators.
+In addition, I thank Harold Soh and Reza Shokri, who examine my thesis with utmost attention.
+Their feedback has been a driving force for me to reflect positively on my thesis, building a mindset for a broader view on research in general.
 
@@
 Priyanka Golia, Teodora Baluta, Yash Pote, Arijit Shaw, Suwei Yang, Jiong Yang, Mate Soos, Tim van Bremen, Anna Latour, Yacine Izza, Gunjan Kumar, Paulius Dilkas, Shubham Sharma, Rahul Gupta, and Zhanzhong Pang have been sources of inspiration and collaboration.
-I truly believe that it takes an entire village to achieve something incredible, and I am fortunate to have such a supportive community.
-Additionally, I extend my thanks to all the staff members in SOC, NUS for their exceptional efficiency in addressing students' needs.
+Additionally, I would like to extend a special thank you to my childhood friends, Kalyan Roy, Sukanto Kundu, Nayem Ul Haque, and Niranjan Chandra Roy.
+I am fortunate to have such a supportive community.
+I extend my thanks to all the staff members in SOC, NUS for their exceptional efficiency in addressing out needs.
 
@@
 Her presence has made the road to completion much smoother, and I am grateful for her love and encouragement.
-Additionally, I would like to extend a special thank you to my childhood friends, Kalyan Roy, Sukanto Kundu, Nayem Ul Haque, and Niranjan Chandra Roy.
-Their friendship has been a constant source of strength and support throughout my life.
+%Their friendship has been a constant source of strength and support throughout my life.
 

### 2023-08-10 chapters/conclusion.tex
thesis conclusio

@@
 Additionally, we discuss a global sensitivity analysis-based method, {\fairXplainer}, that interprets group fairness metrics by computing fairness influences of individual and intersectional features.
-Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local explainability-based approach.
+Notably, {\fairXplainer} approximates bias more accurately using fairness influence functions (FIFs) and demonstrates a higher correlation of FIFs with fairness intervention than the local interpretability-based approach.
 
-In future, our research on fair and interpretable machine learning will focus on several areas, which we elaborate in the following.
+Our future research is dedicated to developing practical and scalable algorithms for trustworthy machine learning.
+Machine learning and artificial intelligence have been compared to the new electricity, with the potential to transform various aspects of human life, evident from the overwhelming response to generative AI.
+Ensuring fairness and interpretability in deployed machine learning is now more necessary than ever.
+To accomplish this, we aim to work in a collaborative environment, gaining insights into real-world challenges and leveraging advances in the field, alongside formal methods, to make significant progress.
+We have identified key research themes that will guide our work towards this vision.
+
+\paragraph{Fairness and Interpretability As a Service.} 
+The goal of modern machine learning extends beyond learning patterns from large-scale historical data to ensuring responsible decision-making through careful regulation to establish trustworthiness.
+For instance, in a job application scenario, a machine learning algorithm must be fair across different demographic groups, resilient to non-actionable changes in candidate profiles, and interpretable to allow candidates to understand the decision-making process.
+The long-term research goal of this thesis is to offer fairness and interpretability as a service with machine learning-based decision-making.
 
 \begin{itemize}
-	\item 
-	\textbf{Post-hoc Explainability.} Black-box classifiers are hard to understand and trust, especially in safety-critical domains.
-In this thesis, we have designed interpretable classifiers that are explainable by design.
-A natural application of interpretable models is to apply them to explain the working of black-box classifiers.
-For example, rule-based classifiers can be leveraged to provide explanations in the form of classification rules, such as abduction-based explanations~\cite{ignatiev2019abduction}, Anchor~\cite{ribeiro2018anchors}, and LORE~\cite{guidotti2018local}.
-As such, we can use interpretable classifiers {\imli} and {\crr} to generate post-hoc explanations for black-box classifiers.
-In future, we plan to test how well {\imli} can synthesize explanations with large samples and how well {\crr} can generate more expressive explanations.
 
-	\item \textbf{Efficiency through Incremental Solving.} We aim to apply incremental solving, which we used for interpretable classification learning with {\imli} and {\crr}, to other optimization problems, even beyond machine learning.
-This can help us scale up solutions without losing much accuracy.
-For example, we aim to apply incremental solving in online learning of interpretable models, such as decision trees, decision lists, and decision sets, where the classifier can be routinely updated with new samples.
-Furthermore, we aim to try different objective functions in the incremental solving, apart from the hamming distance-based heuristic objective, which  works well in the incremental learning of CNF classification rules.
-One idea is to learn different rule-based classifiers, possibly in parallel, on different mini-batches or on different subsets of features, and then construct a rule merging procedure to combine different classification rules into one.
+	\item \textbf{Interpretability with Guarantees.} Our research in interpretable machine learning spans \textbf{two-fold directions}.
+(i) \textit{Interpretability by design:} There is a growing interest for interpretable machine learning in safety-critical domains~\cite{rudin2019stop}.
+Building upon our interpretable rule-based classifier $ \mathsf{IMLI} $, we aim to enhance learning algorithms for interpretable models in large-scale datasets across supervised, semi-supervised, and unsupervised settings.
+(ii) \textit{Post-hoc interpretability:} To explain black-box predictions, we focus on explanations with formal guarantees~\cite{ignatiev2019abduction,ribeiro2018anchors,guidotti2018local}.
+For example, an explanation model must be robust, learned in a privacy-preserving manner, and provide the confidence level of explanations to increase transparency and trust in the decision-making process.
 
@@
 
-		Our work on {\fairXplainer} is a stepping stone towards bridging the gap between fairness and interpretability. We plan to explore more directions in the future.
-
 	\end{enumerate}
-
 \end{itemize}
 
-To conclude, the long-term vision of this thesis is to enhance machine learning in interpretability, fairness, and beyond.
-To achieve this goal, we aim to integrate formal methods and automated reasoning with statistics and algorithm design.
+\paragraph{Verifiable Machine Learning with Formal Methods.} In safety-critical and high-stake domains, the verification of machine learning before deployment is crucial.
+To this end, formal methods provide a template to verify different checkable properties of machine learning.
+In particular, SAT, SMT and their variants allow to concentrate on synthesizing constraints from real-world use cases and delegate the solution finding to respective solvers, thanks to the dedicated community in formal methods.
+Building upon our ongoing research, which involves applying SSAT and MaxSAT to address fairness and interpretability challenges, we plan to explore additional formulations in formal methods, such as functional analysis, abstract interpretation, and solvers with expressive theory, to further enhance the verification of machine learning models.
 
-\begin{comment}
-	\begin{itemize}
-		\item Fairness repair
-		\item Incremental solving
-		\item Fairness and interpretability as a service to more complex models.
-	\end{itemize}
-\end{comment}
+\paragraph{Benchmarking Machine Learning: Improvement in Formal Methods.} Interdisciplinary research is an ongoing process that yields valuable contributions to different disciplines.
+Our work on fairness and interpretability in machine learning has had a positive impact on the improvement of solvers in formal methods.
+For instance, when tackling interpretable classification problems, we realized that MaxSAT alone was insufficient for large-scale rule-based classification, prompting the need for an incremental MaxSAT solver.
+As a result, we contribute to the MaxSAT evaluation competition in 2019 with our MaxSAT benchmarks of interpretable classification.
+Subsequently, the competition introduced an incremental MaxSAT solving track.
+This observation serves as strong motivation for us to continue contributing to the formal methods community by creating additional machine learning verification benchmarks.

### 2023-08-25 chapters/conclusion.tex
final thesis

@@
 
-\paragraph{Verifiable Machine Learning with Formal Methods.} In safety-critical and high-stake domains, the verification of machine learning before deployment is crucial.
-To this end, formal methods provide a template to verify different checkable properties of machine learning.
-In particular, SAT, SMT and their variants allow to concentrate on synthesizing constraints from real-world use cases and delegate the solution finding to respective solvers, thanks to the dedicated community in formal methods.
-Building upon our ongoing research, which involves applying SSAT and MaxSAT to address fairness and interpretability challenges, we plan to explore additional formulations in formal methods, such as functional analysis, abstract interpretation, and solvers with expressive theory, to further enhance the verification of machine learning models.
-
-\paragraph{Benchmarking Machine Learning: Improvement in Formal Methods.} Interdisciplinary research is an ongoing process that yields valuable contributions to different disciplines.
-Our work on fairness and interpretability in machine learning has had a positive impact on the improvement of solvers in formal methods.
-For instance, when tackling interpretable classification problems, we realized that MaxSAT alone was insufficient for large-scale rule-based classification, prompting the need for an incremental MaxSAT solver.
-As a result, we contribute to the MaxSAT evaluation competition in 2019 with our MaxSAT benchmarks of interpretable classification.
-Subsequently, the competition introduced an incremental MaxSAT solving track.
-This observation serves as strong motivation for us to continue contributing to the formal methods community by creating additional machine learning verification benchmarks.
+	\paragraph{A New Paradigm for Scalability: Trustworthy Machine Learning with Formal Methods and Beyond.} The SAT and SMT revolution has accelerated the field of formal methods and automated reasoning with powerful solvers  for various problem domains such as SAT/SMT for decision problems, MaxSAT/MaxSMT for optimization problems, and SSAT for hybrid optimization and counting problems.
+By leveraging the clear distinction between the modeling and solving aspects of SAT, MaxSAT, or SSAT, our goal is to enhance fairness and explainability in machine learning through an efficient translation into formal methods.
+Furthermore, my aim is to explore alternative formulations within formal methods, including functional analysis, abstract interpretation, and solvers with expressive theories.
+These explorations seek to further improve the verification process of machine learning models.
+Concurrently, the outcomes of our research have a broader impact, potentially inspiring advancements in the field of formal methods by benchmarking trustworthy machine learning.

### 2023-08-25 chapters/introduction.tex
final thesis

@@
 Therefore, we discuss an incremental learning approach by wrapping traditional optimization solvers such as MaxSAT and MILP to efficiently learn rule-based classifiers in a mini-batch learning setting.
-Here we summarize the contributions of the thesis on interpretable rule-based machine learning\footnote{We publish a python library (\url{https://github.com/meelgroup/mlic}) for learning interpretable rule-based classification rules in a supervised learning setting.}.
+Here we summarize the contributions of the thesis on interpretable rule-based machine learning\footnote{Corresponding Python library is at \url{https://github.com/meelgroup/mlic}.}.
 
@@
 There is also study on fairness attack algorithms to worsen the fairness of a classifier, such as by adding poisoned data samples~\cite{solans2020poisoning}.
-In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness\footnote{We publish a python library (\url{https://github.com/meelgroup/justicia}) for probabilistic fairness verification.
-} and (ii) interpreting sources of unfairness\footnote{We publish a python library (\url{https://github.com/ReAILe/bias-explainer}) for interpreting group-based fairness metrics in machine learning.}.
+In the presence of multiple fairness metrics and algorithms, in this thesis, we contribute to two fundamental problems in fairness: (i) probabilistic verification of fairness\footnote{Corresponding Python library is at \url{https://github.com/meelgroup/justicia}.} and (ii) interpreting sources of unfairness\footnote{Corresponding Python library is at \url{https://github.com/ReAILe/bias-explainer}.}.
 
